---
title: "Rating the 2026 Senate Races"
type: brief
subtitle: "Thirty-five races, rated in August, and a number you cannot revise in November"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("senate-2026-code.R")
```

```{r setup, include=FALSE}
```

`r fx("n_forecasters")` professional forecasters have already rated every Senate race on the 2026 ballot. Their ratings are public, they disagree with each other in `r as.numeric(fx("n_races")) - as.numeric(fx("n_unanimous"))` of the `r fx("n_races")` races, and not one of them publishes what a rating means as a number. So what does a rating actually claim, and how much of November is already settled in August?

The way to find out is to rate the races yourself first. A prediction made after the fact is not a prediction, and the human capacity to remember having believed whatever happened is close to unlimited. The only defence is a record with a date on it.

## The test

This brief reads `r fx("n_ratings")` published ratings of `r fx("n_races")` Senate races, made by `r fx("n_forecasters")` forecasters between `r fx("ratings_first_date")` and `r fx("ratings_last_date")`, together with the candidates who have qualified for those ballots. All of it comes from one Wikipedia article, "2026 United States Senate elections" (<https://en.wikipedia.org/wiki/2026_United_States_Senate_elections>), read as wikitext rather than as a web page. On the rendered page a rating is a coloured table cell, and a colour cannot be read back out.

That article answers the question because nowhere else do the `r fx("n_forecasters")` sit side by side. They publish no common file and share no vocabulary: Cook says "Solid Republican" where Sabato says "Safe Republican". Wikipedia is not the authority on any of it, and is not treated as one here. Every rating is footnoted to the forecaster who published it and every candidate to a state election official, so what the article supplies is the assembly rather than the facts.

Three sources are joined, and the key is the state. The predictions table gives each state its ratings, the race-summary tables give each state its candidates, and `legislators-current.csv` gives each state its sitting senator and Senate class. The third is used only as a test: the 33 non-special races must be exactly the states with a Class 2 senator. Every join takes on the assumptions of both files plus one more about how they line up. That is the subject of [Join Keys and Crosswalks](../data-sources/data-sources-brief.html), which also covers the crosswalk, a table that translates one set of identifiers into another.

## The data

Start with the arithmetic your `r fx("n_races")` answers get added to, which is in `seat_math.csv`.

```{r seat-math}
```

Read the last three rows together. Democrats hold `r D_HOLD` seats that nobody votes on this year. A majority is `r mq("seats for a Democratic majority")` rather than 50, because the Vice President breaks ties and the Vice President is a Republican. So Democrats need `r D_NEED` of the `r fx("n_races")` races on the map, and they currently hold `r fx("d_up")` of them, which makes the requirement a net gain of `r fx("d_net_gain")`.

`r note(paste0("Two of the ", fx("n_races"), " are special elections, in Florida and Ohio, for seats whose regular term does not end until 2028. Marco Rubio and JD Vance both left the Senate in January 2025, and both seats were filled by appointment. A count of only the 33 expiring terms would leave out Ohio, one of the most contested races in the country."))`

`races.csv` holds one row per race, carrying `seat_party`, `senator`, `status`, `dem_cand` and `rep_cand`. It also carries the Cook Partisan Voting Index in `pvi`, the winner's share at the seat's last election in `last_share`, and the forecasters' average, spread and range in `fc_mean`, `fc_sd`, `fc_min` and `fc_max`. `ratings_long.csv` holds every forecaster's rating of every race, with `rating`, `published` and a numeric `score`. `forecasters.csv` gives each of the twelve a `mean_gap`, and `class_ratings.csv` collects the maps other readers have submitted.

That numeric `score` is the one construction decision worth arguing with. Forecasters do not publish probabilities; they publish one of five words, and a word is a category rather than a measurement.

```{r scale-table}
```

Averaging words is impossible, so the words were turned into the numbers in the middle column: positive for the Democrat, symmetric around a toss-up. The two forecasters who use "Tilt" have it scored at half a point. **Nothing makes those gaps equal.** Saying that "Likely" is exactly twice "Lean" is an assumption rather than a finding, and a different set of numbers gives a different average. The scale is printed here so that you can disagree with it.

There is a second problem with the scale, and it is worse. It has seven boxes, every one of which assumes a Democrat is running against a Republican. In `r fx("n_no_dem")` of the `r fx("n_races")` races no Democrat filed at all.

```{r no-dem}
```

Nebraska is the one to look at. Dan Osborn is an independent who came within a few points of a Republican in 2024, with no Democrat on the ballot against him. A scale running from Safe D to Safe R cannot say what a strong independent looks like. You will still have to rate Nebraska, and whatever you put there will be wrong in a way the map cannot show.

> **Before you go on, commit to two numbers.** Write down how many of the 100
> Senate seats the Democrats will hold in January 2027, and how confident you
> are, as a percentage. Put them somewhere you will not edit. The second number
> is the one that will surprise you.

## What it says about democracy

Rate all `r fx("n_races")` first. Figure 1 is the instrument: pick a rating from the row of buttons, then click states to paint them.

```{r rate-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

```{r rate-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Your map.** One square per state, all the same size, because every state elects two senators no matter how many people live in it. A geographic map would size Wyoming against California by land, which is the wrong quantity for a chamber apportioned by state. The running total adds your `r fx("n_races")` calls to the `r D_HOLD` Democratic seats not on the ballot, and a toss-up counts as neither, which is why the total is often a range.

Two buttons will fill the map for you, one from each seat's last election and one from the forecasters' average. Use them second. Starting from someone else's map and adjusting is a different act from building your own, and it will pull your answer toward theirs.

```{r worksheet}
```

**Table 1. The `r fx("n_races")` races.** The paper version of Figure 1. Fill the last column, then read it down to get your code: 1 for Safe D through 7 for Safe R.

`r note(paste0("The incumbent is not the ballot. ", fx("n_retiring"), " of these senators are retiring, ", fx("n_lost_primary"), " lost their own party's primary, and ", fx("n_appointed"), " were appointed rather than elected. Only ", fx("n_incumbent_running"), " of the ", fx("n_races"), " races have the sitting senator running again."))`

Now look at what everybody else committed to.

```{r compare-static, eval=!knitr::is_html_output(), fig.height=8.4}
```

```{r compare-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same `r fx("n_races")` squares, four ways.** Your map, the `r fx("n_forecasters")` forecasters averaged, each seat's last election, and the average of any codes other readers have pasted in. Nothing here is a result: three of the four layers are somebody's opinion and the fourth is a different election.

Read the forecaster layer by the total it implies rather than by the colours. Averaged across all `r fx("n_forecasters")`, the ratings call `r fx("fc_d")` of the `r fx("n_races")` races for the Democrats. Added to the `r D_HOLD` seats not on the ballot, that is **`r fx("fc_d_total")` Democratic seats** — one short of a majority, in a chamber where a 50-50 split leaves the Republicans in charge. The consensus of `r fx("n_forecasters")` professionals, in late August, lands on the one number that resolves nothing.

`r lede(paste0("On ", fx("n_unanimous"), " of the ", fx("n_races"), " races all ", fx("n_forecasters"), " forecasters give the same answer. The interesting races are the other ", as.numeric(fx("n_races")) - as.numeric(fx("n_unanimous")), "."))`

```{r spread-static, eval=!knitr::is_html_output(), fig.height=6.6}
```

```{r spread-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Every rating, one row per race.** Grey dots are the `r fx("n_forecasters")` forecasters and the diamond is their average. A row where the dots sit on top of each other is a race nobody argues about. A row where they spread out is a race the professionals disagree about, which is not the same thing as a close race.

Two rows are worth finding by hand. **`r fx("closest_state")`** has the average nearest to dead even, at `r fx("closest_mean")`, so the twelve collectively cannot call it. **`r fx("most_disagreed")`** has the widest spread, running from `r fx("most_disagreed_min")` to `r fx("most_disagreed_max")`, which is the difference between a race that tilts one way and a race that is over. Those are different kinds of uncertainty. A toss-up everyone agrees is a toss-up is a statement about the race; a race where the forecasters split is a statement about the forecasters.

The last column below is how far each forecaster's ratings sit, on average, from the mean of the other eleven.

```{r forecaster-table}
```

**Table 2. The `r fx("n_forecasters")`.** `r fx("outlier_forecaster")` sits furthest from the rest, at `r fx("outlier_gap")` on a scale six wide, and `r fx("closest_forecaster")` sits closest. Neither fact says who is right. A forecaster who agrees with everyone else adds nothing to an average, while one who disagrees adds the only thing an average has to work with.

The dates matter too. These ratings were published between `r fx("ratings_first_date")` and `r fx("ratings_last_date")`, a spread of more than two weeks in a campaign where two races changed category during it. Some of the twelve are answering a slightly older question than the others.

## What this brief cannot tell you

**Whether any of these ratings are any good.** Twelve forecasters agreeing is not evidence. They read the same polls, the same fundraising reports and each other, so their agreement is close to one opinion counted twelve times. The spread in Figure 3 is a lower bound on the real uncertainty rather than a measure of it.

**What a rating means as a probability.** "Lean D" does not mean 65%, or any other number, and no forecaster here publishes what it does mean. The average is a summary of language rather than a measurement of chance, and the ratings have polling baked into them without saying how much.

**Whether the candidates on Table 1 will be the candidates.** The list was current when the article was read. Maine's Democratic nominee in this file is the party's second choice, after the first left the race in August, and nothing stops that happening again.

**Whether `r fx("n_no_dem")` races have been rated honestly at all.** Idaho, Nebraska and South Dakota have no Democrat on the ballot, and in two of them the challenger is an independent. Every forecaster still put those races on a Democrat-against-Republican scale, and so did you, because the instrument gave you nowhere else to put them.

## What you should have learned

- Your `r fx("n_races")` calls are added to `r D_HOLD` seats nobody votes on, and a majority needs `r mq("seats for a Democratic majority")` rather than 50 because the Vice President breaks ties. That makes the requirement `r D_NEED` of `r fx("n_races")`, a net gain of `r fx("d_net_gain")`.
- Averaging expert ratings requires inventing a number line for words nobody attached numbers to, and the answer depends on the numbers you invent.
- The `r fx("n_forecasters")` averaged call `r fx("fc_d")` of `r fx("n_races")` races for the Democrats, which is `r fx("fc_d_total")` seats: the one total that decides nothing.
- Agreement and closeness are different. All `r fx("n_forecasters")` agree on `r fx("n_unanimous")` races; `r fx("closest_state")` is the closest at `r fx("closest_mean")`, and `r fx("most_disagreed")` is the most disputed, running from `r fx("most_disagreed_min")` to `r fx("most_disagreed_max")`.
- A seven-point party scale cannot represent `r fx("n_no_dem")` of these `r fx("n_races")` races, because no Democrat filed in them, and every published rating still uses it.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `ratings_long.csv` has every forecaster's `rating` and `score` for every race. Find the races they disagree about most. Are those the closest races, or just the least covered?
- `forecasters.csv` gives each one a `mean_gap`. Work out who is boldest and who hedges, then decide which you would rather read in August.
- `races.csv` carries `pvi` and `last_share` beside `fc_mean`. Find the races where the ratings depart furthest from the partisan lean, and say what the forecasters see that the past result does not.
- `seat_math.csv` fixes the target at `r mq("seats for a Democratic majority")`. Recompute it on the assumption that the Vice President is a Democrat.
- **Stretch.** Write down now how you will score yourself in November, and date the rule before you know anything.
- **Stretch.** After the election, put the result beside `class_ratings.csv` and `ratings_long.csv` and score both against your own rule.

## Sources

**Data**

- Wikipedia, "2026 United States Senate elections"
  (<https://en.wikipedia.org/wiki/2026_United_States_Senate_elections>), fetched
  as wikitext through the MediaWiki interface, currently
  `r format(fxn("wiki_chars"), big.mark = ",")` characters. Two tables are
  read: *Predictions*, for each race's partisan index, incumbent, last result
  and ratings; and *Race summary*, for the qualified candidates. The version
  read is kept beside the numbers it produced.
- The `r fx("n_forecasters")` forecasters whose ratings that table collects,
  each cited there to its own page and carrying the date its ratings were last
  changed: `r { f <- FCS[order(FCS$forecaster), ]; paste0(paste(sprintf("%s (%s)", f$forecaster, format(as.Date(f$published), "%e %B %Y")), collapse = ", ")) }`.
  Their published ratings are recorded here, not recomputed.
- `@unitedstates/congress-legislators`, `legislators-current.csv`
  (<https://unitedstates.github.io/congress-legislators/legislators-current.csv>),
  used only as a check.

**The data itself**

Every figure above is drawn from these tables, which are plain CSV.

[`class_ratings.csv`](data/derived/class_ratings.csv), [`forecasters.csv`](data/derived/forecasters.csv), [`races.csv`](data/derived/races.csv), [`ratings_long.csv`](data/derived/ratings_long.csv), [`seat_math.csv`](data/derived/seat_math.csv)

**Choices made here, which are in no source**

- The seven-point scale and the numbers behind it: Safe 3, Likely 2, Lean 1,
  Tilt 0.5, Toss-up 0, signed positive for the Democrat.
- The rule turning a past election into a rating for Figure 2's last-election
  layer: a winner's share of `r fx("base_cut_safe")`% or more is Safe,
  `r fx("base_cut_likely")` to `r fx("base_cut_safe")` is Likely,
  `r fx("base_cut_lean")` to `r fx("base_cut_likely")` is Lean, and under
  `r fx("base_cut_lean")` is a toss-up.
- Counting the two independent senators who caucus with the Democrats among the
  47, which is how every seat count in the coverage is stated and is in no file.

**Also cited**

- Balk, T. "Can Democrats Win Back the Senate? Why Their Chances Are
  Improving." *The New York Times*, 24 August 2026, for the six-state core
  battlefield and the seat arithmetic reproduced above.

```{r ai-prompt, echo=FALSE, results="asis"}
```
