---
title: "Presidential Returns Across 150 Years"
type: brief
subtitle: "A century and a half of presidential returns, and a statistic that returns to where it started without meaning the same thing"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("historical-campaigns-code.R")
```

```{r setup, include=FALSE}
```

A voter in Alabama who went to the polls in November 1948 intending to vote for Harry Truman found that she could not. There was no Democratic slate on the ballot; the state party had replaced it with the Dixiecrats. Sixteen years later she found the same thing again, when Alabama Democrats withheld the national ticket rather than pledge their electors to Lyndon Johnson. In the file of presidential returns those two elections appear as blanks, and a blank is a strange way to record the disappearance of a major party from a ballot.

Campaigns are the mechanism by which citizens are addressed. Where a state is not contested, nobody is addressed: no persuasion, no organizing, no reason for a candidate to learn what the place wants. **Uncontestedness is not a neutral condition. It is a withdrawal of political attention**, and it settles on particular places for long stretches.

Most accounts of how American campaigning changed between 1864 and 2024 are accounts of technique — party machines giving way to candidate organizations, conventions to primaries, newspapers to television to whatever this is. What follows is the other half of that story. Campaigns did not transform in a vacuum. They transformed on **a map that was itself transforming**, and for most of the period a third of the country was not contested at all.

One row of the file is **one state in one presidential election**.

```{r one-row}
```

`democrat` and `republican` are percentages of the total vote. Note the `winner` column, and note that in 1948 it says **Strom Thurmond** while the Democratic column says `r pc(sv("MS", 1948, "democrat"))`. Both are true: the Democratic ticket took most of the Democrat-versus-Republican vote and neither of them carried the state, because the Dixiecrat did. Hold on to that; it comes back shortly.

## Four hands between the courthouse and this table

The `r length(unique(p$year))` elections here come from *PresElectionResults*, an R data package Jason Timm publishes on GitHub. Its state series is compiled from Wikipedia's per-election articles, which rest in turn on Dave Leip's *Atlas of U.S. Presidential Elections* and on the official state canvasses. Read that chain backwards and notice what is missing from it. There is no federal source.

The United States has never had a national election authority, so no agency holds a record of how the country voted in 1876. The canvass belongs to the state that conducted it. Every layer above the canvass is a private compilation, made by somebody who decided it was worth making. The campaign-finance chapters begin with a statute compelling disclosure. This one begins with a hobby that became indispensable.

On the spectrum this book keeps returning to, that puts these numbers about as far from *made and published by the same body* as anything here gets. They were **assembled and cleaned by compilers from primary records**, and not by one set of hands but by four in a row. Each made decisions, and none wrote them into the file.

By the time a row reaches the table, somebody has already made several decisions. Which canvass to believe where two disagree, how to treat a fusion ticket, and what to do with a state whose returns were disputed. Those decisions are baked into the number, and there is no column recording them.

Getting the file was almost free: one `download.file()` against a public GitHub address, no key, no account, no request form. **The cost is entirely on the other side of that ease.** There is no authority to appeal to when a figure looks wrong, because the authority is fifty-one nineteenth-century state archives that nobody has re-checked. The most modern link in the chain is a `master` branch that can change without notice or a version number. And the file's uniform formatting actively conceals how differently its rows were made. Committing the output here freezes the version being argued about, which is the only part of that any reader downstream can fix.

That is also why the file is the right one. A question about how the map changed across a century and a half can only be answered by a series that is consistent across a century and a half. Consistency is precisely what a compiler supplies, and what fifty-one separate archives do not. Better evidence would be the canvasses themselves, county by county, with turnout attached. For the full period nobody has assembled that, and the reason is not secrecy but expense.

The cost of a compilation is that it inherits the reliability of its weakest link and does not mark which link that was. A row for Louisiana in 1872 and a row for Michigan in 2020 are formatted identically, and are not the same kind of object at all.

Nor does the file carry votes. These are shares, so there is no national total in it and no turnout. The only national figure this chapter can build is an unweighted average across states, which is said out loud wherever it is used. And beneath all of that sits the problem the arithmetic cannot reach. A vote share presumes an electorate, and who was permitted to be part of the electorate is the thing that changed most over these years.

The cleaning was small and worth knowing. What is downloaded is an `.rda` from the package's `master` branch. That is a branch and not a tagged release, which makes the most modern link in the chain the least stable one. The build then checks the expected row count and column names before it trusts anything, keeps seven columns, sorts by year and state, and writes the file.

Crucially it leaves `NA` as `NA` instead of turning a missing party into a zero, which is the only reason the `r nrow(absent)` absent-ballot rows are recoverable at all. The two-party share that most of this chapter runs on is not in the file. It is computed in this document, on the page, so that the choice stays visible at the point where it is made.

## What comes down the wire

The source is not a CSV and not an API. It is an `.rda`, a compressed R object, sitting in a public repository. Fetched by address and loaded into memory, it turns out to be a table of `r nrow(p)` rows and seven columns, in this order:

| Position | Column | What it holds |
|---:|:---|:---|
| 1 | year | the presidential election year |
| 2 | state_abbrev | the state's two-letter code |
| 3 | winner | who carried the state |
| 4 | party_win | that winner's party |
| 5 | democrat | the Democratic share of the vote, as a percentage |
| 6 | other | the share going to everyone else |
| 7 | republican | the Republican share |

Two rows of it, as they arrive:

| year | state_abbrev | winner | party_win | democrat | other | republican |
|---:|:---|:---|:---|---:|---:|---:|
| 1864 | CA | Abraham Lincoln | republican | 41.40 | NA | 58.60 |
| 1948 | AL | Strom Thurmond | other | NA | 79.75 | 19.04 |

The 1864 row is unremarkable and the 1948 row is the whole chapter. **Alabama's Democratic share is `NA`.** Not zero, not small — absent. A reader who opened this file, saw the empty cells and dropped those rows as damaged would have deleted the elections in which a major party was not on the ballot at all. Those are precisely the elections where the party system was breaking. And a reader who did the opposite and filled the blanks with zero would have recorded that no Alabamian wanted to vote for Harry Truman, which is a different claim entirely and an untrue one.

Here is what the same two rows look like afterwards:

```{r clean-hist}
```

Almost nothing happened, and that is the point worth drawing out. The columns were reordered so the two major parties sit together, and the rows were sorted. The row count and the column names were checked against what the script expected, and then the file was written out. **The `NA` was left alone.** That single restraint is what makes the `r nrow(absent)` ballot-absence rows recoverable later in this chapter; every other treatment of them destroys the finding.

### What was decided, and what was refused

**The missing values were not repaired.** Neither the `r nrow(absent)` rows where a party had no ballot line nor the `r format(sum(is.na(p$other)), big.mark = ",")` rows with no third-party figure were filled in. In a file of percentages the temptation to make rows add to 100 is strong and every way of doing it invents something.

**The two-party share is not in the source.** It is computed in this document rather than in the build script, on purpose. That way the choice to ignore everybody outside the two major parties is made where a reader can see it, rather than baked into a column called `d2` that arrives looking like a fact.

**The row count is asserted before the data is used.** The script checks for `r nrow(p)` rows and the seven expected column names and stops if either is wrong. That is the only defense available against a source that is a moving branch rather than a released version: it cannot tell you the numbers changed, but it will tell you the shape did.

**Nothing marks which rows are trustworthy.** A row for Louisiana in 1872 and a row for Michigan in 2020 are formatted identically, carry the same number of decimal places, and are not the same kind of object. The file has no column for that and this chapter cannot add one.

## An absence nobody annotated

```{r coverage}
```

`r sum(p$year == min(p$year))` states in 1864, `r sum(p$year == max(p$year))` now. That is not a data problem; it is the country growing. But it means **every comparison across time is a comparison between different sets of states**, and we will have to keep saying so.

There were thirty-six states in the Union in 1864. This file has `r sum(p$year == 1864)`\. **Eleven states are missing because they were in armed rebellion, and the file does not say so. It just has fewer rows.** That is the first thing to understand about historical data: absence is never annotated.

## A choice with a cost

`r format(sum(is.na(p$other)), big.mark = ",")` of `r format(nrow(p), big.mark = ",")` rows carry no third-party number at all, so the two major-party percentages do not add to 100 and the gap is not meaningful. The standard fix is to ignore everyone else and work in **two-party share**: Democratic votes as a percentage of the Democratic-plus-Republican vote.

This is a real choice with a real cost. In 1912, 1968 and 1992, the years of Roosevelt, Wallace and Perot, a two-party share hides most of what happened. In Mississippi in 1948 it produces the collision in the row above, where a party holds most of the two-party vote and loses the state anyway.

There is a second kind of missingness, and it is not a gap in the data:

```{r absent}
```

`r nrow(absent)` rows have no Democratic or no Republican share, and each records a state where a major party **had no line on the ballot**. In 1912 the Progressives kept Taft off California's ballot. **In 1948 and again in 1964 there was no Democratic slate in Alabama** — replaced by the Dixiecrats in 1948, and withheld in 1964 when Alabama Democrats refused to pledge to Johnson. A voter who wanted to vote for Harry Truman, or for Lyndon Johnson, could not.

The ballot is not a neutral instrument.

## The longest-running fact in American politics

Define the eleven states of the Confederacy, which is a choice and one worth watching being made, then average the Democratic two-party share inside and outside. Figure 1 is the result, for every election in the file.

```{r south-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r south-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The Democratic two-party share, South and rest of the country, `r min(p$year)` to `r max(p$year)`.** Each line is the unweighted average across states, one for the eleven states of the Confederacy and one for everywhere else. The Southern line sits above the other for eighty years, crosses it in `r first_neg(g11)`, and stays below from `r perm_neg(g11)` onward.

```{r gaps}
```

For eighty years the Southern line sits far above the other one. **The gap peaks at `r pc(max(g11$gap, na.rm = TRUE))` points in `r g11$year[which.max(g11$gap)]`**, and is still `r pc(gv(g11, 1948))` in 1948. It **goes negative for the first time in `r first_neg(g11)`**: the year the Civil Rights Act passed, Johnson won a national landslide, and Goldwater carried the Deep South.

It does not vanish overnight. Carter, a Georgian, pulls it back positive in 1976 (`r pc(gv(g11, 1976))`) and 1980 (`r pc(gv(g11, 1980))`). **From `r perm_neg(g11)` onward it is negative in every single election**, reaching `r pc(gv(g11, 2024))` in 2024. The sign has flipped and stayed flipped.

## Move the definition and see what survives

Eleven states was a choice, and a finding that depends on a choice is worth testing against a different one. Here is the same comparison with the eleven plus the border states that permitted slavery and did not secede.

```{r robust}
```

The table takes seven years. Figure 2 takes all of them.

```{r robust-static, eval=!knitr::is_html_output(), fig.height=4.3}
```

```{r robust-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same gap, under two definitions of the South.** The Southern minus non-Southern difference in Democratic two-party share, computed once for the eleven Confederate states and once for those eleven plus six border states. The wider definition dilutes the peak from `r pc(max(g11$gap, na.rm = TRUE))` points to `r pc(max(g17$gap, na.rm = TRUE))`, and leaves the crossing year and the year of the permanent flip exactly where they were.

**The magnitudes shrink; the story does not move.** Adding the border states cuts the `r g11$year[which.max(g11$gap)]` peak from `r pc(max(g11$gap, na.rm = TRUE))` to `r pc(max(g17$gap, na.rm = TRUE))` and the 1964 trough from `r pc(gv(g11, 1964))` to `r pc(gv(g17, 1964))`. But the first negative year is **`r first_neg(g17)` either way** and the permanent flip is **`r perm_neg(g17)` either way**.

That is a robustness check the finding passed, and the dilution is itself evidence: the border states were always less Democratic than the Deep South, so adding them pulls both extremes toward the middle. The effect is concentrated where the theory says it should be.

## A label that cannot be right

While we are here, look at who this file says won states in 1960.

```{r truman}
```

**Harry Truman was not a candidate in 1960.** He had left office eight years earlier.

What happened is more interesting than a typo. Alabama and Mississippi ran slates of **unpledged electors** — Democrats who appeared on the ballot promising *not* to support the national ticket, hoping a close election would let the South bargain over civil rights. Mississippi's slate won outright. Whoever compiled this file needed a name for it and reached for a Democrat.

Notice that this did **not** corrupt the numbers we have been using. The vote shares are fine; only the label is wrong. **Bad data is rarely uniformly bad, which is exactly what makes it dangerous.**

## How far apart are the states?

If every state voted alike, the spread between them would be zero. Standard deviation of the two-party share across states, election by election:

```{r spread}
```

The spread was enormous in the era of the Solid South. It bottoms out at **`r pc(min(sp$sd), 2)` in `r sp$year[which.min(sp$sd)]`**, when the states never agreed more. Then it climbs again, roughly doubling to `r pc(sp$sd[sp$year == 2024], 2)` by 2024.

A standard deviation is one number standing in for a whole shape. Figure 3 shows the shapes themselves: every state, in the widest year, the narrowest year, and now.

```{r shapes-static, eval=!knitr::is_html_output(), fig.height=3.9}
```

```{r shapes-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Every state's Democratic two-party share, in three elections.** The year the states were furthest apart, the year they were closest, and 2024. The spread of `r sd_hi` and the spread of 2024 are of comparable width; the distributions are not comparable objects.

The obvious reading is that the country has returned to something like its old level of division: `r pc(sp$sd[sp$year == 2024], 1)` today against `r pc(max(sp$sd), 1)` at the `r sd_hi` peak, more than half the way back.

## The same number, two different countries

The spread is one number, but it is made of parts. Some of the variation between states is variation **between regions**, and some is variation **within** them. Split the total sum of squares in each year into the part explained by the South-versus-rest division and the part that is not.

> **Before you read the table, put two numbers down.** In `r sd_hi`, with the
> Solid South intact, what percentage of all the state-to-state variation was
> simply the South differing from everyone else? And what is that percentage in
> 2024? If the country has genuinely gone partway back to where it was — the
> reading the standard deviations invite — then the second number ought to be a
> solid fraction of the first. Write both down, and be specific: a percentage
> each, not "high" and "lower".

```{r decomp}
```

**In `r vd$year[which.max(vd$pct_between)]`, `r pc(max(vd$pct_between))`% of all state-to-state variation in presidential voting was the South differing from everyone else. In 2024 the figure is `r pc(vb(2024))`%.**

That is the answer to the prediction, and it is decisive. The first number is usually guessed too low and the second far too high. `r sd_hi` comes in at `r pc(vb(sd_hi))`%. Most readers put 2024 somewhere in the thirties or forties, on the reasonable-sounding logic that a spread half the way back should be half a regional story. It is `r pc(vb(2024))`%.

The 1924 spread was one region voting unlike the rest of the country. The 2024 spread is `r pc(100 - vb(2024))`% variation *within* regions: states differing from their neighbors, everywhere, along lines that no regional label captures. It is the sorting usually called polarization, and it is a different object with a similar standard deviation.

Figure 4 tracks that split across every election in the file.

```{r decomp-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r decomp-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 4. How much of the spread between states is one region differing from the rest.** The share of the total state-to-state variance in Democratic two-party share attributable to the South-versus-everywhere-else division, in each election. It peaks at `r pc(max(vd$pct_between))`% in `r vd$year[which.max(vd$pct_between)]` and stands at `r pc(vb(2024))`% in 2024. The spread that remains is variation between neighbors rather than between regions.

**A number going up does not tell you why.**

## The vanishing battleground

Then there is the fact that matters most for how campaigns behave. Count the states within five points of an even split.

```{r close}
```

**`r cv(1960,"close")` competitive states in 1960; `r cv(2024,"close")` in 2024** — out of more states than existed in 1960. As a share it falls from `r pc(cv(1960,"pct"))`% to `r pc(cv(2024,"pct"))`%.

Figure 5 puts that count on the same axes as the elections it came from.

```{r battle-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

```{r battle-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 5. The share of states decided within five points, election by election.** Filled points mark the `r sum(ncl$even)` elections in which the average state sat within three points of an even split, so that the national result cannot be blamed for an empty battleground. Among those elections, `r pc(ev_pre)`% of states were competitive before 2000 and `r pc(ev_post)`% since.

Most of the sawtooth is the national result: a landslide leaves nothing close anywhere, and the four emptiest elections in the series (`r paste(sort(ncl$year[order(ncl$pct)][1:4]), collapse = ", ")`) are all blowouts. The filled points strip that explanation out. They are the `r sum(ncl$even)` elections in which the average state sat within three points of an even split — and among those, `r pc(ev_pre)`% of states were competitive before 2000 against `r pc(ev_post)`% since. **The battleground shrank in elections that were themselves as close as they had ever been.**

A campaign in 1960 had to think about most of the country. A campaign in 2024 can think about a dozen states and safely ignore the rest. **Nothing about consultants or television explains as much about modern campaign behavior as this does.**

## What a century and a half of returns will support

The Southern realignment happened, it is datable to 1964 with a permanent flip by `r perm_neg(g11)`, and it survives changing the definition of "the South" from eleven states to seventeen.

Sharper than that: the rise in state-to-state spread since 1960 is not a return to the old division. The regional component of that spread fell from `r pc(max(vd$pct_between))`% to `r pc(vb(2024))`%, so the same statistic now measures a different country. **A trend line is not a mechanism.**

What these returns cannot say is *why* the South changed. They date the change; they cannot attribute it, and the attribution is exactly what is contested. Nor can they say that a 1900 Mississippi return measures what a 2024 Mississippi return measures. Post-1877 Southern totals are accurate counts of elections conducted under systematic disenfranchisement: the number is real, and what it measures is not what it appears to measure.

## Why the South changed, and what a state average can settle

**The realignment's timing is agreed; its cause is not**, and the dispute is one of the longest-running in American political science. On one side, Carmines and Stimson (1989) argue that racial policy realigned the parties from the top down, beginning with elite positioning in the early 1960s. Kuziemko and Washington (2018) revisited the question with individual survey records. They found that defection among racially conservative white Southerners accounts for essentially the whole decline in Democratic support between 1958 and 1980, with no role for income growth. On the other, a competing tradition emphasizes Southern economic development, the growth of a suburban middle class, and in-migration from outside the region.

Kuziemko and Washington is the strongest recent evidence on the racial side, and it is strong precisely because it goes below the aggregate. The two lines in Figure 1 are consistent with almost any of these stories, and only individual-level data can separate them. That is a general fact about aggregate returns rather than a fact about this file.

**A second dispute is about whether "the South" is a unit at all.** V. O. Key's *Southern Politics in State and Nation* (1949) is still the founding text. It insisted on state-by-state analysis precisely because a regional average conceals enormous variation. Figure 2 shows the average holds up; it does not show the average is meaningful. Mississippi and Virginia have not moved together for fifty years.

**A third is about polarization.** The rise in spread is routinely reported as evidence that the country is coming apart. Figure 4 shows that the spread is real and that its regional component is not, which means the sorting is happening at some other level — county, neighborhood, household. That claim cannot be tested with state-level data, and this file has nothing smaller than a state.

## What the returns can testify to

The strengths are those of a long consistent series: `r length(unique(p$year))` elections in one format, the same two columns meaning the same thing throughout. The file is also honest about its own gaps. The `r nrow(absent)` rows where a party was absent from a ballot are `NA` rather than zero, which is more than most sources manage. It is the reason the Alabama ballots of 1948 and 1964 are visible at all.

The weaknesses run in the other direction. The set of states changes, from `r sum(p$year == 1864)` to `r sum(p$year == 2024)`, so early years are noisier than they look and every comparison across time is a comparison between different sets of states. Two-party share hides third parties, and 1912, 1968 and 1992 all look tamer here than they were. Shares are not votes: a five-point margin in Wyoming and a five-point margin in California are the same number, and nothing like the same event.

Labels can be wrong while numbers are right, as 1960 demonstrates. Nineteenth-century totals rest on state records of varying quality, and post-1877 Southern totals rest on an electorate constructed by exclusion. And there is no turnout, no demographics, and nothing below the level of a state.

Three questions stay open. Why the South realigned, and how much of it was racial policy against economic change, which is not answerable from aggregates of any kind. What level the modern sorting operates at, which Figure 4 rules out being regional without saying what it is. And why the battleground shrank in elections that themselves stayed close, which Figure 5 establishes without explaining.

The voter in Alabama who could not vote for Truman was living in the extreme case of a general condition. For eighty years an entire region was conceded in advance, and the returns record that concession as ordinary arithmetic. The concession has since moved: it is no longer regional, it no longer has a name, and by 2024 it covered a larger share of the states than it did in 1960.

## Sources

**Data**

- Jason Timm, *PresElectionResults*
  (<https://github.com/jaytimm/PresElectionResults>), state series compiled from
  per-election sources resting ultimately on Dave Leip's *Atlas of U.S.
  Presidential Elections* and the official state canvasses.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`pres_states_1864_2024.csv`](data/derived/pres_states_1864_2024.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `pres_states_1864_2024.csv` has an `other` column that is often missing rather than zero. Count the missing years. What is the difference between a third party getting no votes and nobody recording them?
- Follow one state from the beginning of the series to the end. Write down every point where the party labels stop meaning what they meant before.
- Compute the national two-party share for the earliest and latest years. They are close. Explain to somebody why that sameness is not evidence of stability.
- Find the states missing from the early years. What has to be true about a state before it appears in a table of presidential returns?

**Academic literature**

- Carmines, E. G. and Stimson, J. A. (1989). *Issue Evolution: Race and the
  Transformation of American Politics.* Princeton: Princeton University Press.
- Key, V. O. Jr. (1949). *Southern Politics in State and Nation.* New York:
  Alfred A. Knopf.
- Kuziemko, I. and Washington, E. (2018). "Why Did the Democrats Lose the South?
  Bringing New Data to an Old Debate." *American Economic Review* 108(10):
  2830–2867.



```{r ai-prompt, echo=FALSE, results="asis"}
```
