---
title: "HOLC Redlining Maps and Present-Day Geography"
type: brief
subtitle: "A form filled in on the edge of Cleveland in 1939, and the ground it still describes"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("redlining-code.R")
```

```{r setup, include=FALSE}
```

An appraiser working for the federal government walked a few streets on the southern edge of Cleveland in the late summer of 1939 and then filled in a form. It had spaces for the terrain, the class of the inhabitants, the age of the buildings, the availability of mortgage money. Under *Clarifying Remarks* he wrote:

> `r QT$remarks`

The area is `r QT$label`, *`r QT$neighborhood`*, and the sheet is dated `r QT$dated`. Two other lines on that form matter as much as the paragraph. One is headed *Infiltration of*, and on it he wrote *`r QT$infiltration`*. The other asks what share of the residents were Black, and on that line he wrote `r QT$negro`, meaning none. He gave the area a grade of `r QT$grade`.

Race was not standing in for something else on that sheet. **It was the stated criterion**, applied to a street where by the appraiser's own count no Black family yet lived. Grades like this one told a federal corporation where mortgage lending was safe, and lending is how most American families come to own anything. School money is raised on property, and services, transit and policing are all handed out on a map. So the question is worth asking plainly: ninety years later, on the actual ground, does that line still show?

## The test

Answering it means putting two files beside each other. The first is *Mapping Inequality*, the digitized residential security maps of the Home Owners' Loan Corporation. It holds `r n(SR$features)` neighborhood areas traced from `r SR$city_surveys` city surveys of 1935 to 1940, `r n(SR$graded)` of them carrying a grade, and it is published by the Digital Scholarship Lab at the University of Richmond (<https://dsl.richmond.edu/panorama/redlining/>). The second is the 2020 Census, which counts every resident of every tract (<https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/>).

The 1930s file is the right instrument because it is not a study of the period. It is the working paperwork of the agency that did the sorting, written to be acted on rather than read later. The men who wrote it had no reason to overstate the part race played, because they did not think it needed justifying. A modern index of disfavored neighborhoods would carry its author's theory of what disfavor looks like. Nobody graded on these sheets was asked anything, and most never learned a federal agency had given their street a letter. That is the kind of record [Section III](../admin-records-source/admin-records-source-brief.html) is about.

The census is doing a second job here that is easy to miss. The grade file holds shapes and letters and no people at all. Every share below is a count of Black residents over a count of all residents, and both numbers are borrowed from the census. The record supplies only the letter, and somebody else counted everyone standing under it.

## The data

The table the figures rest on is `tracts.csv`: one row per 2020 census tract, `r n(nrow(tr))` rows, covering `r length(unique(tr$city))` cities in `r length(unique(tr$state))` states. Each row carries `GEOID`, the tract's eleven-digit identifier, with `state` and `city`; `grade`, the letter the ground under it was given in the 1930s; and `total`, `white`, `black` and `pct_black` from the census.

```{r one-record}
```

The letter is the strangest thing in that row. The corporation used four of them:

```{r grades}
```

The D areas were printed in red, which is where the word *redlining* comes from, and the paragraph above is what a D looked like from the inside. Figure 1 is Cleveland's whole sheet.

```{r holc-map-d3, results='asis', eval=knitr::is_html_output()}
```

```{r holc-map-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

**Figure 1. Cleveland as the Home Owners' Loan Corporation graded it in 1939.** `r sum(hm$n)` graded areas over `r pc(sum(hm$area_km2), 0)` square kilometers, filled with the corporation's own four colors read out of the source file rather than chosen here. A map is the right form because the grade was applied to ground. Red sits in the industrial core, an average of `r pc(hm$km_out[hm$grade=="D"])` km from the center of the surveyed area; green sits `r pc(hm$km_out[hm$grade=="A"])` km out at the edge. The concentric rings of creditworthiness are visible before any census number arrives. The dashed rectangle marks the ground enlarged in Figure 2.

Getting the census onto that map takes a decision, because the two sets of shapes have nothing to do with each other. A 1937 boundary was drawn with no knowledge of a 2020 census tract, and tracts have been redrawn every decade since. There is no shared identifier and no lookup table. The rule used here is the simplest defensible one: take each tract's center point, and ask which graded area contains it.

```{r join}
```

```{r zoom-d3, results='asis', eval=knitr::is_html_output()}
```

```{r zoom-static, eval=!knitr::is_html_output(), fig.height=4.9}
```

**Figure 2. A `r pc(ZW$x1 - ZW$x0, 1)` km window of Cleveland, with 1939 underneath and 2020 on top.** Colored surfaces are HOLC areas, thin outlines are 2020 tracts, and each dot is the center point the rule consults. The heavily outlined tract is the rule getting it wrong: its center falls in a `r fo$centre_grade` area, so the whole tract is coded `r fo$centre_grade`, but its land is `r mixstr(fo)`.

That cost is worth stating in numbers. The rule grades `r n(nrow(asg))` tracts in Cleveland and Philadelphia, and `r nrow(mis)` of them get a letter that is not the letter covering most of their own land. Another `r n(nrow(drp))` are dropped, because their center falls outside every graded area. It is at least reproducible: re-running it from the polygons returns the grade the file already carries for all but `r nrow(agree) - sum(agree$centre_grade == agree$grade)` of `r n(nrow(agree))` tracts. A crosswalk is a table that translates one set of areas into another, and one weighted by population would grade some of these tracts differently. Choosing between such rules is Section IV's subject, in [Join Keys and Crosswalks](../../06-putting-data-together/data-sources/data-sources-brief.html).

Three cleaning decisions sit under all of it. The grade lives in a field called `grade`; the natural guess, `holc_grade`, exists in other distributions of these maps and returns nothing here, which would produce a complete, plausible, entirely ungraded map. `GEOID` has to be read as text, because the Californian identifiers begin with a zero a numeric read deletes. And the source's own `area` field is in square degrees, so every area here is computed after reprojecting to metres.

> **Before you read on, commit to two numbers.** The file holds `r nrow(ct)`
> cities with both A-graded and D-graded ground. **In the average city, how many
> percentage points more Black is the D-graded ground than the A-graded
> ground today?** And **how many of the `r nrow(ct)` cities run the other
> way?** Write both down.

## What it says about democracy

Pool all `r n(nrow(tr))` tracts by grade and the obvious comparison appears at once.

```{r national}
```

Areas graded A are `r pc(gv("A","pct"))`% Black today and areas graded D are `r pc(gv("D","pct"))`%, a gap of `r pc(gv("D","pct") - gv("A","pct"))` points. After the appraisal sheet above, that is deflating, and the ordering does not even behave: C areas are `r pc(gv("C","pct"))`% Black, less than the `r pc(gv("B","pct"))`% in B areas.

**The pooled table is not measuring grades. It is measuring cities.** Cleveland's D areas and Oakland's A areas sit in it together, and American cities differ enormously in overall Black population for reasons unconnected to the corporation. San Francisco's D neighborhoods sit in a city about 5% Black; Cleveland's A neighborhoods sit in a city roughly half Black. Differences between cities swamp the difference within them, and only the difference within them is the question.

So make the comparison inside each city, where the same housing market, the same city government and the same regional history apply to both grades.

```{r within}
```

In Cleveland, whose 1939 sheet opens this brief, neighborhoods graded A are `r pc(cv("Cleveland","a_pct_black"))`% Black today and neighborhoods graded D are `r pc(cv("Cleveland","d_pct_black"))`%. That is `r pc(cv("Cleveland","gap"))` points between two sets of neighborhoods separated by nothing except a letter assigned before the Second World War. Cleveland is the extreme case, and the extreme case is never the argument. Across all `r nrow(ct)` cities holding both grades the average gap is `r pc(mean(ct$gap))` points, and `r sum(ct$gap > 0)` run in the direction the history predicts.

```{r slope-static, eval=!knitr::is_html_output(), fig.height=5.2}
```

```{r d3-slope, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Within each city, from its A-graded ground to its D-graded ground.** One line per city, joining the share of residents who are Black in the tracts graded A to the share in the tracts graded D, 2020 Census. Of the `r nrow(ct)` cities with both grades, `r sum(ct$gap > 0)` rise and `r sum(ct$gap < 0)` fall; the steepest rise is `r ct$city[which.max(ct$gap)]`, at `r pc(max(ct$gap))` points. The average line rises `r pc(mean(ct$gap))` points.

The `r sum(ct$gap < 0)` downward lines are the most useful thing in the figure, and they are two different things wearing the same sign.

```{r reversals}
```

**Philadelphia is a real reversal.** Its D areas are `r pc(cv("Philadelphia","d_pct_black"))`% Black and its A areas `r pc(cv("Philadelphia","a_pct_black"))`%, backwards by `r pc(abs(cv("Philadelphia","gap")))` points across `r n(cv("Philadelphia","d_tracts"))` D tracts holding `r n(cv("Philadelphia","d_pop"))` people. North Philadelphia was graded D and has been substantially gentrified, and several historically Black neighborhoods there were never graded D at all. Urban renewal, deindustrialization, highway building and very uneven reinvestment all happened in between.

**Joliet is not a real reversal.** Its `r pc(abs(cv("Joliet","gap")))`-point gap rests on `r cv("Joliet","a_tracts")` A tracts against `r cv("Joliet","d_tracts")` D tract, and a percentage computed on a handful of tracts is not a stable quantity. Printing it beside Cleveland's `r cv("Cleveland","a_tracts")`-against-`r cv("Cleveland","d_tracts")` as equal weight would mislead a reader with accurate numbers. Several other reversals are small in the same way.

So the sentence this evidence will bear is the middle of three. The grade is associated with a neighborhood's racial composition today, on average, within a city. It did not determine it, because Philadelphia went the other way, and nothing here shows that it caused it.

## What this brief cannot tell you

**It cannot tell you whether the maps did anything.** The appraisers were describing segregation that already existed, much of it produced by restrictive covenants, by violence and by private lending that predated the corporation. On that reading these maps are an excellent record of 1930s segregation and no evidence about its causes. Separating the two needs a different design, not a bigger version of this one. Aaronson, Hartley and Mazumder (2021) compared cities that just barely were mapped and just barely were not, and found effects that lasted decades.

**It cannot speak for the country.** The corporation surveyed only cities above a population threshold, so there was never a national frame, and this file holds `r length(unique(tr$city))` cities in `r length(unique(tr$state))` states.

**It cannot show a path.** This is one snapshot. These neighborhoods may have been converging for fifty years, or diverging, or crossing over, and 2020 looks the same in all three cases.

**And a grade is not a measurement.** It is one appraiser's letter, with no second rater, no units and no standard error, used here as though it were an assignment made at random. Grade was never the only thing that differed between an A and a D neighborhood in 1937.

## What you should have learned

- Within a city, neighborhoods graded "Hazardous" in the 1930s are on average `r pc(mean(ct$gap))` points more Black today than neighborhoods the same agency graded "Best", across `r nrow(ct)` cities and `r n(nrow(tr))` tracts.
- The pooled national comparison is the wrong test and it understates: it puts A areas in one city against D areas in another, returning a gap of `r pc(gv("D","pct") - gv("A","pct"))` points with the middle grades out of order.
- `r sum(ct$gap < 0)` cities run the other way, and not for the same reason: Philadelphia reverses across `r n(cv("Philadelphia","d_tracts"))` D tracts, Joliet on `r cv("Joliet","d_tracts")`.
- The link between 1939 and 2020 is one sentence of rule with a measurable cost: `r nrow(mis)` of `r n(nrow(asg))` graded tracts in two cities carry a letter that is not the letter covering most of their land.
- The record supplies the grade and nothing else. Everyone in every share above was counted by the census, not by the corporation.

## Extensions

None of these is answered above, and every one can be answered from the tables below.

- `cities.csv` gives each city a `gap`, the share Black in A areas against the share in D areas. Sort by it and read both ends. Do the `a_tracts` and `d_tracts` counts make the smallest gap worth anything?
- `tracts.csv` has `grade` beside `total` and `pct_black`. Group by `city` and `grade` and compare the population under each letter. Which cities put most of their people under a C?

- `fig_mix.csv` carries `own_share`, how much of a tract's graded land carries the letter it was given. Sort it and read the lowest rows. Those tracts are the ones any map of this data draws wrong.
- `fig_quote.csv` holds one appraiser's own words, with the sheet's `label` and `dated`. Read it, then look up that neighborhood now.
- **Stretch.** *Mapping Inequality* publishes the description sheets for every graded city. Read three D sheets and three A sheets, and count how many name a group of people rather than a state of the buildings.
- **Stretch.** The 2020 census also publishes Hispanic origin by tract. Attach it to `tracts.csv` by `GEOID` and ask whether the within-city gap survives.

## Sources

**Data**

- Nelson, R. K., Winling, L., Marciano, R., Connolly, N. D. B., et al.
  *Mapping Inequality: Redlining in New Deal America.* Digital Scholarship Lab,
  University of Richmond, CC BY-NC-SA.
  <https://dsl.richmond.edu/panorama/redlining/> — **the polygons, the grade
  colors and the appraisal text quoted above are theirs.** Retrieved
  `r SR$fetched`: the national area file, `r n(SR$features)` areas across
  `r SR$city_surveys` city surveys, and the area description sheet for
  `r SR$sheet_key` area `r SR$sheet_label`. The maps and forms underneath are
  1930s US federal records.
- U.S. Census Bureau, 2020 Decennial Census, the P.L. 94-171 redistricting file.
  <https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/>
- U.S. Census Bureau, TIGER/Line 2020 tract boundaries, the outlines in Figure 2.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they open
in a spreadsheet, and they are yours to take further.

[`cities.csv`](data/derived/cities.csv), [`fig_holc_attr.csv`](data/derived/fig_holc_attr.csv), [`fig_holc_meta.csv`](data/derived/fig_holc_meta.csv), [`fig_holc_rings.csv`](data/derived/fig_holc_rings.csv), [`fig_mix.csv`](data/derived/fig_mix.csv), [`fig_quote.csv`](data/derived/fig_quote.csv), [`fig_source.csv`](data/derived/fig_source.csv), [`fig_zoom_attr.csv`](data/derived/fig_zoom_attr.csv), [`fig_zoom_holc.csv`](data/derived/fig_zoom_holc.csv), [`fig_zoom_tracts.csv`](data/derived/fig_zoom_tracts.csv), [`fig_zoom_win.csv`](data/derived/fig_zoom_win.csv), [`tracts.csv`](data/derived/tracts.csv)

**Academic literature**

- Aaronson, D., Hartley, D. and Mazumder, B. (2021). "The Effects of the 1930s
  HOLC 'Redlining' Maps." *American Economic Journal: Economic Policy* 13(4):
  355–392.
- Rothstein, R. (2017). *The Color of Law.* Liveright.

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
