---
title: "Abortion Opinion, 1980-2024"
type: brief
subtitle: "The most recent study first, and the fifty years underneath it"
author: "Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("abortion-opinion-code.R")
```

```{r setup, include=FALSE}
```

This chapter runs backwards from the usual order. It begins with what Americans said in the most recent study, then splits that by the two identities the companion chapters built, and only then asks what moved. The reason is that a trend line is unreadable until you know what a single row means.

It also depends on those two chapters rather than repeating them. Party identification is kept at seven points because the [party-id chapter](../party-id/party-id-brief.html) shows that collapsing it is where the leaners disappear. Ideological self-placement keeps the people who said *haven't thought much about it* as their own group, because the [ideology chapter](../ideology/ideology-brief.html) shows the codebook files that answer under **Valid**.

## What Americans said in `r LAST`

```{r nowtab}
```

Half of the people asked, `r p1(ALW)`%, take the fourth option: abortion as a matter of personal choice. `r p1(NEV)`% take the first, never permitted.

The interesting one is the third. Read it again as the respondent hears it:

```{r raw, comment="", class.output="rawblock"}
```

Option 3 is not a midpoint between never and always. It is a *permission with a gatekeeper* — abortion allowed, but only once someone has established that the need is real. A person choosing it has not split the difference; they have chosen a procedural answer to a question the other three options answer in principle. Keep it in view: it is where the last four years went.

## The same study, by party

```{r partytab}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.0}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The seven-point party scale against the abortion item, most recent study.** The buttons switch which of the four options is drawn; hovering a bar gives that group's whole distribution and its number of respondents. Option 4 is the one that gets reported, and the leaners are the third and fifth bars. Switch between the options and watch how much of the party gradient each one carries: Option 4 runs from `r p1(min(bp$pct_4))`% to `r p1(max(bp$pct_4))`%, a spread of `r p1(diff(range(bp$pct_4)))` points. Option 3 runs only from `r p1(min(bp$pct_3))`% to `r p1(max(bp$pct_3))`%, which is `r p1(diff(range(bp$pct_3)))` points. The gatekeeper answer is not flat across the scale, but it is roughly a quarter as steep, which is how it can absorb movement without registering as a partisan story.

The gradient runs the way you would expect, with one thing worth stopping on. **Leaning Republicans are further from the pro-choice position than weak Republicans**, at `r p1(pv("Independent-Republican"))`% against `r p1(pv("Weak Republican"))`%. The people who decline the party label hold the more conservative view on this issue than the people who accept it. That is the same pattern the party-id chapter found in vote choice, showing up in a second place.

Fold the leaners into "independent," as the file's own three-category column invites you to do, and you would report an independent group sitting near `r p1(pv("Independent-Independent"))`% while the actual unaffiliated middle is a much smaller group. The choice made two chapters ago changes what this table says.

## The same study, by ideology

```{r ideotab}
```

Ideology sorts abortion opinion more sharply than party does, from `r p1(iv("Extremely liberal"))`% among the extremely liberal to `r p1(iv("Extremely conservative"))`% among the extremely conservative.

Now the row that only exists because the ideology chapter refused to throw it away. People who said they **haven't thought much about** liberal and conservative are `r p1(DK_IDE)`% in favour of the always option. That sits between moderates (`r p1(iv("Moderate, middle of the road"))`%) and the slightly conservative (`r p1(iv("Slightly conservative"))`%), and nowhere near the midpoint of the scale.

They have a position. They have a fairly definite one. What they decline is the *vocabulary*. A common habit, dropping them as missing, would have removed `r nn(bi$respondents[bi$group == "Haven't thought much about it"])` people with opinions from a table about opinions.

> **Before you read on, commit to a view.** The party gradient and the ideology
> gradient both look strong. **Which of the two do you expect was already
> strong in 1980, and which one had to be built?** Write down your answer.

## What moved

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. All four options, every study year in which the item was asked.** Scrub across the plot to read all four shares at any year. The buttons below mute a series so the remaining ones can be compared without it. Hollow points mark 2008, where a documented split half-sample leaves the item largely missing, at `r p1(MISS08)`% missing that year. The line is drawn through it rather than bridged silently, because a gap that has been filled in looks exactly like data.

The top line is the one that gets reported, and it is the one that has moved least. *Always permitted* was `r p1(o(FIRST, "always"))`% in `r FIRST` and `r p1(ALW)`% in `r LAST` — a rise of `r p1(ALW - o(FIRST, "always"))` points across forty-four years, most of it arriving by the early 1990s.

Underneath, between `r PREV` and `r LAST`:

```{r changetab}
```

**The absolutist position roughly halved**, and the mass did not go to the opposite pole — *always permitted* fell slightly. It went into option 3, the permission with a gatekeeper, which gained `r p1(o(LAST, "need_established") - o(PREV, "need_established"))` points.

A reader watching only the headline number would conclude that nothing much happened between `r PREV` and `r LAST`. Something quite large happened, and it happened in the categories nobody reports.

## The gaps that had to be built

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. The party gap and the ideology gap in the share saying always permitted.** Democrats minus Republicans, leaners included on both sides; liberals minus conservatives. Higher means the two groups are further apart. Hovering reports the four shares behind the two differences, which is where the interesting part is.

Between `r FIRST` and `r LAST` the Democratic share saying *always* rose by `r p1(gp$dem_always[gp$year == LAST] - gp$dem_always[gp$year == FIRST])` points. The Republican share fell by `r p1(abs(gp$rep_always[gp$year == LAST] - gp$rep_always[gp$year == FIRST]))`. The gap widened mostly because one side moved.

In `r FIRST` the party gap on abortion was **`r p1(PG1)` points**. Democrats and Republicans held almost the same distribution of views. Abortion was an issue that cut across the parties rather than between them.

By `r LAST` the gap is **`r p1(PG2)` points**.

The ideology gap started higher, at `r p1(IG1)` points in `r FIRST`. Liberal and conservative already meant something about abortion when the party labels did not. It too widened, to `r p1(IG2)`.

This is the finding the chapter's order was designed to deliver. The aggregate looks like a country whose mind changed slowly. Split it by party and you see a country whose mind changed much less than its *sorting* did: roughly the same distribution of opinions, redistributed onto different teams.

## What this chapter cannot tell you

**2008 is `r p1(MISS08)`% missing on this item**, and that is documented rather than broken: a random half sample received a branching series of questions instead of the four-option card. Read the codebook note and your sample is halved for a known reason; miss it and 2008 looks like a collapse in response. The year is kept in the figures and drawn hollow rather than dropped.

Four options are not a scale. The distance from *never* to *rape, incest or life* is not the same quantity as the distance from *need established* to *always*, so this chapter reports shares and differences of shares, and never a mean. A mean of these four numbers would be arithmetic on labels.

Nothing here is weighted; `VCF0009z` sits in the file unused, so every percentage describes ANES respondents rather than the country.

The 2020-to-2024 movement invites a *Dobbs* story and this chapter does not tell one. Two studies four years apart cannot separate a decision's effect from everything else that happened, and the codebook records no wording change that would explain it either. What can be said is what the table says: the movement was out of the absolute prohibition and into the conditional permission, and not toward personal choice.

Finally, the party and ideology gradients are not independent of each other, and this chapter does not try to hold one constant while varying the other. Which of the two is doing the work is a real question and a different chapter.

## Sources

- American National Election Studies. *ANES Time Series Cumulative Data File
  (1948–2024)*, version of 5 February 2026. Ann Arbor, MI and Palo Alto, CA:
  University of Michigan and Stanford University.
  <https://electionstudies.org/data-center/anes-time-series-cumulative-data-file/>
- ANES. *Time Series Cumulative Data File Codebook*, same release. The four
  options and the 2008 split-sample note are quoted from it directly.
- Adams, Greg D. (1997). "Abortion: Evidence of an Issue Evolution."
  *American Journal of Political Science* 41(3): 718–737.
- Carmines, Edward G. and James A. Stimson (1989). *Issue Evolution: Race and
  the Transformation of American Politics.* Princeton University Press.
- Layman, Geoffrey C. (2001). *The Great Divide: Religious and Cultural
  Conflict in American Party Politics.* Columbia University Press.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_ideology.csv`](data/derived/by_ideology.csv), [`by_party.csv`](data/derived/by_party.csv), [`gaps.csv`](data/derived/gaps.csv), [`now.csv`](data/derived/now.csv), [`overtime.csv`](data/derived/overtime.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `overtime.csv` runs the same four answer categories across the whole series. Find the year each one peaks. Does the country look like it changed its mind, or like it stayed still?
- `gaps.csv` carries both `party_gap` and `ideology_gap` for every year. Plot them together. Which one opened first, and which is wider now?
- `overtime.csv` has a `pct_missing` column. Find the years where it is highest. Does a question get harder to answer when the politics get hotter?
- `by_party.csv` and `by_ideology.csv` split the most recent study two ways. Which split separates people more cleanly? Then ask which one a pollster would rather report.

