---
title: "The Same Question for Fifty-One Years"
subtitle: "84-355 Democracy's Data | Confidence in institutions, GSS 1973–2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.8)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `gss-confidence-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"Trust in institutions has collapsed."*

You have met surveys three times now: one instrument up close, one question
checked against a public record, and many polls at once. This is the fourth
thing a survey can be, and the rarest.

Since **1973**, the General Social Survey has read the same sentence to
Americans:

> *"I am going to name some institutions in this country. As far as the people
> running these institutions are concerned, would you say you have a great deal
> of confidence, only some confidence, or hardly any confidence at all in
> them?"*

Same words. Same three options. Thirteen institutions. **Fifty-one years, in 33
of which the question was asked.**

That is worth pausing on. Every other trend in this course has an asterisk —
the question changed, the mode changed, the sampling frame changed. Here almost
nothing changed, which means **when the number moves, something in the country
moved.**

# Steps 1-3 — The question, one record, and the scope

```{r load}
cf <- read.csv("data/confidence.csv",   stringsAsFactors = FALSE)
in_ <- read.csv("data/institutions.csv", stringsAsFactors = FALSE)

c(rows = nrow(cf), institutions = nrow(in_),
  first = min(cf$year), last = max(cf$year),
  survey_years = length(unique(cf$year)))
head(cf[cf$institution == "Congress", c("year", "n", "great_deal", "hardly_any")], 5)
```

One row per institution per survey year: the weighted share giving each answer,
and how many people were asked. **75,699 respondents, and 33 survey years in which
the battery was asked**, sit behind these 427 rows.

# Steps 4-5 — One series, and the conclusion you are about to draw

```{r congress}
g <- cf[cf$institution == "Congress", ]
g[g$year %in% c(1973, 1990, 2010, 2024), c("year", "great_deal", "only_some", "hardly_any")]
```

```{r congress-plot}
plot(g$year, g$great_deal, type = "b", pch = 19, col = "#C41230", ylim = c(0, 45),
     xlab = "", ylab = "% with a great deal of confidence in Congress")
lines(g$year, g$hardly_any, type = "b", pch = 1, col = "grey50")
legend("topright", c("a great deal", "hardly any"), col = c("#C41230", "grey50"),
       pch = c(19, 1), bty = "n", cex = 0.85)
```

**In 1973, 24.1% of Americans had a great deal of confidence in the people
running Congress. In 2024, 5.6%.**

Roughly one in four became roughly one in eighteen. Meanwhile the share saying
"hardly any" went the other way. This is the sentence everybody quotes, and here
it is with the receipts.

# Steps 6-7 — All thirteen, and the exception

> **Stop and commit to an answer first.** You have just seen Congress fall from
> 24.1% to 5.6%. The obvious summary is *Americans have lost confidence in their
> institutions.* Thirteen institutions were asked in the same interview. **How
> many of the thirteen do you expect to have fallen?**

**The interesting part is what did *not* happen**, and you only see it because
the same question was asked about thirteen things at once.

```{r change}
in_[order(in_$change), c("institution", "first_year", "last_year", "change")]
```

```{r change-plot}
o <- in_[order(in_$change), ]
par(mar = c(4, 12, 1, 2))
barplot(o$change, horiz = TRUE, names.arg = substr(o$institution, 1, 34), las = 1,
        cex.names = 0.72, col = ifelse(o$change > 0, "#2c7fb8", "#C41230"),
        xlab = "change in % 'a great deal', 1973 to 2024")
abline(v = 0)
```

- **Medicine fell furthest: −28.6 points.** More than Congress.
- **Organized religion −20.2**, Congress **−18.5**, the press and education
  close behind.
- **The military ROSE 10.6 points**, the only substantial gain.
- **Organized labor barely moved** (+1.3) — and it started near the bottom.
- **The scientific community is almost flat** (−4.8) across half a century.

```{r military}
m <- cf[cf$institution == "Military", ]
m[m$year %in% c(1973, 1980, 1991, 2002, 2010, 2024), c("year", "great_deal")]
```

Look at when the military moved. **32.7% in 1973**, at the end of Vietnam — and
**63.1% in 1991**, the year of the Gulf War. It has drifted down since without
ever returning to where it started.

**"Trust in institutions collapsed" is not what this data says.** Confidence in
*most* institutions fell, one rose sharply, and two hardly moved. A single
sentence about "institutions" averages over a pattern with structure in it.

# Steps 8-9 — Where everything stands now, and fifty-one years at once

```{r n}
n <- aggregate(n ~ year, data = cf[cf$institution == "Congress", ], sum)
plot(n$year, n$n, type = "h", lwd = 3, col = "#2c7fb8",
     xlab = "", ylab = "respondents asked about Congress")
```

Two things to notice before trusting any of the above.

**The survey is not annual.** GSS ran yearly, then biennially, and the gaps are
not evenly spaced. A line drawn through these points implies data between them
that does not exist.

**And 2021 was not a normal year.** GSS could not interview face to face during
the pandemic and moved online — the one break in an otherwise unbroken method.
The answers from that year come from a different instrument wearing the same
words, which is exactly the problem the rest of this lab is celebrating the
absence of.

```{r modes}
cf[cf$institution == "Congress" & cf$year >= 2018,
   c("year", "n", "great_deal", "hardly_any")]
```

# Step 10 — What you can now say

**You can say** that confidence in the people running most American
institutions fell substantially between 1973 and 2024, that the decline is not
confined to government or to politics, and that **Medicine — not Congress — lost
the most**.

**You can say it is neither universal nor irreversible.** The military rose,
and its series responds visibly to events: it peaks in 1991, the year of the
Gulf War. Whatever is driving the other eleven is not an irreversible drift in
national character.

**You cannot say** whether institutions became less trustworthy, whether people
became less trusting, or whether the meaning of "a great deal of confidence"
drifted over fifty years. The question measures a report, not a relationship.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different institution.** Pick one that interests you and plot its
whole series. Find the year it moved most, and go and find out what happened
that year. Does the event explain the move, or did you just find a story to fit?

**Option B — The bottom, not the top.** Everything above uses "a great deal."
Redo Step 6 with `hardly_any`. Do the institutions rank the same way? An
institution can lose enthusiasts without gaining enemies, and the two columns
tell different stories.

**Option C — Ceiling and floor.** Organized labor started at 16.1% and ended at
17.5% — but it may have moved a great deal in between. Plot it. What does a flat
first-to-last comparison hide, and what does that imply about the table in
Step 6?

**Option D — Your own question.** 13 institutions, 35 survey years, one
unchanged sentence.

# What this lab cannot tell you

- **Why anyone answered as they did.** No reasons here, only the distribution.
- **Whether the institutions changed or the standards did.** A falling number
  can mean worse performance or higher expectations, and the question cannot
  separate them. This is the deepest problem with any long attitude series.
- **What "the people running these institutions" means to a respondent.** The
  wording is about *people*, not the institution — and whether respondents honour
  that distinction is unknown.
- **Anything about non-respondents.** GSS response rates have fallen sharply
  over fifty years, like every survey's. The weights adjust for what is known;
  the people who stopped answering may differ in ways nobody measured.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Confidence in Congress went from one in four to one in eighteen. Which of the
  other twelve series surprised you more?
- The military rose while everything else fell. What would you need to know to
  explain that?
- A question asked identically for fifty-one years is rare and valuable. What is
  the cost of never updating the wording?

***

### Where the data came from

**General Social Survey**, NORC at the University of Chicago — cumulative file
1972–2024, release 3a. 75,699 respondents.
<https://gss.norc.org/us/en/gss/get-the-data.html>

The confidence battery has been asked with identical wording since 1973.
Estimates are weighted with `wtssps`, the only weight covering every year in the
file.

**The build script uses one package**, `haven`, and it is the only place in this
course that does: the cumulative file is a Stata 13+ `.dta`, which base R cannot
read. The script converts it once to the CSVs this lab reads.
