---
title: "Key: The Same Question for Fifty-One Years"
subtitle: "84-355 Democracy's Data | GSS confidence | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
cf  <- read.csv("data/confidence.csv",   stringsAsFactors = FALSE)
in_ <- read.csv("data/institutions.csv", stringsAsFactors = FALSE)
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`gss-confidence-brief.Rmd`) is now
a chapter rather than a parallel copy of the lab: prose organised by argument,
under idea headings, with three numbered figures and no numbered steps. The lab
(`gss-confidence.Rmd`) still runs Steps 1-10 with the code exposed. **Students
should read the brief first.** **Step numbers below refer to the lab**; the
chapter is cited by figure number or section title.


**This is the fourth survey lab and the one that completes the set.** One
instrument up close; one question checked against a public record; many polls at
once; and now **one question held still for fifty-one years.** Name that
sequence when you open — students should see that "a survey" has been four
different objects this term.

Every number below was run against the data files.

***

# The session in one sentence

Confidence in Congress fell from one in four Americans to one in eighteen, most
other institutions fell too, the military rose, and the reason we can say any of
that is that nobody touched the wording.

***

# The findings

```{r headline}
g <- cf[cf$institution == "Congress", ]
c(congress_1973 = g$great_deal[g$year == 1973],
  congress_2024 = g$great_deal[g$year == 2024],
  institutions  = nrow(in_),
  survey_years  = length(unique(cf$year)),
  respondents   = 75699)
in_[order(in_$change), c("institution", "change")]
```

- **Congress: 24.1% → 5.6%.**
- **Medicine fell furthest, −28.6.** Organized religion −20.2. Congress −18.5.
- **The military rose +10.6**, peaking at **63.1% in 1991**.
- **Organized labor +1.3**, the scientific community −4.8: essentially flat.

***

# The three things to get said

**1. The value here is the constancy, not the numbers.** Every other trend in
this course carries an asterisk — the question changed, the mode changed, the
frame changed. **Ask the room what it would take to run the same survey for
fifty-one years**, and what an organisation gives up by refusing to modernise a
question. The answer is comparability, and it is expensive.

**2. "Trust collapsed" is a sentence the data refuses.** Most fell; one rose
sharply; two barely moved. **The single-number version of this story is wrong in
an interesting way**, and it is wrong because it averages across institutions
with different histories. Same aggregation error as EAVS jurisdictions and
pooled redlining tracts, in a new costume.

**3. Medicine, not Congress, is the biggest fall.** Students assume politics is
the story. It is not: −28.6 for medicine against −18.5 for Congress. **Let them
be surprised, then ask what a fall in confidence in "the people running
medicine" is actually measuring** — hospitals, insurers, pharmaceutical
companies and doctors are all in there, and the respondent decides which.

**The line for the board:** a question that never changes is the only way to
know the country did.

***

# Handling the room

Lower-temperature than the discrimination unit, but two live wires:

- **The military result** will please some students and unsettle others. Keep it
  on the timing: it moves with wars, and the 1991 spike is unmistakable.
- **Confidence in science and medicine** invites 2020s arguments. The data ends
  in 2024 and the series is smoother than students expect. **Show the whole line
  before anyone theorises about the pandemic** — medicine was falling steadily
  from 1973, long before.

***

***

# Walking the steps

| Steps | What they do | Min | Watch for |
|---|---|---|---|
| **1-3** | The question wording, one record, and the scope | 8 | **'The people running these institutions'** — the ambiguity that runs through everything. |
| **4-5** | Congress, 1973 to 2024; then the conclusion they are about to draw | 8 | Let somebody say 'Americans have lost faith in everything.' You need it. |
| **6-7** | All thirteen ranked; then the military's actual series | 12 | **The pivot.** Medicine fell most, not Congress. The military rose. |
| **8-9** | Where everything stands now; fifty-one years at once | 9 | The trusted institutions are the technical ones. Ask why. |
| **10** | What you can say | 5 | A report, not a relationship. |

**Steps 6-7 are the pivot.** The military's 1991 peak is the single most useful number in the file: confidence responds to events, so this is not an irreversible drift in national character.

**If you are short on time**, protect the pivot and the closing step. The numbers are what make the argument concrete; the sequence is the argument.


# Option A — A different institution

Grade on method, not choice.

**3/4:** Plots one series, identifies its biggest move, offers an event.

**4/4:** Checks whether the "event" is plausible given the survey timing — GSS
fields early in the year, so an event in November shows up in the *next* survey,
which may be two years later. **Matching a spike to a headline requires knowing
when the interviews happened.**

**4/4, best:** Notices the danger in the exercise as set. With 13 institutions,
33 survey years and hindsight, **you can always find a story for any move**, and
a student who says they cannot distinguish explanation from post-hoc fitting has
given the right answer to a slightly unfair question. Reward that.

***

# Option B — The bottom, not the top

## Verified output

```{r b}
last <- do.call(rbind, lapply(split(cf, cf$institution), function(x) {
  a <- x[x$year == min(x$year), ]; b <- x[x$year == max(x$year), ]
  data.frame(institution = a$institution[1],
             d_great = round(b$great_deal - a$great_deal, 1),
             d_hardly = round(b$hardly_any - a$hardly_any, 1))
}))
last[order(last$d_great), ]
```

**3/4:** Redoes the ranking on `hardly_any` and reports it differs.

**4/4:** Names the asymmetry. An institution can shed enthusiasts into "only
some" without gaining enemies, or it can polarise — and **the two columns
separate indifference from hostility.** Congress does both; some institutions do
only the first.

**4/4, best:** Notices the middle category is doing hidden work. "Only some
confidence" absorbs everything, so **a three-point scale cannot tell a mild
critic from a fence-sitter**, and a fifty-year commitment to three options is
the price of comparability. That is the trade the whole lab is about, seen from
inside the instrument.

***

# Option C — Ceiling and floor

## Verified output

```{r c}
l <- cf[cf$institution == "Organized labor", c("year", "great_deal")]
c(first = l$great_deal[1], last = l$great_deal[nrow(l)],
  min = min(l$great_deal), max = max(l$great_deal),
  range = round(max(l$great_deal) - min(l$great_deal), 1))
```

Organized labor starts at `r round(l$great_deal[which.min(l$year)], 1)` and ends
at `r round(l$great_deal[which.max(l$year)], 1)` — **a net change of
+`r round(diff(range(c(l$great_deal[which.min(l$year)], l$great_deal[which.max(l$year)]))), 1)`
that conceals a range of more than ten points.**

**And it is the one change in the file that sampling error alone can account
for.** Figure 2 of the chapter, under *All thirteen, ranked by how far they
moved*, draws all thirteen institutions as a first-to-last dumbbell
with 95% intervals on each reading; organised labour is the only row whose
change is inside its own interval. The defensible sentence about that ranking is
therefore *eleven fell, one rose, and one did not move* — not *two rose*. The
intervals in that figure assume simple random sampling, which the GSS is not, so
they are if anything too narrow.

**3/4:** Plots it and observes the flat endpoints hide movement.

**4/4:** Draws the general lesson — **the ranked first-to-last table (the
chapter's *All thirteen, ranked by how far they moved*, Step 6 in the lab) is a
first-and-last comparison, which is the crudest possible summary of a 33-point
series** — and
notes that a different start or end year would reorder the rankings.

**4/4, best:** Proposes a better summary and defends it: a trend slope, or an
average of the first three and last three surveys, both of which are robust to
one odd year. **A student who notices that 2021 is the online-mode year and that
it sits near the end of every series has found the specific reason to distrust
endpoints here.**

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is refusing the single-sentence version — noticing
  that the institutions diverge and that averaging them is the error.
- **A 2** is "Americans have lost trust in everything." Two of thirteen series
  say otherwise and one goes the other way.
- **Reward anyone who flags the 2021 mode change** without being told.

## Anticipated questions

- *"Why does the series skip years?"* — GSS moved from annual to biennial in
  1994 for funding reasons, and 2020 was disrupted. The battery ran in **33** of
  the 51 years.
- *"What happened in 2021?"* — Face-to-face interviewing was impossible, so GSS
  went online. **The one methodological break in the series**, and it lands right
  where students want to draw conclusions about recent years.
- *"Is 'a great deal' the right measure?"* — It is one of three. Option B is
  exactly this question; do not settle it for them.
- *"Does this predict anything?"* — Not from here. Nothing in this file links
  confidence to behaviour — not turnout, not compliance, not vote choice.
- *"Why is medicine in a politics course?"* — Because it is the control. It
  shows that whatever happened, it was not confined to politics.

## Connections

- **The CES lab** — one instrument, up close, and the weighting problem.
- **The CPS turnout lab** — one question checked against a public record; there
  the survey could be *graded*, here it cannot.
- **Polls, models and markets** — many instruments at once.
- **EAVS and redlining** — averaging across units with different histories, and
  what it destroys.
