---
title: "Levels of Aggregation in Election Data"
type: chapter
subtitle: "Which rung of the ladder answers which question"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("levels-of-aggregation-code.R")
```

```{r setup, include=FALSE}
```

Somewhere in Alaska a voter ranked the Republican first and the Democrat second. `r n(cv("ranked Begich first and Peltola second"))` others made the same pair of marks in the same order, and `r n(cv("ranked Peltola first and Begich second"))` made it the other way round. Both numbers sit in a public file. Neither can be recovered from the vote totals of the precinct that voter lives in, or of the county, or of the state — not approximately, not with a better method, not ever. Add up the paper and the two numbers are gone.

Every claim about an election is a claim at some level of aggregation. *"Two-thirds of Alaskans used only one ranking"* is about ballots. *"This precinct swung twelve points"* is about precincts. *"Rural counties moved right"* is about counties. *"Trump won 312 electoral votes"* is about states.

These are not four resolutions of one picture. **Each step up the ladder permanently destroys a kind of question and permanently creates a new way to be wrong.** A photograph you zoom out of can be zoomed back in. An aggregate cannot. This chapter walks the ladder rung by rung, and then works one example in full: an estimate made across the gap, scored against a truth that is almost never available.

## The secret ballot is the deliberate destruction of individual-level data

When American states adopted the Australian ballot in the late nineteenth century they were not solving a data-management problem badly. They were making it impossible to know how any particular person voted, on purpose, because the alternative was a country in which employers and party bosses could check.

So the finest-grained thing that exists for almost every American election is **the precinct** — a few hundred to a few thousand ballots, added up. Everything we claim to know about *who* votes how, including everything a court is told about racially polarized voting under the Voting Rights Act, is an inference made across that gap. **The gap is the point of the institution and the problem of the method.**

## Where the data comes from, and what it is for

**No American institution publishes the same election at more than one level of aggregation**, because no American institution runs elections. The rungs of the ladder are produced by different bodies, with different obligations and no relationship to one another. A chapter about the relationship between levels therefore has to be assembled, and this one is assembled entirely from files other chapters of this book already downloaded, kept and described. If a number on this page is wrong, it is wrong in the chapter it came from too, which is the property worth wanting.

**Who is in it, and how they got there.** Every ballot cast, but almost always as somebody's total. A state publishes precinct returns because its own law says so; here that is Georgia, for 2020 and 2024. Counties reach the public through a private compiler who stapled fifty-one state files together. Congressional-district results are an academic reconstruction, and the count of precincts and jurisdictions comes from a federal survey of election offices. And the only file in which a row is a single ballot exists because Alaska adopted ranked-choice voting and cannot count the election without publishing one.

**Who it was made for.** Certification, not analysis. A return exists so that an official can sign a total and an outcome can take legal effect. Nobody who produced any rung of this ladder was trying to describe voters, and no column in any of these files was designed for the questions this chapter asks of them.

**What it costs to get.** No fees anywhere, and no single place either. The upper rungs are easy; the precinct rung must be assembled state by state in fifty-one formats; the ballot rung exists for one state. What this chapter inherits, it inherits whole: every file arrives from a folder next door with its acquisition history stripped off, and every judgment call those chapters made is silently stacked underneath this one.

**What has already been done to it.** Four decisions, and one of them removes rows. Candidate-name columns (`Kamala D. Harris`, typed by `r n(ld$units_in_georgia[ld$level=="county"])` county tabulation systems) become party roles, `dem` and `rep`, because no two elections share a column name. Shares are **two-party** shares, so third-party votes count toward no denominator (the bottom number of the fraction) anywhere on this page. The precinct key is county and precinct pasted together, because precinct names repeat across counties. And the Alaska ballots were collapsed to distinct ranking sequences, with precinct and tabulator identifiers dropped, before anything was saved — even the bottom rung here is already an aggregate.

Here are three Georgia precincts after those decisions, every column a role rather than a name:

```{r clean-levels}
```

```{r level-rows}
```

**`r n(gac - as.integer(tb[["precinct"]]))` precincts disappear** between arriving and being kept. They cast no two-party votes at all, so their Democratic share is a division by zero rather than a small number. Dropping them is right and it is not free: every precinct count on this page counts precincts *where somebody voted for a major party*, and the difference is `r n(gac - as.integer(tb[["precinct"]]))` places where people live.

## Seven rungs, from a mark on paper to a number on the evening news

`r n(ld$units_nationally[ld$level=="ballot"])` ballots were cast in 2024; `r n(ld$units_nationally[ld$level=="precinct"])` precincts counted them; `r n(ld$units_nationally[ld$level=="election jurisdiction"])` offices ran them; `r n(ld$units_nationally[ld$level=="county"])` counties, `r ld$units_nationally[ld$level=="congressional district"]` districts and `r ld$units_nationally[ld$level=="state"]` states report them; one nation sums them. **Eight orders of magnitude, and every step is one way.**

```{r ladder-tbl}
```

The bottom rung deserves a close look, because it is the only American election data in which a row is a person's ballot. Alaska's **cast vote record** carries every ranking each ballot contained, and `r n(cv("ballots in the file"))` ballots collapse to only `r n(cv("distinct ranking sequences"))` distinct sequences — `r pc(cv("bullet-voting rate among ballots that ranked anyone (%)"))`% of voters who ranked anyone ranked exactly one candidate and stopped.

```{r cvr-second}
```

The pair of numbers that opens this chapter is a crossing of first choice with second choice: a property of two marks on one piece of paper. **No total of any kind contains it.** A precinct that reports how many votes each candidate received has already thrown away which candidates appeared together on a ballot, in a way no later arithmetic undoes.

What the upper rungs buy is everything else. **Coverage**: the whole Alaska file is `r pc(100*cv("ballots in the file")/ld$units_nationally[ld$level=="ballot"], 3)`% of the country's 2024 ballots, two in every thousand, and there is no national cast-vote-record file and will not be one. **Comparability**: a presidential series running from 1864 exists at the state level and nowhere finer, because no smaller unit keeps its identity that long. **The unit of the institution**: the electoral college is counted in states and a House seat in districts, so for those questions the total is not a compromise but the object. And **stability**: a precinct of four hundred voters is mostly noise, and a county is not.

The climb itself is exact where it can be checked and quietly odd where it cannot. Summing all `r n(ld$units_in_georgia[ld$level=="precinct"])` Georgia precincts reproduces the state's county file to the vote. Summing the country's county rows lands `r n(abs(as.numeric(gsub("[^0-9-]","",cn$difference[4]))))` votes short of the certified national total — a seam between compilations that the [county-returns brief](../county-returns/county-returns-brief.html) takes apart. The "county" file is not even all counties: Alaska contributes `r nc$rows_that_are_not_counties[1]` state house districts and the District of Columbia `r nc$rows_that_are_not_counties[2]` wards.

And the district rung cannot be reached by adding at all, because precinct returns record no district. It is *rebuilt* by somebody assigning census blocks to districts — an estimate wearing the clothes of a count.

Read the same election at four rungs and one number survives while another dies:

```{r level-stats}
```

**The mean does not move** — `r pc(LS("precinct","mean_dem_share"),2)`% across `r n(LS("precinct","units"))` precincts, `r pc(LS("county","mean_dem_share"),2)`% across `r LS("county","units")` counties, `r pc(LS("state","mean_dem_share"),2)`% for the state. **The spread collapses**, from a standard deviation of `r pc(LS("precinct","sd_dem_share"))` points to `r pc(LS("county","sd_dem_share"))`: Georgia holds precincts at `r pc(LS("precinct","min_dem_share"))`% and `r pc(LS("precinct","max_dem_share"))`% Democratic, counties only at `r pc(LS("county","min_dem_share"))`% and `r pc(LS("county","max_dem_share"))`%. From precincts to counties, `r pc(100 - LS("county","variance_retained_pct"))`% of the variation in Democratic share disappears. That vanished variation is exactly what the next section tries, and fails, to get back.

## Delete the answer, then try to recover it

Georgia is unusual: it publishes the **joint distribution** of candidate by vote method, so a question like *what share of mail voters chose Biden* is not a guess. It is a table.

```{r truth}
```

Read the columns the other way round and **`r pc(EE("precinct","truth_by_mail"))`% of Georgia's 2020 mail votes went to Biden**, against `r pc(EE("precinct","truth_not_by_mail"))`% of everything else. That is the answer key.

**Now delete the table.** Keep only what every other state publishes: for each precinct, the share of votes cast by mail and the share cast for Biden. Fit the best line through that scatter and read it off at 0% mail and at 100% mail. That line is Goodman's **ecological regression**, the workhorse of voting-rights litigation since 1953.

```{r eco-d3, results='asis', eval=knitr::is_html_output()}
```

```{r eco-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

**Figure 1. Two lines through the same cloud of precincts.** Each dot is one of Georgia's `r n(nrow(ep))` precincts in 2020, placed by the share of its votes cast by mail and the share cast for Biden, with dot area following the precinct's votes. A scatter with two fitted lines is the shape this problem has, because the method's whole claim is that a slope across places reveals the behavior of people. So the honest picture is the cloud, the line Goodman fits to it (dashed), and the line the published truth draws through it (solid). The shaded band is the range a share can occupy. The dashed line leaves it.

**Goodman's regression, run on the precinct margins, says `r pc(EE("precinct","est_by_mail"))`% of Georgia's mail voters chose Biden** — against a published truth of `r pc(EE("precinct","truth_by_mail"))`%, off by `r pc(EE("precinct","error_by_mail"))` points, and a share that cannot exist. Nothing in the aggregate data announces this. The correlation is `r pc(EE("precinct","correlation"), 2)` and the fitted line runs straight through the middle of the cloud. But so does the true line. The two differ only in slope, and the slope is the entire answer.

The reason is not a bug. Precincts with a lot of mail voting were Democratic precincts, and Democratic precincts were more Democratic *among mail voters and election-day voters alike*. The regression cannot tell "mail voters are Democrats" from "Democratic places mailed more," so it charges the whole between-precinct difference to the method of voting. That is **the ecological fallacy**, guessing what individuals did from group totals, which Robinson (1950) demonstrated on literacy and immigration. Here it is in an election, with the answer key attached.

Climb one more rung and it gets worse. Aggregate the same ballots into counties and run the same method:

```{r eco-table}
```

**`r pc(EE("county","est_by_mail"))`%.** The same ballots, grouped more coarsely, and the answer moves a further `r pc(EE("county","est_by_mail") - EE("precinct","est_by_mail"))` points from a truth that never changed. The correlation *rises*, from `r pc(EE("precinct","correlation"),3)` to `r pc(EE("county","correlation"),3)` — the standard result that ecological correlations exceed individual ones, which is why they tempt. Is that a fact about voters, or about where Georgia drew its county lines? Regroup the same `r n(nrow(ep))` precincts into `r nrow(ec)` *random* groups, 500 times over:

```{r maup}
```

Random groups of the same size leave the correlation where it was, at `r pc(mp$correlation[3],3)` on average against `r pc(mp$correlation[1],3)` for the ungrouped precincts. Georgia's real counties push it to `r pc(mp$correlation[2],3)`, which none of the 500 regroupings reached. So the lift is produced not by adding up as such but by adding up along *these particular lines* — Openshaw's **modifiable areal unit problem**, the subject of [its own chapter](../../01-census-bureau/census-geography/census-geography-brief.html) in this book.

One thing the aggregate still supports honestly. Duncan and Davis (1953) showed the margins alone put hard bounds on the truth with no model at all, here `r pc(EE("precinct","bound_lower"))`%–`r pc(EE("precinct","bound_upper"))`% at the precinct level. Wide, useless for litigation, and **correct**, which the regression was not.

And the failure appears where it counts. In a Georgia county whose voter file records race, `r n(V(rp,"Black voters who took a Democratic ballot"))` of `r n(V(rp,"Black voters in the 17 precincts"))` Black voters took a Democratic primary ballot — a truth of `r pc(V(rp,"the truth: Democratic share of Black voters (%)"))`%. Goodman's regression on the same precincts returns `r pc(V(rp,"Goodman's ecological regression on the same precincts (%)"))`%: impossible again, this time by `r pc(V(rp,"overshoot, in percentage points"))` points, on exactly the kind of precinct evidence that *Thornburg v. Gingles* (1986) lets courts rest a racially-polarized-voting finding on. Freedman and colleagues (1991) pressed this objection in court; King (1997) built the standard method that tries to answer it; the argument is not settled, and the [RPV chapter](../../06-putting-data-together/rpv/rpv-brief.html) runs it in full.

## Choosing a rung

Everything above compresses into a single working rule: **name the rung before naming the finding**, and say why it is the rung the question lives on. Then check whether the answer survives being asked one rung up. Goodman's estimate did not — it moved `r pc(EE("county","est_by_mail") - EE("precinct","est_by_mail"))` points from precincts to counties, on ballots that never changed.

```{r decision}
```

## What this data cannot tell you

**It cannot tell you how any person voted.** That is not a limitation to engineer around; it is the secret ballot working. The demand for individual-level election data is a demand to partly undo a democratic achievement. Texas's 2024 emergency guidance on cast-vote-record releases, and the litigation ordering such releases elsewhere, are both defensible positions in that live conflict.

**One state carries most of the walk.** Georgia publishes precinct returns by vote method and most states do not. Whether Goodman's regression fails this badly elsewhere is not shown here and should not be assumed — what is shown is that the failure mode is real, large, and silent.

**One contest carries the bottom rung.** Alaska's cast vote record exists because the election was ranked-choice, so nothing here says what such a record would reveal about an ordinary plurality race.

**The district rung is borrowed rather than built.** It arrives as percentages with no vote counts, so its mean in the table above is unweighted while every other rung's is weighted. That is a difference in construction, not a finding about districts.

**And vote method is not a demographic.** "Mail voter" is a behavior with no census count behind it. The racially polarized case, where the group is fixed, the stakes are a federal remedy, and the answer key almost never exists, is harder than the case worked here, not easier.

## What you should have learned

- Election data comes on a ladder (ballot, precinct, jurisdiction, county, district, state, nation), and each step up is irreversible: the `r n(cv("ranked Begich first and Peltola second"))` Begich-then-Peltola ballots of the opening cannot be recovered from any total.
- Aggregation preserves the mean and destroys the spread: Georgia's Democratic share is `r pc(LS("precinct","mean_dem_share"),2)`% at every rung, while `r pc(100 - LS("county","variance_retained_pct"))`% of the precinct-level variation vanishes by the county rung.
- Ecological regression on Georgia's precinct margins estimated that `r pc(EE("precinct","est_by_mail"))`% of mail voters chose Biden, against a published truth of `r pc(EE("precinct","truth_by_mail"))`% — an impossible share, delivered without warning, and coarser grouping made it worse, not better.
- The grouping lines matter as much as the grouping size: 500 random regroupings of the same precincts averaged a correlation of `r pc(mp$correlation[3],3)`, and none reached the `r pc(mp$correlation[2],3)` produced by Georgia's actual county lines.
- The honest aggregate answer is a bound, not a point: the margins alone put Georgia's mail vote between `r pc(EE("precinct","bound_lower"))`% and `r pc(EE("precinct","bound_upper"))`% Democratic, which is wide, nearly useless, and correct.

## Where this data appears in this book

[What Election Returns Can and Cannot Show](../returns-source/returns-source-brief.html) leads this cluster and states the instrument's limits; this chapter supplies its ladder and its graded ecological exhibit. The cluster's briefs each work on one or two rungs:

- [Fifty-One Publishers, One Election](../county-returns/county-returns-brief.html) — what it costs to assemble the county file no agency publishes.
- [What the Panhandle Did](../panhandle-claim/panhandle-claim-brief.html) — a claim about race and a primary, tested against returns at several rungs.
- [Where Electoral Votes Are, and Where They Are Contested](../electoral-map/electoral-map-brief.html) — the state rung, where the electoral college lives.
- [Mapping Votes Against Land Area](../mapping/mapping-brief.html) — the county rung drawn four ways.
- [Mapping Vote Swing Between Elections](../wind-map/wind-map-brief.html) — county-level change, drawn as arrows.
- [A Table That Is Also Fifty-One Charts](../sparklines/sparklines-brief.html) — the state rung, fifty-one small trends.
- [Presidential Returns Across 150 Years](../historical-campaigns/historical-campaigns-brief.html) — the state rung, the only one with a 150-year series.
- [Distributions, and the Summaries That Hide Them](../distributions/distributions-brief.html) — the district rung, four hundred districts against one average.
- [How Much of 2024 Was Already Decided in 2020](../crossover/crossover-brief.html) — one election against the next at the state and district rungs.
- [One Election, Not Fifty](../nationalization/nationalization-brief.html) — nationalization, measured across rungs.
- [Bellwether Counties, 1960–2024](../bellwether/bellwether-brief.html) — the county rung, and why a perfect record predicts nothing.
- [The Arithmetic of Vote Targeting](../vote-targeting/vote-targeting-brief.html) — the precinct rung, put to campaign work.
- [Georgia Precinct Returns and the 2020 Recount](../ga-precinct-returns/ga-precinct-returns-brief.html) — the precinct file this chapter reads, and its vote-method columns.
- [Comparing Precincts Across Two Elections](../precinct-geography/precinct-geography-brief.html) — why the precinct rung does not hold still between elections.
- [Cast Vote Records and Ranked-Choice Counting](../cast-vote-records/cast-vote-records-brief.html) — the ballot rung, recounted from the Alaska file used here.
- [Residual Votes and Ballot Design](../residual-votes/residual-votes-brief.html) — the jurisdiction rung, and the subtraction no agency performs.
- [The Election Administration and Voting Survey](../eavs/eavs-brief.html) — the survey behind this chapter's counts of precincts and jurisdictions.

## Sources

**Data.** Nothing here was downloaded. Every number above is re-derived from
files other chapters of this book already downloaded and kept, and each of
those chapters says where its file came from. The publishers underneath are these.

- State of Alaska, Division of Elections, 2024 general election cast vote
  record — the bottom rung, one row per distinct ranking sequence.
  <https://www.elections.alaska.gov/results/24GENR/>
- Georgia Secretary of State, certified precinct returns for the 2020 and 2024
  general elections, with the 2020 vote-method breakdown.
  <https://sos.ga.gov/page/historical-elections-results>
- Tony McGovern, *United States General Election Presidential Results by
  District, Ward, or County from 2008 to 2024* — county and state returns for
  the whole country.
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>
- Jason Timm, *PresElectionResults* — results by congressional district for
  2024, and the national series back to 1864.
  <https://github.com/jaytimm/PresElectionResults>
- U.S. Election Assistance Commission, *Election Administration and Voting
  Survey* — the national counts of each kind of unit.
  <https://www.eac.gov/research-and-data/datasets-codebooks-and-surveys>
- Houston County, Georgia, General Primary of 21 May 2024, with self-reported
  race — the graded ecological case. A public record in Georgia, obtained as
  working material rather than downloaded, and described where this book first
  uses it, in the [RPV chapter](../../06-putting-data-together/rpv/rpv-brief.html).

Also cited: Texas Secretary of State, *Election Advisory No. 2024-20*, emergency guidance on voter privacy, 6 June 2024.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`consistency.csv`](data/derived/consistency.csv), [`cvr_facts.csv`](data/derived/cvr_facts.csv), [`cvr_second.csv`](data/derived/cvr_second.csv), [`decision.csv`](data/derived/decision.csv), [`eco_county.csv`](data/derived/eco_county.csv), [`eco_estimates.csv`](data/derived/eco_estimates.csv), [`eco_precinct.csv`](data/derived/eco_precinct.csv), [`ga_levels.csv`](data/derived/ga_levels.csv), [`ladder.csv`](data/derived/ladder.csv), [`level_stats.csv`](data/derived/level_stats.csv), [`maup_draws.csv`](data/derived/maup_draws.csv), [`maup_null.csv`](data/derived/maup_null.csv), [`method_truth.csv`](data/derived/method_truth.csv), [`not_counties.csv`](data/derived/not_counties.csv), [`rpv_check.csv`](data/derived/rpv_check.csv), [`seat_vote.csv`](data/derived/seat_vote.csv), [`suitability.csv`](data/derived/suitability.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `suitability.csv` scores each `question` against each `level` of the ladder in its `fit` column. Pick a question you care about and find the rung that fits. Then check what the published data actually offers at that rung.
- `eco_estimates.csv` shows what an ecological estimate gets wrong, with `truth_by_mail` beside `est_by_mail`. Work out the size of the error at each level. Would you have known it was wrong without the truth column?
- `maup_draws.csv` recomputes the same `correlation` on many random groupings. Look at the spread. How much of a published correlation is the grouping rather than the people?
- `ladder.csv` gives `share_of_voters_pct` and what each rung is `published_for`. Find the rung where publication stops. What questions die there?

**Academic literature**

- Robinson, W. S. (1950). "Ecological Correlations and the Behavior of
  Individuals." *American Sociological Review* 15(3): 351–357.
- Goodman, L. A. (1953). "Ecological Regressions and Behavior of Individuals."
  *American Sociological Review* 18(6): 663–664.
- Duncan, O. D. and Davis, B. (1953). "An Alternative to Ecological Correlation."
  *American Sociological Review* 18(6): 665–666.
- Goodman, L. A. (1959). "Some Alternatives to Ecological Correlation."
  *American Journal of Sociology* 64(6): 610–625.
- Openshaw, S. (1984). *The Modifiable Areal Unit Problem.* Concepts and
  Techniques in Modern Geography 38. Norwich: Geo Books.
- Freedman, D. A., Klein, S. P., Sacks, J., Smyth, C. A. and Everett, C. G.
  (1991). "Ecological Regression and Voting Rights." *Evaluation Review* 15(6):
  673–711.
- King, G. (1997). *A Solution to the Ecological Inference Problem:
  Reconstructing Individual Behavior from Aggregate Data.* Princeton: Princeton
  University Press.

- *Thornburg v. Gingles*, 478 U.S. 30 (1986).
