---
title: "Four Million People Who Are Not in the Denominator"
subtitle: "84-355 Democracy's Data | Felony disenfranchisement, 2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.8)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `disenfranchisement-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"Turnout was 63% of eligible voters."*

You have computed turnout rates in this course several times. Every one of them
was a fraction, and every fraction has a denominator, and **somebody decided who
goes in it.**

Today we look at the four million people that decision is about.

**Every state except Maine, Vermont and the District of Columbia takes the vote
away from at least some people with felony convictions.** The rules vary
enormously — some restore it at release, some at the end of parole, some after
probation, some never without a governor's pardon — and the result is that
whether you can vote depends on which side of a state line you were convicted
on.

# Steps 1-3 — A PDF source, the check that validates it, and one record

```{r load}
s <- read.csv("data/states.csv",   stringsAsFactors = FALSE)
n <- read.csv("data/national.csv", stringsAsFactors = FALSE)

nrow(s)
n
```

One row per state, and one for the country. The columns split the
disenfranchised population by where they are in the correctional system:
**prison, parole, felony probation, jail, and post-sentence.**

**4,049,978 Americans could not vote in 2024 because of a felony conviction —
1.70% of the voting-eligible population.**

# Steps 4-5 — The national total, and the column holding two fifths of it

> **Predict before you look.** Roughly four million adult citizens cannot vote
> because of a felony conviction. **What share of them do you think are
> currently in prison or jail?** Write down a number.

```{r post}
c(total = n$total,
  post_sentence = n$post_sentence,
  pct_of_disenfranchised = round(100 * n$post_sentence / n$total, 1))
```

```{r post-plot}
v <- c(n$prison, n$parole, n$probation, n$jail, n$post_sentence)
names(v) <- c("prison", "parole", "probation", "jail", "post-sentence")
barplot(sort(v), horiz = TRUE, las = 1, cex.names = 0.85,
        col = c(rep("#2c7fb8", 4), "#C41230"),
        xlab = "people disenfranchised, 2024")
```

**1,616,437 of them — 40% — have completed their sentences entirely.**

Not in prison. Not on parole. Not on probation. They have served the sentence a
court gave them, in full, and they still cannot vote. **The largest single
category of disenfranchised Americans is people who are no longer being
punished by any other means.**

That is not a fact about crime. It is a fact about **what each state decided to
do afterwards**, and it is the reason the totals below vary so wildly.

# Steps 6-7 — The range across states, and the whole country at once

```{r states}
head(s[, c("state", "total", "post_sentence", "pct")], 8)
tail(s[, c("state", "total", "pct")], 6)
```

```{r state-plot}
t10 <- head(s, 12)
par(mar = c(4, 9, 1, 2))
barplot(rev(t10$pct), horiz = TRUE, names.arg = rev(t10$state), las = 1,
        cex.names = 0.8, col = "#C41230",
        xlab = "% of the voting-eligible population disenfranchised")
abline(v = n$pct, lty = 2)
```

**Tennessee 7.68%. Florida 6.13%. Alabama 5.95%.** In Tennessee, roughly **one
adult in thirteen** cannot vote.

At the other end, **Maine and Vermont disenfranchise nobody at all** — people
vote from prison — and their rate is zero.

```{r drivers}
s$pct_post <- round(100 * s$post_sentence / pmax(s$total, 1), 1)
head(s[order(-s$total), c("state", "total", "post_sentence", "pct_post", "pct")], 8)
```

Look at what drives the high rates. Florida disenfranchises 961,757 people and
**729,858 of them are post-sentence.** Tennessee, 300,166 of 399,684. The states
at the top of the table are not the states with the most crime. **They are the
states that keep the penalty running after the sentence ends.**

# Steps 8-9 — Check the denominator too

Now return to where we started.

```{r denom}
c(national_rate = n$pct,
  tennessee = s$pct[s$state == "Tennessee"],
  florida = s$pct[s$state == "Florida"])
```

A turnout rate is votes divided by *some* population. Three are in common use:

- **Voting-age population (VAP)** — everyone 18+. Includes non-citizens and
  everyone disenfranchised.
- **Citizen voting-age population (CVAP)** — removes non-citizens. **Still
  includes all four million people here.**
- **Voting-eligible population (VEP)** — removes them too. This is the
  denominator the table above uses.

**The difference between CVAP and VEP turnout in Tennessee is not a rounding
error.** Excluding 7.68% of adults from the denominator raises the turnout rate
noticeably — and a state can look more or less civically healthy depending on a
choice made in a methods note.

You met this in the CPS turnout lab, where 2024 turnout was 64.4% or 58.5%
depending on the denominator. **This is where a chunk of that gap comes from.**

# Step 10 — What you can now say

**You can say** that roughly **four million adult citizens** cannot vote
because of a felony conviction, that **most of them are not incarcerated**, and
that **two fifths have completed their sentences entirely**.

**You can say the variation is a policy choice rather than a crime rate.** Maine
and Vermont face the same criminal justice system and disenfranchise nobody, and
ten states account for the entire post-sentence category.

**You cannot say who those people are.** This file carries no demographics, so
the racial disparities that motivate most of the literature are outside it. And
you cannot say how many would vote if permitted — that is a separate estimation
problem.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Rank by the right thing.** The table is sorted by rate. Sort by
`total` instead. Which states move most, and which number would you use to argue
that this is a national problem rather than a Southern one?

**Option B — Abolish post-sentence disenfranchisement.** Suppose every state
restored the vote at the end of a sentence. Recompute each state's rate using
`total - post_sentence`. How many states drop below 1%, and which stay high?
What does that tell you about where the policy bites?

**Option C — Put it against a margin.** Pick a state and find a statewide race
decided by less than the number of disenfranchised people there. Then say
carefully what that comparison does and does not establish — this is the
argument that gets made constantly and it needs handling with care.

**Option D — Your own question.** Fifty states, five correctional categories.

# What this lab cannot tell you

- **How any of these people would have voted.** The comparison in Option C is a
  comparison of magnitudes, not a claim about outcomes. Disenfranchised people
  are not a random sample and many would not have turned out.
- **Whether the estimates are right.** These are modeled figures, built from
  correctional data and state law by researchers, not a count of named people.
  The report documents the method; the uncertainty is real and not published as
  an interval.
- **Who is affected.** The report has separate tables for Black and Latino
  Americans and for men and women, showing rates several times the overall
  figure. **We used only Table 2.** The disparities are in the same PDF.
- **What the rules actually are.** "Post-sentence" bundles states that require a
  waiting period, a fee, a pardon, or nothing at all. The counts are comparable;
  the laws behind them are not.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Forty percent of disenfranchised Americans have finished their sentences.
  Should completing a sentence restore the vote automatically?
- Maine and Vermont let people vote from prison, and Tennessee bars one adult in
  thirteen. Is it defensible that this varies by state?
- Which denominator should a turnout rate use, and who should decide?

***

### Where the data came from

Uggen, Larson, Shannon and Stewart, **"Locked Out 2024: Four Million Denied
Voting Rights Due to a Felony Conviction"**, The Sentencing Project, October
2024 — Table 2.
<https://www.sentencingproject.org/reports/locked-out-2024-four-million-denied-voting-rights-due-to-a-felony-conviction/>

**The Sentencing Project publishes a report, not a dataset.** These numbers
exist only as a table on page 18 of a PDF, and `data/build-data.py` extracts
them. The parse is checked rather than trusted: **every state's five categories
must sum to its total, and all fifty do.** That check caught a real error —
Texas reports numbers wide enough to overflow their column, and a naive
fixed-width read splits them in half.

The District of Columbia is not in Table 2 and so is not here.
