---
title: "Counting Talk at the Supreme Court"
type: brief
subtitle: "Six defensible answers to who talks the most, and one number that turns out to be about a rule"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("oral-argument-code.R")
```

```{r setup, include=FALSE}
```

> *"You could tell from the questioning which way it was going."*

Reporters write that sentence every term, and there is research behind it: the side that draws the harder questions tends to lose. Any version of the claim rests on a count, and before anybody can count, somebody has to decide what a question is, whose turn it was, and how long it lasted. This brief asks how much of an apparent fact about nine people survives that decision.

It matters because oral argument is the only part of the Supreme Court's work an ordinary citizen can watch happen. The Court publishes no video, but it does publish the audio. Every claim about a hot bench or a silenced advocate is a count somebody made from that tape, and the counts get written up as facts about the justices.

Most of them are not. Ask who talks the most in `r NMEAS` defensible ways and `r length(FIRSTS)` different justices come first. And the sharpest movement anywhere in the file, a number going from zero to a hundred and a twenty-year silence ending, was not a change in anybody's behaviour at all.

## The test

Everything below comes from [Oyez](https://api.oyez.org), a free public archive of the Court's audio run by Cornell's Legal Information Institute, Justia and Chicago-Kent College of Law. The build reads `r n(NARG)` arguments across the `r T0` to `r T1` terms, which is every argument the Roberts Court has heard, and `r n(NTURN)` speaking turns inside them.

This source is the right one for what it is not. The [roll-call chapter](../rollcalls-source/rollcalls-source-brief.html) sets out the clean case, a record in which the act is the data, and names the Supreme Court Database as the courts' version of it. Almost nothing in that database is an observation: whether a decision was liberal or conservative is a coder's reading of an opinion. Oral argument is the exception. Oyez publishes the audio lined up with the transcript, one row per speaking turn, with a start and a stop time on every row.

Nobody had to read anything and reach a conclusion to know that `r EXNAME` first asked a question `r nm(EXT$start, 2)` seconds into the argument in *`r EXCASE`*. Oyez is a teaching archive, built so the public could hear the Court rather than read about it. Using it as a measuring instrument is a repurposing, and a fortunate one: an archive made for listeners keeps exactly the details a count needs.

## The data

The file is not a table. It arrives as nested text, one turn inside a section inside a case, and this is one turn as it comes down.

```{r rawturn}
```

`start` and `stop` are seconds. **Nobody decided them.** A machine lined the words up against the tape. That is a real difference from a coded column, and it is smaller than it looks.

**A turn is a decision, and a person made it.** The timestamps are attached to units a court reporter drew. Somebody decided where one turn stops and the next begins, whether an interruption counts as a turn of its own, and which of two people talking across each other got the words. Oyez then attached a name to every turn, which is a second judgment on top of the first. None of that is sloppiness: there is no way to turn a recording of nine people interrupting each other into rows without deciding what a row is. **The judgments moved upstream**, out of the column you want to analyse and into the definition of the row.

Two kinds of justice turn are not participation, and this build drops them from every count below: gavelling an argument in or out, and courtesy, which is "Thank you, counsel" and "No questions." The Chief Justice opens and closes every argument, so counting those would make whoever holds that office the busiest questioner on the Court by arithmetic rather than by behaviour. Dropping them removes `r nm(ck$value[ck$check == "share of turns dropped as administrative or courtesy is under 10%"], 2)`% of all turns here.

Seven tables come out of that. `arguments.csv` holds one row per argument, with `minutes`, `turns`, `justice_turns`, `justice_words` and a `regime` naming the rules the argument was heard under. `measures.csv` ranks the justices `r NMEAS` ways, with a `measure`, a `value` and a `rank`. `order.csv` gives `pct_in_seniority_order` by term, `silence.csv` gives `pct_silent` per justice per term, and `speakers.csv` gives each justice a `seniority_rank`.

Here are the `r NMEAS` ways of asking who talks the most, over the `r RW[1]` to `r RW[2]` terms, the window with the current nine justices in it. Every one of them is defensible, and every one is computed from the same turns.

```{r measure-list}
```

> **Before you look at the figure, commit to a number.** Take the nine justices
> and rank them by how many words they say in an average argument. Now rank
> them instead by how many seconds of the clock they personally hold. Those
> sound like the same question asked twice. Across all `r NMEAS` measures,
> **how many places does the biggest mover travel**, from their best rank to
> their worst? Then say how many of the nine hold exactly the same rank on all
> `r NMEAS`.

## What it says about democracy

```{r bump-d3, results='asis', eval=knitr::is_html_output()}
```

```{r bump-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 1. The same nine justices, ranked `r NMEAS` ways, on the `r RW[1]` to `r RW[2]` terms.** A bump chart fits this question because what matters is not any one number but whether a justice's *rank* holds still. Rank 1 is at the top, each line is one justice and keeps its identity across every column, so a line that slopes is a justice whose answer depends on which question you asked. Nine lines is more than the eye can separate by colour, so names sit at both ends of each one. On screen the lines start grey and hovering raises one; in print the three the text discusses are red.

```{r measure-table}
```

`r MOVER` is the whole argument in one line. Ranked by `r MOVER_HI`, `r short(MOVER)` is **`r ord(BEST)`** on the Court. Ranked by `r MOVER_LO`, the same justice over the same terms is **`r ord(WORST)`**. Nothing about the underlying turns changed, and the question changed by a few words.

**`r MOVE` places is the widest swing, and `r NMOVED` of the `r NJ` justices move at all.** Taking the floor often and holding it a long time are different habits. A justice who fires twenty short questions and one who asks three long ones can end with the same word count and opposite rankings on everything else.

**The ends of the table do not move**, and that matters more. `r sub(",.*", "", TOP3)` is in the top three on every measure, and `r sub(",.*", "", LAST)` is last on every measure. The scrambling is all in the middle, where the differences between justices are small enough that the choice of measure decides the order.

There is an unusually clean outside check on that. Lee Epstein and Eric Posner built the same kind of file from Westlaw transcripts rather than from Oyez. They ran 18 measures rather than `r NMEAS`, and reported the same justice in the top three on 17 of them.

```{r epcompare}
```

Two teams, two sources, two parsers written by people who never spoke, and the justice word count for the `r T0` term differs by `r abs(round(WRD0) - 3866)` word in `r n(round(WRD0))`. The thing being counted survives who counts it, which is not true of most measures in this book. The share of decisions coded conservative in the [Supreme Court chapter](../scdb/scdb-brief.html) would not survive a second team recoding it. **Word counts survive because a word is not a judgment.**

Everything so far has been a fact about nine people. The next number is not.

Start with the sentence everybody writes: the justices are talking more than they used to. It is true. An argument ran `r nm(MIN0)` minutes on average in the `r T0` term and `r nm(MIN1)` in `r T1`. But the justices speak at `r nm(RAT0)` words a minute in `r T0` and `r nm(RAT1)` in `r T1`, and that rate barely moves across the whole `r NTERM` terms. The bench did not get hotter. The container got bigger, by `r nm(GROWTH, 0)`% since the Court stopped meeting in its courtroom.

On 4 May 2020 the Court heard argument by telephone for the first time, and gave itself a rule it had never had. After the advocate's opening, the justices would question one at a time, in order of seniority, with the Chief going first. That rule leaves a mark anybody can measure. If the justices ask their first questions in seniority order, the list of who spoke first, second and third is sorted.

```{r order-d3, results='asis', eval=knitr::is_html_output()}
```

```{r order-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

**Figure 2. Two shares of each term's arguments.** In blue, the share where the justices asked their first questions in exact seniority order. In red, the share where Justice Thomas asked nothing. Both are divided by the same bottom number, the arguments heard that term, which is why they belong on one axis. The shaded band is the `r TELT` term, argued by telephone.

**For `r NZERO` straight terms the blue line is exactly zero. In the telephone term it is `r nm(ORD(2020))`%.** Then the Court went back to its courtroom and it fell to `r nm(ORD(2021))`% again. The `r SPLIT` term is the seam and reads `r nm(ORD(SPLIT))`%, because it was argued in the courtroom until March and by telephone in May.

Nothing about the nine people produced that. They did not become more deferential to seniority in 2020 and abandon it in 2021. **An administrative decision about how to run a hearing on a conference call moved a social statistic from zero to a hundred and back.**

The red line is Justice Thomas. He asked nothing in `r nm(100 * TH_SILENT_PRE / TH_ON_PRE)`% of the arguments he sat for through 2018, and in `r TH_TERMS_100` of those terms he asked nothing in a single argument. He was famous for it, and the fame came with explanations about temperament and about what he thought argument was for. Then the format handed every justice a turn by name, and his silence went to `r nm(TH_2020)`%. It has stayed near there since, at `r nm(100 * sum(TH_POST$arguments_silent) / sum(TH_POST$arguments_on_bench))`% across every argument.

**Whatever that silence had been, it was at least partly a fact about a room in which you had to interrupt to be heard.** A measure of a person turned out to be measuring an arrangement of chairs. **Ask of any statistic about people: could a rule have produced this?** Here the answer was yes, and it hid behind a sentence everybody found easy to believe.

## What this brief cannot tell you

**Nothing here measures persuasion.** Talking is not the same as changing an outcome, and the justice who says least may be the one who decides the case. This brief can tell you who held the floor. It cannot tell you who won the argument. The claim it opened with, that the questioning shows which way a case is going, needs the outcomes joined to the transcripts and a design this file does not supply.

**A turn is a transcription decision**, drawn by a court reporter, so every count of turns inherits a definition nobody here chose. The rule for dropping gavel and courtesy turns is this build's own, and it removes a different share than Epstein and Posner's does.

**The bench roster is a period, not a day.** Counting silence needs a roster, and a transcript has none: a justice who says nothing leaves no row in it. The roster here is Oyez's record of who was sitting over a stretch of terms, so a justice recused from one case reads as one who kept quiet in it.

**The seniority measure looks only at first questions.** The Court kept a seniority round when it returned to the courtroom, but put it at the end of each advocate's time rather than the start. A measure reading zero in 2021 and zero in 2015 is not saying the same thing about those two terms.

## What you should have learned

- Six reasonable ways to ask who talks the most produce `r length(FIRSTS)` different answers to the question, and the widest swing by one justice is `r MOVE` places.
- The extremes hold and the middle does not: `r sub(",.*", "", TOP3)` is top three on every measure, `r sub(",.*", "", LAST)` is last on every one, and `r NMOVED` of `r NJ` justices move in between.
- Arguments got longer without the justices getting faster, from `r nm(MIN0)` minutes in `r T0` to `r nm(MIN1)` in `r T1` at a near-constant `r nm(RAT1)` words a minute.
- A rule change moved the share of arguments questioned in seniority order from zero to `r nm(ORD(2020))`% and back, and ended a silence that had run `r nm(100 * TH_SILENT_PRE / TH_ON_PRE)`% of one justice's arguments.
- Two teams counting different transcripts agreed to within `r abs(round(WRD0) - 3866)` word in `r n(round(WRD0))`, because a word count is not a coding judgment.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `measures.csv` ranks the justices `r NMEAS` ways, in `measure`, `value` and `rank`. Find the justice whose `rank` changes most. Which measure would you use in a sentence, and what would you be claiming?
- `arguments.csv` has `minutes`, `turns` and `justice_words` per case, with a `regime` column. Average each of them inside each `regime`. What did the rule change actually change?
- `silence.csv` has `pct_silent` per justice per `term`. Sort it, find the justices who said nothing most often, and check whether that changed under the new format.
- `order.csv` gives `pct_in_seniority_order` by `term`, and `speakers.csv` gives each justice a `seniority_rank`. Who gains from a fixed speaking order, and who loses?

Two stretch ideas point past the spreadsheet. Oyez covers argument back to 1955, so download a term from before the Roberts Court and ask whether argument length was already growing. And the Court's own site posts same-day audio, so count a term of it yourself and compare your totals against `arguments.csv`.

## Sources

**Data**

- **Oyez**, oral argument audio aligned to transcripts, `r T0`-`r T1` terms.
  <https://api.oyez.org> — keyless and public. Each transcript is broken into
  turns, and every turn carries a start and stop time in seconds.
  `r n(NARG)` arguments and `r n(NTURN)` turns are read. The bench roster is
  Oyez's record of the natural court.

**Compared against, but not used**

- Epstein, L. and Posner, E. A. (2026). *Two Decades of Oral Argument in the
  Supreme Court, 2005-2025 Terms.* Report prepared for the *New York Times*,
  12 August 2026. <https://epstein.wustl.edu/s/oralargumentreport.pdf> The
  underlying corpus is built from Westlaw transcripts and is not public, so the
  figures quoted from it are read off the report and labelled in the table that
  uses them.

**Academic literature**

- Epstein, L. and Posner, E. A. (2026). "What Is Oral Argument in the Supreme
  Court For?" <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=7227018>
- Epstein, L., Landes, W. M. and Posner, R. A. (2010). "Inferring the Winning
  Party in the Supreme Court from the Pattern of Questioning at Oral Argument."
  *Journal of Legal Studies* 39(2): 433-467.
- Johnson, T. R., Wahlbeck, P. J. and Spriggs, J. F. (2006). "The Influence of
  Oral Arguments on the U.S. Supreme Court." *American Political Science Review*
  100(1): 99-113.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`arguments.csv`](data/derived/arguments.csv), [`exhibit.csv`](data/derived/exhibit.csv), [`measures.csv`](data/derived/measures.csv), [`order.csv`](data/derived/order.csv), [`silence.csv`](data/derived/silence.csv), [`speakers.csv`](data/derived/speakers.csv), [`totals.csv`](data/derived/totals.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
