---
title: "Counting Talk at the Supreme Court"
type: brief
subtitle: "Six defensible answers to who talks the most, and one number that turns out to be about a rule"
author: "Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("oral-argument-code.R")
```

```{r setup, include=FALSE}
```

> *"The justices are talking more than they used to."*

That sentence has been true for five years and it has been written about the wrong thing. This chapter is about the difference.

The sibling chapter in this part built a score for the Supreme Court out of votes. It had to say, over and over, that almost nothing in the Supreme Court Database is an observation. Whether a decision was liberal or conservative is a coder's reading of an opinion. So is the issue area, and so is the reason the Court took the case. **The Court publishes opinions, and opinions are not data until somebody makes them into data.**

Oral argument is the exception, and it is why this chapter sits beside that one. The Court records the argument itself. Oyez publishes that audio lined up with the transcript, one row per speaking turn, with a start time and a stop time on every row. Nobody had to read anything and reach a conclusion to know that `r EXNAME` first asked a question `r nm(EXT$start, 2)` seconds into the argument in *`r EXCASE`*.

So here is the same institution, measured twice. One file is made of judgments. The other is made of a clock. This chapter asks what the second kind buys you, and the answer is less than you would think and something better than you expected.

## What arrived

Everything below comes from Oyez, a free public archive of the Court's audio run by Cornell's Legal Information Institute, Justia and Chicago-Kent College of Law. The build reads `r n(NARG)` arguments across the `r T0` to `r T1` terms: every argument the Roberts Court has heard, and `r n(NTURN)` speaking turns inside them.

None of it was assembled with counting in mind. Oyez is a teaching archive, built so that students and the public could hear the Court rather than read about it. Its editorial product is precisely the object measured here: the transcript aligned against the audio, with a speaker's name on every turn. Using that as a measurement instrument is a repurposing, and a fortunate one. An archive made for listeners keeps exactly the details a listener needs, and they turn out to be the details a count needs too.

The file is not a table. It arrives as nested text, and the shape of it is the first thing worth seeing. Here is one turn, exactly as it comes down:

```{r rawturn}
```

Read down that and compare it with the row of the Supreme Court Database printed in the sibling chapter. There, every value except the name and the date was an integer standing for a word somebody chose. Here, `start` and `stop` are seconds. **Nobody decided them. A machine lined the words up against the tape.**

That is a real difference and it is worth naming. It is also smaller than it looks.

## The clock is honest. The turn is not.

**A turn is a decision, and a person made it.** The timestamps are attached to units that a court reporter drew. Somebody decided where one turn stops and the next begins. Somebody decided whether an interruption counts as a turn of its own. When two people talk across each other, somebody decided which of them got the words.

Oyez then attached a name to every turn, which is a second judgment on top of the first. None of that is sloppiness. There is no way to turn a recording of nine people interrupting each other into rows without deciding what a row is.

So the honest description is narrower than "this file has no judgments in it." It is this: **the judgments moved upstream.** In the Supreme Court Database they sit in the column you want to analyse. Here they sit in the definition of the row, which is easier to forget about and just as consequential.

One more decision is this build's own. Two kinds of justice turn are not participation and are dropped from every count below.

```{r drops}
```

The Chief Justice opens and closes every argument. Counting those turns would make whoever holds that office the busiest questioner on the Court by arithmetic rather than by behaviour. Dropping them removes `r nm(ck$value[ck$check == "share of turns dropped as administrative or courtesy is under 10%"], 2)`% of all turns here. Lee Epstein and Eric Posner (2026) drop the same two categories and report that it removes 4% of theirs. The two rules are close enough to be looking for the same thing, and far enough apart to be worth saying out loud.

## Who talks the most?

This is the question every story about a hot bench is really asking, and it has no single answer. Below are `r NMEAS` ways to ask it, over the `r RW[1]` to `r RW[2]` terms, which is the window with the current nine justices in it. Every one of them is defensible. Every one is computed from the same turns.

```{r measure-list}
```

> **Before you look at the figure, commit to a number.** Take the nine justices
> and rank them by how many words they say in an average argument. Now rank them
> instead by how many seconds of the clock they personally hold. Those sound like
> the same question asked twice. Across all `r NMEAS` measures, how far does one
> justice move? Say how many places, from best rank to worst, the biggest mover
> travels. Then say how many of the nine hold exactly the same rank on all
> `r NMEAS`.

```{r bump-d3, results='asis', eval=knitr::is_html_output()}
```

```{r bump-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 1. The same nine justices, ranked `r NMEAS` ways, on the `r RW[1]` to `r RW[2]` terms.** Rank 1 is at the top. Each line is one justice and keeps its identity across every column, so a line that slopes is a justice whose answer depends on which question you asked. Nine is more than the eye can separate by colour, so nobody gets a colour of their own and names sit at both ends of every line. On screen every line starts grey and hovering raises one. In print the three justices the text discusses are the ones drawn in red.

```{r measure-table}
```

`r MOVER` is the whole argument in one line. Ranked by `r MOVER_HI`, `r short(MOVER)` is **`r ord(BEST)`** on the Court. Ranked by `r MOVER_LO`, the same justice over the same terms is **`r ord(WORST)`**. Nothing about the underlying turns changed. The question changed by a few words.

**`r MOVE` places is the widest swing, and `r NMOVED` of the `r NJ` justices move at all.** Almost nobody stands still. If you have ever read that some justice is the most active questioner on the Court, that sentence had a measure behind it, and the writer probably did not tell you which one.

The reason is not mysterious once you see it. Taking the floor often and holding it a long time are different habits. A justice who fires twenty short questions and one who asks three long ones can end up with the same word count and opposite rankings on everything else.

## Where the measures agree

Now the other half, which matters more.

```{r stable}
```

**The ends of the table do not move.** `r sub(",.*", "", TOP3)` is in the top three on every measure. `r sub(",.*", "", LAST)` is last on every measure. The scrambling is all in the middle, where the differences between justices are small enough that the choice of measure decides the order.

This is the same lesson the sibling chapter reached from votes, arriving from a different direction. There, an ordering of the justices was stable at the level of blocs and unstable at the level of neighbours. Here, a ranking is stable at the extremes and unstable in the middle. **The coarse structure is in the data. The fine structure is in the choice of measure.**

There is an outside check available on that, and it is unusually clean. Epstein and Posner ran 18 measures of participation rather than `r NMEAS`, on a different set of transcripts, and reported that the same justice placed in the top three on 17 of them. Two teams, two sources, two lists of measures, and the same shape.

## The number that is not about the justices

Everything so far has been a fact about nine people. This section is not, and telling the difference is the skill worth taking out of this chapter.

Start with the sentence at the top: the justices are talking more than they used to. It is true. In the `r T0` term an argument ran `r nm(MIN0)` minutes on average. In `r T1` it ran `r nm(MIN1)`. The justices got through `r n(WRD0)` words in an average `r T0` argument and `r n(WRD1)` in an average `r T1` one, which is half as much again.

```{r longview-d3, results='asis', eval=knitr::is_html_output()}
```

```{r longview-static, eval=!knitr::is_html_output(), fig.height=4.0}
```

**Figure 2. How long an argument runs, and how fast the justices fill it.** The top panel is the mean length of an argument in each term. The bottom panel is the number of words the justices got through per minute of argument. The shaded band is the `r TELT` term, argued by telephone. The two panels are drawn one above the other rather than on a shared frame. Minutes and words a minute are different units, and putting them on one plot would suggest a relationship the data does not contain.

Look at the bottom panel. **The justices speak at `r nm(RAT0)` words a minute in `r T0` and `r nm(RAT1)` in `r T1`.** That rate barely moves across the whole `r NTERM` terms. They are not talking faster, and in any given minute they are not talking more. The argument simply lasts longer, by `r nm(GROWTH, 0)`% on average since the Court stopped meeting in its courtroom.

The bench did not get hotter. The container got bigger.

## What changed on 4 May 2020

On that date the Court heard argument by telephone for the first time, and gave itself a rule it had never had. After the advocate's opening, the justices would question one at a time, in order of seniority, with the Chief going first.

That rule leaves a mark you can measure. If the justices ask their first questions in seniority order, the list of who spoke first, second and third is sorted. So count the arguments where it is.

```{r order-d3, results='asis', eval=knitr::is_html_output()}
```

```{r order-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

**Figure 3. Two shares of each term's arguments.** In blue, the share where the justices asked their first questions in exact order of seniority. In red, the share where Justice Thomas asked no question at all. Both are divided by the same bottom number, the arguments heard that term, which is why they belong on one axis.

```{r orderfacts}
```

**For `r NZERO` straight terms the number is exactly zero. In the telephone term it is `r nm(ORD(2020))`%.** Then the Court went back to its courtroom and it fell to `r nm(ORD(2021))`% again.

The `r SPLIT` term is the seam, and it reads `r nm(ORD(SPLIT))`%. That term was argued in the courtroom until March and by telephone in May, so it is the only one in the file with both rules inside it.

Nothing about the nine people produced that. They did not become more deferential to seniority in 2020 and abandon it in 2021. **An administrative decision about how to run a hearing on a conference call moved a social statistic from 0 to 100 and back.**

The fall back to zero is worth a second look, because the rule did not go away. The Court kept a seniority round when it returned, but put it at the *end* of each advocate's time, after the open questioning. So the first questions went back to being a free-for-all while the ordered part moved somewhere this measure does not look. **A measure that reads 0% in 2021 and 0% in 2015 is not saying the same thing about those two terms.**

## The other thing the rule did

The red line in Figure 3 is Justice Thomas.

```{r thomas}
```

Thomas asked nothing in `r nm(100 * TH_SILENT_PRE / TH_ON_PRE)`% of the arguments he sat for through 2018. In `r TH_TERMS_100` of those terms he asked nothing in every single argument. He was famous for it, and the fame came with explanations about temperament and about what he thought argument was for.

Then the Court went to the telephone, the format handed every justice a turn by name, and his silence went to `r nm(TH_2020)`%. It has stayed near there: `r nm(100 * sum(TH_POST$arguments_silent) / sum(TH_POST$arguments_on_bench))`% across every argument since.

**Whatever that silence had been, it was at least partly a fact about a format in which you had to interrupt to be heard.** A measure of a person turned out to be measuring an arrangement of chairs. That is the single most useful thing in this chapter, and it generalises well past the Supreme Court.

One caution belongs with those numbers. Counting silence needs a roster, and a transcript has none: a justice who says nothing leaves no row in it. The bench list here comes from Oyez's record of who was sitting in that stretch of the Court's history, so a justice recused from one case reads as a justice who kept quiet in it. The first version of this build did not notice, and reported that no justice had ever been silent, which is what a file that only records talking will always tell you.

## Checking it against somebody else's file

The measures above have an unusual advantage: somebody else built the same thing from different paper, three days before this chapter was written.

Epstein and Posner assembled the Roberts Court Oral Argument Corpus from transcripts bought from Westlaw and published a report on it for the *New York Times* on 12 August 2026. **Their file is not public and cannot be**, because Westlaw licenses it, so nothing in this chapter is built on their data. What follows is a comparison, and the left column is the only place in this chapter where a number came from somewhere other than this book's own build.

```{r epcompare}
```

Two teams, two sources, two parsers written by people who never spoke. The justice word count for the `r T0` term differs by `r abs(round(WRD0) - 3866)` word in `r n(round(WRD0))`.

That is worth more than it looks. It says the thing being counted is robust to who counts it, which is not true of most measures in this book. The share of decisions coded conservative in the sibling chapter would not survive a second team recoding it. **Word counts survive because a word is not a judgment.**

The differences that remain are also informative rather than embarrassing. This build finds `r n(NARG)` arguments against their 1,425, because Oyez is missing audio for a handful. The turn counts differ by a fraction of a percent for the same reason.

## What this evidence will bear

The record is complete and it is not a sample. Every argument the Roberts Court has heard is here, and the counts inside them are reproducible by anybody with the same file. Two independent teams got the same numbers, which is the strongest form of check available and one most chapters in this book cannot run.

The limits bound different claims, and each has appeared above in its place. **A turn is a transcription decision**, drawn by a court reporter, so every count of turns inherits a definition nobody in this chapter chose. **The rule for dropping gavel and courtesy turns is this build's own**, and it removes a different share than Epstein and Posner's does. **The bench roster is a period, not a day**, so a recusal reads as a silence. And **the seniority measure looks only at first questions**, which is why it reads zero for two terms that are nothing alike.

The largest limit is not about the data at all. **Nothing here measures persuasion.** Talking is not the same as changing an outcome, and the justice who says least may be the one who decides the case. This chapter can tell you who held the floor. It cannot tell you who won the argument, and the temptation to slide from one to the other is exactly what a clean, complete, timestamped file makes easy.

Two questions stay open. Whether the longer post-2020 argument produces better decisions is not answerable from a transcript, and nobody has proposed a way to answer it. And whether the seniority round improved the hearing or merely made it orderly is a question about what oral argument is for, which Epstein and Posner take up in the paper this report accompanies.

What can be said is worth the trouble. **The same nine justices, the same `r n(NARG)` arguments, and six reasonable ways to ask who talks the most produce `r length(FIRSTS)` different answers to that question.** The extremes hold and the middle does not. And the sharpest movement in the whole file, a number going from zero to a hundred and a justice's twenty-year silence ending, was not a change in anybody's behaviour. It was a change in the rules of the room.

**Ask of any statistic about people: could a rule have produced this?** For most of them the answer is no. For this one it was yes, and it was hiding behind a sentence everybody found easy to believe.

## Sources

**Data**

- **Oyez**, oral argument audio aligned to transcripts, `r T0`-`r T1` terms.
  <https://api.oyez.org> — keyless and public. Three endpoints are read: the case
  list for each term, each case, and each case's argument transcript. The
  transcript is turn-level, and every turn carries a start and stop time in
  seconds. `r n(NARG)` arguments and `r n(NTURN)` turns are read; two kinds of
  justice turn, gavelling and courtesy, are excluded from every count, by a rule
  the build script states.
- The bench roster used for the silence table is Oyez's record of the natural
  court, which is the set of justices sitting over a stretch of terms rather than
  on a given day.

**Compared against, but not used**

- Epstein, L. and Posner, E. A. (2026). *Two Decades of Oral Argument in the
  Supreme Court, 2005-2025 Terms.* Report prepared for the *New York Times*,
  12 August 2026. <https://epstein.wustl.edu/s/oralargumentreport.pdf> The
  underlying Roberts Court Oral Argument Corpus is built from Westlaw
  transcripts and is not publicly available, so the six figures quoted from it
  here are read off the published report and are labelled in the table that uses
  them.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`arguments.csv`](data/derived/arguments.csv), [`exhibit.csv`](data/derived/exhibit.csv), [`measures.csv`](data/derived/measures.csv), [`order.csv`](data/derived/order.csv), [`silence.csv`](data/derived/silence.csv), [`speakers.csv`](data/derived/speakers.csv), [`totals.csv`](data/derived/totals.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `measures.csv` ranks the justices six different ways. Find a justice whose rank changes a lot between measures. Which measure would you use, and what would you be claiming?
- `arguments.csv` has `minutes`, `turns` and `justice_words` per case, with a `regime` column. Compare the regimes. What did the rule change actually change?
- `silence.csv` has `pct_silent` per justice per term. Find the justices who said nothing most often. Then check whether that changed under the new format.
- `order.csv` gives `pct_in_seniority_order` by term. Follow it across the regime change. Who benefits from a fixed speaking order?

**Academic literature**

- Epstein, L. and Posner, E. A. (2026). "What Is Oral Argument in the Supreme
  Court For?" <https://papers.ssrn.com/sol3/papers.cfm?abstract_id=7227018>
- Epstein, L., Landes, W. M. and Posner, R. A. (2010). "Inferring the Winning
  Party in the Supreme Court from the Pattern of Questioning at Oral Argument."
  *Journal of Legal Studies* 39(2): 433-467.
- "How a Covid-Era Scramble Transformed Supreme Court Arguments." *New York
  Times*, 13 August 2026, the article the report above was prepared for.


**Further reading**

- Johnson, T. R., Wahlbeck, P. J. and Spriggs, J. F. (2006). "The Influence of
  Oral Arguments on the U.S. Supreme Court." *American Political Science Review*
  100(1): 99-113.
```{r ai-prompt, echo=FALSE, results="asis"}
```
