---
title: "Lab 1: The 2024 Electoral Map"
subtitle: "84-355 Democracy's Data | Week 1"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

You do not need to know any R. Every block of code below already works. Your job
today is to **run it, read it, and then change it**.

Two things to know about this document:

- The grey blocks are **code chunks**. Click the little green ▶ arrow at the
  top-right of a chunk to run it. Or put your cursor in the chunk and press
  **Cmd+Shift+Enter** (Mac) / **Ctrl+Shift+Enter** (Windows).
- Text outside the chunks is just writing. This whole file is *R Markdown*: code
  and prose in one document. When you press **Knit**, R runs every chunk in order
  and produces a finished report.

Nothing here needs any add-on packages. It is all base R.

**Work in order.** Later chunks depend on objects created earlier.

***

# Part 1: Read the data, and look at it

Our dataset is the 2024 presidential election, one row per state (plus the
District of Columbia). Read it in:

```{r read}
elec <- read.csv("data/pres2024_states.csv", stringsAsFactors = FALSE)
```

That line did something quiet but important: it took a file on disk and made it
an object in R called `elec`. Nothing printed, because we didn't ask it to.

Let's ask. `head()` shows the first few rows:

```{r head}
head(elec)
```

How big is it?

```{r dim}
dim(elec)
nrow(elec)
```

51 rows — 50 states and DC. And nine columns:

```{r names}
names(elec)
```

Here is what they mean:

| Column | Meaning |
|---|---|
| `state`, `abbrev` | State name and two-letter abbreviation |
| `ev` | Electoral votes (based on the 2020 Census) |
| `harris`, `trump`, `other` | Percent of the vote for each |
| `winner` | Who carried the state |
| `col`, `row` | Position on a grid — we'll use this in Part 4 |
| `margin` | `trump` minus `harris`. Positive = Trump won it. |

`summary()` gives a quick numerical portrait of every column:

```{r summary}
summary(elec[, c("ev", "harris", "trump", "margin")])
```

> **Pause and read that output.** The `margin` column runs from about $-80$ to
> about $+45$. What kind of place do you suppose is at $-80$? Check your guess in
> the next section.

***

# Part 2: Asking the data questions

## Pulling out one column

A dollar sign gets you a single column:

```{r column}
elec$ev
```

And square brackets get you specific rows. The pattern is
`data[rows, columns]`:

```{r subset-one}
elec[elec$abbrev == "PA", ]
```

Read that literally: "give me the rows of `elec` where the `abbrev` column equals
`PA`, and all columns."

## Sorting

`order()` returns the positions that would sort a column. Feed that back in as
the row index and you get a sorted table. Which states were closest?

```{r closest}
closest <- elec[order(abs(elec$margin)), ]
head(closest[, c("state", "harris", "trump", "margin", "ev")], 10)
```

Those are the states the campaigns actually fought over. Compare against where
the candidates spent their time — we'll do exactly that in Week 6.

And the least competitive:

```{r furthest}
furthest <- elec[order(-abs(elec$margin)), ]
head(furthest[, c("state", "harris", "trump", "margin", "ev")], 5)
```

There is your $-80$: the District of Columbia.

## Filtering on a condition

Which states did Harris win?

```{r harris-states}
harris_states <- elec[elec$winner == "Harris", ]
nrow(harris_states)
harris_states$state
```

Twenty jurisdictions — 19 states plus DC.

***

# Part 3: A first picture

A table of 51 rows is hard to see. Let's draw the ten closest states.

```{r barplot-close, fig.width = 7, fig.height = 5}
top10 <- head(elec[order(abs(elec$margin)), ], 10)
top10 <- top10[order(top10$margin), ]

barplot(top10$margin,
        names.arg = top10$abbrev,
        col = ifelse(top10$margin > 0, "#B2182B", "#2166AC"),
        border = NA,
        ylim = c(-6, 6),
        ylab = "Margin (Trump % - Harris %)",
        main = "The ten closest states, 2024")
abline(h = 0)
```

Every bar is under six points. Seven of these are the states that decided the
election.

***

# Part 4: Building a map out of rectangles

Now the main event. We are going to draw a map — but not a geographic one.

A normal map of the United States draws states at their true size. That is
honest about *land* and misleading about *votes*: Wyoming is enormous on the page
and worth 3 electoral votes; Rhode Island is invisible and worth 4. Land does not
vote.

So we'll draw every state as an **equal-sized square**, laid out roughly where it
belongs geographically. This is called a tile grid. The `col` and `row` columns
in our data give each state a position.

First, a colour for each state. We build a ramp from blue through white to red,
then pick a colour based on the margin:

```{r colours}
pal <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(101)

# Squash margins bigger than +/- 30 points so one blowout doesn't wash out
# the scale, then convert to a position from 1 to 101 along the ramp.
m <- pmax(pmin(elec$margin, 30), -30)
elec$fill <- pal[round((m + 30) / 60 * 100) + 1]
```

Now draw it. `plot(NULL, ...)` opens an empty canvas; `rect()` adds the squares;
`text()` labels them.

```{r tilemap, fig.width = 8, fig.height = 6}
plot(NULL,
     xlim = c(0.4, 11.6), ylim = c(8.6, 0.4),   # y reversed: row 1 at the top
     asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "2024 presidential election: every state the same size")

rect(elec$col - 0.46, elec$row - 0.46,
     elec$col + 0.46, elec$row + 0.46,
     col = elec$fill, border = "white", lwd = 2)

text(elec$col, elec$row, elec$abbrev, cex = 0.75)
```

That is a map of the *election*, not of the land.

## Now size the squares by electoral votes

One more version. Instead of equal squares, make each state's square proportional
to its electoral votes — its actual weight in choosing a president:

Because *area* is what the eye reads, the side of each square is proportional to
the **square root** of the electoral votes. That way a state with four times the
votes gets four times the area, not sixteen.

```{r tilemap-ev, fig.width = 8, fig.height = 6}
s <- 0.46 * sqrt(elec$ev / max(elec$ev))

plot(NULL,
     xlim = c(0.4, 11.6), ylim = c(8.6, 0.4),
     asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "Same map, sized by electoral votes")

rect(elec$col - s, elec$row - s,
     elec$col + s, elec$row + s,
     col = elec$fill, border = "white", lwd = 2)

# Shrink the labels along with the squares, and drop them where they won't fit.
text(elec$col, elec$row,
     ifelse(s > 0.14, elec$abbrev, ""),
     cex = pmin(1.7 * s, 0.8))
```

> **Compare the two maps.** Which states grew? Which shrank?
>
> Notice that most of the labels disappeared. Roughly half the states are too
> small to name. It is tempting to conclude that those states barely matter — and
> for the *outcome*, largely true. But hold that thought, because the opposite is
> also true: measured **per person**, a voter in one of those tiny squares has
> more say in choosing a president than a voter in California. Both statements
> come from the same data.
>
> In Week 3 we will compute exactly how much more a Wyoming voter is worth.

***

# Part 5: Counting to 270

The map shows who won where. It does not show who won. For that we add up
electoral votes:

```{r ev-total}
tapply(elec$ev, elec$winner, sum)
```

Trump 312, Harris 226. That is the certified result, and 538 in total.

But look closely at what we just did. We assigned **every** one of a state's
electoral votes to whoever carried that state. That rule is called
*winner-take-all*, and it is how 48 states do it.

**Maine and Nebraska are different.** They give two electoral votes to the
statewide winner and one to the winner of each congressional district. In 2024
both of them split:

- **Maine** went to Harris statewide, but Trump carried Maine's 2nd district.
- **Nebraska** went to Trump statewide, but Harris carried Nebraska's 2nd
  district — the "blue dot" around Omaha.

So our winner-take-all arithmetic *should* be wrong. Why did it give the right
answer?

```{r me-ne}
elec[elec$abbrev %in% c("ME", "NE"), c("state", "winner", "ev")]
```

Work it through. Winner-take-all hands Harris all 4 of Maine's votes and none of
Nebraska's. Reality gave her 3 from Maine and 1 from Nebraska. Four either way.
**The two errors cancelled.**

That is luck, not a rule. In 2020 the same calculation would have been off, and
in a close election it could flip the outcome — Nebraska's single district vote
was one of the reasons the 2024 race was watched so carefully in Omaha.

> **The lesson, and it is the lesson of this whole course:** a model that
> produces the right number is not necessarily a model that is right. Always ask
> *why* it worked before you trust it on the next case.

***

# Part 6: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option below is a working chunk with one thing to change. Run it, look at what
comes out, and tell me what it means.

Pick **one**.

***

### Option A — Colour the map by vote share instead of margin

Everything above coloured states by the *margin*. This colours by Harris's
**share**. Run it, then compare against the map in Part 4.

```{r opt-a, fig.width = 8, fig.height = 5.5}
# The map in Part 4 coloured the margin on a ramp 30 points wide either side of
# a tie. To compare the two maps you must hold the ramp fixed, so this one uses
# the same +/- 30 points -- but measured as Harris's distance from 50 instead.
dev_a  <- pmax(pmin(50 - elec$harris, 30), -30)
fill_a <- pal[round((dev_a + 30) / 60 * 100) + 1]

plot(NULL, xlim = c(0.4, 11.6), ylim = c(8.6, 0.4),
     asp = 1, axes = FALSE, xlab = "", ylab = "",
     main = "Harris share of the vote, 2024")
rect(elec$col - 0.46, elec$row - 0.46, elec$col + 0.46, elec$row + 0.46,
     col = fill_a, border = "white", lwd = 2)
text(elec$col, elec$row, elec$abbrev, cex = 0.75)
```

**Write about:** the country looks less divided on this map than on the margin
map. Which states changed the most between the two? Which map would you use if
you wanted to argue America is deeply polarised — and which if you wanted to
argue the opposite? Is either map lying?

***

### Option B — Where the electoral votes actually are

Change the `10` below to any number you like and re-run.

```{r opt-b, fig.width = 7, fig.height = 5}
n <- 10                                    # <- change this number

top_n <- head(elec[order(-elec$ev), ], n)
top_n <- top_n[order(top_n$ev), ]

par(mar = c(5, 8, 4, 2))
barplot(top_n$ev, names.arg = top_n$state, horiz = TRUE, las = 1,
        col = ifelse(top_n$winner == "Trump", "#B2182B", "#2166AC"),
        border = NA, xlab = "Electoral votes",
        main = paste("The", n, "largest states"))

sum(top_n$ev)                              # their combined electoral votes
round(100 * sum(top_n$ev) / 538)           # as a percent of 538
```

**Write about:** how few states does it take to reach 270? Does that number
surprise you? What does it imply about where a campaign should spend its money —
and is that the same as where it *should* spend its attention?

***

### Option C — What if we abolished winner-take-all?

This splits each state's electoral votes in proportion to its vote instead of
giving them all to the winner.

```{r opt-c}
h_share  <- elec$harris / (elec$harris + elec$trump)
h_prop   <- round(h_share * elec$ev)
t_prop   <- elec$ev - h_prop

c(Harris = sum(h_prop), Trump = sum(t_prop))   # proportional allocation
tapply(elec$ev, elec$winner, sum)              # actual, winner-take-all
```

**Write about:** who wins under each rule, and by how much? Proportional
allocation is often proposed as a fairer system. Looking at these two numbers,
what does "fairer" actually mean here — and fairer to whom? (If you are feeling
sharp: add up `sum(h_prop) + sum(t_prop)` and check it still comes to 538. Should
it always? That question is the subject of the Week 3 lab.)

***

### Option D — Ask your own question

If something in this lab made you curious, follow it. Change whatever you like,
or just interrogate a table we already made. A clear question answered with a
plot you barely modified beats an elaborate chart with nothing to say.

***

**What to submit.** Knit this document to HTML or PDF and upload it to Canvas,
with your chosen chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a result and say something true and interesting
about American democracy.

***

## Where the data came from

State-level returns are from the `PresElectionResults` dataset
(<https://github.com/jaytimm/PresElectionResults>), which compiles certified
state results. Electoral vote counts reflect the 2020 Census apportionment. The
tile-grid coordinates are a standard layout. The script that assembled
`pres2024_states.csv` is in `data/build-data.R`, so you can see exactly how it
was made — you should always be able to ask that question of any dataset.
