---
title: "Opening Session Key: The 2024 Electoral Map"
subtitle: "84-355 Democracy's Data | Intro: the three layers (Tue Aug 25) | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
elec <- read.csv("data/pres2024_states.csv", stringsAsFactors = FALSE)
pal <- colorRampPalette(c("#2166AC", "#F7F7F7", "#B2182B"))(101)
```

**All code in Part 6 is given to students and runs as-is.** They are not writing
it. What you are grading is the two-to-three paragraphs of interpretation
underneath. These notes are therefore about *what a good answer says*, not about
whether the code worked.

Every number below has been run against the actual data.

***

# Option A — Colour by share instead of margin

## What they will see

Both maps use the same data and the same palette, but very different scales:

- **Margin** runs from $-83.8$ (DC) to $+45.8$ (Wyoming) — a spread of ~130
  points.
- **Harris's share** runs from 25.8% (Wyoming) to 90.3% (DC) — a spread of ~65.

Halving the spread halves the visual drama. The share map is noticeably paler and
the blowout states stop screaming.

## What a good answer says

**3/4:** Notices the share map looks less polarised, and identifies that the
lopsided states (Wyoming, West Virginia, Idaho, Vermont, Maryland, DC) are where
the two maps differ most, while the battlegrounds barely move.

**4/4:** Gets to the real point — **neither map is lying, and the choice of scale
is an argument.** Margin exaggerates: a 70–30 state and a 60–40 state sit 20
points apart on a margin scale but only 10 apart on a share scale, so the margin
map draws the same gap twice as wide. (Both maps in the lab now use the identical
ramp — 30 points either side of a tie — so the paleness of the share map is a
real comparison and not an artefact of two different scales.) A student who says "I would use the margin map for
a story about polarisation and the share map for a story about a divided but
overlapping country, and I should tell my reader which I chose" has understood
the entire lab.

**Push in discussion:** ask which map a cable news channel would use, and why.

***

# Option B — Where the electoral votes are

## Verified output

```{r optb-check}
o <- elec[order(-elec$ev), ]
cum <- cumsum(o$ev)
k <- which(cum >= 270)[1]
k                                    # states needed to reach 270
data.frame(state = o$state[1:k], ev = o$ev[1:k], running = cum[1:k])
```

At the default `n <- 10`: those ten hold **254** electoral votes, or **47%** of
538. Trump carried seven of them (153 EV) and Harris three (101 EV).

**The headline finding is that just 12 states reach 270** — California, Texas,
Florida, New York, Illinois, Pennsylvania, Ohio, Georgia, North Carolina,
Michigan, New Jersey, Virginia. Twelve of fifty-one.

```{r optb-split}
tapply(o$ev[1:k], o$winner[1:k], sum)
```

## What a good answer says

**3/4:** Reports the number (12), registers surprise, and connects it to campaign
resource allocation.

**4/4:** Sees the catch. Those twelve states are *not* the campaign's target list,
because most of them were never competitive — California, New York, Illinois,
Texas, Ohio were all decided by double digits. The states that decide elections
are the intersection of *large* and *close*, which in 2024 meant Pennsylvania,
Georgia, North Carolina, and Michigan. A student who separates "where the votes
are" from "where the votes are in play" has found the distinction that Chapter 5
is built on.

**Common stumble:** some will change `n` to 51 and report that all states hold
100% of the electoral votes. Treat this generously and redirect — they ran the
code correctly and asked a trivial question. That is a 2, and a conversation.

***

# Option C — Proportional allocation

## Verified output

```{r optc}
h_share <- elec$harris / (elec$harris + elec$trump)
h_prop  <- round(h_share * elec$ev)
t_prop  <- elec$ev - h_prop

c(Harris = sum(h_prop), Trump = sum(t_prop))
tapply(elec$ev, elec$winner, sum)
sum(h_prop) + sum(t_prop)
```

Proportional: **Harris 264, Trump 274**. Winner-take-all: **Harris 226, Trump
312**. The margin collapses from 86 to 10. Trump still wins.

## What a good answer says

**3/4:** Reports both numbers and observes that proportional allocation tracks
the national vote far more closely.

**4/4:** Interrogates "fairer." Proportional allocation is fairer to *voters in
landslide states* — a Republican in California and a Democrat in Wyoming would
finally count for something. It is emphatically **not** fairer to small states,
which keep their two-senator bonus under either rule. And it does not touch the
malapportionment at the heart of the system. A student who notices that the
reform changes *which* inequity you get, rather than removing inequity, is
thinking like a political scientist.

**The bonus question** (`sum(h_prop) + sum(t_prop)`) comes to 538 here. It need
not in general — each state is rounded independently. Anyone who spots this has
independently discovered the **apportionment problem** and should be told so by
name, and pointed at Session 2 (the decennial census).

***

# Option D — Their own question

Grade on the quality of the question, not its ambition. "Why does DC have three
electoral votes when it has no senators?" answered with one line of code is a 4.
An elaborate unmotivated chart is a 2.

***

# Grading

The four-point scale from the syllabus. On this first submission specifically:

- **Do not penalise anything code-related.** The code is yours. If a chunk failed
  to run, that is a setup problem to fix in office hours, not a grade.
- **Do penalise description-without-a-point.** "The chart shows California is the
  biggest state" is a 2 no matter how clean the output.
- Expect a wide spread on this first one and grade generously; the point is to establish
  that the *writing* is what counts. Several students will submit code-heavy,
  interpretation-light work out of habit from other courses. Say so in the
  feedback and the following week's submissions will improve.

## Anticipated questions

- *"Why are there 51 rows?"* — DC has had three electoral votes since the 23rd
  Amendment (1961), while having no voting representation in Congress. Good
  Option D material.
- *"Why is `other` zero almost everywhere?"* — **Only 9 of 51 states report a
  nonzero third-party vote in this dataset.** That is a reporting artifact, not
  reality: zero means "not separately tabulated," not "nobody voted third party."
  Worth raising with the whole class, because it is the first live example of the
  provenance problem that Session 1 is entirely about.
- *"Can I use ggplot2 / the tidyverse?"* — yes if they already know it, but make
  clear it earns nothing extra. The labs are base R so that nobody has to install
  anything.
- *"Is this a coding class?"* — no. Say it early and often; the syllabus says it
  too. Some students will arrive braced for a programming course and need to hear
  that they are not behind.

## Connections forward

- The **two maps** (equal-size vs. EV-sized) set up malapportionment in **Session 2**.
- The **`other` column** sets up data provenance in **Session 1**.
- **Rounding in Option C** sets up apportionment in **Session 2**.
- The **ME/NE cancellation** in Part 5 is the course's first instance of a model
  that is right for the wrong reason — a theme worth naming again in **Session 9**
  (polls, models, and markets).
