---
title: "Dec 3 Wrap Key: Finding a Gerrymander Without Looking at a Map"
subtitle: "84-355 Democracy's Data | Thu Dec 3 — Test 3, then the wrap | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
cd <- read.csv("data/pres_by_cd_2024.csv", stringsAsFactors = FALSE)
st <- do.call(rbind, lapply(split(cd, cd$state), function(s) {
  data.frame(state = s$state[1], n = nrow(s),
             vote = round(mean(s$dem_share), 1),
             seats = sum(s$dem_share > 50),
             seat_pct = round(100 * mean(s$dem_share > 50), 1))
}))
st$eg <- round((st$seat_pct - 50) - 2 * (st$vote - 50), 1)
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

A gerrymander is visible in the *distribution* of district partisanship — a hole
where the competitive districts should be — and you can find it without ever
looking at a map, which is fortunate, because the shapes do not reliably tell you
anything.

***

# The findings

```{r headline}
c(districts = nrow(cd), mean_dem = round(mean(cd$dem_share), 1),
  dem_seats = sum(cd$dem_share > 50),
  competitive = sum(abs(cd$dem_share - 50) <= 5))
big <- st[st$n >= 6, ]
rbind(head(big[order(big$eg), ], 4), head(big[order(-big$eg), ], 4))
```

- **49.9%** mean district Democratic share; Democrats carry **205 of 434**
  (47.2%).
- Only **91 districts (21%)** are within five points of even.
- Worst Republican-favouring: **NC −25.6**, **WI −25.4**.
- Worst Democratic-favouring: **MA +22.4**, **IL +19.0**.
- **PA −1.7** — nearly unbiased.

***

# The idea to teach: the empty middle

```{r empty-middle}
for (ab in c("NC", "WI", "PA")) {
  s <- sort(cd$dem_share[cd$state == ab], decreasing = TRUE)
  cat(ab, ":", paste(sprintf("%.1f", s), collapse = "  "), "\n")
}
```

North Carolina: three districts at **73.5, 73.0, 67.5**, then a cliff to **48.4**
and eleven districts between 39.0 and 48.4. Wisconsin: **76.2, 70.4**, then
nothing until **47.7**.

Pennsylvania has districts at 52.8, 50.2, 48.4, 47.4, 45.7 — the middle is
populated.

**Packing** wastes the other side's surplus votes in a few overwhelming
districts. **Cracking** wastes the rest in many narrow losses. The signature is
bimodality with a gap, and students can see it in a printed row of numbers.

This is the answer to the assigned NYT reading: compactness is not the test.
North Carolina's districts are not notably ugly.

***

# Option A — Diagnose a state

## Verified output for the default (Ohio)

```{r a}
s <- cd[cd$state == "OH", ]
s <- s[order(-s$dem_share), ]
s[, c("district", "dem_share", "house_rep_party")]
c(vote = round(mean(s$dem_share), 1), seats = sum(s$dem_share > 50),
  eg = round((100*mean(s$dem_share > 50) - 50) - 2*(mean(s$dem_share) - 50), 1))
```

Ohio: **44.9%** Democratic, **4 of 15** seats, **EG −13.1**. Three packed
districts (OH-11 Cleveland, OH-03 Columbus, OH-09 Toledo) then a long
Republican-leaning tail.

## What a good answer says

**3/4:** Identifies the packed districts, reports the efficiency gap, and says
whether the distribution looks bimodal.

**4/4:** Supplies a genuine alternative explanation and says what would
distinguish it. The strongest candidates:

- **Political geography.** Democratic voters cluster in cities at very high
  densities. Any map that keeps cities whole produces packed Democratic
  districts, with no intent required. This is the Chen–Rodden argument and it is
  the serious objection to reading every negative efficiency gap as a
  gerrymander.
- **The Voting Rights Act.** Majority-minority districts are sometimes legally
  required, and they are packed by construction. **Update the framing here.**
  *Louisiana v. Callais*, 608 U.S. ___ (2026), decided 29 April 2026, narrowed
  that requirement sharply — it updated rather than overruled *Gingles*, and
  Section 2 now reaches far fewer maps. For a 2024 map the explanation still
  holds: the districts in this file were drawn under the old rule. Going
  forward it holds much less often, and a student who separates "why the
  district exists" from "whether the law would require it today" is doing the
  right thing.
- **Court-ordered maps.** Several states' 2024 maps were drawn or redrawn by
  courts.

The evidence that distinguishes them is **simulation**: generate thousands of
legal maps at random and ask where the enacted plan falls in that distribution.
That is the Duchin & Walch chapter students read, and it is the method that has
actually persuaded state courts. A student who arrives at "you need a baseline of
possible maps, not just this map's numbers" has understood the state of the art.

**Common stumble:** treating a large efficiency gap as proof of intent. Push
back every time.

***

# Option B — How much does the threshold matter?

## Verified output

```{r b}
for (band in c(2, 5, 10, 15)) {
  cat("within", band, "points:", sum(abs(cd$dem_share - 50) <= band),
      "districts =", round(100 * mean(abs(cd$dem_share - 50) <= band), 1), "%\n")
}
```

| band | districts | share of House |
|---|---|---|
| ±2 | 35 | 8.1% |
| ±5 | 91 | 21.0% |
| ±10 | 176 | 40.6% |
| ±15 | 269 | 62.0% |

## What a good answer says

**3/4:** Reports that the count rises with the band — and notices how much it
matters where you draw the line. At ±2 only 8% of the House is competitive; at
±15 a clear majority is. A student should say which band they think is the
honest one and why.

**4/4:** Connects it to campaign attention properly. The two facts are related but not
identical: presidential campaign attention concentrates because of the Electoral
College's winner-take-all rule at the *state* level, while House
uncompetitiveness is produced by district lines plus residential sorting. Both
produce the same experience for most voters — living somewhere the outcome is not
in doubt — by different mechanisms.

**4/4, best:** answers the primary question. If the general election is safe, the
only threat to a member is a primary, and primaries are low-turnout contests
dominated by the most committed partisans. That is a mechanism connecting
districting to the elite polarization students measured with DW-NOMINATE in
Session 12. The lines help produce the Congress.

***

# Option C — Who is packed?

## Verified output

```{r c}
head(cd[order(-cd$dem_share), c("district", "dem_share")], 6)
head(cd[order(cd$dem_share),  c("district", "dem_share")], 6)
c(dem_over_70 = sum(cd$dem_share >= 70), rep_over_70 = sum(cd$dem_share <= 30))
```

**PA-03 (Philadelphia) is the most Democratic district in the country at 89.1%.**
The most Republican is AL-04 at 16.1%.

The asymmetry is the finding: far more districts are overwhelmingly Democratic
than overwhelmingly Republican.

## What a good answer says

**3/4:** Notices that extreme Democratic districts are urban and that there are
more of them.

**4/4:** Draws the consequence. Democratic votes are *self-packed* by
urbanisation: a party whose voters live at 90% density in cities wastes enormous
numbers of votes before anybody draws a line. That inflates the efficiency gap
against Democrats even under neutral maps, which is exactly why the measure is
contested and why Massachusetts scores +22.4 despite nobody gerrymandering it.

**4/4, best:** holds both things at once — self-packing is real *and* North
Carolina and Wisconsin are still outliers beyond what geography explains. A
student who says "geography explains some of it, and the empty middle in NC
explains the rest" has the right answer.

***

# Option D — Dave's Redistricting

Expect discrepancies. DRA uses actual votes rather than shares, averages several
elections, and reports seat and vote bias separately from the efficiency gap.
Reward students who can say *why* the numbers differ rather than assuming one is
wrong.

This option also preserves the hands-on DRA exercise from previous years; if you
want the whole class in DRA, assign it to everyone and use the other options as
the written component.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is distinguishing *bias* from *intent*. Every
  number in this lab measures the first. None measures the second. Students who
  slide from "EG is −25" to "they cheated" have skipped the step that matters,
  and it is the step courts actually litigate.
- **Reward the geography objection**, even when a student uses it to argue
  against the lab's own framing. That is the argument working.

## Anticipated questions

- *"Is North Carolina's map illegal?"* — No. *Rucho v. Common Cause* (2019) held
  partisan gerrymandering claims non-justiciable in federal court. State courts
  can still act under state constitutions, which is where the fight now is —
  including Pennsylvania in 2018.
- *"Why is Massachusetts as biased as North Carolina?"* — It is not, in any
  meaningful sense, and this is the lab's most important caveat. The measure
  cannot separate a drawn advantage from a geographic one.
- *"Why use presidential votes for House districts?"* — Because it isolates the
  map. Say this more than once; it is the methodological core.
- *"What about NY-21?"* — Genuinely missing from the source. 434 of 435.
- *"Could a computer draw fair maps?"* — This is the Duchin & Walch chapter.
  Simulation does not draw the map; it tells you whether the enacted one is
  unusual. Worth distinguishing.

## Connections back

This is the last lab of the term, and it collects most of the course:

- **The opening session (Aug 25)** — students computed proportional Electoral College allocation and
  met the idea that a rule converts votes to seats. Same idea, different
  institution.
- **Session 1 (election returns)** — the wasted-vote question flagged forward from the first lab lands here.
  Packing and cracking *are* wasted votes.
- **Session 4 (returns over time)** — 13 competitive states in 2024; here, 91 competitive districts.
  The same shrinking middle.
- **`campaign-visits`** (retired, on disk) — campaign visits concentrate for the same structural reason.
- **Session 2 (the decennial census)** — these districts are drawn from the PL 94-171 file students
  parsed, group quarters and all. Ask what Forest County does to a Pennsylvania
  district.
- **Sessions 3 and 7** — §203 coverage and BISG are the evidentiary machinery of
  the *racial* gerrymandering claims that remain justiciable after *Rucho*.
  *Louisiana v. Callais* (29 April 2026) narrowed that route considerably: it
  updated *Gingles* rather than overruling it, and Section 2 plaintiffs must now
  offer race-neutral illustrative maps that meet all the State's stated goals
  and show polarized voting net of party. The RPV lab carries the fuller note.

If you want one closing question for the final session: *we have spent a semester
measuring American democracy. Which of the numbers we computed would you actually
put in front of a judge, and what would you say about it?*
