---
title: "Lab 15: Finding a Gerrymander Without Looking at a Map"
subtitle: "84-355 Democracy's Data | Week 15"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Everyone knows what a gerrymander looks like. It is the district shaped like a
salamander, the one that reaches a tentacle across three counties to swallow a
neighbourhood.

The reading for today argues that this is wrong — that gerrymandered districts
are often perfectly tidy, and that some of the ugliest districts in America are
ugly for entirely innocent reasons, because rivers and county lines and the
Voting Rights Act are all shapes too.

> ### Note — the Voting Rights Act shape got smaller on 29 April 2026
>
> Some packed districts in this file exist because Section 2 of the Voting
> Rights Act was understood to require them. *Louisiana v. Callais*, 608 U.S.
> ___ (2026), narrowed that requirement sharply. *Thornburg v. Gingles* was not
> overruled but updated: a plaintiff's illustrative map may use no race and must
> meet every legitimate goal the State names, including its political ones;
> racially polarized voting must now be shown net of party; and old
> discrimination counts for much less. Section 2 still exists, but it reaches
> far fewer maps than it did in 2024.
>
> **Nothing in the data below changes** — these are the 2024 maps and the 2024
> presidential vote inside them. What changes is how much weight "the VRA
> required it" carries when you reach for it as an innocent explanation in
> Option A.

So today we will not look at a single map. We will find the gerrymanders anyway,
from nothing but a column of numbers, and by the end you will be able to say
which states have biased plans and roughly how much bias they carry.

Everything is base R. Work in order.

***

# Part 1: The right instrument

```{r read}
cd <- read.csv("data/pres_by_cd_2024.csv", stringsAsFactors = FALSE)
dim(cd)
head(cd[, c("district", "dem_share", "pres_party", "house_rep_party")], 4)
```

This is the **2024 presidential vote, counted inside each congressional
district**. Not the House result — the presidential result.

That choice is the whole methodology. If you want to know whether a *map* is
biased, you have to hold the voters constant and change only the lines. House
results cannot do that: they mix the map together with incumbency, money,
scandal, and seats nobody contested. The presidential race is the same contest
in all 434 districts, so any difference in how those votes turn into seats is a
property of the districting plan.

This is how the question is asked in litigation, and it is what Dave's
Redistricting App is doing when it reports partisan metrics.

***

# Part 2: The national picture

```{r national}
c(mean_district_dem_share = round(mean(cd$dem_share), 1),
  democratic_districts    = sum(cd$dem_share > 50),
  share_of_districts      = round(100 * mean(cd$dem_share > 50), 1))
```

> The average congressional district was **49.9%** Democratic. Democrats carried
> **205 of 434** — **47.2%** of them.
>
> Roughly a tied electorate, distributed into districts in a way that gives one
> side slightly fewer than half.

That gap is small nationally. It is not small everywhere, and the national figure
hides two large biases pointing in opposite directions.

First, how many districts are actually in play?

```{r competitive}
c(competitive_45_to_55 = sum(cd$dem_share >= 45 & cd$dem_share <= 55),
  pct = round(100 * mean(cd$dem_share >= 45 & cd$dem_share <= 55), 1))
```

```{r hist, fig.width = 8, fig.height = 5}
hist(cd$dem_share, breaks = 30, col = "#BBBBBB", border = "white",
     xlab = "Democratic share of the two-party presidential vote (%)",
     ylab = "Districts",
     main = "All 434 congressional districts")
abline(v = 50, lty = 2, lwd = 2)
abline(v = c(45, 55), lty = 3, col = "grey40")
```

Only **91 districts — 21%** — sit within five points of even. Four out of five
seats in the House of Representatives were decided before anybody campaigned.

***

# Part 3: The empty middle

Here is the trick. Compare three states.

```{r three-states}
show_state <- function(ab) {
  s <- cd[cd$state == ab, ]
  s <- s[order(-s$dem_share), ]
  cat(ab, "-- statewide", round(mean(s$dem_share), 1), "% Dem,",
      sum(s$dem_share > 50), "of", nrow(s), "seats\n")
  cat(paste(sprintf("%.0f", s$dem_share), collapse = "  "), "\n\n")
}
show_state("NC"); show_state("WI"); show_state("PA")
```

Read those rows of numbers slowly.

**North Carolina** was 48.5% Democratic and elected Democrats in 3 of 14
districts. Its three Democratic districts sit at 73.5, 73.0 and 67.5. Its eleven
Republican districts sit between 39.0 and 48.4. **There is no district between
48.4 and 67.5.**

**Wisconsin** was 50.2% — a tied state — and elected Democrats in 2 of 8. Two
districts at 76.2 and 70.4; six between 38.6 and 47.7. **Nothing between 47.7 and
70.4.**

**Pennsylvania** was 49.4% and split 8–9. Its districts run smoothly from 27 up
to 89, with five of them between 45 and 53.

```{r gap-plot, fig.width = 8, fig.height = 5.5}
par(mfrow = c(3, 1), mar = c(2.5, 4, 2.5, 1))
for (ab in c("NC", "WI", "PA")) {
  s <- cd[cd$state == ab, ]
  plot(s$dem_share, rep(1, nrow(s)), pch = 19, cex = 2,
       col = ifelse(s$dem_share > 50, "#2166AC", "#B2182B"),
       xlim = c(15, 92), yaxt = "n", ylab = "", xlab = "",
       main = paste0(ab, ": ", round(mean(s$dem_share), 1), "% Democratic, ",
                     sum(s$dem_share > 50), "/", nrow(s), " seats"))
  abline(v = 50, lty = 2, lwd = 2)
}
par(mfrow = c(1, 1))
```

> **That hole in the middle is the gerrymander.**
>
> To win most of the seats with fewer than half the votes, you do two things.
> You **pack** the other side's voters into a few districts they win
> overwhelmingly, wasting their surplus. Then you **crack** the rest across many
> districts where they lose narrowly, wasting all of it.
>
> The result is a map with no competitive districts and a bimodal distribution:
> a small clump way up high, a large clump just below half, and nothing between.
> That is what North Carolina and Wisconsin look like. Pennsylvania does not look
> like it.

We have not looked at a single district's shape.

***

# Part 4: Putting a number on it

The **efficiency gap** compares a party's share of seats to what proportionality
would give it, in a form that counts wasted votes. In its simplest version:

$$\text{EG} = (\text{seat share} - 50\%) - 2 \times (\text{vote share} - 50\%)$$

A plan with no bias scores zero. Negative means the map favours Republicans;
positive means it favours Democrats.

```{r eg}
st <- do.call(rbind, lapply(split(cd, cd$state), function(s) {
  data.frame(state = s$state[1], n = nrow(s),
             vote  = round(mean(s$dem_share), 1),
             seats = sum(s$dem_share > 50),
             seat_pct = round(100 * mean(s$dem_share > 50), 1))
}))
st$eg <- round((st$seat_pct - 50) - 2 * (st$vote - 50), 1)

big <- st[st$n >= 6, ]
big[order(big$eg), c("state", "n", "vote", "seats", "seat_pct", "eg")]
```

```{r eg-plot, fig.width = 8, fig.height = 6}
b <- big[order(big$eg), ]
par(mar = c(5, 5, 4, 2))
barplot(b$eg, names.arg = b$state, las = 2, border = NA,
        col = ifelse(b$eg < 0, "#B2182B", "#2166AC"),
        ylab = "Efficiency gap (points)",
        main = "States with 6+ districts, 2024 presidential vote")
abline(h = 0, lwd = 2)
```

> The two worst Republican-favouring maps are **North Carolina (−25.6)** and
> **Wisconsin (−25.4)**. The two worst Democratic-favouring are
> **Massachusetts (+22.4)** and **Illinois (+19.0)**.
>
> **Massachusetts is 63.8% Democratic and elects Democrats in 9 seats out of 9.**
> Nobody drew that to be unfair; there is simply no way to carve a Republican
> seat out of Massachusetts without deliberately trying to.
>
> **Pennsylvania scores −1.7** — very nearly the fairest large map in the
> country.

That Massachusetts result is why the efficiency gap is contested. It cannot
distinguish a map drawn to disadvantage a party from a state where that party's
voters are too few and too evenly spread to win anywhere. Political geography
produces bias all by itself.

***

# Part 5: The lines are not the whole story

```{r crossover}
cross <- cd[!is.na(cd$house_rep_party) & cd$pres_party != cd$house_rep_party, ]
nrow(cross)
cross[order(cross$state), c("district", "dem_share", "pres_party",
                            "house_rep", "house_rep_party")]
```

Only **15 districts of 434** elected a House member from the party that lost the
district's presidential vote. Every one of them is close: all fifteen fall
between 46.3% and 52.3%.

> A generation ago there were dozens of these. Their near-disappearance is the
> same polarization you measured in Week 4 and Week 7, and it has a consequence
> for redistricting: **when almost nobody splits their ticket, drawing the lines
> is very close to choosing the winner.** A map-drawer in 1980 was making a bet.
> A map-drawer today is making a decision.

Pennsylvania's 1st district is on that list: Brian Fitzpatrick holds a seat
Harris carried.

***

# Part 6: What this cannot tell you

**The efficiency gap here is an approximation.** The proper formula uses actual
wasted votes, which needs turnout by district. We have vote *shares*, so this
version assumes districts cast equal numbers of votes. They do not — districts
are equal in *population*, and turnout varies a great deal.

**One election is not a map's record.** A plan that looks fair against 2024 may
not against 2020 or 2028. This is why Dave's Redistricting reports metrics across
several elections, and why you should look at more than one.

**Bias is not always intent.** Massachusetts, again.

**And one district is missing entirely.**

```{r missing}
nrow(cd)
```

New York's 21st has no presidential figures in the source file. Every national
number in this lab is computed over 434 districts, not 435. It does not change
any conclusion, and you should know it anyway.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — Diagnose a state

```{r opt-a, fig.width = 8, fig.height = 4.5}
this_state <- "OH"           # <- change to any state with several districts

s <- cd[cd$state == this_state, ]
s <- s[order(-s$dem_share), ]

plot(s$dem_share, seq_len(nrow(s)), pch = 19, cex = 2,
     col = ifelse(s$dem_share > 50, "#2166AC", "#B2182B"),
     xlim = c(10, 95), yaxt = "n", ylab = "",
     xlab = "Democratic share of two-party presidential vote (%)",
     main = paste0(this_state, ": ", round(mean(s$dem_share), 1),
                   "% Democratic, ", sum(s$dem_share > 50), " of ",
                   nrow(s), " seats"))
abline(v = 50, lty = 2, lwd = 2)
text(s$dem_share, seq_len(nrow(s)), s$cd, pos = 3, cex = 0.7)

s[, c("district", "dem_share", "house_rep_party")]
round((100 * mean(s$dem_share > 50) - 50) - 2 * (mean(s$dem_share) - 50), 1)
```

**Write about:** is there a hole in the middle? Where are the packed districts,
and what is in them? Report the efficiency gap and say whether the shape of the
distribution agrees with it. Then the harder part: name at least one explanation
other than deliberate gerrymandering that could produce what you see — political
geography, the Voting Rights Act, a court-ordered map — and say what evidence
would let you tell the difference.

***

### Option B — How much does the threshold matter?

```{r opt-b, fig.width = 8, fig.height = 5}
band <- 5                    # <- change this. Try 2, 5, 10, 15.

cd$comp <- abs(cd$dem_share - 50) <= band
by_state <- tapply(cd$comp, cd$state, sum)
by_state <- sort(by_state[by_state > 0], decreasing = TRUE)

c(competitive_districts = sum(cd$comp),
  pct_of_house = round(100 * mean(cd$comp), 1))

par(mar = c(5, 6, 4, 2))
barplot(head(by_state, 15), las = 2, col = "#54278F", border = NA,
        ylab = "Competitive districts",
        main = paste("Districts within", band, "points of even"))
```

**Write about:** how many House seats are genuinely in play, and does that number
survive changing the threshold? Then connect it to Week 6: candidates visited 7
states and ignored 26. Here, four out of five House districts are not
competitive. Are these the same phenomenon? What does it mean for a member of
Congress if the only election that can end their career is a primary?

***

### Option C — Who is packed?

```{r opt-c}
cutoff <- 70                 # <- change this

packed <- cd[cd$dem_share >= cutoff | cd$dem_share <= (100 - cutoff), ]
c(safe_districts = nrow(packed),
  very_dem = sum(packed$dem_share >= cutoff),
  very_rep = sum(packed$dem_share <= (100 - cutoff)))

head(cd[order(-cd$dem_share), c("district", "dem_share")], 8)
head(cd[order(cd$dem_share),  c("district", "dem_share")], 8)

table(packed$state[packed$dem_share >= cutoff])
```

**Write about:** the most Democratic district in the country is **PA-03, at
89.1%** — Philadelphia. Compare the very-Democratic list to the very-Republican
one: are they the same *kind* of place? You should find that extreme Democratic
districts are urban and extreme Republican districts are much less extreme.
Explain what that asymmetry does to the efficiency gap, and whether it is
evidence of gerrymandering or of where people choose to live.

***

### Option D — Take it to Dave's Redistricting

Go to <https://davesredistricting.org/maps> and open the official congressional
map for a state you analysed above. Under **Advanced**, find its reported seat
bias, vote bias and efficiency gap, across several elections.

**Write about:** does DRA's verdict match yours? Where it differs, work out why —
DRA uses actual votes rather than shares, and averages several elections. Which
approach would you trust in court, and what would you want to see before
concluding that a map was drawn to produce the result it produced?

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a column of numbers and tell the difference
between a map that is unfair and a country that is unevenly settled.

***

## Where the data came from

**Presidential results by congressional district, 2024** — Jason Timm,
*PresElectionResults* (<https://github.com/jaytimm/PresElectionResults>), the
same compilation used in Week 4. 434 of 435 districts; New York's 21st carries no
presidential figures in the source.

**Why not House results?** Because they cannot answer the question. A House
result confounds the map with incumbency, money and uncontested seats. Holding
the election constant and varying only the lines is the entire method, and it is
what courts and analysts do.

**On the efficiency gap.** The version computed here —
$(\text{seat share} - 50) - 2(\text{vote share} - 50)$ — assumes equal turnout
across districts. The full version uses wasted vote counts. The measure was
introduced by Nicholas Stephanopoulos and Eric McGhee and was argued before the
Supreme Court in *Gill v. Whitford* (2018); the Court declined to adopt it, and
in *Rucho v. Common Cause* (2019) held that partisan gerrymandering claims are
not justiciable in federal court. The measure remains central in **state** court
litigation, which is now where these cases are fought — including in
Pennsylvania, where the state Supreme Court struck down the congressional map in
2018.

**On the Voting Rights Act.** *Louisiana v. Callais*, 608 U.S. ___ (2026),
decided 29 April 2026, held that Section 2 did not require Louisiana to draw an
additional majority-minority district and that drawing one anyway was an
unconstitutional racial gerrymander. It updated rather than overruled
*Thornburg v. Gingles*, 478 U.S. 30 (1986). Section 2 claims are still available;
they are much harder to win than they were when these maps were drawn.

The script that downloaded, checked and wrote this file is `data/build-data.R`.
It documents the missing district rather than filling it in.
