---
title: "The Election Administration and Voting Survey"
type: brief
subtitle: "A survey the country's election offices fill in about themselves, and the empty box that decides who can be compared"
author: "Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("eavs-code.R")
```

```{r setup, include=FALSE}
```

Somewhere in a county courthouse, after an election is over and the argument about it has moved on, somebody sits down with a questionnaire from Washington and starts filling in boxes.

Item **C1a** asks how many mail ballots the office transmitted to voters. Item **C1b** asks how many came back. Four pages further on, item **C9a** asks how many of those were rejected, and items C9b through C9e ask why. Four blank boxes and a list of reasons, one office at a time, `r n(JUR)` times over.

```{r instrument, out.width="100%", fig.align="center"}
```

**Figure 1. The instrument, and the two boxes everything else rests on.** Items C1a and C1b from page 30 of the 2024 questionnaire and items C9a to C9e from page 34. Every quantity in this chapter is one of these boxes, filled in by a local election official about their own office. **C1b is the denominator of every rejection rate below, and C9a is the numerator**: the bottom number and the top number of the fraction. They are two separate boxes, filled in by two separate acts of clerical attention, and nothing on the form requires either of them. U.S. Election Assistance Commission, retrieved 10 August 2026.

Add the boxes up across the country and the answer to C9a is `r n(g("C9a"))`: mail ballots that a voter filled in, sealed, and sent, and that were not counted. That is more people than live in Wyoming.

## The only service record American elections have

Every two years the **Election Assistance Commission** sends this survey to every election office in the United States. It asks how many ballots went out, how many came back, and how many were rejected and why. It asks how many provisional ballots were issued, how many precincts were run, and how many poll workers could be found. The result is the **Election Administration and Voting Survey**, and it is the closest thing America has to a service record for its own elections.

Every other election dataset records the *output* of an election. This one records whether the machinery worked — and it has a property none of the others have: **the people being measured are the people filling in the form.**

The question it exists to answer is how often the system fails to count a ballot somebody cast, and whether that depends on where you live. **A ballot that is cast and not counted is a distinct kind of failure.** The voter did everything they were asked to do. Somewhere between their kitchen table and the canvass, the transaction did not complete — and in most states nobody is required to tell them.

The failure is also not evenly distributed. Baringer, Herron and Smith (2020) examined millions of Florida mail ballots and found rejection concentrated among younger voters, voters not registered with a major party, and voters who needed assistance. Merivaki and Smith (2020) found the same shape in North Carolina's provisional ballots. **A procedural failure that lands harder on particular people is a franchise problem, not a clerical one.**

And there is no national election administrator. Rules on witnessing, signatures, postmarks, and above all whether you get a chance to fix a problem, are set by fifty-one separate authorities. This survey is the only place where the consequences of those choices sit in one table.

## A questionnaire with no enforcement behind it

The Election Assistance Commission was created by the Help America Vote Act in 2002, and the survey exists because Congress obliged the Commission to report on how federal election law is working on the ground. That is the whole of the authority behind it. Congress did not give anyone the power to make a county clerk answer item C9b, or answer any item accurately, or answer at all. The form goes out, and what comes back comes back. Everything in this chapter is therefore a self-report by the offices whose performance is being measured, gathered under a mandate to publish rather than a mandate to comply.

Which sounds disqualifying and is not, because the alternative is nothing. A rejected mail ballot leaves no trace anywhere else. It is not in the returns, which count what was counted. It is not in the voter file, which in most states records only whether you voted. And the voter is often never told.

A handful of states publish ballot-level rejection records, and the research literature leans heavily on those few. That is exactly the problem. Florida and North Carolina are where the findings come from, because Florida and North Carolina are where the data is.

This survey is the only table in which all `r nrow(st)` states and territories answer the same question in the same words. Being able to ask a question of the whole country is worth a great deal of imprecision.

The imprecision is structural, not incidental. Nothing on the form makes C1b and C9a consistent with one another, and the number of offices answering the two is not the same. Nothing standardizes what an office counts as a rejection when it lets the voter fix the problem instead. And the word *jurisdiction*, the unit of every row, means one thing in Texas and another in Vermont. The next section is about that, because it is the trap most likely to produce a confident wrong answer.

What none of these are is dirt in the file. They are the shape of a national statistic assembled from `r n(JUR)` independent clerical acts.

The file itself needs handling before any of that. This chapter uses **Version 2 of the 2024 survey, released 12 February 2026, not Version 1**. The Commission corrected the first release and published an errata note saying what changed, so a chapter built on V1 would disagree with the EAC's own report.

The public release is a ZIP that opens to a single wide CSV, and it carries line breaks inside its free-text comment fields. A line count reports 17,013 lines for a file with `r n(JUR)` rows, and any tool that splits on line breaks mangles it. Missing values are coded negative rather than left blank, so adding up a raw column mixes real ballots with placeholder codes.

It has to be opened with a real CSV parser, the row count checked against what the Commission reports, and every total routed through something that strips the negative codes before it is summed. What survives that is four small aggregate tables. None of that is analysis. It is the price of opening the file, and anyone who downloads it pays the same price before seeing a single number.

That download is also the whole of the acquisition story, and it is a short one. Nothing here was scraped, keyed in, or pieced together from parts: it is one ZIP at one public address, no account and no request form. It also passed through nobody's hands on the way. The body that collected these answers is the body that published them, with no research team in between reconciling the awkward cases. So the defects in this file are the Commission's own, and the Commission's errata note is where they are described.

What this chapter cost was paid on either side of that half-second, and none of it in the getting. Before it, `r n(JUR)` offices filled in boxes under no obligation to fill them in correctly. After it, the parser. Which is worth naming, because the ease of the download is exactly what makes the rest of it invisible.

## What the file looks like when it arrives

Everything so far has described the survey. This is the file.

The ZIP opens to a single CSV: `r n(JUR)` rows by 535 columns, one row per election office and one column per box on the form. Here are its first two rows and one row from Georgia, sliced down to six of those 535 columns — the county's code, its name, its state, then C1a, C1b and C9a. The values are exactly as they sit in the file; every column in between is elided, and the gaps are marked.

| FIPSCode | Jurisdiction_Name | … | State_Abbr | … | C1a | C1b | … | C9a | … |
|:---|:---|:---:|:---|:---:|---:|---:|:---:|---:|:---:|
| 0100100000 | AUTAUGA COUNTY | … | AL | … | 1822 | -99 | … | 14 | … |
| 0100300000 | BALDWIN COUNTY | … | AL | … | 8902 | -99 | … | 187 | … |
| 1314500000 | HARRIS COUNTY | … | GA | … | 1153 | 1037 | … | 10 | … |

The column names carry none of their own meaning; the survey's own codebook is what turns them into quantities:

| Column | What the codebook says it holds |
|:---|:---|
| FIPSCode | the jurisdiction's identifier — ten digits, of which the first five are the county |
| Jurisdiction_Name | its name |
| State_Abbr | the state |
| C1a | mail ballots transmitted to voters |
| C1b | mail ballots returned by voters |
| C9a | mail ballots rejected |

Read the Autauga row as arithmetic and it says the county sent out 1,822 mail ballots, got back minus ninety-nine of them, and rejected 14. **The −99 is not a number. It is the survey's way of writing "did not answer."** Add the C1b column up as it stands and every unanswered box drags the national total of returned mail ballots downward by ninety-nine. Harris County, Georgia, four rows of the alphabet away, answered all three boxes and can be divided.

The file also does not respect its own lines. Free-text comment fields carry line breaks inside them, so one record can occupy several lines of the file. This is a single value, of a single column, of a single row, from Adams County, Indiana. It is wrapped here to fit the page, but the blank line in the middle of it is really in the file:

```
"THERE ARE A FEW COUNTIES WHERE SUM OF B18A AND B24A ARE OFF BY
A FEW RECORDS, WHICH THE TEAM BELIEVES CAN BE ATTRIBUTED TO USER
PROCESSING ERROR IN THE SVRS SYSTEM.

THERE ARE A FEW COUNTIES WHERE SUM OF B18B AND B24B ARE OFF BY ..."
```

Count lines and the file looks like it holds seventeen thousand records; parse it as a CSV and it holds `r n(JUR)`. Anything that reads the file a line at a time gets a different dataset from anything that reads it a record at a time, and neither one complains.

Here is what comes out the other side: every negative code thrown away, each item summed within a state, one row per state.

```{r clean-states}
```

Alabama's blank in the right-hand column is the −99 arriving where it was always going to arrive. The state's `r n(st$jurisdictions[st$state=="AL"])` counties between them reported `r n(st$mail_rejected[st$state=="AL"])` rejected ballots and no returned ballots at all, so there is nothing to divide by. That is not a bug in the cleaning; it is the survey telling the truth about what it was told.

```{r shape}
```

Four decisions did all of that, and none of them is a matter of syntax.

**Negative codes become missing, not zero.** −99 means the office did not answer and −88 means the item does not apply to it. Both had to be turned into absence before any column could be added up, and turning them into zero instead would have been the single most damaging thing anyone could do to this file.

**The denominator and the numerator are counted separately.** C1b is reported by `r n(rb("C1b"))` offices and C9a by `r n(rb("C9a"))`. Every rate on the pages that follow is a ratio of two sums taken over slightly different sets of offices, and the file offers no way around that.

**The unit of analysis moves.** The file's row is an election office; almost every question worth asking is about a state or about the country. Rolling up is what makes the numbers comparable — and it is also what makes the next section necessary, because the thing being rolled up is a county in Texas and a township in Wisconsin.

**Nothing is dropped for looking wrong.** All `r n(JUR)` rows survive. An office that reported more rejected ballots than returned ones keeps its row and its numbers, because there is no independent record to correct it against. The cleaning here removes codes, not offices.

## What a row is, and why it is not the same thing twice

One row is one **election jurisdiction**: the office that actually runs an election, receives the ballots, and decides what to count.

```{r rows}
```

`r n(JUR)` rows. And that word, jurisdiction, is the first problem.

```{r units}
```

**Wisconsin reports `r n(u$jurisdictions[u$state=="WI"])` jurisdictions. Texas reports `r n(u$jurisdictions[u$state=="TX"])`.** Wisconsin does not have seven times as much election administration as Texas; it runs elections through **townships**, and Texas runs them through **counties**. New England is the same. Vermont's typical jurisdiction has about `r n(u$voters_per_jurisdiction[u$state=="VT"])` voters. California's has about `r n(u$voters_per_jurisdiction[u$state=="CA"])`.

**So anything averaged over rows, across states, is meaningless** — the rows are not the same kind of thing. Totals are safe. Rates are safe. "The average jurisdiction rejected ninety ballots" is not a sentence about the world.

## The two columns nobody quotes

```{r national}
```

`r n(g("F1a"))` registered voters, `r n(g("D1a"))` precincts, `r n(g("C1a"))` mail ballots sent out.

Now read the two right-hand columns, which are the ones that never appear in coverage. **Not every office answers every question.** Total voters comes from `r pr("F1a")`% of them. Provisional ballots rejected comes from **`r pr("E1d")`%**.

Missing is not blank, either: the source codes it **−99** (not reported) and **−88** (not applicable). Sum the raw column without handling those and you get a number that is confidently, silently wrong.

**Every national total on this page is a floor**, and the gap between the floor and the truth is unknown from this file.

## The journey of a mail ballot, and where it ends

```{r mail}
```

`r pc(MRATE)`% of returned mail ballots were rejected. Nationally that is a small number, and it is the number most often quoted. Drawn to scale against the `r n(g("C1a"))` ballots that were sent out, it is a sliver — which is exactly why it has to be magnified before anyone can see what is inside it.

```{r journey-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

```{r journey-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. What happens to a mail ballot, at one scale, with the failure magnified.** Of `r n(SENT)` ballots transmitted, `r n(RET)` came back and `r n(NORET)` did not; of those returned, `r n(CNT)` were counted and `r n(REJ)` were rejected. The bottom bar blows that last block up about `r pc(MAG, 0)` times and splits it by the reason recorded, largest first: **`r top$reason[1]`** alone accounts for `r pc(100 * top$ballots[1] / REJ, 0)`% of every mail ballot rejected in the country. Widths are exact shares. Every total is a floor, since the four items come from between `r pc(min(nat$pct_reporting), 0)`% and `r pc(max(nat$pct_reporting), 0)`% of jurisdictions.

## The same act, counted differently in different states

Now do the same arithmetic state by state, among the `r nrow(big)` states that handled more than 100,000 returned mail ballots — which is `r pc(100*sum(big$mail_returned)/sum(st$mail_returned), 0)`% of all the mail ballots returned in the country.

```{r by-state}
```

**Oklahoma rejects `r pc(OK)`% of returned mail ballots. Vermont rejects `r pc(VT)`%.** That is a **`r pc(OK/VT, 0)`-fold** difference in the chance that a ballot you mailed is thrown away.

That is not voters behaving differently in Oklahoma. It is rules — witness requirements, signature matching, postmark deadlines, and above all whether the state lets you fix a problem after the fact.

```{r d3-states, results='asis', eval=knitr::is_html_output()}
```

```{r state-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

**Figure 3. The chance that a mailed ballot is thrown away, by state.** The `r nrow(big)` states that handled more than 100,000 returned mail ballots, ranked by the share rejected; the dashed line is the national `r pc(MRATE)`%. Oklahoma sits at `r pc(OK)`% and Vermont at `r pc(VT)`%, a `r pc(OK/VT, 0)`-fold difference in the same act performed in two states. `r nrow(hole)` states and territories cannot appear here at all, for a reason this chapter comes back to.

## The reasons, and how well they are known

```{r reasons}
```

**The single largest reason is a signature that did not match: `r n(rsn$ballots[1])` ballots, `r pc(rsn$pct_of_rejected[1], 1)`% of every mail ballot rejected in the country.** A human being compared your handwriting to a specimen on file and decided it was not you.

Second is arriving late. Third, at `r n(rsn$ballots[3])`, is **"Other."**

```{r reason-check}
```

The categories very nearly add up — a gap of about `r n(g("C9a") - sum(rsn$ballots))` out of `r n(g("C9a"))`. That is reassuring about the categories. It says nothing about the reporters.

```{r reporting}
```

**The total rejected is reported by `r n(rb("C9a"))` jurisdictions. Some reasons are reported by as few as `r n(min(rsn$reported_by))`.**

There is a pattern, and it is not random. **The best-reported reasons are the ones an office records automatically**: the ballot arrived after the deadline, the voter had already voted, the voter was ineligible. Those fall out of the system. **The worst-reported are the ones that need a person to write down a judgment**: no postmark, no resident address, unofficial envelope.

Which puts signature mismatch in an uncomfortable place. It is the **largest** category, at `r n(rsn$ballots[1])` ballots, and it is reported by `r n(rsn$reported_by[1])` of `r n(JUR)` jurisdictions — half the country. **We know least about the category where human discretion matters most.**

Plot the two columns against each other and the pattern is not subtle.

```{r reasons-static, eval=!knitr::is_html_output(), fig.height=4.8}
```

```{r reasons-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 4. How many ballots each reason accounts for, against how many offices reported it.** Up the page is the number of ballots rejected for that reason, on a log scale; across is the share of the country's `r n(JUR)` jurisdictions that reported it at all. The three reasons that fall out of an office's system automatically are reported by `r pc(COVA, 0)`% of jurisdictions on average; the three that require someone to have written down a judgment, by `r pc(COVJ, 0)`%. Signature mismatch, the largest category of all at `r n(rs$ballots[1])` ballots, sits at `r pc(rs$cov[1], 0)`%. **The better a reason is known, the less discretion it involved.**

At this point there is a national rate, a state ranking, and a table of reasons, all built from a survey where item response runs between `r pc(min(nat$pct_reporting), 0)`% and `r pc(max(nat$pct_reporting), 0)`%. The natural assumption about that missing fraction is that it adds noise.

## A ranking of the states that answered the question

Every rate in the state ranking is *rejected ÷ returned*: two different survey items, with two different response rates.

> **Before you read on, answer yes or no.** A missing denominator sounds as
> though it should behave the way missing data usually does — scatter the rates
> about a bit, some too high, some too low, no particular direction. **Does
> it?** Write down yes or no, and if no, write down which way it pushes.

```{r denominator-gap}
```

**No.** Most readers write down yes, because scattering is what missing data usually does. This missing data pushes in one direction, and then it takes whole states out of the table altogether.

**The denominator is measured worse than the numerator.** `r n(NOJOIN)` jurisdictions told the EAC how many ballots they rejected and did not tell it how many came back. Wherever that happens, the denominator is short and **the rate is too high** — not by noise, but in one direction.

At state level it stops being subtle:

```{r hole}
```

**`r nrow(hole)` states and territories report sending mail ballots and rejecting mail ballots, while reporting that **zero** came back.** They hold `r n(sum(hole$voters))` registered voters between them. Alabama sent `r n(hole$mail_sent[hole$state=="AL"])` and rejected `r n(hole$mail_rejected[hole$state=="AL"])` of the none that returned.

You cannot compute a rate from that. So Alabama does not appear in Figure 3, or in any published version of Figure 3. **It is not at the bottom of the ranking and it is not at the top. It is not in the ranking.**

*"Oklahoma is `r pc(OK/VT, 0)` times worse than Vermont"* is therefore a statement about `r nrow(big)` states that answered both questions. **The non-response is not noise around the estimate. It silently decides who is eligible to be compared at all.** And the offices likeliest to skip an item are small, understaffed ones, which differ systematically from the offices that answer.

Notice also what the file does *not* do: it does not flag this, refuse to publish, or carry a warning. It reports a zero. **A zero and a missing value look identical once they are in a spreadsheet**, and only arithmetic that cannot work reveals the difference: a rejection rate with nothing in the denominator. Item C1b was left blank, and by the time it reached anybody's analysis it had become a number.

## The other failure, which is much larger

A provisional ballot is what you get when you turn up and something is wrong: you are not on the list, you lack the right ID, you are in the wrong place. It is the system's safety net.

```{r prov}
```

**About one in four is rejected — `r n(g("E1d"))` people.** That is a far higher failure rate than mail, it is much less discussed, and it is reported by the *worst-covered* item in the whole survey, so the real figure is larger.

```{r prov-state}
```

The spread is **wider than mail**, `r pc(max(pv$prate)/min(pv$prate), 0)`-fold among states issuing more than 10,000. And the denominator problem returns in a different costume. A state with same-day registration hands out almost no provisional ballots, so its rate rests on a tiny, unusual base. **The number of provisional ballots measures how badly the registration system is working, not how well the backup is.**

## Strict, or careless, or neither

There are two familiar readings of the state variation in Figure 3, both available in this data and neither demonstrated by it. One says high rejection rates are suppression: strict rules chosen in the knowledge of who they will catch. The other says they are integrity: a state that verifies signatures will reject more ballots than one that does not, and that is the verification working. Nothing in this file distinguishes them. Both produce the same row. Resolving it requires knowing the rules, and the rules are in fifty-one statute books rather than in this table.

There is a third possibility that neither camp likes, and it is the one the evidence most supports: **states with strict rules and no cure process reject more ballots without anybody intending anything.** The rule was written, the envelope failed the rule, the ballot went in a bin. Richard Hasen's *Election Meltdown* (2020) makes the case that this ordinary administrative variation, rather than fraud or conspiracy, is what erodes confidence in American elections.

The literature has largely moved past the argument by changing the question. Rather than comparing states, Baringer, Herron and Smith look **within** a state and ask which voters are rejected, and there the answer is consistent: younger voters, unaffiliated voters, and first-time mail voters.

The quieter argument is about the survey itself, and it is the more important one. Reporting is voluntary in effect: there is no penalty for skipping an item, and in a small township the clerk is part-time. The Commission could require complete responses, fund the offices that cannot manage them, or publish an explicit missing-data flag instead of a zero. It does none of these, and the result is a national statistic nobody can put a confidence interval on. There is also the lag, and it is by design. The survey covering the next federal election will not exist for another two years, so a country receives its election's service record after the argument about that election is over.

## What the survey will bear

The strengths of this source are real and unusual. It is a census of election offices rather than a sample. It is free, public, documented by an actual codebook, and published as a versioned file with errata. Version 2 is the one to take, because Version 1 was corrected and the Commission says so in writing. It is the only source that puts every jurisdiction's administrative outcomes in one table, and it is the reason comparative election administration exists as a research field at all.

The weaknesses run in one direction, which is what makes them dangerous. Item response runs from `r pc(min(nat$pct_reporting), 0)`% to `r pc(max(nat$pct_reporting), 0)`% and is not random, so every total here is a floor of unknown depth. The denominator of a rejection rate is measured worse than the numerator, so rates are biased upward by an unknown amount that differs by state.

`r nrow(hole)` states and territories report a zero where they mean "not reported", and drop out of every comparison silently. The word "jurisdiction" means a county in Texas and a township in Wisconsin, so nothing averaged over rows is a statement about the world. Missing values arrive coded as −99 and −88, which sum perfectly well as numbers. And there are no demographics anywhere in it: the file cannot say who was rejected, only how many.

So the sentences this survey supports are these. The United States rejected `r n(g("C9a"))` mail ballots and `r n(g("E1d"))` provisional ballots in one federal election. The largest single stated reason for a mail rejection is a handwriting judgment. And here is the sharper claim. **The survey's holes are not a caveat on the finding. In places they are the finding.**

It does not support any statement about whether an individual rejection was correct, about who the rejected voters were or whether anyone told them, or about how two states compare without knowing their two rulebooks.

Three questions stay open, and none of them will be settled by another wave of the same instrument. Whether high-rejection states are strict or careless: the file records the outcome, not the intent. How large the true national totals are, which requires the answers nobody gave. And whether rejection rates would fall if every state offered voters a chance to fix a problem. The natural experiment exists across the states, but this file cannot run it, because it does not record which states have one.

What generalizes beyond elections is the shape of the failure. The people being measured fill in the form themselves. There is no penalty for leaving a box empty, and an empty box arrives at the analyst's desk as a zero. Item C1b is a box on a page. Everything in this chapter that could not be computed, could not be computed because somebody left it alone.

## Sources

**Data**

- U.S. Election Assistance Commission, *Election Administration and Voting
  Survey, 2024*, Version 2 — `r n(JUR)` jurisdictions, 535 columns, with the
  official codebook.
  <https://www.eac.gov/research-and-data/datasets-codebooks-and-surveys>
  Version 2 rather than Version 1: V1 was corrected, and the EAC publishes an
  errata note listing what changed.
- The form in Figure 1 is the EAC's own **2024 EAVS questionnaire**, cropped to
  items C1a/C1b and C9a–C9e.
  <https://www.eac.gov/sites/default/files/2024-04/2024_EAVS_FINAL_508c.pdf>
  Retrieved 10 August 2026, and reproduced here, so this document needs no
  network to read.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`national.csv`](data/derived/national.csv), [`reasons.csv`](data/derived/reasons.csv), [`states.csv`](data/derived/states.csv), [`units.csv`](data/derived/units.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `national.csv` reports `pct_reporting` for every item. Sort by it. Which questions do election offices reliably answer, and which do they skip?
- `units.csv` gives `voters_per_jurisdiction` by state. Compare the extremes. Does 'a jurisdiction' mean the same thing in two different states?
- `reasons.csv` breaks rejected mail ballots down by why. Work out the share for each. Which reasons are about the voter, and which are about the system?
- `states.csv` has `mail_sent`, `mail_returned` and `mail_rejected`. Compute a rejection rate per state and sort it. Then decide whether the differences are voters or rules.

**Academic literature**

- Baringer, A., Herron, M. C. and Smith, D. A. (2020). "Voting by Mail and
  Ballot Rejection: Lessons from Florida for Elections in the Age of the
  Coronavirus." *Election Law Journal* 19(3): 289–320.
- Merivaki, T. and Smith, D. A. (2020). "A Failsafe for Voters? Cast and
  Rejected Provisional Ballots in North Carolina." *Political Research
  Quarterly* 73(1): 65–78.
- Hasen, R. L. (2020). *Election Meltdown: Dirty Tricks, Distrust, and the
  Threat to American Democracy.* Yale University Press.


**Further reading**

- Burden, B. C. and Stewart, C. III, eds. (2014). *The Measure of American
  Elections.* Cambridge University Press.
- Stewart, C. III (2011). "Voting Technologies." *Annual Review of Political
  Science* 14: 353–378.
**How the figures were made**

Download the release and the codebook from the Commission at the address above; everything this chapter shows starts there and needs nothing else.

That `r n(JUR)`-row, 535-column file is reduced here to four small tables, one row per state and one per response category. The −99 and −88 codes are turned into missing values rather than added up, and for every item the number of jurisdictions that actually answered is recorded.

Two details of the reading are worth knowing if you fetch the file yourself. It must be opened with a real CSV parser, because line breaks inside the comment fields make a naive line count report thousands of rows too many. And the negative codes must be removed before anything is added up. Every number printed above is computed from those tables at the moment this document was rendered.

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
