---
title: "The Ballots With Nothing On Them"
subtitle: "84-355 Democracy's Data | Residual votes, 2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `residual-votes-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"The machines worked fine."*

After Florida in 2000, a group of political scientists and engineers asked a
question nobody had been asking: **how many ballots come out of an election with
no valid vote for the top office on them?**

The measure they built is the **residual vote** — ballots cast, minus votes
counted in the presidential race. What is left is a ballot that went in and came
out without a usable presidential choice: left blank on purpose, marked twice,
or filled in a way the scanner could not read. It became the standard way to
compare voting equipment and ballot design, and it is why punch cards
disappeared.

**Nobody publishes it.** To get it you need two numbers that live in different
places, and today you are going to put them together — and then discover that
the seam between them is more interesting than the measure.

# Steps 1-3 — Two sources, one subtraction, and units that do not match

```{r load}
co <- read.csv("data/counties.csv",  stringsAsFactors = FALSE)
st <- read.csv("data/states.csv",    stringsAsFactors = FALSE)
an <- read.csv("data/anomalies.csv", stringsAsFactors = FALSE)
un <- read.csv("data/unusable_states.csv", stringsAsFactors = FALSE)

c(counties_joined = nrow(co) + nrow(an), usable = sum(co$state_usable))
head(co[co$state_usable, c("state_name", "county_name", "ballots",
                           "total_votes", "residual", "residual_rate")], 5)
```

**Ballots cast** comes from **EAVS**, item F1a — the survey you have already
met, where every election office in the country reports what it did.

**Presidential votes** come from the county returns in this course's first lab.

Neither was collected with the other in mind. The join is by county FIPS — which
works because **counties nest**, the fact established in the geography lab. In
the township states, EAVS jurisdictions are summed to their county prefix.

# Step 4 — The distribution

```{r rate}
u <- co[co$state_usable, ]
c(counties = nrow(u),
  median_rate = round(median(u$residual_rate), 2),
  ballots = sum(u$ballots),
  presidential_votes = sum(u$total_votes),
  aggregate_rate = round(100 * (sum(u$ballots) - sum(u$total_votes)) / sum(u$ballots), 2))
round(quantile(u$residual_rate, c(0, .25, .5, .75, .95)), 2)
```

```{r rate-plot}
hist(u$residual_rate[u$residual_rate < 6], breaks = 60, col = "#2c7fb8",
     border = "white", main = "", xlab = "residual vote rate, % of ballots cast")
abline(v = median(u$residual_rate), col = "#C41230", lwd = 2)
```

**A median county has a residual vote rate of 0.93%.** Roughly one ballot in a
hundred arrives without a usable presidential vote — and half of counties sit
between 0.64% and 1.39%.

That is a real number and it matches what the voting-technology literature finds.
In a close state it is also a lot of ballots.

# Steps 5-7 — The floor, the 103 counties that break it, and the other tail

> **Before you run this section, notice one thing about the measure.** A
> residual vote cannot be negative — you cannot count more presidential votes
> than there were ballots to carry them. That is arithmetic, not an assumption.
> **So predict what it means if the subtraction comes out below zero anywhere.**

Here is what makes this lab different from every other measurement in the
course. **The residual vote has a floor: it cannot be negative.** You cannot
count more presidential votes than there were ballots to put them on.

So any county where the subtraction goes negative is not a low residual rate.
**It is a mistake, in one source or the other, that we have just caught.**

```{r anomalies}
nrow(an)
head(an[, c("state_name", "county_name", "ballots", "total_votes", "residual")], 8)
```

**103 counties report more presidential votes than ballots cast.** Two federal
data sources, describing the same election in the same county, disagreeing about
something neither could be vague about.

We cannot tell you which one is wrong. **We can tell you that one of them is**,
and that is more than either source will tell you on its own.

# Steps 8-9 — Three states where it breaks, and what survives

```{r unusable}
un
```

Three states had to be set aside, and each fails differently.

**Mississippi: every single county comes out at exactly 0.000%.** Ballots cast
equals presidential votes, county after county, to the vote. That is not a state
with perfect ballots — it is a state that answered the "ballots cast" question
by reporting its presidential total.

**The District of Columbia: 87.7%.** EAVS reports DC by ward; the returns file
does not line up, so the join compares a citywide total against one ward's
votes. **A geography mismatch, exactly of the kind the census-geography lab
warned about.**

**Vermont: a median of 5.8%**, far above any plausible rate.

```{r outliers}
head(u[order(-u$residual_rate),
       c("state_name", "county_name", "ballots", "total_votes", "residual_rate")], 6)
```

And even inside the 45 usable states, individual counties reach 30% and higher.
**Those are not findings about ballot design.** They are bad rows, visible only
because the measure has a plausible range and they fall outside it.

# Step 10 — What you can now say

**You can say** that residual votes are real and common: the typical county
loses about **1% of its ballots** at the top of the ticket, and in a state
decided by 20,000 votes that is not a rounding error.

**You can say something sharper about the sources.** Two federal datasets
describing the same election disagree in **103 counties** and in three whole
states, and nobody reconciles them — because no agency owns the comparison. That
disagreement is only visible because the quantity we happened to compute has a
floor.

**You cannot say** how much of the residual vote is deliberate abstention and
how much is failure, and you cannot rank equipment without equipment data this
file does not carry.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Which source is wrong?** Take five counties from `anomalies.csv`
and work out which number you would trust. What would you check, and where would
you look? You will not be able to settle it from this data — say what would.

**Option B — The state table.** `states.csv` aggregates counties within states.
Compare a state's aggregate rate to the median of its counties (Missouri is
1.36% by median and 5.26% in aggregate). Which is the better summary, and what
is the aggregate doing that the median is not?

**Option C — Set a filter.** Choose a rule for excluding implausible counties
and defend it — a maximum rate, a minimum county size, something else. How many
counties does it remove, and how much does the national figure move? Then say
what your rule might be throwing away.

**Option D — Your own question.** 2,993 joined counties, two federal sources,
one subtraction.

# What this lab cannot tell you

- **Why any ballot was residual.** Deliberate abstention, a spoiled mark and a
  scanner failure are indistinguishable here. The measure was built precisely
  because you can compute it without knowing which — but it means a high rate is
  a symptom, not a diagnosis.
- **Anything about equipment.** The classic use of this measure compares punch
  cards to optical scan to touchscreen. We have no equipment data joined in, so
  we cannot make that comparison.
- **Which source is right.** See Steps 5-7. We have found 103 disagreements and
  resolved none.
- **The whole country.** Three states are unusable, 103 counties are impossible,
  and roughly 170 counties never joined at all.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- One ballot in a hundred carries no presidential vote. Is that a lot?
- In 103 counties two federal sources contradict each other about the same
  election. Should somebody be reconciling them, and who?
- Mississippi appears to have answered a different question than the one it was
  asked. How would anybody have noticed without doing this subtraction?

***

### Where the data came from

**Ballots cast** — U.S. Election Assistance Commission, **EAVS 2024, Version 2**,
item F1a, aggregated from jurisdiction to county FIPS.

**Presidential votes** — county returns, from this course's Session 1 lab.

**The residual vote is not published by anybody**; this lab constructs it. The
build in `data/build-data.R` identifies unusable states from the data rather
than from a list, and writes every impossible county to `anomalies.csv` instead
of dropping them.
