---
title: "Key: The Ballots With Nothing On Them"
subtitle: "84-355 Democracy's Data | Residual votes | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
co <- read.csv("data/counties.csv",  stringsAsFactors = FALSE)
an <- read.csv("data/anomalies.csv", stringsAsFactors = FALSE)
un <- read.csv("data/unusable_states.csv", stringsAsFactors = FALSE)
u  <- co[co$state_usable, ]
an$gap <- an$total_votes - an$ballots
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`residual-votes-brief.Rmd`) is now
a chapter rather than a parallel copy of the lab: prose organised by argument,
under idea headings, with four numbered figures and no numbered steps. The lab
(`residual-votes.Rmd`) still runs Steps 1-10 with the code exposed. **Students
should read the brief first.** **Step numbers below refer to the lab**; the
chapter is cited by figure number or section title.


**Requires EAVS and the Session 1 returns lab**, and pays off the geography lab
directly. Best taught after all three.

**This is the only lab in the course where students construct a measure nobody
publishes**, and then use arithmetic to audit the sources it came from. Say that
at the start — it is a different activity from every other session.

Every number below was run against the data files.

***

# The session in one sentence

One ballot in a hundred carries no presidential vote, and the subtraction that
tells you so also catches 103 counties where two federal datasets contradict
each other outright.

***

# The findings

```{r headline}
c(counties_usable = nrow(u),
  median_rate     = round(median(u$residual_rate), 2),
  aggregate_rate  = round(100 * (sum(u$ballots) - sum(u$total_votes)) / sum(u$ballots), 2),
  impossible      = nrow(an),
  unusable_states = nrow(un))
```

- **Median county residual rate 0.93%**, quartiles 0.64% and 1.39%.
- **103 counties** report more presidential votes than ballots cast.
- **Three states unusable**: Mississippi (all exact zeros), DC (87.7%), Vermont.

***

# The three things to get said

**1. The measure exists because of Florida 2000.** Punch cards produced residual
rates several times higher than optical scan, and the finding — from exactly
this subtraction — is why they are gone. **A number nobody published turned out
to be the one that mattered**, and building it is a demonstration that useful
measures are sometimes lying around unassembled.

**2. The floor is the audit.** A residual vote cannot be negative. That single
constraint converts a measurement into a **test of the sources**, and it is the
only lab in the course where the data can prove itself wrong. Make sure students
see the difference between "this county has an unusual value" and "this county
has an impossible one."

**3. Three states fail in three different ways, and one is the geography lab.**
Mississippi answered a different question. DC is a unit mismatch — EAVS reports
wards, the returns do not. **Point at DC explicitly and connect it back.** These
were found by the build from the data itself, not from a list of known problems.

**The line for the board:** the subtraction measures the ballots, and the
leftovers measure the bookkeeping.

***

***

# Walking the steps

| Steps | What they do | Min | Watch for |
|---|---|---|---|
| **1-3** | Two sources that were never meant to meet; one joined row; the township problem | 8 | Flag the nesting assumption here — Step 6 is where it detonates. |
| **4** | The distribution: median 0.93% | 5 | One ballot in a hundred. Make them convert it to votes in a close state. |
| **5-7** | The floor, the 103 impossible counties, and the other tail | 14 | **The heart of the session.** A measure that cannot be negative is a test of its own sources. |
| **8-9** | Three states where it breaks; what survives | 8 | Mississippi's exact 0.000% everywhere is the one to linger on. |
| **10** | What you can say about the measure and about the sources | 6 | The finding is about the data system, not about voters. |

**Step 5 is the pivot and it is one sentence:** a residual vote cannot be negative. Everything after it follows. Write it on the board before running Step 6.

**If you are short on time**, protect the pivot and the closing step. The numbers are what make the argument concrete; the sequence is the argument.


# Option A — Which source is wrong?

## Verified output

```{r a}
c(worst_gap_county = an$county_name[which.max(an$gap)],
  state = an$state_name[which.max(an$gap)],
  extra_votes = max(an$gap),
  median_gap = median(an$gap),
  gaps_of_10_or_fewer = sum(an$gap <= 10))
head(an[order(-an$gap), c("state_name", "county_name", "ballots", "total_votes", "gap")], 5)
```

**Cook County, Illinois: 943,822 more presidential votes than ballots cast** —
85% of the reported ballots. The median anomaly is 112 votes; only 12 of the 103
are within 10 votes.

**3/4:** Picks counties, observes the size of the gaps, says the data cannot
settle which source erred.

**4/4:** Separates the two populations. **Twelve counties differ by ten votes or
fewer** — plausibly a late-counted provisional or an amended return, and not very
interesting. **Cook County differs by nearly a million**, which is not an error
of that kind at all.

**4/4, best — and this is the answer to steer toward:** diagnoses Cook County.
**Chicago runs its own elections, separate from suburban Cook County**, so EAVS
carries two jurisdictions where the returns file carries one county. Summing by
FIPS prefix should catch both; if Chicago's rows are coded differently, the
county's ballots are undercounted by roughly the size of Chicago. **A student who
reasons from "who administers elections here" to "what the join must have
missed" has used the geography lab exactly as intended.**

***

# Option B — The state table

## Verified output

```{r b}
agg <- tapply(seq_len(nrow(u)), u$state_name,
              function(i) 100 * (sum(u$ballots[i]) - sum(u$total_votes[i])) / sum(u$ballots[i]))
med <- tapply(u$residual_rate, u$state_name, median)
d <- data.frame(state = names(med), median = round(as.vector(med), 2),
                aggregate = round(as.vector(agg), 2))
d$gap <- round(d$aggregate - d$median, 2)
head(d[order(-abs(d$gap)), ], 6)
```

**Missouri: 1.36% by median, 5.26% in aggregate.**

**3/4:** Notes the two differ and that aggregate is vote-weighted.

**4/4:** Explains the direction. The aggregate is a ratio of sums, so **a handful
of large counties with bad numbers drags it up**, while the median is immune.
Missouri's median says its typical county is unremarkable; its aggregate says
several big ones are broken.

**4/4, best:** Says which to use and when — **the aggregate is the right estimate
of "what share of ballots statewide were residual" if the data are clean, and
the median is the better description of typical county performance when they are
not.** Choosing between them requires knowing whether you trust the tails, which
is the question the whole lab is about.

***

# Option C — Set a filter

## Verified output

```{r c}
for (cut in c(3, 5, 10, 100)) {
  k <- u$residual_rate <= cut
  cat(sprintf("  drop rate > %3d%%   keeps %4d   national %.2f%%\n",
              cut, sum(k), 100 * (sum(u$ballots[k]) - sum(u$total_votes[k])) / sum(u$ballots[k])))
}
for (m in c(0, 1000, 5000)) {
  k <- u$ballots >= m
  cat(sprintf("  ballots >= %5d   keeps %4d   national %.2f%%\n",
              m, sum(k), 100 * (sum(u$ballots[k]) - sum(u$total_votes[k])) / sum(u$ballots[k])))
}
```

**Two rules, two completely different outcomes.** Trimming by rate moves the
national figure from 1.39% to 1.20%. **Trimming by county size does nothing at
all** — 1.39% whether you keep every county or only those with 5,000+ ballots.

**3/4:** Picks a rule, reports how many counties it removes and how the figure
moves.

**4/4:** Notices the size filter is inert, and says why that is informative:
**the bad rows are not small counties.** The instinct that noise lives in small
places is wrong here — Cook County is the worst offender in the file.

**4/4, best:** Sees the circularity in the rate filter. **Excluding counties
because their residual rate is implausible uses the outcome to select the
sample**, which will always tighten the estimate and cannot distinguish bad data
from genuinely bad ballot design. A student who says the honest move is to
report both figures and show the excluded rows — which is what the lab does —
has arrived at the professional standard.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is distinguishing *impossible* from *unusual*, and
  treating the impossible ones as evidence about the sources.
- **A 2** is reporting 0.93% as the answer with no engagement with the 103.
- **Also a 2**: "the data is bad so we can't say anything." 2,803 counties
  behaved.
- **Reward anyone who diagnoses a specific anomaly** rather than listing them.

## Anticipated questions

- *"Which number is the real residual rate?"* — Median 0.93% for a typical
  county; 1.39% across the usable states by volume. Both, with the caveat.
- *"Is 1% bad?"* — Depends entirely on the margin. In 2000 Florida it was
  decisive; in a 20-point state it is invisible. **Refuse to answer without a
  comparison.**
- *"Why doesn't the EAC just fix this?"* — EAVS reports what jurisdictions
  submit, and the county returns come from state canvasses. **No agency owns the
  comparison**, which is why nobody had made it.
- *"Could students report an error?"* — Genuinely yes. The EAC accepts
  corrections and the anomalies file is a list of candidates. Worth mentioning:
  the work is real.

## Connections

- **EAVS** — the source of the ballots figure and the same self-report problem.
- **Session 1 returns** — the other half, and the first "two sources disagree"
  lesson.
- **Census geography** — DC's failure is a unit mismatch; Cook County's is an
  administrative one.
- **Independent expenditures** — the other lab where impossible values exposed
  what nobody had checked.
