---
title: "Apportionment: How a Count Becomes Seats"
type: brief
subtitle: "Equal proportions, the seats the 2020 census moved, and a margin smaller than the instrument"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("apportionment-code.R")
```

```{r setup, include=FALSE}
```

Political power in America is supposed to follow the census. Article I, Section 2 orders a count every ten years, and orders that House seats be apportioned among the states "according to their respective Numbers." Every civics course states it as a fact. What the clause does not say is *how*: dividing `r n(sum(ap$seats))` seats among `r nrow(ap)` states essentially never comes out in whole numbers, and whichever way the fractions are settled will benefit one state at another's cost. So the question here is the how. By what rule does a count become seats, and how finely can that rule cut?

The 2020 answer ends at a very sharp edge. The last of the `r n(sum(ap$seats))` seats went to `r tail(b435$order, 1)`. The state next in line was `r ladder$state[1]`, and the distance between them was `r n(margin)` people, a large lecture hall. Because the count fell that way, New York casts one fewer vote on every bill that passes the House for a decade. It casts one fewer vote for President in two elections, since a state's electors are its seats plus two.

That margin is the number everyone quotes about the 2020 census, almost always to argue that the count needed to be more accurate. Whether the margin can bear that argument is where this brief ends, and the answer is not the obvious one.

## The test

The question is answered from the Bureau's own record of the hand-out: the *2020 Census Apportionment Results*, published in April 2021 at <https://www.census.gov/data/tables/2020/dec/2020-apportionment-data.html>. Table 1 carries each state's apportionment population, its seats, and its change from 2010. Table 3 carries the population counted overseas. A third table, *Priority Values*, publishes the arithmetic itself: every House seat from 51 to `r max(pv$house_seat)`, the state that took it, and the value it was won at.

No other source could serve. The seats in this file are not a measurement of how many seats the states have. They **are** how many seats the states have, reported by the agency that ran the count under a legal deadline. And because the Bureau publishes the queue as well as the result, the rule can be re-run here and checked against the publisher to the last digit. Very little political data can be audited that completely.

The count underneath the file is the decennial census. [The decennial census chapter](../census-decennial/census-decennial-brief.html) is that data's biography: what the count asks, who is in it, and who it fails to find. This brief takes the count as given and follows what the law does with it. The rule in force is the **method of equal proportions**, also called Huntington–Hill. Congress changed methods five times before fixing this one by statute in 1941, and the size of the House, `r n(sum(ap$seats))`, has been a statute of its own since 1929.

## The data

The build reads the Bureau's three workbooks and reduces them to three small tables, keeping the state rows and dropping the title and total rows.

`apportionment_2020.csv` has one row per state and five columns besides the name: `resident_pop`, `overseas`, `app_pop`, `seats`, and `people_per_seat`. Fifty rows, not fifty-one: the District of Columbia has no apportioned Representative and does not appear. Here is one row.

```{r one-row}
```

Read the middle columns slowly, because **the apportionment population is not the population of the state**. Federal employees serving abroad, mostly military, are counted and assigned back to a home state, and `app_pop` is `resident_pop` plus that `overseas` column. The assignment is a rule written in advance, not a measurement, and it could defensibly have been written another way. Nationally it moves `r n(sum(ap$overseas))` people, and this brief returns to what that does.

`priority_values.csv` is the Bureau's published queue: one row per House seat from 51 onward, with `house_seat`, `state`, `state_seat` (which of that state's own seats it was), and `priority`, the value it was won at. The table runs ten seats past the `r n(sum(ap$seats))` the statute allows, which turns out to matter.

`seat_change_states.csv` carries `seats_2010`, `seats_2020`, and `change` for every state, from Table 1's own comparison column.

> **Before the findings, write down two numbers.** First: out of
> `r n(sum(ap$seats))` seats, how many do you think changed states after the
> 2020 count? Second: `r ladder$state[1]` missed the last seat by
> `r n(margin)` people — how far short do you think the *next* state in line
> was? Hundreds, thousands, hundreds of thousands? Written down, not thought
> about: the end of this brief is a comparison against your two guesses.

## What it says about democracy

The rule is short enough to write out in full. Every state starts with the one seat the Constitution guarantees it. The remaining `r n(sum(ap$seats) - nrow(ap))` are handed out one at a time, and before each hand-out every state posts a bid, called its **priority value**. For a state's $k$th seat the bid is

$$\text{priority} \;=\; \frac{\text{population}}{\sqrt{k(k-1)}}$$

and the highest bid takes the seat. That is the whole method. The divisor grows with every seat a state wins, so winning lowers your own next bid, and that is the entire reason the queue self-corrects instead of handing the whole House to California. Here are the first twelve seats, with the arithmetic spelled out.

```{r worked-queue}
```

These are the Bureau's own published priority values. Applying the formula to the population column `r ifelse(pv_ok, "reproduces every one of them to the last digit", "does not reproduce them, which would mean the file or the formula is wrong")`.

```{r pq-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

```{r pq-d3, results='asis', eval=knitr::is_html_output()}
```

```{r pq-fig, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The price of a seat, seat by seat, as the queue empties.** Each tread is one seat, drawn in the order it was won, at the priority value of the state that took it. A staircase fits the question because that is what the method is: the House handed out one seat at a time, at a price that only ever falls. Seat `r pv$house_seat[1]` went to `r pv$state[1]` at `r n(round(pv$priority[1]))`; seat `r n(sum(ap$seats))` went to `r tail(b435$order, 1)` at `r n(round(win_pri))`, about `r pc(pv$priority[1] / win_pri, 0)`-fold cheaper. In the browser, the slider replays the queue and shows the four highest bids at every step. The shaded band is the ten seats the Bureau tabulated past the statute; the open circle is the seat `r ladder$state[1]` would have taken.

Run the queue from the population column alone and the claim can be tested directly.

```{r reproduce}
```

**`r ifelse(ok435, "The recomputation reproduces the official House exactly, state for state.", "The recomputation does not reproduce the House — the file is wrong.")`** There is no committee and no fudge factor. Once the census is counted, the House allocates itself.

### What the 2020 count moved

```{r seatmoves}
```

**`r n(SGAIN)` seats sit in different states than they did after 2010, across `r n(sum(scs$change != 0))` states.** Texas took `r scs$change[scs$name == "Texas"]`, `r n(sum(scs$change == 1))` states took one each, and `r n(sum(scs$change == -1))` states, `r ladder$state[1]` among them, gave one up. No state moved by more than two. The direction is the country's population story made political: the gains sit in the South and the West, the losses mostly in the Northeast and the industrial Midwest. And every one of those moves carries an elector with it, because a state's Electoral College weight is its seats plus two.

### The last seat, and the queue behind it

The final seat is where the arithmetic gets finer than most people imagine a national count can be.

```{r howclose}
```

Had `r ladder$state[1]` been counted `r n(margin)` people higher, seat `r n(sum(ap$seats))` would have been its `r ordn(ny$seats + 1)` rather than `r tail(b435$order, 1)`'s `r ordn(mn$seats)`. This is not a reconstruction: the Bureau kept computing past the statute, and its own table names the states that would have taken seats `r n(sum(ap$seats) + 1)` through `r n(sum(ap$seats) + 5)`.

```{r past435}
```

Now the second of your two guesses, answered for every state at once.

```{r ladder-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r ladder-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. How far every near-miss state was from one more seat.** The eight states closest to a further seat, ordered by how short they fell; each dot is the number of additional residents that state needed. The scale is logarithmic, so a dot rather than a bar: bar length on a log axis would encode distance from an arbitrary origin rather than the quantity. `r ladder$state[1]` is the red dot at `r n(margin)`. The next state, `r ladder$state[2]`, is about `r n(round(runner / margin))` times further away.

Most readers guess that second place was also close — a few thousand, perhaps tens of thousands. It is `r n(runner)`. **New York's near-miss was a freak event, and the distance to second place is the evidence.** Nothing about the 2020 apportionment was generally knife-edged; one comparison out of `r nrow(ap)` happened to be.

Then a number nobody is asked to guess, because it sits inside New York's own row.

```{r overseas}
```

**New York had `r n(ny$overseas)` people counted from abroad, about `r round(ny$overseas / margin)` times the margin.** A rule about where to assign soldiers stationed in Germany moved far more people than the count difference that settled the seat. Change that rule and you change the House, without changing a single person's whereabouts.

So the obvious reading of the famous margin is wrong. `r n(margin)` people is not evidence that the census needed to be counted more carefully. It is evidence that the margin fell below the resolution of the instrument: below the census's own published uncertainty about state totals, and far below a policy choice already sitting inside the number. There is no recount for a census, and the quantity in dispute is `r round(ny$overseas / margin)` times smaller than an administrative rule inside the thing that would have to be recounted. **The near-miss is a fact about the rules, not about the accuracy of the count.**

## What this brief cannot tell you

The file has authority without uncertainty. Every population is printed as an exact whole number, and none is exact. The Bureau publishes its coverage error separately, on a different schedule, and puts state-level error in the hundreds of thousands, far above `r n(margin)`. So this brief cannot say that a more careful count would have moved the seat. Nothing here measures enumeration error, and no second source exists against which to check the first.

It is one decade. Whether *some* state is always within a few thousand people of a seat cannot be settled from a single census; that would take the 1990, 2000, and 2010 files alongside.

It says nothing about districts. Apportionment decides how many seats a state gets; districting decides where the lines go inside it, and the two are confused constantly.

And it cannot judge the rule itself. Congress chose equal proportions over five rivals that are equally defensible and do not agree, and the Supreme Court, in *Department of Commerce v. Montana* (1992), declined to second-guess the choice. Which method is fairest is a value judgment wearing technical clothes, and no column in this file settles it.

## What you should have learned

- A formula short enough to write in one line — population over $\sqrt{k(k-1)}$, highest bid wins — turns the count into the House, and recomputing it here `r ifelse(ok435 && pv_ok, "reproduces the Bureau's published seats and prices exactly", "fails to reproduce the published result")`.
- The 2020 count moved `r n(SGAIN)` seats across `r n(sum(scs$change != 0))` states, and each seat moved an Electoral College vote with it.
- `r ladder$state[1]` missed the final seat by `r n(margin)` people, while the next state in line was `r n(runner)` short: the near-miss was an outlier, not the normal state of affairs.
- The margin was smaller than the instrument. New York's own total includes `r n(ny$overseas)` people assigned to it by a rule about federal employees abroad, roughly `r round(ny$overseas / margin)` times the margin that decided the seat.

## Extensions

- `apportionment_2020.csv` — sort by `overseas`. Which states gain the most from people counted abroad, and how do those counts compare with the `r n(margin)`-person margin?
- `apportionment_2020.csv` — find the largest and smallest `people_per_seat`. How many times over is one person's share of the House worth another's, and are the two extreme states large or small?
- `priority_values.csv` — filter `house_seat` to 430 through 445 and watch the queue cross the statute. How many different states appear in those sixteen rows?
- `seat_change_states.csv` — sort by `change`, then by `seats_2020`. Do the states that lost seats tend to be larger or smaller than the states that gained them?
- **Stretch.** The Bureau publishes the same tables for 2010, 2000, and 1990 on the same page. Fetch one and find that decade's last seat and nearest miss: is a two-digit margin usual, or was 2020 unique?
- **Stretch.** Webster's method uses the divisor $k + 0.5$ instead of $\sqrt{k(k-1)}$. Re-run the queue in a spreadsheet with that one change and see which, if any, of the `r n(sum(ap$seats))` seats move.

## Sources

**Data**

- U.S. Census Bureau, *2020 Census Apportionment Results*, Table 1
  (apportionment population and seats, with change from 2010) and Table 3
  (population counted overseas), and *Priority Values for 2020 Census
  Apportionment*, which lists every seat from 51 to `r max(pv$house_seat)`
  with the value it was won at.
  <https://www.census.gov/data/tables/2020/dec/2020-apportionment-data.html>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`apportionment_2020.csv`](data/derived/apportionment_2020.csv), [`priority_values.csv`](data/derived/priority_values.csv), [`seat_change_states.csv`](data/derived/seat_change_states.csv)

**Statutes and cases**

- U.S. Const. art. I, § 2, cl. 3; amend. XIV, § 2.
- Permanent Apportionment Act of 1929, 46 Stat. 21; method of equal
  proportions fixed by Act of 15 November 1941, 55 Stat. 761 (now 2 U.S.C.
  §2a).
- *Department of Commerce v. Montana*, 503 U.S. 442 (1992).

```{r ai-prompt, echo=FALSE, results="asis"}
```
