---
title: "What a Voter File Records, and Why"
type: chapter
subtitle: "The only instrument that can tell you whether a survey is lying"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("voter-files-source-code.R")
```

```{r setup, include=FALSE}
```

The [surveys chapter](../../02-surveys/surveys-source/surveys-source-brief.html) asks people questions. The [census chapter](../../01-census-bureau/census-source/census-source-brief.html) counts everyone. This one is about a source that does neither, and was not built for us at all.

A voter file exists so that a poll worker can decide whether the person standing in front of them may vote today. That is its entire job. It is an **operational record**, kept under legal obligation by an election office. Every research use of it, including every use in this book, is a borrowing. Nothing about its design anticipated anyone studying it.

Which turns out to be its great advantage. Because it was not built to answer questions, it has no stake in the answers.

## The thing only this instrument can do

Ask Americans whether they voted and they will tell you. Ask a state and it will look it up.

```{r cmp}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=3.9}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. What people say against what the record shows.** The grey bars are the share of ANES respondents who said they voted. The teal bars are the share of one Georgia county's current registrants recorded as having voted. Hover a bar for its denominator, the bottom number its share is computed against.

**These are different populations**, and the gap is not a clean measure of overreporting. The caution below says why. What the figure establishes is narrower and more important: only one of the two numbers can be checked at all.

In the most recent study **`r p1(SAY24)`% of survey respondents said they voted**. In Houston County, Georgia, **`r p1(REC24)`% of the `r nn(INFILE)` people currently registered are recorded as having voted** in the same election.

**A caution, because this comparison is easy to overstate.** These are different denominators over different populations. A national sample of adults, against the current registrants of one Georgia county. Registrants are more likely to vote than adults in general. The gap here is not a clean measure of overreporting.

But notice what the comparison establishes anyway, which is the point of the chapter. Survey turnout is *unfalsifiable from inside the survey*. A respondent who says they voted produces a data point that no other column can contradict. Surveys do overstate turnout, consistently, in every study that has checked.

The only way anyone knows that is that somebody went to an administrative record and looked the respondents up. **The voter file is the answer key.** It is the reason we know self-reports need correcting at all.

> **Before you read on, commit to a view.** Suppose a survey and a voter file
> disagree about whether one particular person voted. **Which do you believe,
> and what would have to be true for the other one to be right?** There is a
> good answer on both sides.

## Fifty states, fifty files

There is no national voter file. The Help America Vote Act of 2002 required every state to build a single statewide computerised list, which is why a modern voter file exists at all. It required nothing whatever about what the list should contain, or what the columns should be called.

```{r raw, comment="", class.output="rawblock"}
```

```{r schtab}
```

Georgia's extract has `r NGA` columns and New Jersey's has `r NNJ`. Georgia records **race**, which is why the [RPV](../../06-putting-data-together/rpv/rpv-brief.html) and [BISG](../../06-putting-data-together/bisg-check/bisg-check-brief.html) chapters are possible at all, and which most states do not. New Jersey records a **full date of birth** where Georgia records only a birth year, which is why the [false-matches](../false-matches/false-matches-brief.html) chapter uses New Jersey.

The consequence is not merely inconvenience. **Which questions you may ask depends on which state you are standing in.** A study of racial turnout disparity is straightforward in Georgia and impossible in New Jersey from the file alone. Not because the thing being studied differs, but because one legislature decided to record a column and another did not.

## Three things between a registration and a number

### One: it contains the registered, who are nobody's population

A voter file is not a list of adults, or of citizens, or of the eligible. It is a list of people who performed an administrative act. Everyone who never registered is absent, and they are absent in a patterned way rather than a random one.

This matters most when the file is used as a denominator. Turnout "among registered voters" and turnout "among eligible adults" are different quantities, and the first is always higher, because registering is itself a filter for the kind of person who votes.

### Two: it is a snapshot with no past tense

```{r atttab}
```

**`r nn(LOST20)` people are recorded as voting in the 2020 general election and are no longer in the file** — `r p1(LOSTP)`% of everyone who voted. They moved, died, or were removed through list maintenance. Their votes were real and are now invisible to anyone holding only the current extract.

So the `r p1(REC20)`% in the table above is not the 2020 turnout rate of Houston County. It is the share of *today's* registrants who are recorded as having voted in 2020, computed on a denominator that has been reshaped by six years of arrivals and removals. The further back you reach with a current file, the more wrong the answer, and nothing in the file warns you.

A file with no past tense cannot be asked about the past. The [voter-files chapter](../voter-files/voter-files-brief.html) is what happens when you try.

### Three: the state decides who counts as on it

```{r sttab}
```

`r nn(ACTIVE)` registrations are **active** and `r nn(INACT)` are **inactive**, which is a legal status rather than a description of a person. An inactive registrant may still vote; the label records that mail was returned, or a change-of-address was matched, or a state believes they may have moved.

Whether to include them is a decision, and it changes every rate computed from the file. Include them and turnout falls; exclude them and it rises. The National Voter Registration Act governs how people move between these categories and when they may finally be removed. So the denominator of a great many published numbers is set by a federal statute about list maintenance, and almost nobody who quotes those numbers knows it.

> **A second prediction.** A state announces that it has removed 200,000
> inactive registrations. **What happens to its published turnout rate at the
> next election, and does that tell you anything about whether more people
> voted?**

## What this chapter cannot tell you

The recorded-turnout figures are one county in one state. Houston County is not Georgia and Georgia is not the country; states differ in registration rules, in list-maintenance practice, and in how faithfully history files are kept. The *shape* of the lesson travels. The numbers do not.

**Nothing here measures overreporting properly.** Doing that requires matching individual survey respondents to individual records — which is exactly what the validated-turnout studies do, and what the [validated-turnout chapter](../../06-putting-data-together/validated-turnout/validated-turnout-brief.html) is about. Figure 1 sets two different quantities beside each other to make a point about which one is checkable, and it should not be read as an estimate of how much people exaggerate.

This chapter also says nothing about accuracy *within* the file. Records are duplicated, names are misspelled, addresses go stale, and matching people across states is the subject of its own chapter for good reason. A record kept for another purpose is kept to that purpose's standard, not to yours.

Finally, a word about what these files are. They are public records, Georgia's and New Jersey's alike, and this book shows real rows from them. A chapter that showed a scrubbed imitation would teach students that the file is more private than it is. That publicness is a policy choice, made by legislatures, and it is the reason both the [false-matches](../false-matches/false-matches-brief.html) chapter's subject and this chapter's evidence exist.

## Sources

- Georgia Secretary of State. Houston County voter registration extract 71754,
  and the statewide voter history files. Public records, obtained for the 2026
  Houston County redistricting matter. Not redistributed here.
- New Jersey Division of Elections. Statewide voter list, 21 county extracts.
  Public records. Not redistributed here.
- American National Election Studies. *ANES Time Series Cumulative Data File
  (1948–2024)*, version of 5 February 2026, for self-reported turnout. Free, and
  downloadable after registering an account.
  <https://electionstudies.org/data-center/anes-time-series-cumulative-data-file/>
- Help America Vote Act of 2002, Pub. L. 107–252, §303 (statewide voter
  registration lists).
- National Voter Registration Act of 1993, 52 U.S.C. §20507 (list maintenance).
- Ansolabehere, Stephen and Eitan Hersh (2012). "Validation: What Big Data
  Reveal About Survey Misreporting and the Real Electorate." *Political
  Analysis* 20(4): 437–459.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`attrition.csv`](data/derived/attrition.csv), [`schemas.csv`](data/derived/schemas.csv), [`selfreport.csv`](data/derived/selfreport.csv), [`status.csv`](data/derived/status.csv), [`validated.csv`](data/derived/validated.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `selfreport.csv` and `validated.csv` measure the same election two ways. Compute both rates. How large is the gap between what people say and what is recorded?
- `attrition.csv` gives `pct_of_recorded` for voters who have since left the file. Work out what a validated turnout study loses by starting from a current file.
- `schemas.csv` puts two states' extracts side by side, field by field. Find what one state records and the other does not. What question can you ask in one state and not the other?
- `status.csv` counts registrations by status. Decide which statuses belong in a denominator for turnout, and write down the rule you would apply to all fifty states.

