---
title: "Key: Who Moved, and Who Never Left"
subtitle: "84-355 Democracy's Data | Migration | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
D  <- "data"
fl <- read.csv(file.path(D, "flows.csv"),        stringsAsFactors = FALSE)
st <- read.csv(file.path(D, "states.csv"),       stringsAsFactors = FALSE,
               colClasses = c(fips = "character"))
ar <- read.csv(file.path(D, "arcs.csv"),         stringsAsFactors = FALSE)
mo <- read.csv(file.path(D, "mobility.csv"),     stringsAsFactors = FALSE)
cy <- read.csv(file.path(D, "county_focus.csv"), stringsAsFactors = FALSE)
mm <- read.csv(file.path(D, "meta.csv"),         stringsAsFactors = FALSE)

mv <- function(k) mm$value[mm$key == k]
mn <- function(k) as.numeric(mv(k))
n  <- function(x) format(round(x), big.mark = ",")
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)

S        <- st[st$is_state, ]          # 50 states + DC; Puerto Rico flagged out
FOCUS    <- mv("focus_state")
CONTRAST <- mv("contrast_state")
CTY      <- mv("focus_county")
NOR      <- nrow(st) - 1               # origins available to any one state
F1 <- S[S$state == FOCUS, ]
C1 <- S[S$state == CONTRAST, ]
# the largest flow in the country that cannot be told apart from zero
bad <- fl[!fl$sig, ][which.max(fl$est[!fl$sig]), ]
# born-in-state, ordered
o  <- S[order(-S$born_state_pct), ]
# CPS endpoints
f  <- mo[1, ]; l <- mo[nrow(mo), ]
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`migration-brief.Rmd`) is now a chapter rather than a
parallel copy of the lab: prose organised by argument, under idea headings, with
seven numbered figures and no numbered steps. The lab (`migration.Rmd`) still
runs Steps 1–11 with the code exposed. **Students should read the brief first.**
**Step numbers below refer to the lab**; the chapter is cited by figure number or
section title.

**This is the course's margin-of-error session.** Every other lab shows students
a number somebody published. This one shows them the interval the same publisher
printed beside it, and what happens to a famous-looking map when you honour it.

Every number below is computed from the built CSVs at knit time. Nothing in this
document is typed in.

***

# The session in one sentence

State-to-state migration is a survey estimate, not a count, and when you apply
the margins of error the Census Bureau publishes beside it,
`r pc(mn("pct_fail_of_pairs"), 0)`% of the flows on the map turn out to be
indistinguishable from nobody moving at all — and they are exactly the thin,
surprising, interesting-looking ones.

***

# The findings

```{r headline}
c(state_pairs        = nrow(fl),
  distinguishable    = sum(fl$sig),
  pct_surviving      = round(100 * sum(fl$sig) / nrow(fl), 1),
  pct_of_movers_lost = round(100 * sum(fl$est[!fl$sig], na.rm = TRUE) /
                             sum(fl$est, na.rm = TRUE), 2),
  net_not_sig_states = sum(!S$net_sig),
  focus_net          = F1$net,
  focus_inflows_ok   = F1$sig_in_n,
  contrast_inflows_ok = C1$sig_in_n,
  county_pct_surviving = round(100 * mn("county_all_sig") /
                               mn("county_all_pairs"), 1))
```

- **`r n(sum(!fl$sig))` of `r n(nrow(fl))` state pairs
  (`r pc(100 * sum(!fl$sig) / nrow(fl), 0)`%) cannot be told apart from zero** —
  and they hold only
  `r pc(100 * sum(fl$est[!fl$sig], na.rm = TRUE) / sum(fl$est, na.rm = TRUE), 2)`%
  of the movers. The uncertainty sits where the pattern is, not where the volume
  is.
- **`r FOCUS` is down `r n(-F1$net)` net to other states**, which is
  `r pc(abs(F1$net) / F1$net_moe, 1)` times its own margin of error and therefore
  real. **`r sum(!S$net_sig)` of the `r nrow(S)` jurisdictions have a net figure
  that is not.**
- **`r FOCUS` can use `r F1$sig_in_n` of its `r NOR` inflows;
  `r CONTRAST` can use `r C1$sig_in_n`.** Same survey, same year. Log population
  and usable inflows correlate at
  r = `r pc(cor(log(S$pop1), S$sig_in_n), 2)` — a flow map of a small state is
  mostly a picture of the sample size.
- **At county level only
  `r pc(100 * mn("county_all_sig") / mn("county_all_pairs"))`% of pairs survive**
  (`r n(mn("county_all_sig"))` of `r n(mn("county_all_pairs"))` feeding `r CTY`),
  and that is after pooling five years of sample rather than one.
- **Born-in-state runs from `r pc(min(S$born_state_pct))`%
  (`r o$state[nrow(o)]`) to `r pc(max(S$born_state_pct))`% (`r o$state[1]`)**,
  nationally `r pc(100 * sum(S$born_state_est) / sum(S$pop_total))`%.
- **Americans move at `r pc(100 * l$movers / f$movers, 0)`% of the `r f$year`
  rate**; interstate moving is at `r pc(100 * l$diff_state / f$diff_state, 0)`%
  of it.

***

# The three things to get said

**1. A margin of error is not decoration.** Step 5 is the whole session. An
estimate smaller than its own margin of error is a number the survey cannot
distinguish from nobody. The published example to put on the board is
**`r bad$from_state` to `r bad$to_state`: `r n(bad$est)` ± `r n(bad$moe)`** — the
largest flow in the country whose own publisher says it is consistent with zero.
**Make somebody say what a newspaper should print instead.** The honest sentence
is "somewhere between `r n(max(0, bad$est - bad$moe))` and
`r n(bad$est + bad$moe)`, and we cannot rule out none."

**2. The uncertainty is not spread evenly, and that is what makes it dangerous.**
The failed pairs hold
`r pc(100 * sum(fl$est[!fl$sig], na.rm = TRUE) / sum(fl$est, na.rm = TRUE), 2)`%
of the movers: median surviving flow `r n(median(fl$est[fl$sig]))` people, median
failed flow `r n(median(fl$est[!fl$sig], na.rm = TRUE))`. So the map is
trustworthy about volume and unreliable about *connections* — and the long thin
exotic arc that makes a flow map worth looking at is precisely the arc that is
not there. **The more distinctive a migration claim sounds, the more likely it is
an artefact of sample size.**

**3. Step 6 is the political consequence.** `r FOCUS` gets `r F1$sig_in_n` usable
inflows and `r CONTRAST` gets `r C1$sig_in_n`, and nothing about that is a fact
about the states. It is a fact about how many of their movers landed in the
sample. **Ask what it means that the states whose migration we can describe
confidently are the large ones**, when the states that gain and lose House seats
at the margin are often not.

**The line for the board:** a survey that cannot see a flow has not measured
zero, and a map that draws it anyway is drawing its own sample size.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** What a statistic is | One question: where did you live a year ago | 5 | **Most moving is local.** Interstate is the small slice everyone argues about. |
| **2** One state | `r FOCUS` in, out, net, abroad | 5 | The `±` column is already on screen. Nobody reads it yet. That is the setup. |
| **3** The arc map | Population-weighted centres, both directions | 6 | The heaviest arc runs both ways. Net is a difference of two large numbers. |
| **4** Every state | Net per 1,000, ranked | 5 | Grey bars are already visible here. Do not explain them yet. |
| **5** None are counts | Apply the published margins | **12** | **The pivot.** Take the prediction first. Protect this step. |
| **6** Whose map survives | `r F1$sig_in_n` vs `r C1$sig_in_n` usable inflows | **8** | Size, not connectedness. The scatter makes it mechanical. |
| **7** Born in state | Stocks, not flows | 6 | `r pc(min(S$born_state_pct))`% to `r pc(max(S$born_state_pct))`%. A century of migration, not this year's. |
| **8** From abroad | A residence category, not a legal one | 5 | **Say this precisely or not at all.** Emigration is absent, not small. |
| **9** One county | `r pc(100 * mn("county_all_sig") / mn("county_all_pairs"))`% survive | 5 | Finer grain, worse estimates, *and* a different instrument. |
| **10** The back-series | CPS, `r f$year`–`r l$year` | 5 | Different survey. Do not put it on an ACS axis. |
| **11** What you can say | Including what you cannot | 4 | "Why" is not in the file. Neither is anyone who left. |

**Where the chapter sits against that table.** The rows are the lab's steps and
the minutes are for running the lab, so they stand as they are. The one place the
two documents do not line up is rows 2 and 3: the chapter treats one state and
its arc map as a single section, *The state with the largest imbalance*, ending
in Figure 1, so a class working from the chapter will not find a seam between
them. Run the pair as one eleven-minute block and take the break wherever the
room's attention gives out; the lab still has a `# Step 3` heading if you want the
old cut. The rest map one to one: row 1 to *Nobody votes on this* and *Most moving
is local*, row 4 to Figure 2, row 5 to *An estimate smaller than its own margin of
error* and Figure 3, row 6 to *Whose map can be drawn at all* and Figures 4–5,
row 7 to Figure 6, row 8 to *A residence category, not a legal one*, row 9 to
*Finer grain, worse estimate*, row 10 to Figure 7, and row 11 to *What a survey of
movers can testify to*.

**Step 5 deserves the prediction, and it is the best prediction in the course.**
Ask the room: *there are `r n(nrow(fl))` ordered pairs of states — how many
describe a flow the survey can tell apart from zero?* Almost everyone guesses
something in the nineties. It is `r pc(mn("pct_sig_of_pairs"), 0)`%. Then ask the
follow-up before revealing it: *and how many of the actual movers are in the
pairs that failed?* Having just been told a third of the map is unusable, the
room expects a large number. It is
`r pc(mn("pct_volume_in_failed_pairs"), 2)`%. **Both surprises are needed** — the
first says the map is mostly noise, the second says the noise is weightless, and
the interesting claim lives in the gap between them.

**If you are short on time**, protect Steps 5, 6 and 9. Everything else is
context for those three.

***

# Handling the room

- **Nobody writes code.** Say it at the start; the lab is dense with output and a
  few students read that as a coding session.
- **The Big Sort is the hook.** Most students have met the claim that Americans
  are sorting themselves politically. The chapter's *The argument this data
  cannot settle* is where to spend any spare minutes: Mummolo and Nall, Martin
  and Webster, Chen and Rodden.
- **Do not let the room settle the sorting argument.** This data has no political
  variable at all. The honest contribution is Step 10's constraint: whatever
  sorting is happening, it is happening while the mechanism runs at
  `r pc(100 * l$diff_state / f$diff_state, 0)`% of its `r f$year` speed.
- **Step 8 is where a class can go wrong fast.** "From abroad" is a statement
  about geography. It counts returning citizens, military families, students and
  people arriving from Puerto Rico and the Island Areas alongside immigrants, and
  the file separates none of them. Say so before anyone infers otherwise.

***

# Option A — Audit a state's map

## Verified output

```{r a}
rbind(head(S[order(-S$sig_in_n), c("state", "pop1", "sig_in_n", "sig_out_n")], 3),
      head(S[order( S$sig_in_n), c("state", "pop1", "sig_in_n", "sig_out_n")], 3))
c(median_usable_inflows = median(S$sig_in_n),
  states_below_half     = sum(S$sig_in_n < NOR / 2),
  cor_logpop_usable     = round(cor(log(S$pop1), S$sig_in_n), 2))
```

**3/4:** Picks a state, reports its `sig_in_n` and `sig_out_n`, and says the map
would be partly unusable.

**4/4:** Compares a large state to a small one and states the mechanism — the
survey samples households, so a state with fewer residents contributes fewer
movers to every cell, and the margin of error is large relative to a small
estimate. Says explicitly what the map would be honest about (total volume, the
handful of heavy neighbouring flows) and what it would fake (every thin
long-distance arc).

**4/4, best:** Notices that `sig_in_n` and `sig_out_n` are close for most states
and asks why, or works out that **no state's flow map is complete**: even the
median jurisdiction loses `r NOR - median(S$sig_in_n)` of its `r NOR` inflows,
and `r sum(S$sig_in_n < NOR / 2)` of `r nrow(S)` lose more than half. A student
who converts this into a rule ("do not draw an arc map below population X") has
understood the step better than the lab states it.

***

# Option B — The league table nobody checks

## Verified output

```{r b}
z <- S[!S$net_sig, c("state", "net", "net_moe", "net_per1k")]
z <- z[order(-abs(z$net)), ]
z$net_moe <- round(z$net_moe)
head(z, 8)
nrow(z)
```

**3/4:** Lists the states and reports that their net migration cannot be
distinguished from zero.

**4/4:** Writes the corrected sentence. Something like: *`r z$state[1]`'s
population change from interstate migration in `r mv("acs_year")` was
`r n(z$net[1])`, with a margin of error of ± `r n(z$net_moe[1])` — the survey
cannot tell whether the state gained or lost.* The move that earns the 4 is
refusing the direction, not just adding an interval.

**4/4, best:** Notices that these are mostly **mid-sized states** — not the
smallest — and works out why: net is a difference of two large numbers, so its
margin of error is the root of the sum of two squared margins and can easily
exceed the difference itself. **A state can have perfectly well-measured inflows
and outflows and a completely unmeasured net.** That is the sharpest available
observation on the whole lab, and it explains why "winners and losers" stories
are worse than the data they are built on.

***

# Option C — Born here versus arrived

## Verified output

```{r c}
round(cor(S$born_state_pct, S$net_per1k), 2)
r <- resid(lm(net_per1k ~ born_state_pct, data = S))
rbind(head(S[order(-r), c("state", "born_state_pct", "net_per1k")], 3),
      head(S[order( r), c("state", "born_state_pct", "net_per1k")], 3))
```

**3/4:** Reports the weak correlation and names a couple of states that do not
fit.

**4/4:** Explains what the two variables are measuring on different clocks.
Born-in-state share is a **stock** — the residue of every migration a state has
ever had, plus its birth rate and its retention of the people born there.
Net migration per 1,000 is a **flow**, one year wide. There is no reason for them
to agree, and the weak correlation is the expected result rather than a puzzle.

**4/4, best:** Picks the outliers apart individually and finds that they are
different stories: a state can be low on born-in-state and *losing* people
(decades of churn that has stopped), or high on born-in-state and *gaining*
(recent growth that has not yet worked through the stock). Best of all is the
student who points out that the outlier list is unstable because
`r sum(!S$net_sig)` of the net figures cannot be told from zero anyway —
**`r pc(100 * sum(!S$net_sig) / nrow(S), 0)`% of the residuals in this scatter
are noise, and the lab's own Step 5 says so.**

***

# Option D — Their own question

No verified output. Grade the question as much as the answer.

**3/4:** A sensible question answered correctly from the files.

**4/4:** A question whose answer required checking a margin of error, or one that
compares two things the lab kept separate and says why the comparison is limited.

**4/4, best:** A question the data **cannot** answer, correctly identified as
such, with a statement of what source would be needed. Emigration and motive are
the two obvious ones and both are named in the lab; credit anyone who finds a
third.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given.
- **The move that earns a 4** is treating the margin of error as part of the
  finding rather than a caveat appended to it.
- **A 2** is "the data has margins of error so we can't really know anything" —
  true-sounding, useless, and the opposite of what Step 5 shows. The point is not
  that nothing is knowable; it is that
  `r pc(100 - mn("pct_volume_in_failed_pairs"), 1)`% of the movers are in flows
  we *can* measure, and the failures are concentrated somewhere specific.
- **Also a 2**: reporting any of these figures as a count of people.
- **Reward anyone who asks about the people who left the country.** Emigration is
  absent from this dataset, not small in it, and noticing the absence of a
  variable is harder than reading one that is present.
- **Reward anyone who separates "from abroad" from immigration** without being
  prompted.

## Anticipated questions

- *"Why doesn't the Census just count everybody?"* — It does, every ten years,
  and that count is what apportions the House. This is the survey that fills the
  nine years in between; annual migration is not something the decennial census
  measures at all.
- *"If the margin is that big, why publish the number?"* — Because the Bureau
  publishes the margin with it. The failure is downstream: newspapers, state
  development offices and think tanks republish the estimate and drop the
  interval. Nothing in this lab is a criticism of the Census Bureau.
- *"Couldn't they just pool more years?"* — That is exactly what the
  county-to-county table does (`r mv("c2c_window")`, five years), and even then
  only `r pc(100 * mn("county_all_sig") / mn("county_all_pairs"))`% of pairs
  survive. Pooling also costs you the ability to say *when*.
- *"Does this mean the flow maps I see online are wrong?"* — Not wrong about the
  heavy flows. Wrong to draw the thin ones with the same confidence, and usually
  silent about which is which.
- *"Why is Puerto Rico in the flow table but not the rankings?"* — It is an
  origin and destination for migration, so it belongs on the map, but its
  place-of-birth categories are constructed differently and it is not a state.
  The build script flags it rather than silently ranking it.
- *"Which state is really winning?"* — Only the ones whose net clears its own
  margin. `r nrow(S) - sum(!S$net_sig)` of `r nrow(S)` do; the top two are
  `r head(S[order(-S$net), "state"], 1)` and
  `r head(S[order(-S$net), "state"], 2)[2]`. For the rest, the answer is that the
  survey does not know.

## Connections

- **`census` and `apportionment`** — where migration ends up:
  seats. This lab is the nine years between the counts, and the reminder that
  the annual numbers everyone argues about are not the ones that move seats.
- **`poll-simulation` and `gss-confidence`** — the other places the course handles
  sampling error. Run one of them before this if you can; students who have
  already met a confidence interval get Step 5 much faster.
- **`levels-of-aggregation`** — the same lesson from the other direction: state
  totals are stable and county cells fall apart. Step 9 is that lab's point with
  a published margin attached.
- **`regional-shift`** — seventy years of population moving between regions,
  measured by the census rather than a survey. Useful contrast: there, the number
  is a count and the hard part is what it means.
- **`bisg-check`** — the other lab where an estimate is routinely
  presented without the uncertainty its own authors published.
