---
title: "Peremptory Strikes and Jury Composition"
type: brief
subtitle: "Peremptory strikes in one Mississippi district, and a pattern no single trial can show"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("jury-selection-code.R")
```

```{r setup, include=FALSE}
```

A jury is the only part of American government staffed by lottery. It exists so that the state's gravest power, taking a person's liberty or life, is used by ordinary people rather than professionals. Its authority comes from being drawn out of the community, not picked out of it.

Each side in a trial may remove a fixed number of jurors without giving any reason. That is a **peremptory strike**, and since *Batson v. Kentucky* (1986) exactly one reason for it is forbidden: race. Does race predict who gets struck? A rule against one reason, inside a procedure built on not requiring reasons, can only be enforced by inference, and inference needs a record.

In a Mississippi courtroom in 1992 a prosecutor removed `r KS` people from a jury panel. `r NB + NW` of the people in that room were his to remove if he wanted them gone, `r NB` Black and `r NW` white. He struck all `r NB` of the Black jurors and `r KS - SB` of the `r NW` white ones, and gave no reason for any of them. None was asked of him.

## The test

Using jury-selection records from `r nrow(t)` trials in Mississippi's Fifth Circuit District, `r YR0` to `r YR1`, published by APM Reports (<https://github.com/APM-Reports/jury-data>). No court keeps this record. A clerk logs that a juror was excused, and nothing in that file says what race the juror was, because no rule requires it to. The evidence exists because a newsroom built it. Reporters for the second season of *In the Dark* pulled the paper files of 225 trials and coded every prospective juror by hand.

Two properties make it the right evidence rather than merely the available evidence. It is juror by juror, so a rate can be computed over the people a side was permitted to remove rather than over everyone in the room. And it records **both** sides' strikes, which supplies a comparison group nobody had to construct. The defense drew from the same panels, in the same trials, under the same rules, wanting the opposite thing.

These are administrative records: paperwork kept while somebody was running a court, about people who never asked to be described. [The section introduction](../admin-records-source/admin-records-source-brief.html) sets out what that kind of file can bear. Such a file usually counts the events an institution acted on and says nothing about the population behind them. This one is the lucky case, because that population is inside the file.

## The data

Everything below reads two small tables. `strikes.csv` has four rows and five columns: `side`, `race`, `eligible`, `struck` and `pct`. `trials.csv` is one row per trial, with `trial`, `black_eligible`, `black_struck`, `white_eligible` and `white_struck`. Both come from APM's `jurors.csv`, one row per prospective juror per trial, `r n(14874)` rows in all.

```{r counts}
```

Three decisions in that reduction set what the rates mean.

**The bottom number is eligibility, not attendance.** A rate needs a bottom number to divide by, and here it is the jurors a given side was entitled to strike. APM's `strike_eligibility` column records that, and nearly half the file reads `n/a`: jurors already gone for cause, or never reached before the panel filled. Counting those would make both sides look far more restrained than they were, and would do it to the two sides unequally. The whole comparison rests on this one column, which most court records do not have.

**Nine outcomes were collapsed into one question.** APM's `struck_by` tells a juror struck by the state from one struck for cause, one excused, one never noted, and one whose fate is unknown. Only "Struck by the state" and "Struck by the defense" count as a strike here. The `r n(571)` jurors struck without notation count as struck by neither side, the conservative choice, and that pushes both rates down.

**Two races were kept and `r n(4756)` jurors were dropped.** Nearly all are jurors whose race the paperwork never recorded, `r pc(100 * 4752 / 14874)`% of the file. They are unlikely to be a random third, because a juror the record never described is not a random juror. Four more were coded Asian or Latino, too few to compute anything from. A two-category world appears in the output that was not in the input.

One column deserves a sentence of its own, because almost nothing else in this book has one. `race_source` records **how the reporters knew**: off a jury strike sheet for most jurors, off a transcript of the questioning for a couple of hundred, and from other documents for a few dozen. Where a value came from is normally something a reader has to guess at. Here it sits beside the value, which makes the gaps arguable rather than merely present.

Here is the whole of `strikes.csv`:

```{r clean-jury}
```

> **Before you read on, commit to two numbers.** The state struck
> `r pc(g("state","Black","pct"))`% of the Black jurors it was allowed to
> strike and `r pc(g("state","White","pct"))`% of the white ones. Now the
> other lawyer in that room. The defense drew from the same panels, wanting the
> opposite outcome. Write down the share of eligible
> **white** jurors you think the defense struck, and the share of eligible
> **Black** ones. Then write down how many of the `r nrow(t)` trials you would
> expect chance alone to make look lopsided.

## What it says about democracy

Pool every strike that office made over `r YR1 - YR0` years, by the juror's race:

```{r state-rates}
```

**The state struck `r pc(g("state","Black","pct"))`% of the eligible Black jurors and `r pc(g("state","White","pct"))`% of the eligible white ones**, a ratio of `r pc(g("state","Black","pct")/g("state","White","pct"))` to one. There is an answer to that table, and it is not frivolous. Struck jurors differ in ways that have nothing to do with race: a prosecuted relative, a doubt about the death penalty, an acquaintance with a witness. If those are spread unevenly across the community, a race-neutral rule produces a racial gap by itself.

What answers the reply is the other lawyer.

```{r mirror-static, eval=!knitr::is_html_output(), fig.height=4}
```

```{r d3-mirror, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The two sides strike the same panels in opposite directions.** Four bars, because the claim is about four rates measured on the same jurors, and a bar makes shares comparable at a glance. Across `r nrow(t)` trials the state struck `r pc(g("state","Black","pct"))`% of eligible Black jurors and `r pc(g("state","White","pct"))`% of eligible white ones; the defense struck `r pc(g("defense","White","pct"))`% of eligible white jurors and `r pc(g("defense","Black","pct"))`% of eligible Black ones. The two tall bars differ by `r pc(abs(g("state","Black","pct") - g("defense","White","pct")))` points and the two short ones by `r pc(abs(g("state","White","pct") - g("defense","Black","pct")))`.

**The pattern reverses almost exactly.** Whatever made a Black juror worth striking to the state made a white juror worth striking to the defense at nearly the same rate. The legitimate-characteristics explanation now has to name a trait that is spread unevenly by race *and* valued in precisely opposite directions by the two parties. None has been proposed, in this litigation or in the research it belongs to.

Now ask the question one trial at a time. Take the Black jurors the state could strike and the Black jurors it struck, and work out the chance of a split that lopsided if the strikes had ignored race.

```{r many-trials}
```

`r obs05` of the `r nrow(t)` trials come in significant at the conventional line. The textbook shortcut says chance alone should produce `r pc(naive05, 1)` of them, making the observed count `r pc(naive_ratio, 0)` times too high. **That shortcut is wrong here, and wrong in the direction that works against the finding.** These tests run on small whole numbers, so a trial with four eligible Black jurors and three strikes can produce only a handful of possible answers, none of them 0.04. A test whose next available value below 0.11 is 0.03 never fires at 0.05.

So the benchmark was simulated rather than assumed. Redraw every trial's strikes from its own numbers, keeping its eligible jurors and its strike count, `r n(NSIM)` times, and count how often that imaginary courthouse produces a lopsided trial. Chance produces `r pc(exp05, 1)` such trials, not `r pc(naive05, 1)`. The observed count is `r pc(ratio05, 0)` times the honest benchmark. Being careful about the benchmark makes the gap `r pc(ratio05/naive_ratio, 0)` times larger, which is the opposite of what most readers write down.

```{r thresh-static, eval=!knitr::is_html_output(), fig.height=4.3}
```

```{r thresh-d3, results='asis', eval=knitr::is_html_output()}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

**Figure 2. The result does not depend on choosing 0.05.** Trials called lopsided at every threshold from `r min(THR)` to `r max(THR)`, because a finding that lives at one arbitrary line is not a finding. What happened in this district is drawn against the mean and the middle 95% of `r n(NSIM)` simulated courthouses. In each of those a trial keeps its own jurors and its own strikes, but race does no work. At the conventional line the counts are `r obs05` against `r pc(exp05, 1)`, and the red line never enters the gray band at any threshold on the axis.

`r sum(t$black_eligible > 0 & t$black_struck == t$black_eligible)` of the `r nrow(t)` trials removed every eligible Black juror. The three most lopsided:

```{r flowers}
```

**The most extreme is the third trial of Curtis Flowers. `r FL$black_eligible` Black jurors were eligible and `r FL$black_struck` were struck; of `r FL$white_eligible` eligible white jurors, `r FL$white_struck` were.** The chance of that split under race-neutral striking is `r formatC(min(t$p), format="e", digits=1)`, roughly one in five billion. Flowers was tried six times for the same crime by the same prosecutor. In *Flowers v. Mississippi* (2019) the Supreme Court threw out his conviction, resting substantially on that pattern across the six trials. The charges were dropped, and he went home after twenty-three years.

Notice what it took. Not a statistical argument in one courtroom, and not the mirror in Figure 1, which no judge had ever seen. It took one defendant tried so many times by one office that the career pattern and the single case became the same object. And it took reporters reading paper, because the court never wrote down the fact the rule depends on.

## What this brief cannot tell you

**It cannot win a single case.** *Batson* is litigated before a judge who has seen one panel. The 1992 trial that opens this brief carries a probability of `r pc(100 * PV, 1)`%, about one chance in `r ONE_IN`, which is roughly what an ordinary motion has to work with. It loses, and the rule is enforced at exactly the level where the evidence is weakest.

**It cannot report a reason.** No row records why anyone was struck, because the procedure was designed not to require reasons. The file shows what was done and never what was said, so no explanation a prosecutor offered or a judge accepted is in it.

**It cannot speak for the country.** Mississippi's Fifth Circuit is one district, read precisely because it produced *Flowers*, which is the strongest possible reason to expect it to be unusual.

**It cannot close the missing third.** Race is unknown for `r n(4752)` of `r n(14873)` jurors, and if that gap were related both to race and to being struck, the rates in Figure 1 would shift. The mirror makes that a stretch, since the same jurors are missing from both sides of it.

**And it says nothing about verdicts.** This is a record of who was excluded, not of what the resulting juries decided.

## What you should have learned

- The state struck `r pc(g("state","Black","pct"))`% of eligible Black jurors against `r pc(g("state","White","pct"))`% of eligible white ones, a ratio of `r pc(g("state","Black","pct")/g("state","White","pct"))` to one.
- A disparity on its own proves little, and this file answers that objection from inside itself: the defense struck `r pc(g("defense","White","pct"))`% of eligible white jurors and `r pc(g("defense","Black","pct"))`% of eligible Black ones, from the same panels on the same mornings.
- Being careful about a benchmark can make a finding larger. `r obs05` of `r nrow(t)` trials look lopsided, against `r pc(naive05, 1)` on the textbook shortcut and `r pc(exp05, 1)` when the null is simulated from each trial's own numbers.
- Aggregate evidence and courtroom evidence are different things: one trial gives `r pc(100 * PV, 1)`%, while `r sum(t$black_eligible > 0 & t$black_struck == t$black_eligible)` of `r nrow(t)` trials removed every eligible Black juror.
- The record-keeping is part of the finding. No court recorded jurors' race, a newsroom did it by hand, and outside these `r nrow(t)` trials the question still cannot be answered.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `strikes.csv` has `eligible` and `struck` for each `side` and `race`. Recompute `pct` for all four rows, then the ratio within each side. Which side produces the larger gap?
- `trials.csv` has `black_eligible` and `black_struck`. Compute the share struck in each trial and sort. How many trials sit at zero, and do they cluster in particular years?
- `trials.csv` also carries `white_eligible` and `white_struck`. Compute both shares per trial and count the trials where the white share is higher. Is the district-wide pattern in every trial, or in most of them?
- The `trial` column begins with the year. Group by year and compute the Black strike share. Did the office change over the period?
- **Stretch.** APM published the full juror file too. Fetch it, count how many jurors each value of `race_source` accounts for, and ask whether the unknowns fall evenly across trials.
- **Stretch.** Studies in Philadelphia, North Carolina and Louisiana measured the same thing. Find one and check whether its strike rates would clear a simulated benchmark rather than the textbook one.

## Sources

**Data**

- APM Reports, *In the Dark* season 2, jury data.
  <https://github.com/APM-Reports/jury-data> — `jurors.csv`, `trials.csv` and
  `voir_dire_answers.csv`, coded from the court records of 225 trials in
  Mississippi's Fifth Circuit District, reduced here to the `r nrow(t)` trials
  with at least ten eligible jurors.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`strikes.csv`](data/derived/strikes.csv), [`trials.csv`](data/derived/trials.csv)

**Cases and literature**

- *Batson v. Kentucky*, 476 U.S. 79 (1986), including Marshall, J.,
  concurring, who argued that requiring reasons only teaches lawyers to
  produce them.
- *Flowers v. Mississippi*, 588 U.S. ___ (2019).
- Grosso, C. M. and O'Brien, B. (2012). "A Stubborn Legacy: The Overwhelming
  Importance of Race in Jury Selection in 173 Post-*Batson* North Carolina
  Capital Trials." *Iowa Law Review* 97: 1531–1559.

```{r ai-prompt, echo=FALSE, results="asis"}
```
