---
title: "How Long a Congressional Career Lasts, 1789–2025"
type: brief
subtitle: "Two-thirds of the newest careers have not finished, and the arithmetic that counts them anyway"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("careers-code.R")
```

```{r setup, include=FALSE}
```

Congress is regularly described as a job nobody leaves: safe seats, career politicians, members who stay for decades. Is that true? How long does a congressional career actually last? The obvious way to answer is to take everybody who has ever served, count each person's Congresses, and average — a one-line calculation over `r n(NCAR)` careers.

The answer that calculation gives is wrong, and wrong in a direction that reverses the finding. Done that way, the record says careers have been collapsing since the 1990s. Done properly, it says the opposite: careers today are the longest they have ever been. The difference is entirely about what to do with the people who have not finished yet.

## The test

The data is Voteview's congressional membership file, `HSall_members.csv`, covering the 1st through `r LASTC`th Congresses (1789 to the present), obtained from [Voteview](https://voteview.com/). One row is one member serving in one Congress, compiled from the official rosters with one permanent identifier per person, so a member who leaves and comes back is still one member.

This is the right kind of source because a career is a fact about membership, not about elections. The [roster chapter](../rosters-source/rosters-source-brief.html) introduces this cluster's records and the distinction they rest on: a roster says who held the seat and for how long, which is exactly the unit a career is made of. A returns file cannot supply it. `r n(NGAP)` careers in this file are not one continuous stretch, and a file of election results has no way to know those terms belong to the same person.

The test runs in two parts. First, compute career length the obvious way and see what that arithmetic claims. Then count the unfinished careers with a method built for exactly this problem, and see whether the claim survives.

## The data

The build strings the membership rows into `careers.csv`, one row per career, where a career is one member in one chamber. Its columns, by name:

```{r varmap}
```

Two decisions are baked into that table. A person who served in both chambers appears as two careers, one per chamber, so a senator's House years do not inflate the Senate numbers. And `tenure` counts Congresses actually served rather than the span from first year to last, so the `r n(NGAP)` careers with a gap in them are not credited with time spent out of office.

The column that decides everything is the last one. `r n(NON)` of the `r n(NCAR)` careers, or `r p1(PCTON)`%, belong to members who were still serving in the `r LASTC`th Congress. Those careers have a beginning and a length so far, and no end. The method that handles them calls them **censored**: the record stops before the event does. Censoring is not missing data. Nothing failed to be recorded; the fact the analysis wants has simply not happened yet.

```{r one-row}
```

Two members of the same entering class. The first served `r e$tenure` Congress`r if (e$tenure > 1) "es"` and left; that career is `r e$tenure` long and always will be. The second has served `r g$tenure` so far and may serve twenty more. **Recording the second as a career of length `r g$tenure` is a claim about the future, not a measurement.**

> **Before you read on, write a number down.** Across all `r n(NCAR)` careers
> in the file, what is the median length, in two-year Congresses? Commit to a
> number. Then commit to a direction: are careers begun since 1995 longer or
> shorter than careers begun a century earlier?

## What it says about democracy

Start with one entering class, small enough to see whole.

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=5.2}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. `r n(ENTN)` members, one entering class, `r yr(ENTRY)`.** One bar per career, all starting at the same left edge, because everybody here began together — the form makes the censoring visible as geometry. A bar ending in a tick is a finished career. A bar ending in an arrow is not: `r n(ENTON)` of these `r n(ENTN)` members were still serving in the `r LASTC`th Congress. Hover any bar for the member.

Now the question. What is the average career length of this class? The bars that end can be averaged. The arrows cannot, and they are not a small remainder to be swept up, because they are systematically the *long* ones. A career still running at fifteen years is exactly the kind of career that will be long.

Take every career that began in the 104th Congress or later: `r n(MODN)` careers, of which `r n(MODON)` (`r p1(MODPCT)`%) are still running. Compute the median three ways.

```{r naivetab}
```

Counting the unfinished careers at their current length says `r MODALL`. Dropping them says `r MODEND`. The estimator that handles them says **`r MODKM`**. The two shortcuts do not bracket the right answer; both sit below it, because both, in different ways, throw away the evidence that careers get long.

That error compounds into a false trend, and the trend is the exception this brief exists to tell. Here is every entering cohort, with how much of it is unfinished and the average length of the careers that *have* ended:

```{r cohorttab}
```

Read the last column down the page and it looks like a finding: careers grew to about six Congresses by the 1980s and then fell off a cliff, to `r p1(LAST_CO$mean_ended)` in the newest cohort. Read the column beside it and the finding evaporates. **`r p1(LAST_CO$pct_ongoing)`% of that cohort has not finished**, so the only careers available to average are the ones that ended early. The collapse is a measurement artifact: the longer a career runs, the less likely the record has seen it end.

The Kaplan–Meier estimator does one thing, and it fits in a sentence. It walks up the lengths in order, asks at each length how many careers were still running and what share of them ended right there, and multiplies those shares together into a survival curve. An unfinished career counts as evidence exactly as far as it is evidence, then leaves without ever being counted as an ending.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The share of careers still running, after each number of Congresses.** The dashed line is a half; where a curve crosses it is the median career: `r MH` Congresses for the House, `r MS` for the Senate, `r MKM` overall. Hovering gives the estimate and the number of careers still at risk, which is the honest way to read the right-hand end.

So Congress is not a lifetime job. Roughly a third of House careers are over after a single Congress — members who arrive in a wave and leave in the next one. The median House career is `r MH` Congresses, about `r 2 * MH` years. The Senate curve starts lower and flattens, because a six-year term is a higher floor.

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. The same estimator, four eras, by the Congress in which the career began.** The legend carries each era's number of careers and how many are unfinished. The newest curve rests on a group that is `r p1(LAST_CO$pct_ongoing)`% incomplete, and a reader is entitled to know that before believing it.

The curves separate in the order the ordinary account of Congress would predict, and the modern one is the *highest*. A career begun since the 104th Congress has a median of **`r MODKM` Congresses**, against `r OLDKM` for one begun between the 50th and the 79th. Careers roughly doubled. Both facts are true at once: the typical career is a few Congresses, not a lifetime, and today's careers are the longest the record has ever held.

## What this brief cannot tell you

**The estimator assumes the unfinished careers are like the finished ones at the same length**, and that is the one assumption it cannot check. Members still serving may differ from members who left after the same number of Congresses: safer seats, better health, a different era. If they do, the curve is wrong in a direction nothing in the file reveals. Kaplan–Meier is not a way of knowing the future. It is a way of not pretending the future has already happened.

**Nothing here says why a career ended.** Retirement, defeat, a run for another office and death are one event in this file. The [retirements](../retirements/retirements-brief.html) and [primary-defeats](../primary-defeats/primary-defeats-brief.html) briefs take that question up, and neither can answer it from the membership file alone.

**And every number is relative to the newest Congress in the file.** Rebuild this brief after the next election and every figure moves, not because the past changed but because the record saw more of it.

## What you should have learned

- The median congressional career is `r MKM` Congresses — `r MH` in the House, `r MS` in the Senate — so the typical career is closer to `r 2 * MKM` years than to a lifetime.
- `r n(NON)` of `r n(NCAR)` careers (`r p1(PCTON)`%) are unfinished, and the unfinished ones are systematically the long ones, which is why they cannot be dropped or frozen at their current length.
- Both shortcuts bias the answer the same way: for careers begun since the 104th Congress, counting the unfinished at current length gives a median of `r MODALL` and dropping them gives `r MODEND`, against `r MODKM` from the estimator.
- The apparent collapse of careers after the 1980s is an artifact of censoring: `r p1(LAST_CO$pct_ongoing)`% of the newest cohort has not finished, so only its short careers can be averaged.
- Counted honestly, careers begun since the 104th Congress run about twice as long as careers begun in the late nineteenth century: `r MODKM` Congresses against `r OLDKM`.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `naive.csv` shows two wrong ways to count career length, with `what_it_assumes` beside each `median`. Work out which direction each error runs. Which one would a journalist be more likely to publish?
- `careers.csv` flags every career as `ongoing` or finished. Count the ongoing ones. Then compute the median `tenure` with them and without them, and see how far apart the answers are.
- `km.csv` handles the unfinished careers properly, one length at a time. Read `surv` down the column. At what point in a career does the risk of it ending drop away?
- `cohorts.csv` groups careers by when they started and reports `pct_ongoing`. Which cohorts can you say anything reliable about, and which are still being written?

Two stretch ideas point past the spreadsheet. When the `r LASTC + 1`st Congress convenes, download the updated membership file from Voteview and rebuild: predict which of Figure 3's curves moves most, and in which direction, before you look. And decide how you would count the members who served in the House and then the Senate — one career or two? Recompute the Senate median under your rule and see what changes.

## Sources

**Data**

- Lewis, Jeffrey B., Keith Poole, Howard Rosenthal, Adam Boche, Aaron Rudkin
  and Luke Sonnet. *Voteview: Congressional Roll-Call Votes Database.*
  `HSall_members.csv`, 1st–`r LASTC`th Congresses. <https://voteview.com/>

**Academic literature**

- Kaplan, E. L. and Paul Meier (1958). "Nonparametric Estimation from
  Incomplete Observations." *Journal of the American Statistical Association*
  53(282): 457–481. The estimator is written out in four lines of base R and
  checked against `survival::survfit` every time it runs; it fails rather than
  publishing a curve that disagrees with the reference implementation.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`careers.csv`](data/derived/careers.csv), [`cohorts.csv`](data/derived/cohorts.csv), [`km.csv`](data/derived/km.csv), [`naive.csv`](data/derived/naive.csv), [`timeline.csv`](data/derived/timeline.csv)

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
