---
title: "Key: A Quarter of a Million Dollars and Nine Weeks"
subtitle: "84-355 Democracy's Data | GOTV field experiments — retired | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
g <- read.csv("data/gotv_tactics.csv", stringsAsFactors = FALSE)
e <- g[g$effective, ]
```

```{r vals, include=FALSE}
# Every figure quoted in this key is computed here and inserted inline, so key
# and brief cannot drift apart. BUDGET, HRS and CPH are the lab's own constants.
BUDGET <- 250000; CPH <- 25

CAN  <- e[which.min(e$contacts_per_vote), ]     # most effective per contact
VOL  <- e[which.min(e$cost_per_vote), ]         # cheapest per vote
ROBO <- g[g$tactic == "Robocalls", ]
NBLANK <- sum(!g$effective)

BUY  <- function(B, cpv = VOL$cost_per_vote) floor(B / cpv)
ALLC <- BUY(BUDGET, CAN$cost_per_vote)

# Option B: volunteer hours -> votes, spending any remainder on canvassing
TOT <- function(h) {
  vv   <- floor(h * CPH / VOL$contacts_per_vote)
  left <- BUDGET - vv * VOL$cost_per_vote
  if (left > 0) vv + floor(left / CAN$cost_per_vote) else BUY(BUDGET)
}
BIND <- function(h) if (TOT(h) < BUY(BUDGET)) "volunteers" else "money"
CROSS <- {                                     # fewest hours at which money binds
  hh <- 1:20000
  hh[which(sapply(hh, function(h)
    floor(h * CPH / VOL$contacts_per_vote) * VOL$cost_per_vote) >= BUDGET)[1]]
}

RHO <- cor(e$contacts_per_vote, e$cost_per_contact, method = "spearman")

pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(round(x), big.mark = ",", trim = TRUE)
dl <- function(x) paste0("$", nn(x))
d2 <- function(x) paste0("$", formatC(x, format = "f", digits = 2))
```

All Part 5 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

Mobilization is the one part of campaigning that has been tested
experimentally, and the experiments say the most *effective* tactic and the
cheapest *per vote* are different tactics — while several of the things
campaigns spend most on do nothing measurable at all.

***

# The findings

```{r headline}
e[, c("tactic", "contacts_per_vote", "cost_per_vote")]
g$tactic[!g$effective]
```

- **Door-to-door is the most effective** (1 additional vote per
  `r CAN$contacts_per_vote` conversations) and **volunteer phones are the
  cheapest per vote** (`r dl(VOL$cost_per_vote)` vs `r dl(CAN$cost_per_vote)`).
  Not the same tactic — the whole lab hangs on this.
- **`r NBLANK` tactics have no measurable effect**: campaign direct mail, email,
  and leaflets left on doors. Not small effects — nothing distinguishable from
  zero.
- Spending the whole `r dl(BUDGET)` on the best-value tactic buys about
  **`r nn(BUY(BUDGET))` votes**. Wisconsin 2024 was decided by roughly 29,000.

***

# The idea to teach

Two ideas, and the second is the harder one.

**First:** effectiveness and cost-efficiency are different axes, and campaigns
optimise the wrong one when they confuse them.

**Second, and this is the one worth the time:** *"no significant effect" is a
finding, not an absence of one.* Students read the blank cells as missing data.
They are results — expensive, replicated, negative results. Getting them to
feel the difference between "we don't know" and "we know it doesn't work" is
the most transferable thing in this lab.

**One housekeeping note before you teach it.** `data/build-data.R` still says in
its header that **four** tactics have no statistically significant effect, while
the file it writes flags `r NBLANK`. The data is what this key and the brief
report; the comment is what is stale. The likeliest candidate for the fourth is
robocalls — see the anticipated questions — so reconcile the two before this lab
comes out of retirement.

***

# Option A — A different budget

## Verified output

```{r a}
for (B in c(25000, 250000, 5000000)) {
  v <- floor(B / e$cost_per_vote)
  cat(sprintf("$%9s -> best %s, %s votes | order: %s\n", format(B, big.mark=","),
      e$tactic[which.max(v)], format(max(v), big.mark=","),
      paste(substr(e$tactic[order(-v)], 1, 12), collapse=" > ")))
}
```

| budget | votes |
|---|---|
| `r dl(25000)` | `r nn(BUY(25000))` |
| `r dl(BUDGET)` | `r nn(BUY(BUDGET))` |
| `r dl(5e6)` | `r nn(BUY(5e6))` |

**The ranking is identical at every budget**, and votes scale linearly.

## What a good answer says

**3/4:** Reports that the numbers scale and the best tactic does not change.

**4/4:** Sees why that is suspicious. **A model in which money converts to votes
at a fixed rate has no diminishing returns in it** — but a $5M program cannot
buy `r nn(BUY(5e6))` votes at `r dl(VOL$cost_per_vote)` each, because it would
run out of volunteers, run out of doors worth knocking, and start contacting
people who were always going to vote. The linearity is an artifact of using an
average cost as if it were a constant.

**4/4, best:** turns that into the right conclusion about money in politics.
This lab looks like it says money buys votes at a known price; what it actually
says is that *the first* `r dl(BUDGET)` is efficient. The question "does money buy
elections" needs the marginal, not the average, and this table has only the
average. A student who says "I would need the experiments' dose-response
curves, not their headline" is doing graduate-level thinking.

***

# Option B — Volunteer supply

## Verified output

```{r b}
for (h in c(1000, 2000, 4000, CROSS, 12000)) {
  vv <- floor(h*CPH/VOL$contacts_per_vote); cost <- vv*VOL$cost_per_vote
  left <- BUDGET - cost
  tot <- if (left > 0) vv + floor(left/CAN$cost_per_vote) else floor(BUDGET/VOL$cost_per_vote)
  cat(sprintf("%5d hrs -> %5d vol votes, $%7s left, TOTAL %5d %s\n",
      h, vv, format(max(0,left), big.mark=","), tot,
      ifelse(left <= 0, "  <- budget binds", "")))
}
```

| volunteer hours | total votes | binding constraint |
|---|---|---|
| `r nn(1000)` | `r nn(TOT(1000))` | `r BIND(1000)` |
| `r nn(2000)` | `r nn(TOT(2000))` | `r BIND(2000)` |
| `r nn(4000)` | `r nn(TOT(4000))` | `r BIND(4000)` |
| **~`r nn(CROSS)`** | **`r nn(TOT(CROSS))`** | **the crossover** |
| `r nn(12000)` | `r nn(TOT(12000))` | `r BIND(12000)` |

For reference: all-canvassing is `r nn(ALLC)` votes; unconstrained all-volunteer
is `r nn(BUY(BUDGET))`.

## What a good answer says

**3/4:** Finds roughly where the crossover is and states it.

**4/4:** Says what to do on each side of it. **Below ~`r nn(round(CROSS, -2))`
hours, an extra $50,000 is worth less than fifty more volunteers**, because the
cheap capacity is already exhausted and marginal money buys
`r dl(CAN$cost_per_vote)` canvassing. Above it, more
volunteers sit idle and money is the constraint. This is why field directors
are measured on recruitment: they are managing the binding constraint, not the
budget.

**4/4, best:** notices that at 2,000 hours the mixed strategy (`r nn(TOT(2000))`)
beats all-canvassing (`r nn(ALLC)`) but falls well short of the unconstrained
ideal (`r nn(BUY(BUDGET))`) —
so the *published* `r dl(VOL$cost_per_vote)`/vote figure is unreachable for almost every real
campaign. A number that is true in the aggregate and unattainable in practice
is worth being suspicious of, and students have now met that pattern twice
(the other is Session 4's vote target).

***

# Option C — Why buy the things that do not work

## What a good answer says

This one has no output. It is graded entirely on the quality of the reason.

**3/4:** Offers one plausible reason — usually name recognition or "reaching
people the other tactics miss."

**4/4:** Gives a reason that survives the objection, and names the evidence that
would test it. The strong candidates:

- **Turnout is not the objective; votes for *you* are.** Every number here is a
  turnout effect. Mail may not raise turnout while still moving *choice* among
  people already voting. The table simply does not measure persuasion.
- **Mail and email are fundraising instruments.** A mail program that loses on
  turnout and returns $1.40 per dollar in donations is not irrational.
- **Signalling to donors, endorsers and press** that the campaign is real.
- **List-building** for later cycles.

**4/4, best:** notices the asymmetry in the evidence itself. The negative
findings are about *turnout*, measured against the voter file, which is public.
Persuasion effects require survey data and are much harder and rarer to
measure. So the tactics that look useless may just be the ones whose payoff is
hardest to observe — and campaigns that keep buying them are not necessarily
being stupid, they may be responding to evidence the table cannot show.

**Do not accept** "campaigns are irrational" or "consultants profit from it" as
a complete answer. Both are partly true and neither is testable as stated. Push
for what evidence would settle it.

***

# Option D — Their own question

Good versions: cost per vote if you value a *net* vote (mobilising both sides
equally is worthless); what happens if door-knocking contact rates are half
what the experiments assumed.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating the averages as estimates with
  conditions attached rather than as prices.
- **Reward anybody who separates turnout from persuasion** without being
  prompted. It is the distinction the whole table rests on and it is not
  labelled.

## Anticipated questions

- *"Does 1 vote per `r CAN$contacts_per_vote` mean
  `r CAN$contacts_per_vote - 1` people didn't vote?"* — No, and this is the most
  common misreading. It means turnout in the contacted group was higher by
  about one vote per `r CAN$contacts_per_vote` people **than in an otherwise
  identical group that was not contacted**. Most of the
  `r CAN$contacts_per_vote` were going to vote anyway. Draw the
  treatment/control picture on the board; it takes two minutes and prevents
  half the bad write-ups.
- *"Why is canvassing more effective but more expensive?"* — Because a
  conversation at a door is more persuasive than one on a phone, and it costs
  far more per conversation (`r d2(CAN$cost_per_contact)` vs
  `r d2(VOL$cost_per_contact)`). Effectiveness is per *contact*; cost is per
  *vote*. Two denominators. Across all `r nrow(e)` effective tactics the two
  orderings are an exact reversal — the rank correlation between contacts per
  vote and cost per contact is `r pc(RHO, 0)`, which is the brief's Step 5.
- *"Are these numbers current?"* — Green and Gerber's 5th edition, 2024. Costs
  move with wages and postage; the effect sizes have been broadly stable across
  editions, which is itself notable.
- *"Do robocalls really work?"* — Barely: 1 per `r ROBO$contacts_per_vote`,
  which is `r round(ROBO$contacts_per_vote / CAN$contacts_per_vote)` times as
  many conversations as door-knocking needs for the same single vote. The row
  clears the cost comparison only because a robocall costs
  `r d2(ROBO$cost_per_contact)`, not because it is good. **Say the caveat out
  loud.** `effective` is a `TRUE`/`FALSE` flag standing in for a significance
  test, and the file carries no standard errors, so a positive finding and a
  practically negligible one are indistinguishable in this column. The brief's
  provenance section works this through and uses robocalls as the case in point.
  Nothing shipped in `data/` lets you re-verify that flag against Green and
  Gerber; treat it as the row most likely to move.
- *"Is this legal / is this what campaigns actually do?"* — Yes and yes. The
  experiments were mostly run with real campaigns and nonpartisan groups.

## Connections

- **Session 9 (polls, models, and markets)** — same underlying problem:
  an average estimated on one population applied to another. There it is
  non-response, here it is external validity.
- **Chapter 12's three factors** — ability, motivation, opportunity. Ask which
  one canvassing acts on. (Motivation, mostly. Registration law acts on
  opportunity. They are not substitutes.)
- **Session 4's vote-targeting lab** — targeting says *where*, this says *how*. Together
  they are a field plan.
- **Tuesday's debate** (political equality) — if mobilization mostly moves
  people who were nearly voting anyway, it may widen participation gaps rather
  than close them. That is the sharpest version of the question and worth
  saving for the last ten minutes.
