---
title: "Lab 10b: A Quarter of a Million Dollars and Nine Weeks"
subtitle: "84-355 Democracy's Data | What actually gets people to vote"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# The question

You are running a mobilization program. You have **$250,000** and nine weeks.
How do you spend it to produce the most additional votes?

This is a real question with a real answer, because unlike almost everything
else in this course, mobilization has been tested experimentally. Since the late
1990s, campaigns and researchers have run hundreds of randomised field
experiments: assign some voters to be contacted, some not, then check the public
voter file to see who actually turned out. Green and Gerber collect the results.

Chapter 12 gives you their summary as a table. Today you spend against it.

# Part 1 — What works

```{r load}
g <- read.csv("data/gotv_tactics.csv", stringsAsFactors = FALSE)
g
```

Three columns matter.

- `contacts_per_vote` — how many people you must actually *speak to* to produce
  one additional vote. Lower is better.
- `cost_per_vote` — what that additional vote costs, all in.
- `effective` — whether the experiments found any measurable effect at all.

**Start with the last column.** Three tactics have no measurable effect on
turnout: mail sent by a campaign, email, and leaflets left on doors. Not "a
small effect" — no effect that the experiments could distinguish from zero.

```{r nothing}
g[!g$effective, c("tactic", "effective")]
```

Think about how much of what arrives in your mailbox each October is on that
list.

# Part 2 — Effective and cheap are different things

```{r ranking}
e <- g[g$effective, ]
data.frame(tactic = e$tactic,
           rank_by_effect = rank(e$contacts_per_vote),
           rank_by_cost   = rank(e$cost_per_vote))
```

Door knocking is the most effective thing anyone has measured — one extra vote
for every 17 conversations — and it is *not* the cheapest way to buy a vote.
Volunteer phone banks are.

```{r plot-tradeoff}
plot(e$contacts_per_vote, e$cost_per_vote,
     pch = 19, cex = 1.4, col = "#C41230",
     xlab = "Contacts needed per additional vote  (lower = more effective)",
     ylab = "Cost per additional vote ($)",
     main = "", xlim = c(0, 460), ylim = c(0, 145))
text(e$contacts_per_vote, e$cost_per_vote, labels = e$tactic,
     pos = c(4, 4, 4, 2, 2), cex = 0.7)
grid(col = "grey85", lty = 3)
```

If effectiveness and cost told the same story, the points would sit on a rising
line. They do not.

# Part 3 — Spend the money

```{r budget}
BUDGET <- 250000                      # <- this is the number you will change

spend_all_on <- function(tactic_name) {
  row <- g[g$tactic == tactic_name, ]
  if (!row$effective) return(0)
  floor(BUDGET / row$cost_per_vote)
}

votes <- sapply(e$tactic, spend_all_on)
res <- data.frame(tactic = e$tactic, votes_if_all_in = votes)
res[order(-res$votes_if_all_in), ]
```

Spending the entire budget on the single best-value tactic buys about
**`r format(max(sapply(e$tactic, function(t) { r <- g[g$tactic==t,]; floor(250000/r$cost_per_vote) })), big.mark=",")`**
additional votes.

Hold that number against something real: in 2024 the presidential margin in
Wisconsin was about 29,000 votes, in Michigan about 80,000, and in Pennsylvania
about 120,000.

# Part 4 — The constraint that changes everything

The table prices volunteer phone calls at $46 a vote — but volunteers are not
money. You cannot buy them, and a campaign that has 40 volunteers has 40
volunteers no matter how large its budget.

```{r constrained}
VOLUNTEER_HOURS <- 2000     # total volunteer hours available all cycle
CALLS_PER_HOUR  <- 25       # completed conversations per hour

vol_contacts <- VOLUNTEER_HOURS * CALLS_PER_HOUR
vol_votes    <- floor(vol_contacts / g$contacts_per_vote[g$tactic == "Phone calls, volunteer"])
vol_cost     <- vol_votes * g$cost_per_vote[g$tactic == "Phone calls, volunteer"]

cat("Volunteer capacity:", format(vol_contacts, big.mark = ","), "conversations\n")
cat("  -> produces about", format(vol_votes, big.mark = ","), "votes\n")
cat("  -> costs about $", format(vol_cost, big.mark = ","), "\n", sep = "")
cat("  -> leaves $", format(BUDGET - vol_cost, big.mark = ","), " unspent\n", sep = "")

leftover  <- BUDGET - vol_cost
next_best <- g[g$tactic == "Door-to-door canvassing", ]
cat("\nSpending the remainder on canvassing adds",
    format(floor(leftover / next_best$cost_per_vote), big.mark = ","), "votes\n")
cat("TOTAL:", format(vol_votes + floor(leftover / next_best$cost_per_vote), big.mark = ","), "votes\n")
```

This is why campaigns are obsessed with volunteers, and why field organisers
are measured on recruitment rather than on calls.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different budget.** Change `BUDGET` in Part 3 to $25,000 (a
state legislative race) or $5,000,000 (a competitive Senate race). Does the best
strategy change, or only the scale? What does that tell you about whether money
buys elections?

**Option B — Volunteer supply.** Change `VOLUNTEER_HOURS` in Part 4. Find
roughly the point where volunteers stop being the binding constraint and money
starts. Above that point, what should a campaign do with an extra $50,000 —
and below it, what should it do with fifty more volunteers?

**Option C — Take the "no effect" tactics seriously.** Campaigns spend heavily
on mail and email anyway. Give one *good* reason a competent campaign might
buy something that does not raise turnout. (There are several. Turnout is not
the only thing a campaign is trying to do.) Then say what evidence would tell
you whether that reason is right.

**Option D — Your own question.** Something the table can answer.

# On the discussion board

By **Friday, 11:59 p.m.** — that evening or the next day, while it is fresh.
No length requirement — a sentence or two is enough, so long as it shows you
engaged with what we did. Any *one* of the usual four: something that surprised
you, something we missed, something you do not understand, or an answer to a
classmate. One post is full credit, and it can be about anything we looked at
this week — this lab is the obvious candidate, but it is not the only one.

If you want a starting point, three that would make good posts:

- Chapter 12 explains participation through ability, motivation and opportunity.
  Which one does door-knocking act on? Which does an easier registration law act
  on? They are not substitutes.
- If mobilization mostly moves people who were nearly voting anyway, does it
  narrow the participation gaps the chapter documents, or widen them?
- Every number here is an *average* from experiments run on particular voters in
  particular places. Name one reason the average might not hold for a campaign
  you would actually run.

***

### Where these numbers come from

Donald P. Green and Alan S. Gerber, *Get Out the Vote: How to Increase Voter
Turnout*, 5th ed. (Brookings Institution Press, 2024), 172–73. Sides et al.
reproduce them as Table 12.1.

Two cautions the summary table cannot carry. **"Contacted" is doing a lot of
work** — most doors do not open and most calls are not answered, so the cost
figures already price in the failures, but the effectiveness figures do not.
And these are **turnout** effects, not persuasion: they tell you how to make
someone vote, not how to make them vote for you. A campaign that mobilises the
wrong neighbourhood efficiently has helped its opponent.
