---
title: "Reading an FEC Candidate Filing"
type: brief
subtitle: "Where a campaign's money comes from, and where the record stops naming it"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
md_extensions: -tex_math_dollars
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("campaign-finance-code.R")
```

```{r setup, include=FALSE}
```

Everybody knows where campaign money comes from. It comes from special interests, from lobbyists, from political action committees with names nobody recognises. That is the sentence a citizen hears every election year, usually attached to a number with a lot of zeros in it.

The sentence is checkable, because every federal campaign has to write the answer down. **So where does a candidate's money actually come from?** Not in the aggregate, and not in a slogan, but on the form a treasurer signs.

The answer decides what a reform would have to change. If organised money is buying elections, the problem is who wins. If it is buying something else, the problem is what happens after somebody has already won, and the two need different remedies.

## The test

The evidence is one filing, read end to end, and then all `r n(nrow(fec))` of them at once. The file is the Federal Election Commission's "All Candidates" bulk download for the 2023–24 cycle: one row per candidate who filed, covering committee activity through 31 December 2024.

A mandatory filing is the right instrument for this question, and for a narrow reason. It is a complete count rather than a sample, so an incumbent's receipts and a challenger's can be set side by side with no sampling assumption in between. The four source categories mean the same thing in both rows, because the same form imposed them on both. What a disclosure filing is, who writes it, and why disclosure is not the same as audit is the subject of this cluster's [campaign-finance chapter](../finance-source/finance-source-brief.html).

One row is one candidate's committee for the whole cycle: everything raised, and a breakdown by source.

```{r one-row}
```

The `ici` column beside each name is the narrowest field in the file. It records whether the candidate was an **I**ncumbent, a **C**hallenger, or running for an **O**pen seat.

## The data

The Commission ships a pipe-separated file with no header row. Here is one complete record, laid out one field per row so it fits the page. Nothing else is changed.

```{r fec-raw}
```

That is one line of `r format(nraw, big.mark = ",")`, carrying `r length(f)` fields and no names for any of them. The values sit in an order the file never states. The Commission publishes that order in a separate document, on a different page of its website. Shift the layout by one and every dollar in this brief lands in the wrong category, with no error and no warning.

A few of the positions, once that separate document is consulted:

```{r fec-fields}
```

```{r fec-clean}
```

Cleaned, the same candidate is a row you can read, and four things happened to get there.

**Names were attached, and then checked.** The build writes down all `r length(f)` column names and stops unless exactly `r length(f)` columns arrived. That assertion is the whole defence against the layout changing in a future cycle.

**Two columns were invented.** Neither `office` nor `own money` is in the download. Office is the first character of the candidate ID, so `r pick` is a candidate for `r fec$office[fec$cand_id == pick]` because it begins with `r substr(pick, 1, 1)`. Own money is `cand_contrib`, what the candidate gave their campaign, added to `cand_loans`, what they lent it. That addition is a definition rather than a measurement, and a reader who thinks a loan is not a gift is entitled to a different column.

**Most of the file was thrown away.** Of `r length(f)` columns the build keeps `r ncol(fec)`: `cand_id`, `cand_name`, `office`, `office_st`, `ici`, `party`, `ttl_receipts`, `ttl_disb`, `indiv_contrib`, `pac_contrib`, `party_contrib`, `self_funding`, `cand_contrib`, `cand_loans`, `cash_on_hand`, `debts_owed_by` and `trans_from_auth`. The discarded ones include the coverage end date, which arrives as the text `r f[28]` rather than as a date.

**Nothing was repaired.** `r n(nrow(fec))` rows went in and `r n(nrow(fec))` came out. The build does not merge candidates who appear more than once, does not fill the `r n(sum(fec$ici == ""))` rows whose incumbency letter is blank, and does not remove the `r n(sum(fec$ttl_receipts == 0))` candidates who reported raising nothing at all. Each of those is a defect somebody would ordinarily clean, and each is evidence about how federal campaign finance is recorded.

Notice what the raw record does not contain. Fields `r paste(21:25, collapse = ", ")` are the election results, and here they are empty. They are empty in most records.

> **Before you read on, commit to a number.** Among the `r n(nrow(ser))` serious
> House campaigns in this file, meaning everyone who raised more than fifty
> thousand dollars, political action committees supply
> `r pc(pacm(ser,"I"))`% of the median incumbent's receipts. **What percentage
> do they supply the median open-seat candidate?** That is the race with no
> incumbent to unseat and a result genuinely in doubt. Write a number down,
> above or below `r pc(pacm(ser,"I"))`%, before you read on.

## What it says about democracy

Start with the whole House, where there are enough candidates for a pattern to exist.

```{r typical}
```

The median House candidate raised `r d(MEDH)`. The average raised `r d(MNH)`, which is `r pc(MNH/MEDH, 0)` times more. Whenever a mean sits that far above a median, a few cases are carrying nearly everything and the average has stopped describing anybody. So every figure below is a median or a share, never an average.

```{r mosaic-prep, include=FALSE}
```

```{r d3-mosaic, results='asis', eval=knitr::is_html_output()}
```

```{r mosaic-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

**Figure 1. Individuals pay for most of every kind of campaign; what differs is what covers the rest.** Serious House campaigns only. A mosaic is the right form because the question has two parts at once. Bar height is composition and bar width is total dollars raised, so the area of a coloured block is real money rather than a share of a share. The PAC block in the incumbent bar is `r dm(GT["I"] * MOS["PACs","I"] / 100)` against `r dm(GT["C"] * MOS["PACs","C"] / 100)` in the challenger bar. "Other" is receipts that are none of the four named sources, mostly transfers from the candidate's own other committees, which is why it is largest for incumbents.

So the slogan is wrong in its first clause. **Individuals supply the majority of every group's money**, `r pc(src["Individuals","I"])`% for incumbents, `r pc(src["Individuals","C"])`% for challengers and `r pc(src["Individuals","O"])`% for open-seat candidates. Organised money is the minority partner everywhere in this file.

Where it is not wrong is in the pattern. Look at what fills the gap when PACs do not.

```{r self}
```

About one in six serious candidates paid for more than half of their own campaign. `r pc(src["Self","C"])`% of all challenger money and `r pc(src["Self","O"])`% of all open-seat money comes from the candidate's own pocket, against `r pc(src["Self","I"])`% for incumbents. That is a barrier to entry with a dollar figure attached.

Now the gap everyone quotes.

```{r by-ici}
```

The median incumbent raised `r d(med(hs,"I"))` and the median challenger `r d(med(hs,"C"))`, a ratio of about **`r n(RATIO)` to one**. On its own that settles nothing. Money might make incumbents safe, or safe incumbents might attract money, and both stories predict exactly this table.

The file will break the tie without any research design. Ask not *how much* each group raises, but **what share of it comes from PACs**. An open-seat race is the most competitive kind there is: no incumbent, a result genuinely uncertain, maximum leverage for anyone trying to influence who wins. Money spent to decide an election should go there.

```{r pac-share}
```

```{r pac-prep, include=FALSE}
```

```{r d3-pac, results='asis', eval=knitr::is_html_output()}
```

```{r pac-static, eval=!knitr::is_html_output(), fig.height=3.6}
```

**Figure 2. PAC money is sorted by who holds the seat, not by how close the race is.** Median PAC share of receipts among candidates who raised more than $50,000. In the House it is `r pc(pacm(ser,"I"))`% for the `r n(sum(ser$ici == "I"))` incumbents, `r pc(pacm(ser,"O"))`% for the `r n(sum(ser$ici == "O"))` open-seat candidates and `r pc(pacm(ser,"C"))`% for the `r n(sum(ser$ici == "C"))` challengers. The Senate runs the same way at a fifth the size, `r pc(pacm(sen,"I"))`% against `r pc(pacm(sen,"C"))`%, because PAC limits are fixed per committee while Senate campaigns are enormous. In HTML the two chambers share one axis; in print they are drawn side by side.

Most readers write down a number near the incumbent's, and many write down a larger one, because that is where money would go if money were trying to decide elections. The open-seat figure is `r pc(pacm(ser,"O"))`%, essentially the same as a challenger's `r pc(pacm(ser,"C"))`%. **PAC money is not sorted by competitiveness. It is sorted by who currently holds the seat.**

That is enough to break the tie. Whatever organised money is buying, it is not the outcome of an election, because it is demonstrably not going where outcomes are decided. Fouirnaies and Hall (2014) reach the same conclusion causally, by comparing candidates who barely won against candidates who barely lost, and find the effect concentrated in access-oriented donors.

There is one more thing to say about the biggest block in Figure 1. **The largest source is the least identified one.** Federal law requires a committee to itemise a contributor by name, address, occupation and employer only once that person's giving passes $200. Below that line the money is reported as a lump sum. That threshold is the Commission's rule rather than a finding of this brief, and it is cited in the sources. So the `r pc(src["Individuals","I"])`% marked "Individuals" in the incumbent bar is a sum over two populations, and the public record names only one of them.

## What this brief cannot tell you

**It cannot tell you who won.** The FEC's layout has a general-election-result column and it is blank in every row. The Commission regulates finance, not elections, and no national election administrator exists to supply the answer.

```{r no-outcome}
```

So the most obvious question in the subject, whether the candidate who raises more wins, cannot be asked of this file without going to fifty-one separate state authorities. That is a fact about how the American state keeps its books rather than a gap in this dataset.

**It cannot be summed.** Adding the receipts column gives $`r pc(NAIVE/1e9, 2)` billion for the cycle, of which roughly $`r pc((NAIVE-DEDUP)/1e9, 2)` billion does not exist. It is the same money counted twice, mostly Biden's presidential committee, which became Harris's in July 2024 and is reported in full against both candidate IDs.

**It cannot name small donors.** The itemisation threshold above means the individual-contribution column mixes named givers with an anonymous remainder, and this summary file reports only the sum.

**It covers candidate committees only.** Independent expenditures, party spending and undisclosed money are somewhere else or nowhere. Since *Citizens United v. FEC* (2010) a large share of what is spent on a competitive race never passes through the candidate's hands.

**It is one cycle**, so there is no trend in it, and the $50,000 threshold used for every share above is a choice this brief made rather than found.

## What you should have learned

- Individuals, not organised committees, supply most of every kind of campaign's money: `r pc(src["Individuals","I"])`% for incumbents, `r pc(src["Individuals","C"])`% for challengers, `r pc(src["Individuals","O"])`% for open-seat candidates.
- PAC money is allocated by office held, not by contest. It is `r pc(pacm(ser,"I"))`% of a typical serious House incumbent's receipts and `r pc(pacm(ser,"O"))`% of an open-seat candidate's, and the open seat is the most competitive race there is.
- Candidates without organised money pay for themselves: `r pc(src["Self","C"])`% of all challenger money is the candidate's own, against `r pc(src["Self","I"])`% for incumbents, and `r n(SELF)` of the `r n(nrow(bigc))` candidates raising over $100,000 covered more than half their own campaign.
- The file is authoritative and incomplete at once. It records every federal candidate committee and no election result, and summing its receipts column overstates the cycle by about $`r pc((NAIVE-DEDUP)/1e9, 2)` billion.
- The biggest category is the least itemised: individual giving under $200 in a cycle is reported as a total with no donor named.

## Extensions

None of these is answered above, and every one can be answered by sorting or grouping in a spreadsheet.

- `fec_candidates_2024.csv` has `ttl_receipts` for every candidate. Compute the mean and the median, and note how far apart they are. Which one belongs in a headline about the cost of a seat?
- Sort `fec_candidates_2024.csv` by `self_funding` and take the top fifty. Compare their `ttl_receipts` against the median for their `office`. Does buying your own campaign buy a big one?
- Group `fec_candidates_2024.csv` by `ici` and compute `pac_contrib` divided by `ttl_receipts`. Then do the same for `indiv_contrib`. Which source separates the three groups more sharply?
- Group by `office` and `party` and compare typical `ttl_receipts`. Where in this file is money least evenly spread?

Two stretch ideas go past the spreadsheet. Download the previous cycle's "All Candidates" file from the same address and check whether the PAC pattern in Figure 2 holds in a year with a different result. And take one candidate's `cand_id`, look up their itemised receipts in the FEC's individual-contributions file, and see how much of their individual total the named donors account for.

## Sources

**Data**

- Federal Election Commission, bulk download, "All Candidates" file for the
  2023–24 cycle, covering every candidate committee's activity through 31
  December 2024. <https://www.fec.gov/data/browse-data/?tab=bulk-data>

**The rule behind the form**

- Itemisation threshold: a committee reports a contributor's name, address,
  occupation and employer only when that person's contributions exceed $200 in
  the calendar year or election cycle. 11 CFR 104.3(a)(4)(i).
  <https://www.ecfr.gov/current/title-11/chapter-I/subchapter-A/part-104/section-104.3>

**Academic literature**

- Ansolabehere, S., de Figueiredo, J. M. and Snyder, J. M. Jr. (2003). "Why Is
  There So Little Money in U.S. Politics?" *Journal of Economic Perspectives*
  17(1): 105–130.
- Fouirnaies, A. and Hall, A. B. (2014). "The Financial Incumbency Advantage:
  Causes and Consequences." *Journal of Politics* 76(3): 711–724.
- *Citizens United v. Federal Election Commission*, 558 U.S. 310 (2010).

**The data itself**

Every figure above is drawn from this table. It is plain CSV, it opens in a
spreadsheet, and it is yours to take further.

[`fec_candidates_2024.csv`](data/derived/fec_candidates_2024.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
