---
title: "Data Session 9, part 2: What Is a 70% Chance?"
subtitle: "84-355 Democracy's Data | Forecasting models and prediction markets"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 5)
options(scipen = 999)

# Every figure quoted in the prose below is read back out of the data, so no
# sentence can drift away from what the chunks beside it print. The two objects
# read here are re-created in view in the chunks that first need them.
d <- read.csv("data/polymarket_resolved.csv", stringsAsFactors = FALSE)
b <- cut(d$p7, breaks = seq(0, 1, 0.1), include.lowest = TRUE)

pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
TB <- function(lv, col) tab[[col]][tab$band == lv]
PV <- function(lv) {
  k <- b == lv
  binom.test(sum(d$outcome[k]), sum(k), mean(d$p7[k]))$p.value
}
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"Candidate X has a 72% chance of winning."*

This morning you took a poll apart. A poll reports a **share** — 48% of
respondents said this. Nobody votes on a share. Between the poll and the
sentence above sits a **model**, and next to the model sits a **market**, and
the two produce the same kind of object: a probability.

Two weeks from today the country votes, and every one of these numbers will be
graded. So: **what would it even mean for "72%" to be right?**

If the candidate loses, the forecast was not refuted. 72% is not 100%. Things
priced at 72% are supposed to fail 28% of the time.

**A single probability cannot be checked.** Many of them can — and that is what
we do today.

# Part 1 — The problem with checking one

```{r one}
p <- 0.72
c(if_it_happens = 1 - (1 - p)^2,
  if_it_doesnt  = 1 - p^2)
```

Whatever happens, a defender of the forecast has a reply. **This is not a
weakness of forecasters — it is a property of probability**, and it is why
"the polls were wrong in 2016" is a much emptier sentence than it sounds.

The only escape is volume. Collect hundreds of forecasts, group them by what
they claimed, and check the groups.

# Part 2 — `r nrow(d)` forecasts with known answers

A prediction market is a forecast with money behind it: people buy contracts
that pay $1 if a thing happens, so the price is what the crowd thinks the
chance is. When it resolves, we find out.

```{r load}
d <- read.csv("data/polymarket_resolved.csv", stringsAsFactors = FALSE)

c(markets = nrow(d),
  resolved_yes = sum(d$outcome),
  priced_under_10pct = sum(d$p7 < 0.10))
```

**`r nrow(d)` resolved markets**, each with the price **seven days before it
closed** and what actually happened.

Note the third number.
**`r pc(100 * sum(d$p7 < 0.10) / nrow(d))`% of these markets were priced under
10%** — most questions people trade are about things that do not happen. Keep
that in mind when you read the table below.

```{r sample}
head(d[order(-d$volume), c("question", "p7", "outcome")], 8)
```

The 2024 presidential market is the first row: seven days out it gave Trump
**`r d$p7[which.max(d$volume)]`**, and the outcome was
`r d$outcome[which.max(d$volume)]`.

# Part 3 — Calibration

Group the markets by the price they carried, then ask what share actually
happened. **If the prices mean anything, the two columns should match.**

```{r calibration}
b <- cut(d$p7, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
tab <- data.frame(
  band     = levels(b),
  n        = as.vector(table(b)),
  priced   = round(as.vector(tapply(d$p7,      b, mean)), 3),
  happened = round(as.vector(tapply(d$outcome, b, mean)), 3))
tab$gap <- round(tab$happened - tab$priced, 3)
tab
```

```{r calib-plot}
plot(tab$priced, tab$happened, pch = 19, cex = sqrt(tab$n) / 4,
     xlim = c(0, 1), ylim = c(0, 1), col = "#2c7fb8",
     xlab = "price seven days out", ylab = "share that actually happened")
abline(0, 1, lty = 2)
text(tab$priced, tab$happened, tab$n, pos = 3, cex = 0.7, col = "grey30")
```

**Read the top of the table first.** Markets priced 0.8–0.9 happened
**`r pc(100 * TB("(0.8,0.9]", "happened"))`%** of the time. Priced 0.9–1.0,
**`r pc(100 * TB("(0.9,1]", "happened"))`%**. Priced 0.7–0.8,
**`r pc(100 * TB("(0.7,0.8]", "happened"))`%**. Those are close enough to the
line that the prices are doing real work.

```{r brier}
c(brier_market   = round(mean((d$p7 - d$outcome)^2), 4),
  brier_always50 = round(mean((0.5 - d$outcome)^2), 4),
  brier_baserate = round(mean((mean(d$outcome) - d$outcome)^2), 4))
```

The **Brier score** is the average squared error of a probability — lower is
better. The market scores **`r pc(mean((d$p7 - d$outcome)^2), 4)`**, against
**`r pc(mean((0.5 - d$outcome)^2), 2)`** for saying "50%" every time and
**`r pc(mean((mean(d$outcome) - d$outcome)^2), 4)`** for always guessing the
overall base rate. **The market is carrying a great deal of information.**

# Part 4 — Where it looks wrong, and whether to believe that

Now the bottom of the table. Markets priced 0.1–0.2 happened
**`r pc(100 * TB("(0.1,0.2]", "happened"))`%** of the time — roughly twice what
they claimed. That looks like a real bias.

**Before believing it, count how many chances we gave ourselves to find one.**

```{r sig}
for (lv in levels(b)) {
  k <- b == lv; n <- sum(k)
  if (n < 5) next
  p <- binom.test(sum(d$outcome[k]), n, mean(d$p7[k]))$p.value
  cat(sprintf("  %-10s n=%3d  expected %5.1f  observed %3d   p=%.3f%s\n",
              lv, n, sum(d$p7[k]), sum(d$outcome[k]), p,
              ifelse(p < 0.05, "  *", "")))
}
```

We tested **`r sum(table(b) >= 5)` bands**. At a 5% threshold, you expect roughly
one false positive from pure chance even if the market is perfectly calibrated.

**Exactly one band clears it: the longshots, [0, 0.1] — and it is not a small
band. It holds `r TB("[0,0.1]", "n")` of the `r nrow(d)` markets**
(p = `r pc(PV("[0,0.1]"), 3)`). The next band up, (0.1, 0.2], has
`r TB("(0.1,0.2]", "n")` markets and p = `r pc(PV("(0.1,0.2]"), 3)`; the one
after that has `r TB("(0.2,0.3]", "n")` and p = `r pc(PV("(0.2,0.3]"), 3)`, which
is no evidence of anything.

So the honest reading is **not** "the market misprices longshots across the
board." It is that the one place a real deviation shows up is the place where
almost all the contracts are, and even there the effect is small in absolute
terms: priced `r TB("[0,0.1]", "priced")`, happened
`r TB("[0,0.1]", "happened")`.

```{r combined}
k <- d$p7 > 0.1 & d$p7 <= 0.3
binom.test(sum(d$outcome[k]), sum(k), mean(d$p7[k]))$p.value
```

Pooling the 0.1–0.3 range gives
**p = `r pc(binom.test(sum(d$outcome[k]), sum(k), mean(d$p7[k]))$p.value, 3)`** —
significant by convention, and not by much, on `r sum(k)` markets.

**The honest summary is:** the market is well calibrated where most of the money
is, and there is a hint that low-probability events happen more often than
priced, which this sample is too small to establish. A forecaster who claimed
either "markets are perfectly calibrated" or "markets systematically underprice
longshots" would be going beyond this evidence.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different moment.** The table uses the price seven days out. Redo
it with `p30` — thirty days out — and drop rows where `p30` is missing. Is the
market worse further from resolution? It should be. Is it?

**Option B — Only the big markets.** Restrict to markets with volume above the
median and redo the calibration. More money should mean better prices. Does it?

**Option C — Score yourself.** The Brier score works on any forecaster. Pick
five of the questions in `d` you could plausibly have answered a week out, write
down your own probabilities *before* looking at the outcome column, and compute
your Brier score against the market's. Report honestly.

**Option D — Your own question.** `r nrow(d)` markets, two snapshots, one outcome
column.

# What this lab cannot tell you

- **Whether models beat markets.** We only have the market. **Forecasting model
  outputs are not published as downloadable data** — no file exists to check
  538, Silver Bulletin or Split Ticket against this the way we just checked
  Polymarket. That absence is the same shape as the ad libraries in Session 5:
  the claim is public, the evidence behind it is not.
- **Anything specific to politics.** These are the highest-volume resolved
  markets, and they include sport, crypto and Fed decisions. What is being
  tested is the **market mechanism**, not American elections.
- **Whether the price was right *earlier*.** Seven days out is one arbitrary
  moment. A market that was badly wrong for a year and correct at the end scores
  well here.
- **Whether anyone can beat it.** Calibration says the prices are honest, not
  that they are unbeatable — and a market that is well calibrated overall can
  still be badly wrong about any single question, including the one you care
  about.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Things priced at 90% happened `r pc(100 * TB("(0.9,1]", "happened"))`% of the
  time and things priced at 15% happened
  `r pc(100 * TB("(0.1,0.2]", "happened"))`% of the time. Which of those two
  numbers should worry you more?
- If a single forecast cannot be wrong, what is a forecast *for*?
- The country votes in two weeks. What number would you want to see published
  alongside every forecast, and why is nobody publishing it?

***

### Where the data came from

**Polymarket**, via two public APIs that need no key: the Gamma markets API for
resolved markets and outcomes, and the CLOB price-history API for the price
before resolution. `r nrow(d)` markets survived the filters, out of 900 pulled.
The build is in `data/build-data.py`.

Markets are included only if they settled cleanly to Yes or No — unresolved,
void and scalar markets cannot be scored and were dropped.
