---
title: "Testing a Famous Political Proxy: Whole Foods vs. Cracker Barrel"
type: brief
subtitle: "The metric replicates to within a point, and then does not survive being asked about people"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("whole-foods-cracker-barrel-code.R")
```

```{r setup, include=FALSE}
```

In September 2011 David Wasserman, the House editor of the *Cook Political Report*, stood in the Cracker Barrel in Manassas, Virginia, then drove twenty-six miles east to the Whole Foods in Clarendon. The column he wrote for the *Washington Post* proposed that the coming election would be fought between those two parking lots.^[David Wasserman, "Will the 2012 election be a contest of Whole Foods vs. Cracker Barrel shoppers?," *Washington Post*, 28 September 2011; the address is in the sources below.] It carried a number quoted ever since:

> In 2008, candidate Barack Obama carried 81 percent of counties with a Whole
> Foods and just 36 percent of counties with a Cracker Barrel — a record
> 45-point gap.

Fifteen years later the shorthand is a fixture of election-night commentary. It says you can read a county's politics off its grocery stores. Is that true? The question is worth your time because a **proxy**, a stand-in like this one for something harder to see directly, is how many people picture the political divide, and one that is wrong in a particular direction moves the picture with it.

## The test

Answering it takes three files joined together. Store locations for both chains come from OpenStreetMap, the world map built by volunteers, where mappers have placed every Whole Foods and Cracker Barrel they know of; the copy here is as the map stood on `r format(as.Date(FV("osm_base")), "%d %B %Y")`. County presidential returns for 2008 through 2024 are the researcher Tony McGovern's compilation from news organizations' results pages, downloadable at <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>. The Census Bureau supplies county boundaries, populations and land area, and the Department of Agriculture a count of college-educated adults by county; addresses for all of them are in the sources.

The returns and census tables are the easy half. **The store list is the problem: neither company publishes where its stores are.** Cracker Barrel is a public company, so its annual report to the Securities and Exchange Commission, the Form 10-K, carries a count: `r FV("cb_10k_stores")` stores in `r FV("cb_10k_states")` states as of `r FV("cb_10k_asof")`. A count is not a list, and carries no addresses. Whole Foods has been an Amazon subsidiary since 2017 and files nothing about itself, so the number of its markets in the country is not a published figure at all.

So the inventory here is OpenStreetMap's: every United States object its volunteers have tagged with the chain's Wikidata identifier, a code that names the brand however the store's name was typed. That gives `r FV("n_wf_osm")` Whole Foods Markets and `r FV("n_cb_osm")` Cracker Barrels. The independent half of a famous political metric is a crowd-sourced guess at a private fact.

The join turns three files into one table. Each store is placed inside a county's boundary on the map, which hands it the county's five-digit federal code, the FIPS code that identifies every county in the country, and that code is the column the returns and census tables also carry. [Join Keys and Crosswalks](../data-sources/data-sources-brief.html) states the rule this section keeps returning to, and explains the crosswalk, a table that translates one set of areas into another. A number built from two sources inherits the assumptions of both, and the join adds one about how they line up. Here that assumption is that a county code means the same ground in the store file and in every year's returns. It does not.

## The data

The main table has one row per county. `counties.csv` carries `has_wf` and `has_cb`, the four-way `cat` combining them, then `pop`, `land_sqmi`, `density`, `ba_pct`, and `win`, `dem`, `gop`, `total` per election. `gap.csv` reduces that to one row per election, with `wf_counties`, `cb_counties`, `wf_votes`, `cb_votes`, `gap_counties` and `gap_votes`. `categories.csv` reduces it the other way, one row per store category, with `county_share`, `vote_share`, `dem_counties_pct`, `dem_votes_pct`, `med_pop`, `med_density` and `med_ba`.

Stores are matched on the brand's Wikidata identifier rather than on the name. Names are typed by hand by thousands of people, as *Whole Foods*, *Whole Foods Market* or *WholeFoods*, and a name match drops whichever spellings nobody anticipated.

**Only one brand can be audited, which is itself a finding.** Against Cracker Barrel's own filing the volunteer map holds up: `r FV("n_cb_osm")` of `r FV("cb_10k_stores")` stores, `r FV("cb_coverage")`%. The state count does not. The map places Cracker Barrels in `r FV("cb_osm_states")` states and the company says `r FV("cb_10k_states")`, and the difference is one restaurant: Maine's only Cracker Barrel closed in January 2025 and is still mapped.^[The South Portland restaurant closed on 21 January 2025: News Center Maine, "Cracker Barrel closes its only Maine location in South Portland," <https://www.newscentermaine.com/article/money/business/cracker-barrel-closes-its-only-maine-location-january/97-42f0ab30-2d97-4f84-9c66-8914769cc8d3>.] Nothing equivalent can be done for Whole Foods, because there is no filing to check.

Four cleaning decisions follow, and each changed a published number.

**Connecticut is two different maps.** In 2022 the Census Bureau, at the state's request, replaced Connecticut's eight counties with its nine planning regions as the units that stand in for counties;^[Census Bureau, "Change to County-Equivalents in the State of Connecticut," *Federal Register*, 6 June 2022, <https://www.federalregister.gov/documents/2022/06/06/2022-12063/change-to-county-equivalents-in-the-state-of-connecticut>.] the 2024 returns use the new shapes and the four earlier elections the old ones. Uncorrected, this file showed no Whole Foods county in Connecticut until 2024, and there were four. Stores are now assigned twice, once per set of boundaries.

**Washington is eight rows in 2024 and one before it.** The compilation reports the District by ward, and no boundary file contains wards, so its `r FV("dc_ward_rows")` Whole Foods are summed back into one row. **Shannon County, South Dakota, does not exist**: renamed Oglala Lakota County in 2015 and given a new code,^[The change was published in the *Federal Register* on 17 July 2015: <https://www.federalregister.gov/documents/2015/07/17/2015-17501/change-to-the-name-of-shannon-county-south-dakota>.] it leaves a careless table carrying two rows for one place. **And Alaska has no counties at all**, reporting votes by State House district, so every share here is of fifty jurisdictions rather than fifty-one. Each of these fails quietly, because a missing match produces a zero and a zero is a valid store count.

> **Before you read on, commit to two numbers.** In 2024 the Democratic
> candidate carried some share of counties containing a Whole Foods and some
> share of counties containing a Cracker Barrel. **How many percentage points
> apart are those two shares?** Now ask the same question of *people*: take
> every vote cast in Whole Foods counties and every vote cast in Cracker
> Barrel counties, and take the Democratic share of each. **How many points
> apart are those?** Write both down.

## What it says about democracy

Start with whether the metric can be reproduced. This store list was captured in `r format(as.Date(FV("osm_base")), "%B %Y")`; Wasserman's 2020 figures, posted a month after that election, came from one five and a half years older.^[Dave Wasserman (@Redistrict), 8 December 2020, giving the four-category breakdown: <https://x.com/Redistrict/status/1336355310789922820>.] Recomputing his four categories:

```{r replicate}
```

**Every category lands within about a point.** The same holds further back: for 2008 the column reports a `r pc(anach$gap_published, 0)`-point gap and this file recomputes `r pc(anach$gap_counties, 1)` points, a drift of `r FV("drift_gap_2008")` across fifteen years of openings and closings. **So nothing that follows can be blamed on the data.**

```{r headline}
```

Measured the way the metric is always measured, one county and one tally, the divide is enormous and has grown: `r FV("gap08")` points in 2008 and `r FV("gap24")` in 2024. Measured in people, the same counties are `r FV("gap24v")` points apart, and that number has barely moved in sixteen years.

Here is how one county enters both rows of that table. Knox County, Tennessee, which holds Knoxville, has `r KNOX$n_wf` Whole Foods and `r KNOX$n_cb` Cracker Barrels on the map, so it counts in both columns. In 2024 its voters cast `r n(KNOX$dem2024 + KNOX$gop2024)` ballots for the two major candidates and Trump carried it, the Democrat taking `r pc(KNOX_D)`% of that two-party vote. In the first row Knox is one tally mark: a Whole Foods county the Democrat lost, and a Cracker Barrel county the Democrat lost, one mark on each side. In the second row Knox is `r n(KNOX$dem2024 + KNOX$gop2024)` ballots at `r pc(KNOX_D)`% Democratic, poured into both sides' totals, where it weighs as much as `r round(KNOX_X)` counties with neither store of typical size. Do that for all `r n(FV("n_counties"))` counties and you have both lines of Figure 1.

```{r gapfig-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r gapfig-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The same metric on two units.** Both readings are drawn on one axis, because the claim under test is a distance, and a distance split across two panels is one the reader will not measure. The upper line is the published version, the share of *counties* carried. The lower line is the same counties, elections and stores, with the Democratic share of the two-party vote in place of the count of counties.

Nothing was done to the data between the two lines. The disagreement comes from a fact about American counties the metric's arithmetic assumes away.

```{r cats}
```

The first reason is a category. **The counties with neither store are `r FV("neither_county_pct")`% of all counties and `r pc(VS("neither"), 0)`% of all votes.** They are the largest block of the electorate and the most Republican group in the table, and the metric has no column for them. Whatever the gap measures, it measures on the `r pc(COVERED, 0)`% of counties where one of two companies opened.

The second reason is the definition. Wasserman's own rule is that a county with both stores counts in both columns, which is nearly harmless for `r FV("n_both")` counties out of `r n(FV("n_counties"))` when the unit is counties. It is not harmless when the unit is votes: those counties are the big metros, they cast `r pc(VS("both"), 0)`% of the national vote, and they appear at full weight on *both* sides of the subtraction. Take the overlap out and the exclusive categories are `r FV("excl_wf_votes")`% to `r FV("excl_cb_votes")`%, a `r FV("excl_gap_votes")`-point gap. That is a real and large difference, describing `r pc(EXCL_SHARE, 0)`% of the electorate rather than all of it.

```{r catfig-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

```{r catfig-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Counties are not people.** Each category carries two shares, one of counties and one of votes, and the bar between them is the distance the metric ignores. Cracker Barrel counties are `r pc(CS("cracker barrel only"), 0)`% of counties and `r pc(VS("cracker barrel only"), 0)`% of votes; Whole Foods counties are the reverse.

The folk reading of the metric is that Whole Foods stands for the urban professional and Cracker Barrel for rural America. The first half is right and the second is wrong. **A median Cracker Barrel county holds `r FV("med_pop_cb")` people; a median county with neither store holds `r FV("med_pop_neither")`.** Cracker Barrels sit on interstate exits outside regional towns, where the traffic is, so Cracker Barrel country is the middle of American county size rather than the bottom. Rural America is the `r n(FV("n_neither"))` counties with neither store, and it is the more Republican: the Democratic candidate carried `r pc(DC("neither"), 1)`% of them against `r pc(DC("cracker barrel only"), 1)`% of Cracker Barrel counties.

One question is left. The metric feels like a reading of culture, but a cheaper explanation is available: Whole Foods opens where there are dense, affluent, highly educated customers, and those are the people who vote Democratic. Take the `r FV("n_wf_counties")` counties selected by "has a Whole Foods", then the same number selected by rules using no store data, and ask each the identical question.

```{r alts}
```

**On counties the store list wins by about two points. On votes it loses.** Density and education each pick a set of counties whose electorate is *more* Democratic than the Whole Foods set. The last column shows why the contest is close: `r FV("overlap_dens_pct")`% of the densest `r FV("n_wf_counties")` counties already contain a Whole Foods. The store list is substantially a density list with a brand name on it.

Cracker Barrel's spokeswoman told Wasserman in 2011 that "politics don't play any role in our site selection process." She was right, and that is the problem. The stores are sited on traffic counts, rooftops and household income, which is why they line up with the vote and why they add so little once you have the rooftops.

## What this brief cannot tell you

**It cannot tell you that the store list is complete.** It is a volunteer file audited against one company's filing and against nothing for the other. `r FV("cb_coverage")`% coverage is reassuring, but no guarantee that the missing `r MISSING_CB` are scattered at random. If volunteers map less thoroughly where fewer volunteers live, the gaps sit where the metric is most sensitive.

**The matched-count test is a comparison, not a regression**, meaning not a best line drawn through a scatter of points. That density picks a similar set of counties does not prove the store adds nothing, only that whatever it adds is small.

**A store is not a shopper.** Every claim here is about counties containing a store, never about people who shop in one. Wasserman's headline was about shoppers; his data, like this data, was about county boundaries. This book makes that argument at length in [levels-of-aggregation](../../03-elections/levels-of-aggregation/levels-of-aggregation-brief.html).

**And it says nothing about 2028.** Every figure here is retrospective, and the [bellwether](../../03-elections/bellwether/bellwether-brief.html) brief is the warning about pointing a county-level regularity at the future.

## What you should have learned

- The proxy replicates. Rebuilt from a volunteer store list captured fifteen years later, Wasserman's four 2020 categories land within about a point, and the 2008 gap drifts by `r FV("drift_gap_2008")` points.
- The same metric gives `r FV("gap24")` points measured in counties carried and `r FV("gap24v")` measured in votes cast, from one file in one year.
- Counties are not people. The `r n(FV("n_neither"))` counties with neither store are `r FV("neither_county_pct")`% of counties and `r pc(VS("neither"), 0)`% of votes, and the metric has no column for them.
- Cracker Barrel country is not rural: its median county holds `r FV("med_pop_cb")` people against `r FV("med_pop_neither")` where there is no store at all.
- A proxy is worth only what it adds to the ordinary numbers it stands for, and `r FV("overlap_dens_pct")`% of the densest `r FV("n_wf_counties")` counties already contain a Whole Foods.

## Extensions

Each can be answered by sorting or grouping in a spreadsheet, and none is answered above.

- `gap.csv` has `gap_counties` and `gap_votes` for each election. Follow both down the years. Which one is the metric usually quoted as, and which describes the electorate?
- `categories.csv` gives `med_density`, `med_pop` and `med_ba` per category. Sort by `med_pop`. Is Cracker Barrel country actually rural, or only not urban?
- `alternatives.csv` swaps in other rules for picking counties, with `overlap_wf_pct`. Find one that does about as well as the store list. What was the metric really measuring?
- `inventory_check.csv` sets a company filing's `stores` beside `stores_osm` from the open map. Work out the discrepancy, and decide which source you would cite.

Two go past the spreadsheet. Pick another chain with a political reputation and see whether it publishes addresses at all. Or rebuild the four categories in `counties.csv` on the 2020 returns, using `has_wf_2020` and `has_cb_2020`, and say which answer you would print.

## Sources

**Data**

- OpenStreetMap contributors. Store locations for Whole Foods Market
  (`brand:wikidata=Q1809448`) and Cracker Barrel (`brand:wikidata=Q4492609`),
  every United States object carrying the tag, as the map stood on
  `r format(as.Date(FV("osm_base")), "%d %B %Y")`, under the Open Database
  Licence. Downloaded through the Overpass API, the public search service for
  the map's data; the downloaded copies are kept in this chapter's data
  folder, because the map changes daily and the same search does not return
  the same answer twice. <https://www.openstreetmap.org/>,
  <https://overpass-api.de/api/interpreter>, and the tag's documentation at
  <https://wiki.openstreetmap.org/wiki/Key:brand:wikidata>
- Cracker Barrel Old Country Store, Inc. *Annual Report on Form 10-K for the
  fiscal year ended 1 August 2025.* Item 1, store and state counts as of
  `r FV("cb_10k_asof")`.
  <https://investor.crackerbarrel.com/static-files/881c3ef3-0a78-4e60-9fc8-6d2a340bdbc8>
- McGovern, Tony. *US County Level Presidential Results 2008–2024*, used for
  all five elections.
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>
- U.S. Census Bureau. *2024 and 2020 Cartographic Boundary Files*, counties,
  for the two geographies the elections use; *2024 Gazetteer Files* for land
  area; *Vintage 2024 Population Estimates*.
  <https://www2.census.gov/geo/tiger/GENZ2024/shp/cb_2024_us_county_500k.zip>,
  <https://www2.census.gov/geo/tiger/GENZ2020/shp/cb_2020_us_county_500k.zip>,
  <https://www2.census.gov/geo/docs/maps-data/data/gazetteer/2024_Gazetteer/2024_Gaz_counties_national.zip>
  and
  <https://www2.census.gov/programs-surveys/popest/datasets/2020-2024/counties/totals/co-est2024-alldata.csv>
- U.S. Department of Agriculture, Economic Research Service. *County-level Data
  Sets: Education*, adults 25 and over with a degree, 2018–22.
  <https://ers.usda.gov/sites/default/files/_laserfiche/DataFiles/48747/Education.csv>
- Wasserman, David. "Will the 2012 election be a contest of Whole Foods vs.
  Cracker Barrel shoppers?" *Washington Post*, 28 September 2011: the 2008
  figures and the spokeswoman's remark. His four-category breakdown of 2020,
  posted 8 December 2020, is the published column in the replication table.
  <https://www.washingtonpost.com/opinions/will-the-2012-election-be-a-contest-of-whole-foods-vs-cracker-barrel-shoppers/2011/09/27/gIQAJdKA5K_story.html>
  and <https://x.com/Redistrict/status/1336355310789922820>
- Hoey, Dennis. "Maine's only Cracker Barrel closes in South Portland."
  *Portland Press Herald*, 21 January 2025, on the closure the map still
  carries.
  <https://www.pressherald.com/2025/01/21/maines-only-cracker-barrel-closes-in-south-portland/>
  (the address as first cited; the paper has since moved the story to
  <https://www.pressherald.com/2025/01/21/cracker-barrel-closes-in-south-portland/>).
  Also reported by News Center Maine,
  <https://www.newscentermaine.com/article/money/business/cracker-barrel-closes-its-only-maine-location-january/97-42f0ab30-2d97-4f84-9c66-8914769cc8d3>
- Census Bureau. "Change to County-Equivalents in the State of Connecticut,"
  *Federal Register*, 6 June 2022, and "Change to the Name of Shannon County,
  South Dakota," *Federal Register*, 17 July 2015: the two boundary changes
  the cleaning corrects for.
  <https://www.federalregister.gov/documents/2022/06/06/2022-12063/change-to-county-equivalents-in-the-state-of-connecticut>
  and
  <https://www.federalregister.gov/documents/2015/07/17/2015-17501/change-to-the-name-of-shannon-county-south-dakota>

**The data itself**

The tables this brief rests on are plain CSV and open in a spreadsheet.

[`alternatives.csv`](data/derived/alternatives.csv), [`anachronism.csv`](data/derived/anachronism.csv), [`categories.csv`](data/derived/categories.csv), [`cb_by_state.csv`](data/derived/cb_by_state.csv), [`counties.csv`](data/derived/counties.csv), [`gap.csv`](data/derived/gap.csv), [`inventory_check.csv`](data/derived/inventory_check.csv), [`missing.csv`](data/derived/missing.csv), [`published_2020.csv`](data/derived/published_2020.csv), [`stores.csv`](data/derived/stores.csv)

One more file is not data about the subject -- it is how this chapter checked its own numbers: [`facts.csv`](data/derived/facts.csv), the individual numbers the text above quotes.

```{r ai-prompt, echo=FALSE, results="asis"}
```
