---
title: "Thirty-Five Races, Rated in August"
subtitle: "Commit to a Senate map now, and to a number you cannot revise in November"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("senate-2026-code.R")
```

```{r setup, include=FALSE}
```

## The question you have to answer before you are allowed to be right

Write down how many Senate seats the Democrats will hold after the November 2026 election. Do it now, before you read any further, and do it in ink.

That instruction is the whole chapter. Everything below exists to make the number you write down more informed than a coin flip, and nothing below will tell you whether it is correct. The answer arrives on 3 November, and by then your guess has to already exist or it does not count.

The reason to insist on this is not suspense. A prediction made after the fact is not a prediction, and the human capacity to remember having believed the thing that happened is close to unlimited. The only defence is a record with a date on it. This chapter builds you one.

> **Before you go on.** Write two numbers on paper, or in a file you will not
> edit. First: how many of the 100 Senate seats the Democrats will hold in
> January 2027. Second: how confident you are, as a percentage. Keep them. The
> second number is the one that will surprise you.

## The arithmetic that decides what your guesses are worth

`r fx("n_races")` Senate seats are on the ballot in November 2026, and `r mq("Democratic seats not on the ballot")` of the Senate's other `r 100 - as.numeric(fx("n_races"))` are already Democratic. So your `r fx("n_races")` answers do not decide the Senate on their own. They are added to a number that is already fixed.

```{r seat-math}
```

Read the last three rows together. Democrats hold `r D_HOLD` seats that nobody votes on this year. A majority is `r mq("seats for a Democratic majority")`, not 50, because the Vice President breaks ties and the Vice President is a Republican. So Democrats need `r D_NEED` of the `r fx("n_races")` races on your map. They currently hold `r fx("d_up")` of those `r fx("n_races")` seats, which makes the requirement a net gain of `r fx("d_net_gain")`.

`r note(paste0("Two of the ", fx("n_races"), " are special elections, in Florida and Ohio, for seats whose regular six-year term does not end until 2028. Marco Rubio and JD Vance both left the Senate in January 2025 for jobs in the executive branch, and both seats were filled by appointment. A chapter that counted only the 33 seats whose terms are actually expiring would leave out Ohio, which is one of the most contested races in the country."))`

## Where the data comes from, and what it is for

The `r fx("n_races")` races, the people in them, and `r fx("n_forecasters")` professional forecasters' current ratings of every one of them all come from a single Wikipedia article, "2026 United States Senate elections". It is read as wikitext rather than as a web page, because the wikitext is where the ratings still exist as words. On the rendered page a rating is a coloured table cell, and a colour cannot be read back out.

**Who is in it, and how they got there.** Every candidate who has qualified for a 2026 Senate ballot, every senator whose seat is up, and every rating twelve named forecasters have published. The candidates come from state election officials' own filing lists, cited line by line. The ratings come from the forecasters' own pages, each carrying the date it was last changed.

**Who it was made for.** Nobody in particular, which is the unusual thing about it. The twelve forecasters publish no common file and use no common vocabulary. Cook says "Solid Republican" where Sabato says "Safe Republican"; two of the twelve use a category called "Tilt" that the other ten do not have. This table is the only place they sit side by side, and it exists because volunteers put them there.

**What it costs to get.** Nothing, and no account. One request to the MediaWiki interface returns the whole article, currently `r format(fxn("wiki_chars"), big.mark = ",")` characters of it.

**What has already been done to it.** Three things, and the third is a choice you should argue with. The candidate lists were reduced to the names and party labels. The ratings were pulled out of the templates that draw them. And the five rating words were turned into numbers so they could be averaged, on a scale set out below.

`r note("Wikipedia is not the authority on any of this and is not treated as one here. Every rating in the table is footnoted to the forecaster who published it, and every candidate to a secretary of state. What the article supplies is the assembly, not the facts. The risk that comes with that is real and worth naming: an article edited daily during a campaign is a moving target, so the version read here is kept on disk beside the numbers it produced.")`

## Seven words, and the number line they were forced onto

Forecasters do not publish probabilities. They publish one of five words, and the word is a category rather than a measurement:

```{r scale-table}
```

Averaging words is impossible, so this chapter turned them into the numbers in the middle column: positive for the Democrat, symmetric around a toss-up. Two of the twelve forecasters also use "Tilt", a category weaker than "Lean", and it is scored at half a point.

**Nothing makes those gaps equal.** Saying that "Likely" is exactly twice "Lean" is an assumption, not a finding, and a different set of numbers would produce a different average. The scale is printed here so you can disagree with it. That is the honest position for any scale that turns words into arithmetic, and most published averages of expert ratings do not print theirs.

There is a second problem with the scale, and it is worse. It has seven squares, all of which assume a Democrat is running against a Republican. In `r fx("n_no_dem")` of the `r fx("n_races")` races, no Democrat filed at all: `r fx("no_dem_states")`.

```{r no-dem}
```

Nebraska is the one to look at. Dan Osborn is an independent who took Republicans to within a few points in 2024, running with no Democrat on the ballot against him. A scale that runs from Safe D to Safe R cannot say what a strong independent looks like. You will still have to rate Nebraska, and whatever you put there will be wrong in a way the map cannot show.

## Now rate all thirty-five

Figure 1 is the instrument. Pick a rating from the row of buttons, then click states to paint them. The running total at the top adds your calls to the `r D_HOLD` Democratic seats that are not on the ballot. It is always a prediction about the whole Senate, not about the part you can see.

Two buttons fill the map for you: one from the result of each seat's last election, one from the twelve forecasters' average. **Use them second, not first.** Starting from someone else's map and adjusting is a different act from building your own, and it will pull your answer toward theirs. If you want to know what you think, paint the map before you look at either.

```{r rate-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

```{r rate-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Your map.** One square per state, all the same size, because every state elects two senators no matter how many people live in it. The running total above the map adds your `r fx("n_races")` calls to the `r D_HOLD` Democratic seats that are not on the ballot. A toss-up is counted as neither, which is why the total is often a range.

```{r worksheet}
```

**Table 1. The `r fx("n_races")` races, in the order the code reads them.** This is the paper version of Figure 1, and it is the one to use if you are reading this as an attachment. Fill the last column, then read your ratings down the column to get your code: 1 for Safe D through 7 for Safe R.

`r note(paste0("The incumbent is not the ballot. ", fx("n_retiring"), " of these senators are retiring, ", fx("n_lost_primary"), " lost their own party's primary, and ", fx("n_appointed"), " hold seats they were appointed to rather than elected to. Only ", fx("n_incumbent_running"), " of the ", fx("n_races"), " races have the sitting senator running for the seat again. Rating a race by how the incumbent has done before is, in most of these, rating somebody who is not running."))`

## Submit it, so the class has an average

Your code is `r fx("n_races")` digits long and it is the whole of your answer. Paste it into the class form, which asks for two things: the code, and your name. Nothing else.

The form feeds a spreadsheet. That spreadsheet is what the second map on this page reads, and until people submit, the map has nothing to average. A class average is not a better forecast than yours. There is no reason a room's mean should beat a room's best guess. But it is a different object, and the gap between it and your map is the most interesting thing on this page.

`r if (HAS_CLASS) note(paste0("This copy was built with ", fx("class_n"), " submitted codes. Figure 2 shows them.")) else note("This copy was built before any codes were submitted, so Figure 2 opens on the last-election map instead. Codes pasted into the box under Figure 2 are averaged in the browser, without rebuilding anything, which is how the class map works in the room on the day.")`

## What everyone else has already committed to

```{r compare-static, eval=!knitr::is_html_output(), fig.height=8.4}
```

```{r compare-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same `r fx("n_races")` squares, four ways.** Your map, the `r fx("n_forecasters")` forecasters averaged, each seat's last election, and the class's average once codes exist. Nothing here is a result. Three of the four layers are somebody's opinion and the fourth is a different election.

Read the forecaster layer first, and read the total it implies rather than the colours. Averaged across all `r fx("n_forecasters")`, the ratings call `r fx("fc_d")` of the `r fx("n_races")` races for the Democrats. Added to the `r D_HOLD` seats not on the ballot, that is **`r fx("fc_d_total")` Democratic seats**. One short of the majority, in a chamber where 50-50 leaves the Republicans in charge because the Vice President breaks ties.

That is the sharpest thing on this page and it is worth sitting with. The consensus of `r fx("n_forecasters")` professional forecasters, in late August, does not resolve the question. It lands on the one number that resolves nothing.

## Where the professionals disagree with each other

`r lede(paste0("On ", fx("n_unanimous"), " of the ", fx("n_races"), " races all ", fx("n_forecasters"), " forecasters give the same answer. The interesting races are the other ", as.numeric(fx("n_races")) - as.numeric(fx("n_unanimous")), "."))`

```{r spread-static, eval=!knitr::is_html_output(), fig.height=6.6}
```

```{r spread-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Every rating, one row per race.** Grey dots are the `r fx("n_forecasters")` forecasters; the diamond is their average. A row where the dots sit on top of each other is a race nobody argues about. A row where they spread out is a race where the professionals do not agree, which is not the same thing as a close race.

Two rows are worth finding by hand. **`r fx("closest_state")`** has the average nearest to dead even, at `r fx("closest_mean")`. That is a race the twelve collectively cannot call. **`r fx("most_disagreed")`** has the widest spread: its ratings run from `r fx("most_disagreed_min")` to `r fx("most_disagreed_max")` on this scale, which is the difference between a race that tilts one way and a race that is over.

Those are different kinds of uncertainty and it is worth separating them. A toss-up everyone agrees is a toss-up is a statement about the race. A race where half the forecasters say one thing and half say another is a statement about the forecasters.

The `r fx("n_forecasters")` are not equally far from each other, either. The last column below is how far each one's ratings sit, on average, from the mean of the other eleven.

```{r forecaster-table}
```

**Table 2. The `r fx("n_forecasters")`.** `r fx("outlier_forecaster")` sits furthest from the rest, at `r fx("outlier_gap")` on a scale whose whole width is six. `r fx("closest_forecaster")` sits closest. Neither of those facts says anything about who is right, and it would be a mistake to read the second as a recommendation. A forecaster who agrees with everyone else adds nothing to an average; a forecaster who disagrees adds the only thing an average has to work with.

The dates matter too. These ratings were published between `r fx("ratings_first_date")` and `r fx("ratings_last_date")`, which is a spread of more than two weeks in a campaign where two races moved category during it. Some of the twelve are answering a slightly older question than the others.

## What a rating cannot do, and what your sheet can

A rating is not a probability. "Lean D" does not mean 65%, or any other number, and no forecaster on this page publishes what it does mean. That is why the scale here had to be invented before anything could be averaged, and it is why the average is a summary of language rather than a measurement of chance.

This matters for how you will grade yourself in November. If you called a race Safe D and it went Republican, you were wrong in a way you cannot argue with. If you called it Lean R and it went Democratic by a point, you were arguably right about the shape and wrong about the outcome. No rule on this page settles which. **Decide now how you will score yourself, while you still have no idea what happened.** That is the second half of the exercise and it is the half people skip.

`r note("A useful rule, and one you may reject: count a race as correctly called if you put it on the side that won, count a toss-up as half credit either way, and count Safe as double weight in both directions. Write down whatever rule you choose, next to your ratings, and date it.")`

## What this chapter cannot tell you

**Whether any of these ratings are any good.** Twelve forecasters agreeing is not evidence. They read the same polls, the same fundraising reports and each other, so their agreement is close to one opinion counted twelve times. The spread in Figure 3 is a lower bound on the real uncertainty, not a measure of it.

**What the polls say.** Deliberately. There is no polling in this chapter, and the ratings that are here have polling baked into them without saying how much. A rating is a judgement that has already digested the evidence, and you cannot take it apart again from the outside.

**Whether the candidates on Table 1 will be the candidates.** The list was current when the article was read. Maine's Democratic nominee in this file is the party's second choice, after the first left the race in August. Nothing stops that happening again between now and November in any of the `r fx("n_races")` races.

**What happens in the House.** The Senate is `r fx("n_races")` decisions and the House is 435. They are not independent, because a national swing moves both, and a chapter about one of them cannot tell you about the other. The seat-forecast chapter takes up that problem directly.

**Whether `r fx("n_no_dem")` races have been rated honestly at all.** Idaho, Nebraska and South Dakota have no Democrat on the ballot, and in two of them the challenger is an independent. Every forecaster on this page still put those races on a Democrat-versus-Republican scale, and so did you, because the instrument gave you nowhere else to put them.

## Sources

**Data**

- Wikipedia, "2026 United States Senate elections"
  (<https://en.wikipedia.org/wiki/2026_United_States_Senate_elections>), fetched
  as wikitext through the MediaWiki API. Two tables are read: the *Predictions*
  table, for each race's Cook Partisan Voting Index, incumbent, last result and
  twelve forecasters' ratings; and the *Race summary* tables, for the qualified
  candidates. The version read is kept beside the numbers it produced, because
  the article is edited daily during a campaign.
- The `r fx("n_forecasters")` forecasters whose ratings that table collects,
  each cited there to its own page and carrying the date its ratings were last
  changed: `r { f <- FCS[order(FCS$forecaster), ]; paste0(paste(sprintf("%s (%s)", f$forecaster, format(as.Date(f$published), "%e %B %Y")), collapse = ", ")) }`.
  Their published ratings are recorded here, not recomputed.
- `@unitedstates/congress-legislators`, `legislators-current.csv`
  (<https://unitedstates.github.io/congress-legislators/legislators-current.csv>),
  maintained from the Senate's own records. It is used only as a check: the 33
  non-special races must be exactly the states with a Class 2 senator, and the
  two special-election seats must not be Class 2. If those two tests fail, the
  parse of the Wikipedia table is wrong.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`class_ratings.csv`](data/derived/class_ratings.csv), [`forecasters.csv`](data/derived/forecasters.csv), [`races.csv`](data/derived/races.csv), [`ratings_long.csv`](data/derived/ratings_long.csv), [`seat_math.csv`](data/derived/seat_math.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `ratings_long.csv` has every forecaster's rating of every race, scored. Find the races they disagree about most. Are those the closest races, or just the least covered?
- `forecasters.csv` gives each one a `mean_gap`. Work out who is boldest and who hedges. Then decide which you would rather read in August.
- `races.csv` carries `pvi` and `last_share` for every seat. Find the races where the ratings depart furthest from the partisan lean. What are the forecasters seeing?
- Commit to your own rating for all the races before November, then work out what your score would be under the same scoring rule. Would you have beaten the forecasters?

**The choices this chapter made, which are not in any source**

- The seven-point scale, and the numbers behind it. Safe 3, Likely 2, Lean 1,
  Tilt 0.5, Toss-up 0, signed positive for the Democrat.
- The rule that turns a past election into a rating for Figure 2's last-election
  layer: a winner's share of `r fx("base_cut_safe")`% or more is Safe,
  `r fx("base_cut_likely")` to `r fx("base_cut_safe")` is Likely,
  `r fx("base_cut_lean")` to `r fx("base_cut_likely")` is Lean, and under
  `r fx("base_cut_lean")` is a toss-up. The share is of all votes cast, third
  parties included, because that is what the source publishes.
- Counting the two independent senators who caucus with the Democrats among the
  47. That is how every seat count in the coverage is stated and it is not in
  any file.

**The article this chapter was built around**

- Balk, T. "Can Democrats Win Back the Senate? Why Their Chances Are Improving."
  *The New York Times*, 24 August 2026. It reports the six-state core
  battlefield (Alaska, Iowa, Maine, Michigan, Ohio and Texas) and the seat
  arithmetic reproduced above.

Every figure is recomputed from the files named here, and checked against the values asserted in `checks.csv`. The table is printed verbatim.

```{r checks}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
