---
title: "Part IV — Records of Political Actors"
type: chapter
subtitle: "What somebody who ran for office leaves behind"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-4-political-actors-code.R")
```

```{r setup, include=FALSE}
```

Everyone in this part put themselves forward. They ran for office, or raised money to help somebody run, or registered to lobby the people who won. **That one fact is what holds the part together**, and it is the line between this part and the next one, where the people in the records never volunteered for anything.

Because they put themselves forward, the rules are different. A member of Congress votes on the record, every time, and the record is published. A campaign files what it raised and who from. A lobbyist registers, and says roughly what they were paid to care about. None of it was collected to help you. All of it is public, and most of it is public because a statute says it must be.

## Disclosure is not audit

Here is the thing to hold onto, because it is the single most common mistake made with this kind of data.

**A disclosure is a claim its subject made about itself.** The Federal Election Commission does not verify that a campaign spent money the way the campaign says it did. It publishes what was filed. The lobbying registry does not check whether the issues listed were the issues discussed. It publishes what was submitted.

So these files are simultaneously the most detailed data in this book and the least independently checked. A number here is excellent evidence about **what somebody was willing to say on a form with their name on it**, and much weaker evidence about what happened.

That is not a reason to distrust it. It is a reason to be precise about what a sentence built on it is actually claiming.

## A roll call is a different animal

The voting record is the exception, and it is why this part opens with it rather than with money. When a member of Congress votes, the vote *is* the event — not a report of an event. Nobody self-reported it, nobody could have misremembered it, and there is no filing deadline to miss.

That makes it the cleanest data in the whole book, and it is worth meeting a clean source early so you have something to compare the messy ones against.

What gets built on top of it is another matter, and Part VI takes that up.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

## Where this part sits

```{r reusetab}
```

## The weakest part of the book, and it is worth saying so

This part is thinner than it should be, and the gaps are real rather than stylistic. Congressional votes, campaign money, lobbying, press attention and what a politician chooses to share are the sources this part does have. Political actors generate many more records that this part does not touch: committee assignments, staff hiring, travel, state legislative voting, and the enormous world of local government.

Nothing here is wrong. There is simply less of it than the subject deserves, and it is the part most likely to grow.

## How to read this part

Start at [the voting record](../rollcalls-source/rollcalls-source-brief.html) for the clean case, then go to [campaign money](../finance-source/finance-source-brief.html) for the self-reported one. The contrast between the two is most of the argument.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens.
- The statutory filings — Federal Election Commission reports and Lobbying
  Disclosure Act registrations — together with the congressional roll-call
  record and the Supreme Court database. Each chapter names its own source and
  the date its copy was retrieved.
