---
title: "Lab 4: A Century and a Half of Campaigns"
subtitle: "84-355 Democracy's Data | Week 4"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Chapter 3 is about how American campaigns changed: from party-run operations to
candidate-run ones, from conventions to primaries, from newspapers to television
to whatever we are in now.

This lab looks at the other half of that story. Campaigns did not transform in a
vacuum — they transformed on a **map that was itself transforming**. For most of
the period we are about to look at, a third of the country was not contested at
all. No amount of television changes a campaign's behaviour as much as a state
becoming winnable.

We have 41 presidential elections, 1864 to 2024, state by state. Everything is
base R. Work in order — later chunks use objects made earlier.

***

# Part 1: 160 years in one table

```{r read}
pres <- read.csv("data/pres_states_1864_2024.csv", stringsAsFactors = FALSE)
dim(pres)
head(pres)
```

Nearly two thousand rows: one per state per election. `democrat` and
`republican` are percentages of the total vote in that state.

How many elections, and how many states in each?

```{r coverage}
length(unique(pres$year))
range(pres$year)
head(table(pres$year), 6)
tail(table(pres$year), 3)
```

Twenty-five states in 1864, fifty-one now. That is not a data problem — it is the
country growing. But it means **every comparison across time is a comparison
between different sets of states**, and we will have to keep saying so.

Look again at 1864: twenty-five states. There were thirty-six in the Union that
year.

> **Eleven states are missing from 1864 because they were in armed rebellion.**
> The file does not say so. It just has fewer rows. This is the first thing to
> understand about historical data: absence is never annotated.

***

# Part 2: Why we will work in two-party shares

Last week's lab found that third-party votes go missing from election files.
This file has the same problem, and is honest enough that we can measure it:

```{r other-na}
sum(is.na(pres$other))
nrow(pres)
```

Most rows have no third-party number at all. So `democrat + republican` will not
add to 100, and the gap is not meaningful:

```{r sums}
p24 <- pres[pres$year == 2024, ]
summary(p24$democrat + p24$republican)
```

The fix is the standard one for historical comparison: ignore everyone else and
ask what share of the **two-party** vote each side won.

```{r two-party}
pres$d2 <- 100 * pres$democrat / (pres$democrat + pres$republican)
round(head(pres$d2), 1)
```

This is a real choice with a real cost. In 1912, 1968 and 1992 — Roosevelt,
Wallace, Perot — a two-party share hides most of what happened. Keep that in
mind when those years look strange later.

## When a missing value is a historical fact

Nineteen rows have no Democratic or no Republican share at all:

```{r na-rows}
absent <- pres[is.na(pres$d2), c("year", "state_abbrev", "democrat",
                                 "republican", "winner")]
absent
```

These are not gaps. Each one records a state where a major party **had no line on
the ballot**. In 1892 the Populists fused with the Democrats across the mountain
west, so there was no separate Democratic ticket. In 1912 the Progressives kept
Taft off California's ballot entirely.

And then:

```{r alabama}
absent[absent$state_abbrev == "AL", ]
```

> **In 1948 and again in 1964, there was no Democratic slate on the Alabama
> ballot.** In 1948 the Dixiecrats replaced it; in 1964 Alabama Democrats
> refused to pledge to Johnson. A voter who wanted to vote for Harry Truman, or
> for Lyndon Johnson, could not do it.
>
> The ballot is not a neutral instrument. Hold on to this — it is the whole
> subject of Part 3.

***

# Part 3: The longest-running fact in American politics

Define the eleven states of the Confederacy. This is a choice, and you should see
it being made:

```{r south}
south <- c("AL", "AR", "FL", "GA", "LA", "MS", "NC", "SC", "TN", "TX", "VA")
pres$region <- ifelse(pres$state_abbrev %in% south, "South", "Rest of country")
```

Now average the Democratic two-party share in each region, election by election:

```{r region-means}
reg <- aggregate(d2 ~ year + region, data = pres, FUN = mean, na.rm = TRUE)
wide <- reshape(reg, idvar = "year", timevar = "region", direction = "wide")
names(wide) <- c("year", "rest", "south")
wide$gap <- round(wide$south - wide$rest, 1)
head(wide, 4)
```

Plot both lines:

```{r south-plot, fig.width = 8, fig.height = 5.5}
plot(wide$year, wide$south, type = "l", lwd = 2.5, col = "#2166AC",
     ylim = c(20, 95), xlab = "", ylab = "Democratic share of two-party vote (%)",
     main = "The Solid South, and its end")
lines(wide$year, wide$rest, lwd = 2.5, col = "#B2182B")
abline(h = 50, lty = 3, col = "grey50")
abline(v = 1964, lty = 2, col = "grey30")
text(1966, 90, "1964", pos = 4, cex = 0.85, col = "grey30")
legend("bottomleft", legend = c("11 Confederate states", "Rest of country"),
       col = c("#2166AC", "#B2182B"), lwd = 2.5, bty = "n", cex = 0.9)
```

> **Pause and read that.** For eighty years the blue line sits far above the red
> one — the South voting 70, 80, sometimes 85 percent Democratic while the rest
> of the country split near even. Then the lines cross, and stay crossed.

When exactly?

```{r crossover}
wide[wide$year >= 1948 & wide$year <= 1976, c("year", "south", "rest", "gap")]
```

The gap peaks around **+42 points in 1924**. It is still nearly +18 in 1948. It
goes negative for the first time in **1964** — the year the Civil Rights Act
passed, when Johnson won a national landslide and Goldwater carried the Deep
South.

It does not vanish overnight. Carter, a Georgian, pulls it back positive in 1976
(+6.8) and it holds in 1980 (+5.5). **From 1984 onward it is negative in every
single election**, and by 2024 it is −7.4.

By 2024 the South is about 7 points *more* Republican than the rest of the
country. The sign has flipped and stayed flipped.

## A dead man wins two states

While we are here, look at who the file says won in 1960:

```{r who-won-1960}
table(pres$winner[pres$year == 1960])
pres[pres$year == 1960 & pres$winner == "Harry S. Truman",
     c("year", "state_abbrev", "democrat", "republican", "other", "winner")]
```

**Harry Truman was not a candidate in 1960.** He had left office eight years
earlier.

What actually happened is more interesting than a typo. Alabama and Mississippi
ran slates of **unpledged electors** — Democrats who appeared on the ballot
promising *not* to support the national ticket, hoping a close election would let
the South bargain over civil rights. Mississippi's unpledged slate won outright.
Whoever compiled this file needed a name for that slate and reached for a
Democrat, and Truman is what came out.

> This is a Week 2 problem wearing 1960 clothes. Somebody made a judgement call
> while assembling a file, wrote nothing down, and the judgement is now sitting
> in your data looking like a fact. Notice that it did **not** corrupt the
> numbers we have been using — the vote shares are fine. Only the label is
> wrong. Bad data is rarely uniformly bad, which is exactly what makes it
> dangerous.

***

# Part 4: How different are the states from one another?

If every state voted alike, the spread between them would be zero. If the country
were made of separate political worlds, the spread would be large. Standard
deviation across states, election by election:

```{r spread}
spread <- aggregate(d2 ~ year, data = pres, FUN = sd, na.rm = TRUE)
names(spread)[2] <- "sd"
spread$sd <- round(spread$sd, 2)
tail(spread, 5)
```

```{r spread-plot, fig.width = 8, fig.height = 5}
plot(spread$year, spread$sd, type = "b", pch = 19, lwd = 2, col = "#4D4D4D",
     xlab = "", ylab = "SD of state Democratic share (points)",
     main = "How far apart the states are")
abline(v = 1960, lty = 2, col = "grey40")
text(1962, 20, "1960: the states\nnever agreed more", pos = 4, cex = 0.8,
     col = "grey30")
```

The most-alike year and the least-alike years:

```{r spread-extremes}
head(spread[order(spread$sd), ], 3)
head(spread[order(-spread$sd), ], 3)
```

The spread was enormous in the era of the Solid South — the South was voting
differently from everyone else, and that shows up as huge variance. It collapses
to a minimum of **5.8 in 1960**. And then it climbs again, roughly doubling by
2024.

Careful here. That one number is made of parts. Some of the distance between
states is a whole region differing from the rest of the country; the rest is
states differing from their own neighbours. The next chunk splits the spread
into those two pieces.

> **Before you run it, write two numbers down.** In 1924, with the Solid South
> intact, **what percentage of all the state-to-state variation was simply the
> South differing from everyone else? And what is that percentage in 2024?** The
> standard deviations invite you to think the country has gone half the way
> back, in which case the second number ought to be a solid fraction of the
> first. Commit to two percentages — not "high" and "lower".

```{r decomp}
between <- function(y) {
  s  <- pres[pres$year == y & !is.na(pres$d2), ]
  g  <- s$region == "South"
  gm <- tapply(s$d2, g, mean); nn <- table(g); tot <- mean(s$d2)
  100 * sum(nn * (gm - tot)^2) / sum((s$d2 - tot)^2)   # between-region share
}

pb <- sapply(c(1924, 2024), between)
names(pb) <- c("1924", "2024")
round(pb, 1)
```

> **`r round(pb[["1924"]], 1)`% in 1924, and `r round(pb[["2024"]], 1)`% in
> 2024.** The first number is usually written down too low and the second far
> too high — most people put 2024 in the thirties or forties, on the
> reasonable-sounding logic that a spread half the way back ought to be half a
> regional story. It is not. The states are further apart today than in 1960,
> but *not* for the 1920s reason. Then, one region was voting unlike the rest.
> Now `r round(100 - pb[["2024"]], 1)`% of the spread is states differing from
> their own neighbours, everywhere, along lines no regional label captures —
> this is the sorting you have heard called polarization. The same statistic,
> two completely different worlds.
>
> A number going up does not tell you why.

***

# Part 5: The vanishing battleground

Here is the fact that matters most for how campaigns behave. Count the states
within five points of an even split:

```{r close}
pres$close <- abs(pres$d2 - 50) < 5
nclose <- aggregate(close ~ year, data = pres, FUN = sum, na.rm = TRUE)
tail(nclose, 6)
```

```{r close-plot, fig.width = 8, fig.height = 5}
plot(nclose$year, nclose$close, type = "h", lwd = 6, col = "#8C6BB1",
     xlab = "", ylab = "States within 5 points of 50-50",
     main = "Competitive states, 1864-2024")
points(nclose$year, nclose$close, pch = 19, col = "#54278F")
```

```{r close-numbers}
nclose[nclose$year %in% c(1960, 1976, 2000, 2012, 2020, 2024), ]
```

Thirty-three competitive states in 1960. Thirteen in 2024 — out of fifty-one, and
with far more states in the union than in 1960.

**This is the sentence to take away from the lab.** A campaign in 1960 had to
think about most of the country. A campaign in 2024 has to think about a dozen
states, and can safely ignore the rest. Nothing in Chapter 3 about consultants or
television explains as much about modern campaign behaviour as this does.

We will see the consequence directly in Week 6, when we look at where candidates
actually go.

***

# Part 6: What this data cannot tell you

Three limits, stated plainly, because you should be able to state them yourself
by now.

**The set of states changes.** Comparing the spread in 1864 (25 states) to 2024
(51) is comparing different objects. The trend is real but the early numbers are
noisier than they look.

```{r states-per-year}
range(table(pres$year))
```

**Two-party share hides third parties.** 1912, 1968 and 1992 all look tamer here
than they were.

```{r third-party-years}
pres[pres$year == 1912 & pres$state_abbrev %in% c("CA", "SD", "WA"),
     c("year", "state_abbrev", "democrat", "republican", "other", "winner")]
```

Theodore Roosevelt won California, South Dakota and Washington in 1912. He does
not appear in our `d2` column at all — he is inside `other`, or missing entirely.

**Shares are not votes.** Every number here is a percentage. A five-point margin
in Wyoming and a five-point margin in California are the same number and
nothing like the same event.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change, or nothing at all. Run it,
look at it, and tell me what it means.

Pick **one**.

***

### Option A — Move the line

Part 3 used the eleven Confederate states. That was a choice. Here it is again
with a different one — the eleven plus the border states that permitted slavery
but did not secede.

```{r opt-a, fig.width = 8, fig.height = 5.5}
# Change this list and re-run. Try adding "KY","MO","MD","WV","OK","DE".
my_south <- c("AL","AR","FL","GA","LA","MS","NC","SC","TN","TX","VA")

pres$my_region <- ifelse(pres$state_abbrev %in% my_south, "South", "Rest")
r <- aggregate(d2 ~ year + my_region, data = pres, FUN = mean, na.rm = TRUE)
w <- reshape(r, idvar = "year", timevar = "my_region", direction = "wide")
names(w) <- c("year", "rest", "south")

plot(w$year, w$south, type = "l", lwd = 2.5, col = "#2166AC", ylim = c(20, 95),
     xlab = "", ylab = "Democratic two-party share (%)",
     main = paste("The South as", length(my_south), "states"))
lines(w$year, w$rest, lwd = 2.5, col = "#B2182B")
abline(h = 50, lty = 3, col = "grey50"); abline(v = 1964, lty = 2, col = "grey30")

w$gap <- round(w$south - w$rest, 1)
w[w$year %in% c(1924, 1948, 1960, 1964, 1976, 2024), ]
```

**Write about:** how much does the story depend on which states you call
Southern? Does the 1964 crossover move, sharpen, or blur — or hold exactly where
it was? Report what you actually find, including "almost nothing changed" if that
is the answer.

Then the real question, whichever way it went: a finding that survives a change
in its own definition is worth more than one that does not. Did this one survive?
And what should you now think of anyone — a journalist, a campaign, a
professor — who reports "the South" as a single number without telling you which
states they meant?

***

### Option B — Pick a state and follow it

```{r opt-b, fig.width = 8, fig.height = 5}
this_state <- "MS"                # <- change to any two-letter abbreviation

s <- pres[pres$state_abbrev == this_state, ]
s <- s[order(s$year), ]

plot(s$year, s$d2, type = "b", pch = 19, lwd = 2, col = "#2166AC",
     ylim = c(0, 100), xlab = "", ylab = "Democratic two-party share (%)",
     main = paste("Presidential voting in", this_state))
abline(h = 50, lty = 3, col = "grey50")
lines(wide$year, wide$rest, lwd = 1.5, col = "grey60", lty = 2)
legend("topright", c(this_state, "Rest of country"),
       col = c("#2166AC", "grey60"), lwd = 2, lty = c(1, 2), bty = "n", cex = 0.85)

s[s$year %in% c(1948, 1960, 1964, 1972, 2024), c("year", "d2", "winner")]
```

**Write about:** narrate the line. Where does your state break from the national
pattern, and when does it rejoin? Try Mississippi, then try Vermont — which ran
the same journey in the opposite direction. Then say what a campaign would have
concluded about your state in 1950, and in 2024. Are those the same campaign?

***

### Option C — When did the battleground shrink?

```{r opt-c, fig.width = 8, fig.height = 5}
margin <- 5                       # <- change this threshold and re-run

pres$is_close <- abs(pres$d2 - 50) < margin
nc <- aggregate(is_close ~ year, data = pres, FUN = sum, na.rm = TRUE)
tot <- as.data.frame(table(pres$year)); names(tot) <- c("year", "n")
nc$year <- as.numeric(nc$year); tot$year <- as.numeric(as.character(tot$year))
nc <- merge(nc, tot, by = "year")
nc$pct <- round(100 * nc$is_close / nc$n, 1)

plot(nc$year, nc$pct, type = "b", pch = 19, lwd = 2, col = "#54278F",
     ylim = c(0, 80), xlab = "", ylab = "% of states that are competitive",
     main = paste("States within", margin, "points of even"))

nc[nc$year %in% c(1960, 1976, 1992, 2008, 2024), ]
```

**Write about:** this version uses the *share* of states rather than the count,
which is the fairer comparison when the number of states keeps changing. Does the
decline still hold? Now change `margin` to 2 and to 10 — does the story survive
all three thresholds, or only some? A finding that appears at one cut-off and
vanishes at another is worth less, and you should say so if that is what you see.

***

### Option D — Ask your own question

Something in here is bound to look wrong. 1912 and 1968 are strange for real
reasons. Vermont and Mississippi swapped places. Follow whatever nags at you.
Re-running an existing chunk with a different state or year is plenty.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at 160 years of election results and say something
true about how campaigning changed.

***

## Where the data came from

**State returns, 1864–2024** — Jason Timm, *PresElectionResults*
(<https://github.com/jaytimm/PresElectionResults>), state series compiled from
per-election sources resting ultimately on Dave Leip's *Atlas of U.S.
Presidential Elections* and the official state canvasses.

**National popular and electoral vote** — the same compilation, in
`data/pres_national.csv`. Not used above; it is there for Option D.

Two warnings the source deserves. Nineteenth-century "popular vote" totals rest
on state records of wildly varying quality, and in much of the South after 1877
they rest on elections conducted under systematic disenfranchisement — a
90 percent Democratic share in 1900 Mississippi is an accurate count of a vote
that most adults in Mississippi were prevented from casting. The number is real.
What it measures is not what it appears to measure.

The script that downloaded, checked and wrote these files is `data/build-data.R`.
It prints its own sanity checks, including the nineteen rows where a major party
was absent from the ballot.
