---
title: "Data Session 4 Key, part 1: A Century and a Half of Campaigns"
subtitle: "84-355 Democracy's Data | Session 4 — Returns over time (self-study) | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
pres <- read.csv("data/pres_states_1864_2024.csv", stringsAsFactors = FALSE)
pres$d2 <- 100 * pres$democrat / (pres$democrat + pres$republican)
gap_for <- function(sts) {
  pres$r <- ifelse(pres$state_abbrev %in% sts, "S", "R")
  a <- aggregate(d2 ~ year + r, data = pres, FUN = mean, na.rm = TRUE)
  w <- reshape(a, idvar = "year", timevar = "r", direction = "wide")
  names(w) <- c("year", "rest", "south"); w$gap <- round(w$south - w$rest, 1); w
}
c11 <- c("AL","AR","FL","GA","LA","MS","NC","SC","TN","TX","VA")
c17 <- c(c11, "KY","MO","MD","WV","OK","DE")
```

```{r vals, include=FALSE}
# Every figure quoted in this key is computed here and inserted inline, using
# the same definitions as the brief, so the two cannot drift apart.
A11 <- gap_for(c11); A17 <- gap_for(c17)
G   <- function(w, y) w$gap[w$year == y]
FIRSTNEG <- function(w) w$year[!is.na(w$gap) & w$gap < 0][1]
FLIP <- function(w) {
  yy <- w$year[!is.na(w$gap)]
  min(yy[sapply(yy, function(y) all(w$gap[w$year >= y & !is.na(w$gap)] < 0))])
}

# state the year, and say "either way" only if the two state lists agree
EITHER <- function(f) {
  x <- f(A11); y <- f(A17)
  if (identical(x, y)) paste(x, "either way")
  else paste0(x, " on the eleven-state list and ", y, " on the seventeen")
}

SP  <- aggregate(d2 ~ year, data = pres, FUN = sd, na.rm = TRUE)
names(SP)[2] <- "sd"
SDY   <- function(y) SP$sd[SP$year == y]
SDMIN <- SP$year[which.min(SP$sd)]; SDMAX <- SP$year[which.max(SP$sd)]

NST <- function(y) sum(pres$year == y)
NCL <- function(y, mg = 5) sum(abs(pres$d2[pres$year == y] - 50) < mg, na.rm = TRUE)
PCL <- function(y, mg = 5) 100 * NCL(y, mg) / NST(y)

# share of state-to-state variation that is South vs rest (the brief's Figure 4)
VB <- function(y) {
  d <- pres[pres$year == y & !is.na(pres$d2), ]
  tot <- mean(d$d2); grp <- d$state_abbrev %in% c11
  gm  <- tapply(d$d2, grp, mean); nn <- table(grp)
  100 * sum(nn * (gm - tot)^2) / sum((d$d2 - tot)^2)
}

ST <- function(s, y) pres$d2[pres$state_abbrev == s & pres$year == y]
VTFIRST <- min(pres$year[pres$state_abbrev == "VT" & pres$d2 > 50], na.rm = TRUE)

pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
sg <- function(x, k = 1) paste0(ifelse(x >= 0, "+", "−"),
                                formatC(abs(x), format = "f", digits = k))
```

All Part 7 code is given to students. Grade the interpretation, not the code.

Every number below was run against the data file.

***

# The lab in one sentence

American campaigns transformed partly because the map they ran on transformed —
a country where a third of the states were uncontested for eighty years became
one where a dozen states decide everything, and that shift explains more about
modern campaign behaviour than any change in technique.

***

# The three findings students should leave with

```{r headline}
a <- gap_for(c11)
a[a$year %in% c(1924, 1948, 1964, 1976, 1980, 1984, 2024), c("year","south","rest","gap")]

sp <- aggregate(d2 ~ year, data = pres, FUN = sd, na.rm = TRUE); names(sp)[2] <- "sd"
sp[sp$year %in% c(1924, 1960, 2024), ]

pres$close <- abs(pres$d2 - 50) < 5
nc <- aggregate(close ~ year, data = pres, FUN = sum, na.rm = TRUE)
nc[nc$year %in% c(1960, 1976, 2024), ]
```

1. **The South flips.** Gap peaks at **`r sg(G(A11, 1924))` in 1924**, is
   **`r sg(G(A11, 1948))` in 1948**, goes negative for the first time in
   **`r FIRSTNEG(A11)`**, returns positive under Carter
   (**`r sg(G(A11, 1976))` in 1976**, **`r sg(G(A11, 1980))` in 1980**), and is
   negative in **every election from `r FLIP(A11)` onward**, reaching
   **`r sg(G(A11, 2024))` in 2024**.
2. **State-to-state spread is U-shaped.** SD bottoms at
   **`r pc(SDY(SDMIN), 2)` in `r SDMIN`** and **roughly
   `r pc(SDY(2024) / SDY(SDMIN))`× that by 2024, at `r pc(SDY(2024), 2)`** — but
   for a different reason than the peak of `r pc(SDY(SDMAX))` in `r SDMAX`. In
   `r SDMAX`, **`r pc(VB(SDMAX))`%** of the state-to-state variation was the
   South differing from everywhere else; in 2024 that is **`r pc(VB(2024))`%**.
   The country is spread out again, but along different seams.
3. **The battleground shrinks.** **`r NCL(1960)`** states within 5 points in
   1960; **`r NCL(2024)`** in 2024, out of `r NST(2024)` states rather than the
   `r NST(1960)` that existed then.

***

# Option A — Move the line

## Verified output

```{r a}
b <- gap_for(c17)
m <- merge(a[, c("year","gap")], b[, c("year","gap")], by = "year",
           suffixes = c("_11", "_17"))
m[m$year %in% c(1924, 1948, 1960, 1964, 1976, 1984, 2024), ]
```

**The magnitudes shrink; the story does not move.** Adding the six border states
cuts the 1924 peak from `r pc(G(A11, 1924))` to `r pc(G(A17, 1924))` and the
1964 trough from `r sg(G(A11, 1964))` to `r sg(G(A17, 1964))`. But the first
negative year is **`r EITHER(FIRSTNEG)`**, and the permanent flip is
**`r EITHER(FLIP)`**.

## What a good answer says

**3/4:** Reports that the numbers changed but the shape didn't, and identifies
1964 as stable.

**4/4:** Draws the right inference — this is a **robustness check the finding
passed**. The realignment is not an artifact of a convenient state list. A
student who says "I tried to break it and couldn't, which makes me believe it
more" is doing science.

**4/4, best:** notices *why* the magnitudes shrink — the border states were
always less Democratic than the Deep South, so adding them dilutes both extremes.
The dilution is itself evidence: it says the effect is concentrated where the
theory predicts.

**Common stumble:** treating "the numbers changed" as evidence the finding is
shaky. Push back — the question is whether the *conclusion* changed, and it
didn't. This distinction is worth a minute of class time.

***

# Option B — Pick a state and follow it

## Verified output

```{r b}
for (s in c("MS", "VT")) {
  x <- pres[pres$state_abbrev == s, ]
  cat("\n", s, "\n"); print(x[x$year %in% c(1948, 1960, 1964, 1972, 2024),
                              c("year", "d2", "winner")])
}
```

**Mississippi:** `r pc(ST("MS", 1948))` Democratic two-party share in 1948 →
`r pc(ST("MS", 1964))` in 1964 → `r pc(ST("MS", 2024))` in 2024.
**Vermont:** `r pc(ST("VT", 1948))` in 1948 → `r pc(ST("VT", 1964))` in 1964 →
`r pc(ST("VT", 2024))` in 2024.

They have swapped places almost exactly. `r VTFIRST` was the **first time
Vermont ever voted Democratic for president**.

## What a good answer says

**3/4:** Narrates the line and identifies the break year for their state.

**4/4:** Notices that Mississippi 1948 shows `d2` of `r pc(ST("MS", 1948))` while
the `winner` column says
**`r pres$winner[pres$state_abbrev == "MS" & pres$year == 1948]`**. Both are
true: Truman took `r pc(ST("MS", 1948), 0)`% of the
Truman-versus-Dewey vote, and neither of them won the state, because the
Dixiecrat took it. That is the two-party-share limitation from Part 2 biting in
the exact place the lab warned it would. Students who catch this have connected
two parts of the lab unprompted.

**4/4, best:** answers the campaign question seriously. A 1950 strategist wrote
Mississippi off as unwinnable-because-already-won and Vermont as
unwinnable-because-already-lost; in 2024 a strategist writes both off for exactly
the opposite reasons. The state changed completely and its *campaign treatment*
did not change at all. Uncontested is uncontested.

**Common stumble:** picking a state that never did anything interesting (Utah,
Nebraska) and concluding the lab is boring. Redirect them to MS, VT, WV, CA, or
GA in office hours.

***

# Option C — When did the battleground shrink?

## Verified output

```{r c}
tot <- as.data.frame(table(pres$year)); names(tot) <- c("year","n")
tot$year <- as.numeric(as.character(tot$year))
for (mg in c(2, 5, 10)) {
  pres$cl <- abs(pres$d2 - 50) < mg
  n <- aggregate(cl ~ year, data = pres, FUN = sum, na.rm = TRUE)
  n <- merge(n, tot, by = "year"); n$pct <- round(100 * n$cl / n$n, 1)
  cat("margin", mg, ":",
      paste0(n$year[n$year %in% c(1960,1976,1992,2008,2024)], "=",
             n$pct[n$year %in% c(1960,1976,1992,2008,2024)], "%", collapse = "  "), "\n")
}
```

| threshold | 1960 | 1976 | 1992 | 2008 | 2024 |
|---|---|---|---|---|---|
| ±2 pts | `r pc(PCL(1960, 2))`% | `r pc(PCL(1976, 2))`% | `r pc(PCL(1992, 2))`% | `r pc(PCL(2008, 2))`% | **`r pc(PCL(2024, 2))`%** |
| ±5 pts | `r pc(PCL(1960, 5))`% | `r pc(PCL(1976, 5))`% | `r pc(PCL(1992, 5))`% | `r pc(PCL(2008, 5))`% | `r pc(PCL(2024, 5))`% |
| ±10 pts | `r pc(PCL(1960, 10))`% | `r pc(PCL(1976, 10))`% | `r pc(PCL(1992, 10))`% | `r pc(PCL(2008, 10))`% | `r pc(PCL(2024, 10))`% |

## What a good answer says

**3/4:** Confirms the decline holds as a percentage, not just a count.

**4/4:** Reports honestly that the finding is **robust at ±5 and ±10 but wobbles
at ±2**, where 2024 (`r pc(PCL(2024, 2))`%) is *higher* than 2008
(`r pc(PCL(2008, 2))`%). At the tightest cut-off
you are counting a handful of states and the series gets noisy. The right
conclusion is "the decline is real and large, and the tightest threshold is too
noisy to add anything" — not "the finding failed."

**4/4, best:** notices the ±10 row barely moves across 1960, 1976 and 1992 and
then drops — and then goes and looks at the years the table does not print,
where it is not flat at all. It falls to `r pc(PCL(1972, 10))`% in 1972 and
`r pc(PCL(1984, 10))`% in 1984, both national landslides, and climbs back to
`r pc(PCL(1992, 10))`% in 1992. At ±10 the measure is picking up **how lopsided
the national result was** as much as how many states were genuinely in play, and
the three early columns in the table all happen to be close national elections.
Compare like with like — close elections only — and the decline survives and is
still late: `r pc(PCL(1992, 10))`% in 1992 against `r pc(PCL(2024, 10))`% in
2024, both close nationally. That is a genuinely open question and good Option D
material.

**Do not award the 4 for "the ±10 row is flat until 2008" on its own.** It reads
that way only because of which five years the table prints; across every
election in the file the ±10 share spans
`r pc(diff(range(sapply(SP$year, PCL, 10))))` points. This is the same lesson as Option A in
reverse — there, sampling the state list did not change the conclusion; here,
sampling the years does.

**Common stumble:** reporting only ±5 because it worked and not mentioning the
others. Ask for all three in feedback.

***

# Option D — Their own question

Reward curiosity over machinery. The 1960 Truman rows, the 1912 Roosevelt states,
and the Vermont–Mississippi swap are all live threads.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **This lab rewards noticing.** The strongest submissions will flag something
  the lab did not point at — an odd year, a state that misbehaves, a label that
  cannot be right. Say so in feedback so the habit spreads.
- **Reward robustness reasoning.** Options A and C are both robustness checks in
  disguise. A student who distinguishes "the number moved" from "the conclusion
  moved" has learned the most transferable thing here.

## Anticipated questions

- *"Why does the file say Truman won states in 1960?"* — Alabama and Mississippi
  ran **unpledged elector** slates; Mississippi's won. The compiler needed a name
  and used a Democrat's. Covered in Part 3, and it is the best single example in
  the course of a data error that leaves the numbers intact and the labels wrong.
- *"Why is the South missing in 1864?"* — Eleven states were in rebellion. The
  file records this only as absent rows.
- *"Are the 19th-century numbers trustworthy?"* — Partly. The bigger problem is
  not accuracy but meaning: post-1877 Southern returns are accurate counts of
  elections conducted under systematic disenfranchisement. Say this out loud;
  it connects back to **Session 3 (§203)** and forward to the
  `disenfranchisement/` lab.
- *"Isn't two-party share cheating?"* — It is a choice with a cost, stated in
  Part 2. 1912, 1968 and 1992 are where it hurts.

## Connections forward

- **`campaign-visits`** (retired, on disk) is the direct sequel: `r NCL(2024)`
  competitive states is *why* candidates go where they go. Refer back to Part 5
  by name.
- **Session 12 (roll calls and the bench)** — the realignment here is the mass-level
  counterpart to the elite polarization students will measure there. Same
  phenomenon, two instruments.
- **Session 3 (the ACS and §203)**, which ran last week — Part 3's disenfranchisement caveat is its historical setup.
  The Solid South was solid partly because most Black Southerners could not vote,
  and the 1965 Act is what changed the electorate underneath these lines.
- **Session 1 (election returns)** — the Truman rows and the sparse `other` column are the same
  provenance lesson, now in a historical file.
