---
title: "Older Than They Used to Be"
subtitle: "84-355 Democracy's Data | The age of government actors, and the denominator that decides the answer"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 3.4,
                      dpi = 96, fig.retina = 1)
options(scipen = 999)
D <- "data"

# Formatting only. Every number printed in the prose below is produced by the
# code chunks you can see; these just control how many decimals it shows.
# R rounds half to even and JavaScript rounds half up, so anything appearing in
# both the interactive and the static figure is formatted here, once, in R.
f0 <- function(x) formatC(x, format = "f", digits = 0)
f1 <- function(x) formatC(x, format = "f", digits = 1)
f2 <- function(x) formatC(x, format = "f", digits = 2)
sgn <- function(x, k = 1) paste0(ifelse(x >= 0, "+", "-"),
                                 formatC(abs(x), format = "f", digits = k))
n  <- function(x) format(round(x), big.mark = ",")
```

Everything here runs as given. **You will not write any code.** Your job is to
read what comes out and say what it means.

> **Read `officeholder-age-brief.Rmd` first.** This lab runs the same steps with
> the code exposed.

***

# Where did that number come from?

> *"Congress is older than it has ever been."*
>
> *"Sure, but people live thirty years longer than they used to."*

The first sentence is checkable. The second sounds like it settles the matter
and does not. This lab is about why.

# Step 1 — One file, and what a birth *year* buys you

```{r load}
m  <- read.csv(file.path(D, "members.csv"),         stringsAsFactors = FALSE)
ac <- read.csv(file.path(D, "age_by_congress.csv"), stringsAsFactors = FALSE)
le <- read.csv(file.path(D, "life_expectancy.csv"), stringsAsFactors = FALSE)
et <- read.csv(file.path(D, "entry_tenure.csv"),    stringsAsFactors = FALSE)
pr <- read.csv(file.path(D, "presidents.csv"),      stringsAsFactors = FALSE)

c(rows        = nrow(m),
  congresses  = length(unique(m$congress)),
  first_year  = min(m$year),
  last_year   = max(m$year),
  no_birth_yr = sum(is.na(m$born)))
table(m$chamber)
```

Every age here comes from one column, `born`, in the Voteview member file, and
`born` holds **a year, not a date**. Age in this lab means *the year the
Congress convened minus the year the member was born*.

```{r biden}
pr[pr$bioname == "BIDEN, Joseph Robinette (Joe), Jr.",
   c("bioname", "born", "year", "age_at_term_start")]
```

Joe Biden was inaugurated in January 2021 aged **78**; his birthday is in
November. Year arithmetic says
`r pr$age_at_term_start[pr$bioname == "BIDEN, Joseph Robinette (Joe), Jr."]`.

**Every age in this lab is approximate to within a year**, and the errors point
in both directions, so medians and trends survive it and individual ages do not.

```{r missing-born}
c(rows_with_no_birth_year = sum(is.na(m$born)),
  of_those_before_1830    = sum(is.na(m$born) & m$year < 1830))

# one person, some rows with a birth year and some without
sp <- split(m$born, m$icpsr)
unique(m$bioname[m$icpsr %in%
  as.integer(names(sp)[sapply(sp, function(b) any(is.na(b)) & any(!is.na(b)))])])
```

`r n(sum(is.na(m$born)))` rows have no birth year at all, almost all of them in
the first twenty Congresses. **And exactly one person has a birth year on some
of his rows and not on others**: Grover Cleveland, recorded on his first term
and blank on his second. Nothing in the file announces that. An average would
skip the blank rows and hand back a number that looked perfectly healthy.

# Step 2 — How old is Congress, and when did it change?

```{r trend}
C <- ac[ac$chamber == "Congress", ]
c(youngest      = min(C$median_age),
  youngest_year = C$year[which.min(C$median_age)],
  oldest        = max(C$median_age),
  now           = C$median_age[C$congress == max(C$congress)],
  senate_now    = ac$median_age[ac$chamber == "Senate" &
                                ac$congress == max(ac$congress)],
  house_now     = ac$median_age[ac$chamber == "House" &
                                ac$congress == max(ac$congress)])
C$year[C$median_age == max(C$median_age)]        # every Congress at the max
```

```{r trough}
TR    <- C[C$year >= 1965 & C$year <= 1995, ]
TR    <- TR[which.min(TR$median_age), ]
PRIOR <- max(C$year[C$year < TR$year & C$median_age <= TR$median_age])
c(trough_year = TR$year, trough_median = TR$median_age,
  last_time_that_young = PRIOR)
```

**The rise is real and it is recent.** The median member was
`r f0(TR$median_age)` in `r TR$year` — a figure Congress had not touched since
`r PRIOR` — and `r f0(max(C$median_age))` by
`r min(C$year[C$median_age == max(C$median_age)])`. Almost all of the increase
people argue about happened after 1980.

The Senate has always been the older chamber:
`r f0(ac$median_age[ac$chamber == "Senate" & ac$congress == max(ac$congress)])`
against
`r f0(ac$median_age[ac$chamber == "House" & ac$congress == max(ac$congress)])`
today.

```{r presidents}
c(n_presidencies = nrow(pr),
  presidencies_with_no_age = sum(is.na(pr$age_at_term_start)),
  mean_before_1900 = round(mean(pr$age_at_term_start[pr$year <  1900], na.rm = TRUE), 1),
  mean_since_1981  = round(mean(pr$age_at_term_start[pr$year >= 1981], na.rm = TRUE), 1))
head(pr[order(-pr$age_at_term_start), c("bioname", "year", "age_at_term_start")], 4)
```

`r nrow(pr)` presidencies is not enough observations to support a trend line,
and this lab does not draw one. The arithmetic is still worth having: the
average age on taking office was
`r f1(mean(pr$age_at_term_start[pr$year < 1900], na.rm = TRUE))` before 1900 and
`r f1(mean(pr$age_at_term_start[pr$year >= 1981], na.rm = TRUE))` since 1981.

# Step 3 — The rebuttal, stated as strongly as it deserves

```{r naive}
sitting <- function(y) floor((y - 1787) / 2)   # the Congress sitting in year y
BY <- 1950; EY <- 2018                         # the window where both life
                                               # expectancy series exist
A0 <- C$median_age[C$congress == sitting(BY)]
A1 <- C$median_age[C$congress == sitting(EY)]
E0a <- le$e0_hus[le$year == BY]; E0b <- le$e0_hus[le$year == EY]

c(congress_median_1950 = A0, congress_median_2018 = A1, gain = A1 - A0,
  life_expect_1950 = E0a, life_expect_2018 = E0b, gain_le = E0b - E0a)
```

Life expectancy rose **`r f1(E0b - E0a)` years** while the median member of
Congress aged **`r f0(A1 - A0)`**. On that comparison Congress has aged *less*
than the country it is drawn from, and the rebuttal wins.

**That comparison is wrong.**

# Step 4 — The denominator that flips the answer

`r f1(E0b)` is life expectancy **at birth** — the average length of a life for a
baby born in `r EY`. It rose so much across the twentieth century mostly because
**babies and children stopped dying**, and a sitting senator has already
survived all of that.

The number that describes an officeholder is **remaining life expectancy at an
age they have reached**. NCHS publishes it in the same table, one section down
the page.

```{r correct}
E6a <- le$e65[le$year == BY]; E6b <- le$e65[le$year == EY]
c(remaining_at_65_1950 = E6a, remaining_at_65_2018 = E6b, gain_e65 = E6b - E6a,
  congress_gain = A1 - A0,
  share_of_headline_gain_after_65 = round(100 * (E6b - E6a) / (E0b - E0a)))
```

```{r fig-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
B <- data.frame(lab = c("Life expectancy AT BIRTH", "Median age of CONGRESS",
                        "Remaining life expectancy AT 65"),
                v   = c(E0b - E0a, A1 - A0, E6b - E6a),
                col = c("#8c8c8c", "#C41230", "#2c7fb8"),
                stringsAsFactors = FALSE)
cat(sprintf('
<div id="yd" style="margin:0.6em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const B=[%s], CONG=%s;
const W=760,H=232,M={t:36,r:120,b:38,l:250};
const svg=d3.select("#yd").append("svg").attr("viewBox",`0 0 ${W} ${H}`)
  .attr("style","max-width:100%%;height:auto;font:12px inherit");
const x=d3.scaleLinear().domain([0,12]).range([M.l,W-M.r]);
const y=d3.scaleBand().domain(B.map(d=>d.lab)).range([M.t,H-M.b]).padding(0.34);
svg.append("g").attr("transform",`translate(0,${H-M.b})`).call(d3.axisBottom(x).ticks(6));
svg.append("text").attr("x",(M.l+W-M.r)/2).attr("y",H-6).attr("text-anchor","middle")
  .attr("font-size","11.5px").attr("fill","#555")
  .text("gain between 1950 and 2018, in years");
svg.selectAll("rect").data(B).join("rect")
  .attr("x",x(0)).attr("y",d=>y(d.lab)).attr("height",y.bandwidth())
  .attr("width",d=>x(d.v)-x(0)).attr("fill",d=>d.col).attr("fill-opacity",0.88);
svg.selectAll("text.v").data(B).join("text").attr("class","v")
  .attr("x",x(d3.max(B,d=>d.v))+9).attr("y",d=>y(d.lab)+y.bandwidth()/2+4)
  .attr("font-size","12.5px").attr("font-weight","600").attr("fill",d=>d.col)
  .text(d=>"+"+d.s+" yrs");
svg.selectAll("text.l").data(B).join("text").attr("class","l")
  .attr("x",M.l-10).attr("y",d=>y(d.lab)+y.bandwidth()/2+4).attr("text-anchor","end")
  .attr("font-size","12px").attr("fill","#333").text(d=>d.lab);
svg.append("line").attr("x1",x(CONG)).attr("x2",x(CONG)).attr("y1",M.t-8)
  .attr("y2",H-M.b).attr("stroke","#C41230").attr("stroke-width",1.2)
  .attr("stroke-dasharray","4,3");
svg.append("text").attr("x",x(CONG)).attr("y",M.t-13).attr("text-anchor","middle")
  .attr("font-size","11px").attr("fill","#C41230").text("Congress");
})();
</script>
', paste(sprintf('{"lab":"%s","v":%s,"col":"%s","s":"%s"}',
                 B$lab, f1(B$v), B$col, f1(B$v)), collapse = ","), f1(A1 - A0)))
```

```{r fig-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=2.7}
B <- data.frame(lab = c("Remaining life expectancy AT 65", "Median age of CONGRESS",
                        "Life expectancy AT BIRTH"),
                v   = c(E6b - E6a, A1 - A0, E0b - E0a),
                col = c("#2c7fb8", "#C41230", "#8c8c8c"),
                stringsAsFactors = FALSE)
par(mar = c(3.4, 15.2, 1.0, 3.2))
bp <- barplot(B$v, horiz = TRUE, col = adjustcolor(B$col, 0.88), border = NA,
              xlim = c(0, 12), names.arg = B$lab, las = 1, cex.names = 0.72,
              cex.axis = 0.8, space = 0.55)
abline(v = A1 - A0, lty = 2, col = "#C41230")
text(max(B$v) + 0.3, bp, paste0("+", f1(B$v), " yrs"), adj = 0, cex = 0.75,
     font = 2, col = B$col, xpd = NA)
mtext("gain between 1950 and 2018, in years", 1, line = 2.2, cex = 0.8)
```

**Congress aged `r f0(A1 - A0)` years. Remaining life expectancy at 65 grew
`r f1(E6b - E6a)`.** Against the at-birth yardstick Congress aged slower than
the country; against the honest one it aged faster. Same source, same table,
different row.

**At most `r f0(100 * (E6b - E6a) / (E0b - E0a))`% of the famous rise in life
expectancy is a rise in how long a 65-year-old lives.** The rest is people who
used to die young and now do not.

> **A comparison has a denominator, the denominator is a choice, and a
> denominator chosen without thinking can reverse the sign of the answer.**

## What the at-65 series will not do

```{r sparse}
c(years_with_e65 = sum(!is.na(le$e65)),
  first_year     = min(le$year[!is.na(le$e65)]),
  last_year      = max(le$year[!is.na(le$e65)]),
  years_with_e0  = sum(!is.na(le$e0_annual)))
le$year[!is.na(le$e65)]
```

The at-birth series is annual back to 1900. **The at-65 series exists for
`r sum(!is.na(le$e65))` years only**, none of them before
`r min(le$year[!is.na(le$e65)])`. Nothing between the published years has been
interpolated here. The comparison in this step can be made for 1950 onward and
**cannot honestly be made for 1900** — which is a pity, because 1900 is where
the at-birth number looks most dramatic.

The two NCHS publications of life expectancy at birth also disagree with each
other:

```{r sources-disagree}
ov <- le[!is.na(le$e0_hus) & !is.na(le$e0_annual), ]
c(shared_years = nrow(ov),
  agree_exactly = sum(ov$e0_hus == ov$e0_annual),
  worst_gap = max(abs(ov$e0_hus - ov$e0_annual)))
ov[ov$e0_hus != ov$e0_annual, c("year", "e0_annual", "e0_hus")]
```

On `r sum(ov$e0_hus == ov$e0_annual)` of `r nrow(ov)` shared years they agree to
the tenth. On two they do not, and the pattern — one series monotone, the other
not — is what a transposed pair looks like. **This lab does not adjudicate it.
It reports it.**

# Step 5 — Entering older, or staying longer?

A rising median age has exactly two possible sources: **new members arrive
older**, or **sitting members stay longer**. For every member, at every
Congress, `age now = age on arriving + years since`. That is an identity, so it
survives averaging.

```{r decomp}
dec <- function(ch, y1, y2) {
  a <- ac[ac$chamber == ch & ac$congress == sitting(y1), ]
  b <- ac[ac$chamber == ch & ac$congress == sitting(y2), ]
  round(c(rise_in_mean_age = b$mean_age - a$mean_age,
          from_arriving_older = b$mean_entry_age - a$mean_entry_age,
          from_staying_longer = b$mean_years_since_entry -
                                a$mean_years_since_entry), 2)
}
rbind(Congress = dec("Congress", BY, EY),
      House    = dec("House",    BY, EY),
      Senate   = dec("Senate",   BY, EY))
```

```{r decomp-share}
DC <- dec("Congress", BY, EY); DH <- dec("House", BY, EY); DS <- dec("Senate", BY, EY)
round(c(Congress = 100 * DC[2] / DC[1],
        House    = 100 * DH[2] / DH[1],
        Senate   = 100 * DS[2] / DS[1]))
```

**`r f0(100 * DC[2] / DC[1])`% of the rise is people arriving older**, not
incumbents hanging on. And the chambers disagree: it is
`r f0(100 * DH[2] / DH[1])`% in the House and `r f0(100 * DS[2] / DS[1])`% in
the Senate, so **most of the Senate's ageing really is senators staying** —
which six-year terms and very high re-election rates would predict.

```{r decomp-eras}
rbind(`1950 to 1980` = dec("Congress", 1950, 1980),
      `1980 to 2025` = dec("Congress", 1980, 2025))
```

Congress got **younger** through the 1960s and 1970s, and did it by electing
younger people: arrival age fell
`r f1(abs(dec("Congress", 1950, 1980)[2]))` years while time in office barely
moved. Then the same lever went the other way.

# Step 6 — The median hides the tail

```{r tail}
q <- c("pct_65plus", "pct_70plus", "pct_80plus")
round(rbind(
  `Congress 1950` = unlist(ac[ac$chamber == "Congress" & ac$congress == sitting(BY), q]),
  `Congress 2025` = unlist(ac[ac$chamber == "Congress" & ac$congress == max(ac$congress), q]),
  `Senate 2025`   = unlist(ac[ac$chamber == "Senate"   & ac$congress == max(ac$congress), q])), 1)
```

```{r edges}
hs <- m[m$chamber %in% c("House", "Senate") & !is.na(m$age), ]
Q <- function(y) quantile(hs$age[hs$congress == sitting(y)], c(0.05, 0.5, 0.95))
round(rbind(`1950` = Q(BY), `2025` = Q(2025), `moved by` = Q(2025) - Q(BY)))
```

**The middle moved; the top moved further.** The share of Congress over 70 went
from
`r f1(ac$pct_70plus[ac$chamber == "Congress" & ac$congress == sitting(BY)])`% to
`r f1(ac$pct_70plus[ac$chamber == "Congress" & ac$congress == max(ac$congress)])`%,
and in the Senate it is now
`r f1(ac$pct_70plus[ac$chamber == "Senate" & ac$congress == max(ac$congress)])`%.
Between 1950 and 2025 the 5th percentile of age rose
`r f0(Q(2025)[1] - Q(BY)[1])` years, the median `r f0(Q(2025)[2] - Q(BY)[2])`,
and the 95th `r f0(Q(2025)[3] - Q(BY)[3])`. **The distribution did not slide. It
stretched, at the top.**

# Step 7 — Careers, and the censoring trap

```{r careers}
E2 <- et[!et$still_serving & !is.na(et$entry_age) & et$entry_year <= 1985, ]
E2$cohort <- cut(E2$entry_year, c(1789, 1849, 1909, 1949, 1986), right = FALSE,
                 labels = c("1789-1848", "1849-1908", "1909-1948", "1949-1985"))
do.call(rbind, lapply(levels(E2$cohort), function(g) {
  z <- E2[E2$cohort == g, ]
  data.frame(entered = g, careers = nrow(z),
             median_entry_age = median(z$entry_age),
             median_years_served = median(z$years_served),
             pct_20yrs_plus = round(100 * mean(z$years_served >= 20), 1))
}))
```

Notice what that table excluded: everybody still serving, and everybody who
entered after 1985.

```{r censoring}
c(careers_in_file   = nrow(et),
  still_serving     = sum(et$still_serving),
  entered_since_1986 = sum(et$entry_year >= 1986))
```

**A career you can measure is a career that is over.** Ask "how long do members
serve?" of everyone in the file and the recent cohorts look short — not because
service got shorter but because it has not finished. That is **right-censoring**,
and it is the commonest way a tenure statistic gets reported backwards.

```{r pipeline}
S <- et[et$chamber == "Senate" & !is.na(et$entry_age), ]
c(pct_with_prior_house_all_time = round(100 * mean(S$prior_house), 1),
  pct_with_prior_house_since_1981 =
    round(100 * mean(S$prior_house[S$entry_year >= 1981]), 1),
  mean_entry_age_with_house =
    round(mean(S$entry_age[S$entry_year >= 1981 &  S$prior_house]), 1),
  mean_entry_age_without =
    round(mean(S$entry_age[S$entry_year >= 1981 & !S$prior_house]), 1))
```

A popular explanation for senators arriving old is that they come up through the
House. They increasingly do —
`r f1(100 * mean(S$prior_house[S$entry_year >= 1981]))`% of those entering since
1981 — but **those senators arrive younger**, not older
(`r f1(mean(S$entry_age[S$entry_year >= 1981 & S$prior_house]))` against
`r f1(mean(S$entry_age[S$entry_year >= 1981 & !S$prior_house]))`). The pipeline
is not the cause.

# Step 8 — Who leaves the data before you can count them

```{r deaths}
died_in_office <- !is.na(hs$died) & hs$died >= hs$year & hs$died <= hs$year + 1
era <- ifelse(hs$year < 1900, "before 1900",
       ifelse(hs$year < 1950, "1900-1949",
       ifelse(hs$year < 2000, "1950-1999", "2000 and after")))
round(100 * tapply(died_in_office, factor(era, c("before 1900", "1900-1949",
                                                 "1950-1999", "2000 and after")),
                   mean), 2)
```

```{r deaths-total}
c(member_terms = length(died_in_office), ended_in_death = sum(died_in_office))
```

Members used to die in office several times as often as they do now. **Each of
those deaths removed an old member from every later Congress in this file**, so
the historical series is biased downward — in exactly the direction that makes
Congress look as though it used to prefer the young.

That bias *is* explained by longevity, and it belongs in the same argument as
Step 4, on the other side of the ledger.

# Step 9 — What you can now say

**You can say** that Congress is at or near its oldest recorded point, and that
the rise is recent rather than gradual.

**You can say that "people just live longer" does not account for it**, and how
much of it that argument could account for at most:
`r f1(E6b - E6a)` years against a `r f0(A1 - A0)`-year rise.

**You can say which mechanism dominates**, and that it differs by chamber:
`r f0(100 * DH[2] / DH[1])`% arrival age in the House,
`r f0(100 * DS[2] / DS[1])`% in the Senate.

**You cannot say why.** Nothing in this file identifies a cause.

**You cannot say it matters.** No measure of performance appears anywhere here.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Use the wrong denominator on purpose.** Redo Step 4 with `e75`
instead of `e65`, and then with `e0_annual`. Report all three answers and say
which one you would put in a news story, and why. There is a defensible case for
more than one.

**Option B — Pick a different window.** Step 4 uses 1950 to 2018 because that is
where both series exist. Redo it 1950-1980 and 1980-2018 separately. Does the
flip survive in both halves? If it does not, say what that means for the
headline claim.

**Option C — One state, or one party.** `members.csv` carries `state_abbrev` and
`party_code` (100 = Democrat, 200 = Republican). Compute median age by party
over time, or for the delegation of a state you care about. Small groups are
noisy; say something about how noisy before you interpret a gap.

**Option D — The decomposition, one chamber at a time.** Step 5 shows the House
and the Senate behaving differently. Find the Congress where they diverged and
say what was happening in American politics around then. The data will not tell
you; you have to bring that.

**Option E — Your own question.** `r n(nrow(m))` member-Congress records, every
Congress since 1789.

# What this lab cannot tell you

- **Why members arrive older or stay longer.** Primaries, money, safe seats,
  party recruitment, longer pre-political careers and the absence of any
  retirement rule are all consistent with these numbers. Nothing here separates
  them.
- **Whether age affects how anyone governs.** No performance measure appears in
  this file, and none of the figures should be read as if one did.
- **Anything about who dies in office**, beyond the raw rate. Members who died
  are missing from every later Congress, which flatters the historical series
  (Step 8), and this lab measures the size of that effect but cannot remove it.
- **Anything about candidates who lost.** Every person in this file won. Whether
  the electorate is choosing older candidates or being offered older candidates
  is invisible here, and it is the most important thing the data does not carry.
- **A single person's exact age.** `born` is a year (Step 1).

# On the discussion board

By **Friday, 11:59 p.m.**

- Step 4 changes one row of one table and reverses the conclusion. Write the
  sentence a journalist would honestly write about Congress and longevity, in
  one sentence, using both numbers.
- Step 5 says most of the rise is people arriving older, not incumbents staying.
  What would you expect to see in the data if the opposite were true?
- Step 8 shows that members used to die in office much more often. Does that
  make the historical comparison in Step 2 better or worse, and in which
  direction?

***

### Where the data came from

- **Members of Congress and presidents, 1789 to present** — Voteview,
  *Congressional Roll-Call Votes Database* (Lewis, Poole, Rosenthal, Boche,
  Rudkin and Sonnet).
  <https://voteview.com/static/data/out/members/HSall_members.csv>
  One row per member per Congress, with `born` and `died` as years. The same
  file the `dw-nominate` lab uses for something completely different, which is
  worth noticing: one well-built file answers many questions.
- **Life expectancy at birth, 1900-2018, annual** — National Center for Health
  Statistics, via CDC open data.
  <https://data.cdc.gov/api/views/w9j2-ggv5/rows.csv?accessType=DOWNLOAD>
- **Remaining life expectancy at 65 and at 75, selected years 1950-2018** —
  National Center for Health Statistics, *Health, United States, 2019*, Trend
  Table 4. <https://www.cdc.gov/nchs/data/hus/2019/004-508.pdf>
  Parsed from the published PDF, because this series is not in CDC's open-data
  catalogue, and checked against the at-birth series above.

The build is `data/build-data.R`.
