---
title: "Names as Data"
type: brief
subtitle: "The field every record about a person carries, and the three jobs it is asked to do"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("names-code.R")
```

```{r setup, include=FALSE}
```

Every administrative record about a person carries a name, and in many it is the only description there is. The voter registration has one, and so do the campaign filing, the court docket, the jury questionnaire and the traffic stop. None was written to describe anybody, so most hold no age, no sex and no race. They hold a name, because a clerk had to write something in the box.

So the name is asked to do work, and always the same three jobs:

1. **Identify.** Pick this person out from everybody else.
2. **Link.** Two records carry the same name, so they are the same person.
3. **Infer.** The file records no sex or race, so guess them from the name.

Each is a claim about how names are spread across the country, and nobody holding a voter file can check it from the voter file. Two administrative tabulations can. How far a name carries decides whether you are matched to the right registration, flagged as somebody else on a purge list, or counted into a disparity a court must rule on.

## The test

The first source is the Census Bureau's tabulation of the name field in the 2020 census. It appeared in April 2026 as *Frequently Occurring First Names in the 2020 Census* and its companion for last names (<https://www.census.gov/topics/population/genealogy/data/2020_names.html>). Together they hold `r n(F("first_names"))` first names covering `r n(F("first_people"))` people, and `r n(F("last_names"))` surnames covering `r n(F("last_people"))`.

The second is the Social Security Administration's national name file: one plain text table per year of birth from `r F("ssa_min")` to `r F("ssa_max")`, three columns wide, name and sex and count (<https://www.ssa.gov/oact/babynames/limits.html>).

These two are the right instruments because they count the field itself. Every other file in this book has a name column and no way to say how common a name is, which is what all three jobs depend on. Neither was made to study names. The agency writes down a name and a birth year because it cannot issue a card otherwise; the Bureau counts residents. Both tabulations are by-products, the kind of record [Section III](../admin-records-source/admin-records-source-brief.html) is about.

## The data

Each row of a census table is one name with a rank and a count. The first-name table adds `pct_female`; both add six shares, `pct_white`, `pct_black`, `pct_aian`, `pct_asian`, `pct_twoplus` and `pct_hispanic`. That is the whole record: no age, no state, no way to join a first name to a surname.

```{r sizes}
```

Three things there decide what can be asked of it. **A name is listed only if about a hundred people had it**, and everyone else, `r n(F("residual_first"))` people by first name, goes into a final row called ALL OTHER NAMES with an empty rank. **A first name was recorded for `r pc(F("first_coverage_pct"))`% of those enumerated.** The other `r n(F("no_first_name"))` are not a random draw; they sit wherever the census fell back on administrative records or a neighbour's answer. **And the counts have been noised**, by the same disclosure-avoidance machinery as the rest of that census.

Before any of it could be counted, the field had to be made. The unedited returns held about 6.0 million distinct first names, and editing removed roughly 800,000. The largest problem was not misspelling. Over ten million entries were not names at all: PERSON, CHILD, DAUGHTER, TENANT, MRS SMITH. Left alone, PERSON would have been one of the country's most common first names.

Two repairs travel into every other file you will ever join. Punctuation, numbers, spaces and suffixes were removed, so MICHAEL, MI CH AEL and MICHAEL III all count as MICHAEL. That is right for a frequency table, and it is what a matching program has to do too, which most do not. And Spanish names holding Ñ reached the Bureau with the character already turned into a question mark, so PEÑA is published as PENA. The damage happened upstream, and the repair is a second decision on top of it.

The noise was run independently into the sex table and the race table, so **`r n(F("disagree_n"))` of `r n(F("first_names"))` first names carry a different total in the two.** Neither is wrong. No fact about how many Americans are named MICHAEL survives publication, because publication is what added the noise.

```{r disagree-tab}
```

The Social Security file has different limits. It counts card applications by the birth year the applicant gave, and drops any name given to fewer than five babies in a year. Cards were first issued in `r F("ssa_card_year")`, so anyone born earlier appears only if they lived long enough to apply.

> **Before you read on, commit to a number.** The census publishes
> `r n(F("last_names"))` surnames. **How many of them, the most common ones,
> would you need to cover half of everybody in the file?** Write it down, then
> do the same for first names, of which there are `r n(F("first_names"))`.

## What it says about democracy

**The first job is to identify, and a name is a distribution rather than a label.**

```{r top-tab}
```

It takes `r n(F("half_first"))` first names to cover half the country, and `r n(F("half_last"))` surnames.

```{r curve-d3, results='asis', eval=knitr::is_html_output()}
```

```{r curve-static, eval=!knitr::is_html_output(), fig.height=4.3}
```

**Figure 1. How much of the country the most common names cover.** Rank runs along the bottom on a log scale, so the first thousand names and the hundred-thousandth are both readable. A cumulative curve fits the question, which is not what one name is worth but how fast coverage piles up. Neither curve reaches the top, and that is the threshold rather than a drawing error: every published first name covers only `r pc(max(curve$cum_pct[curve$which == "first name"]))`% of the file, every published surname `r pc(max(curve$cum_pct[curve$which == "last name"]))`%.

```{r cover-tab}
```

**First names are the more concentrated of the two, and by a lot.** A thousand first names cover `r pc(cover$first_pct[cover$top_n == 1000])`% of Americans where a thousand surnames cover `r pc(cover$last_pct[cover$top_n == 1000])`%. That runs against the intuition that first names are where individuality lives, and the cause is plain. A surname is inherited from a stock the country's whole immigration history filled up, and a first name is chosen from fashion, a much smaller menu.

**The second job is to link, and the number underneath it is small.**

```{r coll-tab}
```

A first name and a surname together are worth about **one in `r n(CL("two people share both, if the two are unrelated"))`**. That is the figure a matching program quietly relies on, and it comes from multiplying the two separate chances. That step is legitimate only if a first name tells you nothing about a surname, and it obviously does. A floor on the error can be measured: correlate the two through the six census race categories alone, and a shared full name gets `r pc(100 * (FN("inflate_race") - 1), 0)`% more likely. The real correlation runs through national origin and generation, which is why NGUYEN and GARCIA do not draw first names from the same pot.

Each is the chance for **one pair of people**, not the chance of a match somewhere in a list, which depends on how long the list is. That is the [false matches brief](../../03-elections/false-matches/false-matches-brief.html).

```{r both-tab}
```

The two fields are not even distinct. `r n(F("both_names"))` names appear in both tables, and **`r pc(F("both_pct_first"))`% of Americans have a first name that is also somebody's surname**. That is why a swapped-column bug is so hard to see. Exchange `first` and `last` for part of a load, and three-quarters of the affected rows still hold an ordinary name.

**The third job is to infer, and it is the one most used and least measured.** Assign every person the majority sex of their first name and you are right **`r pc(F("maj_right_pct"), 2)`% of the time**. That is **`r n(F("maj_wrong"))` people wrong**, a population the size of Massachusetts.

```{r unisex-tab}
```

Those are the most-used first names closest to an even split. JORDAN is `r pc(uni$pct_female[uni$name == "JORDAN"])`% female and TAYLOR is `r pc(uni$pct_female[uni$name == "TAYLOR"])`%, so a majority rule errs by the same amount in opposite directions on the two. The census item is **sex**, asked as two boxes, and nothing here speaks to gender. The table is also a 2020 snapshot, so a name whose usage shifted is described by an average across the shift.

Race is the job the field is asked to do most, and the two halves of a name are not equally good at it. Put both on one scale: how much of your uncertainty about a person's race does the name remove?

```{r race-tab}
```

**A surname removes `r pc(RI("last name", "pct_of_uncertainty_removed"))`% of the uncertainty; a first name removes `r pc(RI("first name", "pct_of_uncertainty_removed"))`%.** That is the whole justification for surname-based inference, measured rather than assumed. It is not enough alone, because a name cannot outrun the base rate. The file's most distinctive surname for American Indians is only `r pc(rext$pct_in_group[rext$which == "last name" & rext$group == "Am. Indian"])`% American Indian, because the group is `r pc(rext$group_share_of_file[rext$which == "last name" & rext$group == "Am. Indian"])`% of the file. Any method has to bring something else, usually where the person lives. That is Section IV's subject: the [surnames brief](../../06-putting-data-together/surnames/surnames-brief.html) builds the method from this same table.

**All three jobs lean on a count of everyone alive, and such a count is slow.** The census file is a stock, how many there are now; the birth file is a flow, how many arrived this year. Amazon put the Echo on sale in `r F("echo_year")`, the best year ALEXA has ever had at rank `r F("alexa_best_rank")`. By `r F("ssa_max")` it went to `r n(F("alexa_last_n"))` girls in the whole country, down `r pc(F("alexa_drop_pct"), 0)`%.

```{r flow-stock, fig.height=3.5}
```

**Figure 2. ALEXA, both sexes, from the Social Security birth file.** The left panel is how many were given the name each year; the right is the same numbers added up, which is what a census counts. Two panels rather than two lines on one pair of axes, because a count per year and a running total share no scale.

A census taken in 2020 would have shown the name **rising by `r pc(F("alexa_stock_growth"), 0)`%**. Everyone given it before the collapse was still alive to be counted, and `r n(F("alexa_born_decade"))` more were added between the censuses. **A stock absorbs a shock and a flow records it**, and that is a rule about records rather than names. Registered voters, houses and licences are counted as stocks, and each shows a change years late and smaller than it was.

## What this brief cannot tell you

**It cannot put a first name and a surname together.** They are tabulated separately, so every statement above about full names rests on multiplying two tables. The `r pc(100 * (FN("inflate_race") - 1), 0)`% correction is a floor on that error, not a measure of it.

**It cannot tell you where anybody is.** Names are not spread evenly and everything here is national, which is why the surname method needs a second file.

**It is not a source for the racial composition of the country**, and the Bureau prints that warning on both products. The shares above describe the people whose names could be used, `r pc(F("first_coverage_pct"))`% of the census.

**It cannot see a name that arrived recently.** To be published from 2020 a first name had to occur a hundred times in 2010 as well, so the names doing something interesting in the 2010s are missing by design.

**And these are edited names.** MICHAEL III is MICHAEL and PEÑA is PENA, which is right for a frequency table and makes the file a poor guide to the raw name field in the records you will actually join.

## What you should have learned

- A name is a distribution, not a label: `r n(F("half_first"))` first names cover half the country, and `r n(F("half_last"))` surnames, out of `r n(F("first_names"))` and `r n(F("last_names"))` published.
- First names are more concentrated than surnames, `r pc(cover$first_pct[cover$top_n == 1000])`% against `r pc(cover$last_pct[cover$top_n == 1000])`% for the top thousand, because a first name is chosen from fashion and a surname inherited.
- A full name is worth about one in `r n(CL("two people share both, if the two are unrelated"))` for a single pair, and that is optimistic: it assumes the two halves are unrelated when they are not.
- Guessing sex from a first name is `r pc(F("maj_right_pct"), 2)`% right and `r n(F("maj_wrong"))` people wrong. A surname removes `r pc(RI("last name", "pct_of_uncertainty_removed"))`% of the uncertainty about race; a first name removes `r pc(RI("first name", "pct_of_uncertainty_removed"))`%.
- A stock absorbs a shock and a flow records it: ALEXA fell `r pc(F("alexa_drop_pct"), 0)`% in births while the census count rose `r pc(F("alexa_stock_growth"), 0)`%.

## Extensions

None is answered above, and every one can be answered from the tables below.

- `explore_first.csv` gives every common first name a `pct_female`. Sort by it and read the names nearest 50. What happens to those people when a file guesses sex from a name?
- `explore_last.csv` has `pct_white` through `pct_hispanic` beside `people`. Sort by `pct_hispanic`, then by `people`. How many Americans do the strongest surnames cover?
- `race_info.csv` reports `bits_before`, `bits_after` and `pct_of_uncertainty_removed` for both kinds of name. Compare the rows. Is either enough to act on alone?
- `ssa_totals.csv` has `census_2020` beside `born_all` and `median_birth_year`. Pick a name once common and now rare, and say whether it is dying out or only ageing.
- **Stretch.** The Social Security file publishes a table for every state. Download two and compare one name in both, which is the geography this file cannot give you.

## Sources

**Data**

- U.S. Census Bureau. *Frequently Occurring First Names in the 2020 Census*, by
  sex and by race and Hispanic origin, and *Frequently Occurring Last Names in
  the 2020 Census*. Spreadsheets, no account required.
  <https://www.census.gov/topics/population/genealogy/data/2020_names.html>
- Comenetz, Joshua. *First Name Data From the 2020 Census*, 2020 Census Briefs
  C2020BR-13, April 2026, and *Last Name Data From the 2020 Census*,
  C2020BR-14. Every statement above about editing, generic entries, the
  question mark and the noise is theirs.
  <https://www2.census.gov/library/publications/decennial/2020/c2020br-13.pdf>
- Social Security Administration. *Beyond the Top 1000 Names*, national data,
  one file per year of birth, `r F("ssa_min")` to `r F("ssa_max")`.
  <https://www.ssa.gov/oact/babynames/limits.html>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they open
in a spreadsheet, and they are yours to take further.

[`both_ways.csv`](data/derived/both_ways.csv), [`collision.csv`](data/derived/collision.csv), [`cover.csv`](data/derived/cover.csv), [`curve.csv`](data/derived/curve.csv), [`disagreement_top.csv`](data/derived/disagreement_top.csv), [`explore_first.csv`](data/derived/explore_first.csv), [`explore_last.csv`](data/derived/explore_last.csv), [`race_extremes.csv`](data/derived/race_extremes.csv), [`race_info.csv`](data/derived/race_info.csv), [`ssa_alexa.csv`](data/derived/ssa_alexa.csv), [`ssa_totals.csv`](data/derived/ssa_totals.csv), [`top25.csv`](data/derived/top25.csv), [`unisex.csv`](data/derived/unisex.csv)

**Related work**

- Word, D. L., Coleman, C. D., Nunziata, R. and Kominski, R. (2008).
  "Demographic Aspects of Surnames from Census 2000." U.S. Census Bureau
  technical documentation, which these products descend from.
- Elliott, M. N. et al. (2009). "Using the Census Bureau's surname list to
  improve estimates of race/ethnicity." *Health Services and Outcomes Research
  Methodology* 9: 69–83, the method the
  [surnames](../../06-putting-data-together/surnames/surnames-brief.html) and
  [BISG](../../06-putting-data-together/bisg-check/bisg-check-brief.html)
  briefs implement by hand.

```{r ai-prompt, echo=FALSE, results="asis"}
```
