---
title: "Names as Data"
type: brief
subtitle: "The field every record about a person carries, and the three jobs it is asked to do"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("names-code.R")
```

```{r setup, include=FALSE}
```

Every administrative record about a person carries a name, and in many it is the only description there is. The voter registration has one, and so do the campaign filing, the court docket, the jury questionnaire and the traffic stop. None was written to describe anybody, so most hold no age, no sex and no race. They hold a name, because a clerk had to write something in the box.

So the name is asked to do work, and always the same three jobs:

1. **Identify.** Pick this person out from everybody else.
2. **Link.** Two records carry the same name, so they are the same person.
3. **Infer.** The file records no sex or race, so guess them from the name.

Each is a claim about how names are spread across the country, and nobody holding a voter file can check it from the voter file. Two administrative tabulations can. How far a name carries decides whether you are matched to the right registration, flagged as somebody else on a purge list, or counted into a disparity a court must rule on.

## The test

The first source is the Census Bureau's tabulation of the name field in the 2020 census. It appeared in April 2026 as *Frequently Occurring First Names in the 2020 Census* and its companion for last names, downloadable as spreadsheets at <https://www.census.gov/topics/population/genealogy/data/2020_names.html>.^[The Bureau's own account of how the tables were made: Joshua Comenetz, *First Name Data From the 2020 Census*, Census Brief C2020BR-13 (April 2026), <https://www2.census.gov/library/publications/decennial/2020/c2020br-13.pdf>, and *Last Name Data From the 2020 Census*, C2020BR-14, <https://www2.census.gov/library/publications/decennial/2020/c2020br-14.pdf>.] Together they hold `r n(F("first_names"))` first names covering `r n(F("first_people"))` people, and `r n(F("last_names"))` surnames covering `r n(F("last_people"))`.

The second is the Social Security Administration's national name file: one plain text table per year of birth from `r F("ssa_min")` to `r F("ssa_max")`, three columns wide, name and sex and count (<https://www.ssa.gov/oact/babynames/limits.html>).

These two are the right instruments because they count the field itself. Every other file in this book has a name column and no way to say how common a name is, which is what all three jobs depend on. Neither was made to study names. The agency writes down a name and a birth year because it cannot issue a card otherwise; the Bureau counts residents. Both tabulations are by-products, the kind of record [Section III](../admin-records-source/admin-records-source-brief.html) is about.

## The data

Twenty-six tables come out of this chapter, two files' worth of source: the
2020 Census surname tables, and the Social Security Administration's name-by-year
registrations back to 1880. Sizes are rows by columns, read off the files.

| Table | What it holds | Size |
|---|---|---|
| [`top25.csv`](data/derived/top25.csv) | The most common first names and last names, ranked, with count and share. | `r dd_dim("top25.csv")` |
| [`cover.csv`](data/derived/cover.csv) | What share of the country the top 10, top N first and last names cover. | `r dd_dim("cover.csv")` |
| [`curve.csv`](data/derived/curve.csv) | The cumulative-share curve: rank against running percentage of the population, for first and last names. | `r dd_dim("curve.csv")` |
| [`collision.csv`](data/derived/collision.csv) | Birthday-problem probabilities: how likely two people share a name, from the same counts. | `r dd_dim("collision.csv")` |
| [`concentration.csv`](data/derived/concentration.csv) | How many names the file actually publishes against how many people that leaves in the unpublished residual, with information-theoretic bounds on what a name still reveals. | `r dd_dim("concentration.csv")` |
| [`race_info.csv`](data/derived/race_info.csv) | How much a first or last name narrows down a person's race: bits of uncertainty removed, and how many names cross 90% modal accuracy. | `r dd_dim("race_info.csv")` |
| [`race_extremes.csv`](data/derived/race_extremes.csv) | The single most race-concentrated first name and last name in each of the six published race/ethnicity groups. | `r dd_dim("race_extremes.csv")` |
| [`explore_first.csv`](data/derived/explore_first.csv) | Every published first name with its full race and sex breakdown -- the table the name-lookup figure reads. | `r dd_dim("explore_first.csv")` |
| [`explore_last.csv`](data/derived/explore_last.csv) | The same, for last names, without the sex columns the Bureau does not publish for surnames. | `r dd_dim("explore_last.csv")` |
| [`both_ways.csv`](data/derived/both_ways.csv) | Names that are common as both a first name and a last name, with each count side by side. | `r dd_dim("both_ways.csv")` |
| [`disagreement_top.csv`](data/derived/disagreement_top.csv) | The names whose count differs most between the Bureau's sex-by-name table and its race-by-name table -- two tables about the same names that do not perfectly agree. | `r dd_dim("disagreement_top.csv")` |
| [`disagreement.csv`](data/derived/disagreement.csv) | That same sex-table-vs-race-table gap, as a distribution: how many names differ by how much. | `r dd_dim("disagreement.csv")` |
| [`negatives.csv`](data/derived/negatives.csv) | How often the Bureau's privacy noise pushed a published cell below zero, by table -- the sign of the disclosure-avoidance system at work. | `r dd_dim("negatives.csv")` |
| [`negatives_top.csv`](data/derived/negatives_top.csv) | The specific names hit hardest by a negative cell: published counts next to the Bureau's own noise-corrected research file. | `r dd_dim("negatives_top.csv")` |
| [`unisex.csv`](data/derived/unisex.csv) | Names closest to an even sex split, ranked by how close to 50/50 they fall. | `r dd_dim("unisex.csv")` |
| [`sex_bands.csv`](data/derived/sex_bands.csv) | How many names cross each threshold of sex skew -- the distribution unisex.csv is drawn from. | `r dd_dim("sex_bands.csv")` |
| [`sex_scatter.csv`](data/derived/sex_scatter.csv) | One row per name: total people and percent female, the scatter the sex-balance figure plots. | `r dd_dim("sex_scatter.csv")` |
| [`ssa_totals.csv`](data/derived/ssa_totals.csv) | Every name in the Social Security birth-registration file since 1880, against its 2020 census count and rank -- the table linking the two eras. | `r dd_dim("ssa_totals.csv")` |
| [`ssa_alexa.csv`](data/derived/ssa_alexa.csv) | One name's (Alexa's) year-by-year SSA registrations, rank and cumulative births -- the worked example of a name's rise and fall. | `r dd_dim("ssa_alexa.csv")` |
| [`ssa_births.csv`](data/derived/ssa_births.csv) | Total SSA-registered births by year and sex, back to 1880 -- the denominator every other SSA table is a share of. | `r dd_dim("ssa_births.csv")` |
| [`ssa_coverage.csv`](data/derived/ssa_coverage.csv) | How completely SSA birth registration actually covers births, by year -- registered against the true count, where known. | `r dd_dim("ssa_coverage.csv")` |
| [`ssa_explore.csv`](data/derived/ssa_explore.csv) | Every published name's year-by-year SSA count by sex -- the raw series the browser's name explorer reads. | `r dd_dim("ssa_explore.csv")` |
| [`ssa_newcomers.csv`](data/derived/ssa_newcomers.csv) | Names with no SSA registrations before a given year: when they first appear, and how many people carry them since. | `r dd_dim("ssa_newcomers.csv")` |
| [`ssa_risers.csv`](data/derived/ssa_risers.csv) | Names whose frequency has climbed most since an earlier peak, then and now, with the peak year. | `r dd_dim("ssa_risers.csv")` |
| [`ssa_shock.csv`](data/derived/ssa_shock.csv) | Names whose SSA registrations spiked in a specific year, tied to the event that year is named for. | `r dd_dim("ssa_shock.csv")` |
| [`ssa_vs_census.csv`](data/derived/ssa_vs_census.csv) | For a name, SSA births carried forward to 2020 against the actual 2020 census count -- how much of a birth cohort the census still finds. | `r dd_dim("ssa_vs_census.csv")` |

Two more files here are not data about names at all -- they are how this chapter checked its own work: [`facts.csv`](data/derived/facts.csv), the individual numbers the text above quotes, and [`checks.csv`](data/derived/checks.csv), the arithmetic verified before anything was drawn.

Each row of a census table is one name with a rank and a count. The first-name table adds `pct_female`; both add six shares, `pct_white`, `pct_black`, `pct_aian`, `pct_asian`, `pct_twoplus` and `pct_hispanic`. That is the whole record: no age, no state, no way to join a first name to a surname.

```{r sizes}
```

Three things there decide what can be asked of it. **A name is listed only if about a hundred people had it**, and everyone else, `r n(F("residual_first"))` people by first name, goes into a final row called ALL OTHER NAMES with an empty rank. **A first name was recorded for `r pc(F("first_coverage_pct"))`% of those counted.** The other `r n(F("no_first_name"))` are not a random draw; they sit wherever the census fell back on administrative records or a neighbour's answer. **And the counts have had noise added to them on purpose.** The Bureau's **disclosure avoidance** system nudges every published 2020 figure by a small random amount so that no individual can be picked out of a table,^[Census Bureau, "Disclosure Avoidance and the 2020 Census," <https://www.census.gov/programs-surveys/decennial-census/decade/2020/planning-management/process/disclosure-avoidance.html>.] and the name tables went through it like everything else.

Before any of it could be counted, the field had to be made. The unedited returns held about 6.0 million distinct first names, and editing removed roughly 800,000.^[The editing figures, the generic entries and the question-mark problem in the next paragraph are all reported in Comenetz, *First Name Data From the 2020 Census*, C2020BR-13.] The largest problem was not misspelling. Over ten million entries were not names at all: PERSON, CHILD, DAUGHTER, TENANT, MRS SMITH. Left alone, PERSON would have been one of the country's most common first names.

Two repairs travel into every other file you will ever join. Punctuation, numbers, spaces and suffixes were removed, so MICHAEL, MI CH AEL and MICHAEL III all count as MICHAEL. That is right for a frequency table, and it is what a matching program has to do too, which most do not. And Spanish names holding Ñ reached the Bureau with the character already turned into a question mark, so PEÑA is published as PENA. The damage happened upstream, and the repair is a second decision on top of it.

The noise was run independently into the sex table and the race table, so **`r n(F("disagree_n"))` of `r n(F("first_names"))` first names carry a different total in the two.** Neither is wrong. No fact about how many Americans are named MICHAEL survives publication, because publication is what added the noise.

```{r disagree-tab}
```

The Social Security file has different limits. It counts card applications by the birth year the applicant gave, and drops any name given to fewer than five babies in a year. Cards were first issued in `r F("ssa_card_year")`, so anyone born earlier appears only if they lived long enough to apply.

> **Before you read on, commit to a number.** The census publishes
> `r n(F("last_names"))` surnames. **How many of them, the most common ones,
> would you need to cover half of everybody in the file?** Write it down, then
> do the same for first names, of which there are `r n(F("first_names"))`.

## What it says about democracy

**The first job is to identify, and a name is a distribution rather than a label.**

```{r top-tab}
```

It takes `r n(F("half_first"))` first names to cover half the country, and `r n(F("half_last"))` surnames.

```{r curve-d3, results='asis', eval=knitr::is_html_output()}
```

```{r curve-static, eval=!knitr::is_html_output(), fig.height=4.3}
```

**Figure 1. How much of the country the most common names cover.** Rank runs along the bottom on a log scale, so the first thousand names and the hundred-thousandth are both readable. A cumulative curve fits the question, which is not what one name is worth but how fast coverage piles up. Neither curve reaches the top, and that is the threshold rather than a drawing error: every published first name covers only `r pc(max(curve$cum_pct[curve$which == "first name"]))`% of the file, every published surname `r pc(max(curve$cum_pct[curve$which == "last name"]))`%.

```{r cover-tab}
```

**First names are the more concentrated of the two, and by a lot.** A thousand first names cover `r pc(cover$first_pct[cover$top_n == 1000])`% of Americans where a thousand surnames cover `r pc(cover$last_pct[cover$top_n == 1000])`%. That runs against the intuition that first names are where individuality lives, and the cause is plain. A surname is inherited from a stock the country's whole immigration history filled up, and a first name is chosen from fashion, a much smaller menu.

**The second job is to link, and the number underneath it is small.**

```{r coll-tab}
```

Pick two Americans at random. The chance they share a first name is one in `r n(CL("two people share a first name"))`, found by adding up, name by name, the chance that both drew that name; the chance they share a surname is one in `r n(CL("two people share a last name"))`. If the two halves of a name had nothing to do with each other, the chance of sharing both would be the two multiplied: `r n(CL("two people share a first name"))` × `r n(CL("two people share a last name"))` ≈ `r n(CL("two people share both, if the two are unrelated"))`. So a first name and a surname together are worth about **one in `r n(CL("two people share both, if the two are unrelated"))`**. That is the figure a matching program quietly relies on, and the multiplying step is legitimate only if a first name tells you nothing about a surname, which it obviously does. A floor on the error can be measured: let the two halves be related only through the six race categories the census publishes, so that a GARCIA is likelier to be a JOSE than a surname picked at random, and a shared full name gets `r pc(100 * (FN("inflate_race") - 1), 0)`% more likely. The real correlation runs through national origin and generation, which is why NGUYEN and GARCIA do not draw first names from the same pot.

Each is the chance for **one pair of people**, not the chance of a match somewhere in a list, which depends on how long the list is. That is the [false matches brief](../../03-elections/false-matches/false-matches-brief.html).

```{r both-tab}
```

The two fields are not even distinct. `r n(F("both_names"))` names appear in both tables, and **`r pc(F("both_pct_first"))`% of Americans have a first name that is also somebody's surname**. That is why a swapped-column bug is so hard to see. Exchange `first` and `last` for part of a load, and three-quarters of the affected rows still hold an ordinary name.

**The third job is to infer, and it is the one most used and least measured.** Assign every person the majority sex of their first name and you are right **`r pc(F("maj_right_pct"), 2)`% of the time**. That is **`r n(F("maj_wrong"))` people wrong**, a population the size of Massachusetts.^[Massachusetts counted 7,029,917 people in the 2020 census: U.S. Census Bureau, state population totals, <https://www.census.gov/data/tables/time-series/demo/popest/2020s-state-total.html>.]

```{r unisex-tab}
```

Those are the most-used first names closest to an even split. JORDAN is `r pc(uni$pct_female[uni$name == "JORDAN"])`% female and TAYLOR is `r pc(uni$pct_female[uni$name == "TAYLOR"])`%, so a majority rule errs by the same amount in opposite directions on the two. The census item is **sex**, asked as two boxes, and nothing here speaks to gender. The table is also a 2020 snapshot, so a name whose usage shifted is described by an average across the shift.

Race is the job the field is asked to do most, and the two halves of a name are not equally good at it. Put both on one scale: how much of your uncertainty about a person's race does the name remove? The table measures that in bits, the unit information theory uses, where one bit is the uncertainty in a single fair coin toss.

```{r race-tab}
```

**A surname removes `r pc(RI("last name", "pct_of_uncertainty_removed"))`% of the uncertainty; a first name removes `r pc(RI("first name", "pct_of_uncertainty_removed"))`%.** That is the whole justification for surname-based inference, measured rather than assumed. One surname shows what the average is made of. The file counts `r n(S1$people)` people named `r S1$name`, the country's most common surname, and says `r pc(S1$pct_white)`% of them are white, `r pc(S1$pct_black)`% Black and `r pc(S1$pct_hispanic)`% Hispanic. Guess "white" for every `r S1$name` and you are right about two times in three. The name has told you something, since Black Americans are `r pc(rext$group_share_of_file[rext$which == "last name" & rext$group == "Black"])`% of the file and `r pc(S1$pct_black)`% of the `r S1$name`s, and it has left most of the uncertainty where it was. `r rext$name[rext$which == "last name" & rext$group == "Black"]` is the other end of the scale: `r pc(rext$pct_in_group[rext$which == "last name" & rext$group == "Black"])`% of the `r n(rext$people[rext$which == "last name" & rext$group == "Black"])` people carrying it are Black. Even that is not enough alone, because a name cannot outrun the **base rate**, the share of the whole file a group makes up. The file's most distinctive surname for American Indians is only `r pc(rext$pct_in_group[rext$which == "last name" & rext$group == "Am. Indian"])`% American Indian, because the group is `r pc(rext$group_share_of_file[rext$which == "last name" & rext$group == "Am. Indian"])`% of the file. Any method has to bring something else, usually where the person lives. That is Section IV's subject: the [surnames brief](../../06-putting-data-together/surnames/surnames-brief.html) builds the method from this same table.

**All three jobs lean on a count of everyone alive, and such a count is slow.** The census file is a stock, how many there are now; the birth file is a flow, how many arrived this year. Amazon put the Echo, the speaker one addresses as "Alexa," on general sale in `r F("echo_year")`,^[Amazon, "Amazon Echo Now Available to All Customers," press release, 23 June 2015, <https://press.aboutamazon.com/2015/6/amazon-echo-now-available-to-all-customers>.] the best year ALEXA has ever had at rank `r F("alexa_best_rank")`. By `r F("ssa_max")` it went to `r n(F("alexa_last_n"))` girls in the whole country, down `r pc(F("alexa_drop_pct"), 0)`%.

```{r flow-stock, fig.height=3.5}
```

**Figure 2. ALEXA, both sexes, from the Social Security birth file.** The left panel is how many were given the name each year; the right is the same numbers added up, which is what a census counts. Two panels rather than two lines on one pair of axes, because a count per year and a running total share no scale.

A census taken in 2020 would have shown the name **rising by `r pc(F("alexa_stock_growth"), 0)`%**. Everyone given it before the collapse was still alive to be counted, and `r n(F("alexa_born_decade"))` more were added between the censuses. **A stock absorbs a shock and a flow records it**, and that is a rule about records rather than names. Registered voters, houses and licences are counted as stocks, and each shows a change years late and smaller than it was.

## What this brief cannot tell you

**It cannot put a first name and a surname together.** They are tabulated separately, so every statement above about full names rests on multiplying two tables. The `r pc(100 * (FN("inflate_race") - 1), 0)`% correction is a floor on that error, not a measure of it.

**It cannot tell you where anybody is.** Names are not spread evenly and everything here is national, which is why the surname method needs a second file.

**It is not a source for the racial composition of the country**, and the Bureau prints that warning on both products. The shares above describe the people whose names could be used, `r pc(F("first_coverage_pct"))`% of the census.

**It cannot see a name that arrived recently.** To be published from 2020 a first name had to occur a hundred times in 2010 as well, so the names doing something interesting in the 2010s are missing by design.

**And these are edited names.** MICHAEL III is MICHAEL and PEÑA is PENA, which is right for a frequency table and makes the file a poor guide to the raw name field in the records you will actually join.

## What you should have learned

- A name is a distribution, not a label: `r n(F("half_first"))` first names cover half the country, and `r n(F("half_last"))` surnames, out of `r n(F("first_names"))` and `r n(F("last_names"))` published.
- First names are more concentrated than surnames, `r pc(cover$first_pct[cover$top_n == 1000])`% against `r pc(cover$last_pct[cover$top_n == 1000])`% for the top thousand, because a first name is chosen from fashion and a surname inherited.
- A full name is worth about one in `r n(CL("two people share both, if the two are unrelated"))` for a single pair, and that is optimistic: it assumes the two halves are unrelated when they are not.
- Guessing sex from a first name is `r pc(F("maj_right_pct"), 2)`% right and `r n(F("maj_wrong"))` people wrong. A surname removes `r pc(RI("last name", "pct_of_uncertainty_removed"))`% of the uncertainty about race; a first name removes `r pc(RI("first name", "pct_of_uncertainty_removed"))`%.
- A stock absorbs a shock and a flow records it: ALEXA fell `r pc(F("alexa_drop_pct"), 0)`% in births while the census count rose `r pc(F("alexa_stock_growth"), 0)`%.

## Extensions

None is answered above, and every one can be answered from the tables below.

- `explore_first.csv` gives every common first name a `pct_female`. Sort by it and read the names nearest 50. What happens to those people when a file guesses sex from a name?
- `explore_last.csv` has `pct_white` through `pct_hispanic` beside `people`. Sort by `pct_hispanic`, then by `people`. How many Americans do the strongest surnames cover?
- `race_info.csv` reports `bits_before`, `bits_after` and `pct_of_uncertainty_removed` for both kinds of name. Compare the rows. Is either enough to act on alone?
- `ssa_totals.csv` has `census_2020` beside `born_all` and `median_birth_year`. Pick a name once common and now rare, and say whether it is dying out or only ageing.
- **Stretch.** The Social Security file publishes a table for every state. Download two and compare one name in both, which is the geography this file cannot give you.

## Sources

**Data**

- U.S. Census Bureau. *Frequently Occurring First Names in the 2020 Census*, by
  sex and by race and Hispanic origin, and *Frequently Occurring Last Names in
  the 2020 Census*. Spreadsheets, no account required.
  <https://www.census.gov/topics/population/genealogy/data/2020_names.html>
- Comenetz, Joshua. *First Name Data From the 2020 Census*, 2020 Census Briefs
  C2020BR-13, April 2026, and *Last Name Data From the 2020 Census*,
  C2020BR-14. Every statement above about editing, generic entries, the
  question mark and the noise is theirs.
  <https://www2.census.gov/library/publications/decennial/2020/c2020br-13.pdf>
- Social Security Administration. *Beyond the Top 1000 Names*, national data,
  one file per year of birth, `r F("ssa_min")` to `r F("ssa_max")`.
  <https://www.ssa.gov/oact/babynames/limits.html>

**The data itself**

All twenty-six tables are described in [The data](#the-data) at the top of
this brief. They are plain CSV, they open in a spreadsheet, and they are yours
to take further: [`top25.csv`](data/derived/top25.csv), [`cover.csv`](data/derived/cover.csv), [`curve.csv`](data/derived/curve.csv), [`collision.csv`](data/derived/collision.csv), [`concentration.csv`](data/derived/concentration.csv), [`race_info.csv`](data/derived/race_info.csv), [`race_extremes.csv`](data/derived/race_extremes.csv), [`explore_first.csv`](data/derived/explore_first.csv), [`explore_last.csv`](data/derived/explore_last.csv), [`both_ways.csv`](data/derived/both_ways.csv), [`disagreement_top.csv`](data/derived/disagreement_top.csv), [`disagreement.csv`](data/derived/disagreement.csv), [`negatives.csv`](data/derived/negatives.csv), [`negatives_top.csv`](data/derived/negatives_top.csv), [`unisex.csv`](data/derived/unisex.csv), [`sex_bands.csv`](data/derived/sex_bands.csv), [`sex_scatter.csv`](data/derived/sex_scatter.csv), [`ssa_totals.csv`](data/derived/ssa_totals.csv), [`ssa_alexa.csv`](data/derived/ssa_alexa.csv), [`ssa_births.csv`](data/derived/ssa_births.csv), [`ssa_coverage.csv`](data/derived/ssa_coverage.csv), [`ssa_explore.csv`](data/derived/ssa_explore.csv), [`ssa_newcomers.csv`](data/derived/ssa_newcomers.csv), [`ssa_risers.csv`](data/derived/ssa_risers.csv), [`ssa_shock.csv`](data/derived/ssa_shock.csv), [`ssa_vs_census.csv`](data/derived/ssa_vs_census.csv).

**Related work**

- Word, D. L., Coleman, C. D., Nunziata, R. and Kominski, R. (2008).
  "Demographic Aspects of Surnames from Census 2000." U.S. Census Bureau
  technical documentation, which these products descend from.
  <https://www2.census.gov/topics/genealogy/2010surnames/surnames.pdf>
- Comenetz, J. (2026). *First Name Data From the 2020 Census* (C2020BR-13) and
  *Last Name Data From the 2020 Census* (C2020BR-14). Census Briefs.
  <https://www2.census.gov/library/publications/decennial/2020/c2020br-14.pdf>
- Elliott, M. N. et al. (2009). "Using the Census Bureau's surname list to
  improve estimates of race/ethnicity." *Health Services and Outcomes Research
  Methodology* 9: 69–83, <https://link.springer.com/article/10.1007/s10742-009-0047-1>, the method the
  [surnames](../../06-putting-data-together/surnames/surnames-brief.html) and
  [BISG](../../06-putting-data-together/bisg-check/bisg-check-brief.html)
  briefs implement by hand.

```{r ai-prompt, echo=FALSE, results="asis"}
```
