---
title: "Key: A Country Nobody Lives In"
subtitle: "84-355 Democracy's Data | Demographics | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
D <- "data"
rd <- function(f, ...) read.csv(file.path(D, f), stringsAsFactors = FALSE, ...)

fc  <- rd("facts.csv")
nat <- rd("nation.csv")
sta <- rd("states.csv",   colClasses = c(st = "character", geoid = "character"))
cou <- rd("counties.csv", colClasses = c(geoid = "character", st = "character"))
lad <- rd("ladder.csv")
reg <- rd("regions.csv")
cts <- rd("categories.csv")
std <- rd("standards.csv")
tld <- rd("tract_largest.csv")
wtr <- rd("wayne_tracts.csv", colClasses = c(geoid = "character"))
pro <- rd("eightmile_profile.csv")

FV <- function(k) { v <- fc$value[fc$name == k]
                    if (!length(v)) stop("no such fact: ", k); v }
FN <- function(k) as.numeric(FV(k))
n  <- function(x) format(round(as.numeric(x)), big.mark = ",")
pc <- function(x, k = 1) formatC(as.numeric(x), format = "f", digits = k)
mag <- function(k) sub("^[+-]", "", FV(k))

CAT <- c("hispanic", "nh_white", "nh_black", "nh_asian", "nh_two",
         "nh_aian", "nh_sor", "nh_nhpi")
RACE6 <- c("white", "black", "aian", "asian", "nhpi", "sor")
cou$div <- round(100 * cou$entropy / log(8), 1)
reg$div <- round(100 * reg$entropy / log(8), 1)
rg4 <- reg[reg$kind == "region", ]
sta$div <- round(100 * sta$entropy / log(8), 1)
WRANK <- as.integer(rank(-cou$entropy)[cou$geoid == "26163"])
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`demographics-brief.Rmd`) is a chapter, "A Country
Nobody Lives In," organised under prose headings and carrying eight numbered
figures: Figure 1, the eight categories in true proportion with the small tail
magnified; Figure 2, which box you could be counted in by census year, with the
two OMB revisions drawn as rules; Figure 3, the six races counted alone and
alone-or-in-combination; Figure 4, 2010 against 2020 with the processing
discontinuity shaded; Figure 5, six small-multiple maps of the states; Figure
6, the diversity staircase with its random-shuffle null; Figure 7, Wayne County
at three scales; Figure 8, the blocks along Eight Mile Road and the composition
profile across it. The lab (`demographics.Rmd`) runs Steps 1–11 with the code
exposed. **Students should read the brief first.** **Step numbers below refer
to the lab.** The chapter covers the same ground in the same order, so each lab
Step has a section facing it: Step 1 = "What the census counted" (Figure 1);
Step 2 = "Nobody at the Census Bureau chose these boxes" (Figure 2); Step 3 =
the prediction prompt and Figure 3; Step 4 = "A decade that mostly did not
happen" (Figure 4); Steps 5 and 6 = "Where people are" (Figure 5); Steps 7 and
8 = "The number that keeps moving" (Figure 6); Step 9 = "One county with no
majority" (Figure 7); Step 10 = "One street" (Figure 8); Step 11 = "What this
data can and cannot support."

**What this lab is for, and what it deliberately is not.** Three labs in this
course already own parts of this territory and this one stays out of all three.
`census` owns *how the census defines race* — Questions 8 and 9, the write-in
boxes, "Some Other Race." `areal-units` owns *the modifiable areal unit
problem*. `census-geography` owns *the nesting spine*. This lab is the
substantive one: what the country actually looks like, and how differently it
looks at different scales. Where it touches those labs it says so and points at
them; it does not re-teach them.

Every number below is computed from the built files at knit time.

***

# The session in one sentence

The United States scores `r FV("div_nation")` on this chapter's diversity scale
and the average American's census tract scores `r FV("div_censustract")` — and
neither number means anything until you know which of the Census Bureau's two
correct counts of race it was built from, and which revision of an OMB
directive created the categories it counts.

***

# The findings

```{r headline}
c(us_population        = FN("us_pop"),
  diversity_nation     = FN("div_nation"),
  diversity_state      = FN("div_state"),
  diversity_county     = FN("div_county"),
  diversity_tract      = FN("div_censustract"),
  diversity_block      = FN("div_censusblock"),
  H_tract              = FN("H_censustract"),
  H_tract_null         = FN("H0_censustract"),
  pct_in_tract_80plus  = FN("pop_tract_80plus"))

c(white_alone          = FN("us_white_alone"),
  white_any            = FN("us_white_any"),
  aian_alone           = FN("us_aian_alone"),
  aian_any             = FN("us_aian_any"),
  aian_ratio           = FN("us_aian_ratio"),
  white_ratio          = FN("us_white_ratio"))

c(chg_white_alone_2010_2020 = FV("chg_white_alone"),
  chg_white_any_2010_2020   = FV("chg_white_any"),
  chg_two_plus              = FV("chg_two_plus"),
  chg_sor_any               = FV("chg_sor_any"))
```

- **The White population both fell and rose over the decade.** Alone:
  `r FV("chg_white_alone")`%. Alone or in any combination:
  `r FV("chg_white_any")`%. Same censuses, same forms, same answers. This is
  the chapter's cold open and the single best thing in the session.
- **American Indian and Alaska Native goes from `r n(FV("us_aian_alone"))` to
  `r n(FV("us_aian_any"))` — a factor of `r FV("us_aian_ratio")`.** White moves
  by `r n(FN("us_white_gap"))` people and a factor of only
  `r FV("us_white_ratio")`. Largest absolute mover and largest proportional
  mover are different categories; both answers to "which is most affected" are
  right.
- **The staircase: `r FV("div_nation")` → `r FV("div_state")` →
  `r FV("div_county")` → `r FV("div_censustract")`.** The random-shuffle null
  stays at `r pc(100 * lad$ebar0[lad$level == "Census tract"] / log(8))`
  through the tract rung, so this is sorting, not small-unit arithmetic.
- **`r FV("pop_tract_80plus")`% of Americans live in a tract that is at least
  80% one category**; `r FV("pop_tract_90plus")`% in one at least 90%.
- **Wayne County, Michigan is in the most diverse
  `r pc(100 * WRANK / nrow(cou))`% of `r n(nrow(cou))` counties
  (rank `r n(WRANK)`), and `r FV("wayne_pop_80_pct")`% of its residents live in
  a tract that is at least 80% one group.** Both sentences are true.
- **Across Eight Mile Road the Black share falls `r FV("prof_jump_black")`
  points** — `r FV("prof_in500_black")`% in the last 500 m inside Detroit,
  `r FV("prof_out500_black")`% in the first 500 m outside. Flat, step, flat.

## Wayne County, the inventory

```{r wayne-inventory}
table(largest_group_in_tract = wtr$largest)

c(tracts               = nrow(wtr),
  tracts_with_people   = FN("wayne_tracts_pop"),
  median_pct_one_group = FN("wayne_tract_median_largest"),
  at_least_80_black    = FN("wayne_tracts_80_black"),
  at_least_80_white    = FN("wayne_tracts_80_white"),
  people_in_those_pct  = FN("wayne_pop_80_pct"),
  at_least_90_any      = FN("wayne_tracts_90"),
  people_in_those_pct  = FN("wayne_pop_90_pct"))
```

**`r n(FV("wayne_tracts_white_largest"))` tracts where the largest group is
White, `r n(FV("wayne_tracts_black_largest"))` where it is Black, and almost
nothing that is genuinely mixed.** That near-tie at the county level is exactly
why the county looks integrated and exactly why it is not. Say the two numbers
out loud together; the room usually hears the tie before it hears the split.

***

# The three things to get said

**1. Every count of race has at least two correct values, and the reason is
dated.** Step 3 is where this lands. Make somebody say *why* it is worse than
either number alone: if the two counts differed only occasionally you would
call it rounding, but the gap is systematically largest for the smallest
categories — `r FV("us_aian_ratio")`× for American Indian and Alaska Native,
`r FV("us_white_ratio")`× for White — which means the choice of definition
moves the groups that civil-rights enforcement is written about, and barely
touches the group it is not. Then give the date: the ambiguity did not exist
before the 1997 revision of OMB Directive 15, because the form allowed one
mark.

**2. A decade of demographic change that mostly was not one.** Step 4. The
Bureau's own analysts attribute the jump in Some Other Race "largely" to the
improved form, processing and coding. Push the class past "so the data is
unreliable," which is the failure mode, and onto the harder point: **2010 and
2020 were the easy pair.** Same standard, same categories, same file layout,
same field order. Everything that could be held constant was, and the
comparison still fails. Ask what that implies for any series crossing 1997, or
any series that will cross the 2024 revision.

**3. Step 8 is the session.** Two lines on one chart. The observed staircase
falls from `r FV("div_nation")` to `r FV("div_censustract")`; the
random-shuffle null does not move. **The lesson is not that America is
segregated** — students arrive believing that. The lesson is that a national
statistic and a neighbourhood statistic are answers to different questions, and
that the gap between them is measurable, is `r FV("H_censustract")` on Theil's
index, and survives the obvious objection.

**The line for the board:** the country is not a place anybody lives.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** The eight categories | `r n(FV("us_pop"))` people, 8 boxes, no remainder | 4 | Somebody will ask why Hispanic is not a race. That is `census`'s lab; give the one-sentence answer and move on. |
| **2** Who chose the boxes | Three issues of OMB Directive 15 | **8** | **Do not skip.** The "Two or More Races" row being empty before 2000 is the whole point of the figure. |
| **3** Alone or in combination | `r FV("us_aian_ratio")`× against `r FV("us_white_ratio")`× | **12** | **The prediction.** Take the vote before revealing. |
| **4** Two censuses | `r FV("chg_white_alone")`% and `r FV("chg_white_any")`% | **10** | The trap is "so the data is unreliable." Push to "the easy pair still failed." |
| **5** Regions | `r rg4$name[which.max(rg4$div)]` `r pc(max(rg4$div))`, `r rg4$name[which.min(rg4$div)]` `r pc(min(rg4$div))` | 4 | Regions are too big to be interesting. Say so and keep moving. |
| **6** States | Hawaii `r pc(max(sta$div))`, Maine `r pc(min(sta$div))` | 4 | Only `r sum(sta$largest != "White")` states have a largest group that is not White. |
| **7** The measure | Why entropy, not dissimilarity | **8** | The null is introduced here, not in Step 8. Make sure they know what it controls for before they see it. |
| **8** The ladder | `r FV("div_nation")` → `r FV("div_censustract")` | **12** | **The pivot.** Ask what the dashed line would look like if the fall were an artefact. |
| **9** One county | Wayne: no majority, and `r FV("wayne_pop_80_pct")`% in a lopsided tract | **10** | Both facts, in the same breath. The tie between `r n(FV("wayne_tracts_white_largest"))` and `r n(FV("wayne_tracts_black_largest"))` tracts is the moment. |
| **10** One street | `r FV("prof_jump_black")` points across one road | 5 | The *shape* — flat, step, flat — not the levels. |
| **11** What you can say | Including what you cannot | 3 | Make them say the definitional caveat unprompted. |

Total **80 minutes**.

**Step 3 deserves the prediction, and it is the best one in the course.** Ask
the room: *when you let people mark more than one race, which category grows
most in proportion?* Almost everyone says Black or White, because those are the
two categories the first two steps put in their heads. It is **American Indian
and Alaska Native**, which nearly triples, and the mechanism — the smaller the
group, the larger the share of its identifying population that is also
something else — is obvious in hindsight and essentially never guessed.

**If you are short on time**, protect 3, 4 and 8. The two-definitions reveal,
the decade that did not happen, and the staircase with its null are the
argument. Steps 5 and 6 can be read aloud from the tables in ninety seconds.

***

# Two things that will come up, and the honest answers

**"Isn't this just the modifiable areal unit problem again?"** Yes and no, and
the distinction is worth eight minutes because a sharp student will raise it.
`areal-units` shows that the *statistic* changes with the unit — variance
collapses, extremes vanish, a correlation flips sign — and concludes that the
unit is a choice with consequences. That is a fact about the instrument. Step 8
runs the same shape of ladder and then **subtracts the instrument**: shuffle the
same people at random into units of the identical sizes and the staircase
disappears.

```{r maup-control}
lad[, c("level", "units", "diversity", "H", "H0")]
```

Look at `H0`. It is essentially zero at every rung down to block groups
(`r pc(FN("H0_blockgroup"), 4)`), and only at the block rung does it reach
`r FV("H0_censusblock")`, against an observed `r FV("H_censusblock")`. **So
roughly `r pc(100 * (1 - FN("H0_censusblock") / FN("H_censusblock")))`% of even
the block-level figure is real.** MAUP is a genuine problem and this is what it
looks like when you measure how much of it you have.

**"Isn't the block level ruined by differential privacy?"** Partly, and the
answer should be precise rather than hand-waving. The 2020 disclosure-avoidance
system held **state** total populations invariant and did not alter them; it
did **not** hold block totals invariant. Block counts carry the most noise of
any level, and this is a second reason — beyond small-unit arithmetic — to
treat the last rung of Figure 6 as indicative rather than exact. The tract rung
is the one to quote. If a student notices that the null in Step 8 does not
separate DP noise from small-N noise, that is a genuinely sophisticated
observation and should be rewarded: it does not, and the two are confounded at
the block level.

***

# Option A — A different county

**Scope: all `r n(nrow(cou))` counties, in `data/counties.csv`.**

## Verified output

```{r a}
big <- cou[cou$total >= 250000, ]
head(big[order(-big$div), c("name", "st", "total", "div", "largest")], 6)
head(big[order(big$div),  c("name", "st", "total", "div", "largest")], 5)
```

**3/4:** Picks a high-scoring county and a low-scoring one, reports both
scores, and says the county score does not tell you whether people live
together.

**4/4:** Names what would settle it — the tract-level distribution inside the
county, or the same staircase computed within it — and notices that the lab
already did this for Wayne and that Wayne's county score
(`r pc(cou$div[cou$geoid == "26163"])`) is high.

**4/4, best:** Notices the **Webb County trap**, which is the reason this option
exists. The least diverse large county in the United States is
`r big$name[which.min(big$div)]` County, Texas, at
`r pc(min(big$div))` — and its largest group is
`r tolower(big$largest[which.min(big$div)])`, not White. **A low diversity score
does not mean "white."** It means "one group," and the entropy index is
deliberately blind to which. A student who spots that has understood what the
measure does and does not encode, and has independently rediscovered why the
chapter refused to use a two-group dissimilarity index.

**Watch for:** a student reading the diversity score as a moral scale, so that
Webb County comes out "worse" than a county that is 60% White. Correct it on
the spot: the index measures evenness across eight administrative categories
and nothing else.

***

# Option B — A different definition

**Scope: the state table from Step 6, recomputed with `black_any` in place of
`nh_black`.**

## Verified output

```{r b}
sta$pb_nh  <- round(100 * sta$nh_black  / sta$total, 2)
sta$pb_any <- round(100 * sta$black_any / sta$total, 2)
sta$gap    <- round(sta$pb_any - sta$pb_nh, 2)
sta$ratio  <- round(sta$black_any / sta$nh_black, 2)

head(sta[order(-sta$gap),   c("name", "pb_nh", "pb_any", "gap", "ratio")], 5)
head(sta[order(-sta$ratio), c("name", "pb_nh", "pb_any", "gap", "ratio")], 5)
```

**The two orderings have almost nothing in common, and that is the finding.**
By percentage points the biggest mover is
`r sta$name[which.max(sta$gap)]` (`r pc(max(sta$gap), 2)` points). By ratio it
is `r sta$name[which.max(sta$ratio)]` (`r pc(max(sta$ratio), 2)`×), where the
Black share is small enough that a modest number of second marks nearly
triples it.

**3/4:** Recomputes, reports that the numbers rise everywhere, and names a few
states that move.

**4/4:** Distinguishes the two orderings and explains why they differ —
absolute change tracks where Black populations are large, proportional change
tracks where they are small. This is the same structure as Step 3's national
result, applied one level down, and saying so is the transfer worth rewarding.

**4/4, best:** Takes a position on the legal question and defends it. Voting
rights litigation under Section 2 generally uses **any-part Black**, and there
is a real argument for it: the question is who could form a politically
cohesive group, not who ticked exactly one box. There is also a real argument
against, since the any-part definition double-counts people across groups and
the shares no longer sum to 100. A student who notices **that the any-part
columns do not sum to the population** — and that this is a feature of the
definition rather than a bug in the file — has read the table properly.

**Watch for:** adding `black_any` and `nh_white` together and being surprised
that they exceed the total. Also watch for a student comparing `black_any`
against `nh_black`, which mixes the multiple-race axis with the ethnicity axis;
the clean comparison is `black_alone` against `black_any`, and `nh_black` is a
third thing again. That confusion is worth ten minutes if it appears, because
it is precisely the seam the `census` lab is about meeting the seam this lab is
about.

***

# Option C — A different threshold

**Scope: `data/tract_largest.csv`, the whole distribution.**

## Verified output

```{r c}
o <- tld
o$pct <- round(100 * o$people / sum(o$people), 1)
o$at_or_above <- round(100 * rev(cumsum(rev(o$people))) / sum(o$people), 1)
o
```

Reading down the last column: **`r pc(o$at_or_above[o$bin == "(40,50]"])`% of
Americans live in a tract whose largest group is over 40%,
`r pc(o$at_or_above[o$bin == "(70,80]"])`% over 70%,
`r pc(o$at_or_above[o$bin == "(80,90]"])`% over 80%, and
`r pc(o$at_or_above[o$bin == "(90,100]"])`% over 90%.**

**3/4:** Reports the distribution and picks a threshold.

**4/4:** Notices that the answer can be made to sound like almost anything —
"a third of Americans live in a near-segregated neighbourhood" and "two thirds
of Americans live in a tract with no dominant group" are both defensible from
this one column — and says which they would use and why.

**4/4, best:** Sees that this is why the chapter leads with an index rather
than a threshold. The entropy score has no threshold in it; it is a single
number per unit with no cut point to shop for. Then notices the cost of that
choice: the index is harder to explain, and "`r FV("pop_tract_80plus")`% of
Americans" is the sentence that will actually be quoted. **The best answers
treat the trade-off between a defensible measure and a repeatable sentence as
the real problem, not as a rhetorical flourish.**

***

# Option D — The 2024 categories

**Conceptual; no verified output. Grade the reasoning.**

```{r d}
std[, c("key", "issued", "citation", "categories", "question_form")]
cts[cts$c2030 == 1 | cts$c2020 == 1, c("category", "dimension", "c2020", "c2030")]
```

**3/4:** Lists what changes — a combined question,
`r std$categories[std$key == "2024"]` categories, MENA added — and says the
numbers would be different.

**4/4:** Sorts the lab's numbers into three piles correctly. **Unchanged:** the
total population; the ladder's *structure*, since it depends on geography
rather than categories. **Changed and predictable in direction:** the diversity
score, because it is computed over the number of categories — more categories
mechanically raise the ceiling, so `r FV("E_max")` becomes
$\ln `r std$categories[std$key == "2024"]`$ and every score is rescaled.
**Unpredictable:** everything about White and MENA, because nobody knows how
people who were counted as White in 2020 will answer a form that offers them a
different box.

**4/4, best:** Sees that the last pile is the interesting one and states the
consequence: **a change in the count of the White population between 2020 and
2030 will be uninterpretable in exactly the way Step 4's 2010-to-2020 change
already was, only more so.** The strongest answers connect this to Step 9 and
notice that Wayne County contains Dearborn, home to the largest Arab-American
concentration in the United States, whose residents the 2020 form counted as
White — so the county the lab zooms into is one whose *2030* numbers will move
for purely definitional reasons.

**Watch for:** a student assuming the 2024 standard is already in the data. It
is not. It was issued in 2024, agencies' action plans are now due in 2027 after
two extensions, full compliance is due in 2029, and the Census Bureau has said
it is *preparing* to implement it in the 2027 American Community Survey and the
2030 Census. Nothing in this lab's data uses it.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given.
- **The move that earns a 4** is treating a definitional choice as a
  substantive decision with consequences, not as a technical footnote.
- **Also a 4**: quoting a number *with its universe attached* — "White alone,"
  "any-part Black," "total population, not voting-age." The documents model
  this everywhere; using it unprompted is the habit the lab exists to build.
- **Also a 4**: noticing that the random-shuffle null in Step 8 is what
  licenses the whole argument, and saying what the figure would look like
  without it.
- **A 2** is "race data is socially constructed so the numbers do not mean
  anything." True in its first clause, false in its second, and the lab is
  built to move past it: the numbers mean something quite precise, and what
  they mean is a joint fact about people and about a directive.
- **Also a 2**: reporting the national composition as though it described
  anywhere.
- **Also a 2**: quoting Wayne County as a national finding, or the national
  staircase as a fact about Detroit. The documents make the scope explicit;
  ignoring it is the failure mode this lab introduces.
- **Reward anyone who asks what the numbers would be under the 2024
  categories.** They cannot be computed, and knowing that is the right answer.

## Anticipated questions

- *"Why is Hispanic not a race?"* — Because OMB said so in 1977 and again in
  1997, and the Bureau implements OMB's standard. As of the 2024 revision it is
  a category in the same list. The full answer is the `census` lab; the
  one-sentence answer is that this was an administrative decision with a date
  on it.
- *"Why Wayne County?"* — Because its county-level number is misleading in an
  instructive direction: no majority group, top `r pc(100 * WRANK / nrow(cou))`%
  of counties for diversity, and most residents in a tract that is at least
  four-fifths one group. Then say what the national staircase is for: if the
  county were doing the work, Figure 6 would not be there.
- *"Is Wayne County typical?"* — On diversity, no: it is unusually high, which
  is the point. On the gap between its county score and its tract scores, it is
  a strong case rather than a unique one. The honest answer is that the lab
  does not compute the within-county gap for all `r n(nrow(cou))` counties, and
  that doing so would be a good final project.
- *"Couldn't you just use the dissimilarity index?"* — Yes, for two groups. The
  chapter says why it did not: with eight categories, choosing one contrast is
  a substantive claim. If a student wants Black-White dissimilarity for a
  particular metro, that is a legitimate extension, not a correction.
- *"Do the block numbers mean anything at all?"* — Individually, no. In
  aggregate, yes: the null in Step 8 shows most of the block-level effect
  survives. Say both halves.
- *"Has anything changed since 2020?"* — The categories have, on paper. The
  data has not. That gap — a standard in force and no data collected under it —
  is itself worth a minute.
- *"Why does the White population show up as both falling and rising?"* — Step
  3 and Step 4 together. If a student can answer this question to the room, the
  session has worked.

## Rebuilding the data

`data/build-data.R` downloads the 2020 national roll-up and all 51 state P.L.
94-171 files (about 1.3 GB, processed and discarded), the 2010 national file,
and the TIGER layers not already cached elsewhere in this course. It reuses
`../redlining/data/tiger/26/` for Michigan tracts and
`../residual-votes/data/tiger/st/` for state outlines rather than
re-downloading them. It writes about 1.7 MB of CSVs, records its source URLs
and fetch date, and **stops** if the eight categories fail to partition the
population at any level of any state, or if a state's blocks do not sum to the
state. Set `DEMOG_CACHE` to a directory of already-downloaded zips to skip the
downloads. Rerun it and reknit all three documents; every number in them is
computed at knit time, so a stale build is the only way they can go out of
date.

## Connections

- **`census`** — the two questions on the 2020 form, the write-in boxes, and
  "Some Other Race." That lab explains where this lab's Table P2 comes from.
  Run it first if you can; Step 1 assumes it.
- **`areal-units`** — the modifiable areal unit problem, on Fulton County. Step
  8's null is the control that lab's ladder does not have, and the contrast is
  worth teaching explicitly.
- **`census-geography`** — the nesting spine. It is why Step 8's ladder is
  assumption-free: a county's population *is* the sum of its tracts, by
  construction, so nothing in Figure 6 required a crosswalk.
- **`redlining`** — one mechanism that could produce the step in Step 10, and
  the only lab in the course that puts a named document behind a boundary.
- **`surnames` and `bisg-check`** — what happens when race must be *inferred*
  because nobody recorded it. This lab is the case where somebody did.
- **`section-203`** — a federal formula that turns counts of specific language
  and ancestry groups into legal obligations, and therefore a place where the
  definitional choices in Steps 2 and 3 have direct consequences.
