---
title: "A Country Nobody Lives In"
subtitle: "84-355 Democracy's Data | Race in the 2020 census, from the nation to the block"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.6,
                      dpi = 96, fig.retina = 1)
options(scipen = 999)
D <- "data"
rd <- function(f, ...) read.csv(file.path(D, f), stringsAsFactors = FALSE, ...)

# GEOIDs are CHARACTER. A tract identifier is 11 digits and a block identifier
# 15; read as numbers the block becomes 2.6163e+14 and any identifier in a
# state whose FIPS code starts with a zero loses that zero. Every read below
# that touches an identifier pins the column.
fc  <- rd("facts.csv")
nat <- rd("nation.csv")
reg <- rd("regions.csv")
sta <- rd("states.csv",   colClasses = c(st = "character", geoid = "character"))
cou <- rd("counties.csv", colClasses = c(geoid = "character", st = "character"))
lad <- rd("ladder.csv")
dec <- rd("decades.csv")
cts <- rd("categories.csv")
std <- rd("standards.csv")
wtr <- rd("wayne_tracts.csv",      colClasses = c(geoid = "character"))
wtg <- rd("wayne_tract_rings.csv", colClasses = c(id = "character"))
wbg <- rd("wayne_bg.csv",          colClasses = c(geoid = "character"))
wbr <- rd("wayne_bg_rings.csv",    colClasses = c(id = "character"))
wou <- rd("wayne_outline.csv",     colClasses = c(id = "character"))
wtx <- rd("wayne_transect.csv",    colClasses = c(GEOID = "character"))
wbk <- rd("wayne_block_rings.csv", colClasses = c(id = "character"))
wdt <- rd("wayne_detroit_rings.csv")
pro <- rd("eightmile_profile.csv")

FV <- function(k) { v <- fc$value[fc$name == k]
                    if (!length(v)) stop("no such fact: ", k); v }
FN <- function(k) as.numeric(FV(k))
n  <- function(x) format(round(as.numeric(x)), big.mark = ",")
pc <- function(x, k = 1) formatC(as.numeric(x), format = "f", digits = k)
mag <- function(k) sub("^[+-]", "", FV(k))

CAT <- c("hispanic", "nh_white", "nh_black", "nh_asian", "nh_two",
         "nh_aian", "nh_sor", "nh_nhpi")
SHORT <- c(hispanic = "Hispanic", nh_white = "White", nh_black = "Black",
           nh_asian = "Asian", nh_two = "Two or More", nh_aian = "AIAN",
           nh_sor = "Some Other Race", nh_nhpi = "NHPI")
COL <- c(hispanic = "#E08214", nh_white = "#7F9BB3", nh_black = "#C41230",
         nh_asian = "#4D9221", nh_two = "#8073AC", nh_aian = "#B35806",
         nh_sor = "#999999", nh_nhpi = "#2C7FB8")
RACE6 <- c("white", "black", "aian", "asian", "nhpi", "sor")
RLAB  <- c(white = "White", black = "Black or African American",
           aian = "American Indian and Alaska Native", asian = "Asian",
           nhpi = "Native Hawaiian and Other Pacific Islander",
           sor = "Some Other Race")

# ---- geometry helpers, shared by both renderers ----------------------------
onepath <- function(X, Y) {
  dx <- X[-1] - X[-length(X)]; dy <- Y[-1] - Y[-length(Y)]
  seg <- paste0(dx, ifelse(dy < 0, "", ","), dy)
  sep <- c("", ifelse(substr(seg[-1], 1, 1) == "-", "", " "))
  paste0("M", X[1], ",", Y[1], "l", paste0(sep, seg, collapse = ""), "Z")
}
ringpaths <- function(d, sx, sy) {
  if (!nrow(d)) return(character(0))
  k <- interaction(d$id, d$part, drop = TRUE)
  p <- vapply(split(d, k), function(z) {
    X <- round(sx(z$x)); Y <- round(sy(z$y))
    keep <- c(TRUE, X[-1] != X[-length(X)] | Y[-1] != Y[-length(Y)])
    X <- X[keep]; Y <- Y[keep]
    if (length(X) < 3) return(paste0("M", X[1], ",", Y[1], "h1v1h-1Z"))
    onepath(X, Y)
  }, character(1))
  p[nzchar(p)]
}
ringids <- function(d) {
  k <- interaction(d$id, d$part, drop = TRUE)
  vapply(split(d, k), function(z) z$id[1], character(1))
}
drawrings <- function(d, fill, border = NA, lwd = 0.3) {
  if (!nrow(d)) return(invisible())
  k <- interaction(d$id, d$part, drop = TRUE)
  for (l in levels(k)) {
    z <- d[k == l, ]
    polygon(z$x, z$y, col = fill[[as.character(z$id[1])]], border = border,
            lwd = lwd)
  }
}
jstr <- function(x) paste0("[", paste0('"', x, '"', collapse = ","), "]")
jnum <- function(x) paste0("[", paste0(x, collapse = ","), "]")
win <- function(d, pad = 0.04) {
  rx <- range(d$x); ry <- range(d$y)
  s <- max(diff(rx), diff(ry)) * (1 + pad) / 2
  list(x = mean(rx) + c(-s, s), y = mean(ry) + c(-s, s), s = 2 * s)
}
spread <- function(y, gap) {
  o <- order(y); v <- y[o]
  for (i in 2:length(v)) if (v[i] - v[i - 1] < gap) v[i] <- v[i - 1] + gap
  v <- v + (mean(y) - mean(v))
  out <- numeric(length(y)); out[o] <- v; out
}

knit_print.data.frame <- function(x, ...) {
  nm <- names(x); nm <- gsub("_", " ", nm)
  nm <- sub("^(.)", "\\U\\1", nm, perl = TRUE)
  knitr::knit_print(knitr::kable(x, col.names = nm, row.names = FALSE), ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

Everything here runs as given. **You will not write any code.** Your job is to
read what comes out and say what it means.

> **Read `demographics-brief.Rmd` first.** This lab runs the same argument as
> numbered steps with the code exposed. The chunks that *compute* are shown;
> the chunks that only *draw* are folded away, because their job is pictures,
> not arithmetic.

***

# The claim under examination

> *"The United States is `r pc(100 * as.numeric(nat["nh_white"]) / nat$total)`%
> White."*

True, on one of at least three defensible definitions, of a country in which
almost nobody lives in a place that looks like that.

This lab holds the census fixed — one day, one enumeration,
`r n(FV("us_pop"))` people — and changes two things: **which definition** you
count with, and **how big a box** you compute the average over.

# Step 1 — The eight categories

```{r step1}
c(us_population = FN("us_pop"),
  categories    = FN("n_categories"))

d <- data.frame(category = SHORT[CAT],
                people   = as.numeric(nat[CAT]),
                pct      = round(100 * as.numeric(nat[CAT]) / nat$total, 2))
d <- d[order(-d$people), ]
d$people <- format(d$people, big.mark = ",")
d

# the eight must partition the population exactly -- no remainder, no overlap
sum(as.numeric(nat[CAT])) == nat$total
```

These come from **Table P2** of the 2020 redistricting file, which cross-cuts
the race question by the Hispanic-origin question and reports every resident
exactly once. `r n(FV("us_pop"))` people, `r FV("n_categories")` boxes, no
remainder.

Two questions on the form, one table out. The `census` lab in this course is
about those two questions and the write-in boxes behind them; this lab starts
from the table they produce.

# Step 2 — Nobody at the Census Bureau chose these boxes

```{r step2}
o <- std[, c("issued", "citation", "categories", "question_form",
             "first_census")]
o$issued <- format(as.Date(o$issued), "%d %b %Y")
o
```

The categories are set by the **Office of Management and Budget** in
Statistical Policy Directive No. 15, and they bind every federal agency that
collects data on race. The directive has been issued three times, and each
version changed what a person could be counted as:

```{r step2b}
YC <- paste0("c", c(1980, 1990, 2000, 2010, 2020, 2030))
o <- cts[cts$dimension == "race", c("category", YC)]
names(o) <- c("category", c(1980, 1990, 2000, 2010, 2020, "2030 (planned)"))
o
```

Read the row for **Two or More Races**. It is empty until 2000, and not because
nobody was of two races before 2000. Until the 1997 revision the form allowed
one mark. **The category was created by a sentence in the Federal Register**,
and the people it counted in 2000 had been in the country all along, in other
boxes.

Read the row for **Middle Eastern or North African**. It appears only in the
planned 2030 column. The 1977 directive defined White as covering origins in
"Europe, North Africa, or the Middle East," so until the 2024 revision the
category absorbed them. Hold that fact: Step 9 lands on a county that contains
the largest Arab-American concentration in the United States.

> **Before you run the next chunk, commit to a number.** Step 3 counts each of
> the six races two ways — alone, and alone-or-in-any-combination. **Which
> category grows the most, in proportion, when the second mark is allowed, and
> by what factor?** Write down a name and a multiplier before you run it.

# Step 3 — Alone, or in any combination

```{r step3}
o <- data.frame(
  race  = RLAB[RACE6],
  alone = vapply(RACE6, function(r) FN(paste0("us_", r, "_alone")), 0),
  any   = vapply(RACE6, function(r) FN(paste0("us_", r, "_any")), 0))
o$difference <- o$any - o$alone
o$ratio      <- round(o$any / o$alone, 2)
o[order(-o$ratio), ]
```

```{r fig-dumbbell-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
R6 <- data.frame(k = RACE6, lab = RLAB[RACE6],
                 a = vapply(RACE6, function(r) FN(paste0("us_", r, "_alone")), 0),
                 b = vapply(RACE6, function(r) FN(paste0("us_", r, "_any")), 0),
                 stringsAsFactors = FALSE)
R6 <- R6[order(-R6$b / R6$a), ]
xmax <- sqrt(max(R6$b) * 1.02)
W3 <- 740; L3 <- 232; R3 <- 118; T3 <- 40; RH3 <- 32
H3 <- T3 + nrow(R6) * RH3 + 42
rows <- paste(vapply(seq_len(nrow(R6)), function(i) sprintf(
  '{"l":"%s","a":%.4f,"b":%.4f,"r":"%s"}', R6$lab[i],
  sqrt(R6$a[i]) / xmax, sqrt(R6$b[i]) / xmax,
  FV(paste0("us_", R6$k[i], "_ratio"))), character(1)), collapse = ",")
tk <- c(1e6, 1e7, 5e7, 1e8, 2e8)
tks <- paste(sprintf('{"x":%.4f,"t":"%s"}', sqrt(tk) / xmax,
                     c("1m", "10m", "50m", "100m", "200m")), collapse = ",")
cat(paste0('
<div id="db" style="margin:1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const R=[', rows, '],TK=[', tks, '];
const W=', W3, ',H=', H3, ',L=', L3, ',RR=', R3, ',T=', T3, ',RH=', RH3, ';
const svg=d3.select("#db").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const x=v=>L+v*(W-L-RR);
TK.forEach(t=>{
  svg.append("line").attr("x1",x(t.x)).attr("x2",x(t.x)).attr("y1",T-14)
    .attr("y2",T+R.length*RH-8).attr("stroke","#eee");
  svg.append("text").attr("x",x(t.x)).attr("y",T-20).attr("text-anchor","middle")
    .attr("font-size","10px").attr("fill","#aaa").text(t.t);});
R.forEach((d,i)=>{
  const y=T+i*RH;
  svg.append("text").attr("x",L-12).attr("y",y+4).attr("text-anchor","end")
    .attr("font-size","11.5px").attr("fill","#333").text(d.l);
  svg.append("line").attr("x1",x(d.a)).attr("x2",x(d.b)).attr("y1",y).attr("y2",y)
    .attr("stroke","#C41230").attr("stroke-width",3).attr("stroke-opacity",0.5);
  svg.append("circle").attr("cx",x(d.a)).attr("cy",y).attr("r",5.2)
    .attr("fill","#fff").attr("stroke","#555").attr("stroke-width",1.6);
  svg.append("circle").attr("cx",x(d.b)).attr("cy",y).attr("r",5.2)
    .attr("fill","#C41230");
  svg.append("text").attr("x",W-RR+10).attr("y",y+4).attr("font-size","12px")
    .attr("font-weight","600").attr("fill","#C41230").text("x "+d.r);});
const lg=svg.append("g").attr("transform","translate("+L+","+(H-14)+")");
lg.append("circle").attr("r",5.2).attr("fill","#fff").attr("stroke","#555")
  .attr("stroke-width",1.6);
lg.append("text").attr("x",11).attr("y",4).attr("font-size","11px")
  .attr("fill","#666").text("alone");
lg.append("circle").attr("cx",84).attr("r",5.2).attr("fill","#C41230");
lg.append("text").attr("x",95).attr("y",4).attr("font-size","11px")
  .attr("fill","#666").text("alone or in any combination  (square-root scale)");
})();
</script>'))
```

```{r fig-dumbbell-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=2.9}
R6 <- data.frame(k = RACE6, lab = RLAB[RACE6],
                 a = vapply(RACE6, function(r) FN(paste0("us_", r, "_alone")), 0),
                 b = vapply(RACE6, function(r) FN(paste0("us_", r, "_any")), 0),
                 stringsAsFactors = FALSE)
R6 <- R6[order(R6$b / R6$a), ]
xmax <- sqrt(max(R6$b) * 1.02)
par(mar = c(2.4, 14.0, 1.8, 5.0))
plot(NA, xlim = c(0, 1), ylim = c(0.5, nrow(R6) + 0.5), axes = FALSE, ann = FALSE)
tk <- c(1e6, 1e7, 5e7, 1e8, 2e8)
abline(v = sqrt(tk) / xmax, col = "#eeeeee")
axis(3, sqrt(tk) / xmax, c("1m", "10m", "50m", "100m", "200m"), tick = FALSE,
     cex.axis = 0.55, col.axis = "#aaa", line = -1.2)
for (i in seq_len(nrow(R6))) {
  a <- sqrt(R6$a[i]) / xmax; b <- sqrt(R6$b[i]) / xmax
  segments(a, i, b, i, col = adjustcolor("#C41230", 0.5), lwd = 3)
  points(a, i, pch = 21, bg = "white", col = "#555555", cex = 1.1, lwd = 1.4)
  points(b, i, pch = 19, col = "#C41230", cex = 1.1)
  mtext(R6$lab[i], side = 2, at = i, las = 1, line = 0.3, cex = 0.55)
  text(1.02, i, paste0("x ", FV(paste0("us_", R6$k[i], "_ratio"))), adj = 0,
       cex = 0.6, font = 2, col = "#C41230", xpd = NA)
}
legend("bottomleft", c("alone", "alone or in any combination"),
       pch = c(21, 19), col = c("#555555", "#C41230"),
       pt.bg = c("white", "#C41230"), bty = "n", cex = 0.6, horiz = TRUE,
       inset = c(0, -0.12), xpd = NA)
```

**`r RLAB[["aian"]]` goes from `r n(FN("us_aian_alone"))` to
`r n(FN("us_aian_any"))`, a factor of `r FV("us_aian_ratio")`.** White moves by
the largest absolute amount — `r n(FN("us_white_gap"))` people — and by the
smallest proportion, `r FV("us_white_ratio")`.

Almost everyone predicts Black or White, because those are the categories the
first two steps put in your head. The rule is the opposite: **the smaller the
category, the more the second mark matters.** And the small categories are
exactly the ones that voting-rights law, tribal funding formulas and
civil-rights enforcement are written about.

# Step 4 — Two censuses, and why they do not compare

```{r step4}
K <- c("white_alone", "white_any", "black_alone", "black_any",
       "aian_any", "sor_any", "two_plus")
data.frame(
  measure   = K,
  y2010     = vapply(K, function(k) FN(paste0("d10_", k)), 0),
  y2020     = vapply(K, function(k) FN(paste0("d20_", k)), 0),
  pct_change = vapply(K, function(k) FV(paste0("chg_", k)), ""))
```

**White alone falls `r mag("chg_white_alone")`% and White in any combination
rises `r mag("chg_white_any")`%.** Two or More Races rises
`r mag("chg_two_plus")`%.

Do not read those as demographic change. Between the two censuses the Bureau
added dedicated write-in areas under "White" and under "Black or African
American," raised the characters captured per write-in from 30 to 200, and
began assigning up to six detailed codes per response instead of two. Of the
jump in Some Other Race the Bureau's own analysts wrote that it "could be
attributed to a number of factors, including demographic change since 2010. But
we expect they were largely due to the improvements to the design of the two
separate questions for race and ethnicity, data processing and coding."

Notice how hard this failure is. **2010 and 2020 are the easy pair**: same OMB
standard, same categories, same file format, same fields in the same order.
Everything that could be held constant was, and the comparison still cannot be
trusted. A comparison crossing 1997 has no chance at all.

# Step 5 — Regions and divisions

```{r step5}
o <- reg[, c("kind", "name", "total", CAT[1:4], "entropy")]
o$diversity <- round(100 * o$entropy / log(8), 1)
o$entropy <- NULL
o[order(o$kind, -o$total), ]
```

Composition varies enormously by region. But the smallest of the four regions
holds `r n(min(reg$total[reg$kind == "region"]))` people, and an average over a
container that size describes nowhere in particular.

# Step 6 — States

```{r step6}
sta$diversity <- round(100 * sta$entropy / log(8), 1)
sta$white_pct <- round(100 * sta$nh_white / sta$total, 1)

head(sta[order(-sta$diversity), c("name", "total", "white_pct", "largest",
                                  "diversity")], 6)
head(sta[order(sta$diversity),  c("name", "total", "white_pct", "largest",
                                  "diversity")], 6)

# the states whose largest group is not White
sta[sta$largest != "White", c("name", "largest", "diversity")]
```

`r sta$name[which.max(sta$entropy)]` scores
`r pc(100 * max(sta$entropy) / log(8))` and
`r sta$name[which.min(sta$entropy)]` scores
`r pc(100 * min(sta$entropy) / log(8))`. Only
`r sum(sta$largest != "White")` states have a largest group that is not White.

# Step 7 — The measure, and the control

The **diversity score** used from here on is the multigroup entropy index,
expressed as a percentage of its maximum. For a unit with shares $p_r$ over the
eight categories, $E = \sum_r p_r \ln(1/p_r)$; it is 0 when everyone is in one
category and $\ln 8$ when the eight are perfectly even.

```{r step7}
c(E_united_states = FN("E_us"),
  E_maximum       = FN("E_max"),
  diversity_score = FN("div_nation"))
```

Entropy rather than the more famous dissimilarity index, for one reason:
dissimilarity takes exactly **two** groups. With eight categories that means
either 28 numbers or a decision that one contrast — nearly always Black against
White — is *the* contrast. That decision is a claim about the country, and the
measure should not make it silently.

The measure has a real problem, and Step 8 controls for it. **Smaller units
always look more segregated, even if people were sprinkled at random**, because
a unit of twelve people cannot be evenly mixed across eight categories. So the
build script also computes a **null**: the same population shuffled at random
into units of the identical sizes.

```{r step7b}
o <- lad[, c("level", "units", "diversity", "H", "H0")]
o$H <- round(o$H, 4); o$H0 <- round(o$H0, 4)
o
```

`H` is the share of national diversity missing from the average person's unit.
`H0` is the same number for the shuffled population. **Read them as a pair.**

# Step 8 — The ladder

```{r fig-ladder-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
LV <- lad$level
d0 <- 100 * lad$ebar0 / log(8); d1 <- lad$diversity
W6 <- 740; L6 <- 58; R6M <- 150; T6 <- 42; B6 <- 58; H6 <- 320
SW <- (W6 - L6 - R6M) / length(LV)
yl <- c(30, 63)
ys <- function(v) T6 + (yl[2] - v) / diff(yl) * (H6 - T6 - B6)
steps <- function(v) {
  s <- character(0)
  for (i in seq_along(v)) {
    x0 <- L6 + (i - 1) * SW; x1 <- L6 + i * SW
    s <- c(s, sprintf("%s%.1f,%.1f L%.1f,%.1f", if (i == 1) "M" else "L",
                      x0, ys(v[i]), x1, ys(v[i])))
    if (i < length(v)) s <- c(s, sprintf("L%.1f,%.1f", x1, ys(v[i + 1])))
  }
  paste(s, collapse = " ")
}
labs <- paste(vapply(seq_along(LV), function(i) sprintf(
  '{"l":"%s","u":"%s","x":%.1f,"y":%.1f,"d":"%s"}', LV[i], n(lad$units[i]),
  L6 + (i - 0.5) * SW, ys(d1[i]), pc(d1[i])), character(1)), collapse = ",")
cat(paste0('
<div id="ld" style="margin:1em 0"></div>
<script>
(function(){
const L=[', labs, '],P1="', steps(d1), '",P0="', steps(d0), '";
const W=', W6, ',H=', H6, ',LX=', L6, ',T=', T6, ',B=', B6, ';
const svg=d3.select("#ld").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
[35,40,45,50,55,60].forEach(v=>{
  const y=T+(', yl[2], '-v)/', diff(yl), '*(H-T-B);
  svg.append("line").attr("x1",LX).attr("x2",W-150).attr("y1",y).attr("y2",y)
    .attr("stroke","#f0f0f0");
  svg.append("text").attr("x",LX-8).attr("y",y+4).attr("text-anchor","end")
    .attr("font-size","10px").attr("fill","#aaa").text(v);});
svg.append("path").attr("d",P0).attr("fill","none").attr("stroke","#bbb")
  .attr("stroke-width",1.6).attr("stroke-dasharray","5,3");
svg.append("path").attr("d",P1).attr("fill","none").attr("stroke","#C41230")
  .attr("stroke-width",2.6);
L.forEach(d=>{
  svg.append("circle").attr("cx",d.x).attr("cy",d.y).attr("r",3.6)
    .attr("fill","#C41230");
  svg.append("text").attr("x",d.x).attr("y",d.y-10).attr("text-anchor","middle")
    .attr("font-size","11.5px").attr("font-weight","600").attr("fill","#C41230")
    .text(d.d);
  svg.append("text").attr("x",d.x).attr("y",H-B+16).attr("text-anchor","middle")
    .attr("font-size","11px").attr("fill","#333").text(d.l);
  svg.append("text").attr("x",d.x).attr("y",H-B+29).attr("text-anchor","middle")
    .attr("font-size","9.5px").attr("fill","#999").text(d.u);});
svg.append("text").attr("x",LX).attr("y",T-16).attr("font-size","11.5px")
  .attr("fill","#C41230").attr("font-weight","600")
  .text("diversity score of the unit the average American lives in");
const lx=W-144;
svg.append("line").attr("x1",lx).attr("x2",lx+22).attr("y1",T+8).attr("y2",T+8)
  .attr("stroke","#C41230").attr("stroke-width",2.6);
svg.append("text").attr("x",lx+28).attr("y",T+12).attr("font-size","11px")
  .text("observed");
svg.append("line").attr("x1",lx).attr("x2",lx+22).attr("y1",T+26).attr("y2",T+26)
  .attr("stroke","#bbb").attr("stroke-width",1.6).attr("stroke-dasharray","5,3");
svg.append("text").attr("x",lx+28).attr("y",T+30).attr("font-size","11px")
  .text("shuffled at random");
})();
</script>'))
```

```{r fig-ladder-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.4}
LV <- lad$level
d0 <- 100 * lad$ebar0 / log(8); d1 <- lad$diversity
par(mar = c(4.0, 3.4, 2.6, 0.8))
plot(NA, xlim = c(0.5, length(LV) + 0.5), ylim = c(30, 63), axes = FALSE,
     ann = FALSE)
abline(h = seq(35, 60, 5), col = "#f0f0f0")
axis(2, seq(35, 60, 5), cex.axis = 0.6, col = "#dddddd", col.axis = "#aaa",
     tck = -0.015, mgp = c(2, 0.4, 0), las = 1)
stepx <- rep(seq_along(LV), each = 2) + c(-0.5, 0.5)
lines(stepx, rep(d0, each = 2), col = "#bbbbbb", lwd = 1.6, lty = 2)
lines(stepx, rep(d1, each = 2), col = "#C41230", lwd = 2.4)
points(seq_along(LV), d1, pch = 19, col = "#C41230", cex = 0.8)
text(seq_along(LV), d1 + 1.6, pc(d1), cex = 0.6, font = 2, col = "#C41230")
axis(1, seq_along(LV), LV, tick = FALSE, cex.axis = 0.6, line = -0.4,
     col.axis = "#333")
mtext(n(lad$units), side = 1, at = seq_along(LV), line = 1.1, cex = 0.5,
      col = "#999")
title("Diversity score of the unit the average American lives in",
      cex.main = 0.85, adj = 0, col.main = "#C41230", line = 1.2)
legend("bottomleft", c("observed", "shuffled at random"),
       col = c("#C41230", "#bbbbbb"), lwd = c(2.4, 1.6), lty = c(1, 2),
       bty = "n", cex = 0.6)
```

```{r step8}
c(nation      = FN("div_nation"),
  state       = FN("div_state"),
  county      = FN("div_county"),
  tract       = FN("div_censustract"),
  block       = FN("div_censusblock"))

# the same climb, with the null beside it
o <- lad[, c("level", "H", "H0", "H_net")]
o[, 2:4] <- round(o[, 2:4], 4)
o
```

**The country scores `r FV("div_nation")`; the average American's census tract
scores `r FV("div_censustract")`.** The dashed line is flat at the national
value all the way down to block groups — so the fall is **not** an artefact of
using smaller boxes. Only at the block rung, where a unit holds a few dozen
people, does randomness start producing apparent segregation by itself, and
even there most of the effect survives the control.

Without any index at all:

```{r step8b}
c(pct_in_tract_at_least_80_pct_one_group = FN("pop_tract_80plus"),
  pct_in_tract_at_least_90_pct_one_group = FN("pop_tract_90plus"),
  pct_in_tract_whose_largest_group_under_half = FN("pop_tract_under50"))
```

**`r FV("pop_tract_80plus")`% of Americans live in a census tract where at
least four residents in five are in the same category.**

Compare this with `areal-units`, which runs a similar ladder to make the
*opposite* point: that aggregation destroys tails and flips correlations. That
lab is about the measuring instrument. Step 8 is what is left when you subtract
the instrument.

# Step 9 — One county

```{r step9}
c(wayne_population = FN("wayne_pop"),
  tracts           = FN("wayne_tracts"),
  block_groups     = FN("wayne_bg"),
  blocks           = FN("wayne_blocks"))

# the county row: no majority group at all
round(100 * as.numeric(cou[cou$geoid == "26163", CAT]) /
      cou$total[cou$geoid == "26163"], 1) -> v
setNames(v, SHORT[CAT])

c(diversity_score = FN("wayne_div"),
  rank_of_3143_counties = as.integer(rank(-cou$entropy)[cou$geoid == "26163"]))
```

**Wayne County, Michigan** contains Detroit, has no majority group, and sits in
the most diverse `r pc(100 * rank(-cou$entropy)[cou$geoid == "26163"] / nrow(cou))`%
of American counties. From the county row alone you would call it integrated.

```{r fig-wayne-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
PWs <- 232L; GAPs <- 12L; TOPs <- 34L; BOTs <- 40L
w <- win(wou, pad = 0.03); sc <- PWs / w$s
KEYC <- c(White = "#7F9BB3", Black = "#C41230", Hispanic = "#E08214",
          Asian = "#4D9221", `Two or More` = "#8073AC", AIAN = "#B35806",
          `Some Other Race` = "#999999", NHPI = "#2C7FB8")
lev <- list(
  list(t = "the county", sub = "1 unit", rings = wou, id = "26163",
       largest = FV("wayne_largest"), pct = FN("wayne_largest_pct")),
  list(t = "its tracts", sub = paste(n(FV("wayne_tracts")), "units"),
       rings = wtg, id = wtr$geoid, largest = wtr$largest,
       pct = wtr$largest_pct),
  list(t = "its block groups", sub = paste(n(FV("wayne_bg")), "units"),
       rings = wbr, id = wbg$geoid, largest = wbg$largest,
       pct = wbg$largest_pct))
mkw <- function(i) {
  L <- lev[[i]]; off <- (i - 1L) * (PWs + GAPs)
  fx <- function(x) off + (x - w$x[1]) * sc
  fy <- function(y) TOPs + (w$y[2] - y) * sc
  P <- ringpaths(L$rings, fx, fy); ids <- ringids(L$rings)
  m <- match(ids, L$id)
  cl <- KEYC[L$largest[m]]; cl[is.na(cl)] <- "#eeeeee"
  op <- 0.22 + 0.78 * pmin(pmax((L$pct[m] - 30) / 60, 0), 1)
  op[is.na(op)] <- 0.1
  paste0("[", jstr(P), ",", jstr(unname(cl)), ",",
         jnum(sprintf("%.2f", op)), ",", off, ',"', L$t, '","', L$sub, '"]')
}
PANW <- vapply(1:3, mkw, character(1))
KL <- names(KEYC)[names(KEYC) %in% unique(c(wtr$largest, wbg$largest))]
keyj <- paste(sprintf('{"n":"%s","c":"%s"}', KL, unname(KEYC[KL])),
              collapse = ",")
Ww <- 3L * PWs + 2L * GAPs; Hw <- TOPs + PWs + BOTs
cat(paste0('
<div id="wy" style="margin:1em 0"></div>
<script>
(function(){
const P=[', paste(PANW, collapse = ","), '],K=[', keyj, '];
const W=', Ww, ',H=', Hw, ',PW=', PWs, ';
const svg=d3.select("#wy").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
P.forEach(p=>{
  const paths=p[0],cl=p[1],op=p[2],off=p[3],t=p[4],s=p[5];
  svg.append("g").selectAll("path").data(paths).join("path").attr("d",d=>d)
    .attr("fill",(d,j)=>cl[j]).attr("fill-opacity",(d,j)=>op[j])
    .attr("stroke","#ffffff").attr("stroke-width",0.25);
  svg.append("text").attr("x",off+PW/2).attr("y",15).attr("text-anchor","middle")
    .attr("font-size","12.5px").attr("font-weight","600").text(t);
  svg.append("text").attr("x",off+PW/2).attr("y",27).attr("text-anchor","middle")
    .attr("font-size","10px").attr("fill","#888").text(s);});
let kx=8;
K.forEach(k=>{
  svg.append("rect").attr("x",kx).attr("y",H-24).attr("width",11)
    .attr("height",11).attr("fill",k.c);
  svg.append("text").attr("x",kx+15).attr("y",H-15).attr("font-size","10.5px")
    .attr("fill","#555").text(k.n);
  kx+=26+k.n.length*5.9;});
svg.append("text").attr("x",8).attr("y",H-3).attr("font-size","10px")
  .attr("fill","#999")
  .text("colour = the largest category in the unit; stronger colour = a larger share");
})();
</script>'))
```

```{r fig-wayne-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.0}
w <- win(wou, pad = 0.03)
KEYC <- c(White = "#7F9BB3", Black = "#C41230", Hispanic = "#E08214",
          Asian = "#4D9221", `Two or More` = "#8073AC", AIAN = "#B35806",
          `Some Other Race` = "#999999", NHPI = "#2C7FB8")
lev <- list(
  list(t = "the county", sub = "1 unit", rings = wou, id = "26163",
       largest = FV("wayne_largest"), pct = FN("wayne_largest_pct")),
  list(t = "its tracts", sub = paste(n(FV("wayne_tracts")), "units"),
       rings = wtg, id = wtr$geoid, largest = wtr$largest,
       pct = wtr$largest_pct),
  list(t = "its block groups", sub = paste(n(FV("wayne_bg")), "units"),
       rings = wbr, id = wbg$geoid, largest = wbg$largest,
       pct = wbg$largest_pct))
layout(matrix(1:3, 1, 3))
op <- par(mar = c(0.4, 0.2, 2.6, 0.2), oma = c(2.6, 0, 0, 0))
for (L in lev) {
  cl <- KEYC[L$largest]; cl[is.na(cl)] <- "#eeeeee"
  al <- 0.22 + 0.78 * pmin(pmax((L$pct - 30) / 60, 0), 1); al[is.na(al)] <- 0.1
  fill <- as.list(mapply(adjustcolor, unname(cl), al)); names(fill) <- L$id
  plot(NA, xlim = w$x, ylim = w$y, asp = 1, axes = FALSE, ann = FALSE)
  drawrings(L$rings, fill, border = "#ffffff", lwd = 0.15)
  title(L$t, cex.main = 0.95, line = 1.1)
  mtext(L$sub, side = 3, line = 0.25, cex = 0.55, col = "#888")
}
KL <- names(KEYC)[names(KEYC) %in% unique(c(wtr$largest, wbg$largest))]
par(fig = c(0, 1, 0, 1), oma = c(0, 0, 0, 0), mar = c(0, 0, 0, 0), new = TRUE)
plot(NA, xlim = c(0, 1), ylim = c(0, 1), axes = FALSE, ann = FALSE)
legend("bottom", KL, fill = unname(KEYC[KL]), border = NA, bty = "n",
       cex = 0.55, horiz = TRUE, inset = c(0, 0.005))
par(op); layout(1)
```

```{r step9b}
table(largest_group_in_tract = wtr$largest)

c(median_tract_pct_one_group = FN("wayne_tract_median_largest"),
  tracts_at_least_80_pct_black = FN("wayne_tracts_80_black"),
  tracts_at_least_80_pct_white = FN("wayne_tracts_80_white"),
  pct_of_county_in_those_tracts = FN("wayne_pop_80_pct"),
  tracts_at_least_90_pct_one_group = FN("wayne_tracts_90"),
  pct_of_county_in_those = FN("wayne_pop_90_pct"))
```

**The median Wayne County tract is `r FV("wayne_tract_median_largest")`% a
single category**, and `r FV("wayne_pop_80_pct")`% of the county's residents
live in a tract that is at least 80% one group. The county row and the tract
rows answer different questions, and only the tract rows are about anybody's
street.

# Step 10 — One street

The boundary you can see in the middle panel is Eight Mile Road, the northern
city limit of Detroit. Along this stretch it is **also the county line**, which
means no county-level statistic can show it.

```{r fig-eightmile-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
MW <- 380L; PWp <- 290L; GP <- 34L
rxk <- range(wbk$x); ryk <- range(wbk$y)
scm <- min(MW / diff(rxk), 290 / diff(ryk))
MH <- round(diff(ryk) * scm); T8 <- 42L; B8 <- 40L
W8 <- MW + GP + PWp + 74; H8 <- T8 + MH + B8
fx <- function(x) 8 + (x - rxk[1]) * scm
fy <- function(y) T8 + (ryk[2] - y) * scm
BP <- ringpaths(wbk, fx, fy); bid <- ringids(wbk)
m <- match(bid, wtx$GEOID)
pb <- 100 * wtx$nh_black[m] / ifelse(wtx$total[m] == 0, 1, wtx$total[m])
pb[is.na(pb) | wtx$total[m] == 0] <- NA
DP <- ringpaths(wdt, fx, fy)
PX <- 8 + MW + GP
pro$pw <- 100 * pro$nh_white / pro$total
pro$pb <- 100 * pro$nh_black / pro$total
prj <- paste(sprintf('{"y":%.1f,"b":%.2f,"w":%.2f}',
                     T8 + (ryk[2] - (mean(ryk) + pro$mid)) * scm,
                     pro$pb, pro$pw), collapse = ",")
cat(paste0('
<div id="em" style="margin:1em 0"></div>
<script>
(function(){
const BP=', jstr(BP), ',V=', jnum(ifelse(is.na(pb), -1, round(pb, 1))), ';
const DP=', jstr(DP), ',PR=[', prj, '];
const W=', W8, ',H=', H8, ',PX=', PX, ',PW=', PWp, ',T=', T8, ',MH=', MH, ';
const svg=d3.select("#em").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const ramp=d3.interpolateRgb("#f7f7f7","#C41230");
svg.append("g").selectAll("path").data(BP).join("path").attr("d",d=>d)
  .attr("fill",(d,j)=>V[j]<0?"#e8e8e8":ramp(0.06+0.94*V[j]/100))
  .attr("stroke","#ffffff").attr("stroke-width",0.15);
svg.append("g").selectAll("path").data(DP).join("path").attr("d",d=>d)
  .attr("fill","none").attr("stroke","#111").attr("stroke-width",1.8);
svg.append("text").attr("x",8).attr("y",16).attr("font-size","12.5px")
  .attr("font-weight","600").text("blocks along Eight Mile Road");
svg.append("text").attr("x",8).attr("y",30).attr("font-size","10px")
  .attr("fill","#888").text("shading = share Black");
const x=v=>PX+v/100*PW;
[0,25,50,75,100].forEach(v=>{
  svg.append("line").attr("x1",x(v)).attr("x2",x(v)).attr("y1",T).attr("y2",T+MH)
    .attr("stroke","#eee");
  svg.append("text").attr("x",x(v)).attr("y",T+MH+15).attr("text-anchor","middle")
    .attr("font-size","10px").attr("fill","#aaa").text(v+"%");});
const ln=k=>d3.line().x(d=>x(d[k])).y(d=>d.y).curve(d3.curveStepBefore)(PR);
svg.append("path").attr("d",ln("b")).attr("fill","none").attr("stroke","#C41230")
  .attr("stroke-width",2.4);
svg.append("path").attr("d",ln("w")).attr("fill","none").attr("stroke","#7F9BB3")
  .attr("stroke-width",2.4);
const yline=', sprintf("%.1f", T8 + (ryk[2] - mean(ryk)) * scm), ';
svg.append("line").attr("x1",PX).attr("x2",PX+PW).attr("y1",yline).attr("y2",yline)
  .attr("stroke","#111").attr("stroke-width",1.4).attr("stroke-dasharray","4,3");
svg.append("text").attr("x",PX+PW).attr("y",yline-6).attr("text-anchor","end")
  .attr("font-size","10.5px").attr("fill","#111").text("the city limit");
svg.append("text").attr("x",PX).attr("y",16).attr("font-size","12.5px")
  .attr("font-weight","600").text("composition, by distance from it");
svg.append("text").attr("x",PX+8).attr("y",T+MH-10).attr("font-size","11px")
  .attr("fill","#C41230").attr("font-weight","600").text("Black");
svg.append("text").attr("x",PX+8).attr("y",T+22).attr("font-size","11px")
  .attr("fill","#7F9BB3").attr("font-weight","600").text("White");
})();
</script>'))
```

```{r fig-eightmile-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.4}
rxk <- range(wbk$x); ryk <- range(wbk$y)
layout(matrix(1:2, 1, 2), widths = c(1.25, 1))
op <- par(mar = c(2.6, 0.4, 2.6, 0.4))
ramp <- colorRampPalette(c("#f7f7f7", "#C41230"))(101)
pbv <- 100 * wtx$nh_black / ifelse(wtx$total == 0, 1, wtx$total)
pbv[wtx$total == 0] <- NA
fill <- as.list(ifelse(is.na(pbv), "#e8e8e8", ramp[round(6 + 0.94 * pbv) + 1]))
names(fill) <- wtx$GEOID
plot(NA, xlim = rxk, ylim = ryk, asp = 1, axes = FALSE, ann = FALSE)
drawrings(wbk, fill, border = "#ffffff", lwd = 0.1)
for (p in unique(wdt$part)) {
  z <- wdt[wdt$part == p, ]
  lines(z$x, z$y, col = "#111111", lwd = 1.4)
}
title("blocks along Eight Mile Road", cex.main = 0.85, adj = 0, line = 1.2)
mtext("shading = share Black; black line = Detroit city limit", side = 3,
      line = 0.2, cex = 0.5, col = "#888", adj = 0)
pro$pw <- 100 * pro$nh_white / pro$total
pro$pb <- 100 * pro$nh_black / pro$total
par(mar = c(2.6, 2.6, 2.6, 1.6))
plot(NA, xlim = c(0, 100), ylim = c(-2.5, 2.5), axes = FALSE, ann = FALSE)
abline(v = seq(0, 100, 25), col = "#eeeeee")
ys2 <- c(pro$mid[1] - 0.125, rep(pro$mid, each = 2)[-c(1, 2 * nrow(pro))],
         pro$mid[nrow(pro)] + 0.125)
lines(rep(pro$pb, each = 2), ys2, col = "#C41230", lwd = 2.2)
lines(rep(pro$pw, each = 2), ys2, col = "#7F9BB3", lwd = 2.2)
abline(h = 0, col = "#111111", lty = 2, lwd = 1.2)
axis(1, seq(0, 100, 25), paste0(seq(0, 100, 25), "%"), cex.axis = 0.55,
     col = "#dddddd", col.axis = "#aaa", tck = -0.015, mgp = c(2, 0.3, 0))
axis(2, c(-2.5, 0, 2.5), c("2.5 km south", "the city limit", "2.5 km north"),
     cex.axis = 0.5, col = "#dddddd", col.axis = "#888", tck = -0.01,
     mgp = c(2, 0.4, 0), las = 1)
text(70, -2.0, "Black", col = "#C41230", cex = 0.7, font = 2)
text(70, 2.0, "White", col = "#7F9BB3", cex = 0.7, font = 2)
title("composition, by distance from it", cex.main = 0.85, adj = 0, line = 1.2)
par(op); layout(1)
```

```{r step10}
c(blocks_in_the_profile = FN("profile_blocks"),
  people_in_the_profile = FN("profile_pop"),
  pct_black_south_of_the_line = FN("prof_south_black"),
  pct_black_north_of_the_line = FN("prof_north_black"),
  pct_black_last_500m_inside   = FN("prof_in500_black"),
  pct_black_first_500m_outside = FN("prof_out500_black"),
  the_step_in_points = FN("prof_jump_black"))
```

**Across one road the Black share falls `r FV("prof_jump_black")` points.**

Look at the *shape* rather than the levels. It is flat, then a step, then flat
again. Composition does not fade with distance from Detroit the way something
diffusing would; it changes at the boundary and stops. Whatever produced this
was operating on the line itself — on which side of a municipal border a house
sits — and not on distance from a city centre.

# Step 11 — What you can now say

**You can say** what the United States looked like in 2020, in eight
categories, and give the numbers.

**You can say** that "the Black population" and "the White population" are
quantities with at least two correct values each, name the gap
(`r FV("us_aian_ratio")`× for American Indian and Alaska Native,
`r FV("us_white_ratio")`× for White), and explain which OMB revision created
the ambiguity and when.

**You can say** that the 2010-to-2020 change in the multiracial and Some Other
Race counts is largely a measurement change, and cite the Bureau saying so.

**You can say** that the diversity of the unit the average American lives in
falls from `r FV("div_nation")` to `r FV("div_censustract")` between the nation
and the census tract, **and** that a random-shuffle control rules out the
obvious objection, because the null stays flat at `r FV("H0_censustract")`
through the tract rung.

**You cannot say** why anyone lives where they live. Nothing here has a
mechanism in it.

**You cannot say** that any of these numbers is definition-free. Change the
categories — which OMB has now done — and every figure moves.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different county.** `data/counties.csv` holds all
`r n(FV("n_counties"))` counties with the eight categories and a diversity
score. Find one whose county-level score is high and say what you would need in
order to know whether it is actually integrated. Then find a county whose score
is *low* and ask whether the county row is misleading in the other direction.

**Option B — A different definition.** Step 3 gives every race two counts.
Recompute the state table in Step 6 using `black_any` instead of `nh_black`,
and report which states move most. Which definition would you defend in a
voting-rights case, and why?

**Option C — A different threshold.** Step 8 reports the share of Americans in
a tract that is at least 80% one group. `data/tract_largest.csv` has the whole
distribution in ten-point bins. At what threshold does the statistic stop
sounding alarming, and what does the fact that you can choose tell you about
the statistic?

**Option D — The 2024 categories.** The revised Directive 15 has
`r std$categories[std$key == "2024"]` categories rather than eight, and moves
Hispanic origin into the race list. Using `data/categories.csv`, work out which
of this lab's numbers would change, which would not, and which are impossible
to predict without new data.

# What this lab cannot tell you

- **Whether the categories are the right ones.** They are administrative
  categories set by OMB and implemented by the Census Bureau. Every figure here
  is downstream of them.
- **How the country has changed.** See Step 4. The data will not carry a clean
  time series and the honest thing is to say so.
- **Anything exact about a single block.** The 2020 census applied a
  disclosure-avoidance system based on differential privacy. State totals were
  held invariant; **block-level counts were not**, and the noise is
  proportionally largest where units are smallest.
- **Anything about voters.** These are residents, including children and
  non-citizens. Voting-age population is in the same file and gives different
  numbers.
- **Whether Wayne County is typical.** It was chosen because its county row is
  misleading in an instructive direction. Step 8 is the national check that
  stops the county example doing work it cannot do.

# On the discussion board

By **Friday, 11:59 p.m.**

- Step 3 shows the same population producing two correct counts. If a news
  story reports "the Black population of the United States," what is the one
  question you now have to ask before believing the number?
- Step 4 shows a decade of apparent demographic change that the Census Bureau
  attributes largely to a better form. Whose fault is a headline like "the
  White population declined for the first time in history," and what should the
  Bureau have done differently — if anything?
- Step 8 shows the country scoring `r FV("div_nation")` and the average
  American's neighbourhood scoring `r FV("div_censustract")`. Which of those
  two numbers should appear in a news story about American diversity, and does
  your answer change if the story is about school assignment rather than
  national identity?

***

### Where the data came from

- **Race and Hispanic origin at every level** — U.S. Census Bureau, *2020
  Census Redistricting Data (P.L. 94-171) Summary File*, legacy format, Tables
  P1 and P2. The national roll-up plus all 51 state files, giving the nation,
  4 regions, 9 divisions, 51 states, `r n(FV("n_counties"))` counties,
  `r n(FV("n_tracts"))` tracts, `r n(FV("n_blockgroups"))` block groups and
  `r n(FV("n_blocks"))` blocks. Fetched `r FV("fetched")`. Keyless; **this lab
  uses no Census API and contains no API key.**
  <https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/>
- **The 2010 comparison** — the same file for 2010, national roll-up only.
  <https://www2.census.gov/census_2010/redistricting_file--pl_94-171/National/>
- **Boundaries** — Census TIGER/Line 2020: states, and for Michigan the tracts,
  block groups, blocks and place boundaries.
  <https://www2.census.gov/geo/tiger/TIGER2020/>
- **The categories** — Office of Management and Budget, Statistical Policy
  Directive No. 15 (1977), `r std$citation[std$key == "1997"]` (1997), and
  `r std$citation[std$key == "2024"]` (2024).
- **The measure** — Theil, H. (1972), *Statistical Decomposition Analysis*;
  Reardon, S. F. and Firebaugh, G. (2002), "Measures of Multigroup
  Segregation," *Sociological Methodology* 32(1).

The build is `data/build-data.R`. It records its source URLs, fetch date and
row counts, checks that the eight categories partition the population at every
level of every state, and stops if a state's blocks do not sum to the state.
