---
title: "The Age Structure of the Electorate, 1966–2024"
type: brief
subtitle: "Who is old enough to vote, who says they voted, and sixty years of the gap between the two"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("age-structure-code.R")
```

```{r setup, include=FALSE}
```

Older Americans run American elections. You have heard the sentence, or one of its relatives: the electorate is grey, the young do not vote, politicians answer to the people who show up. It is repeated as though it were a permanent fact about the country. Is it true — and if it is, how much more do older Americans actually vote, and has the gap always been there?

Start with a distinction the word "electorate" hides. It sometimes means everybody entitled to vote, and it sometimes means everybody who did. Those are different populations. In `r LP`, **`r mil(VAPL)`** Americans were old enough to vote and **`r mil(VOTL)`** told the Census Bureau they had — `r p1(TURNL)`%. The clearest way to see how different the two groups are is to draw them as two pyramids on one axis, and to watch the pyramids for sixty years.

## The test

The data is the Census Bureau's [Current Population Survey, Voting and Registration Supplement](https://www2.census.gov/programs-surveys/cps/tables/time-series/voting-historical-time-series/hst_vote01.xlsx) — historical Table A-1, whose age blocks cover `r NEL` federal elections from `r FIRST` to `r LAST`. The Current Population Survey is the Bureau's monthly survey of households, best known as the source of the unemployment rate. After every federal election since 1964, it has asked the roughly 60,000 households in that month's sample whether the adults in them voted, and the Bureau publishes what they said, by age.^[U.S. Census Bureau, "About the Current Population Survey," <https://www.census.gov/programs-surveys/cps/about.html>, and the voting supplement's own page, <https://www.census.gov/topics/public-sector/voting/about.html>.]

Be clear about what kind of source that is. The same Bureau runs the decennial census, which is an enumeration: an attempt to count everyone, one by one. The CPS is the other instrument: a survey, a sample of households chosen to stand in for the country. Its answers are weighted, which means counting some responses more than others so the sample matches the population.

Nobody's voting answer is checked against a record; every "voted" in this brief is a self-report — a person's own account, taken at its word. This brief sits in the survey section because that is what the source is, whoever publishes it. [The surveys chapter](../surveys-source/surveys-source-brief.html) lays out what a sample can establish that a count cannot, and this series is a sixty-year example.

Why this source, of all sources? Because nothing else reaches. No administrative record pairs a national age profile of the adult population with an age profile of its voters, election after election, back to the 1960s. The CPS supplement does exactly that, in one continuously published table, from the agency that also measures the population itself.

## The data

The workbook is not one table. It is ten tables stacked in a single spreadsheet column: five age groups (the total, then four bands), each appearing twice, once for voting and once for registration. Pulled apart and reassembled, those ten blocks become two flat files.

`bands.csv` holds one row per year per age band:

```{r varmap}
```

The Bureau publishes the population and the percentage; the count of voters is the two multiplied, which is arithmetic on its own figures rather than an estimate of ours. Here is `r LP`:

```{r one-row}
```

`shares.csv` is derived from it: each band's `pop_share` of everybody old enough to vote, its `vote_share` of everybody who said they voted, and `gap`, the second minus the first. The four bands are published separately from the total, so they can be checked against it:

```{r checks}
```

> **Before you read on, commit to a view.** People aged 65 and over are
> `r p1(OPOP)`% of American adults. **What share of the votes do you think they
> cast?** Write it down, and write down what you would have guessed for 1968.

## What it says about democracy

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The adult population and the voters, `r LP`.** Two pyramids on one axis, because the question is whether the second population is a scale model of the first, and a pyramid shows the answer band by band. Grey to the left is each band's share of everybody old enough to vote; teal to the right is its share of everybody who said they voted. Drag the year; the button switches from shares to counts of people.

The two sides are not the same shape. **People aged 65 and over are `r p1(OPOP)`% of the adult population and `r p1(OVOTE)`% of the voters.** People aged 18 to 24 are `r p1(YPOP)`% of adults and `r p1(YVOTE)`% of voters. Compare that with what you wrote down.

The mechanism is not mysterious — it is turnout. `r p1(OTURN)`% of the oldest band said they voted against `r p1(YTURN)`% of the youngest, a ratio of `r p1(RATIO)` to one. Some of that gap opens before the ballot: `r p1(OREG)`% of the oldest band were registered against `r p1(YREG)`% of the youngest, so roughly half the difference is a registration difference rather than a turnout one.

So the claim survives its first test. The people who show up are older than the country that could have shown up, by a margin large enough to see from across the room. The second question is whether it was always so, and the answer is the finding of this brief.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Over-representation, presidential elections `r min(PRES)` to `r LP`.** One line per band: its share of the voters minus its share of the adults, so zero means a band cast exactly its share of the votes. A line, rather than more pyramids, because the question here is change over time.

The oldest band starts at **`r p1(OGAP1)`** in `r min(PRES)`, meaning it voted almost exactly in proportion to its size. It ends at **+`r p1(OGAP)`**. The over-representation of older voters is not an old feature of American elections. It is roughly as old as the people currently complaining about it.

Two things drove it and they are easy to confuse. The band grew: 65-and-over went from `r p1(OPOP1)`% of adults in `r min(PRES)` to `r p1(OPOP)`% now, which is demography and would have raised their share of the votes even if nothing about their behaviour had changed. And on top of that their turnout advantage widened. **The first is a fact about the country. The second is a fact about its politics.** A count of votes cast cannot separate them, which is why Figure 2 is drawn as a difference between two shares.

The youngest band, meanwhile, is roughly where it was. It has been between `r p1(min(sh$gap[sh$band == "18 to 24 Years"]))` and `r p1(max(sh$gap[sh$band == "18 to 24 Years"]))` points under-represented in every presidential election in the series. Whatever has changed about young Americans and voting, their standing relative to their own numbers is not it.

One property of Figure 2 is worth knowing before you lean on it. Suppose turnout rose by ten points in every age band at once: nothing in the figure would move, because a uniform rise cancels out of a share. That is what makes the lines readable across high-turnout and low-turnout years, and it is also what makes them blind to which years those were. Figure 1's counts button is there for that.

## What this brief cannot tell you

**These are people who said they voted.** The survey asks after the election, and nobody checks. The [validated-turnout chapter](../../06-putting-data-together/validated-turnout/validated-turnout-brief.html) establishes that the answer is consistently too high — in `r LP` the survey implies more voters than there were ballots. If over-reporting differs by age, and there is good reason to think it does,^[When survey answers are matched against the official record of who voted, the people who claim a vote they did not cast turn out to resemble actual voters — more educated, more partisan, more engaged — so the over-report is not spread evenly across groups. Stephen Ansolabehere and Eitan Hersh, "Validation: What Big Data Reveal About Survey Misreporting and the Real Electorate," *Political Analysis* 20, no. 4 (2012): 437–459, <https://doi.org/10.1093/pan/mps023>.] then the right-hand pyramid is distorted in a way none of these figures can correct.

**Four bands is a coarse pyramid.** A conventional population pyramid has five-year bands and splits by sex; this one has four unequal bands and no sex, because that is what the historical table publishes. The 25-to-44 band spans twenty years and the 65-and-over band has no upper bound, so both hide structure — the oldest band in particular mixes people at very different turnout rates. And the youngest band changes its meaning once. Before 1972 the voting age was 21 in most states, so the youngest block for the earlier elections counts mostly 21-to-24-year-olds; the Twenty-sixth Amendment, ratified in 1971, lowered the age to 18 everywhere,^[U.S. Const. amend. XXVI, text at <https://www.law.cornell.edu/constitution/amendmentxxvi>.] and the band's population jumps from `r mil(bd$vap[bd$year == 1968 & bd$band == "18 to 24 Years"])` in 1968 to `r mil(bd$vap[bd$year == 1972 & bd$band == "18 to 24 Years"])` in 1972 for that reason. Figure 2's first point for that band describes a narrower group than its label.

**Nothing here is about a state.** These are national figures. The [turnout-denominator chapter](../../06-putting-data-together/turnout-denominator/turnout-denominator-brief.html) shows how much states differ on the denominator alone, the bottom number a turnout rate is divided by. An age pyramid for Florida would not look like one for Utah.

**Being over-represented is not the same as being decisive.** A band that casts `r p1(OVOTE)`% of the votes decides nothing on its own, and this brief has said nothing about how any band voted. Who shows up is prior to, and separate from, what they do when they get there.

## What you should have learned

- In `r LP`, people 65 and over were `r p1(OPOP)`% of American adults and `r p1(OVOTE)`% of the people who said they voted; people 18 to 24 were `r p1(YPOP)`% of adults and `r p1(YVOTE)`% of voters.
- The mechanism is turnout — `r p1(OTURN)`% against `r p1(YTURN)`%, a ratio of `r p1(RATIO)` to one — and roughly half of that gap is a registration gap.
- The oldest band's over-representation is new: `r p1(OGAP1)` points in `r min(PRES)`, +`r p1(OGAP)` points in `r LP`. Part of that is the band growing, part is its turnout advantage widening, and the two are different kinds of fact.
- The CPS supplement is a survey: a weighted sample of self-reports, published by the Census Bureau but not a count, and every "voter" in it is a respondent's claim.

## Extensions

- `shares.csv` carries a `gap` for every year and age band. Which band's gap has moved the most across the series, and which has barely moved?
- `bands.csv` reports `pct_registered` and `pct_voted` separately. Work out what share of the registered actually voted, band by band. Do the youngest fail to register, or fail to turn out?
- Pick a presidential year in `bands.csv` and the midterm beside it. Which age band falls away the most between the two?
- Everything here is people saying they voted. Add up `voters` across the bands in `bands.csv` and set the total against a turnout figure from anywhere else. Which way does the error run?
- *Stretch:* the Bureau publishes the same table for single states in recent years. Build Florida's two pyramids and Utah's, and see how far a national figure can mislead about either.
- *Stretch:* the CPS microdata records age in years, not bands. Sketch what Figure 2 would look like with a 65-to-74 line and a 75-and-over line, and say which result above you would expect to change.

## Sources

**Data**

- U.S. Census Bureau. *Current Population Survey, Voting and Registration
  Supplement*, Table A-1, historical time series, November 1964 to 2024.
  <https://www2.census.gov/programs-surveys/cps/tables/time-series/voting-historical-time-series/hst_vote01.xlsx>
- The same workbook is read for its national totals by this book's
  [validated-turnout chapter](../../06-putting-data-together/validated-turnout/validated-turnout-brief.html), which
  documents what the table does and does not publish. The four age blocks
  further down the same sheet are read here, and checked to see that they
  reconstruct the published total.
- U.S. Census Bureau. "About the Current Population Survey."
  <https://www.census.gov/programs-surveys/cps/about.html>
- Ansolabehere, Stephen, and Eitan Hersh (2012). "Validation: What Big Data
  Reveal About Survey Misreporting and the Real Electorate." *Political
  Analysis* 20(4): 437–459. <https://doi.org/10.1093/pan/mps023>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`bands.csv`](data/derived/bands.csv), [`shares.csv`](data/derived/shares.csv)

Two more files here are not data about the subject at all -- they are how this chapter checked its own work: [`facts.csv`](data/derived/facts.csv), the individual numbers the text above quotes, and [`checks.csv`](data/derived/checks.csv), the arithmetic verified before anything was drawn.

```{r ai-prompt, echo=FALSE, results="asis"}
```
