---
title: "Two Denominators for Turnout"
type: brief
subtitle: "One numerator, two denominators, and a thirty-year decline that is not in the data"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("turnout-denominator-code.R")
```

```{r setup, include=FALSE}
```

`r n(N24$TOTAL_BALLOTS_COUNTED)` ballots were counted in the November 2024 general election. Nobody disputes that number, and it will not change again. Divide it by one population and American turnout in 2024 was **`r f1(N24$rate_vap_tb)`%**. Divide it by another and it was **`r f1(N24$rate_vep_tb)`%**.

So what *is* the turnout rate? The question is not pedantic. A turnout rate is how a country grades its own democracy, and the American figure is the one quoted in every league table that ranks the United States against other rich democracies. Both populations below are published by serious people, both are called something reasonable, and both describe the United States in November 2024. The gap between the two answers is `r f1(GAP24)` points, worth `r n(EXTRA)` ballots.

## The test

This brief joins one numerator, the top number of a fraction, to two competing denominators. A denominator is the bottom number, the thing you are dividing by. The ballots and the populations come from the same two files, keyed on the election year for the national series and on state and year for the fifty states.

The 1948–2000 half is Michael McDonald and Samuel Popkin's Table 1 from *The Myth of the Vanishing Voter*, printed on page 966 of the December 2001 *American Political Science Review* (<https://www.jstor.org/stable/3117725>). The 1980–2024 half is McDonald's maintained series at the United States Elections Project, published as CSV under a Creative Commons licence (<https://election.lab.ufl.edu/data-downloads/>).

Those two sources answer the question because they are the only ones that print the ballots next to *both* candidate denominators for the same elections. Anyone can recompute either rate and compare. The join runs on a well-behaved key: a year and a state name are about as safe as keys get, which is why the interesting failure here is not in the join. [Join Keys and Crosswalks](../data-sources/data-sources-brief.html) is the chapter on what a key can promise, and on the crosswalk, a table that translates one set of identifiers into another when two files share no key.

## The data

Five tables carry this brief. `mp1948.csv` holds the article's own table, `r cv("Table 1 data rows parsed")` rows covering `r cv("Table 1 election years")`, with `year`, `vote_highest`, `vap`, `vap_rate`, the three adjustments (`noncitizens`, `felons`, `overseas`), `vep`, `vep_rate` and `inelig_pct`. `national.csv` and `states.csv` hold the modern series, `r cv("Elections Project rows, 1980-2024")` rows in all, with `YEAR`, `STATE`, `TOTAL_BALLOTS_COUNTED`, `VOTE_FOR_HIGHEST_OFFICE`, `VAP`, `VEP`, `NONCITIZEN_PCT`, `INELIGIBLE_FELONS_TOTAL`, `ELIGIBLE_OVERSEAS` and the four rates this brief computes from them. `state_trends.csv` fits both trends per state and flags `sign_flip`; `pairs2024.csv` compares every pair of states on both denominators.

The voting-eligible population is not a mysterious quantity. It is the voting-age population, minus the people in it who cannot vote, plus the people who can vote and are not in it.

```{r vep-def}
```

Two construction decisions matter more than the rest.

**The older half is typesetting, not a file.** There is no dataset behind the 1948–2000 series. The numbers exist as an arrangement of characters on a printed page, and the build recovers them by reading the converted text. Public is not the same as usable: a number can be authoritative, cited for twenty-five years, and still not be a dataset. Two independent properties of the published table are checked against each other, and all `r cv("Table 1 data rows parsed")` rows reconstruct within `r cv("largest disagreement between rebuilt and printed VEP rate, pts")` of a percentage point.

**The modern file carries two numerators, and they are never spliced.** `VOTE_FOR_HIGHEST_OFFICE` counts votes for the top statewide contest; `TOTAL_BALLOTS_COUNTED` counts every ballot, including those where the voter skipped that contest. The second runs two to three per cent higher, and the two cover different years. Joining them end to end inserts a jump larger than most of the effects below, so every trend here is computed on one numerator at a time. That is why the state analysis stops at 2016 rather than 2024.

> **Before you read on, commit to a number.** Divided by the voting-age
> population, presidential turnout fell **`r f2(abs(B_VAP[1]))` points per
> election** across 1972–2000. Divided by the voting-*eligible* population,
> with the same elections and the same ballots, **how much did it fall?**
> Half as fast? A third?

## What it says about democracy

**`r sg(B_VEP[1], 3)` points per election.** Not half, and not a third. The corrected series falls by `r f2(abs(B_VEP[1] * (nrow(W72) - 1)))` of a percentage point across the whole of 1972 to 2000, with a standard error `r n(abs(B_VEP[2]/B_VEP[1]))` times the estimate itself. It cannot be told apart from a flat line.

```{r trend-tab}
```

That matters because of what the falling version was used for. By the late 1990s the decline of American voter participation was, in the words of the standard text on the subject, the most analyzed trend in recent American political history. Turnout had run near `r f1(max(PRES$vap_rate))`% in `r PRES$year[which.max(PRES$vap_rate)]` and was down to `r f1(PRES$vap_rate[PRES$year == 1996])`% by 1996. Television had done it, or negative advertising, or the collapse of party organizations, or the decline of social capital. Every one of those explanations was an explanation of the same measurement.

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

**Figure 1. Presidential turnout, 1948–2000, the same ballots over two denominators.** Two lines on one axis is the right form here because the two series share a numerator exactly: any daylight between them is the denominator and nothing else. Before 1972 they are nearly the same line. After it they separate, and the red line goes down while the blue one does not.

The reason the choice changes a trend rather than only a level is that the error is not a constant. If the voting-age population had always overstated the electorate by the same proportion, every year's rate would sit too low by the same amount and the slope would be untouched.

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

**Figure 2. The share of the voting-age population that cannot vote, 1948–2024.** There are no ballots in this figure at all; it is a statement about who lives here. The share runs about `r f1(min(SH_MP$inelig_pct))`% at the low point in the 1960s and `r f1(SH_EP$inelig_pct[SH_EP$year == 2024])`% in 2024. A growing correction subtracted from a flat series *is* a declining series.

The vanishing voter was never a fact about voters. It was a fact about immigration, incarceration and Census arithmetic, entering the literature through the bottom of a fraction that nobody was reading. Here is the same arithmetic for one recent election.

```{r decomp}
```

Noncitizens are `r f1(100 * NC24 / (NC24 + N24$INELIGIBLE_FELONS_TOTAL))`% of everything subtracted, which is why the correction tracks immigration so closely and why it grew when it did.

Noncitizens are also not spread evenly. In 2024 they are `r f1(max(st$NONCITIZEN_PCT[st$YEAR == 2024]))`% of the voting-age population in `r st$STATE[st$YEAR == 2024][which.max(st$NONCITIZEN_PCT[st$YEAR == 2024])]` and `r f1(min(st$NONCITIZEN_PCT[st$YEAR == 2024]))`% in `r st$STATE[st$YEAR == 2024][which.min(st$NONCITIZEN_PCT[st$YEAR == 2024])]`. Fit the same two lines for each state across the `r TX$n` presidential elections from 1980 to 2016 and `r cv("states whose 1980-2016 presidential trend changes sign")` states come out with turnout falling on one denominator and rising on the other.

```{r flip-tab}
```

Texas is the clearest case, falling `r f1(abs(TX$fit_vap))` points against everyone of voting age and rising `r f1(TX$fit_vep)` points against everyone allowed to vote. Now the honest part, which is the more useful lesson of the two. **Not one of the `r 2 * nrow(FLIP)` trends in that table is statistically distinguishable from zero.** `r P_SIG` of the `r 2 * nrow(tr)` state trends across all `r nrow(tr)` jurisdictions do clear the conventional threshold, so this is not a claim that ten elections can never establish anything. The sign flips where neither slope is well determined, which is exactly where you would expect it.

That is the finding, not a caveat on it. The denominator decides which of two unsupported stories the data appears to tell, and a reader shown only one line has no way to know the other exists.

One comparison carries no such uncertainty, because it is two divisions rather than two estimates of a slope. Of the `r cv("state pairs compared in 2024")` distinct pairs of states in 2024, `r cv("pairs that reverse under the two denominators")` rank one way under one denominator and the opposite way under the other.

```{r pair-tab}
```

`r TOP$state_b` beats `r TOP$state_a` by `r f1(TOP$vap_b - TOP$vap_a)` points, or loses to it by `r f1(TOP$vep_a - TOP$vep_b)` points, depending on which population goes underneath. Both statements are computed from the same certified ballot counts.

```{r fig3-d3, results="asis", eval=knitr::is_html_output()}
```

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 3. Every state in 2024, measured both ways.** Every state sits above the diagonal, because every state has somebody in it who cannot vote. What varies is how far above, from `r f1(min(S$rate_vep_tb - S$rate_vap_tb))` points in `r S$STATE[which.min(S$rate_vep_tb - S$rate_vap_tb)]` to `r f1(max(S$rate_vep_tb - S$rate_vap_tb))` points in `r S$STATE[which.max(S$rate_vep_tb - S$rate_vap_tb)]`. A ranking compares vertical positions, and the vertical positions have been shifted by different amounts.

## What this brief cannot tell you

**The denominator is an estimate, not a count.** Voting-age population comes from Census projections between the decennial counts, and the noncitizen share from the American Community Survey. The felon population comes from Bureau of Justice Statistics series, and the overseas figure is projected forward from a single survey. None of them carries anything saying how far off it could be. The voting-eligible population is a better denominator, not a measured one.

**None of the five sign changes rests on a significant trend.** They show how sensitive a conclusion is to a choice of denominator. They do not establish that turnout rose anywhere, and the claim that Texas turnout fell is not supported either.

**The better denominator is still wrong.** It subtracts people in prison, on probation and on parole where state law bars them from voting. It does not subtract people who have completed a sentence in full and remain barred, because the maintainer reports having no reliable way to estimate that population and says so in the documentation. Residency requirements are not adjusted for by either source.

**And the 2024 figures are a version 0.x release.** Some state ballot counts in that file are visibly rounded estimates rather than certifications, and the publisher still calls the file preliminary.

## What you should have learned

- The same `r n(N24$TOTAL_BALLOTS_COUNTED)` ballots give a 2024 turnout rate of `r f1(N24$rate_vap_tb)`% or `r f1(N24$rate_vep_tb)`%, `r f1(GAP24)` points apart, worth `r n(EXTRA)` ballots, depending only on which population sits underneath.
- Presidential turnout from 1972 to 2000 falls `r f2(abs(B_VAP[1]))` points per election on the voting-age population and `r sg(B_VEP[1], 3)` on the voting-eligible population, so an entire research literature was explaining a denominator.
- The correction is not a level shift: the share of voting-age adults who cannot vote runs from about `r f1(min(SH_MP$inelig_pct))`% in the 1960s to `r f1(SH_EP$inelig_pct[SH_EP$year == 2024])`% in 2024, and a growing correction turns a flat series into a declining one.
- `r cv("states whose 1980-2016 presidential trend changes sign")` states change the direction of their turnout trend with the denominator, and `r cv("pairs that reverse under the two denominators")` of `r cv("state pairs compared in 2024")` state pairs swap places in a 2024 ranking.
- A denominator is a claim about who counts, every such claim is contestable, and the honest move is to say which one you used and why.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `state_trends.csv` carries `b_vap`, `b_vep` and `sign_flip`. Sort by `divergence` and read the twenty states where the two denominators disagree most, whether or not the sign changes.
- `pairs2024.csv` has `vap_a`, `vap_b`, `vep_a`, `vep_b` and `swing`. Sort by `swing` and list the ten pairs the choice of denominator reverses most sharply. Do the same states keep appearing?
- `mp1948.csv` shows the adjustments one at a time in `noncitizens`, `felons` and `overseas`. Work out which one matters most in 1960 and which matters most in 2000.
- `states.csv` has `NONCITIZEN_PCT` and `inelig_pct` for every state and year. Rank the states by how much their correction grew between 1980 and 2024.
- **Stretch.** Several other democracies publish turnout over a registered-voter denominator rather than an age or eligibility one. Find two, and work out which American figure their number should be compared against.
- **Stretch.** The Sentencing Project counts Americans who have completed a felony sentence and remain barred from voting. Download that number, subtract it from the voting-eligible population, and see how far the 2024 rate moves.

## Sources

**Data**

- McDonald, Michael P., and Samuel L. Popkin. 2001. "The Myth of the Vanishing
  Voter." *American Political Science Review* 95 (4): 963–974.
  <https://www.jstor.org/stable/3117725>
  Table 1, page 966, captured with `pdftotext -layout` into
  `mp2001-table1.txt`. `r cv("Table 1 data rows parsed")` election
  years, `r cv("Table 1 election years")`. See `README.txt` for what
  was and was not altered.
- McDonald, Michael P. *1980–2022 General Election Turnout Rates*, version 1.2.
  United States Elections Project, University of Florida. CC BY 4.0.
  <https://election.lab.ufl.edu/data-downloads/turnoutdata/Turnout_1980_2022_v1.2.csv>
- McDonald, Michael P. *2024 General Election Turnout Rates*, version 0.4.
  United States Elections Project, University of Florida.
  <https://election.lab.ufl.edu/data-downloads/turnoutdata/Turnout_2024G_v0.4.csv>
  Both files' documentation is downloaded alongside them and quoted above.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`mp1948.csv`](data/derived/mp1948.csv), [`national.csv`](data/derived/national.csv), [`pairs2024.csv`](data/derived/pairs2024.csv), [`state_trends.csv`](data/derived/state_trends.csv), [`states.csv`](data/derived/states.csv)

**Also cited**

- Rosenstone, Steven J., and John Mark Hansen. 1993. *Mobilization,
  Participation, and Democracy in America.* New York: Macmillan — the
  characterisation of the decline as the most analyzed trend in recent American
  political history, quoted at the opening of the 2001 article.
- Teixeira, Ruy. 1992. *The Disappearing American Voter.* Washington, DC:
  Brookings Institution — the case for the voting-age population as the
  denominator, which the 2001 article argues against.
- Putnam, Robert D. 2000. *Bowling Alone.* New York: Simon and Schuster —
  one of the explanations offered for the decline that turns out not to need
  explaining.

```{r ai-prompt, echo=FALSE, results="asis"}
```
