---
title: "The Cost of Voting Index"
type: brief
subtitle: "Fifty-six election laws compressed into one number, and the decisions that decide what it says"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("cost-of-voting-code.R")
```

```{r setup, include=FALSE}
```

A voter in Oregon is mailed a ballot. It arrives unasked for, it carries prepaid postage, and it can go in a drop box or in the mail. Nobody checks identification, because nobody meets the voter.

A voter in `r HARD` is not mailed a ballot. Voting absentee there means qualifying under a fixed list of excuses, requesting the ballot, enclosing a copy of an identification document, and getting the envelope notarized. Voting in person means producing a photo ID, having registered thirty days earlier.

Those are two different acts wearing the same name. Is it genuinely harder to vote in some states than in others, and how would you measure that? Both voters are casting the same federal vote, and the difference between them is written entirely in state law.

## The test

Somebody has already tried to measure it. The Cost of Voting Index reads the election law of all fifty states, scores it against a fixed list of considerations, and compresses the result into one value per state per election: an **index**, many rules folded into one number. It was built by the political scientists Quan Li, Michael Pomante and Scot Schraufnagel, who published it in 2018 with scores for every presidential election back to 1996,^[Quan Li, Michael J. Pomante II and Scot Schraufnagel, "Cost of Voting in the American States," *Election Law Journal* 17, no. 3 (2018): 234–247, doi:10.1089/elj.2017.0478.] and Pomante has since extended it through 2024.^[Michael J. Pomante II, "Cost of Voting in the American States: 2024," *Election Law Journal* 24, no. 1 (2025): 52–61, doi:10.1089/elj.2024.0037.] The scores, the laws behind them and the codebook are published by the authors at <https://costofvotingindex.com>.

This brief joins three sources on one key: the state, in an election year. The index supplies the laws and the score. Turnout comes from the United States Elections Project's turnout tables, maintained by the political scientist Michael McDonald at the University of Florida and downloadable at <https://election.lab.ufl.edu/data-downloads/>, measured over the voting-eligible population, meaning adult citizens less those a state's felony law excludes. Party control of state government is not in the index at all. For 1996 through 2008 it is the political scientist Carl Klarner's state partisan balance measures, as carried in Michigan State University's Correlates of State Policy collection, downloadable at <https://ippsr.msu.edu/public-policy/correlates-state-policy>; later elections are coded from governors and chamber majorities. A *trifecta*, below, is one party holding the governorship and both chambers of the legislature.

This index can answer the question, where most cannot, because it publishes its own ingredients. The laws, the codebook and the component weights are all on the site, so the number can be rebuilt rather than taken on trust. Every join here inherits the assumptions of both files, plus one more about how they line up. [Join Keys and Crosswalks](../data-sources/data-sources-brief.html) is the chapter on that third assumption, and on the crosswalk, a table that translates one set of identifiers into another.

## The data

`r fact("n_state_years")` rows: fifty states, `r fact("n_years")` elections. Each row carries the state's `final` score, its `final_rank`, and the ten *issue areas* the score is built from. `covi_2024.csv` holds one election with `vep_turnout` and `control` joined on; `covi_panel.csv` holds them all.

```{r one-row}
```

Each of those columns is a different kind of thing, which is the first thing to notice about the file. **`reg. deadline` and `early voting` are numbers of days.** **`voter ID` is a five-point scale**, from no identification required to strict photo. **`absentee` and `inconveniences` are counts of restrictions**, each one a law that either exists in a state or does not. Everything is oriented so that bigger is worse. Underneath the ten columns sit `r nrow(it)` individual yes-or-no items, listed in `items_2024.csv`, and the additive issue areas are those items added up.

The recipe for turning them into one number is published. Standardize the ten issue areas, which means putting each on the same scale by subtracting its fifty-state average and dividing by its spread, so that a column measured in days and a column measured on a five-point scale count alike. Run a principal component analysis, which takes ten correlated columns and finds the directions along which states differ most, and keep the first four of those directions, the components. Then add them up, weighting each by how much of the variation it accounts for, which means counting some parts of the score for more than others.

Standardizing is the step to hold onto. Take `r HARD`, the hardest state in the file. Its registration deadline is `r MSR$reg_deadline` days before the election, against a fifty-state average of `r pc(mean(d$reg_deadline), 1)` days with a spread, a standard deviation, of `r pc(sd(d$reg_deadline), 1)`; subtract the average and divide by the spread and the deadline becomes `r pc(Z("reg_deadline", MSR))`, about one spread above the average state. Its voter-identification score is `r MSR$voter_id` on the five-point scale, against an average of `r pc(mean(d$voter_id), 2)` and a spread of `r pc(sd(d$voter_id), 2)`, so standardized it is `r pc(Z("voter_id", MSR))`. On their own scales a `r MSR$reg_deadline` dwarfs a `r MSR$voter_id`; standardized, both say the same kind of thing, how unusual the state is, in the same units. The principal component analysis then combines the ten standardized columns into four weighted sums, adds those, and for `r HARD` the arithmetic gives `r pc(MSR$final)`, the highest score of the fifty.

```{r rebuild-check}
```

**The index rebuilds exactly** from the published laws. Every state lands in the same rank, and the largest disagreement is `r fact("repro_max_difference")` on a scale running from about −2.5 to 2. That matters below, because when this brief re-weights the index it re-weights the real thing rather than a model of it.

> **Before you read on, commit to a number.** The index weights its four
> components by variance. Suppose you weighted them equally instead, a rule at
> least as easy to defend. **How many places would the worst-affected state
> move?** Write down a number out of 50.

## What it says about democracy

The states really are far apart. `r EASY_NOW` scores `r pc(min(d$final))` and `r HARD` scores `r pc(max(d$final))`, and the spread between them is the largest thing in the file.

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=6.4}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The 2024 Cost of Voting Index, all fifty states.** Lower is easier. A sorted bar per state is the right form because the data is fifty named cases with one value each, and the question is who sits where in the order. The colour is party control, joined on from outside; the index does not contain it.

Now the answer to the prediction. Equal weights move the average state `r pc(mean(abs(rw$equal_move)), 1)` places and the worst-affected state `r max(abs(rw$equal_move))`. Leaving the ten issue areas unstandardized moves one state `r max(abs(rw$unstandardized_move))` places.

```{r reweight-table}
```

The reason the unstandardized version behaves so badly is worth naming. Two issue areas are measured in **days**, the registration deadline and early voting, and their spreads, the standard deviations from above, run around `r pc(max(LOAD$sd), 0)` against about `r pc(min(LOAD$sd), 1)` for automatic registration. Skip the standardizing step and those two columns write the index nearly on their own, not because days matter more but because days are numbered higher.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same fifty states, ranked four ways.** Each line is a state, carried from its published rank across three alternative weighting rules. No law changed between the columns.

The published rule is not wrong. The point is that the rule is a choice, the choice is invisible in the number, and a state's rank is partly an artifact of it. Rhode Island ranks `r rw$published_rank[rw$state == "Rhode Island"]` of 50 under the published rule and `r rw$unstandardized_rank[rw$state == "Rhode Island"]` unstandardized. A reader given either number has been told something about Rhode Island's laws and something about a decision made in a spreadsheet, in unmarked proportions.

Variance weighting has one further consequence that is easy to state and hard to unsee. *A law that nearly every state has cannot matter much, no matter how burdensome it is.*

```{r near-universal}
```

**`r max(it$states_scored)` of 50 states forbid people in prison from registering to vote.** As a barrier that is absolute. It is the difference between voting and not voting for everyone it touches: the Sentencing Project's count puts roughly four million American adult citizens barred from voting over a felony conviction, about three in ten of them in prison or jail.^[Christopher Uggen, Ryan Larson, Sarah Shannon and Robert Stewart, *Locked Out 2024: Four Million Denied Voting Rights Due to a Felony Conviction* (Washington: The Sentencing Project, 2024), <https://www.sentencingproject.org/reports/locked-out-2024-four-million-denied-voting-rights-due-to-a-felony-conviction/>.] As a *variable* it is nearly worthless, because it barely separates one state from another, and it sits in the index carrying a standard deviation of `r pc(it$sd[it$states_scored == max(it$states_scored)], 2)`. That is what an index buys and what it hides: a single comparable number, at the price of everything the states agree about.

That leads directly to the assumption the whole enterprise rests on, which is that cost affects behaviour.

```{r turnout-cases}
```

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.8}
```

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Cost against turnout, 2024.** Each point is a state, and the dashed line is the best straight line through them. It slopes the way the theory says it should, with a correlation of `r fact("turnout_correlation")` on a scale where −1 would mean cost predicts turnout perfectly and 0 not at all, and explains `r pc(100 * as.numeric(fact("turnout_r_squared")), 1)`% of the variation between states.

Hawaii is the `r d$final_rank[d$state == "Hawaii"]`rd easiest state in which to vote and turns out `r pc(d$vep_turnout[d$state == "Hawaii"], 1)`% of its eligible adults. `r HARD` is the hardest of all `r nrow(d)` and turns out `r pc(d$vep_turnout[d$state == HARD], 1)`%, which is `r pc(d$vep_turnout[d$state == HARD] - d$vep_turnout[d$state == "Hawaii"], 1)` points *higher*. Wisconsin ranks `r d$final_rank[d$state == "Wisconsin"]`th on cost and turns out `r pc(d$vep_turnout[d$state == "Wisconsin"], 1)`%.

Three readings of that scatter are available. Cost may genuinely matter little at the margins the states actually occupy. The index may be measuring the wrong thing, scoring statutes rather than the experience of voting. Or turnout may be driven by something stronger and absent here, such as how competitive the state is. What the figure establishes is narrower and worth having: whatever drives American turnout, it is not mostly the ease of voting.

One pattern the index carries easily, and it is the largest in the data.

```{r control-trend}
```

In 1996 the average Republican-trifecta state ranked `r fact("r_rank_1996")` of 50 and the average Democratic-trifecta state ranked `r fact("d_rank_1996")`. Voting was easier, on average, in the states Republicans controlled. The lines cross in `r CROSS`, and by 2024 the gap has inverted and widened to `r pc(as.numeric(fact("gap_2024")), 1)` places the other way.

```{r fig4-static, eval=!knitr::is_html_output(), fig.height=4.8}
```

```{r fig4-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 4. Mean cost-of-voting rank by party control of state government, 1996–2024.** Ranks closer to 1 are more accessible. Divided governments sit between the two lines and move least.

That pattern survives every re-weighting tried above, which is what makes it worth trusting. Note what kind of claim it is. It is a trend in ranks rather than in laws, because every year is a separate cross-section ranked within itself.

## What this brief cannot tell you

**It measures statutes, not experience.** Every quantity in it is a property of a state's law, so two counties in the same state have the same score. A five-hour line in one and a five-minute line in the next are the same observation. Where election administration actually varies, county by county, this index is silent by design.

**Its weights come from variance, not from burden.** It would not notice if all fifty states adopted the same harsh new requirement tomorrow, because nothing would have changed *between* states. An index that cannot detect a nationwide restriction is measuring dispersion rather than cost.

**Its nine years are not one instrument.** The index was built from `r Y96$n_issue_areas` issue areas in 1996 and `r Y24$n_issue_areas` in 2024, and the components are re-estimated on each year's set. The back series has also been rewritten: the 1996 index as first published and as the current file gives it correlate `r pc(Y96$cor_init_final, 3)`, with the average state moving `r pc(Y96$mean_rank_move, 1)` places and `r Y96$max_move_state` moving `r Y96$max_rank_move`.

**Its yes-or-no items flatten laws that are not yes-or-no.** Which photo IDs count, and how long a voter has to cure a provisional ballot, differ by state and decide who actually votes.

**And it has no people in it.** No race, no age, no income. The claim that most motivates this literature is that voting costs fall hardest on people with the least slack in their lives, and this file cannot test it.

## What you should have learned

- American states impose genuinely different costs to cast the same federal vote: `r EASY_NOW` scores `r pc(min(d$final))` and `r HARD` scores `r pc(max(d$final))` on the 2024 index.
- An index is a weighting rule wearing a number's clothing. Weighting the four components equally moves the average state `r pc(mean(abs(rw$equal_move)), 1)` places and one state `r max(abs(rw$equal_move))`; leaving the issue areas unstandardized moves one state `r max(abs(rw$unstandardized_move))`.
- Variance weighting measures how unusual a state is, not how hard it is to vote there. The item `r max(it$states_scored)` of 50 states share carries a standard deviation of `r pc(it$sd[it$states_scored == max(it$states_scored)], 2)` and is nearly invisible in the score.
- Cost explains `r pc(100 * as.numeric(fact("turnout_r_squared")), 1)`% of the variation in 2024 turnout between states, and `r HARD`, the hardest state of all, turns out `r pc(d$vep_turnout[d$state == HARD] - d$vep_turnout[d$state == "Hawaii"], 1)` points above the `r d$final_rank[d$state == "Hawaii"]`rd easiest.
- The parties swapped ends. The average Republican-trifecta state ranked `r fact("r_rank_1996")` of 50 in 1996 against `r fact("d_rank_1996")` for Democratic trifectas, and `r fact("r_rank_2024")` against `r fact("d_rank_2024")` in 2024.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `reweight_2024.csv` re-ranks every state three other ways in `equal_rank`, `pc1_rank` and `unstandardized_rank`. Sort by `unstandardized_move` and read the ten biggest movers. Do they have anything in common?
- `covi_2024.csv` puts `vep_turnout` next to `final_rank` and `control`. Group the states by `control` and compare each group's average turnout against its average cost rank.
- `items_2024.csv` has `states_scored` and `sd` for every law item. Sort by `states_scored` and list the ten laws nearly every state has. Which would you call the heaviest burden?
- `covi_panel.csv` carries `initial` and `final` for every state in every year. Pick a state and follow both columns across nine elections. Does its score move when its party control changes, or before?
- **Stretch.** Check three states' current law on one issue area against a state legislature's own website. How many still match what the index says?
- **Stretch.** Find a county-level source for wait times or polling-place closures in one state. Is the variation inside that state larger than the variation between states this index measures?

## Sources

**Data**

- Michael J. Pomante II, *Cost of Voting Index*, <https://costofvotingindex.com>
  — the Data page, which publishes the COVI values and ranks for 1996 through
  2024, the master spreadsheet of state voting requirements, and the codebook
  defining every item. All three are kept as they were downloaded in August
  2026, because the announced 2026 edition will re-estimate the whole series.
- Turnout is the 2024 rate over the voting-eligible population, from the United
  States Elections Project's general election turnout series (Michael McDonald,
  University of Florida), <https://election.lab.ufl.edu/data-downloads/>. That
  file was still a preliminary release when it was taken.
- Party control of state government came from two places. For 1996 through 2008
  it is Carl Klarner's state partisan balance measures, in the Correlates of
  State Policy Project (Grossmann, Jordan and McCrain, Michigan State
  University), <https://ippsr.msu.edu/public-policy/correlates-state-policy>.
  The 2012 through 2024 elections were classified from each state's governor
  and chamber majorities and tested against Pomante (2025) Table 2; the Klarner
  years differ from that test on `r fact("control_states_differing")` states.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`control_check.csv`](data/derived/control_check.csv), [`control_panel.csv`](data/derived/control_panel.csv), [`control_trend.csv`](data/derived/control_trend.csv), [`covi_2024.csv`](data/derived/covi_2024.csv), [`covi_panel.csv`](data/derived/covi_panel.csv), [`items_2024.csv`](data/derived/items_2024.csv), [`rebuild_2024.csv`](data/derived/rebuild_2024.csv), [`reweight_2024.csv`](data/derived/reweight_2024.csv), [`year_structure.csv`](data/derived/year_structure.csv)

Two more files here are not data about the subject at all -- they are how this chapter checked its own work: [`facts.csv`](data/derived/facts.csv), the individual numbers the text above quotes, and [`checks.csv`](data/derived/checks.csv), the arithmetic verified before anything was drawn.

**Academic literature**

- Pomante, M. J. II (2025). "Cost of Voting in the American States: 2024."
  *Election Law Journal* 24(1): 52–61. doi:10.1089/elj.2024.0037; the index's
  own site links the article from <https://costofvotingindex.com/publications>.
- Li, Q., Pomante, M. J. II and Schraufnagel, S. (2018). "Cost of Voting in the
  American States." *Election Law Journal* 17(3): 234–247. The paper that
  built the 1996–2016 series. doi:10.1089/elj.2017.0478.
- Klarner, C. (2013). "State Partisan Balance Data, 1937–2011." Harvard
  Dataverse, hdl:1902.1/20403.
- Uggen, C., Larson, R., Shannon, S. and Stewart, R. (2024). *Locked Out 2024:
  Four Million Denied Voting Rights Due to a Felony Conviction.* The
  Sentencing Project — the four million figure quoted above.
  <https://www.sentencingproject.org/reports/locked-out-2024-four-million-denied-voting-rights-due-to-a-felony-conviction/>

```{r ai-prompt, echo=FALSE, results="asis"}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
