---
title: "Data Session 6 Key: The File With Your Name On It"
subtitle: "84-355 Democracy's Data | Session 6 — Voter registration files | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
s  <- read.csv("data/schema.csv", stringsAsFactors = FALSE)
rp <- read.csv("data/race_party.csv", stringsAsFactors = FALSE, check.names = FALSE)
st <- read.csv("data/status.csv", stringsAsFactors = FALSE)
to <- read.csv("data/turnout.csv", stringsAsFactors = FALSE, check.names = FALSE)
lost <- read.csv("data/lost.csv", stringsAsFactors = FALSE)
```

All code is given to students. Grade the interpretation.

**Where this sits.** Data Session 6, and it exists because of an ordering
mistake in the old schedule: BISG was being taught before students had ever seen
a voter file. **This session introduces the source; Session 7 infers from it;
Session 14 tries to recover from aggregates what this file states outright.**
Same county throughout, deliberately — three sessions, one file, escalating
distance from the truth.

**Say the arc out loud at the start and again at the end.** It is the spine of
the back half of the course.

Every number below was run against the data files.

***

# The session in one sentence

The most consequential dataset in American campaigning is a routing table for
poll workers, it has no party column worth the name, and it forgets people —
13,803 of them since 2020.

***

# The findings

```{r headline}
c(voters        = 127560,
  columns       = nrow(s),
  ballot_cols   = sum(s$purpose == "Which ballot you get"),
  inactive      = sum(st$voters[st$status == "INACTIVE"]),
  lost_2020     = lost$voters_no_longer_in_file[lost$election == "2020 general"])
```

- **21 of 53 columns** exist to decide which ballot you get.
- **70% of the file** has no usable party signal — Georgia has no party
  registration.
- **9,386 inactive** registrants, removed for postal reasons, not for anything
  they did.
- **13,803 people voted in 2020 and are not in the 2026 file.**
- 2024 turnout: **White 69.2, Black 55.1, Asian 55.2, Hispanic 47.7** — and
  every one of those is computed on the wrong denominator, on purpose.

***

# The three things to get said

**1. The file is a routing table, not a description of the electorate.** This is
the framing that makes everything else make sense. Twenty-one columns place you
in overlapping districts — water board, fire district, "Combo" — because a clerk
must hand you one specific ballot. Researchers are *scavenging* an
administrative artifact. Ask the room what a file built *for* research would
look like instead, and how you would ever get one.

**2. "Registered Democrat" is not a thing in Georgia.** Students arrive certain
that party registration is universal; roughly half the country has none. The
`Last Party Voted` column looks like party ID and is not — it is the last
primary ballot requested, blank for 40,663 people. **This is the single most
useful factual correction in the session**, because the misreading is everywhere
in campaign journalism.

**3. The file has no memory, and the forgetting is not random.** Part 4 is the
finding to spend time on. The 13,803 missing 2020 voters are disproportionately
people who move — renters, younger voters, lower-income voters — so any
retrospective analysis is biased in a predictable direction. **Ask what a
researcher who did not check would have concluded.**

**The line for the board:** the data is not wrong, it is answering a different
question than the one you asked.

***

# Part 5 — the denominator, which is the teachable error

```{r p5}
to$rate <- round(100 * to$`2024 general` / to$in_file_2026, 1)
to[to$in_file_2026 > 1000, c("race", "in_file_2026", "2024 general", "rate")]
```

**Do not let students write these down as turnout rates.** They are a 2024
numerator over a 2026 denominator. Three separate problems, and students should
be able to name all three:

1. People who voted in 2024 and have since left are in **neither** number.
2. People who registered in 2025–26 are in the denominator and **could not have
   voted**.
3. "Registered" is itself ambiguous — 127,560 or 118,174, a 7.4% swing, decided
   by the inactive flag.

The correct denominator is the file as it stood in November 2024. **Nobody kept
it.** Most states overwrite. That is the honest ending: the right analysis is
not merely hard here, it is impossible after the fact, which is why serious work
uses archived snapshots and why those are expensive.

***

# Option A — Active voters only

## Verified output

```{r a}
act <- 118174 / 127560
data.frame(race = to$race, rate_all = to$rate,
           rate_if_scaled = round(to$rate / act, 1))[to$in_file_2026 > 1000, ]
```

**3/4:** Recomputes and notes every rate rises by about 8%.

**4/4:** Notices the flaw in the scaling — inactive voters are **not** spread
evenly across groups, so applying one countywide ratio is itself an assumption.
The file has the group-level counts; a student who says "I would want inactive
broken down by race before believing any of this" has the right instinct. (We
did not build that table. **Good — let them find the limit.**)

**4/4, best:** Connects it to Part 3. Being marked inactive is triggered by
returned mail and change-of-address records, which track residential mobility,
which tracks age, tenure and income. So the inactive flag is *correlated with
the thing being measured*. Dropping inactives does not clean the data, it tilts
it.

***

# Option B — The midterm drop

## Verified output

```{r b}
to$drop <- round(100 * (1 - to$`2022 general` / to$`2024 general`), 1)
to[to$in_file_2026 > 1000, c("race", "2022 general", "2024 general", "drop")]
```

**3/4:** Reports that turnout fell for every group between the presidential and
midterm election, and identifies which fell furthest.

**4/4:** Refuses to call it a fact about voters. The 2022 column is matched
against a 2026 file that has already lost **5,677** of that election's voters,
and the 2024 column has lost only 3,798 — so **the two columns are not measured
on comparable populations.** Part of any apparent difference is decay, not
behaviour.

**4/4, best:** Proposes the fix — you would need the file as it stood in each
November, and then notes that this makes the comparison a data-availability
problem rather than a political one. That is the correct and slightly deflating
answer.

***

# Option C — The categories

## Verified output

```{r c}
o <- to[to$race %in% c("OTHER", "UNKNOWN"), ]
o$rate <- round(100 * o$`2024 general` / o$in_file_2026, 1)
o[, c("race", "in_file_2026", "2024 general", "rate")]
```

**9,923 people** are OTHER or UNKNOWN — more than the Hispanic and Asian counts
combined — and UNKNOWN records **29.7%** turnout, the lowest of any group.

**3/4:** Offers a voter-side explanation (these voters are less engaged) and a
form-side one (the category is a residue of the form).

**4/4:** Argues the form side is more likely and says why. UNKNOWN largely means
*the race field was never filled in*, which happens with particular registration
channels — motor-voter, third-party drives, older records — and those channels
correlate with everything. A student who says "this is not a group of people, it
is a group of **records**" has it.

**4/4, best:** Notices this is Session 2's finding again from the other side.
There, 94.5% of "Some Other Race" census respondents were Hispanic — a real
population failing a category. Here it is mostly a missing-data artefact wearing
the same label. **Same word, two completely different phenomena**, and only
knowing how the data was collected tells them apart.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is naming the file's purpose and reasoning from
  it. Any student who writes "this file was built to hand people ballots, so…"
  and completes the sentence correctly has understood the session.
- **A 2** is reporting the turnout gap as a finding about voters without
  touching the denominator problem, after Part 5 has spelled it out.
- **Reward discomfort about the privacy question.** Their names are in a file
  like this, it is public, and it is sold. There is no clean answer and they
  should not be given one.

## Anticipated questions

- *"Is my own registration public?"* — Yes, in every state, though what is
  released varies. Pennsylvania sells a statewide file. Do not dodge this; it
  makes the session land.
- *"Why does Georgia record race?"* — Section 5 of the Voting Rights Act
  required covered jurisdictions to demonstrate that changes did not have a
  discriminatory effect, which requires knowing who voters are. The coverage
  formula was struck down in *Shelby County* (2013); the practice remained.
- *"Can campaigns see how I voted?"* — **No.** They can see *that* you voted, and
  which primary ballot you requested. The difference matters enormously and is
  the whole premise of Session 14.
- *"Is this legal to use in class?"* — It is a public record. Note that we
  committed no individual rows, and say why.
- *"Why not just use the whole state?"* — Size, and because Sessions 7 and 14
  reuse this county. Repetition across three sessions is doing pedagogical work.

## Connections

- **Session 2 (census)** — where the block demographics come from, and the same
  one-box race problem.
- **Session 5 (money and message)** — the targeting operations were aimed at
  these rows.
- **Session 7 (surnames and BISG)** — *next week*, and it needs this file. Most
  states do not record race, so people infer it. **Say the handoff explicitly.**
- **Session 8 (surveys)** — the CPS *asks* people whether they voted and gets a
  different answer. Administrative record vs self-report, same quantity.
- **Session 14 (RPV)** — will try to recover the party-by-race numbers from
  Part 2 using only precinct aggregates, and will get them wrong by nine points.
  **Show Part 2's table again that day.**
