---
title: "Data Session 6: The File With Your Name On It"
subtitle: "84-355 Democracy's Data | Voter registration files"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number came from?

> *"Turnout among registered Black voters was up six points."*

You have read a sentence like that. Today you meet the file it came from.

Every state maintains a list of everyone registered to vote. **It is a public
record**, it has your name on it if you are registered, and it is the single
most consequential dataset in American campaigning — every targeting operation
in Session 5 was aimed at rows in a file like this one.

**But it was not built for research.** It was built so that a poll worker can
find you in a book and hand you the right ballot. Almost everything odd about it
follows from that.

We are using **Houston County, Georgia**: 127,560 registered voters. Georgia
because it is one of a handful of states that records **race** on the
registration form — collected for Voting Rights Act compliance, not for anybody's
convenience. That makes this file unusually revealing, and it is why Sessions 7
and 14 both come back to it.

# Part 1 — What one row is

```{r schema}
s <- read.csv("data/schema.csv", stringsAsFactors = FALSE)
nrow(s)
table(s$purpose)
```

Fifty-three columns per person. Look at what they are for:

```{r schema-ballot}
s$column[s$purpose == "Which ballot you get"]
```

**Twenty-one of the fifty-three columns exist to place you in a district.**
Congressional, state senate, state house, judicial, county commission, school
board, city council, municipal school board, water board, fire — plus two kinds
of precinct, and a "Combo" field that bundles the whole stack into one code.

That is the file telling you what it is for. It is not a description of the
electorate. **It is a routing table.** Everything a researcher wants from it —
who registered when, who turned out, who lives where — is a by-product of the
county's need to hand you the correct piece of paper.

```{r schema-rest}
s[s$purpose %in% c("Who you are", "Demographics", "When things happened"),
  c("column", "purpose")]
```

Note what is here: your **birth year**, your **race**, your **gender**, the date
you registered, and the date you last voted. Note what is *not* here: how you
voted, ever, in anything.

# Part 2 — The party column that isn't

```{r party}
rp <- read.csv("data/race_party.csv", stringsAsFactors = FALSE, check.names = FALSE)
rp
```

The column is called **"Last Party Voted"** and it is the one people misread most.

**Georgia has no party registration.** You cannot register as a Democrat there.
What this column records is the last party's *primary ballot* you happened to
request — which is why **40,663 voters have nothing in it at all**, and another
48,603 are marked non-partisan.

```{r party-share}
tot   <- rowSums(rp[, -1])
blank <- rp[, "no primary ballot on record"] + rp[, "NON-PARTISAN"]
data.frame(race = rp$race, voters = tot,
           pct_no_party_signal = round(100 * blank / tot, 1))

cat(sprintf("\nwhole file: %.1f%% carries no party signal\n", 100 * sum(blank) / sum(tot)))
```

**So seventy percent of this file carries no usable party signal**, and it is
worst for the groups with fewest voters.
Anybody who tells you how many registered Democrats there are in Georgia is
either using a different state's data or making it up.

Now look at the voters who *do* have a signal:

```{r party-ratio}
d <- rp$DEMOCRAT; r <- rp$REPUBLICAN
data.frame(race = rp$race, dem = d, rep = r,
           ratio = ifelse(r > 0, round(d / r, 1), NA))
```

Among voters who ever took a partisan primary ballot, Black voters here broke
Democratic over Republican by more than **thirty to one**; white voters broke
Republican by nearly five to one. Hold on to that — **Session 14 will try to
recover exactly these numbers without being allowed to see them.**

# Part 3 — Who counts as registered

```{r status}
st <- read.csv("data/status.csv", stringsAsFactors = FALSE)
st
```

**9,386 people on this list are marked INACTIVE**, and the file records why.
"NCOA" is the Postal Service's National Change of Address database. "CROSS
STATE" means another state reported registering them. "RETURNED MAIL" means a
letter came back.

None of those people did anything. They are being removed from a list because a
piece of mail failed, and the reasons are a record of an administrative process
rather than of anybody's intentions.

**This is the column that decides whether "registered voters" means 127,560 or
118,174** — a difference of 7.4% — and every turnout rate you will ever read
depends on which one somebody used, usually without saying.

# Part 4 — The file has no memory

Here is the thing that makes voter files genuinely dangerous to use casually.

```{r lost}
lost <- read.csv("data/lost.csv", stringsAsFactors = FALSE)
lost
```

```{r lost-plot}
barplot(lost$voters_no_longer_in_file, names.arg = lost$election,
        col = "#C41230", ylab = "voters who voted, and are not in the 2026 file")
```

We took the current file and matched it against the state's record of who
actually voted. **13,803 people voted in the 2020 general election and are not in
the 2026 file at all.** For 2022 it is 5,677; for 2024, 3,798.

They died, moved away, or were purged. **A voter file is a snapshot of a live
system, not an archive.** It tells you about *now*, and the further back you
reach with it, the more people have quietly vanished from your denominator —
and they are not a random sample of anybody.

# Part 5 — Turnout by race, and why it is not quite right

```{r turnout}
to <- read.csv("data/turnout.csv", stringsAsFactors = FALSE, check.names = FALSE)
to$rate_2024 <- round(100 * to$`2024 general` / to$in_file_2026, 1)
to[, c("race", "in_file_2026", "2024 general", "rate_2024")]
```

```{r turnout-plot}
k <- to$race %in% c("WHITE","BLACK","HISPANIC/LATINO","ASIAN/PACIFIC ISLANDER")
barplot(to$rate_2024[k], names.arg = c("White","Black","Hispanic","Asian/PI"),
        col = "#2c7fb8", ylim = c(0, 80), ylab = "2024 turnout (%)")
```

White **69.2%**, Black **55.1%**, Asian **55.2%**, Hispanic **47.7%** — a
twenty-one-point gap between the top and the bottom.

**Every one of those numbers is slightly wrong, and you now know why.** The
numerator is 2024 votes; the denominator is who is on the list in 2026. Anyone
who voted in 2024 and has since moved away is in neither. Anyone who registered
in 2025 is in the denominator and could not possibly have voted.

The right denominator is *the file as it stood in November 2024*, which nobody
kept. This is the ordinary situation. **Published turnout-by-race figures from
voter files almost always have this problem**, and almost none of them say so.

Compare it against Session 8, where the Census Bureau will *ask* people whether
they voted and get a different answer again.

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Use active voters only.** Recompute Part 5's rates against 118,174
active registrants instead of all 127,560 (`status.csv` has the counts). Which
group's turnout rate moves most, and what does that tell you about who ends up
marked inactive?

**Option B — The midterm drop.** Compare 2022 to 2024 by race. Turnout falls for
everybody, but not equally. Which group falls furthest, and what would you need
to know before calling that a fact about those voters rather than about the
file?

**Option C — The categories.** 7,748 people are "OTHER" and 2,175 are "UNKNOWN" —
nearly 10,000 who fit no box, more than the Hispanic and Asian counts combined.
Their recorded turnout is the lowest of any group. Give two explanations, one
about voters and one about the form, and say how you would tell them apart.

**Option D — Your own question.** 53 columns, 127,560 people, three elections.

# What this lab cannot tell you

- **How anybody voted.** The ballot is secret. This file knows *that* you voted
  and *which primary ballot you asked for*; it will never know who you chose.
  Session 14 is entirely about what people do with that gap.
- **One county in one state.** Most states record no race, and many record party
  registration that Georgia does not. The *shape* of the file generalises; the
  columns do not.
- **Race here is self-reported on a form**, one box, at the moment of
  registration, possibly decades ago — with all the flattening Session 2 showed
  in the census.
- **We used the public file.** Commercial vendors — L2, Catalist, TargetSmart —
  sell versions with hundreds of appended columns: modeled party, modeled
  turnout, consumer data, estimated income. Campaigns use those. The modeling
  in Session 7 is the cheapest of them.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough, so long as it shows you
engaged with what we did.

Three that would make good posts:

- Twenty-one of fifty-three columns exist to route you to a ballot. What does
  that tell you about what this file is, and about what researchers are doing
  when they use it?
- 13,803 people voted in 2020 and are gone from the current file. Does that
  change how much you trust a turnout statistic?
- Your name is in a file like this, and it is public. Should it be?

***

### Where the data came from

Houston County, Georgia registration extract 71754, and the state's voter-history
files for the 2020, 2022 and 2024 general elections. Both are public records in
Georgia. Working data for the 2026 Houston County redistricting matter.

**Nothing individual-level is in this folder.** 127,560 records were collapsed to
a schema listing and four small tables before anything was saved — the column
names and the margins are all this lab needs. The build is in
`data/build-data.R`.
