---
title: "Lab 1b: Where Election Data Comes From"
subtitle: "84-355 Democracy's Data | Week 1 — Election returns"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

On Tuesday you drew a map of the 2024 election from a file that was simply handed
to you, and we named the **four families of evidence** this course runs on:

| Family | Made by | Made how | Weeks |
|---|---|---|---|
| **Election returns** | nobody in particular | counted, canvassed, certified — separately, by every county and state | **this week** |
| **The census** | one agency, on a legal schedule | a constitutionally required count, plus a rolling survey | 2, 12–15 |
| **Surveys** | pollsters and agencies | asking a sample and weighting the answers | 3, 10 |
| **Administrative records** | a bureaucracy, as a by-product | filings and logs kept for some other purpose | 5, 7 |

Today is the first family, and it is the strangest of the four. The other three
have a producer you can name. Returns do not.

Here is the chain that produces an official presidential result. A voter marks a
ballot in a precinct. The precinct's count goes to a county clerk, who canvasses
it. The county sends a certified return to the secretary of state, who canvasses
again and certifies a statewide total, usually a few weeks after election day.
Nobody in that chain publishes a tidy national file. **There is no national
election administrator in the United States, and therefore no official national
returns file.** Every dataset you have ever seen of American election results was
assembled by somebody — a news organization, an academic project, a hobbyist —
who went out and collected fifty-one separate things and stapled them together.

Today you will do the stapling yourself, and watch it fail in four different
ways. Keep asking, as each failure appears, whether it is a *returns* problem or
a problem any data would have. Some of what breaks today breaks because nobody
is in charge — and that is specific to this family.

The same rules as last week: grey blocks are code chunks, run them with the green
▶ arrow or **Cmd/Ctrl+Shift+Enter**, everything is base R, and **work in order**
because later chunks depend on earlier ones.

***

# Part 1: Four files, and a trap in the first one

The `data/` folder has four files.

| File | Rows | What it is |
|---|---|---|
| `pres2024_counties.csv` | 3,160 | 2024 presidential votes by county |
| `pres2020_counties.csv` | 3,152 | the same, for 2020 |
| `census_counties.csv` | 3,144 | the Census Bureau's official list of counties |
| `pres2024_states.csv` | 51 | last week's state file, unchanged |

Start with the county returns:

```{r read-naive}
cty <- read.csv("data/pres2024_counties.csv", stringsAsFactors = FALSE)
head(cty)
```

Everything looks fine. It is not.

Look at `county_fips`. A **FIPS code** is the federal government's unique
identifier for a place: two digits for the state, three for the county. Autauga
County, Alabama is `01001` — state `01`, county `001`. That is a *label*, like a
phone number. It is not a quantity.

But `read.csv()` looks at a column of digits and decides it is a number:

```{r fips-class}
class(cty$county_fips)
head(cty$county_fips)
```

The leading zero is gone. `01001` has become the number `1001`. How much damage?

```{r fips-damage}
table(nchar(cty$county_fips))
```

```{r fips-damage-states}
short <- cty[nchar(cty$county_fips) == 4, ]
sort(unique(short$state_name))
```

> **Pause and read that output.** 328 of 3,160 rows are broken, and they are not
> random — they are exactly the seven states whose FIPS code begins with a zero.
> Alabama, Alaska, Arizona, Arkansas, California, Colorado, Connecticut. If you
> later try to match these against any other government file, none of them will
> match, and the failure will be silent.

This is the single most common bug in real election-data work. Fix it by telling
`read.csv()` that the column is text, not a number:

```{r read-fixed}
cty <- read.csv("data/pres2024_counties.csv",
                stringsAsFactors = FALSE,
                colClasses = c(county_fips = "character"))

table(nchar(cty$county_fips))
head(cty$county_fips)
```

`colClasses` takes a named vector: the name is the column, the value is the type
you want. Now every code is five characters, as it should be.

Read the other three files the same careful way:

```{r read-rest}
old <- read.csv("data/pres2020_counties.csv",
                stringsAsFactors = FALSE,
                colClasses = c(county_fips = "character"))

cen <- read.csv("data/census_counties.csv",
                stringsAsFactors = FALSE,
                colClasses = c(fips = "character"))

st  <- read.csv("data/pres2024_states.csv", stringsAsFactors = FALSE)

dim(cty); dim(old); dim(cen); dim(st)
```

One more thing to notice before we go on. Compare the two 2024 files:

```{r counts-vs-pct}
head(cty[, c("state_name", "county_name", "votes_dem", "votes_gop")], 3)
head(st[, c("state", "harris", "trump")], 3)
```

The county file is **counts of ballots**. The state file is **percentages**. You
cannot add a percentage to a count, and you cannot compare them until you convert
one to the other. That sounds obvious written down; it is a genuine source of
published errors.

***

# Part 2: Adding the counties up

If the county file is right and the state file is right, then adding up a state's
counties should reproduce that state's result. Let's find out.

`aggregate()` is the base R tool for "add up this column, grouped by that
column." The formula reads *left side, broken out by right side*:

```{r aggregate}
agg <- aggregate(cbind(votes_dem, votes_gop, total_votes) ~ state_name,
                 data = cty, FUN = sum)

nrow(agg)
head(agg)
```

51 rows — 50 states and DC, which is what we want. Now convert counts to
percentages so the two files speak the same language:

```{r shares}
agg$harris_c <- round(100 * agg$votes_dem / agg$total_votes, 2)
agg$trump_c  <- round(100 * agg$votes_gop / agg$total_votes, 2)
```

The `_c` suffix is a reminder: these came from the *counties*.

Now join the two tables. `merge()` matches rows between two data frames using a
shared column. Ours are named differently, so we say which is which:

```{r merge-states}
cmp <- merge(agg, st, by.x = "state_name", by.y = "state")
nrow(cmp)
```

51 rows out, 51 rows in. The join worked. Now the moment of truth:

```{r diffs}
cmp$dh <- round(cmp$harris_c - cmp$harris, 2)
cmp$dt <- round(cmp$trump_c  - cmp$trump,  2)

sum(cmp$dh == 0 & cmp$dt == 0)
```

Eighteen states agree exactly. Thirty-three do not. Here are the worst:

```{r worst}
worst <- cmp[order(-abs(cmp$dh)), ]
head(worst[, c("state_name", "harris", "harris_c", "dh",
               "trump", "trump_c", "dt")], 8)
```

A picture of all 51:

```{r discrepancy-plot, fig.width = 7.5, fig.height = 6}
top <- cmp[order(-abs(cmp$dh)), ][1:15, ]
top <- top[order(abs(top$dh)), ]

par(mar = c(5, 10, 4, 2))
barplot(top$dh,
        names.arg = top$state_name,
        horiz = TRUE, las = 1, border = NA,
        col = ifelse(abs(top$dh) > 1, "#B2182B", "#999999"),
        xlim = c(0, 2),
        xlab = "County sum minus state file (Harris, percentage points)",
        main = "Where the two files disagree")
```

> **Pause and read that plot.** Two things are strange. First, Alaska and New
> York are in a class of their own. Second — and this is the real clue — look at
> the direction. Check it:

```{r direction}
c(higher = sum(cmp$dh > 0), equal = sum(cmp$dh == 0), lower = sum(cmp$dh < 0))
```

**Every single disagreement runs the same way.** The county sums never give
Harris a *lower* share than the state file; they give her a higher one, or the
same one. Random errors do not behave like that. A one-sided error means
something systematic, and systematic errors have findable causes.

Before we chase it: does any of this change who won?

```{r winner-check}
cmp$winner_c <- ifelse(cmp$harris_c > cmp$trump_c, "Harris", "Trump")
sum(cmp$winner_c != cmp$winner)
```

No. Zero states flip. That is worth saying plainly, because it is the honest
summary: **the disagreement is real, it is systematic, and it does not change the
outcome.** All three parts of that sentence matter.

***

# Part 3: The vote that isn't there

If Harris's share is too *high* in the county sums, either her count is too big
or the denominator is too small. Let's look at the denominator.

Every county's `total_votes` should be Harris plus Trump plus everybody else.
So everybody else is:

```{r other}
cty$other <- cty$total_votes - cty$votes_gop - cty$votes_dem
summary(cty$other)
```

No negatives, so the arithmetic is at least internally consistent. But look at
the minimum. How many counties report *zero* third-party votes?

```{r other-zero}
sum(cty$other == 0)
```

In 2024 more than two million people voted for someone other than Harris or
Trump. In how many counties did literally nobody? Almost certainly none. So a
zero here is not a measurement. Where are they?

```{r other-zero-states}
zeros <- tapply(cty$other, cty$state_name, function(x) sum(x == 0))
zeros[zeros > 0]
```

```{r other-zero-frac}
units <- table(cty$state_name)
names(zeros)[zeros == units[names(zeros)]]
```

> **There it is.** In Alaska and New York, *every* reporting unit shows zero
> third-party vote. Those are exactly the two states at the top of our
> disagreement plot. The data does not say "no one voted third party in New
> York." It says the person compiling this file never got New York's third-party
> numbers.

**A zero and a missing value are not the same thing, and spreadsheets cannot tell
them apart.** This is the most important sentence in the lab.

Now watch the same disease in the *other* file. The state file has an `other`
column too:

```{r state-other}
sum(st$other == 0)
```

42 of 51 states report zero third-party vote. But those three percentages should
add to 100. Do they?

```{r state-sums}
st$total_pct <- st$harris + st$trump + st$other
range(st$total_pct)
sum(abs(st$total_pct - 100) <= 0.01)
```

Not one state sums to 100. Every state is short by one to four points — and that
shortfall *is* the third-party vote, sitting there unlabelled. We can recover it:

```{r implied-other}
st$other_implied <- round(100 - st$harris - st$trump, 2)
head(st[order(-st$other_implied), c("state", "other", "other_implied")], 5)
```

So now we can estimate how many ballots the county file never saw. If the county
file's Harris *count* is right, and the state file's Harris *share* is right, then
the true statewide total must be the count divided by the share:

```{r missing}
cmp$implied_total <- round(cmp$votes_dem / (cmp$harris / 100))
cmp$missing       <- cmp$implied_total - cmp$total_votes
cmp$missing_pct   <- round(100 * cmp$missing / cmp$implied_total, 2)

head(cmp[order(-cmp$missing_pct), c("state_name", "total_votes",
                                    "implied_total", "missing", "missing_pct")], 6)
```

Alaska is missing about 13,000 ballots — 3.9% of everything cast there — and New
York roughly 182,000, or 2.2%. Those are not rounding errors. They are entire
candidates.

We will test this diagnosis properly in the exercises.

***

# Part 4: Joining on names is a trap

Everything so far compared *states*, where there are only 51 things and the names
are unambiguous. Counties are where joins go to die.

Bring in the Census Bureau's official county list. **This is the first time two
families meet in this course**: returns assembled by a news organisation, checked
against geography defined by a federal agency. Watch which one you end up
trusting, and notice that you did not decide it on evidence — you decided it on
who produced the file.

```{r census-head}
head(cen)
nrow(cen)
```

3,144 county-equivalents. Our returns file has 3,160 rows. Close, and we will
come back to the gap.

Suppose we want to attach these official records to our returns. The obvious move
is to match on the county name. Try it:

```{r name-join}
by_name <- merge(cty, cen, by.x = "county_name", by.y = "name")
nrow(by_name)
```

> **Pause.** We put in 3,160 rows of returns and got out more than 14,000. Where
> did they come from?

`merge()` matches *every* row on the left against *every* row on the right that
shares the key. If a name appears in several states, you get every combination:

```{r washington}
sort(table(cty$county_name), decreasing = TRUE)[1:6]
```

```{r washington-join}
c(returns = sum(cty$county_name == "Washington County"),
  census  = sum(cen$name == "Washington County"),
  joined  = sum(by_name$county_name == "Washington County"))
```

Thirty Washington Counties on each side, matched every-which-way: $30 \times 30 =
900$ rows, of which 30 are right and 870 are garbage. County names are not unique
and were never meant to be.

The usual first fix is to match on state *and* name together. `paste()` glues
strings into one key:

```{r state-name-join}
cty$key <- paste(cty$state_name, cty$county_name)
cen$key <- paste(cen$state_name, cen$name)

nrow(merge(cty, cen, by = "key"))
```

Much better. But we started with 3,160 rows and only 3,112 came through:

```{r state-name-miss}
missed <- cty[!cty$key %in% cen$key, ]
table(missed$state_name)
```

Alaska and the District of Columbia, entirely. Hold that thought.

***

# Part 5: The join that succeeds and is still wrong

Names are bad keys. Codes are what codes are for. Join on FIPS instead:

```{r fips-join}
by_fips <- merge(cty, cen, by.x = "county_fips", by.y = "fips")
nrow(by_fips)
```

3,116 matched, up from 3,112. Which returns still found no home?

```{r fips-miss}
table(cty$state_name[!cty$county_fips %in% cen$fips])
```

And which official counties got no returns?

```{r fips-miss-other}
table(cen$state_name[!cen$fips %in% cty$county_fips])
```

Now here is the part that should make you uneasy. We joined on a code, so
whatever matched should be the same place. Check it — compare the name that came
from the returns against the name that came from the Census:

```{r wrong-match}
wrong <- by_fips[by_fips$county_name != by_fips$name, ]
wrong[, c("county_fips", "state_name.x", "county_name", "name", "total_votes")]
```

> **Read that table slowly.** Four rows joined *successfully* to the wrong place.
> Alaska's State House District 13 is now labelled Aleutians East Borough.
> Anchorage Municipality — a city of nearly 300,000 — has been handed 6,965 votes
> from a state legislative district. And the District of Columbia, all 700,000
> residents of it, is now Ward 1.
>
> Nothing errored. Nothing warned. If you had joined and moved on, you would have
> published it.

## Three places that don't have counties

The four wrong matches and the 44 orphans all come from the same place: three
jurisdictions that do not report presidential results by county at all.

**Alaska** has 30 boroughs and census areas, but does not tabulate votes in them.
It reports by **state house district** — 40 of them. There is no federal code for
"Alaska House District 7," so whoever built this file invented one by tacking the
district number onto Alaska's state code `02`. That is a reasonable improvisation
right up to the moment it collides with real codes `02013`, `02016` and `02020`,
which belong to actual boroughs.

```{r alaska}
head(cty[cty$state_name == "Alaska", c("county_fips", "county_name", "total_votes")], 4)
head(cen[cen$state_abbrev == "AK", c("fips", "name")], 4)
```

**The District of Columbia** is one county-equivalent, `11001`. This file splits
it into eight **wards** and numbers them `11001` through `11008` — so Ward 1
inherits the code that means "all of DC."

```{r dc}
cty[cty$state_name == "District of Columbia",
    c("county_fips", "county_name", "total_votes")]
cen[cen$state_abbrev == "DC", c("fips", "name")]
```

**Hawaii** is the small mystery in the other direction — one official county with
no returns:

```{r hawaii}
cen[cen$state_abbrev == "HI", c("fips", "name")]
cty[cty$state_name == "Hawaii", c("county_fips", "county_name", "total_votes")]
```

Kalawao County has about eighty residents. It is the former leprosy settlement on
Molokai, it is administered by the state health department, and its ballots are
counted with Maui's. It is a real county that never appears in election returns.

***

# Part 6: The same source changed its mind

One more failure, and it is the sneakiest. Both county files came from the *same
repository*, built by the *same person*, four years apart. Surely those can be
compared.

```{r compare-years}
c(rows_2020 = nrow(old), rows_2024 = nrow(cty))
```

```{r compare-fips}
sum(!cty$county_fips %in% old$county_fips)   # in 2024, not in 2020
sum(!old$county_fips %in% cty$county_fips)   # in 2020, not in 2024
```

```{r compare-where}
table(cty$state_name[!cty$county_fips %in% old$county_fips])
table(old$state_name[!old$county_fips %in% cty$county_fips])
```

Three states, three different stories.

**Connecticut abolished its counties.** Connecticut's eight counties stopped
functioning as governments in 1960 and survived only as statistical lines on a
map. The state eventually asked the Census Bureau to replace them with the nine
**planning regions** its councils of government actually use, and the Bureau
agreed in a Federal Register notice dated June 6, 2022. New boundaries, new
names, new codes:

```{r ct}
old[old$state_name == "Connecticut", c("county_fips", "county_name")]
cty[cty$state_name == "Connecticut", c("county_fips", "county_name")]
```

```{r ct-overlap}
length(intersect(old$county_fips[old$state_name == "Connecticut"],
                 cty$county_fips[cty$state_name == "Connecticut"]))
```

Zero codes in common. And the regions are not the counties regrouped — the lines
themselves moved, so there is no lookup table that turns one into the other
without going back to town-level data.

**Alaska got renumbered by the compiler.** In 2020 this source coded Alaska's
districts `02901`–`02940`, deliberately staying clear of real borough codes. In
2024 it switched to `02001`–`02040`, which does not:

```{r ak-recode}
head(old$county_fips[old$state_name == "Alaska"], 3)
head(cty$county_fips[cty$state_name == "Alaska"], 3)
```

That was not a change in the world. That was a change in one person's convention,
and nothing in the file announces it.

**DC changed shape.** In 2020 this source had one DC row. In 2024 it has eight.

```{r dc-2020}
old[old$state_name == "District of Columbia", c("county_fips", "county_name")]
```

Now put it together. Suppose you want the swing from 2020 to 2024 in every
county. You join on FIPS:

```{r swing}
sw <- merge(old[, c("county_fips", "state_name", "county_name",
                    "votes_dem", "votes_gop")],
            cty[, c("county_fips", "votes_dem", "votes_gop")],
            by = "county_fips", suffixes = c("_20", "_24"))
nrow(sw)
```

```{r swing-calc}
sw$r20 <- 100 * sw$votes_gop_20 / (sw$votes_gop_20 + sw$votes_dem_20)
sw$r24 <- 100 * sw$votes_gop_24 / (sw$votes_gop_24 + sw$votes_dem_24)
sw$swing <- round(sw$r24 - sw$r20, 2)

head(sw[order(-sw$swing), c("state_name", "county_name", "r20", "r24", "swing")], 6)
```

That is a real and famous result — the biggest Republican shifts of 2024 were in
heavily Hispanic border counties in south Texas and in Imperial County,
California, with the New York City boroughs close behind.

And then there is this row:

```{r swing-dc}
sw[sw$county_fips == "11001",
   c("county_fips", "county_name", "r20", "r24", "swing")]
```

> **Look at what that says.** All of the District of Columbia in 2020, compared
> against Ward 1 in 2024, producing a swing of $-0.25$ points. The number is not
> absurd. It is not flagged. It sits in a table of 3,104 rows and looks exactly
> like its neighbours. It is also meaningless.
>
> **This is the lesson of the lab.** A join that produces plausible output is not
> a join that is correct. The only way to know is to check the join itself — count
> your rows, look at what dropped, and look at what matched.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option below is a working chunk with one thing to change, or nothing at all to
change. Run it, look at what comes out, and tell me what it means.

Pick **one**.

***

### Option A — Test the diagnosis

Part 3 blamed the disagreement on the county file's *denominator*: it misses
third-party ballots, so its Harris share comes out too high. That is a claim
about the world, and it makes a prediction we can check. If the counts are fine
and only the denominator is wrong, then throwing third parties out of *both*
files should make the disagreement disappear.

```{r opt-a}
# All-party share (what Part 3 compared)
cmp$dh <- round(100 * cmp$votes_dem / cmp$total_votes - cmp$harris, 2)

# Two-party share — third parties removed from both sides
cmp$h2_cty <- 100 * cmp$votes_dem / (cmp$votes_dem + cmp$votes_gop)
cmp$h2_st  <- 100 * cmp$harris    / (cmp$harris    + cmp$trump)
cmp$d2     <- round(cmp$h2_cty - cmp$h2_st, 2)

c(median_gap_all_party = median(abs(cmp$dh)),
  median_gap_two_party = median(abs(cmp$d2)))

c(states_agreeing_all_party = sum(abs(cmp$dh) < 0.05),
  states_agreeing_two_party = sum(abs(cmp$d2) < 0.05))

head(cmp[order(-abs(cmp$d2)), c("state_name", "dh", "d2", "missing_pct")], 5)
```

**Write about:** did the prediction hold? Say what happened to the number of
states that agree. Then look at the handful that still disagree — check their
`missing_pct` column. Does the leftover disagreement have the same explanation as
the original, or a different one? A diagnosis that survives a test it could have
failed is worth more than one that was never tested; say whether this one earned
that.

***

### Option B — Rank the states by data quality

Which states' county returns are least complete?

```{r opt-b, fig.width = 7.5, fig.height = 5.5}
n <- 12                                  # <- change this number

q <- cmp[order(-cmp$missing_pct), ][1:n, ]
q <- q[order(q$missing_pct), ]

par(mar = c(5, 10, 4, 2))
barplot(q$missing_pct, names.arg = q$state_name, horiz = TRUE, las = 1,
        col = "#4D4D4D", border = NA,
        xlab = "% of ballots the county file never saw",
        main = "Where the county returns are least complete")

# Is incompleteness about how big a state is?
round(cor(cmp$missing_pct, log(cmp$total_votes)), 3)
```

**Write about:** name the worst states. Then take the correlation seriously — it
is about $-0.19$, which is nearly nothing. So incompleteness is *not* explained by
state size. Try to say what it is explained by, and be willing to end at "these
are individual state reporting decisions, not a pattern." That is a real finding,
and it is more honest than inventing a story. Why would it matter to a journalist
which states these are?

***

### Option C — Where did the country move, and who is missing?

The famous 2024 result — big Republican swings along the Texas border and in the
Bronx — with the question the headline never asks.

```{r opt-c, fig.width = 7.5, fig.height = 5.5}
top_sw <- head(sw[order(-sw$swing), ], 10)
top_sw <- top_sw[order(top_sw$swing), ]

par(mar = c(5, 12, 4, 2))
barplot(top_sw$swing, names.arg = paste0(top_sw$county_name, ", ",
        substr(top_sw$state_name, 1, 2)), horiz = TRUE, las = 1,
        col = "#B2182B", border = NA,
        xlab = "Republican two-party swing, 2020 to 2024 (points)",
        main = "Largest Republican swings")

# Now the bookkeeping the headline skips
c(counties_in_2024_file = nrow(cty),
  counties_in_2020_file = nrow(old),
  counties_that_matched = nrow(sw))

setdiff(cty$state_name, sw$state_name)     # states with no rows at all
sw[sw$county_fips == "11001", c("county_name", "r20", "r24", "swing")]
```

**Write about:** describe the pattern in the figure. Then do the part the
headline skips — two whole states are missing from this analysis entirely, and
one row in the table is comparing two different things. Name them, explain why
each happened, and say whether their absence changes how much you would trust a
national story built on this table.

***

### Option D — Ask your own question

If something in this lab made you suspicious, chase it. You do not need new code
— re-running an existing chunk with a different state, a different year, or a
different threshold is enough. A sharp question answered with a chunk you barely
touched beats an elaborate figure with nothing to say.

***

**What to submit.** Knit this document to HTML or PDF and upload it to Canvas,
with your chosen chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a result and tell me what it does and does not
support.

One answer is always available to you and is always worth full marks when it is
true: *the data cannot support this claim, and here is the check that convinced
me.*

***


# What today was really about

Four things broke, and it is worth sorting them by cause.

**Two were about this family specifically.** The missing third-party votes and
the county that changed meaning between 2020 and 2024 happen because **no single
body produces American election returns**. Fifty-one jurisdictions certify
separately, a scraper stitches them together, and the stitching inherits every
inconsistency underneath. A census or an FEC filing cannot fail this way,
because somebody is accountable for the whole file.

**Two were general.** Joining on names rather than codes, and a join that
succeeds while still being wrong, will bite you in any dataset in any field.
You will meet both again in Week 12 with census geography.

So the lesson is not "election data is bad." It is that **each family fails in
its own characteristic way**, and knowing the family tells you what to check
first:

- **Returns** — who assembled this, from how many separate sources, and does it
  match the certified count? (Ours was about 8,000 votes off for one candidate.)
- **The census** — what exactly was counted, on what date, and what is the
  margin of error? *(Week 2)*
- **Surveys** — who was asked, how many refused, and how were the rest
  weighted? *(Week 3)*
- **Administrative records** — what was this collected *for*, and what does that
  purpose leave out? *(Weeks 5 and 7)*

Those four questions are most of what this course teaches. Everything after
today is practice.

***

## Where the data came from

**County returns, 2020 and 2024** — Tony McGovern, *United States General
Election Presidential Results by District, Ward, or County from 2008 to 2024*
(<https://github.com/tonmcg/US_County_Level_Election_Results_08-24>,
DOI 10.5281/zenodo.14223604). The repository's README is worth reading precisely
because it is candid: the 2024 figures were **scraped from Fox News's results
pages**, the 2020 figures from Fox News, Politico and the New York Times, and the
README states plainly that the results "are not authoritative." Almost every
county-level analysis you will read in the press rests on a file like this one.
Its national totals land about 8,000 votes below the certified count for Trump
and 11,000 below for Harris — impressively close, and still not the official
number.

**Official county list** — U.S. Census Bureau, *2024 Gazetteer Files, Counties*
(<https://www2.census.gov/geo/docs/maps-data/data/gazetteer/2024_Gazetteer/>),
trimmed to codes and names for the 50 states and DC. This is the legal geography:
what officially counts as a county in 2024.

**State returns** — the identical file you used in Week 1, built from jaytimm's
`PresElectionResults` compilation. Its vote columns are percentages, and its
`other` column is zero wherever the source did not separately tabulate
third-party votes — which, as Part 3 showed, is most of the country.

The script that downloaded, trimmed and checked all of this is
`data/build-data.R`. Run it and it will rebuild every file in the folder from the
original sources. You should be able to ask that of any dataset you are handed,
and you should be suspicious when you cannot.
