---
title: "Key: The Ground Moves Under the Data"
subtitle: "84-355 Democracy's Data | Precinct geography | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
D  <- "../ga-precinct-returns/data"
HC <- "HOUSTON"
a  <- read.csv(file.path(D, "assign_point.csv"),      stringsAsFactors = FALSE)
ba <- read.csv(file.path(D, "block_assign.csv"),      stringsAsFactors = FALSE)
ar <- read.csv(file.path(D, "crosswalk.csv"),         stringsAsFactors = FALSE)
po <- read.csv(file.path(D, "crosswalk_pop.csv"),     stringsAsFactors = FALSE)
cp <- read.csv(file.path(D, "crosswalk_compare.csv"), stringsAsFactors = FALSE)
ck <- read.csv(file.path(D, "crosswalk_check.csv"),   stringsAsFactors = FALSE)
ts <- read.csv("fig_twosource.csv",                   stringsAsFactors = FALSE)
wm <- read.csv("fig_rozr_meta.csv",                   stringsAsFactors = FALSE)

# 30 rows of the 2020 shapefile (27 of 2024) carry no county and no precinct
# name, so the build key collapses them into a single pseudo-precinct "NA|NA".
# Left in, it masquerades as a real precinct and inflates the extremes -- the
# "worst disagreement" reads 0.894 and is really 0.890. Drop it from EVERY file
# it appears in, not just the comparison table.
ok   <- function(x) !is.na(x) & x != "NA|NA"
junk <- function(d) d[ok(d$from_2020) & ok(d$to_2024), ]
a <- junk(a); ar <- junk(ar); po <- junk(po); cp <- junk(cp)

cty <- function(x) sub("\\|.*$", "", x)
nm  <- function(x) sub("^[^|]*\\|", "", x)
inC <- function(d) d[cty(d$from_2020) == HC, ]

p20 <- unique(a$from_2020);  p24 <- unique(a$to_2024)
h20 <- sort(unique(a$from_2020[cty(a$from_2020) == HC]))
h24 <- sort(unique(a$to_2024[cty(a$to_2024)   == HC]))
h24all <- sort(unique(ba$precinct_2024[!is.na(ba$precinct_2024) &
                                       cty(ba$precinct_2024) == HC]))
arH <- inC(ar); poH <- inC(po); cpH <- inC(cp)
baH <- ba[!is.na(ba$precinct_2020) & cty(ba$precinct_2020) == HC, ]
sp  <- function(d, f = 0) { d <- d[d$weight >= f, ]; round(100*mean(table(d$from_2020) > 1), 1) }
sv  <- function(A, B) round(100 * length(intersect(A, B)) / length(A), 1)
gap <- abs(wm$ballots_area[wm$target == "PEC"] - wm$ballots_pop[wm$target == "PEC"])
spl <- names(which(table(poH$from_2020) > 1))   # Houston precincts really split
ren <- setdiff(h20, h24)                        # Houston precincts merely renamed
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`precinct-geography-brief.Rmd`) is now a chapter, "The
Ground Moves Under the Data": it is organised by argument under prose headings and
carries four numbered D3 figures rather than numbered steps — Figure 1, Houston
County's precincts in 2020 and 2024; Figure 2, disaggregate-then-reaggregate in
one sketch; Figure 3, the one piece of Houston ground that changed precinct;
Figure 4, where land and people give different answers across Georgia. The lab
(`precinct-geography.Rmd`) is unchanged and still runs Steps 1–10 with the code
exposed. **Students should read the brief first.** **Step numbers below refer to
the lab.** The chapter covers the same ground in the same order, so each lab Step
has a heading facing it: Step 1 = "Nobody in Atlanta drew these lines"; Step 2 =
"Sixteen precincts you can name" (Figure 1); Step 3 = "The repair that deletes the
wrong precincts"; Steps 4 and 5 = "A unit that both maps contain" (Figure 2);
Step 6 = "Land or people", worked out in "One precinct, taken apart" (Figure 3)
and scaled up in "How bad it gets elsewhere" (Figure 4); Steps 7, 8 and 9 =
"Carrying the votes, and checking that they arrived"; Step 10 = "What the
crosswalk licenses".

**The worked example is Houston County, and the statewide numbers stay.** Every
step is illustrated in one county you can talk about from memory, then checked
against Georgia so nobody thinks the county was chosen for its answer. Where the
two differ, the difference is the lesson — most sharply in Step 6, where Houston
is a *stronger* case than the state, and in Option B, where it is a much weaker
one.

**This is a methods session, and it is the only one in the course that teaches a
procedure students could be asked to perform.** Everything else asks them to read
data somebody else assembled. Here they learn a thing redistricting analysts,
expert witnesses and election reporters actually do, and why it is contested.

Every number below is computed from the built files at knit time.

***

# The session in one sentence

Precinct boundaries move between elections, so comparing precincts across years
means putting both years on one map — and the reasonable ways of doing that
disagree with each other about something as basic as whether a precinct was
split.

***

# The findings

```{r headline}
rbind(
  HOUSTON = c(name_survival_pct = sv(h20, h24),
              precincts_2020    = length(h20),
              blocks            = nrow(baH),
              population        = sum(baH$pop),
              areal_split_pct   = sp(arH),
              pop_split_pct     = sp(poH),
              pairs_over_10pts  = sum(abs(cpH$diff) > 0.10),
              worst_gap         = round(max(abs(cpH$diff)), 3)),
  GEORGIA = c(sv(p20, p24), length(p20), nrow(ba), sum(ba$pop, na.rm = TRUE),
              sp(ar), sp(po), sum(abs(cp$diff) > 0.10), round(max(abs(cp$diff)), 3)))
```

- **`r sv(h20, h24)`% of Houston's precinct names survive 2020→2024;
  `r sv(p20, p24)`% of Georgia's do.** Statewide, precinct *codes* do worse still
  — about 75%. Two keys, two answers.
- **Houston: `r format(nrow(baH), big.mark=",")` blocks,
  `r format(sum(baH$pop), big.mark=",")` people.** Georgia:
  `r format(nrow(ba), big.mark=",")` blocks,
  `r format(sum(ba$pop, na.rm=TRUE), big.mark=",")` — the census count exactly.
- **Areal weighting calls `r sp(arH)`% of Houston's precincts split; population
  weighting calls `r sp(poH)`%.** Statewide: `r sp(ar)`% and `r sp(po)`%. **The
  county's contrast is the sharper one** — that is why the example moved there.
- **`r round(100*mean(abs(cpH$diff) > 0.10), 1)`% of Houston's weight pairs
  differ by more than ten points, worst `r sprintf("%.3f", max(abs(cpH$diff)))`.**
  Statewide `r round(100*mean(abs(cp$diff) > 0.10), 1)`%, worst
  `r sprintf("%.3f", max(abs(cp$diff)))`. **Do not quote the statewide extreme as
  Houston's** — see the note under Option B.
- The carry loses **`r ck$value[ck$check == "votes lost"]` votes of
  `r ck$value[ck$check == "votes before"]`**.

## The Houston inventory, which is the new centre of the session

```{r inventory}
aH <- inC(a); aH <- aH[order(aH$from_2020), ]
data.frame(precinct_2020 = nm(aH$from_2020), became_2024 = nm(aH$to_2024))

setdiff(nm(h24all), nm(h24))          # the 2024 precinct no interior point sees
```

**`r length(h20)` lines, and they tell three different stories.**

1. **`r length(intersect(h20, h24))` precincts are unchanged** in name and, as
   the crosswalk confirms, in ground.
2. **`r length(ren)` are renames**: HCTC→VFW, NSES→NORTH HOUSTON SPORT COMPLEX,
   RECR→WELLSTON CENTER. Population weighting sends **100%** of each old
   precinct's people to the new name; areal overlap is above 99.7%. **The
   polling place moved and the boundary did not.**
3. **`r length(spl)` is a real split**: ROZR shed its western half to a
   brand-new precinct, PEC.

```{r renames}
po[po$from_2020 %in% ren, ]                   # the renames: weight is exactly 1
po[po$from_2020 %in% spl, ]                   # the one real split
```

**This is the single biggest teaching win of the county version, and it is worth
protecting time for.** The statewide statistic — `r sv(p20, p24)`% of names
survive — silently merges two completely different events. In Houston you can separate them by
name, out loud, in ninety seconds: **`r length(ren)` of the
`r length(ren) + length(spl)` changes are clerical and `r length(spl)` is real.**
Ask the room which of them a name-matched comparison drops.
The answer is *the `r length(ren)` that did not matter* — the rule deletes the
harmless changes and keeps the one that would have broken the comparison.

***

# The three things to get said

**1. An identifier that is not stable is not an identifier.** Step 2 gives two
different answers depending on whether you key on name or on code, and both are
computed correctly. **Make somebody say why that is worse than either number
alone**: if the two keys disagreed only occasionally you would call it noise, but
they drift independently, which means neither is tracking the thing you care
about — a piece of ground. Houston makes it concrete: `r sv(h20, h24)`% of names
survived, and only `r length(spl)` of the county's `r length(h20)` precincts
changed its boundary at all.

**2. The dropped precincts are not missing at random** (Step 3). This is the step
students will skate over, and it is the one with real consequences. Statewide the
argument is that counties split precincts where population grew, so dropping
non-matching precincts deletes the fastest-growing places. **In Houston the
mechanism is visible and it points the other way**: name-matching drops the three
renames and keeps ROZR, the only precinct whose ground moved. Either way,
**name-matching is not a filter on boundary change at all** — it is a filter on
clerical churn. Ask what a turnout-change map looks like when the filter is
uncorrelated with the thing it is supposed to control for.

**3. Step 6 is the session.** Two defensible methods, computed on the same data,
disagreeing about whether `r sp(arH)`% or `r sp(poH)`% of a county's precincts
were split. Neither is a bug. Area answers "how much land moved"; population
answers "how many people moved." **The lesson is not which to pick — it is that a
methodological choice this invisible moves the answer this much.**

New this version, and worth two minutes: **most of what area calls a split is a
sliver.**

```{r slivers}
c(houston_area = sp(arH), houston_area_1pct_floor = sp(arH, 0.01),
  houston_pop  = sp(poH), houston_pop_1pct_floor  = sp(poH, 0.01),
  georgia_area = sp(ar),  georgia_area_1pct_floor = sp(ar, 0.01),
  georgia_pop  = sp(po),  georgia_pop_1pct_floor  = sp(po, 0.01))
```

Ignore any fragment smaller than 1% of a precinct and Houston's areal figure
falls from `r sp(arH)`% to `r sp(arH, 0.01)`%, exactly matching population.
Statewide the same correction takes area from `r sp(ar)`% to `r sp(ar, 0.01)`%
and population from `r sp(po)`% to `r sp(po, 0.01)`%; the
`r round(sp(ar, 0.01) - sp(po, 0.01), 1)`-point residue is genuine empty land.
**This links Step 5 to Step 6** — the slivers Step 5 says the interior-point
method avoids are the same slivers inflating area's answer in Step 6.

**The line for the board:** the map is not the ground, and the label is not
the map.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** Who owns a boundary | The state's own "best fit" language | 6 | **Read the quotation aloud.** Students assume the state draws precincts. It does not. Houston's board does. |
| **2** How much moved | `r sv(h20, h24)`% of Houston's names, `r sv(p20, p24)`% of Georgia's | 7 | The two-key disagreement is the finding, not a nuisance. |
| **2b** The inventory | `r length(h20)` lines: `r length(intersect(h20,h24))` same, `r length(ren)` renamed, `r length(spl)` split | **8** | **New, and the best thing in the lab.** Do it by name, out loud. |
| **3** Why you cannot drop them | Non-random missingness | 7 | Houston drops the three harmless ones and keeps the real one. |
| **4** The recipe | Target, disaggregate, reaggregate | 5 | Houston: `r format(nrow(baH), big.mark=",")` blocks, `r round(nrow(baH)/length(h20))` per precinct. |
| **5** How a block is assigned | Interior point, not intersection | 6 | PEC is invisible to a precinct-level interior point. That is why blocks. |
| **6** Two weightings | `r sp(arH)`% vs `r sp(poH)`% in Houston | **12** | **The pivot.** Take a vote before revealing the second number. |
| **7** Carry and check | `r ck$value[ck$check == "votes lost"]` votes of `r ck$value[ck$check == "votes before"]` | 8 | The 9.8% failure story lives here. Do not rush it. |
| **8** Was the shortcut legitimate | `r format(ts$identical[ts$scope=="Georgia"], big.mark=",")` of `r format(ts$matched[ts$scope=="Georgia"], big.mark=",")` identical | 4 | Check *before* relying, not after. |
| **9** What you can say | Including what you cannot | 6 | Population is not voters. |

**Step 6 deserves the prediction.** Ask the room: *areal weighting says
`r round(sp(arH))`% of Houston's precincts are split by the boundary change —
what will population weighting say?* Almost everyone predicts a similar number.
It is `r sp(poH)`% — **`r length(spl)` precinct** — and the reason (most splits
are slivers or empty land) is obvious in hindsight and almost never guessed.
Houston makes the reveal sharper than the statewide version did, because
"`r length(spl)` precinct out of `r length(h20)`" is a fact about a place rather
than a percentage.

**If you are short on time**, protect 2b, 6 and 7. The inventory, the two
weightings, and the join that lost a tenth of the state are the argument.

***

# The mistake in Step 7, and how to teach it

**Tell this story, because it happened while building the lab.**

The first version attached votes to the map by matching precinct names. It
reached 90.4% and lost **489,876 votes — 9.8% of Georgia** — concentrated in
Chatham (92 precincts) and DeKalb (48), where the returns and the shapefile spell
precincts differently.

The fixes attempted, in order:

1. Strip a leading ballot-order number: `"01 BETHLEHEM COMMUNITY CENTER"` →
   `"BETHLEHEM COMMUNITY CENTER"`. Loss fell from 11.1% to 9.8%.
2. Allow a trailing letter on the code: `"01C FIRST PRESBYTERIAN CHURCH"`. Loss
   did not move at all.

**Neither was the answer, and the third attempt was not another regex.** The
precinct shapefile already carried the vote totals in its own attribute table,
attached to the geometry. There was never a join to make.

The pedagogy: **ask the class what they would have tried third.** Most will
propose fuzzy matching or a manual alias table — both reasonable, both more work,
both unnecessary. The instinct to improve a join is strong enough that it
crowds out the question of whether the join is needed.

Then note what made it safe to abandon: Step 8 shows the two state sources agree
on every precinct the join reached.

```{r twosource}
ts
round(100 * ts$matched / ts$shapefile_precincts, 1)     # what the join reached
```

**The shortcut was verified before it was relied on**, which is the difference
between a shortcut and a guess. And note the second line: the join reached
`r round(100*ts$matched[ts$scope=="Georgia"]/ts$shapefile_precincts[ts$scope=="Georgia"], 1)`%
of Georgia's precincts and **`r round(100*ts$matched[ts$scope==HC]/ts$shapefile_precincts[ts$scope==HC], 1)`%
of Houston's** — the data was never wrong, the labels were.

> **A note on a number you may remember.** Earlier drafts of this key quoted
> "2,401 of 2,401" for the two-source check. That figure came from a
> normalisation of the returns that is no longer in the pipeline. Recomputed from
> the files as they now stand, the plain `COUNTY|PRECINCT` join reaches
> `r format(ts$matched[ts$scope=="Georgia"], big.mark=",")` precincts and
> **agreement is still exact on every one of them**, which is the part that
> matters. The chunk above is the live number; use it.

***

# Handling the room

This lab is more technical than the rest of the course and a few students will
decide it is "the coding one" and disengage. Three things help:

- **Nobody writes code here either.** Say so. The procedure is conceptual —
  target, disaggregate, reaggregate — and the code is given.
- **Anchor it in a place, not a statistic.** The county version exists for this.
  "Warner Robins" and "the precinct at the tech college" beat "78.3% of
  precincts" every time; the statewide figure is there to stop anyone concluding
  that Houston is weird.
- **The disagreement in Step 6 is the interesting part**, not the machinery. If
  the room only takes away that two reasonable methods said `r sp(arH)`% and
  `r sp(poH)`% about the same `r length(h20)` precincts, that is a successful
  session.

***

# Option A — Use the areal crosswalk instead

**Scope: Houston, `r length(h24all)` precincts.** Only one of them moves at all,
which is the point.

## Verified output

```{r a}
big <- cpH[order(-abs(cpH$diff)), c("from_2020","to_2024","weight_area","weight_pop","diff")]
big[, 3:5] <- round(big[, 3:5], 3)
head(big, 5)

wm     # ballots, both ways, for the one precinct that actually splits
```

**3/4:** Swaps the file, recomputes, and reports that
`r length(h20) - length(spl)` of Houston's `r length(h20)` precincts do not move
at all.

**4/4:** Identifies ROZR as the only precinct where the choice matters, quantifies
it (**`r format(gap, big.mark=",")` ballots** out of
`r format(wm$ballots_total[1], big.mark=",")` cast), and explains *why*: the
western half is large and thinly settled, so area over-weights it and population
does not.

**4/4, best:** Generalises from the one case — area and population agree wherever
population density is uniform and diverge wherever it is not, so **the choice of
method is effectively a choice about how much sparsely settled ground counts.**
A student who then checks that claim against the statewide scatter in the brief
(Figure 4, "How bad it gets elsewhere"),
and notices that the off-diagonal points are the low-density ones, has done the
whole exercise.

***

# Option B — Find the worst disagreement in Georgia

> **Scope is statewide, and the lab now says so explicitly.** Houston's worst
> disagreement is only `r sprintf("%.3f", max(abs(cpH$diff)))` — the ROZR split — so a
> student who searches only the county will not find anything dramatic and should
> not be marked down for reporting that. **The `r sprintf("%.3f", max(abs(cp$diff)))`
> case does not exist in Houston.** If a student's answer quotes
> `r sprintf("%.3f", max(abs(cp$diff)))` as a Houston number, that is the error to
> correct.

## Verified output

```{r b}
w <- cp[which.max(abs(cp$diff)), ]                 # cp already has "NA|NA" dropped
data.frame(from = w$from_2020, to = w$to_2024,
           area = round(w$weight_area, 3), pop = round(w$weight_pop, 3),
           gap  = round(w$diff, 3))

# for contrast, the same query inside the county
cpH[which.max(abs(cpH$diff)), ]
```

**Watch for the junk row.** A student who forgets the `"NA|NA"` guard will report
0.894 from the pseudo-precinct instead of `r sprintf("%.3f", max(abs(cp$diff)))` from
`r nm(w$from_2020)`. Noticing the guard is itself worth credit; the lab and the
brief both print it.

**3/4:** Reports the pair and the size of the gap.

**4/4:** Works out what the gap means concretely — one method sends about
`r round(100*w$weight_area)`% of that precinct's votes to the target, the other
sends about `r round(100*w$weight_pop)`%. Says which is more likely correct and
why.

**4/4, best:** Recognises that a single pair cannot be adjudicated from the
table, and that settling it needs a map or local knowledge — then says what they
would look for (a river, an industrial tract, a park). **Knowing that the data
cannot answer it is the right answer.** The strongest version notices the
asymmetry the county example sets up: *in Houston we could adjudicate, because
somebody in the room knows the ground; in Chatham nobody in the room does, and
the method has no way to tell you that it is out of its depth.*

***

# Option C — Turnout, honestly

**Scope: Houston.** No fully verified output; grade the reasoning and the
comparison.

**3/4:** Computes turnout change both ways and reports that the numbers differ.

**4/4:** Quantifies the difference and names the three precincts the naive
method drops (`r paste(nm(setdiff(h20, h24)), collapse = ", ")`), noting that
the naive method is not merely noisier but **structurally wrong** — it deletes
precincts for a clerical reason unrelated to whether the ground moved.

**4/4, best:** Notices that even the crosswalked version has a problem the lab
warned about: **2020 turnout carried on 2020 population weights, compared against
2024 turnout, attributes to "turnout change" some of what is really population
change.** In a county that split a precinct precisely because it was growing,
that is not a hypothetical. A student who sees that has understood the limitation
better than the method does.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given.
- **The move that earns a 4** is treating the weighting choice as a substantive
  decision with consequences, not a technical detail.
- **Also a 4**: separating "the name changed" from "the boundary changed."
  Houston hands them the distinction; using it on a number they did not compute
  themselves — the statewide `r sv(p20, p24)`% — is the transfer worth rewarding.
- **A 2** is "the boundaries changed so the data is unreliable" — true, useless,
  and the thing the whole lab is designed to move past.
- **Also a 2**: reporting the crosswalked numbers as though they were counted.
- **Also a 2**: quoting a Houston illustration as though it were a statewide
  finding, or the reverse. The documents are built to make the scope explicit;
  ignoring it is the failure mode the county version introduces.
- **Reward anyone who asks for an error bar.** There isn't one, and noticing its
  absence is the most sophisticated available response.

## Anticipated questions

- *"Why Houston?"* — Because it is small enough to check by hand and because the
  instructor knows the ground. Then say what the statewide columns are for: if
  the county were doing the work, they would not be there.
- *"Is Houston typical?"* — On name survival, almost exactly (`r sv(h20, h24)`%
  against `r sv(p20, p24)`%). On the area-versus-population split it is more
  extreme, and the honest answer is that this is partly a small-county artefact:
  with `r length(h20)` precincts, `r length(spl)` real split is `r sp(poH)`%.
- *"Why not just use counties?"* — You can, and it is stable. You also lose all
  the resolution that makes precinct data worth having. This is a real position
  some analysts hold; say so.
- *"Couldn't we weight by registered voters?"* — Yes, and it would be better.
  Nobody publishes registered voters by census block. Worth naming as the gap.
- *"Is this what expert witnesses do?"* — Yes. Effectively this procedure, with
  more care about which small unit to use.
- *"How would I know if a precinct changed?"* — By noticing that it did. There is
  no notification, no changelog, and no requirement to publish one. That answer
  usually lands harder than anything else in the session.

## If you want to check a Houston number against a third source

Houston County publishes its own precinct-level results at
<https://www.houstoncountyga.gov/residents/election-results.cms>. Nothing in the
lab is built on it and no chunk fetches it. It is worth knowing about for two
reasons: it is a genuinely independent copy of the same totals, so it settles any
question about a Houston figure on the spot; and it is a live example of the
Step 8 point that three offices publish the same numbers separately and the only
thing that ever disagrees is the labels.

## Rebuilding the figures

`build-houston-figures.R` in this folder reads the shapefiles and block
assignments in `../ga-precinct-returns/data/` and writes the small coordinate
tables the maps draw from (`fig_houston*.csv`, `fig_rozr*.csv`,
`fig_twosource.csv`). **Rerun it after any rebuild of the crosswalks**, then
reknit all three documents; every number in them is computed at knit time, so a
stale figure file is the only way they can go out of date.

## Connections

- **`census-geography`** — the nesting spine, and why precincts sit off it. This
  lab is that lab's abstract point made concrete and consequential.
- **`ga-precinct-returns` and `rpv`** — both use precinct data as though the
  unit were fixed. Run this first if you can.
- **`redlining`** — point-in-polygon, the same uniform-distribution assumption,
  and there we could not measure the error. Here we can.
- **`apportionment`** — the other lab where a technical rule with no
  obvious politics turns out to allocate something people care about.
