---
title: "The Ground Moves Under the Data"
subtitle: "84-355 Democracy's Data | Precinct boundaries, and how to compare two elections"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4,
                      dpi = 96, fig.retina = 1)
options(scipen = 999)
D  <- "../ga-precinct-returns/data"
HC <- "HOUSTON"
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `precinct-geography-brief.Rmd` first.** This lab runs the same steps
> with the code exposed.

**The plan.** Every step is done twice: once in **Houston County**, which is
small enough to check by hand, and once across Georgia, to show the county is
representative rather than convenient. Watch for the places where the two
answers differ — those are the interesting ones.

***

# Where did that number come from?

> *"Turnout in this precinct fell 9 points between 2020 and 2024."*

A sentence like that is in every post-election story, and it contains a hidden
claim: **that "this precinct" means the same piece of ground in both years.**

It usually does not. This lab is about what to do instead.

```{r helpers}
# Two helpers used throughout. Precinct keys look like "HOUSTON|ROZR".
cty <- function(x) sub("\\|.*$", "", x)      # county part
nm  <- function(x) sub("^[^|]*\\|", "", x)   # precinct part

# 30 rows of the 2020 shapefile carry no county and no precinct name, so the
# build script collapses them into one pseudo-precinct "NA|NA". It is not a
# precinct. Drop it before any statistic.
ok   <- function(x) !is.na(x) & x != "NA|NA"
junk <- function(d) d[ok(d$from_2020) & ok(d$to_2024), ]
```

# Step 1 — Who owns a precinct boundary

Read this, from Georgia's Legislative and Congressional Reapportionment Office,
which publishes the statewide precinct map:

> *"Each individual county across the State of Georgia maintains the official
> record for the boundaries of their voting precinct lines. County election
> officials are responsible for any adjustments to these boundaries as well as
> assigning voters to the specific precinct in which they will vote. The Office
> of Legislative and Congressional Reapportionment compiles all 159 voting
> precinct maps provided by the counties into one master shape file...*
>
> *This file is a **best fit** version of the maps provided by county election
> offices, created using current census geography of that decade."*

Three facts, stated by the state about its own data:

1. **159 counties are the authority.** Not the state.
2. **They adjust boundaries whenever they like.** No cycle, no notice.
3. **The statewide file is a "best fit"** to census geography — which is an
   admission that precincts do not naturally align to it.

Houston County is one of the 159, and the Houston County Board of Elections is
the only body that has ever drawn its precinct lines.

# Step 2 — How much moved

```{r stability}
a <- junk(read.csv(file.path(D, "assign_point.csv"), stringsAsFactors = FALSE))

# statewide
p20 <- unique(a$from_2020); p24 <- unique(a$to_2024)
# one county
h20 <- sort(unique(a$from_2020[cty(a$from_2020) == HC]))
h24 <- sort(unique(a$to_2024[cty(a$to_2024)   == HC]))

rbind(
  HOUSTON = c(precincts_2020 = length(h20), precincts_2024 = length(h24),
              in_both = length(intersect(h20, h24)),
              only_2020 = length(setdiff(h20, h24)),
              only_2024 = length(setdiff(h24, h20)),
              pct_surviving = round(100 * length(intersect(h20, h24)) / length(h20), 1)),
  GEORGIA = c(length(p20), length(p24), length(intersect(p20, p24)),
              length(setdiff(p20, p24)), length(setdiff(p24, p20)),
              round(100 * length(intersect(p20, p24)) / length(p20), 1)))
```

Houston has the same *number* of precinct names in both years and they are not
the same names: `r length(setdiff(h20, h24))` go, `r length(setdiff(h24, h20))` arrive. Statewide the survival rate is very
close to Houston's, so this county is a fair sample of the problem, not an
outlier.

Measured on the counties' precinct *codes* rather than their names, statewide
survival is only about 75%. **Two keys, two answers, and neither is wrong** —
names and codes drift independently, which is the first sign that neither is a
stable identifier for a piece of ground.

Either way, a sixth to a quarter of any 2020-to-2024 precinct comparison is
comparing a label to itself and nothing else. **Do not fix this by dropping the
non-matching precincts** — they are not missing at random.

## Every Houston precinct, by name

The county had `r length(h20)` precincts in 2020, few enough to print in full.
This is all of them.

```{r inventory}
aH <- a[cty(a$from_2020) == HC, ]
aH <- aH[order(aH$from_2020), ]
data.frame(precinct_2020 = nm(aH$from_2020),
           became_2024   = nm(aH$to_2024),
           row.names = NULL)
```

**`r length(intersect(h20, h24))` rows say the same thing twice.** The other
`r length(setdiff(h20, h24))` are name changes:
HCTC became VFW, NSES became NORTH HOUSTON SPORT COMPLEX, RECR became WELLSTON
CENTER. Step 6 will show that in every one of those cases the *ground* did not
move at all — only the polling place, and therefore the label.

Then there is a fourth change this table cannot show you, because a
precinct-to-precinct table is not able to record a split. Step 5 comes back to
it.

# Step 3 — The instinct that fails

The obvious repair is to match on precinct name and accept the losses.

```{r naive}
c(HOUSTON = round(100 * length(intersect(h20, h24)) / length(union(h20, h24)), 1),
  GEORGIA = round(100 * length(intersect(p20, p24)) / length(union(p20, p24)), 1))

# which Houston precincts does name-matching throw away?
setdiff(nm(h20), nm(h24))
```

Matching on name keeps about three quarters of the names in either scope and
silently discards the rest. In Houston the discarded precincts are named above,
and by Step 6 you will be able to say something sharper about them: **every one
of them is a precinct whose boundary did not move.** The rule deletes the places
where nothing happened.

Statewide the bias runs the other way as well — growing suburbs split precincts,
so fast-changing places also fall out. Either way the discard is not random.

**We need a way to put 2020's votes onto 2024's map.**

# Step 4 — Pick a target, and a common currency

The recipe is always the same:

1. **Choose the newest boundaries as the target.** They are the ones future data
   will use.
2. **Disaggregate the old data onto a smaller unit that both years contain.**
3. **Reaggregate that unit up to the target.**

The smaller unit here is the **census block** — small enough that precincts
rarely cut through one, and nested inside exactly one county.

```{r blocks}
ba  <- read.csv(file.path(D, "block_assign.csv"), stringsAsFactors = FALSE)
baH <- ba[!is.na(ba$precinct_2020) & cty(ba$precinct_2020) == HC, ]

rbind(
  HOUSTON = c(blocks = nrow(baH), population = sum(baH$pop, na.rm = TRUE),
              placed_2020 = sum(!is.na(baH$precinct_2020)),
              placed_2024 = sum(!is.na(baH$precinct_2024))),
  GEORGIA = c(nrow(ba), sum(ba$pop, na.rm = TRUE),
              sum(!is.na(ba$precinct_2020)), sum(!is.na(ba$precinct_2024))))
```

Houston's `r length(h20)` precincts are built out of
`r format(nrow(baH), big.mark = ",")` blocks holding
`r format(sum(baH$pop), big.mark = ",")` people, or about
`r round(nrow(baH) / length(h20))` blocks per precinct. That is the resolution
everything below depends on.

Statewide, six blocks fell outside the 2020 map, almost certainly along the
coast where the two agencies drew the shoreline differently. None of them is in
Houston. The state population total is Georgia's 2020 census count, to the
person.

# Step 5 — How a block gets assigned

Not by overlaying polygons. **Reduce each block to a point guaranteed to lie
inside it, then ask which precinct contains that point.**

That is the method in `assignPolys` (Cervas, *R-Functions*), and it is more
robust than intersecting polygons: no sliver fragments where two agencies drew
the same line slightly differently, no fractional pieces, and every block gets
exactly one answer.

```
block  --interior point-->  which 2020 precinct contains it?
                            which 2024 precinct contains it?
```

## What the interior point cannot see

The inventory in Step 2 applied the same trick to whole precincts, and that is
where its blind spot comes from.

```{r unseen}
h24all <- sort(unique(ba$precinct_2024[!is.na(ba$precinct_2024) &
                                       cty(ba$precinct_2024) == HC]))
c(named_by_interior_points = length(h24),
  actually_on_the_2024_map = length(h24all))

setdiff(nm(h24all), nm(h24))
```

**PEC is a real 2024 precinct that the interior-point table cannot name.** It
was carved out of ROZR, and no 2020 precinct has its interior point inside it,
so nothing in a precinct-to-precinct table points at it.

That is the whole argument for the block. A precinct reduced to one point can
only land in one place, so it can never record a split. Thousands of blocks,
each reduced to one point, can.

Blocks are now the common currency. A 2020 precinct's votes can travel to 2024
precincts in proportion to **where its people are**.

# Step 6 — Two ways to weight, and they disagree

```{r weights}
ar <- junk(read.csv(file.path(D, "crosswalk.csv"),     stringsAsFactors = FALSE))  # area
po <- junk(read.csv(file.path(D, "crosswalk_pop.csv"), stringsAsFactors = FALSE))  # people
arH <- ar[cty(ar$from_2020) == HC, ]
poH <- po[cty(po$from_2020) == HC, ]

split_pct <- function(d, floor = 0) {
  d <- d[d$weight >= floor, ]
  round(100 * mean(table(d$from_2020) > 1), 1)
}
rbind(
  HOUSTON = c(areal_pct_split = split_pct(arH), pop_pct_split = split_pct(poH),
              areal_at_1pct   = split_pct(arH, 0.01),
              pop_at_1pct     = split_pct(poH, 0.01)),
  GEORGIA = c(split_pct(ar), split_pct(po), split_pct(ar, 0.01), split_pct(po, 0.01)))
```

**Areal weighting says almost every Houston precinct straddles a 2024 boundary.
Population weighting says one does.** The statewide gap points the same way and
is less extreme, so the county sharpens the finding rather than inventing it.

The last two columns say what the gap is made of. Ignore any piece smaller than
1% of a precinct and area's answer collapses onto population's. **Most of what
area calls a "split" is a sliver** — a hairline where the 2020 and 2024 files
digitised the same road a metre apart, exactly the artefact Step 5 said the
interior-point method avoids. What is left over after the slivers are removed is
**empty land**: a boundary clipping a corner of farmland with nobody on it.

```{r rozr}
# the one Houston precinct that population weighting agrees was split
po[po$from_2020 == "HOUSTON|ROZR", ]
ar[ar$from_2020 == "HOUSTON|ROZR", ]

# the same split expressed in ballots, from build-houston-figures.R
wm <- read.csv("fig_rozr_meta.csv", stringsAsFactors = FALSE)
wm
```

```{r rozr-gap, include=FALSE}
gapv <- abs(wm$ballots_area[wm$target == "PEC"] - wm$ballots_pop[wm$target == "PEC"])
```

ROZR is the only piece of Houston ground that changed precinct in four years.
Area puts the line in one place; people put it in another. On the
`r format(wm$ballots_total[1], big.mark = ",")` ballots ROZR cast in 2020 that
gap is `r format(gapv, big.mark = ",")` votes moved from one precinct to another
by nothing but a methodological choice.

```{r compare}
cp  <- junk(read.csv(file.path(D, "crosswalk_compare.csv"), stringsAsFactors = FALSE))
cpH <- cp[cty(cp$from_2020) == HC, ]

summarise <- function(d) c(pairs = nrow(d),
                           median_abs_diff = round(median(abs(d$diff)), 3),
                           pct_over_10_points = round(100 * mean(abs(d$diff) > 0.10), 1),
                           worst = round(max(abs(d$diff)), 3))
rbind(HOUSTON = summarise(cpH), GEORGIA = summarise(cp))
```

**The rate generalises. The extreme does not.** Houston and Georgia disagree by
more than ten points of weight at a similar rate — but Houston's worst
disagreement is the ROZR split, and Georgia's worst is far uglier:

```{r worst}
cp[which.max(abs(cp$diff)), ]
```

One method sends almost none of that precinct's votes to the target; the other
sends nearly all of them. **Nothing like it happens in Houston.** A county is
where you learn the mechanism; the state is where you find out how bad it gets.

**Areal weighting assumes people are spread evenly across a precinct.** They are
not. That is the same assumption the redlining lab makes with point-in-polygon,
wrong in the same way, and here we can measure how wrong.

## The county, both years

The drawing code for this figure is hidden; it is plumbing, not analysis. The
geometry it reads is built by `build-houston-figures.R` in this folder.

```{r map-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
fg <- read.csv("fig_houston.csv",     stringsAsFactors = FALSE)
fl <- read.csv("fig_houston_lab.csv", stringsAsFactors = FALSE)

ren   <- poH[poH$weight > 0.999 & nm(poH$from_2020) != nm(poH$to_2024), ]
spl   <- names(which(table(poH$from_2020) > 1))
RDR   <- unique(c(nm(spl), nm(poH$to_2024[poH$from_2020 %in% spl])))
fl$cat <- ifelse(fl$name %in% RDR, "redrawn",
          ifelse(fl$name %in% c(nm(ren$from_2020), nm(ren$to_2024)), "renamed", "same"))

W <- 760; PH <- 400; PW <- 368; PAD <- 10; GAP <- 24
rx <- range(fg$x); ry <- range(fg$y)
s  <- min((PW - 2*PAD) / diff(rx), (PH - 2*PAD) / diff(ry))
mkx <- function(off) function(x) off + (PW - diff(rx)*s)/2 + (x - rx[1]) * s
mky <- function(y) PAD + 22 + (ry[2] - y) * s
ring <- function(d, by, sx, sy) {
  k <- interaction(d[[by]], d$part, drop = TRUE)
  vapply(split(d, k), function(z)
    paste0("M", paste(sprintf("%.1f,%.1f", sx(z$x), sy(z$y)), collapse = "L"), "Z"),
    character(1))
}
J <- c(); LAB <- c()
for (i in 1:2) {
  Y <- c(2020, 2024)[i]
  d <- fg[fg$yr == Y, ]; l <- fl[fl$yr == Y, ]
  fx <- mkx((i - 1) * (PW + GAP))
  P  <- ring(d, "name", fx, mky)
  nn <- sub("\\.[0-9]+$", "", names(P))
  J  <- c(J, sprintf('{"c":"%s","d":"%s"}', l$cat[match(nn, l$name)], P))
  LAB <- c(LAB, sprintf('{"x":%.1f,"y":%.1f,"t":"%s"}', fx(l$x), mky(l$y), l$code))
}
cat(sprintf('
<div id="hou" style="margin:1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const G=[%s], L=[%s], W=%d, H=%d;
const FILL={same:"#f2f2f2",renamed:"#fbe3c2",redrawn:"#cfe3f1"};
const EDGE={same:"#9a9a9a",renamed:"#c07d20",redrawn:"#2c7fb8"};
const svg=d3.select("#hou").append("svg").attr("viewBox",`0 0 ${W} ${H}`)
  .attr("style","max-width:100%%;height:auto;font:12px inherit");
[[0,"2020 \\u2014 %d precincts"],[1,"2024 \\u2014 %d precincts"]].forEach(([i,t])=>
  svg.append("text").attr("x",i*%d+%d).attr("y",14).attr("font-size","13px")
     .attr("font-weight","600").attr("fill","#333").text(t));
const g=svg.append("g");
g.selectAll("path").data(G).join("path").attr("d",d=>d.d)
  .attr("fill",d=>FILL[d.c]).attr("stroke",d=>EDGE[d.c]).attr("stroke-width",0.8);
g.selectAll("text").data(L).join("text").attr("x",d=>d.x).attr("y",d=>d.y)
  .attr("text-anchor","middle").attr("font-size","8.5px").attr("fill","#444").text(d=>d.t);
})();
</script>
', paste(J, collapse = ","), paste(LAB, collapse = ","), W, PH + 22,
   length(h20), length(h24all), PW + GAP, PAD))
```

```{r map-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=5.0}
fg <- read.csv("fig_houston.csv",     stringsAsFactors = FALSE)
fl <- read.csv("fig_houston_lab.csv", stringsAsFactors = FALSE)

ren <- poH[poH$weight > 0.999 & nm(poH$from_2020) != nm(poH$to_2024), ]
spl <- names(which(table(poH$from_2020) > 1))
RDR <- unique(c(nm(spl), nm(poH$to_2024[poH$from_2020 %in% spl])))
fl$cat <- ifelse(fl$name %in% RDR, "redrawn",
          ifelse(fl$name %in% c(nm(ren$from_2020), nm(ren$to_2024)), "renamed", "same"))
CAT  <- c(same = "#f2f2f2", renamed = "#fbe3c2", redrawn = "#cfe3f1")
CATB <- c(same = "#9a9a9a", renamed = "#c07d20", redrawn = "#2c7fb8")

op <- par(mfrow = c(1, 2), mar = c(0.2, 0.2, 1.8, 0.2))
for (Y in c(2020, 2024)) {
  d <- fg[fg$yr == Y, ]; l <- fl[fl$yr == Y, ]
  plot(NA, xlim = range(fg$x), ylim = range(fg$y), asp = 1, axes = FALSE, ann = FALSE)
  k <- interaction(d$name, d$part, drop = TRUE)
  for (i in seq_along(levels(k))) {
    z <- d[k == levels(k)[i], ]; j <- match(z$name[1], l$name)
    polygon(z$x, z$y, col = CAT[l$cat[j]], border = CATB[l$cat[j]], lwd = 0.7)
  }
  text(l$x, l$y, l$code, cex = 0.42, col = "#333333")
  title(sprintf("%d - %d precincts", Y, nrow(l)), cex.main = 0.95, line = 0.3)
}
par(op)
```

**Amber**: the name changed and the ground did not — the
`r length(ren)` renames. **Blue**: the ground itself changed — ROZR split, and
its western half became the new precinct PEC. **Grey** is unchanged in both name
and boundary, which is `r length(intersect(h20, h24)) - length(spl)` of
`r length(h20)` precincts.

# Step 7 — Carry the votes

```{r carry}
est <- read.csv(file.path(D, "precincts_2024_pop.csv"), stringsAsFactors = FALSE)
eH  <- est[cty(est$to_2024) == HC, ]
eH  <- eH[order(-eH$VOTED20), c("to_2024", "TRUMP20", "BIDEN20", "REG20", "VOTED20")]
eH

c(houston_precincts = nrow(eH),
  houston_trump = sum(eH$TRUMP20), houston_biden = sum(eH$BIDEN20),
  georgia_precincts = nrow(est),
  georgia_trump = sum(est$TRUMP20), georgia_biden = sum(est$BIDEN20))
```

2020's presidential votes, expressed on 2024's precinct map — including PEC,
which did not exist when those votes were cast. Now a precinct-level
2020-to-2024 comparison is a comparison of the same ground.

# Step 8 — Check what survived

A crosswalk that loses votes is worse than no crosswalk, because it looks fine.

```{r check}
chk <- read.csv(file.path(D, "crosswalk_check.csv"), stringsAsFactors = FALSE)
chk
```

**`r chk$value[chk$check == "votes lost"]` votes out of
`r chk$value[chk$check == "votes before"]` —
`r chk$value[chk$check == "share lost (%)"]`%.**

That number was not always this good. An earlier build joined the votes to the
shapefile **by precinct name** and lost 489,876 votes — 9.8% of Georgia,
concentrated in Chatham and DeKalb, where the two sources spell precincts
differently. Two rounds of increasingly clever name-matching moved the loss from
11.1% to 9.8% and then not at all.

**The join was never necessary.** The precinct shapefile already carries the
vote totals in its own attribute table, attached to the geometry. Nothing to
match, nothing to lose.

> **When a join is losing rows, ask whether you needed the join.** That is the
> transferable lesson of this lab, and it cost the person who built it two
> wrong guesses to learn.

# Step 9 — Two state sources, one answer

Was the shapefile's copy of the votes trustworthy? It is published by the
reapportionment office; the returns come from the Secretary of State.

```{r twosource}
ts <- read.csv("fig_twosource.csv", stringsAsFactors = FALSE)
ts

round(100 * ts$matched / ts$shapefile_precincts, 1)   # what share the join reached
```

**On every precinct where the two could be compared, they agreed exactly** —
`r format(ts$identical[ts$scope == "Georgia"], big.mark = ",")` statewide, and
all `r ts$identical[ts$scope == HC]` of Houston's, which is every precinct the
county had. Two offices, two publications, the same numbers. That is what
licensed using the shapefile's copy and abandoning the name join.

Look at what the join *reached*, though:
`r round(100 * ts$matched[ts$scope == "Georgia"] / ts$shapefile_precincts[ts$scope == "Georgia"], 1)`%
of Georgia's precincts, against
`r round(100 * ts$matched[ts$scope == HC] / ts$shapefile_precincts[ts$scope == HC], 1)`%
of Houston's. **The data was never wrong. The labels were.**

# Step 10 — What you can now say

**You can say** that precinct boundaries are not stable, that
`r round(100 - 100 * length(intersect(h20, h24)) / length(h20), 1)`% of
Houston's precinct names and
`r round(100 - 100 * length(intersect(p20, p24)) / length(p20), 1)`% of
Georgia's did not survive four years, and that comparing precincts across years
requires putting both on one map rather than matching labels.

**You can say that a changed name and a changed boundary are different events.**
In Houston you can count them separately: `r length(setdiff(h20, h24))` renames
and `r length(spl)` real split.
Statewide nobody separates them, and the name-survival statistic quietly
conflates the two.

**You can say how to do it**: pick the newest boundaries, disaggregate to census
blocks by interior point, reaggregate weighted by population — and check that
the votes survive.

**You can say that the weighting choice matters**, and by how much: statewide it
changes `r round(100 * mean(abs(cp$diff) > 0.10), 1)`% of the weights by more
than ten points, and in Houston it moves
`r format(gapv, big.mark = ",")` ballots out of one precinct into another.

**You cannot say** that population weighting is right. Population is not voters
— it counts children and non-citizens, and turnout varies. Registered voters by
block would be better, and nobody publishes it.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Use the areal crosswalk instead.** Swap `crosswalk_pop.csv` for
`crosswalk.csv` in Step 7 and recompute Houston's `r length(h24all)` precincts.
Which
ones move, and can you tell from the map in Step 6 why?

**Option B — Find the worst disagreement in Georgia.** This one is
**statewide, not Houston**: the county's worst gap is only the ROZR split, and
the interesting case is elsewhere. `crosswalk_compare.csv` is sorted by how far
the two methods diverge. Take the top row *after* dropping the `"NA|NA"`
pseudo-precinct, look up where it is, and work out what is on that land.

**Option C — Turnout, honestly.** Using `precincts_2024_pop.csv` and the 2024
returns, compute the change in turnout for Houston's precincts. Then do it the
naive way, by name-matching, and report how different the two answers are — and
which `r length(setdiff(h20, h24))` precincts the naive method silently drops.

**Option D — Your own question.** Two maps, and a block-level bridge between
them.

# What this lab cannot tell you

- **Whether any boundary change was intended to matter.** Counties redraw
  precincts for polling-place logistics far more often than for anything else,
  and three of Houston's four changes were exactly that.
- **Where voters actually live within a block.** The block is the floor here.
- **Anything about 2022.** The state publishes 2012, 2014, 2016, 2018, 2020,
  2022 and 2024; this lab uses two of them.
- **Whether population weights are right for turnout.** See Step 10.

# On the discussion board

By **Friday, 11:59 p.m.**

- Step 6 shows area and population disagreeing about whether a Houston precinct
  is "split". Which question is each answering?
- Houston lost three precinct names in four years and only one precinct's ground
  actually changed. What does that do to a claim built on name survival?
- Step 8 describes a join that lost 9.8% of the votes and looked fine. How would
  you have caught it?
- If a county redraws precincts every cycle, what does that do to the idea of a
  "bellwether precinct"?

***

### Where the data came from

- **Precinct boundaries**, 2020 and 2024 — Georgia Legislative and Congressional
  Reapportionment Office. <https://www.legis.ga.gov/joint-office/reapportionment>
  Seven vintages, 2012–2024, all downloadable by script.
- **Census blocks with population** — Census TIGER/Line 2020, Georgia,
  `tl_2020_13_tabblock20`.
- **Precinct returns** — Georgia Secretary of State, November 2020.
- **Houston County's own results**, precinct by precinct, are published at
  <https://www.houstoncountyga.gov/residents/election-results.cms>. Nothing here
  uses it. It is listed because it is a genuinely independent third copy of the
  same totals, so any Houston number in this lab can be checked against it by
  hand.
- **Assignment method** — Cervas, *R-Functions*, `assignPolys`.
  <https://github.com/jcervas/R-Functions/tree/main/assignPolys>

Builds are in `../ga-precinct-returns/data/`: `build-crosswalk.R` (areal, plus the
interior-point assignment) and `build-block-crosswalk.R` (population, via
blocks). The Houston map is built by `build-houston-figures.R` in this folder.
