---
title: "Section 203 Language Coverage Determinations"
type: brief
subtitle: "Section 203 of the Voting Rights Act, and the survey that decides who gets a ballot they can read"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("section-203-code.R")
```

```{r setup, include=FALSE}
```

A ballot a citizen cannot read is a barrier of the same kind as a literacy test. Congress decided exactly that in 1975, when it added Section 203 to the Voting Rights Act and made the translated ballot a right rather than a courtesy. Where enough voting-age citizens of a language minority speak English less than very well, the county **must** provide election materials in their language, with a federal court behind the obligation. The right switches on at a number, and the number comes from a survey.

In December 2021 the Census Bureau published that number for Solano County, California. It said that `r n(sol$VACLEP)` of the county's voting-age citizens were of Hispanic origin and spoke English less than very well. In the next column of the same row, the Bureau published the margin of error, which says how far off that number could be: plus or minus `r n(sol$MVACLEP)`.

The statute names a threshold of ten thousand. Solano County came in `r n(10000 - sol$VACLEP)` people short, so Solano County owes its Spanish-speaking citizens nothing. A few rows away sits `r sub(" County", "", phl$NAMELSAD)` County, whose estimate for its Chinese-speaking citizens is `r n(phl$VACLEP)`, over the line by `r n(phl$VACLEP - 10000)`, with a margin of plus or minus `r n(phl$MVACLEP)`. Philadelphia must translate its ballots, instructions, notices and polling-place assistance into Chinese for the next five years.

The law draws a bright line. Can the instrument that feeds it draw one too? This brief opens the file the determinations were made from and reads the one column the statute ignores.

## The test

The test uses the decision itself: the Census Bureau's *2021 Section 203 Determinations* public use file, downloadable from the address in the sources below. The Bureau published it on 8 December 2021, beside the notice in the *Federal Register*. Nobody reconstructed this arithmetic. The Director of the Census is required to run the test, and this file is the run, with the flags the Bureau set and the estimates it set them from.

Every estimate in it is an American Community Survey figure pooled over 2015 to 2019. The [American Community Survey chapter](../census-acs/census-acs-brief.html) is that survey's biography: roughly 3.5 million addresses a year, weighted up to the country, a margin of error beside every estimate. Asking whether a legal line falls inside its own measurement error takes a file that publishes both, and almost no administrative file does. This one does, in adjacent columns, in every row, which is why this source, of all sources, can answer the question.

## The data

The statute's rule comes first. A jurisdiction is covered for a language group if **either**

- more than **10,000** of its voting-age citizens are in that group and have
  limited English, **or**
- more than **5%** of its voting-age citizens are,

**and** that group's illiteracy rate exceeds the national rate, itself a survey estimate recomputed each cycle. The legal facts here are the statute's, at 52 U.S.C. §10503; the arithmetic about them is measured here and can be re-measured by anyone.

Before any threshold applies, the statute decides who can be counted at all. A language minority group means, at 52 U.S.C. §10310(c)(3), persons who are American Indian, Asian American, Alaskan Native or of Spanish heritage: four groups, listed in 1975, unchanged since. The file reports `r N_LANG` language categories and every one sits inside the four. A county with fifty thousand limited-English Arabic speakers is not covered and cannot become covered, because only Congress can revise the list.

Two extracts of the Bureau's archive are read here: `sect203_counties.csv`, every county-language pair with any limited-English population (`r n(nrow(cty))` rows), and `sect203_determined.csv`, every jurisdiction the Bureau issued a determination about.

```{r determinations}
```

That is `r n_juris` jurisdictions carrying `r nrow(cv)` determinations, some places being covered for more than one language; `r n_spanish` determinations are for Spanish. The counts match the Bureau's announcement of `r lev[["County"]] + lev[["MCD"]]` counties and minor civil divisions plus `r lev[["State"]]` states.

One row is one county and one language group. Here is the largest such row in the country:

```{r one-row}
```

Six columns carry the whole determination, and the last is the only one with legal force: a `1` under *covered* is an obligation on a county government. `VACLEP` is the estimate the statute turns on. **`MVACLEP` is the margin of error the Bureau publishes next to it**, computed by the same office, and it is the column nothing in the law consults.

The cleaning was modest, and one decision is worth naming. The archive expands to roughly 120 MB across nine levels of geography; the build keeps the county level, drops county-language pairs with no limited-English population, and reads blank coverage flags as zeros. That last step is almost certainly what the Bureau means, and it is still a decision made on the reader's behalf. Read the blanks as missing instead and every count of covered places below changes, because a convention did.

> **Before you read on, commit to an answer.** You have both counties'
> numbers: `r n(sol$VACLEP)` give or take `r n(sol$MVACLEP)` in Solano,
> `r n(phl$VACLEP)` give or take `r n(phl$MVACLEP)` in Philadelphia. Turn
> each into a range and write down two things: do the ranges miss each
> other, touch at the edges, or mostly overlap; and how many contain ten
> thousand — neither, one, or both?

## What it says about democracy

Start with what the file establishes: American language access reaches far past the Spanish-language ballot most people picture.

```{r langs-static, eval=!knitr::is_html_output(), fig.height=5.0}
```

```{r langs-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Every language group that triggers coverage anywhere in the United States.** One dot per group on a logarithmic scale, which steps by multiples rather than equal amounts, because a linear axis would flatten everyone behind Spanish to nothing. Spanish accounts for `r n_spanish` of the `r nrow(cv)` determinations, and `r sum(lang$determinations == 1)` of the `r nrow(lang)` groups are covered in a single jurisdiction each. `r n_native` groups are covered somewhere under the American Indian or Alaska Native provisions rather than the numeric tests.

Aleut. Hmong. Seminole. Shoshone. **Somewhere in America a county must print its ballots in Shoshone**, because enough Shoshone speakers live there and Congress decided this was a right and not a gesture.

The arithmetic behind it is honest. The file carries a flag for each statutory test, and the flag can be checked against the estimate it summarizes:

```{r rule}
```

No discretion, no rounding, no thumb near the scale: above ten thousand the flag is on, below it off, in all `r n(nrow(cty))` rows. The difficulty that follows is not the Bureau's. It sits between a rule that demands an exact answer and a survey that produces a range.

Now the prediction. Solano's interval runs from `r n(sol$lo)` to `r n(sol$hi)`. Philadelphia's runs from `r n(phl$lo)` to `r n(phl$hi)`. Most readers write down that the ranges clear each other, or at worst brush at the edges. Instead they overlap along almost their whole length, and both contain ten thousand. On the evidence the federal government itself published, these two counties cannot be told apart.

They are not alone. Here is every county whose published interval contains the statutory line:

```{r moe}
```

```{r straddle-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r straddle-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Every county whose published interval contains the statutory line.** The dot is the Bureau's estimate, the bar is the margin it published alongside, and the dashed line is the number in the statute. Every bar crosses it. The color is the legal consequence, decided by where the dot happens to fall: `r sum(s10$FLAG_COV == 1)` of these `r nrow(s10)` counties are covered and `r sum(s10$FLAG_COV == 0)` are not.

Philadelphia's margin is `r pc(phl$MVACLEP / (phl$VACLEP - 10000))` times the distance by which it clears the line. Solano's is `r pc(sol$MVACLEP / (10000 - sol$VACLEP))` times the distance by which it misses. The law sorts these rows cleanly into two kinds; the survey declines to sort them at all. **The law has no category for "we cannot tell."**

The percentage test is quietly worse.

```{r pct}
```

**`r nrow(p5)` counties cannot be placed on the 5% line by their own estimate**, about `r pc(nrow(p5) / nrow(s10))` times as many as on the count rule. A percentage carries an estimate on top, in its numerator, the top number of the fraction, and another underneath, in its denominator, the number being divided by. Both come from a sample of the county, and the smaller the county, the wider the margin on both.

```{r funnel-static, eval=!knitr::is_html_output(), fig.height=4.8, dev="png", dpi=200}
```

```{r funnel-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. The smaller the jurisdiction, the wider the margin of error as a share of the estimate.** One dot per county-language row, both axes logarithmic. Among the smallest jurisdictions here, near `r n(round(fmed$mid[1]))` voting-age citizens, the published margin runs to a median `r pc(fmed$med[1])`% of the estimate itself; near `r n(round(fmed$mid[nrow(fmed)]))` citizens it falls to `r pc(fmed$med[nrow(fmed)])`%. The highlighted rows straddle the 5% line, on the wide side of the funnel. The rule is least able to say what it means exactly where a translated ballot would be noticed most.

The statute's third condition, the illiteracy test, turns out to be no condition at all. `r n(n_edu_fail)` of the file's `r n(nrow(cty))` rows fail it, yet not one row clearing either numeric bar is stopped by it. A group that large essentially always has an illiteracy rate above the national one.

Congress has already conceded the deeper point once. The most interesting rows in the file are covered while failing every numeric test:

```{r odd}
```

There are `r nrow(odd)` such rows, and the smallest estimate among them is `r min(odd$VACLEP)`. For American Indian areas and Alaska Native regions the statute works on the reservation or the Native village, not the county containing it. Where the population is very small, it stops trying to measure and simply covers. That is not a sentimental carve-out; it is a correction for a known defect in the instrument. The defect it corrects is the funnel in Figure 3, and the funnel is not confined to Alaska.

Last, the threshold itself. If ten thousand captured some real feature of American language communities, the distribution of estimates should show it. It does not.

```{r hist-static, eval=!knitr::is_html_output(), fig.height=3.9}
```

```{r hist-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 4. There is no cliff at ten thousand.** County-language rows near the threshold, in bins a thousand wide. The distribution decays smoothly through the statutory line: the bin just below it holds `r hbar$n[hbar$lo == 9000]` rows and the bin just above holds `r hbar$n[hbar$lo == 10000]`. The color changes where the statute says so, and the data does not.

Lower the line to nine thousand and `r thtab$added[thtab$threshold==9000]` county-language pairs join the covered side; raise it to twelve and a half thousand and `r thtab$dropped[thtab$threshold==12500]` drop off. A round number chosen in 1975 decides coverage for dozens of jurisdictions, and the distribution offers no reason to prefer it to its neighbors.

The strongest defense of the rule is that any alternative is worse. A discretionary standard invites litigation and hands the decision to the least accountable actor; a bright line administers itself, which Tucker (2009) argues is why the provision has survived reauthorization. The Act's other formula, Section 4(b), died in *Shelby County v. Holder* (2013) for resting on 1960s data. **Section 203 is the formula that is updated**, every five years, and the cost of that virtue is the instability documented above. A county in the straddle band can change legal status between cycles without anybody moving, and the next determination is due in 2026.

## What this brief cannot tell you

**Whether any county was misclassified, or whether coverage works.** A margin of error is not an error, and every estimate here may be very close to the truth; the file cannot say, and the statute proceeds as though it could. No column measures registration, turnout, or whether a translated ballot ever changed a vote. A `1` in the covered column is an obligation on a county government, not a ballot in anybody's hand.

**The geography the hardest determinations were made on.** The extract is the county level. `r n(nrow(det) - sum(det$SUMLVL == 50))` of the `r n(nrow(det))` determined rows sit at other levels: minor civil divisions, whole states, and the Native areas where the survey struggles most.

**Whether the straddle counties actually flip.** This is a single determination cycle. Whether a county changes sides at the next one without anyone moving needs earlier cycles laid beside this one, in a matched form the public files do not offer.

**What was actually measured.** Nobody has ever tested whether an American can read a ballot. The estimate counts people whose household said they speak English less than very well, one member answering for all. And the determinations are made for "Hispanic" people rather than Spanish speakers, because the statute defines its groups by heritage. The two are close enough that the mismatch rarely bites, and far enough apart that a careful reader should know it is there.

## What you should have learned

- A civil right switches on at a survey estimate: `r n_juris` jurisdictions
  carry `r nrow(cv)` language coverage determinations, computed by the Census
  Bureau with no hearing and no appeal.
- The Bureau's arithmetic is faithful to the statute in all `r n(nrow(cty))`
  county-language rows. The difficulty is the statute's: it reads an estimate
  as though it were a count.
- `r nrow(s10)` counties straddle the 10,000-person line and `r nrow(p5)`
  straddle the 5% line, so `r nrow(s10) + nrow(p5)` legal determinations turn
  on where inside its own margin an estimate happened to land.
- The uncertainty is largest exactly where the statute aims: among the
  smallest jurisdictions in the file the median margin runs to
  `r pc(fmed$med[1])`% of the estimate itself.
- Congress has already corrected for the instrument once, covering
  `r nrow(odd)` small Native communities without measuring them, and the
  defect that correction admits is not confined to them.

## Extensions

- `sect203_counties.csv` has `LEPPCT` and `MLEPPCT` beside the coverage
  flags. Sort by how close `LEPPCT` sits to 5. What separates two counties on
  opposite sides of the line?
- In `sect203_counties.csv`, compute `MVACLEP` as a share of `VACLEP` and
  sort by it. Would you switch a right on or off at the top of that list?
- `sect203_determined.csv` holds every covered jurisdiction. Group by
  `LANGUAGE`: which languages carry the most obligations, and which appear
  only once or twice?
- Look up your own county in `sect203_counties.csv` by `NAMELSAD`. Which
  language rows does it have, and what would have to change for a flag to
  change?
- **Stretch.** The Bureau published determinations in 2016 as well. Download
  that cycle's file and find the counties that changed status between the two
  cycles. How many sat in this brief's straddle band?
- **Stretch.** Pick one covered county and find its translated election
  materials online: what did the obligation actually produce?

## Sources

**Data**

- U.S. Census Bureau, Redistricting and Voting Rights Data Office, *2021
  Section 203 Determinations — Public Use File*.
  <https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/voting-rights-determination-file.html>
  Published 8 December 2021, from American Community Survey estimates for
  2015–2019 and 2020 Census population counts.

**Law and literature**

- Voting Rights Act §203, 52 U.S.C. §10503.
  <https://www.law.cornell.edu/uscode/text/52/10503>
- Voting Rights Act §14(c)(3), 52 U.S.C. §10310(c)(3), the four-group
  definition of a language minority group.
  <https://www.law.cornell.edu/uscode/text/52/10310>
- "Voting Rights Act Amendments of 2006, Determinations Under Section 203,"
  *Federal Register* 86: 69611 (8 December 2021).
- *Shelby County v. Holder*, 570 U.S. 529 (2013).
  <https://supreme.justia.com/cases/federal/us/570/529/>
- Tucker, J. T. (2009). *The Battle Over Bilingual Ballots: Language
  Minorities and Political Access Under the Voting Rights Act.* Farnham:
  Ashgate.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`sect203_counties.csv`](data/derived/sect203_counties.csv), [`sect203_determined.csv`](data/derived/sect203_determined.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
