---
title: "Public by Law, and You Still Cannot Have It"
type: brief
subtitle: "Fifty-one jurisdictions, one downloadable voter file, and the only instrument in this book that cannot be handed to you"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("voter-file-access-code.R")
```

```{r setup, include=FALSE}
```

This book has now gone and got three kinds of data, and each was hard in a different way.

The [census](../../01-census-bureau/census-access/census-access-brief.html) gives you too much: three open routes, and more than a dozen published figures for a single county's population, no two of them answering quite the same question. [Election returns](../county-returns/county-returns-brief.html) have no door at all — no federal agency counts votes, so you assemble fifty-one publications yourself.

The voter file is the third kind, and it is the one the other two do not prepare you for. **The list is public by law in every state.** There is no argument about that. And you still, mostly, cannot have it.

## Where the addresses come from

The National Conference of State Legislatures keeps a fifty-state survey of who may obtain a voter list, what it contains, what is confidential and what it may be used for. For each jurisdiction it also carries a link labelled *"&lt;State&gt; Access/Purchase List"*, pointing at that state's own page.

Those **`r FV("n_states")` links are the addresses used here.** On **`r FV("scan_date")`** all `r FV("n_states")` of those pages were opened, one after another. What happened at each was written down: whether it answered at all, what it turned out to be, how big it was, and whether it offered the file for download.

The division of labour matters and holds throughout. **The legal facts in this chapter are NCSL's**, and are cited to NCSL. **The retrievability facts are measured here**, and can be re-measured by anyone.

> **Before you read on, commit to a number.** Every state's registration list is
> a public record. Fifty-one jurisdictions each publish a page telling you how
> to get theirs. **On how many of those pages can you actually download the
> file?** Write the number down.

## The funnel

```{r funnelfig, fig.cap="What happened at fifty-one addresses on one day."}
```

```{r funnel}
```

**`r FV("n_ok")` of `r FV("n_states")` answered.** `r FV("n_blocked")` refused a script outright with HTTP 403 — the page exists and a person with a browser can read it, but an automated request is turned away. And `r FV("n_rot")` did not resolve at all:

```{r rot}
```

That is **`r pc(FV("rot_pct"), 1)`% of the survey's own links** dead or unreachable on the day they were tried. The survey is careful work; the pages underneath it move.

## Asking is the normal case

Of the `r FV("n_ok")` addresses that answered, here is what they are:

```{r shape}
```

**`r FV("n_app")` of `r FV("n_ok")` present an application or a request form**, `r FV("n_fee")` mention a fee or a purchase, `r FV("n_login")` want you to log in, and `r FV("n_pdf")` are a PDF to print and post. This is the ordinary condition of the American voter file: not secret, not published — **applied for.**

## Three links, and what is behind them

Only `r FV("n_linked")` of the answering pages link a data file of any kind. A hyperlink ending in `.txt` is a fact about HTML, so all three were opened:

```{r claims}
```

**Two of the three are not the voter file.** Virginia's is a directory of candidates and committees. Maine's has one row per absentee ballot, carrying request type, issue date, return date and status. That is a genuine and interesting file, and it is not a registration list.

So the answer to the question above is **`r FV("n_true_download")`.** One jurisdiction of fifty-one puts its statewide voter list at an address a script can fetch: **North Carolina**, `r n(FV("nc_mb"))` MB, no login, no form, no fee. It is the file the [vote-targeting chapter](../vote-targeting/vote-targeting-brief.html) is built on, and now you know why that chapter is about North Carolina.

## The file nobody asked about

Maine's is worth a paragraph of its own, because it is the opposite failure.

Nothing here was looking for absentee records. The scan asked every state for its *registration list*. Maine's page answered with **`r n(FV("me_bytes"))` bytes** of person-level absentee data: one row per ballot, carrying a voter record number, a party, a municipality, and the date the ballot came back. It is posted openly, requires nothing of the person downloading it, and is not what the page was asked for.

The lesson is not that Maine did something wrong; the state publishes what its law tells it to publish. It is that **"what is available" and "what you went looking for" are different questions**, and a scan that only counted whether a link existed would have scored Maine as a success.

## What it costs, and why nobody can tell you exactly

**The price of a statewide voter file runs from nothing to tens of thousands of dollars, and the two published tables of it disagree.** `r FV("n_fee")` of the pages above mention a fee, and nothing measured here says how much. Two compilations do name them, for all fifty-one jurisdictions: the U.S. Election Assistance Commission, and Ballotpedia as of July 2026. Read side by side they are a lesson in what a price is.

```{r price-spread}
```

**They agree on `r FV("n_same")` jurisdictions and differ on `r FV("n_differ")` of the `r FV("n_comparable")` where both print one plain number.** Some of those differences are a few dollars. Others are not.

```{r price-gaps}
```

`r FV("eac_max_state")` is the extreme case, and it matters more than the others because it is the number everybody quotes. The federal document opens by saying the price of a statewide voter file ranges from `$0` to `$`r n(FV("eac_max"))``, and that upper figure is `r FV("eac_max_state")`'s. The 2026 compilation prices the same list at `$`r n(PCMP$bp_price_usd[PCMP$state == FV("eac_max_state")])``.

**The federal table prints no date anywhere on it.** Eleven pages, a stable government address, no year. The file itself records one: it was created on `r format(as.Date(FV("eac_made")), "%d %B %Y")`, which the chapter can say only because the date is in the document's own properties rather than on any page a reader sees. A price list with no date on it does not look stale, and this one is six years old.

**And a price is often not a price.** For `r FV("bp_not_a_number")` jurisdictions the 2026 table cannot give a single amount, because the state does not charge one.

```{r price-prose}
```

Each of those is a rule rather than a figure, and the federal table has one number for every one of them. Arizona and Texas charge by the record. Michigan charges for staff time, Tennessee prices paper and digital differently, and two states sell a subscription rather than a file. Flatten any of that into a dollar amount and you have made the amount up.

**One row of these tables this chapter can settle itself.** The EAC prices North Carolina at nothing and the 2026 table at `$`r n(price$bp_price_usd[price$state == "North Carolina"])``. The scan above went and got the file: `r n(FV("nc_mb"))` MB, no login, no form, nothing paid. On that one jurisdiction the measurement beats both compilations, which is the whole argument for making one.

## What this book had to do

This chapter cannot end with instructions, because this book did not follow any.

Four chapters in this book use a voter file: [voter-files](../voter-files/voter-files-brief.html), [false-matches](../false-matches/false-matches-brief.html), [bisg-check](../../06-putting-data-together/bisg-check/bisg-check-brief.html) and [rpv](../../06-putting-data-together/rpv/rpv-brief.html). It is a Georgia county extract that arrived as working data in a redistricting case. It was not downloaded. It could not have been: Georgia is one of the `r FV("n_ok")` that answers with a form.

Two consequences, and both are stated here rather than hidden:

**Those four chapters do not rebuild anywhere but on the machine that made them.** Their scripts read an absolute path into a private working directory. Every other chapter in this book fetches its own data or reads a committed capture; these four cannot, because the file is a list of named people and is not ours to redistribute. The ANES chapters have the same shape for a gentler reason. That survey asks people to take the file from ANES, and those chapters say so in the same way.

**This is the only instrument in the book that cannot be handed to a reader.** Everything else here is public, free and re-fetchable: the census files, the returns, the FEC filings, the roll calls. The voter file is public too, in the strict legal sense that made it obtainable in litigation. It is the one where that word does not mean you can have it.

## What this chapter cannot tell you

**It does not price anything itself.** The two tables above are other people's counting, reproduced and compared. Nothing here was paid, and no state was asked what it charges. The amounts sit in statutes, fee schedules and request forms. Two careful compilations went to those sources and came back with different answers for `r FV("n_differ")` jurisdictions, which is the honest reading of the section above.

**It is not a legal summary.** Who may request a list, what a state redacts, and what uses are permitted are NCSL's subject and are cited to NCSL. Restating a statute is exactly the kind of claim this book asks you not to take on trust, and it is not taken on trust here.

**It is one day.** The scan is dated `r FV("scan_date")`, and `r pc(FV("rot_pct"), 1)`% of the links were already broken when it ran, which tells you how fast this surface moves. The results are frozen for that reason: a chapter whose numbers change every time somebody knits it is reporting the weather, not a measurement. Re-running the scan and reading what changed would itself be a finding.

**And a 403 is not a refusal of access.** Ten states turned away an automated request. A person with a browser may well be able to read those pages and follow the process on them. What is measured here is what a *script* meets, which is the honest scope of a scan and not a claim about what a citizen can do.

## Sources

- National Conference of State Legislatures. *Access To and Use Of Voter
  Registration Lists.* The fifty-state survey, its five narrative columns, and
  the per-jurisdiction "Access/Purchase List" links used as the scan's
  addresses.
  <https://www.ncsl.org/elections-and-campaigns/access-to-and-use-of-voter-registration-lists>
- The `r FV("n_states")` state and District election-agency pages that survey
  links to, each requested on `r FV("scan_date")`. Every address, status,
  content type and byte count is recorded per jurisdiction alongside this
  chapter.
- U.S. Election Assistance Commission. *Availability of State Voter File and
  Confidential Information.* Eleven pages, captured 29 August 2026. The price
  and availability table used above is its page 2. **The document prints no
  date**; the `r format(as.Date(FV("eac_made")), "%B %Y")` given for it above
  is the creation date recorded in the file's own properties. Its own source
  line reads "National Conference of State Legislatures, United States
  Elections Project, State Election Officials", so it is a compilation of
  compilations rather than a federal collection.
  <https://www.eac.gov/sites/default/files/voters/Available_Voter_File_Information.pdf>
- Ballotpedia. *Availability of state voter files*, the version dated July
  2026, captured 29 August 2026. Fifty-one jurisdictions with a price column
  that is prose rather than a number, which is why it is kept as prose here.
  <https://ballotpedia.org/Availability_of_state_voter_files>
- North Carolina State Board of Elections. *Statewide voter registration file*,
  `ncvoter_Statewide.zip`, `r n(FV("nc_mb"))` MB.
  <https://s3.amazonaws.com/dl.ncsbe.gov/data/>
- Maine Department of the Secretary of State. Absentee-ballot file, posted as a
  plain-text export, `r n(FV("me_bytes"))` bytes.
- Help America Vote Act of 2002, § 303, which required every state to build a
  single computerised statewide list and said nothing about what it should
  contain — the condition the
  [voter-files-source chapter](../voter-files-source/voter-files-source-brief.html)
  takes as its subject.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`claimed_downloads.csv`](data/derived/claimed_downloads.csv), [`funnel.csv`](data/derived/funnel.csv), [`link_rot.csv`](data/derived/link_rot.csv), [`page_shape.csv`](data/derived/page_shape.csv), [`price.csv`](data/derived/price.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `funnel.csv` narrows fifty-one jurisdictions down stage by stage. Read it top to bottom. At which stage does the public's access actually disappear?
- `page_shape.csv` counts how many official pages present a form rather than a file. What is the difference, for a citizen, between public and available?
- `link_rot.csv` records what happened at each address. Count the dead ones. What does that say about how these files are maintained?
- `claimed_downloads.csv` lists the links that really do hand you a file, with `bytes`. Only a few states do this. Work out what someone studying all fifty would have to do instead.
- `price.csv` puts two published price tables side by side. Sort by the gap between `eac_price_usd` and `bp_price_usd`. Which jurisdictions would you have to go and check yourself before quoting either number?
- `price.csv` marks the jurisdictions whose price is not a single amount in `bp_price_is_one_number`, and `bp_price` says what they charge instead. Read those rows. What would a researcher have to decide before putting any one of them in a column of dollars?

```{r ai-prompt, echo=FALSE, results="asis"}
```
