---
title: "Who Can Actually Get a State Voter File"
type: brief
subtitle: "Public by law, and you still cannot have it: fifty-one official pages, one downloadable statewide list"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("voter-file-access-code.R")
```

```{r setup, include=FALSE}
```

Every state's voter registration list — the **voter file**, the list of everyone registered to vote, kept so that each person can be handed the right ballot — is a public record. That is the law in all fifty states and the District of Columbia, and it is repeated so often that it sounds like an instruction: go and download it. This brief asks the question the sentence skips. Each jurisdiction publishes an official page telling you how to get its list. If you went to all fifty-one of those pages on one day, at how many could you actually get the file?

## The test

The addresses come from the National Conference of State Legislatures (NCSL), the bipartisan organization that serves the fifty state legislatures and tracks their laws. It keeps a fifty-state survey of who may obtain a voter list, what it contains, and what it may be used for, at <https://www.ncsl.org/elections-and-campaigns/access-to-and-use-of-voter-registration-lists>. For each jurisdiction the survey carries a link labelled *"&lt;State&gt; Access/Purchase List"*, pointing at that state's own election-agency page. Those **`r FV("n_states")` links are the addresses tested here.**

On **`r FV("scan_date")`** every one of them was requested by a script — an automated program standing in for a person at a browser — one after another, and what happened at each was written down. Did it answer at all? What did the page turn out to be? Did it offer the file for download?

This is the right test because availability is the one question about voter files that reading the law cannot answer. What the file itself is, and the ballot-handling job its columns do, is the subject of the [voter-file chapter](../voter-files-source/voter-files-source-brief.html); this brief measures only the door. One division of labour holds throughout. **The legal facts in this brief are NCSL's**, and are cited to NCSL. **The retrievability facts are measured here**, and can be re-measured by anyone.

## The data

The scan produced five small tables, and every number below is computed from them. `funnel.csv` narrows the `r FV("n_states")` jurisdictions stage by stage, in `stage` and `states`. `link_rot.csv` names each address that failed, with the `outcome` it produced. `page_shape.csv` counts what the answering pages actually are, in `reading` and `states`. `claimed_downloads.csv` follows every link that claims to be a data file, recording `what_the_link_is`, `is_the_voter_list` and `bytes`. Behind them all sits `scan.csv`, one row per jurisdiction: the HTTP status (the three-digit code a web server answers with — 200 for a page, 403 for a refusal, 404 for nothing there), the size, and whether the page shows a login, an application or a fee.

One row shows how those columns were read. North Carolina's: status `r NC$http`, so the page answered; a web page of `r n(NC$bytes)` bytes rather than a PDF; the words *application* and *fee* appear on it, as they do on most; and `r n(NC$data_links)` of its links end in a data-file extension, `.zip`, `.csv` and the like — a count of markup, which says nothing about what any of those files is. Ohio's row reads `r OH$http`, a refusal. Something came back, `r n(OH$bytes)` bytes of it, but it is the server's notice turning the script away, not the election office's page, so the flags read from it describe the wrong document and the scan counts Ohio as blocked rather than as answering. Notice that nothing in North Carolina's row says the list is downloadable — a page can mention a fee and still give the file away. Settling that took opening the links, which is the last stage of the funnel below.

One more table was not scanned but compiled. `price.csv` lays the two published price lists for a statewide file side by side: the U.S. Election Assistance Commission's (the EAC, the federal agency HAVA created to help states run elections), in `eac_price_usd`, and the July 2026 table kept by Ballotpedia, a nonprofit encyclopedia of American politics, in `bp_price` and `bp_price_usd`. Ballotpedia's column is kept as the prose it prints, because for `r FV("bp_not_a_number")` jurisdictions the price is a rule rather than a single amount: per record, per staff hour, by subscription. Flattening a rule into a dollar figure would invent the figure.

Two cleaning decisions matter. Nothing was retried or repaired: a page that refused an automated request on the day is recorded as refusing, because that refusal is a finding rather than a nuisance. And the results are frozen at the scan date. A brief whose numbers changed every time it was rebuilt would be reporting the weather, not a measurement.

> **Before you read on, commit to a number.** Every state's registration list is
> a public record. Fifty-one jurisdictions each publish a page telling you how
> to get theirs. **On how many of those pages can you actually download the
> file?** Write the number down.

## What it says about democracy

```{r funnelfig}
```

**Figure 1. What happened at fifty-one addresses on one day.** A funnel fits this question because the answer narrows in stages: every jurisdiction, then the pages that answered, then the pages linking any file, then the file that really is the voter list. The finding is how few survive each cut.

```{r funnel}
```

**`r FV("n_ok")` of `r FV("n_states")` addresses answered.** `r FV("n_blocked")` refused an automated request outright with HTTP 403: the page exists, and a person with a browser can probably read it, but a script is turned away. And `r FV("n_rot")` failed before any page came back — a dead link, or a server that never answered:

```{r rot}
```

That is `r pc(FV("rot_pct"), 1)`% of the survey's own links dead or unreachable on the day they were tried. The survey is careful work; the pages underneath it move.

Of the `r FV("n_ok")` pages that answered, here is what they are:

```{r shape}
```

**Asking is the normal case.** `r FV("n_app")` of the `r FV("n_ok")` answering pages present an application or a request form, `r FV("n_fee")` mention a fee or a purchase, `r FV("n_login")` want a login, and `r FV("n_pdf")` are a PDF to print and post. This is the ordinary condition of the American voter file: not secret, not published — applied for.

Only `r FV("n_linked")` of the answering pages link a data file of any kind, and a link is only a promise about what is behind it, so all `r FV("n_linked")` were opened:

```{r claims}
```

`r FN("n_linked") - FN("n_true_download")` of the `r FV("n_linked")` are not the voter list. Virginia's is a directory of candidates and committees. Maine's carries one row per absentee ballot — a real and interesting file, posted openly, and not what the page was asked for. A scan that only counted whether a link existed would have scored both as successes.

So the answer to the question above is **`r FV("n_true_download")`**. One jurisdiction of fifty-one puts its statewide voter list at an address anyone can download: **North Carolina**, `r n(FV("nc_mb"))` MB, no login, no form, no fee. That gap between *public* and *obtainable* is not a paperwork curiosity. It decides who can check claims about the electorate. Campaigns and vendors, who pay and wait, can; a citizen or a student mostly cannot. Much of what nobody knows about American elections lives in that gap.

What asking costs is its own measurement problem, because the two published price tables disagree:

```{r price-spread}
```

They agree on `r FV("n_same")` jurisdictions and differ on `r FV("n_differ")` of the `r FV("n_comparable")` where both print one plain number. Some differences are a few dollars. Others are not:

```{r price-gaps}
```

`r FV("eac_max_state")` matters most, because its `$`r n(FV("eac_max"))`` is the number everybody quotes as the top of the national range. The 2026 compilation prices the same list at `$`r n(PCMP$bp_price_usd[PCMP$state == FV("eac_max_state")])``. The federal table prints no date anywhere on it; the file's own properties record that it was made in `r format(as.Date(FV("eac_made")), "%B %Y")`, which makes it `r EAC_AGE` years old and not visibly stale. And for `r FV("bp_not_a_number")` jurisdictions a price is not a price at all:

```{r price-prose}
```

One row of these tables the scan can settle itself. The EAC prices North Carolina at nothing and the 2026 table at `$`r n(price$bp_price_usd[price$state == "North Carolina"])``. The scan went and got the file: `r n(FV("nc_mb"))` MB, nothing signed, nothing paid. On that one jurisdiction the measurement beats both compilations, which is the whole argument for making one.

## What this brief cannot tell you

**It prices nothing itself.** The two tables above are other people's counting, reproduced and compared. Nothing here was paid and no state was asked what it charges, which is why the `r FV("n_differ")` disagreements are reported rather than resolved.

**It is not a legal summary.** Who may request a list, what a state redacts, and what uses are permitted are NCSL's subject and are cited to NCSL, not restated on this brief's authority.

**It is one day.** The scan is dated `r FV("scan_date")`, and `r pc(FV("rot_pct"), 1)`% of the links were already broken when it ran, which tells you how fast this surface moves. Re-running it and reading what changed would itself be a finding.

**And a 403 is not a refusal of access.** `r FV("n_blocked")` states turned away an automated request. A person with a browser may well be able to read those pages and follow the process on them. What is measured here is what a script meets — the honest scope of a scan, not a claim about what a determined citizen can do.

## What you should have learned

- A voter file is public by law in all `r FV("n_states")` jurisdictions and downloadable in `r FV("n_true_download")`: North Carolina, `r n(FV("nc_mb"))` MB, no login, no form, no fee.
- Asking is the normal case: `r FV("n_app")` of the `r FV("n_ok")` answering pages present an application or request form, and `r FV("n_fee")` mention a fee.
- `r FV("n_rot")` of the survey's `r FV("n_states")` links — `r pc(FV("rot_pct"), 1)`% — were dead or unreachable on the day of the scan.
- Of `r FV("n_linked")` pages linking any data file, only `r FV("n_true_download")` link the voter list; a link that exists is not evidence of what is behind it.
- The two published price tables disagree on `r FV("n_differ")` of the `r FV("n_comparable")` jurisdictions where both give one plain number, and the federal one carries no date.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `funnel.csv` narrows fifty-one jurisdictions stage by stage in `stage` and `states`. Read it top to bottom. At which stage does the public's access actually disappear?
- `page_shape.csv` counts, in `reading` and `states`, how many official pages present a form rather than a file. What is the difference, for a citizen, between public and available?
- `link_rot.csv` records each failed address in `outcome`. Count the kinds of failure. What does each kind say about how these pages are maintained?
- `claimed_downloads.csv` marks each linked file in `is_the_voter_list`, with its `bytes`. Work out what someone studying all fifty states would have to do instead of downloading.
- `price.csv` puts two price tables side by side. Sort by the gap between `eac_price_usd` and `bp_price_usd`. Which jurisdictions would you check yourself before quoting either number?
- `price.csv` also marks, in `bp_price_is_one_number`, the jurisdictions whose charge is a rule rather than an amount, spelled out in `bp_price`. What would you have to decide before putting those rows in a column of dollars?

Two stretch ideas point past the spreadsheet. Open your own state's page, at the address recorded for it in `scan.csv`, and note what a person with a browser meets that a script did not. That is one row of the measurement, re-run. And download North Carolina's file yourself, then compare its size against the `bytes` recorded here to see whether the list has grown.

## Sources

**Data**

- National Conference of State Legislatures. *Access To and Use Of Voter
  Registration Lists.* The fifty-state survey, its narrative columns, and
  the per-jurisdiction "Access/Purchase List" links used as the scan's
  addresses.
  <https://www.ncsl.org/elections-and-campaigns/access-to-and-use-of-voter-registration-lists>
- The `r FV("n_states")` state and District election-agency pages that survey
  links to, each requested on `r FV("scan_date")`. Every address, status,
  content type and byte count is recorded per jurisdiction in `scan.csv`.
- U.S. Election Assistance Commission. *Availability of State Voter File and
  Confidential Information.* Eleven pages, captured 29 August 2026. **The
  document prints no date**; the `r format(as.Date(FV("eac_made")), "%B %Y")`
  given above is the creation date in the file's own properties. Its source
  line names NCSL and others, so it is a compilation of compilations rather
  than a federal collection.
  <https://www.eac.gov/sites/default/files/voters/Available_Voter_File_Information.pdf>
- Ballotpedia. *Availability of state voter files*, the version dated July
  2026, captured 29 August 2026. Fifty-one jurisdictions with a price column
  that is prose rather than a number, which is why it is kept as prose here.
  <https://ballotpedia.org/Availability_of_state_voter_files>
- North Carolina State Board of Elections. *Statewide voter registration file*,
  `ncvoter_Statewide.zip`, `r n(FV("nc_mb"))` MB.
  <https://s3.amazonaws.com/dl.ncsbe.gov/data/>; the Board's own page
  describing the file is <https://www.ncsbe.gov/results-data/voter-registration-data>.
- Maine Department of the Secretary of State. Absentee-ballot file, posted as a
  plain-text export, `r n(FV("me_bytes"))` bytes.
- Help America Vote Act of 2002, § 303, codified at 52 U.S.C. §21083,
  <https://www.law.cornell.edu/uscode/text/52/21083>, which required every
  state to build a single computerised statewide list and said nothing about
  what it should contain — the condition the
  [voter-files-source chapter](../voter-files-source/voter-files-source-brief.html)
  takes as its subject.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`checks.csv`](data/derived/checks.csv), [`claimed_downloads.csv`](data/derived/claimed_downloads.csv), [`facts.csv`](data/derived/facts.csv), [`funnel.csv`](data/derived/funnel.csv), [`link_rot.csv`](data/derived/link_rot.csv), [`page_shape.csv`](data/derived/page_shape.csv), [`price.csv`](data/derived/price.csv), [`scan.csv`](data/derived/scan.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
