---
title: "Key: The Finest Grain, Counted Twice"
subtitle: "84-355 Democracy's Data | Georgia precinct returns | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)

D   <- "data"
p   <- read.csv(file.path(D, "precincts.csv"), check.names = FALSE, stringsAsFactors = FALSE)
co  <- read.csv(file.path(D, "counties.csv"),  check.names = FALSE, stringsAsFactors = FALSE)
rc  <- read.csv(file.path(D, "recount.csv"),   stringsAsFactors = FALSE)
vm  <- read.csv(file.path(D, "ga2020_vote_methods.csv"), stringsAsFactors = FALSE)
s20 <- read.csv(file.path(D, "ga2020_structure.csv"), stringsAsFactors = FALSE)
s24 <- read.csv(file.path(D, "ga2024_structure.csv"), stringsAsFactors = FALSE)
v24 <- read.csv(file.path(D, "ga2024_vote_methods.csv"), stringsAsFactors = FALSE)
chk <- read.csv(file.path(D, "crosswalk_check.csv"), stringsAsFactors = FALSE)

DEM <- "Joseph R. Biden"; REP <- "Donald J. Trump"; LIB <- "Jo Jorgensen"

n  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)

TOTV   <- sum(p$total)
MARGIN <- sum(p[[DEM]]) - sum(p[[REP]])
np     <- p[p$total > 0, ]

# vote method, 2020
mm     <- tapply(vm$votes, list(vm$candidate, vm$method), sum)
mtot   <- colSums(mm)
mshare <- 100 * mm[DEM, ] / (mm[DEM, ] + mm[REP, ])
ED <- grep("Election Day",     names(mtot), value = TRUE)
MB <- grep("Absentee by Mail", names(mtot), value = TRUE)
AV <- grep("Advanced",         names(mtot), value = TRUE)

# recount
tot      <- aggregate(cbind(original, recount) ~ candidate, rc, sum)
tot$chg  <- tot$recount - tot$original
RCMARGIN <- tot$recount[tot$candidate == DEM] - tot$recount[tot$candidate == REP]
CHANGED  <- length(unique(rc$county[rc$change != 0]))
NCTY     <- length(unique(rc$county))
MOVED    <- sum(abs(rc$change))

# does the precinct file aggregate to the county file, exactly?
ag  <- aggregate(cbind(p[[DEM]], p[[REP]], p$total), list(county = p$county), sum)
names(ag) <- c("county", "d", "r", "t")
mg  <- merge(ag, co, by = "county")
MAXRESID <- max(abs(mg$t - mg$total), abs(mg$d - mg[[DEM]]), abs(mg$r - mg[[REP]]))
EXACT    <- sum(mg$t == mg$total)

# where the biggest precincts actually are
ONEPREC  <- co$county[co$precincts == 1]
METRO    <- c("Fulton", "DeKalb", "Gwinnett", "Cobb", "Clayton")
BIGROW   <- np[which.max(np$total), ]
METROMAX <- max(np$total[np$county %in% METRO])
METROMED <- median(np$total[np$county %in% METRO])
RESTMED  <- median(np$total[!np$county %in% METRO])
EMPTYCO  <- paste(unique(p$county[p$total == 0]), collapse = ", ")

# quintile composition, recomputed here so the Option C prose is not typed
qq    <- cut(np$total, quantile(np$total, 0:5 / 5), include.lowest = TRUE,
             labels = paste0("Q", 1:5))
Q5ONE <- 100 * mean(np$county[qq == "Q5"] %in% ONEPREC)
Q1FUL <- 100 * mean(np$county[qq == "Q1"] == "Fulton")

# 2024 vote-method coverage
v24tot <- tapply(v24$votes, v24$method, sum)
NOTREP <- v24tot[["(not reported)"]]

getv <- function(s, k) s$value[s$item == k]
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`precincts-brief.Rmd`) walks ten numbered steps in prose
with a D3 figure; the lab (`precincts.Rmd`) runs the same material with the code
exposed and ends in four options. **Students should read the brief first.** Step
numbers below refer to the brief.

**This lab is the data spine of the course.** Its parsed Georgia returns and the
census-block shapefile in `data/` are read directly by `precinct-geography`,
`levels-of-aggregation`, `areal-units` and `vote-dilution`, and `rpv` runs on the
same county with the same registration logic. That changes how you should teach
it: **an error here is not contained.** If a student mistrusts something in this
session, they are entitled to mistrust four later ones, and the honest answer is
that they are right to.

Every number below was computed from the built files at knit time.

***

# The session in one sentence

Georgia's precinct returns are the finest-grained public record of American
voting and the primary source for it — and the same state counted the same
ballots a second time, got a different answer in `r CHANGED` of `r NCTY`
counties, and certified the second one, which means the finest data in the
country is also the count that was superseded.

***

# The findings

```{r headline}
c(precincts            = nrow(p),
  counties             = nrow(co),
  presidential_votes   = TOTV,
  margin_precinct_file = MARGIN,
  margin_recount       = RCMARGIN,
  counties_changed     = CHANGED,
  votes_moved          = MOVED,
  pct_election_day     = round(100 * mtot[[ED]] / sum(mtot), 1),
  mail_minus_eday_pts  = round(mshare[[MB]] - mshare[[ED]], 1),
  largest_precinct     = max(p$total),
  empty_precincts      = sum(p$total == 0),
  county_residual      = MAXRESID)
```

- **`r n(nrow(p))` precincts in `r nrow(co)` counties, `r n(TOTV)` presidential
  votes.** The complete state, from the state's own publication.
- **The precinct file says the margin was `r n(MARGIN)`. The recount says
  `r n(RCMARGIN)`.** The second is Georgia's certified number.
- **`r CHANGED` of `r NCTY` counties returned a different total the second time**
  — `r pc(100 * CHANGED / NCTY)`% of them — moving `r n(MOVED)` votes in
  absolute terms, `r pc(100 * MOVED / TOTV, 3)`% of the ballots cast.
- **Only `r pc(100 * mtot[[ED]] / sum(mtot))`% of votes were cast in person on
  election day.** Mail and election-day electorates differ by
  `r pc(mshare[[MB]] - mshare[[ED]])` points of two-party share.
- **Precincts aggregate to counties with a residual of exactly `r MAXRESID`** —
  `r EXACT` of `r nrow(mg)` counties reproduced to the vote.

***

# The three things to get said

**1. Precision and authority are different properties.** This is the most precise
election data the United States produces, and it does not agree with the official
result. Not because anyone erred — because "official" is a decision the state
makes at a coarser resolution than the one it publishes. **Make somebody say the
uncomfortable version out loud**: every racially polarized voting estimate in the
second half of this course is built on the superseded count, because the recount
was never published at precinct resolution. That is not a reason to distrust the
analysis; it is a reason to know what it rests on.

**2. A source that requires you to identify yourself is a source that can
disappear.** This lab used to run on VEST via a Harvard Dataverse guestbook, and
on MEDSL before that. It now reads the Secretary of State directly. **The course
now has no guestbooked sources at all, down from three.** The point is not that
guestbooks are wrong — VEST's work is careful and the form is reasonable. The
point is that a dependency you cannot script is a dependency somebody else
controls, and going upstream past it bought something VEST does not carry: the
vote-method column that Steps 5 and 6 are entirely about.

**3. Step 7 is the session.** Two counts, same ballots, same state, published as
two archives. `r n(MOVED)` votes moved and the outcome held. **Ask the room what
that proves before you tell them.** Half will say it proves the count was
accurate; half will say it proves hand and machine counts differ. Both readings
use the same table, and both are defensible. The honest statement is the boring
one — recounts move small numbers, and the margin was much larger than the
movement.

**The line for the board:** the count is not a fact the data carries; it is a
procedure the state performed, and Georgia performed two of them.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** What one record is | County, precinct, registration, ballots, votes | 5 | **The precinct is named, not coded.** First sign this file came from 159 offices. |
| **2** What the file contains | `structure.csv`, and the vote-method column | 6 | The lab's Part 2 goes further here — same file, three labels for one office. |
| **3** Not a standard size | 1 vote to `r n(max(p$total))` | **9** | The biggest precincts are **rural** — both documents now say so. Take the guess first. |
| **4** The state total | `r n(MARGIN)`-vote margin; exact aggregation | 6 | The clean aggregation is the setup for Steps 7 and 9. Say "hold onto that." |
| **5** The vote-method column | `r pc(100 * mtot[[ED]] / sum(mtot))`% cast on election day | 8 | Students expect election day to dominate. It is fifth of the total. |
| **6** Two electorates | `r pc(mshare[[MB]] - mshare[[ED]])`-point gap between mail and election day | **12** | The "red mirage" is arithmetic here, not allegation. Do that slowly. |
| — Predict, before Step 7 | Counties changed, and by how much | 3 | **Collect the guesses in writing.** The step does not land without them. |
| **7** Counted twice | `r CHANGED` of `r NCTY` counties; margin `r n(MARGIN)` to `r n(RCMARGIN)` | **14** | The pivot. Nobody made a mistake; that is the hard part to convey. |
| **8** Where the movement was | Fulton dominates the change | 6 | Largest movements in the largest counties is what *no* anomaly looks like. |
| **9** The join that fails | Precincts and blocks do not nest | 7 | This is the bridge to `precinct-geography` and `rpv`. Name them. |
| **10** What you can now say | Including what you cannot | 4 | End on the transferable finding, not on the recount. |

That is 80 minutes with the prediction included.

**Step 3 deserves the prediction too, and it is cheaper — ask before you show
the table.** Both student documents now print this evidence themselves (the
`size-truth` chunk in the lab, the top-six table in the brief), so collect the
guess first or the surprise is spent.

```{r size-truth}
big <- head(np[order(-np$total), c("county", "precinct", "total")], 6)
big$one_precinct_county <- big$county %in% ONEPREC
print(big, row.names = FALSE)
c(largest_in_metro_five = METROMAX,
  one_precinct_counties = length(ONEPREC),
  votes_in_them         = sum(co$total[co$precincts == 1]))
```

Ask where the largest precinct in Georgia is. The room will say Atlanta. It is
**`r BIGROW$precinct`, in `r BIGROW$county` County, `r n(BIGROW$total)` votes** —
and `r BIGROW$county` is one of the `r length(ONEPREC)` counties that report the
entire county as a single precinct. **The largest precinct anywhere in the metro
five (Fulton, DeKalb, Gwinnett, Cobb, Clayton) is `r n(METROMAX)`** — a third of
the state record.

Be precise about what that does and does not show, because the seductive version
is wrong. Metro precincts are *not* small: their median is `r n(METROMED)` against
`r n(RESTMED)` in the rest of the state. What metro counties do is **cap** them,
because a county with `r n(max(co$precincts))` precincts is managing polling
places. The extreme tail belongs to counties that never subdivided at all, so the
maximum is a fact about administration and the median is a fact about population.
**Two different things are being called "precinct size".** Option C depends on the
class having heard this.

**If you are short on time, protect Steps 3, 6 and 7.** The unit that is not a
unit, the two electorates, and the count that was superseded are the argument.
Steps 1, 8 and 10 can be read rather than worked.

***

# Where this data comes from, and the two histories worth telling

## The source moved upstream, past a gated one

**Told as a sequence, this is the course's provenance lesson in one lab.** The
lab ran first on MEDSL, then on VEST through a Harvard Dataverse guestbook —
which returns, to a script, `"You may not download this file without the required
Guestbook response for guestbookID 458."` It now reads the Georgia Secretary of
State directly: the XML county archives for 2020, and the 2024 JSON export at
`results.sos.ga.gov/cdn/results/Georgia/export-2024NovGen.json`, parsed across all
`r nrow(co)` counties by `data/parse-ga-sos.py`.

**What forced the move was Cloudflare.** `sos.ga.gov` returns `403` to a script
and `200` to a browser on the identical URL with identical headers — verified,
not assumed, and recorded in `data/README.md`. The irony is worth stating: the
bot check is why the outer ZIP is still a manual download, and it is also why
going upstream was worth doing at all, because it exposed how much of the course
depended on somebody else's willingness to keep serving a file.

**Two things to be precise about in front of students**, because both are easy to
overstate:

- **Only the 2020 route is gated.** The 2024 JSON is on a CDN path that is *not*
  behind the bot check and is script-fetchable (`--json-url` fetches it
  unattended). So the lab's "everything is gated" framing is true of the archive
  and false of the newer platform. Georgia is in the middle of fixing this
  without saying so.
- **The route does not generalise.** It exists because one state publishes
  structured county exports. Forty-nine others do something else, which is why
  VEST existed in the first place and why the volunteer project is not obsolete.

## The parser had to be taught what the race is called

```{r contest-names}
data.frame(
  source        = c("XML archives (2012 - June 2024)", "JSON export (Nov 2024)"),
  contest_label = c(getv(s20, "contest"),   getv(s24, "contest")),
  precincts     = c(getv(s20, "precincts"), getv(s24, "precincts")),
  candidates    = c(getv(s20, "candidates"), getv(s24, "candidates")),
  stringsAsFactors = FALSE)
```

```{r method-labels}
cat(paste0("2020 vote-method labels: ", getv(s20, "vote methods"), "\n",
           "2024 vote-method labels: ", getv(s24, "vote methods"), "\n"))
```

**The same state, the same race, two labels.** The XML calls it *President of the
United States*; the 2024 JSON calls it *President of the US*. The vote-method
labels drift too — `Election Day Votes` becomes `Election Day`. The parser's
default contest pattern is the longest prefix common to both.

**This is the best concrete argument in the course against joining on
human-readable names**, and it is better than the abstract version because the
failure it prevents is documented in the script. An earlier pattern matched the
bare substring `President` and swept in
`Co Commission Chair/Presidente de la Comisión del Condado` — County Commission
Chair, in Spanish, from Georgia's bilingual counties — putting a Gwinnett
commission candidate into the presidential totals with 233,110 votes. **A join
key that is a sentence in a natural language is a join key that changes when
somebody edits a label.**

**One correction to make if you tell this story.** The parser handles a
candidate string carrying more than one parenthetical tag by reading the *last*
one, and strips `(I)` first because in this source `(I)` means **incumbent, not
Independent**. That is a real repair and worth a sentence. But **the scripts in
this folder do not contain fusion-voting handling or footnote repair** — Georgia
does not permit fusion voting, and no footnote markers appear in these exports.
If you have heard those two attached to this lab, they belong to a different
state's parser. Do not repeat them; a key that oversells its own provenance story
teaches the opposite of the lesson.

***

# The 489,876-vote join, and whether this lab would have caught it

**Tell this story here, not only in `precinct-geography`.** The sibling lab's
first crosswalk attached votes to the 2020 shapefile by matching precinct names.
It reached 90.4% and lost **489,876 votes, 9.8% of Georgia**, concentrated in
Chatham and DeKalb. Two rounds of name normalisation moved the loss from 11.1% to
9.8% and then not at all. The shapefile already carried the vote totals in its own
attribute table; there was never a join to make.

*(That figure is history, recorded in the header of `data/build-crosswalk.R`. It
is not recomputable from any current file, and it is the only number in this key
that is typed rather than computed. Flag it as such if a student asks.)*

**Would this lab's checks have caught it? Half yes, and the half matters.**

- **The automated checks would not have.** They are internal to the parse:
  `parse-ga-sos.py` reconciles each candidate's stated total against the sum of
  their precinct-by-method rows, and `build-data.R` asserts `nrow(co) == 159`.
  Both would pass while an entirely separate join to a shapefile threw away a
  tenth of the state. **A check inside one file cannot see a mistake made
  between two files**, and that is the general form of the lesson.
- **The habit the lab teaches would have caught it instantly.** Step 4's whole
  purpose is to sum the parts and compare against an independently published
  total:

```{r step4-check}
c(counties_reproduced_exactly = EXACT, of = nrow(mg), worst_residual = MAXRESID)
```

  A residual of `r MAXRESID` across `r nrow(mg)` counties is what a correct
  aggregation looks like. Nobody ran that comparison on the name join. Had they
  totalled the joined file once and put it next to `r n(TOTV)`, the 489,876 would
  have announced itself in one line.

**So the answer to give the class is:** the check that would have caught it is
the one this lab performs in Step 4 and the sibling lab did not perform until
after the damage — and the reason it went unnoticed is that the joined file
*looked fine*, which is what 90.4% always looks like.

The carry that shipped loses `r chk$value[chk$check == "votes lost"]` votes of
`r chk$value[chk$check == "votes before"]`, a share of
`r chk$value[chk$check == "share lost (%)"]`%.

***

# Before you teach this: one defect, and two repairs

**Written down so you are not surprised in the room.** Two of the three problems
this section used to list are fixed in the student documents; the third is in a
script and is still live.

**Repair one: the lab's Part 2.** It used to filter
`structure.csv` on a column named `kind` that a rebuild no longer produces. It
did not error — it printed `vote_columns 0` and `<0 rows>` — and then taught
VEST's `G20PRERTRU` convention, a file this lab has not read since the move to
the Secretary of State. Part 2 now reads the real `structure.csv`
(`r nrow(s20)` rows, columns `r paste(names(s20), collapse = ", ")`) and teaches
**the contest-name and vote-method drift from the section above**, computed at
knit time from `ga2020_structure.csv`, `ga2020_vote_methods.csv` and
`ga2024_structure.csv`, with the Gwinnett trap printed straight out of the
header of `data/parse-ga-sos.py`. `G20PRERTRU` survives in the lab as an
explicitly labelled historical contrast: two designs, two failure modes.

**So the lab and this key now overlap on purpose.** Students meet the label
drift in Part 2; you have the same material here plus the parenthetical
corrections. **Do not teach it twice** — in the room, go straight to the
question the lab sets up and this key answers: a code cannot drift but cannot
explain itself, a sentence explains itself but drifts, and *both* files are
unusable by somebody who was not told.

**Repair two: the largest-precinct sentence.** Both documents used to follow
`r n(max(p$total))` with "the word means one thing in rural Georgia and something
else in metropolitan Atlanta," which students hear as *metro precincts are big* —
the opposite of what the tail shows. Both now lead with the rural tail and the
metropolitan cap, computed at knit time, and land on the version this key has
always given: **the word means one thing in a county with
`r n(max(co$precincts))` precincts and another in a county with one.** The lab
prints the top six and the metro comparison in its `size-truth` chunk; the brief
prints the same six as a table before its county spread. **The correction is no
longer yours to make in the room** — what is still yours is the prediction, and
Option C still depends on the class having heard the distinction out loud.

**The defect that is still live: the 2024 file is 18 votes short at precinct
level, and the script says it is not.** The header of `data/parse-ga-sos.py` reports that 84 rows have precinct
sums falling short of the candidate total and states "**No statewide contest is
affected.**" That is wrong. Summing the 2024 JSON's presidential precinct rows
gives Trump 2,663,110 and Harris 2,548,006 against county-level totals of
2,663,117 and 2,548,017 — **7 and 11 votes, all in Fulton County.** It changes
nothing substantive (18 votes in 5.25 million) and it matters anyway, because
the claim in the script is exactly the kind of reassurance the lab teaches
students not to accept unverified. **If you use the 2024 file, say the number.**

***

# Option A — The empty precincts

## Verified output

```{r a}
p[p$total == 0, c("county", "precinct", "registered", "ballots_cast",
                  "total", "dem_two_party_pct")]
c(empty_precincts = sum(p$total == 0),
  na_dem_share    = sum(is.na(p$dem_two_party_pct)),
  mean_dem_share_naive = mean(p$dem_two_party_pct),
  mean_dem_share_narm  = round(mean(p$dem_two_party_pct, na.rm = TRUE), 2),
  statewide_two_party  = round(100 * sum(p[[DEM]]) / sum(p[[DEM]] + p[[REP]]), 2))
```

**3/4:** Finds all `r sum(p$total == 0)`, reports that they are in
`r paste(unique(p$county[p$total == 0]), collapse = ", ")` County with almost no
registered voters and no ballots cast, and concludes they are administrative
rather than real.

**4/4:** Notices they carry a registration count but zero ballots, works out that
dropping them cannot change a single vote total, and identifies what they *do*
change — `r EMPTYCO`'s precinct count and the state's, and therefore every
per-precinct average.

**4/4, best:** Uses them to catch a live bug in the file's own design. The naive
`mean(dem_two_party_pct)` is **`r ifelse(is.na(mean(p$dem_two_party_pct)), "NA", "a number")`**
because a zero-vote precinct has an undefined share; the `na.rm = TRUE` version is
`r pc(mean(p$dem_two_party_pct, na.rm = TRUE), 2)`, and the statewide two-party
share is `r pc(100 * sum(p[[DEM]]) / sum(p[[DEM]] + p[[REP]]), 2)`. **Three
different answers to "how Democratic was Georgia", and the gap between the last
two has nothing to do with the empty precincts** — it is the difference between
averaging precincts and averaging voters, which is the same "per what?" problem
Step 3 raised about size. A student who gets from three blank rows to that
distinction has done the best available work on this option.

***

# Option B — Aggregate up

## Verified output

```{r b}
mismatch <- mg[mg$t != mg$total, c("county", "t", "total")]
c(counties           = nrow(mg),
  reproduced_exactly = EXACT,
  mismatched         = nrow(mismatch),
  worst_residual     = MAXRESID,
  precinct_counts_match = sum(as.vector(table(p$county)) ==
                              co$precincts[order(co$county)]))
head(mg[order(-mg$total), c("county", "precincts", "t", "total")], 5)
```

**3/4:** Confirms the aggregation is exact — `r EXACT` of `r nrow(mg)` counties,
residual `r MAXRESID` — and says so.

**4/4:** Answers the harder half. Precincts nest inside counties **because the
county drew them inside its own border**; blocks and precincts do not nest
because two different authorities drew them at two different times for two
different purposes, and neither is required to respect the other. The
distinction is authorship, not geometry.

**4/4, best:** Notices that the exactness is weaker evidence than it looks.
Both files were produced by the same parser from the same source, so the check
verifies the arithmetic and nothing about the underlying count — it is
**internal consistency, not validation.** The real external check in this lab is
Step 7's recount, and it *fails*: the same election, independently recounted,
disagrees. A student who says "Step 4 passing does not license trusting Step 4's
totals, and Step 7 proves it" has understood something the brief only gestures
at.

***

# Option C — Size and politics

## Verified output

```{r c}
q  <- cut(np$total, quantile(np$total, 0:5 / 5), include.lowest = TRUE,
          labels = paste0("Q", 1:5))
wt <- tapply(seq_len(nrow(np)), q,
             function(i) 100 * sum(np[[DEM]][i]) / sum(np[[DEM]][i] + np[[REP]][i]))
data.frame(quintile          = levels(q),
           precincts         = as.vector(table(q)),
           median_size       = as.vector(round(tapply(np$total, q, median))),
           mean_precinct_dem = round(as.vector(tapply(np$dem_two_party_pct, q, mean)), 1),
           vote_weighted_dem = round(as.vector(wt), 1))
c(correlation_size_vs_dem_share =
    round(cor(np$total, np$dem_two_party_pct, use = "complete.obs"), 3))

# who is actually in the two tails
sapply(c("Q1", "Q5"), function(k) names(head(sort(table(np$county[q == k]),
                                                  decreasing = TRUE), 4)))
```

**3/4:** Plots it, reports that the correlation is
`r round(cor(np$total, np$dem_two_party_pct, use = "complete.obs"), 3)` — near
zero — and concludes there is no relationship between precinct size and vote
share.

**4/4:** Notices that "no correlation" is the wrong summary because the pattern
is not linear. The middle quintiles are the most Democratic and both tails are
less so; a single correlation coefficient averages an arch down to nothing.
Says what a scatterplot shows that `cor()` hides.

**4/4, best:** Says what the variable is actually measuring. Sorting precincts by
size mostly sorts **counties**: the smallest quintile is `r pc(Q1FUL)`% Fulton —
the county that subdivides most finely — mixed with rural south Georgia, and the
largest quintile is the metro ring, Gwinnett and Cobb and Cherokee. **Fulton
appears in both tails**, which is the tell: a county with
`r n(max(co$precincts))` precincts contains the whole range by itself, so the
variable is not sorting places, it is sorting *within and across* counties at the
same time. **The arch is a composition effect.** It runs from urban-plus-rural,
through the Democratic inner suburbs, out to the less Democratic outer ones, and
precinct size is the incidental variable that happens to order them that way.
**The apparent finding is a fact about who lives where and how counties draw
polling places, wearing the costume of a fact about precinct size.** Anyone who
reaches that, and connects it to Step 3's warning that the unit is not comparable
across counties, has the best answer available. Reward equally a student who
declines to interpret it at all on those grounds.

**One thing to correct if a student asserts it:** the *largest single precincts*
are rural (Step 3), but the *largest quintile* is suburban. Only `r pc(Q5ONE)`%
of precincts in Q5 sit in a one-precinct county. Both facts are in the data and
they are about different parts of the distribution.

***

# Option D — Your own question

No fully verified output; grade the reasoning.

**3/4:** Asks a question the file can answer and answers it correctly.

**4/4:** States in advance what the data would have to look like for the answer
to be different, then checks. Any question about vote method, roll-off between
ballots cast and presidential votes, or 2020-vs-2024 county movement is well
served by these files.

**4/4, best:** Picks a question the file *cannot* answer and demonstrates why
from the file itself — no boundaries, no demographics, no individual ballots, one
state, one election, and the count that was superseded. **Establishing a limit
empirically is worth more than answering an easy question**, and this is the one
option where a student can do that without leaving the folder.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code — it is given.** Every chunk in the lab runs as
  distributed; a student who reports what came out has done the assigned work.
- **The move that earns a 4** is treating a property of the data as a
  consequence of who produced it and how. Precinct size is county practice; the
  method split is administrative channel; the recount gap is procedure.
- **A 2** is "the data is messy so we cannot trust it." True, useless, and the
  thing the lab exists to move past.
- **Also a 2**: quoting the `r n(MARGIN)` margin as *the* result after Step 7 has
  explained that the certified figure is `r n(RCMARGIN)`.
- **Reward anyone who asks which count the downstream labs use.** The answer is
  the precinct file, because it is the only one published at that resolution, and
  noticing the question is more sophisticated than most of the available answers.

## Anticipated questions

- *"So which count is right?"* — The recount is certified, so legally the
  recount. Neither is "right" in the sense the question means. A recount is a
  second performance of a procedure, not an appeal to a ground truth, and the
  ballots are the only ground truth there is.
- *"Why did `r pc(100 * CHANGED / NCTY, 0)`% of counties change?"* — Because
  hand-counting five million ballots produces different handling errors than
  machine-counting them. `r n(MOVED)` votes across `r n(TOTV)` is
  `r pc(100 * MOVED / TOTV, 3)`%. That rate is the finding; the direction is not.
- *"Isn't `r pc(100 * mtot[[ED]] / sum(mtot))`% on election day suspicious?"* —
  No, and this is the question to prepare for. 2020 was a pandemic election with
  three weeks of in-person early voting in Georgia. **Advanced Voting alone is
  `r pc(100 * mtot[[AV]] / sum(mtot))`% of the total.** The file shows the split
  without any claim attached to it, which is precisely why it is worth having.
- *"Could we do this for another state?"* — Not this way. The route exists
  because Georgia publishes structured county exports. That is the honest answer
  and it is also the argument for why VEST existed.
- *"Why does the state publish 159 files instead of one?"* — Because the counties
  are the authority and the state is compiling, not counting. The same fact
  drives `precinct-geography`'s Step 1.

## Connections

- **`precinct-geography`** — runs on this folder's shapefiles and crosswalks and
  is the direct sequel. Teach this one first; that lab assumes the returns are
  understood.
- **`levels-of-aggregation`** — reads `ga2020_precincts.csv`,
  `ga2020_vote_methods.csv` and the 2024 files directly. The same votes at four
  resolutions.
- **`rpv`** and **`vote-dilution`** — both need returns and demographics in the
  same place, which is Step 9's unsolved join. `vote-dilution` reads the census
  blocks that live in this folder.
- **`areal-units`** — shares the block shapefile and generalises Step 9 into the
  modifiable areal unit problem.
- **`residual-votes`** — the roll-off visible here between ballots cast and
  presidential votes (`r n(sum(p$ballots_cast) - TOTV)` ballots,
  `r pc(100 * (sum(p$ballots_cast) - TOTV) / sum(p$ballots_cast), 2)`%) is that
  lab's whole subject, measured at this lab's resolution.
- **`data-sources`** — the provenance argument in miniature. This is the lab
  where the course stopped depending on a form.
- **`election-night`** — the red mirage in Step 6 is the arithmetic behind that
  session's reporting sequence. They pair well in a single board post.
