---
title: "The Age Structure of the Electorate"
subtitle: "Who is old enough to vote, who says they voted, and sixty years of the gap between the two"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("age-structure-code.R")
```

```{r setup, include=FALSE}
```

"The electorate" is a phrase that does two jobs and hides the difference. It sometimes means everybody entitled to vote, and it sometimes means everybody who did. Those are different populations, and the clearest way to see how different is to draw them as two pyramids on one axis.

In `r LP`, **`r mil(VAPL)`** Americans were old enough to vote and **`r mil(VOTL)`** told the Census Bureau they had — `r p1(TURNL)`%. The people in the second group are not a scale model of the first. They are older, and they have been getting relatively older for sixty years.

## One row is a year and an age band

```{r one-row}
```

```{r varmap}
```

The Bureau publishes the population and the percentage; the count of voters is the two multiplied, which is arithmetic on its own figures rather than an estimate of ours. The four bands are published separately from the total, so they can be checked against it:

```{r checks}
```

> **Before you read on, commit to a view.** People aged 65 and over are
> `r p1(OPOP)`% of American adults. **What share of the votes do you think they
> cast?** Write it down, and write down what you would have guessed for 1968.

## Two pyramids

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The adult population and the voters, `r LP`.** Bands are the four the Census Bureau publishes. Grey to the left is each band's share of everybody old enough to vote; teal to the right is its share of everybody who said they voted. Drag the year; the button switches from shares to counts of people.

The two sides are not the same shape. **People aged 65 and over are `r p1(OPOP)`% of the adult population and `r p1(OVOTE)`% of the voters.** People aged 18 to 24 are `r p1(YPOP)`% of adults and `r p1(YVOTE)`% of voters. Compare that with what you wrote down.

The mechanism is not mysterious — it is turnout. `r p1(OTURN)`% of the oldest band said they voted against `r p1(YTURN)`% of the youngest, a ratio of `r p1(RATIO)` to one. Some of that gap opens before the ballot: `r p1(OREG)`% of the oldest band were registered against `r p1(YREG)`% of the youngest, so roughly half the difference is a registration difference rather than a turnout one.

## The gap is not a constant

Now drag Figure 1 back to the 1960s, or read it here.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Over-representation, presidential elections `r min(PRES)` to `r LP`.** Zero means a band cast exactly its share of the votes. Above the line it is over-represented among voters; below it, under-represented.

The oldest band starts at **`r p1(OGAP1)`** in `r min(PRES)`, meaning it voted almost exactly in proportion to its size. It ends at **+`r p1(OGAP)`**. That is the finding, and it is not the one most people expect. The over- representation of older voters is not an old feature of American elections. It is roughly as old as the people currently complaining about it.

Two things drove it and they are easy to confuse. The band grew: 65-and-over went from `r p1(OPOP1)`% of adults in `r min(PRES)` to `r p1(OPOP)`% now, which is demography and would have raised their share of the votes even if nothing about their behaviour had changed. And on top of that their turnout advantage widened. **The first is a fact about the country. The second is a fact about its politics.** A figure of votes cast alone cannot separate them, which is why this one is drawn as a difference between two shares rather than as a count.

The youngest band, meanwhile, is roughly where it was. It has been between `r p1(min(sh$gap[sh$band == "18 to 24 Years"]))` and `r p1(max(sh$gap[sh$band == "18 to 24 Years"]))` points under-represented in every presidential election in the series. Whatever has changed about young Americans and voting, their standing relative to their own numbers is not it.

> **A second prediction.** Suppose turnout rose by ten points in every age band
> at once. **What would happen to the lines in Figure 2?** Answer before you
> read the next paragraph.

Nothing would happen to them. A uniform rise cancels out in a share. That is what makes Figure 2 readable, and it is also what makes it blind. It cannot tell you whether an election was high-turnout or low, only who was over-represented within it. Figure 1's counts button is there for that.

## What this chapter cannot tell you

**These are people who said they voted.** The Current Population Survey asks, after the election, and nobody checks. The [validated-turnout chapter](../../02-surveys/validated-turnout/validated-turnout-brief.html) establishes that the answer is consistently too high — in `r LP` the survey implies more voters than there were ballots. If over-reporting differs by age, and there is good reason to think it does, then the right-hand pyramid is distorted in a way none of these figures can correct. **Every "voter" in this chapter is a survey respondent's claim.**

**Four bands is a coarse pyramid.** A conventional population pyramid has five-year bands and splits by sex; this one has four unequal bands and no sex, because that is what the historical table publishes. The 25-to-44 band spans twenty years and the 65-and-over band has no upper bound, so both hide structure — the oldest band in particular mixes people at very different turnout rates.

**Nothing here is about a state.** These are national figures. The [turnout-denominator chapter](../../03-elections/turnout-denominator/turnout-denominator-brief.html) shows how much states differ on the denominator alone, the bottom number a turnout rate is divided by. An age pyramid for Florida would not look like one for Utah.

**Being over-represented is not the same as being decisive.** A band that casts `r p1(OVOTE)`% of the votes decides nothing on its own, and this chapter has said nothing about how any band voted. It is a chapter about who shows up, and who shows up is prior to, and separate from, what they do when they get there.

## Sources

- U.S. Census Bureau. *Current Population Survey, Voting and Registration
  Supplement*, Table A-1, historical time series, November 1964 to 2024.
  <https://www2.census.gov/programs-surveys/cps/tables/time-series/voting-historical-time-series/hst_vote01.xlsx>
- The same workbook is read for its national totals by this corpus's
  [validated-turnout chapter](../../02-surveys/validated-turnout/validated-turnout-brief.html), which
  documents what the table does and does not publish. The four age blocks
  further down the same sheet are read here, and checked to see that they
  reconstruct the published total.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`bands.csv`](data/derived/bands.csv), [`shares.csv`](data/derived/shares.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `shares.csv` carries a `gap` for every year and age band: the share of the population set against the share of the voters. Which band's gap has closed the most across the series, and which has barely moved?
- `bands.csv` reports `pct_registered` and `pct_voted` separately. Work out what share of the registered actually voted, band by band. Do the youngest fail to register, or fail to turn out?
- Pick a presidential year in `bands.csv` and the midterm beside it. Which age band falls away the most between the two?
- Everything here is people saying they voted, not records of them voting. Add up `voters` across the bands and set it against a turnout figure from anywhere else. Which way does the error run?

```{r ai-prompt, echo=FALSE, results="asis"}
```
