---
title: "Residual Votes and Ballot Design, 2000–2024"
type: brief
subtitle: "A punch card from Palm Beach County, and the one subtraction no agency performs"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("residual-votes-code.R")
```

```{r setup, include=FALSE}
```

A ballot is just paper, and paper is not supposed to decide elections. Does the design of a ballot actually change election outcomes? Palm Beach County, Florida, put that question beyond argument on 7 November 2000. Its presidential ballot ran the tickets down two facing pages with one column of punch holes between them. On the left-hand page, Bush was first and Gore second. In the column of holes down the middle, the second hole belonged to Pat Buchanan.

```{r butterfly-d3, results='asis', eval=knitr::is_html_output()}
```

```{r butterfly-static, eval=!knitr::is_html_output(), fig.width=7.2, fig.height=3.6}
```

**Figure 1. The Palm Beach County presidential ballot, November 2000.** A sketch, not a facsimile: only the arrangement the U.S. Commission on Civil Rights recorded is drawn. The tickets sit on two facing pages with the punch holes down the centre, Bush and Gore first and second on the left page, and the Reform Party hole second in the column. A voter who counted names down the left page and punched the matching hole voted for a candidate they had not read.

That sheet of paper produced two failures, and only one leaves a trace. A voter who took Gore's position in the list and punched the second hole cast a counted vote, for Buchanan, and no arithmetic will ever find it. A voter who caught the mistake and punched Gore as well spoiled the ballot; by the Commission's count, roughly 19,000 Palm Beach ballots carried more than one presidential punch. Those ballots were counted as ballots and contain no president. The presidency turned on 537 votes in Florida that year.

The second failure is a thing you can count. Take the ballots cast, subtract the votes counted in the top race, and what is left over after you subtract is the **residual vote**: ballots that entered the system without a usable choice for president. After 2000, Ansolabehere and Stewart computed exactly that and showed that residual rates varied by voting technology in a patterned way. Punch cards were worst; hand-counted paper and optical scan far better. That evidence helped drive the Help America Vote Act and the end of punch-card voting.

Design moves votes measurably, and the residual-vote rate is how you see it. So this brief computes that rate for the whole country in 2024 — and finds that the harder problem is now the record itself.

## The test

The measure needs two numbers per place: ballots cast and presidential votes counted. No agency publishes both, which is why nobody publishes the residual vote. The ballots come from the [U.S. Election Assistance Commission's](https://www.eac.gov/research-and-data/datasets-codebooks-and-surveys) *Election Administration and Voting Survey 2024*, Version 2, item F1a — a federal questionnaire every election office answers about itself. The votes come from a public county-level compilation of the fifty-one state canvasses, [Tony McGovern's returns file](https://github.com/tonmcg/US_County_Level_Election_Results_08-24), as prepared by this book's [data-sources chapter](../../06-putting-data-together/data-sources/data-sources-brief.html).

```{r sources}
```

Why these sources, when the question is about individual ballots? The record that answers it properly is the cast vote record, which shows what each ballot actually carried — and it exists for a few dozen jurisdictions, which is not a country. Certified returns cover everywhere but are totals by construction, silent about any single ballot; [the returns chapter](../returns-source/returns-source-brief.html) sets out that limit. The subtraction is the one measure that works everywhere, and it buys that reach by giving up the reason a ballot came up empty.

The subtraction also has a property most measures lack. You cannot count more presidential votes than there were ballots to carry them, so the residual vote has a hard floor at zero. A negative residual vote is not a low rate; it is proof that one of the two sources is wrong. That turns the measure into a test of the record as well as a measurement of the ballots.

## The data

The survey arrives as one CSV, one row per election jurisdiction — 6,461 of them. Jurisdictions are counties in most states, but Wisconsin, Michigan and most of New England run elections by township. Those township rows are summed into counties using the first five digits of the FIPS code, which are the county code. The returns arrive one row per county. The two are joined on the FIPS code, carried as text throughout, because `r n(sum(substr(co$fips, 1, 1) == "0"))` of the `r n(nrow(co))` county codes begin with a zero that a numeric column would silently drop. One joined row looks like this:

```{r one-row}
```

Allegheny County took in `r n(ex$ballots)` ballots and counted `r n(ex$total_votes)` presidential votes. The difference, `r n(ex$residual)` ballots, is the residual vote: `r pc(ex$residual_rate)`% of everything cast.

Two cleaning decisions matter enough to state. First, the survey writes "did not answer" and "does not apply" as negative codes in the same columns as the counts. Every negative value is therefore set to missing before anything is added up; summed as they stand, those codes subtract real ballots from real counties. Second, where the join produced an impossible answer, more presidential votes than ballots, the row was not dropped. A disagreement between two official sources is a finding, and tidying it away would destroy it.

The result is four tables. `counties.csv` holds one row per county: `fips`, `ballots` (item F1a, summed to the county), `total_votes` (the presidential total), `residual` (the subtraction), `residual_rate` (the same thing as a percentage of ballots), and `state_usable`. That last column is a judgment this brief attaches, not a fact from either source: it flags three whole states whose answers imply they read "ballots cast" to mean something else. `anomalies.csv` holds the `r nrow(an)` impossible rows, with the same columns. `unusable_states.csv` names the three set-aside states and gives each a `reason`. `states.csv` rolls the usable counties up to state level.

> **Before you read on, commit to a number.** Two official sources describe the
> same `r n(nrow(co) + nrow(an))` counties, and the subtraction between them
> cannot come out below zero unless one of them is wrong. **In how many
> counties does it?** Write down a count out of `r n(nrow(co) + nrow(an))`
> before you look.

## What it says about democracy

```{r summary}
```

A median county has a residual vote rate of `r pc(median(u$residual_rate))`% — about one ballot in a hundred arriving without a usable presidential vote. Half of all counties sit between `r pc(quantile(u$residual_rate,.25))`% and `r pc(quantile(u$residual_rate,.75))`%. One percent sounds small. In a state decided by twenty thousand votes it is not. And it is the size of number that design moves: one bad butterfly pushed one county's spoiled ballots into the tens of thousands.

Most readers write down zero for the prompt above; an official count is not the sort of thing one expects to come out impossible. **`r nrow(an)` counties report more presidential votes than ballots cast.** Figure 2 puts them on the same axis as everybody else.

```{r floor-setup, include=FALSE}
```

```{r floor-d3, results='asis', eval=knitr::is_html_output()}
```

```{r floor-static, eval=!knitr::is_html_output(), fig.height=3.7}
```

**Figure 2. The arithmetic floor, and `r nrow(an)` counties on the wrong side of it.** A histogram with the floor drawn in, because this measure's defining property is a boundary: nothing can sit left of the heavy black line and still be a measurement. The blue bars are the `r n(nrow(u))` counties whose two sources are at least consistent, with the median marked in red at `r pc(median(u$residual_rate))`%. Each red dot is a county reporting more presidential votes than ballots, spread vertically so overlapping cases stay countable. `r ffar` fall left of the frame entirely, the worst at `r pc(an$residual_rate[fwi], 1)`% in `r an$county_name[fwi]`, `r an$state_name[fwi]`.

```{r anomalies}
```

The largest gap is Cook County, Illinois, off by `r n(max(an$gap))` votes — and it is diagnosable. Chicago runs its own elections separately from suburban Cook County, so the survey row and the returns row describe different administrative units. The township-to-county summing assumed that smaller units nest cleanly inside counties, and here they do not, in the second most populous county in America. Which source is wrong cannot be determined from here; that one of them is wrong can be, and that is more than either source discloses about itself.

```{r map-d3, results='asis', eval=knitr::is_html_output()}
```

```{r map-static, eval=!knitr::is_html_output(), fig.width=7.2, fig.height=5.4}
```

**Figure 3. Every county in the United States, colored by residual vote rate.** A map, because the audit's last question is *where*: `r n(nrow(ma))` counties from Census TIGER 2020 boundaries, the blue ones passing the arithmetic test and shaded by rate. The `r n(MIM)` in red report more presidential votes than ballots; they fall in `r mn("impossible_states")` states, most heavily in `r mv("top_impossible_state")` with `r mn("top_impossible_n")`, a clustering that looks like reporting practice rather than like voters. The `r n(MSE)` in mid gray belong to the three states set aside below. The `r n(MNO)` in pale gray never entered the join at all, including every county in `r knitr::combine_words(mb$state)` and `r mn("n_va_nodata")` of Virginia's independent cities.

The exceptions have their own stories, and each was found by the arithmetic failing rather than by any documentation.

```{r unusable}
```

Mississippi returns exactly `r pc(un$median_rate[un$state == "Mississippi"], 3)`% in every county — ballots cast equal to presidential votes, to the vote, statewide. That is not a state with perfect ballots; it is a state that answered the ballots-cast question with its presidential total. The District of Columbia's `r pc(un$median_rate[un$state == "District of Columbia"], 1)`% comes from a survey that reports by ward joined to returns that do not. Vermont's `r pc(un$median_rate[un$state == "Vermont"], 1)`% in every county at once is a signal about reporting, not about Vermont's voters.

The floor also catches errors in one direction only. Wabash County, Illinois survives the test at `r pc(max(u$residual_rate), 1)`%, nearly two thirds of its ballots recording no president, which no plausible mechanism produces. An error that inflates the rate passes the test and looks like data.

What survives is still most of the country: `r n(nrow(u))` of the `r n(nrow(co) + nrow(an))` counties joined, or `r pc(100*nrow(u)/(nrow(co)+nrow(an)), 0)`%. On those, about one ballot in a hundred is cast and records no president. And the audit is the other half of the finding. Two official datasets describing the same election disagree in `r nrow(an)` counties and three whole states, and nobody reconciles them, because no agency owns the comparison. The number that detected the Palm Beach ballot's kind of failure still has to be assembled, by hand, from two files that were never built to meet.

## What this brief cannot tell you

**It cannot say why any ballot is blank.** A deliberate abstention, a spoiled mark and a scanner failure all count the same in the subtraction. That is why the classic studies compare equipment types across otherwise similar places rather than reading levels directly: the level in one county alone attributes nothing. No equipment data is joined here, so the comparison that made this measure famous is not run in these pages.

**It cannot say which source is wrong** in any of the `r nrow(an)` impossible counties — only that one of them is. Settling it would take the underlying certified canvasses, county by county.

**It cannot catch errors on the survivable side of zero.** Wabash County is almost certainly as broken as the red counties, and it passes. A floor is a one-way test.

**And it is one election.** A single year is a photograph. Whether rates still vary by equipment as they did in 2000 needs more years and the equipment join, and both are extensions below.

## What you should have learned

- A residual vote is ballots cast minus votes counted in the top race — a countable trace of ballot failure. In 2024 the median county's rate was `r pc(median(u$residual_rate))`% of ballots cast.
- Ballot design moves that number measurably: after the 2000 election, residual rates sorted by equipment type, punch cards worst, and the finding helped end punch-card voting by law.
- The measure has a floor at zero, and `r nrow(an)` of `r n(nrow(co) + nrow(an))` counties fall below it — proof that one of two official sources is wrong in each, though not which.
- Three whole states, one of them reporting exactly `r pc(un$median_rate[un$state == "Mississippi"], 3)`% everywhere, answered the ballots-cast question with something else and had to be set aside.
- No agency publishes both halves of the subtraction, so the number capable of detecting the next bad ballot design is published by nobody.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `counties.csv` has `residual_rate` for every county, with `state_usable` beside it. Sort the usable counties by rate. Which places lose the most votes between the ballot and the count, and do they share anything?
- `anomalies.csv` holds the impossible rows. Sort by `residual` and read the worst ten. How many look like Cook County — a city counted separately from its county?
- `unusable_states.csv` gives each excluded state a `reason`. Decide whether you agree with each exclusion, and what including the state would do to the national median.
- `states.csv` has a `residual_rate` per state. Compare the highest and lowest usable states. Is a high state rate more plausibly voters, ballot design, or reporting?

Two stretch ideas point past the spreadsheet. The survey this brief takes its ballots from also asks what voting equipment each jurisdiction uses. Download the release from the Commission's address below, join the equipment items onto `counties.csv`, and rerun the 2000-era comparison on 2024 data. And for one jurisdiction that publishes cast vote records, count its blank and overvoted presidential ballots directly and check the subtraction's answer against the ballots' own.

## Sources

**Data**

- U.S. Election Assistance Commission, *Election Administration and Voting
  Survey 2024*, Version 2, item F1a. Version 2 rather than Version 1: the
  first release was corrected, and the Commission's errata note says what
  changed.
  <https://www.eac.gov/sites/default/files/2026-02/2024_EAVS_for_Public_Release_nolabel_V2_csv.zip>
- County-level presidential returns for 2024, from Tony McGovern, *United
  States General Election Presidential Results by District, Ward, or County
  from 2008 to 2024*, by way of
  [`pres2024_counties.csv`](../../06-putting-data-together/data-sources/data/derived/pres2024_counties.csv)
  in the [data-sources chapter](../../06-putting-data-together/data-sources/data-sources-brief.html),
  which documents the download and what the file is and is not.
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>
- U.S. Census Bureau, TIGER/Line 2020 county and state boundaries, for the
  map.
  <https://www2.census.gov/geo/tiger/TIGER2020/COUNTY/tl_2020_us_county.zip>
  and <https://www2.census.gov/geo/tiger/TIGER2020/STATE/tl_2020_us_state.zip>

**Academic literature**

- Ansolabehere, S. and Stewart, C. III (2005). "Residual Votes Attributable
  to Technology." *Journal of Politics* 67(2): 365–389.
- Caltech/MIT Voting Technology Project (2001). *Voting: What Is, What Could
  Be.* <https://vote.caltech.edu/reports/1>
- U.S. Commission on Civil Rights (2001). *Voting Irregularities in Florida
  During the 2000 Presidential Election.* The Palm Beach ballot's arrangement
  in Figure 1 is drawn from its account.
  <https://www.usccr.gov/files/pubs/vote2000/report/main.htm>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`anomalies.csv`](data/derived/anomalies.csv), [`counties.csv`](data/derived/counties.csv), [`states.csv`](data/derived/states.csv), [`unusable_states.csv`](data/derived/unusable_states.csv)

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
