---
title: "Key: A Reason You Are Not Required to Give"
subtitle: "84-355 Democracy's Data | Jury selection | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
s <- read.csv("data/strikes.csv", stringsAsFactors = FALSE)
t <- read.csv("data/trials.csv",  stringsAsFactors = FALSE)
t$ns <- t$black_struck + t$white_struck
t$N  <- t$black_eligible + t$white_eligible
t$p  <- mapply(function(bs, be, n, N)
                 if (be > 0 && n > 0 && n < N)
                   phyper(bs - 1, be, N - be, n, lower.tail = FALSE) else NA,
               t$black_struck, t$black_eligible, t$ns, t$N)
k <- !is.na(t$p) & t$black_eligible >= 3
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`jury-selection-brief.Rmd`) is now
a chapter rather than a parallel copy of the lab: prose organised by argument,
under idea headings, with four numbered figures and no numbered steps. The lab
(`jury-selection.Rmd`) still runs Steps 1–10 with the code exposed. **Students
should read the brief first.** **Step numbers below refer to the lab**; the
chapter is cited by figure number or section title.


**One of three labs in the discrimination unit**, alongside redlining and
policing. Run in parallel and have the groups compare inferences.

**This is the strongest inference of the three**, because the control group is
inside the file, and it is worth telling students that once they have found it.

Every number below was run against the data files.

***

# The session in one sentence

Prosecutors struck half the Black jurors they could and a ninth of the white
ones; the defense did the mirror image; and no court found a violation until
journalists assembled 225 trials in one place.

***

# The findings

```{r headline}
s
c(state_ratio = round(s$pct[1] / s$pct[2], 1),
  trials_above_the_diagonal = sum(t$black_struck[k] / t$black_eligible[k] >
                                  t$white_struck[k] / pmax(t$white_eligible[k], 1)),
  trials_tested = sum(k))
```

- State struck **49.8%** of eligible Black jurors, **11.2%** of white — **4.4×**.
- Defense struck **47.3%** of white, **15.0%** of Black — the mirror.
- **188 of 202 trials (93%)** above the diagonal.
- 11 trials where the state struck *every* eligible Black juror; 59 where it
  struck *none* of the white ones.

***

# The three things to get said

**1. Peremptory means "no reason required."** Students will not know this and it
is the hinge of the whole lab. A prosecutor may strike almost anyone for almost
anything. *Batson* forbids only race, and only if you can show it — and a
race-neutral explanation is easy to produce. **Say plainly: in the thirty years
after *Batson*, the North Carolina Supreme Court never once found a
discriminatory strike.**

**2. The mirror is the finding, not the disparity.** Make the room construct the
objection at Step 4 first — *maybe the Black jurors differed*. Then reveal the
defense column. **The objection now has to hold that white jurors were equally
different in the opposite direction to the opposite side.** One variable explains
both; the objection needs two coincidences.

**3. Aggregate evidence, individual law.** *Batson* operates inside one trial,
where numbers are small and each strike gets an individual explanation. The
pattern is only visible from above. **This mismatch is the lab's real subject**,
and Option A demonstrates it arithmetically rather than rhetorically.

**The line for the board:** the pattern is illegal, and no single strike in it
is.

***

# Handling the room

Charged material, and the case has a name and a person attached. Two things:

- **Curtis Flowers was released in 2019 and the charges dropped in 2020.** Say
  so. Students otherwise leave assuming the story ends badly, and the actual
  ending is relevant: the remedy came from journalism and the Supreme Court, not
  from the trial-level process that was supposed to catch it.
- **Do not let it become a discussion of whether he was guilty.** This lab is
  about who was allowed to sit on the jury. That question is independent of
  guilt, and keeping them separate is itself worth teaching.

***

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** One trial | The four counts on one row | 4 | Make them read a single row before any total. |
| **2** Scope | 211 trials, one prosecutor, 26 years | 3 | The narrowness is the point — this is about a decision-maker. |
| **3** The disparity | 49.8% vs 11.2%, a 4.4× ratio | 5 | Students think this is the finding. It is not. |
| **4** Why it is not an answer | The "jurors differed" reply | **7** | **Give the reply its full force.** Do not let anyone wave it away. |
| **5** The control group | The defense strikes the same pools | 5 | **Take a show of hands on the prediction before Step 6.** |
| **6** The mirror | 47.3% white / 15.0% Black | **6** | The pivot. Let the reversal sit before explaining it. |
| **7** One trial's probability | `phyper`, p = 0.044 | **7** | **It must feel weak.** That is the point about how *Batson* is litigated. |
| **8** All 211 | 99 significant where the simulated null gives 3.5 | 6 | About 28x the chance rate. **The shortcut 0.05x211=10.6 is wrong**: discrete tests on small counts are conservative, so being careful makes the disparity look larger, not smaller. |
| **9** Trial by trial, and Flowers III | 15 of 15 struck, 0 of 21 | 5 | One chance in five billion. |
| **10** What you can say | Pattern, and the law's blind spot | 6 | The aggregate/individual gap is the transferable lesson. |

**Steps 5–7 are the spine of the session.** Ask the room to predict the
defense's numbers before Step 6 — the mirror is far more persuasive to somebody
who has just committed to a wrong guess. Then Step 7 has to *disappoint*: a 4.4%
probability is what a real *Batson* motion works with, and it loses.

**If you are short on time**, protect Steps 4, 6 and 7. The objection, the
mirror that answers it, and the single trial that cannot carry the evidence are
the argument; the rest is detail.


# Option A — What are the odds?

## Verified output

```{r a}
c(testable_trials = sum(k),
  p_below_05 = sum(t$p[k] < 0.05),
  pct        = round(100 * mean(t$p[k] < 0.05)),
  p_below_01 = sum(t$p[k] < 0.01),
  median_p   = round(median(t$p[k]), 3))
```

```{r a2}
o <- t[k, ][order(t$p[k]), ][1:4, ]
o[, c("trial", "black_eligible", "black_struck",
      "white_eligible", "white_struck", "p")]
```

**Ninety-eight of 199 testable trials — 49% — are individually significant at
p < 0.05, and the median trial sits at p = 0.051.**

**And look at the top row.** The single most extreme trial in twenty-six years
is ***Curtis Flowers III*: 15 eligible Black jurors, all 15 struck; 21 eligible
white jurors, none struck. p = 1.8 × 10⁻¹⁰.**

**3/4:** Computes `phyper()` for a few trials and reports that some are extreme.

**4/4:** Answers the question actually asked — *is any single trial damning on
its own?* — and finds that the median one is not, at p = 0.051. Half the trials
individually clear the conventional bar and half do not, which is exactly the
position a defendant is in.

**4/4, best:** Notices that Flowers' own trial is the most extreme in the
dataset, and draws the right conclusion: **it took a case that extreme, tried six
times, plus a year of journalism, to get relief.** A pattern that is obvious in
aggregate produced no remedy for the 98 other trials that were individually
significant.

***

# Option B — The pooled test

## Verified output

```{r b}
prop.test(c(902, 372), c(1811, 3318))$p.value
```

**p = 7.7 × 10⁻²⁰⁵.**

**3/4:** Runs it and observes the number is absurdly small.

**4/4:** Says what the contrast with Option A means. The evidence exists and is
overwhelming — but it only exists at a level no court considers. **Batson asks
about one trial; the answer lives in 225.** A student who says the law is
looking through the wrong end of the telescope has it.

**4/4, best:** Notices that a vanishingly small p-value is not the interesting
part, because with 5,129 jurors almost any real disparity would clear
significance. **The effect size is what matters — 49.8 against 11.2** — and a
student who says so has absorbed the lesson from the prediction-market lab about
not confusing significance with importance.

***

# Option C — Missing race

No verified output; grade the reasoning.

**4,752 of 14,873 jurors — 32% — have race Unknown.**

**3/4:** Gives one scenario each way.

**4/4:** Makes the scenarios specific. *Inflating*: if race was easier to
determine for jurors who were struck (more courtroom discussion, more record),
and if struck jurors were disproportionately Black, the observed pool
over-represents exactly the cases that show the pattern. *Hiding*: if race went
unrecorded more often in the rural, whiter counties of the district, those
trials drop out and with them a set of low-disparity trials.

**4/4, best:** Notices that the direction is testable from the file itself —
compare the Unknown share across trials with high and low measured disparity —
and says that a third of the data missing non-randomly is a bigger threat than
anything in Options A or B. **Being able to rank the threats to your own finding
is the 4/4 skill across this whole unit.**

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is explaining why the defense column changes the
  strength of the inference — not merely noticing it is there.
- **A 2** is reporting 49.8 vs 11.2 as the finding and stopping.
- **Also a 2**: "the p-value proves it." See Option B, best answer.
- **Reward anyone who separates the pattern claim from the guilt question**
  without being told to.

## Anticipated questions

- *"Isn't the defense doing the same thing?"* — Yes, and say so plainly. Both
  sides used race. *Batson* applies to both in principle and is enforced almost
  entirely against the defense's opponent in practice. Students find this
  genuinely surprising and it is a good place to end.
- *"What happened to Flowers?"* — Conviction vacated in 2019, charges dropped in
  2020 after 23 years, most of it on death row.
- *"What happened to Doug Evans?"* — He remained District Attorney. Worth
  stating without comment.
- *"Could you use this in court?"* — This kind of statistical evidence has been
  offered and largely rejected at the trial level; the Supreme Court in
  *Flowers* used the history of one prosecutor against one defendant. **That
  distinction is the lab.**

## Connections

- **The other two labs in this unit** — policing (no control group, so it needs
  the outcome test) and redlining (intent documented, effect contested).
  **Ask which of the three inferences the room would most want to take to
  court.**
- **Session 12 (SCOTUS)** — *Flowers* is in the database students used there.
- **RPV** — statistical evidence of discrimination, and what courts do with it.
