---
title: "A Reason You Are Not Required to Give"
subtitle: "84-355 Democracy's Data | Peremptory strikes in Mississippi, 1992–2017"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **One of three labs in this unit.** All three ask the same question of
> different data: **can you show discrimination when nobody wrote down a
> reason?** The others are redlining and policing. Whichever you did, you will
> be asked to explain your inference to somebody who did a different one.

> **Read `jury-selection-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"The prosecutor struck Black jurors because they were Black."*

That is a constitutional violation. *Batson v. Kentucky* (1986) forbids striking
a juror on account of race, and gives defendants a way to object.

It is also almost impossible to prove one strike at a time. A **peremptory
strike** is the one a lawyer may use *without giving a reason* — that is what
"peremptory" means. Challenged under *Batson*, a prosecutor need only offer some
race-neutral explanation, and almost any explanation will do: the juror seemed
inattentive, had a relative with a conviction, lived near the defendant.

**In the thirty years after *Batson*, the North Carolina Supreme Court never
once found a strike to be racially discriminatory.**

So how would you ever show it? Not from one strike. From thousands.

# Step 1 — What one record is

```{r load}
s <- read.csv("data/strikes.csv", stringsAsFactors = FALSE)
t <- read.csv("data/trials.csv",  stringsAsFactors = FALSE)

c(trials = nrow(t))
s
```

Journalists at APM Reports spent a year reading court records from **225 trials
over 26 years** in Mississippi's Fifth Circuit — every trial prosecuted by
District Attorney Doug Evans's office. **14,873 jurors**, each with a race and a
record of what happened to them.

They did it because of one defendant: **Curtis Flowers, tried six times for the
same crime by the same prosecutor.** In *Flowers v. Mississippi* (2019) the
Supreme Court threw out his conviction, and the majority relied heavily on the
pattern of who Evans struck.

# Step 2 — How many, and over what

```{r scope}
c(trials = nrow(t),
  jurors_eligible = sum(t$black_eligible + t$white_eligible),
  struck_by_state = sum(t$black_struck + t$white_struck))
```

One prosecutor's office, 26 years. That narrowness is the point: a national
average would tell you about American prosecution, this tells you about a
**decision-maker**.

# Step 3 — The disparity

```{r state}
st <- s[s$side == "state", ]
st
round(st$pct[st$race == "Black"] / st$pct[st$race == "White"], 1)
```

**The state struck 49.8% of the Black jurors it could strike, and 11.2% of the
white ones** — **4.4 times** the rate.

Note the denominator, because it matters. A juror already removed for cause, or
never reached, could not be struck by anybody. These rates count only jurors the
**state was actually able to strike** — which is a stricter and fairer
comparison than dividing by everyone in the room.

# Step 4 — Why that is not yet an answer

**A disparity is not proof.** The reply is ready, and it is not frivolous:
perhaps the Black jurors in these pools differed in some legitimate way — more
of them had relatives with convictions, or expressed doubts about the death
penalty. If those characteristics are unevenly distributed in a county with this
history, a race-neutral striking rule produces a racial disparity on its own.

**You cannot refute that with the table above**, because it has no control
group. What you need is a group of people who saw exactly the same jurors, in
the same trials, under the same rules, and wanted the *opposite* outcome.

# Step 5 — The control group is in the file

**The defense strikes too.** Same trials, same jury pools, same *Batson* rules,
opposite incentives.

If Black jurors really were disproportionately strike-worthy for reasons both
sides would recognise, the defense should strike them too — or at least not
strike them dramatically *less*.

**Before running the next chunk, write down what you expect the defense's two
numbers to look like under each explanation.** The chunk is two lines and it
will not wait for you.

# Step 6 — The mirror

```{r both}
s
```

```{r mirror-plot}
m <- matrix(c(s$pct[s$side == "state"   & s$race == "Black"],
              s$pct[s$side == "state"   & s$race == "White"],
              s$pct[s$side == "defense" & s$race == "Black"],
              s$pct[s$side == "defense" & s$race == "White"]), nrow = 2)
barplot(m, beside = TRUE, names.arg = c("struck by the STATE", "struck by the DEFENSE"),
        col = c("#C41230", "#2c7fb8"), ylab = "% of eligible jurors struck",
        legend.text = c("Black jurors", "white jurors"), ylim = c(0, 60))
```

**The state struck 49.8% of Black jurors and 11.2% of white.
The defense struck 47.3% of white jurors and 15.0% of Black.**

It is a mirror. Whatever made Black jurors worth striking to the prosecution
made white jurors worth striking to the defense — at almost exactly the same
rate, in the same courtrooms, from the same pools.

**This is what makes the inference strong.** The "Black jurors were different"
explanation now has to explain why white jurors were *equally* different, in the
opposite direction, to the opposite side. There is one variable that does that
easily, and both sides were clearly using it.

Only one side's use of it is unconstitutional in the way *Batson* was written to
catch.

# Step 7 — Now try to prove it in a single trial

Here is the difficulty at the centre of the law. ***Batson* is litigated one
trial at a time**, in front of a judge who sees one jury pool. Aggregate
patterns are not before the court.

So take the first trial in the file on its own terms and ask: if the state had
chosen its strikes at random, how often would the split have come out this
lopsided or worse? That is an urn problem, and R has it built in.

```{r one-trial}
o <- t[1, ]
p1 <- phyper(o$black_struck - 1, o$black_eligible, o$white_eligible,
             o$black_struck + o$white_struck, lower.tail = FALSE)
c(black_eligible = o$black_eligible, black_struck = o$black_struck,
  white_eligible = o$white_eligible, white_struck = o$white_struck,
  probability = round(p1, 3))
```

**About one chance in twenty-three.** Suggestive, and nowhere near enough. A
judge would be right to say that unlikely things happen in single trials.

**A single trial almost never carries the evidence.** That is not a flaw in this
data — it is a description of how the rule works.

# Step 8 — Twenty-six years of single trials

Compute that same probability for every trial and count.

```{r all-trials}
t$p <- phyper(t$black_struck - 1, t$black_eligible, t$white_eligible,
              t$black_struck + t$white_struck, lower.tail = FALSE)
# The shortcut 0.05 * 211 assumes p-values spread evenly over 0 to 1. These are
# hypergeometric tests on small counts, so their p-values are discrete and the
# tests are conservative -- they fire LESS often than 5% of the time. Simulate
# the real null instead: redraw each trial's strikes from its own margins.
set.seed(20260810)
sim <- replicate(2000, {
  bs <- rhyper(nrow(t), t$black_eligible, t$white_eligible,
               t$black_struck + t$white_struck)
  sum(phyper(bs - 1, t$black_eligible, t$white_eligible,
             t$black_struck + t$white_struck, lower.tail = FALSE) < 0.05)
})
c(trials = nrow(t),
  significant_05 = sum(t$p < 0.05),
  expected_shortcut = round(0.05 * nrow(t), 1),
  expected_simulated = round(mean(sim), 1),
  significant_01 = sum(t$p < 0.01))
```

**Ninety-nine of 211 trials clear the 0.05 threshold, where chance alone would
produce about 3.5.** That is about twenty-eight times as many. Note that the
textbook shortcut, 0.05 x 211 = 10.6, is wrong here and wrong in the direction
that *understates* the finding: discrete tests on small counts are conservative,
so the real chance rate is lower than the nominal one.

# Step 9 — Trial by trial, and the one at the far corner

```{r trials}
t$black_rate <- ifelse(t$black_eligible > 0, 100 * t$black_struck / t$black_eligible, NA)
t$white_rate <- ifelse(t$white_eligible > 0, 100 * t$white_struck / t$white_eligible, NA)
k <- t$black_eligible >= 3 & t$white_eligible >= 3

c(trials = sum(k),
  struck_every_black_juror = sum(t$black_struck[k] == t$black_eligible[k]),
  struck_no_white_juror    = sum(t$white_struck[k] == 0))
```

```{r trial-plot}
plot(jitter(t$white_rate[k], 2), jitter(t$black_rate[k], 2), pch = 19,
     col = adjustcolor("#C41230", 0.5), xlim = c(0, 100), ylim = c(0, 100),
     xlab = "% of eligible white jurors struck by the state",
     ylab = "% of eligible Black jurors struck by the state")
abline(0, 1, lty = 2)
```

Every point above the dashed line is a trial where the state struck Black
jurors at a higher rate than white ones. **188 of 202 trials — 93% — are above
it.** In 11 trials the state struck *every* Black juror it could; in 59 it
struck *no* white juror at all.

This is also the level at which the law operates. A *Batson* challenge happens
inside **one** trial, where the numbers are small and a prosecutor can explain
each strike individually. The pattern is only visible from above — and no
individual defendant has 225 trials to point at.

**Curtis Flowers did, eventually, because journalists built this file.**

```{r worst}
w <- t[order(t$p), ][1, ]
c(trial = w$trial, black_eligible = w$black_eligible, black_struck = w$black_struck,
  white_eligible = w$white_eligible, white_struck = w$white_struck)
format(w$p, digits = 2)
```

The most extreme trial in the file is **the third trial of Curtis Flowers**:
fifteen Black jurors eligible and fifteen struck, twenty-one white jurors
eligible and none. Roughly one chance in five billion.

# Step 10 — What you can now say

**You can say** that in this district, over 26 years, the state's strikes were
not race-neutral — and that the conclusion does not rest on assuming which
jurors *should* have been struck, because the defense's mirror-image behaviour
supplies the comparison.

**You can also say something about the law.** The pattern is overwhelming in
aggregate and nearly invisible one trial at a time. Step 7 was not a detour: a
4.4% probability is what a typical *Batson* motion actually has to work with,
and it loses. **The rule is enforced at the only level where the evidence is
weakest.**

**You cannot say** why any individual juror was struck. No reasons are recorded,
which is the whole design of a peremptory challenge.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — What are the odds?** Take a trial with, say, 5 eligible Black
jurors and 20 white, where the state struck 5 and 3. If strikes were made
without regard to race, the number of Black jurors struck follows a
hypergeometric distribution — `phyper()` in R will give you the probability of a
result that extreme. Compute it for a few trials. Is any *single* trial damning
on its own?

**Option B — The pooled test.** Do the same thing on the whole dataset at once
(`prop.test` on the two strike rates). Compare the p-value to what you got from
one trial in Option A, and say what that difference means for how *Batson* is
enforced.

**Option C — Missing race.** 4,752 of 14,873 jurors have race recorded as
Unknown — about a third — and they are excluded from everything above. Give one
scenario in which their exclusion inflates the disparity and one in which it
hides it. Which do you think is likelier, and why?

**Option D — Your own question.** 211 trials, two sides, one variable.

# What this lab cannot tell you

- **Why any individual juror was struck.** Peremptory strikes require no
  reason, and the reasons given when challenged are not in this file. The
  pattern is aggregate; the law is individual, and that mismatch is the whole
  problem.
- **Whether Doug Evans is unusual.** We have one district and one prosecutor.
  Studies in North Carolina and Philadelphia found similar patterns, but this
  file cannot establish that.
- **What race the jurors would have claimed.** Race here was coded by
  journalists from court records and other sources — not self-report. For a
  third of jurors they could not determine it at all.
- **Whether the juries were wrong.** Nothing here says anything about verdicts.
  It is about who got to sit.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- The disparity alone was not proof; the defense's mirror-image strikes made it
  much stronger. Why?
- A pattern visible across 225 trials is invisible inside one. What should a
  court do about that?
- If you did one of the other two labs in this unit, which inference did you
  find more convincing, and why?

***

### Where the data came from

**APM Reports**, *In the Dark* season 2 — peremptory strikes in Mississippi's
Fifth Circuit Court District, 1992–2017, compiled from court records across 225
trials. <https://github.com/APM-Reports/jury-data>

Strike rates use only jurors the relevant side was **eligible** to strike.
Jurors whose race the reporters could not determine — 4,752 of 14,873 — are
excluded. The build is in `data/build-data.py`.

The Supreme Court's decision is *Flowers v. Mississippi*, 588 U.S. ___ (2019).
