---
title: "Measuring Migration with ACS Flow Estimates"
type: brief
subtitle: "One question on an American survey form, the map it draws, and how much of that map is not there"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("migration-code.R")
```

```{r setup, include=FALSE}
```

Every few months a headline announces that Americans are fleeing one state for another. Fleeing California for Texas, fleeing New York for Florida, fleeing taxes, rents, and each other. The claim arrives with maps full of confident arrows. Is it true? Who actually moves, between which states, and how would anyone know?

The United States does not record moves. There is no register that learns your address when you change it, because the country has repeatedly declined to build one. What stands in its place is a single question on the American Community Survey: **Where did you live one year ago?** Cross the answers against where each household lives now, scale the sample up, and out comes a `r nrow(st)` × `r nrow(st)` table — `r n(mn("n_pairs"))` ordered pairs of places, a number printed in every cell.

One of those cells is `r mv("largest_fail_pair")`. It reads `r pm(mn("largest_fail_est"), mn("largest_fail_moe"))`. The second number is the Bureau's published 90 percent margin of error: how far off the estimate could be and still be doing its job. Read the two together and the federal government is saying that the number of people who made that move was somewhere between nobody at all and about `r n(mn("largest_fail_est") + mn("largest_fail_moe"))`. **The estimate is smaller than its own uncertainty.** Every figure in this brief carries a margin like that, and almost none of them is ever printed with it.

## The test

The test is the Census Bureau's own flow table: *State-to-State Migration Flows* from the `r mv("acs_year")` American Community Survey one-year estimates, downloadable from the address in the sources below; the copy read here is from `r format(as.Date(mv("fetched")), "%d %B %Y")`. For the long view it is joined by the Current Population Survey's mover series, which has asked the same question every year since `r mn("cps_first")`. The ACS is a survey, and the [American Community Survey chapter](../census-acs/census-acs-brief.html) is its biography: roughly 3.5 million addresses a year, weighted up to the country, a margin of error published beside every estimate.

It is the right source because there is no better one, and the near misses are instructive. The Internal Revenue Service builds county-to-county migration from the addresses on consecutive tax returns, an administrative record with no margin of error at all. But it covers only tax filers and the people claimed on their returns, which leaves out the poor and the young, and they are the most mobile people in the country. The Postal Service knows every forwarding order and publishes no flow table. A survey with honest margins is what remains, and the margins turn out to be the story.

## The data

The workbook the Bureau publishes was built to be looked at, not computed with. Its first eight rows are headers, one of which explains in a sentence where the headers are. Column names arrive with footnote markers welded on, so `Residence 1 year ago2` is a name carrying a footnote. And every state-to-itself cell holds the letter `X` in a numeric column: people who did not leave their state are suppressed, not counted.

The build script parses that workbook into `flows.csv`, one row per ordered pair of states. Its columns are `from_state`, `to_state`, the estimate `est`, its margin `moe`, the interval ends `lo` and `hi`, and a flag `sig`. Here is the top of it:

```{r clean-mig}
```

```{r shape-mig}
```

Three cleaning decisions sit in those columns, and each is a judgment. The `X` cells were read as absence rather than zero: `r mv("n_suppressed")` of the `r n(mn("n_pairs"))` pairs carry no usable estimate, and turning them into zeros would assert that nobody made those moves. The interval ends are the estimate plus and minus its margin, with the bottom held at zero, because a negative number of movers is not a thing. That clamp quietly makes some intervals lopsided. And `sig` is this brief's arithmetic, not the Bureau's: it asks only whether the interval clears zero.

Two companion tables travel with the flows. `states.csv` has one row per state: arrivals `in_est`, departures `out_est`, the difference `net`, each with its margin, plus `born_state_pct`, the share of residents born in the state they live in. `mobility.csv` is the Current Population Survey series: for each year since `r mn("cps_first")`, the share of Americans who moved at all (`movers`) and the share who crossed a state line (`diff_state`).

Before any arrows are drawn, the proportions:

```{r what-it-is}
```

`r n(mn("us_movers_within_state"))` people moved inside their own state against `r n(mn("us_movers_between_states"))` who crossed a state line. The interstate migration everyone argues about is roughly `r pc(100 * mn("us_movers_between_states") / (mn("us_movers_within_state") + mn("us_movers_between_states")), 0)`% of domestic moving, and about `r pc(100 * mn("us_movers_between_states") / mn("us_pop1"), 1)`% of the population in a year. The flows are real, and they are small beside the stayers.

> **Before you read on, commit to two numbers.** There are
> `r n(mn("n_pairs"))` ordered pairs of states in this table. **How many of
> them describe a flow the survey can tell apart from zero?** Write down a
> count, or a share of `r n(mn("n_pairs"))`. Then a second number: **of all
> the people who moved between states that year, what share are in the pairs
> that fail?**

## What it says about democracy

```{r gainers, include=FALSE}
```

Migration is apportionment, and apportionment is power. Every ten years the House of Representatives is redistributed by where people live: a state that gains people gains seats and electoral votes, and one that loses them loses both. Nobody votes on the transfer. It is settled by millions of private decisions about jobs, rent, and elderly parents, which is why this table gets argued over for the nine years between counts.

Start with the state with the biggest gap between arrivals and departures. The data picks it, not us: **`r FOCUS`**.

```{r focus-row}
```

**`r FOCUS` lost `r n(-F1$net)` more people to other states than it gained from them**, the largest net domestic outflow of any state. That figure is larger than its own margin of error, so it is not a trick of the survey. The biggest gainers are `r G$state[1]` and `r G$state[2]`, up `r n(G$net[1])` and `r n(G$net[2])`. Those loud claims stand; the trouble begins with the quiet ones.

```{r arcwidth-facts, include=FALSE}
```

```{r arcmap-d3, results='asis', eval=knitr::is_html_output()}
```

```{r arcmap-static, eval=!knitr::is_html_output(), fig.width=7.2, fig.height=2.8}
```

**Figure 1. Interstate migration into and out of `r FOCUS`, drawn between population-weighted state centers.** An arc map fits because the data is pairs of places: each arc is one pair, its width the number of people, on an absolute square-root scale keyed at lower right. Color carries a separate fact — blue means the estimate clears its own margin of error, gray means the survey cannot tell the flow apart from zero. In the browser the buttons switch direction; in print the two directions sit side by side.

The heaviest arc runs both ways, and it is the same arc. `r FOCUS` and `r PART` exchange more people than any other pair of states in the country: `r n(OUTMX)` one way, `r n(INMX)` back. A net migration figure is a small difference between two very large numbers, which is exactly the situation where a margin of error matters most.

Now the prediction. The Bureau prints a margin beside every cell, so the test is arithmetic on the Bureau's own two columns, applied to all `r n(mn("n_pairs"))` pairs:

```{r moe-accounting}
```

Most readers write something in the nineties for the first number. It is `r pc(mn("pct_sig_of_pairs"), 0)`%: **`r n(mn("n_fail"))` of the `r n(mn("n_pairs"))` pairs fail**, or `r pc(mn("pct_fail_of_pairs"), 0)`% of the table. Having read that, most readers revise the second number upward, and it goes the other way. **The failed pairs hold `r pc(mn("pct_volume_in_failed_pairs"), 2)`% of the people who moved.** A margin of error is set, roughly, by how many sampled households made a given move, so it is close to a fixed amount. It wipes out thin flows and barely grazes thick ones: the middle surviving flow is `r n(mn("median_est_sig"))` people, the middle failed one `r n(mn("median_est_fail"))`.

```{r cater-setup, include=FALSE}
```

```{r cater-static, eval=!knitr::is_html_output(), fig.height=3.3}
```

```{r cater-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Twelve flows with the 90 percent interval the Bureau publishes beside each one.** The top four are the largest flows in the country, where the interval is a rounding error beside the estimate. The eight below, in red, are the largest flows whose interval reaches zero: the survey cannot rule out that nobody at all made the move. The largest of them is `r mv("largest_fail_pair")`, at `r pm(mn("largest_fail_est"), mn("largest_fail_moe"))`.

So the table is wrong in a particular way, and it is not the way people usually fear. It is reliable about **volume**: nearly all interstate movement travels along flows large enough to measure. It is unreliable about **connections**, and the long, thin, surprising arc is precisely the one that makes a flow map look like a discovery. Nothing about the failing pairs is exotic. They are ordinary pairs of mostly small states, whose moves are rare in the sample rather than rare in the country. The pattern runs all the way up: for `r nrow(S) - mn("n_states_net_sig")` of these `r nrow(S)` jurisdictions, even the statewide net migration figure cannot be told apart from zero.

One more exhibit reframes all of it. The Current Population Survey has asked the same question since `r mn("cps_first")`. It is a different survey with a different sample, so its levels do not belong on the ACS's axis, but its direction is unambiguous.

```{r mob-facts, include=FALSE}
```

```{r mobility-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

```{r mobility-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. The share of Americans who moved in the previous year, `r mo$year[1]` to `r mo$year[nrow(mo)]`.** One line per kind of move, because the question is change over time. Every line falls: moving at all drops from `r pc(f$movers)`% to `r pc(l$movers)`%, and interstate moving from `r pc(f$diff_state)`% to `r pc(l$diff_state)`% — now `r pc(100 * l$diff_state / f$diff_state, 0)`% of its `r f$year` rate.

Every argument about Americans sorting themselves across the country is taking place against this decline. The flows are real. They are also small, shrinking, and at their edges unmeasurable.

## What this brief cannot tell you

**Why anybody moved.** There is no political column and no reason recorded anywhere in the file. The popular claim since Bishop's *The Big Sort*, that Americans move toward politically like-minded places, cannot be tested here. The research that can test it finds the effect far weaker than advertised (Mummolo and Nall 2017).

**Who left the country.** The survey interviews only people living in the United States, so there is no matching figure for anyone who departed. Emigration in this data is not small and not zero; it is unmeasured. And "moved from abroad" is a statement about geography, not legal status: it mixes returning citizens, military families, students and immigrants, and separates nothing further.

**Anything finer than a state, reliably.** The county version of this table pools five years of sample (`r mv("c2c_window")`), and even then only `r n(mn("county_all_sig"))` of the `r n(mn("county_all_pairs"))` county flows into `r CTY` clear their own margins. Finer grain means worse estimates.

**A trend, from the flow table alone.** It is a one-year snapshot: someone who moved in March and returned in November is invisible, and someone who moved three times counts once. The long decline in Figure 3 comes from a different instrument, trusted here for its direction and never for cell-by-cell comparison.

## What you should have learned

- One retrospective survey question is the entire national measurement of
  moving. In `r mv("acs_year")`, it found `r n(mn("us_movers_between_states"))`
  people crossing state lines while `r n(mn("us_movers_within_state"))` moved
  within their own state.
- `r pc(mn("pct_fail_of_pairs"), 0)`% of the `r n(mn("n_pairs"))` state pairs
  fail their own published margins of error, but those pairs carry only
  `r pc(mn("pct_volume_in_failed_pairs"), 2)`% of the movers: the table is
  reliable about volume and unreliable about connections.
- The loud findings survive: `r FOCUS` lost `r n(-F1$net)` people net, several
  times its margin, while `r G$state[1]` gained `r n(G$net[1])`.
- The uncertainty sits exactly where the interesting-looking pattern is: the
  more specific a flow claim sounds, the more likely it is a product of sample
  size.
- Interstate moving has fallen to `r pc(100 * l$diff_state / f$diff_state, 0)`%
  of its `r f$year` rate, so whatever sorting is happening is happening while
  the mechanism slows.

## Extensions

- Every pair of states appears in `flows.csv` twice, once in each direction
  of `from_state` and `to_state`. Find a pair where the two directions are
  wildly unequal, and a pair where they almost cancel. What kind of state
  pair gives you each?
- Count the FALSE rows in the `sig` column of `flows.csv`, then look at which
  states they connect. Are the unusable pairs scattered, or concentrated on
  the small states?
- Sort `states.csv` by `born_state_pct`, then compare it with `net_per1k`.
  Are the states people are least likely to be born in also the states
  gaining the most people now?
- In `mobility.csv`, follow `diff_state` down the years and find the decade
  where the fall is steepest. Then go and read what else was happening then.
- **Stretch.** Download the previous year's release of the same table from the
  Bureau's server and rerun the margin test: do the same pairs fail, or does
  the set of failures churn from year to year?
- **Stretch.** The IRS publishes county-to-county migration built from tax
  returns, with no margins at all. Compare its picture of one state with this
  one, and work out which differences come from who each source can see.

## Sources

**Data**

- U.S. Census Bureau, *State-to-State Migration Flows*, `r mv("acs_year")`
  American Community Survey 1-year estimates, Table T13.
  <https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx>
- U.S. Census Bureau, *State of Residence by Place of Birth*, `r mv("acs_year")`
  ACS 1-year estimates, Table T13.
  <https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/place-of-birth-acs-2024/State_of_Residence_By_Place_of_Birth_Table_2024_T13.xlsx>
- U.S. Census Bureau, *County-to-County Migration Flows*, `r mv("c2c_window")`
  ACS 5-year estimates.
  <https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2020/county-to-county-migration-2016-2020/county-to-county-migration-flows/county-to-county-2016-2020-ins-outs-nets-gross.xlsx>
- U.S. Census Bureau, *Geographical Mobility*, CPS historical Table A-1,
  `r min(mo$year)` to `r max(mo$year)`.
  <https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/time-series/historic/tab-a-1.xls>
- U.S. Census Bureau, *Centers of Population by State*, 2020 Census; and
  *Cartographic Boundary Files*, 2020.
  <https://www2.census.gov/geo/docs/reference/cenpop2020/CenPop2020_Mean_ST.txt> and
  <https://www2.census.gov/geo/tiger/GENZ2020/shp/cb_2020_us_state_20m.zip>

Margins of error throughout are the Bureau's published 90 percent margins.

**Academic literature**

- Bishop, Bill, with Robert G. Cushing (2008). *The Big Sort: Why the Clustering
  of Like-Minded America Is Tearing Us Apart.* Houghton Mifflin.
- Mummolo, Jonathan, and Clayton Nall (2017). "Why Partisans Do Not Sort: The
  Constraints on Political Segregation." *The Journal of Politics* 79(1): 45–59.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`arcs.csv`](data/derived/arcs.csv), [`county_focus.csv`](data/derived/county_focus.csv), [`flows.csv`](data/derived/flows.csv), [`map_insets.csv`](data/derived/map_insets.csv), [`map_states.csv`](data/derived/map_states.csv), [`meta.csv`](data/derived/meta.csv), [`mobility.csv`](data/derived/mobility.csv), [`states.csv`](data/derived/states.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
