---
title: "Key: Which Way Did It Blow"
subtitle: "84-355 Democracy's Data | The election wind map | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 96, fig.retina = 1)
options(scipen = 999)
D  <- "data"
us <- read.csv(file.path(D, "wind_us.csv"), colClasses = c(county_fips = "character"))
us <- us[us$in_frame == "TRUE" | us$in_frame == TRUE, ]
ga <- read.csv(file.path(D, "wind_ga.csv"), colClasses = c(fips = "character"))
og <- read.csv(file.path(D, "office_gap.csv"))
ff <- read.csv(file.path(D, "facts.csv"))
FV <- setNames(ff$value, ff$key)
F  <- function(k) { z <- FV[[k]]; n <- suppressWarnings(as.numeric(z))
                    if (is.na(n)) z else n }
c20 <- read.csv("../data-sources/data/pres2020_counties.csv",
                colClasses = c(county_fips = "character"))
c24 <- read.csv("../data-sources/data/pres2024_counties.csv",
                colClasses = c(county_fips = "character"))
agg <- function(d, g) 100 * (sum(g) - sum(d)) / (sum(g) + sum(d))
V   <- us$votes_dem_24 + us$votes_gop_24
O   <- order(-V)
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`wind-map-brief.Rmd`) and the lab (`wind-map.Rmd`) run
the same numbered steps — the brief in prose with three D3 figures, the lab with
the code exposed. **Students should read the brief first.** Step numbers below
refer to the lab.

**This session has two halves and the second one has not happened yet.** The
2020-to-2024 map is fully built. The 2026 map is a stub with the 2022 baseline
already in it, waiting on an election that falls on **Tuesday 3 November 2026**,
inside the semester. See "Running the 2026 map live" at the end.

Every number below was run against the built files.

***

# The session in one sentence

A wind map is the most persuasive picture of an election anyone draws, and every
arrow rests on a join that reports its successes and never its meanings — so the
lab draws the map, then audits the join, and finds an arrow pointing the wrong
way in a place where nothing failed.

***

# The findings

```{r headline}
c(national_arrows      = nrow(us),
  median_county_swing  = round(median(us$swing), 2),
  national_vote_swing  = round(agg(us$votes_dem_24, us$votes_gop_24) -
                               agg(us$votes_dem_20, us$votes_gop_20), 2),
  dc_naive_swing       = round(F("naive_dc_swing"), 2),
  dc_true_swing        = round(F("true_dc_swing"), 2),
  rows_either_way      = as.integer(F("naive_matched")),
  georgia_counties     = nrow(ga),
  georgia_state_swing  = round(F("ga_state_swing"), 2),
  office_gap_same_ballot = round(F("og_state_gap"), 2),
  precinct_join_ceiling  = F("pj_best_pct"))
```

- **`r round(F("us_share_R"), 1)`% of counties moved toward Trump**, median
  `r sprintf("%+.2f", F("us_swing_median"))`. The map is unambiguous and that is
  part of the problem: it looks like it cannot be wrong.
- **The District of Columbia's arrow reverses**, from
  `r sprintf("%+.2f", F("naive_dc_swing"))` to
  `r sprintf("%+.2f", F("true_dc_swing"))`, and **both joins return
  `r format(F("naive_matched"), big.mark = ",")` rows.**
- **The counties moved `r sprintf("%+.2f", F("us_swing_median"))`; the country
  moved `r sprintf("%+.2f", F("us_agg_swing"))`.** One arrow per county is an
  editorial position.
- **Two offices on the same Georgia ballots differ by
  `r round(F("og_state_gap"), 2)` points** — as large as Georgia's entire real
  2020-to-2024 shift of `r round(F("ga_state_swing"), 2)`.
- **The precinct-level join stalls at `r F("pj_best_pct")`%**, and
  `r F("pj_worst_county")` County loses all `r F("pj_worst_total")` precincts.

***

# The three things to get said

**1. The row count cannot warn you.** This is the session. Two joins of the same
two files, `r format(F("naive_matched"), big.mark = ",")` rows either way, and
one of them produces an arrow pointing the opposite direction. **Make somebody
say out loud what check would have caught it**, and keep pushing until the answer
is "read the `county_name` column, not the count." Every instinct students have
about data quality — match rate, missingness, row counts — is silent here.

**2. Loud failures are the safe ones.** Alaska and Connecticut break the join and
the rows fall out; you can count what you lost and say so. The District of
Columbia does not break anything. **The ranking students will get backwards is
that the loud failure is worse.** Ask which of the three they would rather have
in a dataset they were about to publish from.

**3. One arrow per county is a choice about whose movement is visible.** The
counties moved `r sprintf("%+.2f", F("us_swing_median"))` and the country moved
`r sprintf("%+.2f", F("us_agg_swing"))`. Nothing is miscomputed. The map answers
"what did the typical county do", which almost nobody reading it thinks is the
question. **This is not a caveat; it is what the figure is.**

**The line for the board:** a join reports whether it matched, never whether it
meant anything.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** The encoding | Angle and length, declared as constants | 6 | Have someone read an arrow back off the legend. If they cannot, nothing later lands. |
| **2** The national map | 3,100 arrows, 88.5% pointing right | 7 | Let them enjoy it first. The audit is more effective after they have believed it. |
| **3** Who published this | No federal county-level source exists | 6 | **Most students assume the government publishes this.** It does not. Say the FEC line explicitly. |
| **4** Audit the units | Alaska, DC, Connecticut | 9 | Only Connecticut is a real unit change. Do not let the CT story swallow the other two. |
| **5** The silent failure | DC's arrow reverses; row counts identical | **12** | **The pivot.** Take a prediction before revealing the harmonised number. |
| **6** What harmonising cost | Two of three repairs are deletions | 4 | A counted hole beats an invented arrow. |
| **7** One arrow per county | 3.06 vs 6.04 | 9 | Not a caveat. The figure *is* the choice. |
| **8** Georgia, officially sourced | Same encoding, defensible provenance | 6 | The only difference from Step 2 is the publisher. Say so plainly. |
| **9** Why not precincts | The join stalls at 90.9% | 6 | Chatham loses **all** its precincts. The hole would look like quiet. |
| **10** The office problem | 2.08 points, same ballots | 9 | This is what licenses using 2022 rather than 2024 as the 2026 baseline. |
| **11** The 2026 slot | 159 tails, no heads | 6 | Show the stub file on screen. Two empty columns is the whole design. |

**Step 5 deserves the prediction.** Put the two DC rows on the screen — same
FIPS, "District of Columbia" and "Ward 1" — and ask: *the join will match this
row. Will the resulting arrow be roughly right?* Most of the room says yes,
reasoning that DC's wards are all overwhelmingly Democratic so the margin will be
similar. **They are almost right, and that is the trap**: the naive margin is
`r round(F("naive_dc_margin_24"), 1)` against a true
`r round(F("true_dc_margin_24"), 1)`, close enough to look fine, and the *swing*
built from it points the wrong way.

**If you are short on time**, protect Steps 5, 7 and 10. The silent join, the
editorial weighting, and the measured office effect are the argument. Steps 1
and 2 can be compressed to a demonstration.

***

# The mistake this lab made, and how to teach it

**Tell this, because it happened while building the lab.**

The first plan was a Georgia **precinct** map — seventeen times the resolution,
with a companion lab having already solved the boundary problem by crosswalking
2020's votes onto 2024's precinct lines. All that remained was attaching the 2024
returns by precinct name.

```{r ladder}
read.csv(file.path(D, "precinct_join_ladder.csv"))
```

Each rung is a reasonable repair. Together they reach `r F("pj_best_pct")`% and
stop. And the residual is not scattered:

```{r resid}
head(read.csv(file.path(D, "precinct_join_residual.csv")), 5)
```

**`r F("pj_worst_county")` County — Savannah — loses every one of its
`r F("pj_worst_total")` precincts**, because the state's returns identify
precincts by codes like `1-05C` and the boundary file identifies them by the name
of the polling place. There is no regex for that.

The pedagogy: **ask what they would have tried next.** Fuzzy matching and a
hand-built alias table are the usual answers, and both are more work in service
of a map that would have had an invisible hole where a majority-Black city is.
The decision was to drop to county level, where nothing is lost. **Resolution you
cannot trust everywhere is not resolution**, and this is the same lesson
`precinct-geography` reaches from the other direction.

***

# Handling the room

The figure is genuinely beautiful and that is a teaching hazard.

- **Let them like it.** Five minutes of just looking at Step 2 buys the audit.
- **The audit is not debunking.** The map is basically right. The point is that
  nothing in the process would have told you if it were not.
- **Nobody writes code.** Say so at the top; the D3 in the brief makes some
  students assume otherwise.
- **The 2026 half makes this the one session with a sequel.** Flag on day one
  that they will redraw this map with data that does not exist yet.

***

# Option A — Weight the map by votes

## Verified output

```{r a}
big <- O[1:100]; sml <- O[(length(O) - 999):length(O)]
data.frame(
  group = c("largest 100 counties", "smallest 1,000 counties"),
  share_of_votes = round(c(100*sum(V[big])/sum(V), 100*sum(V[sml])/sum(V)), 1),
  median_swing = round(c(median(us$swing[big]), median(us$swing[sml])), 2),
  arrows_on_the_map = c(100, 1000))
round(agg(us$votes_dem_24[big], us$votes_gop_24[big]) -
      agg(us$votes_dem_20[big], us$votes_gop_20[big]), 2)
```

**3/4:** Reports that big and small counties moved differently and that the map
gives them equal weight.

**4/4:** Puts numbers on it — the largest 100 counties cast about 40% of the vote
and swung a median of +5.6, the smallest 1,000 cast 2.3% and swung +3.1, and the
map draws ten times as many arrows for the second group. Concludes that the
national swing of +6.0 is invisible on a map whose median arrow is +3.1.

**4/4, best:** Notices that **there is no weighting that makes the map correct**,
because a wind map's job is to show geography and a vote-weighted map is a
cartogram that has stopped being a map. Says which question each version answers
and picks one on purpose. A student who proposes drawing both and labelling them
has understood the figure better than the figure does.

***

# Option B — Break the harmonisation on purpose

## Verified output

```{r b}
t20 <- table(substr(c20$county_fips, 1, 2))
t24 <- table(substr(c24$county_fips, 1, 2))
k <- union(names(t20), names(t24))
d <- data.frame(state_fips = k, rows_2020 = as.integer(t20[k]),
                rows_2024 = as.integer(t24[k]))
d[!is.na(d$rows_2020) & !is.na(d$rows_2024) & d$rows_2020 != d$rows_2024, ]
# states whose row COUNT is unchanged but whose codes all changed
same <- d$state_fips[d$rows_2020 == d$rows_2024]
z <- sapply(same, function(s)
  length(setdiff(c20$county_fips[substr(c20$county_fips,1,2) == s],
                 c24$county_fips[substr(c24$county_fips,1,2) == s])))
z[z > 0]
```

**3/4:** Finds the two states whose row counts differ — Connecticut and the
District of Columbia — and describes them correctly.

**4/4:** Notices that **Alaska has exactly 40 rows in both years and every single
code is different**, so a row-count diff misses it completely. States the general
rule: comparing counts detects changes in cardinality, not changes in identity.

**4/4, best:** Turns it round and asks what a state that changed its units
*without* changing its count would look like — which is Alaska — and then points
out that a state which renamed its units while keeping both count and codes would
be invisible to every check in this lab. **Recognising that the audit has a floor
is the most sophisticated response available.**

***

# Option C — Audit Georgia against the national file

## Verified output

```{r c}
n24 <- c24[substr(c24$county_fips, 1, 2) == "13", ]
n20 <- c20[substr(c20$county_fips, 1, 2) == "13", ]
m <- merge(ga[, c("fips","county","dem_20","gop_20","dem_24","gop_24","swing")],
           n24[, c("county_fips","votes_dem","votes_gop")],
           by.x = "fips", by.y = "county_fips")
m <- merge(m, n20[, c("county_fips","votes_dem","votes_gop")],
           by.x = "fips", by.y = "county_fips", suffixes = c("_n24", "_n20"))
mg <- function(d, g) 100 * (g - d) / (g + d)
m$nat_swing <- (mg(m$votes_dem_n24, m$votes_gop_n24) -
                mg(m$votes_dem_n20, m$votes_gop_n20))
c(counties            = nrow(m),
  gop_2024_difference = sum(m$votes_gop_n24) - sum(m$gop_24),
  dem_2024_difference = sum(m$votes_dem_n24) - sum(m$dem_24),
  dem_2020_difference = sum(m$votes_dem_n20) - sum(m$dem_20),
  max_swing_gap       = round(max(abs(m$nat_swing - m$swing)), 3),
  median_swing_gap    = round(median(abs(m$nat_swing - m$swing)), 4),
  arrows_that_flip    = sum(sign(m$nat_swing) != sign(m$swing)))
```

**3/4:** Reports that the two sources agree closely and concludes the national
file is fine.

**4/4:** Quantifies it — totals differ by a handful of votes out of five million,
the median county's swing differs by six thousandths of a point, and **no arrow
changes direction** — and then draws the right conclusion, which is *not* that
the national file is fine. It is that the file is **accurate here**, and that
this could only be established because Georgia happens to publish a check.

**4/4, best:** States the actual problem: **for 49 other states there is no
comparable check available at this cost**, so the correct summary is "verified in
one state, unverifiable in the rest" rather than "verified." A student who
observes that a single-state audit passing is weak evidence about the other
fifty — and that the states most likely to be scraped badly are the ones least
likely to publish conveniently — has the argument.

***

# Option D — Argue with the encoding

## Verified output

```{r d}
ga$turnout_change <- round(100 * (ga$tot_24 - ga$tot_20) / ga$tot_20, 2)
c(state_total_2020 = sum(ga$tot_20),
  state_total_2024 = sum(ga$tot_24),
  state_change_pct = round(100*(sum(ga$tot_24)-sum(ga$tot_20))/sum(ga$tot_20), 2),
  median_county    = round(median(ga$turnout_change), 2),
  counties_losing_votes = sum(ga$turnout_change < 0),
  correlation_with_margin_swing = round(cor(ga$turnout_change, ga$swing), 3))
ga[c(which.min(ga$turnout_change), which.max(ga$turnout_change)),
   c("county", "turnout_change", "swing")]
```

**3/4:** Builds the turnout map and observes that it looks nothing like the
margin map.

**4/4:** Notes the negative correlation of about -0.44 — counties where turnout
grew swung *less* toward Trump — and says what that implies: the margin map and
the turnout map are telling two halves of one story and neither is complete
alone.

**4/4, best:** Confronts the identification problem head on. **A margin swing
cannot distinguish persuasion from differential turnout**, and the two have
completely different implications for what a campaign should do next. Says that a
front page needs the margin map and a campaign needs both, and that the wind map
is popular partly *because* it collapses a distinction its readers do not know
is there.

***

# Option E — Your own question

No verified output. Grade the question as much as the answer. **A 4 requires the
question to be answerable from the files given and the student to say what would
have changed their mind.**

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given.
- **The move that earns a 4** is treating an encoding or a join as a decision
  with consequences that can be measured, and then measuring it.
- **A 2** is "the data has problems so the map is unreliable" — true, useless,
  and precisely what the lab is built to move past. The map is mostly *right*;
  the point is that nothing in the process would have said so.
- **Also a 2**: reporting the national numbers as though a government published
  them.
- **Reward anyone who asks for an error bar on an arrow.** There isn't one, there
  is no agreed way to draw one, and noticing the absence is the most
  sophisticated available response.

## Anticipated questions

- *"Why is Alaska missing?"* — Because Alaska does not report presidential votes
  by county, both files invent their own codes for State House Districts, and the
  districts were redrawn in between. Three separate reasons, any one of which is
  sufficient.
- *"Couldn't we just fix Connecticut with a crosswalk?"* — In principle, with
  town-level returns, which Connecticut does publish. It is a real project, not a
  line of code, and it is a good final-paper topic.
- *"Isn't the map basically right?"* — Yes. Say so. Then ask how they know, and
  watch the answer turn out to be Option C, which covers one state.
- *"Why not use the FEC data?"* — Because the FEC publishes by state and
  congressional district. There is no county sheet. This surprises people every
  time.
- *"Who decides that up means no change?"* — Whoever draws it. There is no
  standard. Some published wind maps put no-change at horizontal instead, which
  changes what the map looks like without changing a number.

## Connections

- **`data-sources`** — the same two county files, and the lab that establishes
  their provenance. **Run that one first if you can**; this lab depends on
  students already distrusting the file.
- **`precinct-geography`** — boundaries moving between elections. This lab is the
  other failure mode: the boundary held and the *label* moved.
- **`areal-units`** — the unit changes the answer. Step 7 is that lab's point
  rendered as a picture.
- **`migration`** — the other arrow map in the corpus, and the reason the length
  scale here is a declared constant with a printed legend.
- **`regional-shift`** — the aggregate version of the same question, and a useful
  contrast: it cannot be wrong in this way because it has no join.

***

# Running the 2026 map live

## What to have ready before class

```{r stub}
st <- read.csv(file.path(D, "wind_input_2026.csv"),
               colClasses = c(unit_id = "character"))
c(rows = nrow(st), filled = sum(!is.na(st$votes_dem_to)))
names(st)
```

- **The baseline is already built.** `wind_ga_counties.csv` holds all 159
  Georgia counties with their 2022 governor margins;
  `r F("ga22_election")`, snapshot `r F("ga22_snapshot")`.
- **One file changes**: `data/wind_input_2026.csv`, two empty columns.
- **One command runs**:

  ```
  Rscript -e 'source("data/build-data.R"); build_2026()'
  ```

  It fetches `export-2026NovGen.json` from `results.sos.ga.gov`, fills the two
  columns, recomputes every swing, and rewrites the baseline file. Re-knit and
  the map appears.

## The office decision, and why it is not negotiable

The class map is **2022 governor against 2026 governor**, not 2024 president
against 2026 governor. Step 10 measures why: two offices on the *same Georgia
ballots* in November 2020 differed by `r round(F("og_state_gap"), 2)` points,
which is the size of Georgia's entire real shift between two presidential
elections — and that is with the electorate held exactly constant. A
presidential-to-midterm arrow adds a turnout change of roughly a third on top of
a floor that already exceeds the signal.

**If a student proposes the 2024 comparison, do not just refuse it.** Have them
compute what it would show, then put the two side by side. The version of this
lesson that sticks is the one where they draw the bad arrow themselves.

## What will break

| Failure | Likelihood | What to do |
|---|---|---|
| `export-2026NovGen.json` 404s | **Moderate.** The URL pattern is inferred from 2022 and 2024, both of which return HTTP 200. The state may rename it. | Look at <https://results.sos.ga.gov/> in a browser, find the export link, pass it: `build_2026(url = "...")`. |
| The export exists but has no `localResults` | Low | 2022 and 2024 both have it. If not, county totals are on the results page; fill the two columns by hand and call `build_2026(from_stub = TRUE)`. |
| The contest is not called `Governor` | Moderate | 2022 called it exactly `Governor`. Pass `office = "^Governor"` or whatever the export uses. |
| Counties still reporting | **Certain, if you run this early** | `build_2026()` prints how many of 159 counties have a result. Draw it anyway and label it. |
| Nothing is posted at all | Low by December | Fall back to the 2022 baseline figure in Step 11 and run the session as a design exercise: *what would you need to see before you drew the arrow?* That is a legitimate 80 minutes. |

## If the data is not out

**Do not cancel the step.** Run it as a pre-registration exercise: have the room
predict the 2026 map before it exists — direction of the statewide arrow, which
counties move most, whether metro Atlanta and rural Georgia move together — and
write the predictions down. Then redraw it when the data lands, in office hours
or by email. **A prediction recorded before the data arrives is worth more than
the map**, and it is the only exercise in this course that can do that, because
it is the only one whose data has not happened yet.

## A note on the figures

The interactive (D3) and printed (base-R) versions of every map read the same
`x`, `y`, `dx`, `dy` columns out of the same CSV. Neither one projects or
rescales anything, so they cannot drift. Checked mechanically against the
knitted HTML: the two renderers place arrow tips within **1.0 pixel** on the
national map (the whole-pixel rounding floor of the D3 payload) and within
**0.1 pixel** on the Georgia map, on figures 780 and 720 pixels wide.

```{r parity}
read.csv(file.path(D, "parity.csv"))
```

## The provisional label

Whatever gets drawn in early December will be **unofficial**. Certification runs
into December in many states, close races will still be open, and precinct-level
returns will not exist. Put that on the figure, in words, on the slide. **Every
wind map published on election night has this property and almost none of them
say so**, which is itself the last thing this lab has to teach.
