---
title: "Which Way Did It Blow"
subtitle: "84-355 Democracy's Data | The election wind map, and what an arrow promises"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2,
                      fig.height = 4.4, dpi = 96, fig.retina = 1)
options(scipen = 999)
D <- "data"
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `wind-map-brief.Rmd` first.** This lab runs the same steps with the
> code exposed.

***

# The claim an arrow makes

> *"Miami-Dade swung eleven points."*

Every wind map is thousands of sentences like that at once. Each one says:
**this is one place, I observed it twice, and here is the difference.**

That is three claims, and any of them can be false while the arrow still gets
drawn.

# Step 1 — The encoding, fixed once

```{r encoding}
ff <- read.csv(file.path(D, "facts.csv"))
FV <- setNames(ff$value, ff$key)
F  <- function(k) { z <- FV[[k]]; n <- suppressWarnings(as.numeric(z))
                    if (is.na(n)) z else n }

data.frame(
  channel = c("direction", "direction", "length", "length", "thickness"),
  rule = c("straight up = no change",
           paste0(F("deg_per_point"), " degrees clockwise per point toward the Republican"),
           paste0(F("km_per_point_us"), " km per point (national map)"),
           paste0("stops growing past ", F("len_cap_pts"), " points"),
           "square root of votes cast"))
```

Direction and length carry the same number twice. That redundancy is the point:
the map survives being printed in black and white.

**The scale is a declared constant, not something fitted to the data on the
page.** A map that rescales its arrows to fill its own frame is a picture of
the frame.

# Step 2 — The national map, and what it is made of

```{r national}
us <- read.csv(file.path(D, "wind_us.csv"),
               colClasses = c(county_fips = "character"))   # leading zeros!
us <- us[us$in_frame == "TRUE" | us$in_frame == TRUE, ]
c(counties        = nrow(us),
  median_swing    = round(median(us$swing), 2),
  share_toward_R  = round(100 * mean(us$swing > 0), 1),
  largest_R_shift = round(max(us$swing), 1),
  largest_D_shift = round(min(us$swing), 1))
```

**Read `county_fips` as character.** Read as a number, `01001` becomes `1001`
and Alabama's counties join to nothing. A sibling lab lost 1,549 of 4,489
identifiers to exactly this.

# Step 3 — Who published this?

There is **no federal source for county-level presidential returns.** The
Federal Election Commission publishes official results by state and by
congressional district. County returns are published by 51 separate state
election offices in 51 formats.

So the national file used above is somebody's compilation. Its own README says
the 2024 figures were scraped from a television network's results pages and
that the numbers **"are not authoritative."**

That does not make it useless. It makes it a file you have to audit.

# Step 4 — Audit the units

```{r audit}
c20 <- read.csv("../data-sources/data/pres2020_counties.csv",
                colClasses = c(county_fips = "character"))
c24 <- read.csv("../data-sources/data/pres2024_counties.csv",
                colClasses = c(county_fips = "character"))
c(rows_2020   = nrow(c20),
  rows_2024   = nrow(c24),
  in_both     = length(intersect(c20$county_fips, c24$county_fips)),
  only_2020   = length(setdiff(c20$county_fips, c24$county_fips)),
  only_2024   = length(setdiff(c24$county_fips, c20$county_fips)))
```

Eight more rows in 2024 than in 2020. **The country did not gain eight
counties.** Take the non-matching codes apart:

```{r audit2}
aud <- read.csv(file.path(D, "provenance_audit.csv"),
                colClasses = c(fips = "character"))
table(sub("^([A-Za-z ]+):.*", "\\1", aud$reason), aud$appears_in)
```

Three different things, only one of which is a change on the ground:

- **Alaska** does not report presidential votes by county at all. Both files
  carry 40 State House Districts under **different made-up FIPS codes** —
  `r F("ak_2020_range")` in one file, `r F("ak_2024_range")` in the other. Same
  kind of unit, different arbitrary labels. And Alaska redrew those districts
  after the 2020 census, so matching by district number would compare
  different ground anyway.
- **The District of Columbia** is one row in 2020 and eight wards in 2024.
  Nothing moved; somebody reported it finer.
- **Connecticut** is the real unit change. The Census Bureau retired its eight
  counties and adopted **nine planning regions** as county equivalents, with
  new codes, effective for 2022 and after.

# Step 5 — The failure that does not show up in a set difference

Alaska and Connecticut fail loudly: the codes do not match and the rows drop
out. Watch what the District of Columbia does.

```{r dc}
rbind(c20[c20$county_fips == "11001", 1:6],
      c24[c24$county_fips == "11001", 1:6])
```

**Same FIPS. Different place.** The join succeeds and an arrow gets drawn
comparing a whole city to one eighth of itself.

```{r dc2}
mg <- function(d, g) 100 * (g - d) / (g + d)
naive <- mg(c24$votes_dem[c24$county_fips == "11001"],
            c24$votes_gop[c24$county_fips == "11001"]) -
         mg(c20$votes_dem[c20$county_fips == "11001"],
            c20$votes_gop[c20$county_fips == "11001"])
dc24 <- c24[substr(c24$county_fips, 1, 2) == "11", ]
true  <- mg(sum(dc24$votes_dem), sum(dc24$votes_gop)) -
         mg(c20$votes_dem[c20$county_fips == "11001"],
            c20$votes_gop[c20$county_fips == "11001"])
c(naive_swing = round(naive, 2), true_swing = round(true, 2),
  error = round(true - naive, 2))
```

**The arrow reverses.** Naively the District moved toward the Democrats; summed
properly it moved toward the Republicans, by more than the median county moved
at all.

And both joins return the same number of rows:

```{r dc3}
c(naive_join_rows = nrow(merge(c20, c24, by = "county_fips")),
  harmonised_rows = as.integer(F("matched_after_harmonise")))
```

> **The row count cannot warn you.** Nothing about a successful join tells you
> whether the two sides mean the same thing.

# Step 6 — What harmonising cost

```{r harm}
data.frame(
  state = c("Alaska", "District of Columbia", "Connecticut"),
  action = c("dropped all 40 units",
             "eight wards summed back to one District",
             "dropped"),
  cost = c("no presidential arrows for Alaska",
           "none; this is a repair",
           paste0(format(F("ct_votes_lost"), big.mark = ","), " votes")))
```

Two of the three repairs are deletions. **A wind map cannot draw an arrow for a
place that did not survive**, and a hole you have counted is better than an
arrow you have invented.

# Step 7 — One arrow per county, or one per voter?

```{r weight}
v <- us$votes_dem_24 + us$votes_gop_24
agg <- function(d, g) 100 * (sum(g) - sum(d)) / (sum(g) + sum(d))
c(median_county_swing = round(median(us$swing), 2),
  mean_county_swing   = round(mean(us$swing), 2),
  national_vote_swing = round(agg(us$votes_dem_24, us$votes_gop_24) -
                              agg(us$votes_dem_20, us$votes_gop_20), 2),
  largest_county      = max(v), smallest_county = min(v))
```

**The counties moved about three points. The country moved about six.** Both
are correct and they answer different questions.

One arrow per county gives the smallest county in the file the same visual
weight as Los Angeles, which cast roughly
`r format(F("us_size_ratio"), big.mark = ",")` times as many votes. Thickness
compensates a little. It does not compensate much.

**That is an editorial choice, not a technical one**, and this lab has made it.

# Step 8 — Georgia, from the state's own record

```{r ga}
ga <- read.csv(file.path(D, "wind_ga.csv"), colClasses = c(fips = "character"))
c(counties       = nrow(ga),
  margin_2020    = round(F("ga_state_margin_20"), 2),
  margin_2024    = round(F("ga_state_margin_24"), 2),
  statewide_swing= round(F("ga_state_swing"), 2),
  median_county  = round(median(ga$swing), 2),
  share_toward_R = round(100 * mean(ga$swing > 0), 1))
```

Both years are the **Georgia Secretary of State's own record** — the 2020
general election archive and the state's 2024 results export. Georgia has had
159 counties since 1945, so nothing was recoded, nothing was renamed, and no
arrow here can be fictional.

Same encoding, different provenance. That is the only difference between this
map and the national one, and it is the whole difference.

# Step 9 — Why not precincts?

Georgia has 2,701 precincts — seventeen times the resolution. A companion lab
has already carried the 2020 votes onto the 2024 precinct boundaries, so the
boundary problem is solved. All that remains is attaching the 2024 returns by
precinct **name**.

```{r ladder}
read.csv(file.path(D, "precinct_join_ladder.csv"))
```

```{r resid}
head(read.csv(file.path(D, "precinct_join_residual.csv")), 6)
```

The ladder stalls at `r F("pj_best_pct")`%, and the loss is not scattered:
**`r F("pj_worst_county")` County loses all `r F("pj_worst_total")` of its
precincts**, because the returns identify precincts by codes like `1-05C` and
the boundary file identifies them by the name of the polling place.

A precinct map would have a hole where Savannah is, and **the hole would look
like a quiet neighbourhood.** That is why Step 8 is drawn at county level.

# Step 10 — The 2026 arrow, and the office problem

The midterm is **Tuesday, 3 November 2026**, during this semester. The plan is
to draw the same map in class once returns exist.

A 2024-to-2026 arrow would compare a presidential race to a midterm: different
office, different electorate, different turnout. How much does that matter?

Georgia's November 2020 ballot carried a presidential race **and** a US Senate
race. Same voters, same day, same piece of paper. Any disagreement is a pure
office effect with **zero** turnout difference.

```{r office}
og <- read.csv(file.path(D, "office_gap.csv"))
c(precincts          = nrow(og),
  statewide_pres     = round(F("og_state_pres"), 2),
  statewide_senate   = round(F("og_state_sen"), 2),
  office_gap         = round(F("og_state_gap"), 2),
  median_precinct_gap= round(median(abs(og$gap)), 2),
  pct_over_2_points  = round(100 * mean(abs(og$gap) > 2), 1),
  georgia_2020_2024_swing = round(F("ga_state_swing"), 2))
```

**The office effect alone is about two points — the same size as the entire
real shift Georgia experienced across two presidential elections.** And that is
the best case, with the electorate held exactly fixed.

**So this lab does not draw a 2024-to-2026 arrow.** The class map is
**Georgia's 2022 governor's race against its 2026 governor's race**: same
office, both midterms, same 159 counties, same publisher.

# Step 11 — The half that is already built

```{r base}
g22 <- read.csv(file.path(D, "wind_ga_counties.csv"),
                colClasses = c(fips = "character"))
c(counties    = nrow(g22),
  dem_2022    = F("ga22_dem"),
  gop_2022    = F("ga22_gop"),
  margin_2022 = round(F("ga22_margin"), 2),
  counties_with_a_2026_result = sum(!is.na(g22$margin_to)))
```

`r F("ga22_election")`, snapshot `r F("ga22_snapshot")`, straight from
`results.sos.ga.gov`. Every county has a tail. None has a head yet.

## What has to happen on class day

**One file changes.** `data/wind_input_2026.csv` has ten columns and 159 rows;
two of the columns are empty:

```{r stub}
str(read.csv(file.path(D, "wind_input_2026.csv"),
             colClasses = c(unit_id = "character")))
```

| Column | Meaning | Rule |
|---|---|---|
| `unit_id` | 5-digit county FIPS | **character**, leading zeros kept |
| `unit_name` | county name | must match the baseline |
| `office` | the office compared | **identical in both years, or stop** |
| `year_from`, `year_to` | 2022, 2026 | both midterms |
| `votes_dem_from`, `votes_gop_from` | 2022 governor | already filled |
| `votes_dem_to`, `votes_gop_to` | 2026 governor | **the two empty columns** |

The refresh is one call:

```
Rscript -e 'source("data/build-data.R"); build_2026()'
```

which fetches `export-2026NovGen.json` from the Secretary of State, fills the
two columns, recomputes every swing, and rewrites `wind_ga_counties.csv`. If
the state has not posted the export, edit the two columns by hand from any
source and call `build_2026(from_stub = TRUE)`.

## What will not be knowable

A month after 3 November 2026, expect **unofficial** returns. Certification
deadlines run into December in many states; close races will still be open;
precinct-level returns rarely appear this fast. **The map drawn in class will
be provisional and should be labelled so.** Every wind map published on
election night has this property and most do not admit it.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Weight the map by votes.** Step 7 shows the counties moving about
three points while the country moved about six. Redraw the national map with
arrows sized by votes cast instead of one per county — or simply report the
swing separately for the largest 100 counties and the smallest 1,000. Which
map is the honest one, and honest about what?

**Option B — Break the harmonisation on purpose.** Re-run Step 5 for every
state, not just the District of Columbia: for each state, compare the number of
rows in the two files. Which other states change their reporting grain between
elections, and would you have caught them without being told to look?

**Option C — Audit Georgia against the national file.** The national file has
Georgia counties too, from a press compilation. Compare its Georgia numbers to
the Secretary of State's in `wind_ga.csv`, county by county. Where do they
disagree, by how much, and does the disagreement change any arrow's direction?

**Option D — Argue with the encoding.** The map draws margin change and
nothing else. Build the turnout-change map instead, using `tot_20` and
`tot_24` in `wind_ga.csv`, and say which of the two you would put on a front
page.

**Option E — Your own question.** `r format(nrow(us), big.mark = ",")` national
arrows, 159 Georgia arrows, and one that has not been drawn yet.

# What this lab cannot tell you

- **Whether the national numbers are right.** No federal publisher exists and
  the compilation says of itself that it is not authoritative. Georgia's are
  the state's own record; the rest are not.
- **Anything about Alaska.** Two files, two arbitrary coding schemes, and
  redrawn districts in between.
- **What happened in Connecticut.** Its eight counties became nine planning
  regions and those votes are simply absent from the national map.
- **Whether a swing was caused by persuasion or by turnout.** Margin change
  cannot separate them. A county where nobody changed their mind and a third of
  one side stayed home gets the same arrow as a county that was persuaded.
- **How uncertain an arrow is.** There is no error bar on a wind map and no
  agreed way to draw one.

# On the discussion board

By **Friday, 11:59 p.m.**

- Step 5 shows a join that succeeded, reported the same row count as the
  correct one, and produced an arrow pointing the wrong way. **What check would
  have caught it?** Be specific enough that someone could run it.
- Step 7 says one arrow per county is an editorial choice. Whose interests does
  that choice serve, and is the alternative any less of a choice?
- Step 10 measures a two-point gap between two offices on the *same ballots*.
  Given that, what would you need before you believed any cross-year swing
  number smaller than two points?

***

### Where the data came from

- **Georgia returns, all three elections** — Georgia Secretary of State.
  2020 general election archive at
  <https://sos.ga.gov/page/historical-elections-results>; 2022 and 2024 results
  exports at <https://results.sos.ga.gov/>, e.g.
  `https://results.sos.ga.gov/cdn/results/Georgia/export-2022NovGen.json`
  (fetched 2026-08-10, HTTP 200).
- **County population centroids** — U.S. Census Bureau, *Centers of Population
  by County: 2020*,
  <https://www2.census.gov/geo/docs/reference/cenpop2020/county/CenPop2020_Mean_CO.txt>
  (fetched 2026-08-10, HTTP 200, 3,221 rows).
- **National county presidential returns** — Tony McGovern, *US County Level
  Election Results 08-24*, as committed in `../data-sources/`.
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>
  **Scraped from news organisations; not authoritative, by its own account.**
- **Official federal returns** — Federal Election Commission,
  <https://www.fec.gov/introduction-campaign-finance/election-results-and-voting-information/>
  Published by state and congressional district, **not by county**, which is
  why this lab needs the compilation above.
- **The wind-map idiom** — Observable, *Plot: election wind map*.
  <https://observablehq.com/@observablehq/plot-election-wind-map>

The build is `data/build-data.R`. It records every URL, the fetch date and the
row counts, and it contains `build_2026()`, the one function that has to run in
November.
