---
title: "Data Session 11: Did the Machinery Work?"
subtitle: "84-355 Democracy's Data | The Election Administration and Voting Survey"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"Thousands of mail ballots were rejected."*

Every session so far has used the **output** of an election — who won, who voted,
who gave money. None of them asked whether the machinery worked.

Somebody does ask. Every two years the **Election Assistance Commission** sends
a survey to every election office in the country and asks: how many ballots did
you mail out, how many came back, how many did you reject and why, how many
provisional ballots did you issue, how many precincts did you run, how many poll
workers did you find.

It is the closest thing America has to a service record for its own elections,
and it is free. Today we open it.

# Part 1 — What one row is

```{r load}
nat <- read.csv("data/national.csv", stringsAsFactors = FALSE)
st  <- read.csv("data/states.csv",   stringsAsFactors = FALSE)
u   <- read.csv("data/units.csv",    stringsAsFactors = FALSE)

c(jurisdictions = sum(st$jurisdictions), states_and_territories = nrow(st))
```

**6,461 rows. One per election jurisdiction.** And that word is the first
problem.

```{r units}
rbind(head(u[, c("state","jurisdictions","voters","voters_per_jurisdiction")], 4),
      tail(u[, c("state","jurisdictions","voters","voters_per_jurisdiction")], 4))
```

**Wisconsin reports 1,851 jurisdictions. Texas reports 254.** Wisconsin does not
have seven times as much election administration as Texas — it runs elections
through **townships**, and Texas runs them through **counties**. New England is
the same: Vermont's typical jurisdiction has about 1,500 voters; California's has
about 279,000.

So a sentence like "the average jurisdiction rejected 90 ballots" is close to
meaningless, because the average is taken over units that are not the same kind
of thing. **Anything you compute per jurisdiction, across states, is wrong.**
Totals are fine. Rates are fine. Averages over rows are not.

# Part 2 — The national picture

```{r national}
nat[, c("item", "total", "reported_by", "pct_reporting")]
```

**158 million voters, 177,708 precincts, 66.9 million mail ballots sent out.**

Now look at the last two columns, which are the ones nobody quotes. **Not every
jurisdiction answers every question.** Total voters is reported by 99.8% of them;
provisional ballots rejected, by **79.9%**.

That means the 436,258 figure is a sum over four-fifths of the country, and the
missing fifth is not random — small offices with less staff are likelier to skip
items. **Every national total on this page is a floor, not a count.**

```{r missing-codes}
# Missing is coded -99 (not reported) and -88 (not applicable), not blank.
# Summing the raw column would quietly mix real counts with these codes.
nat[nat$variable %in% c("E1a", "E1d"), c("item", "total", "pct_reporting")]
```

# Part 3 — Rejected mail ballots

```{r mail}
sent <- nat$total[nat$variable == "C1a"]
ret  <- nat$total[nat$variable == "C1b"]
rej  <- nat$total[nat$variable == "C9a"]

c(transmitted = sent, returned = ret, never_returned = sent - ret,
  rejected = rej, pct_of_returned = round(100 * rej / ret, 2))
```

**584,463 mail ballots were rejected — 1.22% of those returned.** That sounds
small, and nationally it is. Now do it by state.

```{r by-state}
st$rate <- round(100 * st$mail_rejected / st$mail_returned, 2)
big <- st[st$mail_returned > 100000, ]
rbind(head(big[order(-big$rate), c("state","mail_returned","mail_rejected","rate")], 5),
      head(big[order( big$rate), c("state","mail_returned","mail_rejected","rate")], 4))
```

```{r state-plot}
b <- big[order(-big$rate), ]
barplot(b$rate, names.arg = b$state, las = 2, cex.names = 0.55,
        col = ifelse(b$rate > 2, "#C41230", "#2c7fb8"),
        ylab = "% of returned mail ballots rejected")
abline(h = 100 * rej / ret, lty = 2)
```

**Oklahoma rejects 4.35% of returned mail ballots. Vermont rejects 0.20%.**
Among states handling more than 100,000 mail ballots, that is a **twenty-two-fold
difference** in the chance your ballot is thrown out.

That is not voters behaving differently in Oklahoma. It is **rules** — witness
requirements, signature matching, postmark deadlines, and whether the state lets
you fix a problem after the fact. The same act, performed identically, is
counted in one state and not in another.

Now the part of that ranking nobody checks. Every rate in it is *rejected ÷
returned*: two separate survey items, each with its own response rate, neither
answered by the whole country.

> **Before you run the next chunk, answer yes or no.** A missing denominator
> sounds as though it should behave the way missing data usually does — scatter
> the rates about a bit, some too high, some too low, no particular direction.
> **Does it?** Write down yes or no, and if no, write down which way it pushes.

```{r denominator}
nat[nat$variable %in% c("C9a", "C1b"),
    c("item", "total", "reported_by", "pct_reporting")]

# States reporting mail ballots sent, but zero returned
hole <- st[st$mail_sent > 0 & st$mail_returned == 0,
           c("state", "voters", "mail_sent", "mail_returned", "mail_rejected")]
hole
```

> **No. It pushes one way, and then it takes states out of the table
> altogether.**
>
> The denominator is measured worse than the numerator. The rejection count
> comes from `r nat$pct_reporting[nat$variable == "C9a"]`% of jurisdictions, the
> returned count from only `r nat$pct_reporting[nat$variable == "C1b"]`%.
> Wherever an office reports the one and not the other, the denominator is short
> and the rate comes out **too high** — not noisily, always upward.
>
> At state level it stops being subtle. Some `r nrow(hole)` states and
> territories, with `r format(sum(hole$voters), big.mark = ",")` registered
> voters between them, report sending mail ballots while reporting that **zero**
> came back — and `r sum(hole$mail_rejected > 0)` of them go on to report
> rejecting some of the none that returned. Alabama sent
> `r format(hole$mail_sent[hole$state == "AL"], big.mark = ",")` and rejected
> `r format(hole$mail_rejected[hole$state == "AL"], big.mark = ",")`.
>
> You cannot compute a rate from that, so Alabama is not at the top of the
> ranking above and not at the bottom. **It is not in the ranking.** A zero and
> a missing value look identical once they are in a spreadsheet, and only
> arithmetic that cannot work reveals the difference. "Oklahoma is
> `r round(max(big$rate) / min(big$rate))` times worse than Vermont" is a
> statement about the `r nrow(big)` states that answered both questions.
> **Non-response is not noise around the estimate; it decides who is eligible to
> be compared at all** — and the offices likeliest to skip an item are the small,
> understaffed ones.

# Part 4 — Why they were rejected, and who says

```{r reasons}
r <- read.csv("data/reasons.csv", stringsAsFactors = FALSE)
head(r[, c("reason", "ballots", "pct_of_rejected", "reported_by")], 8)
```

**The single largest reason is a signature that did not match: 197,979 ballots,
34% of every mail ballot rejected in the country.** A human being compared your
handwriting to a specimen on file and decided it was not you.

Second is arriving late — 103,441. Third, at 82,366, is **"Other."**

```{r reasons-plot}
top <- head(r, 6)
barplot(rev(top$ballots), horiz = TRUE, names.arg = rev(substr(top$reason, 1, 26)),
        las = 1, cex.names = 0.7, col = "#2c7fb8", xlab = "ballots rejected")
```

Now the column that matters most, and it is not the count:

```{r reporting}
r[, c("reason", "ballots", "reported_by")][c(1:3, 17:19), ]
```

**The total rejected is reported by 6,361 jurisdictions. The reasons are
reported by as few as 845.** You can find out *how many* ballots were thrown out
far more reliably than *why* — which is exactly backwards from what anyone
trying to fix the problem would need.

```{r reason-sum}
c(stated_total = rej, sum_of_reasons = sum(r$ballots),
  difference = rej - sum(r$ballots))
```

The reasons do very nearly add up — a gap of about 3,000 out of 584,463. **That
is a good sign and it does not rescue the point above.** The categories account
for the total; the *jurisdictions* reporting each category do not cover the
country.

# Part 5 — Provisional ballots

```{r prov}
cast <- nat$total[nat$variable == "E1a"]
full <- nat$total[nat$variable == "E1b"]
prej <- nat$total[nat$variable == "E1d"]
c(cast = cast, counted_in_full = full, rejected = prej,
  pct_rejected = round(100 * prej / cast, 1))
```

A provisional ballot is what you get when you turn up and something is wrong —
you are not on the list, you lack the right ID, you are at the wrong place. It
is the system's safety net.

**One in four is rejected. 436,258 people.**

That is a far higher failure rate than mail, and it is much less discussed,
because provisional ballots are issued to exactly the people least likely to
have anybody advocating for them. Note also that this figure is reported by only
**79.9%** of jurisdictions, so the real number is larger.

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Provisional by state.** Compute `100 * prov_rejected / prov_cast`
for states with more than 10,000 provisional ballots. The spread is wider than
mail. Which states stand out, and what would you need to know before blaming
either the voters or the rules?

**Option B — Mail take-up.** Part 3 shows 66.9 million ballots sent and 48.0
million returned — 19 million never came back. Compute that share by state
(`1 - mail_returned/mail_sent`). Where is it highest, and is an unreturned mail
ballot evidence of anything?

**Option C — The reporting gap.** Rank the reasons by `reported_by` instead of
by `ballots`. Which reasons are best reported and which worst? Is there a
pattern in what election offices find easy to say?

**Option D — Your own question.** 6,461 jurisdictions, 535 columns in the source
file, and a codebook that tells you what every one of them means.

# What this lab cannot tell you

- **Whether a rejection was correct.** The file records that a ballot was
  rejected and the reason recorded by the office. It does not record whether the
  signature really was different, or whether the voter was told, or whether they
  could fix it.
- **Who the rejected voters were.** No demographics. Research linking rejection
  to race and age exists and uses voter files — Session 6's data — matched to
  administrative records, not this survey.
- **A clean national total.** Response is 80–99% depending on the item, and the
  non-response is not random. Everything here is a floor.
- **Comparisons across states without care.** Different rules, different units,
  different definitions of the same word. That is not a defect in the survey; it
  is a fact about American elections that the survey is honest enough to expose.
- **What happened in 2026.** This is 2024. The 2026 EAVS will not be published
  until mid-2027 — long after the election we watched in Session 10.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Your ballot is 22 times likelier to be rejected in Oklahoma than in Vermont.
  Is that a problem, and whose?
- One in four provisional ballots is rejected. Should the safety net work better
  than that, or is that what a safety net looks like?
- The count of rejections is well reported; the reasons are not. Who benefits
  from that asymmetry, if anyone?

***

### Where the data came from

U.S. Election Assistance Commission, **Election Administration and Voting
Survey, 2024, Version 2** (released 12 February 2026), 6,461 jurisdictions and
535 columns, with the official codebook.
<https://www.eac.gov/research-and-data/datasets-codebooks-and-surveys>

**Version 2, not Version 1** — V1 was corrected, and the EAC publishes an errata
note listing what changed. A lab built on V1 would disagree with the EAC's own
report.

Three traps handled in `data/build-data.R`: missing values are coded **−99 and
−88** rather than left blank, so anything summing a raw column gets nonsense;
the file contains **embedded newlines** inside comment fields, so counting lines
gives 17,013 for a file with 6,461 rows; and **"jurisdiction" means different
things in different states.**
