---
title: "Data Session 11 Key: Did the Machinery Work?"
subtitle: "84-355 Democracy's Data | Session 11 — Election administration (EAVS) | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
nat <- read.csv("data/national.csv", stringsAsFactors = FALSE)
st  <- read.csv("data/states.csv",   stringsAsFactors = FALSE)
r   <- read.csv("data/reasons.csv",  stringsAsFactors = FALSE)
u   <- read.csv("data/units.csv",    stringsAsFactors = FALSE)
g   <- function(v) nat$total[nat$variable == v]
st$rate <- 100 * st$mail_rejected / st$mail_returned
big <- st[st$mail_returned > 100000, ]
```

All code is given to students. Grade the interpretation.

**Where this sits.** Session 11, the week after the election. Every prior session
used the *output* of an election; this one asks whether the machinery worked.
It is also the session that most reliably connects to whatever is in the news
that week, because administration failures are what get reported in November.

**It runs on 2024 data, deliberately.** The 2026 EAVS will not exist until
mid-2027. Say so early — students will ask, and the two-year lag is itself a
fact worth teaching about administrative data.

Every number below was run against the data files.

***

# The session in one sentence

The country rejected 584,463 mail ballots and 436,258 provisional ballots, your
odds of that happening to you vary twenty-two-fold by state, and the survey that
tells us so is missing a fifth of its answers.

***

# The findings

```{r headline}
c(jurisdictions   = sum(st$jurisdictions),
  mail_rejected   = g("C9a"),
  mail_pct        = round(100 * g("C9a") / g("C1b"), 2),
  prov_rejected   = g("E1d"),
  prov_pct        = round(100 * g("E1d") / g("E1a"), 1),
  worst_state     = big$rate[big$state == "OK"],
  best_state      = big$rate[big$state == "VT"])
```

- **584,463 mail ballots rejected**, 1.22% of those returned.
- **Oklahoma 4.35%, Vermont 0.20% — a 22-fold gap.**
- **436,258 provisional ballots rejected: one in four.**
- The largest single reason is **a signature that did not match — 197,979, 34%.**
- The *total* is reported by 6,361 jurisdictions; some *reasons* by as few as
  **845**.

***

# The three things to get said

**1. "Jurisdiction" is not one thing, so most averages are meaningless.**
Wisconsin reports 1,851 and Texas 254 because Wisconsin runs elections through
townships. **Make the class say out loud what is safe to compute**: totals yes,
rates yes, per-jurisdiction averages across states no. This is the "units"
lesson from Session 4's midterm lab arriving in a new costume, and naming that
connection is worth thirty seconds.

**2. Variation between states is about rules, not voters.** The 22-fold gap is
the number students will remember. Push them past outrage to mechanism: witness
requirements, signature matching, postmark rules, and — most decisive —
**whether the state has a cure process** letting a voter fix a problem after the
fact. The same act, performed identically, is counted in Vermont and discarded
in Oklahoma.

**3. The survey's holes are the lesson, not a footnote.** Provisional rejections
are reported by 79.9% of jurisdictions. Ask directly: *is the missing fifth
random?* It is not — smaller offices with less staff skip more items, and
smaller offices are systematically different. So every national total here is a
**floor**, and the gap between floor and truth is unknown and unknowable from
this file.

**The line for the board:** this is the only dataset in the course where the
people being measured are the same people filling in the form.

***

# Part 3 — where most of the discussion will go

```{r p3}
rbind(head(big[order(-big$rate), c("state","mail_returned","mail_rejected","rate")], 5),
      head(big[order( big$rate), c("state","mail_returned","mail_rejected","rate")], 3))
```

**Anticipate the wrong inference.** Students will read a high rate as either
voter incompetence or deliberate suppression. Both are available, neither is
demonstrated, and the file cannot separate them from a third possibility:
**states with strict rules and no cure process reject more ballots without
anybody intending anything.**

The strongest move available to a student is to notice that **Michigan rejects
3.05% on 2.08 million returned ballots** — the largest absolute count of any
high-rate state, 63,561 ballots — while Indiana rejects 0.21% on 1.6 million.
Two large mail states, fourteen times apart.

***

# Option A — Provisional by state

## Verified output

```{r a}
s <- st[st$prov_cast > 10000, ]
s$prate <- round(100 * s$prov_rejected / s$prov_cast, 1)
rbind(head(s[order(-s$prate), c("state","prov_cast","prov_rejected","prate")], 5),
      head(s[order( s$prate), c("state","prov_cast","prov_rejected","prate")], 4))
```

**3/4:** Computes the rates, reports the spread, and notes it is wider than mail.

**4/4:** Refuses to compare the top and bottom directly, because **provisional
ballots mean different things in different states.** Where election-day
registration exists, almost nobody needs one; where they are handed out freely
for any discrepancy, many are issued and many fail. A high rejection rate can
mean a generous issuing policy. **Denominators again.**

**4/4, best:** Notices that states with the *fewest* provisional ballots per
voter are often those with same-day registration — that is, the safety net is
smallest where the main system works. A student who says "the number of
provisional ballots is a measure of how badly the registration system is
performing, not of how well the backup is" has understood the instrument.

***

# Option B — Mail take-up

## Verified output

```{r b}
s <- st[st$mail_sent > 100000, ]
s$unret <- round(100 * (1 - s$mail_returned / s$mail_sent), 1)
rbind(head(s[order(-s$unret), c("state","mail_sent","mail_returned","unret")], 5),
      head(s[order( s$unret), c("state","mail_sent","mail_returned","unret")], 3))
```

**3/4:** Computes the share and identifies the extremes.

**4/4:** Distinguishes the two worlds inside this number. In **all-mail states**
every registered voter is sent a ballot whether they asked or not, so an
unreturned ballot mostly means "this person did not vote" — ordinary
non-voting, arriving by post. In **request-only states**, somebody actively
asked for a ballot and then did not return it, which is a different and more
interesting failure.

**4/4, best:** Says the comparison should not be made at all without splitting
the states into those two groups first, and notes that this file does not label
which is which — you would need to bring that knowledge from outside. **Knowing
what you need and not having it is a legitimate finding.**

***

# Option C — The reporting gap

## Verified output

```{r c}
r[order(-r$reported_by), c("reason","reported_by","ballots")][c(1:4, 16:19), ]
```

**3/4:** Ranks by reporting and observes that lateness is well reported while
several others are not.

**4/4:** Finds the pattern. **The best-reported reasons are the ones an office
records automatically** — a ballot arrived after the deadline, the voter had
already voted, the voter was ineligible. These are byproducts of the system.
**The worst-reported are the ones requiring a human to have written down a
judgement** — no postmark (845), no resident address (1,959), unofficial
envelope (2,072).

**4/4, best:** Draws the conclusion for the signature category, which sits
awkwardly in between: signature mismatch is the **largest** reason, at 197,979,
but is reported by only 3,222 of 6,461 jurisdictions. So the most consequential
discretionary judgement in mail voting is documented by half the country. A
student who says "we know least about the category where discretion matters
most" has the session.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating the response rate as data rather than
  as a caveat. Anyone who writes "this total is a floor because…" and completes
  it correctly has the point.
- **A 2** is reporting the 22-fold gap as evidence of suppression without
  mentioning rules, cure processes, or the possibility that the states differ in
  what they count.
- **Reward anyone who is bothered by the two-year lag.** We watched an election
  in Session 10 and cannot evaluate its administration until 2027. That is a
  real democratic problem and it is nobody's fault.

## Anticipated questions

- *"Is 1.22% a lot?"* — It is 584,463 people, and the 2024 presidential margin in
  several states was smaller than the number of ballots that state rejected. Do
  not assert that rejections would have changed an outcome — rejected ballots
  are not a random sample of anything — but the comparison of magnitudes is fair
  and worth making.
- *"Why do offices not answer?"* — No penalty, no staff, and in small townships
  the clerk is often part-time. This is voluntary reporting by the people being
  reported on.
- *"Could someone reject ballots on purpose?"* — The file cannot answer that.
  Point at the signature category and note that discretion concentrated in a
  poorly documented category is where anyone worried about it would look.
- *"Why is 2026 not here?"* — Fielded in early 2027, published mid-2027.
- *"Why does Wisconsin have so many jurisdictions?"* — Townships. Then ask what
  else in this course changed meaning by state (Session 6: party registration).

## Connections

- **Session 6 (voter files)** — the registration side of the same administrative
  system, and the source research uses to find *who* gets rejected, which this
  survey cannot say.
- **Session 10 (live returns)** — we watched the election; this is its service
  record, two years late.
- **Session 4 / Oct 20** — "per what?" The units problem, third appearance.
- **Session 3 (ACS margins)** — both sessions are about a published number that
  arrives with its own uncertainty attached, and both reward reading the column
  nobody quotes.
