---
title: "Data Session 10 Key: The Midterms, Two Days Later"
subtitle: "84-355 Democracy's Data | Session 10 — Live returns | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
sen <- read.csv("data/senate_2026_landscape.csv", stringsAsFactors = FALSE)
reh <- read.csv("data/rehearsal_house_2024.csv", stringsAsFactors = FALSE)
sen$baseline_party <- ifelse(sen$pres24_winner == "Trump", "Republican", "Democrat")
```

***

# READ THIS FIRST — what this lab is and is not

**This lab was built in August 2026, three months before the election it is
about.** The results did not exist. Neither did any honest analysis of them.

So it ships in two halves:

- **Parts 1–3 are complete and final.** The 2026 map, the baseline, and the
  rehearsal are all real data and all verified.
- **Parts 4–7 are written, tested, and dormant.** Every chunk has been rendered
  against a synthetic results file to confirm it runs. They are switched off by
  `REHEARSAL <- TRUE`.

## Your checklist for the morning of 5 November 2026

1. Get results into `data/results_senate_2026.csv` with exactly three columns:
   `state` (two-letter), `winner_party` ("Democrat"/"Republican"), `dem_pct`
   (number). A template with the right shape is already in `data/`.
   **Typing 33 rows by hand from a news site takes about ten minutes and is
   entirely defensible.** Tell the class you did it; it is a better lesson than
   pretending a pipeline existed.
2. Open `election-night.Rmd`, find Part 4, change `REHEARSAL <- TRUE` to
   `REHEARSAL <- FALSE`.
3. Knit. That is the whole procedure.
4. Check the landscape file against reality — see the retirement caveat below.

## The one thing that will be wrong

`senator_last` is whoever held the seat in August 2026. **Retirements, primary
defeats and appointments between then and November are not in this file.**
Several will have happened. Before class, skim the 33 rows and note which are
actually open seats; an open seat behaves very differently from an incumbent
defending, and a student who spots a name that is not on the ballot has found a
real defect worth praising.

***

# The lab in one sentence

A prediction rule that uses no information about the candidates, the campaigns,
the money or the polls is right about 96% of the time — and everything anyone
says about an election has to be measured against that.

***

# Parts 1–3: verified now

```{r verified}
c(seats_up = nrow(sen),
  republican_defending = sum(sen$party == "Republican"),
  democrat_defending   = sum(sen$party == "Democrat"),
  seats_on_hostile_turf = sum(sen$hostile_turf))

sen[sen$hostile_turf, c("state", "senator_last", "party",
                        "pres24_winner", "pres24_margin")]

c(rehearsal_districts = nrow(reh),
  baseline_correct = sum(reh$as_expected),
  pct = round(100 * mean(reh$as_expected), 1))
```

- **33 seats: 20 Republican, 13 Democratic.** The asymmetry is an accident of
  2020, not a strategy.
- **Only 3 seats sit on hostile turf**: Peters (MI, Trump +1.4), Ossoff (GA,
  Trump +2.2), Collins (ME, Harris +6.9).
- **The presidential baseline was right in 418 of 433 House districts in 2024 —
  96.5%.**

That 96.5% is the number to put on the board. It is the whole intellectual
content of the lab.

***

# What to say on the day, depending on what happened

Three scenarios. You will know which by Thursday morning.

## If the baseline held (roughly 30+ of 33 seats as implied)

The story is **turnout and calendar, not persuasion**. Say plainly that the
election was largely determined before the campaign, and push students to
distinguish "nothing happened" from "nothing changed" — a huge amount of money
and effort produced an outcome a rule from 2024 already knew.

Good question for the room: *if the result was this predictable, what were two
years of campaigning for?* Acceptable answers include holding the line (the
baseline holds because someone worked to make it hold) and the seats that did
move being the entire point.

## If it broke toward one party in several states

That is a **national swing**, and it is the interesting case. Direct students to
the *spread*, not the average — Option B. A large mean shift with a small
standard deviation is one country moving together, which is the signature of a
nationalised midterm and the thing Rodden and Hopkins argue has been increasing.

Watch for the classic error: reading three dramatic races as a national verdict.
Three exceptions out of 33 is 91% baseline accuracy, which is *worse* than 2024
but still very high.

## If it broke in both directions

Then it was **incumbency and local factors**, not a national message. This is the
hardest case to write about well and the most likely to be misreported in the
press that week. Reward students who notice that scattered exceptions are
evidence *against* a national story.

***

# Option A — Judge your own prediction

The point is not accuracy. It is **calibration against the null**.

**3/4:** Compares their Session 9 prediction to the result honestly, including
where they were wrong.

**4/4:** Compares their prediction to the **baseline** as well, and reckons with
the possibility that their extra knowledge subtracted value. Most students will
have deviated from the baseline in at least one race and been wrong to. A
student who writes "I knew more than the rule and did worse than the rule" has
learned the most useful thing available in this lab.

**4/4, best:** connects it to Session 5's media-attention demo. Attention concentrated on a handful of
dramatic races; those are exactly the races where the baseline is least reliable
and where a student's intuition felt most informed. Feeling informed and being
informed came apart, and the data shows where.

**Do not reward being right.** With 33 races and three genuinely competitive
ones, being right is mostly luck, and saying so is part of the lesson.

***

# Option B — Nationalised or local?

The two statistics do different jobs and students routinely report only the
first:

- **mean shift** — which way, and how far, the country moved
- **standard deviation of the shift** — whether it moved *together*

**3/4:** Reports both and interprets the mean.

**4/4:** Interprets the spread. Small SD = one national verdict; large SD =
thirty-three local contests sharing a date. Then says which one the phrase "the
voters sent a message" requires, and whether the data supports using it.

**4/4, best:** notes that the mean can be near zero while enormous shifts cancel,
and checks for that rather than assuming.

**In rehearsal mode** this option runs on 2024 House districts, where the SD of
district Democratic share around 50 is large — because districts differ hugely
from one another, which is a different quantity than a *shift*. Flag this if you
teach the rehearsal version: the interpretation only lines up cleanly once real
2026 results are loaded.

***

# Option C — The seats that were never in doubt

```{r c}
safe <- sen[sen$abs_margin >= 15, ]
c(safe_seats = nrow(safe), of_total = nrow(sen),
  pct = round(100 * nrow(safe) / nrow(sen)))
head(safe[order(-safe$abs_margin), c("state", "senator_last", "party",
                                     "pres24_margin")], 8)
```

**3/4:** Counts the safe seats and connects them to the earlier weeks.

**4/4:** Answers the question at the end, which is genuinely hard. House
uncompetitiveness is partly *drawn* — someone chose the lines. Senate
uncompetitiveness is not: nobody drew Wyoming. So the same symptom has two
different causes and only one of them has a remedy anyone has proposed. A
student who separates "unfair because someone made it so" from "unfair because
of where people live" has the distinction the Dec 3 wrap builds toward.

**4/4, best:** notices that the Senate's problem is *worse* in one specific
sense — the malapportionment is constitutionally entrenched and effectively
unamendable, whereas district lines are redrawn every ten years.

***

# Grading

Four-point scale from the syllabus. Grade generously this week: the material is
two days old and nobody has had time to think properly, including you.

- **Never penalise code.** It is your code, and this week it is code you switched
  on that morning.
- **Reward hedging.** This is the one lab where "I cannot tell yet, and here is
  what I would need" is close to the ideal answer. Part 6 exists to license it.
- **Penalise confident causal stories.** Two days after an election, nobody knows
  why anything happened. Students who write "voters rejected X" without evidence
  should be marked down and told exactly why.

## Anticipated questions

- *"Are these the final numbers?"* — Almost certainly not. Mail ballots in
  several states count for days or weeks, and late-counted ballots have leaned
  more Democratic than election-day ballots. Margins cited on Thursday move.
- *"Why isn't Senator X on this list?"* — Either their class is not up, or they
  retired after this file was built. Both are worth explaining.
- *"Did the forecast get it wrong?"* — Usually unanswerable from one election.
  A 70/30 call that loses is not a failed forecast. Calibration needs many
  predictions; you have three competitive races. Say so, and connect to Session 9.
- *"Why is there no live results feed?"* — Session 1's answer: no national election
  administrator exists. This is the lab where that fact costs *you* something in
  front of the class, which makes it land.

## Connections back

This is the hinge of the term — everything before it was a prediction, and
everything after it works on what the election actually produced.

- **Session 4 (returns over time)** — 13 competitive states, down from 33 in 1960. Today's map is that
  fact applied to one Senate class.
- **`campaign-visits`** (retired, on disk) — the campaign visits map predicted where attention would go.
  Compare it to where the Senate races actually turned out to be close.
- **Session 5 (media attention)** — attention concentrated on a few races; the baseline says most
  were settled. Attention and consequence were not the same set.
- **Session 9 (polls, models, and markets)** — the forecast students read is now testable, and the honest answer
  about testing it is "not with 33 races."
- **Session 2 (the decennial census)**, looking back — the Senate's
  malapportionment is that count's consequence. Wyoming and California appear on
  the same list with the same two seats, and students have already counted the
  people behind them.
- **Sessions 11 to 14** — from here the course moves to how the votes cast
  tonight were administered, recorded, and will be argued about.
