---
title: "Lab 11: The Midterms, Two Days Later"
subtitle: "84-355 Democracy's Data | Week 11"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

The country voted on Tuesday. It is Thursday.

Every other lab this term has looked at an election that finished long ago, with
the answer already known and the arguments about it already settled. Today the
count is barely finished, some races are not called, and nobody yet knows what
the story is. You are going to work it out from the numbers, in the same two
days that the people writing the analyses you will read next week are working on
theirs.

You have one advantage over them: you wrote down what you expected **before** it
happened. That is the only way to find out whether you understood anything.

Everything is base R. Work in order.

***

# Part 1: The map, as it looked in advance

```{r read-landscape}
sen <- read.csv("data/senate_2026_landscape.csv", stringsAsFactors = FALSE)
nrow(sen)
table(sen$party)
```

The Senate is split into three classes and one faces the voters every two years.
**Class 2 was up in 2026: 33 seats, of which Republicans were defending 20 and
Democrats 13.**

Which side is defending more seats is not a strategic decision. It is an accident
of which states happened to elect whom six years ago, compounded by a rule
written in 1789. Before a single vote was cast, one party had 20 chances to lose
something and the other had 13.

```{r landscape}
sen[order(sen$abs_margin),
    c("state", "senator_last", "party", "pres24_winner", "pres24_margin")]
```

***

# Part 2: The baseline

The simplest possible prediction: **each seat goes to the party whose
presidential candidate carried that state in 2024.** No polling, no fundraising,
no candidate quality — just the last time everyone voted.

```{r baseline}
sen$baseline_party <- ifelse(sen$pres24_winner == "Trump", "Republican", "Democrat")

c(baseline_R = sum(sen$baseline_party == "Republican"),
  baseline_D = sum(sen$baseline_party == "Democrat"))

sum(sen$hostile_turf)
sen[sen$hostile_turf,
    c("state", "senator_last", "party", "pres24_winner", "pres24_margin")]
```

> Only **three** of the 33 seats were held by the party that lost that state in
> 2024: **Peters in Michigan** (Trump +1.4), **Ossoff in Georgia** (Trump +2.2),
> and **Collins in Maine** (Harris +6.9).
>
> Everywhere else, the seat and the state already agreed. If nothing else
> happened, almost nothing would change.

That is the null hypothesis for tonight. Elections are interesting exactly to the
extent that they defeat it.

```{r near-misses}
sen[sen$abs_margin < 8,
    c("state", "senator_last", "party", "pres24_winner", "pres24_margin")]
```

***

# Part 3: How good is that baseline, really?

Before trusting it, test it on an election we already know the answer to. In
2024, did each House district elect the party that won it for president?

```{r rehearsal}
reh <- read.csv("data/rehearsal_house_2024.csv", stringsAsFactors = FALSE)

c(districts = nrow(reh),
  went_as_baseline_implied = sum(reh$as_expected),
  pct = round(100 * mean(reh$as_expected), 1))
```

> **The baseline was right in 418 of 433 districts — 96.5% of the time.**

A prediction that requires no information about the candidates, the campaigns,
the money or the polls is right nineteen times in twenty. This is what
polarization means operationally, and it is why the fifteen exceptions are worth
knowing by name:

```{r exceptions}
reh[!reh$as_expected,
    c("district", "dem_share", "pres_party", "house_rep", "house_rep_party")]
```

Every one of them is a district close to even, and most are held by an incumbent
who has survived by being personally known. **Incumbency is what the baseline
does not contain**, and it is almost the only thing that beats it.

Keep this number in your head for the rest of the lab: **96.5%**. Anything you
conclude tonight has to be measured against a stupid rule that is right that
often.

***

# Part 4: The returns

```{r results-switch}
# ============================================================
# ON 5 NOVEMBER: change this to FALSE and make sure
# data/results_senate_2026.csv exists.
# ============================================================
REHEARSAL <- TRUE
```

```{r load-results}
if (REHEARSAL) {
  res <- NULL
  cat("Running in rehearsal mode -- no 2026 results loaded.\n")
} else {
  res <- read.csv("data/results_senate_2026.csv", stringsAsFactors = FALSE)
  res <- merge(sen, res, by = "state")
  res$as_expected <- res$winner_party == res$baseline_party
  cat("Loaded", nrow(res), "Senate results.\n")
}
```

```{r results-summary, eval = !REHEARSAL}
# Seats won by each party
table(res$winner_party)

# Net change against the seats as they were held going in
c(held_before_R = sum(res$party == "Republican"),
  won_now_R     = sum(res$winner_party == "Republican"))

# Where the baseline failed
res[!res$as_expected,
    c("state", "senator_last", "party", "pres24_margin",
      "winner_party", "dem_pct")]

round(100 * mean(res$as_expected), 1)
```

```{r results-plot, eval = !REHEARSAL, fig.width = 8, fig.height = 6}
r <- res[order(res$pres24_margin), ]
plot(r$pres24_margin, r$dem_pct, pch = 19, cex = 1.4,
     col = ifelse(r$winner_party == "Democrat", "#2166AC", "#B2182B"),
     xlab = "2024 presidential margin (positive = Trump)",
     ylab = "Democratic share of the 2026 Senate vote (%)",
     main = "2026 Senate results against the 2024 baseline")
abline(h = 50, lty = 2); abline(v = 0, lty = 2)
text(r$pres24_margin, r$dem_pct, r$state, pos = 3, cex = 0.65)
```

**In rehearsal mode the chunks above do not run.** They are written, checked
against the 2024 rehearsal file, and waiting. On Thursday morning your
instructor will drop the results file into `data/` and flip one switch.

***

# Part 5: What to look for when the numbers land

Four questions, in the order worth asking them.

**1. Did the baseline hold?** Start with the count of seats that went the way
2024 implied. If it is near 96%, the election was not about persuasion; it was
about turnout and the calendar. If it is much lower, something real moved.

**2. Where did it fail, and in which direction?** A handful of exceptions all
favouring one party is a national swing. A handful scattered in both directions
is noise plus incumbency. These look completely different and are easy to
confuse when you are reading about only the exciting races.

**3. Did the states move together?** In a nationalised election, nearly every
state shifts by a similar amount in the same direction. Compare each state's
Democratic Senate share to its 2024 presidential share and look at the spread of
those differences, not just the average. A tight cluster means one national
verdict; a wide scatter means thirty-three local elections.

**4. Was the forecast any good?** You read the Senate Forecast Explorer before
the election. A forecast that said a race was 70/30 and lost is not thereby
wrong — 30% things happen 30% of the time. The only fair test is across many
races at once: of everything it called 70/30, did roughly 70% come in? You will
not have enough races to settle that. Say so.

***

# Part 6: What you cannot say on Thursday

**The count is not finished.** Several states take days or weeks. California and
Arizona in particular count large numbers of mail ballots after election day, and
those ballots have historically leaned more Democratic than election-day ballots
— so early returns can move steadily in one direction as counting proceeds. A
margin you cite today may not be the final margin.

**Turnout figures are provisional.** The denominator — how many people were
eligible — is an estimate that gets revised for months.

**Exit polls are not results.** They are surveys, with all of Week 10's problems,
and they are weighted to the reported outcome as it comes in.

**And you are two days in.** The people who study elections for a living will
publish their real analyses months from now, using precinct data that does not
exist yet. What you can do today is describe what happened and be honest about
the parts you cannot yet see. That is a genuinely useful skill and it is what
most published election-night commentary fails at.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.**

Pick **one**.

***

### Option A — Judge your own prediction

In Week 10 you wrote down what you expected: how many Senate seats each party
would hold, and which specific races you thought would be close.

```{r opt-a}
# Fill these in from what you wrote in Week 10
my_predicted_R_seats <- 22          # <- your number
my_predicted_D_seats <- 11          # <- your number

# The naive baseline, for comparison
c(baseline_R = sum(sen$baseline_party == "Republican"),
  baseline_D = sum(sen$baseline_party == "Democrat"))
```

**Write about:** compare your prediction to the actual result **and** to the
baseline. The interesting question is not whether you were right — it is whether
you beat the stupid rule. If you did not, what were you adding when you deviated
from it? Most published forecasting adds very little to the baseline and charges
a great deal of attention for it.

***

### Option B — Nationalised or local?

```{r opt-b}
# In rehearsal mode this runs on the 2024 House districts.
# After the election, set REHEARSAL <- FALSE above and it will run on the Senate.
if (REHEARSAL) {
  x <- reh
  x$shift <- x$dem_share - 50
  cat("Rehearsal: spread of district Democratic share around 50\n")
} else {
  x <- res
  x$shift <- x$dem_pct - (50 - x$pres24_margin / 2)
  cat("2026: each state's Senate result minus its 2024 presidential baseline\n")
}

c(mean_shift = round(mean(x$shift, na.rm = TRUE), 2),
  sd_shift   = round(sd(x$shift, na.rm = TRUE), 2),
  n          = sum(!is.na(x$shift)))

hist(x$shift, breaks = 20, col = "#BBBBBB", border = "white",
     xlab = "Shift relative to baseline (points)",
     main = "How much did places move, and did they move together?")
abline(v = 0, lty = 2, lwd = 2)
```

**Write about:** the **mean** tells you which way the country moved; the
**standard deviation** tells you whether it moved as one country. Report both.
A large mean with a small spread is a national verdict. A mean near zero with a
large spread is thirty-three separate contests that happened to share a date.
Which did we just have, and what does your answer imply about whether "the
voters sent a message"?

***

### Option C — The seats that were never in doubt

```{r opt-c}
safe <- sen[sen$abs_margin >= 15, ]
c(safe_seats = nrow(safe), of_total = nrow(sen))
safe[order(-safe$abs_margin),
     c("state", "senator_last", "party", "pres24_margin")]
```

**Write about:** how many of the 33 seats were realistically settled before the
campaign began? Connect it to Week 4 (13 competitive states), Week 6 (26 states
that got no candidate visits at all) and Week 15 (four out of five House seats
uncompetitive). You have now measured the same phenomenon four times with four
different instruments. Say what it adds up to — and whether a Senate seat being
safe is the same kind of problem as a House seat being safe, given that nobody
draws state lines.

***

### Option D — Ask your own question

Something happened on Tuesday that nobody expected. Find it in the numbers and
explain it.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at an election two days old and say something true
about it while being honest about what you cannot yet know.

***

## Where the data came from

**The 2026 Senate map** — Senate class membership from
`@unitedstates/congress-legislators`
(<https://unitedstates.github.io/congress-legislators/legislators-current.csv>),
joined to the 2024 presidential results by state used in Weeks 1, 2 and 6.

One caveat that matters. The `senator_last` column is **whoever held the seat
when this file was built**, not necessarily whoever appeared on the ballot.
Retirements, primary defeats and appointments all break that assumption, and
some of them will have happened. Treat the column as "the seat currently held
by" and check the actual candidates before relying on any individual row.

**The rehearsal file** — 2024 House results next to the presidential result in
the same district, from the Week 15 file. Used so that every chunk in the
post-election half of this lab could be written and tested before an election
that had not yet happened.

**The 2026 results** — added on the morning of 5 November. There is no national
source of live returns, because, as Week 2 established, the United States has no
national election administrator. `data/build-data.R` lists the options in order
of preference and writes a template with the required columns. Typing thirty-three
numbers in by hand from a news site is a completely defensible method, and if
that is what happened, the lab should say so out loud.
