---
title: "Proving a Threshold You Cannot See"
subtitle: "84-355 Democracy's Data | Traffic stops and searches, San Francisco"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `policing-brief.Rmd` first.** This lab runs the same ten steps with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

> **One of several labs in this unit.** They ask the same question of different
> data: **can you show discrimination when nobody wrote down a reason?** The
> others are redlining and jury selection. Whichever you did, you will be asked
> to explain your inference to somebody who did a different one.

***

# Where did that number come from?

> *"Black drivers are pulled over more often."*

Easy to say, hard to prove — and the reason is worth being precise about. To
know whether drivers of one race are stopped *more than they should be*, you
would need to know how many were driving, and how many were doing something
stop-worthy. **Nobody records that.** The denominator does not exist.

So the honest version of the question has to be built differently. There is a
way, and it does not need the denominator at all. Ten steps.

***

# Step 1 — What one record is

The source file has one row per stop. We work from a collapsed version, but the
fields it was collapsed from decide everything below:

```{r schema}
data.frame(
  field = c("date, time", "location", "subject_race", "reason_for_stop",
            "search_conducted", "contraband_found", "outcome"),
  holds = c("when", "district or intersection",
            "the OFFICER'S PERCEPTION, entered on a form",
            "the violation cited", "TRUE / FALSE",
            "TRUE / FALSE — recorded only if a search happened",
            "warning, citation, arrest"))
```

Two of those matter enormously.

**`subject_race` is what the officer believed**, not self-report. For a question
about the officer's behaviour that is arguably the right variable — but it is not
the census's variable, and the two must never be mixed.

**`contraband_found` exists only when a search happened.** You never learn what
an unsearched driver was carrying. That absence is the whole problem.

# Step 2 — How many, and over what

```{r load}
d <- read.csv("data/by_race.csv", stringsAsFactors = FALSE)
y <- read.csv("data/by_year.csv", stringsAsFactors = FALSE)
c(stops = sum(d$stops), searches = sum(d$searched),
  contraband = sum(d$contraband_found),
  years = length(unique(y$year)))
round(100 * sum(d$searched) / sum(d$stops), 2)
```

**905,070 stops over ten years in one city**, and searches happen in about 1.6%
of them. Everything interesting is inside that small slice.

# Step 3 — The first table anybody makes

```{r stops}
d$share <- round(100 * d$stops / sum(d$stops), 1)
d[order(-d$stops), c("race", "stops", "share")]
```

Black drivers were **16.8% of stops**. San Francisco's Black population over
this period was roughly 5–6% of residents.

It is very tempting to put those two numbers next to each other.

# Step 4 — Why that comparison does not work

Do not put them next to each other. **The residential population is not the
population at risk of being stopped.**

San Francisco is a commuter city — a large share of weekday cars belong to
people who live elsewhere. Driving is not distributed like residence: it varies
with car ownership, work schedules, and which roads run through which
neighbourhoods.

And even a perfect count of drivers would not be enough. You would also need to
know **how many drivers of each group were doing something that warranted a
stop**. Nobody records that. Nobody could.

```{r denominator}
data.frame(
  you_would_need = c("Who was driving", "Where and when",
                     "Who committed a violation", "Who was carrying contraband"),
  does_it_exist  = c("No", "No", "No", "Only for those searched"))
```

**This is not a funding problem.** There is no census of who was on the road.
The disparity in Step 3 — real as the counts are — proves nothing on its own,
and most public arguments about policing data stop right here, with both sides
asserting a denominator neither can produce.

# Step 5 — Condition on the stop

Change the question. Take the stop as given, and ask what happened next.

```{r rates}
d$search_rate <- round(100 * d$searched / d$stops, 2)
d[order(-d$search_rate), c("race", "stops", "searched", "search_rate")]
```

**Black drivers were searched in 15.52% of stops. White drivers, 3.14%** —
nearly five times the rate.

Better than Step 3, because the comparison is now *within* a population the
officer had already selected. **Still not enough**, and the objection is the
standard one: perhaps the stopped groups genuinely differed. If officers were
encountering more grounds for suspicion in one group, a higher search rate would
follow from correct policing rather than a lower bar.

You cannot answer that with search rates. You would have to know how suspicious
each driver actually was — the unobservable thing.

# Step 6 — The move that gets around it

In 2001 three economists — **Knowles, Persico and Todd** — proposed a way out
that needs no denominator at all. Condition one step further. **Take the search
as given, and ask what the officer found.**

> An officer searches when suspicion crosses some internal threshold. If the
> same threshold applies to everyone, the searches it authorises should succeed
> at roughly equal rates. A *lower* threshold means more searches on weaker
> evidence, and weaker evidence means a lower success rate. **So unequal hit
> rates imply unequal thresholds.**

The comparison is now between searches, not populations.

**Before running the next chunk, predict.** Given a search rate five times
higher for Black drivers, should the share of searches finding contraband be
higher, lower, or about the same?

# Step 7 — What the searches found

```{r hit}
d$hit_rate <- round(100 * d$contraband_found / d$searched, 1)
d[order(-d$search_rate), c("race", "searched", "contraband_found", "hit_rate")]
```

**Contraband was found in 9.2% of searches of Black drivers and 24.2% of
searches of white drivers.**

Searched **4.9 times as often**, and found with something **2.6 times less
often**. Under the outcome test, that is evidence of a lower evidentiary bar.

# Step 8 — Both numbers at once

One group could be a coincidence. Put every group on the same picture.

```{r hit-plot}
plot(d$search_rate, d$hit_rate, pch = 19, col = "#C41230", cex = 1.6,
     xlim = c(0, 17), ylim = c(0, 40),
     xlab = "% of stops that led to a search",
     ylab = "% of searches that found contraband")
text(d$search_rate, d$hit_rate, d$race, pos = 4, cex = 0.75)
abline(h = weighted.mean(d$hit_rate, d$searched), lty = 3, col = "grey50")
```

The relationship is **monotone across all five groups**: the more often a group
was searched, the less often those searches found anything. Asian and Pacific
Islander drivers were searched least — 1.81% — and had the *highest* hit rate at
36.1%.

That ordering is what makes an innocent explanation hard. It would have to cover
the Black and Hispanic results **and** the reverse pattern for Asian drivers,
simultaneously, across 905,070 stops.

Notice what this argument never needed: **who was driving.**

# Step 9 — Is it just one bad year?

Ten years are stacked into every number above. A pattern that comes from one
unusual year is a different finding from one that repeats.

```{r trend}
y$hit <- round(100 * y$contraband_found / pmax(y$searched, 1), 1)
w <- y[y$race == "white", c("year", "hit")]
b <- y[y$race == "black", c("year", "hit")]
m <- merge(w, b, by = "year", suffixes = c("_white", "_black"))
m$white_higher <- m$hit_white > m$hit_black
m
```

```{r trend-plot}
ks <- c("white", "black", "hispanic", "asian/pacific islander")
cl <- c("#2c7fb8", "#C41230", "#e08214", "#4d9221")
plot(NA, xlim = range(y$year), ylim = c(0, 50), las = 1, xlab = "",
     ylab = "% of searches finding contraband")
for (i in seq_along(ks)) {
  s <- y[y$race == ks[i], ]; s <- s[order(s$year), ]
  lines(s$year, s$hit, col = cl[i], lwd = 2.2)
}
legend("topright", ks, col = cl, lwd = 2.2, bty = "n", cex = 0.8)
```

**In all ten years, searches of white drivers found contraband more often than
searches of Black drivers.** The levels move around a great deal — everyone's
hit rate falls through 2012 and rises after — but the ordering never flips.

# Step 10 — What you can now say

**You can say:** within this department, over these ten years, searches of Black
and Hispanic drivers were authorised on weaker evidence than searches of white
drivers. That is a claim about the bar an officer applied, reached without ever
knowing who was driving.

**You cannot say** how much of the gap is individual animus, how much is
departmental policy, and how much is where officers were deployed. The data
cannot separate them, and the distinction matters enormously for what anyone
would do about it.

## The objection that survives

The outcome test has one real weakness and you should be able to state it.

It compares **average** hit rates, and the theory is about the **marginal**
search — the last, weakest one an officer decides to make. If the distribution
of actual suspiciousness differs between groups, average hit rates can differ
even when the bar is identical. Economists call this **infra-marginality**, and
it is why the test is strong evidence rather than proof.

```{r arrests}
d$arrest_rate <- round(100 * d$arrests / d$stops, 2)
d[order(-d$search_rate), c("race", "search_rate", "hit_rate", "arrest_rate")]
```

**Take the objection seriously, then notice how much it has to do.** To explain
this table innocently you need a story in which Black drivers are searched five
times as often for good reasons, *and* those reasons turn out wrong far more
often, *and* the same holds for Hispanic drivers, *and* the reverse holds for
Asian drivers. Each piece is possible. Together they are a lot to carry.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Did it change?** Step 9 shows the ordering never flips. Plot the
*gap* between the Black and white hit rates over time instead of the levels. Is
the disparity shrinking, and would a policy change show up in the search rate,
the hit rate, or both?

**Option B — Equalise the bar.** If Black drivers had been searched at the white
rate (3.14%), how many fewer searches would that be? Compute it. Then ask what
the department loses — how much contraband goes unfound — and say whether that
is a trade you would make.

**Option C — The other direction.** Asian and Pacific Islander drivers are
searched least and have the highest hit rate. Under the outcome test, what does
that imply, and is it a claim you are comfortable making from this data?

**Option D — Your own question.** Five rows, 905,070 stops behind them.

# What this lab cannot tell you

- **Who was driving.** The denominator problem never went away; we routed around
  it. Nothing here supports a claim about stop rates *per driver*.
- **Whether any individual officer discriminated.** This is an aggregate
  pattern — it says something about a department and nothing about a person, the
  same gap between group and individual that runs through the RPV lab.
- **One city, and a decade old.** San Francisco, through 2016. Do not
  generalise the numbers. The *method* generalises; Stanford has comparable
  files for dozens of jurisdictions.
- **Race as recorded** — the officer's perception, not self-report.
- **Whether "contraband found" is itself neutral.** What counts, and whether a
  find gets recorded, are decisions by the same officer whose behaviour is being
  measured.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

- Step 3 proves nothing and Step 7 proves quite a lot. What is the difference
  between them?
- Infra-marginality is a real objection. Does it change your reading of the
  table, or just your confidence in it?
- If you did one of the other labs in this unit, which inference did you find
  more convincing, and why?

***

### Where the data came from

**Stanford Open Policing Project**, San Francisco, CA — 905,070 traffic stops,
2020-04-01 release. <https://openpolicing.stanford.edu/data/>

The source is a 177 MB file with one row per stop. Nothing individual-level is
committed here: it is collapsed to five rows by race, plus a small year-by-race
table, before anything is saved. The build is in `data/build-data.py`.
