---
title: "Position-Taking in House Primaries: 14 Issues, 1,662 Candidates"
type: brief
subtitle: "What candidates would not say, and the code that records the silence"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("primary-positions-code.R")
```

```{r setup, include=FALSE}
```

A primary is supposed to let a party's voters read the candidates and choose. Is there anything to read? In 2014, `r n(NCAND)` people ran in a congressional primary, and a team at the Brookings Institution opened every one of their campaign websites. They wrote down where each candidate stood on fourteen issues of the day: immigration, the Affordable Care Act, guns, abortion, the minimum wage, climate change and eight more.

`r ALL14` of those candidates said nothing about any of the fourteen. That is not a gap in the data; it is the data. The coders gave themselves four codes for each issue, and the fourth one is **"Candidate Provides No Information."** Read across all fourteen issues, the fourth code is the most common answer in the file. So this brief asks the question the file was built to answer: how much of a platform does a primary candidate actually publish, where a voter can read it?

## The test

The data is the **Primaries Project** database, compiled at the Brookings Institution by Elaine Kamarck and Alexander Podkul and reported in 2014, obtained from [Brookings](https://www.brookings.edu/articles/the-2014-primaries-project-data/). Their question was about factions: whether the Tea Party was really taking over the Republican Party. To answer it they needed to place every candidate, not just the famous ones, so they coded nearly everybody who filed — `r n(NCAND)` candidates, the whole field rather than a sample of it.

This source fits the question because of one fact about how it was made. **The coding was done from the candidates' own websites.** Nobody interviewed these candidates and nobody sent them a questionnaire; a researcher read what each campaign chose to publish, and coded it. A survey would have offered "no opinion" and a candidate would have picked something. Here nobody was asked, so silence survives as a recorded value, which makes this the rare file where *not saying* is measured.

This brief belongs to the cluster the [roster chapter](../rosters-source/rosters-source-brief.html) introduces: records of who ran and who won, which say nothing about what any candidate stood for. This file holds the missing half, what the candidates would say, read against the same 2014 field. The test: count the silence, per issue and per candidate, and see whether it looks like an accident or like a strategy.

## The data

The workbook holds one row per candidate and `r ck("columns in the workbook")` columns. The ones this brief reads, by name: fourteen issue columns, each coded 1 to 4; `party` and `party_side`; `prim_result`, whether the candidate won or lost; and `faction_code`. A derived count, `silent_on`, gives the number of issues coded 4 for each candidate. The four codes:

```{r codes}
```

Codes 1 and 2 are positions. Code 3 is a position the coder could not resolve into a side. **Code 4 is not a position at all**: the researcher read the website and found the subject was not raised.

```{r scope}
```

The cleaning is where two programs disagree about the same file. The party column holds `r ck("distinct party strings the file holds")` distinct strings for what are really about a dozen parties — three candidates filed under `D` followed by a space, one Republican as `G.O.P.`, one lowercase Libertarian.

```{r labels}
```

Whether that trailing space exists depends on which program opens the file. R's reader trims whitespace by default, so the three dirty cells arrive clean and nothing warns you; Python's reader leaves them alone.

```{r readers}
```

Neither program is wrong. But a student who counts parties in R and one who counts in Python will disagree, and neither will see why. This brief reads the column both ways on purpose: the clean version is what the tables use, and the `r ck("cells readxl's trim_ws default silently rewrites")` cells that differ are counted and printed rather than assumed.

One variable is used only with a warning attached. The faction variable has ten codes; the codebook prints four of them, under the heading *examples*, and for codes 11, 33, 55, 66 and 99 there is no published definition anywhere.

```{r factions}
```

Altogether `r ck("candidates whose faction code the codebook does not define")` candidates carry a faction code with no published meaning, and this brief does not guess what the codes are. Getting the file costs nothing: no account, no key, no fee. What is missing is any promise of permanence — both files sit in a content-management system's upload folder and carry no DOI, which is why copies are committed beside this brief.

> **Before you read on, write a number down.** The file codes `r NISS` issues
> for each of `r n(NCAND)` candidates. **On how many of those `r NISS` do you
> think the typical candidate said nothing?** Pick a number between 0 and
> `r NISS`, and commit to it before you look at the figures.

## What it says about democracy

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 1. Where `r n(NCAND)` candidates stood on fourteen issues, sorted by how many stayed off the subject.** Stacked bars because every row is the same `r n(NCAND)` candidates split four ways, so the rows are directly comparable; the figure at the right of each row is the share coded "provides no information."

The range is wide and it is not random. `r QUIET$issue` is the quietest subject at `r pc(QUIET$silent_pct)` percent and `r LOUD$issue` the loudest at `r pc(LOUD$silent_pct)` percent. Read the top and bottom rows together. Benghazi was the story conservative media ran hardest in 2014, and `r n(QUIET$silent)` of `r n(NCAND)` candidates did not put it on their website. The Affordable Care Act was the most-discussed subject in the file, and `r n(LOUD$silent)` candidates still said nothing about it.

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

**Figure 2. How many of the `r NISS` issues each candidate left unmentioned.** `r NONE` candidates addressed all `r NISS`, which is `r pc(100 * NONE / NCAND)` percent of the field; `r ALL14` addressed none of them.

Most readers guess something between two and four, because the candidates they can picture are the ones with a full issues page. The median is `r MED`. And the shape matters more than the middle: this is not a hump around a typical candidate who covers most subjects. The distribution leans heavily toward silence, and its single tallest bar sits at the far end, where a candidate's website mentioned none of the fourteen. The `r ALL14` are not all fringe candidates, either — some are serious people whose sites were a biography and a donate button.

```{r outcome}
```

House candidates who won their primary left `r pc(WON)` of the `r NISS` issues unmentioned on average; the ones who lost left `r pc(LOST)`. **Do not read that as a finding about what wins primaries.** Winners include hundreds of incumbents running unopposed, who have the least to gain from a policy page and the most staff to build one. The comparison is here because it shows the variable is not noise: it moves with something. What it moves with, this file cannot say.

Why does the silence matter for democracy? There is a book about exactly this. *Parties on the Ground* (2026), by Kathleen Bawn, John Zaller and four colleagues, studies 55 winnable open-seat nominations from this same 2013–14 cycle through 346 interviews. Its argument is that primaries do not work the way the reformers who created them intended. Most voters pay too little attention to use them; the actors who matter are organised groups with intense policy demands, who vet candidates, vouch for them, and supply or withhold money.

**Figure 1 is the public half of that argument.** If a voter wanted to do the job the primary was designed to give them, the website is where they would start. On the median candidate they would find `r NISS - MED` of `r NISS` subjects addressed, and on `r ALL14` candidates none. The groups in Bawn's account are not reading websites — they meet candidates and get answers that are never published. So the fourth code is doing more than flagging missing values. It measures how much of a nomination contest happens where the public cannot see it.

## What this brief cannot tell you

**It cannot tell you what candidates believed.** It records published claims. A candidate who opposed the Affordable Care Act privately and never mentioned it is coded identically to one who had no view.

**It cannot tell you why a subject was left out.** Silence on immigration might be strategy in a district where the issue splits your coalition. It might be a website built in a weekend, or a candidate who talks about the subject constantly at events and never wrote it down. The code is the same in all three cases.

**It cannot separate "no experience" from "we could not find out."** The prior-office variable is a 0/1 flag whose zero the codebook defines as *"`r ck("CandidateElectoralExperience, meaning of 0")`"*. Two very different candidates share that number.

**It is one election.** 2014 was a midterm with an unpopular Democratic president and a Republican faction fight in progress. Nothing here establishes that campaign websites are usually this empty.

**And the websites are gone.** The coding was done in 2014 from live sites, and most of those domains now expire or redirect. The file is the only surviving record of what it describes, which means its coding decisions cannot be audited by anyone, including its authors.

## What you should have learned

- The most common answer in a file of `r n(NCAND)` coded candidates is silence: the median candidate stated a position on `r NISS - MED` of `r NISS` issues, and `r ALL14` candidates stated none at all.
- Silence varies by subject in ways that look strategic: from `r pc(LOUD$silent_pct)` percent on `r LOUD$issue` to `r pc(QUIET$silent_pct)` percent on `r QUIET$issue`.
- A blank on a document the candidate chose to write is a decision, and this file is unusual for recording it as a value instead of discarding it.
- The same workbook yields different party counts in R and Python, because one reader trims a trailing space and the other does not — `r ck("cells readxl's trim_ws default silently rewrites")` cells differ between the two reads.
- `r ck("candidates whose faction code the codebook does not define")` candidates carry a faction code the codebook never defines, and an undefined code is a column you can count but not interpret.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `by_issue.csv` gives `silent_pct` for every issue. Sort it. Which issues do candidates most avoid, and is avoidance a position?
- `by_count.csv` counts candidates by `silent_on`. Find the typical candidate. How much of a platform is actually stated?
- `by_outcome.csv` compares `mean_silent` for winners and losers. Does saying less help you win a primary, or is something else moving both?
- `factions.csv` has a `documented` flag and codes with no label at all. Count the undocumented candidates. What do you do with a variable whose codebook is incomplete?
- `party_labels.csv` records every raw party string with its `characters` count. Find the labels that are really the same party. How many candidates would a naive count misfile?

Two stretch ideas point past the spreadsheet. Pick five candidates from `candidates.csv` and look their 2014 campaign sites up in the Internet Archive's Wayback Machine. Check the coding of one issue against what the archived page says, and note how many sites survive at all. And read this file beside the [`nomination-anchors`](../nomination-anchors/nomination-anchors-brief.html) brief, which rebuilds Bawn and Zaller's 55 open-seat contests from the same cycle. Do the candidates groups organised behind say more or less on their websites than the field?

## Sources

**Data**

- Brookings Institution, *The Primaries Project* database, compiled by Elaine C.
  Kamarck and Alexander R. Podkul.
  <https://www.brookings.edu/articles/the-2014-primaries-project-data/>
  — `r n(NCAND)` candidates in the 2014 congressional primaries,
  `r ck("columns in the workbook")` columns. The workbook's own metadata records
  it as last saved on 5 January 2015. **The fourteen issue positions are
  Brookings' coding, made by reading each candidate's campaign website during
  the 2014 cycle.**
- Brookings Institution, *Primaries Project Database Codebook*, December 2014.
  Distributed beside the data. Documents four of the ten values of the faction
  variable and states that one is unlabeled. Its own pointer to the companion
  report no longer resolves.

Both files are committed beside this brief, because they are served from a content management system's upload folder rather than an archive, and neither carries a DOI. The party column is read twice, once with whitespace trimming turned off, and the difference between the two reads is reported rather than resolved silently; the validation results are reproduced in `checks.csv`.

**Academic literature**

- Bawn, K., Brown, K., Ocampo, A. X., Patterson, S. Jr., Ray, J. L. and Zaller,
  J. (2026). *Parties on the Ground: A Study of Nominations for the House of
  Representatives.* University of Chicago Press.
  doi:10.7208/chicago/9780226853314. The 55 contests, 346 interviews and the
  account of vetting are theirs.
- Kamarck, E. C. and Podkul, A. R. (2014). "The 2014 Congressional Primaries:
  Who Ran and Why." Brookings Institution.
- Hall, A. B. (2019). *Who Wants to Run? How the Devaluing of Political Office
  Drives Polarization.* University of Chicago Press.
- Boatright, R. G. (2013). *Getting Primaried: The Changing Politics of
  Congressional Primary Challenges.* University of Michigan Press.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_count.csv`](data/derived/by_count.csv), [`by_issue.csv`](data/derived/by_issue.csv), [`by_outcome.csv`](data/derived/by_outcome.csv), [`candidates.csv`](data/derived/candidates.csv), [`factions.csv`](data/derived/factions.csv), [`party_labels.csv`](data/derived/party_labels.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
