---
title: "Lab 7: Measuring Ideology Without Reading a Single Bill"
subtitle: "84-355 Democracy's Data | Week 7"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Here is a hard problem. You want to know how liberal or conservative a member of
Congress is. You could read every bill they voted on and score it — but that
means *you* are deciding what counts as liberal, and somebody will reasonably
disagree with every call you make.

DW-NOMINATE solves this by refusing to look at the bills at all.

It takes nothing but the pattern of who voted with whom, across every roll call
in American history, and asks: what arrangement of members on a line best
predicts those votes? The answer falls out of the arithmetic. Nobody read a
bill. Nobody labelled a policy. And the line it produces turns out to be the one
we call left and right.

Today you will use it to watch Congress come apart.

Everything is base R. Work in order.

***

# Part 1: A scale nobody designed

```{r read}
nom <- read.csv("data/nominate_members.csv", stringsAsFactors = FALSE)
dim(nom)
head(nom[, c("congress","year","chamber","state","bioname","party","dim1")], 4)
```

Every member of the House and Senate since 1879. `dim1` is the left–right
coordinate: negative is liberal, positive is conservative, and the scale runs
roughly −1 to +1.

Look at the current House:

```{r current}
now <- max(nom$congress)
h <- nom[nom$chamber == "House", ]
cur <- h[h$congress == now, ]

c(congress = now, year = cur$year[1], members = nrow(cur))
aggregate(dim1 ~ party, data = cur, FUN = function(x) round(median(x), 3))
```

```{r current-plot, fig.width = 8, fig.height = 4}
par(mar = c(4, 1, 3, 1))
plot(cur$dim1, jitter(rep(1, nrow(cur)), 8), pch = 19, cex = 0.7,
     col = ifelse(cur$party == "Democrat", "#2166AC66", "#B2182B66"),
     yaxt = "n", ylab = "", xlab = "DW-NOMINATE dimension 1 (liberal → conservative)",
     xlim = c(-1, 1), main = paste0("The House, ", cur$year[1]))
abline(v = 0, lty = 3, col = "grey60")
```

> Two clouds, and nothing between them. That separation is not something anybody
> put into the method — it is what the voting produced.

***

# Part 2: It was not always like this

The same picture, sixty years earlier.

```{r sixties, fig.width = 8, fig.height = 6}
par(mfrow = c(3, 1), mar = c(3.5, 1, 2.5, 1))
for (cg in c(90, 105, now)) {
  x <- h[h$congress == cg, ]
  plot(x$dim1, jitter(rep(1, nrow(x)), 8), pch = 19, cex = 0.7,
       col = ifelse(x$party == "Democrat", "#2166AC66", "#B2182B66"),
       yaxt = "n", ylab = "", xlab = "", xlim = c(-1, 1),
       main = paste0(cg, "th Congress (", x$year[1], ")"))
  abline(v = 0, lty = 3, col = "grey60")
}
par(mfrow = c(1, 1))
```

In 1967 the two clouds overlap in the middle. There were liberal Republicans and
conservative Democrats, and they sat in the same ideological space.

***

# Part 3: The gap, measured

Take the median of each party in each Congress and subtract.

```{r gap}
med <- function(cg, p) median(h$dim1[h$congress == cg & h$party == p])
cgs <- sort(unique(h$congress))

gap <- data.frame(
  congress = cgs,
  year = 1789 + 2 * (cgs - 1),
  dem  = sapply(cgs, med, p = "Democrat"),
  rep  = sapply(cgs, med, p = "Republican"))
gap$gap <- gap$rep - gap$dem

tail(gap, 4)
```

```{r gap-plot, fig.width = 8.5, fig.height = 5}
plot(gap$year, gap$gap, type = "l", lwd = 2.5, col = "#54278F",
     xlab = "", ylab = "Distance between party medians",
     main = "How far apart the two parties are, House, 1879-present")
abline(h = 0, col = "grey70")
```

```{r gap-numbers}
gap[gap$congress %in% c(90, 100, 110, now), c("congress","year","dem","rep","gap")]
```

> The gap was **0.584** in 1967. It is **0.925** now — the widest in this entire
> series, which reaches back to 1879.
>
> But look at the left half of that plot before you conclude anything. The 1890s
> reached **0.852**, not far below today. The genuine outlier is the middle of
> the twentieth century: the narrowest gap on record is **0.498**, in 1947. The
> shape is a U, and the era that produced most of our intuitions about how
> Congress is supposed to work sits at the bottom of it.

***

# Part 4: The overlap, and the year it ran out

A sharper question than "how far apart are the medians?" is: **is there anybody
in between?**

Count the Republicans who sit to the left of the most conservative Democrat.

```{r overlap}
overlap <- function(cg) {
  x <- h[h$congress == cg, ]
  sum(x$dim1[x$party == "Republican"] < max(x$dim1[x$party == "Democrat"]))
}
ov <- data.frame(congress = cgs, year = gap$year,
                 overlapping = sapply(cgs, overlap))

ov[ov$congress %in% 100:112, ]
```

```{r overlap-plot, fig.width = 8.5, fig.height = 4.5}
plot(ov$year, ov$overlapping, type = "h", lwd = 2, col = "#B2182B",
     xlab = "", ylab = "Republicans left of the rightmost Democrat",
     main = "The vanishing middle")
```

> **The 107th Congress — 2001 to 2002 — was the last one in which a single
> Republican sat to the left of a single Democrat.** There were two of them.
> Since the 108th, in 2003, the number has been zero every time.

***

# Part 5: Both parties moved. They did not move equally.

```{r asymmetry}
c(dem_1967 = round(med(90, "Democrat"), 3),
  dem_now  = round(med(now, "Democrat"), 3),
  dem_moved = round(med(now, "Democrat") - med(90, "Democrat"), 3))

c(rep_1967 = round(med(90, "Republican"), 3),
  rep_now  = round(med(now, "Republican"), 3),
  rep_moved = round(med(now, "Republican") - med(90, "Republican"), 3))
```

```{r asymmetry-plot, fig.width = 8.5, fig.height = 5}
plot(gap$year, gap$rep, type = "l", lwd = 2.5, col = "#B2182B",
     ylim = c(-0.6, 0.7), xlab = "", ylab = "Party median, dimension 1",
     main = "Each party's median over time")
lines(gap$year, gap$dem, lwd = 2.5, col = "#2166AC")
abline(h = 0, col = "grey70", lty = 3)
legend("topleft", c("Republicans", "Democrats"),
       col = c("#B2182B", "#2166AC"), lwd = 2.5, bty = "n")
```

> Since 1967 the Democratic median has moved **0.053** to the left. The
> Republican median has moved **0.289** to the right — **five and a half times as
> far**.
>
> This is the finding called **asymmetric polarization**, and it is one of the
> most contested claims in the study of Congress. Part 6 is about why.

***

# Part 6: What the number cannot tell you

**The scale is relative, not absolute.** DW-NOMINATE places members relative to
the votes actually held. If Congress stops voting on some issues and starts
voting on others, a member who never changed her mind can move on the scale.
"Republicans moved right" and "the agenda moved and Republicans stayed put"
produce the same picture here, and this data cannot separate them. That is the
serious objection to the asymmetry finding, and you should take it seriously.

**It measures voting, not belief.** A member who votes with her party while
privately disagreeing looks identical to a true believer.

**And it is elites, not voters.** Everything here is about 535 people. Whether
the public has polarized the same way is a *different question with different
data* — which is exactly the debate we set up on Tuesday.

```{r dim2}
sapply(c(80, 90, 100, now), function(cg) round(sd(h$dim2[h$congress == cg]), 3))
```

One more thing the method knows: there is a **second dimension**, and it used to
matter. It captured the split over civil rights that cut across both parties.
Its spread has fallen from 0.561 in 1947 to 0.315 today. American politics used
to be two-dimensional. It is now very nearly one.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — The Senate, or another era

```{r opt-a, fig.width = 8.5, fig.height = 5}
which_chamber <- "Senate"        # <- change to "House"

ch <- nom[nom$chamber == which_chamber, ]
cg2 <- sort(unique(ch$congress))
m2 <- function(cg, p) median(ch$dim1[ch$congress == cg & ch$party == p])
g2 <- data.frame(year = 1789 + 2*(cg2-1),
                 dem = sapply(cg2, m2, p = "Democrat"),
                 rep = sapply(cg2, m2, p = "Republican"))
g2$gap <- g2$rep - g2$dem

plot(g2$year, g2$gap, type = "l", lwd = 2.5, col = "#54278F",
     xlab = "", ylab = "Distance between party medians",
     main = paste("Party distance,", which_chamber))

g2[g2$year %in% c(1967, 1987, 2007, max(g2$year)), ]
```

**Write about:** does the Senate tell the same story as the House? Report the
1967 gap and the current one for your chamber. The Senate is smaller, its
members serve six years, and its states are whole states rather than drawn
districts — if districting were the main driver of polarization, the Senate
should look different. Does it? What does your answer imply about the
gerrymandering explanation you will meet in Week 15?

***

### Option B — Has this happened before?

```{r opt-b, fig.width = 8.5, fig.height = 5}
start_year <- 1879              # <- change this. Try 1930, 1950, 1980.

g <- gap[gap$year >= start_year, ]
plot(g$year, g$gap, type = "l", lwd = 2.5, col = "#54278F",
     xlab = "", ylab = "Distance between party medians",
     main = paste("Party distance from", start_year))
abline(h = gap$gap[gap$congress == 90], lty = 2, col = "grey50")

# The most and least polarized Congresses on record
rbind(head(gap[order(-gap$gap), c("congress","year","gap")], 4),
      head(gap[order( gap$gap), c("congress","year","gap")], 4))

# Congresses with no overlap at all
ov$year[ov$overlapping == 0]
```

**Write about:** start at 1980 and the story is a straight line upward. Start at
1879 and it is a U. Which is the honest picture? Note that **zero overlap is not
new** — it also happened repeatedly between 1879 and 1921. Then the harder
question: if the mid-twentieth century was the *unusual* period rather than the
current one, what does that do to the way polarization is usually discussed?

***

### Option C — Who is in the middle?

```{r opt-c}
which_congress <- 90            # <- change this. Try 100, 107, 119.

x <- h[h$congress == which_congress, ]
dmax <- max(x$dim1[x$party == "Democrat"])
dmin <- min(x$dim1[x$party == "Republican"])

cat("Congress", which_congress, "(", x$year[1], ")\n")
cat("most conservative Democrat: ", round(dmax, 3), "\n")
cat("most liberal Republican:    ", round(dmin, 3), "\n")
cat("members in the overlap:     ", sum(x$dim1 > dmin & x$dim1 < dmax), "\n\n")

mid <- x[x$dim1 > dmin & x$dim1 < dmax, c("bioname","party","state","dim1")]
head(mid[order(mid$dim1), ], 12)
```

**Write about:** who were these people, and where were they from? Run it for the
90th and you will get a long list; run it for the 107th and you will get almost
nobody. Look at the **states** in the 90th — the conservative Democrats and the
liberal Republicans were not randomly scattered. Connect that to Week 4's Solid
South: what happened to those members, and was it that they changed their minds
or that they were replaced?

***

### Option D — Ask your own question

Look up a member you have heard of. Compare your own state's delegation across
two eras. Follow whatever nags.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can read a measurement carefully enough to say what it does
and does not establish.

***

## Where the data came from

**Voteview** (<https://voteview.com>) — Jeffrey B. Lewis, Keith Poole, Howard
Rosenthal, Adam Boche, Aaron Rudkin and Luke Sonnet, *Congressional Roll-Call
Votes Database*. Every member of the House and Senate from the 46th Congress
(1879) to the present, with DW-NOMINATE coordinates.

Three things about this file worth holding on to:

- **Presidents are in the source and are dropped here.** They receive scores
  based on announced positions rather than roll calls, and folding them into a
  chamber median would compare unlike things.
- **The current Congress is still voting.** Its scores are estimated from the
  roll calls held so far and will shift slightly.
- **The scale is estimated, not observed.** DW-NOMINATE is a model fitted to
  votes; the coordinates have uncertainty that this file does not carry, and
  small differences between two members are not meaningful.

The script that downloaded, filtered and checked the file is `data/build-data.R`.
It prints the party medians, the asymmetry, and the overlap series, so the
numbers in this lab can be checked against the source in one command.
