---
title: "Data Session 12 Key: Measuring Ideology Without Reading a Single Bill"
subtitle: "84-355 Democracy's Data | Session 12 — Roll calls and the bench | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
nom <- read.csv("data/nominate_members.csv", stringsAsFactors = FALSE)
h   <- nom[nom$chamber == "House", ]
now <- max(h$congress)
med <- function(cg, p, x = h) median(x$dim1[x$congress == cg & x$party == p])
cgs <- sort(unique(h$congress))
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

DW-NOMINATE recovers a left–right scale from nothing but roll-call patterns —
nobody read a bill — and that scale shows the two parties separating until, in
2003, the last member standing between them disappeared.

***

# The findings

```{r headline}
gap <- sapply(cgs, function(c_) med(c_, "Republican") - med(c_, "Democrat"))
names(gap) <- cgs
round(c(gap_1967 = gap["90"], gap_now = gap[as.character(now)],
        widest = max(gap), narrowest = min(gap)), 3)
c(widest_year = 1789 + 2*(as.integer(names(which.max(gap))) - 1),
  narrowest_year = 1789 + 2*(as.integer(names(which.min(gap))) - 1))
```

- Party gap **0.584** in 1967 → **0.925** today: the **widest in the whole series
  back to 1879**.
- The **narrowest is 0.498, in 1947.** The shape is a U, and the mid-century is
  the outlier — not the present.
- **Asymmetry:** Democrats moved 0.053 left since 1967, Republicans 0.289 right —
  **5.5× further**.
- **The 107th Congress (2001–02) was the last with any overlap.** Two
  Republicans. Zero in every Congress since.

***

# The idea to teach

Students arrive believing polarization is new. The plot says otherwise: the
1890s reached 0.852, not far below today. What is genuinely unprecedented is not
the *distance* between the parties but the **emptiness between them** — zero
overlap has happened before (1879–1921) but never before alongside a gap this
wide.

Push on the difference between those two measures. The medians moving apart is
compatible with a populated middle; the middle emptying is a different fact and
arguably the one that matters for whether anything passes.

***

# Option A — The Senate, or another era

## Verified output

```{r a}
for (ch in c("House", "Senate")) {
  x <- nom[nom$chamber == ch, ]; cg <- sort(unique(x$congress))
  g <- sapply(cg, function(c_) med(c_, "Republican", x) - med(c_, "Democrat", x))
  names(g) <- cg
  cat(sprintf("  %-7s 1967 %.3f | now %.3f | change %+.3f\n",
              ch, g["90"], g[as.character(max(cg))], g[as.character(max(cg))] - g["90"]))
}
```

| | 1967 | now | change |
|---|---|---|---|
| House | 0.584 | 0.925 | +0.342 |
| **Senate** | **0.587** | **0.917** | **+0.331** |

**They are nearly identical.**

## What a good answer says

**3/4:** Reports both chambers and notices the similarity.

**4/4:** Draws the inference the prompt sets up. **Senators represent whole
states. Nobody draws state lines.** If gerrymandering were the main engine of
polarization, the Senate should have polarized markedly less than the House. It
did not — the two chambers moved in near-lockstep. That is real evidence against
the districting explanation, and it is worth a student's discomfort, because the
gerrymandering story is the one they arrive with.

**4/4, best:** does not overclaim in the other direction either. This shows
districting is not *sufficient* to explain polarization; it does not show
districting has no effect. Both chambers face primaries, both face the same
national donors and media, and those are shared causes that would produce this
pattern. A student who says "this rules out one story, and does not establish
another" has it right.

**Set up for the Dec 3 wrap**, where students will find the House map's efficiency gaps.
Having met this result first makes them careful about what those gaps prove.

***

# Option B — Has this happened before?

## Verified output

```{r b}
g <- data.frame(congress = cgs, year = 1789 + 2*(cgs-1), gap = round(gap, 3))
rbind(head(g[order(-g$gap), ], 4), head(g[order(g$gap), ], 4))
```

Top four are all recent (2025, 2023, 2017, 2015). The bottom four are all
mid-century. The 1890s peak was **0.852** — close to, but below, today.

## What a good answer says

**3/4:** Notices that starting in 1980 gives a rising line and starting in 1879
gives a U, and says the U is the honest picture.

**4/4:** Takes the consequence seriously. If the mid-twentieth century was the
anomaly, then "returning to normal" is not a coherent goal — the normal to which
we would return is the 1890s. And the bipartisan era that produced most of our
institutional expectations (committee power, cross-party coalitions, the idea
that the Senate works by unanimous consent) rested on a party system that
included conservative Southern Democrats, which the civil-rights realignment
ended. You cannot want the comity without the coalition that produced it.

**4/4, best:** notices that **zero overlap is not new** — it occurred repeatedly
between 1879 and 1921 — but that the *combination* of zero overlap with a
record gap is. Two measures, two different stories, and the student separates
them.

**Common stumble:** treating the choice of start year as a trick or a gotcha.
It is a real analytical decision and both plots are honest; what is dishonest is
choosing one without saying why.

***

# Option C — Who is in the middle?

## Verified output

```{r c}
for (cg in c(90, 107)) {
  x <- h[h$congress == cg, ]
  dmax <- max(x$dim1[x$party == "Democrat"]); dmin <- min(x$dim1[x$party == "Republican"])
  mid <- x[x$dim1 > dmin & x$dim1 < dmax, ]
  cat("\n", cg, "th Congress: ", nrow(mid), " members in the overlap\n", sep = "")
  if (nrow(mid)) {
    cat("  Democrats from: ",
        paste(names(head(sort(table(mid$state[mid$party=="Democrat"]), decreasing=TRUE), 6)), collapse=" "), "\n")
    cat("  Republicans from: ",
        paste(names(head(sort(table(mid$state[mid$party=="Republican"]), decreasing=TRUE), 6)), collapse=" "), "\n")
  }
}
```

**90th Congress: 204 members in the overlap.**

- Democrats in the middle: **TX, GA, NC, FL, MS, SC** — the Solid South.
- Republicans in the middle: **NY, CA, MI, PA, OH, IL** — the industrial North.

**107th Congress: 2 members.** One Texas Democrat, one New York Republican — the
last survivors of exactly those two species.

## What a good answer says

**3/4:** Notices the overlap shrinks and that the states are not random.

**4/4:** Names the two groups — conservative Southern Democrats and liberal
Northeastern Republicans — and connects to **Session 4**. The middle of Congress in
1967 was made of the two regional exceptions to the national party alignment. As
the South realigned (the 1964 crossover students measured in Session 4), both
species went extinct: Southern Democrats were replaced by Republicans, Northern
Republicans by Democrats.

**4/4, best:** answers the "changed their minds or replaced?" question and gets
it mostly right — **replacement**, not conversion. Individual members rarely
move much on DW-NOMINATE; the party medians move because different people arrive.
That is a genuinely important mechanism and it means polarization is a story
about *elections*, not about incumbents becoming more extreme in office.

**Be precise about this if a student pushes.** In `nominate_members.csv` a member
does not "rarely" move — a member *cannot* move. The file carries Voteview's
career-constant DW-NOMINATE estimate, so each member has one `dim1` for their
whole career; of the 6,936 House members in the file, 8 have more than one value,
and every one of those left Congress and later came back. Replacement therefore
accounts for **100%** of the movement in any party median computed from this
file, by construction, not as a finding. Voteview also publishes the
Nokken-Poole variant, which re-estimates each member in each Congress and does
let positions drift; a student who wants to test conversion needs that file, and
the brief says so at Step 9.

***

# Option D — Their own question

Looking up a familiar member works well. Warn against over-reading small
differences: the coordinates are model estimates and two members half a point
apart are not meaningfully distinguishable.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating DW-NOMINATE as a *measurement* with
  properties rather than a fact — noticing that it is relative to the agenda,
  that it captures voting rather than belief, and that it describes 535 people
  and not the country.
- **Reward the Senate result being taken seriously**, especially by students who
  came in confident that gerrymandering explains polarization.

## Anticipated questions

- *"How can a computer know what's liberal?"* — It cannot, and it does not try.
  It finds the arrangement of members that best predicts the votes. That the
  first dimension looks like left–right is a *finding*, not an assumption. This
  is the most important thing in the lab and is worth five minutes.
- *"Why does the current Congress look slightly off?"* — It is still voting.
  Scores are estimated from roll calls so far and will shift.
- *"Did Republicans really move 5.5× further?"* — On this measure, yes. Whether
  that means what it appears to is Part 6: if the agenda moved, a party that
  never changed can appear to move. Do not let students report the ratio without
  the caveat, and do not let them dismiss it because of the caveat either.
- *"Where are independents?"* — Dropped. Only party codes 100 and 200 are kept,
  so Sanders and King are not here. Worth mentioning; a student may notice.
- *"Are presidents in this?"* — Not in our file. They are in the source, scored
  on announced positions rather than roll calls, and mixing them into a chamber
  median would compare unlike things.

## Connections

- **Session 4 (returns over time)** — the Solid South realignment is what emptied the middle. Option C
  makes that concrete with state names.
- **The Dec 3 wrap (redistricting)** — the Senate result here should make students cautious about
  reading efficiency gaps as proof that districting causes polarization.
- **Tuesday's debate** (*Elite-Driven vs. Mass-Driven Polarization*) — this lab
  measures the elite side only, and says so. The mass side needs survey data,
  which is what Sessions 8 and 9 supply. Ask which one the causal claim requires.
