---
title: "The Age of Members of Congress"
type: brief
subtitle: "The answer depends on which row of the life table you read"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("officeholder-age-code.R")
```

```{r setup, include=FALSE}
```

Congress is older than it has ever been. That is the sentence everybody says, and it arrives with its own rebuttal attached: *of course they are older, everybody is older, life expectancy has gone up thirty years.*

Both halves are testable. **Is Congress at its oldest, and does longer life explain it?**

Age is not a proxy for competence and nothing here treats it as one. It matters for a narrower reason. **Age is a rough guide to how long ago somebody's politics were formed, and to how much of the future they will personally be present for.** The Constitution sets a floor of 25 for a representative, 30 for a senator and 35 for a president, and no ceiling at all.

## The test

The record is Voteview's member table: one row per member per Congress, from the first to the `r n(max(m$congress))`th, maintained at UCLA by Lewis, Poole, Rosenthal, Boche, Rudkin and Sonnet. It carries a `born` column, a permanent numeric identifier for each person, and `r n(nrow(m))` rows in all.

This is the right instrument because it is a complete enumeration rather than a sample. Every member of every Congress is in it, so a median is the actual middle of the chamber and not an estimate of one. The table exists to serve a scaling of roll-call votes, and what that record can bear is the subject of this cluster's [roll-call chapter](../rollcalls-source/rollcalls-source-brief.html).

Testing the rebuttal needs a second kind of evidence, from the National Center for Health Statistics. Life expectancy at birth comes from CDC's open-data catalogue as an annual series. Remaining life expectancy at 65 and at 75 had to be read out of Trend Table 4 of *Health, United States, 2019*, a published PDF. Two rows of one government table, about to give two different answers.

## The data

One row is one member in one Congress. Here are the first Congress's opening rows, after the build:

```{r cleanrow}
```

Four of those columns are not in the download. `year` is the claim that Congress *n* begins in 1787 + 2*n*, checked against two fixed points rather than trusted: the first Congress convenes in `r ck$value[ck$check == "Congress 1 begins in 1789"]` and the `r n(max(m$congress))`th in `r ck$value[ck$check == "Congress 119 begins in 2025"]`. `age` is `year` minus `born`. `entry_year` is the earliest Congress that identifier appears in that chamber, and `years_since_entry` is the difference.

**`born` is a year, not a date**, so every age here is right to within twelve months and no better. For anybody with a December birthday the subtraction overstates their age on opening day by one. Across `r n(nrow(m))` rows those errors run about half a year and point in no direction, so medians and trends survive them.

Nothing was dropped and nothing was de-duplicated. `r ck$value[ck$check == "Voteview rows"]` rows arrived and `r n(nrow(m))` remain, because **a member who serves twenty years is supposed to appear ten times**. One row is a member in one Congress, not a member. George Washington keeps his row and carries four missing values, which is why `r ck$value[ck$check == "member-Congress rows with no birth year"]` rows still have no age.

The cleaning is otherwise assertion. No person carries two conflicting birth years, a count that comes to `r ck$value[ck$check == "people carrying two different birth years"]`. Exactly `r ck$value[ck$check == "people with a birth year on some rows and not others"]` person has one on some rows and not others, and the build names him rather than patching him: `r ck$value[ck$check == "who that is"]`.

Three more tables come out of it. `age_by_congress.csv` collapses `r n(sum(m$chamber %in% c("House", "Senate")))` member-Congress rows into one per chamber per Congress, with `median_age`, `mean_age`, `mean_entry_age`, `mean_years_since_entry` and the shares aged 65, 70 and 80 or over. It also carries a third chamber called `Congress`, the two pooled, which appears nowhere in the download. `entry_tenure.csv` has one row per chamber career, `r n(nrow(et))` of them, with `entry_age` and `years_served`. `life_expectancy.csv` holds `year`, `e0_hus` and `e0_annual` for life expectancy at birth, and `e65` and `e75` for the years remaining at those ages.

The join is the calendar year, stated outright rather than left to a merge, because a Congress convenes in an odd year and a life table covers a calendar one.

> **Before you read on, commit to a number.** Between `r BY` and `r EY` the
> median member of Congress aged `r f0(dA)` years, while life expectancy at
> birth rose `r f1(dE0)` years. There is a third way to measure how long people
> live, and it is the one that applies to a sitting senator: the years a
> 65-year-old has left. **Over those same `r EY - BY` years, did it rise by
> more than `r f1(dE0)`, by something between `r f0(dA)` and `r f1(dE0)`, or by
> less than `r f0(dA)`?** Write a number down before you read on.

## What it says about democracy

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 1. Two and a third centuries of congressional age.** Median age of each chamber at every Congress, with the shaded band covering the middle half of the chamber. A line over time is the right form because the claim is about a trend, and the band is there because a median alone hides how spread out a chamber is. Each green dot is a president at the start of a term.

```{r step2-tab}
```

**The first half of the claim holds.** Congress is at or near its oldest recorded point, and the median senator is about `r f0(val("Senate", NOW) - val("House", NOW))` years older than the median representative. That gap is old: the Senate has run above the House for most of two centuries, doing what six-year terms and a higher age floor were meant to do.

But the rise is not a long slide. Congress stood at `r f0(TR$median_age)` in `r TR$year`, the first time it had been that young since `r PRIOR`. It reached `r f0(max(CC0$median_age))` by `r min(OLDEST)`. Almost the whole increase people argue about happened inside a single working lifetime.

Now the rebuttal, stated as strongly as it deserves. Between `r BY` and `r EY` American life expectancy at birth rose from **`r f1(E0a)`** years to **`r f1(E0b)`**, a gain of **`r f1(dE0)` years**. Over the same span the median member of Congress aged from `r f0(A0)` to `r f0(A1)`, a gain of **`r f0(dA)` years**. On that comparison Congress aged *less* than the country it governs, and the brief could stop here.

It is a fair comparison and it is not the only one. `r f1(E0b)` is life expectancy **at birth**: the average length of a life for a baby born in `r EY`. It rose so far across the twentieth century mostly because babies and children stopped dying. Antibiotics, sanitation, vaccination and obstetrics are enormous achievements, all of them already behind a sitting senator.

```{r step4-tab}
```

The quantity that describes an officeholder's remaining life is the years left at an age they have reached, and NCHS prints it one section further down the same page.

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=3.0}
```

**Figure 2. The same `r EY - BY` years, measured against two rows of one government table.** Life expectancy at birth rose `r f1(dE0)` years, the median age of Congress rose `r f0(dA)`, and the remaining life expectancy of a 65-year-old rose `r f1(dE6)`.

```{r step4-share}
```

**At most `r f0(REL)`% of the famous rise in life expectancy is a rise in how long a 65-year-old lives.** The rest is people who used to die young and now do not, which transforms the average life and changes a senator's prospects not at all.


So the honest answer to the rebuttal is two answers, and both belong on the page. Measured against life expectancy at birth, Congress aged `r f0(dA)` years while the yardstick grew `r f1(dE0)`, and Congress aged slower than the country. Measured against the years a 65-year-old has left, the yardstick grew `r f1(dE6)`, and Congress aged faster. Same data, same source, same printed table, opposite conclusions.

The choice between them is a choice of denominator, the bottom number you divide by. Nothing in either file makes it, and a denominator picked without thinking can reverse the sign of an answer. The at-65 row is the one that describes people who have already reached the ages members of Congress are, which is why this brief leads with it and reports the other beside it.

A median can rise two ways: new members arrive older, or sitting members stay longer. These have different causes and different remedies, and people assert one or the other without checking. Checking is possible because the two add up exactly, for every member at every Congress:

$$\text{age now} \;=\; \text{age when they first arrived} \;+\; \text{years since}$$

That is an identity rather than a model, so it survives averaging.

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

**Figure 3. Mean age of Congress, split into its only two components.** The lower band is the average age at which sitting members arrived, the upper band the average time they have been there since. The two sum exactly to the mean age, the dark line. Between `r BY` and `r EY` the mean rose `r sgn(DC["total"], 2)` years, of which `r sgn(DC["entry"], 2)` is arrival age and `r sgn(DC["stay"], 2)` is time served.

```{r step5-tab}
```

**For Congress as a whole, `r f0(100 * DC["entry"] / DC["total"])`% of the rise between `r BY` and `r EY` is people arriving older.** The popular explanation, that incumbents cling on, is the smaller half. And the two chambers disagree: in the House `r f0(100 * DH["entry"] / DH["total"])`% of the rise is arrival age, in the Senate `r f0(100 * DS["entry"] / DS["total"])`%. Most of the Senate's ageing really is senators staying, which six-year terms and very high re-election rates would predict.

## What this brief cannot tell you

**Ages are approximate to within a year**, because `born` is a year, and `r ck$value[ck$check == "member-Congress rows with no birth year"]` rows have none at all. Those sit in the first twenty Congresses, so the earliest part of Figure 1 rests on fewer people than it appears to.

**The at-65 series is sparse and begins in `r ck$value[ck$check == "e(65) series first year"]`.** It is published for `r ck$value[ck$check == "years with a published e(65)"]` years up to `r ck$value[ck$check == "e(65) series last year"]`, and nothing between them has been filled in. So the most tempting version of the comparison, 1900 against today, cannot honestly be made and is not made here.

**Two NCHS sources disagree in `r ck$value[ck$check == "years where they disagree"]` years.** On `r ck$value[ck$check == "of those, years where the two agree exactly"]` of `r ck$value[ck$check == "years where HUS and CDC both give e(0)"]` shared years they agree to the tenth. That is small and real, and reported rather than resolved.

**Members used to die in office far more often**, and every such death removed an old member from every later Congress here. That biases the historical series downward, in the direction that flatters the reading that Congress used to be young.

**No measure of what the institution does appears in this file.** Seniority concentrates power, so a chamber's leadership may matter more than its median member, and neither is measured here.

## What you should have learned

- Congress is at or near its oldest recorded point, with a median member of `r f0(val("Congress", NOW))` and a median senator of `r f0(val("Senate", NOW))`, and the rise is recent rather than a long slide.
- "People live longer now" gives two opposite answers from one table. Between `r BY` and `r EY` Congress aged `r f0(dA)` years, life expectancy at birth rose `r f1(dE0)`, and the years remaining to a 65-year-old rose `r f1(dE6)`.
- Only `r f0(REL)`% of the gain in life expectancy is about people who already reached 65, so the yardstick that flatters Congress is the one describing deaths it never faced.
- The rise is mostly people arriving older, not staying longer: `r f0(100 * DC["entry"] / DC["total"])`% of it for Congress as a whole, and `r f0(100 * DH["entry"] / DH["total"])`% in the House against `r f0(100 * DS["entry"] / DS["total"])`% in the Senate.
- The top of the distribution moved further than the middle: the share of Congress aged 70 or over went from `r f1(val("Congress", BY, "pct_70plus"))`% to `r f1(val("Congress", NOW, "pct_70plus"))`%, and in the Senate it is now `r f1(val("Senate", NOW, "pct_70plus"))`%.

## Extensions

None of these is answered above, and every one can be answered by sorting or grouping in a spreadsheet.

- `age_by_congress.csv` has `median_age` and `mean_entry_age`. Filter `chamber` to Congress and compare the two columns decade by decade. Which one turns first?
- Join `age_by_congress.csv` to `life_expectancy.csv` on `year`. Compare the rise in `median_age` against `e0_annual`, then against `e65`. Which yardstick would you put in a headline?
- `entry_tenure.csv` has `entry_age` and `years_served` for every career. Group by the decade of `entry_year` and take the median of each. Has the path into Congress changed?
- `age_by_congress.csv` splits by `chamber`. Subtract the House `median_age` from the Senate's, Congress by Congress. Is the gap growing, shrinking, or steady?

Two stretch ideas point past the spreadsheet. `members.csv` has `died` as well as `born`, so you can count member-terms ending in a death in office by decade and ask how much of the historical series that explains. And no series here measures the age of the country itself: download the Census Bureau's median age by year and ask whether Congress has aged faster than its electorate.

## Sources

**Data**

- Lewis, J.B., Poole, K., Rosenthal, H., Boche, A., Rudkin, A. and Sonnet, L.
  *Voteview: Congressional Roll-Call Votes Database.*
  <https://voteview.com/static/data/out/members/HSall_members.csv>
  `r ck$value[ck$check == "Voteview rows"]` member-Congress records,
  Congresses 1 to `r max(m$congress)`.
- National Center for Health Statistics. *Life expectancy at birth, by race and
  sex, 1900-2018.* CDC open data.
  <https://data.cdc.gov/api/views/w9j2-ggv5/rows.csv?accessType=DOWNLOAD>
- National Center for Health Statistics. *Health, United States, 2019*, Trend
  Table 4: life expectancy at birth, at age 65 and at age 75.
  <https://www.cdc.gov/nchs/data/hus/2019/004-508.pdf>
  Read out of the published PDF and checked against the open-data series above
  across the `r ck$value[ck$check == "years where HUS and CDC both give e(0)"]` years they share.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`age_by_congress.csv`](data/derived/age_by_congress.csv), [`entry_tenure.csv`](data/derived/entry_tenure.csv), [`life_expectancy.csv`](data/derived/life_expectancy.csv), [`members.csv`](data/derived/members.csv), [`presidents.csv`](data/derived/presidents.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
