---
title: "Campaign Finance Disclosure Is Not an Audit"
subtitle: "The only source in this book written by the people it is about"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("finance-source-code.R")
```

```{r setup, include=FALSE}
```

Every other source in this book is a record made about its subject by somebody else. A census counts you. A state records that you voted. A clerk writes down how a member of Congress voted. The subject does not hold the pen.

Campaign finance data is the exception, and the exception is structural:

> **The subject of the record writes the record.**

Committees file reports about themselves. The Federal Election Commission receives those reports, publishes them, and does not verify them first. *Disclosure* is a legal obligation to report. It is not an audit, and the difference between those two things is the subject of this chapter.

## What that difference costs

Here is the 2024 independent-expenditure file, meaning money spent by outside groups for or against federal candidates. First exactly as it arrives from the FEC, then again with sixteen rows removed.

```{r filedtab}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=3.6}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The same file, before and after removing sixteen rows.** Not a change of size. A change of sign. Hover a segment for the dollars behind it, which is where the reversal becomes readable.

The opposing total is identical in both bars (`r bn(fil$opposing[1])`). It is the supporting total that collapses, from `r bn(fil$supporting[1])` to `r bn(fil$supporting[2])`. The dashed line is 50%.

As filed, the file says outside spending was **`r p1(RAWP)`% supportive** — `r bn(RAWS)` for candidates against `r bn(fil$opposing[1])` opposing them. That is a finding. It is the kind of sentence that appears in a news story about the positivity of a campaign.

Remove `r NJUNK` rows and the same file says outside spending was **`r p1(OPPP)`% opposing**.

The junk does not inflate the total and leave the shape intact. **It reverses the conclusion.**

```{r outtab}
```

`r NJUNK` rows account for `r bn(JUNK)` — more than the entire real total. The largest is `r usd(TOPJ)`, filed by a committee calling itself *`r TOPS`*, on behalf of a candidate who is not a national figure.

Nothing about this is hidden or hard to find. It is in the official bulk download, published by the agency, in the same columns as every real transaction. There is no flag, no footnote, and no quality code. The file is a faithful record of **what was filed**, which is exactly what a disclosure regime promises and is not the same thing as a record of what happened.

> **Before you read on, commit to a view.** The FEC could refuse to publish
> filings that look like these until they are checked. **Name one cost of that
> policy** — and say who bears it.

## An average that describes nobody

The candidate file has the same character in a quieter way.

```{r disttab}
```

Mean receipts are `r usd(MEANR)`. Median receipts are `r usd(MEDR)`. The mean is **`r p1(RATIO)` times** the median.

The reason is that the file contains everyone who filed, and filing is not the same as running a campaign. `r nn(ZERO)` of the `r nn(NCAND)` candidates, about a quarter, raised **nothing at all**. Meanwhile the top one per cent account for `r p1(TOP1)`% of all money raised.

"The average candidate raised `r usd(MEANR)`" is arithmetically correct and describes no one in the file. This is the book's recurring denominator problem wearing a dollar sign. The denominator is the bottom number you divide by, and here it is the set of candidates being averaged over. That set was defined by an administrative act, and it is not the population anyone has in mind.

## What the regime does and does not require

```{r regtab}
```

Three boundaries are worth naming, because each one makes something structurally unobservable rather than merely unreported.

**The itemisation threshold.** Contributions from a person aggregating $200 or less in a cycle are not itemised. Small donors exist in the file only as a lump sum. So *how many people gave to this campaign* cannot be answered from the itemised records, and the itemised records are what everybody uses.

**Donor disclosure for 501(c)(4) organisations.** A social welfare organisation that makes independent expenditures reports the spending and need not report who funded it. The money appears; its origin does not.

**Coordination.** Whether spending was coordinated with a campaign is a *legal question*, decided after the fact if at all. It is not a field. A file that separates "independent" from "coordinated" spending is reporting a filer's characterisation, not an established fact.

> **A second prediction.** Suppose the itemisation threshold were lowered from
> $200 to $0. **What new question could you answer** — and what would happen
> to the file's usefulness for the questions you can answer now?

## What this chapter cannot tell you

**Nothing here is evidence of fraud, and the sixteen rows are almost certainly not a scheme.** They look like what they probably are: nonsense filings, or a filer's error, sitting unremarked in an unverified system. The lesson is not that someone tried to deceive anybody. It is that a system which publishes what it receives will publish whatever it receives, and the reader is the only quality control there is.

This chapter also does not measure **influence**. The file records money moved between entities. What that money bought is not in it: access, attention, a changed vote, or nothing at all. No amount of transaction detail gets there. The [lobbying chapter](../lobbying/lobbying-brief.html) is the closest this book comes, and it is careful to say the same thing.

The figures are one cycle, 2024, and federal only. State campaign finance is fifty more regimes with fifty more thresholds, and the boundaries described here are federal ones.

Finally, a word in the FEC's defence, because the chapter has been hard on the file. Disclosure was the reform. Before the Federal Election Campaign Act there was no obligation to say who paid for a campaign at all. That anyone can now download every itemised contribution in a federal election is an achievement of transparency law, not a failure of it. That the resulting file requires care is the price of it being public quickly and without a gatekeeper deciding what is fit to publish. **A slower, cleaner, verified file would be a different and less democratic instrument.**

## Sources

- Federal Election Commission. *Independent Expenditure* bulk file, 2024
  cycle, via this book's
  [independent-expenditures chapter](../independent-expenditures/independent-expenditures-brief.html).
  <https://www.fec.gov/data/browse-data/?tab=bulk-data>
- Federal Election Commission. *All Candidates* bulk file (`weball24`), 2024
  cycle, via this book's
  [campaign-finance chapter](../campaign-finance/campaign-finance-brief.html).
  <https://www.fec.gov/data/browse-data/?tab=bulk-data>
- Federal Election Campaign Act of 1971, as amended, 52 U.S.C. §30101 et seq.
- *Citizens United v. Federal Election Commission*, 558 U.S. 310 (2010).
- *SpeechNow.org v. FEC*, 599 F.3d 686 (D.C. Cir. 2010).
- La Raja, Raymond J. and Brian F. Schaffner (2015). *Campaign Finance and
  Political Polarization: When Purists Prevail.* University of Michigan Press.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`distribution.csv`](data/derived/distribution.csv), [`filed.csv`](data/derived/filed.csv), [`outliers.csv`](data/derived/outliers.csv), [`regime.csv`](data/derived/regime.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `outliers.csv` holds filings with implausible amounts. Read a few. Nobody checked these before publication — work out what that means for any total computed from the raw file.
- `filed.csv` shows how the file reads as it arrives versus what it means. Work out the share supporting and opposing. Which reading would a journalist publish by mistake?
- `regime.csv` says who must report and `what_is_invisible`. List the money that never appears in this file at all. How much of campaign finance is being disclosed?
- `distribution.csv` summarises the spread of candidate receipts. Find a summary statistic in it that would mislead somebody, and say who it would mislead.

