---
title: "Part V — Records of Ordinary People"
subtitle: "A record about you that you never volunteered for"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-5-ordinary-people-code.R")
```

```{r setup, include=FALSE}
```

The records in this part look like the ones in Part IV. An institution wrote something down about a person while doing its job, and the writing survived.

**One thing is different, and it changes everything: nobody in these records volunteered.** A member of Congress chose to run for office. The person in a traffic-stop record was driving to work. The people graded on a 1937 map of Philadelphia were not consulted, were mostly not told, and in many cases never learned that a federal agency had assigned their neighbourhood a letter.

## Made to run a system, not to describe anybody

Every record here exists to move a process along.

A stop record exists so an officer can account for their shift. A jury summons exists so a court can seat twelve people. A lending grade existed so a government corporation could decide where to insure mortgages. In none of those cases was the goal to describe the person — the description is a by-product, and the by-product is what we are reading.

This is why administrative records fail in a particular way. They are extremely good evidence about **what the institution did**, and much weaker evidence about **who the people were**. A police department's stop data tells you a great deal about that department's behaviour and rather little about the population of drivers, because the department only wrote down the people it stopped.

## A numerator with no denominator

That last point is the single most useful idea in this part, so it is worth stating plainly.

An administrative record almost always gives you a **top number** and no **bottom number**. The bottom number, the thing you divide by, is called the **denominator**. You will need one for every rate in this part. You get the people who were stopped, not the people who were driving. The people summoned, not the people eligible. The neighbourhoods graded, not the neighbourhoods that existed.

Every serious question you can ask of this data needs a bottom number. You have to go somewhere else to get one, usually to Part I. Which bottom number you pick can change the answer by a factor of several, and one chapter here shows exactly that happening.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

## Where this part sits

```{r reusetab}
```

## This part is four chapters, and that is a gap rather than a design

The title of this part promises more than it currently delivers, and you should know that going in.

The clearest modern case is **commercial location data**: the pings a phone sends out, bought and sold by companies most people have never heard of. It is used to estimate who visited a protest, a church or a clinic. Nobody whose phone did the pinging volunteered for any of it. It belongs in this part, and it is not here yet.

What is here is the government half: policing, courts, and a historical lending map. It is enough to teach the idea. It is not yet enough to cover the subject.

## How to read this part

Read [the source chapter](../admin-records-source/admin-records-source-brief.html) first. It sets up the missing-denominator problem that all three of the others run into. Then take whichever of the three cases interests you. They do not depend on each other.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens.
- The individual records — police stop data, court and jury records, and the
  Home Owners' Loan Corporation security maps — are named, dated and cited in
  the chapters that use them.
