---
title: "Data Session 12, part 2: Building a Score From Scratch"
subtitle: "84-355 Democracy's Data | The Supreme Court's votes, 1946–2023"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"The most conservative Court since the 1930s."*

This morning you used DW-NOMINATE: a number that places each member of Congress
on a left–right line, computed from roll calls. You took the number as given.

**Today you build one.**

The Supreme Court is the easiest place to do it. Nine people, every vote
recorded, every case public, and a database going back to 1946 that has counted
all of it. And unlike Congress, the number of people is small enough that you
can see the whole thing at once.

There *are* published scores for the Court — Martin-Quinn is the standard one.
We are not using them, for a reason worth stating: **a published score is the
answer, and this session is about how the answer gets made.**

# Part 1 — What one row is

```{r load}
ag <- read.csv("data/agreement.csv", stringsAsFactors = FALSE)
jj <- read.csv("data/justices.csv",  stringsAsFactors = FALSE)

nrow(jj)
jj
```

Ten justices sat between the 2020 and 2023 terms. For each we have how many
cases they voted in, what share of their votes the database codes as
**conservative**, and how often they were **in the majority**.

Note that Breyer and Jackson have about half as many cases as the others —
Breyer retired during the window and Jackson replaced him. That will matter in
Part 4.

# Part 2 — Who agrees with whom

The raw material for any ideal-point score is embarrassingly simple: **how often
does each pair of people vote the same way?**

```{r agreement}
head(ag[order(-ag$pct), c("a", "b", "pct", "cases")], 5)
head(ag[order( ag$pct), c("a", "b", "pct", "cases")], 5)
```

**Kavanaugh and Roberts agreed 96.2% of the time.** Kagan and Sotomayor, 92.3%.
At the other end, **Thomas and Sotomayor agreed 53.0%** — barely better than a
coin.

That is the whole dataset. Forty-five numbers, one per pair. **No ideology has
been supplied to it.** Nobody has told this table who is liberal.

Except one of the forty-five is not a number at all:

```{r missing}
ag[ag$cases == 0, ]
```

**Breyer and Jackson never sat on a case together** — she replaced him. Their
agreement is not low or high; it is **undefined**, and no amount of data will
fix it. Hold that thought for thirty seconds.

# Part 3 — The score falls out

Now the move that makes DW-NOMINATE, Martin-Quinn, and every other ideal-point
score work. Take the agreement rates as a measure of **closeness**, treat
disagreement as **distance**, and ask: if these ten people had to be arranged
along a single line so that people who agree sit near each other, what line
would that be?

```{r scaling}
js <- sort(unique(c(ag$a, ag$b)))
M <- matrix(NA_real_, length(js), length(js), dimnames = list(js, js))
diag(M) <- 1
for (i in seq_len(nrow(ag))) {
  if (ag$cases[i] == 0) next          # Breyer and Jackson: no shared cases
  r <- ag$agree[i] / ag$cases[i]
  M[ag$a[i], ag$b[i]] <- r
  M[ag$b[i], ag$a[i]] <- r
}

# The scaling cannot take a hole. We fill the one missing pair with the average
# agreement across every other pair -- which is a GUESS, and the most
# conservative one available: it assumes nothing about whether they would have
# agreed. Part 6 Option B asks what happens if you refuse to guess.
M[is.na(M)] <- mean(M[upper.tri(M)], na.rm = TRUE)

score <- cmdscale(as.dist(1 - M), k = 1)[, 1]
round(sort(score), 3)
```

**That fill is a modeling decision, made by us, in one line.** Option B has you
remove it, and when you do, **two adjacent pairs swap places** — Thomas and
Alito, and Roberts and Kavanaugh. Published scores make dozens of choices like
this one. The difference is that ours is visible and reversible.

```{r scale-plot}
o <- sort(score)
plot(o, rep(0, length(o)), pch = 19, cex = 1.6, col = "#2c7fb8",
     yaxt = "n", ylab = "", xlab = "position recovered from agreement alone",
     ylim = c(-1, 1), xlim = range(o) * 1.25)
text(o, 0.25, names(o), srt = 45, adj = 0, cex = 0.75)
abline(h = 0, col = "grey70")
```

**Read the order.** Thomas, Alito, Gorsuch, Barrett, Roberts, Kavanaugh — then a
gap — then Jackson, Kagan, Breyer, Sotomayor.

That is the ideological ordering of the Roberts Court as every court reporter in
America would give it. **It came out of forty-five agreement rates.** We never
told the model that Thomas is conservative, that these are political questions,
or even that the people are judges.

```{r validate}
jj$score <- score[jj$justice]
round(cor(jj$score, jj$pct_conservative), 3)
```

**Correlation −0.975** with each justice's share of conservative votes — a
variable the scaling never saw. That is what an ideal-point score *is*: a
summary of agreement that turns out to line up with something we already had a
word for.

**And notice the gap.** The six on the left span **0.145**. The distance from
Kavanaugh to Jackson is **0.224**, larger than the entire conservative bloc.
**This is not a spectrum with people spread along it. It is two clusters.**

# Part 4 — The number that surprises people more

```{r majority}
jj[order(-jj$pct_in_majority), c("justice", "pct_conservative", "pct_in_majority")]
```

```{r maj-plot}
plot(jj$score, jj$pct_in_majority, pch = 19, col = "#C41230", cex = 1.3,
     xlab = "position (left = conservative here)", ylab = "% of the time in the majority")
text(jj$score, jj$pct_in_majority, jj$justice, pos = 3, cex = 0.65)
```

**Kavanaugh is in the majority 95.5% of the time. Roberts, 94.3%.** The lowest
are Sotomayor at 70.7% and Breyer at 72.1% — and, on the other side, Thomas at
79.3%.

The two justices nearest the middle of that line win almost every case. The
justices at either end lose most often. **"Moderate" here is not a description
of anyone's opinions — it is a description of how often you are on the winning
side**, and it is a mechanical consequence of where you sit.

# Part 5 — Seventy-eight terms

```{r long}
bt <- read.csv("data/by_term.csv", stringsAsFactors = FALSE)
c(terms = nrow(bt), from = min(bt$term), to = max(bt$term))
```

```{r long-plot}
plot(bt$term, bt$pct_conservative, type = "l", col = "grey60",
     xlab = "term", ylab = "% of decisions coded conservative")
lines(bt$term, filter(bt$pct_conservative, rep(1/5, 5), sides = 2),
      col = "#C41230", lwd = 2)
abline(h = 50, lty = 3)
```

The same database goes back to **1946**, so the long series is free. The red line
is a five-term moving average.

**Before believing any story you see in it**, notice what is being plotted: the
share of decisions the database's coders labelled conservative. That is a human
judgement, applied case by case, for seventy-eight years, by people following a
codebook. It is a good codebook. It is still a judgement, and it is the thing the
whole series is made of.

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Two dimensions.** Change `k = 1` to `k = 2` in the scaling and plot
both. Does the second dimension separate anybody the first one merged? If it
does, what might it be picking up — and how would you tell that from noise?

**Option B — Refuse to guess.** Part 3 filled the Breyer–Jackson hole with the
average. Instead, drop **both** of them from `ag` and re-run the scaling on the
eight justices who overlap completely. Does the order of those eight change?
What does the comparison tell you about how much the fill was doing?

**Option C — The close cases.** `close_cases.csv` has the 30 one-vote decisions
in this window and who was in the majority for each. Count how often each
justice was in the majority *in those cases only*, and compare to Part 4. Does
the middle still win when it is close?

**Option D — Your own question.** 45 pairs, 10 justices, 78 terms.

# What this lab cannot tell you

- **Whether "conservative" means the same thing across seventy-eight years.**
  The codebook is consistent; the politics are not. A 1962 case coded
  conservative and a 2022 one may have almost nothing in common.
- **Why anyone voted as they did.** Agreement is not agreement *about reasons*.
  Two justices can reach the same result from incompatible doctrines, and this
  data cannot see the difference — which is most of what law professors argue
  about.
- **That the line is real.** We *chose* one dimension and it fit well. That does
  not prove judicial disagreement is one-dimensional; it proves a single
  dimension explains much of the variation in these votes, in this window.
- **What Martin-Quinn would say.** The published scores use a more careful model
  that lets positions move over time. Ours is a snapshot, and it cannot show
  anyone changing.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- We recovered the standard left–right ordering of the Court without ever
  mentioning ideology. Is that impressive, or is it circular?
- The two justices in the middle win 95% of cases and the ones at the ends lose
  a third of them. Should that bother anybody?
- Every number in Part 5 depends on a coder deciding what "conservative" means.
  Does that make the series useless, or just human?

***

### Where the data came from

**The Supreme Court Database**, Washington University in St. Louis, 2024 release
01, justice-centered by citation — every justice's vote in every case since
1946. <http://scdb.wustl.edu/data.php>

Note the protocol: **the site serves over `http` only**, and `https` fails.

Agreement and scaling use the **2020–2023 terms**, and cases where the database
records no ideological direction are excluded, because "voted the same way" is
undefined without a direction to compare. The build is in `data/build-data.R`.
