---
title: "Key: Building a Score From Scratch"
subtitle: "84-355 Democracy's Data | Session 12, part 2 — the Supreme Court's votes | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)

ag <- read.csv("data/agreement.csv",   stringsAsFactors = FALSE)
jj <- read.csv("data/justices.csv",    stringsAsFactors = FALSE)
bt <- read.csv("data/by_term.csv",     stringsAsFactors = FALSE)
cc <- read.csv("data/close_cases.csv", stringsAsFactors = FALSE)

js    <- sort(unique(c(ag$a, ag$b)))
NJ    <- length(js)
NPAIR <- nrow(ag)
HOLE  <- ag[ag$cases == 0, ]

nice <- function(x) sub("^[A-Z]+(?=[A-Z][a-z])", "", x, perl = TRUE)
nm   <- function(x, k = 3) formatC(x, format = "f", digits = k)
pc   <- function(x, k = 1) formatC(x, format = "f", digits = k)

# --- the scaling exactly as the lab runs it --------------------------------
build <- function(a, jsx = js) {
  m <- matrix(NA_real_, length(jsx), length(jsx), dimnames = list(jsx, jsx))
  diag(m) <- 1
  for (i in seq_len(nrow(a))) {
    if (a$cases[i] == 0) next
    r <- a$agree[i] / a$cases[i]
    m[a$a[i], a$b[i]] <- r; m[a$b[i], a$a[i]] <- r
  }
  m
}
M    <- build(ag)
FILL <- mean(M[upper.tri(M)], na.rm = TRUE)
M[is.na(M)] <- FILL
sol   <- cmdscale(as.dist(1 - M), k = 2, eig = TRUE)
score <- sol$points[, 1]
ord   <- sort(score)
R     <- cor(score[jj$justice], jj$pct_conservative)

CONS  <- names(ord)[1:6]; LIBS <- names(ord)[7:NJ]
SPAN  <- as.vector(ord[6] - ord[1])
CHASM <- as.vector(ord[7] - ord[6])
GAPS  <- diff(ord)

# --- the scaling refusing to guess -----------------------------------------
OUT <- c(HOLE$a, HOLE$b)
a2  <- ag[!(ag$a %in% OUT | ag$b %in% OUT), ]
js2 <- sort(unique(c(a2$a, a2$b)))
sol2   <- cmdscale(as.dist(1 - build(a2, js2)), k = 2, eig = TRUE)
score2 <- sol2$points[, 1]
ord2   <- sort(score2)
R2     <- cor(score2[js2], jj$pct_conservative[match(js2, jj$justice)])
keep   <- names(ord)[names(ord) %in% js2]
MOVED  <- sum(keep != names(ord2))

# --- close cases -----------------------------------------------------------
blocs <- strsplit(cc$majority_bloc, " ")
inmaj <- table(unlist(blocs))
jj$close_maj <- as.vector(inmaj[jj$justice])
# a justice's real denominator is the one-vote cases in the terms they sat for.
# Tenure is inferred from the terms in which they appear in any one-vote
# majority; every justice here appears in both their first and last term.
jj$close_n <- sapply(jj$justice, function(j) {
  t <- cc$term[sapply(blocs, function(v) j %in% v)]
  sum(cc$term >= min(t) & cc$term <= max(t))
})
jj$close_pct <- 100 * jj$close_maj / jj$close_n
# did the recovered blocs decide the one-vote cases?
nc <- sapply(strsplit(cc$majority_bloc, " "), function(v) sum(v %in% CONS))
BLOC_WINS <- sum(nc == 5)

# --- how much of the ordering is sampling noise ----------------------------
set.seed(2026)
B <- 2000; ord0 <- names(ord); same <- 0; flipTA <- 0; flipRK <- 0
cor1 <- cor2 <- numeric(B)
for (b in seq_len(B)) {
  a <- ag; ok <- a$cases > 0
  a$agree[ok] <- rbinom(sum(ok), a$cases[ok], a$agree[ok] / a$cases[ok])
  mm <- build(a); mm[is.na(mm)] <- mean(mm[upper.tri(mm)], na.rm = TRUE)
  pp <- cmdscale(as.dist(1 - mm), k = 2, eig = TRUE)$points
  o  <- names(sort(pp[, 1]))
  if (identical(o, ord0)) same <- same + 1
  if (which(o == "SAAlito")     < which(o == "CThomas"))   flipTA <- flipTA + 1
  if (which(o == "BMKavanaugh") < which(o == "JGRoberts")) flipRK <- flipRK + 1
  cor1[b] <- abs(cor(pp[, 1], sol$points[, 1]))
  cor2[b] <- abs(cor(pp[, 2], sol$points[, 2]))
}
SAME <- 100 * same / B; FTA <- 100 * flipTA / B; FRK <- 100 * flipRK / B
D1COR <- median(cor1); D2COR <- median(cor2)
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`scdb-brief.Rmd`) is a chapter, "Forty-Four Numbers and
a Guess". It is organised by argument under prose headings and carries four
numbered figures — Figure 1, the agreement matrix in both orders, with the
toggling D3 panel; Figure 2, the scaling with the guess and without it; Figure 3,
majority rates against position; Figure 4, the by-term series. It no longer has
numbered steps. The lab (`scdb.Rmd`) runs the same sequence in Parts 1–6 with the
code exposed and ends with four options. **Students should read the brief first.**
**The Step numbers below are this key's own walking order for the session**, the
eleven rows of the timing table, and the lab's Parts map onto them (Part 1 =
Step 1, Part 2 = Steps 2 and 6, Part 3 = Steps 3–7, Part 4 = Step 8, Part 5 =
Step 10, Part 6 = the options). Where the brief is meant it is named, and cited by
figure number or heading.

**This is the second half of Session 12.** The morning lab (`dw-nominate`) hands
students a published ideal-point score and asks them to read it. This one builds
one in front of them, from `r NPAIR - nrow(HOLE)` numbers, and then breaks it.
The pairing is the point: the same method, once as a consumer and once as a
manufacturer.

Every number below was computed from the files in `data/` at knit time. **The
brief computes its numbers the same way; the lab does not.** `scdb.Rmd` prints
live output but states the figures in prose by hand — `r pc(max(ag$pct, na.rm = TRUE))`%,
`r nm(R)`, `r nm(CHASM)`, and the rest. They are all correct against the current
files, checked. If `data/build-data.R` is ever re-run against a newer SCDB
release, **the lab's prose is the thing that goes stale**, and this key and the
brief will not warn you, because they will simply report the new numbers.

***

# The session in one sentence

Feed nothing but "how often did these two vote the same way" into a
one-dimensional scaling and the conventional left–right ordering of the Roberts
Court falls out — but one of the `r NPAIR` cells contains no data, and what you
put there decides the fine ordering, so the blocs are in the data and the ranking
is in the assumptions.

***

# The findings

```{r headline}
c(justices          = NJ,
  pairs             = NPAIR,
  pairs_with_data   = NPAIR - nrow(HOLE),
  fill_value        = round(FILL, 3),
  cor_with_fill     = round(R, 3),
  cor_refusing      = round(R2, 3),
  bloc_span         = round(SPAN, 3),
  chasm             = round(CHASM, 3),
  positions_moved   = MOVED,
  one_vote_cases    = nrow(cc),
  bloc_only_wins    = BLOC_WINS,
  terms             = nrow(bt))
```

```{r headline-2}
# all nine adjacent gaps; the fill only ever decides the two smallest
round(GAPS, 4)

# how often the recovered order survives resampling the votes
c(exact_order_pct   = SAME,
  alito_before_thomas_pct    = FTA,
  kavanaugh_before_roberts_pct = FRK)
```

- **The ordering is `r paste(nice(CONS), collapse = ", ")`, then a gap, then
  `r paste(nice(LIBS), collapse = ", ")`** — recovered from
  `r NPAIR - nrow(HOLE)` agreement rates with no ideology supplied.
- **It correlates `r nm(R)`** with each justice's share of conservative votes, a
  variable the scaling never saw. Refusing to guess, on `r length(js2)`
  justices: **`r nm(R2)`**.
- **It is two clusters, not a spectrum.** The six on one side span
  `r nm(SPAN)`; the single step across the middle is **`r nm(CHASM)`**.
- **One cell of `r NPAIR` has no data.** `r nice(HOLE$a)` and `r nice(HOLE$b)`
  never sat together. Filling it with the mean moves `r MOVED` positions.
- **And those positions were never determined anyway.** The two pairs that swap
  are separated by `r nm(GAPS[1], 4)` and `r nm(GAPS[5], 4)` on a scale spanning
  `r nm(diff(range(ord)))`. Resample the votes and the full order comes back
  **`r pc(SAME)`% of the time**; Alito and Thomas trade places on essentially a
  coin flip.
- **Kavanaugh is in the majority `r pc(max(jj$pct_in_majority))`% of the time**
  and Roberts `r pc(jj$pct_in_majority[jj$justice == "JGRoberts"])`%. In the
  `r nrow(cc)` one-vote decisions both fall to
  **`r pc(max(jj$close_pct))`%**.
- **Only `r BLOC_WINS` of the `r nrow(cc)` one-vote decisions** were won by a
  majority drawn entirely from the six-justice bloc the scaling found. Two thirds
  of the closest cases cut across the line the lab just built.
- The long series runs **`r nrow(bt)` terms, `r min(bt$term)` to
  `r max(bt$term)`**, every value a coder's judgement about direction.

***

# The three things to get said

**1. The blocs are in the data; the fine ordering is in the assumptions. This is
the pivot.** Steps 6 and 7 are the session. A student who leaves knowing only
that one empty cell reorders the justices has got the course's transferable
lesson: **every published score has a fill**, a cell with nothing in it and a
decision about what to put there, and the question to ask of any such number is
not whether it made assumptions but which ones and what happens if you change
them. Make somebody state the asymmetry out loud — the coarse structure survives
refusing to guess (`r nm(R2)` on `r length(js2)` justices, blocs unchanged) and
the fine structure does not.

**And go one step past the brief.** The two pairs that swap are separated by
`r nm(GAPS[1], 4)` and `r nm(GAPS[5], 4)`. They were never distinguished in the
first place. The fill is not overturning a finding; it is breaking a tie that the
data never broke. That is worse than it sounds and better than it sounds at the
same time, and saying which is the discussion.

**2. "Moderate" is a description of the winning side, not of anyone's opinions.**
Step 8 lands hard: the two justices nearest the middle of the recovered line win
`r pc(max(jj$pct_in_majority))`% and `r pc(jj$pct_in_majority[jj$justice == "JGRoberts"])`%
of cases, and the ones at the ends lose a quarter to a third. It is a mechanical
consequence of position, not a personality trait. **Then take it back with Step
9**: restricted to the `r nrow(cc)` one-vote decisions those same two fall to
`r pc(max(jj$close_pct))`%, and only `r BLOC_WINS` of the `r nrow(cc)` were won by
the conservative bloc alone. The `r pc(max(jj$pct_in_majority))`% figure is
mostly a fact about unanimous and lopsided cases, which are most cases.

**3. Every number in the `r nrow(bt)`-term series is a human coding decision, and
every row of the underlying file is a person's act.** The direction codes are a
codebook applied case by case since `r min(bt$term)`. That is not a reason to
discard the series; it is a reason to say what it is. And note the level: this is
**individual-level data** — Thomas's vote is Thomas's vote, not a rate computed
over a group — which is why the agreement matrix is even possible. Aggregate data
cannot support this analysis at all.

**The line for the board:** the score is a measurement of nine people and a
measurement of one decision we made, and the arithmetic cannot tell them apart.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** What one record is | Ten justices, cases, direction, majority | 5 | The two short tenures are planted here. Point at them; do not explain yet. |
| **2** Who agrees with whom | `r NPAIR` pairs, `r pc(max(ag$pct, na.rm = TRUE))`% to `r pc(min(ag$pct, na.rm = TRUE))`% | 6 | "No ideology has been supplied to it" is the sentence that has to land. |
| **3** The score falls out | `cmdscale` on 1 minus agreement | 7 | Students think a model was fitted. Nothing was fitted; distances were arranged. |
| **4** Check against something unseen | Correlation `r nm(R)` | 4 | The sign is arbitrary (see below). Do not let anyone read meaning into it. |
| **5** Not a spectrum | Span `r nm(SPAN)` vs chasm `r nm(CHASM)` | 5 | Ask for the prediction *before* showing the chasm. |
| **6** One cell is not a number | `r nice(HOLE$a)` and `r nice(HOLE$b)`, `r HOLE$cases` shared cases | 8 | **Undefined, not zero, not missing.** No future data fixes it. |
| **7** Refuse to guess | `r MOVED` positions move; blocs do not | **12** | **The pivot.** Take the prediction in the block quote seriously — most rooms guess "nothing changes" or "everything changes", and both are wrong. |
| **8** The number that surprises | `r pc(max(jj$pct_in_majority))`% in the majority | 7 | Someone will call it power. It is geometry. |
| **9** Close cases | `r nrow(cc)` one-vote decisions | **10** | Denominators differ by justice — see the caution below. This is Step 6's problem wearing a different hat. |
| **10** Seventy-eight terms | `r min(bt$term)`–`r max(bt$term)`, coded direction | 8 | The series stops at `r max(bt$term)`. Say so out loud. |
| **11** What you can now say | And cannot | 6 | Push for "the blocs are in the data" in a student's own words. |

Total 78 minutes; Steps 7 and 9 absorb overruns.

**If you are short on time, protect Steps 6, 7 and 9.** The undefined cell, what
happens when you refuse to fill it, and the collapse of the majority rates in
close cases are the argument. Steps 1–5 can be walked in ten minutes as a
narrated demonstration; Step 10 can be a single sentence over the plot.

***

# What the brief now shows before any of this

Figure 1 of the brief, under "Who agrees with whom", draws all `r NPAIR`
agreement rates as a 10 x 10 heatmap, with the one undefined pair left blank and
outlined in red, and a button labelled *re-sorted by the recovered order* (two
panels in the PDF) that re-sorts the same matrix. **Use it as the cold open.**
Students see the block structure and the hole with no method in between, which
makes Steps 3 to 6 an explanation of something they have already noticed rather
than a sequence of results they have to take on trust. Figure 3, under "Does the
middle still win when it is close?", adds the majority rates plotted against
position: an arch for all cases, and a flatter, partly inverted shape for the
one-vote cases.

# Two cautions about the numbers themselves

## The axes have no meaning and no orientation

Classical multidimensional scaling recovers **distances**, not coordinates. A
scaling solution can be reflected, rotated, or both without changing its fit by
any amount at all: if you multiply every position by -1 the pairwise distances are
identical, so the fit is identical. **"Left" and "right" on that plot are
conventions the analyst imposes after the fact, not findings.**

Three practical consequences:

- The correlation with conservative voting is `r nm(R)` **and it is negative only
  because the solution came out with conservatives on the negative side.** A
  reflected solution would give `r nm(-R)`. The magnitude is the finding; the
  sign is not.
- **The brief now states this under "The score falls out"**, immediately after the recovered
  ordering, and both figures that use the scale label the axis as arbitrary in
  direction. If a student quotes the brief at you, it agrees with this page.
- The lab's plot is labelled `position (left = conservative here)`. The word
  "here" is doing real work. **If the plot ever renders mirrored on a student's
  machine, nothing is wrong** — eigen-decomposition sign conventions can differ.
  Say this before somebody reports a bug.
- In two dimensions (Option A) rotation is free as well, so "the second dimension
  is X" is a claim about a coordinate system you chose, not about the Court.

This is not a defect peculiar to our version. It applies to Martin-Quinn, to
NOMINATE, and to every ideal-point estimate students will ever read about. The
orientation is always supplied by the analyst.

## The close-case denominators are not all `r nrow(cc)`

`close_cases.csv` records the term, so each justice's real denominator is the
one-vote decisions in the terms they actually sat for.

```{r denoms}
d <- jj[order(-jj$close_pct), c("justice", "close_maj", "close_n", "close_pct")]
d$justice <- nice(d$justice); d$close_pct <- pc(d$close_pct)
print(d, row.names = FALSE)
```

`r nice(HOLE$b)` sat for `r jj$close_n[jj$justice == HOLE$b]` of the one-vote
decisions and `r nice(HOLE$a)` for `r jj$close_n[jj$justice == HOLE$a]`. Dividing
either by `r nrow(cc)` — which is what a student will do — gives
`r pc(100 * jj$close_maj[jj$justice == HOLE$b] / nrow(cc))`% and
`r pc(100 * jj$close_maj[jj$justice == HOLE$a] / nrow(cc))`%, both badly wrong.
Barrett, by contrast, was confirmed before the first one-vote decision in the
window was handed down, so her `r jj$close_maj[jj$justice == "ACBarrett"]` of
`r nrow(cc)` is comparable even though she has fewer coded cases than the rest.

**This is the single most likely arithmetic error in Option C, and catching it is
a 4.**

***

# Option A — Two dimensions

## Verified output

```{r a}
round(sol$points[order(sol$points[, 1]), ], 4)
round(sol$eig[1:5], 4)                       # scree: how much each axis carries
round(c(eig2_over_eig1 = sol$eig[2] / sol$eig[1],
        eig2_over_eig3 = sol$eig[2] / sol$eig[3],
        gof_1d = cmdscale(as.dist(1 - M), k = 1, eig = TRUE)$GOF[1],
        gof_2d = sol$GOF[1]), 3)
round(c(median_cor_dim1_under_resampling = D1COR,
        median_cor_dim2_under_resampling = D2COR), 3)
```

Dimension 1 carries an eigenvalue of `r nm(sol$eig[1])`; dimension 2 carries
`r nm(sol$eig[2])` and dimension 3 carries `r nm(sol$eig[3])`. **The second
dimension is barely larger than the third**, which is the cleanest available
signature of noise: a real second dimension stands apart from the ones below it,
and this one does not. Resampling the votes, the recovered second dimension
correlates a median `r nm(D2COR)` with the one in the data — dimension 1 comes
back at `r nm(D1COR)` by comparison.

**Who it separates matters too.** On dimension 2 the extremes are
`r nice(names(which.max(sol$points[, 2])))` and
`r nice(names(which.min(sol$points[, 2])))` — which are exactly the two justices
joined by the invented cell from Step 6. **The second dimension in the ten-justice
solution is substantially manufactured by the fill.** Run it on the eight
overlapping justices and a different second dimension appears, with Gorsuch at
one end and Roberts and Kavanaugh at the other:

```{r a2}
round(sol2$points[order(sol2$points[, 1]), ], 4)
round(sol2$eig[1:4], 4)
```

**3/4:** Runs `k = 2`, plots both, reports that the second dimension does not
change the left–right ordering and separates one or two justices from the rest.

**4/4:** Names who moves and asks whether it is real. Any of these is the move:
notices the second dimension's extremes are Breyer and Jackson and connects that
to the fill; or compares eigenvalue 2 to eigenvalue 3 and concludes they are
indistinguishable; or proposes a substantive reading (Gorsuch on criminal
procedure and Indian law is the standard one) **and says what evidence would
distinguish it from noise**.

**4/4, best:** Says that the question as posed cannot be settled from the
agreement matrix at all, because the axes of a scaling solution are only defined
up to rotation — dimension 2 is whatever is left over after dimension 1, in a
coordinate system nobody chose on substantive grounds. To interpret it you would
need to leave the matrix and go look at which cases those justices split on.
**Knowing the data cannot answer it is the right answer**, and it is the same
move rewarded in Option B of the `precinct-geography` lab.

***

# Option B — Refuse to guess

## Verified output

```{r b}
data.frame(rank = seq_along(ord2),
           filled  = nice(keep),
           refusing = nice(names(ord2)),
           changed = ifelse(keep == names(ord2), "", "swapped"))
round(c(cor_with_fill = R, cor_refusing = R2), 3)
round(ord[keep], 4)   # positions with the guess
round(ord2, 4)        # positions refusing to guess
```

Two adjacent pairs trade places, `r MOVED` positions in all. Nothing else moves,
the blocs are identical, and the correlation with conservative voting **improves**
from `r nm(R)` to `r nm(R2)`.

**3/4:** Re-runs it, reports that two pairs swapped and that the blocs held.

**4/4:** Draws the general conclusion — coarse structure is robust, fine ordering
is not — and says what that implies for reading any published ranking of the
justices. Bonus for noticing that the correlation went *up* without the guess,
and for resisting the conclusion that the eight-justice version is therefore
"better": it answers a different question, about a Court that never existed.

**4/4, best:** Notices *which* pairs swapped and checks the gaps. Thomas and
Alito are `r nm(GAPS[1], 4)` apart and Roberts and Kavanaugh `r nm(GAPS[5], 4)`,
on a scale spanning `r nm(diff(range(ord)))` — so the fill did not overturn a
result, it broke a tie. A student who says "those two were never distinguished by
this data in the first place, with or without the guess" has understood the
method better than the brief states it. The resampling above puts a number on it:
the full order survives `r pc(SAME)`% of the time and Alito precedes Thomas in
`r pc(FTA)`% of resamples.

*(That resampling treats each pair's agreement count as an independent binomial,
which it is not — the same votes enter many pairs — so read it as illustrative
rather than as a confidence interval. A proper resample would be at the case
level and the aggregated files do not support one. Worth saying if a student
asks; do not require it.)*

***

# Option C — The close cases

## Verified output

```{r c}
o <- jj[order(-jj$close_pct),
        c("justice", "pct_in_majority", "close_maj", "close_n", "close_pct")]
o$justice <- nice(o$justice); o$close_pct <- pc(o$close_pct)
print(o, row.names = FALSE)

# how many one-vote majorities came entirely from the six-justice bloc
table(conservatives_in_majority = nc)
```

**3/4:** Counts correctly, reports that Kavanaugh and Roberts are still top but
much lower — `r pc(max(jj$close_pct))`% against
`r pc(max(jj$pct_in_majority))`% — and that the ordering below them scrambles.

**4/4:** Handles the denominator (see the caution above): either restricts to the
justices who sat the whole window, or uses each justice's own terms. A student
who divides Breyer's `r jj$close_maj[jj$justice == "SGBreyer"]` by `r nrow(cc)`
and reports `r pc(100 * jj$close_maj[jj$justice == "SGBreyer"] / nrow(cc))`% has
made exactly the Step 6 error in a new place, and should be told so rather than
marked down hard — it is the most instructive mistake in the lab.

**4/4, best:** Notices that only `r BLOC_WINS` of the `r nrow(cc)` one-vote
decisions were won by a majority drawn entirely from the conservative bloc, and
draws the conclusion: **the dimension the scaling recovered explains the Court's
agreement in general but not its closest cases.** Two thirds of the one-vote
decisions cut across it. That is a genuine limit on what an ideal-point score
buys you, and it is not stated anywhere in the brief — reward it warmly.

***

# Option D — Your own question

No fully verified output; grade the reasoning and the fit between question and
data.

**3/4:** Asks a question the four files can actually answer and answers it
without error — most-agreeing pair by term, the shape of the by-term series
around a particular appointment, which justices appear together most often in
one-vote majorities.

**4/4:** States what would have to be true for the answer to mean what the
student wants it to mean, and checks at least one of those things. The commonest
good version: a claim about change over time that runs into the fact that this
is a snapshot and the by-term series measures a different quantity (case
direction) from the scaling (justice agreement).

**4/4, best:** Asks a question the data *cannot* answer and explains precisely
why — then says what data would. Examples that have worked: "did Roberts move
after 2018?" (needs a dynamic model, which is exactly why Martin-Quinn exists);
"do these blocs disagree about outcomes or about reasons?" (the file has no
opinions in it); "is the docket selected by the same nine people whose agreement
we are measuring?" (yes, and it makes the denominator endogenous). **Rejecting a
question for a stated reason is worth as much as answering one.**

Common failures: comparing "% conservative" across distant terms without any
argument about the codebook; reading a trend off the raw grey line in Step 10
rather than the smoothed one; treating the `r nrow(bt)`-term series and the
four-term scaling as if they measured the same thing.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code — it is given.** A wrong number from correctly running
  the given code is a reading error, not a coding error, and should be graded as
  interpretation.
- **The move that earns a 4** is treating the fill as a substantive decision with
  consequences rather than a technical detail, and being explicit about which
  parts of the answer survive changing it.
- **A 2** is "the score is subjective so it doesn't mean anything." True in the
  trivial sense, useless, and the exact conclusion Step 7 is designed to prevent —
  the whole point is that *some* of it survives and you can say which.
- **Also a 2:** reporting the recovered ranking as though the positions were
  measured quantities, e.g. "Thomas is the most conservative justice" as a finding
  rather than as a coin flip with Alito.
- **Also a 2:** dividing every close-case count by `r nrow(cc)` and not noticing.
- **Reward anyone who asks how precise the positions are.** The lab gives no
  error bars and there is no honest way to attach them without going back to the
  case-level data. Noticing their absence is the most sophisticated available
  response, and it is what the resampling above exists to answer for you rather
  than for them.

## Anticipated questions

- *"Isn't this circular? You got the answer everyone already knew."* — The
  strongest thing you can say is that the ordering was not put in: the scaling saw
  `r NPAIR - nrow(HOLE)` numbers and no labels, and the dimension it found is the
  one the culture already names. The strongest objection is that agreement and
  ideology may be the same phenomenon measured twice, in which case `r nm(R)` is a
  tautology with extra steps. **Both are defensible and neither is refutable from
  this table.** Say that; do not resolve it.
- *"Why not just use Martin-Quinn?"* — Because a published score is the answer,
  and this session is about how the answer gets made. Also, practically, the
  Martin-Quinn site is behind a bot check and cannot be fetched by a script; the
  build note in `data/build-data.R` says so.
- *"Why is conservative negative?"* — It is not. See the orientation caution
  above. The solution could be reflected at no cost.
- *"Couldn't you just leave the missing cell out?"* — Not in classical scaling,
  which needs a complete distance matrix. Option B's answer is to drop the two
  justices instead, which is the honest version of leaving it out and costs you
  two justices.
- *"Why does the data stop at `r max(bt$term)`?"* — Because the SCDB release used
  here is the 2024 release, covering terms through `r max(bt$term)`. **Nothing
  after the `r max(bt$term)` term is in these files**, so no claim in this lab is
  about the current Court. Anyone reaching for a 2025 or 2026 decision is off the
  edge of the data.
- *Is* Allen v. Milligan, *which is sitting in `close_cases.csv`, still good
  law?* — Largely not, and it is a gift of an example. *Allen v. Milligan*, 599
  U.S. 1 (2023), appears in the file as a
  `r cc$majVotes[cc$caseName == "ALLEN v. MILLIGAN"]`–`r cc$minVotes[cc$caseName == "ALLEN v. MILLIGAN"]`
  decision whose majority was **Roberts and Kavanaugh joined with Kagan,
  Sotomayor and Jackson** — one of the `r nrow(cc) - BLOC_WINS` one-vote cases
  that cut across the recovered line. Then came *Louisiana v. Callais*, 608 U.S.
  ___ (2026), decided April 29, 2026, in which the Court substantially reworked
  Voting Rights Act § 2 and held that § 2 did not require Louisiana to draw an
  additional majority-minority district. Alito wrote for a six-justice majority
  that included **both Roberts and Kavanaugh**, with Kagan dissenting for
  herself, Sotomayor and Jackson. **Two of the justices who made the** Milligan
  **majority were on the other side three years later** — a concrete
  demonstration of the lab's own limitation that a snapshot cannot show anyone
  changing. If you use this, get the neighbours right too: *Shelby County v.
  Holder*, 570 U.S. 529 (2013), struck the § 4(b) coverage formula, not § 5
  itself, and *Bartlett v. Strickland*, 556 U.S. 1 (2009), is a plurality, not a
  majority.
- *"Do the justices know these numbers exist?"* — Yes. Several have commented on
  them, usually unfavourably. That is a good thirty seconds, not more.

## Connections

- **`dw-nominate`** — the other half of Session 12 and the direct pair. There
  students consume a published score; here they build one. **Run them in that
  order**, and ask at the end of this one what DW-NOMINATE's fill is.
- **`levels-of-aggregation`** — this lab sits on the **individual** side of the
  course's individual-versus-aggregate axis. A justice's vote is a person's act,
  recorded as such, and that is precisely why a pairwise agreement matrix can
  exist at all; no aggregate table of "the Court voted conservative 60% of the
  time" could support it. Worth one sentence at the top of the session.
- **`cast-vote-records`** — the other lab built on individual acts rather than
  rates, and the closest analogue: an anonymous ballot there, a named vote here.
  The contrast between the two is a good exam question.
- **`rpv`** — where the *absence* of individual-level data forces inference from
  aggregates, and everything gets harder. The cleanest possible foil for this
  session.
- **`precinct-geography`** — the same shape of lesson in a different domain: two
  defensible methods, one invisible choice, a moved answer.
- **`surveys-source`** and **`gss-confidence`** — the other places in the course
  where a headline number turns out to rest on a coding decision made by somebody
  the reader never meets.
