---
title: "Part I — The Census Bureau"
type: chapter
subtitle: "One agency, three instruments, and the geography everything else is published on"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-1-census-bureau-code.R")
```

```{r setup, include=FALSE}
```

Almost every number you will meet in this book has a population underneath it. A turnout rate is votes divided by people. A claim that a district is majority Black is a count of people. A stop rate, a registration rate, a share of anything: all of them need somebody to have said how many people are there.

That bottom number, the thing you divide by, is called the **denominator**, and this book will come back to it constantly. Choosing the wrong one is the most common way a true number becomes a false sentence.

In the United States, that somebody is the **Census Bureau**, and this part is about what it does and how to read what it publishes.

The Bureau goes first for a plain reason. **`r nn(GOES)` of the `r nn(OUT)` chapters in the rest of this book go to it for a number.** Not because those chapters are about the census. They are about traffic stops, turnout, district maps. They come here because their question needs a denominator, and this is where denominators come from.

## One agency, three instruments, endlessly confused

The Bureau is not one survey. Its own catalogue lists 134 surveys and censuses. Three of them get almost all of this part, and it is worth saying now why those three and not the others.

It is because they are the three that get mistaken for each other. Each one answers some version of *how many people live here*. Each publishes a figure that looks like a population. Each carries the same agency's name and appears on the same website in the same typeface. Nobody has ever confused the Commodity Flow Survey with a headcount, because nothing about them looks alike. These three look exactly alike, and they are not the same:

- the **decennial census**, which tries to count everyone, once every ten years;
- the **American Community Survey**, which asks a sample of about 3.5 million
  addresses a year and publishes estimates with a margin of error — a stated
  range for how far off the estimate could be while still doing its job;
- the **population estimates**, which are a model that carries the last count
  forward using births, deaths and moves.

Ask the three of them for the population of one county and you will get three different numbers. None of them is wrong. They are answers to three different questions, and which one you quoted decides what your sentence means.

## A survey that lives in the counting part

One thing here will look misfiled, and the explanation is worth having early.

The American Community Survey is a **survey** — it asks a sample of people questions, which is exactly what Part II is about. It is in this part anyway, because **answering it is required by law.** The same statute that compels the decennial count compels the ACS, and the envelope says so. Every survey in Part II is voluntary: you may hang up, and most people do.

That difference is not a technicality. It runs through everything the two kinds of data can be made to say, and holding the two side by side is one of the more useful things this book asks you to do.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

Read the length of the fourth row. `r nn(bt$chapters[bt$beat == "4 use"])` of the `r NCH` chapters here are chapters that *use* the data rather than introduce it. A part is mostly its applications; the opening chapters exist so that the applications can be read.

## You will meet this data again

```{r reusetab}
```

Two numbers, and the gap between them is the honest shape of this part's importance.

**`r nn(GOES)` chapters elsewhere in this book go to the Census Bureau themselves.** They come for a turnout denominator, a benchmark for traffic stops, or a check on a guess about somebody's race. Or for a population to compare a survey against.

But only **`r nn(DEPS)`** of them read *this part's* files. It would be easy, and wrong, to claim that everything else rests on Part I. It does not.

The true version is narrower and more useful. **This is the source the rest of the book keeps going back to, independently, on its own terms.** That is exactly why it is worth a part of its own. You meet it here, rather than first meeting it as somebody else's denominator.

## How to read this part

Start with [what a census is](../census-source/census-source-brief.html), then take the three instruments in order. They are ordered by how often they are mistaken for each other, not by importance.

If you only have time for two chapters, read the source chapter and the one on the instrument you are actually about to use. If you already know which file you need, the [access chapter](../census-access/census-access-brief.html) is the fastest route. Read its finding first, because the Bureau has more than a dozen published answers in print for one county's population, and naming which one you mean is the whole skill.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens; reorganise the part and rebuild, and the tables above follow.
- U.S. Constitution, Article I, Section 2; the Census Act, 13 U.S.C., including
  § 221, which makes answering compulsory; Public Law 94-171 (1975); and
  Section 203 of the Voting Rights Act. Between them these four obligations
  produce almost everything in this part.
