---
title: "The Modifiable Areal Unit Problem: Fulton County, 2020"
type: brief
subtitle: "One county's census blocks, and what moving the boxes does to the share"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("areal-units-code.R")
```

```{r setup, include=FALSE}
```

Here is a claim you have met in map form. One map of a city shades each neighborhood by its Black share and shows a place sharply divided. Another shades the same city as a single unit and shows one even tone. Both can be drawn from the same census, taken on the same day, with no error in either. Two honest maps of the same people can tell opposite stories. How?

The how has a name worth learning: the **modifiable areal unit problem**, named by the geographer Stan Openshaw in 1984. It is the fact that a statistic computed over areas depends on the boxes the people were sorted into, and not only on the people. This brief asks it in its sharpest form. Fulton County, Georgia is `r pc(FV("ful_black_pct"), 2)`% Black; cut the county into six districts of equal population, and how many are majority-Black? That sounds like a question about Fulton's people. Whether it is one is the test.

## The test

The test holds the people still and moves only the boxes. The data is the [2020 Census Redistricting Data (P.L. 94-171) Summary File](https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/Georgia/ga2020.pl.zip) for Georgia, downloaded whole from the Census Bureau's own server, with block boundaries from the Bureau's TIGER/Line 2020 release. A 1975 statute obliges the Bureau to deliver this file to every state legislature so the legislature can draw districts. So these pages use the redistricting numbers themselves, in the form Georgia's map-drawers received them.

Three properties make this file, of all files, the right instrument. It is a complete count rather than a sample, so there are no margins of error below, no allowance for how far off an estimate could be, because nothing is estimated. Its atom, the census block, is the smallest unit the Bureau publishes, so everything larger can be built out of it. And census units nest exactly. A block's identifier contains the identifiers of its block group, tract, county and state, so climbing that ladder is arithmetic on a string, with no map operation to introduce error. How that machinery works, and where it stops, is the subject of the [census geography chapter](../census-geography/census-geography-brief.html).

Two experiments follow. The first climbs the ladder: the same blocks, added into block groups, then tracts, then the county, watching what each rung does to the spread of the Black share. That is the **scale effect**, what happens to a statistic as the units it is computed over grow. The second holds the number and size of units fixed and changes only their shapes. That is the **zoning effect**, and it is the one that decides elections, because redrawing shapes while holding population constant is the literal definition of redistricting.

## The data

```{r p1-layout}
```

Everything below rests on one table built here from the Bureau's raw files: [`ga_block_race.csv`](data/derived/ga_block_race.csv), one row per Georgia census block. Its columns, by name: `GEOID20`, the block's fifteen-character identifier; `pop`, everyone counted there in 2020; `black_alone`; `black_any`; `vap`, the voting-age population; and `black_any_vap`, the voting-age count of any part Black.

Building it was the ordinary work of opening a government file never designed to be convenient. One decision in that work changes what you can trust. Every identifier is held as text, because a fifteen-digit identifier read as a number collapses into scientific notation, and a tract identifier silently drops its leading zero. The build stops outright if any identifier is not exactly fifteen characters wide.

Keeping only the rows whose summary level means "block" leaves `r n(FV("ga_blocks"))` of them. Nothing was imputed, meaning no value nobody recorded was filled in by guessing. No row was dropped for being empty either: blocks with nobody living on them stay in, all `r n(FN("ful_blocks") - FN("ful_blocks_pop"))` of Fulton's, because how many there are is part of the record.

Two decisions in the table are choices, and they are named. `black_any` counts a person as Black if their answer is any of the `r n_black` of the census's `r n_cells` race combinations that include Black. Counting `black_alone` instead gives Georgia `r pc(FV("ga_black_alone_pct"), 1)`% rather than `r pc(FV("ga_black_pct"), 2)`%; voting-rights litigation generally uses any-part, so this brief does too. And every statistic below is population-weighted: a unit counts as many times as it holds people, so a block of three cannot pull as hard as a tract of six thousand. What the file holds, at the two scales this brief uses:

```{r step1}
```

The rest of the folder was made here rather than downloaded, and made data can answer only the question its maker thought to ask. The ladder tables, [`ladder_fulton.csv`](data/derived/ladder_fulton.csv) and [`ladder_state.csv`](data/derived/ladder_state.csv), add the same blocks up four ways, from block to county. The sweep, [`zoning_summary.csv`](data/derived/zoning_summary.csv), divides Fulton's `r n(FV("ful_blocks"))` blocks into `r FV("K")` units of `r n(FV("quota"))` people each, one sixth of the county apiece, `r n(FV("n_plans"))` different ways. The ways come from two rules, each rotated through `r FV("n_angles")` one-degree steps:

```{r rules}
```

Neither rule looks at race, or at anything except where people live and which way the ruler points. Every unit of every plan holds a sixth of Fulton to within `r pc(FV("max_dev_pct"), 2)`% of the quota, tighter than any real districting plan in the United States is required to be.

> **Before you read on, commit to a number.** The same `r n(FV("ful_blocks"))`
> blocks, the same `r n(FV("ful_pop"))` people, `r FV("K")` units of equal
> population, and two rules neither of which reads race. **How many of the
> `r FV("K")` units will be majority-Black?** Write your number down before
> you look at the first figure.

## What it says about democracy

```{r plans-d3, results='asis', eval=knitr::is_html_output()}
```

```{r plans-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

**Figure 1. The same dots, twice, with different lines drawn over them.** Each dot is `r n(FV("dot_people"))` people; red dots are residents recorded as any part Black. A paired map fits the question because the dots are identical in both panels, the same census, and only the boundaries differ. On the left, the `r FV("lo_family")` rule at `r FV("lo_angle")` degrees gives `r FV("K")` units running `r pc(FV("lo_min"))`% to `r pc(FV("lo_max"))`% Black, and `r FV("lo_maj")` majority-Black. On the right, the `r FV("hi_family")` rule at `r FV("hi_angle")` degrees gives units running `r pc(FV("hi_min"))`% to `r pc(FV("hi_max"))`%, and `r FV("hi_maj")` of `r FV("K")` majority-Black.

Most readers commit to two or three, because the county is `r pc(FV("ful_black_pct"), 2)`% Black and a sixth of it ought to look like the whole. **Across the `r n(FV("n_plans"))` plans, the answer runs from `r FV("maj_min")` to `r FV("maj_max")`.**

The left panel is what political scientists call **cracking**, arrived at with no intent. The slabs run across the county's grain, so each one picks up part of the Black south and west and part of the white north. Every unit lands near the county figure and not one is a majority. The right panel cuts with the grain instead. A ruler and a protractor did this; no demographic fact entered either rule.

```{r zone-d3, results='asis', eval=knitr::is_html_output()}
```

```{r zone-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

**Figure 2. Every plan in the sweep, sorted by how many of its units are majority-Black.** Each row is one plan and each dot one of its `r FV("K")` units, placed by the unit's Black share; dots right of the red 50% line are majority-Black units. The two heavily marked rows are the extremes drawn in Figure 1.

`r n(FV("plans_zero_maj"))` of the `r n(FV("n_plans"))` plans produce no majority-Black unit at all; `r n(sum(sm$n_majority == FN("maj_max")))` produce `r FV("maj_max")`; `r n(FV("plans_two_maj"))` produce exactly two. Same census, same day, same county, same `r FV("K")` equal populations. The mistake worth naming is not any particular guess. It is assuming the question has *an* answer.

The honest objection is that these are not districts, so test it. Counted strictly, `r FV("plans_strict_contig")` of the `r n(FV("n_plans"))` plans have all `r FV("K")` units in one connected piece. That is Georgia's block layer at work: the riverbed, the interstate and the airport apron get blocks of their own, so any cut anywhere strands a few near-empty blocks.

The criterion that means something is whether a unit's people form one connected mass. Call a plan effectively contiguous when every unit has at least `r FV("contig_thr")`% of its population in one piece: `r n(FV("plans_contig"))` plans pass, and among them the answer still runs from `r FV("contig_maj_min")` to `r FV("contig_maj_max")` majority-Black units. The zero-majority plans do not survive the test; the ambiguity does. What is left is a one-seat swing decided by which of two respectable rules you happened to use, and on a `r FV("K")`-seat county commission that is a sixth of the government.

The zoning effect is half of the problem. The other half appears when the same blocks are simply added up the census ladder, with no redrawing at all.

```{r scale-d3, results='asis', eval=knitr::is_html_output()}
```

```{r scale-static, eval=!knitr::is_html_output(), fig.height=3.5}
```

**Figure 3. The same county, the same census, four ideas of what counts as a place.** Fulton County, 2020, shaded by the share of each unit that is any part Black, on one color ramp across all four panels; the four-panel form fits because only the unit changes between them. The left panel is `r n(LF("Census block", "units"))` inhabited shapes and the right is one shape, holding exactly the same `r n(FV("ful_pop"))` people. "Spread" beneath each panel is the standard deviation of the units' Black shares, a measure of how far units sit from the average, weighted by population; it falls from `r pc(LF("Census block", "sd_share"))` points to `r pc(LF("County", "sd_share"))`.

The population-weighted mean is the control in this experiment: `r pc(LF("Census block", "mean_share"), 2)`% at every rung, because the same people are being averaged, only bundled differently. Everything else moves. The spread falls from `r pc(LF("Census block", "sd_share"))` points across blocks to `r pc(LF("Census tract", "sd_share"))` across tracts and to `r pc(LF("County", "sd_share"))` across the county, where one unit has nothing left to vary. Statewide, the same climb does more than shrink variation:

```{r ladder-state}
```

Two columns of that table are worth more than they look. The most Black unit in Georgia falls from `r pc(LG("Census block", "max_share"))`% of a block to `r pc(LG("County", "max_share"))`% of a county. No county in Georgia is as Black as thousands of its blocks are, because extremes are a property of the unit. And the correlation between a unit's Black share and its share of residents under 18 runs `r pc(LG("Census block", "corr_child"), 3)` across blocks and `r pc(LG("County", "corr_child"), 3)` across counties. It changes sign on the way up. A correlation is not a fact about people; it is a fact about people *and* the boxes they were sorted into.

What this says about democracy is direct. Any claim about "places", a segregated city, a poor neighborhood, a Black district, inherits a choice of box that somebody made, usually for another purpose, usually long ago. There is no true value underneath for a better method to recover. "How many majority-Black districts can Fulton County support?" is asked and answered under Section 2 of the Voting Rights Act, as though it were a fact about people. It is a fact about partitions of people, and the same people have just been shown to admit partitions giving `r FV("maj_min")` and partitions giving `r FV("maj_max")`.

And gerrymandering is this fact weaponized. Neither rule here can see race, yet one plan produces `r FV("lo_maj")` majority-Black units and another `r FV("hi_maj")`. So "the algorithm doesn't look at race" defends nothing: a race-blind procedure has a racial output, settled by settings that look purely technical, an angle, a splitting order. The zoning effect is not evidence of gerrymandering. It is the reason gerrymandering is possible, and equally the reason that a plan you dislike is not, by itself, proof of intent.

## What this brief cannot tell you

The range is a floor, not a measurement. Two rule families are not the space of all partitions, and neither imposes compactness, respects a city line or keeps a community together. Real plans are more constrained in ways that narrow the range, and drawn by people with motives in ways that widen it. This is one county, in one state, in one decade: the mechanism is general and these numbers are not.

The ground truth is itself a set of boxes. A census block was drawn by the Bureau, along visible features, for the convenience of the counting, so the finest line drawn here is a line somebody at the Bureau already drew. The published block counts also carry privacy noise added on purpose, and that noise matters more at block level than at county level. Race here is what a person said about themselves on one day. And population is not the electorate: Fulton is `r pc(FV("ful_black_pct"), 2)`% Black by population and `r pc(FV("ful_bvap_pct"), 2)`% by voting-age population.

Nothing here says what Fulton should have. No plan is optimized for anything, and a real plan must respect city lines, communities of interest and the Voting Rights Act. Nor does anything here say how individuals voted or lived. Reading individual behavior off area averages is its own famous mistake, the **ecological fallacy**, guessing what individuals did from group totals; this brief stays at the level of areas.

## What you should have learned

- A statistic about areas is a joint fact about the people and the boxes: the same Fulton residents show `r pc(LF("Census block", "sd_share"))` points of spread across blocks and `r pc(LF("County", "sd_share"))` across the county, around a mean pinned at `r pc(LF("Census block", "mean_share"), 2)`%.
- Adding up destroys extremes by a measurable amount: Georgia's most Black unit falls from `r pc(LG("Census block", "max_share"))`% of a block to `r pc(LG("County", "max_share"))`% of a county, and one correlation changes sign between those rungs.
- Holding scale fixed and rotating shapes moved the answer from `r FV("maj_min")` to `r FV("maj_max")` majority-Black units across `r n(FV("n_plans"))` race-blind plans, and `r FV("contig_maj_min")` to `r FV("contig_maj_max")` among the effectively contiguous ones.
- A race-blind rule can produce a racial result, which is why the zoning effect is the mechanism that makes gerrymandering possible rather than evidence of it.

## Extensions

Every question below can be answered by sorting or grouping in a spreadsheet, using the tables handed over under Sources.

- `ladder_fulton.csv` and `ladder_state.csv` hold the same four levels, for one county and for the whole state. Compare the `sd_share` column down each ladder: where does Fulton's spread barely move, and where does it collapse?
- In `zoning_summary.csv`, sort by `n_majority` and read off `family` and `angle`. Which rotations make the most majority-Black units, which make the fewest, and do the two rule families behave differently?
- `zoning_summary.csv` also carries `max_dev_pct` and `units_contig`. Do the plans producing the most majority-Black units stay equal in population, and how many of their units are literally one piece?
- `ga_block_race.csv` has both `black_alone` and `black_any`. Count the blocks where the two disagree, and work out how much of Fulton's Black share the choice between them moves.

Two questions point past the spreadsheet. The sweep counted majorities of total population, and `zoning_summary.csv` carries `n_majority_bvap`, the same count on voting-age population. Comparing the two columns measures how much the choice of denominator, the bottom number you divide by, moves the answer. And the P.L. 94-171 file covers every state, so you can fetch another county's blocks from the address below and ask whether its answer is as movable as Fulton's.

## Sources

**Data**

- U.S. Census Bureau, *2020 Census Redistricting Data (P.L. 94-171) Summary
  File*, legacy format, Georgia. Tables P1, P3.
  <https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/Georgia/ga2020.pl.zip>
- U.S. Census Bureau, TIGER/Line 2020: Georgia tabulation blocks
  (`tl_2020_13_tabblock20`), block groups
  (<https://www2.census.gov/geo/tiger/TIGER2020/BG/tl_2020_13_bg.zip>) and tracts
  (<https://www2.census.gov/geo/tiger/TIGER2020/TRACT/tl_2020_13_tract.zip>).

**Concepts**

- Openshaw, S. (1984). *The Modifiable Areal Unit Problem.* CATMOG 38.
- Robinson, W. S. (1950). "Ecological Correlations and the Behavior of
  Individuals." *American Sociological Review* 15(3).

**Law**

- Voting Rights Act § 2, 52 U.S.C. § 10301 — the provision under which "how
  many majority-minority districts are possible" is asked and answered.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`county_outline.csv`](data/derived/county_outline.csv), [`dens_fulton.csv`](data/derived/dens_fulton.csv), [`dots_fulton.csv`](data/derived/dots_fulton.csv), [`ga_block_race.csv`](data/derived/ga_block_race.csv), [`ladder_fulton.csv`](data/derived/ladder_fulton.csv), [`ladder_state.csv`](data/derived/ladder_state.csv), [`map_units.csv`](data/derived/map_units.csv), [`map_vals.csv`](data/derived/map_vals.csv), [`zoning_plan_geo.csv`](data/derived/zoning_plan_geo.csv), [`zoning_plan_units.csv`](data/derived/zoning_plan_units.csv), [`zoning_summary.csv`](data/derived/zoning_summary.csv), [`zoning_sweep.csv`](data/derived/zoning_sweep.csv)

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
