---
title: "Data Session 4 Key, part 2: Where the Votes Have to Come From"
subtitle: "84-355 Democracy's Data | Session 4 — Returns over time (self-study) | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
nc <- read.csv("data/nc_gov_county.csv", stringsAsFactors = FALSE)
tr <- tapply(nc$rep, nc$year, sum); td <- tapply(nc$dem, nc$year, sum)
nc$rs <- nc$rep / tr[as.character(nc$year)]
nc$ds <- nc$dem / td[as.character(nc$year)]
ar <- sort(tapply(nc$rs, nc$county, mean), decreasing = TRUE)
ad <- sort(tapply(nc$ds, nc$county, mean), decreasing = TRUE)
```

```{r vals, include=FALSE}
# Every figure quoted in this key is computed here and inserted inline, using
# the same definitions as the brief, so the two cannot drift apart.
SW <- aggregate(cbind(dem, rep, oth, total) ~ year, nc, sum)
SW$rep_pct <- 100 * SW$rep / SW$total
SW$oth_pct <- 100 * SW$oth / SW$total

CUM   <- function(v, i) 100 * sum(v[i])
HALF  <- function(v) unname(which(cumsum(v) >= .5)[1])
YR    <- function(y, k) SW[[k]][SW$year == y]

BEST   <- max(SW$rep)
BESTYR <- SW$year[which.max(SW$rep)]
REP24  <- YR(2024, "rep")
DEM24  <- YR(2024, "dem")
OTHX   <- YR(2024, "oth") / YR(2020, "oth")

# TARGET is the textbook's 2028 goal (Ch. 5), not a figure from data/.
TARGET <- 3200000

TG   <- function(T, cty) unname((ar * T)[cty])
A24  <- function(cty) nc$rep[nc$county == cty & nc$year == 2024]
SHORT <- function(cty) TG(TARGET, cty) - A24(cty)

nc$pct <- nc$rep / nc$total
bp <- sort(tapply(nc$pct, nc$county, mean), decreasing = TRUE)
OVER <- length(intersect(names(ar)[1:5], names(bp)[1:5]))

TT <- function(x) {                    # WAKE -> Wake, NEW HANOVER -> New Hanover
  s <- tolower(x)
  gsub("(^|[ -])(\\w)", "\\1\\U\\2", s, perl = TRUE)
}
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(round(x), big.mark = ",", trim = TRUE)
```

All Part 5 code is given to students. Grade the interpretation.

**Note the timing.** This lab runs in **Session 4**, the self-study week (Sep
15/17), before Chapter 5 explains vote targeting. That is deliberate — Ch. 3 is
about campaigns becoming professional, and this is what professional means — but
it does mean students meet the method before the book's account of it. Expect
some to say so; that is a fair observation and worth a point, not a deduction.

Every number below was run against the data file.

***

# The lab in one sentence

Vote targeting turns one statewide number into a hundred county numbers by
assuming the past distribution of your vote still holds — and 2024 is sitting
in the data to show you what happens when it does not.

***

# The findings

```{r headline}
st <- aggregate(cbind(dem, rep, oth) ~ year, nc, sum); st
round(c(top10 = 100*sum(ar[1:10]), top25 = 100*sum(ar[1:25]),
        bottom50 = 100*sum(ar[51:100])), 1)
```

- **Republican votes are concentrated but not extremely so**: the top 10 counties
  supply `r pc(CUM(ar, 1:10))`%, the top 25 supply `r pc(CUM(ar, 1:25))`%, and it
  takes **`r HALF(ar)` counties** to reach half the party's vote.
- **The book's 3.20M target is unreachable on this evidence.** The best
  Republican performance in the data is **`r nn(BEST)` (`r BESTYR`)**. The target
  is **`r nn(TARGET - BEST)` above that**, and `r nn(TARGET - REP24)` above 2024.
- **2024 is a candidate-specific collapse, not a trend.** The Republican share
  falls to `r pc(SW$rep_pct[SW$year == 2024])`% and the "other" column runs at
  `r pc(OTHX)`× its 2020 level.

***

# The idea to teach

The arithmetic is trivial and completely correct. What makes it a lab is that
**every step is an assumption wearing a number's clothing**. Averaging four
elections assumes they are exchangeable. Using share-of-party-vote assumes the
county's *relative* contribution is the stable thing. Picking 3.20M assumes a
performance nobody has achieved.

Students who say "the method is fine, the inputs are guesses" have it.

***

# Option A — A different statewide goal

## Verified output

```{r a}
# the three targets the lab offers: the book's 2028 goal, the winning
# Democratic total in 2024, and the best Republican total in the data
for (T in c(TARGET, DEM24, BEST)) {
  t <- ar * T
  cat(sprintf("target %9s | Wake %6.0f  Meck %6.0f  Durham %5.0f | top-5 order same: %s\n",
      format(T, big.mark=","), t["WAKE"], t["MECKLENBURG"], t["DURHAM"],
      identical(names(sort(t, decreasing=TRUE))[1:5], names(ar)[1:5])))
}
```

| target | Wake | Mecklenburg | Durham |
|---|---|---|---|
| `r nn(TARGET)` (book's 2028 goal) | `r nn(TG(TARGET, "WAKE"))` | `r nn(TG(TARGET, "MECKLENBURG"))` | `r nn(TG(TARGET, "DURHAM"))` |
| `r nn(DEM24)` (2024 Democratic total) | `r nn(TG(DEM24, "WAKE"))` | `r nn(TG(DEM24, "MECKLENBURG"))` | `r nn(TG(DEM24, "DURHAM"))` |
| `r nn(BEST)` (best Republican total, `r BESTYR`) | `r nn(TG(BEST, "WAKE"))` | `r nn(TG(BEST, "MECKLENBURG"))` | `r nn(TG(BEST, "DURHAM"))` |

**The ranking never changes.** Only the magnitudes scale.

## What a good answer says

**3/4:** Reports the new targets and notices every county moves proportionally.

**4/4:** Draws the conclusion the prompt is fishing for — **changing the target
does not change where the campaign goes, only how hard it has to work
everywhere.** The target sets ambition; the *shares* set geography, and the
shares are computed from history. A campaign that wants a different map of
priorities has to change the shares, which means arguing the past is not a good
guide — and that is a political judgment, not an arithmetic one.

**4/4, best:** compares the target to 2024 actuals and sees how large the gap
is per county — Wake short by `r nn(SHORT("WAKE"))`, Mecklenburg by
`r nn(SHORT("MECKLENBURG"))`. Asking a county organiser to find
`r nn(round(SHORT("WAKE"), -3))` additional votes is not a plan, it is a wish. A
student who notices that the method produces *numbers* rather than *strategies*
has understood something real about how campaigns talk to themselves.

***

# Option B — Target the other party

## Verified output

```{r b}
rbind(REP = round(100*c(top10=sum(ar[1:10]), top25=sum(ar[1:25]), bottom50=sum(ar[51:100])),1),
      DEM = round(100*c(top10=sum(ad[1:10]), top25=sum(ad[1:25]), bottom50=sum(ad[51:100])),1))
c(counties_to_half_REP = which(cumsum(ar) >= .5)[1],
  counties_to_half_DEM = which(cumsum(ad) >= .5)[1])
```

| | top 10 | top 25 | bottom 50 | counties to reach half |
|---|---|---|---|---|
| Republican | `r pc(CUM(ar, 1:10))`% | `r pc(CUM(ar, 1:25))`% | `r pc(CUM(ar, 51:100))`% | **`r HALF(ar)`** |
| Democratic | `r pc(CUM(ad, 1:10))`% | `r pc(CUM(ad, 1:25))`% | `r pc(CUM(ad, 51:100))`% | **`r HALF(ad)`** |

Democratic top six: `r paste(TT(names(ad)[1:6]), collapse = ", ")`.
Republican top six: `r paste(TT(names(ar)[1:4]), collapse = ", ")`,
**`r paste(TT(names(ar)[5:6]), collapse = ", ")`**.

## What a good answer says

**3/4:** Notices the Democratic vote is more concentrated.

**4/4:** Says what that implies operationally. **`r HALF(ad)` counties versus
`r HALF(ar)` is a different campaign**, not a different emphasis: fewer field offices, denser
media buys, a smaller travel radius, and far more exposure to one bad night in
one metro. The Republican coalition is more spread out, which is more expensive
to organise and more robust to a local collapse.

**4/4, best:** notices the top three counties are *the same for both parties*,
because big counties are big. The lists diverge at positions 4–6, where
Democrats add Durham and Buncombe and Republicans add Union and Gaston. Then
asks the right question: is a county worth visiting because it has many of your
voters, or because it has many *persuadable* voters? This method cannot tell
you, and that is Option C.

***

# Option C — Share of votes cast, not share of party votes

## Verified output

```{r c}
nc$pct <- nc$rep / nc$total
bp <- sort(tapply(nc$pct, nc$county, mean), decreasing = TRUE)
data.frame(by_contribution = names(ar)[1:5], by_percentage = names(bp)[1:5])
length(intersect(names(ar)[1:5], names(bp)[1:5]))
```

| by contribution | by Republican percentage |
|---|---|
| `r TT(names(ar)[1])` | `r TT(names(bp)[1])` |
| `r TT(names(ar)[2])` | `r TT(names(bp)[2])` |
| `r TT(names(ar)[3])` | `r TT(names(bp)[3])` |
| `r TT(names(ar)[4])` | `r TT(names(bp)[4])` |
| `r TT(names(ar)[5])` | `r TT(names(bp)[5])` |

**`r c("Zero","One","Two","Three","Four","Five")[OVER + 1]` overlap.** Not "a
different order" — a disjoint list.

## What a good answer says

**3/4:** Notices the lists are completely different and says the two measures
answer different questions.

**4/4:** Assigns each ranking to a decision. **Contribution tells you where the
votes are; percentage tells you where you are liked.** If the job is to bank
raw votes, go to `r TT(names(ar)[1])`. If the job is to find a friendly
audience, go to `r TT(names(bp)[1])` — which is `r pc(100 * bp[1])`% Republican
and casts `r pc(100 * mean(nc$total[nc$county == names(bp)[1]]) / mean(nc$total[nc$county == names(ar)[1]]), 1)`%
as many votes as `r TT(names(ar)[1])`. A rural county at
`r pc(100 * bp[1])`% Republican may deliver fewer additional votes than a
`r pc(100 * mean(nc$pct[nc$county == names(ar)[1]]))`% share of a metro.

**4/4, best:** connects this to the campaign-visits material. Campaigns visit big
competitive states, not the states that like them most, and this is the same
logic one level down. Also notices that neither ranking mentions **margin** or
**persuadability** — both are about where your existing voters live, so both
implicitly treat the election as a turnout problem rather than a persuasion
problem. That is a real strategic assumption, and Chapter 5 makes it mostly
without argument.

***

# Option D — Their own question

Good versions: which counties moved most between 2020 and 2024 (the Robinson
collapse is not uniform); whether third-party voting in 2024 is concentrated
where Robinson underperformed; whether county rank order is stable across all
four elections.

Warn against comparing raw vote counts across counties of wildly different size
without normalising.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating the target as an *assumption* rather
  than a goal. Students who audit the inputs beat students who extend the
  arithmetic.
- **Reward anybody who checks the book.** The lab closes on two errors in
  Table 5.3; a student who verifies rather than takes my word for it is doing
  exactly what the course is for.

## Anticipated questions

- *"Why average across four elections instead of using the most recent?"* —
  Because one election is noisy. But averaging assumes exchangeability, and
  2024 was not exchangeable. There is no clean answer; there is a defensible
  choice you have to state. Good discussion, five minutes.
- *"Is the book wrong?"* — Table 5.3's 2016 statewide figure is wrong
  (transposed digits), and 2020 is off by one. The five printed counties are
  right. Nothing in the chapter's argument depends on either. Say all three of
  those things — the point is the habit of checking, not scoring on a textbook.
- *"Why the governor's race rather than president?"* — Because the chapter uses
  it, and because a state race isolates state-level strategy from presidential
  coattails. Also because 2024 gives us a natural experiment in candidate
  collapse.
- *"Where is turnout?"* — Table 5.1 is registration and turnout; we work in
  votes because targeting works in votes. A student who asks for turnout as a
  denominator is thinking well — that is the ratio that tells you whether a
  county is maxed out.

## Connections

- **`campaign-visits`** (currently retired for lack of a slot) — same
  strategic question at the state level. Run them together if you reinstate it.
- **Session 2 (apportionment)** — both are about converting a total into parts by
  a rule. There the rule is law; here it is a campaign's own choice.
- **Session 10 (live returns)** — this is a baseline-versus-actual exercise done
  in advance; that one does it after the fact with real returns.
- **This week's Tuesday debate** (deliberation, across two centuries) — Ch. 3 asks whether
  the transformation of campaigns raised the volume of information while lowering
  its quality. Targeting is the mechanism: it decides who hears anything at all.
- **Session 5 (money, message, and attention)** — when Ch. 5 arrives, students will already have done its worked
  example. Refer back to it explicitly; the chapter reads differently once the
  arithmetic is in your hands.
