---
title: "Lab 4b: Where the Votes Have to Come From"
subtitle: "84-355 Democracy's Data | Vote targeting in North Carolina, 2012–2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means. Nothing asks you to write code.

***

# The question

A campaign has one statewide number it needs to hit, and a hundred counties to
get it from. **Where does it look for the votes?**

This is called *vote targeting*, and it is the least glamorous and most
consequential thing a campaign does. Before any advertisement is written,
somebody decides which counties are worth a field office. That decision is
arithmetic, and today you do the arithmetic.

**You are meeting this two weeks before the chapter that explains it.** Sides et
al. take up vote targeting in Chapter 5, which we read in Week 6. That is
deliberate: Chapter 3, which you are reading this week, is the story of campaigns
becoming candidate-centred and professional, and this is what "professional"
turned out to mean in practice. Do the arithmetic now; you will meet the argument
about it later, and it will land harder for having done it.

We use the same four elections Chapter 5 uses — the North Carolina governor's
race in 2012, 2016, 2020 and 2024 — but we build it from the state's own
returns rather than retyping the book's table. That turns out to matter.

# Part 1 — The data

```{r load}
nc <- read.csv("data/nc_gov_county.csv", stringsAsFactors = FALSE)
str(nc)
head(nc, 3)
```

One row per county per election. `dem`, `rep` and `oth` are votes for the
Democratic, Republican and all other candidates.

```{r statewide}
st <- aggregate(cbind(dem, rep, oth, total) ~ year, nc, sum)
st$rep_pct <- round(100 * st$rep / st$total, 1)
st
```

**Read this before going on.** Republicans won in 2012 and lost the next three,
and 2024 is not a normal loss — the Republican share falls off a cliff and the
"other" column quadruples. Mark Robinson's campaign collapsed late after a
news investigation, and voters who would not vote for him went somewhere else.
Hold that thought; it comes back in Part 4.

# Part 2 — Each county's share of the party's votes

The targeting method is simple. For each election, ask what fraction of the
party's *statewide* votes came from each county. Then average those fractions
across elections. That average is the county's expected contribution.

```{r shares}
tot_rep <- tapply(nc$rep, nc$year, sum)
nc$share <- nc$rep / tot_rep[as.character(nc$year)]

avg <- tapply(nc$share, nc$county, mean)
avg <- sort(avg, decreasing = TRUE)

head_tbl <- data.frame(county = names(avg)[1:10],
                       avg_share_pct = round(100 * avg[1:10], 2))
head_tbl
```

Ten counties out of a hundred. Notice how concentrated this is.

```{r concentration}
cat("Top 10 counties supply", round(100 * sum(avg[1:10]), 1), "% of the Republican vote\n")
cat("Top 25 counties supply", round(100 * sum(avg[1:25]), 1), "%\n")
cat("The bottom 50 counties supply", round(100 * sum(avg[51:100]), 1), "%\n")
```

# Part 3 — Turning a statewide goal into county goals

Now pick a statewide target and push it down to the counties. The chapter uses
**3.20 million** votes as a 2028 goal.

```{r targets}
STATEWIDE_TARGET <- 3200000          # <- this is the number you will change

target <- avg * STATEWIDE_TARGET
out <- data.frame(county = names(target)[1:10],
                  avg_share_pct = round(100 * avg[1:10], 2),
                  target_2028   = round(target[1:10]))
out
```

That is a campaign plan: Wake County must produce about 267,000 Republican
votes, Mecklenburg about 228,000, and so on down the list.

**Now compare the target to what has actually happened.**

```{r reality-check}
data.frame(year = st$year,
           rep_votes = st$rep,
           short_of_target = STATEWIDE_TARGET - st$rep)
```

# Part 4 — One election can move the whole plan

The average treats all four elections as equally informative. 2024 was not a
normal election. Watch what happens when it is left out.

```{r drop-2024}
sub  <- nc[nc$year != 2024, ]
tot2 <- tapply(sub$rep, sub$year, sum)
sub$share <- sub$rep / tot2[as.character(sub$year)]
avg2 <- tapply(sub$share, sub$county, mean)

cmp <- data.frame(county = names(avg),
                  with_2024 = round(100 * avg, 2),
                  without   = round(100 * avg2[names(avg)], 2))
cmp$change <- round(cmp$without - cmp$with_2024, 2)
rbind(head(cmp[order(-cmp$change), ], 5),
      head(cmp[order(cmp$change), ], 5))
```

The counties that gain when 2024 is dropped are the big metros. The counties
that lose are smaller and more rural. **That is the shape of the Robinson
collapse**, and if you leave 2024 in the average you are planning the next
campaign around a candidate-specific disaster.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different goal.** In the chunk in Part 3, change
`STATEWIDE_TARGET` to a number you can defend: the winning Democratic total in
2024 (3,069,496), or the best Republican total in the data (2,586,604, in 2020). Re-run
Parts 3. Which counties now look reachable, and which look hopeless? Does the
choice of target change *where* a campaign should go, or only how hard it has
to work everywhere?

**Option B — Target the other party.** In Part 2, change `nc$rep` to `nc$dem`
in both places, and `tot_rep` accordingly. Where does the Democratic vote come
from, and how does its geographic concentration compare? A party whose votes
are concentrated in ten counties and a party whose votes are spread over sixty
do not run the same campaign.

**Option C — Share of votes cast, not share of party votes.** Replace
`nc$rep / tot_rep[...]` with `nc$rep / nc$total` to get each county's
Republican *percentage* rather than its contribution. Now rank the counties.
The list looks completely different. Which ranking should a campaign manager
use, and for what decision?

**Option D — Your own question.** The data has 400 rows and every county in the
state. Ask something and answer it.

# On the discussion board

By **Friday, 11:59 p.m.** — that evening or the next day, while it is fresh.
No length requirement — a sentence or two is enough, so long as it shows you
engaged with what we did. Any *one* of the usual four: something that surprised
you, something we missed, something you do not understand, or an answer to a
classmate. One post is full credit, and it can be about anything we looked at
this week — this lab is the obvious candidate, but it is not the only one.

If you want a starting point, three that would make good posts:

- The arithmetic tells a campaign where the votes *were*. Does that tell it
  where the votes *will be*? Pick a side.
- Chapter 3 describes campaigns becoming candidate-centred and professional.
  Having now done a piece of that professional work: what did it gain, and what
  did it cost?
- Something in Part 4 you did not expect. The Robinson collapse does not move
  every county the same way, and the pattern is worth a sentence.

***

### A note on the textbook's table

Ch. 5 prints five counties and a statewide row. Our five counties match the
book exactly, which is reassuring. The statewide row does not: the book's own
Tables 5.2 and 5.3 give two different figures for the 2016 Republican total
(2,298,880 and 2,289,880), and the state's returns say the first is right.
The 2020 figure is off by one vote.

Nothing important turns on either. That is the point — you could not have known
that without checking, and checking took one line of code. `data/build-data.R`
does it automatically every time it runs.
