---
title: "The Gender Gap Is a Subtraction"
type: brief
subtitle: "Fifty-two years of party identification, the one number everybody quotes, and the two numbers underneath it"
author: "Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("gender-gap-code.R")
```

```{r setup, include=FALSE}
```

In 1972 the General Social Survey asked Americans which party they identified with. Men came out `r sg(F("men_first"))` net Democratic. Women came out `r sg(F("women_first"))`.

Fifty-two years later the same question, asked the same way, returns `r sg(F("men_last"))` for men and `r sg(F("women_last"))` for women. The difference between those two numbers is `r p1(F("gap_last"))` points, and that difference has a name. **The gender gap** is one of the most-quoted numbers in American politics.

It is also a subtraction. Somebody computes a number for women, computes a number for men, and prints the gap between them. What almost never gets printed is the pair.

## Before you read on

`r mnote("Write both numbers down. Most readers get the direction of one of them wrong.")`

In 1972 men and women were within a point and a half of each other, both about `r p1(F("women_first"))` points net Democratic. Today the gap between them is about `r p1(F("gap_last"))` points.

**Which side moved, and by how much?** Give a number for each.

## Where this comes from, and why it is the GSS

Everything below is the General Social Survey cumulative file, release 3a: `r n(FN("src_rows"))` respondents interviewed between `r F("first_year")` and `r F("last_year")`, weighted with `wtssps`.

The GSS is here partly because it is right for the question and partly because it is the one that opens. A fifty-two-year series on an unchanged question is exactly what a claim about change requires, and the [confidence chapter](../gss-confidence/gss-confidence-brief.html) explains why that consistency exists at all.

The other reason is less flattering to the field. **Two better-known sources would answer this question and neither can be fetched by a script.** The American National Election Studies sits behind a registration page. The Cooperative Election Study sits on Harvard Dataverse, which answers an automated request with a success code and an empty body, a refusal this book's [survey access chapter](../survey-access/survey-access-brief.html) measures directly. The GSS hands the file to anybody who asks.

That is a fact about access rather than about quality. It decides which survey a student can actually work with, which is a different thing from which survey is best, and the difference is worth noticing every time it happens.

## The one number

Here is the gap, drawn the way it is usually drawn.

```{r gap-only-static, fig.height=3.1}
```

*Figure 1. Women's net Democratic identification minus men's, every GSS year from `r F("first_year")` to `r F("last_year")`.*

It starts near zero. It widens through the 1980s, peaks at `r p1(F("gap_max"))` points in `r F("gap_max_year")`, and sits at `r p1(F("gap_last"))` today.

Read that figure and one story suggests itself so strongly that most people do not notice they have supplied it. The gap opened, so **women must have moved left**. Nothing in Figure 1 says that. Nothing in Figure 1 can.

## The two numbers

```{r two-d3, results='asis', eval=knitr::is_html_output()}
```

```{r two-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

*Figure 2. The same years, with the subtraction undone. In the browser, the buttons switch between the two series and the gap alone.*

**Both lines fall.** Men went from `r sg(F("men_first"))` to `r sg(F("men_last"))`, a move of `r p1(abs(FN("men_move")))` points. Women went from `r sg(F("women_first"))` to `r sg(F("women_last"))`, a move of `r p1(abs(FN("women_move")))`.

Women did not become more Democratic. They became less Democratic, by more than twenty points. The gap opened because **`r F("faster_side")` moved `r F("move_ratio")` times as far in the same direction.**

```{r decade-table}
```

The decade averages say the same thing more slowly. Every column falls. The last one rises because the first two fall at different speeds.

## Why this matters beyond one number

A gap is a difference, and a difference discards the level. Two series moving together by twenty points and two series moving apart by twenty points can produce the identical gap, and no amount of care in reporting the gap will recover which happened.

That is not a fault in the gender gap. It is a property of every difference this book reports, and several of them run the same way. The [turnout chapter](../validated-turnout/validated-turnout-brief.html) reports a gap between what people say and what they did, and that gap shrank while both of its halves moved. The [cost of voting chapter](../../06-putting-data-together/cost-of-voting/cost-of-voting-brief.html) ranks states against each other, which is a difference by construction, and its levels are not comparable across years at all.

**Ask for the pair. It is one extra number and it changes the sentence.**

## A second subtraction, larger than the first

Split women by marital status and a bigger gap appears inside the smaller one.

```{r marital-table}
```

In `r LM$year`, unmarried women were `r sg(LM$unmarried_women)` and married women `r sg(LM$married_women)`. That is `r p1(F("marriage_gap_last"))` points between two groups of women, against `r p1(F("sex_gap_same_year"))` points between women and men.

**The split inside the category is wider than the split between categories.** Anybody reporting one number for women is averaging across a difference larger than the one they are reporting.

## What this chapter cannot tell you

**`sex` here is two boxes.** The GSS records sex, not gender identity, and for most of this series it was coded by the interviewer. The file gained a different item, `sexnow1`, which was asked in `r F("sexnow_years")` of `r n(FN("sexnow_answers"))` respondents. That is too few years to carry a fifty-two-year comparison, so it is counted here and not used. Nothing in this chapter speaks to people for whom the two items differ.

**Party identification is not a vote.** It is what somebody says they are, not what they did, and the two come apart. Whether the leaners belong with the party they lean toward is a decision this chapter made. The [party identification chapter](../party-id/party-id-brief.html) takes it apart, and finds the same choice moving a headline number by a factor of five.

**These are cross-sections, not the same people.** A series that falls can fall because individuals changed their minds, or because the people who held the old view died and were replaced. Nothing here separates those, and they are different countries.

**The composition of each group changed underneath the line.** "Women" in 1972 and "women" in 2024 differ in education, marital status, race and age. Some of the movement in Figure 2 is those compositions changing rather than anybody's politics.

**And a gap is not an explanation.** This chapter shows which side moved. It does not say why, and the literature on why is large and unsettled.

## What it does establish

Both sexes moved away from the Democratic Party across this series. Men moved `r p1(abs(FN("men_move")))` points and women `r p1(abs(FN("women_move")))`, in the same direction, on the same question, weighted the same way.

The gap between them grew from `r p1(F("gap_1970s"))` points in the 1970s to `r p1(F("gap_2020s"))` in the 2020s. **Every point of that growth is a difference in speed, not a difference in direction.**

And the most-quoted version of the number hides that completely, because one line cannot show two movements.

## Sources

- **General Social Survey**, cumulative file, release 3a (2024), NORC at the
  University of Chicago. `r n(FN("src_rows"))` respondents,
  `r F("first_year")`–`r F("last_year")`, weight `wtssps`.
  <https://gss.norc.org/us/en/gss/get-the-data.html>
- **Build.** `data/build-data.R` downloads the 47 MB zip, reads nine columns
  out of the 570 MB Stata file, and writes the small tables this chapter
  reads. The source is not committed; the derived tables are. Every number
  above is computed at the moment this document was rendered.
- **Related chapters.** [Party identification](../party-id/party-id-brief.html)
  on the leaner decision; [confidence in institutions](../gss-confidence/gss-confidence-brief.html)
  on the same file and the same weight;
  [survey access](../survey-access/survey-access-brief.html) on why this is the
  GSS and not the CES.

### Checks

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_decade.csv`](data/derived/by_decade.csv), [`partyid_by_sex.csv`](data/derived/partyid_by_sex.csv), [`partyid_by_sex_marital.csv`](data/derived/partyid_by_sex_marital.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `partyid_by_sex.csv` has `men` and `women` separately, and the `gap` between them. Plot all three. Which of the two lines is doing the moving?
- `partyid_by_sex_marital.csv` splits the same people again by marital status. Compare the gap between married and unmarried women against the gap between women and men. Which subtraction is bigger?
- `by_decade.csv` collapses the series into decades. Compare it against the year-by-year file. What does the smoothing hide?
- `partyid_by_sex.csv` carries `n_men` and `n_women` for every year. Find the thinnest years. Would you report a gap that size off a sample that size?

```{r checks-table}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
