---
title: "The Races Nobody Contested"
subtitle: "84-355 Democracy's Data | House elections, `r {yy <- read.csv('data/by_year.csv')$year; paste0(min(yy), '–', max(yy))}`"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)

# Every number in the prose below is computed from the two built files, so it
# cannot drift when the data is rebuilt. The visible `load` chunk re-reads them
# in front of the students.
d  <- read.csv("data/races.csv",  stringsAsFactors = FALSE)
by <- read.csv("data/by_year.csv", stringsAsFactors = FALSE)

n1 <- function(x) format(x, big.mark = ",")
p1 <- function(x) sprintf("%.1f", x)

# decade mean of a by_year column (each election weighted equally)
dc <- function(v, decade) p1(mean(by[[v]][(by$year %/% 10) * 10 == decade]))
# decade mean of a race-level flag (each race weighted equally)
rd <- function(x, yr, decade) p1(100 * mean(x[(yr %/% 10) * 10 == decade]))

jac  <- d$source == "Jacobson"          # the two sources spliced in races.csv
cont <- d[!d$uncontested, ]             # contested races only
unc     <- d[d$uncontested, ]           # uncontested races
unc_inc <- unc[!is.na(unc$incwin), ]    # ... where the file records incumbency

ns_dec  <- tapply(by$pct_uncontested_non_south, (by$year %/% 10) * 10, mean)
lnd_dec <- tapply(by$pct_landslide,             (by$year %/% 10) * 10, mean)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `house-competition-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"Only about forty House seats are really in play."*

You have seen district partisanship measured from presidential votes. This is
the other way to ask: **what actually happened in the House elections
themselves**, in every district, in every election, since `r min(by$year)`.

The dataset is Gary Jacobson's, assembled over a career and still the standard
source for congressional elections, carried past his last year with the Clerk of
the House's official returns. **`r nrow(by)` elections, `r round(mean(by$races))`
seats each.**

And before we can use it, we have to survive it.

# Steps 1-3 — What one record is, how many, and the three definitions

```{r load}
d  <- read.csv("data/races.csv",  stringsAsFactors = FALSE)
by <- read.csv("data/by_year.csv", stringsAsFactors = FALSE)

c(races = nrow(d), elections = nrow(by),
  first = min(by$year), last = max(by$year),
  seats_per_election = round(mean(by$races)))
```

Jacobson's file as distributed has **30,005 rows. 14,777 of them are empty** — no
year, no votes, no candidate, nothing. Padding.

Drop them and you have **`r n1(sum(jac))` real races**:
`r round(mean(by$races))` seats × `r length(unique(d$year[jac]))` elections,
`r min(d$year[jac])` through `r max(d$year[jac])`. Compute anything before that
drop and every number you produce is wrong by a factor of two, silently.

**This is the ordinary condition of an old research dataset.** It was built for
its author, who knew what was in it. Nothing warns you.

Jacobson stops at `r max(d$year[jac])`. The `r n1(nrow(d))` rows you just loaded
run to `r max(by$year)` because the last `r n1(sum(!jac))` races were parsed from
the Clerk of the House's official returns and spliced on. That splice, not the
end of the file, is where this data's real limit sits — hold onto it; we come
back to it in *What this lab cannot tell you*.

# Steps 4-5 — The first number, and who holds the uncontested seats

```{r missing}
c(races = nrow(d),
  no_vote_share_recorded = sum(d$uncontested),
  pct = round(100 * mean(d$uncontested), 1))
```

**`r n1(nrow(unc))` races have no Democratic vote share.** There is no
"uncontested" flag in this file. There is simply a hole where the number should
be.

The hole *is* the finding. `dv` is the Democratic share of the **two-party**
vote, and if only one party fielded a candidate there is no two-party share to
record.

Two pieces of evidence that this is what missingness means:

```{r evidence}
table(uncontested = d$uncontested, incumbent_won = d$incwin, useNA = "ifany")
round(100 * tapply(d$uncontested, d$south, mean), 1)
```

**Where the file records incumbency at all, the incumbent won
`r n1(sum(unc_inc$incwin == 1))` of those `r n1(nrow(unc_inc))` races —
`r p1(100 * mean(unc_inc$incwin == 1))`%.** And missingness runs at
**`r p1(100 * mean(d$uncontested[d$south == 1]))`% in the South against
`r p1(100 * mean(d$uncontested[d$south == 0]))`% elsewhere.**

That is what an unopposed incumbent looks like in a spreadsheet.

Read the `NA` column of that table, not just the two you were looking for.
`r n1(sum(is.na(unc$incwin)))` of the uncontested races carry no incumbency flag
at all — they are the post-`r max(d$year[jac])` rows, where the source changes
and the flag stops existing. **Divide `r n1(sum(unc_inc$incwin == 1))` by
`r n1(nrow(unc))` instead of by `r n1(nrow(unc_inc))` and you have quietly
counted every one of them as "no incumbent".** The honest denominator is the
races where the answer was recorded.

# Step 6 — Split the country in two

> **Before running this, take the flat national trend seriously.** Uncontested
> races ran at `r p1(by$pct_uncontested[which.min(by$year)])`% in
> `r min(by$year)` and `r p1(by$pct_uncontested[which.max(by$year)])`% in
> `r max(by$year)` — and the widest single-election jump in between is
> `r p1(max(abs(diff(by$pct_uncontested))))` points, larger than the whole
> `r max(by$year) - min(by$year)`-year drift. That looks like noise around a
> line that went nowhere. **Predict what you will see if you compute the same
> share separately for the South and for everywhere else.**

```{r south}
by[by$year %in% c(1946, 1958, 1972, 1990, 2010, 2024),
   c("year", "pct_uncontested_south", "pct_uncontested_non_south")]
```

```{r south-plot}
plot(by$year, by$pct_uncontested_south, type = "b", pch = 19, col = "#C41230",
     ylim = c(0, max(by$pct_uncontested_south, by$pct_uncontested_non_south) + 5),
     xlab = "", ylab = "% of races with no opponent")
lines(by$year, by$pct_uncontested_non_south, type = "b", pch = 1, col = "#2c7fb8")
legend("topright", c("South", "rest of the country"),
       col = c("#C41230", "#2c7fb8"), pch = c(19, 1), bty = "n", cex = 0.85)
```

**In the 1950s, `r dc("pct_uncontested_south", 1950)`% of Southern House races
had only one candidate** — `r p1(max(by$pct_uncontested_south))`% in
`r by$year[which.max(by$pct_uncontested_south)]`, the worst year. Not close
races — *no* races. The Democratic primary was the election, and November was a
formality.

By the 2010s it is **`r dc("pct_uncontested_south", 2010)`%**, and
**`r dc("pct_uncontested_south", 2020)`%** so far in the 2020s. Meanwhile the
rest of the country sat between `r p1(min(ns_dec))`% and `r p1(max(ns_dec))`% in
every decade and never moved much.

**The most dramatic democratisation in modern American electoral history is
visible here as a column filling in.**

# Steps 7-8 — Contested is not competitive, and a fourth measure

```{r comp}
by[by$year %in% c(1946, 1964, 1988, 2004, 2014, 2024),
   c("year", "pct_competitive", "pct_landslide")]
```

```{r comp-plot}
plot(by$year, by$pct_competitive, type = "b", pch = 19, col = "#2c7fb8",
     ylim = c(0, max(by$pct_competitive, by$pct_landslide) + 5),
     xlab = "", ylab = "% of all races")
lines(by$year, by$pct_landslide, type = "b", pch = 1, col = "#C41230")
legend("topleft", c("won by under 5 points", "won by 20+ points"),
       col = c("#2c7fb8", "#C41230"), pch = c(19, 1), bty = "n", cex = 0.85)
```

**Close races fell from `r dc("pct_competitive", 1940)`% of all seats in the
1940s to `r dc("pct_competitive", 2000)`% in the 2000s**, recovering to
`r dc("pct_competitive", 2010)`% in the 2010s and
`r dc("pct_competitive", 2020)`% in the 2020s — still well below where they
started. **Landslides went the other way** — `r dc("pct_landslide", 1940)`% in
the 1940s to a peak of `r p1(max(lnd_dec))`% in the
`r names(lnd_dec)[which.max(lnd_dec)]`s, and
`r dc("pct_landslide", 2020)`% now.

So the South stopped being uncontested, and the country as a whole got *less*
competitive. Those are not contradictory: one-party dominance stopped taking the
form of empty ballots and started taking the form of lopsided ones.

# Step 9 — All four at once

Everything in Steps 7-8 divides by **all** races, uncontested ones included. That
is a choice, and it changes the answer.

```{r denom}
cont <- d[!d$uncontested, ]
dec <- function(x, y) round(100 * tapply(x, 10 * (y %/% 10), mean), 1)

rbind(all_races        = dec(d$competitive, d$year),
      contested_only   = dec(cont$competitive, cont$year))
```

**Among all races, competitive seats fell from
`r rd(d$competitive, d$year, 1940)`% in the 1940s to
`r rd(d$competitive, d$year, 2000)`% in the 2000s. Among contested races only,
from `r rd(cont$competitive, cont$year, 1940)`% to
`r rd(cont$competitive, cont$year, 2000)`%.**

Both are true. They answer different questions — *what share of seats were in
play* versus *when somebody bothered to run, how close was it* — and an
uncontested race is arguably the least competitive race there is, which argues
for keeping it in.

**Nobody who quotes one of these numbers tells you which one it is.** You have
now met this exact problem in the midterm-penalty lab, in the turnout labs, and
in election administration. It is the same question every time: *per what?*

# Step 10 — What you can now say

**You can say** that American House elections did not become less competitive
in a straight line, and did not become more competitive either. **One region's
one-party system was dismantled while the rest of the country drifted toward
having one of its own**, and the two motions roughly cancel in the national
number.

**You can say non-competition is an elite decision, not a voter verdict**: of
the uncontested races whose incumbency the file records,
`r p1(100 * mean(unc_inc$incwin == 1))`% had one — which means a party looked at
a seat with a sitting member in it and declined to run anybody.

**You cannot say why.** This file has no candidate quality, no fundraising and
no district boundaries, so redistricting, geographic sorting and party
recruitment cannot be separated here.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Midterms.** `races.csv` has a `midterm` flag. Are midterm elections
more or less competitive than presidential-year ones? Before running it, write
down what you expect and why — then check. (`incwin` is empty after
`r max(d$year[jac])`, so keep this one to `competitive` and `uncontested`, which
are recorded for all `r n1(nrow(d))` races.)

**Option B — Split districts.** `split_district` is TRUE where the district's
presidential and House winners came from different parties. Plot its share over
time (`pct_split` in `by_year.csv`). What does the trend say about how
nationalised congressional elections have become, and when did it change?

> **Note the source change at `r max(d$year[jac])`.** Everything from
> `r min(d$year[!jac])` on is parsed from the Clerk of the House's official
> statistics rather than Jacobson's file. The `source` column in `races.csv`
> says which is which, and `dvp` is empty for every one of those rows — so the
> denominator shifts at the splice as well as with coverage.
>
> **Two traps in this one, both deliberate.** First, `pct_split` is `NA` for
> `r sum(is.na(by$pct_split))` elections — check `split_coverage` before you
> plot, and say in your write-up what you did with the gaps. Second, the file
> has both a `dvp` and a `dpres` column. **`dvp` is "Democratic vote,
> *previous* election", not "presidential"**; an earlier version of this lab
> used it by mistake and reported seat turnover as split districts for a term.
> Use `dpres`.

**Option C — The South, defined.** The `south` flag is one researcher's list of
states. Look at which states it includes (`unique(d$state[d$south == 1])`), and
say whether the Step 6 finding depends on that choice. What would you do with
border states?

**Option D — Your own question.** `r n1(nrow(d))` races, `r nrow(by)` elections,
`r max(by$year) - min(by$year)` years.

# What this lab cannot tell you

- **Who ran.** No candidate names, no parties of challengers, no quality
  measures. "Uncontested" here is inferred from a missing number, not read from
  a list of candidates.
- **Why nobody ran.** A seat can go uncontested because it is hopeless, because
  the party is disorganised, or because a deal was struck. This data cannot
  distinguish them.
- **Who was sitting in the seat, after `r max(d$year[jac])`.** The file does not
  end there — it runs to `r max(by$year)` — but it *changes source* there.
  Jacobson supplies `r n1(sum(jac))` races through `r max(d$year[jac])`; the
  Clerk of the House supplies the remaining `r n1(sum(!jac))` from
  `r min(d$year[!jac])` on, and the Clerk's tables report votes, not careers.
  **`incwin` and `dvp` are empty for all `r n1(sum(!jac))` of those rows.** Vote
  shares, margins, uncontested races and competitiveness cross the splice
  intact; incumbency and the previous election's result do not. A claim about
  incumbency computed over the whole file will silently answer a question about
  `r min(d$year[jac])`–`r max(d$year[jac])` and label it
  `r min(by$year)`–`r max(by$year)`. That is the boundary in this data, and it
  is invisible unless you look at `source`.
- **Whether `dv` missing *always* means uncontested.** It is an inference from
  the incumbent-won pattern and the regional concentration, not a documented
  flag. A handful of the `r n1(nrow(unc))` may be missing for other reasons.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- In the 1950s, `r dc("pct_uncontested_south", 1950)`% of Southern House races
  had one candidate. What does that do to your idea of what "an election" is?
- Competition fell even as uncontested races became rarer. How can both be true?
- The uncontested races are a hole in a column, not a label. What else might you
  have missed in a dataset because absence was the signal?

***

### Where the data came from

**Gary C. Jacobson's House elections dataset,
`r min(d$year[jac])`–`r max(d$year[jac])`** — the standard source for
congressional elections research, held locally in this repository, and
`r n1(sum(jac))` of the `r n1(nrow(d))` races here.

**The Clerk of the House's official election statistics,
`r min(d$year[!jac])`–`r max(d$year[!jac])`** — parsed from the published PDFs
in `data/`, and the remaining `r n1(sum(!jac))` races. The two sources overlap
for several elections and agree closely on vote shares, which is why the splice
is usable; they do not carry the same columns, which is why `incwin` and `dvp`
stop at `r max(d$year[jac])`.

**14,777 of Jacobson's 30,005 rows are empty padding and must be dropped**, and
**uncontested races appear as a missing vote share rather than a flag.** Both
are handled and documented in `data/build-data.R`. `split_district` comes from
the file's own `split` column, which was checked against a recomputation from
`dv` and `dpres` before being trusted — an earlier build had it backwards, and
`data/build-data.R` records how that was settled. Where `dpres` does not exist
the series is masked rather than reported, which is why `pct_split` is `NA` for
`r sum(is.na(by$pct_split))` elections.
