---
title: "One Election, Not Fifty"
type: brief
subtitle: "Nationalization measured in three records, and the count that hides two causes"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("nationalization-code.R")
```

```{r setup, include=FALSE}
```

American elections used to be decided in different places for different reasons. The textbook says they are not any more: today's partisan elections are less about local politics and more about national issues and debates. Elections have been **nationalized**.

That is a claim about a trend, so it needs a series, and a series needs a record that reaches back. This chapter measures the same claim three times, in three records that are not the same kind of thing, and then takes the measure apart.

## What would settle it

A voter who decides on national grounds votes the same way all the way down the ballot. A voter who decides on local grounds does not. Nobody watches voters fill in a ballot, so the measure has to be built from outcomes, and the standard one is simple.

**A split outcome is a place that voted one way for president and the other way for another office on the same ballot.** A state that backed the Republican candidate for president and elected a Democrat to the Senate is a split outcome. So is a congressional district that voted for the Democratic candidate for president and sent a Republican to the House.

If elections have been nationalized, split outcomes should be getting rarer. They are. The rest of this chapter is about what that sentence is worth.

## Where the numbers here come from

**Three records answer the same question, and only one of them is an election return.**

**A roster.** Who sat in the Senate for each state, Congress by Congress, from 1879. A state whose two senators belong to different parties has been answering the same question two different ways. The [rosters-source chapter](../rosters-source/rosters-source-brief.html) is about what a roster is and what it cannot say: it records a membership, never a vote.

**Returns for two offices on one page.** After every general election the Clerk of the House publishes *Statistics of the Presidential and Congressional Election*, and it prints the presidential vote and the Senate vote for a state on the same pages. The [clerk-source chapter](../clerk-source/clerk-source-brief.html) is about that document and what makes it hard to read.

**Returns for two offices, one of which nobody official publishes.** The same measure for the House needs the presidential vote recomputed inside each congressional district. No government produces that number. The [house-competition chapter](../house-competition/house-competition-brief.html) fetched it from The Downballot and built the file this chapter reads.

**Why both congressional numbers come off the Clerk's page.** The measure is a subtraction: how a state voted for the Senate, minus how it voted for president. A subtraction is exactly where two differently-built files leak their differences into the answer. The [county-returns chapter](../county-returns/county-returns-brief.html) sets two compilations of the same election side by side and finds them disagreeing about how many votes were cast in half the country. Both numbers here come off one page, compiled by one office from one set of state canvasses on one date, so that cannot happen.

## One row is a state in a Congress

```{r one-row}
```

The Voteview file holds `r n(fn("vv_rows"))` rows, one for each person serving in each Congress since 1789. Reading the Senate rows from 1879 gives `r n(fn("del_state_congresses"))` state-Congresses, and the question asked of each is whether its senators ran on the same party line.

**Party is read as the line the senator ran on, and nothing else.** An independent is not a Democrat. Two of the states below are represented by independents who sit with the Democrats in Washington, and this chapter does not quietly file them as Democrats. The [crossover chapter](../crossover/crossover-brief.html) shows what that decision is worth in a single election.

## A century and a half of the same question

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. States whose two senators belonged to different parties, `r f("del_first_year")` to `r f("del_last_year")`.** Each point is one Congress. The heavy line uses every state. The light line uses only the states that sent exactly two senators to that Congress, and it is there for a reason given in the next section.

The shape is a hill. Split delegations were ordinary for most of the twentieth century. They ran to a peak in `r f("del_peak_year")`, when **`r n(fn("del_peak_n"))` of `r n(50)` states, `r p1(DPEAK)`% of them, sent two senators from different parties**. Then they fell for forty-five years. In `r f("del_last_year")` there are **`r n(fn("del_last_n"))`**: `r f("del_last_states")`.

That is one measure, one record, and a clean answer. Now the reasons to be careful with it.

## A roster is not a photograph

A state has two Senate seats, so it should contribute two rows to each Congress. **`r n(fn("del_three_plus"))` state-Congresses contribute three or more.** Somebody died, resigned or was appointed, and the file records people who were never in the chamber at the same time. Read as a photograph of a moment, the roster shows a state with three senators, and a state with three senators is much more likely to look split.

The fix is not to argue about it. The light line in Figure 1 is the same measure computed only over states that sent exactly two senators, which is the only case where the photograph reading is safe.

**The two lines agree.** They correlate at `r f("del_r_readings")` and differ by a median of `r p1(fn("del_medgap"))` points. Their worst disagreement is `r p1(fn("del_maxgap"))` points, in `r f("del_maxgap_year")` — wartime, and the Congress with the most mid-term arrivals in the series. Since 1959 they never differ by more than `r p1(fn("del_gap59"))`.

So the mid-term arrivals are not driving the answer. That sentence cost one extra line on a figure, and without it the chapter would be resting on a measure that had never been asked whether it works.

## The same question, one election at a time

The roster describes a moment. Six years of elections produced that moment, so a fall in the roster is six years of separate news arriving together. The Clerk's volumes let you ask the question one election day at a time.

```{r senate-table}
```

Two things in that table are worth stopping on, and they point in opposite directions.

**The last column falls and keeps falling.** It is the typical distance between a state's Senate result and its presidential result, in points of vote share. That is how far the Senate candidate finished from their own party's presidential candidate: same state, same ballot, same day. It was `r p1(fn("sen_wobble_2004"))` points in 2004 and `r p1(fn("sen_wobble_2024"))` in 2024.

**The count of split outcomes does not.** It runs `r paste(sy$split, collapse = ", ")` across the six elections. There were **none at all in 2016** — the first presidential election since 1913 in which every state that elected a senator elected one from the party that carried it. Eight years later there were `r n(fn("sen_split_2024"))`.

Nationalization did not reverse between 2016 and 2024. Thirty-odd contests is simply too few for a count of crossings to be steady, and a count of crossings is a different thing from the distance underneath it. Hold on to that; it is the whole of the last half of this chapter.

## The three contests the document cannot hand you

```{r runoff-table}
```

A state that sends its Senate election to a second round prints **two columns of votes**, one for each round. The two columns overlap when the page is turned into text. What survives is the first digit of the first-round total, plus the whole second-round total stranded on a line of its own with no name attached.

Read literally, Raphael Warnock received **one vote** in Georgia in 2020, and a Libertarian carried the state.

**Nothing about that reads as an error.** The number parses. The line is well formed. The file that comes out is a file of statewide Senate results with two states quietly wrong, and no check on the shape of the file would find it. What found it was a result that could not be true.

The repair is a rule rather than a list of states. Look for a named candidate whose total is immediately followed by a nameless line holding a number more than a hundred times larger. That candidate's vote has been split across two columns, and the larger number is the one that decided the seat. All `r n(fn("sen_runoffs"))` of them keep their winner and are marked. A runoff is a different day and a different electorate, so its share cannot be set beside a presidential share from November.

> **Before you read on, write down a prediction.** In `r f("h_peak_year")`,
> `r p1(HPEAK)`% of congressional districts voted one way for president and
> the other way for the House. In `r f("h_last_year")` it was `r p1(HLAST)`%.
> **Which of these two things do you think did more of that work?** That House
> candidates stopped finishing far from their party's presidential candidate,
> or that districts themselves moved further from the point where the country
> divides evenly. Write down which, and roughly how much of the fall you
> would give it.

## The House, the long way

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=6.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. `r n(fn("h_years"))` presidential elections, `r n(fn("h_districts"))` district-elections.** Only presidential years are drawn, so every point compares a House election with the presidential election held on the same day. The top panel is the share of districts that split. The bottom panel holds the two quantities the top panel is made of.

The top panel is the finding everybody quotes. Split districts run to **`r p1(HPEAK)`% in `r f("h_peak_year")`** and fall to **`r p1(HLAST)`% in `r f("h_last_year")`**.

The bottom panel is why. A district splits when two things are true at once, and each has its own line.

**The distance from the ticket.** How far the House candidate finished from their own party's presidential candidate in the same district. It was `r p1(fn("h_wobble_peak"))` points in `r f("h_peak_year")` and is `r p1(fn("h_wobble_last"))` now — smaller by a factor of `r p1(fn("h_wobble_ratio"))`.

**The spread of the districts.** How far a typical district sits from the middle of that year's distribution — measured around the year's own median, so that a landslide does not read as districts moving apart. It was `r p1(fn("h_spread_peak"))` points in `r f("h_peak_year")` and is `r p1(fn("h_spread_last"))` now. Districts are further apart than they were, by nearly half again.

A district splits when the first number beats the second, in the direction that crosses the fifty-per-cent line. Both lines moved, and both moved the way that makes splitting harder. The count cannot say which did the work.

## Two ingredients, and the count that hides them

You can find out by swapping one ingredient and leaving the other alone.

Take `r f("h_peak_year")`'s districts. Keep their presidential vote exactly as it was — the same country, the same lines, the same landslide. Now give those districts `r f("h_last_year")`'s candidates: each district keeps its rank, so the district whose House candidate ran furthest ahead of the ticket still does, but the distances are today's. Count the splits again.

Nothing here is simulated and nothing is random. The districts are ranked by their own distance from the ticket and handed the distance at the same rank in the other year. Handing a year its own distances back reproduces its own count exactly, and that is checked before anything is written.

```{r fig3-static, fig.height=4.4}
```

```{r swap-table}
```

**Figure 3. The same districts, other years' candidates.** The two dark bars are the elections as they happened. The two light bars are each year's districts run with the other year's candidates.

**Changing the candidates alone takes `r f("h_peak_year")` from `r p1(HPEAK)`% down to `r p1(fn("swap_old_new"))`%.** That is almost the entire fall, produced without moving a single district.

**Changing the districts alone takes `r f("h_last_year")` from `r p1(HLAST)`% up to `r p1(fn("swap_new_old"))`%.** Which is also a large move, in the other direction, and it is why the honest answer is not "one cause".

Read them together. Either change on its own would have produced most of the collapse, so the two are not additive and no clean percentage can be handed to each. But they are not equal either. The candidates alone land within `r p1(fn("swap_old_new") - HLAST)` points of where the country actually is. The districts alone land `r p1(fn("swap_new_old") - HLAST)` points away from it.

**The count did not tell you any of this.** "Split districts fell from `r p1(HPEAK)`% to `r p1(HLAST)`%" is true, it is quoted constantly, and it is a threshold — a line drawn at fifty per cent, with a yes or a no on either side. Two very different quantities moved underneath it, and a threshold reports their combination as one number.

**The rule of thumb.** When a share of cases crosses a line, ask what would have to change for a case to land on the other side. There is usually more than one answer, and the count gives you none of them.

## Where the ballot goes and the data does not

Everything above is three offices: president, Senate, House. That is the whole of what the Clerk's volume contains: `r n(51)` presidential slates, `r n(fn("sen_n_2024"))` Senate contests and 435 House races in 2024. It is the complete national record of partisan federal elections for that year.

The textbook's claim does not stop at those three. It is a claim about state and local races too, and there the argument bites hardest. A voter who knows nothing about a county commissioner but knows the party label is voting nationally, whether they mean to or not.

**No federal office publishes those results, and no state is required to send them anywhere.** There is no *Statistics of the Gubernatorial, Legislative and Municipal Election*. Governors, state legislators, mayors, county commissioners, school boards, judges — every one of them is elected by somebody, counted by somebody and published by somebody, and none of it arrives at a national desk. What exists are compilations built by researchers one office at a time, and they are not what this chapter is made of.

So the claim is best measured exactly where it is least surprising, and least measurable exactly where it matters most. That is not a flaw in the claim. It is a fact about the record, and the reason the nationalization argument took decades to establish while looking obvious the whole time.

## What this chapter cannot tell you

**No voter appears anywhere in this chapter.** Every number here is a total for a place. Take a district that voted for one party's presidential candidate and elected the other party's representative. Many people there may have split their ballots. Or two different sets of people turned out for the two contests. Guessing what individuals did from group totals is called ecological inference, and it has [its own chapter](../../06-putting-data-together/rpv/rpv-brief.html).

**The roster measure and the returns measures are not the same measurement.** Figure 1 counts states by who was sitting; the Clerk table counts contests by who was elected. A state can appear split in the roster for six years on the strength of one election, and the two series are not expected to agree year for year.

**The Senate series is six elections long.** The Clerk's volumes exist much further back, but only the recent ones were fetched and converted here, and the older typesetting is not the same. Six points is a series in the sense that it has an order. It is not enough to establish a trend on its own, which is why the House series carries that weight.

**The swap is a re-ranking, not a world.** It asks what the count would be if the distances between House candidates and their tickets were distributed as they were in another year. It does not claim that any particular district would have gone the other way. It holds turnout, candidates, districts and the national result fixed, while changing one thing that in life was never separate from them.

**Party is the ballot line throughout.** Read it as the two sides in Washington instead and the last Congress in Figure 1 holds `r n(fn("del_last_caucus"))` split states rather than `r n(fn("del_last_ballot"))`.

```{r last-congress}
```

**In `r n(fn("sen_no_two_party"))` Senate contests there is no distance from the ticket at all**, because one of the two major parties did not put up a candidate. They keep their winner and drop out of the last column of the Senate table. They are not a random ten. A party that leaves a Senate seat uncontested has usually decided the state is out of reach.

## Sources

- Lewis, Jeffrey B., Keith Poole, Howard Rosenthal, Adam Boche, Aaron Rudkin
  and Luke Sonnet. *Voteview: Congressional Roll-Call Votes Database*,
  `HSall_members.csv`. One row per member per Congress since 1789. Fetched
  fresh here; the [rosters-source
  chapter](../rosters-source/rosters-source-brief.html) is about what a roster
  is and is not.
  <https://voteview.com/static/data/out/members/HSall_members.csv>
- U.S. House of Representatives, Office of the Clerk. *Statistics of the
  Presidential and Congressional Election*, the volumes for 2004, 2008, 2012,
  2016, 2020 and 2024. Read from the copies the [house-competition
  chapter](../house-competition/house-competition-brief.html) fetched and
  converted to text. Every presidential and Senate number here comes from
  those documents.
  <https://clerk.house.gov/Votes/ElectionStatistics>
- The House series, 1952 to 2024, is the house-competition chapter's file,
  built from Gary Jacobson's House elections data, the Clerk's own volumes and
  The Downballot's presidential results by congressional district. No
  government agency computes that last one.
  <https://www.the-downballot.com/p/the-downballots-calculations-of-presidential>
- The 2024 Senate contests were parsed a second time, independently, by the
  [crossover chapter](../crossover/crossover-brief.html). The two parses agree
  on every contest, which is one of the checks recorded before this chapter
  wrote a single table.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`delegation_states.csv`](data/derived/delegation_states.csv), [`delegations.csv`](data/derived/delegations.csv), [`house_years.csv`](data/derived/house_years.csv), [`senate_contests.csv`](data/derived/senate_contests.csv), [`senate_years.csv`](data/derived/senate_years.csv), [`swap.csv`](data/derived/swap.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `delegations.csv` gives `pct_split` for every Congress across a century and a half. Follow it. Was there ever a time when states routinely sent two senators of different parties?
- `house_years.csv` carries `wobble` and `spread` alongside `pct_split`. Which of those three is the best measure of nationalization, and why?
- `senate_contests.csv` has `gap` between the Senate and presidential result in each state. Sort by it. Who still outruns their party, and where?
- `delegations.csv` distinguishes `pct_split` from `pct_split_two`, which excludes independents. Compare them. How much does the treatment of a few senators change the story?

```{r ai-prompt, results='asis'}
```
