---
title: "The Nationalization of Congressional Elections"
type: brief
subtitle: "One election, not fifty"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("nationalization-code.R")
```

```{r setup, include=FALSE}
```

Political scientists have long observed that congressional elections have nationalized. Voters once weighed the person in front of them: the incumbent who fixed a bridge, the challenger with a familiar name. Now, the observation goes, they vote the national fight from the top of the ballot to the bottom, and a Senate or House race is mostly a referendum on the president.

That matters to anyone who votes. If congressional elections are national, your representative's own record barely enters your verdict on them, and accountability moves from the person to the party. The textbook states the change as a settled fact. This brief does what this book always does with a settled fact: turns it into a question. Is it true — and if it is, when did it happen, and what exactly moved?

## The test

A split outcome measures the claim. A split outcome is a place that voted one way for president and the other way for another office on the same ballot. A state that backed the Republican for president and elected a Democratic senator is a split outcome; so is a House district that did the reverse. A voter deciding on national grounds votes the same way all the way down, so if elections have nationalized, split outcomes should be getting rarer.

Three records carry that measure back in time. Voteview's roster of the Senate says who sat for each state in every Congress since 1879, so a state whose two senators belong to different parties has answered the same question two ways. The Clerk of the House's *Statistics of the Presidential and Congressional Election* prints each state's presidential and Senate vote on the same page, for 2004 through 2024. And the [house-competition chapter's](../house-competition/house-competition-brief.html) file carries the presidential vote inside each congressional district back to 1952, a number no government computes.

All three rest on certified returns, the complete and official record of who won each contest, and [the returns chapter](../returns-source/returns-source-brief.html) explains what such a record can bear. One care is taken with the Senate measure. It is a subtraction, how a state voted for the Senate minus how it voted for president, and a subtraction is exactly where two differently built files leak their differences into the answer. So both numbers come off the Clerk's one page, compiled by one office from one set of state canvasses on one date.

## The data

Three tables hold the series, and each row is a place at a moment.

`delegations.csv` has one row per Congress from `r f("del_first_year")` to `r f("del_last_year")`, built by reading the Senate rows out of Voteview's `r n(fn("vv_rows"))` member-Congress records. Its columns are `year`, `states`, `split_states` and the share `pct_split`. Party is read as the line the senator ran on and nothing else: an independent is not quietly filed as a Democrat, however they vote in Washington. One complication is cleaned into a companion column. A state that lost a senator mid-term can contribute three people to one Congress, and `r n(fn("del_three_plus"))` state-Congresses do. `pct_split_two` recomputes the share over states that sent exactly two senators, the only case where reading the roster as a photograph is safe.

`senate_years.csv` and `senate_contests.csv` come from the Clerk's six volumes: `r n(fn("sen_contests"))` Senate contests, 2004 through 2024, each with the state's Senate share, its presidential share and the `gap` between them. The cleaning made two decisions. The `r n(fn("sen_runoffs"))` contests decided in runoffs are repaired and marked, because a runoff is a different day and a different electorate, and its share cannot sit beside a November presidential share. And in `r n(fn("sen_no_two_party"))` contests one major party ran nobody, so there is no distance to measure; they keep their winner and drop out of the `gap` column.

`house_years.csv` is the House series: `r n(fn("h_districts"))` district-elections across the `r n(fn("h_years"))` presidential years from `r f("h_first_year")` to `r f("h_last_year")`. Each year carries `pct_split` and the two quantities it is made of. `wobble` is how far the typical House candidate finished from their own party's presidential candidate in the same district. `spread` is how far the typical district sits from the middle of that year's distribution, measured around the year's own median so a landslide does not read as districts moving apart. `swap.csv` holds one small experiment described below.

Here is the roster series at three points, a century and a half apart:

```{r one-row}
```

> **Before you read on, write down a prediction.** In `r f("h_peak_year")`,
> `r p1(HPEAK)`% of congressional districts voted one way for president and
> the other way for the House. In `r f("h_last_year")` it was `r p1(HLAST)`%.
> **Which of these two things do you think did more of that work?** That House
> candidates stopped finishing far from their party's presidential candidate,
> or that districts themselves moved further from the point where the country
> divides evenly. Write down which, and roughly how much of the fall you
> would give it.

## What it says about democracy

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. States whose two senators belonged to different parties, `r f("del_first_year")` to `r f("del_last_year")`.** Two lines on a time axis, because the claim is a claim about change and the answer is a shape. The heavy line uses every state; the light line uses only states that sent exactly two senators. The two agree, correlating at `r f("del_r_readings")` and differing by a median of `r p1(fn("del_medgap"))` points, so the mid-term replacements are not driving the answer.

The shape is a hill. Split delegations were ordinary for most of the twentieth century, and they ran to a peak in `r f("del_peak_year")`, when `r n(fn("del_peak_n"))` of 50 states, `r p1(DPEAK)`% of them, sent two senators from different parties. Then they fell for forty-five years. In `r f("del_last_year")` there are `r n(fn("del_last_n"))`: `r f("del_last_states")`.

A roster describes a moment that six years of elections produced. The Clerk's volumes ask the same question one election day at a time:

```{r senate-table}
```

Two columns of that table point in opposite directions. The last column, the typical distance between a state's Senate result and its presidential result on the same ballot, falls from `r p1(fn("sen_wobble_2004"))` points in 2004 to `r p1(fn("sen_wobble_2024"))` in 2024. The count of split outcomes does not fall with it: it runs `r paste(sy$split, collapse = ", ")` across the six elections.

The exceptions are worth their story. In 2016 there were none at all, the first presidential election since 1913 in which every state that elected a senator elected one from the party that carried it. Eight years later there were `r n(fn("sen_split_2024"))`. Nationalization did not reverse between 2016 and 2024. Thirty-odd contests is too few for a count of crossings to be steady, and a count of crossings is a different thing from the distance underneath it.

The House series has enough contests to hold that distinction up to the light.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=6.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. `r n(fn("h_years"))` presidential elections, `r n(fn("h_districts"))` district-elections.** Only presidential years are drawn, so every point compares a House election with the presidential election held the same day. The top panel is the share of districts that split. The bottom panel holds the two quantities that count is made of.

The top panel is the finding everybody quotes: split districts run to **`r p1(HPEAK)`% in `r f("h_peak_year")`** and fall to **`r p1(HLAST)`% in `r f("h_last_year")`**. The bottom panel is why. The distance from the ticket was `r p1(fn("h_wobble_peak"))` points in `r f("h_peak_year")` and is `r p1(fn("h_wobble_last"))` now, smaller by a factor of `r p1(fn("h_wobble_ratio"))`. The spread of the districts went the other way, from `r p1(fn("h_spread_peak"))` points to `r p1(fn("h_spread_last"))`. A district splits when the first number beats the second in the direction that crosses the fifty-per-cent line. Both moved, and both moved the way that makes splitting harder, so the count alone cannot say which did the work.

The swap experiment in `swap.csv` can. Take `r f("h_peak_year")`'s districts and keep their presidential vote exactly as it was: the same country, the same lines, the same landslide. Now give those districts `r f("h_last_year")`'s candidates: each district keeps its rank, but takes the distance from the ticket at the same rank in the other year. Count the splits again. Nothing is simulated and nothing is random, and handing a year its own distances back reproduces its own count exactly, which is checked before anything is written.

```{r fig3-static, fig.height=4.4}
```

**Figure 3. The same districts, other years' candidates.** The two dark bars are the elections as they happened. The two light bars are each year's districts run with the other year's candidates.

Changing the candidates alone takes `r f("h_peak_year")` from `r p1(HPEAK)`% down to `r p1(fn("swap_old_new"))`%, almost the entire fall, without moving a single district. Changing the districts alone takes `r f("h_last_year")` from `r p1(HLAST)`% up to `r p1(fn("swap_new_old"))`%, a large move the other way. So the honest answer is not one cause. But the two are not equal: the candidates alone land within `r p1(fn("swap_old_new") - HLAST)` points of where the country actually is, and the districts alone land `r p1(fn("swap_new_old") - HLAST)` points away.

The count told you none of this. "Split districts fell from `r p1(HPEAK)`% to `r p1(HLAST)`%" is true and quoted constantly, and it is a threshold: a line drawn at fifty per cent with a yes or a no on either side. Two different quantities moved underneath it, and the threshold reports their combination as one number. When a share of cases crosses a line, ask what would have to change for a case to land on the other side. There is usually more than one answer, and the count gives you none of them.

## What this brief cannot tell you

**No voter appears anywhere in it.** Every number is a total for a place. In a district that split, many people may have split their ballots, or two different sets of people turned out for the two contests. Guessing what individuals did from group totals is called ecological inference, and it has [its own chapter](../../06-putting-data-together/rpv/rpv-brief.html).

**The roster measure and the returns measures are not the same measurement.** Figure 1 counts states by who was sitting; the Clerk's table counts contests by who was elected. A state can look split in the roster for six years on the strength of one election, so the series are not expected to agree year for year. And the Senate returns series is six elections long, an order rather than a trend, which is why the House series carries the weight.

**The swap is a re-ranking, not a world.** It says what the count would be if the distances between House candidates and their tickets were distributed as in another year. It does not claim any particular district would have gone the other way.

**Party is the ballot line throughout.** Read the Senate by who sits with whom in Washington instead, and the last Congress holds `r n(fn("del_last_caucus"))` split states rather than `r n(fn("del_last_ballot"))`.

**The claim reaches further than the record does.** Nationalization is also a claim about governors, legislatures and school boards, where a voter leaning on the party label is voting nationally whether they mean to or not. No national office collects those results, so the claim is best measured exactly where it is least surprising.

## What you should have learned

- Split Senate delegations peaked in `r f("del_peak_year")` at `r n(fn("del_peak_n"))` of 50 states and have fallen to `r n(fn("del_last_n"))` in `r f("del_last_year")` — the textbook's claim is true in the roster, and the fall took decades.
- Split House districts fell from `r p1(HPEAK)`% in `r f("h_peak_year")` to `r p1(HLAST)`% in `r f("h_last_year")`, across `r n(fn("h_districts"))` district-elections.
- Two quantities moved under that count: the typical House candidate's distance from the ticket shrank by a factor of `r p1(fn("h_wobble_ratio"))`, while the spread of the districts grew from `r p1(fn("h_spread_peak"))` to `r p1(fn("h_spread_last"))` points.
- Swapping in today's candidates alone takes `r f("h_peak_year")` from `r p1(HPEAK)`% to `r p1(fn("swap_old_new"))`%; a count of threshold crossings hides which ingredient did the work.
- A count of split outcomes can be noisy while the distance underneath it moves steadily: the Senate produced `r n(0)` splits in 2016 and `r n(fn("sen_split_2024"))` in 2024 without nationalization reversing.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `delegations.csv` gives `pct_split` for every Congress across a century and a half. Follow it backward. Was there ever a time when states routinely sent two senators of different parties?
- `house_years.csv` carries `wobble` and `spread` alongside `pct_split`. Which of the three is the best single measure of nationalization, and why?
- `senate_contests.csv` has a `gap` between the Senate and presidential result in each state. Sort by it. Which senators still outrun their party, and where?
- `delegations.csv` distinguishes `pct_split` from `pct_split_two`. Compare them decade by decade. How much does the handling of a few mid-term replacements change the story?

Two stretch ideas point past the spreadsheet. Pick one state, download its certified returns for a down-ballot office such as attorney general, and compute that office's distance from the presidential ticket — the measure this brief could not extend below Congress. And the Clerk's volumes run much further back than 2004: download an older one and see whether the Senate distance was ever as large as the House distance of the 1970s.

## Sources

**Data**

- Lewis, Jeffrey B., Keith Poole, Howard Rosenthal, Adam Boche, Aaron Rudkin
  and Luke Sonnet. *Voteview: Congressional Roll-Call Votes Database*,
  `HSall_members.csv`, one row per member per Congress since 1789. Downloaded
  fresh here; the [rosters-source chapter](../rosters-source/rosters-source-brief.html)
  is about what a roster is and is not.
  <https://voteview.com/static/data/out/members/HSall_members.csv>
- U.S. House of Representatives, Office of the Clerk. *Statistics of the
  Presidential and Congressional Election*, the volumes for 2004 through 2024,
  read from the copies the [house-competition
  chapter](../house-competition/house-competition-brief.html) downloaded and
  converted to text. The 2024 Senate contests were parsed a second time,
  independently, by the [crossover chapter](../crossover/crossover-brief.html),
  and the two parses agree on every contest.
  <https://clerk.house.gov/Votes/ElectionStatistics>
- The House series, 1952 to 2024, is the house-competition chapter's file,
  built from Gary Jacobson's House elections data, the Clerk's volumes and The
  Downballot's presidential results by congressional district. No government
  agency computes that last number.
  <https://www.the-downballot.com/p/the-downballots-calculations-of-presidential>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`delegations.csv`](data/derived/delegations.csv), [`house_years.csv`](data/derived/house_years.csv), [`senate_contests.csv`](data/derived/senate_contests.csv), [`senate_years.csv`](data/derived/senate_years.csv), [`swap.csv`](data/derived/swap.csv)

```{r ai-prompt, results='asis'}
```
