---
title: "Key: The Door That Revolves, On the Record"
subtitle: "84-355 Democracy's Data | Lobbying | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
f  <- read.csv("data/filings.csv",   stringsAsFactors = FALSE)
lo <- read.csv("data/lobbyists.csv", stringsAsFactors = FALSE)
is <- read.csv("data/issues.csv",    stringsAsFactors = FALSE)
en <- read.csv("data/entities.csv",  stringsAsFactors = FALSE)
lo$has <- nchar(lo$former_government_position) > 0
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`lobbying-brief.Rmd`) is now a
chapter rather than a parallel copy of the lab: prose organised by argument,
under idea headings, with three numbered figures and no numbered steps. The lab
(`lobbying.Rmd`) still runs Steps 1-10 with the code exposed. **Students should
read the brief first.** **Step numbers below refer to the lab**; the chapter is
cited by figure number or section title.


**Pairs with the campaign-finance labs.** Those are money spent to decide who
wins; this is money spent on what happens next, and it is larger. Teaching them
in the same week makes the contrast do the work.

Every number below was run against the data files.

***

# The session in one sentence

Congress passed a law requiring lobbyists to disclose that they used to work for
Congress, more than a quarter of them do, and the field recording it is more
precise than the field recording the money.

***

# The findings

```{r headline}
c(filings = nrow(f), lobbyists = nrow(lo),
  with_former_gov_job = sum(lo$has),
  pct = round(100 * mean(lo$has), 1),
  median_filing = median(f$amount[f$amount > 0]))
```

- **216 of 773 lobbyists (27.9%)** disclose a previous government position.
- Median filing reports **$30,000** for the quarter.
- Top issue: **Budget/Appropriations**. Most-lobbied: **the House and Senate**,
  then HHS and the White House.
- The sample is **1,500 of 19,016** Q2 2024 filings — see below.

***

# The three things to get said

**1. The revolving door is not an exposé here — it is a form field.** Students
expect this kind of fact to come from investigative journalism. It comes from
`covered_position`, a box on a federal disclosure form, because the **Lobbying
Disclosure Act of 1995** requires it. **Read three or four of the entries
aloud.** "U.S. Representative Nancy Pelosi (D-CA)" lands harder than any
percentage.

**2. Then ask what disclosure accomplishes.** Requiring somebody to announce a
conflict is not the same as preventing it. Congress chose the weaker of the two
tools, and the data exists precisely *because* it chose the weaker one. **That
is worth five minutes and has no clean answer.**

**3. Compare the precision of the two fields.** `covered_position` is specific
enough to name a member of Congress. The money is a single number per filing,
undivided across issues, with no meetings, dates or people named. **Both were
written by the same Congress in the same statute.** Ask which lobbyists would
have objected to which.

**The line for the board:** everything here is public because somebody decided
disclosure was enough.

***

# The sampling problem — teach it, do not hide it

**We have 1,500 of 19,016 filings from one quarter.** The Senate's API
rate-limits anonymous users to 25 records a request and returns 429 if pushed.

This is not a footnote; **it is the third instance of the course's running
theme** — the census API needing a key, VEST behind a guestbook, and now a
public database that hands out its contents 25 rows at a time. Say the pattern
out loud: **"public" and "usable" are different properties.**

And be strict about what survives it. Rates (the 27.9%) are fine. Rankings
(Budget/Appropriations first) are probably fine. **Totals are meaningless** and
students should not quote the $45 million.

***

***

# Walking the steps

| Steps | What they do | Min | Watch for |
|---|---|---|---|
| **1-3** | One filing; the 7.9% sample; the money column and its zeros | 10 | **Public is not usable.** The rate limit is a finding, not an excuse. |
| **4-6** | The number that does not exist; issues; entities | 12 | **The pivot.** One dollar figure, 17 issue areas, no division. |
| **7-8** | The covered-position field, then reading the entries | 12 | Precise enough to name a senator — and loose enough to contain 'President and CEO'. |
| **9** | The asymmetry | 6 | Same statute, same Congress: one field too specific to aggregate, one too coarse to attribute. |
| **10** | What you can say | 5 | The finding is about the regime, not the lobbyists. |

**Figure 2 of the chapter is the form itself**, under *The question that cannot
be asked*. It is one filed LD-2 — Barker Leavitt, PLLC for Millcreek, Utah, Q2
2024, which is a row of `filings.csv`, not an illustration of one — with four
boxes marked on it: the single income figure (line 12), the two general issue
area codes on separate pages (line 15), the institutions contacted (line 17), and
the covered-position field (line 18). Put it on the screen before arguing about
the data, because every complaint the lab raises in its Steps 4 to 9 is a
property of that piece of paper. Two
details worth pointing at: the same lobbyist is listed twice under two different
descriptions of his own career, and on the second page his surname is misspelt.
Nothing on the form is validated. The fetch, the redaction of the filer's
telephone number and e-mail, and the annotations are in
`data/fetch-form-image.R`.

**The form and the asymmetry are the pair that matters** — Figure 2 and *The same statute, in two failing directions* in the chapter, Steps 4 and 9 in the lab. Ask the class how much was spent lobbying on health care. Let them look for it. The answer is that the form does not ask.

**If you are short on time**, protect the pivot and the closing step. The numbers are what make the argument concrete; the sequence is the argument.


# Option A — Firms versus causes

## Verified output

```{r a}
r <- sort(table(f$registrant), decreasing = TRUE)
head(data.frame(registrant = names(r), filings = as.vector(r)), 8)
c(unique_registrants = length(unique(f$registrant)),
  unique_clients     = length(unique(f$client)))
```

**3/4:** Reports the most frequent registrants and notices a few firms file many
times.

**4/4:** Reads the client lists and sees that a single firm represents clients
with nothing in common — a dispatch centre, a university, a trade association.
**That is not subject-matter expertise. It is access being rented**, and the
lobbyist's asset is the relationship, not the policy knowledge.

**4/4, best:** Connects it to *A field precise enough to name a senator* — if the asset is access, the
covered-position field is a **price list**, and a firm's value is partly the sum
of its employees' former employers. A student who says the disclosure form
inadvertently documents the business model has it.

***

# Option B — Search the door

## Verified output

```{r b}
p <- lo$former_government_position[lo$has]
c(mentions_Senator        = sum(grepl("Senator", p, ignore.case = TRUE)),
  mentions_Representative = sum(grepl("Representative|Congressman|Congresswoman",
                                      p, ignore.case = TRUE)),
  mentions_Chief_of_Staff = sum(grepl("Chief of Staff", p, ignore.case = TRUE)),
  mentions_Counsel        = sum(grepl("Counsel", p, ignore.case = TRUE)),
  total_with_position     = length(p))
```

**3/4:** Counts the categories and reports which is commonest.

**4/4:** Notices that **staff roles vastly outnumber members**. Former members
are the famous case; the volume is chiefs of staff, legislative directors and
committee counsel — the people who actually wrote the bills. A student who says
the door revolves mostly at staff level has corrected a popular misconception.

**4/4, best:** Notices the field is **free text with no controlled vocabulary** —
"U.S. Representative Nancy Pelosi (D-CA)" and "former Legislative Director for
Congressman Mike Turner (R-OH)" are formatted completely differently — so any
count is a guess about strings. **The law requires disclosure and does not
require it to be machine-readable**, which is a very old trick.

***

# Option C — What the sample costs you

No verified output; grade the reasoning.

**3/4:** Says rates are safer than totals.

**4/4:** Distinguishes three things — **rates** (fine, if the sample is not
selected on the outcome), **rankings** (usually fine, since it takes a big error
to reorder the top of a skewed distribution), and **totals** (hopeless, off by
roughly 13×). Then notes the sample is not random: it is whatever the API
returned first, which is likely near-chronological.

**4/4, best:** Identifies the specific danger — **the largest filings are rare**,
so an 8% sample of a heavily skewed distribution will usually miss them
entirely. Any statement about "the biggest lobbying spenders" from this file is
unsupported no matter how carefully hedged. **Knowing which claims are
unavailable is the skill.**

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating disclosure as a policy choice with
  consequences rather than as a neutral source of data.
- **A 2** is "lobbying is corrupt," which the data does not show and was not the
  question.
- **Also a 2**: quoting a total from an 8% sample.
- **Reward anyone who notices the free-text field** and what it implies.

## Anticipated questions

- *"Is any of this illegal?"* — No. Everything here is legal, disclosed, and
  filed on time. That is what makes it interesting.
- *"Who checks the filings?"* — The Secretary of the Senate and Clerk of the
  House refer non-compliance to the U.S. Attorney; enforcement is rare. Compare
  the FEC's independent-expenditure junk filings, where nobody checked either.
- *"What about people who don't register?"* — The definition requires multiple
  contacts and a share of working time. **Strategic advisers who never contact
  an official file nothing**, and that is the largest hole in the regime.
- *"Can I look up a specific person?"* — Yes, the database is searchable at
  lda.senate.gov. Students find this genuinely startling.

## Connections

- **Campaign finance** — the same money, before the election instead of after.
- **Independent expenditures** — the other disclosure regime, and the other
  place where the government publishes what was filed without checking it.
- **EAVS** — self-reporting by the people being measured.
- **The Census API and VEST** — the recurring gap between public and usable.
