---
title: "The Door That Revolves, On the Record"
subtitle: "84-355 Democracy's Data | Federal lobbying disclosures, 2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `lobbying-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"He left Congress and became a lobbyist."*

Campaign finance is money spent to decide **who wins**. This is money spent on
what happens **afterwards** — and it is much larger, much less covered, and in
one respect much better documented.

The **Lobbying Disclosure Act of 1995** requires anyone paid to lobby the
federal government to file quarterly: who paid them, how much, what they lobbied
about, which parts of the government they contacted, and — the part that makes
today possible — **whether the lobbyist previously held a government position,
and which one.**

Congress wrote a law requiring lobbyists to disclose that they used to work for
Congress. Today we read what they filed.

> **A warning about how this lab is built.** Two of the steps ahead are about
> things this data cannot do: a question it cannot answer, and a money column
> that cannot be attributed to anything. **Those are not digressions.** They are
> not throat-clearing to be got through before the real findings arrive — they
> are the findings. A disclosure regime is worth what its records can be made to
> say, and the most important thing these records say is what they decline to.

# Steps 1-3 — One filing, how much of the quarter this is, and the money column

```{r load}
f  <- read.csv("data/filings.csv",   stringsAsFactors = FALSE)
lo <- read.csv("data/lobbyists.csv", stringsAsFactors = FALSE)
is <- read.csv("data/issues.csv",    stringsAsFactors = FALSE)
en <- read.csv("data/entities.csv",  stringsAsFactors = FALSE)

c(filings = nrow(f), unique_lobbyists = nrow(lo), issue_codes = nrow(is))
head(f[order(-f$amount), c("client", "registrant", "amount")], 6)
```

One row per **filing**: a lobbying firm (the *registrant*) reporting what it did
for a *client* in one quarter.

**A note on the sample, because it matters.** There were **19,016** filings in
the second quarter of 2024. We have **1,500** — about 8% — because the Senate's
API rate-limits anonymous users to 25 records at a time and refuses to go faster.
The data is public and the pipe is narrow. **Every rate below is a sample
estimate; every total is a fraction of the true total.**

# Steps 4-6 — The number that does not exist, what is lobbied on, and whom

> **Try to answer an obvious question first:** how much was spent lobbying on
> health care this quarter? Look at the columns you have and work out whether
> you can get there. **The answer to that attempt is the finding.**

```{r issues}
head(is, 10)
```

**Look at the column heading.** The quantity is `mentions`, not dollars. That is
where your attempt died, and it died for a reason you can see on the form: a
filing reports one amount for the quarter and a list of issue areas, with no
division of the money between them. Step 9 is that money column, refusing to be
split.

```{r issue-plot}
top <- head(is, 10)
barplot(rev(top$mentions), horiz = TRUE, names.arg = rev(substr(top$issue, 1, 28)),
        las = 1, cex.names = 0.7, col = "#2c7fb8", xlab = "activity mentions")
```

**Budget and appropriations first**, then health, defence, transportation. That
ordering is not about ideology. It is about **where money is decided** — the
appropriations process moves more dollars than anything else Congress does, so
that is where the lobbyists are.

```{r entities}
head(en, 8)
```

The House and the Senate dwarf everything else, and then it is the agencies that
write cheques and rules: HHS, the White House, USDA, Energy.

# Steps 7-8 — The unusual field, and reading the entries

```{r door}
lo$has_position <- nchar(lo$former_government_position) > 0
c(lobbyists = nrow(lo),
  with_former_gov_job = sum(lo$has_position),
  pct = round(100 * mean(lo$has_position), 1))
```

**216 of 773 lobbyists in this sample — 27.9% — disclose a previous government
position.** More than one in four.

> **Before you run the next chunk, write down what you expect to see.** The Act
> requires a lobbyist to disclose a previous government position. A requirement
> like that could be satisfied with two words — "federal service" — and nobody
> could say it had been broken. **How specific are these twelve entries going to
> be?** Commit to an answer before you look: vague job categories, or named
> members of Congress and named committees.

```{r door-sample}
set.seed(84355)
sample(lo$former_government_position[lo$has_position], 12)
```

Read them. **They are not vague, and hardly anybody writes that down.** They
name the member, the committee, the agency — sometimes several in sequence, with
dates — because the form asks for it.

**This is the fact the lab exists to deliver.** The revolving door is not an
inference, a leak, or an investigation. **It is a disclosure requirement**, and
the government publishes the answers in a searchable database. One in four paid
federal lobbyists used to be on the other side of the table, and they said so on
a form.

# Step 9 — The asymmetry

```{r money}
p <- f[f$amount > 0, ]
c(filings_reporting_money = nrow(p),
  total = sum(p$amount),
  median = median(p$amount),
  largest = max(p$amount))
```

```{r money-plot}
hist(log10(p$amount[p$amount > 0]), breaks = 30, col = "#C41230", border = "white",
     main = "", xlab = "log10( dollars reported in the quarter )")
```

**A median filing reports $30,000 for the quarter.** The distribution is
enormously skewed — most filings are small, a few are very large.

Now notice what you cannot do with this. **The amount is per filing, not per
issue and not per contact.** A firm reporting $150,000 across nine issue areas
does not tell you how much went to any one of them. There is no requirement to
say, so nobody does.

You also cannot see **who was met, or when**. The filing names the House, the
Senate, an agency — not a person, not a date, not a meeting. Compare that to the
precision of the covered-position field, and ask why one was written strictly
and the other loosely.

# Step 10 — What you can now say

**You can say** that more than a quarter of registered federal lobbyists claim
a prior government position, and that many of those claims **name a specific
member of Congress** — self-reported evidence of a revolving door, in a federal
database.

**You can say lobbying concentrates on appropriations and on Congress itself**,
in that order.

**You cannot say** how much was spent on any issue, who was actually met, when,
or what was asked for. And you cannot report a national total from this sample.

**The most transferable finding is about the regime, not the lobbyists.**
Disclosure was chosen instead of prohibition, and it was designed so that the
two facts people most want — how much, on what — cannot be recovered from it.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Firms versus causes.** Group `filings.csv` by `registrant` and by
`client`. Which registrants appear most often, and what does it mean that the
same handful of firms represents dozens of unrelated clients? Is that expertise
or access?

**Option B — Search the door.** The `former_government_position` field is free
text. Count how many mention "Senator", "Representative", "Chief of Staff", or a
named agency (`grepl`). Which kind of former job is most common, and which would
you expect to be most valuable to a client?

**Option C — What the sample costs you.** We have 8% of one quarter. Which of
the numbers in this lab would you still trust, and which would you not? Be
specific: rates versus totals versus rankings.

**Option D — Your own question.** 1,500 filings, 773 lobbyists, 76 issue areas.

# What this lab cannot tell you

- **Whether lobbying works.** No outcomes here at all. The filings record effort
  and expenditure, never results.
- **The real totals.** 8% of one quarter of one year, and not a random 8% —
  it is whatever the API returned first. Rankings and rates are informative;
  sums are not.
- **Who was actually met.** The law requires naming the chamber or agency, not
  the person, meeting or date.
- **Anything about the unregistered.** People who influence policy without
  meeting the legal definition of "lobbyist" — strategic advisers, consultants
  under the time threshold — file nothing. **The disclosure regime measures the
  part of the activity that agreed to be measured.**

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- More than one in four lobbyists disclosed a former government job. Is the
  disclosure requirement a solution to the revolving door, or a substitute for
  one?
- The covered-position field is precise and the money field is vague. Who
  benefits from that asymmetry?
- The data is public and the API hands it over 25 rows at a time. Does that
  count as public?

***

### Where the data came from

**United States Senate, Lobbying Disclosure Act database**, filings API —
second quarter of 2024. <https://lda.senate.gov/api/v1/filings/>

No key required, and **rate-limited to 25 records per request for anonymous
users**, which is why this is a 1,500-filing sample of 19,016. The build is in
`data/build-data.py`, which throttles deliberately and says so.

Reporting is required by the **Lobbying Disclosure Act of 1995**, as amended by
the Honest Leadership and Open Government Act of 2007.
