---
title: "Primary Defeats of House Incumbents, 1946–2024"
type: brief
subtitle: "The election that is supposed to substitute for the one nobody contests"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("primary-defeats-code.R")
```

```{r setup, include=FALSE}
```

On the tenth of June 2014, the Majority Leader of the United States House of Representatives lost a primary election. It was in Virginia's seventh district, to an economics professor who had spent about two hundred thousand dollars. Eric Cantor was the second-ranking Republican in the House and, by most accounts, the next Speaker. No sitting Majority Leader had ever been denied renomination since the office was created in 1899.

The commentary drew the obvious lesson: incumbents live in fear of being "primaried." This brief asks the question the lesson skips. How often does a sitting member actually lose a primary?

The number answers a specific argument. [About one House race in seven](../house-competition/house-competition-brief.html) has a single name on the November ballot, and the standard reply is that a safe seat has not removed accountability but moved it upstream, into the summer. Between `r Y1` and `r Y2`, `r n(NRAN)` sitting members stood on a primary ballot. `r NDEN` of them were refused — `r pc(RATE, 2)` percent, Cantor among them.

## The test

The main source is the **Federal Election Commission's** *Federal Elections* series, the biennial compilations for `r Y1` through `r Y2`, obtained from the [Commission](https://www.fec.gov/introduction-campaign-finance/election-results-and-voting-information/). For every federal candidate it reports the primary, runoff and general-election vote, assembled from the states' official counts. Three properties make it the right source: it is one step from the authoritative record, it is candidate-level, so a losing challenger appears with their votes, and it marks incumbents. The obvious alternative does not exist: the Clerk's *Statistics of the Congressional Election* contains no primary results in any edition.

The machine-readable workbooks cover only ten elections, and ten elections holding `r NDEN` events is not many. So a second series runs alongside. Table 2-7 of **Brookings'** *Vital Statistics on Congress* counts primary defeats for every election from 1946 to 2024, compiled from the *Biographical Directory* and *CQ Almanac* rather than from returns. A third source, congressional reporter **Greg Giroux's** hand-kept tracker of members defeated for renomination, is used only as an independent check. And who counts as an incumbent is decided against the roster: Voteview's membership file, the record the [roster chapter](../rosters-source/rosters-source-brief.html) introduces. The FEC's own incumbent marker turns out to be wrong in both directions.

The test: derive the denial rate from the workbooks, check it against two independent hand-built counts, and then ask whether the primary looks like the site of accountability the argument needs.

## The data

The clean file is `incumbents.csv`: one row per sitting member per election, an *incumbent-cycle*, whether or not anybody ran against them. Its columns, by name: `year`, `stab` and `dnum` for the district, `party`, `unopposed`, `prim_votes` and `prim_share`, `rivals`, `won_nom`, and `denied`. Here is Cantor's career in it:

```{r cleaninc}
```

Four cycles where nobody filed against him, one challenger beaten four to one, and `r max(IN$year[IN$last == "Cantor" & IN$denied])`. The column this brief is about, `denied`, exists in no source file: the workbooks hold a row per candidate per ballot line, and Cantor's defeat is recorded nowhere. **It is inferred, from an absence** — a primary vote total, a rival with more votes, and an empty general-election cell. `denied` is `ran_prim & !won_nom`, computed here.

Getting from ballot lines to that table meant deciding, for each member, which contest was actually at stake and whether the incumbent came first. Every rule below was found by a number coming out wrong.

**A blank is not always missing.** Cantor's empty general-election cell means he was not on the ballot; elsewhere in the same column a blank means the FEC wrote the word `Unopposed` instead of a number. Reading those as missing would have deleted `r n(sum(IN$unopposed))` incumbent candidacies from the denominator, the bottom number being divided by.

**A ballot line is not a candidacy.** Joe Crowley lost his 2018 primary and was still on New York's Working Families line in November; asked of all his rows at once, "was this member nominated?" answers yes. Louisiana is the same problem in state form: in most of these years its "primary" *is* its general election.

**And the file's own flags cannot be trusted alone.** The incumbent marker is checked against the roster: of `r ck("incumbent-cycles in the FEC files")` marked candidacies, `r ck("incumbent-cycles matched to a Voteview member")` were confirmed, `r ck("not matched at all, and dropped")` dropped, and `r ck("incumbents the FEC did not mark, recovered by id")` unmarked members recovered by identifier. The nomination rule is checked against the FEC's own winner flag, printed from 2020: `r ck("same test restricted to incumbents: candidacies compared")` incumbent candidacies, `r ck("of those, incumbent candidacies where the two disagree")` disagreements — a winner who died, one who resigned, one who withdrew before a runoff.

The result agrees with the people who counted by hand. Across the `r nrow(CM)` overlapping elections, Brookings and the Giroux tracker agree with each other exactly, and this derivation matches them in `r sum(CM$gap_brookings == 0)` of `r nrow(CM)`, never off by more than `r max(abs(CM$gap_brookings))`. Every miss is this count running short for a named reason: a missing incumbent marker, Louisiana's blanket ballot, or a convention with no primary to lose.

> **Before you read on, write a number down.** In `r Y2`,
> `r by(Y2, "ran_primary")` House incumbents stood on a primary ballot.
> **How many of them had even one opponent on that ballot?** Any name at all
> besides their own. Commit to a count before you look at the figures.

## What it says about democracy

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=3.9}
```

**Figure 1. Every House member-cycle from `r Y1` to `r Y2`, narrowed one condition at a time.** A funnel, because the argument is about survival at each stage rather than a trend; widths are on a square-root scale so the last band is visible at all. Of the members on a primary ballot, `r pc(100 * NCON / NRAN)` percent had at least one rival on it, and `r pc(100 * NDEN / NCON)` percent of those were refused the nomination.

`r NDEN` in `r n(NRAN)` is `r pc(RATE, 2)` percent. The Brookings series, over forty elections rather than ten, gives `r sum(VS$lost_primary)` defeats against `r n(sum(VS$seeking))` members seeking re-election — `r pc(VSR, 2)` percent, the same number counted by different people from different documents. **A sitting member of the House who seeks renomination gets it about ninety-eight and a half times in a hundred.** That is the fact the accountability argument has to be made against.

It would be easy to conclude that nobody bothers. That would be wrong.

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

**Figure 2. Every contested incumbent primary from `r Y1` to `r Y2`, one tick per member.** A barcode rather than a histogram, because the point is the individual members in the left tail. A tick appears only for incumbents who had a rival, at the share of the primary vote they took; red ticks are the `r NDEN` members refused renomination.

The count rose from `r by(Y1, "contested")` of `r by(Y1, "ran_primary")` in `r Y1`, or `r pc(by(Y1, "pct_contested"), 0)` percent, to `r by(Y2, "contested")` of `r by(Y2, "ran_primary")` in `r Y2`, or `r pc(by(Y2, "pct_contested"), 0)` percent. Most readers guess near a quarter. **Being challenged and being beaten have come apart.** Challenges roughly doubled; the denial rate did not. Incumbents held under sixty percent of their own primary went from `r by(Y1, "under60")` to `r by(Y2, "under60")`, so the pressure is real and measurable. It simply almost never finishes the job.

One caution: the earlier workbooks record fewer minor candidates, so the trend's direction is better supported than its size.

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

**Figure 3. House incumbents denied renomination at every election from `r min(VS$year)` to `r max(VS$year)`, from the Brookings series.** The columns in the accent color are the exceptions with a story: the first election held on each decade's new districting map. Those eight average `r pc(RD$lost_primary[RD$redist])` denials; the other `r sum(!VS$redist)` elections average `r pc(RD$lost_primary[!RD$redist])`, and the tallest column is `r VS$year[which.max(VS$lost_primary)]` at `r max(VS$lost_primary)`.

**The clustering is a redistricting effect, and its mechanism is not what the word "primary" suggests.** A new map does not usually abolish a member's district; it puts two members into one, and two sitting members in one district produce one primary that one of them must lose. The count doubles without anything changing about how voters feel. Strip those years out and the series is flat and low. There is no long rise in members being turned out by their own voters — the rise a theory of primaries as the new site of accountability would predict.

So where does the fear fit? The literature's answer: upstream of the ballot. Hall (2019) argues the deeper effect is on *who runs*: if the only plausible candidates in a safe district are ones a primary electorate would accept, the primary chose the field before anybody voted. Brady, Han and Pope (2007) show that members who drift toward the centre of their district pay a price in primary vote share — a price paid in share, and share is not a seat.

And there is a mechanical answer this file can support: **anticipation**. A member who expects a serious primary can retire instead, and `r pc(100 * sum(RE$not_running) / sum(RE$left), 0)` percent of House departures are members who did not seek re-election at all. No file separates the ones who left because they wanted to from the ones who counted the votes first. **A primary that removes members by frightening them leaves exactly the same trace as no primary at all.**

The finding has two halves: challenge has become ordinary, and defeat has not. Either the fear does the work upstream or nothing does — and these files, which record only what reached a ballot, cannot say which.

## What this brief cannot tell you

**Ten elections is a small window for a rare event.** `r NDEN` denials is enough to establish a rate and not enough to model one; the Brookings column extends the rate to forty elections but is not built from returns.

**Three kinds of nomination are invisible here.** A convention, Louisiana's blanket ballot, and the top-two primaries of California, Washington and Alaska make "denied renomination" either impossible or something else. `r sum(DN$nonpartisan)` of the `r NDEN` denials happened under a top-two system and are flagged in the data.

**A share of the primary vote is not a measure of danger.** An incumbent held to sixty percent by a strong challenger looks identical to one held there by three unknowns; the file has no candidate quality, no spending, no endorsements. Turnout is missing too: these are shares of whoever voted, and primary electorates are a small, unrepresentative slice of a district.

**And the counterfactual is unobservable.** Whether the primary holds members to account turns on what members do because they fear one, and fear is not a column. This brief can say how often the mechanism fires. It cannot say how often the threat was enough.

## What you should have learned

- A sitting House member who seeks renomination is refused about `r pc(RATE, 1)` percent of the time: `r NDEN` of `r n(NRAN)` in the FEC years, and `r pc(VSR, 2)` percent across the Brookings series back to 1946.
- The event this brief counts exists in no source file: `denied` is inferred from an absence — a primary vote, a rival with more, and an empty general-election cell.
- Being challenged and being beaten are different series: the share of incumbents facing any primary opponent roughly doubled between `r Y1` and `r Y2`, while the denial rate stayed flat.
- The spikes in primary defeats are a redistricting effect: the first election on a new map averages `r pc(RD$lost_primary[RD$redist])` denials against `r pc(RD$lost_primary[!RD$redist])` otherwise, because a new map pairs two members into one district.
- A derived count is only as good as its checks: this one matches two independent hand-built counts in `r sum(CM$gap_brookings == 0)` of `r nrow(CM)` elections, every miss for a named reason.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `compare.csv` puts three independent counts of the same thing side by side, with `gap_brookings` and `gap_giroux`. Find the years they disagree. Whose count would you use?
- `by_year.csv` has `pct_unopposed` and `pct_contested`. Follow them across the ten cycles. If the general election is uncontested and the primary is unopposed too, who chose the member?
- `denied.csv` lists the incumbents who lost renomination, with `prim_share`. Sort by it. How badly do incumbents lose when they lose?
- `vsoc.csv` splits every year since 1946 into `retired`, `lost_primary`, `lost_general` and `reelected`. Compute the share leaving by each route. Which way out of Congress is most common?

Two stretch ideas point past the spreadsheet. The Giroux tracker covers the current cycle, which the FEC will not report for two more years: download it and see whether this cycle looks like a redistricting year or an ordinary one. And the same workbooks carry the Senate — work out whether senators are denied renomination more or less often than House members.

## Sources

**Data**

- Federal Election Commission, *Federal Elections*, biennial compilations for
  2004 through 2022.
  <https://www.fec.gov/introduction-campaign-finance/election-results-and-voting-information/>
  — retrieved 2026-08-11, `r ck("FEC cycles read")` workbooks,
  `r ck("FEC House candidate rows, 2004-2022")` House candidate rows. **The
  vote totals in this brief are the Commission's, compiled from the official
  canvasses of the state election officials.**
- Lewis, J. B., Poole, K., Rosenthal, H., Boche, A., Rudkin, A. and Sonnet, L.
  *Voteview: Congressional Roll-Call Votes Database*, member file
  `HSall_members.csv`. <https://voteview.com/> — retrieved 2026-08-11,
  `r ck("Voteview rows")` rows. Used to decide who was an incumbent.
- Brookings Institution, *Vital Statistics on Congress*, Chapter 2, Table 2-7,
  edition of April 2026.
  <https://www.brookings.edu/articles/vital-statistics-on-congress/>
  — 1946 to 2024, compiled from the *Biographical Directory*, *CQ Almanac* and
  *National Journal*.
- Giroux, G. *Members of Congress Defeated For Renomination, 1966–2026.*
  <https://docs.google.com/spreadsheets/d/1X97S6hUby6PPHAMoEwvAxkAUo8EUwLk8xMosx4hVd8c/>
  — retrieved 2026-08-11, `r ck("Giroux tracker rows")` rows in
  `r ck("Giroux year-blocks parsed")` year-blocks. A hand-kept tracker, cited
  as an independent check.
- Office of the Clerk, U.S. House of Representatives, *Statistics of the
  Congressional Election*, 1946 to 2024.
  <https://history.house.gov/Institution/Election-Statistics/Election-Statistics/>
  — checked and **not** used: it contains no primary results in any edition.

The assembly is shared with the [`retirements`](../retirements/retirements-brief.html) brief, which downloaded and keeps the workbooks; the validation results are reproduced in `checks.csv`.

**Academic literature**

- Boatright, R. G. (2013). *Getting Primaried: The Changing Politics of
  Congressional Primary Challenges.* University of Michigan Press.
- Hall, A. B. (2019). *Who Wants to Run? How the Devaluing of Political Office
  Drives Polarization.* University of Chicago Press.
- Brady, D. W., Han, H. and Pope, J. C. (2007). "Primary Elections and
  Candidate Ideology: Out of Step with the Primary Electorate?" *Legislative
  Studies Quarterly* 32(1): 79–105.
- Mayhew, D. R. (1974). *Congress: The Electoral Connection.* Yale University
  Press.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_year.csv`](data/derived/by_year.csv), [`compare.csv`](data/derived/compare.csv), [`denied.csv`](data/derived/denied.csv), [`giroux.csv`](data/derived/giroux.csv), [`incumbents.csv`](data/derived/incumbents.csv), [`vsoc.csv`](data/derived/vsoc.csv), [`exits_by_year.csv`](../retirements/data/derived/exits_by_year.csv)
