---
title: "Lines Drawn in 1937"
subtitle: "84-355 Democracy's Data | HOLC grades against the 2020 census"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6,
                      dpi = 96, fig.retina = 1)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **One of three labs in this unit.** All three ask the same question of
> different data: **can you show discrimination when nobody wrote down a
> reason?** The others are policing and jury selection. Whichever you did, you
> will be asked to explain your inference to somebody who did a different one.

> **Read `redlining-brief.Rmd` first.** This lab works through the same material with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"You can still see the redlining maps in the census data."*

In the late 1930s the **Home Owners' Loan Corporation** — a federal agency —
graded neighbourhoods in 200-odd American cities for mortgage risk. **A** was
"Best", green on the map. **D** was "Hazardous", red. The written appraisals
name the presence of Black residents as a reason for a low grade, in so many
words.

The maps were drawn ninety years ago. The claim above says they are still
visible. **That is a testable claim**, and unlike the other two labs in this
unit, here the discriminatory intent is not hidden at all — it is written on the
document. What has to be shown is that it *still matters*.

# Steps 1-3 — Two sources, the join that has to be built, and one record

Two datasets, drawn eighty years apart, that were never meant to meet:

- **HOLC grades**, digitised by the University of Richmond: 10,154 areas drawn
  on the 1930s security maps, 9,324 of them carrying a grade from A to D.
- **The 2020 census**, by tract.

Neither carries the other's identifier. To connect them you have to decide **when
a tract counts as being in a graded area** — and we did the simplest thing:
take the tract's centre point and ask which polygon it falls inside.

```{r load}
# GEOID is text, never a number. A tract identifier is 11 digits and a third of
# the ones here start with a zero (California is state 06). Read as a number
# they come back 10 digits long and match nothing.
tr <- read.csv("data/tracts.csv", stringsAsFactors = FALSE,
               colClasses = c(GEOID = "character"))
ci <- read.csv("data/cities.csv", stringsAsFactors = FALSE)

c(tracts_matched = nrow(tr), cities = length(unique(tr$city)))
table(tr$grade)
```

## The map the grades came off

Before any of that, look at the thing itself. These are the 1939 HOLC areas for
Cleveland, in the four colours the HOLC printed them in — the colour of each
area is read out of the source file, not chosen by us.

```{r holc-map, fig.height=4.4}
rings <- read.csv("data/fig_holc_rings.csv", stringsAsFactors = FALSE)
areas <- read.csv("data/fig_holc_attr.csv", stringsAsFactors = FALSE)
cols  <- setNames(areas$fill[match(c("A", "B", "C", "D"), areas$grade)],
                  c("A", "B", "C", "D"))

par(mar = c(0.1, 0.1, 0.1, 0.1))
plot(NA, xlim = range(rings$x), ylim = range(rings$y), asp = 1,
     axes = FALSE, ann = FALSE)
for (i in areas$id) {
  d <- rings[rings$id == i, ]
  for (q in unique(d$part))
    polygon(d$x[d$part == q], d$y[d$part == q],
            col = cols[[areas$grade[areas$id == i]]], border = "white", lwd = 0.4)
}
legend("topleft", sprintf("%s  %s (%d areas)", names(cols),
       c("Best", "Still Desirable", "Definitely Declining", "Hazardous"),
       as.vector(table(areas$grade)[names(cols)])),
       fill = cols, border = "#999999", bty = "n", cex = 0.7,
       title = "HOLC grade, Cleveland 1939", title.adj = 0)
```

**Red in the middle, green at the edge.** Every number in this lab is a
comparison between the red parts of maps like this one and the green parts,
eighty years later. The polygons come from Mapping Inequality; the brief draws
the same map with the 2020 tract boundaries laid over it, which is where you can
see how badly the two geographies fit.

The crosswalk matches **11,274 tracts across 215 cities** to a HOLC grade
nationally. **4,489 of them, in 56 cities, also have census data here**, because
we pulled the 2020 tract file for six states — PA, IL, MI, OH, MO and CA.

Every one of those matches is a judgement call: a tract whose centre falls just
outside a graded area gets nothing, and a tract straddling an A and a D gets
whichever polygon contains its centre.

**Hold on to that.** It is the same allocation problem as precincts and census
blocks, and there is no correct answer here either.

# Steps 4-5 — The first comparison, and what is wrong with it

```{r national}
g <- aggregate(cbind(total, black) ~ grade, data = tr, sum)
g$pct_black <- round(100 * g$black / g$total, 1)
g
```

```{r nat-plot}
barplot(g$pct_black, names.arg = g$grade, col = c("#2c7fb8","#7fcdbb","#fec44f","#C41230"),
        ylab = "% Black, 2020", xlab = "1937 HOLC grade", ylim = c(0, 30))
```

Tracts graded **D** in 1937 are **25.6%** Black today. Tracts graded **A** are
**16.7%**.

**That is a real gap and a disappointing one** — a ratio of about 1.5. If you
expected the redlining maps to leap out of the census, this looks like weak
evidence. Plenty of published claims stop here, in one direction or the other.

**They should not.** The number is wrong, and it is wrong for a reason this
course has met before.

# Steps 6-7 — Compare within cities, and every city at once

> **Before running this, work out what was wrong with the last comparison.**
> It put A-graded areas in one city beside D-graded areas in another. Cities
> differ enormously in overall Black population for reasons that have nothing to
> do with the HOLC. **What is the comparison that holds the city constant?**

That table adds up cities. Chicago and Detroit and Los Angeles have very
different racial compositions **overall**, for reasons that have nothing to do
with which tracts got which grade. Pooling them compares a D-graded tract in one
city against an A-graded tract in another.

**The comparison has to happen inside a city.**

```{r cities}
head(ci[, c("city", "a_pct_black", "d_pct_black", "gap")], 8)
tail(ci[, c("city", "a_pct_black", "d_pct_black", "gap")], 5)
```

```{r city-plot}
k <- head(ci[order(-abs(ci$gap)), ], 14)
k <- k[order(k$gap), ]
barplot(k$gap, horiz = TRUE, names.arg = k$city, las = 1, cex.names = 0.7,
        col = ifelse(k$gap > 0, "#C41230", "#2c7fb8"),
        xlab = "percentage-point difference, D-graded minus A-graded (% Black, 2020)")
abline(v = 0)
```

Now it is visible, and it is enormous.

- **Cleveland:** A-graded tracts are 21.3% Black. D-graded, **68.1%**.
- **Pittsburgh:** 4.2% against **34.1%** — eight times.
- **Chicago:** 3.2% against **36.6%** — eleven times.

**The national number said 1.5×. Within cities it reaches ten.** Pooling
did not weaken the finding slightly; it nearly erased it.

# Steps 8-9 — Take the reversals seriously

```{r reversed}
ci[ci$gap < 0, c("city", "a_pct_black", "d_pct_black", "gap")]
```

In some cities the relationship runs **backwards**. Philadelphia's A-graded
tracts are **40.7%** Black and its D-graded tracts **34.3%**.

**Do not explain this away.** It is the most interesting row in the lab, and the
explanation is history rather than error: in cities that lost enormous
white population after 1950, the neighbourhoods graded "Best" in 1937 — leafy,
desirable, close in — are exactly the ones that later became majority Black. The
1937 grade stopped predicting because the city around it changed completely.

**A finding that holds in most places and reverses in some is not a broken
finding.** It is a finding with a mechanism, and the reversals are where you can
see the mechanism working.

# Step 10 — What you can now say

**You can say** that within cities, neighbourhoods the federal government
graded "Hazardous" in the 1930s are on average **14.5 points more Black today**
than neighbourhoods it graded "Best" — visible eighty years and several complete
rebuildings of American cities later.

**You can say the pooled national comparison is misleading**, and that the
correct unit of comparison is the city.

**You cannot say the grades caused it.** You cannot say the relationship is
uniform — six cities reverse, and Philadelphia reverses because North
Philadelphia gentrified. And you cannot generalise beyond the cities in this
file.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Weight by population.** The city table treats each city's A-graded
and D-graded tracts as blocks. Recompute the national figure as the *average of
the within-city gaps* instead of the pooled ratio from Step 4. How different is
it, and which number should a newspaper print?

**Option B — B and C.** Parts 3 and 4 use only A and D — the extremes. Bring in
the middle grades and see whether the gradient is smooth. If B and C sit neatly
between, that is evidence the grade itself mattered; if they do not, say what
else might be going on.

**Option C — Attack the join.** Every result here depends on centroid-in-polygon.
Name two ways that choice could produce the Step 6 result even if HOLC grades
had no lasting effect at all. Then say what data you would need to rule them out.

**Option D — Your own question.** 11,274 tracts, 215 cities, one 1937 map.

# What this lab cannot tell you

- **That redlining caused any of this.** The maps recorded segregation that
  already existed as well as entrenching it. Separating the two needs a research
  design this lab does not have — and it is the central difficulty of the whole
  literature.
- **Anything about people.** Tract composition is not any individual's
  experience, and inferring one from the other is the ecological fallacy the
  RPV lab is about.
- **The full country.** Census data here covers six states, chosen for their
  HOLC cities. The crosswalk covers 215 cities; the join covers fewer.
- **What "A" meant.** The grades bundled housing age, construction quality,
  income and race together. The written appraisals separate them; the letter
  does not.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- The pooled number said 1.5×; within Chicago it is eleven. What does that do to
  your trust in national averages?
- In Philadelphia the pattern reverses. Is that evidence against the redlining
  story, or evidence for a more specific version of it?
- If you did one of the other two labs in this unit, which inference did you
  find more convincing, and why?

***

### Where the data came from

**Mapping Inequality**, Digital Scholarship Lab, University of Richmond — the
digitised HOLC residential security maps, 10,154 areas of which 9,324 carry a
grade, published CC BY-NC-SA.
<https://dsl.richmond.edu/panorama/redlining/>
The polygons drawn above, the four grade colours, and the appraisal text quoted
in the brief are all theirs, retrieved **2026-08-10** from
`static/mappinginequality.json`. The maps and the appraisal forms underneath are
1930s US federal government records. `data/build-brief-figures.R` is the script
that fetched them and measured them against the tract boundaries.

**2020 Census PL 94-171**, tract level, for six states (PA, IL, MI, OH, MO, CA).

**The crosswalk between them is ours.** A published tract-level redlining score
exists — Meier & Mitchell, openICPSR 141121 — but it is behind a login. So
`data/build-data.py` builds the join directly: tract centre points from the
Census Gazetteer, tested against each HOLC polygon by ray casting, in about
fifteen lines. You can read exactly what the join assumes, which is the point.
