---
title: "Key: Lines Drawn in 1937"
subtitle: "84-355 Democracy's Data | Redlining | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", dpi = 96, fig.retina = 1)
options(scipen = 999)
tr <- read.csv("data/tracts.csv", stringsAsFactors = FALSE,
               colClasses = c(GEOID = "character"))
ci <- read.csv("data/cities.csv", stringsAsFactors = FALSE)
g  <- aggregate(cbind(total, black) ~ grade, data = tr, sum)
g$pct <- 100 * g$black / g$total
# the HOLC geometry the brief and the lab now draw, so the numbers quoted below
# are read from the same built files rather than typed
MX <- read.csv("data/fig_mix.csv", stringsAsFactors = FALSE,
               colClasses = c(GEOID = "character"))
HM <- read.csv("data/fig_holc_meta.csv", stringsAsFactors = FALSE)
asg <- MX[MX$centre_grade != "", ]
mis <- asg[asg$centre_grade != asg$modal_grade, ]
drp <- MX[MX$centre_grade == "" & MX$graded > 0, ]
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`redlining-brief.Rmd`) is now a
chapter, "Hazardous": it is organised by argument under prose headings and
carries four numbered D3 figures rather than numbered steps — Figure 1, Cleveland
as the HOLC graded it in 1939; Figure 2, a window of that city with 2020 tracts
laid over the 1939 areas; Figure 3, how much of each tract is really one grade;
Figure 4, the within-city A-to-D gap for every city. The lab (`redlining.Rmd`) is
unchanged and still runs Steps 1–10 with the code exposed. **Students should read
the brief first.** **Step numbers below refer to the lab.** The chapter covers the
same ground in the same order, so each block of lab Steps has a heading facing it:
Steps 1–3 = "Best, Still Desirable, Definitely Declining, Hazardous" and "Two
geographies that were never meant to meet" (Figures 1 and 2), with "The cost of
the rule, measured" (Figure 3); Steps 4–5 = "The answer that arrives first is the
wrong one"; Steps 6–7 = "Holding the city fixed" (Figure 4); Steps 8–9 = "The
cities that run the other way" and "Did the maps do anything, or only record what
was already there?"; Step 10 = "What the ground can testify to".

**The maps are now in the documents.** The chapter opens on the appraisal sheet
for area D9 quoted verbatim — the *Clarifying Remarks*, the *Infiltration of*
line, and the appraiser's own count of 0% Black residents — and Figure 1 then
draws the real 1939 HOLC map of Cleveland in the HOLC's own four colours.
Figure 2 lays the 2020 tract boundaries over the same polygons and shows the rule
failing on a named tract. The lab draws the Cleveland map too. All of it comes
from Mapping Inequality's digitised polygons via `data/build-brief-figures.R`.
**You no longer have to pull the maps up on a projector — but do it anyway if the
room is engaged, and read a second area description aloud.**


**One of three labs in the discrimination unit**, alongside policing and jury
selection. Run in parallel, then compare.

**This one is different from the other two in a way worth naming out loud.**
There, the discriminatory intent is unrecorded and has to be inferred. Here the
intent is written on the document — HOLC appraisers named Black residents as a
reason for a low grade. What has to be shown is not intent but **persistence**,
and that turns out to be the harder problem.

Every number below was run against the data files.

***

# The session in one sentence

The comparison that gets published says the 1937 maps barely matter; doing it
one city at a time says they matter enormously; and the cities where it reverses
explain why both are true.

***

# The findings

```{r headline}
round(setNames(g$pct, g$grade), 1)
c(pooled_gap_D_minus_A = round(g$pct[4] - g$pct[1], 1),
  mean_within_city_gap = round(mean(ci$gap), 1),
  cities = nrow(ci),
  cities_reversed = sum(ci$gap < 0))
```

- Pooled: D-graded tracts **25.6%** Black, A-graded **16.7%** — a gap of 8.9
  points, ratio 1.5.
- **Mean within-city gap: 14.5 points.** Median 11.4.
- Cleveland **+46.9**, Chicago **+33.4**, Pittsburgh **+29.9**.
- **6 of 33 cities run backwards**, Philadelphia among them.

***

# The three things to get said

**1. Step 4 is designed to disappoint, and students must feel that.** A ratio of
1.5 looks like weak evidence for a famous claim, and some will be ready to
conclude the redlining story is overstated. **Let them sit with it for a
minute.** The reveal in Step 6 only works if Step 4 was believed.

**2. The pooling problem is the transferable lesson.** Adding cities together
compares a D tract in Chicago against an A tract in Los Angeles — cities with
different racial compositions for reasons unrelated to HOLC grades. **This is
the aggregation problem the course has met three times** (EAVS jurisdictions,
precinct units, ecological inference), arriving with the largest consequence:
the pooled estimate is barely half the within-city one.

**3. The reversals are evidence, not noise.** Six cities run backwards, and the
explanation is mid-century white flight — the "Best" 1937 neighbourhoods in
Philadelphia and Detroit are exactly the ones that later became majority Black.
**A student who wants to drop those cities is doing the wrong thing**, and
saying so is the most useful thirty seconds in the session.

**The line for the board:** an average over places is not a finding about any
place.

***

# Handling the room

Less volatile than the other two labs — the intent is documented, so there is
nothing to argue about on that front. Two risks instead:

- **Overclaiming causation.** The maps recorded existing segregation as well as
  entrenching it. The lab says so; enforce it. "Redlining caused this" is not
  supported and is not necessary.
- **Fatalism.** Ninety-year-old lines still visible can read as immovable. The
  reversals are the antidote — in some cities the pattern did not persist, and
  the reason was that everything else changed.

***

***

# Walking the steps

| Steps | What they do | Min | Watch for |
|---|---|---|---|
| **1-3** | Two sources 80 years apart; the join that has to be built; one record | 10 | **Point-in-polygon is an assumption.** Say so before any result. The close-up in the brief, Figure 2, shows it failing; use it. |
| **4-5** | The pooled comparison, and what is wrong with it | 10 | 16.7% vs 25.6% is deliberately underwhelming. Let it disappoint. |
| **6-7** | Within cities; then every city at once | 12 | **The pivot.** Cleveland: 21.3 vs 68.1 in one housing market. |
| **8-9** | The six reversals, and what they mean for the claim | 10 | Philadelphia is gentrification, not absence of effect. Small-denominator cities are noise. |
| **10** | What you can say — and what needs Aaronson et al. | 6 | Association within cities, yes. Causation, not from this design. |

**Steps 4-6 are the pivot.** The pooled national comparison must be believed for a minute before it is broken. The lesson is the unit of comparison, not the number.

**If you are short on time**, protect the pivot and the closing step. The numbers are what make the argument concrete; the sequence is the argument.


# Option A — Weight by population

## Verified output

```{r a}
c(pooled_gap        = round(g$pct[4] - g$pct[1], 1),
  mean_of_city_gaps = round(mean(ci$gap), 1),
  median_city_gap   = round(median(ci$gap), 1))
```

**Pooling gives 8.9 points. Averaging within-city gaps gives 14.5** — 63%
larger. The median city is 11.4.

**3/4:** Computes both and reports that the within-city figure is bigger.

**4/4:** Answers the second half — which should a newspaper print — and defends
it. The strongest case is for the within-city average, because the question
"does a 1937 grade predict who lives there now" is a question *about
neighbourhoods within a housing market*, and the pooled number answers a
different question nobody asked.

**4/4, best:** Notices that the mean of city gaps weights Joliet equally with
Chicago, which is its own distortion, and proposes a population-weighted mean of
within-city gaps as the estimator that answers the question without either
defect. **Recognising that both published numbers are wrong in different
directions is the strongest available answer.**

***

# Option B — B and C

## Verified output

```{r b}
full <- names(which(tapply(tr$grade, tr$city, function(x) length(unique(x))) == 4))
s <- tr[tr$city %in% full, ]
gg <- aggregate(cbind(total, black) ~ grade + city, data = s, sum)
gg$pct <- 100 * gg$black / gg$total
c(cities_with_all_four = length(full))
round(tapply(gg$pct, gg$grade, mean), 1)
```

Within the 32 cities that have all four grades, the mean tract composition rises
monotonically: **A 12.7, B 20.1, C 23.5, D 27.6.**

**3/4:** Adds B and C and observes they fall between A and D.

**4/4:** Says why a smooth gradient matters. If only the extremes differed, you
would suspect the A and D areas were simply different kinds of places —
downtown against suburb. **A monotone gradient across four grades is much harder
to produce by accident**, and it is the strongest single piece of evidence in
the lab.

**4/4, best:** Notices the steps are uneven — the A-to-B jump is 7.4 points and
C-to-D is 4.1 — and asks whether the grades were equally spaced in whatever they
were measuring. They were not: HOLC's categories were qualitative, and treating
A/B/C/D as an interval scale is an assumption nobody has justified. **Same
lesson as DW-NOMINATE: the number is a model.**

***

# Option C — Attack the join

No verified output; grade the reasoning.

**3/4:** Names centroid-in-polygon as an approximation and observes tracts get
misassigned.

**4/4:** Produces two mechanisms that could generate the Step 6 result without any lasting
HOLC effect. The best two: **(i)** D-graded areas were downtown, and downtown
tracts are small, so a centroid rule assigns them differently from sprawling
suburban A tracts — the join could be picking up geography, not grade; **(ii)**
tract boundaries are themselves drawn partly along the same physical features
(highways, rivers, rail) that the graders used, so the two are correlated by
construction.

**4/4, best:** Says what would settle it — area-weighted allocation rather than
centroids, so a tract split between grades contributes proportionally — and
notes that this is precisely the published crosswalk we could not download. **A
student who arrives at "the fix exists and is behind a login" has understood
something true about the field.**

**The measured version, if you need it.** The brief reports it under "The cost of
the rule, measured", the section Figure 3 sits in:
`r nrow(mis)` of `r nrow(asg)` tracts in Cleveland and Philadelphia carry a
grade that is not the grade covering most of them, `r nrow(drp)` overlapping
tracts are dropped outright, and the median Cleveland tract has
`r round(median(MX$own_share[MX$city=="Cleveland" & MX$centre_grade!=""]))`% of
its graded land under the grade it was given. **A student who says "the rule is
probably fine" can be shown a number.** What none of it settles is the direction
of the resulting bias, and saying so is worth a mark.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is explaining *why* pooling destroyed the finding,
  not just noticing that it did.
- **A 2** is "redlining still affects neighbourhoods today," which is the
  conclusion they arrived with.
- **Also a 2**: treating the reversals as errors or dropping them.
- **Reward anyone who distinguishes persistence from causation** without
  prompting.

## Anticipated questions

- *"Did redlining cause segregation?"* — Partly, and it also recorded
  segregation that already existed. The honest answer is that this data cannot
  separate the two, and that the best research on it uses boundary
  discontinuities rather than the comparison we made.
- *"Why only six states?"* — Because we downloaded tract-level census data for
  six. The crosswalk itself covers 215 cities; extending it is a download, not a
  method change.
- *"Isn't this just income?"* — Very likely entangled with it. We have race and
  not income; a student who wants income should be told the ACS has it and the
  join already exists.
- *"Are the maps online?"* — Yes, Mapping Inequality, with the original
  appraisal text, and the brief now draws Cleveland's from those very polygons.
  **Read the D9 sheet the chapter opens with out loud.** The written justification is more
  persuasive than any table here, and it takes ninety seconds. Note what it
  says about a neighbourhood that was, on the appraiser's own form, 0% Black:
  the grade is about who might arrive.
- *"How wrong is the centroid rule?"* — Measurably. In the two cities the brief
  draws, `r nrow(mis)` of `r nrow(asg)` tracts get a grade that is not the grade
  covering most of their land, and `r nrow(drp)` more tracts overlap a graded
  area and are dropped because their centre point missed. **That is Option C's
  answer arriving as a picture rather than as a worry**, and a student who has
  seen Figures 2 and 3 should be held to a higher standard on it.

## Connections

- **The other two labs in this unit** — jury selection (a control group in the
  file) and policing (conditioning to escape a missing denominator).
  **Ask which inference the room would most want to defend, and why.**
- **Session 2 (census)** — the tract data, and the race categories.
- **Session 14 (precincts)** — the identical allocation problem: units drawn by
  different people for different purposes, joined by an assumption.
- **EAVS** — the first time pooling across incomparable units broke an average.
