---
title: "The Dataset That Was Never Released"
subtitle: "Fifty-five nominations, four tables, and a column that exists in none of them"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("nomination-anchors-code.R")
```

```{r setup, include=FALSE}
```

Six political scientists spent a decade studying how candidates get nominated for the House of Representatives. They did 346 interviews across `r NR` contests, and in 2026 they published what they found.

They did not publish a data file. There is no download link, no repository, no spreadsheet. **The evidence is printed as twenty-three tables scattered across eleven chapters of a book.**

This chapter is about what it takes to get that evidence back into a form you can work with, and what you find on the way. The answer to the second part is two mistakes in the book and one finding the authors never printed, because none of the three is visible until the tables sit side by side.

## A source with no file

The book is *Parties on the Ground* (2026), by Kathleen Bawn, John Zaller and four colleagues. Its subject is the **nomination** — the contest that decides whose name appears on the November ballot for each party.

That contest matters because most House seats are not really contested in November. The [`house-competition`](../house-competition/house-competition-brief.html) chapter of this book finds that about one race in seven has a single name on the ballot. Where a district reliably votes one way, the primary is the election that decides who goes to Congress.

The authors studied `r NR` of them: the nominations that were both **open**, with no sitting member running, and **winnable**, meaning the winner had a real chance in November. Those are the contests that send new people to Congress.

```{r scope}
```

Four of those twenty-three tables describe the same `r NR` contests and label each one the same way, with a code like `PA-13D` — Pennsylvania's thirteenth district, Democratic side. Because they share that code, they can be joined.

```{r sources}
```

Read the third row again. **Tables 6.2 and 6.3 give totals and no rows.** They report that 28 winners were carried by coordinated groups, 9 by money, 6 by volunteers and 4 by professional campaign help. They never say which contest was which.

That column is the one this chapter needs, and it is in no table.

## Recovering a column from prose

The column can be rebuilt, and the way it is rebuilt is the method worth learning. It comes out of sentences rather than cells.

**The first step is an annotation almost small enough to miss.** Table 5.2 lists 32 contests with coordinated support. Beside three of them it prints `(1st and 2nd)` or `(2nd)`, marking races where the group's candidate came second.

Expand those annotations and the 32 contests become `r ck("candidates with coordinated support")` candidates, of whom `r ck("of those, who won the nomination")` won. Chapter 6 says the same thing in words, and the two agree. That is the first sign the reading is right.

**The second step is a sentence that moves two names.** Chapter 6 takes two winners out of the coordination column and puts them in the money column, naming both and saying why: the authors "judged the money to have had a bigger impact." Thirty minus two is twenty-eight, which is exactly what the totals table prints.

**The third step is a set of lists.** Chapter 6 names the other seven candidates backed mainly by money, the four backed by professional campaign help, and the volunteer-driven campaigns, all in ordinary paragraphs.

**Then whatever is left over is the answer.** `r NRES` contests match none of those lists, and the totals table says `r NRES` winners had no backing the authors could identify.

```{r rebuild}
```

**Four counts, arrived at four different ways, land on four numbers the book already printed.** None of them was adjusted to fit. That agreement is the whole warrant for the column, and the derivation stops if any of the four ever disagrees.

It is still a derived column, and every row says so. `r NRES` of the `r NR` carry the note *left over: stated in no table and no sentence*, because for those the answer is an absence rather than a statement.

## What the join found

Joining tables does something that reading them cannot. It puts two claims about the same thing next to each other, where they can contradict.

**Two district codes in the book are wrong, and both were found this way.**

```{r errata}
```

Neither mistake is visible in the table it appears in. A district code looks fine on its own. It only goes wrong beside another district code, when the same person turns up in two places or a race turns up with two winners.

The check that caught them is one line: **every race code in a smaller table must match a row in the big one.** Two did not.

**And two tables disagree with each other about `r NDIS` contests.**

```{r disagree}
```

One table names a backer. The other counts the same candidate among those with no backer the authors could identify. Both are the same six authors in the same book.

The explanation is a threshold. To count as a real backer, a group had to supply at least $250,000, or $400,000 in races with a runoff. These three candidates had a clear group behind them that gave less than that. Chapter 8 says so plainly: had one group "spent as much on Mimi Walters as Club for Growth regularly spent on its candidates," the race would have been counted the other way.

**The disagreement is not an error. It is a cutoff becoming visible.** Both readings are kept here, side by side, rather than one being picked.

## What the tables say together

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=3.6}
```

**Figure 1. Why each group got involved, against what it actually supplied.** Rows are the reason the authors judged a group to have backed its candidate. Columns are the resource that backing consisted of. Cells are counts of nominations, and blank cells are zero.

Most of the grid is empty, and the empty cells are the point. Every contest where the goal was simply winning in November was carried by coordinated groups, never by money or volunteers alone. Every volunteer-driven campaign sits in one row, the one for groups motivated by values rather than material interest.

Read down the tallest column instead and you get the book's headline. `r sum(NM$anchor_form == "coordination")` of `r NR` winners were carried by groups that worked together, which is more than all other kinds of backing combined.

## Who gets to be the outsider

The book divides Republican candidates into two camps. **Establishment** candidates were backed by business or by party organizations. **Insurgents** ran against the party leadership, usually with support from national conservative groups, churches or local Tea Party chapters.

The authors report, in one chapter, how often coordination appeared in a district. They do not report how often it appeared behind a candidate of each camp, because faction and coordination are printed in different chapters.

> **Before you read on, write a number down.** Of the
> `r fac(INS, "nominees")` Republican insurgents who won a nomination, **how many
> won a contest where groups had coordinated behind somebody?** Any coordination
> at all, for any candidate, including against them. Commit to a count out of
> `r fac(INS, "nominees")` before you look at Figure 2.

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=2.9}
```

**Figure 2. Coordination behind each kind of Republican nominee.** Each bar is all the winners of that type. The dark part is those who won a contest where groups had coordinated behind some candidate.

Most readers guess four or five insurgents, because coordination sounds like something that happens in most competitive races. It is **`r ck("insurgents in a coordinated race")`**.

At the other end, **`r ck("party-backed establishment nominees in a coordinated race")`** party-backed establishment nominees won a contest with coordination in it. Every one.

**Those two rows are the same fact seen from both ends.** Where groups organised behind a candidate, an insurgent almost never emerged as the winner. Where no group organised, insurgents won regularly.

The money runs the same way, and it is the second thing the join shows.

```{r pacs}
```

A Republican carried by volunteers took a median of **`r ck("median business PAC money, grassroots-anchored Republicans")`** from business PACs. A Republican carried by coordinated groups took **`r ck("median business PAC money, coordination-anchored Republicans")`**. The book argues that volunteer campaigns are a real alternative to money. Here that argument is a number.

## What this cannot tell you

**It cannot tell you that coordination defeats insurgents.** Figure 2 shows two things happening in the same contests. It cannot show which caused which, and the authors say so directly: groups may organise behind candidates who were going to win anyway.

**Worse, the two columns in Figure 2 were not decided independently.** The same researchers judged whether coordination happened and whether a candidate was an insurgent, in the same pass, from the same interviews. They write that they "acknowledge the potential for bias in this situation." Figure 2 is the picture that admission is about.

**Every classification here is somebody's judgment.** Insurgent or establishment, backed or not backed, and the reason a group cared — none of these is a fact you could look up. They are readings of 346 interviews by the six people who did them. The dollar amounts and vote counts are facts. The labels are not.

**The thresholds are arbitrary, and the authors call them that.** A group had to give $250,000 to count as a backer, or $400,000 where a runoff was possible. One candidate missed by $20,000 and is filed among those with no backer. Move the line and Figure 1 changes shape.

**`r NDIS` of the `r NR` rows carry two answers**, because two tables in one book disagree. Nothing here resolves that.

**And these are open seats only.** All `r NR` had no sitting member running, in a single election cycle, in districts where one party was expected to win. Nothing here describes a race against an incumbent, which is most races.

## What the reassembly was worth

The joined table holds `r NR` rows and nineteen columns. Every column came from the book. Not one of them came from a file.

What the joining bought was three things the book does not contain: a column that appears in no table, `r NERR` corrected district codes, and a comparison between two chapters that was never drawn.

None of that required new evidence. It required putting the existing evidence in one place, where the pieces could be checked against each other.

That is worth remembering the next time a source arrives as a document rather than a download. **A book is a data format.** It is a bad one — the machine cannot read it, the columns are scattered, and nothing warns you when two pages disagree. But the data is in there, and getting it out is ordinary work.

The harder question is the one this chapter cannot settle. The authors could have released a file. They chose to publish tables. Should a reader have to do this work to check a finding?

## Sources

**Data**

- Bawn, K., Brown, K., Ocampo, A. X., Patterson, S. Jr., Ray, J. L. and Zaller,
  J. (2026). *Parties on the Ground: A Study of Nominations for the House of
  Representatives.* University of Chicago Press.
  <https://doi.org/10.7208/chicago/9780226853314> — tables 5.2, 6.2, 6.3, 9.1
  and 10.1, covering `r NR` winnable open-seat House nominations in the 2013–14
  cycle. **Every figure, classification and dollar amount above is theirs**,
  transcribed from the printed pages and cited to the table it came from. The
  authors state that the study is exploratory and that none of its conclusions
  has been tested as a hypothesis.
- The book announces an online appendix at <https://osf.io/s5982>. That project
  was not publicly readable when this was written; the address returns a sign-in
  page and the public catalogue reports no such record. Everything here
  therefore comes from the printed tables.

Facts are not owned, and these are facts: race codes, names, dollar amounts, counts. The judgments attached to them are the authors' scholarly work, are used here to describe and check their own book, and are attributed throughout.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_anchor.csv`](data/derived/by_anchor.csv), [`by_faction.csv`](data/derived/by_faction.csv), [`nominations.csv`](data/derived/nominations.csv), [`pac_by_anchor.csv`](data/derived/pac_by_anchor.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `nominations.csv` has an `anchor_form` for each nomination and an `anchor_form_source` saying where that came from. Count how many were recovered from prose rather than a table. How much of this dataset had to be rebuilt by hand?
- `by_faction.csv` reports `pct_coordinated` by faction. Which faction is most coordinated, and does that match how those candidates describe themselves?
- `pac_by_anchor.csv` gives median and mean PAC money by anchor form. The two are far apart. Which would you report, and what is the gap telling you?
- `by_anchor.csv` crosses motive against the form of support. Find the combinations with no cases at all. Is that a fact about politics, or about how the categories were drawn?

**On why the nomination matters**

- Hall, A. B. (2019). *Who Wants to Run? How the Devaluing of Political Office
  Drives Polarization.* University of Chicago Press.
- Boatright, R. G. (2014). *Congressional Primary Elections.* Routledge.
- Ocampo, A. X. (2018). "The Politics of Inclusion: A Sense of Belonging to U.S.
  Society and Latino Political Participation." Cited by the authors for the
  finding that who is vetted decides what representation a district gets.

**Related chapters in this book**

[`primary-defeats`](../primary-defeats/primary-defeats-brief.html) counts how often sitting members lose a primary. [`primary-positions`](../primary-positions/primary-positions-brief.html) reads what candidates in the same 2014 cycle would say about fourteen issues on their own websites. [`house-competition`](../house-competition/house-competition-brief.html) is the source of the claim that about one House race in seven has a single name on the ballot.

**How the figures were made**

The five transcribed tables are kept beside this chapter exactly as they were read off the page, because there is no file to re-fetch and a transcription that cannot be rechecked is worth little. Every number printed above is computed from them at the moment this document was rendered. The derived column is rebuilt each time and compared against the four totals the book printed; if any of the four disagrees, nothing is written. The validation results are reproduced in `checks.csv`.
