---
title: "Data Session 14 Key: Inferring the Secret Ballot"
subtitle: "84-355 Democracy's Data | Session 14 — Precinct returns, maps, and racial bloc voting | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
p <- read.csv("data/houston_primary.csv", stringsAsFactors = FALSE)
p$pct_black <- p$black / p$voters
p$pct_dem   <- p$dem   / p$voters
fit <- lm(pct_dem ~ pct_black, data = p, weights = voters)
truth_black <- 100 * sum(p$black_dem) / sum(p$black)
truth_white <- 100 * sum(p$white_dem) / sum(p$white)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
cf <- coef(fit)
g_black <- 100 * (cf[[1]] + cf[[2]]); g_white <- 100 * cf[[1]]
r2 <- summary(fit)$r.squared
bl <- p$pct_black; dm <- p$pct_dem
lo <- pmax(0, (dm - (1 - bl)) / bl); hi <- pmin(1, dm / bl)
bnd_lo <- 100 * max(lo); bnd_hi <- 100 * min(hi)
max_blk <- 100 * max(p$pct_black)
n80 <- sum(p$pct_black >= 0.80)
thr <- sapply(c(.80, .70, .60, .50), function(t) {
  h <- p[p$pct_black >= t, ]
  if (nrow(h)) 100 * sum(h$dem) / sum(h$voters) else NA_real_
})
names(thr) <- c("t80", "t70", "t60", "t50")
p2 <- p[p$precinct != "WELLSTON CENTER", ]
f2 <- lm(pct_dem ~ pct_black, data = p2, weights = voters); c2 <- coef(f2)
g2_black <- 100 * (c2[[1]] + c2[[2]]); r2_2 <- summary(f2)$r.squared
max_blk2 <- 100 * max(p2$pct_black)
p$true_blk <- 100 * p$black_dem / p$black
tb_lo <- min(p$true_blk); tb_hi <- max(p$true_blk)
tb_lo_nm <- p$precinct[which.min(p$true_blk)]; tb_hi_nm <- p$precinct[which.max(p$true_blk)]
fit_err <- abs(100 * fitted(fit) - 100 * p$pct_dem)
fu <- lm(pct_dem ~ pct_black, data = p); cu <- coef(fu)
u_white <- 100 * cu[[1]]; u_black <- 100 * (cu[[1]] + cu[[2]])
```

All code is given to students. Grade the interpretation.

**Framing, Dec 2026.** This is the last lab of the term and the payoff of the
arc that runs from Session 2 to here. Session 2 asked what the census can
measure; Session 3 asked who gets counted as what; Session 7 graded a guess
about *who somebody is*; today
grades a guess about *what they did*. Same county, same registration file, same
move — find the rare case where the truth is recoverable and use it to audit a
method that is normally unfalsifiable.

**It is also the lab that most directly touches a live case.** Section 2
liability turns on the numbers students are about to watch fail.

**And the law under it moved on 29 April 2026.** *Louisiana v. Callais*, 608
U.S. ___ (2026), left *Gingles* standing but rewrote preconditions two and
three: a plaintiff must now "provide an analysis that controls for party
affiliation" and show racial bloc voting "that cannot be explained by partisan
affiliation," and "simply pointing to inter-party racial polarization proves
nothing." Neither method in this lab does that, and the data cannot — the
outcome variable is which party's ballot a voter requested. A marked note has
been added to both the lab and the brief; the grading bullets below have been
updated to match. **The pedagogical payoff is better than before, not worse:**
this county is the cleanest possible illustration of why the requirement is hard
to meet in the states where Section 2 cases are actually filed.

Every number below was run against the data file.

***

# The lab in one sentence

Ecological regression fits the data almost perfectly, reports that
`r pc(g_black, 0)`% of Black voters took a Democratic ballot, and is wrong by
`r pc(g_black - truth_black)` points — and the only reason anybody could tell is
that Georgia happens to record the answer.

***

# The findings

```{r headline}
b <- coef(fit)
bl <- p$pct_black; dm <- p$pct_dem
lo <- pmax(0, (dm - (1 - bl)) / bl); hi <- pmin(1, dm / bl)
c(goodman_black = round(100 * (b[1] + b[2]), 1),
  goodman_white = round(100 * b[1], 1),
  bound_lo      = round(100 * max(lo), 1),
  bound_hi      = round(100 * min(hi), 1),
  truth_black   = round(truth_black, 1),
  truth_white   = round(100 * sum(p$white_dem) / sum(p$white), 1),
  r_squared     = round(summary(fit)$r.squared, 3))
```

- **Truth: `r pc(truth_black)`% of Black voters and `r pc(truth_white)`% of white
  voters** requested a Democratic ballot. Near-total polarization.
- **Goodman's regression says `r pc(g_black)`%** — impossible — with an
  **R² of `r pc(r2, 3)`**.
- **The bounds are [`r pc(bnd_lo)`%, `r pc(bnd_hi)`%]**, correct by construction
  and `r pc(bnd_hi - bnd_lo)` points wide.
- **Homogeneous precinct analysis cannot run at all**:
  `r if (n80 == 0) "zero" else n80` precincts are ≥80% Black. The most Black
  precinct in the county is `r pc(max_blk)`%.
- Every method reached the right conclusion about *polarization*. Every method
  got the *numbers* wrong. And after *Callais*, none of them reaches the legal
  conclusion at all — see the framing note above.

***

# The idea to teach

**Not "the model was bad."** The model was excellent by every diagnostic
available to the person running it. That is the point, and it is the hardest one
to land.

Three things to get said out loud:

1. **Fitting the aggregate is not recovering the individual.** The regression
   predicts each precinct's Democratic share to within `r pc(max(fit_err))`
   points at worst, `r pc(mean(fit_err))` on average — look at `fit_dem` against
   `act_dem` in Option C — and still misstates the group rate by
   `r pc(g_black - truth_black)`. Predicting the total and decomposing the total
   are different problems. **This is the ecological fallacy, shown rather than
   asserted.**
2. **The estimate came from where there is no data.** The Black estimate is the
   line extended `r pc(100 - max_blk)` points beyond the most Black precinct in
   the county. Regression will always hand you that number and will never flag it.
3. **The impossibility was a gift.** `r pc(g_black, 0)`% is self-refuting, so somebody would
   catch it. Had the truth been 60% and Goodman returned 51%, the report would
   have been wrong, plausible, and unchallengeable. **Ask the room how they
   would have caught that.** The honest answer is: they would not have.

**The line worth writing on the board:** the method was most confident exactly
where it had the least evidence.

***

# Option A — Move the homogeneity threshold

## Verified output

```{r a}
for (t in c(.80, .70, .60, .50)) {
  h <- p[p$pct_black >= t, ]
  cat(sprintf("  >=%2.0f%%  n=%d  Dem share %s\n", 100 * t, nrow(h),
      if (nrow(h)) sprintf("%.1f%%  (error %+.1f)",
                           100 * sum(h$dem) / sum(h$voters),
                           100 * sum(h$dem) / sum(h$voters) - truth_black)
      else "-- cannot compute"))
}
```

## What a good answer says

**The intuitive fix makes it worse, and students should be made to sit with
that.** Loosening the threshold to get an estimate produces
**`r pc(thr[["t70"]])`% at ≥70% (`r pc(truth_black - thr[["t70"]])` points low)**
and **`r pc(thr[["t50"]])`% at ≥50% (`r pc(truth_black - thr[["t50"]])` points
low)**. Every relaxation admits more white voters, who broke
`r pc(truth_white)`% Democratic, and drags the estimate down.

**3/4:** Runs the thresholds, reports that the estimate gets worse as the
threshold loosens, and identifies contamination as the cause.

**4/4:** Notices that ≥70% and ≥60% give the *identical* answer, because both
capture exactly one precinct — WELLSTON CENTER. The "method" at that point is
one polling place, and calling it homogeneous-precinct analysis dresses up a
sample size of one.

**4/4, best:** Turns it into the cross-examination question. There is no
principled threshold; 80% is convention. **A witness who reports the 70%
threshold and not the 80% one has not lied about anything** — the 80% simply
returns nothing. Students who see that the *choice of threshold is where the
discretion lives*, and that the discretion is invisible in the final number,
have got the lesson.

***

# Option B — Drop the biggest precinct

## Verified output

```{r b}
p2 <- p[p$precinct != "WELLSTON CENTER", ]
f2 <- lm(pct_dem ~ pct_black, data = p2, weights = voters); b2 <- coef(f2)
rbind(
  `all 17`      = c(black = 100*(b[1]+b[2]),   white = 100*b[1],
                    R2 = summary(fit)$r.squared, max_black = 100*max(p$pct_black)),
  `drop 1`      = c(black = 100*(b2[1]+b2[2]), white = 100*b2[1],
                    R2 = summary(f2)$r.squared,  max_black = 100*max(p2$pct_black))
) |> round(1)
```

## What a good answer says

Removing one precinct out of `r nrow(p)` moves the Black estimate **from
`r pc(g_black)`% to `r pc(g2_black)`%** — further into impossibility — and the
R² stays at **`r pc(r2_2, 3)`**.

**3/4:** Reports the shift and notes the fit is still excellent.

**4/4:** Connects it to the extrapolation. Without WELLSTON the most Black
precinct is **`r pc(max_blk2)`%**, so the estimate is now the line extended
*`r pc(100 - max_blk2, 0)`* points past the data. The estimate got worse in
exact proportion to how far it was
extrapolated — which tells you the number was never really an estimate of Black
voters at all, but a property of the line's slope.

**4/4, best:** Asks the fragility question the prompt sets up. **A
fifty-year-old courtroom standard, in this county, rests on one polling place.**
A student who says they would want to know, for any RPV report, how many
precincts anchor the minority end — and that this is never reported — is
thinking like a reviewer.

***

# Option C — Predict, then check

## Verified output

```{r c}
p$true_blk <- 100 * p$black_dem / p$black
p$fit_dem  <- 100 * fitted(fit)
p$act_dem  <- 100 * p$pct_dem
o <- p[order(p$pct_black), c("precinct","pct_black","act_dem","fit_dem","true_blk")]
print(o, row.names = FALSE, digits = 3)
cat(sprintf("\nTRUE Black Democratic rate ranges %.1f%% to %.1f%% — spread of %.1f points.\n",
            min(p$true_blk), max(p$true_blk), max(p$true_blk) - min(p$true_blk)))
```

## What a good answer says

**This is the best of the five options and worth steering strong students
toward.** It exposes the actual mechanism.

**3/4:** Notices `fit_dem` tracks `act_dem` closely — the model predicts
precinct outcomes well — while the group estimate is wrong.

**4/4:** Reads the `true_blk` column and sees that **Black Democratic support is
not constant across precincts: it runs from `r pc(tb_lo)`% in `r tb_lo_nm` to
`r pc(tb_hi)`% in `r tb_hi_nm`, a spread of `r pc(tb_hi - tb_lo)` points.**
Goodman's regression assumes that rate is the
*same everywhere* — that is precisely what a single intercept and slope encode.
The assumption is false in the data, and we can prove it only because we have
the column.

**4/4, best:** Notices the pattern in the violation. Black voters are *more*
Democratic in precincts with more Black voters. That correlation between group
behaviour and group share is exactly the condition under which ecological
regression is known to break, and it is not detectable from aggregates — it
looks like a slightly steeper line. A student who gets here has independently
rediscovered why King wrote the book.

***

# Option D — Would it transfer?

No output; graded on reasoning.

**3/4:** Lists real differences between a May primary and a November general —
smaller electorate, more partisan, different composition.

**4/4:** Names the specific bridging assumption. To carry today's validation to
November you must assume the *method's error* behaves the same in both, which is
a claim about the geography–behaviour correlation, not about turnout. Nothing we
did tests that.

**4/4, best:** Observes that the validation is available only in states that
record race *and* run partisan primaries, which is a small and unrepresentative
set of mostly Southern states — the same places most likely to be in Section 2
litigation. **The evidence about whether these methods work is drawn from
exactly the counties where they are most often used, and that is either
reassuring or circular depending on how you look at it.** Either answer, argued,
is a 4.

***

# Option E — Own question

Grade on whether the question could be answered with the columns present.
Common good ones: does precinct size predict error; does the Republican side
mirror the Democratic side (it must, with two parties — worth letting a student
discover); what happens with an unweighted regression (white `r pc(u_white)`%,
black `r pc(u_black)`% — the weighting is not the problem).

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is refusing to conclude "the method is broken."
  Every method got the polarization answer right. The correct reading is
  narrower and harder: the methods are reliable for *direction* and unreliable
  for *magnitude*, and until this year Section 2 mostly needed direction — which
  was the strongest available defence of fifty years of practice, and students
  should be able to state it before they attack it.
- **Do not let a student stop there now.** *Callais* took away the second half
  of that defence: direction alone is no longer sufficient, because the
  direction has to be net of party. A student who notices that the lab's
  outcome variable *is* party, and that this makes the county an illustration of
  the problem rather than a solution to it, has done better than the document
  did before the note was added. **A 4 either way** — the pre-*Callais* defence
  stated accurately as history is still a 4.
- **A 2** is "statistics can prove anything" or "the ecological fallacy means
  you can't do this." Both skip the work.
- **Reward anyone who is bothered that we validated on the easy case.**
  Polarization here is near-total; the errors did not matter. That is the right
  thing to be uneasy about.

## Anticipated questions

- *"Why not just use King's method?"* — We could, and it does better here.
  It is still a model, it still cannot be checked in the counties where it is
  actually used, and its assumptions are harder to state than Goodman's rather
  than weaker. Worth being straight about: this lab is not an argument for any
  particular method, it is an argument about what validation is available.
- *"Isn't `r pc(g_black, 0)`% just a rounding problem?"* — No. It is
  `r pc(g_black - truth_black)` points past the truth and
  `r pc(g_black - 100)` points past possibility. Say plainly that a percentage over 100
  means an assumption failed, not that a number was rounded.
- *"If race is on the file, why infer at all?"* — Because race is on the file in
  a handful of states, and how people **voted** is nowhere. Today's shortcut is
  the party-ballot request, which exists only in partisan primaries. In November
  nobody knows anything.
- *"Did we just prove Houston County is polarized?"* — We showed its **May
  primary electorate** was. Do not let this become a finding about the county;
  it is a finding about a method.
- *"Does* Callais *make this lab pointless?"* — The opposite, and say so. The
  statistical content is untouched: ecological inference is still the only route
  from precinct aggregates to group behaviour, and Steps 6–9 still show it
  failing. What changed is that the estimate no longer carries the case by
  itself. If a student asks what a party-controlled analysis would look like
  here, the honest answer is that the Court required one and did not specify it,
  and that in a partisan-primary file the natural fixes all reduce to comparing
  Black and white voters *inside* one party's primary — which is what *Gingles*
  itself rested on, 478 U.S., at 59. We do not run that today.
- *"Is `r nrow(p)` precincts too few?"* — It is on the small side but entirely ordinary
  for a county-level Section 2 case. Do not let students dismiss the result as a
  small-sample artefact: the failure is structural — extrapolation past the data
  and a false constancy assumption — and more precincts of the same shape would
  not fix either.

## Connections

- **Session 7 (surnames and BISG)** — the direct predecessor and the same county. Both labs
  grade a normally-ungradeable inference against Georgia's registration data.
  Say so explicitly; the pairing is the argument.
- **Session 3 (the ACS and §203)** — the statute this evidence is produced for.
- **Session 2 (the decennial census)** — where the denominators in every RPV case come from.
- **Session 8 (CPS turnout)** — the term's first "the measurement is not the thing"
  lab, and the first appearance of a number that looked fine and was not.
- **None of this material is on a test** — all three tests cover the textbook only,
  and Test 3 (same day as the wrap) is Sides et al. Ch. 9–11. Say so early. The
  risk is that students read "not on the test" as "not important" in the four
  weeks the labs were actually built for, so the board post is the whole
  assessment here and worth treating as such.
