---
title: "Lab 15: Inferring the Secret Ballot"
subtitle: "84-355 Democracy's Data | Racially polarized voting, Houston County, Georgia"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(x, big.mark = ",", trim = TRUE)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Before we start

Section 2 of the Voting Rights Act turns on a factual question: **do voters of
different races prefer different candidates?** If they do not, there is no
claim. Everything you read this morning — King, Monmonier — exists because of
one obstacle in the way of answering it.

> ### Note — the law under this changed on 29 April 2026
>
> *Louisiana v. Callais*, 608 U.S. ___ (2026), did not overrule *Gingles*. It
> updated it, and the update lands on the two preconditions this lab is about. A
> plaintiff must now "provide an analysis that controls for party affiliation"
> and show racial bloc voting "that cannot be explained by partisan
> affiliation." Simply pointing to inter-party polarization, the Court said,
> "proves nothing." Five weeks later it put it flatly: "the mere fact that
> voters of different races vote for different parties is not relevant to
> proving racially polarized voting patterns," *Allen v. Milligan*, 608 U.S.
> ___ (2026) (per curiam) (June 2, 2026).
>
> **What we estimate today is which party's ballot each racial group asked
> for** — the inter-party comparison, exactly. The statistical problem is
> unchanged and everything below still runs. What changed is what the answer
> would be worth in court, and that is worth holding in mind from Part 6 on.

**The ballot is secret.** Nobody records how you voted. What gets recorded is
how a *precinct* voted, and separately, who *lives* in that precinct. From those
two aggregates an expert witness must infer how each racial group behaved — and
then a federal judge decides a case on the answer.

That inference is called **ecological inference**, and it is the thing every
statistics course warns you about: reasoning from groups to individuals. Do it
badly and you commit the ecological fallacy.

Today we do it, and then we find out whether it worked.

**Why we can check.** In Georgia, two facts are public that are private almost
everywhere else: the race a voter reports on the registration form, and **which
party's primary ballot they requested**. So for one election — the General
Primary of 21 May 2024 — we know, person by person, the exact quantity
ecological inference is trying to estimate.

This is the same trick as last week. Week 14 graded a guess about *who someone
is*. Today we grade a guess about *what they did*.

For the next four parts, **pretend we do not know the answer.** That is the
honest position of every expert who has ever testified on this.

# Part 1 — What you observe

```{r load}
p <- read.csv("data/houston_primary.csv", stringsAsFactors = FALSE)

p$pct_black <- p$black / p$voters
p$pct_dem   <- p$dem   / p$voters

obs <- p[order(p$pct_black), c("precinct", "voters", "pct_black", "pct_dem")]
print(obs, row.names = FALSE, digits = 3)
```

Seventeen precincts, `r nn(sum(p$voters))` voters. Two columns matter: what
share of each precinct's primary electorate was Black, and what share of its
ballots were Democratic.

**This table is the entire evidentiary record.** An expert witness in a Section
2 case gets essentially this and nothing more.

Notice the range. The least Black precinct
(`r p$precinct[which.min(p$pct_black)]`) is `r pc(100 * min(p$pct_black))`%
Black; the most (`r p$precinct[which.max(p$pct_black)]`) is
`r pc(100 * max(p$pct_black))`%. **There is no precinct in this county that is
overwhelmingly Black.** Hold on to that — it matters more than it looks.

# Part 2 — The eyeball

```{r scatter}
plot(p$pct_black, p$pct_dem, pch = 19, cex = sqrt(p$voters) / 22,
     col = "#2c7fb8", xlim = c(0, 1), ylim = c(0, 1),
     xlab = "Share of the precinct's primary voters who are Black",
     ylab = "Share of ballots that were Democratic")
abline(0, 1, lty = 3, col = "grey60")
grid()
```

The relationship is about as clean as social science ever gets. Precincts with
more Black voters cast more Democratic ballots, and the points sit close to the
45-degree line — which would be exactly what you would see if **every** Black
voter took a Democratic ballot and **no** white voter did.

That is a hypothesis, not a finding. Two different worlds produce this picture:
one where Black and white voters are perfectly polarized, and one where
Democratic-leaning people of every race happen to live near each other. The
scatter cannot tell them apart. **Neither can any method in this lab.**

# Part 3 — Homogeneous precincts

The oldest method in voting-rights litigation, and still the one courts find
most intuitive: **find precincts that are nearly all one race, and just look at
how they voted.** No modeling, no assumptions.

```{r homogeneous}
hb <- p[p$pct_black >= 0.80, ]
hw <- p[p$pct_black <= 0.20, ]

cat("precincts at least 80% Black:", nrow(hb), "\n")
cat("precincts at most  20% Black:", nrow(hw), "\n\n")

if (nrow(hw))
  cat(sprintf("white-homogeneous precincts, Democratic share: %.1f%%\n",
              100 * sum(hw$dem) / sum(hw$voters)))
if (nrow(hb))
  cat(sprintf("Black-homogeneous precincts, Democratic share: %.1f%%\n",
              100 * sum(hb$dem) / sum(hb$voters)))
```

**Read that first number again: there are none.** Not one precinct in Houston
County is 80% Black, so the most trusted method in the field cannot produce an
estimate for Black voters here *at all*. Not an imprecise one — none.

This is not a quirk of our threshold. It is a fact about the county, and it is
the ordinary situation in most of the United States. The method works best
exactly where segregation is most severe.

And the number it *does* give us is worth distrusting. The `r nrow(hw)`
"white-homogeneous" precincts return
**`r pc(100 * sum(hw$dem) / sum(hw$voters))`%** Democratic — but those precincts
are up to 20% Black, and those Black voters cast ballots too. The estimate for
white voters is contaminated by the very group we are trying to hold out.

# Part 4 — Goodman's ecological regression

The standard alternative, and for fifty years the workhorse of Section 2
litigation. Fit a line through the scatter, then read off its ends: where the
line crosses `pct_black = 0` is the estimate for white voters, and where it
crosses `pct_black = 1` is the estimate for Black voters.

```{r goodman}
fit <- lm(pct_dem ~ pct_black, data = p, weights = voters)
b   <- coef(fit)

est_white <- 100 * b[1]
est_black <- 100 * (b[1] + b[2])

cat(sprintf("R-squared: %.3f\n\n", summary(fit)$r.squared))
cat(sprintf("  estimated Democratic support, white voters: %5.1f%%\n", est_white))
cat(sprintf("  estimated Democratic support, Black voters: %5.1f%%\n", est_black))
```

```{r goodman-plot}
plot(p$pct_black, p$pct_dem, pch = 19, cex = sqrt(p$voters) / 22,
     col = "#2c7fb8", xlim = c(0, 1), ylim = c(0, 1.1),
     xlab = "Share of the precinct's primary voters who are Black",
     ylab = "Share of ballots that were Democratic")
abline(fit, col = "#C41230", lwd = 2)
abline(h = 1, lty = 2, col = "grey40")
rect(max(p$pct_black), -1, 1.2, 2, col = adjustcolor("grey70", alpha.f = 0.25),
     border = NA)
text(0.88, 0.45, "no data\nout here", col = "grey30", cex = 0.9)
points(c(0, 1), c(b[1], b[1] + b[2]), pch = 21, bg = "#C41230", cex = 1.6)
```

**The model fits beautifully — R² of `r pc(summary(fit)$r.squared, 3)` — and
returns an impossible answer.** It estimates that **`r pc(est_black)`%** of
Black voters took a Democratic ballot.

There is no gentle reading of this. A percentage above 100 is not a close call
or a rounding problem; it is a signal that the method has been asked to do
something it cannot do. And notice that nothing in the regression *complained*.
The fit statistics are superb. A `r pc(summary(fit)$r.squared, 2)` R² is the
kind of number that gets quoted in a report.

The grey band shows why. **The most Black precinct in the county is
`r pc(100 * max(p$pct_black))`% Black**, so the estimate for Black voters comes
from extending the line `r pc(100 - 100 * max(p$pct_black))` points past the
last observation. The model is not summarising data out there. There is no data
out there.

# Part 5 — What is knowable without a model

Before King, Duncan and Davis pointed out something simple: even with no model
at all, the precinct totals *constrain* the answer. If a precinct is 40% Black
and cast 45% Democratic ballots, then even if every single white voter went
Democratic, Black voters must supply the rest — and that puts a floor and a
ceiling on their behaviour.

```{r bounds}
bl <- p$pct_black
dm <- p$pct_dem

lo <- pmax(0, (dm - (1 - bl)) / bl)   # if every white voter went Democratic
hi <- pmin(1, dm / bl)                # if no white voter did

cat(sprintf("Tightest single precinct: [%.1f%%, %.1f%%]  (%s)\n",
            100 * max(lo), 100 * min(hi), p$precinct[which.min(hi - lo)]))
cat(sprintf("Intersection of all 17:   [%.1f%%, %.1f%%]\n",
            100 * max(lo), 100 * min(hi)))
cat(sprintf("width: %.1f points\n", 100 * (min(hi) - max(lo))))
```

**Black Democratic support is between `r pc(100 * max(lo))`% and
`r pc(100 * min(hi), 0)`%.** That statement requires no assumptions whatsoever —
it is arithmetic on the marginals, and it cannot be wrong.

It is also **`r pc(100 * (min(hi) - max(lo)))` points wide**, which is why
nobody testifies with it alone.

Here is the trade the whole field runs on. The bounds are *guaranteed correct
and too vague to decide a case*. The regression is *precise, decisive, and
impossible*. King's method — the reading for today — is an attempt to get
something in between: use the bounds as a hard constraint, then use a
statistical model to say where inside them the answer probably sits.

# Part 6 — The answer

We know the truth. Every voter in that table requested a party ballot by name,
and reported their race on the registration form.

```{r truth}
truth_black <- 100 * sum(p$black_dem) / sum(p$black)
truth_white <- 100 * sum(p$white_dem) / sum(p$white)

res <- data.frame(
  method = c("Homogeneous precincts", "Goodman regression",
             "Bounds (lower)", "Bounds (upper)", "TRUTH"),
  black  = c(NA, est_black, 100 * max(lo), 100 * min(hi), truth_black),
  white  = c(100 * sum(hw$dem) / sum(hw$voters), est_white, NA, NA, truth_white))

print(res, row.names = FALSE, digits = 3)
```

```{r truth-plot}
barplot(c(Goodman = est_black, `Bounds low` = 100 * max(lo),
          `Bounds high` = 100 * min(hi), TRUTH = truth_black),
        col = c("#C41230", "grey70", "grey70", "#2c7fb8"),
        ylim = c(0, 110), ylab = "Black Democratic support (%)")
abline(h = 100, lty = 2); abline(h = truth_black, lty = 3, col = "#2c7fb8")
```

**The truth is `r pc(truth_black)`% for Black voters and `r pc(truth_white)`%
for white voters.** This county
is about as racially polarized as American politics gets, and every method in
this lab said so. On the question these methods were built to answer — *is
voting polarized here?* — they all got it right.

On the question *Callais* now asks — is it polarized by **race** rather than by
**party**? — none of them took a position. They could not. What this file
records is the party of the ballot each voter requested.

Now look at how they got there.

- **The bounds contained the truth**, as they must. They always do. That is
  their entire virtue and it cost them
  `r pc(100 * (min(hi) - max(lo)))` points of precision.
- **Goodman missed by `r pc(est_black - truth_black)` points** and landed
  outside the range of possible numbers. Its white estimate,
  `r pc(est_white)`%, is less than half the true `r pc(truth_white)`%.
- **Homogeneous precincts could not answer at all** for Black voters, and
  overstated white Democratic support by
  `r pc(100 * sum(hw$dem) / sum(hw$voters) - truth_white)` points —
  `r pc(100 * sum(hw$dem) / sum(hw$voters))`% against `r pc(truth_white)`%.

The uncomfortable part is that the method that was *most confident* was the one
that was wrong, and its confidence was legible: R² of
`r pc(summary(fit)$r.squared, 3)`, a tight line, a clean plot. Nothing in the
output warned anybody. **The impossibility was the only warning, and only
because the truth happened to lie near a boundary.** In a county where Black
Democratic support were 60%, Goodman could have missed by the same
`r pc(est_black - truth_black)` points and returned a perfectly plausible
`r pc(60 - (est_black - truth_black), 0)`%, and no one would ever have known.

# Part 7 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Move the homogeneity threshold.** Part 3 used 80%. Try 70%, or 60%
(`hb <- p[p$pct_black >= 0.60, ]`). At some threshold you get a Black estimate.
How close is it, and how much did you have to loosen the definition of
"homogeneous" to get it? Would you defend that threshold under
cross-examination?

**Option B — Drop the biggest precinct.** WELLSTON CENTER is the only precinct
above 50% Black, so it is doing nearly all the work of anchoring the right-hand
end of the line. Refit without it (`p2 <- p[p$precinct != "WELLSTON CENTER", ]`)
and compare. What does the change tell you about how much of a fifty-year-old
courtroom method rests on one polling place?

**Option C — Predict, then check, precinct by precinct.** The regression
implies a Democratic share for each precinct (`fitted(fit)`). Compare those to
what each precinct actually did (`p$pct_dem`), and then to what the truth
columns say Black voters in that precinct actually did
(`p$black_dem / p$black`). Where is the model worst, and is there a pattern to
where it fails?

**Option D — Turn it around.** We validated the method where we could check it.
Write down what would have to be true for today's result to transfer to the
November general election, where nobody can check it. Then say how confident you
would be, and why.

**Option E — Your own question.** `r nrow(p)` precincts,
`r nn(sum(p$voters))` voters, and a column called `black_dem` that real analysts
never get.

# What this lab cannot tell you

- **This would not establish *Gingles* preconditions two and three.** Before 29
  April 2026 it would have. Since *Callais* a plaintiff has to show racial bloc
  voting that partisanship cannot explain, and the only outcome in this file is
  which party's ballot a voter asked for. We measured the confound. We did not
  break it.
- **A primary ballot is not a vote.** Requesting a Democratic ballot in May is
  not the same as voting for a Democrat in November, and primary electorates are
  smaller and more partisan than general ones — 18,133 people here against
  82,114 in the general. We validated the *method*, on a real electorate, for a
  real choice. We did not measure how Houston County votes.
- **This is the easy case.** Polarization here is near-total, so every method
  reached the correct legal conclusion despite bad numbers. In a county at 55/45
  the same errors would flip the answer. **Do not conclude that these methods
  are fine because they passed today.**
- **Two groups only.** 1,135 primary voters — Hispanic, Asian, American Indian,
  Other, Unknown — were dropped, because two-group ecological inference has
  nowhere to put them. They voted. The framework, not the data, excluded them.
- **Race on the form is still a measurement**, with the same one-box problem
  Weeks 12 and 14 raised.
- **We never ran King's method**, only the two it was designed to replace. It
  would do better here. It would also still be a model, and it would still be
  unfalsifiable in every county but this kind.

# On the discussion board

By **Friday, 11:59 p.m.** No length requirement; a sentence or two is enough, so
long as it shows you engaged with what we did. Any *one* of the usual four.

Four that would make good posts:

- A method returned `r pc(est_black, 0)`% and had an R² of
  `r pc(summary(fit)$r.squared, 3)`. What is the general lesson about trusting
  model fit?
- The bounds were guaranteed right and `r pc(100 * (min(hi) - max(lo)), 0)`
  points wide; the regression was
  precise and wrong. If you were the judge, which would you rather be handed,
  and what would you ask for instead?
- Every method got the *polarization* answer right and the *numbers* wrong. Does
  that make this a success story or a warning?
- *Callais* says a difference in party between racial groups "proves nothing."
  Look again at the two truth numbers in Part 6 and ask what evidence could tell
  race and party apart here. If you cannot think of any, is the Court's
  requirement a demand for rigour or a demand that cannot be met?

***

### Where the data came from

Houston County, Georgia, General Primary of 21 May 2024. Party of the primary
ballot requested comes from the state's voter-history file; race is
self-reported on the voter registration form; precinct is from registration
extract 71754. All three are public records in Georgia. Working data for the
2026 Houston County redistricting matter.

**Nothing individual-level is in the committed file** — `r nn(sum(p$voters))`
voter records were collapsed to `r nrow(p)` precinct rows before anything was
saved. **Rows whose race
had been filled in by BISG rather than reported by the voter were dropped**
before the collapse; grading one inference against another would have proved
nothing. The build is in `data/build-data.R`.
