---
title: "Where Nobody Lives"
type: brief
subtitle: "Building a national map of the census blocks that hold no people"
author: "Democracy's Data"
date: "`r local({ m <- max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))); sprintf('Last updated %s at %s%s %s', gsub('  ', ' ', format(m, '%B %e, %Y')), sub('^0', '', format(m, '%I:%M')), tolower(substr(format(m, '%p'), 1, 1)), format(m, '%Z')) })`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("zero-population-blocks-code.R")
```

```{r setup, include=FALSE}
```

```{r geometry-helpers, include=FALSE}
```

## Where nobody lives

```{r map-d3, results='asis', eval=knitr::is_html_output()}
```

```{r map-static, eval=!knitr::is_html_output(), fig.height=4.5}
```

**Figure 1. Every census block in the United States that recorded no residents in 2020.** Red is empty, white is inhabited land. Click a state to zoom to it; click again, or press Escape, to come back. Hovering gives that state's population, its block count, the share of its blocks that are empty and the share of its land.

This is a map of where nobody lives in the United States. **`r pc(PCTLAND)`% of the country's land area** is in census blocks that recorded not one resident — an area of about `r n(ZLAND)` square miles — more than the combined land area of `r andlist(BIGST)`, the `r n(NBIG)` largest states in the country.

And it understates the emptiness, because the threshold is one person. A block counts as inhabited if a single resident is counted anywhere on it, however many acres it covers.

West Virginia is the case worth holding onto. It is a thinly settled state — `r pc(ST("WV")$dens, 0)` people per square mile — and yet only `r pc(ST("WV")$pct_land)`% of its land is empty here, so it reads on the map as one of the fuller states in the country. Its people are spread in ones and twos along the hollows and the ridges, and a scattered population is very good at putting at least one person on every block. Maine is thinner still, `r pc(ST("ME")$dens, 0)` people per square mile, and `r pc(ST("ME")$pct_land)`% of its land is empty, because Maine's people are on the coast and the North Woods have nobody at all.

So what is red on this map is not thinly settled land. It is land with nobody, and the two are different things.

## What a census block is

A census block is the smallest piece the Census Bureau cuts the country into, and the reason it exists is arithmetic. Every other geography the Bureau publishes — a tract, a city, a school district, a congressional district — has to be expressible as some pile of blocks, because counts add upward but cannot be divided downward. Make the smallest piece too big and whole classes of question stop having answers. The block is the Bureau's answer to how small the smallest piece has to be.

The Bureau draws them, not the states, and it draws them along things a person could stand next to: streets, streams, railroad tracks, shorelines, and the legal boundaries of counties, cities and reservations.^[The rule and the definition are in the Bureau's glossary entry for [census block](https://www.census.gov/programs-surveys/geography/about/glossary.html); the visible-feature principle is why a river or an interstate so often has blocks of its own on either side of it.] States do get a say. Through the [Redistricting Data Program](https://www.census.gov/programs-surveys/decennial-census/about/rdo/program-management.html) they can suggest where block boundaries should fall, so that blocks nest inside the precincts they actually run elections in. But the pen is federal, and a block is drawn for the convenience of counting rather than to mark anything a resident would recognize.

There are `r n(TOT)` of them and they are small. The median block with any land on it covers `r n(FV("median_block_sqft"))` square feet — about `r FS("median_block_acres")` acres — and holds `r n(MED)` people.

The extremes run a long way from that median, in both directions and not together. The largest block in the country by area covers `r n(FV("biggest_area_sqmi"))` square miles in `r FS("biggest_area_where")`, more land than `r n(nrow(SMALL))` states and state equivalents including `r SMALL$state[1]`, whose `r n(SMALL$land_sqmi[1])` square miles it exceeds. Its population is `r FS("biggest_area_pop")`: the biggest block in the country is an empty one. The largest by population is `r n(FV("biggest_pop"))` people in `r FS("biggest_pop_where")`, on `r FS("biggest_pop_acres")` acres.

That pairing is this brief's subject. `r n(ZERO)` blocks — `r pc(PZ)`% of every block in the country — recorded nobody at all in 2020. That is not a defect in the count. It is what happens when you cut a country into pieces small enough that most pieces are a lake, an interchange, a runway, a cemetery, or a stretch of national forest.

This brief is a walkthrough. It shows where the data came from, what was done to it, and what the resulting map does and does not say. The point is not the map at the end; it is that you could rebuild it.

## Where the data came from

One download, and it holds both halves of the problem.

**TIGER/Line.** The Bureau's [TIGER/Line](https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html) files are its digital outlines of every census unit. The layer used here is `TABBLOCK20`, the 2020 tabulation blocks, one zipped shapefile per state. They are large -- about 10 GB for the country -- so the build downloads them rather than keeping them beside this chapter.

**The count rides along with the shape.** Every block record in those files carries a `POP20` field: the total population the 2020 decennial census counted in that block. It is not an estimate and not a sample. It is the decennial count, the same number published in the [P.L. 94-171 Redistricting Data Summary File](https://www.census.gov/programs-surveys/decennial-census/about/rdo/summary-files.html) that every state legislature receives for redistricting. The Bureau prepends it to the geography, so a block's boundary and its population arrive together, in one file, already joined.

That is the whole data step. There is no second source to fetch, no key to register, and no join to get wrong -- and a join on a fifteen-character block identifier is the commonest way this work goes wrong, because an identifier read as a number becomes `1.3121e+14` and loses its leading zero. Nothing here has to read one.

Three properties make this the right count. It is complete rather than sampled, so there are no margins of error below and no allowance for how far off an estimate could be, because nothing is estimated. Its unit, the census block, is the smallest the Bureau publishes, so everything larger can be built out of it. And it is the operative number rather than an academic one: it is what district maps are drawn from.

The same counts are also served, one variable at a time, by the [Census Data API](../census-api/census-api-brief.html), which has a chapter of its own. For a map of every block in the country it is much the slower road: blocks cannot be asked for a whole state at once, so covering the country takes some 3,100 requests against the 52 files downloaded here.

## How the map was made

The pipeline is four steps, and each one throws something away on purpose.

1. **Read** each state's block file and keep the blocks whose `POP20` is zero. One state at a time, so that a single state is in memory rather than the whole country.
2. **Drop the water.** A block that is all water reports zero people truthfully, but drawing those fills the Great Lakes and every coastal bay with solid color. Blocks with no land area are set aside here, and counted separately.
3. **Dissolve** the surviving blocks into one shape. Touching empty blocks become one region, which is honest here: the boundary between two adjacent empty blocks is not a feature of the world, it is an artifact of where the Bureau put its lines. Every count in this brief is taken before this step, or carried through it as a summed field, because after a dissolve a count of shapes is a count of regions and not of blocks.
4. **Simplify** with [mapshaper](https://github.com/mbloch/mapshaper), project to Albers USA, and round the coordinates to whole units of the figure's own grid.

Simplification is the step that deserves suspicion, so here are its terms. The national layer at full resolution is about 200 MB, far too large to put in a web page. What ships is 2% of the original vertices: `r n(nrow(MZ))` points in `r n(length(unique(MZ$poly)))` polygons, about 1 MB. At the width this figure is drawn, one unit of the grid is roughly 2.3 kilometers, so detail finer than that could not be seen even if it were kept. What is lost is the exact shape of any individual block. What survives is where the empty land is, which is the only claim the figure makes.

Two things in steps 3 and 4 are easy to get wrong and were got wrong here first. Dissolving produces polygons with **holes** — a populated town inside an empty county is an interior ring — and a ring drawn on its own is painted as fill rather than subtracted, which turns Nevada into a solid rectangle. Both renderings below therefore draw one path per polygon with all of its rings together, under an even-odd fill rule. And rounding coordinates has to happen *after* projecting, not before: rounding longitude and latitude to whole numbers rounds them to whole degrees, and the country dissolves.

The built tables are small and readable: [`state_blocks.csv`](data/derived/state_blocks.csv), one row per state; [`block_pop_hist.csv`](data/derived/block_pop_hist.csv), a count of blocks at each population from 0 to `r n(CAP)`; [`zero_land.csv`](data/derived/zero_land.csv), land area of the empty blocks by state; [`block_size.csv`](data/derived/block_size.csv), how big a block is and the two extremes; [`dc_blocks.csv`](data/derived/dc_blocks.csv) with [`dc_frame.csv`](data/derived/dc_frame.csv) and [`dc_facts.csv`](data/derived/dc_facts.csv), one city at full block resolution; [`facts.csv`](data/derived/facts.csv), the national totals quoted in this text; and the figure geometry in [`map_zero.csv`](data/derived/map_zero.csv), [`map_states.csv`](data/derived/map_states.csv) and [`map_frame.csv`](data/derived/map_frame.csv), stored as integer figure coordinates so the web version and the print version draw the same shapes from the same numbers.

```{r data-table}
```

## How big is a census block, anyway?

```{r hist-d3, results='asis', eval=knitr::is_html_output()}
```

```{r hist-static, eval=!knitr::is_html_output(), fig.height=5.0}
```

**Figure 2. How many people live in a census block.** Blocks grouped by population, from empty to a thousand and over; the red bar is the `r n(ZERO)` with nobody. Bins widen toward the tail because block populations do: binned this way the whole range fits one axis that starts at zero, so every bar's length is its count and nothing has to be logged or broken. The panel beneath is the running share — the portion of all blocks at or below each bin. Hover a bin in the web edition for its exact count.

The shape says the block is a small thing. The median block in the country holds `r n(MED)` people. Among blocks with anybody at all, the median is `r n(MEDINH)`. `r pc(UNDER10)`% of all blocks hold ten people or fewer, and the running share passes half by the 20-49 bin. A census block is not a neighborhood; it is closer to a city block, which is what it was named for.

This matters for reading any block-level statistic. A percentage computed on a unit holding fourteen people is mostly noise, and the Bureau's [disclosure-avoidance noise](https://www.census.gov/programs-surveys/decennial-census/decade/2020/planning-management/process/disclosure-avoidance.html) is deliberately largest, in relative terms, exactly here.

## Which states are emptiest

```{r state-d3, results='asis', eval=knitr::is_html_output()}
```

```{r state-static, eval=!knitr::is_html_output(), fig.height=5.7}
```

**Figure 3. The share of each state's blocks that recorded nobody.** Sorted rather than alphabetical, because the ordering is the finding. Hovering a bar in the web edition gives the counts underneath the percentage.

The range runs from `r MOST$state` at `r pc(MOST$pct_zero)`% to `r LEAST$state` at `r pc(LEAST$pct_zero)`%. It is tempting to read this as a map of emptiness, and it is partly that. But it is also a map of how the Bureau cut each state. Blocks are drawn along visible features, so a state with a dense grid of section-line roads gets many small blocks whether or not anyone lives on them, and a state with a lot of water gets a lot of water blocks. `r MOST$state` is genuinely empty; it also has an enormous coastline and river network, each stretch of which is its own block.

## Is this just a map of density?

```{r dens-d3, results='asis', eval=knitr::is_html_output()}
```

```{r dens-static, eval=!knitr::is_html_output(), fig.height=6.4}
```

**Figure 4. Emptiness against population density, measured two ways.** One dot per state on a shared log axis, because state densities span four orders of magnitude, from `r pc(FV("dens_min"))` people per square mile in Alaska to `r n(FV("dens_max"))` in the District of Columbia. The top panel is the share of each state's *land* that sits in empty blocks; the bottom is the share of its *blocks* that are empty. Dashed lines are least-squares fits. Only the extremes and the states furthest from each line are labelled; hover any dot in the web edition for the rest.

The two panels disagree, and the disagreement is the point. Density predicts the share of a state's **blocks** that are empty rather well: `r pc(FV("dens_r_log"), 2)`, or about `r pc(100 * FV("dens_r2_log"), 0)`% of the variation between states. It predicts the share of a state's **land** that is empty much less well: `r pc(FV("land_r_log"), 2)`, about `r pc(100 * FV("land_r2_log"), 0)`%.

That ordering is backwards from what you would want, and the reason is the thing this brief keeps returning to. A block count is partly a record of how the Bureau cut a state, and the Bureau cuts dense places into small pieces. So the block measure is contaminated by the very thing it is being compared against, and correlates well for a reason that has nothing to do with land. The land measure is the substantive one, and it is loose: `r FS("land_hi_st")` is `r pc(FV("land_hi_pct"))`% unoccupied land and `r FS("land_lo_st")` is `r pc(FV("land_lo_pct"))`%, a difference that is about federal ownership and settlement pattern rather than about people per square mile. The two measures agree with each other only to `r pc(FV("measures_r"), 2)`.

On the raw scale neither relationship survives: `r pc(FV("dens_r_raw"), 2)` and `r pc(FV("land_r_raw"), 2)`. Whatever emptiness tracks, it tracks the *order of magnitude* of a state's density, so the distance from Wyoming to Ohio matters far more than the distance from Ohio to New Jersey.

And then there is the point stranded at the right of both panels.

## One city, block by block

```{r dc-d3, results='asis', eval=knitr::is_html_output()}
```

```{r dc-static, eval=!knitr::is_html_output(), fig.height=6.2}
```

**Figure 5. Every census block in the District of Columbia.** Red is a block with no residents, pale grey is a block with residents, and blue-grey is a block that is entirely water. Water is drawn as its own class rather than folded into the red, because a block with nobody on it *because it is the Potomac* is a different fact about a city than a block with nobody on it because it is a park, and merging the two would let the rivers argue the case.

The District is the densest place in the country, `r n(FV("dc_density"))` people per square mile of land. `r n(FV("dc_zero"))` of its `r n(FV("dc_blocks"))` blocks — `r pc(FV("dc_pct_zero"))`% — recorded nobody, and `r n(FV("dc_zero_land"))` of those are dry land. They cover `r pc(FV("dc_empty_sqmi"))` of the District's `r pc(FV("dc_land_sqmi"))` square miles, `r pc(FV("dc_empty_land_pct"))`% of it.

The shapes are recognizable if you know the city, and that is the argument. The long wedge running north from the middle is Rock Creek Park. The band across the centre, west of the Capitol, is the Mall. The eastern edge is the Anacostia and the parkland along it; the large form in the south is a military installation. None of this is unpopulated in the sense of being remote. It is unpopulated because it is a park, a monument, a rail yard, a base, a river — land in a dense city that no one is counted as living on.

That is why the District sits where it does in Figure 4: it is not a failure of the relationship between density and emptiness so much as a demonstration of what emptiness is actually measuring. **A place can be as dense as anywhere in America and still be a quarter empty**, because emptiness is a fact about how land is used and how it was cut, not only about how many people live nearby.

## What this brief cannot tell you

The `r pc(PZ)`% is a fact about blocks, not about people or land, and the three do not track each other. Nothing here says how much of the country is *uninhabitable*, only which units recorded no usual residents on 1 April 2020. A block containing a large prison, a dormitory closed for the pandemic, or a seasonal resort counted empty that day may be busy most of the year.

The map is drawn from the land blocks. `r n(ZWBLK)` of the empty blocks are water and are not drawn, because including them fills the Great Lakes and every coastal bay with solid color and makes the map a map of hydrography. That is a presentation choice, and it is reversible; the land-area figures are unaffected either way, because the Bureau's `ALAND20` already excludes water.

Empty is not the same as thinly settled, and this is not a population density map, though Figure 4 shows how much of one it is. Density is people divided by area, a continuous quantity every block has some value of. Emptiness is a threshold applied to a unit: zero, or not zero. A solid red county on this map means its blocks crossed that threshold, not that nobody lives nearby.

Simplification means no individual block on this map is the right shape. Do not measure anything off it. The [built tables](data/derived/state_blocks.csv) carry the counts; the figure carries the pattern.

And zero is a published number, not a raw one. The 2020 counts passed through disclosure avoidance before release, which can move a small count by a little. At the block level that noise is proportionally largest, though a block that is a reservoir was not going to be populated either way.

## What you should have learned

- The census block is small: the median holds `r n(MED)` people, and `r pc(UNDER10)`% hold ten or fewer. Percentages computed on units that size are fragile.
- `r pc(PZ)`% of blocks are empty but they cover about `r pc(PCTLAND)`% of the land. Counting units and measuring area are different operations, and "most of the country is empty" and "most blocks are empty" are different claims.
- A join that matches 100% is a test you passed, not a formality. Identifiers are text.
- Every step that made this map — filtering, dissolving, simplifying — discarded something. A figure is trustworthy when you can say what it discarded and why, not when nothing was discarded.

## Extensions

- `state_blocks.csv` — divide `pop` by `blocks` for each state. Which states put the fewest people in a block, and is that the same list as the emptiest states in Figure 3?
- `state_blocks.csv` joined to `zero_land.csv` — rank states by `pct_zero`, then by `zero_land_sqmi`. Name a state that is high on one list and low on the other, and say what the Bureau did to produce that gap.
- `block_pop_hist.csv` — the last row is a bucket, every block of `r n(CAP)` people or more. How many blocks are in it, and what kind of place holds that many people inside one block?
- `block_pop_hist.csv` — compute the share of the national population living in blocks of ten people or fewer. Compare it with the `r pc(UNDER10)`% of *blocks* that size. Explain the difference to someone in one sentence.
- **Stretch.** Download one state's [block file](https://www2.census.gov/geo/tiger/TIGER2020/TABBLOCK20/) and check its `POP20` against the same counts served by the API. Report how many disagree, and decide which you would trust in print.
- **Stretch.** Rebuild Figure 1 for one state without the dissolve step, drawing each empty block separately. Say what the dissolved version hides and what it clarifies.

## Sources

**Data**

- U.S. Census Bureau, TIGER/Line Shapefiles 2020, `TABBLOCK20` layer — the
  boundary of every 2020 tabulation block, one file per state, carrying
  `GEOID20`, `ALAND20`, `AWATER20` and `POP20`. This is the only file the
  build downloads: `POP20` is the 2020 decennial count for that block, so the
  population arrives prepended to the geography.
  <https://www2.census.gov/geo/tiger/TIGER2020/TABBLOCK20/>
- U.S. Census Bureau, 2020 Census Redistricting Data (P.L. 94-171) Summary
  File — the decennial release `POP20` reports, and the file state
  legislatures receive. Not downloaded here; cited because it is what the
  number in the shapefile is.
  <https://www.census.gov/programs-surveys/decennial-census/about/rdo/summary-files.html>
- U.S. Census Bureau, Cartographic Boundary Files 2020, state layer at
  1:500,000 — generalized outlines drawn for mapping rather than measurement,
  clipped to the shoreline, used here only as the frame of Figure 1.
  <https://www.census.gov/geographies/mapping-files/time-series/geo/cartographic-boundary.html>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`block_pop_hist.csv`](data/derived/block_pop_hist.csv), [`block_size.csv`](data/derived/block_size.csv), [`dc_blocks.csv`](data/derived/dc_blocks.csv), [`dc_facts.csv`](data/derived/dc_facts.csv), [`dc_frame.csv`](data/derived/dc_frame.csv), [`facts.csv`](data/derived/facts.csv), [`map_frame.csv`](data/derived/map_frame.csv), [`map_states.csv`](data/derived/map_states.csv), [`map_zero.csv`](data/derived/map_zero.csv), [`state_blocks.csv`](data/derived/state_blocks.csv), [`zero_land.csv`](data/derived/zero_land.csv)

**Method**

- Bloch, M. *mapshaper* — the simplification, dissolve and projection steps in
  Figure 1 are all mapshaper commands; its Visvalingam implementation is what
  "2% of the vertices" means here.
  <https://github.com/mbloch/mapshaper>
- Bostock, M. *D3* — the three figures in the HTML edition are plain SVG built
  with D3 scales and axes.
  <https://d3js.org/>

**Documentation**

- U.S. Census Bureau, "Glossary: Census Block" — the definition, and the rule
  that blocks are bounded by visible features and by legal boundaries.
  <https://www.census.gov/programs-surveys/geography/about/glossary.html>
- U.S. Census Bureau, "Disclosure Avoidance and the 2020 Census" — why a
  published block count is not a raw block count.
  <https://www.census.gov/programs-surveys/decennial-census/decade/2020/planning-management/process/disclosure-avoidance.html>

```{r ai-prompt, echo=FALSE, results="asis"}
```
