---
title: "Where Campaigns Travelled, and Where They Did Not"
type: brief
subtitle: "Eight months of one campaign's travel log, and the half of America that never appears in it"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("campaign-visits-code.R")
```

```{r setup, include=FALSE}
```

Between `r format(min(v$date), "%d %B")` and `r format(max(v$date), "%d %B %Y")`, five people made `r n(nrow(v))` public campaign appearances in the United States: the two presidential nominees, their running mates, and the president who withdrew in July. `r nrow(z)` of the `r nrow(m)` jurisdictions that cast electoral votes received none of them. Not a rally, not a fundraiser, not a fifteen-minute stop at a diner. In half the country, across eight months, nobody came.

A presidential campaign has more money than it can spend well, more staff than it can supervise and more advice than it can use. It has exactly one resource that cannot be manufactured: **the candidate's own time.** There are a fixed number of days, and the candidate can be in one place on each of them. The travel schedule is therefore a ranked list of the campaign's own judgments about where the election will be decided, compiled by the people with the best polling in the country and revised daily. It is a strategy document that a campaign publishes by executing it.

Start with what one line of it is.

```{r one-record}
```

One row is **one public appearance, by one person, in one city, on one day.** Not a day of campaigning, not a state visit, not an advertising buy — an event.

That definition is doing more work than it looks. A rally of twenty thousand and a fifteen-minute stop at a diner are both exactly one row, and nothing in this file distinguishes them. Every count below is a count of *occasions*.

## How much of a campaign this is

```{r scope}
```

**Note who compiled this.** Campaign-finance data exists because federal law compels its filing. Nothing compelled any of this. A news organization watched the campaigns and wrote down where they went, and the file exists at the Associated Press's discretion. That is why it carries a version number, and why the version number turns out to matter.

Nothing in the Constitution says a presidential campaign should visit some places and not others. The concentration comes from one rule, adopted state by state and written nowhere in federal law: **the winner of a state takes all of its electoral votes.** Under that rule a vote is worth campaigning for only if it might change which candidate crosses 270, which means only where the outcome is in doubt.

Whether any of that matters depends on whether attention does anything, and it does. **Enos and Fowler (2018)** compared targeted and untargeted states in 2012. They estimated that the presidential campaigns raised turnout in heavily targeted states by **about 7 to 8 percentage points**, mostly through traditional ground campaigning. Being campaigned in is not a symbolic good. People in battleground states are mobilized, informed and argued with; people elsewhere are left to watch.

## The only witness there is

The travel log is an Associated Press interactive, *Tracking the 2024 presidential campaign trail*, published at an embed address that carries a version number. The table above is its whole extent: `r n(nrow(v))` appearances by `r length(unique(v$candidate))` people over eight months. It is a wire cooperative's byproduct. Reporters covering a beat wrote down where the candidates were, and somebody decided the accumulation was worth publishing. There is no methodology note, no codebook, no correction log, and nobody signed it.

It was never published as data, either. What can be fetched is the CSV that feeds the interactive graphic, the file the page loads in order to draw itself. It is not a dataset with a landing page and a citation. That is where the version number in the address comes from. It means the download is one easy line of code that will stop working, with no announcement, on the day the graphic is retired.

That is a thinner record than anything in the campaign-finance chapters, and it is still the best evidence there is, because the alternative is none. No agency collects candidate itineraries. A campaign's internal schedule is disclosed to nobody, and the FEC file records what a campaign spent without recording where the candidate stood while spending it. The question here has no official record and is never going to acquire one: where did the people with the best polling in the country believe the election would be decided? A newsroom's log is not a second-best substitute for a filing. It is the only witness.

What that costs is a definition nobody published. Someone at AP decided what counts as a public campaign appearance, and the decision is invisible from here. A fundraiser in a private home, a local television interview, an official White House event that is also a campaign event: each is in or out by a judgment made off the page. The very first row of this file is an infrastructure announcement, which the notes column says in as many words. Coverage is unlikely to be even either, since the principals were watched more closely than their running mates and a wire service's attention is not spread uniformly over fifty states. None of that is falsifiable from inside the file, which is a different position from a filing that can be checked against the document it summarizes.

One more thing belongs here rather than in the source, because it is a property of the build. **A state that received no attention is not a row in this file.** The tracker records events, and a place with no events leaves nothing to record. The `r nrow(z)` jurisdictions that got no stop enter this chapter only because the state results are kept as a separate table and joined back to the visits. That join keeps every state and fills the missing counts with zero.

The opening claim of this chapter is in that sense manufactured by a join. The silence had to be built, because absence does not write itself down. The file's other instability, that it went on changing after it had been cited, gets a section of its own further on and is worth waiting for.

## A file built to draw a graphic

The capture is kept exactly as it arrived. Here is its first record, every column of it, with a star against the ones this chapter keeps.

```{r ap-raw}
```

That is record 1 of `r format(nrow(rawv), big.mark = ",")`, all `r ncol(rawv)` columns of it.

`r ncol(rawv)` columns arrive and `r length(kept)` are kept, and the `r ncol(rawv) - length(kept)` that go are the most interesting thing in the file. `biden-total-visits-city`, `total-state-visits`, `county-text`, `reg-voters`: none of those is a fact about the event on this row. They are the contents of the tooltip that appears when a reader hovers over a dot on AP's map. This is not a dataset with a graphic attached. It is a graphic's payload, and the rows are events only incidentally.

That has consequences a codebook would have warned about. The counts in those columns are cumulative over the whole campaign, so the same total is repeated on every row of the same city and would be badly wrong if anyone summed the column. `reg-voters` arrives as the text `r r1[["reg-voters"]]`, quoted, with a thousands separator inside it, which is a number formatted for a reader rather than a number. `reg-voters-as-of-date` is `r r1[["reg-voters-as-of-date"]]` with no year in it. And the date the chapter runs on arrives as `r r1[["date"]]`, which sorts as text into an order no calendar recognizes.

The kept version is eleven columns and one event per row:

```{r ap-clean}
```

Three judgments, and the third is the one that makes the chapter.

**Take the columns that describe the row, and drop the ones that describe the page.** Everything cumulative goes, and it goes because it is precomputed. Any total this chapter wants is recomputed from the events themselves, which is slower and is the only way the total can be checked.

**Parse the date, and refuse to continue if any of them fails.** `r r1[["date"]]` becomes `r format(v$date[1])`. Two-digit years are ambiguous by construction, and the build stops outright if a single row fails to convert rather than letting one silent `NA` fall out of a chart.

**Manufacture the zeros.** This is the important one. AP's file records events, so a state where nothing happened contributes no row. Counted straight, the file has `r length(unique(v$state))` states in it and the United States has `r nrow(m)` jurisdictions with electoral votes. The `r nrow(z)` places nobody went exist in this chapter only because the build writes a separate file of state results. The setup joins the two, keeps every state whether or not it matched, and fills the empty side with zero rather than with `NA`.

Absence does not write itself down. The central claim of this chapter is a fact about rows that had to be created, from a second source, in order to be counted at all.

## Where they went

```{r top-states}
```

```{r pointmap-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

```{r pointmap-d3, results='asis', eval=knitr::is_html_output()}
```

```{r conc}
```

**Figure 1. Every logged stop of the 2024 campaign, placed where it happened.** One mark per appearance, `r n(nrow(v))` in all, by five people over eight months. The clustering is not an artifact of population: the empty regions include some of the largest cities in the country.

**Seven states holding `r pc(100 * sum(t7$ev) / sum(m$ev))`% of the electoral college absorbed `r pc(100 * sum(t7$visits) / nrow(v))`% of the campaign.**

The empty half of that map is the finding. Pennsylvania alone took `r m$visits[m$state == "Pennsylvania"]` stops across `r length(unique(v$city[v$state == "Pennsylvania"]))` different towns — a stop somewhere in one state every `r pc(as.numeric(max(v$date) - min(v$date)) / m$visits[m$state == "Pennsylvania"], 1)` days, for eight months.

## Where they did not go

That table was built by counting rows in the visit file. A state that never appears in the file has no row to count, and so does not appear in the answer either. To see the states that got nothing, the counting has to start from a list of all the states and attach the visits to it, rather than the other way round.

```{r zeros}
```

**`r nrow(z)` of `r nrow(m)` jurisdictions, half the country, received not one stop** from any of the five people on the two tickets, in eight months. Between them they hold `r sum(z$ev)` electoral votes, `r pc(100 * sum(z$ev) / sum(m$ev))`% of the total.

```{r cartogram-static, eval=!knitr::is_html_output(), fig.height=4.1}
```

```{r cartogram-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The country as a grid, shaded by how many stops each state received.** One tile per jurisdiction, positioned roughly where the state belongs. The `r nrow(z)` blank tiles are the states nobody visited; between them they hold `r sum(z$ev)` electoral votes, `r pc(100 * sum(z$ev) / sum(m$ev))`% of the total.

The cartogram shows *which* states were skipped. It cannot show what they were owed, because every tile is the same size whatever the state's electoral weight. Put the two quantities on the same axes and the disproportion becomes a single shape. Figure 3 orders all `r nrow(m)` jurisdictions from least to most attention per electoral vote, then walks along them accumulating both their electoral votes and their campaign stops.

> **Before you turn to that figure, put a number to it.** Walk up that ordering
> until you have half the electoral college in hand — the `r nrow(z)`
> states that got nothing, which between them hold
> `r pc(100 * zero_ec)`% of the college, and then however many
> lightly-visited states it takes to reach fifty. **What share of the
> `r n(nrow(v))` stops did that bottom half of the country receive?**
> Write down a percentage now, while the only thing you have to go on is the
> map.

```{r lorenz-static, eval=!knitr::is_html_output(), fig.height=4.0}
```

```{r lorenz-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. How the campaign's time was distributed across the electoral college.** Each point is one state, added in order of how much attention it got per electoral vote. The horizontal axis is the running share of electoral votes, and the vertical axis the running share of stops. The dashed diagonal is the country a campaign would visit if it weighted every electoral vote equally. The red segment along the bottom is the `r nrow(z)` states that received nothing: `r pc(100 * zero_ec)`% of the electoral college and none of the campaign. The Gini coefficient of the curve is `r r2(gini)`.

**The least-visited half of the electoral college received `r pc(100 * approx(lx, ly, 0.5)$y)`% of all campaign stops.** Most people guess somewhere in the teens, reasoning that the zero-stop bloc is only `r pc(100 * zero_ec)`% of the college and that the states sitting just above it must have picked up a respectable remainder. They did not: the whole bottom half of the country, weighted the way the Constitution weights it, is worth about `r pc(approx(lx, ly, 0.5)$y * nrow(v), 0)` of the `r n(nrow(v))` stops. The Gini coefficient of that curve is `r r2(gini)` — a level of concentration that would be reported as extreme if the quantity being divided were income rather than a candidate's time.

**This is where most analyses of this kind go wrong, and it comes down to one word.** The join behind these figures keeps every state whether or not it matched a visit (`all.x = TRUE`). The default behavior keeps only the states appearing in both files. Run the default and the `r nrow(z)` blank tiles vanish silently: no warning, no error, a perfectly clean table describing a country that does not exist.

## Attention per unit of electoral weight

```{r per-ev}
```

**Nevada has `r m$ev[m$state == "Nevada"]` electoral votes to California's `r m$ev[m$state == "California"]`, and received `r m$visits[m$state == "Nevada"]` stops to California's `r m$visits[m$state == "California"]`.** Per electoral vote, Nevada got about **`r pc(per("Nevada") / per("California"), 0)` times** the attention California did. Texas, the second-largest prize in the country, received `r m$visits[m$state == "Texas"]`\.

## The two obvious explanations, and what happens to them

Candidates go where the votes are. That is the first thing anyone would guess, and it is testable in one line.

```{r size}
```

**`r r2(cor(m$visits, m$ev))`.** Essentially nothing. `r lg$state[1]`, `r lg$state[2]` and `r lg$state[3]` hold `r sum(lg$ev[1:3])` electoral votes between them and received `r sum(lg$visits[1:3])` stops in total — fewer than any one of the seven most-visited states, the smallest of which got `r min(t7$visits)`.

Size is not the answer. The natural second guess is closeness.

```{r close}
```

**`r r2(cor(m$visits, m$absm))`** — better, and in the right direction: the closer the state, the more stops. But it is a moderate correlation, not a mechanism, and the exceptions are conspicuous.

```{r close-tab}
```

`r cl$state[which.min(cl$visits)]` was decided by `r pc(cl$absm[which.min(cl$visits)], 2)` points and got `r cl$visits[which.min(cl$visits)]` stops; Minnesota, decided by `r pc(m$absm[m$state == "Minnesota"], 2)`, got `r m$visits[m$state == "Minnesota"]`\. Both were close and both were left alone.

So neither variable works. Size explains almost nothing, and closeness explains something but not much. The two most obvious accounts of campaign strategy have both failed a simple test, on a file that ought to be the ideal case for them.

## Neither variable works, and both do

Ask the size question again, but only of the close states. Then ask the closeness question again, but only of the large states.

```{r interaction}
```

**Among the `r nrow(cl)` states decided by under five points, size and stops correlate at `r r2(cor(cl$visits, cl$ev))`. Among the ten largest states, closeness and stops correlate at `r r2(cor(lg$visits, lg$absm))`.**

Marginally, each variable is nearly useless. Conditionally, each is powerful. That is the signature of an **interaction**: the effect of one variable depends on the value of the other, and averaging over the other destroys it.

Build the product directly: electoral votes multiplied by whether the state was within five points, an index with no tuning in it. Then correlate that with stops:

```{r index}
```

**`r r2(cor(m$visits, m$prize))`**, using no information that was not already in the two failed variables.

A campaign is not asking "where are the voters?" and it is not asking "where is it close?" It is asking one question that multiplies the two: *where might a day of my time change who wins?* A large safe state fails that test. A tiny close state passes it barely. A large close state is worth `r m$visits[m$state == "Pennsylvania"]` visits.

**This is the winner-take-all rule expressed as a travel schedule**, and Figure 4 is what it looks like when every jurisdiction is plotted at once.

```{r d3-scatter, results='asis', eval=knitr::is_html_output()}
```

```{r static-scatter, eval=!knitr::is_html_output(), fig.height=3.8}
```

**Figure 4. Stops against the prize a state represented.** One point per jurisdiction: electoral votes on one axis, the campaign's stops on the other, with the states decided by under five points distinguished from the rest. Neither size nor closeness orders this picture on its own; their product correlates with stops at `r r2(cor(m$visits, m$prize))`.

## Two rows that do not mean what the column says

Look at the far right of that plot. One jurisdiction was decided by `r pc(m$absm[m$state == "District of Columbia"], 2)` points and still logged `r m$visits[m$state == "District of Columbia"]` stops.

```{r dc}
```

**Nobody was campaigning in Washington for Washington's `r m$ev[m$state == "District of Columbia"]` electoral votes.** They were speaking to national conventions of interest groups, in the city where the studios, the donors and the press corps already are. For two of the five people in this file, so is the office. The audience was not in the room, and it was not in the District.

The row satisfies every filter you would write. It does not belong to the question you are asking. And it is not harmless:

```{r dc-effect}
```

**Removing one row of `r nrow(m)` moves the correlation from `r r2(cor(m$visits, m$absm))` to `r r2(cor(mn$visits, mn$absm))`.** One misclassified case was suppressing about a quarter of the relationship.

Now the opposite error. Nebraska is a safe state, decided by `r pc(m$absm[m$state == "Nebraska"], 2)` points, so the index above predicts nothing there. It got `r m$visits[m$state == "Nebraska"]` stops:

```{r nebraska}
```

Both in Omaha, because **Nebraska does not award its electoral votes winner-take-all.** One goes to the winner of the Omaha-centered congressional district, and that district was genuinely in play. The results file has one row per state and no row for a district, so the prize the campaign was chasing is not represented anywhere in the data.

**DC is a row that satisfies the filter and does not belong to the question. Nebraska is a prize that belongs to the question and has no row.** Both are the same failure: the unit of the file is not the unit of the thing being contested.

## The source kept moving after it was cited

The AP tracker lives at a URL carrying a **version number**, and AP went on publishing new versions after the election. The figures above were originally built against version `r OLD$version` and are now built against version `r NEW$version`. Both are still served, so the two can be put side by side.

```{r snapshots}
```

Two different things happened, and only one of them is obvious.

**The tracker gained the ending.** Version `r OLD$version` stops on `r format(LAST_OLD, "%d %B")`, four days before the election. Version `r NEW$version` runs to `r format(as.Date(NEW$last_event), "%d %B")` and adds `r added_late` stops in those final days. That is `r pc(100 * added_late / nrow(v))`% of the entire eight-month campaign packed into three days. It is exactly the closing rush you would expect, and exactly what the older file could not show.

**The tracker also rewrote the part it had already covered.** Both versions claim everything up to `r format(LAST_OLD, "%d %B")`. For that window the old capture has `r n(OLD$rows_in_shared_window)` stops and the new one has `r n(NEW$rows_in_shared_window)`\. **`r added_back` events were added to a period the file already claimed to describe.** One of them put a state on the map that had none before. New Mexico went from zero stops to `r m$visits[m$state == "New Mexico"]`, which moves a tile on the cartogram and changes the count of skipped states by one.

**A public dataset with no statutory deadline has no moment at which it is finished.** The census closes; a campaign-finance filing has a due date and an amendment history; this has neither. It is complete when a newsroom stops updating it, and nothing announces that day. **A citation to this source without a version number does not identify what was read** — which is why one version is pinned here, both are fetched, and the comparison is kept in `ap_snapshots.csv`.

## Two more edges, and the calendar that shows them

```{r edges}
```

**It still ends the day before the election**, so every count remains a floor — much closer to the truth than version `r OLD$version` allowed, and still short.

**The candidate changed mid-file.** Biden's last logged stop and Harris's first are `r as.numeric(min(v$date[v$candidate == "Harris"]) - max(v$date[v$candidate == "Biden"]))` days apart in July, so any comparison of "the Democrat" to "the Republican" over time compares a continuous campaign to a broken one. As whole tickets the two sides are nearly matched, `r sum(v$candidate %in% c("Biden","Harris","Walz"))` to `r sum(v$candidate %in% c("Trump","Vance"))`. That surprises people who expect one side to have obviously more.

Every figure so far has thrown the `date` column away. Put it back and both of those edges, and the closing surge the new version added, become visible at once.

```{r calendar-static, eval=!knitr::is_html_output(), fig.height=3.8}
```

```{r calendar-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 5. The whole campaign as a calendar.** One lane per person, one tick per logged stop, taller where somebody made several appearances in a day; red is the Republican ticket and blue the Democratic. The gray band is the week between Biden's last stop and Harris's first. The pink band at the right is the final `r added_late` stops, `r pc(100 * added_late / nrow(v))`% of the entire eight-month campaign, which version `r OLD$version` never saw.

The lanes make three claims visible at once. **Biden's lane simply stops** in July and Harris's starts a week later, so no continuous "Democratic campaign" exists to compare against Trump's. **Walz has no lane at all before `r format(min(v$date[v$candidate == "Walz"]), "%d %B")`,** so his `r sum(v$candidate == "Walz")` stops are `r pc(100 * sum(v$candidate == "Walz") / nrow(v))`% of the file compressed into its last third. And **every lane thickens into the pink band**: `r added_late` of `r n(nrow(v))` stops, `r pc(100 * added_late / nrow(v))`% of the whole campaign, fall in the last three days — the part version `r OLD$version` was missing.

## Where the presidency was actually contested

**The 2024 presidential campaign was conducted in seven states.** They took `r pc(100*sum(t7$visits)/nrow(v))`% of all stops and hold `r pc(100*sum(t7$ev)/sum(m$ev))`% of the electoral college, while `r nrow(z)` states saw nobody at all.

The mechanism is not a slogan. Neither size nor closeness predicts attention on its own, at `r r2(cor(m$visits, m$ev))` and `r r2(cor(m$visits, m$absm))`; their product does, at `r r2(cor(m$visits, m$prize))`. Campaigns behave as though they are maximizing the chance that a day of the candidate's time changes an outcome, which is exactly what winner-take-all rewards.

It follows that **the Electoral College does not reward smallness.** Nevada and New Hampshire are small and were visited; Wyoming, Vermont, the Dakotas, Delaware and Alaska are small and were not. The system rewards closeness, and a few small states happen to be close.

None of it says anything about effort, money or audience, since an arena and a diner are one row each. Nor about what turnout or votes any of this produced. Nothing here measures a consequence. And no number from this file can be cited without saying which version of it was read.

## Whether the concentration is a defect

**The standard defense of the Electoral College is that it protects small states.** This data neither supports that defense nor quite refutes it. It replaces it with a sharper claim. The system rewards *competitiveness*, not size in either direction. A small safe state is the most ignored place in America, and a small close state is the most courted. Which one a given state is has nothing to do with the Electoral College, and everything to do with how its voters happen to be divided in a particular cycle.

**The serious reply is that the counterfactual is not equality.** Under a national popular vote attention would be reallocated, not spread out: campaigns would go where votes are cheap to move in bulk. Whether that concentrates attention more or less than the present arrangement is an empirical question, and `r n(nrow(v))` rows from one election cannot answer it.

**A second dispute is about what a visit is for.** **Chen and Reeves (2011)** studied county-level appearances in 2008 and found the two tickets following different logics. McCain-Palin concentrated on counties where the previous Republican ticket had done well, which is a base strategy. Obama-Biden went to fast-growing counties on the edges of metropolitan areas. If mobilization and persuasion imply different maps, then "attention" is not one quantity, and a state-level count of stops averages over the distinction. Figure 1 is at the right level of detail to test that; this chapter does not test it.

**A third is about consequences for citizens rather than candidates.** **Gimpel, Kaufmann and Pearson-Merkowitz (2007)** compared battleground and "blackout" states and found differences in political engagement between them. On that account Figure 2 describes not only where campaigns go but where political life is more intense. That makes the `r nrow(z)` blank tiles a claim about the quality of democratic life, rather than about travel budgets.

## What a travel log can testify to

The strengths are real. This is a near-complete record of public appearances by both tickets, geocoded and dated. Both sides were watched by the same observer applying the same rule, which makes a ticket-to-ticket comparison mean something, in a way that comparisons across differently sourced files usually do not.

The weaknesses begin with what kind of document it is. **It is a news organization's log, not a legal filing.** Nobody can be penalized for an omission, and coverage of the running mates is probably less complete than coverage of the principals. It is versioned, and the version moved: `r OLD$rows` stops became `r NEW$rows`, `r added_back` of them backfilled into a window the earlier file already claimed to cover, and nothing guarantees version `r NEW$version` is the last. It ends on `r format(max(v$date), "%d %B")`, the day before the election, so every count in this chapter is a floor.

Then there are the limits of the unit. **An event is an event**: no crowd size, no time spent, no money, which is the single largest weakness of the file. `r nrow(z)` states have no rows in it at all, so any analysis beginning from the visit file rather than from a list of states will never see them. The five-point cut that defines a close state is a convention rather than a fact about the world. The pattern survives other thresholds, but the exact correlations move.

And `r r2(cor(m$visits, m$ev))` is a correlation computed on a badly lopsided variable. The rank correlation between size and stops is `r r2(cor(m$visits, m$ev, method = "spearman"))`, appreciably larger. So "size does not matter" is too strong, and *size alone does not order the map* is right. The unit is the state, which is the wrong unit for Maine and Nebraska and for the cities that Figure 1 shows and everything after it throws away. And it is one election, in which only `r sum(m$absm < 10)` states finished within ten points; nothing here establishes that 2024 was typical.

Three questions stay open. Whether attention caused anything in this cycle: Enos and Fowler found large turnout effects in 2012, and this file has no turnout column. What a national popular vote would actually look like on the ground, as opposed to in the abstract. And whether the two tickets were running the same play at all, which needs county-level detail rather than state counts.

The travel log is a strategy document that nobody wrote down as a strategy. Read as one, it says that the presidency of a country of three hundred and forty million people was contested in seven states. The rule producing that result is not in the Constitution. It is in fifty separate statutes, each of which could be amended by the legislature that passed it.

## Sources

**Data**

- Associated Press, "Tracking the 2024 presidential campaign trail," version
  `r NEW$version`\. <https://interactives.ap.org/embeds/ziEfd/54/> — public
  campaign appearances by both tickets, `r format(min(v$date), "%d %B")` to
  `r format(max(v$date), "%d %B %Y")`\. Version `r OLD$version`, against which the
  earlier release of this lab was built, is retained for comparison in
  `ap_snapshots.csv`.
- State-level 2024 presidential returns, the same file used in Weeks 1 and 2.
- The state-name join is checked, and the two published versions of the tracker
  are compared, rather than either being taken on trust.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`ap_snapshots.csv`](data/derived/ap_snapshots.csv), [`campaign_visits_2024.csv`](data/derived/campaign_visits_2024.csv), [`pres2024_states.csv`](data/derived/pres2024_states.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `campaign_visits_2024.csv` has a state on every event. Count events per state, then join `pres2024_states.csv` and compare against `ev` and `margin`. How much of the country got no visit at all?
- Compute visits per electoral vote for the states that were visited. Which states got more attention than their weight, and which got less?
- `ap_snapshots.csv` records two versions of the same file, with different row counts and a different `last_event`. Work out what the earlier version was missing. What would you have published from it?
- Group the events by `event_type` and by candidate. Do the two campaigns travel the same way?

**Academic literature**

- Chen, L. J. and Reeves, A. (2011). "Turning Out the Base or Appealing to the
  Periphery? An Analysis of County-Level Candidate Appearances in the 2008
  Presidential Campaign." *American Politics Research* 39(3): 534–556.
- Enos, R. D. and Fowler, A. (2018). "Aggregate Effects of Large-Scale Campaigns
  on Voter Turnout." *Political Science Research and Methods* 6(4): 733–751.
- Gimpel, J. G., Kaufmann, K. M. and Pearson-Merkowitz, S. (2007). "Battleground
  States versus Blackout States: The Behavioral Implications of Modern
  Presidential Campaigns." *The Journal of Politics* 69(3): 786–797.

**Further reading**

- Shaw, D. R. (1999). "The Effect of TV Ads and Candidate Appearances on
  Statewide Presidential Votes, 1988–96." *American Political Science Review*
  93(2): 345–361.
- Shaw, D. R. (2006). *The Race to 270: The Electoral College and the Campaign
  Strategies of 2000 and 2004.* University of Chicago Press.



```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
