---
title: "Where Campaigns Travelled, and Where They Did Not"
type: brief
subtitle: "Eight months of one campaign's travel log, and the half of America that never appears in it"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("campaign-visits-code.R")
```

```{r setup, include=FALSE}
```

Presidential campaigns describe themselves as national. Both tickets in 2024 said they were competing for the whole country, and both raised and spent money in every state. So where did the candidates themselves actually go?

A campaign has more money than it can spend well and more advice than it can use. It has exactly one resource nobody can manufacture: the candidate's own time. There are a fixed number of days before an election, and a person can stand in one place on each of them. A travel schedule is therefore a ranked list of a campaign's own judgments about where the presidency will be decided, kept by the people with the best polling in the country and revised daily. It is a strategy document that a campaign publishes by carrying it out.

This brief reads the 2024 schedule and asks one question with a checkable answer. How much of the United States appears in it?

## The test

The record is the Associated Press's log of the 2024 campaign trail. It holds every public appearance by the two presidential nominees, their running mates, and the president who withdrew in July, from `r format(min(v$date), "%d %B")` to `r format(max(v$date), "%d %B %Y")`. AP publishes it as the file behind an interactive graphic at <https://interactives.ap.org/embeds/ziEfd/54/>.

```{r scope}
```

Why this source and not a government filing? No agency collects candidate itineraries. A campaign's internal schedule is shown to nobody, and the Federal Election Commission records what a campaign spent without recording where the candidate stood while spending it. A newsroom's log is not a second-best substitute for a filing here. It is the only witness there is.

That makes it a record of a familiar kind: something written down by people who were doing another job. Reporters covering a beat noted where the candidates were, and somebody decided the accumulation was worth publishing. [The section introduction](../../05-ordinary-people/admin-records-source/admin-records-source-brief.html) is about that whole family of files, and about what you gain and lose when nobody involved was measuring anything.

Nothing compelled this one. Campaign-finance data exists because federal law requires the filing, and a filer who leaves something out can be penalised. This file exists at AP's discretion, and it carries no codebook, no correction log and no signature.

## The data

One row is one public appearance, by one person, in one city, on one day. It is not a day of campaigning, not a state visit, and not an advertising buy.

```{r one-record}
```

That definition does more work than it looks. A rally of twenty thousand people and a fifteen-minute stop at a diner are both exactly one row, and nothing in the file separates them. Every count below is a count of occasions.

`r ncol(rawv)` columns arrive and `r length(kept)` are kept. What goes is telling: `biden-total-visits-city`, `total-state-visits`, `county-text`, `reg-voters`. None of those is a fact about the event on the row. They are the contents of the box AP shows a reader who hovers over a dot on its map. This is a graphic's payload, and the rows are events only incidentally.

The columns kept are the ones describing the event: `date`, `city`, `county`, `state`, `postal`, `lat`, `long`, `fips`, `candidate`, `event_type` and `notes`. Dates arrive looking like `r r1[["date"]]`, which sorts as text into an order no calendar recognises, so every one of them is converted to a real date. The build stops outright if a single row fails to convert. One silent missing value is easier to publish than to find.

The third decision is the one that makes this brief. **A state that received no attention is not a row in this file.** The log records events, and a place with no events leaves nothing to record. Counted straight, the file holds `r length(unique(v$state))` states, while the country has `r nrow(m)` jurisdictions that cast electoral votes.

The `r nrow(z)` places nobody visited enter here through a second table. `pres2024_states.csv` carries the 2024 presidential result for every state and the District: `state`, `abbrev`, `ev` for electoral votes, `margin`, `winner`, and a grid position for the map below. The two tables are joined so that every state is kept whether or not it matched a visit, and the empty side is filled with zero rather than left blank. Absence does not write itself down.

> **Before you read on, commit to a number.** Order all `r nrow(m)` jurisdictions
> from the least attention each got per electoral vote to the most, then walk up
> that order until you hold half of the electoral college. That bottom half
> starts with the `r nrow(z)` places that got nothing, which are
> `r pc(100 * zero_ec)`% of the college by themselves. **What share of the
> `r n(nrow(v))` stops did the bottom half receive?** Write a percentage down
> now, while the only thing you have to go on is the question.

## What it says about democracy

```{r pointmap-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

```{r pointmap-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Every logged stop of the 2024 campaign, drawn where it happened.** One mark per appearance, `r n(nrow(v))` in all, coloured by ticket. A dot map fits this question because the question is about where a fixed number of days went, and a map lets you read the empty space as easily as the crowded space. The clustering is not population: the blank areas hold several of the largest cities in the country.

```{r top-states}
```

**Seven states holding `r pc(100 * sum(t7$ev) / sum(m$ev))`% of the electoral college absorbed `r pc(100 * sum(t7$visits) / nrow(v))`% of the campaign.** Pennsylvania alone took `r m$visits[m$state == "Pennsylvania"]` stops in `r length(unique(v$city[v$state == "Pennsylvania"]))` different towns. That is a stop somewhere in one state every `r pc(as.numeric(max(v$date) - min(v$date)) / m$visits[m$state == "Pennsylvania"], 1)` days, for eight months.

```{r zeros}
```

**`r nrow(z)` of `r nrow(m)` jurisdictions received not one stop** from any of the five people on the two tickets. Between them they hold `r sum(z$ev)` electoral votes, `r pc(100 * sum(z$ev) / sum(m$ev))`% of the total. Not a rally, not a fundraiser, not a fifteen-minute stop at a diner.

```{r cartogram-static, eval=!knitr::is_html_output(), fig.height=4.1}
```

```{r cartogram-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The country as a grid of equal tiles, shaded by the stops each place received.** One tile per state and the District, set roughly where it belongs, so no state's area on the page depends on its size on the ground. The `r nrow(z)` palest tiles are the places nobody came to.

A grid shows which states were skipped. It cannot show what they were owed, because every tile is the same size whatever the state's electoral weight. Put both quantities on one pair of axes and the disproportion becomes a single shape.

```{r lorenz-static, eval=!knitr::is_html_output(), fig.height=4.0}
```

```{r lorenz-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. How the campaign's time was spread across the electoral college.** Each point adds one state, in order of the attention it got per electoral vote. The horizontal axis is the running share of electoral votes and the vertical axis the running share of stops. The dashed diagonal is the country a campaign would visit if it treated every electoral vote alike, and the red segment along the bottom is the `r nrow(z)` states that got nothing.

**The least-visited half of the electoral college received `r pc(100 * half_y)`% of all campaign stops.** Most readers guess somewhere in the teens, reasoning that the states just above the empty ones must have picked up a respectable remainder. They did not. The whole bottom half of the country, weighted the way the Constitution weights it, is worth about `r pc(half_y * nrow(v), 0)` of the `r n(nrow(v))` stops.

The gap between the curve and the diagonal has a standard summary. A Gini coefficient of 0 would mean every electoral vote drew the same attention, and 1 that a single state took everything. Here it is `r r2(gini)`, a figure that would be reported as extreme if the thing being divided were income rather than a candidate's time.

Nothing in the Constitution produces that shape. It comes from one rule, adopted state by state and written nowhere in federal law: the winner of a state takes all of its electoral votes. Under that rule a vote is worth chasing only where it might change which candidate reaches 270.

You can watch the rule work in the file. Start with the obvious guess, that candidates go where the voters are. A correlation is one number between -1 and 1 saying how tightly two quantities rise and fall together, and for stops against electoral votes it is `r r2(cor(m$visits, m$ev))`. That is close to nothing. Closeness does better and in the right direction, at `r r2(cor(m$visits, m$absm))`, and it still leaves most of the map unexplained.

Now multiply the two. Give each state its electoral votes if it finished within five points, and zero otherwise. The index has no tuning in it.

```{r index}
```

**`r r2(cor(m$visits, m$prize))`**, from no information that was not already in the two weak measures. Each one is nearly useless alone and powerful once the other is held still. A campaign is not asking where the voters are, and it is not asking where the race is close. It is asking where a day of the candidate's time might change who wins. **That is winner-take-all expressed as a travel schedule.**

Two rows show what the columns still hide. Washington, DC was decided by `r pc(m$absm[m$state == "District of Columbia"], 2)` points and logged `r m$visits[m$state == "District of Columbia"]` stops anyway, none of them about its `r m$ev[m$state == "District of Columbia"]` electoral votes. They were addresses to national conventions of interest groups, in the city where the donors and the press corps already are. Removing that single row moves the correlation between stops and closeness from `r r2(cor(m$visits, m$absm))` to `r r2(cor(mn$visits, mn$absm))`.

Nebraska is the opposite mistake. It is a safe state, decided by `r pc(m$absm[m$state == "Nebraska"], 2)` points, and it got `r m$visits[m$state == "Nebraska"]` stops, both of them in Omaha. Nebraska does not award its electoral votes winner-take-all. One goes to the winner of the Omaha-centred district, and that district was genuinely contested, but the results table has a row per state and no row for a district. The prize being chased appears nowhere in the data.

Whether any of this matters to a citizen depends on whether attention does anything, and there is evidence that it does. Enos and Fowler (2018) compared targeted and untargeted states in 2012. They estimated that presidential campaigning raised turnout in the heavily targeted states by about 7 to 8 percentage points, mostly through ordinary ground organising. On that account the `r nrow(z)` blank tiles are not a travel budget. They are a description of where political life is thinner.

## What this brief cannot tell you

**It cannot tell you how much campaigning happened.** An event is an event, with no crowd size, no hours spent and no money attached. An arena and a diner are one row each, which is the largest single limit of the file.

**It cannot tell you the log is complete.** Nobody is penalised for an omission from a newsroom's file, and the running mates were watched less closely than the principals. The log also ends on `r format(max(v$date), "%d %B")`, the day before the election, so every count here is a floor.

**It cannot be cited without a version number.** AP kept publishing after the election: version `r OLD$version` held `r n(OLD$rows)` stops and version `r NEW$version` holds `r n(NEW$rows)`. `r added_back` of the additions were backfilled into days the older file already claimed to cover. A public dataset with no statutory deadline has no moment at which it is finished.

**It cannot see anything smaller than a state, and it is one election.** Maine and Nebraska award electoral votes by district, and no row here does. Only `r sum(m$absm < 10)` states finished within ten points in 2024, and nothing in this file shows that 2024 was typical.

**It measures no consequence.** There is no turnout column and no vote column. What the visits produced is a different question, and it needs a different file.

## What you should have learned

- Seven states took `r pc(100 * sum(t7$visits) / nrow(v))`% of the `r n(nrow(v))` logged stops while holding `r pc(100 * sum(t7$ev) / sum(m$ev))`% of the electoral college, and `r nrow(z)` of `r nrow(m)` jurisdictions received none at all.
- The least-visited half of the electoral college got `r pc(100 * half_y)`% of the campaign, a concentration of `r r2(gini)` on the Gini scale.
- Neither size nor closeness orders the map alone, at `r r2(cor(m$visits, m$ev))` and `r r2(cor(m$visits, m$absm))`. Their product does, at `r r2(cor(m$visits, m$prize))`, which is winner-take-all doing exactly what it rewards.
- The Electoral College does not reward smallness. Nevada and New Hampshire are small and were visited; Wyoming, Vermont and the Dakotas are small and were not.
- A file that records events cannot record a place where nothing happened. The `r nrow(z)` empty states exist in this brief only because a second table supplied them, and because the join kept every row on both sides.

## Extensions

- `campaign_visits_2024.csv` names a `state` on every event. Count the events per state, bring in `pres2024_states.csv`, and compare each count against `ev` and `margin`. Which states got more attention than their weight, and which got less?
- `campaign_visits_2024.csv` also has `city`. Sort the cities by how many stops each received. Is a battleground state campaigned in all over, or in three or four places?
- `campaign_visits_2024.csv` has `candidate` and `date`. Count the stops per candidate per month. Do the two tickets travel at the same rate, and does either change gear late?
- `ap_snapshots.csv` records two versions of the same log, with different `rows`, a different `last_event`, and a different `rows_in_shared_window`. Work out what the earlier version was missing, and what you would have published from it.
- **Stretch.** Outside trackers publish state-level television advertising totals for 2024. Set a state's advertising against its stop count, and ask whether money and time are spent in the same places.
- **Stretch.** Maine and Nebraska award electoral votes by congressional district. Rebuild the index with those districts as units of their own, and see how much of the Nebraska result it absorbs.

## Sources

**Data**

- Associated Press, "Tracking the 2024 presidential campaign trail," version
  `r NEW$version`\. <https://interactives.ap.org/embeds/ziEfd/54/> — public
  campaign appearances by both tickets, `r format(min(v$date), "%d %B")` to
  `r format(max(v$date), "%d %B %Y")`\. Version `r OLD$version`, against which
  the earlier release of this brief was built, is kept for comparison in
  `ap_snapshots.csv`.
- State-level 2024 presidential returns and electoral votes, from Jason Timm's
  `PresElectionResults` compilation
  (<https://github.com/jaytimm/PresElectionResults>), which rests on the
  fifty-one certified state canvasses.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`ap_snapshots.csv`](data/derived/ap_snapshots.csv), [`campaign_visits_2024.csv`](data/derived/campaign_visits_2024.csv), [`pres2024_states.csv`](data/derived/pres2024_states.csv)

**Academic literature**

- Enos, R. D. and Fowler, A. (2018). "Aggregate Effects of Large-Scale Campaigns
  on Voter Turnout." *Political Science Research and Methods* 6(4): 733–751.
- Chen, L. J. and Reeves, A. (2011). "Turning Out the Base or Appealing to the
  Periphery? An Analysis of County-Level Candidate Appearances in the 2008
  Presidential Campaign." *American Politics Research* 39(3): 534–556.
- Gimpel, J. G., Kaufmann, K. M. and Pearson-Merkowitz, S. (2007). "Battleground
  States versus Blackout States: The Behavioral Implications of Modern
  Presidential Campaigns." *The Journal of Politics* 69(3): 786–797.
- Shaw, D. R. (2006). *The Race to 270: The Electoral College and the Campaign
  Strategies of 2000 and 2004.* University of Chicago Press.

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
