---
title: "Draw the Line Somewhere Else"
subtitle: "84-355 Democracy's Data | The modifiable areal unit problem, on one county's census blocks"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)

D  <- "data"
fc <- read.csv(file.path(D, "facts.csv"),          stringsAsFactors = FALSE)
ls_ga <- read.csv(file.path(D, "ladder_state.csv"), stringsAsFactors = FALSE)
lf <- read.csv(file.path(D, "ladder_fulton.csv"),  stringsAsFactors = FALSE)
dn <- read.csv(file.path(D, "dens_fulton.csv"),    stringsAsFactors = FALSE)
mu <- read.csv(file.path(D, "map_units.csv"),      stringsAsFactors = FALSE)
sm <- read.csv(file.path(D, "zoning_summary.csv"), stringsAsFactors = FALSE)
sw <- read.csv(file.path(D, "zoning_sweep.csv"),   stringsAsFactors = FALSE)
pu <- read.csv(file.path(D, "zoning_plan_units.csv"), stringsAsFactors = FALSE)
co <- read.csv(file.path(D, "county_outline.csv"), stringsAsFactors = FALSE)
mv <- read.csv(file.path(D, "map_vals.csv"),                 # GEOIDs are text
               colClasses = c(unit = "character"), stringsAsFactors = FALSE)

FV <- function(k) { v <- fc$value[fc$name == k]; if (!length(v)) stop("no fact: ", k); v }
FN <- function(k) as.numeric(FV(k))
n  <- function(x) format(round(as.numeric(x)), big.mark = ",")
pc <- function(x, k = 1) formatC(as.numeric(x), format = "f", digits = k)
LF <- function(l, cc) lf[[cc]][lf$level == l]
LG <- function(l, cc) ls_ga[[cc]][ls_ga$level == l]
LEVS <- lf$level
RAMP <- colorRampPalette(c("#f7f7f7", "#d9e6f2", "#8fbedd", "#3f8dc0",
                           "#1f5d8c", "#12395a"))(101)
RX <- range(mu$x); RY <- range(mu$y)

onepath <- function(X, Y) {
  dx <- X[-1] - X[-length(X)]; dy <- Y[-1] - Y[-length(Y)]
  seg <- paste0(dx, ifelse(dy < 0, "", ","), dy)
  sep <- c("", ifelse(substr(seg[-1], 1, 1) == "-", "", " "))
  paste0("M", X[1], ",", Y[1], "l", paste0(sep, seg, collapse = ""), "Z")
}
ringpaths <- function(d, by, sx, sy) {
  k <- interaction(d[[by]], d$part, drop = TRUE)
  p <- vapply(split(d, k), function(z) {
    X <- round(sx(z$x)); Y <- round(sy(z$y))
    keep <- c(TRUE, X[-1] != X[-length(X)] | Y[-1] != Y[-length(Y)])
    X <- X[keep]; Y <- Y[keep]
    if (length(X) < 3) paste0("M", X[1], ",", Y[1], "h1v1h-1Z") else onepath(X, Y)
  }, character(1))
  p[nzchar(p)]
}
jstr <- function(x) paste0("[", paste0('"', x, '"', collapse = ","), "]")
jnum <- function(x) paste0("[", paste0(x, collapse = ","), "]")
drawrings <- function(d, by, fill, border = NA, lwd = 0.4) {
  k <- interaction(d[[by]], d$part, drop = TRUE)
  for (lv in levels(k)) {
    z <- d[k == lv, ]
    polygon(z$x, z$y, col = fill[[as.character(z[[by]][1])]], border = border, lwd = lwd)
  }
}

# ---- render every data.frame in this document as a TABLE, not code output ----
# A data.frame printed the ordinary way comes out as a "#>"-prefixed code block,
# which reads as machinery rather than as a result. Registering knit_print for
# data.frame turns all of them into real tables in HTML and PDF alike.
knit_print.data.frame <- function(x, ...) {
  nm <- names(x)
  nm <- gsub("_", " ", nm)
  nm <- sub("^(.)", "\\U\\1", nm, perl = TRUE)
  knitr::knit_print(knitr::kable(x, col.names = nm, row.names = FALSE), ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

Everything here runs as given. **You will not write any code.** Your job is to
read what comes out and say what it means.

> **Read `areal-units-brief.Rmd` first.** This lab runs the same numbered steps
> with the code exposed. The chunks that *compute* are shown; the chunks that
> only *draw* are folded away, because their job is pictures, not arithmetic.

***

# The claim under examination

> *"Fulton County, Georgia is `r pc(FV("ful_black_pct"))`% Black."*

True. Also a fact about a shape somebody drew. This lab holds the census fixed —
one day, one count, `r n(FV("ful_pop"))` people — and changes only the shape.

Two effects, and the whole lab depends on keeping them apart:

- **Scale effect** — same points, bigger boxes. Blocks to block groups to tracts
  to counties.
- **Zoning effect** — same number of boxes, same size boxes, different shapes.
  This is what redistricting is.

# Step 1 — The same people, all the way down

```{r step1}
c(ga_blocks   = FN("ga_blocks"),
  ga_pop      = FN("ga_pop"),
  ful_blocks  = FN("ful_blocks"),
  ful_blocks_with_people = FN("ful_blocks_pop"),
  ful_pop     = FN("ful_pop"),
  ful_black   = FN("ful_black"))
c(ful_black_pct = FN("ful_black_pct"), ful_black_vap_pct = FN("ful_bvap_pct"))
```

**`r n(FV("ful_pop"))` people on `r n(FV("ful_blocks"))` blocks**, of which
`r n(FV("ful_blocks_pop"))` have anyone living on them. "Black" here means **any
part Black** — any of the 32 of the census's 63 race combinations that include
Black or African American, which is the definition voting-rights litigation
uses.

Every statistic below is **population-weighted**, so a block holding three people
does not count as much as a tract holding six thousand.

# Step 2 — Census geography is a string, not a map

```{r step2}
eg <- mv[mv$lev == "b", ]
eg <- eg$unit[which.max(eg$pop)]      # the most populous Fulton block
c(block = eg, block_group = substr(eg, 1, 12), tract = substr(eg, 1, 11),
  county = substr(eg, 1, 5), state = substr(eg, 1, 2))
```

**Climbing the census hierarchy is `substr()`.** No overlay, no intersection, no
interior points, no assumption about how people are spread inside a polygon. So
nothing that changes in Steps 3–5 can be blamed on a geometric approximation:
there isn't one.

GEOIDs are **character**. Read that block as a number and it becomes
`1.312101e+14`; read a tract as a number and it silently loses its leading zero.
Every read in this document that touches an identifier pins the column.

# Step 3 — The scale effect, seen

```{r scale-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
PW <- 208L; GAP <- 10L; TOP <- 42L; BOTPAD <- 40L
s  <- PW / diff(RX); PH <- round(diff(RY) * s)
W  <- 4L * PW + 3L * GAP; H <- TOP + PH + BOTPAD
KEYS <- c("b", "g", "t", "c")
mkpanel <- function(i) {
  off <- (i - 1L) * (PW + GAP)
  fx <- function(x) off + (x - RX[1]) * s
  fy <- function(y) TOP + (RY[2] - y) * s
  v  <- mv[mv$lev == KEYS[i], ]
  P  <- ringpaths(mu[mu$lev == KEYS[i], ], "uid", fx, fy)
  uid <- as.integer(sub("\\.[0-9]+$", "", names(P)))
  ci  <- match(v$fill[match(uid, v$uid)], RAMP) - 1L
  ci[is.na(ci)] <- -1L
  base <- ringpaths(mu[mu$lev == "c", ], "uid", fx, fy)
  paste0("[", jstr(c(base, P)), ",", jnum(c(rep(-1L, length(base)), ci)), "]")
}
PAN <- vapply(seq_along(KEYS), mkpanel, character(1))
sub1 <- paste0(vapply(LEVS, function(l) n(LF(l, "units")), character(1)),
               ifelse(LEVS == "County", " unit", " units"))
sub2 <- paste0("median ", vapply(LEVS, function(l) n(LF(l, "median_pop")),
                                 character(1)), " people")
sub3 <- paste0("spread ", vapply(LEVS, function(l) pc(LF(l, "sd_share")),
                                 character(1)), " pts")
lab <- paste0("[", paste0('{"t":"', LEVS, '","a":"', sub1, '","b":"', sub2,
                          '","c":"', sub3, '"}', collapse = ","), "]")
cat(paste0('
<div id="scl" style="margin:1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const PAN=[', paste(PAN, collapse = ","), '],L=', lab, ',R=', jstr(RAMP), ';
const W=', W, ',H=', H, ',PW=', PW, ',GAP=', GAP, ',TOP=', TOP, ',PH=', PH, ';
const svg=d3.select("#scl").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const col=i=>i<0?"#e6e6e6":R[i];
PAN.forEach((P,i)=>{
  const off=i*(PW+GAP);
  svg.append("g").selectAll("path").data(P[0]).join("path").attr("d",d=>d)
    .attr("fill",(d,j)=>col(P[1][j])).attr("stroke","none");
  svg.append("text").attr("x",off+PW/2).attr("y",15).attr("text-anchor","middle")
    .attr("font-size","13px").attr("font-weight","600").attr("fill","#222").text(L[i].t);
  [[L[i].a,27],[L[i].b,37]].forEach(([t,y])=>
    svg.append("text").attr("x",off+PW/2).attr("y",y).attr("text-anchor","middle")
      .attr("font-size","9.5px").attr("fill","#777").text(t));
  svg.append("text").attr("x",off+PW/2).attr("y",TOP+PH+12).attr("text-anchor","middle")
    .attr("font-size","9.5px").attr("fill","#777").text(L[i].c);
});
const KW=210,KX=(W-KW)/2,KY=TOP+PH+21;
R.forEach((c,i)=>svg.append("rect").attr("x",KX+i*KW/101).attr("y",KY)
  .attr("width",KW/101+0.5).attr("height",7).attr("fill",c));
[[0,"0%"],[50,"50%"],[100,"100%"]].forEach(([v,t])=>
  svg.append("text").attr("x",KX+v*KW/100).attr("y",KY+8).attr("text-anchor","middle")
    .attr("dy","0.9em").attr("font-size","8px").attr("fill","#888").text(t));
svg.append("text").attr("x",KX-8).attr("y",KY+6).attr("text-anchor","end")
  .attr("font-size","8.5px").attr("fill","#888").text("share of the unit that is Black");
})();
</script>
'))
```

```{r scale-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.5}
layout(matrix(c(1, 2, 3, 4, 5, 5, 5, 5), 2, 4, byrow = TRUE), heights = c(8, 1))
op <- par(mar = c(1.6, 0.2, 3.0, 0.2), xpd = NA)
for (i in seq_along(LEVS)) {
  lv <- c("b", "g", "t", "c")[i]
  v  <- mv[mv$lev == lv, ]
  fill <- as.list(v$fill); names(fill) <- as.character(v$uid)
  plot(NA, xlim = RX, ylim = RY, asp = 1, axes = FALSE, ann = FALSE)
  polygon(co$x, co$y, col = "#e6e6e6", border = NA)
  drawrings(mu[mu$lev == lv, ], "uid", fill, border = NA)
  title(LEVS[i], cex.main = 1.0, line = 1.9)
  mtext(paste0(n(LF(LEVS[i], "units")),
               if (LEVS[i] == "County") " unit" else " units"),
        side = 3, line = 0.95, cex = 0.55, col = "#777")
  mtext(paste0("median ", n(LF(LEVS[i], "median_pop")), " people"),
        side = 3, line = 0.25, cex = 0.55, col = "#777")
  mtext(paste0("spread ", pc(LF(LEVS[i], "sd_share")), " pts"),
        side = 1, line = 0.2, cex = 0.55, col = "#777")
}
par(mar = c(0, 0.2, 0, 0.2))
plot(NA, xlim = c(0, 1), ylim = c(0, 1), axes = FALSE, ann = FALSE)
bx <- seq(0.40, 0.62, length.out = 102)
rect(bx[-102], 0.42, bx[-1], 0.72, col = RAMP, border = NA)
text(c(0.40, 0.51, 0.62), 0.42, c("0%", "50%", "100%"), pos = 1,
     cex = 0.52, col = "#888", offset = 0.25)
text(0.385, 0.57, "share of the unit that is Black", pos = 2, cex = 0.52, col = "#888")
par(op); layout(1)
```

`r n(LF("Census block", "units"))` shapes on the left, one shape on the right,
**exactly the same `r n(FV("ful_pop"))` people.**

# Step 4 — The scale effect, measured

```{r ladder}
lf[, c("level", "units", "median_pop", "mean_share", "sd_share",
       "p10_share", "p90_share", "units_majority")]
```

The **mean is constant at `r pc(LF("Census block", "mean_share"), 2)`% on every
rung** — it has to be, the same people are being averaged. That is the control.
Everything else moves:

```{r ladder-moves}
c(sd_blocks = LF("Census block", "sd_share"),
  sd_tracts = LF("Census tract", "sd_share"),
  sd_county = LF("County", "sd_share"),
  p10_blocks = LF("Census block", "p10_share"),
  p10_tracts = LF("Census tract", "p10_share"),
  p90_blocks = LF("Census block", "p90_share"),
  p90_tracts = LF("Census tract", "p90_share"))
```

Statewide the same climb goes further, because Georgia contains between-county
variation that Fulton cannot:

```{r ladder-state}
ls_ga[, c("level", "units", "sd_share", "max_share", "units_majority",
          "pct_black_in_majority", "corr_child")]
```

Two results in that table are worth stopping for.

**No county in Georgia is as Black as thousands of its blocks are**:
`r pc(LG("Census block", "max_share"))`% falls to
`r pc(LG("County", "max_share"))`%. And the **share of Black Georgians who live
in a majority-Black unit** falls from `r pc(FV("ga_maj_block"))`% (blocks) to
`r pc(FV("ga_maj_tract"))`% (tracts) to `r pc(FV("ga_maj_county"))`% (counties).
Nobody moved.

**The correlation with the under-18 share changes sign** —
`r pc(LG("Census block", "corr_child"), 3)` across blocks,
`r pc(LG("County", "corr_child"), 3)` across counties. A correlation is a fact
about people *and* the boxes you sorted them into.

# Step 5 — Watch the distribution collapse

```{r dens-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
DL <- c("Census block", "Block group", "Census tract")
COLS <- c("#9ec6e0", "#3f8dc0", "#12395a")
W <- 900; H <- 330; L <- 54; R <- 22; T0 <- 46; B <- 54
ymax <- max(dn$y[dn$level %in% DL], na.rm = TRUE)
ser <- vapply(seq_along(DL), function(i) {
  z <- dn[dn$level == DL[i], ]
  paste0('{"c":"', COLS[i], '","p":[',
         paste0("[", formatC(z$x, format = "f", digits = 1), ",",
                formatC(z$y, format = "f", digits = 5), "]", collapse = ","), ']}')
}, character(1))
leg <- vapply(seq_along(DL), function(i)
  paste0('{"n":"', DL[i], '","c":"', COLS[i], '","s":"SD ',
         pc(LF(DL[i], "sd_share")), ' pts"}'), character(1))
cat(paste0('
<div id="dns" style="margin:1em 0"></div>
<!-- d3 v7 is loaded once, by the first figure above -->
<script>
(function(){
const S=[', paste(ser, collapse = ","), '],LG=[', paste(leg, collapse = ","), '];
const W=', W, ',H=', H, ',L=', L, ',R=', R, ',T0=', T0, ',B=', B, ',YM=', ymax, ';
const CV=', formatC(LF("County", "mean_share"), format = "f", digits = 3), ';
const CVL="the County: one value, ', pc(LF("County", "mean_share"), 2), '%";
const svg=d3.select("#dns").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const x=d3.scaleLinear().domain([0,100]).range([L,W-R]);
const y=d3.scaleLinear().domain([0,YM*1.06]).range([H-B,T0]);
svg.append("line").attr("x1",x(50)).attr("x2",x(50)).attr("y1",T0-6).attr("y2",H-B)
  .attr("stroke","#C41230").attr("stroke-width",1.2).attr("stroke-dasharray","4 3");
svg.append("text").attr("x",x(50)+5).attr("y",T0+4).attr("font-size","12px")
  .attr("fill","#C41230").attr("font-weight","600").text("50%");
const ar=d3.area().x(d=>x(d[0])).y0(y(0)).y1(d=>y(d[1])).curve(d3.curveBasis);
const ln=d3.line().x(d=>x(d[0])).y(d=>y(d[1])).curve(d3.curveBasis);
S.forEach(s=>{svg.append("path").attr("d",ar(s.p)).attr("fill",s.c).attr("fill-opacity",0.16);
  svg.append("path").attr("d",ln(s.p)).attr("fill","none").attr("stroke",s.c)
    .attr("stroke-width",2.2);});
svg.append("line").attr("x1",x(CV)).attr("x2",x(CV)).attr("y1",T0-6).attr("y2",H-B)
  .attr("stroke","#b3651a").attr("stroke-width",2.6);
// The label text is formatted in R, not by JS. R rounds half to even and
// JavaScript rounds half up, so 44.775 prints as 44.78 in the PDF and 44.77
// in the browser if each side formats it for itself.
svg.append("text").attr("x",x(CV)+7).attr("y",T0+(H-B-T0)*0.40).attr("text-anchor","start")
  .attr("font-size","13px").attr("font-weight","600").attr("fill","#b3651a").text(CVL);
d3.range(0,101,10).forEach(v=>svg.append("text").attr("x",x(v)).attr("y",H-B+18)
  .attr("text-anchor","middle").attr("font-size","12px").attr("fill","#aaa").text(v+"%"));
svg.append("line").attr("x1",L).attr("x2",W-R).attr("y1",H-B).attr("y2",H-B).attr("stroke","#ddd");
svg.append("text").attr("x",L).attr("y",H-10).attr("font-size","12.5px").attr("fill","#888")
  .text("share of the unit that is Black");
const lg=svg.append("g").attr("transform","translate("+(W-R-210)+","+(T0+14)+")");
LG.forEach((d,i)=>{
  lg.append("rect").attr("y",i*20-9).attr("width",16).attr("height",4).attr("fill",d.c);
  lg.append("text").attr("x",23).attr("y",i*20-3).attr("font-size","13px")
    .attr("fill","#333").text(d.n+"  ("+d.s+")");});
})();
</script>
'))
```

```{r dens-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.4}
DL <- c("Census block", "Block group", "Census tract")
COLS <- c("#9ec6e0", "#3f8dc0", "#12395a")
ymax <- max(dn$y[dn$level %in% DL], na.rm = TRUE)
par(mar = c(3.4, 3.0, 2.4, 1.0))
plot(NA, xlim = c(0, 100), ylim = c(0, ymax * 1.10), axes = FALSE, ann = FALSE)
abline(v = 50, col = "#C41230", lty = 2, lwd = 1.1)
for (i in seq_along(DL)) {
  z <- dn[dn$level == DL[i], ]
  polygon(c(z$x, rev(z$x)), c(z$y, rep(0, nrow(z))),
          col = adjustcolor(COLS[i], 0.16), border = NA)
  lines(z$x, z$y, col = COLS[i], lwd = 2)
}
cv <- LF("County", "mean_share")
abline(v = cv, col = "#b3651a", lwd = 2.2)
text(cv + 1.8, ymax * 0.60, sprintf("the County: one value, %.2f%%", cv),
     adj = 0, cex = 0.6, font = 2, col = "#b3651a")
axis(1, seq(0, 100, 10), paste0(seq(0, 100, 10), "%"), cex.axis = 0.6,
     col = "#ddd", col.axis = "#999", tck = -0.02, mgp = c(2, 0.3, 0))
mtext("share of the unit that is Black", side = 1, line = 1.7, cex = 0.6, col = "#888")
mtext("share of Fulton's people in a unit of that Black share", side = 2,
      line = 0.6, cex = 0.6, col = "#888")
legend("topleft", paste0(DL, "   (SD ", vapply(DL, function(l) pc(LF(l, "sd_share")),
       character(1)), " pts)"), col = COLS, lwd = 2, bty = "n", cex = 0.62)
title(sprintf("Same people, four groupings, common bandwidth of %s points",
              FV("dens_bw")), cex.main = 0.85, adj = 0, line = 0.8)
```

**Two humps, and they stay two humps.** Fulton is segregated at a spatial scale
*larger than a census tract*, so climbing three rungs barely touches it. The
textbook version of the scale effect promises smooth collapse; this is a
counterexample, and the reason is worth holding on to: **how much the scale
effect bites depends on how big the real spatial structure is compared with the
unit.**

# Step 6 — The zoning effect: same size units, different shapes

Fulton elects a Board of Commissioners from `r FV("K")` districts. So take the
same blocks and cut them into `r FV("K")` units of `r n(FV("quota"))` people
each — `r n(FV("n_plans"))` different ways, from two rules, each rotated through
`r FV("n_angles")` one-degree steps. **Neither rule can see race.**

```{r sweep}
c(plans = nrow(sm), units_per_plan = FN("K"), people_per_unit = FN("quota"),
  worst_population_deviation_pct = max(sm$max_dev_pct))
table(family = sm$family, majority_Black_units = sm$n_majority)
```

**`r FV("maj_min")` to `r FV("maj_max")`.** `r n(FV("plans_zero_maj"))` plans put
no unit over 50%. `r n(sum(sm$n_majority == FN("maj_max")))` put four over.
`r n(FV("plans_two_maj"))` land on exactly two.

```{r zone-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
ord <- order(sm$n_majority, sm$max_share)
key <- paste(sm$family[ord], sm$angle[ord]); swk <- paste(sw$family, sw$angle)
rows <- vapply(seq_along(key), function(i)
  paste0("[", i - 1L, ",", paste0(formatC(sort(sw$share[swk == key[i]]),
         format = "f", digits = 1), collapse = ","), "]"), character(1))
i_lo <- which(key == paste(FV("lo_family"), FV("lo_angle"))) - 1L
i_hi <- which(key == paste(FV("hi_family"), FV("hi_angle"))) - 1L
lab_lo <- paste0(FV("lo_maj"), " majority-Black units · ", FV("lo_family"),
                 " rule at ", FV("lo_angle"), "°")
lab_hi <- paste0(FV("hi_maj"), " majority-Black units · ", FV("hi_family"),
                 " rule at ", FV("hi_angle"), "°")
brk <- cumsum(table(factor(sm$n_majority[ord], levels = sort(unique(sm$n_majority)))))
blab <- paste0("[", paste0('{"y":', as.integer(brk), ',"t":"', names(brk),
        ifelse(names(brk) == "1", " majority-Black unit", " majority-Black units"),
        '","n":', as.integer(table(sm$n_majority)[names(brk)]), '}', collapse = ","), "]")
cat(paste0('
<div id="zon" style="margin:1em 0"></div>
<!-- d3 v7 is loaded once, by the first figure above -->
<script>
(function(){
const P=[', paste(rows, collapse = ","), '],BR=', blab, ';
const NR=P.length,LO=', i_lo, ',HI=', i_hi, ';
const RH=1.55,L=228,R=32,T0=54,B=48;
const W=900,H=Math.round(T0+NR*RH+B);
const svg=d3.select("#zon").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const x=d3.scaleLinear().domain([0,100]).range([L,W-R]);
const yr=i=>T0+i*RH;
svg.append("rect").attr("x",x(50)).attr("y",T0-6).attr("width",W-R-x(50))
  .attr("height",NR*RH+6).attr("fill","#C41230").attr("fill-opacity",0.045);
BR.forEach(b=>{
  svg.append("line").attr("x1",L-6).attr("x2",W-R).attr("y1",yr(b.y)).attr("y2",yr(b.y))
    .attr("stroke","#ddd");
  svg.append("text").attr("x",L-12).attr("y",yr(b.y)-b.n*RH/2+4).attr("text-anchor","end")
    .attr("font-size","13px").attr("fill","#555").text(b.n+" plans → "+b.t);});
const FLAT=[];P.forEach(p=>{for(let j=1;j<7;j++)FLAT.push([p[0],p[j]]);});
svg.append("g").selectAll("circle").data(FLAT).join("circle")
  .attr("cx",d=>x(d[1])).attr("cy",d=>yr(d[0])+RH/2).attr("r",1.7)
  .attr("fill",d=>d[1]>50?"#C41230":"#3f8dc0").attr("fill-opacity",0.55);
[[LO,"', lab_lo, '"],[HI,"', lab_hi, '"]].forEach(([i,t])=>{
  svg.append("line").attr("x1",L-6).attr("x2",W-R).attr("y1",yr(i)+RH/2)
    .attr("y2",yr(i)+RH/2).attr("stroke","#111").attr("stroke-width",1);
  const p=P[i];
  for(let j=1;j<7;j++)
    svg.append("circle").attr("cx",x(p[j])).attr("cy",yr(i)+RH/2).attr("r",3.6)
      .attr("fill",p[j]>50?"#C41230":"#1f5d8c").attr("stroke","#fff").attr("stroke-width",1);
  svg.append("text").attr("x",W-R).attr("y",yr(i)-7).attr("text-anchor","end")
    .attr("font-size","13px").attr("font-weight","600").attr("fill","#111").text(t);});
svg.append("line").attr("x1",x(50)).attr("x2",x(50)).attr("y1",T0-14).attr("y2",T0+NR*RH)
  .attr("stroke","#C41230").attr("stroke-width",1.4);
svg.append("text").attr("x",x(50)).attr("y",T0-20).attr("text-anchor","middle")
  .attr("font-size","14px").attr("font-weight","600").attr("fill","#C41230").text("50%");
d3.range(0,101,10).forEach(v=>svg.append("text").attr("x",x(v)).attr("y",T0+NR*RH+20)
  .attr("text-anchor","middle").attr("font-size","12px").attr("fill","#aaa").text(v+"%"));
svg.append("text").attr("x",L).attr("y",T0+NR*RH+40).attr("font-size","13px").attr("fill","#888")
  .text("Black share of a unit · every unit of every plan · "+(6*NR).toLocaleString()
        +" units in "+NR+" plans");
})();
</script>
'))
```

```{r zone-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=5.6}
ord <- order(sm$n_majority, sm$max_share)
key <- paste(sm$family[ord], sm$angle[ord]); swk <- paste(sw$family, sw$angle)
par(mar = c(3.6, 9.6, 2.6, 1.0))
plot(NA, xlim = c(0, 100), ylim = c(nrow(sm), 0), axes = FALSE, ann = FALSE)
rect(50, 0, 100, nrow(sm), col = adjustcolor("#C41230", 0.045), border = NA)
for (i in seq_along(key)) {
  v <- sw$share[swk == key[i]]
  points(v, rep(i, length(v)), pch = 19, cex = 0.24,
         col = adjustcolor(ifelse(v > 50, "#C41230", "#3f8dc0"), 0.55))
}
brk <- cumsum(table(factor(sm$n_majority[ord], levels = sort(unique(sm$n_majority)))))
cnt <- as.integer(table(sm$n_majority)[names(brk)])
segments(0, brk, 100, brk, col = "#dddddd")
text(-2, brk - cnt / 2, paste0(cnt, " plans -> ", names(brk),
     ifelse(names(brk) == "1", " majority-Black unit", " majority-Black units")),
     adj = 1, cex = 0.6, col = "#555", xpd = NA)
abline(v = 50, col = "#C41230", lwd = 1.4)
for (tg in c("lo", "hi")) {
  i <- which(key == paste(FV(paste0(tg, "_family")), FV(paste0(tg, "_angle"))))
  v <- sw$share[swk == key[i]]
  segments(0, i, 100, i, col = "#111111", lwd = 0.7)
  points(v, rep(i, length(v)), pch = 21, cex = 0.62, lwd = 0.5,
         bg = ifelse(v > 50, "#C41230", "#1f5d8c"), col = "white")
  lb <- paste0(FV(paste0(tg, "_maj")), " majority-Black units - ",
               FV(paste0(tg, "_family")), " rule at ", FV(paste0(tg, "_angle")), " deg")
  yy <- if (tg == "lo") i - 5 else i - 6
  rect(100 - strwidth(lb, cex = 0.58) - 1, yy - strheight(lb, cex = 0.58) * 0.9,
       100.5, yy + strheight(lb, cex = 0.58) * 0.9, col = "white", border = NA)
  text(100, yy, lb, adj = 1, cex = 0.58, font = 2, col = "#111111")
}
axis(1, seq(0, 100, 10), paste0(seq(0, 100, 10), "%"), cex.axis = 0.6,
     col = "#ddd", col.axis = "#999", tck = -0.012, mgp = c(2, 0.3, 0))
mtext(sprintf("Black share of a unit - every unit of every plan - %s units in %d plans",
              n(6 * nrow(sm)), nrow(sm)), side = 1, line = 1.9, cex = 0.6, col = "#888")
title(sprintf("%s equal-population partitions of the same %s blocks",
              n(nrow(sm)), n(FV("ful_blocks"))), cex.main = 0.9, adj = 0, line = 0.8)
```

# Step 7 — The two extremes, unit by unit

```{r extremes}
z <- pu[pu$plan %in% c("fewest", "most"),
        c("plan", "family", "angle", "unit", "pop", "black", "share", "bvap")]
z
```

The **`r FV("lo_family")`** rule at `r FV("lo_angle")` degrees puts all
`r FV("K")` units between `r pc(FV("lo_min"))`% and `r pc(FV("lo_max"))`% —
every one of them within a couple of points of the county's own
`r pc(FV("ful_black_pct"), 2)`%, and **none of them a majority**. That is
*cracking*, arrived at with no intent at all: the slabs run across the grain of
the county, so each picks up part of the Black south-west and part of the white
north.

The **`r FV("hi_family")`** rule at `r FV("hi_angle")` degrees cuts with the
grain instead: `r pc(FV("hi_min"))`% to `r pc(FV("hi_max"))`%, and
`r FV("hi_maj")` of `r FV("K")` majority-Black.

A ruler and a protractor did this.

# Step 8 — But are they districts? The contiguity test

```{r contig}
c(plans                       = FN("n_plans"),
  strictly_contiguous_plans   = FN("plans_strict_contig"),
  effectively_contiguous_plans = FN("plans_contig"),
  threshold_pct               = FN("contig_thr"))
table(effectively_contiguous = sm$largest_piece_pct >= FN("contig_thr"),
      majority_Black_units   = sm$n_majority)
```

**Strict contiguity — every block of a unit reachable from every other — is
satisfied by `r FV("plans_strict_contig")` plans out of
`r n(FV("n_plans"))`,** and the reason is not inkblot shapes. Georgia's block
layer gives the Chattahoochee's bed, the interstate right-of-way and the airport
apron blocks of their own, and they share no boundary with the land across them.
Any cut anywhere strands a few near-empty blocks.

So the criterion with meaning is whether a unit's **people** form one connected
mass. Requiring `r FV("contig_thr")`% of every unit's population in a single
piece keeps `r n(FV("plans_contig"))` plans — and among them the answer still
runs **`r FV("contig_maj_min")` to `r FV("contig_maj_max")`** majority-Black
units. The most extreme plans do not survive; the ambiguity does. On a
`r FV("K")`-seat board that is one seat, a sixth of the government, decided by
which respectable rule you used.

```{r contig-plans}
pu[pu$plan %in% c("fewest_contig", "most_contig"),
   c("plan", "family", "angle", "unit", "pop", "share", "bvap", "pct_largest")]
```

# Step 9 — What you can now say

**You can say** that a unit's Black share is a joint fact about the population
and the unit, and give the size of the effect:
`r pc(LF("Census block", "sd_share"))` points of spread across blocks,
`r pc(LF("County", "sd_share"))` across the county.

**You can say** that aggregation destroys tails, and by how much: the share of
Black Georgians in a majority-Black unit falls from `r pc(FV("ga_maj_block"))`%
to `r pc(FV("ga_maj_county"))`% between blocks and counties, with nobody moving.

**You can say** that a correlation is a property of the grouping, and point to
one that changes sign.

**You can say** that the zoning effect is real and large: `r FV("K")` equal
units, `r n(FV("n_plans"))` race-blind ways of drawing them,
`r FV("maj_min")`–`r FV("maj_max")` majority-Black units, still
`r FV("contig_maj_min")`–`r FV("contig_maj_max")` once ill-shaped plans go.

**You cannot say** that any of these is the plan Fulton should have. Nothing here
optimises anything. **The point is not that one of the `r n(FV("n_plans"))` is
right; it is that "what is the Black share of a district here?" has
`r n(FV("n_plans"))` answers before anyone has argued about anything.**

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different county.** `data/ga_block_race.csv` holds all
`r n(FV("ga_blocks"))` Georgia blocks. Change the five-character county GEOID
from Fulton's `13121` to another county and rebuild the scale ladder. Chatham
(`13051`), Dougherty (`13095`), Forsyth (`13117`) and Clayton (`13063`) are
usefully different from each other. Does the scale effect bite harder or softer
than in Fulton, and what does that tell you about the spatial scale of
segregation there?

**Option B — A different definition of "Black."** The ladder uses `black_any`.
Rebuild it with `black_alone` and report what moves and what does not. Which
statistic is most sensitive to the definition — the mean, the spread, or the
count of majority-Black units? Which choice would you defend in court?

**Option C — A different threshold.** Fifty percent is a legal convention, not a
natural constant. Using `data/zoning_sweep.csv`, recount how many units of each
plan clear 40%, 45%, 55% and 60%, and do it with `bvap` (voting-age) as well as
`share` (total population). At which threshold does the choice of plan stop
mattering — and why is that the interesting number?

**Option D — Your own question.** `r n(FV("n_plans"))` partitions,
`r n(FV("ga_blocks"))` blocks, four rungs of census geography.

# What this lab cannot tell you

- **Whether any real plan was drawn to do this.** Nothing here has intent.
  The lab shows the zoning effect is *available*, not that it was used.
- **The true range of possible outcomes.** Two rule families are not the space of
  all equal-population partitions. `r FV("maj_min")`–`r FV("maj_max")` is a lower
  bound on the ambiguity, not the answer.
- **Anything about voters.** These are residents. Fulton is
  `r pc(FV("ful_black_pct"), 2)`% Black by population and
  `r pc(FV("ful_bvap_pct"), 2)`% by voting-age population, and neither is
  turnout.
- **Anything about individuals.** Aggregating race to areas and reasoning down
  to people is the **ecological fallacy**; see `levels-of-aggregation`, and see
  `rpv` for the one place in this course where the inference can be
  checked against an answer key.
- **How much of the block counts is noise.** The 2020 census injects
  differential-privacy noise, and it matters more at block level than at county
  level — which is itself a scale effect.

# On the discussion board

By **Friday, 11:59 p.m.**

- Step 6 shows two rules that cannot see race producing `r FV("lo_maj")` and
  `r FV("hi_maj")` majority-Black units. What, if anything, does "the algorithm
  is race-blind" establish?
- Step 4 shows a correlation changing sign between blocks and counties. If you
  read a study reporting a correlation across counties, what is the one question
  you now have to ask?
- Section 2 of the Voting Rights Act asks whether a minority group is
  "sufficiently large and geographically compact" to constitute a majority in a
  district. After this lab, is that a question about people, about maps, or about
  both — and who should get to answer it?

***

### Where the data came from

- **Race by census block** — U.S. Census Bureau, *2020 Census Redistricting Data
  (P.L. 94-171) Summary File*, legacy format, Georgia; tables P1 (total
  population by race) and P3 (voting-age population by race). This is the file
  the Bureau is required by law to deliver to every state legislature so it can
  draw districts.
  <https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/Georgia/ga2020.pl.zip>
- **Block, block group and tract boundaries** — Census TIGER/Line 2020, Georgia.
  <https://www2.census.gov/geo/tiger/TIGER2020/BG/tl_2020_13_bg.zip> and
  <https://www2.census.gov/geo/tiger/TIGER2020/TRACT/tl_2020_13_tract.zip>;
  blocks (`tl_2020_13_tabblock20`) are shared with `ga-precinct-returns`.
- **Interior points** — the Bureau's own `INTPTLAT20`/`INTPTLON20`, the same idea
  as `assignPolys` in Cervas, *R-Functions*.
  <https://github.com/jcervas/R-Functions/tree/main/assignPolys>
- **The concept** — Openshaw, S. (1984), *The Modifiable Areal Unit Problem*,
  CATMOG 38; Robinson, W. S. (1950), "Ecological Correlations and the Behavior of
  Individuals," *American Sociological Review* 15(3).

Builds are in `data/`: `build-block-race.R` reads the P.L. 94-171 file into
`ga_block_race.csv` (`r n(FV("ga_blocks"))` blocks, `r n(FV("ga_pop"))` people);
`build-areal-units.R` produces the ladders, the maps and the
`r n(FV("n_plans"))` partitions. Both record their source URLs, fetch date and
row counts, and both stop if a GEOID is not the width it should be.
