---
title: "Key: Draw the Line Somewhere Else"
subtitle: "84-355 Democracy's Data | Areal units | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)

D  <- "data"
fc <- read.csv(file.path(D, "facts.csv"),           stringsAsFactors = FALSE)
ls_ga <- read.csv(file.path(D, "ladder_state.csv"), stringsAsFactors = FALSE)
lf <- read.csv(file.path(D, "ladder_fulton.csv"),   stringsAsFactors = FALSE)
mu <- read.csv(file.path(D, "map_units.csv"),       stringsAsFactors = FALSE)
sm <- read.csv(file.path(D, "zoning_summary.csv"),  stringsAsFactors = FALSE)
sw <- read.csv(file.path(D, "zoning_sweep.csv"),    stringsAsFactors = FALSE)
pu <- read.csv(file.path(D, "zoning_plan_units.csv"), stringsAsFactors = FALSE)
co <- read.csv(file.path(D, "county_outline.csv"),  stringsAsFactors = FALSE)
mv <- read.csv(file.path(D, "map_vals.csv"),
               colClasses = c(unit = "character"), stringsAsFactors = FALSE)
# the 232,717-block file, read once, GEOID pinned to character
bl <- read.csv(file.path(D, "ga_block_race.csv"),
               colClasses = c(GEOID20 = "character"), stringsAsFactors = FALSE)
bl$bg <- substr(bl$GEOID20, 1, 12)
bl$tract <- substr(bl$GEOID20, 1, 11)
bl$county <- substr(bl$GEOID20, 1, 5)

FV <- function(k) { v <- fc$value[fc$name == k]; if (!length(v)) stop("no fact: ", k); v }
FN <- function(k) as.numeric(FV(k))
n  <- function(x) format(round(as.numeric(x)), big.mark = ",")
pc <- function(x, k = 1) formatC(as.numeric(x), format = "f", digits = k)
LF <- function(l, cc) lf[[cc]][lf$level == l]
LG <- function(l, cc) ls_ga[[cc]][ls_ga$level == l]
LEVS <- lf$level
RAMP <- colorRampPalette(c("#f7f7f7", "#d9e6f2", "#8fbedd", "#3f8dc0",
                           "#1f5d8c", "#12395a"))(101)
RX <- range(mu$x); RY <- range(mu$y)

onepath <- function(X, Y) {
  dx <- X[-1] - X[-length(X)]; dy <- Y[-1] - Y[-length(Y)]
  seg <- paste0(dx, ifelse(dy < 0, "", ","), dy)
  sep <- c("", ifelse(substr(seg[-1], 1, 1) == "-", "", " "))
  paste0("M", X[1], ",", Y[1], "l", paste0(sep, seg, collapse = ""), "Z")
}
ringpaths <- function(d, by, sx, sy) {
  k <- interaction(d[[by]], d$part, drop = TRUE)
  p <- vapply(split(d, k), function(z) {
    X <- round(sx(z$x)); Y <- round(sy(z$y))
    keep <- c(TRUE, X[-1] != X[-length(X)] | Y[-1] != Y[-length(Y)])
    X <- X[keep]; Y <- Y[keep]
    if (length(X) < 3) paste0("M", X[1], ",", Y[1], "h1v1h-1Z") else onepath(X, Y)
  }, character(1))
  p[nzchar(p)]
}
jstr <- function(x) paste0("[", paste0('"', x, '"', collapse = ","), "]")
jnum <- function(x) paste0("[", paste0(x, collapse = ","), "]")
drawrings <- function(d, by, fill, border = NA, lwd = 0.4) {
  k <- interaction(d[[by]], d$part, drop = TRUE)
  for (lv in levels(k)) {
    z <- d[k == lv, ]
    polygon(z$x, z$y, col = fill[[as.character(z[[by]][1])]], border = border, lwd = lwd)
  }
}

# ---- the ladder, recomputed from blocks, for the student options ------------
# Identical arithmetic to data/build-areal-units.R: population-weighted, and
# units with nobody in them dropped rather than counted as 0%.
wm <- function(x, w) sum(x * w) / sum(w)
wsd <- function(x, w) sqrt(sum(w * (x - wm(x, w))^2) / sum(w))
ladder <- function(d, race = "black_any") {
  o <- NULL
  for (k in c("GEOID20", "bg", "tract")) {
    p <- tapply(d$pop, d[[k]], sum); b <- tapply(d[[race]], d[[k]], sum)
    ok <- p > 0; p <- p[ok]; b <- b[ok]; sh <- 100 * b / p
    o <- rbind(o, data.frame(
      level = c(GEOID20 = "block", bg = "block group", tract = "tract")[[k]],
      units = length(p), sd_share = round(wsd(sh, p), 1),
      units_majority = sum(sh > 50),
      pct_black_in_majority = round(100 * sum(b[sh > 50]) / sum(d[[race]]), 1)))
  }
  cbind(o, mean_share = round(100 * sum(d[[race]]) / sum(d$pop), 2))
}

knit_print.data.frame <- function(x, ...) {
  nm <- names(x); nm <- gsub("_", " ", nm)
  nm <- sub("^(.)", "\\U\\1", nm, perl = TRUE)
  knitr::knit_print(knitr::kable(x, col.names = nm, row.names = FALSE), ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`areal-units-brief.Rmd`) and the lab
(`areal-units.Rmd`) run the same numbered steps — the brief in prose with four
figures, the lab with the code exposed. **Students should read the brief first.**
Step numbers below refer to both.

**This is the conceptual spine of the redistricting half of the course.** Every
lab that measures a group by area — `redistricting`, `vote-dilution`,
`rpv`, `redlining` — inherits the problem demonstrated here. Run it before
them if the calendar allows.

Every number below was recomputed against the built files while this document was
knitted. Nothing is typed.

***

# The session in one sentence

The same people, counted on the same day, divided into six units of the same
size, yield anywhere from `r FV("maj_min")` to `r FV("maj_max")` majority-Black
units depending only on the angle of the cuts — so "how many majority-Black
districts does this county support?" is not a question about the county.

***

# The findings

```{r headline}
c(fulton_people          = FN("ful_pop"),
  fulton_black_pct       = FN("ful_black_pct"),
  plans_built            = FN("n_plans"),
  people_per_unit        = FN("quota"),
  worst_deviation_pct    = FN("max_dev_pct"),
  majority_black_min     = FN("maj_min"),
  majority_black_max     = FN("maj_max"),
  contiguous_plans       = FN("plans_contig"),
  contiguous_maj_min     = FN("contig_maj_min"),
  contiguous_maj_max     = FN("contig_maj_max"))
```

- **`r n(FV("ful_pop"))` people, `r n(FV("ful_blocks"))` blocks, one census.**
- **The scale ladder**: weighted spread `r pc(LF("Census block", "sd_share"))`
  points across blocks, `r pc(LF("Census tract", "sd_share"))` across tracts,
  `r pc(LF("County", "sd_share"))` across the county. The **weighted mean never
  moves**: `r pc(LF("Census block", "mean_share"), 2)`% on every rung.
- **Statewide**, the share of Black Georgians living in a majority-Black unit
  falls from `r pc(FV("ga_maj_block"))`% (blocks) to `r pc(FV("ga_maj_tract"))`%
  (tracts) to `r pc(FV("ga_maj_county"))`% (counties). Nobody moved.
- **The correlation with under-18 share changes sign**:
  `r pc(LG("Census block", "corr_child"), 3)` across blocks,
  `r pc(LG("County", "corr_child"), 3)` across counties.
- **The zoning sweep**: `r n(FV("n_plans"))` equal-population partitions,
  worst deviation `r pc(FV("max_dev_pct"), 2)`% of the quota, giving
  `r FV("maj_min")` to `r FV("maj_max")` majority-Black units. Restricted to the
  `r n(FV("plans_contig"))` effectively contiguous plans, still
  `r FV("contig_maj_min")` to `r FV("contig_maj_max")`.

***

# The three things to get said

**1. Scale and zoning are different effects, and the second is the one that
matters.** Students arrive knowing "bigger units smooth things out" and will try
to file the whole lab under that. Steps 3–5 are the scale effect and they are the
warm-up. **Step 6 holds scale exactly fixed** — same number of units, same
population in each, to `r pc(FV("max_dev_pct"), 2)`% — and the answer still moves
by `r FN("maj_max") - FN("maj_min")` seats. Make somebody say back to you why
that is not the same phenomenon.

**2. The mean is the control, and nobody notices it.** In Step 4 the
population-weighted mean is `r pc(LF("Census block", "mean_share"), 2)`% at every
rung, exactly. **Point at that column before anything else.** It is what makes
the rest an experiment rather than an anecdote: one quantity is provably
unchanged, so everything that did change is attributable to the unit. Students
who miss this think the lab is showing that aggregation introduces error. It
isn't. The arithmetic is exact throughout.

**3. A race-blind rule has a racial output.** Neither the strip rule nor the
split rule reads a single race variable. One produces `r FV("lo_maj")`
majority-Black units, the other `r FV("hi_maj")`. **This is the sentence to get
on the board**, because "the algorithm doesn't look at race" is the most common
defence of automated redistricting and this lab shows it establishes nothing at
all about the outcome.

**The line for the board:** *the map is not a measurement of the people; it is
half of the measurement.*

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1** The same people | `r n(FV("ful_pop"))` people, `r n(FV("ful_blocks"))` blocks | 4 | Say "any part Black" out loud and say it is a choice. Someone will ask; it is `r pc(FV("ga_black_pct"), 2)`% vs `r pc(FV("ga_black_alone_pct"), 2)`% statewide. |
| **2** GEOID as a string | Block group and tract are prefixes | 5 | **Do not skip.** It is what makes Steps 3-5 airtight: no spatial operation, so no geometric error to blame. |
| **3** Scale effect, seen | Four maps, one ramp | 8 | Let them look before you talk. Ask what happened to the black neighbourhoods between panel 1 and panel 4. |
| **4** Scale effect, measured | The ladder | 9 | **The constant mean.** Also the sign flip in the correlation column. |
| **5** Distribution collapse | Density by rung | 7 | It **does not** collapse much for Fulton. That is the finding, not a failure. |
| **6** The zoning effect | `r n(FV("n_plans"))` partitions | **14** | **The pivot.** Take the prediction first (below). |
| **7** The two extremes | The maps of both | 9 | Name *cracking* only after they have seen it. Stress that no rule saw race. |
| **8** Are they districts? | Contiguity | 8 | The honest objection, honestly answered. `r FV("plans_strict_contig")` of `r n(FV("n_plans"))` pass strictly, and the reason is rivers, not shapes. |
| **9** Why it is a democracy problem | Section 2 | 8 | "There is no true value to recover" is the hard idea. Give it room. |
| **10** The ecological fallacy | The tension | 5 | Name `rpv` as the answer-key case. Do not resolve it. |
| **11** What you can say | Including what you cannot | 4 | Close on the range, not on a number. |

**Step 6 deserves the prediction.** Before showing the figure, put the setup on
the board — same `r n(FV("ful_blocks"))` blocks, `r FV("K")` units, everyone
within `r pc(FV("max_dev_pct"), 2)`% of `r n(FV("quota"))` people, two rules
neither of which reads race — and ask: **how many of the six will be
majority-Black?** Almost everyone gives one number, usually two or three, because
the county is `r pc(FV("ful_black_pct"), 2)`% Black. The answer is that
`r n(FV("plans_zero_maj"))` plans give `r FV("maj_min")` and
`r n(sum(sm$n_majority == FN("maj_max")))` give `r FV("maj_max")`. The room
reliably assumes the question has an answer.

**If you are short on time**, protect Steps 4, 6 and 8. The constant mean, the
sweep, and the contiguity test that fails to rescue us are the argument.

***

# The figure to project

```{r zone-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
ord <- order(sm$n_majority, sm$max_share)
key <- paste(sm$family[ord], sm$angle[ord]); swk <- paste(sw$family, sw$angle)
rows <- vapply(seq_along(key), function(i)
  paste0("[", i - 1L, ",", paste0(formatC(sort(sw$share[swk == key[i]]),
         format = "f", digits = 1), collapse = ","), "]"), character(1))
i_lo <- which(key == paste(FV("lo_family"), FV("lo_angle"))) - 1L
i_hi <- which(key == paste(FV("hi_family"), FV("hi_angle"))) - 1L
lab_lo <- paste0(FV("lo_maj"), " majority-Black units · ", FV("lo_family"),
                 " rule at ", FV("lo_angle"), "°")
lab_hi <- paste0(FV("hi_maj"), " majority-Black units · ", FV("hi_family"),
                 " rule at ", FV("hi_angle"), "°")
brk <- cumsum(table(factor(sm$n_majority[ord], levels = sort(unique(sm$n_majority)))))
blab <- paste0("[", paste0('{"y":', as.integer(brk), ',"t":"', names(brk),
        ifelse(names(brk) == "1", " majority-Black unit", " majority-Black units"),
        '","n":', as.integer(table(sm$n_majority)[names(brk)]), '}', collapse = ","), "]")
cat(paste0('
<div id="zon" style="margin:1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const P=[', paste(rows, collapse = ","), '],BR=', blab, ';
const NR=P.length,LO=', i_lo, ',HI=', i_hi, ';
const RH=1.55,L=228,R=32,T0=54,B=48;
const W=900,H=Math.round(T0+NR*RH+B);
const svg=d3.select("#zon").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const x=d3.scaleLinear().domain([0,100]).range([L,W-R]);
const yr=i=>T0+i*RH;
svg.append("rect").attr("x",x(50)).attr("y",T0-6).attr("width",W-R-x(50))
  .attr("height",NR*RH+6).attr("fill","#C41230").attr("fill-opacity",0.045);
BR.forEach(b=>{
  svg.append("line").attr("x1",L-6).attr("x2",W-R).attr("y1",yr(b.y)).attr("y2",yr(b.y))
    .attr("stroke","#ddd");
  svg.append("text").attr("x",L-12).attr("y",yr(b.y)-b.n*RH/2+4).attr("text-anchor","end")
    .attr("font-size","13px").attr("fill","#555").text(b.n+" plans → "+b.t);});
const FLAT=[];P.forEach(p=>{for(let j=1;j<7;j++)FLAT.push([p[0],p[j]]);});
svg.append("g").selectAll("circle").data(FLAT).join("circle")
  .attr("cx",d=>x(d[1])).attr("cy",d=>yr(d[0])+RH/2).attr("r",1.7)
  .attr("fill",d=>d[1]>50?"#C41230":"#3f8dc0").attr("fill-opacity",0.55);
[[LO,"', lab_lo, '"],[HI,"', lab_hi, '"]].forEach(([i,t])=>{
  svg.append("line").attr("x1",L-6).attr("x2",W-R).attr("y1",yr(i)+RH/2)
    .attr("y2",yr(i)+RH/2).attr("stroke","#111").attr("stroke-width",1);
  const p=P[i];
  for(let j=1;j<7;j++)
    svg.append("circle").attr("cx",x(p[j])).attr("cy",yr(i)+RH/2).attr("r",3.6)
      .attr("fill",p[j]>50?"#C41230":"#1f5d8c").attr("stroke","#fff").attr("stroke-width",1);
  svg.append("text").attr("x",W-R).attr("y",yr(i)-7).attr("text-anchor","end")
    .attr("font-size","13px").attr("font-weight","600").attr("fill","#111").text(t);});
svg.append("line").attr("x1",x(50)).attr("x2",x(50)).attr("y1",T0-14).attr("y2",T0+NR*RH)
  .attr("stroke","#C41230").attr("stroke-width",1.4);
svg.append("text").attr("x",x(50)).attr("y",T0-20).attr("text-anchor","middle")
  .attr("font-size","14px").attr("font-weight","600").attr("fill","#C41230").text("50%");
d3.range(0,101,10).forEach(v=>svg.append("text").attr("x",x(v)).attr("y",T0+NR*RH+20)
  .attr("text-anchor","middle").attr("font-size","12px").attr("fill","#aaa").text(v+"%"));
svg.append("text").attr("x",L).attr("y",T0+NR*RH+40).attr("font-size","13px").attr("fill","#888")
  .text("Black share of a unit · every unit of every plan · "+(6*NR).toLocaleString()
        +" units in "+NR+" plans");
})();
</script>
'))
```

```{r zone-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=5.6}
ord <- order(sm$n_majority, sm$max_share)
key <- paste(sm$family[ord], sm$angle[ord]); swk <- paste(sw$family, sw$angle)
par(mar = c(3.6, 9.6, 2.6, 1.0))
plot(NA, xlim = c(0, 100), ylim = c(nrow(sm), 0), axes = FALSE, ann = FALSE)
rect(50, 0, 100, nrow(sm), col = adjustcolor("#C41230", 0.045), border = NA)
for (i in seq_along(key)) {
  v <- sw$share[swk == key[i]]
  points(v, rep(i, length(v)), pch = 19, cex = 0.24,
         col = adjustcolor(ifelse(v > 50, "#C41230", "#3f8dc0"), 0.55))
}
brk <- cumsum(table(factor(sm$n_majority[ord], levels = sort(unique(sm$n_majority)))))
cnt <- as.integer(table(sm$n_majority)[names(brk)])
segments(0, brk, 100, brk, col = "#dddddd")
text(-2, brk - cnt / 2, paste0(cnt, " plans -> ", names(brk),
     ifelse(names(brk) == "1", " majority-Black unit", " majority-Black units")),
     adj = 1, cex = 0.6, col = "#555", xpd = NA)
abline(v = 50, col = "#C41230", lwd = 1.4)
for (tg in c("lo", "hi")) {
  i <- which(key == paste(FV(paste0(tg, "_family")), FV(paste0(tg, "_angle"))))
  v <- sw$share[swk == key[i]]
  segments(0, i, 100, i, col = "#111111", lwd = 0.7)
  points(v, rep(i, length(v)), pch = 21, cex = 0.62, lwd = 0.5,
         bg = ifelse(v > 50, "#C41230", "#1f5d8c"), col = "white")
  lb <- paste0(FV(paste0(tg, "_maj")), " majority-Black units - ",
               FV(paste0(tg, "_family")), " rule at ", FV(paste0(tg, "_angle")), " deg")
  yy <- if (tg == "lo") i - 5 else i - 6
  rect(100 - strwidth(lb, cex = 0.58) - 1, yy - strheight(lb, cex = 0.58) * 0.9,
       100.5, yy + strheight(lb, cex = 0.58) * 0.9, col = "white", border = NA)
  text(100, yy, lb, adj = 1, cex = 0.58, font = 2, col = "#111111")
}
axis(1, seq(0, 100, 10), paste0(seq(0, 100, 10), "%"), cex.axis = 0.6,
     col = "#ddd", col.axis = "#999", tck = -0.012, mgp = c(2, 0.3, 0))
mtext(sprintf("Black share of a unit - every unit of every plan - %s units in %d plans",
              n(6 * nrow(sm)), nrow(sm)), side = 1, line = 1.9, cex = 0.6, col = "#888")
title(sprintf("%s equal-population partitions of the same %s blocks",
              n(nrow(sm)), n(FV("ful_blocks"))), cex.main = 0.9, adj = 0, line = 0.8)
```

If you show one image, show this one. The red band is above 50%; each row is one
plan; the two black rows are the extremes. **The whole argument is that every row
contains the same `r n(FV("ful_pop"))` people.**

And the same county at four scales, for Step 3:

```{r scale-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
PW <- 208L; GAP <- 10L; TOP <- 42L; BOTPAD <- 40L
s  <- PW / diff(RX); PH <- round(diff(RY) * s)
W  <- 4L * PW + 3L * GAP; H <- TOP + PH + BOTPAD
KEYS <- c("b", "g", "t", "c")
mkpanel <- function(i) {
  off <- (i - 1L) * (PW + GAP)
  fx <- function(x) off + (x - RX[1]) * s
  fy <- function(y) TOP + (RY[2] - y) * s
  v  <- mv[mv$lev == KEYS[i], ]
  P  <- ringpaths(mu[mu$lev == KEYS[i], ], "uid", fx, fy)
  uid <- as.integer(sub("\\.[0-9]+$", "", names(P)))
  ci  <- match(v$fill[match(uid, v$uid)], RAMP) - 1L
  ci[is.na(ci)] <- -1L
  base <- ringpaths(mu[mu$lev == "c", ], "uid", fx, fy)
  paste0("[", jstr(c(base, P)), ",", jnum(c(rep(-1L, length(base)), ci)), "]")
}
PAN <- vapply(seq_along(KEYS), mkpanel, character(1))
sub1 <- paste0(vapply(LEVS, function(l) n(LF(l, "units")), character(1)),
               ifelse(LEVS == "County", " unit", " units"))
sub2 <- paste0("median ", vapply(LEVS, function(l) n(LF(l, "median_pop")),
                                 character(1)), " people")
sub3 <- paste0("spread ", vapply(LEVS, function(l) pc(LF(l, "sd_share")),
                                 character(1)), " pts")
lab <- paste0("[", paste0('{"t":"', LEVS, '","a":"', sub1, '","b":"', sub2,
                          '","c":"', sub3, '"}', collapse = ","), "]")
cat(paste0('
<div id="scl" style="margin:1em 0"></div>
<!-- d3 v7 is loaded once, by the first figure above -->
<script>
(function(){
const PAN=[', paste(PAN, collapse = ","), '],L=', lab, ',R=', jstr(RAMP), ';
const W=', W, ',H=', H, ',PW=', PW, ',GAP=', GAP, ',TOP=', TOP, ',PH=', PH, ';
const svg=d3.select("#scl").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const col=i=>i<0?"#e6e6e6":R[i];
PAN.forEach((P,i)=>{
  const off=i*(PW+GAP);
  svg.append("g").selectAll("path").data(P[0]).join("path").attr("d",d=>d)
    .attr("fill",(d,j)=>col(P[1][j])).attr("stroke","none");
  svg.append("text").attr("x",off+PW/2).attr("y",15).attr("text-anchor","middle")
    .attr("font-size","13px").attr("font-weight","600").attr("fill","#222").text(L[i].t);
  [[L[i].a,27],[L[i].b,37]].forEach(([t,y])=>
    svg.append("text").attr("x",off+PW/2).attr("y",y).attr("text-anchor","middle")
      .attr("font-size","9.5px").attr("fill","#777").text(t));
  svg.append("text").attr("x",off+PW/2).attr("y",TOP+PH+12).attr("text-anchor","middle")
    .attr("font-size","9.5px").attr("fill","#777").text(L[i].c);
});
const KW=210,KX=(W-KW)/2,KY=TOP+PH+21;
R.forEach((c,i)=>svg.append("rect").attr("x",KX+i*KW/101).attr("y",KY)
  .attr("width",KW/101+0.5).attr("height",7).attr("fill",c));
[[0,"0%"],[50,"50%"],[100,"100%"]].forEach(([v,t])=>
  svg.append("text").attr("x",KX+v*KW/100).attr("y",KY+8).attr("text-anchor","middle")
    .attr("dy","0.9em").attr("font-size","8px").attr("fill","#888").text(t));
svg.append("text").attr("x",KX-8).attr("y",KY+6).attr("text-anchor","end")
  .attr("font-size","8.5px").attr("fill","#888").text("share of the unit that is Black");
})();
</script>
'))
```

```{r scale-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.5}
layout(matrix(c(1, 2, 3, 4, 5, 5, 5, 5), 2, 4, byrow = TRUE), heights = c(8, 1))
op <- par(mar = c(1.6, 0.2, 3.0, 0.2), xpd = NA)
for (i in seq_along(LEVS)) {
  lv <- c("b", "g", "t", "c")[i]
  v  <- mv[mv$lev == lv, ]
  fill <- as.list(v$fill); names(fill) <- as.character(v$uid)
  plot(NA, xlim = RX, ylim = RY, asp = 1, axes = FALSE, ann = FALSE)
  polygon(co$x, co$y, col = "#e6e6e6", border = NA)
  drawrings(mu[mu$lev == lv, ], "uid", fill, border = NA)
  title(LEVS[i], cex.main = 1.0, line = 1.9)
  mtext(paste0(n(LF(LEVS[i], "units")),
               if (LEVS[i] == "County") " unit" else " units"),
        side = 3, line = 0.95, cex = 0.55, col = "#777")
  mtext(paste0("median ", n(LF(LEVS[i], "median_pop")), " people"),
        side = 3, line = 0.25, cex = 0.55, col = "#777")
  mtext(paste0("spread ", pc(LF(LEVS[i], "sd_share")), " pts"),
        side = 1, line = 0.2, cex = 0.55, col = "#777")
}
par(mar = c(0, 0.2, 0, 0.2))
plot(NA, xlim = c(0, 1), ylim = c(0, 1), axes = FALSE, ann = FALSE)
bx <- seq(0.40, 0.62, length.out = 102)
rect(bx[-102], 0.42, bx[-1], 0.72, col = RAMP, border = NA)
text(c(0.40, 0.51, 0.62), 0.42, c("0%", "50%", "100%"), pos = 1,
     cex = 0.52, col = "#888", offset = 0.25)
text(0.385, 0.57, "share of the unit that is Black", pos = 2, cex = 0.52, col = "#888")
par(op); layout(1)
```

***

# Handling the room

**The thing that goes wrong** is that a few students conclude the data is
worthless — "so you can make it say anything." That is the wrong lesson and it is
worth pushing back on directly:

- **The census count is exact and the arithmetic is exact.** Nothing here is
  measurement error. Say so.
- **The ambiguity is bounded and measurable.** `r FV("maj_min")` to
  `r FV("maj_max")`, and `r FV("contig_maj_min")` to `r FV("contig_maj_max")`
  among plausible plans. That is a *result*, not a shrug.
- **Some questions are unaffected.** "How many Black people live in Fulton
  County?" has one answer: `r n(FV("ful_black"))`. Ask the room which of their
  own questions are unit-dependent and which are not. That sorting exercise is
  the most useful five minutes in the session.

**The second thing** is students conflating this with the ecological fallacy,
which they may have met in `levels-of-aggregation`. They are cousins, not twins.
The ecological fallacy is about inferring *individuals* from areas. The
modifiable areal unit problem is about the *area statistic itself* being
non-unique before anyone infers anything. Step 10 names both; keep them apart.

***

# Option A — A different county

## Verified output

```{r a}
CTY <- c("Fulton" = "13121", "Chatham" = "13051", "Dougherty" = "13095",
         "Forsyth" = "13117", "Clayton" = "13063")
o <- do.call(rbind, lapply(names(CTY), function(nm) {
  d <- bl[bl$county == CTY[[nm]], ]
  z <- ladder(d)
  data.frame(county = nm, people = n(sum(d$pop)), black_pct = z$mean_share[1],
             sd_blocks = z$sd_share[1], sd_tracts = z$sd_share[3],
             collapse_pts = round(z$sd_share[1] - z$sd_share[3], 1),
             in_maj_blocks = z$pct_black_in_majority[1],
             in_maj_tracts = z$pct_black_in_majority[3])
}))
o
```

**3/4:** Runs a second county, reports its ladder, notices the spread falls
there too.

**4/4:** Compares the *size* of the collapse. Fulton loses
`r pc(o$collapse_pts[o$county == "Fulton"])` points of spread between blocks and
tracts; Chatham loses `r pc(o$collapse_pts[o$county == "Chatham"])`. Explains
why: **the scale effect bites in proportion to how much of the segregation
happens below the size of a tract.** Fulton's is a north-south divide tens of
kilometres across and survives aggregation; Chatham's is finer-grained and does
not.

**4/4, best:** Notices Clayton, where `pct_black_in_majority` goes the *wrong*
way — `r pc(o$in_maj_blocks[o$county == "Clayton"])`% at block level rising to
`r pc(o$in_maj_tracts[o$county == "Clayton"])`% at tract level. In a county that
is already `r pc(o$black_pct[o$county == "Clayton"])`% Black, coarsening *pulls
people into* majority-Black units instead of out of them. **The scale effect has
no fixed direction; it depends on which side of 50% the surrounding area sits.**
A student who works that out has understood the mechanism rather than memorised
the Fulton result. Forsyth is the mirror image: at
`r pc(o$black_pct[o$county == "Forsyth"])`% Black, exactly
`r n(ladder(bl[bl$county == "13117", ])$units_majority[1])` of its
`r n(ladder(bl[bl$county == "13117", ])$units[1])` populated blocks are
majority-Black, and none of its tracts are.

***

# Option B — A different definition of "Black"

## Verified output

```{r b}
d <- bl[bl$county == "13121", ]
z <- rbind(cbind(measure = "any part Black", ladder(d, "black_any")),
           cbind(measure = "Black alone",    ladder(d, "black_alone")))
z[, c("measure", "level", "units", "mean_share", "sd_share",
      "units_majority", "pct_black_in_majority")]
```

**3/4:** Reports that the numbers are lower under "Black alone" and that the
pattern is the same.

**4/4:** Notices *which* statistic is sensitive. The mean falls
`r pc(z$mean_share[1] - z$mean_share[4], 2)` points, and the **count of
majority-Black blocks falls from `r n(z$units_majority[1])` to
`r n(z$units_majority[4])`** — about
`r pc(100 * (z$units_majority[1] - z$units_majority[4]) / z$units_majority[1])`%
— while the spread barely moves
(`r pc(z$sd_share[1])` to `r pc(z$sd_share[4])` points). **Threshold counts are
far more sensitive to the definition than distributional statistics are**,
because a definition change moves units across a line rather than rearranging
them.

**4/4, best:** Connects it to Step 6. If a `r pc(z$mean_share[1] - z$mean_share[4], 2)`-point
change in *definition* can move the count of majority-Black units, then the
`r FN("maj_max") - FN("maj_min")`-unit range from *zoning* is not the only
non-uniqueness in the answer — and the two compound. Notes that litigants choose
the definition too, and that the choice is not neutral.

***

# Option C — A different threshold

## Verified output

```{r c}
k <- paste(sw$family, sw$angle)
o <- do.call(rbind, lapply(c(40, 45, 50, 55, 60, 70), function(t)
  do.call(rbind, lapply(c("share", "bvap"), function(v) {
    cnt <- tapply(sw[[v]] > t, k, sum)
    data.frame(threshold = paste0(t, "%"),
               measure = c(share = "total population", bvap = "voting age")[[v]],
               fewest = min(cnt), most = max(cnt), range = max(cnt) - min(cnt),
               plans_at_fewest = sum(cnt == min(cnt)))
  }))))
o
```

**3/4:** Reports that the counts change with the threshold, and that voting-age
numbers are a little lower than total-population numbers.

**4/4:** Reads the `range` column, which is the point: **at no threshold does the
choice of plan stop mattering.** Even at
`r o$threshold[which.min(o$range)]` the gap between the best and worst plan is
`r min(o$range)` units. There is no threshold at which the map becomes
irrelevant.

**4/4, best:** Notes that the ambiguity is *largest* right where the law puts its
line. At 50% of total population the range is
`r o$range[o$threshold == "50%" & o$measure == "total population"]` of
`r FV("K")` units; the biggest gap anywhere in the table is
`r max(o$range)`. **The legal threshold sits in the region of maximum
sensitivity**, which is not a coincidence — 50% is near the county's own
`r pc(FV("ful_black_pct"), 2)`%, and a threshold near the population share is
exactly where partitioning has the most leverage. A student who says that has
found something the lab does not state.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given, and this lab gives more of it than most.
- **The move that earns a 4** is treating "the Black share of a district" as a
  joint fact about people and boundaries, and reasoning about the *range* rather
  than reporting a number.
- **A 2** is "so the data can say anything" — nihilism dressed as sophistication.
  The ambiguity here is measured, bounded, and smaller than the student thinks.
- **Also a 2**: describing the scale effect (Steps 3-5) and stopping, having
  missed that Step 6 holds scale fixed.
- **Also a 2**: treating the zoning result as evidence that someone gerrymandered
  Fulton. Nothing in the lab has intent in it.
- **Reward anyone who asks what the range would be over *all* partitions**, not
  just these two rule families. The honest answer is that we do not know and it
  is certainly wider; noticing that `r n(FV("n_plans"))` is a convenience sample
  of an enormous space is the most sophisticated available response.

## Anticipated questions

- *"Isn't this just saying bigger units are smoother?"* — No, and Step 6 is the
  proof: the units there are all the same size. Have this answer ready; it is the
  most common misreading.
- *"Which of the `r n(FV("n_plans"))` plans is the right one?"* — None. That is
  the finding. A real plan is chosen by a political process operating under legal
  constraints, and the constraints do not narrow it to one.
- *"Couldn't a computer just find the fairest map?"* — Only once "fairest" is
  defined, and the definitions conflict. Point at the two rules: both are
  perfectly principled and they disagree by
  `r FN("hi_maj") - FN("lo_maj")` seats.
- *"Why does contiguity fail for every plan?"* — Rivers, interstates and the
  airport are their own blocks and touch nothing across them. The strict test is
  measuring the block layer, not the plan. Hence the
  `r FV("contig_thr")`%-of-population criterion, which is stated in Step 8 rather
  than hidden.
- *"Is `r pc(FV("max_dev_pct"), 2)`% population deviation realistic?"* — It is far
  tighter than required. Congressional plans aim for near-zero; local plans are
  commonly allowed 10% total deviation, which would make the range found here
  **wider**, not narrower.
- *"Does this mean census tracts are wrong?"* — No. Tracts are a reasonable unit
  for many questions. They are the wrong unit for any question whose answer
  depends on where a 50% line falls.

## Connections

- **`precinct-geography`** — the same machinery with **time** as the variable
  instead of the unit: boundaries move between elections. Run the two together
  and the pair covers both ways a unit can betray you.
- **`levels-of-aggregation`** — meets the modifiable areal unit problem in
  passing, with a random-regrouping null. This lab is that observation made the
  whole subject, and adds the zoning effect, which that lab does not isolate.
- **`redistricting`** — the applied version. Everything there assumes a
  plan; this lab is about the space of plans it was drawn from.
- **`vote-dilution` and `rpv`** — both measure race by area and inherit
  this problem wholesale. `rpv` is also the one place in the course where
  the ecological inference can be checked against individual records.
- **`redlining`** — point-in-polygon and the same assumption that people are
  spread evenly inside a polygon, in a setting where the error cannot be measured.
- **`census-geography`** — the nesting spine that Step 2 exploits, and the list of
  things that do not nest.
