---
title: "Junk Filings in FEC Independent Expenditures"
type: brief
subtitle: "A junk filing in the federal record of political spending, and the difference between wrong and plausibly wrong"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
md_extensions: -tex_math_dollars
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("independent-expenditures-code.R")
```

```{r setup, include=FALSE}
```

Outside money is the least accountable part of an American campaign and the best documented part of it. Every dollar a super PAC spends for or against a candidate has to be reported to the Federal Election Commission. The Commission publishes the whole cycle as one file that anybody can download for nothing. So the obvious questions look answerable. How much was there in 2024, and was it spent promoting candidates or attacking them?

Both are questions a citizen has a stake in, because this is the money nobody can be asked about. A campaign that runs a dishonest advertisement can be made to answer for it. A super PAC cannot be voted out, and the candidate it helps can disclaim it truthfully, because coordinating with it would be illegal. *Citizens United v. FEC* (2010) held that such spending cannot be limited, on the reasoning that money spent independently cannot corrupt. Whether it mostly promotes or mostly attacks is one of the few things about it a reader can check.

So you download the file and add up a column. **Can you?**

## The test

The data is the Commission's bulk file of independent expenditures for the 2024 cycle: `r n(ROWS)` rows, one per reportable expenditure, published at fec.gov with no key, no account and no request form. An independent expenditure is money spent for or against a candidate by somebody the candidate does not control. Anyone who spends above a threshold on a message urging the election or defeat of a named federal candidate has to report it. In the closing weeks of a campaign they have to report it within a day or two of writing the check.

This file can answer the question because of one column. Schedule E makes the spender code every line as **supporting** or **opposing** the candidate it names. Almost no other record of political money in the United States carries that field, and without it the only answerable question is how much, which is the less interesting one. What the file does not carry is any checking. Nobody stands between the filer and the reader: the Commission is a disclosure agency rather than an auditor, for the reasons [this cluster's chapter on campaign finance disclosure](../finance-source/finance-source-brief.html) sets out.

The test is to add the column up as the file arrives, look at the largest rows, and add it up again.

## The data

The table every number here rests on is that bulk file: `r NCOL` columns, `r n(ROWS)` rows. Here is every column, with the part worth insisting on — **what each one actually contains across all `r n(ROWS)` rows**, rather than whatever it happened to say in the first one.

```{r schema}
```

```{r schema-helpers, include=FALSE}
```

Three things in that table are worth stopping on.

**The spender's name begins with a space.** `" Go America PAC"`, not `"Go America PAC"`. Group the file by spender without trimming and this committee becomes two committees, sorted to opposite ends of any alphabetical list, with its money split between them. Nothing announces it. It is one invisible character.

**There are three date columns and the obvious one is the emptiest.** `exp_date` is the date of the expenditure, the column anyone would reach for, and it is blank on **`r pc(bl("exp_date"))`% of rows**. `dissem_dt`, the date the message actually went out, is blank on `r pc(bl("dissem_dt"))`%. Only `receipt_dat` is filled in on every row, and that is the date the Commission received the paperwork rather than the date anybody spent anything.

**`sup_opp` is one letter, and this brief's central finding is a count of it.** `S` for support, `O` for oppose, written by the filer and checked by nobody. The whole argument about whether outside money attacks or promotes is that single character added up across `r n(ROWS)` rows.

Two decisions were made before any of that was counted, and both are written down rather than assumed. **Candidate names were uppercased** so that case variants of one person would merge: `Evans, Gabe` becomes `EVANS, GABE`. The key is a name rather than an identifier, and it does not fully work.

**One exclusion rule was applied, stated as a number:** rows above `r ml(CUT)` are dropped. That is a threshold and not a diagnosis. Every row it removes is committed to `ie_outliers.csv`, so a reader can disagree with the cut and recompute from the discarded rows. Nothing else was touched: no amount corrected, no filer contacted, no obviously false row quietly deleted.

One number below cannot be recomputed from the tables. The raw download runs to roughly 19 megabytes and is not committed, so `r n(ROWS)` is the row count the build recorded at the moment it read the file.

> **Before you read on, commit to a number.** `r NOUT` of those `r n(ROWS)` rows
> sit above the cut and come out. A count that small, in a file that large,
> ought to leave a percentage where it was. Write down how far you think the
> share of
> money spent **supporting** candidates moves when they go. A tenth of a point?
> A point? Five? Put an actual number on the page before you turn it.

## What it says about democracy

Here is the file added up exactly as it arrives.

```{r raw}
```

**`r bn(RAW)`, of which `r pc(100*RAWS/RAW)`% supports candidates and `r pc(100*RAWO/RAW)`% opposes them.** Both of those numbers are wrong. One is wrong by a factor of ten. The other is backwards, and it is the dangerous one.

```{r outliers}
```

```{r outlier-share}
```

**`r NOUT` rows, `r pc(100*NOUT/ROWS, 2)`% of the file, carry `r pc(SHARE)`% of the dollars.** Read who filed them. **"THE COURT OF DIVINE JUSTICE."** **"THE COMMITTEE OF 300."** **"The Masonic Illuminati Eye."** Nearly ten billion dollars each, almost all naming one obscure candidate, for purposes including *"DEPOSIT FOR WINNING IN ELECTION"*.

One of them can be read in full, because the Commission publishes the submissions as well as the summary file.

```{r d3-card, results='asis', eval=knitr::is_html_output()}
```

```{r card-static, eval=!knitr::is_html_output(), fig.height=4.7}
```

**Figure 1. One junk filing, as the Federal Election Commission received and published it.** A facsimile of the record fits this question better than any chart would, because the claim is about what a federal agency published rather than about a quantity. Every value on the card is read out of `ie_filing_1707511.fec` rather than retyped: `r n(FLB)` bytes, three records, fields separated by the ASCII file-separator character. The payee is a fictional lawyer, the treasurer a fictional drug distributor, the candidate the protagonist of a television series, and the amount `r d(FEEAMT)`. The single letter in the support-or-oppose field, shown in blue, decides which side of every total below the money lands on.

These are not errors. They are **junk filings**, sitting in the federal government's official record of campaign spending, published exactly as submitted. That is not an oversight. An agency empowered to decide which political filings are real would hold a power over political speech that nobody would want it to have, exercised by commissioners appointed on a partisan basis. **There is no free version of this**, and the cost of the arrangement we have is a national statistic that is wrong by a factor of ten until somebody reads `r NOUT` rows.

Not every bad row is somebody lying, either.

```{r fw}
```

`r FWN` rows, all `r d(FW$amount[1])`, all the same committee, all supporting the same candidate — and **all carrying the same file number**. One filing, exported `r FWN` times, worth `r bn(FWAMT)` of spending that never happened. The junk filings are people lying to the government. This is the government's own export repeating a row. They need different fixes, and calling both of them "bad data" loses the distinction.

Now take all `r NOUT` out.

```{r cleaned}
```

**The total falls from `r bn(RAW)` to `r bn(CLN)`**, which matches what watchdog organizations report for the cycle. **And the headline reverses.** In the raw file, `r pc(100*RAWS/RAW)`% of outside money supports candidates. Once `r NOUT` rows are gone, **`r pc(100*CLNO/CLN)`% of it is spent attacking somebody.** The supportive share does not slip by a point or two. It falls `r pc(100*RAWS/RAW - 100*CLNS/CLN)` points and crosses the halfway line on the way down.

```{r d3-flip, results='asis', eval=knitr::is_html_output()}
```

```{r flip-static, eval=!knitr::is_html_output(), fig.height=2.6}
```

**Figure 2. Removing `r NOUT` rows out of `r n(ROWS)` does not shrink the answer. It reverses it.** Each bar is the share of its own version's dollars spent attacking a candidate: `r pc(100*RAWO/RAW)`% as published, `r pc(100*CLNO/CLN)`% with `r NOUT` rows gone. One bar against a line at half works better here than a pair of totals, because the two versions differ tenfold in size and the point is not size. The point is which side of the line they land on.

Here is why that matters more than the ten-fold error. `r bn(RAW)` is *obviously* wrong, and anyone would query it. **`r pc(100*RAWS/RAW)`% supportive is a perfectly plausible number.** It would pass a referee, sit unremarked in a paper, and be cited. **Wrongness that announces itself is the safe kind.**

Two exceptions have to be told, because a finding that depends on where a threshold sits is not a finding.

```{r threshold}
```

**Nothing sits anywhere near the cut.** The largest row the rule keeps is `r d(KEPT)` and the smallest it removes is `r d(min(o$amount))`, a gap of `r ml(min(o$amount) - KEPT)` with `r INBAND` rows in it. Any threshold between those two figures removes the same `r NOUT` rows and gives the same answer, so the cut could have been set anywhere across a factor of `r pc(min(o$amount) / KEPT)` without changing a number here. That slack is what makes the rule defensible, and it is the thing to check rather than the roundness of the number chosen.

The second exception is sharper. Put back **only** the `r FWN` duplicated rows, the export bug rather than the fantasy, and see whether the finding survives.

```{r restore}
```

**It does not.** Restoring one committee's duplicated filing is enough on its own to flip the conclusion back. So the honest sentence is narrower than "most outside money is negative." It is **"most outside money is negative, and that conclusion depends on handling one export bug correctly."** Which is exactly why the `r NOUT` removed rows are published beside the total instead of being quietly dropped.

## What this brief cannot tell you

**The cleaned file is not clean.** Uppercasing merged `Harris, Kamala` with `HARRIS, KAMALA`, and could not merge `TRUMP, DONALD` with `TRUMP, DONALD J.`. `r nrow(TR)` spellings of Donald Trump and `r nrow(TE)` of Jon Tester survive into `ie_candidates.csv`, each carrying part of the money. Cleaning is not a phase that finishes.

**Nothing here is evidence of fraud, and nobody was prosecuted.** The `r NOUT` rows look like what they probably are: nonsense, or somebody's error, sitting unremarked in a system that publishes what it receives.

**"Independent" is a legal claim rather than an observed property.** Coordination is decided after the fact if at all, and it is not a field. The constitutional argument that this spending cannot corrupt rests on a factual premise no dataset can check.

**There is no persuasion data anywhere in this.** Dollars and direction are here; effects are not. The obvious next source is the platform advertising libraries, and they are closed: Meta requires identity verification and Google reports spending in ranges. This is also one category of spending in one cycle. Electioneering communications, party spending, and groups that never file at all are elsewhere or nowhere.

## What you should have learned

- A federal record can be complete, official and free, and still be wrong: `r NOUT` filings out of `r n(ROWS)` inflate the 2024 outside-spending total from `r bn(CLN)` to `r bn(RAW)`, and they are still in the published file.
- Bad rows do more than inflate a total. Removing these `r NOUT` moves the file from `r pc(100*RAWS/RAW)`% supportive to `r pc(100*CLNO/CLN)`% attacking, a `r pc(100*RAWS/RAW - 100*CLNS/CLN)`-point swing that crosses the halfway line.
- The dangerous error is the plausible one. `r bn(RAW)` invites a query; `r pc(100*RAWS/RAW)`% supportive would pass a referee and be cited.
- A cleaning rule is trustworthy when there is slack around it: the cut here could sit anywhere between `r d(KEPT)` and `r d(min(o$amount))` without changing an answer.
- Two kinds of bad row need two kinds of fix. `r FWN` of the `r NOUT` are one filing exported `r FWN` times by the agency's own system, and restoring just those flips the finding back.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `ie_outliers.csv` lists every filing the rule cut, with `spender`, `amount` and `pur`. Read them. What makes these obviously wrong, and what would a plausibly wrong one look like?
- `ie_columns.csv` gives each column a `blank` percentage. Sort by it and find the columns that are mostly empty. Which of them would you have assumed was reliable?
- `ie_candidates.csv` has `pct_against` for every candidate, with `supporting` and `opposing` beside it. Sort it. Who was money mostly spent against rather than for, and does office explain it?
- `ie_committees.csv` has `pct_against` and `amount` per spender. Which committees exist almost entirely to attack, and how much of the cycle's money do they hold between them?

Two stretch ideas point past the spreadsheet. Take the largest row in `ie_outliers.csv`, look the committee up on fec.gov, and see what else it filed and whether anything ever happened to it. And download the 2022 cycle's file from the same address, apply the same rule, and find out whether junk of this size is a feature of the register or an accident of one year.

## Sources

**Data**

- Federal Election Commission bulk downloads, **independent expenditures, 2024
  cycle** — `r n(ROWS)` filings.
  <https://www.fec.gov/data/browse-data/?tab=bulk-data>
  Every row removed by the cleaning rule is kept in `ie_outliers.csv`, so the
  decision can be checked rather than trusted. The `r n(ROWS)` row count is the
  source file's, recorded by the build; the raw file itself is not committed.
- Federal Election Commission, **electronic filing `r FLN`**, the Form 24
  reproduced in Figure 1 — `ie_filing_1707511.fec`, `r n(FLB)` bytes,
  retrieved 10 August 2026 from
  <https://docquery.fec.gov/dcdev/posted/1707511.fec>.
  This is the original submission behind the `r d(FEEAMT)` row in
  `ie_outliers.csv`, in the FEC's own delimited filing format; every value in
  the figure is parsed from it.

**Academic literature**

- La Raja, R. J. and Schaffner, B. F. (2015). *Campaign Finance and Political
  Polarization: When Purists Prevail.* University of Michigan Press.
- Ansolabehere, S., de Figueiredo, J. M. and Snyder, J. M. Jr. (2003). "Why Is
  There So Little Money in U.S. Politics?" *Journal of Economic Perspectives*
  17(1): 105–130.
- *Citizens United v. Federal Election Commission*, 558 U.S. 310 (2010).
- *Buckley v. Valeo*, 424 U.S. 1 (1976).

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`ie_candidates.csv`](data/derived/ie_candidates.csv), [`ie_columns.csv`](data/derived/ie_columns.csv), [`ie_committees.csv`](data/derived/ie_committees.csv), [`ie_facts.csv`](data/derived/ie_facts.csv), [`ie_outliers.csv`](data/derived/ie_outliers.csv), [`ie_summary.csv`](data/derived/ie_summary.csv)

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
