---
title: "The Decennial Census"
subtitle: "Seven questions asked once a decade, and the people it does not find"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("census-decennial-code.R")
```

```{r setup, include=FALSE}
```

The Constitution orders a count. Article I, Section 2, Clause 3 requires an "actual Enumeration" of the whole population, once at the founding and then "within every subsequent Term of ten Years". That is why the census falls in years ending in zero.

Two words in that clause do most of the work. **Actual Enumeration** means a head count rather than an estimate. The clause then adds "in such Manner as they shall by Law direct," which hands Congress everything else.

The Census Act is what Congress directed. It puts the count in the hands of the Secretary of Commerce and fixes a single day, April 1, as the decennial census date, at 13 U.S.C. §141(a). Where a person was on that day is where they are counted, whatever happened the week before or the week after.

What the count decides is representation. The state totals decide how many of the 435 House seats each state gets, and with them how many electoral votes it casts. Those totals go to the President within nine months of Census Day under 13 U.S.C. §141(b). A separate statute, 2 U.S.C. §2a, turns them into seats.

Federal money follows the count as well. But apportionment is the constitutional job, and it is the reason this is an enumeration, an attempt to find everyone, rather than a survey of some.

Districts are the second job, and they are not constitutional. Inside the states the same count is what maps are drawn from, because a district has to hold about as many people as the one next to it. That delivery is required by Public Law 94-171, passed in 1975 and codified at 13 U.S.C. §141(c). The statute obliges the Bureau to hand every state legislature small-area counts within a year of Census Day. It does not oblige a single state to use them, which is a distinction this chapter comes back to.

This chapter is about that count on its own terms: what the decennial census actually asks, what it actually publishes, what it costs to run, and, in the last third, who it does not find.

Start with the thing people get wrong before anything else. The decennial census is not a rich source of information about Americans. It is an extraordinarily thin one.

## Seven questions

```{r form, comment="", class.output="rawblock"}
```
Nothing about income, education, language, occupation, commuting, disability, veteran status, health insurance or ancestry. Every one of those is the American Community Survey, which is a sample rather than a count, with different properties.

The telephone number is never published. It is there so the Bureau can call back about a missing or contradictory answer.

That is the entire census. The "long form", which asked many other questions, was retired after 2000. Those additional questions and others added since then are now the [American Community Survey](../census-acs/census-acs-brief.html), which is a sample of all persons instead of an enumeration.

So when a claim about income, education, language, commuting, disability or ancestry is attributed to "the census", it did not come from the census. It came from a survey run by the same agency, with a margin of error: a published range for how far off the number could be.

**An eighth question was nearly added, and the fight over it went to the Supreme Court.** In 2018 the Secretary of Commerce directed the Bureau to ask everyone about citizenship. The last time the decennial form asked that of every household was 1950.

Six lawsuits followed. In *Department of Commerce v. New York* (2019) the Court held that the Secretary did have the authority to ask. It also held that the reason he gave for asking did not match the record, and sent the decision back for review. The administration dropped the question in July 2019, and the form above is what went out.

## What seven questions become

The shape of the published file is set by statute, not by the Bureau. Public Law 94-171 tells it to give the states small-area counts of population, of people old enough to vote, of race and Hispanic origin, of housing and of group quarters. That list is the file. Here it is, and how it is produced into 'tables'.

```{r tabtab}
```

**`r NTAB` tables. `r nn(NCELL)` numbers per geography.** That is what the country redraws itself from.

Now watch the width accumulate, because it comes almost entirely from the last question. A count on its own is one number per place. Broken out by race it becomes P1's `r P1N`. Add Hispanic origin and P1 and P2 together are `r nn(W("P1") + W("P2"))`. Apply the 18-and-over line to both and they double to `r nn(RACECELL)` across P1 through P4 — **`r nn(RACECELL)` of the `r nn(NCELL)` numbers in the file exist because of one question.** The remaining `r nn(RESTCELL)` are the housing units and the group quarters.

That last question is also where the file's strangest arithmetic lives. Of P1's `r P1N` cells, nine are the single races and the summary lines. Of the `r COMBO` below them, five are subtotals: *population of two races*, of three, and so on. The other **`r COMBOR` count specific combinations**: 15 pairs, 20 triples, 15 quadruples, 6 quintuples, and one cell for people who marked all six.

That is a fact about the question rather than about the population, since most of those `r COMBOR` cells hold very small numbers. It is the subject of the [race chapter](../census-race/census-race-brief.html).

## What that looks like in one state

Take Georgia. Everything the file reports about the state that is not a race cell fits in two sentences. There were `r nn(POP)` people on census day, `r nn(G("Population 18 and over"))` of them 18 or over, living in `r nn(G("Housing units"))` housing units of which `r nn(G("Vacant housing units"))` stood empty. Another `r nn(G("Population in group quarters"))` were counted in group quarters rather than households. The vacant units are worth noticing. A count of people has to count the places people live, and the empty ones are how the count of the occupied ones can be checked at all.

The race cells are the rest of it, and they are where the width went.

```{r racetab}
```

Note what is *not* here. No income, no education, no language, no household composition. And no age either, past the 18-and-over line, though age *is* asked, as the form above shows. That line exists because Section 2 of the Voting Rights Act, 52 U.S.C. §10301, is enforced by comparing a minority group's voting-age population against the rest of a district. Age is collected and mostly not published, which is a distinction this chapter ends on.

The decennial resolves the country finely in **space** and hardly at all in **subject**. That trade is what the census is.

## Where a person is counted is a rule, not a fact

Every number above depends on a decision about where people live, and the Bureau publishes that decision in advance. For 2020 it is the *Final 2020 Census Residence Criteria and Residence Situations*, issued in February 2018 after public comment. It runs to dozens of living situations. Three principles hold it together.

Count people where they live and sleep most of the time. Count people in certain group facilities at the facility. Count people with no usual residence wherever they are on 1 April.

**The second principle is the one that moves whole neighbourhoods.** A person in a state or federal prison is counted at the prison. A student in a dormitory is counted at the college. A resident of a nursing home is counted at the nursing home, and a soldier in barracks at the barracks. None of those people chose the address the count gives them.

That is why the group-quarters line in the state table above is not a leftover category. `r nn(G("Population in group quarters"))` Georgians were counted at an address a rule assigned them rather than one they gave. A rural legislative district built around a large prison is built partly out of people the count placed there.

Some people are not counted at all, and the rule says so plainly. A foreign citizen visiting on holiday is not counted. A baby born on 2 April 2020 is not counted. A person who died on 31 March is not counted. Neither is the crew of a US-flagged ship sailing between two foreign ports on census day.

The strangest rule is the one for federal employees abroad, because it produces a number that exists for one purpose only. Someone stationed or assigned overseas on census day is counted in their **home state of record**, the state they enlisted from, and counted there *for apportionment purposes only*. They are not in that state's population for any other use of the file. [The apportionment chapter](../apportionment/apportionment-brief.html) watches that column decide a House seat.

### States may undo the prison rule for their own maps

Nothing obliges a legislature to draw districts the way the Bureau counted. New York and Maryland moved prisoners back to their last known address before drawing their 2010-cycle maps, and `r AD("States adding prisoner reallocation for 2020")` more states said they would do it for the 2020 round. That makes `r AD("States reallocating prisoners, both cycles")` in all. Another count puts `r AD("States whose 2010-cycle plans used a modified file")` states in the 2010 round drawing from a file they had altered, reassigning or excluding servicemembers, students or prisoners.

So the count is one file, and what a legislature draws from is not always that file. A published number can be moved after publication, by somebody who is not the Bureau, and the Bureau's table will not record that it happened.

## Both deadlines are statutory. Both were missed.

```{r deadtab}
```

The apportionment counts arrived **`r nn(dead$days_late[1])` days** after the date the statute names. The redistricting data arrived **`r nn(dead$days_late[2])` days** late, in the middle of the year the states had set aside to draw maps with it.

Nothing about that was hidden, and the reasons are well documented: a census already carrying undertested new systems ran into a pandemic that suspended field operations. The point for a reader of files is narrower. A deadline written into the United States Code is a *commitment*, not a guarantee. Every schedule that assumed it had to move, including every state's redistricting calendar in 2021.

It is not cheap to run, either. The 2020 census cost about **\$`r p1(CO("Cost of the 2020 census, through 2023") / 1e9)` billion** through 2023, which is **\$`r p2(CO("Cost per person counted"))` for each person counted**.

> **Before you read on, commit to a number.** The Bureau measures its own
> accuracy afterward and publishes the result. For the 2020 census, **what do
> you think the net undercount was for Black residents, and for Hispanic
> residents — and did those gaps get better or worse since 2010?**

## The count is not complete, and it never has been

A census attempts to count everyone and does not succeed. The Bureau knows this, measures it, and says so. After each census it runs a separate survey, the Post-Enumeration Survey, and estimates how far off the count was, and for whom.

```{r covtab}
```

The last two columns are two different questions, and keeping them apart is the whole discipline of reading this table. **Differs from zero** asks whether the 2020 rate is distinguishable from a perfect count. **Differs from 2010** asks whether the 2020 rate is distinguishable from the one before it. Subtraction will hand you a Change for every row; only the rows marked Yes in the last column have earned one.

A negative number is an undercount. Read the first column down and the shape of it is unmistakable:

- **Black residents, `r p2(abs(CV("Black or African American", "pes_2020")))`% undercounted.**
- **Hispanic residents, `r p2(abs(CV("Hispanic or Latino", "pes_2020")))`% undercounted.**
- **American Indian and Alaska Native residents on reservations, `r p2(abs(CV("American Indian or Alaska Native, on reservation", "pes_2020")))`% undercounted.**
- **White, not Hispanic residents, `r p2(CV("White, not Hispanic", "pes_2020"))`% *over*counted.**

Now read the change column, using the last column as a filter. Against 2010 the Hispanic undercount deepened by `r p2(abs(CV("Hispanic or Latino", "change")))` points, and the non-Hispanic White overcount grew by `r p2(CV("White, not Hispanic", "change"))` — two movements in opposite directions, both of which the Bureau calls real. **The gap did not narrow between 2010 and 2020. It widened at both ends.**

The Black undercount and the undercount on reservations also sit deeper in 2020 than in 2010, by `r p2(abs(CV("Black or African American", "change")))` and `r p2(abs(CV("American Indian or Alaska Native, on reservation", "change")))` points. Resist saying they grew. The Bureau cannot tell either movement apart from 2010. So the honest claim about those two groups is not that the undercount got worse. It is that it was large in 2010 and is large still.

This is the single most important thing to carry out of this chapter. The census is not wrong in a random direction. It is wrong by different amounts for different groups, in the same direction, census after census. Every rate in this book with a census population as its denominator inherits that error silently. The denominator is the bottom number, the thing being divided by.

## What a percentage is, in people

Percentages are easy to nod at. Here is what those rates are in bodies, using Georgia's published counts:

```{r scaletab}
```

The last column is stated in words rather than in signed numbers on purpose. A minus sign has already been used in this chapter to mean an undercount, and here it would have to mean the opposite. Read it as it is written.

The arithmetic is also not the one you would reach for first. A coverage rate is measured against the *true* population rather than the published one, so the implied gap is the published count divided by one plus the rate, minus the published count. For the first row that is `r nn(sc$implied_people[1])` people — not the `r nn(sc$published_in_georgia[1] * 0.033)` you get by taking `r p2(abs(sc$national_rate[1]))`% of the published figure.

**This is an illustration of scale, not a state estimate**, and the difference matters. The Bureau does not publish coverage rates by state and race, so the table applies *national* rates to *Georgia's* counts. What it buys is a sense of scale. Rates of three and five per cent, applied to one medium-sized state, are tens of thousands of people each. And a [House seat can turn on fewer people than that](../apportionment/apportionment-brief.html).

Note which White population is in the table. The Bureau reports its White rate for White residents who are **not** Hispanic, so that is the count it is applied to — `r nn(sc$published_in_georgia[3])`, not the `r nn(R("White alone"))` in Georgia's race table further up. Applying it to the larger figure would have put Georgia's Hispanic White residents under two rows of the same table, at rates of opposite sign.

## Nobody is allowed to correct it

The count is wrong in a patterned direction, the Bureau measures how wrong, and the law will not let anyone fix it. Those three facts sit together, and together they are why the coverage table matters more than it looks.

**The apportionment count may not be adjusted by sampling.** Sampling means measuring part of a population carefully and scaling the answer up to the whole. Title 13 tells the Secretary to use sampling in the Bureau's work "except for the determination of population for purposes of apportionment", at 13 U.S.C. §195. In *Department of Commerce v. U.S. House of Representatives* (1999) the Supreme Court read that section to bar a sampled apportionment count.

The statute is not tidy about this. Section 141(a) tells the Secretary to take the census in whatever "form and content" he determines, "including the use of sampling procedures and special surveys". Section 195 then takes apportionment back out. The two sections point opposite ways, and the Court resolved them against sampling.

**The Bureau is not obliged to adjust even when it holds the measurement.** After the 1990 census the Secretary declined to correct the published count using that decade's post-enumeration survey. In *Wisconsin v. City of New York* (1996) the Court upheld the refusal, citing the wide discretion Congress and Title 13 give over the form of the census.

Filling in a blank is a different thing, and it is allowed. Where nothing usable came back for an address, the Bureau guesses a value from the records around it. Utah argued after 2000 that this was sampling under another name. In *Utah v. Evans* (2002) the Court disagreed, holding that filling in blanks is not the same as estimating a whole from a part. The guessing moved a House seat from Utah to North Carolina, which is [the apportionment chapter's subject](../apportionment/apportionment-brief.html).

So the published count is the count that governs, and the Bureau's own estimate of how far off it is arrives separately, in a document nothing is required to act on. **A census is a legal object first and a measurement second.** The [coverage chapter](../census-coverage/census-coverage-brief.html) works through the measurement in the Bureau's own detail.

## Nothing requires a state to draw with it

The Constitution requires the count for one job only. Article I, Section 2 ties the apportionment of House seats to an "actual Enumeration" and says nothing at all about districts. Public Law 94-171 obliges the Bureau to deliver the redistricting file to every state legislature. It does not oblige any legislature to open it.

So whether a state must draw from the federal count is a question of that state's own constitution and statutes, and the answer is not the same everywhere. The National Conference of State Legislatures read all fifty and sorted them:

```{r usetab}
```

**`r UU("Explicitly required", "legislative")` states have written the federal census into their own law for legislative districts.** Another `r UU("Implied or in practice", "legislative")` rely on it without writing it down. In `r UU("Not required", "legislative")` states nothing requires it at all. Several of those constitutions were written in the nineteenth century, and they say why: the state may run its own enumeration if the federal one is late or unsatisfactory.

The congressional column is short by five states, because Appendix B names only one branch for Delaware, Montana, Nebraska, North Dakota and South Dakota. Four of the five had a single at-large House seat when the book went to press, and so no congressional map to draw.

Courts have allowed the alternatives, within one limit. Hawaii drew its legislative districts from registered voters rather than from the census. In *Burns v. Richardson* (1966) the Supreme Court let that stand. Its reason was that the districts came out close to what total population would have produced. *Ely v. Klahr* (1971) repeated that condition for Arizona. *Kirkpatrick v. Preisler* (1969) struck down a Missouri plan partly because the state never established how many eligible voters each district held.

The recent question runs the other way. In *Evenwel v. Abbott* (2016) Texas voters argued that drawing on total population diluted their votes, because some districts held many residents who could not vote. The Court held that states **may** use total population. It did not hold that they must, and it left the other direction open.

Two states have written the delay itself into their law. New York and Ohio provide for the federal count unless it is unavailable or late, in which case each may look elsewhere. Both provisions long predate 2021, when the redistricting file arrived `r nn(dead$days_late[2])` days late.

## What this chapter cannot tell you

**The coverage rates are keyed in, not fetched.** The Bureau publishes the Post-Enumeration Survey results as PDFs rather than as data, so those six rows were typed in by hand. They are frozen at the values in the March 2022 release, and if the Bureau revised them nothing here would notice.

Everything else is read out of the file itself, or computed from the statutory dates, and moves when they move: the table widths, the state's counts, the days late.

**The state-law table is keyed in too, and it is frozen at 2019.** Those fifty rows were transcribed from the National Conference of State Legislatures' *Redistricting Law 2020*, which is a PDF. A state that has amended its constitution since would still appear here as it stood then, and several redistricting reforms passed after that book went to press.

**That table and its own source do not fully agree.** NCSL's summary paragraph counts 21 states as explicitly requiring the census. Its state-by-state appendix, counted here, gives `r UU("Explicitly required", "legislative")` for legislative districts. Most of the gap is four states where the appendix splits the two chambers and the summary does not. The rest is a swap: the summary lists Mississippi, which the appendix has as implied, and omits Missouri, which the appendix has as explicit. The appendix is the per-state authority, so the appendix is what the table follows.

**Six groups are shown; the Bureau reported more.** It also found an undercount for the Some Other Race population, a rate not typed in here.

**Five residence situations are described; the rule has dozens.** The Bureau's residence criteria run to several pages of living arrangements, and this chapter names the handful that move the most people or that changed for 2020. Nothing here says how many people any one rule moved, because the published file does not break the count out that way.

And one of the six rows shown, Native Hawaiian or Other Pacific Islander, is the Bureau's own "not distinguishable from zero". It is carried rather than dropped, because a coverage table showing only the clear rows reads like a verdict rather than a measurement.

**The Bureau's race groups are broader than the state table's.** Coverage for Black, Asian, American Indian and Pacific Islander residents is reported *alone or in combination*, meaning everyone who marked that race whether or not they marked another beside it. The Georgia counts here are the *alone* categories. Only the Hispanic and the non-Hispanic White rows line up exactly. What separates them is the `r nn(R("Two or more races"))` Georgians who marked more than one race, and for the Black row above it means the implied figure is, if anything, on the low side.

**Net coverage error is not the same as accuracy.** A net rate near zero can hide a large undercount and a large overcount cancelling each other out, and the PES measures the household population — people in group quarters are handled separately. "How close did it come" is a bigger question than one number per group.

**The state statistics are Georgia's.** Housing vacancy, group-quarters share and the racial composition all differ by state, and a rural state would look different. What does not differ is the shape of the count: seven questions, `r NTAB` tables, once a decade.

**The published counts are not the collected counts.** Every figure above is a published one, and in 2020 the Bureau did something to the published figures it had never done at this scale. A file detailed enough to give one census block a count by race and by age is also detailed enough to identify the people living on it. So before release, a formal privacy system added noise to everything below the state level.

State totals were held exact, and Georgia's `r nn(POP)` is the counted figure. The noise beneath them was required to add back up to it.

The consequence worth carrying is not the noise itself but its invisibility. The altered file passes every check that could be run against it: no block with more voting-age people than people, no negative counts, nothing impossible anywhere. A reader who suspected the file had been changed and went looking would find nothing and would conclude, reasonably and wrongly, that nothing had been done to it. **The only reason anyone knows the counts were altered is that the Bureau said so.**

Keep that apart from coverage error, because they are opposite kinds of problem. Noise is deliberate, documented, applied to the file after the count, and undetectable from inside it. An undercount is unintended, happens to the count out in the field, and can only be found by going back out with a second survey and looking.

## Sources

- U.S. Census Bureau. *2020 Census P.L. 94-171 Redistricting Data Summary
  File*, Georgia — the state file as delivered to the states in August 2021.
  Every table width and every Georgia count above is read out of it directly;
  it is a pipe-delimited set of files with no column headers, so the layout
  documentation is not optional if you fetch it yourself.
  <https://www.census.gov/programs-surveys/decennial-census/about/rdo/summary-files.html>
- U.S. Census Bureau. "Census Bureau Releases Estimates of Undercount and
  Overcount in the 2020 Census", release CB22-CN.02, 10 March 2022 — the 2020
  Post-Enumeration Survey against the 2010 Census Coverage Measurement survey.
  <https://www.census.gov/newsroom/press-releases/2022/2020-census-estimates-of-undercount-and-overcount.html>
- U.S. Census Bureau. *2020 Census Questionnaire*, and the 2020 Census
  Apportionment Results, 26 April 2021.
- U.S. Government Accountability Office. *2020 Census*, high-risk series —
  the cost figure through 2023.
  <https://www.gao.gov/highrisk/decennial-census>
- U.S. Census Bureau. *Disclosure Avoidance for the 2020 Census: An
  Introduction*, 2021 — the Bureau's own account of the noise added below the
  state level, and of which totals were held invariant.
  <https://www.census.gov/library/publications/2021/decennial/2020-census-disclosure-avoidance-handbook.html>
- National Conference of State Legislatures. *Redistricting Law 2020*. Denver:
  NCSL, October 2019 — Chapter 1 (The Census), Appendix A (Census Residence
  Concepts) and Appendix B (Redistricting and the Use of Census Data). The
  fifty rows of the state-law table are transcribed from Appendix B; the
  residence rules and the prisoner-reallocation counts come from Chapter 1
  and Appendix C.
- U.S. Census Bureau. *Final 2020 Census Residence Criteria and Residence
  Situations*, 83 Fed. Reg. 5525 (8 February 2018).
  <https://www.federalregister.gov/documents/2018/02/08/2018-02370/final-2020-census-residence-criteria-and-residence-situations>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`adjustments.csv`](data/derived/adjustments.csv), [`census-use.csv`](data/derived/census-use.csv), [`cost.csv`](data/derived/cost.csv), [`coverage.csv`](data/derived/coverage.csv), [`deadlines.csv`](data/derived/deadlines.csv), [`georgia.csv`](data/derived/georgia.csv), [`race.csv`](data/derived/race.csv), [`scale.csv`](data/derived/scale.csv), [`tables.csv`](data/derived/tables.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `deadlines.csv` records what the statute required, what was delivered, and `days_late`. Both deadlines were missed. Work out who was waiting on each one and what they could not do meanwhile.
- `scale.csv` applies a national coverage rate to a state's published count to get `implied_people`. Do that arithmetic yourself for a different group. What does the size of the answer do to your confidence in the published number?
- `tables.csv` says how many cells come out of each published table. Multiply out what seven questions become. Where do all those cells come from, if only seven things were asked?
- `race.csv` gives each group's `share_of_state`. Add the shares. If they do not come to a hundred, work out why before you decide it is an error.

**Statutes**

- U.S. Const. art. I, § 2, cl. 3 (the "actual Enumeration" clause).
- Census Act, 13 U.S.C. § 141(a) (April 1 as the decennial census date),
  § 141(b) (apportionment counts to the President), and § 141(c) (redistricting
  data to the states), the last enacted as Public Law 94-171 (1975).
- 13 U.S.C. § 195 (sampling, except for apportionment).
- 2 U.S.C. § 2a (apportioning the seats from the counts).
- Voting Rights Act § 2, 52 U.S.C. § 10301 (the reason the file carries a
  voting-age line).

**Cases**

- *Burns v. Richardson*, 384 U.S. 73 (1966).
- *Kirkpatrick v. Preisler*, 394 U.S. 526 (1969).
- *Ely v. Klahr*, 403 U.S. 108 (1971).
- *Wisconsin v. City of New York*, 517 U.S. 1 (1996).
- *Department of Commerce v. U.S. House of Representatives*, 525 U.S. 316 (1999).
- *Utah v. Evans*, 536 U.S. 452 (2002).
- *Evenwel v. Abbott*, 578 U.S. 54 (2016).
- *Department of Commerce v. New York*, 588 U.S. 752 (2019).
