---
title: "The Decennial Census"
type: chapter
subtitle: "Seven questions asked once a decade, and the people it does not find"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("census-decennial-code.R")
```

```{r setup, include=FALSE}
```

The Constitution orders a count. Article I, Section 2, Clause 3 requires an "actual Enumeration" of the whole population, once at the founding and then "within every subsequent Term of ten Years". That is why the census falls in years ending in zero. **Actual Enumeration** means a head count rather than an estimate. The clause then hands Congress everything else, "in such Manner as they shall by Law direct."

The Census Act is what Congress directed. It fixes a single day, 1 April, as Census Day, at 13 U.S.C. §141(a). Where a person was living on that day is where they are counted, whatever happened the week before or the week after. So the decennial census is a snapshot, one moment once a decade, and everything about the instrument follows from that.

What the count decides is representation. The state totals fix how many of the 435 House seats each state gets, and with them its votes in the Electoral College. Districts are the second job, because inside each state the same count is what maps are drawn from. Public Law 94-171, passed in 1975, obliges the Bureau to hand every state legislature small-area counts within a year of Census Day. Federal money follows the count as well, but apportionment is the constitutional job, and it is why this is an attempt to find everyone rather than a survey of some.

This chapter is about that count on its own terms: what it asks, what it publishes, what it costs, and, in the last third, who it does not find. Start with the thing people get wrong before anything else. The decennial census is not a rich source of information about Americans. It is an extraordinarily thin one.

## Where the data comes from, and what it is for

The **Census Bureau** runs the count and publishes the results. What this chapter reads is the 2020 **P.L. 94-171 redistricting file** for Georgia, as it was delivered to the states in August 2021: four pipe-delimited files with no header row. Every table width and every Georgia count below is measured out of those files directly, rather than copied from documentation.

**Who is in it, and how they got there.** Everyone living in the United States on 1 April 2020, at the address the Bureau's published residence rules assign them. Three principles hold those rules together. Count people where they live and sleep most of the time. Count people in certain group facilities (prisons, dormitories, barracks, nursing homes) at the facility. Count people with no usual residence wherever they are on Census Day.

The second principle moves whole neighbourhoods. `r nn(G("Population in group quarters"))` Georgians were counted at an address a rule assigned them rather than one they gave, and a rural district built around a large prison is built partly out of people the count placed there. The rules also say who is counted nowhere: a foreign visitor, a baby born on 2 April 2020, a person who died on 31 March. Federal employees stationed overseas are counted in their home state for apportionment only. And a residence rule can be undone after the fact. `r AD("States reallocating prisoners, both cycles")` states now move prisoners back to their home addresses before drawing their own maps, so what a legislature draws from is not always the file the Bureau delivered.

**Who it was made for.** Congress, whose seats the state totals divide, and the fifty legislatures, which redraw their districts from the small-area counts. The file breaks out one demographic line, people 18 and over. That line exists because Section 2 of the Voting Rights Act is enforced by comparing a minority group's voting-age population against the rest of a district. Everyone else who divides by a census number is borrowing a legal record as a statistic.

**What it costs to get.** The file is free: open downloads, no key, no account. Producing it cost about \$`r p1(CO("Cost of the 2020 census, through 2023") / 1e9)` billion through 2023, which is \$`r p2(CO("Cost per person counted"))` for each person counted.

Both statutory delivery deadlines were missed in 2020–21, in a census that ran into a pandemic. The apportionment counts arrived `r nn(dead$days_late[1])` days after the date the statute names. The redistricting data arrived `r nn(dead$days_late[2])` days late, in the middle of the year the states had set aside to draw maps with it. A deadline in the United States Code is a commitment, not a guarantee.

**What has already been done to it.** The published counts below the state level are not the collected counts. A file detailed enough to describe one block by race and age can identify the people on it, so before release a formal privacy system added noise to everything beneath the state totals. Georgia's `r nn(POP)` is the counted figure, and the block counts underneath were required to sum back to it. The alteration is invisible from inside the file — the [chapter on what a census can and cannot do](../census-source/census-source-brief.html) demonstrates that no internal check can find it. Separately, where nothing usable came back from an address, the Bureau fills in a guess from the records around it, which the Supreme Court has held is not sampling (*Utah v. Evans*, 2002).

## Seven questions

```{r form, comment="", class.output="rawblock"}
```

Nothing about income, education, language, occupation, commuting, disability, veteran status, health insurance or ancestry. The "long form" that asked such things was retired after 2000. Its questions now live in the [American Community Survey](../census-acs/census-acs-brief.html), a sample rather than a count. The ACS publishes a **margin of error**, a stated range for how far off each estimate could be, beside every number. So when a claim about income or language is attributed to "the census", it did not come from the census. The telephone number is never published; it is there so the Bureau can call back about a missing or contradictory answer.

An eighth question was nearly added, and the fight over it went to the Supreme Court. In 2018 the Secretary of Commerce directed the Bureau to ask every household about citizenship, which the form last did in 1950. In *Department of Commerce v. New York* (2019) the Court held that the Secretary had the authority to ask, and also that the reason he gave did not match the record. The administration dropped the question in July 2019, and the form above is what went out.

## What seven questions become

The shape of the published file is set by statute, not by the Bureau. Public Law 94-171 tells it to give the states small-area counts of population, of people old enough to vote, of race and Hispanic origin, of housing, and of group quarters. That list is the file:

```{r tabtab}
```

**`r NTAB` tables. `r nn(NCELL)` numbers per geography.** That is what the country redraws itself from. The file repeats it at every level of census geography, from the state down to each census block — the [demographics brief](../demographics/demographics-brief.html) reads the same counts at eight of those levels.

Now watch the width accumulate, because it comes almost entirely from the last question on the form. A count on its own is one number per place. Broken out by race it becomes P1's `r P1N` cells. Add Hispanic origin, and P1 and P2 together are `r nn(W("P1") + W("P2"))`. Apply the 18-and-over line to both and they double to `r nn(RACECELL)` across P1 through P4 — **`r nn(RACECELL)` of the `r nn(NCELL)` numbers in the file exist because of one question.**

Of P1's `r P1N` cells, `r COMBOR` count specific *combinations* of races, most holding very small numbers. That is a fact about the question rather than about the population, and it is the [race brief's](../census-race/census-race-brief.html) subject.

Take Georgia. Everything the file reports about the state that is not a race cell fits in two sentences. There were `r nn(POP)` people on Census Day, `r nn(G("Population 18 and over"))` of them 18 or over, living in `r nn(G("Housing units"))` housing units of which `r nn(G("Vacant housing units"))` stood empty. Another `r nn(G("Population in group quarters"))` were counted in group quarters rather than households. The race cells are the rest of it, and they are where the width went.

```{r racetab}
```

The decennial resolves the country finely in **space** and hardly at all in **subject**. That trade is what the census is.

One more thing the statute does not do. It obliges the Bureau to deliver this file to every legislature, and it obliges no legislature to open it. By the National Conference of State Legislatures' state-by-state reading, `r UU("Explicitly required", "legislative")` states have written the federal count into their own law for legislative districts. Another `r UU("Implied or in practice", "legislative")` rely on it without writing it down, and in `r UU("Not required", "legislative")` states nothing requires it at all.

## The count is not complete, and it never has been

A census attempts to count everyone and does not succeed. The Bureau knows this, measures it, and says so. After each census it runs a separate survey, the Post-Enumeration Survey, and estimates how far off the count was, and for whom.

```{r cov-static, eval=!knitr::is_html_output(), fig.height=3.2}
```

```{r cov-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Net coverage error by group: the 2010 census against the 2020 census.** A negative number is an undercount. Each group has exactly two measurements, a decade apart, and that is why the chart is a pair of dots joined by a line rather than a pair of bars. The dots place each census against a perfect count at zero, and the line between them is the movement — what to read is its direction. Black, Hispanic and American Indian residents sit below zero in both censuses; non-Hispanic White residents sit above it in both. And between 2010 and 2020 the ends moved apart, not together.

The figure hides one discipline the numbers demand, so here is the same table with the Bureau's own significance tests attached:

```{r covtab}
```

The last two columns are two different questions, and keeping them apart is the whole skill of reading this table. **Differs from zero** asks whether the 2020 rate is distinguishable from a perfect count. **Differs from 2010** asks whether the 2020 rate is distinguishable from the one before it. Subtraction will hand you a Change for every row; only the rows marked Yes in the last column have earned one.

Read the 2020 column first: **Black residents, `r p2(abs(CV("Black or African American", "pes_2020")))`% undercounted; Hispanic residents, `r p2(abs(CV("Hispanic or Latino", "pes_2020")))`% undercounted; American Indian residents on reservations, `r p2(abs(CV("American Indian or Alaska Native, on reservation", "pes_2020")))`% undercounted; White, not Hispanic residents, `r p2(CV("White, not Hispanic", "pes_2020"))`% *over*counted.** Now read the change column through the filter of the last column. Against 2010 the Hispanic undercount deepened by `r p2(abs(CV("Hispanic or Latino", "change")))` points and the non-Hispanic White overcount grew by `r p2(CV("White, not Hispanic", "change"))` — two movements in opposite directions, both of which the Bureau calls real. **The gap did not narrow between 2010 and 2020. It widened at both ends.**

The Black undercount and the undercount on reservations also sit deeper in 2020, but the Bureau cannot tell either movement apart from 2010. So the honest claim there is not that they grew. It is that they were large in 2010 and are large still.

This is the single most important thing to carry out of this chapter. The census is not wrong in a random direction. It is wrong by different amounts for different groups, in the same direction, census after census. Every rate in this book with a census population as its **denominator**, the bottom number of the fraction, inherits that error silently. The [coverage brief](../census-coverage/census-coverage-brief.html) reads the Bureau's estimates in their own detail, state by state; this chapter carries only the headline.

Percentages are easy to nod at, so here is what those rates are in bodies, using Georgia's published counts:

```{r scaletab}
```

**This is an illustration of scale, not a state estimate.** The Bureau does not publish coverage rates by state and race, so the table applies national rates to Georgia's counts. The arithmetic is also not the one you would reach for first. A coverage rate is measured against the *true* population rather than the published one. So the implied gap is the published count divided by one plus the rate, minus the published count — `r nn(sc$implied_people[1])` people for the first row. Rates of three and five per cent, applied to one medium-sized state, are tens of thousands of people each, and a [House seat can turn on fewer people than that](../apportionment/apportionment-brief.html).

The count is wrong in a patterned direction, the Bureau measures how wrong, and the law will not let anyone fix it. **The apportionment count may not be adjusted by sampling** — measuring part of a population carefully and scaling the answer up. The Supreme Court read 13 U.S.C. §195 to bar that, in *Department of Commerce v. U.S. House of Representatives* (1999).

And the Bureau is not obliged to adjust even when it holds the measurement. After the 1990 census the Secretary declined to correct the published count with that decade's survey, and *Wisconsin v. City of New York* (1996) upheld the refusal. So the published count is the count that governs, and the Bureau's own estimate of how far off it is arrives separately, in a document nothing is required to act on. **A census is a legal object first and a measurement second.**

## What this data cannot tell you

**The coverage rates are keyed in, not fetched.** The Bureau publishes the Post-Enumeration Survey results as PDFs rather than as data, so those six rows were typed in by hand from the March 2022 release. If the Bureau revised them, nothing here would notice. Everything else is read out of the file itself or computed from the statutory dates, and moves when they move.

**Six groups are shown; the Bureau reported more.** It also found an undercount for the Some Other Race population, a rate not typed in here. And one of the six rows, Native Hawaiian or Other Pacific Islander, is the Bureau's own "not distinguishable from zero". It is carried rather than dropped, because a coverage table showing only the clear rows reads like a verdict rather than a measurement.

**The Bureau's race groups are broader than the state table's.** Coverage for Black, Asian, American Indian and Pacific Islander residents is reported *alone or in combination*: everyone who marked that race, whether or not they marked another beside it. The Georgia counts here are the *alone* categories, so for the Black row the implied figure is, if anything, on the low side.

**Net coverage error is not the same as accuracy.** A net rate near zero can hide a large undercount and a large overcount cancelling each other out, and the PES measures the household population; people in group quarters are handled separately. "How close did it come" is a bigger question than one number per group.

**The state-law counts are frozen at 2019.** They were transcribed from NCSL's *Redistricting Law 2020*, whose own summary paragraph disagrees with its state-by-state appendix; the appendix is the per-state authority, so the appendix is what the numbers follow. A state that has amended its law since would still appear here as it stood then.

**Five residence situations are described; the rule has dozens.** The Bureau's residence criteria run to pages of living arrangements, and nothing in the published file says how many people any one rule moved.

**The state statistics are Georgia's.** Housing vacancy, group-quarters share and racial composition all differ by state. What does not differ is the shape of the count: seven questions, `r NTAB` tables, once a decade.

Finally, keep the two kinds of wrongness apart, because they are opposite problems. The privacy noise is deliberate, documented, applied to the file after the count, and undetectable from inside it. An undercount is unintended, happens out in the field, and can only be found by going back out with a second survey and looking.

## What you should have learned

- The whole instrument is seven questions, and the published redistricting file is `r NTAB` tables — `r nn(NCELL)` numbers per place, `r nn(RACECELL)` of them produced by the race question alone.
- The count is a snapshot of one day, taken at addresses the residence rules assign: `r nn(G("Population in group quarters"))` Georgians were counted at group quarters rather than at a home they chose.
- The count misses people unevenly and in the same direction each time. In 2020 it undercounted Black residents by `r p2(abs(CV("Black or African American", "pes_2020")))`%, Hispanic residents by `r p2(abs(CV("Hispanic or Latino", "pes_2020")))`%, and overcounted non-Hispanic White residents by `r p2(CV("White, not Hispanic", "pes_2020"))`% — and the 2010–2020 gap widened at both ends.
- The law makes the published count the count that governs: no sampling for apportionment, and no obligation to adjust even when the Bureau holds the measurement.
- Counting the country cost about \$`r p1(CO("Cost of the 2020 census, through 2023") / 1e9)` billion, \$`r p2(CO("Cost per person counted"))` per person counted, and both statutory delivery deadlines were still missed — by `r nn(dead$days_late[1])` and `r nn(dead$days_late[2])` days.

## Where this data appears in this book

This chapter anchors the cluster on the decennial census. Each brief in the cluster asks one question of the count introduced here:

- [Apportionment: How a Count Becomes Seats](../apportionment/apportionment-brief.html) — the constitutional purpose carried out: the formula, and a seat decided by a margin smaller than the instrument's error.
- [Regional Population Shift and House Seats](../regional-shift/regional-shift-brief.html) — six reapportionments laid end to end, and the region that lost its majority of the House.
- [Measuring Race at Eight Geographic Scales](../demographics/demographics-brief.html) — the same race counts read at every level the file publishes, and why only the smallest describes anyone.
- [When a Scatter Plot Has Too Many Points](../overplotting/overplotting-brief.html) — every populated Georgia block on one chart, and what that density does to a scatter plot.

## Sources

**Data**

- U.S. Census Bureau. *2020 Census P.L. 94-171 Redistricting Data Summary
  File*, Georgia — the state file as delivered to the states in August 2021.
  Every table width and every Georgia count above is read out of it directly;
  it is a pipe-delimited set of files with no column headers, so the layout
  documentation is not optional if you fetch it yourself.
  <https://www.census.gov/programs-surveys/decennial-census/about/rdo/summary-files.html>
- U.S. Census Bureau. "Census Bureau Releases Estimates of Undercount and
  Overcount in the 2020 Census", release CB22-CN.02, 10 March 2022 — the 2020
  Post-Enumeration Survey against the 2010 Census Coverage Measurement survey.
  <https://www.census.gov/newsroom/press-releases/2022/2020-census-estimates-of-undercount-and-overcount.html>
- U.S. Census Bureau. *2020 Census Questionnaire*, and the 2020 Census
  Apportionment Results, 26 April 2021.
- U.S. Government Accountability Office. *2020 Census*, high-risk series —
  the cost figure through 2023.
  <https://www.gao.gov/highrisk/decennial-census>
- U.S. Census Bureau. *Disclosure Avoidance for the 2020 Census: An
  Introduction*, 2021 — the Bureau's own account of the noise added below the
  state level, and of which totals were held invariant.
  <https://www.census.gov/library/publications/2021/decennial/2020-census-disclosure-avoidance-handbook.html>
- National Conference of State Legislatures. *Redistricting Law 2020*. Denver:
  NCSL, October 2019 — Chapter 1 (The Census), Appendix A (Census Residence
  Concepts) and Appendix B (Redistricting and the Use of Census Data). The
  state-law counts are computed from a transcription of Appendix B; the
  residence rules and the prisoner-reallocation counts come from Chapter 1
  and Appendix C.
- U.S. Census Bureau. *Final 2020 Census Residence Criteria and Residence
  Situations*, 83 Fed. Reg. 5525 (8 February 2018).
  <https://www.federalregister.gov/documents/2018/02/08/2018-02370/final-2020-census-residence-criteria-and-residence-situations>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`adjustments.csv`](data/derived/adjustments.csv), [`census-use.csv`](data/derived/census-use.csv), [`cost.csv`](data/derived/cost.csv), [`coverage.csv`](data/derived/coverage.csv), [`deadlines.csv`](data/derived/deadlines.csv), [`georgia.csv`](data/derived/georgia.csv), [`race.csv`](data/derived/race.csv), [`scale.csv`](data/derived/scale.csv), [`tables.csv`](data/derived/tables.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `deadlines.csv` records what the statute required, what was delivered, and `days_late`. Both deadlines were missed. Work out who was waiting on each one and what they could not do meanwhile.
- `scale.csv` applies a national coverage rate (`national_rate`) to a state's published count to get `implied_people`. Do that arithmetic yourself for a different group. What does the size of the answer do to your confidence in the published number?
- `tables.csv` says how many `cells` come out of each published table. Multiply out what seven questions become. Where do all those cells come from, if only seven things were asked?
- `race.csv` gives each group's `share_of_state`. Add the shares. If they do not come to a hundred, work out why before you decide it is an error.
- `census-use.csv` counts states by what their own law says about using the federal count (`legislative`, `congressional`). Which arrangement surprises you more — a state obliged by its own constitution, or a state obliged by nothing? What would the second kind of state need, to draw maps without the Bureau?

**Statutes**

- U.S. Const. art. I, § 2, cl. 3 (the "actual Enumeration" clause).
- Census Act, 13 U.S.C. § 141(a) (April 1 as the decennial census date),
  § 141(b) (apportionment counts to the President), and § 141(c) (redistricting
  data to the states), the last enacted as Public Law 94-171 (1975).
- 13 U.S.C. § 195 (sampling, except for apportionment).
- 2 U.S.C. § 2a (apportioning the seats from the counts).
- Voting Rights Act § 2, 52 U.S.C. § 10301 (the reason the file carries a
  voting-age line).

**Cases**

- *Wisconsin v. City of New York*, 517 U.S. 1 (1996).
- *Department of Commerce v. U.S. House of Representatives*, 525 U.S. 316 (1999).
- *Utah v. Evans*, 536 U.S. 452 (2002).
- *Department of Commerce v. New York*, 588 U.S. 752 (2019).
