---
title: "Part III — Elections"
subtitle: "The result, and the machinery that produced it"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-3-elections-code.R")
```

```{r setup, include=FALSE}
```

This is the largest part of the book, `r NCH` chapters. It is large because "election data" is two different things that share a name.

**One is the result.** Who won, by how much, certified by the officials whose job is to certify it. It is the most authoritative data in this book: somebody signed their name to it.

**The other is the machinery.** Who was registered, who was removed from the rolls, how many ballots arrived late or were rejected, what a precinct is and where its boundary runs. Nobody signed this. It is the paperwork of running an election, published unevenly, in whatever format the office that kept it happened to use.

Both get called election data. A surprising number of published claims come from quietly swapping one for the other — using registration records to talk about who voted, or certified totals to talk about who was allowed to.

## The result is authoritative and still not a dataset

Here is the thing that surprises people about certified returns: **there is no national file.** No federal agency collects election results and publishes them as a table. Elections are run by states, counted by counties, and certified in fifty-one different formats — a PDF here, a spreadsheet there, a web page somewhere else.

So every national number you have ever seen about an American election was assembled by somebody, privately, from those pieces. Assembly involves judgment, and judgment can be wrong. Several chapters here are about what happens at the seams.

## The machinery is public by law and still hard to get

The second half runs into the opposite problem. The voter file is public in every state — a list of registered voters, their addresses, and whether they voted in each election. Public by law. **Downloadable in almost none of them.** You may be asked to sign something, pay something, explain yourself, or appear in person.

The chapters in this run keep hitting a version of that wall, and it is worth noticing as a pattern rather than as bad luck. *Public* and *obtainable* are different words, and the gap between them is where a lot of what we do not know about American elections actually lives.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

The chapter list is long. It runs the result first, all the way through, and then starts again with the machinery — which is why a second set of introductory chapters appears partway down rather than at the top.

## Where this part sits

```{r reusetab}
```

**`r nn(RU("Chapters elsewhere that read this part\'s files"))` chapters outside this part read its files**, the most of any part in the book. Election returns turn out to be the thing everything else wants to be checked against: a survey's claim about turnout, a guess about who voted which way, a map's effect on an outcome. If you want to know whether a method works, you point it at a result somebody certified and see whether it gets the right answer.

## How to read this part

The two runs can be read independently. If you want the result, start at [what a return is](../returns-source/returns-source-brief.html). If you want the machinery, start at [the voter file](../voter-files-source/voter-files-source-brief.html) and read forward from there.

One chapter here is fixed to a date rather than a place in the sequence. [Election night](../election-night/election-night-brief.html) runs on the night of the election, with results that do not exist until then, and it grades predictions made earlier in the term against what the country actually did.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens.
- Certified returns come from the state and county officers who publish them,
  and from the Clerk of the U.S. House for federal races; the voter-file and
  administration chapters name their state sources individually. Each chapter
  states where its own copy came from and when it was retrieved.
