---
title: "Who Held the Seat, and Who Won the Election"
type: chapter
subtitle: "A roster is not a return, and the gap between them is 2,344 people"
author: "Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("rosters-source-code.R")
```

```{r setup, include=FALSE}
```

This part is called the record of an outcome, and most of it earns the name: certified county returns, precinct returns, the ballots themselves. Each one records a **contest** — on one day, in one place, these people got these votes.

Four chapters in this part do not use returns at all. `midterm-loss` counts seats each party held. `retirements` and `careers` follow members in and out of office. `primary-defeats` asks which incumbents were denied renomination. What they run on is a **roster**: a list of who was a member of Congress, and when.

A roster and a return look interchangeable, and they are not. This chapter is about the difference, which is measurable.

## A return records a contest; a roster records a membership

```{r rostab}
```

The [returns chapter](../returns-source/returns-source-brief.html) built this part on a single limit: a return is complete, official, and **permanently silent about who voted**. The ballot is secret, so no return links a vote to a voter.

A roster has the mirror-image limit, and it is easy to miss because it points the other way. **A roster is silent about votes.** It will tell you exactly who sat in the 47th Congress and for how long, and it does not contain a single ballot. The party-divisions series that `midterm-loss` runs on is the clean case: `r cv("Columns in the party-divisions series carrying a vote total")` of its columns carry a vote total. It counts seats.

That is a fair trade as long as seats and elections correspond. Mostly they do.

> **Before you read on, commit to a number.** Across
> `r NCONG` Congresses there have been `r nn(SEATS)` seat-terms to fill.
> **How many different people do you think have filled them?** The same number?
> A few dozen more? Write it down.

## More people than there are seats

```{r exctab}
```

Across `r NCONG` Congresses, **`r nn(MEM)` people have served in `r nn(SEATS)` seats** — an excess of **`r nn(XS)`**, or `r XSP`% more people than there were seats for. The worst single case is the `r WW` Congress, which seated `r WM` members in `r WS` seats.

Every one of those `r nn(XS)` arrived **mid-Congress**, and that is not an approximation — it is the same number counted a second way, which the build asserts rather than assumes. They were appointed, or won a special election, or were seated once a contested result was resolved. **No general election for that Congress put them there.** `r nn(DIED)` of the vacancies they filled were created by a death in office.

This is the practical consequence: a file of general-election returns, however complete, cannot account for `r XSP`% of the people who have held a seat in the House. Ask a returns file who represented a district in 1815 and it will give you the person who won the election, which in `r WW` was frequently not the person doing the job.

## Most departures are not decided by voters

The gap runs the other way too. Ask why a member stopped serving, and the election explains less than you would expect.

```{r exttab}
```

Over the `r NCEX` most recent Congresses, `r nn(DEP)` members stopped serving. A voter decided **`r nn(BYV)`** of those departures — losing a general election, or being denied renomination in a primary. The other **`r nn(NOV)`**, or **`r NOVP`%**, were not decided by any voter at all: the member retired, ran for the Senate, or died.

**Roughly two thirds of the turnover in the House is not an electoral event.** That is a fact about the institution which no return can report. It is exactly the fact that `retirements` and `careers` exist to study.

## What a roster can settle that a return cannot

The point is not that rosters are inferior. They answer questions returns cannot touch.

A return is a snapshot of one day. A roster is a **duration**, and duration is what a career is made of. Of `r nn(NCAR)` careers in the file, **`r nn(NGAP)` are not one continuous stretch**, or `r p1(GAPP)`% of them. The member left and came back. A returns file records each of those wins separately and has no way to know they belong to the same person's career.

A roster is also the only way to answer *who was actually in the room*. Whether a bill passed depends on who held the seat that week, not on who won the election two years earlier. That is why the roll-call chapters in Part IV begin from a membership file rather than from returns.

## What this chapter cannot tell you

**It cannot tell you why anyone left.** The outcome column is a classification built from what the file records — appeared on a ballot, did not, died. A retirement that was really a pre-emption of certain defeat is recorded as a retirement, because that is all the roster knows.

**It cannot reconcile the two kinds of file for you.** Matching a roster to a returns file is its own problem, and `primary-defeats` is where it is done. That chapter matches `r cv("Incumbent-primary rows available for matching")` incumbent-primary rows to members by name and district, and reports the failures rather than dropping them.

**It cannot be extended backwards indefinitely.** The membership series runs across `r NCONG` Congresses; the departure classification depends on candidate filings and runs over `r NCEX`. The two-thirds finding is a modern one, and this chapter does not claim it held in 1815.

**And it cannot tell you who the voters wanted.** That is the returns file's question, and the point of putting the two beside each other is that neither one answers both.

## Sources

Nothing in this chapter was downloaded. Every figure is read from the derived tables four other chapters already built:

- [`retirements`](../retirements/retirements-brief.html) — `departures.csv`,
  `exits.csv`. Members and seats by Congress, and why each member stopped
  serving.
- [`careers`](../careers/careers-brief.html) — `careers.csv`. One row per
  career, with the gaps.
- [`midterm-loss`](../midterm-loss/midterm-loss-brief.html) —
  `house_midterms.csv`. Party divisions by election.
- [`primary-defeats`](../primary-defeats/primary-defeats-brief.html) —
  `incumbents.csv`. Incumbents facing a primary, matched to the roster.

Three files sit underneath them, and each is published free of charge:
Voteview's membership file (<https://voteview.com/>), the Office of the
Historian's Party Divisions tables
(<https://history.house.gov/Institution/Party-Divisions/Party-Divisions/>) and
the Federal Election Commission's *Federal Elections* compendium
(<https://www.fec.gov/introduction-campaign-finance/election-results-and-voting-information/>).
The four chapters above document what each one cost to get and what it left
out; this chapter's subject is what the three have in common.

Every figure above is recomputed from those tables, and checked against the values asserted in `checks.csv`. The table is printed here verbatim.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`careers.csv`](data/derived/careers.csv), [`excess.csv`](data/derived/excess.csv), [`exits.csv`](data/derived/exits.csv), [`rosters.csv`](data/derived/rosters.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `excess.csv` shows more members than seats in almost every Congress, with `mid_congress` and `died` beside it. Follow `excess` across the series. Is the churn falling?
- `exits.csv` flags whether each departure was `voter_decided`. Work out what share of exits the voters actually chose. Does that match how congressional turnover is usually described?
- `rosters.csv` says what each source `answers` and what it is `silent_about`. Pick a question you would want answered and decide whether any roster can settle it.
- A roster and a return disagree about how many people held a seat. Using `excess.csv`, work out where the extra people come from.

```{r checks}
```
