---
title: "Data Session 13: The Ballots Themselves"
subtitle: "84-355 Democracy's Data | Alaska's cast vote record"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"He won in the final round after transfers."*

Every session so far has used **totals** — how many people voted for X. Totals
are what elections publish, because the ballot is secret and a total is what is
left once you throw the ballots away.

Alaska does not throw them away.

Since adopting ranked-choice voting, Alaska publishes its **cast vote record**:
one row per scanned ballot, showing every ranking on it. No name, no address —
but the actual choices, one voter at a time. It is the closest thing to a public
record of individual votes that exists in this country, and almost nowhere else
has one.

**Today we recount an election from the ballots up**, and then check our answer
against the state's.

The contest is the 2024 race for Alaska's single seat in the U.S. House:
**Mary Peltola**, the Democratic incumbent, against **Nick Begich**, with two
minor candidates. Voters could rank up to four.

# Part 1 — What one row is

```{r load}
s <- read.csv("data/sequences.csv", stringsAsFactors = FALSE)

nrow(s)
sum(s$ballots)
head(s, 6)
```

**340,778 ballots, collapsed into 926 distinct ranking sequences.** Every
different *pattern* of ranking that any Alaskan produced, and how many people
produced it.

Read the first two rows. The most common ballot in Alaska was **"Begich, and
nobody else"** — 121,602 of them. The second most common was **"Peltola, and
nobody else"** — 86,278.

```{r bullets}
ranks_used <- rowSums(s[, grep("^rank", names(s))] != "")
blank <- sum(s$ballots[ranks_used == 0])
one   <- sum(s$ballots[ranks_used == 1])
voted <- sum(s$ballots) - blank

c(blank_on_this_race = blank,
  ranked_only_one = one,
  pct_of_voters = round(100 * one / voted, 1))
```

**Given four rankings, 65% of Alaskans used one.**

That is the first real finding, and it is a fact about ranked-choice voting that
neither its advocates nor its opponents advertise. The system's whole promise is
that you can express a second preference without wasting your vote. Two-thirds
of voters declined.

```{r ranks-plot}
u <- tapply(s$ballots, ranks_used, sum)
barplot(u, names.arg = names(u), col = "#2c7fb8",
        xlab = "number of candidates ranked", ylab = "ballots")
```

# Part 2 — Who ranked whom

```{r pairs}
top <- aggregate(ballots ~ rank1 + rank2, data = s[s$rank2 != "", ], sum)
top <- top[order(-top$ballots), ]
head(top, 6)
```

The most common pair is **Peltola then Hafner** — 39,684 ballots — followed by
**Begich then Howe**, 24,386. Voters who bothered with a second choice mostly
moved from a major candidate to a minor one.

```{r crossover}
maj <- c("Begich, Nick", "Peltola, Mary S.")
cross <- top[top$rank1 %in% maj & top$rank2 %in% maj, ]
cross
```

Two things in that table, and the second one is not what we went looking for.

**First, the crossover.** 10,771 ballots ranked Peltola then Begich, and 6,663
ranked Begich then Peltola — **17,434 in all, about 5% of the electorate.**
Ranked-choice voting is supposed to reward candidates who can be somebody's
second choice across the divide. One voter in twenty took the offer.

**Second, look at rows three and four.** **2,455 ballots rank Begich first and
Begich second.** Another 1,049 do the same for Peltola. The same person, twice.

```{r dupes}
dup <- top[as.character(top$rank1) == as.character(top$rank2), ]
c(begich = dup$ballots[dup$rank1 == "Begich, Nick"],
  peltola = dup$ballots[dup$rank1 == "Peltola, Mary S."],
  all_candidates = sum(dup$ballots))
```

**3,790 ballots in all name the same candidate twice.**

Nothing stopped them: the scanner records the marks it sees, and a voter who
wants to make very sure can fill the same column twice. **The tabulator ignores
the duplicate** and moves on to the next distinct name — so these ballots count
normally — but the behaviour is in the record, and you would never see it in a
total. **This is what individual-level data buys you.**

Which brings the argument back to the two-thirds. If ranking one candidate and
stopping is *confusion* — people not understanding the offer at the top of the
ballot — then the rate should be about the same whichever candidate they marked.
Confusion has no reason to prefer a front-runner.

> **Before you run the next chunk, commit to an answer.** It splits that
> `r round(100 * one / voted, 1)`% by who the voter put first. **Do Begich's
> voters and Hafner's voters look alike?** Write down two things: whether the
> gap between them is small — within a few points, which is what confusion
> would give you — or large; and if it is large, whose supporters stopped after
> one ranking more often.

```{r bullet-by-cand}
cands <- c("Begich, Nick", "Peltola, Mary S.", "Howe, John Wayne", "Hafner, Eric")

bul <- data.frame(
  first_choice = sub(",.*", "", cands),
  voters = sapply(cands, function(c_) sum(s$ballots[s$rank1 == c_])),
  pct_bullet = sapply(cands, function(c_)
    round(100 * sum(s$ballots[s$rank1 == c_ & ranks_used == 1]) /
                sum(s$ballots[s$rank1 == c_]), 1)),
  row.names = NULL)

bul <- bul[order(-bul$voters), ]
bul
```

> **They do not look alike at all.** `r bul$pct_bullet[1]`% of Begich's voters
> ranked nobody else; `r bul$pct_bullet[nrow(bul)]`% of Hafner's did. The gap is
> `r round(bul$pct_bullet[1] - bul$pct_bullet[nrow(bul)])` points wide, and it
> runs the opposite way from the guess most people write down when they assume
> the smallest candidate's supporters were the most lost. The pattern runs
> cleanly down the ballot: the bigger your candidate, the likelier you were to
> stop after one.
>
> **That is not confusion. It is arithmetic voters appear to have done.** If
> your candidate is the front-runner, a second ranking only ever gets read if
> your first choice is eliminated — and your first choice is not going to be
> eliminated. If your candidate is polling at one percent, a second ranking is
> the only part of your ballot that will do anything. There is a second reading
> this file cannot separate from the first — **campaign instruction**, since
> "rank one and stop" is a documented tactic in these races — and either way it
> is something no published total would ever show you.

# Part 3 — Count it yourself

Instant-runoff is simple enough to write out. Count everyone's top *continuing*
choice; if nobody has a majority, eliminate the lowest and look again at those
ballots' next choice; repeat.

```{r irv}
cols <- grep("^rank", names(s), value = TRUE)

irv <- function(dat, eliminated = character(0)) {
  pick <- rep(NA_character_, nrow(dat))
  for (i in seq_len(nrow(dat))) {
    for (cl in cols) {
      v <- dat[i, cl]
      if (v == "") next
      if (v == "OVERVOTE") break          # an overvote stops the ballot
      if (v %in% eliminated) next
      pick[i] <- v; break
    }
  }
  tapply(dat$ballots, pick, sum)
}

elim <- character(0)
for (round in 1:5) {
  cnt <- sort(irv(s, elim), decreasing = TRUE)
  active <- sum(cnt)
  cat(sprintf("\nRound %d  (continuing ballots: %s)\n", round,
              format(active, big.mark = ",")))
  print(round(100 * cnt / active, 2))
  if (cnt[1] > active / 2) { cat("  WINNER:", names(cnt)[1], "\n"); break }
  elim <- c(elim, names(cnt)[length(cnt)])
  cat("  eliminate:", names(cnt)[length(cnt)], "\n")
}
```

**Begich wins with 51.22%.** You just recounted a congressional election from
the ballots.

# Part 4 — Check it against the state

```{r official}
o <- read.csv("data/official_rounds.csv", stringsAsFactors = FALSE)
o[o$round == "Round 3", ]
```

Alaska's official final round: **Begich 51.22%, Peltola 48.78%.** Ours matched
to two decimal places.

The **counts** do not match exactly — ours are about 150 ballots higher out of
322,000, or 0.05%. That gap is not a mistake to fix; it is the lab's real
lesson. The state's tabulator applies rules ours does not, published on the
front page of its own report: *exclude unresolved write-ins*, *exhausted on two
or more ranks skipped*, and a specific treatment of overvotes.

**A different, defensible rule about skipped rankings changes the count.** It
did not change the winner here. In a closer race it could.

# Part 5 — Where the losers' votes went

```{r transfers}
tr <- read.csv("data/official_transfers.csv", stringsAsFactors = FALSE)
tr
```

```{r transfer-plot}
h <- tr[tr$eliminated == "Howe, John Wayne", ]
barplot(h$ballots, names.arg = h$transferred_to, col = "#C41230",
        ylab = "ballots", main = "Where Howe's 13,871 ballots went")
```

When John Wayne Howe was eliminated, his 13,871 ballots split: **4,817 to
Begich, 2,724 to Peltola, and 6,295 exhausted.**

**Forty-five percent of his voters had ranked nobody else.** Their ballots left
the count entirely. By the final round **7,595 ballots were exhausted** — people
who turned up, voted, and had no say in the decisive round, because the
candidate they backed was gone and they had named no alternative.

That is the trade at the heart of ranked-choice voting. It lets you express more
than one preference. It does not make you.

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Abolish the runoff.** Who wins under plain plurality — just
`rank1`? Compare to the RCV winner and say which result better reflects what
Alaskans wrote on their ballots. Both answers are defensible; the argument is
the work.

**Option B — Exhaustion by candidate.** For each first-choice candidate, what
share of their voters ranked *only* them? (Group `s` by `rank1` and compare
`ranks_used == 1` to the rest.) Whose supporters were most likely to bullet
vote, and what does that suggest about how each campaign told people to vote?

**Option C — A different overvote rule.** The `irv()` function stops a ballot at
an overvote. Change it to *skip* the overvoted rank and continue to the next one
(`if (v == "OVERVOTE") next`). How many ballots does that rescue, and does
anything change? Then say which rule you would write into law.

**Option D — Your own question.** 926 sequences, 340,778 ballots, four
candidates.

# What this lab cannot tell you

- **Who any of these people are.** No names, no addresses. But a cast vote
  record is still individual-level: in a small precinct an unusual ranking can
  narrow down who cast it, which is why the file you have been given is
  collapsed to sequences and counts, with precinct and tabulator identifiers
  removed.
- **The certified result.** Alaska says plainly that the CVR **excludes ballots
  counted by hand** and does not show what write-in voters actually wrote. Our
  340,778 is not the official turnout for this contest.
- **Why anyone ranked as they did.** A ballot records a choice and nothing about
  a reason. Whether 65% bullet voting means satisfaction, confusion, distrust of
  the system, or a campaign telling people to "rank the red" is not in here —
  Session 8's survey data is where that question goes.
- **Whether RCV is good.** One race, one state, one year. What this shows is
  what the mechanism *does* to a set of ballots, which is a narrower and more
  useful thing than an opinion about the system.

# On the discussion board

By **Mon, Nov 30, 11:59 p.m.** — Thursday is Thanksgiving, so this week's post
runs long. A sentence or two is enough.

Three that would make good posts:

- Given four rankings, two-thirds of Alaskans used one. What do you make of
  that?
- 6,295 of Howe's voters had their ballots exhausted before the decisive round.
  Were they disenfranchised, or did they choose that?
- Our count matched the official percentages exactly and the raw counts by 150
  ballots, because of a rule about skipped rankings. Who should decide that
  rule?

***

### Where the data came from

State of Alaska, 2024 General Election, **cast vote record**, published by the
Division of Elections as a Dominion JSON export — 2,043 files, 1.58 GB,
one record per scanned ballot.
<https://www.elections.alaska.gov/results/24GENR/CVR_Export_20241130154411.zip>

Official round-by-round results and transfer detail are transcribed from the
state's own **RCV Detailed Report**,
<https://www.elections.alaska.gov/results/24GENR/RCV-USRep.pdf>

**The committed file contains no individual ballots.** 340,778 records were
collapsed to 926 distinct ranking sequences with counts, and every precinct and
tabulator identifier was dropped, before anything was saved. The build is in
`data/build-data.py` — Python rather than R, because base R has no JSON parser
and the source is 1.58 GB of JSON.
