---
title: "Data Session 13 Key: The Ballots Themselves"
subtitle: "84-355 Democracy's Data | Session 13 — Cast vote records | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
s    <- read.csv("data/sequences.csv", stringsAsFactors = FALSE)
cols <- grep("^rank", names(s), value = TRUE)
ru   <- rowSums(s[, cols] != "")
tr   <- read.csv("data/official_transfers.csv", stringsAsFactors = FALSE)
```

All code is given to students. Grade the interpretation.

**Where this sits.** Session 13, and it is the only session in the course where
students see **individual votes**. Everything else has been totals, samples, or
inferences about people from aggregates. Say that out loud at the start: today
the unit of observation is one ballot, and that is extraordinary.

**Logistics.** This is the **remote session** and the week before Thanksgiving.
Board posts are due **Mon, Nov 30**, not Friday. There is no draw this week.
The lab is well suited to Zoom — it is one small file and a lot of reading.

Every number below was run against the data file.

***

# The session in one sentence

Alaska publishes the ballots, so we recount a congressional election ourselves,
match the official percentages exactly, and discover that two-thirds of voters
declined the one thing ranked-choice voting exists to offer.

***

# The findings

```{r headline}
c(ballots        = sum(s$ballots),
  sequences      = nrow(s),
  bullet_votes   = sum(s$ballots[ru == 1]),
  pct_bullet     = round(100 * sum(s$ballots[ru == 1]) /
                         (sum(s$ballots) - sum(s$ballots[ru == 0])), 1),
  howe_exhausted = tr$ballots[tr$eliminated == "Howe, John Wayne" &
                              tr$transferred_to == "Exhausted"])
```

- **340,778 ballots** collapse to **926 distinct ranking sequences.**
- **65% of voters who voted in this race ranked exactly one candidate.**
- Our recount gives **Begich 51.22%, Peltola 48.78%** — the official figures to
  two decimals.
- **45% of Howe's 13,871 ballots exhausted** when he was eliminated.
- **3,790 ballots name the same candidate twice.**

***

# The three things to get said

**1. This file exists because Alaska decided it should.** No federal law
requires publishing a cast vote record, and most states do not. Ask what would
change if every state did — and then ask what the risk is, which is the privacy
point below. **The availability of data is a political choice**, which is the
argument the whole course has been circling.

**2. Sixty-five percent bullet voting is the finding, and it is not a failure.**
Students will want it to mean voters were confused. Push back: ranking one
candidate is a coherent, sincere ballot. It does mean the *mechanism* did less
work than its advocates suggest, and it means a voter's influence ends when
their candidate does. **Both readings are defensible; an essay that only holds
one is a 2.**

**3. Matching to two decimals and missing by 150 ballots is the methodological
lesson.** We reproduced the percentages exactly. We did not reproduce the counts,
because the official tabulator applies documented rules ours does not —
*exclude unresolved write-ins*, *exhausted on two or more ranks skipped*, and
its own overvote handling, all printed on page 1 of the state's report. **The
rules are not in the data. They are a separate document, and they change the
answer.** In this race by 0.05%; in a closer one, by the outcome.

**The line for the board:** the ballots do not count themselves.

***

# The privacy point — make it explicitly

A CVR carries **no name** and is a public record. It is also
**individual-level**, and a distinctive ranking in a small precinct can narrow
down who cast it. That is why the committed file is collapsed to sequences with
counts and why precinct and tabulator identifiers were dropped before anything
was saved.

Worth five minutes: **anonymous is not the same as unidentifiable.** This is the
same lesson as Session 7's BISG, arriving from the opposite direction — there,
we inferred an identity nobody recorded; here, an identity could be inferred
from a record with no name on it.

***

# Option A — Abolish the runoff

## Verified output

```{r a}
p <- tapply(s$ballots, s$rank1, sum)
p <- sort(p[names(p) != ""], decreasing = TRUE)
round(100 * p / sum(p), 2)
c(plurality_winner = names(p)[1],
  margin = unname(p[1] - p[2]))
```

**Begich wins under plurality too**, by 6,816 first-preference votes; the
instant-runoff margin is 7,876. **RCV did not change the outcome of this race.**

**3/4:** Reports that the winner is the same either way.

**4/4:** Notices that "the same winner" is not "no difference." Under plurality
Begich wins with **48.3%** — a plurality, not a majority. Under RCV he finishes
above 50% of continuing ballots. The systems produce the same *person* and a
different *claim to legitimacy*, and the second is what the argument is usually
about.

**4/4, best:** Notices that 7,595 ballots were exhausted by the final round, so
Begich's 51.22% is a majority **of the ballots still in the count**, not of the
people who voted. Of all 340,778 ballots, his final-round total is about 48%.
**A student who works out that the majority is manufactured by shrinking the
denominator has found the sharpest criticism of IRV there is** — and should also
be asked what the alternative denominator would even mean.

***

# Option B — Exhaustion by candidate

## Verified output

```{r b}
for (c in c("Begich, Nick", "Peltola, Mary S.", "Howe, John Wayne", "Hafner, Eric")) {
  k <- s$rank1 == c
  cat(sprintf("  %-18s %8s voters   %5.1f%% ranked only that candidate\n",
              c, format(sum(s$ballots[k]), big.mark = ","),
              100 * sum(s$ballots[k & ru == 1]) / sum(s$ballots[k])))
}
```

**This is the best option and the one to steer strong students toward.**

**Begich voters bullet-voted at 76.1%. Peltola voters at 56.4%.** A twenty-point
asymmetry between the two major candidates' supporters.

**3/4:** Reports the gap and notes it is large.

**4/4:** Offers a mechanism. The most likely is **campaign instruction** —
"rank the red" style messaging telling supporters that ranking anyone else helps
the other side. A student who says the ballots are showing us the *campaigns'*
strategies, not just the voters' preferences, has understood what individual
data reveals.

**4/4, best:** Notices the pattern runs down the ballot — Howe 43.4%, Hafner
35.4% — so **the smaller your candidate, the likelier you are to rank a second.**
That is exactly what a rational voter would do, and it means bullet voting is
concentrated among front-runners' supporters, who lose nothing by it. The
behaviour is strategic, not confused.

***

# Option C — A different overvote rule

## Verified output

```{r c}
has_ov <- apply(s[, cols], 1, function(r) any(r == "OVERVOTE"))
c(sequences_with_overvote = sum(has_ov),
  ballots = sum(s$ballots[has_ov]),
  overvote_at_rank1 = sum(s$ballots[s$rank1 == "OVERVOTE"]))
```

**1,472 ballots contain an overvote**, 610 of them at the first rank. Changing
the rule from *stop* to *skip* rescues some of these into the count.

**3/4:** Makes the change, reports how many ballots move, notes the winner does
not change.

**4/4:** Observes that 1,472 is small **here** and would not be in a closer
race — and that whether an overvoted ballot is discarded or partially honoured
is a policy question with no technically correct answer. Alaska's own report
names its choice; the choice is not in the data.

**4/4, best:** Connects it to Part 4. We already know our count differs from the
official one by ~150 ballots because of rules like this. **Option C is a
demonstration that the count is a function of the rulebook**, and a student who
says they would want the rulebook published alongside every result has arrived
at the professional standard.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating bullet voting as behaviour to be
  explained rather than as a mistake to be deplored.
- **A 2** is "RCV is confusing, look how many people got it wrong." Nothing here
  shows anyone got anything wrong.
- **Reward anyone who notices the denominator** in the final round.

## Anticipated questions

- *"Is this legal to publish?"* — Yes, and Alaska does it deliberately, for
  transparency. Note that most states do not, and that reasonable people
  disagree about whether they should.
- *"Could I find my own ballot?"* — In principle a sufficiently distinctive
  ranking in a small precinct narrows things down, which is why our file has no
  precinct. Do not brush this off; it is the real objection to publishing CVRs.
- *"Why doesn't our total match the certified result?"* — Two reasons, both
  from Alaska: hand-counted ballots are excluded, and our tabulation rules are
  simpler. Say both.
- *"Did RCV cost Peltola the seat?"* — No. She lost under both systems on this
  data. Do not let the room turn this into a referendum on RCV using one race.
- *"Why can Alaska rank the presidential race?"* — It does; contest 4 allows
  eight ranks. Worth mentioning as an aside about how far the system extends.

## Connections

- **Session 1 (returns)** — totals, which is all any other state publishes.
- **Session 7 (BISG)** — anonymity and identifiability, opposite directions.
- **Session 10 (live returns)** — election night reports the first-round
  numbers; this is what was underneath them.
- **The wrap (Dec 3)** — the four-standards debate explicitly asks students to
  score electoral systems. **They will have handled RCV's ballots by then**, and
  should be told to use what they saw.
