---
title: "Fifty-One Publishers, One Election"
type: brief
subtitle: "What it costs to assemble the file no agency publishes, and the half of the country where the answer changes"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("county-returns-code.R")
```

```{r setup, include=FALSE}
```

The [census chapters](../../01-census-bureau/census-access/census-access-brief.html) had a problem of choosing between doors. This one has the opposite problem. For county-level presidential returns there is no door, because **no federal agency counts votes.** The Federal Election Commission regulates money. The Election Assistance Commission surveys administration. Neither is given the returns, and neither publishes them.

What exists instead is fifty-one separate publications by fifty-one separate officers, each certifying their own state, in whatever form that office happens to use. The [returns-source chapter](../returns-source/returns-source-brief.html) makes that argument. This chapter is what happens when you take it literally and go and get all fifty-one.

## What fifty-one publishers actually costs

**`r n(FV("prov_rows"))` documents, `r FV("prov_states")` jurisdictions, two elections, `r FV("prov_formats")` distinct file formats.**

```{r formats}
```

`r n(FV("prov_pdf"))` of the `r n(FV("prov_rows"))` are PDFs, and one of those has **no text layer at all** — it is a photograph of a document, and the only way through it is optical character recognition. Three of the `r n(FV("prov_rows"))` sit on pages that do not say the returns are certified, so they are recorded as `unknown` rather than promoted.

The result covers **`r n(FV("off_rows"))` reporting units and `r n(FV("off_total"))` votes for president**. Every row carries the address it came from, the date it was fetched and the format it arrived in. With fifty-one sources, a citation at the level of the file would be a claim about nothing.

## Fifty of those rows are not counties

The target was "county-level returns," and fifty rows refuse the category.

```{r notcty}
```

Rhode Island publishes by **town**; its five counties have no government and the state does not report to them. The District of Columbia publishes by **ward**. Kansas City reports separately from Jackson County, which contains most of it. A Maine unit is not a county either.

None of these is a defect in the source. They are what those places publish, and the assembly keeps them as they are — **with no county code**, because inventing one would be the error rather than the fix.

> **Before you read on, commit to two numbers.** You now have the certified
> returns from every state, and the same GitHub compilation several other
> chapters of this book read too, among them
> [`mapping`](../mapping/mapping-brief.html) and
> [`wind-map`](../wind-map/wind-map-brief.html). Both describe the 2024
> presidential election.
> **In how many of the roughly 3,100 counties do you expect them to disagree
> about the Democratic vote? And about the total number of votes cast?**
> Write both down.

## They agree about who won

```{r agree}
```

On the major-party vote the two files are in near-perfect agreement: **`r FV("diff_dem")` counties** of `r n(FV("matched"))` differ on the Democratic vote and `r FV("diff_gop")` on the Republican. Anybody using either file to say who carried a county will say the same thing almost everywhere.

## The turn: they do not agree about how many people voted

**`r n(FV("diff_total"))` counties report a different number of votes cast.** That is `r pc(FV("diff_total_pct"), 1)`% of them, half the country.

The two files agree about the numerator and disagree about the denominator — the top number and the bottom number of every rate built from them. That is the more dangerous kind of disagreement, because turnout, margin, share and every rate in this book are computed against it, and nothing in either file announces the difference.

```{r magfig, fig.cap="How large the disagreement is, where there is one. Most are a handful of votes; a few are enormous."}
```

The median disagreement is **`r FV("dt_median")` votes** — small enough to look like rounding, and it is not rounding. It exceeds a hundred votes in **`r n(FV("dt_over_100"))` counties** and a thousand in **`r n(FV("dt_over_1000"))`**. Nationally the two totals are **`r n(FV("nat_gap"))` votes** apart.

## Three named reasons, and they do not add up to all of it

**Dropped third parties.** In `r FV("third_zero")` counties the compilation's total is exactly the Democratic plus the Republican vote, meaning no other candidate is in the denominator. In **`r FV("third_zero_real")`** of those the certified return does record third-party votes. They are not scattered:

```{r tpz}
```

New York and Alaska between them are nearly all of it — the same two states, and the same defect, that the `returns-source` chapter finds from the other direction. This explains only **`r pc(FV("third_explains"), 1)`%** of the disagreeing counties.

**A city that is not in its county.** The single largest disagreement is `r FV("kc_county")`:

```{r kc}
```

`r n(FV("kc_dem_gap"))` Democratic votes move depending on which file you opened. Kansas City certifies separately, so the certified assembly keeps it separate and the compilation folds it in. Neither is wrong. Only one of them matches a census population for the same code, and that is the one you need if you are computing turnout.

**Write-ins, over-votes and under-votes.** What remains is the long tail of small differences, and it is a question of definition rather than of data quality. Is a spoiled ballot, an over-vote or a write-in for a non-candidate a vote cast for president? States answer it differently and a compiler must answer it once.

## The unit itself is not stable

Worse than a differing number is a differing geography, and Connecticut has one:

```{r shapes}
```

**Connecticut appears as `r FV("ct_certified")` counties in one file and `r FV("ct_compilation")` planning regions in the other.** The state replaced county-equivalents with councils of government for statistical purposes. These are not two codings of one geography. They are different shapes on the ground. A 2020-to-2024 comparison in Connecticut is comparing different objects unless somebody decided otherwise on purpose, and neither file records such a decision.

Rhode Island shows the same seam more quietly: **`r FV("ri_certified")` towns in the certified file, `r FV("ri_compilation")` counties in the compilation.** The compilation is not wrong to aggregate. It is only silent about having done it.

```{r gaps}
```

## What this is for

The compilation is careful work by somebody under no obligation to anyone: no correction path, no custodian, no undertaking that it will exist next year. Three chapters of this book read it. That is a reasonable thing to do and this chapter is not an argument against it.

It is an argument for **saying which file you opened**. "The 2024 vote in Jackson County" is not a fact. "The certified Missouri return for Jackson County, excluding Kansas City" is one, and it differs from the other true statement by `r n(FN("kc_com") - FN("kc_off"))` votes.

## What arrives

Every document is kept as it came: `r n(FV("prov_rows"))` files across `r FV("prov_formats")` formats, including the PDFs and the one scan. A record file carries a row per jurisdiction-year with the address, the retrieval date, the format, whether the page claimed certification, and a note on every judgement made while parsing it. That file is the real output of this chapter. The returns themselves are downstream of it.

## What this chapter cannot tell you

**It does not say which file is right**, and mostly the question is malformed. Where the two disagree about Kansas City they are answering different questions; where they disagree about a write-in they are applying different definitions. Only the third-party omissions are a defect rather than a choice, and those are `r pc(FV("third_explains"), 1)`% of the disagreement.

**It covers 2024, and 2020 alongside it, and no further back.** How much further back this could go is a separate question with an expensive answer. The good long compilations are licensed rather than published. The public-domain route is the one nobody has assembled, because assembling it is the expensive part. That is the subject the redesign notes set aside, and it is not settled here.

**It does not audit the states against themselves.** A certified return is the state's own claim, and this chapter takes it at its word. What a canvass can and cannot establish, and the fact that the compilations disagree with the certified national total in both directions — belongs to [returns-source](../returns-source/returns-source-brief.html).

**And the parsing is ours.** `r n(FV("prov_pdf"))` PDFs were read by script; a figure recovered from a PDF column is a claim about layout as much as about votes. The state-level crosscheck against an independent wire-service tally is in `derived/` for exactly that reason, and it is the check that would catch a parser that quietly went wrong.

## Sources

- The certified returns of all `r FV("prov_states")` jurisdictions for the 2020
  and 2024 presidential elections, published by each state's chief election
  officer. Every address, retrieval date and format is recorded per
  jurisdiction-year in this chapter's `provenance.csv`; they are too many to
  list here and that is the chapter's point.
- McGovern, Tony. *US County Level Presidential Results 2008–2024.*
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>
- U.S. Census Bureau. *2024 Gazetteer Files*, counties, for the FIPS
  resolution and county-name normalisation used when matching state
  publications to codes.
- Connecticut Office of Policy and Management, on the replacement of counties
  by councils of government as county-equivalent geography.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`formats.csv`](data/derived/formats.csv), [`largest_gaps.csv`](data/derived/largest_gaps.csv), [`not_counties.csv`](data/derived/not_counties.csv), [`third_party_zero.csv`](data/derived/third_party_zero.csv), [`unit_shapes.csv`](data/derived/unit_shapes.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `largest_gaps.csv` holds the counties where the state's certified number and
  the national compilation disagree. Take the biggest gap and work out which
  of the two sources is wrong.
- `third_party_zero.csv` counts counties reporting no third-party votes at
  all. Some states let one candidate appear on several party lines. Is that
  zero a fact about voters, or a choice about formatting?
- `not_counties.csv` lists the jurisdictions that do not report by county at
  all. What happens to a national county map if you include them, and what
  happens if you drop them?
- `formats.csv` counts how many of these publications arrive as PDFs. Estimate
  how many hours it would take you to collect all fifty states by hand.

```{r ai-prompt, echo=FALSE, results="asis"}
```
