---
title: "Reading a Lobbying Disclosure Filing"
type: brief
subtitle: "A quarterly lobbying disclosure, and the two things the form was built not to record"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("lobbying-code.R")
```

```{r setup, include=FALSE}
```

A law firm filed a form with the Secretary of the Senate on behalf of a small city in Utah, covering the second quarter of 2024. On line 12 of that form, in the box headed *income relating to lobbying activities*, somebody wrote `r d(LD2$amount)`.

That is the whole of what the federal government knows about how much Millcreek, Utah, spent influencing Congress in those three months. Not a range, not an invoice, not a rate: one figure, and the form's own instruction is to round it to the nearest ten thousand dollars.

Two pages later the same form names the subjects the money was spent on. There are two of them, on separate sheets, and the form asks for no division of the money between them.

Hold those two facts together, because everything else in this chapter follows from them. **The federal record of lobbying reports what was paid and what it was about, and it never joins the two.**

> **A warning about how this chapter is built.** Two of the sections ahead are
> about things this data cannot do: a question it cannot answer, and a field
> that cannot be counted. **Those are not digressions.** They are not
> throat-clearing to be got through before the real findings arrive — they are
> the findings. A disclosure regime is worth what its records can be made to
> say, and the most important thing these records say is what they decline to.

## A law requiring lobbyists to name the members they used to work for

Campaign finance is money spent deciding who holds office. **Lobbying is money spent on what they do afterwards**, and it is the larger of the two.

Since the **Lobbying Disclosure Act of 1995**, anyone paid to lobby the federal government files quarterly: who is paying, roughly how much, what issues, and which parts of government were contacted. The Act also requires something more unusual.

> **A lobbyist must disclose whether they previously held a government
> position, and name it.**

Congress wrote a law requiring lobbyists to declare that they used to work for Congress. The rest of this chapter reads what they filed.

It matters because **access is a scarce good and it is distributed unequally.** Legislators have finite attention; whoever gets it shapes what is considered. That is not corruption in any legal sense — it is the ordinary mechanism by which organized interests convert money into influence.

The revolving door concentrates the problem. If what a client buys is a personal relationship with a specific office, then **the value of public service becomes partly a private asset**, realizable on exit. That changes the incentives of people currently in government, in ways no disclosure requirement addresses.

The strongest evidence that this is what is being sold comes from **Blanes i Vidal, Draca and Fons-Rosen (2012)**, who tracked lobbyists whose former employer left the Senate. Revenue generated by those lobbyists **fell by about 24% when their senator departed** — immediately, discontinuously, and durably. The asset was the connection, not the expertise.

## What the statute counts as lobbying

Every number in this chapter comes from the Lobbying Disclosure Act database, which the Secretary of the Senate and the Clerk of the House maintain jointly and publish in full. That is `r n(nrow(fl))` filings from the second quarter of 2024, and the `r n(nrow(lo))` individual lobbyists named on them. Nobody compiled it. Registrants typed it, because since 1995 the law has obliged them to. Since the Honest Leadership and Open Government Act of 2007 they have had to do it four times a year rather than twice. Enforcement runs by referral to the United States Attorney for the District of Columbia, and referral is rare — which is part of why the forms look the way they do.

The reason to use this file rather than one of the commercial services that resell it is narrow and decisive. This is the only record in which the *person* doing the lobbying is named, and named beside the government job he used to hold. No vendor creates that field; it exists because a statute made a lobbyist write it down about himself. A chapter asking whether public service becomes a private asset needs a record in which the seller identifies the office he used to work in, and there is exactly one.

The harder trouble sits upstream of the file. "Lobbying" here is a legal category, not a description of behavior. The Act reaches a person who makes more than one lobbying contact with covered officials and spends at least a fifth of their time for that client doing it. Registration is self-triggered: the filer decides whether they crossed the line.

Campaigns aimed at persuading constituents to call their representative fall outside. So does strategic advice that never touches a covered official, and so does the former member who talks to their old colleagues and calls it something other than lobbying. A record can be complete inside its boundary and silent about the most consequential influence, and nothing in the rows will tell a reader which case is in front of them.

Cleaning began before the first table, because what the database returns is not a table. Each filing arrives as a nested record carrying a list of named lobbyists and a list of issue-area codes. So the `n_issues` and `n_lobbyists` columns below are lengths of those lists, rather than anything a filer wrote. The build flattens all of it into four files. Two of them, the issue and entity tables, arrive already counted up to mentions, so this chapter cannot go back beneath them. One flattening decision costs more than the others.

`lobbyists.csv` carries a name and a former position, and no filing key, so a lobbyist cannot be joined to the client paying them. Whether former staffers command higher fees is therefore a question this extract cannot be asked, and that one is not the form's fault.

One object in this chapter did not come through that database at all, and how it was got is worth a sentence. The API returns filings; it does not return the *form*, and several of the observations below are properties of the printed page rather than of any field in a row. So the public print view of a single filing is requested directly. The build checks that three particular strings are still on it; if they are not, the layout has changed and the crop is wrong. Then it drives a browser to render the page, replaces the filer's telephone number and e-mail with `[redacted]`, and photographs the result.

That filing is also one of the rows in `filings.csv`. So Figure 2 and the tables are the same object seen twice: once as the database hands it over, and once as the person filling it in saw it. It is the only thing in this book obtained by taking a picture of a web page, and a screenshot is exactly as fragile as any other scrape. The next redesign of `lda.gov` breaks it, and nothing will announce that it did.

## A filing is a tree; the table is a shadow of it

One filing, three ways. Here it is as the Senate's API returns it. This is the same Millcreek filing the form in Figure 2 reproduces, with the address blocks and the filer's telephone number left out and the descriptions cut short:

| Field | Value |
|:---|:---|
| filing_uuid | 302fd116-100b-475c-9d3b-37c3c78dcdc5 |
| filing_type | Q2 |
| filing_type_display | 2nd Quarter - Report |
| income | 20000.00 |
| expenses | null |
| client.name | MILLCREEK, UTAH |
| client.state | UT |
| client.general_description | Municipal Government |
| registrant.name | BARKER LEAVITT, PLLC (SKA MR. JAMES C. BARKER) |
| registrant.state | DC |
| lobbying_activities[1].general_issue_code | BUD |
| lobbying_activities[1].general_issue_code_display | Budget/Appropriations |
| lobbying_activities[1].description | FY24 Appropriations Bill DHS/FEMA. Community Project Funding Request for Neff's … |
| lobbying_activities[1].lobbyists[1].lobbyist.first_name | RYAN |
| lobbying_activities[1].lobbyists[1].lobbyist.last_name | LEAVITT |
| lobbying_activities[1].lobbyists[1].covered_position | Legislative Director- Sen. Josh Hawley |
| lobbying_activities[1].lobbyists[2].lobbyist.first_name | JAMES |
| lobbying_activities[1].lobbyists[2].lobbyist.last_name | BARKER |
| lobbying_activities[1].lobbyists[2].covered_position | null |
| lobbying_activities[2].general_issue_code | DIS |
| lobbying_activities[2].general_issue_code_display | Disaster Planning/Emergencies |
| lobbying_activities[2].description | Advocacy for a FEMA BRIC Grant for Millcreek, UT; Debris Retention Basin Project … |
| lobbying_activities[2].lobbyists | *…the same two people again…* |

And here is that filing as a row:

```{r clean-lobby}
```

**A tree became six numbers and strings, and the tree is where the content was.** The table has no room for anything a person would actually want to know: that this was about a debris retention basin, or that one of the two lobbyists used to run a senator's office. That is not a failure of the extract. It is what a table *is*, and it is why this chapter has to keep going back to the form.

### Four decisions live inside those six columns

**`n_lobbyists` is 2, and the obvious arithmetic gives 4.** The same two people are listed under both issue areas, because the form asks who lobbied on each issue and the honest answer repeated. Counting the entries would double them. Somebody had to decide that a person named twice is one lobbyist — which is right, and which also means this column cannot tell you how much work was done.

**`n_issues` is a length, not a subject.** It records that there were two activity blocks. The issue *codes* went to `issues.csv` and the descriptions went nowhere, so a filing about appropriations and a filing about disaster planning are indistinguishable in this table beyond the count.

**One of two `state` fields was kept.** The client sits in Utah and the registrant in Washington, D.C., and the column called `state` is the client's. That is the right choice for asking who buys influence and the wrong one for asking where the industry is, and nothing in the column name says which question it was built for.

**`income` arrives as the text `"20000.00"`, not as a number.** It has to be converted before it can be added, and the conversion is where a filing reporting nothing and a filing with no income box at all can quietly become the same thing. Registrations carry `null` here rather than a figure, and they are a different `filing_type` from the quarterly reports. That is why the type has to be filtered before any dollar sign is believed.

There is one more, and it is the sort that never shows up in a row count. The API gives Ryan Leavitt's former job as `Legislative Director- Sen. Josh Hawley`; `lobbyists.csv` carries it as `r lo$former_government_position[toupper(lo$name) == "RYAN LEAVITT"][1]`. Same fact, reordered into a house style so that the office and the officeholder can be told apart. **Somebody made that consistent by hand-writing a rule about somebody else's free text**, and every reordering of that kind is a small opportunity to be wrong about what the filer meant.

## Two parties, one dollar figure

```{r one-filing}
```

A filing has two parties. The **client** is the organization buying influence; the **registrant** is the firm selling it. When the two names are the same, an organization lobbies for itself.

Then one dollar figure, a count of issue areas, and a count of named lobbyists. Notice that the dollar figure does not vary with the second column: a filing covering five issue areas reports the same single number as one covering one.

## The data is public and the pipe is narrow

```{r sample}
```

**This is `r pc(100*nrow(fl)/QTOTAL)`% of the quarter, and that is a finding rather than a design choice.** The Senate's database is entirely public: no key, no registration, no fee. It hands out records twenty-five at a time and refuses to go faster.

The data is public; the pipe is narrow. That distinction has a practical consequence which runs through every number below: **rates and rankings are informative here, totals are not.** No total dollar figure appears anywhere in this chapter. The denominator, the bottom number every rate is divided by, would be an accident of how many requests somebody was willing to make.

## Estimates dressed as accounts

```{r money}
```

**`r n(sum(fl$amount == 0))` of `r n(nrow(fl))` filings report zero.** Not missing — zero.

And look at the smallest positive figure: **exactly `r d(min(paid$amount))`.** That is not a coincidence. It is the statutory reporting threshold, and filers below it report nothing rather than a small number.

> **Before you read the next table, write a percentage down.** You are looking
> at `r n(nrow(paid))` filings reporting a positive dollar figure, running from
> `r d(min(paid$amount))` to `r d(max(paid$amount))`. If these were accounts —
> salaries, hours, expenses, added up — almost none of them would land on a
> round number. **What share of them are exact multiples of $10,000?**
>
> Commit to a figure before you look. Whatever you write, the gap between it and
> the answer is the distance between an account and an estimate, and that
> distance is the subject of this chapter.

```{r rounding}
```

**`r pc(100*mean(paid$amount %% 10000 == 0))`% of amounts are round multiples of ten thousand dollars.** These are not invoices. They are estimates, reported to the nearest convenient figure. The median filing, `r d(median(paid$amount))`, should be read as approximately that rather than exactly that.

Plot every reported amount and the point stops being a statistic.

```{r heap-prep, include=FALSE}
```

```{r heap-d3, results='asis', eval=knitr::is_html_output()}
```

```{r heap-static, eval=!knitr::is_html_output(), fig.height=3.6}
```

**Figure 1. Every dollar figure reported in the quarter, and how many filings carry it.** Green stems are multiples of ten thousand dollars, gray are everything else; the dashed line is `r d(THRESH)`, the statutory reporting threshold, below which filers report nothing rather than a small number. `r NDIST` distinct figures cover all `r n(nrow(paid))` paid filings, and the three tallest spikes alone account for `r pc(100 * sum(TOP3$n) / nrow(paid))`% of them: `r d(TOP3$amt[1])`, `r d(TOP3$amt[2])` and `r d(TOP3$amt[3])`. `r NABOVE` filings above `r XMAXLAB` are off the right of the axis.

**Nothing that anybody actually invoiced would look like this.** A real price distribution is smooth, because real work is billed in hours and real hours do not come in units of five thousand dollars. This one lands on the numbers a person reaches for when asked to estimate. Every figure in this chapter that carries a dollar sign is that kind of number.

## The question that cannot be asked

So: *how much was spent lobbying on health care?*

It is the first question anyone puts to this data, and it cannot be answered. The reason is the one the opening paragraph set out. **A filing reports one dollar figure and a list of issues, with no division between them.**

Every row in the tables above was typed into a form like this one. It is one of the `r n(nrow(fl))` filings retrieved here: row `r LD2ROW`, Barker Leavitt, PLLC, reporting for the city of Millcreek, Utah. It was fetched from the public database and reproduced exactly as filed.

```{r ld2, out.width="80%", fig.align="center"}
```

**Figure 2. One filed LD-2, the object every number in this chapter is made of.** **1** is line 12, the income box: a single figure, `r d(LD2$amount)`, for the whole quarter's work, rounded to the nearest ten thousand because the form says to. **2** is line 15, the general issue area code, filled in twice on separate pages, `BUD` and `DIS`. The form asks for a page per code and no split of the money between them, so the box marked 1 belongs to both boxes marked 2 and to neither.

**3** is line 17, who was contacted: two institutions, no person and no date. **4** is line 18, the covered-position field, where a lobbyist declares a previous government job. The filer's telephone number and e-mail have been blacked out. Nothing else has been altered.

Look at box 4 for a moment; the chapter returns to it. The same lobbyist is entered twice, under two different descriptions of his own career, and on the second page of the filing his surname is misspelt. **Nothing on this form is validated by anyone.**

```{r multi}
```

**`r pc(100*mean(paid$n_issues > 1))`% of paid filings cover more than one issue.** One of them reports `r d(mx$amount)` spread across `r max(paid$n_issues)` different issue areas, and the form does not ask how it divided.

So any published figure for "lobbying spending on issue X" was produced by someone making an assumption — usually splitting evenly, sometimes assigning the whole amount to every issue it touches. **Neither is in the data.** The next time a dollar total appears attached to a policy area, this is what stands behind it.

## What is lobbied on, and who is lobbied

Since money cannot be attributed, count mentions instead.

```{r issues}
```

**`r is$issue[1]` leads**, and it is not close. This is not about ideology; it is about where money is decided. The appropriations process moves more dollars than anything else Congress does, so that is where lobbying concentrates.

The other question the form asks is which parts of government were contacted.

```{r entities}
```

**The House and Senate dwarf everything**, at `r n(en$mentions[1])` and `r n(en$mentions[2])` mentions against `r n(en$mentions[3])` for the third-place agency.

But read what this field actually is. **It names an institution, not a person and not a date.** A filing saying "SENATE" has told you that somebody in a hundred-member body was contacted at some point in three months. **There is no meeting, no office, no calendar.** The most important fact about a lobbying contact, who and when, is not collected.

```{r issues-static, eval=!knitr::is_html_output(), fig.height=5}
```

```{r entities-static, eval=!knitr::is_html_output(), fig.height=5}
```

```{r d3-toggle, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. What lobbyists said they worked on, and which parts of government they said they contacted.** The `r nrow(head(is, 14))` most-mentioned issue areas and the `r nrow(head(en, 14))` most-mentioned institutions, in the `r n(nrow(fl))` filings retrieved. `r is$issue[1]` leads the first list with `r n(is$mentions[1])` mentions; the House and Senate lead the second with `r n(en$mentions[1])` and `r n(en$mentions[2])`, against `r n(en$mentions[3])` for the third-placed agency. Both are counts of mentions, not of dollars, and neither list carries a date or a name.

## A field precise enough to name a senator

Everything so far is what anyone would expect a disclosure regime to collect. The next field is not.

```{r revolving}
```

**`r n(sum(lo$has))` of `r n(nrow(lo))` lobbyists disclose a previous government position, or `r pc(100*mean(lo$has))`%.** More than one in four.

Before reading the entries, it is worth guessing how specific they will be. A requirement like this could be satisfied with "federal service."

```{r positions}
```

*(Selected from the `r n(sum(lo$has))` entries to show the range, not sampled.)*

They are not vague. **They name the member**, and sometimes four of them in sequence with dates — the first row above runs through a senator and three representatives. One entry records a Senate Finance Committee detail by year. This is a federal database in which lobbyists write down, unprompted and by statute, the names of the legislators whose offices they used to work in.

And then the field is free text with no controlled vocabulary, and some filers put something else in it entirely.

```{r junk}
```

**None of those is a government position.** They are private-sector job titles, filed into a field asking about public service.

So the `r pc(100*mean(lo$has))`% above is not clean. It is the share who wrote *something* in the box. Counting how many are genuinely former officials would mean reading `r n(sum(lo$has))` strings and making judgment calls one at a time. **There is no code, no flag, and no standard vocabulary to do it with.** A field precise enough to name a senator is also loose enough to be unusable in aggregate.

## The same statute, in two failing directions

Set the money column and the covered-position column side by side.

```{r asym}
```

**The same statute produced one field too specific to aggregate and one too coarse to attribute.** Both were drafted at the same time, by the same Congress, under lobbying from the people who would have to file them.

Neither is an oversight in the ordinary sense. A regime that reported dollars per issue and coded former employers against a controlled list would answer the questions people actually ask. This one does not, and it is worth asking who benefits from that.

## Whether any of it works

Before drawing a conclusion from a disclosure regime, it is worth knowing whether the activity being disclosed does anything, and that is genuinely disputed. The largest systematic study is Baumgartner and colleagues (2009), which followed 98 randomly selected policy issues. It found that resources predicted outcomes far more weakly than expected, and that the strongest predictor of success was **defending the status quo**. On that account lobbying mostly buys maintenance rather than change, and the sums involved purchase much less than they appear to.

The counterargument is that studying visible legislative fights misses where influence operates: in agency rulemaking, in the drafting of technical provisions, and in keeping items off agendas entirely. None of those produce the roll-call votes that outcome studies measure. That is not a rebuttal so much as a claim that the measurable part is the unimportant part, which is a hard thing to test and an easy thing to assert.

A second dispute concerns the regime rather than the activity. The Act's definition of "lobbyist" turns on making contacts and on a share of working time, so people who advise on strategy without personally contacting officials file nothing at all. How large that unregistered population is, and whether registration has fallen because of it, is contested — but everyone agrees the regime measures only the part of the activity that agreed to be measured.

## What the filings will bear

The evidence assembled here has one unusual strength and a set of weaknesses that all point the same way, and both are worth stating without hedging.

The strength is that filing is mandatory and enforced by statute, so coverage of *registered* lobbying is close to complete. At its best, the covered-position field is specific enough to name individual members of Congress. Nothing in this chapter is a modeled estimate. It is what people wrote on a form under legal obligation, read back to them.

The weaknesses begin with the sample. The `r pc(100*nrow(fl)/QTOTAL)`% of the quarter retrieved here was not drawn at random. It is whatever the database handed out first, twenty-five records at a time. That is why only rates and rankings appear above, and no total.

Within those records the money column is one estimate per filing, rounded to a convenient figure and never divided across the issues it covers. So no dollar amount in this chapter can be tied to a subject.

The contact field names institutions rather than people or dates. The two facts that would make a lobbying contact meaningful, who and when, are absent by design rather than by accident. The covered-position field, the most remarkable thing in the whole regime, is uncontrolled free text in which private-sector job titles sit beside senators' names, so it cannot be counted without being read. Everything that never registers is invisible by construction. And this is one quarter, which supports no trend of any kind.

What survives all that is real. More than a quarter of registered federal lobbyists claim a prior government position, and many of those claims name a specific member of Congress. That is direct, self-reported evidence of a revolving door, sitting in a federal database. Lobbying concentrates on appropriations and on Congress itself, in that order. Neither statement needs a dollar figure, which is fortunate, because the dollar figures will not support one.

Three questions are left genuinely open, and more filings would settle none of them. Whether lobbying changes outcomes at all: Baumgartner and colleagues found weaker effects than anyone expected, and their critics argue the design misses where influence operates. How much activity goes unregistered, which by definition cannot be counted from the register. And whether disclosure is a remedy or a substitute for one.

Requiring somebody to announce a conflict is not the same as preventing it. Congress chose the weaker tool, which is also the only reason any of this data exists.

The transferable finding is about the regime rather than the lobbyists. Disclosure was chosen instead of prohibition. Then the disclosure was drafted so that the two facts people most want from it, how much and on what, cannot be recovered from it. Both fields were written at the same time, by the same Congress, under lobbying from the people who would have to file them. The form reproduced above is what that compromise looks like when somebody fills it in.

## Sources

**Data**

- United States Senate, Lobbying Disclosure Act database, filings API, Q2 2024.
  <https://lda.senate.gov/api/>
- The filed LD-2 in Figure 2 is the public print view of filing
  `302fd116-100b-475c-9d3b-37c3c78dcdc5` (Barker Leavitt, PLLC for Millcreek,
  Utah, Q2 2024), retrieved 10 August 2026 from
  <https://lda.gov/filings/public/filing/302fd116-100b-475c-9d3b-37c3c78dcdc5/print/>.
  The image was fetched from that address; the filer's telephone number and
  e-mail were redacted before it was photographed, and four annotations added.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`entities.csv`](data/derived/entities.csv), [`filings.csv`](data/derived/filings.csv), [`issues.csv`](data/derived/issues.csv), [`lobbyists.csv`](data/derived/lobbyists.csv), [`quarter.csv`](data/derived/quarter.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `lobbyists.csv` has a `former_government_position` column, empty for most
  rows. Count the ones that are filled in. What share of these lobbyists used
  to work in government?
- `filings.csv` reports an `amount` of zero on many rows. Look at those rows.
  Is zero a real number here, or a way of not answering?
- `issues.csv` and `entities.csv` rank what gets lobbied on and who gets
  lobbied. Does that ranking look like what the news says lobbying is about?
- `quarter.csv` records how many filings were taken and how many existed, and
  the ones taken are the first pages the system returned rather than a random
  draw. Which of your answers above would change if you had all of them?

**Academic literature**

- Blanes i Vidal, J., Draca, M. and Fons-Rosen, C. (2012). "Revolving Door
  Lobbyists." *American Economic Review* 102(7): 3731–3748.
- Baumgartner, F. R., Berry, J. M., Hojnacki, M., Kimball, D. C. and Leech, B. L.
  (2009). *Lobbying and Policy Change: Who Wins, Who Loses, and Why.* University
  of Chicago Press.

**How the figures were made**

The LD-2 in Figure 2 was retrieved from the print view above, the filer's telephone number and e-mail redacted, and four annotations drawn on it; nothing else on that image was altered. The tables and the remaining figures are computed from `filings.csv`, `lobbyists.csv`, `issues.csv` and `entities.csv`, all four flattened out of the Senate's API. Every number printed above is computed from those files at the moment this document was rendered.

That includes the one number that cannot be derived from the sample: the `r n(QTOTAL)` filings the Senate received for the quarter. It is the count the database reports for the whole query, carried out of the API response into `quarter.csv` and read from there rather than typed. It supplies the denominator of the `r pc(100*nrow(fl)/QTOTAL)`% figure.

The sixty pages the API returned are kept exactly as they arrived, and the tables are rebuilt from them offline, so the sample is fixed even though the database is not. Asking the API again will move it: the quarter's count is live, and filings are amended for years after the quarter closes.

```{r ai-prompt, echo=FALSE, results="asis"}
```
