---
title: "Reading a Lobbying Disclosure Filing"
type: brief
subtitle: "A quarterly lobbying disclosure, and the two things the form was built not to record"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("lobbying-code.R")
```

```{r setup, include=FALSE}
```

A law firm filed a form with the Secretary of the Senate on behalf of a small city in Utah, covering the second quarter of 2024. On line 12 of that form, in the box headed *income relating to lobbying activities*, somebody wrote `r d(LD2$amount)`. That is the whole of what the federal government knows about how much Millcreek, Utah, spent influencing Congress in those three months. Not a range, not an invoice, not a rate: one figure, rounded to the nearest ten thousand dollars because the form says to.

Two pages later the same form names the subjects the money was spent on. There are two of them, on separate sheets, and the form asks for no division of the money between them.

Now scale that up. The question everybody puts to the lobbying register is **how much was spent lobbying on a given issue** — health care, defence, immigration. It is the right question to ask. Lobbying is money spent on what officeholders do after the election, and it is the larger of the two kinds of political money. Legislators have finite attention, and whoever gets it shapes what gets considered. So: can the register answer it?

## The test

Every number here comes from the Lobbying Disclosure Act database, which the Secretary of the Senate and the Clerk of the House maintain jointly and publish in full. That is `r n(nrow(fl))` filings from the second quarter of 2024, and the `r n(nrow(lo))` individual lobbyists named on them, out of `r n(QTOTAL)` filings the quarter received.

This is the record to ask, because it is the only one where the money and the subjects appear on the same document. Since the **Lobbying Disclosure Act of 1995**, anyone paid to lobby the federal government files quarterly: who is paying, roughly how much, what issues, and which parts of government were contacted. Nobody compiled it. Registrants typed it, because the law obliges them to. It is also self-reported and unchecked, the same arrangement as the campaign-finance filings [this cluster's chapter on disclosure](../finance-source/finance-source-brief.html) describes.

The test is to take one filing, read it as the database hands it over and again as the filer saw it, and try to attach a dollar figure to a subject.

## The data

The `r n(nrow(fl))` filings are the first pages the database returned rather than a random draw:

```{r sample}
```

**That is `r pc(100*nrow(fl)/QTOTAL)`% of the quarter, and it is a fact about the pipe rather than about lobbying.** The database is entirely public: no key, no registration, no fee. It hands out records twenty-five at a time and refuses to go faster. The consequence runs through every number below — **rates and rankings are informative here, totals are not.** No total dollar figure appears anywhere in this brief, because its denominator, the bottom number a rate is divided by, would be an accident of how many requests somebody was willing to make.

What the database returns is not a table. Each filing arrives as a nested record carrying a list of named lobbyists and a list of issue-area codes. Here is the Millcreek filing as the Senate's API returns it, with the address blocks and the filer's telephone number left out and the descriptions cut short:

| Field | Value |
|:---|:---|
| filing_uuid | 302fd116-100b-475c-9d3b-37c3c78dcdc5 |
| filing_type | Q2 |
| income | 20000.00 |
| expenses | null |
| client.name | MILLCREEK, UTAH |
| client.state | UT |
| registrant.name | BARKER LEAVITT, PLLC (SKA MR. JAMES C. BARKER) |
| registrant.state | DC |
| lobbying_activities[1].general_issue_code_display | Budget/Appropriations |
| lobbying_activities[1].description | FY24 Appropriations Bill DHS/FEMA. Community Project Funding Request for Neff's … |
| lobbying_activities[1].lobbyists[1].lobbyist.last_name | LEAVITT |
| lobbying_activities[1].lobbyists[1].covered_position | Legislative Director- Sen. Josh Hawley |
| lobbying_activities[1].lobbyists[2].lobbyist.last_name | BARKER |
| lobbying_activities[1].lobbyists[2].covered_position | null |
| lobbying_activities[2].general_issue_code_display | Disaster Planning/Emergencies |
| lobbying_activities[2].description | Advocacy for a FEMA BRIC Grant for Millcreek, UT … |
| lobbying_activities[2].lobbyists | *…the same two people again…* |

Flattened into `filings.csv`, that whole tree is one row of six columns:

```{r clean-lobby}
```

The `client` is the organization buying influence and the `registrant` is the firm selling it; when the two names match, an organization is lobbying for itself. **A tree became six numbers and strings, and the tree is where the content was.** Four decisions live inside those six columns.

**`n_lobbyists` is 2, and the obvious arithmetic gives 4.** The same two people are listed under both issue areas, because the form asks who lobbied on each issue and the honest answer repeated. Somebody had to decide that a person named twice is one lobbyist, which also means this column cannot tell you how much work was done.

**`n_issues` is a length, not a subject.** The issue *codes* went to `issues.csv` and the descriptions went nowhere, so a filing about appropriations and one about disaster planning are indistinguishable here beyond the count.

**One of two `state` fields was kept.** The client sits in Utah and the registrant in Washington, D.C., and `client_state` is the client's. That is right for asking who buys influence and wrong for asking where the industry is.

**`amount` arrives as the text `"20000.00"`, not as a number.** Registrations carry `null` there rather than a figure, and they are a different `filing_type` from the quarterly reports, so the type has to be filtered before any dollar sign is believed.

One more table matters. `lobbyists.csv` carries a `name` and a `former_government_position` and no filing key, so a lobbyist cannot be tied to the client paying them.

> **Before you read the next table, write a percentage down.** `r n(nrow(paid))`
> of these filings report a positive dollar figure, running from
> `r d(min(paid$amount))` to `r d(max(paid$amount))`. If these were accounts —
> salaries, hours and expenses, added up — almost none would land on a round
> number. **What share of them are exact multiples of $10,000?** Commit to a
> figure before you look.

## What it says about democracy

Start with the money column, because the first thing it says is that it is not a ledger.

```{r money}
```

**`r n(sum(fl$amount == 0))` of `r n(nrow(fl))` filings report zero.** Not missing: zero. And the smallest positive figure is exactly `r d(min(paid$amount))`, which is not a coincidence. It is the statutory reporting threshold, and filers below it report nothing rather than a small number.

```{r rounding}
```

**`r pc(100*mean(paid$amount %% 10000 == 0))`% of amounts are round multiples of ten thousand dollars.** Plot every reported figure and the point stops being a statistic.

```{r heap-prep, include=FALSE}
```

```{r heap-d3, results='asis', eval=knitr::is_html_output()}
```

```{r heap-static, eval=!knitr::is_html_output(), fig.height=3.6}
```

**Figure 1. Every dollar figure reported in the quarter, and how many filings carry it.** One spike per distinct figure is the right form here because the question is about *which* numbers filers reach for, and a histogram would smooth exactly that away by pooling neighbouring values into bins. Green stems are multiples of ten thousand dollars, gray are everything else, and the dashed line is the statutory threshold of `r d(THRESH)`. `r NDIST` distinct figures cover all `r n(nrow(paid))` paid filings, and the three tallest spikes alone carry `r pc(100 * sum(TOP3$n) / nrow(paid))`% of them: `r d(TOP3$amt[1])`, `r d(TOP3$amt[2])` and `r d(TOP3$amt[3])`. `r NABOVE` filings above `r XMAXLAB` are off the right of the axis.

**Nothing anybody actually invoiced would look like this.** A real price distribution is smooth, because real work is billed in hours and real hours do not come in units of five thousand dollars. Every figure in this brief carrying a dollar sign is an estimate of that kind, including the median filing of `r d(median(paid$amount))`.

**Now the question this brief opened with, and it cannot be answered.** A filing reports one dollar figure and a list of issues, with no division between them. That is easier to see on the form than in any table. Here is the Millcreek filing again, this time as the filer saw it.

```{r ld2, out.width="80%", fig.align="center"}
```

**Figure 2. One filed LD-2, the object every number in this brief is made of.** **1** is line 12, the income box: a single figure, `r d(LD2$amount)`, for the whole quarter's work. **2** is line 15, the general issue area code, filled in twice on separate pages, `BUD` and `DIS`. The form asks a page per code and no split of the money, so the box marked 1 belongs to both boxes marked 2 and to neither. **3** is line 17, who was contacted: two institutions, no person and no date; **4** is line 18, where a lobbyist declares a previous government job. The filer's telephone number and e-mail have been blacked out, and nothing else altered.

```{r multi}
```

**`r pc(100*mean(paid$n_issues > 1))`% of paid filings cover more than one issue area.** One of them reports `r d(mx$amount)` spread across `r max(paid$n_issues)` of them, and the form does not ask how it divided. So any published figure for "lobbying spending on issue X" was produced by somebody making an assumption, usually splitting evenly or assigning the whole amount to every issue it touches. **Neither is in the data.** This is not a cleaning problem that a better extract would fix. It is what Congress wrote on the form.

Since the money cannot be attributed, count mentions instead. That much the register does well.

```{r issues}
```

**`r is$issue[1]` leads**, and it is not close. This is not about ideology but about where money is decided: the appropriations process moves more dollars than anything else Congress does. The other question the form asks is which parts of government were contacted.

```{r entities}
```

**The House and Senate dwarf everything**, at `r n(en$mentions[1])` and `r n(en$mentions[2])` mentions against `r n(en$mentions[3])` for the third-place agency. But read what that field is. **It names an institution, not a person and not a date.** A filing saying "SENATE" has told you that somebody in a hundred-member body was contacted at some point in three months.

One field on the form is far more specific than any of that, and it is the most remarkable thing in the regime. The Act requires a lobbyist to disclose whether they previously held a government position, and to name it.

```{r revolving}
```

**`r n(sum(lo$has))` of `r n(nrow(lo))` lobbyists disclose a previous government post, or `r pc(100*mean(lo$has))`%.** More than one in four. A requirement like that could be satisfied with "federal service." It is not.

```{r positions}
```

*(Selected from the `r n(sum(lo$has))` entries to show the range, not sampled.)*

**They name the member**, and sometimes four of them in sequence with dates. This is a federal database in which lobbyists write down, by statute and unprompted, the names of the legislators whose offices they used to work in. What that connection is worth has been measured elsewhere. Blanes i Vidal, Draca and Fons-Rosen (2012) tracked lobbyists whose former employer left the Senate, and found their revenue fell by about 24% when their senator departed.

Then the field turns out to be free text with no controlled vocabulary.

```{r junk}
```

**None of those is a government position.** They are private-sector job titles, filed into a field asking about public service. So the `r pc(100*mean(lo$has))`% above is the share who wrote *something* in the box, and separating the real ones would mean reading `r n(sum(lo$has))` strings and judging them one at a time. **A field precise enough to name a senator is also loose enough to be unusable in aggregate.**

**The same statute produced one field too coarse to attribute and one too specific to count.** Both were drafted at the same time, by the same Congress, under lobbying from the people who would have to file them.

## What this brief cannot tell you

**It cannot give you a total for anything.** The `r pc(100*nrow(fl)/QTOTAL)`% of the quarter here was not drawn at random, so only the rates and rankings are safe to carry away.

**It cannot see anything that never registers.** The Act reaches a person who makes more than one lobbying contact with covered officials and spends at least a fifth of their time on it. Registration is self-triggered, and campaigns urging constituents to call their representative fall outside entirely.

**It cannot tell you whether lobbying works.** Baumgartner and colleagues (2009) followed 98 policy issues and found that resources predicted outcomes far more weakly than expected. Nothing in this register settles that argument.

**And this is one quarter**, which supports no trend.

## What you should have learned

- The federal lobbying register reports what was paid and what it was about, and never joins the two: `r pc(100*mean(paid$n_issues > 1))`% of paid filings cover more than one issue area with one undivided dollar figure.
- Any published figure for spending on a policy area therefore rests on somebody's assumption, not on a column. The largest multi-issue filing here spreads `r d(mx$amount)` across `r max(paid$n_issues)` subjects with no split recorded.
- The dollar figures are estimates rather than accounts. `r pc(100*mean(paid$amount %% 10000 == 0))`% are exact multiples of ten thousand dollars, and the smallest positive figure in the file is the statutory threshold of `r d(min(paid$amount))`.
- The register is excellent at the thing nobody asks it for: `r pc(100*mean(lo$has))`% of the `r n(nrow(lo))` lobbyists named here disclose a former government post, often naming the member of Congress they worked for.
- A field can be too precise to aggregate. That covered-position box is free text with no vocabulary, so private-sector job titles sit in it beside senators' names.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `filings.csv` reports an `amount` of zero on many rows. Look at those rows and their `n_issues`. Is zero a real number here, or a way of not answering?
- `filings.csv` also has `client_state`. Group by it. Which states buy the most federal lobbying, and does the ranking look like population, or like something else?
- `lobbyists.csv` has a `former_government_position` column, empty for most rows. Read the filled ones and sort them into genuine public posts and everything else. What share of the `r pc(100*mean(lo$has))`% survives?
- `issues.csv` and `entities.csv` rank what gets lobbied on and who gets lobbied, by `mentions`. Does either ranking look like what the news says lobbying is about?

Two stretch ideas point past the spreadsheet. `quarter.csv` records how many filings existed against how many were taken; say which of your answers above would change if you had all `r n(QTOTAL)`. And take one client, find its next quarter's filing on the Senate's site, and see whether the figure moved or simply repeated.

## Sources

**Data**

- United States Senate, Lobbying Disclosure Act database, filings API, Q2 2024.
  <https://lda.senate.gov/api/>
- The filed LD-2 in Figure 2 is the public print view of filing
  `302fd116-100b-475c-9d3b-37c3c78dcdc5` (Barker Leavitt, PLLC for Millcreek,
  Utah, Q2 2024), retrieved 10 August 2026 from
  <https://lda.gov/filings/public/filing/302fd116-100b-475c-9d3b-37c3c78dcdc5/print/>.
  The filer's telephone number and e-mail were redacted before it was
  photographed, and four annotations added. Nothing else on the image was
  altered.

**Academic literature**

- Blanes i Vidal, J., Draca, M. and Fons-Rosen, C. (2012). "Revolving Door
  Lobbyists." *American Economic Review* 102(7): 3731–3748.
- Baumgartner, F. R., Berry, J. M., Hojnacki, M., Kimball, D. C. and Leech, B. L.
  (2009). *Lobbying and Policy Change: Who Wins, Who Loses, and Why.* University
  of Chicago Press.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`entities.csv`](data/derived/entities.csv), [`filings.csv`](data/derived/filings.csv), [`issues.csv`](data/derived/issues.csv), [`lobbyists.csv`](data/derived/lobbyists.csv), [`quarter.csv`](data/derived/quarter.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
