---
title: "Tue Oct 20 Key: The Most Reliable Thing in American Politics"
subtitle: "84-355 Democracy's Data | The base rate — test 1 day, revisited in Session 10 | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
h <- read.csv("data/house_midterms.csv", stringsAsFactors = FALSE)
m <- h[h$midterm & !is.na(h$pres_party_change), ]
m$pct <- 100 * m$pres_party_change / m$seats
```

All Part 6 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

One sentence — *the president's party loses House seats at the midterm* —
forecasts the direction of a national election 88% of the time across 41 tries,
which is more than any poll, model or pundit in this course manages.

***

# The findings

```{r headline}
c(midterms = nrow(m),
  lost     = sum(m$pres_party_change < 0),
  pct_lost = round(100 * mean(m$pres_party_change < 0)),
  mean     = round(mean(m$pres_party_change), 1),
  median   = median(m$pres_party_change))
```

- **41 midterms since 1858. The president's party lost seats in 36 — 88%.**
- **Mean −32.5 seats, median −28.**
- Five exceptions: **1866 (+37), 1902 (+7), 1934 (+9), 1998 (+4), 2002 (+8)**.
- Worst: **1894 (−125)**, 1874 (−96), 1890 (−93), 1922 (−77), 1938 (−72).

***

# The idea to teach

Two, and the second is the one that will still be useful in ten years.

**First:** this is the single most reliable regularity students will meet all
term, and it requires no data collection, no model and no money. Set it against
Session 9 next week — a poll of 1,000 costs real money and gets you a few points of
precision on one question in one state. This gets you the direction of a
national election from a sentence.

**Second, and harder:** **Part 4 gives two different answers depending on the
unit.** Raw seats say the penalty has collapsed since the nineteenth century.
As a share of the chamber it has fallen too, but far less, and it *rises* again
in the most recent era. The House had 226 seats in 1866 and has 435 now;
comparing raw counts across that change is the same error as comparing county
vote totals without normalising, which they met in Session 4.

If a student leaves able to ask "per what?" before reading a trend, the lab has
done its job.

***

# Option A — Drop Reconstruction

## Verified output

```{r a}
a <- m[m$election_year >= 1878, ]
rbind(all   = c(n = nrow(m), pct_lost = round(100*mean(m$pres_party_change < 0)),
                mean = round(mean(m$pres_party_change), 1)),
      from_1878 = c(n = nrow(a), pct_lost = round(100*mean(a$pres_party_change < 0)),
                mean = round(mean(a$pres_party_change), 1)))
```

| | n | lost | mean |
|---|---|---|---|
| all | 41 | 88% | −32.5 |
| 1878+ | 37 | 89% | −32.8 |

**Almost nothing moves.** That is the answer, and the lab tells students to
expect it.

## What a good answer says

**3/4:** Reports that the exclusion barely changes anything.

**4/4:** Says why that is informative rather than disappointing. One observation
out of 41 cannot move a mean much — 1866 is a genuinely bad case on the merits
(Congress was refusing to seat Southern delegations, so the chamber is not
comparable), and it *still* does not matter arithmetically. **Being right to
exclude a case and finding the exclusion changes nothing are two separate
things**, and a student who separates them is doing real methodological work.

**4/4, best:** notices the corollary — if one bad case does not move the result,
then the result is also not being rescued by any single good case. Robustness
cuts both ways.

**Common stumble:** concluding "so it didn't matter, I shouldn't have bothered."
The check is what licenses the claim. You do not know it is robust until you
look.

***

# Option B — First term or second? ("the six-year itch")

## Verified output

```{r b}
h2 <- h[order(h$election_year), ]
run <- integer(nrow(h2)); r <- 0
for (i in seq_len(nrow(h2))) {
  r <- if (i == 1 || h2$pres_party[i] != h2$pres_party[i-1]) 0 else r + 2
  run[i] <- r
}
h2$tenure <- run
mm <- h2[h2$midterm & !is.na(h2$pres_party_change), ]
mm$pct <- 100 * mm$pres_party_change / mm$seats
early <- mm[mm$tenure <= 2, ]; later <- mm[mm$tenure >= 6, ]
rbind(early = c(n = nrow(early), seats = round(mean(early$pres_party_change),1),
                pct = round(mean(early$pct),1)),
      later = c(n = nrow(later), seats = round(mean(later$pres_party_change),1),
                pct = round(mean(later$pct),1)))
```

| | n | mean seats | mean % |
|---|---|---|---|
| party in power ≤2 yrs | 18 | −35.6 | −9.4 |
| party in power ≥6 yrs | 11 | **−40.3** | **−11.3** |

**The itch is real in direction and weak in evidence.**

## What a good answer says

**3/4:** Finds that later midterms are worse and reports the gap.

**4/4:** Refuses to oversell it. **Eleven observations**, and the difference is
about five seats — well inside what you would expect from chance with groups
this small. A student who says "the folk wisdom points the right way but this
cannot confirm it" is exactly right, and is doing something most published
commentary does not.

**4/4, best:** notices that the operationalisation is a choice. "Second midterm"
could mean the second midterm of one president, or six years of one *party*
holding the White House — and Ford, Johnson and Truman make those different. The
lab uses party tenure. A student who spots that the answer depends on a
definition nobody stated has found the real issue.

***

# Option C — Change the eras

## Verified output

```{r c}
aggregate(cbind(pres_party_change, pct) ~ era, transform(m,
  era = cut(election_year, c(1857,1900,1950,1980,2000,2030),
            labels = c("1858-1900","1902-1950","1952-1980","1982-2000","2002-2024"))),
  function(x) round(mean(x), 1))
```

| era | mean seats | mean % |
|---|---|---|
| 1858–1900 | −40.7 | −12.8 |
| 1902–1950 | −37.8 | −8.8 |
| 1952–1980 | −27.9 | −6.4 |
| 1982–2000 | −17.6 | −4.0 |
| **2002–2024** | **−25.0** | **−5.7** |

## What a good answer says

**3/4:** Moves the cut points and reports that the trend is broadly stable.

**4/4:** Notices the last row goes back **up** on both measures, so "the penalty
is shrinking" was never the right description — it fell through the twentieth
century and has partly rebounded since 2002. Then connects that to **Chapter 10**:
a rebound in the era of nationalised congressional elections is what the
nationalisation argument predicts, and it is the raw-seats table that hides it.

**4/4, best:** says the honest thing about era boundaries — with 41 cases split
five ways, some cells hold three or four elections, and almost any story can be
produced by moving a line. The right response is not to pick better boundaries
but to distrust the whole exercise and look at the year-by-year plot instead.

***

# Option D — Their own question

Good versions: is the penalty larger when the president won narrowly; does it
correlate with how many seats the party gained in the prior presidential year
(regression to the mean); do the five exceptions share anything (they mostly
involve a crisis or a scandal aimed at the *other* party).

***

# Part 5 — the lab runs twice

**This lab is deliberately split across two sessions**, and the gap between them
is the lesson.

**Oct 20, after Test 1.** `CHANGE_2026` is `NA`, so the chunk prints only the
historical distribution — which is genuinely all anybody has two weeks out.
**Make every student write a number down and keep it.** Collect them if you can:
a photo of the board, a Canvas quiz with one numeric field, anything you can
produce again on Nov 5. If the numbers are not recorded, the second half of this
lab does not work, because people reliably misremember having predicted whatever
happened.

**Nov 5, in the election post-mortem.** Reopen this file, enter the seat change,
re-run. The chunk reports the percentile against all 41 midterms. Ten seconds,
and it is the payoff for two weeks of waiting.

**The question to ask on Nov 5 is not "who was right."** It is *how many of you
beat the median?* If the class as a whole did not beat simply reading the median
off a table, that is the finding — and it is the same lesson as Session 9's
polling lab the following week, arriving from the other direction. Sophistication is not free, and
a base rate is very hard to beat.

**If the count is unresolved on Nov 5**, use the best available estimate and say
out loud that you are doing so. A provisional number with a stated caveat is a
better lesson than waiting for certainty.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is asking "per what?" — noticing that seats and
  shares tell different stories and saying which is honest.
- **Reward calibrated language.** This lab makes it very easy to overclaim.
  "The president's party always loses" is wrong; 88% is not always.

## Anticipated questions

- *"Why does the president's party lose?"* — Several explanations compete:
  regression to the mean after presidential coattails, the out-party's supporters
  being angrier and turning out, and voters balancing power deliberately. The lab
  does not settle it and should not pretend to. Worth five minutes precisely
  because the pattern is so clear and the cause is not.
- *"Is 1866 cheating?"* — Yes, and Option A shows it does not matter. Both things
  are true.
- *"Where do the seat counts come from?"* — The House's own party-division table.
  Say why we did not count members from the DW-NOMINATE file (Session 12): replacements make 449
  "members" in a 435-seat chamber. This is a live example of a unit-of-analysis
  error that would have gone unnoticed.
- *"Could you have forecast 2026 with this?"* — Yes, directionally, and better
  than most of what students read. That is the uncomfortable point.
- *"Why start at 1858?"* — Stable two-party Democratic-versus-Republican
  competition. Earlier Congresses have different party systems and the columns in
  the source table change names.

## Connections

- **Session 10's election-night lab (Nov 5)** — this one sets the prior;
  that one delivers the result. Run this first, two weeks ahead, and revisit it
  live on the night.
- **Chapter 10** — the nationalisation claim is the week's debate, and Option C
  is the closest thing to a test of it that we have.
- **Session 4's vote-targeting lab** — the "per what?" lesson, second appearance: county
  vote totals there, chamber size here.
- **Session 9 (polls, models, and markets)** — a one-sentence rule outperforming a 1,000-person poll is
  worth naming out loud. It is not an argument against polling; it is an argument
  about what each instrument is for.
