---
title: "Lab 11a: The Most Reliable Thing in American Politics"
subtitle: "84-355 Democracy's Data | The midterm penalty, 1858–2026"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# The question

**In two weeks the country votes.** Before it does, we are going to write down
what *should* happen — because there is a pattern here so regular that you can
forecast a midterm knowing nothing about the candidates, the issues, the money,
or the year.

**The president's party loses House seats at the midterm.** That is it. That is
the whole rule.

Today you find out how reliable it is, how big it is, and whether it is still
true. **Then you commit to a number, in writing, before a single vote is
counted.** On **Nov 5** we open this lab again, put the real result in, and find
out how the rule did.

That order is the whole point. A prediction made after the fact is not a
prediction, and by Nov 5 you will not be able to remember what you would have
guessed — nobody ever can. So write it down today.

# Part 1 — The data

```{r load}
h <- read.csv("data/house_midterms.csv", stringsAsFactors = FALSE)
str(h)
tail(h[, c("election_year", "seats", "dem", "rep", "pres_party",
           "midterm", "pres_party_change")], 6)
```

One row per House election since 1858. `pres_party` is the party of the sitting
president at that election; `pres_party_change` is how many seats that party
gained or lost compared with the previous election.

**A warning about where this came from.** The Week 7 lab has a file listing
every House member's party for every Congress, and it looks like counting those
rows would give you seat totals. It does not. Members who die, resign or are
replaced mid-term appear next to their replacements, so that file "has" 449
members in a 435-seat chamber. Seats and people are different units. These
numbers come from the House's own official party-division table instead.

# Part 2 — How reliable is it?

```{r reliability}
m <- h[h$midterm & !is.na(h$pres_party_change), ]

cat("Midterm elections in the data:", nrow(m), "\n")
cat("President's party LOST seats in", sum(m$pres_party_change < 0),
    sprintf("(%.0f%%)\n", 100 * mean(m$pres_party_change < 0)))
cat("Mean change:  ", round(mean(m$pres_party_change), 1), "seats\n")
cat("Median change:", median(m$pres_party_change), "seats\n")
```

Almost nothing in political science is this consistent. Hold that against the
polling lab: a poll of 1,000 people gets you within a few points if you are
lucky. This gets you the direction of a national election, nine times in ten,
from a rule you can say in one sentence.

```{r plot-history}
plot(m$election_year, m$pres_party_change, type = "h", lwd = 3,
     col = ifelse(m$pres_party_change < 0, "#C41230", "#2c7fb8"),
     xlab = "", ylab = "Seat change, president's party", main = "")
abline(h = 0, lwd = 1)
points(m$election_year, m$pres_party_change, pch = 19, cex = 0.6,
       col = ifelse(m$pres_party_change < 0, "#C41230", "#2c7fb8"))
grid(col = "grey88", lty = 3)
```

# Part 3 — The exceptions

Five midterms went the other way. They are worth knowing by name.

```{r exceptions}
m[m$pres_party_change >= 0,
  c("election_year", "pres_party", "pres_party_change")]
```

```{r worst}
head(m[order(m$pres_party_change),
       c("election_year", "pres_party", "pres_party_change")], 5)
```

**1866 is not what it looks like.** Congress was refusing to seat delegations
from the former Confederacy, so the chamber that year is not comparable to the
one before it. Any pattern that includes Reconstruction is measuring an argument
about who counts as a state, not an ordinary election.

# Part 4 — Is the penalty shrinking?

Chapter 10 argues congressional elections have been nationalised, which should
make the midterm penalty *bigger* — if a House vote is really a vote about the
president, the president's party should suffer more when he is unpopular. But
the raw numbers seem to say the opposite.

```{r era-seats}
m$era <- cut(m$election_year, c(1857, 1900, 1950, 1980, 2000, 2030),
             labels = c("1858-1900", "1902-1950", "1952-1980",
                        "1982-2000", "2002-2024"))
aggregate(pres_party_change ~ era, m, function(x) round(mean(x), 1))
```

Now look again, in a unit that survives the comparison.

```{r era-pct}
m$pct <- 100 * m$pres_party_change / m$seats
aggregate(pct ~ era, m, function(x) round(mean(x), 1))
```

**The House had 226 seats in 1866 and 435 today.** Losing 40 seats out of 226 is
not the same event as losing 40 out of 435, and the first table quietly compares
them as though it were. The second table divides by chamber size. The story
changes — not completely, but enough that you would say something different.

# Part 5 — And 2026?

Today this prints the historical distribution and nothing else, because the
election has not happened. **That is what a forecast looks like from the
inside:** a range, a centre, and no answer.

**Write your number down now** — in the box below, on the board post, anywhere
you will still have it on Nov 5. One number: how many House seats the
president's party gains or loses. You are allowed to just use the median. You
are also allowed to depart from it, but if you do, say why in one sentence.

```{r twenty-six}
# Left as NA until the seats are called on Nov 5, when we reopen this lab and
# put the real number in. Until then the chunk prints the historical
# distribution only -- which is all anybody actually has before an election.
CHANGE_2026 <- NA        # <- seat change for the president's party in 2026

cat("Historical midterm distribution (president's party seat change):\n")
print(summary(m$pres_party_change))
cat("\nAs a share of the chamber:\n")
print(round(summary(m$pct), 1))

if (!is.na(CHANGE_2026)) {
  pctile <- 100 * mean(m$pres_party_change <= CHANGE_2026)
  cat("\n2026:", CHANGE_2026, "seats\n")
  cat("That is worse than", round(pctile), "% of midterms since 1858.\n")
} else {
  cat("\n2026: not entered yet. Put the number in and re-run.\n")
}
```

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Drop Reconstruction.** In Part 2, use `m <- m[m$election_year >= 1878, ]`
before computing the summary. I told you 1866 was not a comparable election.
Now find out whether excluding it actually changes any conclusion — and be ready
for the answer to be no. If a bad case turns out not to matter, that is worth
knowing, and it tells you something about how much one observation can move an
average built from forty-one.

**Option B — First term or second?** A president's *second* midterm is supposed
to be worse than his first — the "six-year itch." Add
`m$second <- m$election_year %% 4 == 2` is not the test; you need to know when
each presidency began. Using the list in `data/build-data.R`, check a handful by
hand and say whether the folk wisdom survives.

**Option C — Change the eras.** The cut points in Part 4 are mine and they are
arbitrary. Move them. Does the "shrinking penalty" conclusion hold under
different boundaries, or did I choose the boundaries that produce it?

**Option D — Your own question.** 84 elections, one row each.

# On the discussion board

By **Friday, 11:59 p.m.** — that evening or the next day, while it is fresh.
No length requirement; a sentence or two is enough, so long as it shows you
engaged with what we did. Any *one* of the usual four: something that surprised
you, something we missed, something you do not understand, or an answer to a
classmate.

Three that would make good posts:

- We just forecast a national election from one sentence and got the direction
  right 88% of the time. What does that say about everything else campaigns do?
- Part 4 gave two different answers depending on whether we counted seats or
  shares. Which is the honest one, and how would you have known before seeing
  both?
- Chapter 10 says elections are nationalising. Does the midterm penalty support
  that, contradict it, or fail to speak to it?

***

### Where these numbers come from

Office of the Historian, U.S. House of Representatives, *Party Divisions of the
House of Representatives, 1789 to Present*,
<https://history.house.gov/Institution/Party-Divisions/Party-Divisions/>. The
party of the president at each election is a lookup table written by hand in
`data/build-data.R`; it is the one part of this lab not machine-read from a
source, so it is the part most likely to contain a mistake. Check it if a year
looks wrong.
