---
title: "Lab 10: A Poll That Is Precise and Wrong"
subtitle: "84-355 Democracy's Data | Week 10"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
set.seed(2026)
```

# Before we start

The country votes on Tuesday. Between now and then you will see a great many
numbers with a **±3** attached to them, and it is worth knowing exactly what that
number promises — and, more importantly, what it does not.

Every other lab this term handed you real data and asked what it showed. Today
we do the opposite. We are going to **build a population where we already know
the answer**, and then poll it. That is the only way to watch a poll miss,
because in real life the truth arrives after the poll and nobody can go back.

The truth we will use is the real one. In 2024, of everyone who voted for one of
the two major candidates:

- **Trump 50.75%**
- **Harris 49.25%**

A margin of 1.5 points. Remember that number; everything today is measured
against it.

Everything is base R. Work in order.

***

# Part 1: A country we can see all of

```{r population}
truth <- 0.4925                      # Harris share of the two-party vote
N     <- 1000000                     # a million voters

population <- c(rep("Harris", round(N * truth)),
                rep("Trump",  N - round(N * truth)))

table(population)
round(100 * mean(population == "Harris"), 2)
```

We know the answer. No pollster ever does. That is the only thing that is
unrealistic here, and it is the point.

***

# Part 2: One poll

```{r one-poll}
poll <- sample(population, 1000)
table(poll)
round(100 * mean(poll == "Harris"), 2)
```

Not exactly 49.25, but close. Run it again and you get a different answer:

```{r three-polls}
for (i in 1:3) {
  p <- sample(population, 1000)
  cat("poll", i, ":", round(100 * mean(p == "Harris"), 2), "\n")
}
```

That wobble is **sampling error** — the only kind of error the ±3 describes.

***

# Part 3: What the margin of error actually promises

Run a thousand polls and look at the spread.

```{r many-polls, fig.width = 8, fig.height = 5}
estimates <- replicate(1000, 100 * mean(sample(population, 1000) == "Harris"))

hist(estimates, breaks = 30, col = "#BBBBBB", border = "white",
     xlab = "Poll estimate of Harris's share (%)", ylab = "Polls",
     main = "One thousand honest polls of 1,000 people each")
abline(v = 100 * truth, lwd = 3, col = "#B2182B")
```

```{r moe-check}
moe <- 1.96 * sqrt(0.25 / 1000) * 100
c(margin_of_error = round(moe, 2),
  within_moe = round(100 * mean(abs(estimates - 100*truth) <= moe), 1))
```

> The margin of error is **±3.1 points**, and about **95%** of the polls land
> inside it. The promise is kept exactly.
>
> But read the promise carefully. It says: *if my only problem is that I asked
> 1,000 people instead of all of them, I will usually be within 3 points.* It
> says nothing whatever about any other problem.

***

# Part 4: More people, less error

```{r moe-curve, fig.width = 8, fig.height = 5}
n <- seq(200, 20000, by = 100)
plot(n, 1.96 * sqrt(0.25 / n) * 100, type = "l", lwd = 2.5, col = "#54278F",
     xlab = "People polled", ylab = "Margin of error (points)",
     main = "Bigger polls are more precise")
abline(h = 3, lty = 3, col = "grey50")
```

```{r moe-table}
for (k in c(500, 1000, 5000, 100000)) {
  cat(sprintf("n = %6d   margin of error = %.2f points\n",
              k, 1.96 * sqrt(0.25 / k) * 100))
}
```

Familiar, and true. Now we break it.

***

# Part 5: The problem that is not sample size

Everything above assumed that when you call somebody, they answer. In 2024 the
response rate to a typical telephone poll was **under 1%**.

That would be survivable if the 1% who answer were a random 1%. They are not.
Suppose Harris voters are somewhat likelier to pick up the phone — 20% likelier.

```{r biased-poll}
answers <- ifelse(population == "Harris", 1.20, 1.00)   # who picks up

biased <- sample(population, 1000, prob = answers, replace = TRUE)
round(100 * mean(biased == "Harris"), 2)
```

That is not sampling noise. Do it a thousand times:

```{r biased-many, fig.width = 8, fig.height = 5}
biased_est <- replicate(1000,
  100 * mean(sample(population, 1000, prob = answers, replace = TRUE) == "Harris"))

hist(biased_est, breaks = 30, col = "#F4A58255", border = "white",
     xlim = c(45, 60),
     xlab = "Poll estimate of Harris's share (%)", ylab = "Polls",
     main = "One thousand polls, when Harris voters answer more often")
abline(v = 100 * truth, lwd = 3, col = "#B2182B")
abline(v = mean(biased_est), lwd = 3, lty = 2, col = "#2166AC")
legend("topright", c("the truth", "average poll"), col = c("#B2182B","#2166AC"),
       lwd = 3, lty = c(1, 2), bty = "n")
```

```{r biased-summary}
c(truth = 100 * truth,
  average_poll = round(mean(biased_est), 2),
  bias = round(mean(biased_est) - 100 * truth, 2))
```

> The polls are still tightly clustered. They are just clustered **around the
> wrong number**. Averaging them together — which is what a polling average
> does — does not help at all, because they are all wrong in the same direction.

***

# Part 6: The bit that should worry you

Bias does not care how many people you ask.

```{r the-demonstration}
for (k in c(1000, 10000, 100000)) {
  s <- sample(population, k, prob = answers, replace = TRUE)
  est <- 100 * mean(s == "Harris")
  cat(sprintf("n = %6d   estimate %.2f   reported +/- %.2f   actually wrong by %+.2f\n",
              k, est, 1.96 * sqrt(0.25/k) * 100, est - 100 * truth))
}
```

> Look at the last line. A poll of **100,000 people** reports a margin of error
> of **±0.31** and is wrong by about **4.5 points** — roughly **fifteen times**
> its own stated precision.
>
> And it does not drift toward the truth as $n$ grows. It converges — on
> **53.8%**, which is not the answer. More data makes you more confident about
> the wrong number.

This is the argument in Bailey's *Polling at a Crossroads*, and it is why the
polling industry's problem in the last three presidential elections has not been
one that a bigger sample could fix.

***

# Part 7: Weighting, and why it is not a cure

Pollsters know this, and they correct for it. If you know Harris voters are
over-represented, you can count their answers for less. The catch is that you
have to **know which variable is doing the damage**, and be able to see it.

```{r weighting-setup}
# A population with two attributes: how you vote, and something else about you
# — call it trust in institutions — that we will let drive who responds.
N2 <- 200000
harris <- rbinom(N2, 1, truth)
trust  <- rbinom(N2, 1, ifelse(harris == 1, 0.75, 0.45))
true_share <- 100 * mean(harris)

responds <- ifelse(trust == 1, 1.00, 0.35)   # low-trust people rarely answer
i <- sample(N2, 5000, prob = responds, replace = TRUE)
s <- data.frame(harris = harris[i], trust = trust[i])

round(c(truth = true_share, raw_poll = 100 * mean(s$harris)), 2)
```

Weight on **trust** — the thing actually driving non-response:

```{r weight-right}
target <- mean(trust)                            # the true share, if you knew it
w <- ifelse(s$trust == 1, target / mean(s$trust),
                          (1 - target) / (1 - mean(s$trust)))
round(c(weighted = 100 * weighted.mean(s$harris, w),
        error    = 100 * weighted.mean(s$harris, w) - true_share), 2)
```

Now weight on something real, measurable, and **unrelated** to who answers:

```{r weight-wrong}
irrelevant <- rbinom(nrow(s), 1, 0.5)
w2 <- ifelse(irrelevant == 1, 0.5 / mean(irrelevant), 0.5 / (1 - mean(irrelevant)))
round(c(weighted = 100 * weighted.mean(s$harris, w2),
        error    = 100 * weighted.mean(s$harris, w2) - true_share), 2)
```

> Weighting on the right variable takes the error from about **+5.5 points to
> under one**. Weighting on the wrong variable changes **nothing** — the estimate
> does not move.
>
> So weighting works, exactly and only when you already know what is wrong. In
> 2016 the industry discovered it had not been weighting by education, and fixed
> it. In 2020 it missed again, in the same direction. The variable you have not
> thought of is invisible by construction, and no amount of arithmetic finds it.

***

# Part 8: What to carry into Tuesday

**A margin of error is a statement about sample size and nothing else.** It is
honest and it is narrow. It does not cover bias, and bias is the thing that has
been going wrong.

**Polling averages do not fix a shared error.** If every poll is wrong in the
same direction, so is the average of them.

**And the errors have been correlated across states.** When the polls miss, they
tend to miss the same way everywhere at once, which is why the states you care
about most are the ones where being wrong matters most.

On Thursday next week we will have the answers, and you can check the forecast
you read against what actually happened.

***

# Part 9: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — How big does the bias have to be?

```{r opt-a, fig.width = 8, fig.height = 5}
gap <- 0.20                # <- change this. Try 0.05, 0.10, 0.40.

ans <- ifelse(population == "Harris", 1 + gap, 1)
est <- replicate(300, 100 * mean(sample(population, 1000, prob = ans,
                                        replace = TRUE) == "Harris"))

hist(est, breaks = 25, col = "#F4A58255", border = "white", xlim = c(44, 62),
     xlab = "Poll estimate (%)", ylab = "Polls",
     main = paste0("Harris voters ", round(100*gap), "% likelier to respond"))
abline(v = 100 * truth, lwd = 3, col = "#B2182B")

c(response_gap = gap, average_poll = round(mean(est), 2),
  bias = round(mean(est) - 100 * truth, 2),
  moe_claimed = 3.1)
```

**Write about:** how small a response gap does it take to matter? Start at 0.05
— a 5% difference in willingness to answer, which sounds like nothing — and
report the bias it produces. Compare that to the ±3.1 the poll would advertise.
Then say what this implies for reading a poll that shows a 1.5-point race, which
is what 2024 actually was.

***

### Option B — Does averaging help?

```{r opt-b}
n_polls <- 20              # <- change this. Try 5, 20, 200.

ans <- ifelse(population == "Harris", 1.20, 1)

honest <- replicate(n_polls, 100 * mean(sample(population, 1000) == "Harris"))
skewed <- replicate(n_polls, 100 * mean(sample(population, 1000, prob = ans,
                                               replace = TRUE) == "Harris"))

round(c(truth          = 100 * truth,
        avg_of_honest  = mean(honest),
        avg_of_skewed  = mean(skewed),
        spread_honest  = sd(honest),
        spread_skewed  = sd(skewed)), 2)
```

**Write about:** averaging many honest polls works — the average closes in on the
truth as you add more. Averaging many biased polls does not, however many you
add. Report both. Then the question that matters this week: polling *aggregators*
average dozens of polls and report a very tight interval. What is that interval
actually measuring, and what would have to be true for it to mean what readers
take it to mean?

***

### Option C — Weighting on the wrong thing

```{r opt-c}
strength <- 0.35           # <- how strongly trust drives response. Try 0.1, 0.35, 0.8.

resp <- ifelse(trust == 1, 1.00, strength)
j <- sample(N2, 5000, prob = resp, replace = TRUE)
ss <- data.frame(harris = harris[j], trust = trust[j])

tgt <- mean(trust)
wr <- ifelse(ss$trust == 1, tgt/mean(ss$trust), (1-tgt)/(1-mean(ss$trust)))

round(c(truth       = true_share,
        raw         = 100 * mean(ss$harris),
        weighted    = 100 * weighted.mean(ss$harris, wr),
        raw_error   = 100 * mean(ss$harris) - true_share,
        wtd_error   = 100 * weighted.mean(ss$harris, wr) - true_share), 2)
```

**Write about:** as `strength` falls toward 1, non-response stops mattering and
the raw poll is fine. As it falls toward 0.1, almost nobody low-trust answers and
the raw poll is badly wrong — but weighting still rescues it, *because we can see
the variable*. State plainly what would happen if the variable driving
non-response were something the pollster does not measure. Then connect it to
Week 9: attention and response are both behaviours we can observe, and belief is
not.

***

### Option D — Ask your own question

Change the truth to a landslide instead of a 1.5-point race and see whether bias
still matters. Or find out how large a poll would have to be for its margin of
error to shrink below the bias — a trick question worth answering.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can read a number with a ± after it and say what that symbol
does and does not cover.

***

## Where the data came from

Mostly, it did not. **This lab is a simulation**, and that is the design rather
than a compromise: you cannot watch a poll fail unless you know the true answer,
and with real polling data you never do until afterwards.

The truth the simulation is built around is real — the 2024 national two-party
vote, **Trump 50.75% to Harris 49.25%**, from the compilation used in Week 4.
`data/pres2024_states.csv` is the state-level file from Weeks 1, 2 and 6, copied
here so the lab stands alone and so Option D can use it.

Two honest limits on what a simulation can show:

- **The response gaps here are invented.** Twenty percent is a plausible-looking
  number, not a measured one. The *shape* of the argument is what transfers, not
  the size of any particular bias.
- **Real non-response is not one variable.** It is many, correlated, and shifting
  between elections. Part 7 uses a single tidy variable so the mechanism is
  visible; reality gives pollsters no such favour.

If you want the real thing, the **Cooperative Election Study**
(<https://doi.org/10.7910/DVN/X11EP6>) publishes tens of thousands of interviews
per cycle with its weights attached, and its documentation of how those weights
were built is worth reading alongside Bailey.
