---
title: "Comparing Precincts Across Two Elections"
type: brief
subtitle: "Precinct names and precinct lines drift independently, so a precinct that swung may be a precinct that moved"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("precinct-geography-code.R")
```

```{r setup, include=FALSE}
```

Every account of an election contains sentences like *turnout in this precinct fell nine points*, or *this precinct swung twelve to the challenger*. They read like measurements, and each carries a second claim that is never stated and almost never checked: that "this precinct" is the same piece of ground in both elections. **Can you compare a precinct with itself four years later?**

Often you cannot, and Houston County, Georgia (Warner Robins, Perry, and `r n(HPOP)` people in `r n(length(h20))` precincts) shows why. Between the 2020 and 2024 presidential elections, `r n(length(setdiff(h20, h24)))` of its precinct names disappeared and a precinct that had never existed appeared. Nothing in either year's file records what any change was: the 2024 map does not say what it replaced, and no county has to announce that it moved a line.

The stakes are larger than one county's bookkeeping. The precinct is the smallest unit at which American voting can be observed, and everything fine-grained is built on it: voting-rights analysis, campaign targeting, journalism about which neighborhoods shifted. And counties split precincts where population grows, so the places most likely to be miscompared are exactly the fast-changing suburbs coverage cares about most. An error of this kind does not look like an error. It looks like a finding.

## The test

Three files answer it, from three different publishers. The precinct boundaries belong to Georgia's Legislative and Congressional Reapportionment Office, which posts a statewide precinct map for each general election (<https://www.legis.ga.gov/joint-office/reapportionment>). The two read here are the `r n(ts$shapefile_precincts[ts$scope == "Georgia"])`-precinct map of 2020 and the map that replaced it in 2024. The population belongs to the Census Bureau: `r n(nrow(ba))` Georgia census blocks from TIGER/Line 2020, carrying the 2020 census count. The votes belong to the Secretary of State's certified precinct returns for November 2020.

Why this data can answer: the precinct sits at the bottom of the ladder of election reporting, the finest rung published statewide, and [the levels-of-aggregation chapter](../levels-of-aggregation/levels-of-aggregation-brief.html) walks that ladder. This brief asks the question that chapter leaves open — whether the bottom rung holds still between elections. Georgia is the test state because one office publishes both vintages in one format, across the interval coverage actually compares.

## The data

A shapefile is not a picture: beside the geometry sits an attribute table, one row per shape. The 2020 table:

```{r raw20}
```

`r n(D20[2])` columns, and the names are not the names anybody chose. The format cuts every column name off at ten characters, so `JORGENSON2` is a truncated candidate. `VOTED20` and `VOTED201` are two different quantities, told apart by a digit the format appended. The votes sit bolted straight to the geometry; hold onto that. The 2024 release from the same office is a different kind of object:

```{r raw24}
```

`r n(D24[2])` columns and not one of them is a vote. Neither file carries a field pointing at the other — no prior name, no successor code. The only shared handle is county name plus precinct name, so the build manufactures a key that exists in no source file: `HOUSTON|ROZR`. About thirty rows a year carry neither name and collapse to a single non-precinct, `NA|NA`, which is dropped everywhere before anything is counted.

One more caution sits upstream. The counties hold the official lines and may redraw them at any time; the office itself calls its statewide file a "best fit" of the 159 county maps, snapped to census geography.

The census blocks are what make comparison possible. Each block carries `GEOID20`, a fifteen-digit identifier the Bureau maintains and does not reuse — a real key. Each block is reduced to a point guaranteed to lie inside it, and that point is asked two questions: which 2020 precinct contains you, and which 2024 precinct? Three of ROZR's blocks:

```{r cleanblocks}
```

Two blocks stayed and one moved; that third row is this brief's whole finding, written out once. Grouping the `r n(nrow(ba))` blocks by their pair of precincts and summing population turns the assignments into a **crosswalk**, a table that translates one set of areas into another, with a weight on every pair. `crosswalk_pop.csv` holds `from_2020`, `to_2024`, the `weight`, and the `pop` behind it; `block_assign.csv` holds every block's `GEOID20`, `pop`, and two precinct keys. The running example:

```{r cleanxwalk}
```

The votes are read straight off the 2020 map's own attribute table rather than joined on by name. That shortcut is legitimate only if the map's copy is right, so it was checked against the Secretary of State's certified returns:

```{r twosource}
```

On every precinct where the two publications could be compared, they agreed to the vote. An earlier build had joined the votes on by precinct name instead, and silently lost 489,876 votes (9.8% of Georgia) to spelling differences. The data was never wrong. The labels were.

> **Before you read on, commit to a number.** Between 2020 and 2024, Houston
> County's `r n(length(h20))` precincts produced
> `r n(length(REN20) + length(RDR20))` changes:
> `r n(length(setdiff(h20, h24)))` names disappeared and
> `r n(length(setdiff(nm(h24all), nm(h20))))` name appeared that had not
> existed before. **How many of those changes moved actual ground from one
> precinct into another**, and how many are a polling place renamed while
> every resident stayed put? Write down a number.

## What it says about democracy

Start with the names, which are all most analyses ever see:

```{r moved}
```

Only `r pc(surv(p20, p24))`% of Georgia's precinct names survive the four years, and Houston, at `r pc(surv(h20, h24))`%, is slightly harder than average. Measured on the counties' precinct *codes* instead of names, statewide survival drops to about three quarters. Two keys, two answers, and neither is wrong. An identifier that changes when the thing it names does not, and holds still when the thing it names changes, is not an identifier.

Houston is small enough to show every change in full: every 2020 precinct and what became of it, then the new 2024 name, with the precinct that supplied its people:

```{r inventory}
```

```{r inventory-new}
```

Most readers guess that most of the changes are real. **The ground moved in `r n(length(RDR20))` of the `r n(length(REN20) + length(RDR20))`.** The other `r n(length(REN20))` are renames: HCTC became VFW, NSES became NORTH HOUSTON SPORT COMPLEX, RECR became WELLSTON CENTER. In each case, weighting by population (counting some blocks for more than others, by the people living in them) sends 100% of the old precinct's people to the new name. The polling place moved; the ground did not. The one real change is ROZR, split in two, its western half becoming the new precinct PEC — and there, two defensible ways of dividing the old precinct's votes disagree about `r n(PECB)` ballots.

```{r houston-d3, results='asis', eval=knitr::is_html_output()}
```

```{r houston-static, eval=!knitr::is_html_output(), fig.height=5.0}
```

```{r houston-static-legend, eval=!knitr::is_html_output(), fig.height=0.55}
```

**Figure 1. Houston County's precincts in 2020 and 2024, with the two kinds of change separated.** A pair of maps, because the claim is about ground and only a map can show ground holding still while names change. Amber areas are the `r n(length(REN20))` renamed precincts whose boundaries stayed put; blue is ROZR, the one precinct whose ground actually changed; gray kept both name and shape. The 2024 panel counts `r n(length(h24all))` precincts where the table lists `r n(length(h24))`; the missing one is `r unseen`, explained below.

The tempting repair, keeping only the precincts whose names match in both years, is exactly wrong. In Houston, name-matching throws away `r n(length(dropped))` precincts (`r paste(nm(dropped), collapse = ", ")`, carrying `r n(sum(poH$pop[poH$from_2020 %in% dropped]))` people), and all three are renames where nothing happened. It keeps ROZR, the one place where something did. Name-matching is a filter on clerical churn, not on boundary change, and a rule whose sign can flip between a county and its state is not a rule to build on.

The fix is the crosswalk built above: newest boundaries as the target, old data broken down onto blocks, blocks added back up inside the new lines. A whole precinct reduced to one interior point can never record a split, which is why `r unseen` is invisible to the precinct-level table; `r n(nrow(baH))` blocks, each reduced to a point, can.

Then the choice that moves votes: when a precinct's ballots travel to more than one successor, in what proportion — by land, or by people?

```{r weights}
```

The two rules disagree even about how many precincts split at all. Most of area's extra "splits" are slivers: hairline overlaps where the two files digitized the same road a meter apart. Raise the floor to 1% of a precinct and area's Houston figure falls from `r pc(sp_arH)`% to `r pc(sp1(arH))`%, exactly matching population. Statewide, a residue of `r pc(sp1(ar) - sp1(po))` points survives the correction: boundaries clipping empty farmland, a split in land that no vote can come from. Where the choice bites is a real split:

```{r rozr-d3, results='asis', eval=knitr::is_html_output()}
```

```{r rozr-static, eval=!knitr::is_html_output(), fig.height=4.0}
```

**Figure 2. The one piece of Houston ground that changed precinct, and the `r n(WBAL)` ballots that have to be divided.** ROZR's `r n(nrow(wp))` census blocks, each dot sized by population and colored by where its interior point landed in 2024. The western half is big and thinly settled, so land gives PEC `r pc(100 * PECA)`% of ROZR while people give it `r pc(100 * PECP)`%. The gap is `r n(PECB)` ballots, moved from one precinct to another by a methodological choice, with nothing miscounted.

```{r d3, results='asis', eval=knitr::is_html_output()}
```

```{r scatter-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

**Figure 3. Where land and people give different answers, across Georgia.** Each point is one 2020-to-2024 precinct pair, placed by its weight under each rule, because the question is agreement and a scatter shows agreement as distance from the diagonal. Of the state's `r n(nrow(cp))` pairs, `r n(sum(abs(cp$diff) > 0.10))` (`r pc(100*mean(abs(cp$diff) > 0.10))`%) differ by more than ten points of weight. The labeled extreme, `r nm(cp$from_2020[which.max(abs(cp$diff))])` in `r cty(cp$from_2020[which.max(abs(cp$diff))])` County, is sent `r pc(100*cp$weight_area[which.max(abs(cp$diff))])`% of its precinct's votes by one rule and `r pc(100*cp$weight_pop[which.max(abs(cp$diff))])`% by the other. Houston's `r n(nrow(cpH))` pairs are green; its worst case is the ROZR split.

With a rule chosen, the carry itself is mechanical. It ends with the comparison the opening asked for: Houston's 2020 presidential vote expressed on Houston's 2024 map, including PEC, a precinct that did not exist when those votes were cast:

```{r carry}
```

A crosswalk that loses votes is worse than none, because it looks fine. So the carry is checked statewide against the totals it started from:

```{r check}
```

`r ck$value[ck$check == "votes lost"]` votes lost out of `r ck$value[ck$check == "votes before"]`. The trip can be audited, and this one arrives whole.

## What this brief cannot tell you

**Population is not voters.** The weights count children and non-citizens, and turnout varies inside a precinct in ways a headcount cannot see. Registered voters by block would be the right weight, and nobody publishes it: that is the binding limit here.

**The blocks are older than the boundaries.** Block populations are from the 2020 census; the 2024 lines came later, so growth in between is invisible to the weights — in Houston, precisely the growth that caused the split.

**No number above carries an error bar.** A carried-over total prints with the same authority as a counted one, and in the Houston table `r n(length(RDR24))` of the `r n(nrow(eH))` rows are estimates while nothing in the output says which.

**And one county is one county.** Houston shows the mechanism cleanly because it is simple. The statewide figures above say the rate travels; the anatomy of any other county has to be read the same way, by hand.

## What you should have learned

- Only `r pc(surv(p20, p24))`% of Georgia's precinct names survived from 2020 to 2024, and a changed name is usually not a changed boundary: Houston's `r n(length(REN20) + length(RDR20))` changes were `r n(length(REN20))` renames and `r n(length(RDR20))` real split.
- Keeping only name-matched precincts filters on the wrong thing: in Houston it deletes `r n(length(dropped))` precincts where nothing happened and keeps the one where something did.
- The weighting choice is not a technicality: land and people disagree by more than ten points of weight on `r pc(100*mean(abs(cp$diff) > 0.10))`% of Georgia's precinct pairs, and in ROZR the choice moves `r n(PECB)` ballots.
- A carry can be checked: pushed through the population crosswalk, Georgia's 2020 vote arrives on the 2024 map with `r ck$value[ck$check == "votes lost"]` votes lost.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `fig_twosource.csv` compares the shapefile's precincts against the certified list. Find the precincts in one and not the other. What is a precinct that exists on a map but not in the returns?
- `fig_houston_lab.csv` flags each Houston precinct as kept, renamed or split. Count each kind. What share of the county's precincts would a name-matcher have compared correctly?
- `fig_rozr_meta.csv` divides ROZR's ballots two ways and reports both. Compare `ballots_area` against `ballots_pop`. Which assumption would you defend to an editor, and why?
- `fig_rozr_blocks.csv` assigns each of ROZR's blocks a target precinct and a `pop`. Find the blocks with no people in them. What do empty blocks do to a weighting by area?
- `crosswalk_pop.csv` covers every county, with a `weight` per precinct pair. Pick a county you know and count its precincts that split. Quieter or louder than Houston?

Two stretch ideas point past the spreadsheet. The reapportionment office posts precinct maps back to 2012; fetch an older pair and measure whether this churn rate is new or normal. And find one county's own notice of a precinct change, in a board of elections agenda, and check it against what the two maps imply.

## Sources

**Data**

- Georgia Legislative and Congressional Reapportionment Office, *Statewide
  Voting Precincts*, 2020 and 2024 shapefiles.
  <https://www.legis.ga.gov/joint-office/reapportionment>
- U.S. Census Bureau, TIGER/Line 2020, Georgia tabulation blocks
  (`tl_2020_13_tabblock20`), with 2020 Census population.
  <https://www2.census.gov/geo/tiger/TIGER2020/TABBLOCK20/>
- Georgia Secretary of State, certified precinct returns, November 2020 —
  the independent check on the shapefile's vote columns.
  <https://sos.ga.gov/page/historical-elections-results>
- Houston County Board of Elections publishes its own precinct results at
  <https://www.houstoncountyga.gov/residents/election-results.cms> — a third
  copy of the same numbers, an independent check on any Houston figure above.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`fig_houston.csv`](data/derived/fig_houston.csv), [`fig_houston_lab.csv`](data/derived/fig_houston_lab.csv), [`fig_rozr_blocks.csv`](data/derived/fig_rozr_blocks.csv), [`fig_rozr_meta.csv`](data/derived/fig_rozr_meta.csv), [`fig_rozr_outline.csv`](data/derived/fig_rozr_outline.csv), [`fig_rozr_pts.csv`](data/derived/fig_rozr_pts.csv), [`fig_twosource.csv`](data/derived/fig_twosource.csv), [`crosswalk_pop.csv`](../ga-precinct-returns/data/derived/crosswalk_pop.csv), [`block_assign.csv`](../ga-precinct-returns/data/derived/block_assign.csv)

**Method**

- Cervas, J. `assignPolys`, *R-Functions*.
  <https://github.com/jcervas/R-Functions/tree/main/assignPolys>

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
