---
title: "The American Community Survey"
type: chapter
subtitle: "Three windows onto the same country, one of them closed, and a margin beside every number that has one"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("census-acs-code.R")
```

```{r setup, include=FALSE}
```

The [decennial chapter](../census-decennial/census-decennial-brief.html) ended on a short list of things the census does not ask: income, education, language, occupation, commuting, disability, ancestry. Every one of them is asked here.

The American Community Survey is where nearly everything people call "census data" actually comes from. It is published by the Census Bureau, on the Bureau's website, in the Bureau's house style — and it is **a survey**, with all the properties the [source chapter](../census-source/census-source-brief.html) gave that word. Roughly 3.5 million addresses a year, continuously, forever.

## Two instruments, side by side

```{r insttab}
```

Read the last row first. The ACS cannot apportion the House, and not for any practical reason. The Census Act tells the Bureau it may use sampling except in determining the population for apportionment, and in 1999 the Supreme Court read that section to bar a sampled apportionment count. The Constitution's "actual Enumeration" clause is the usual shorthand for this, but the ruling was about the statute and the Court did not reach the constitutional question. That is the legal line between the two instruments, and every other difference follows from which side of it a number was produced on.

Then read the second row. It answers a question this chapter has been inviting since its first sentence. *If the ACS is a survey, why is it here and not with the other surveys?*

**Because you have to answer it.** Refusing to answer the American Community Survey is punishable under 13 U.S.C. § 221, the same statute that compels the decennial count. The mail package says so on the envelope: *your response is required by law.* Nothing that arrives later in this book can say that. The American National Election Studies asks about eight thousand people whether they would like to participate, and most of them decline; nobody is fined for hanging up on the General Social Survey.

That difference is not a technicality about enforcement. It decides what each instrument can be made to say.

A survey you may refuse has to persuade its sample to stay. So it can ask what you believe, and it cannot promise that the people who answered resemble the people who did not. A survey you may not refuse can be asked of a random three and a half million addresses. It will still return something the Bureau will publish a margin of error for, a stated range for how far off the estimate could be. It can also be run every year forever, because it is funded as an obligation rather than a grant.

**Method puts the ACS with the surveys. Where the data came from puts it here.** This book is organised by where the data came from. That is why a sample of 3.5 million addresses is in *Counting People*, and a sample of eight thousand is in [Asking People](../../02-surveys/surveys-source/surveys-source-brief.html). Hold the two side by side when you get there: same country, same people, opposite obligation.

## Why there is more than one window

A single year of sample is thin. Spread 3.5 million addresses over a country of 3,222 counties and most counties get too few to say anything about. So the Bureau pools years, and publishes more than one product from the same survey.

```{r wintab}
```

**The one-year file covers `r nn(C1)` counties. The five-year file covers all `r nn(C5)`.** For **`r nn(MISS)` counties there is no one-year estimate at all**, which is `r p1(MISSP)`% of the country.

That is the answer to why anyone uses the five-year series. It is not a preference, and it is not caution. For nearly three-quarters of American counties it is the only thing published.

The floor is visible in the file. The smallest county with a one-year estimate holds **`r nn(FLOOR)` people** — just under the 65,000 the Bureau states as its threshold, because the threshold is not applied to the estimate you are reading. A stated rule and a delivered file rarely line up to the person.

## The window that was closed

There used to be three windows. Between 2007 and 2013 the Bureau also published a **three-year** series for places above 20,000. It was the middle ground between a thin single year and a five-year average reaching a long way into the past.

It was discontinued. You do not have to take that on faith, and this chapter does not assert it: it asks the server.

```{r vinttab}
```

The three-year directory answers for `r LAST3` and returns **404** for every year after it. The five-year directory answers for all of them. The discontinuation is a fact you can check with a URL, and the last three-year release was the `r LAST3` vintage.

This is worth more than a historical note. A script written in 2014 that pointed at the three-year series did not start returning wrong numbers — it started returning nothing, which is the good case. The bad case is the analyst who quietly substituted the five-year series and did not adjust the claim, because those two files answer different questions.

## The same county, two populations

For the `r nn(CM("Counties in both files"))` counties that appear in both files, you can put the two windows side by side. Same agency, same survey, same release, same county.

```{r cmptab}
```

**`r nn(CM("Counties where the two differ by more than 1%"))` of `r nn(CM("Counties in both files"))` differ by more than one per cent**, and `r nn(CM("Counties where the two differ by more than 5%"))` differ by more than five. The largest gap is `r p2(CM("Largest absolute difference, per cent"))`%.

Neither number is wrong. The one-year figure is 2023. The five-year figure is an average over 2019–2023, which is a different question with a different answer. The trouble is that both are labelled with the same year and quoted interchangeably.

Look at where the gaps are largest:

```{r dvgtab}
```

Every one of the eight is a fast-growing exurban county, and in **every one the five-year figure is lower**. The direction is not noise. A five-year average includes four years when the place was smaller, so where growth is fast the five-year window *systematically understates the present* — and where a place is shrinking it systematically overstates it.

Choosing a window chooses a finding. If you want to argue a county is booming, the one-year file helps you; if you want to argue it is not, the five-year file does. Both are official.

> **Before you read on, commit to a view.** Every ACS estimate is published with
> a margin of error beside it. **What fraction of the 3,222 county rows in the
> ACS table for total population do you think carries a usable margin** — and
> what would it mean if a Census Bureau table printed no margin at all?

## Whether a number has a margin depends on which table it came from

Here is the thing that decides how the rest of this chapter had to be written.

```{r ctrltab}
```

**`r nn(SENT)` of the `r nn(C5)` county rows for total population carry no margin of error**, or `r p1(SENTP)`% of them. What they carry instead is `-555555555`, which is not a margin of minus half a billion people. It is the Bureau's code for an estimate that was **controlled**: not estimated freely from the sample at all, but forced to agree with a figure from somewhere else.

That somewhere else is the Population Estimates Program, the subject of the [`census-pep`](../census-pep/census-pep-brief.html) chapter. So the ACS county population you look up is, in a real sense, not an ACS number.

Median household income is not controlled, and every one of its `r nn(ctrl$county_rows[2])` county rows carries a real margin. The lesson is not that one table is better. It is that **"does this number have a margin" is a question about the table, not about the ACS** — and the only way to know is to look. Arithmetic that treats the sentinel as a number produces nonsense quietly. How far the control goes is measured in the [access chapter](../census-access/census-access-brief.html): for a county's total population the one-year estimate is not merely controlled, it is the Population Estimates figure exactly.

## The margin is not decoration

So, using a table that actually carries margins:

```{r margtab}
```

In the largest counties the median margin on household income is **±`r p1(BIG)`%** — tight enough to compare places with. In counties under a thousand people it is **±`r p1(SMALL)`%**, which is wide enough that a county at the national median is statistically indistinguishable from one well above or below it.

And now put that beside the earlier finding. The places with the widest margins are the small ones — which are exactly the places that have no one-year estimate, and for which the five-year file is the only source. **The uncertainty is largest precisely where there is no alternative.**

## What arrives

```{r raw, comment="", class.output="rawblock"}
```

## What this chapter cannot tell you

**One table, one year, one geography.** Everything above is total population and median household income, 2023 vintage, at the county level. The ACS publishes thousands of tables down to the block group, and margins at block-group level are far wider than anything shown here. The shape of the finding does not depend on the table: margins widen as places shrink, and windows disagree where places change. The specific numbers do depend on it.

**The three-year probe is a check on the server, not on the archive.** A 404 means the Bureau does not publish that directory today. Data released between 2007 and 2013 still exists in other forms, and the chapter's claim is only that the series stopped, which is what the file system shows.

**Why an estimate was controlled is not visible here.** The file records *that* a number was controlled, not the reasoning or the source figure it was controlled to. That the target is the Population Estimates Program is documentation, not something recovered from this file.

**Nothing here measures whether the ACS is right.** Margins of error describe sampling uncertainty — how much the estimate would move if you drew a different sample. They say nothing about who did not answer, who was not asked, or whether the question meant the same thing to everyone. A narrow margin on a badly measured quantity is still narrow.

## Sources

- U.S. Census Bureau. *American Community Survey Summary File*, 2023, table-based
  format: table B01003 (total population), one-year and five-year; table B19013
  (median household income), five-year.
  <https://www2.census.gov/programs-surveys/acs/summary_file/2023/table-based-SF/>
- U.S. Census Bureau. *American Community Survey Summary File* archive, vintages
  2011–2015, used to establish when the three-year series stopped being
  published. <https://www2.census.gov/programs-surveys/acs/summary_file/>
- 13 U.S.C. § 221, which makes refusal to answer either instrument an offence,
  and § 141, which authorises the surveys the Bureau conducts between counts.
  The obligation those two sections create is what separates this instrument
  from every survey later in the book.
  <https://www.law.cornell.edu/uscode/text/13/221>
- Every published figure for one county's population, and the exact equality
  between a one-year total and a population estimate, are laid out in this
  book's [access chapter](../census-access/census-access-brief.html).
- ACS state-to-state migration flows, and what their margins do to a finding,
  are in the [migration chapter](../migration/migration-brief.html).

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`compare.csv`](data/derived/compare.csv), [`controlled.csv`](data/derived/controlled.csv), [`diverge.csv`](data/derived/diverge.csv), [`instrument.csv`](data/derived/instrument.csv), [`margins.csv`](data/derived/margins.csv), [`vintages.csv`](data/derived/vintages.csv), [`windows.csv`](data/derived/windows.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `margins.csv` groups counties by size and reports the median margin on median income. Read `median_margin_pct` down the column. At what size of county does the margin stop being a detail and start being the story?
- `diverge.csv` sets the one-year estimate against the five-year estimate for the same county. Sort by `difference_pct`. What kind of county disagrees with itself the most, and which of the two windows would you trust there?
- `controlled.csv` shows that most county rows for total population arrive with no margin at all. Read the `why` column. Is a number with no margin more certain, or just less honest about it?
- `vintages.csv` records which series were published in which year and what the server said. One window is closed. Work out what a researcher lost when it shut.

```{r ai-prompt, echo=FALSE, results="asis"}
```
