---
title: "Interstate Migration as a Chord Diagram"
type: brief
subtitle: "Everybody moved, almost nobody left: a year of state-to-state flows in one circle, and the grouping the circle demands"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("chord-code.R")
```

```{r setup, include=FALSE}
```

Americans tell the story of their own migration as an exodus. The Northeast empties into the South, California empties into Texas, and whole regions are winners or losers. The story matters to democracy because House seats and electoral votes really do follow the movers. But is the movement actually one-way? Or do the flows mostly run in both directions at once and cancel?

The record that can answer this is a survey. In the year the American Community Survey last measured it, `r n(BETWEEN)` people in the United States lived in a different state than they had lived in twelve months earlier. Another `r n(WITHIN_ST)` moved within their own state, and `r n(ABROAD)` arrived from abroad. The between-state moves are published as a matrix: `r NUNIT` origins by `r NUNIT` destinations, `r n(NPAIR)` ordered pairs. Each pair carries an estimate and a margin of error, which says how far off that estimate could be.

That matrix is the cleanest flow data in American demography, and it is almost impossible to look at. A **chord diagram** is the form built for exactly this shape of data. It is a circle whose rim is the set of places, with a ribbon joining each pair, wide where the flow is heavy. It will not take `r n(NPAIR)` ribbons. Getting the matrix small enough to draw is most of the work in this brief, and it is where the arguing should happen.

## The test

The flows are the Census Bureau's *State-to-State Migration Flows* table, from the `r YR` American Community Survey one-year estimates. The ACS reaches roughly 3.5 million addresses a year and asks each household where it lived one year ago; that one question is the country's entire measurement of moving. The [American Community Survey chapter](../census-acs/census-acs-brief.html) is the survey's biography: who is in it, how it is weighted up to the country, and why a margin of error stands beside every estimate it publishes.

The numbers here are the same tables the [migration brief](../migration/migration-brief.html) hands over, suppressed cells and all, so the two documents cannot silently disagree about a number. The test itself is the one the exodus story begs for. Group the states, add up the flow in each direction between every pair of groups, and measure how much is net and how much cancels.

## The data

One row of the flow table is an ordered pair of states. Direction is a column, not a sign: `r BA` to `r BB` and `r BB` to `r BA` are two different rows holding two different numbers, and this brief turns on the fact that both are large.

```{r one-row}
```

After grouping, the working table keeps six columns:

```{r varmap}
```

The states are grouped into the Census Bureau's own **nine divisions**, the clusters of neighbouring states — New England, the Middle Atlantic, the Pacific and so on — that the Bureau uses to summarise the country, with Puerto Rico carried as a tenth group.^[The regions and divisions are defined in chapter 6 of the Bureau's *Geographic Areas Reference Manual*, <https://www2.census.gov/geo/pdfs/reference/GARM/Ch6GARM.pdf>.] The Bureau's divisions cover only the fifty states and the District of Columbia, and `r n(PROUT)` people left Puerto Rico for a state that year. The list of which state belongs to which division is the one this book's [regional-shift brief](../regional-shift/regional-shift-brief.html) uses, checked there against the Bureau's published region totals for every decade since 1910.

```{r grouptab}
```

The grouping is a decision, not a fact about the country. **The Bureau fixed these divisions in 1910.** They put Delaware and Florida in one group, Maryland in the South, Missouri in the Midwest, and Texas with Arkansas and Oklahoma. Every sentence below about "the South" is a sentence about that list of states, and a different list would draw a different circle from identical data.

Two cleaning decisions travel with the grouping, and both push in a known direction. First, **`r n(NSUPP)` of the `r n(NPAIR)` state pairs are not published at all**: the Bureau releases no estimate when too few sampled households made a move. Aggregation adds them as zeros, because there is nothing else to add, so the group totals run slightly low. `matrix.csv` therefore carries a count of suppressed pairs in every cell, and `r n(DSUPP)` of the `r n(DCELL)` division-to-division cells contain at least one. A cell assembled mostly out of suppressed pairs cannot pass itself off as a measurement.

Second, **margins of error on the grouped cells are added in quadrature**, which is the Bureau's own instruction for a sum of published estimates:^[U.S. Census Bureau, *Understanding and Using American Community Survey Data: What All Data Users Need to Know* (2020), appendix 3, <https://www.census.gov/programs-surveys/acs/library/handbooks/general.html>.] square each margin, add the squares, and take the square root of the total. The largest exchange in the country shows the arithmetic. `r BA` to `r BB` carries a margin of ±`r n(MAB)`, and the return flow ±`r n(MBA)`. The net between them is the difference of the two estimates, and its margin is the square root of `r n(MAB)`² plus `r n(MBA)`², which comes to ±`r n(sqrt(MAB^2 + MBA^2))` — less than the two added outright, because two errors are unlikely to both fall the same way. That rule assumes the errors are independent, and cells from a single survey are not quite independent. So the margins below are approximate, and if they are wrong they are too small.

> **Before you read on, commit to a number.** Roughly `r mil(BETWEEN)` Americans
> changed state in a year. **What share of them do you think crossed a Census
> division boundary** — left the Northeast for the South, say — rather than
> moving to a neighbouring state inside the same division? Write down a
> percentage.

## What it says about democracy

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=6.4}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. A year of interstate migration, `r YR`.** A chord diagram fits because the data is a square grid of flows among the same ten groups, and the circle is the one form that shows every pair at once. Each rim segment is one division, sized by how many people left it. A ribbon's width where it meets the rim is how many people that division sent, so **the taper of a ribbon is the net flow**; ribbons take the colour of whichever division sent more. Click a rim segment to isolate it; the buttons switch to the four-region grouping and drop the within-division ribbons.

Now the prediction. **`r p1(WPCT)`% of interstate moves do not cross a division boundary.** Of `r n(WD + BD)` moves in the grid, `r n(WD)` were to another state in the same division: New York to New Jersey, Virginia to Maryland, Ohio to Michigan. They are the fat lens-shaped ribbons sitting on the rim in Figure 1. The button that hides them is worth pressing twice, because with them gone the circle looks busier and the country looks more mobile than it is.

That is the first thing the grouping does. Every one of those moves was a real change of state, in a matrix built to record changes of state, and grouping turns them into a shape that reads as *staying put*.

The second finding is what the taper of the ribbons is showing, and it is the answer to the exodus story.

```{r pairtab}
```

The largest exchange in the country is `r BA` with `r BB`: `r n(BAB)` people one way, `r n(BBA)` back, `r n(BG)` in total and a net of `r n(BN)`. That net is the exception that stands. It is `r p1(BNP)`% of the exchange, comfortably clear of its margin, and it is the *most* directional big exchange in the country. Even here, `r p1(100 - BNP)`% of the movement between the two divisions cancelled out.

The typical case cancels harder. **Across all `r n(DP)` pairs of divisions the median net is `r p1(MNP)`% of the gross flow**, and adding every pair together gives the same answer: `r p1(ONP)`% of all movement between divisions is net. The other `r p1(100 - ONP)`% is two people passing each other in opposite directions. `r CA` and `r CB` exchanged `r n(CG)` people for a net of `r n(abs(CN))`, against a margin of `r n(CM)`, which is statistically nothing. `r TA` and `r TB` exchanged `r n(TG)` people, a third of a million, for a net of `r n(TN)` with a margin of `r n(TM)`.

**Only `r n(DPS)` of the `r n(DP)` division pairs have a net larger than its own margin of error.** Group up to the four regions and it is `r n(RPS)` of `r n(RP)`. A net is a difference between two survey estimates and inherits the uncertainty of both, which is why a net of twenty thousand people can be indistinguishable from zero.

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same matrix, as a matrix.** Thousands of people moving from the row's division to the column's, shaded on a log scale, so each step of shading is a tenfold increase rather than a fixed number of people; the diagonal is moves that changed state without leaving the division. Hover any cell for the estimate, its margin of error, and how many suppressed state pairs are inside it. The button switches to net flow, and blanks every pair whose net is smaller than its own margin of error.

Press it. **Most of the matrix goes blank**, and what is left is a small number of genuinely directional exchanges sitting in a field of churn. That is the honest picture of American migration, and no chord diagram can draw it, because a ribbon has no way to say *this taper is not distinguishable from parallel*. The trade runs the other way too: the matrix cannot show at a glance that the Pacific division sends people almost everywhere while East South Central mostly trades with its neighbours. Figure 1 hands you that in a second and will not surrender a single value.

The regions hold one more exception worth telling. The Northeast-to-South net, `r n(NESN)` people out of `r n(NESG)` exchanged, runs the way the Sun Belt story says. But the South and the West exchanged `r n(SWG)` people, and their net of `r n(abs(SWN))` ran *toward the West*, which is not the direction the story predicts. The button on Figure 1 will show you both.

Why does any of this matter to democracy? Because House seats are apportioned by population, and population moves.

```{r nettab}
```

The `r n(LDIV)` divisions losing people on net hold `r n(LSEAT)` House seats between them; the `r n(WDIV)` gaining hold `r n(WSEAT)`. `r TLOSE` loses `r p1(abs(TLP))` people per thousand residents a year and `r TGAIN` gains `r p1(TGP)`, and a decade of that is what the [apportionment brief](../apportionment/apportionment-brief.html) is recomputing when the seat totals change. The rates are small, and that is the point worth carrying away. Nothing in Figure 1 looks like an exodus, yet a net flow of a few people per thousand per year, sustained for a decade, reorganises the House of Representatives. That is why this table is published annually and read cumulatively.

## What this brief cannot tell you

**Anything the grouping did not decide first.** Nine divisions is `r n(DRIB)` ribbons; fifty-two units would be `r n(NPAIR)` and unreadable. Every finding above would change under a different list: how much cancels, which pairs stand out, what counts as "within". The Census divisions are a 1910 committee decision, not a natural kind, and a grouping by metropolitan area or by cost of living would draw a different circle from identical numbers.

**A trend.** This is a single ACS release. The `r p1(WPCT)`% of moves that stay inside a division and the `r p1(MNP)`% median net are one year's estimates. The [migration brief](../migration/migration-brief.html) carries the long series, in which American mobility has been falling for decades.

**The suppressed cells.** The `r n(NSUPP)` missing state pairs were added as zeros, and they are the smallest flows between the smallest states, so the rim segments for the least populated divisions are the least trustworthy. They cannot be recovered from the published state totals either, because the ACS estimates its cells and its totals separately and they do not add up.

**Why anybody moved, or whether they moved back.** Somebody who left Pennsylvania for Florida and returned the next year appears twice, in opposite directions, and contributes nothing to any net. That is not an error in the data; it is what the data is. The word for it is churn, and it is `r p1(100 - MNP)`% of what the median pair of divisions did to each other.

## What you should have learned

- **`r p1(WPCT)`% of interstate moves never cross a division boundary**, so a
  circle drawn from nine groups reads as staying put even though every one of
  its `r n(WD + BD)` moves changed state. Aggregation is a decision with
  consequences, made before the first pixel is drawn.
- Migration is mostly churn: across the `r n(DP)` division pairs the median
  net is `r p1(MNP)`% of the gross flow, and only `r p1(ONP)`% of all
  between-division movement is net.
- Only `r n(DPS)` of the `r n(DP)` division nets, and `r n(RPS)` of the
  `r n(RP)` region nets, are larger than their own margins of error, because
  a net inherits the uncertainty of both estimates it is built from.
- The one-way exodus is the exception, not the rule: `r BA`–`r BB` is the most
  directional large exchange in the country, and even there
  `r p1(100 - BNP)`% of the movement cancelled.
- A chord diagram shows the shape of a whole flow grid at a glance and will
  not surrender a number; the matrix does the opposite. Neither replaces the
  other.

## Extensions

- `pairs.csv` has `a_to_b`, `b_to_a`, `gross`, `net` and `net_pct_of_gross`.
  Sort by the last one. Which pairs are a genuine one-way move, and which are
  two large flows that almost cancel?
- Count the TRUE rows in the `net_sig` column of `pairs.csv`. What is left of
  the story about Americans moving between regions once the nets that cannot
  be told apart from zero are set aside?
- `matrix.csv` records how many state pairs were `suppressed` in each cell.
  Does suppression fall evenly across the grid, or does it land on particular
  kinds of place?
- `states.csv` maps each state to both a `division` and a `region`. Regroup
  the flows by region and compare: does the picture change, and would you have
  known if you had only ever seen one grouping?
- **Stretch.** Download the previous year's release of the same Bureau table,
  from the address in the sources below with the year changed, and rebuild the
  division pairs. Do the same nets clear their margins, or does the set churn?
- **Stretch.** Invent a grouping of your own, by which party carried the
  state or by cost of living, and reaggregate the state flows in a
  spreadsheet using your list. Does the exodus story look stronger or weaker
  than under the Bureau's 1910 map?

## Sources

**Data**

- U.S. Census Bureau. *State-to-State Migration Flows*, American Community
  Survey `r YR` one-year estimates, table T13. The figures here read the
  tables this book's [migration brief](../migration/migration-brief.html)
  hands over, which describes the `r n(NSUPP)` suppressed pairs.
  <https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx>
- U.S. Census Bureau. *Geographic Areas Reference Manual*, chapter 6, on the
  four regions and nine divisions. The state-to-division assignment is read
  from this book's
  [regional-shift brief](../regional-shift/regional-shift-brief.html);
  `states.csv` writes the assignment out again so it can be checked state by
  state.
  <https://www2.census.gov/geo/pdfs/reference/GARM/Ch6GARM.pdf>
- U.S. Census Bureau. *2020 Census Apportionment Results*, as committed by
  this book's [apportionment brief](../apportionment/apportionment-brief.html).
  <https://www.census.gov/data/tables/2020/dec/2020-apportionment-data.html>
- U.S. Census Bureau (2020). *Understanding and Using American Community
  Survey Data: What All Data Users Need to Know*, Appendix 3, on aggregating
  estimates and their margins of error — the source of the quadrature rule
  used here.
  <https://www.census.gov/programs-surveys/acs/library/handbooks/general.html>

**Academic literature**

- Krzywinski, Martin et al. (2009). "Circos: an information aesthetic for
  comparative genomics." *Genome Research* 19(9): 1639–1645. The paper that
  made the chord diagram a standard form, and the source of the layout in
  Figure 1. <https://pmc.ncbi.nlm.nih.gov/articles/PMC2752132/>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`arcs.csv`](data/derived/arcs.csv), [`facts.csv`](data/derived/facts.csv), [`groups.csv`](data/derived/groups.csv), [`matrix.csv`](data/derived/matrix.csv), [`pairs.csv`](data/derived/pairs.csv), [`states.csv`](data/derived/states.csv)

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
