---
title: "Which Rung?"
subtitle: "84-355 Democracy's Data | Levels of aggregation in election data"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.2)
options(scipen = 999)

# Formatting helpers for the inline numbers in the prose below. Every figure in
# this lab is read out of the CSV that produced it, so nothing here can drift
# away from the data the chunks print.
n  <- function(x) format(round(x), big.mark = ",")
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
V  <- function(d, k, col = "value", key = names(d)[1]) d[[col]][d[[key]] == k]
num <- function(x) abs(as.numeric(gsub("[^0-9-]", "", x)))
LS <- function(k, col) lst[[col]][lst$level == k]
EE <- function(k, col) ee[[col]][ee$level == k]
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `levels-of-aggregation-brief.Rmd` first.** This lab runs the same
> numbered steps with the code exposed. The brief tells you what the question is
> and why it is hard; this shows you where each number came from — and in Step 6
> it *runs* the method that fails, rather than reporting that it failed.

***

# Which unit is your claim about?

> *"Turnout was up in the suburbs."*

Every sentence like that names a level of aggregation without saying so. The
suburbs are made of precincts, which are made of ballots, and they sit inside
counties, districts, states and a country. **Six of those seven levels are
available to you. One almost never is.**

This lab walks a single election up the ladder and shows what each step costs.

# Step 1 — The ladder

```{r ladder}
ld <- read.csv("data/ladder.csv", stringsAsFactors = FALSE)
ld[, c("rung", "level", "units_nationally", "units_in_georgia", "a_row_is")]
```

**`r n(ld$units_nationally[ld$level == "ballot"])` ballots.
`r n(ld$units_nationally[ld$level == "precinct"])` precincts.
`r n(ld$units_nationally[ld$level == "election jurisdiction"])` election
jurisdictions. `r n(ld$units_nationally[ld$level == "county"])` counties.
`r n(ld$units_nationally[ld$level == "congressional district"])` congressional
districts. `r n(ld$units_nationally[ld$level == "state"])` states. One nation.**

Read the `published_for` column too — it is the column that decides what you can
actually do.

```{r ladder-pub}
ld[, c("level", "published_for")]
```

# Step 2 — The bottom rung, which barely exists

Alaska runs ranked-choice elections. To run them it must publish a **cast vote
record**: one row per ballot, carrying every ranking on it.

```{r cvr}
cf <- read.csv("data/cvr_facts.csv", stringsAsFactors = FALSE)
cf
```

**`r n(V(cf, "ballots in the file"))` ballots.
`r n(V(cf, "distinct ranking sequences"))` distinct ranking sequences.**
`r pc(V(cf, "bullet-voting rate among ballots that ranked anyone (%)"))`% of
voters who ranked anyone ranked exactly one candidate and stopped, which is why
so few distinct sequences cover so many ballots.

Now the numbers that matter for this lab:

```{r second}
c2 <- read.csv("data/cvr_second.csv", stringsAsFactors = FALSE)
c2 <- c2[c2$second != "(no second choice)", ]
head(c2[order(-c2$ballots), ], 8)
```

**`r n(V(cf, "ranked Begich first and Peltola second"))` Alaskans ranked Begich
first and Peltola second — a Republican first and a Democrat second, on one
ballot.** `r n(V(cf, "ranked Peltola first and Begich second"))` did the reverse.

No total, at any level above this one, contains those numbers. They are
properties of a *pair* of marks on one piece of paper, and adding the paper up
destroys the pair.

# Step 3 — One rung up

```{r coverage}
BALLOTS <- ld$units_nationally[ld$level == "ballot"]
AK      <- cf$value[cf$quantity == "ballots in the file"]
c(alaska_cvr_ballots = AK, all_2024_ballots = BALLOTS,
  pct_of_national = round(100 * AK / BALLOTS, 3))
```

**The whole Alaska file is `r pc(100 * AK / BALLOTS, 3)`% of the ballots cast in
2024.** There is no
national cast-vote-record file. If you want to say something about the country,
you must leave the bottom rung.

# Steps 4 and 5 — Climb, and watch what holds

One election, the 2024 presidential vote in Georgia, read at four levels:

```{r levels}
lst <- read.csv("data/level_stats.csv", stringsAsFactors = FALSE)
lst
```

**The vote-weighted mean does not move — `r pc(LS("precinct", "mean_dem_share"), 2)`%
across precincts, `r pc(LS("county", "mean_dem_share"), 2)`% across counties,
`r pc(LS("state", "mean_dem_share"), 2)`% for the state.** The district row reads
`r pc(LS("congressional district", "mean_dem_share"), 2)` because the district
file carries percentages and no vote counts, so that one mean is unweighted. The
spread collapses: the standard deviation of Democratic share falls from
`r pc(LS("precinct", "sd_dem_share"), 2)` points across
`r n(LS("precinct", "units"))` precincts to `r pc(LS("county", "sd_dem_share"), 2)`
across `r n(LS("county", "units"))` counties. `variance_retained_pct` says that
**`r pc(LS("county", "variance_retained_pct"))`% of the precinct-level variation
survives the trip to counties**, so
`r pc(100 - LS("county", "variance_retained_pct"))`% of it does not.

Does the arithmetic hold?

```{r consistency}
cs <- read.csv("data/consistency.csv", stringsAsFactors = FALSE)
cs[, c("comparison", "from", "against", "difference")]
```

Summing all `r n(ld$units_in_georgia[ld$level == "precinct"])` precincts
reproduces the county file **exactly**. Against a different compiler the same
Georgia counties are `r num(cs$difference[2])` votes off, and the whole country's
counties land `r n(num(cs$difference[4]))` votes short of the certified national
total — `r pc(abs(cs$pct_difference[4]), 4)`% low.

Part of the reason is that a rung is not a rung everywhere:

```{r notcounties}
nc <- read.csv("data/not_counties.csv", stringsAsFactors = FALSE)
nc
```

**Alaska has no counties.** The national county file carries
`r nc$rows_that_are_not_counties[nc$state == "Alaska"]` state house districts in
their place, and DC contributes
`r nc$rows_that_are_not_counties[nc$state == "District of Columbia"]` wards.

And one rung cannot be reached by adding at all — precinct returns record no
congressional district. But it is the only rung that can answer this:

```{r seatvote}
sv <- read.csv("data/seat_vote.csv", stringsAsFactors = FALSE)
sv
```

**Georgia's median congressional district is
`r pc(V(sv, "Georgia: median district share", key = "measure"))`% Democratic
while the state is
`r pc(V(sv, "Georgia two-party Democratic share, from the precincts", key = "measure"))`%.**
Trump won `r pc(V(sv, "Trump's share of the national popular vote (%)", key = "measure"))`%
of the national popular vote and
`r pc(V(sv, "Trump's share of the electoral college (%)", key = "measure"))`% of
the electoral college. Neither fact is visible from any other level.

> **Predict before you run Step 6.** Georgia publishes its precinct returns
> broken out by vote method. Step 6 deletes that breakdown, keeps only each
> precinct's two margins, and runs the standard method used in voting-rights
> litigation to recover what share of mail voters chose Biden. **Write down your
> guess — and write down the largest value the answer could possibly take.**

# Step 6 — Run the method that fails

First the truth, which Georgia published:

```{r truth}
mt <- read.csv("data/method_truth.csv", stringsAsFactors = FALSE)
mt
TRUE_MAIL <- 100 * mt$by_mail[mt$candidate == "Joseph R. Biden"] / sum(mt$by_mail)
TRUE_NOT  <- 100 * mt$not_by_mail[mt$candidate == "Joseph R. Biden"] /
                   sum(mt$not_by_mail)
round(c(dem_share_of_mail_votes = TRUE_MAIL,
        dem_share_of_everything_else = TRUE_NOT), 2)
```

Now throw the crosstab away. Keep only what every other state publishes — one
row per precinct, two margins — and fit Goodman's ecological regression:

```{r goodman}
ep <- read.csv("data/eco_precinct.csv", stringsAsFactors = FALSE)
head(ep[, c("unit", "votes", "mail_pct", "dem_pct")], 4)

fit <- lm(dem_pct ~ mail_pct, data = ep, weights = ep$votes)
b <- coef(fit)
round(c(estimate_at_0_pct_mail = b[[1]],
        estimate_at_100_pct_mail = b[[1]] + 100 * b[[2]]), 1)
```

**`r pc(b[[1]] + 100 * b[[2]])`%.**

The regression, run on real precinct returns with a respectable method, says
that `r pc(b[[1]] + 100 * b[[2]])`% of Georgia's mail voters chose Biden. The
truth is `r pc(TRUE_MAIL)`%. Nothing in
the aggregate data announces the failure — the fit is fine and the line runs
through the middle of the cloud.

```{r eco-d3, echo=FALSE, results='asis', eval=knitr::is_html_output()}
set.seed(3)
z <- ep[sample(nrow(ep), 1200), ]
pts <- paste(sprintf('[%.1f,%.1f,%.2f]', z$mail_pct, z$dem_pct,
                     sqrt(z$votes / max(ep$votes))), collapse = ",")
cat(sprintf('
<div id="eco" style="margin:1.1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const P=[%s],G0=%.2f,G1=%.2f,T0=%.2f,T1=%.2f;
const W=760,H=380,L=52,R=180,TOP=32,B=42;
const svg=d3.select("#eco").append("svg").attr("viewBox",`0 0 ${W} ${H}`)
  .attr("style","max-width:100%%;height:auto;font:12px inherit");
const x=d3.scaleLinear().domain([0,100]).range([L,W-R]);
const y=d3.scaleLinear().domain([-30,210]).range([H-B,TOP]);
const T=(a,b,s,o)=>{const t=svg.append("text").attr("x",a).attr("y",b)
  .attr("text-anchor",(o&&o.a)||"start").attr("font-size",(o&&o.s)||"11px")
  .attr("fill",(o&&o.c)||"#444");if(o&&o.b)t.attr("font-weight","600");return t.text(s);};
svg.append("rect").attr("x",L).attr("y",y(100)).attr("width",W-R-L)
  .attr("height",y(0)-y(100)).attr("fill","#f4f4f4");
T(L+6,y(100)-5,"no share can be above this line",{s:"10px",c:"#999"});
[-25,0,25,50,75,100,125,150,175,200].forEach(v=>{
  svg.append("line").attr("x1",L).attr("x2",W-R).attr("y1",y(v)).attr("y2",y(v))
    .attr("stroke",v===0||v===100?"#bbb":"#eee");
  T(L-7,y(v)+3.5,v,{a:"end",s:"9.5px",c:"#aaa"});});
[0,25,50,75,100].forEach(v=>T(x(v),H-B+15,v,{a:"middle",s:"9.5px",c:"#aaa"}));
svg.append("g").selectAll("circle").data(P).join("circle")
  .attr("cx",d=>x(d[0])).attr("cy",d=>y(d[1])).attr("r",d=>1+3.2*d[2])
  .attr("fill","#2c7fb8").attr("fill-opacity",0.22);
const ln=(a,b,c,dash)=>svg.append("line").attr("x1",x(0)).attr("y1",y(a))
  .attr("x2",x(100)).attr("y2",y(b)).attr("stroke",c).attr("stroke-width",2.6)
  .attr("stroke-dasharray",dash||null);
ln(T0,T1,"#4d9221"); ln(G0,G1,"#C41230","7,4");
[[T1,"#4d9221","the truth: "+T0.toFixed(1)+"%% to "+T1.toFixed(1)+"%%"],
 [G1,"#C41230","Goodman: "+G0.toFixed(1)+"%% to "+G1.toFixed(1)+"%%"]]
 .forEach(d=>{svg.append("circle").attr("cx",x(100)).attr("cy",y(d[0])).attr("r",4.5)
   .attr("fill",d[1]);T(x(100)+9,y(d[0])+4,d[2],{c:d[1],b:1,s:"11.5px"});});
svg.append("circle").attr("cx",x(0)).attr("cy",y(T0)).attr("r",4.5).attr("fill","#4d9221");
svg.append("circle").attr("cx",x(0)).attr("cy",y(G0)).attr("r",4.5).attr("fill","#C41230");
T(8,18,"%s Georgia precincts, 2020",{b:1,s:"12.5px",c:"#222"});
T(L,H-8,"share of the precinct\\u2019s votes cast by mail (%%)",{s:"10.5px",c:"#888"});
})();
</script>
', pts, b[[1]], b[[1]] + 100 * b[[2]], TRUE_NOT, TRUE_MAIL, n(nrow(ep))))
```

```{r eco-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.6}
par(mar = c(3.4, 3.6, 2.0, 9.4))
plot(NA, xlim = c(0, 100), ylim = c(-30, 210), axes = FALSE, ann = FALSE)
rect(0, 0, 100, 100, col = "#f4f4f4", border = NA)
abline(h = c(0, 100), col = "#bbb")
axis(1, seq(0, 100, 25), cex.axis = 0.62, col = "#bbb", col.axis = "#888",
     tck = -0.012, mgp = c(2, 0.3, 0))
axis(2, seq(-25, 200, 25), las = 1, cex.axis = 0.6, col = "#bbb",
     col.axis = "#888", tck = -0.012, mgp = c(2, 0.45, 0))
set.seed(3); z <- ep[sample(nrow(ep), 1200), ]
points(z$mail_pct, z$dem_pct, pch = 19,
       cex = 0.18 + 1.0 * sqrt(z$votes / max(ep$votes)),
       col = adjustcolor("#2c7fb8", 0.22))
segments(0, TRUE_NOT, 100, TRUE_MAIL, col = "#4d9221", lwd = 2.4)
segments(0, b[[1]], 100, b[[1]] + 100 * b[[2]], col = "#C41230", lwd = 2.4, lty = 2)
points(c(0, 100, 0, 100), c(TRUE_NOT, TRUE_MAIL, b[[1]], b[[1]] + 100 * b[[2]]),
       pch = 19, cex = 1.1, col = c("#4d9221", "#4d9221", "#C41230", "#C41230"))
text(102, b[[1]] + 100 * b[[2]], sprintf("Goodman: %.1f%% to %.1f%%",
     b[[1]], b[[1]] + 100 * b[[2]]), adj = 0, cex = 0.62, font = 2,
     col = "#C41230", xpd = NA)
text(102, TRUE_MAIL, sprintf("the truth: %.1f%% to %.1f%%", TRUE_NOT, TRUE_MAIL),
     adj = 0, cex = 0.62, font = 2, col = "#4d9221", xpd = NA)
text(3, 106, "no share can be above this line", adj = 0, cex = 0.55, col = "#999")
title(sprintf("%s Georgia precincts, 2020", n(nrow(ep))),
      cex.main = 0.84, adj = 0, line = 0.6)
mtext("share of the precinct's votes cast by mail (%)", side = 1, cex = 0.58,
      col = "#888", line = 1.7)
mtext("Democratic share of the precinct's votes (%)", side = 2, cex = 0.58,
      col = "#888", line = 2.3)
```

Both lines pass through the same cloud. **They differ only in slope, and the
slope is the whole answer.** Precincts that mailed heavily were Democratic
precincts, and Democratic precincts were more Democratic among mail voters *and*
election-day voters alike. The regression charges that entire between-precinct
difference to the method of voting.

# Step 7 — Do it again one rung higher

```{r county}
ec <- read.csv("data/eco_county.csv", stringsAsFactors = FALSE)
b2 <- coef(lm(dem_pct ~ mail_pct, data = ec, weights = ec$votes))
round(c(precinct_level = b[[1]] + 100 * b[[2]],
        county_level   = b2[[1]] + 100 * b2[[2]],
        the_truth      = TRUE_MAIL), 1)
```

**Coarser units, worse answer: `r pc(b2[[1]] + 100 * b2[[2]])`%.** The full
comparison, with Duncan and Davis's assumption-free bounds alongside:

```{r eco-table}
ee <- read.csv("data/eco_estimates.csv", stringsAsFactors = FALSE)
ee
```

The bounds are the honest thing the aggregate *can* support. They are also nearly
useless — `r pc(EE("precinct", "bound_lower"))`% to
`r pc(EE("precinct", "bound_upper"))`% — and they get wider, not narrower, one
rung up: `r pc(EE("county", "bound_lower"))`% to
`r pc(EE("county", "bound_upper"))`%.

Is the county correlation a fact about voters or about county lines? Regroup the
same precincts into the same number of *random* groups, 500 times:

```{r maup}
mp <- read.csv("data/maup_null.csv", stringsAsFactors = FALSE)
mp
```

**Random groups leave the correlation where it was
(`r pc(mp$correlation[3], 3)` against `r pc(mp$correlation[1], 3)` ungrouped).
Real counties push it to `r pc(mp$correlation[2], 3)`, which none of the 500
draws reached.** The lift comes from these
particular lines — Openshaw's modifiable areal unit problem, which `areal-units`
takes on by itself.

# Step 8 — The same failure, where it counts

```{r rpv}
rp <- read.csv("data/rpv_check.csv", stringsAsFactors = FALSE)
rp
```

`rpv` runs this on the question ecological inference is actually used for:
**Goodman says
`r pc(V(rp, "Goodman's ecological regression on the same precincts (%)"))`% of
Black voters took a Democratic ballot; the voter file says
`r pc(V(rp, "the truth: Democratic share of Black voters (%)"))`%.** Same method,
same impossibility, smaller margin. `vote-dilution`
takes the same county on into the first *Gingles* precondition.

# Steps 9 and 10 — What the aggregate is for, and how to choose

```{r decision}
dc <- read.csv("data/decision.csv", stringsAsFactors = FALSE)
dc[, c("question", "right_rung")]
```

```{r why}
dc[, c("right_rung", "why", "the_trap")]
```

# Step 11 — What you can now say

**You can say** that election data comes on a ladder, that the arithmetic of
climbing is exact, and that each step destroys something specific —
`r pc(100 - LS("county", "variance_retained_pct"))`% of the variation between
precincts, and every pair of marks that sat on one ballot.

**You can say it sharper than "aggregate data is risky,"** because you have a
case with an answer key: `r pc(b[[1]] + 100 * b[[2]])`% against a published
`r pc(TRUE_MAIL)`%.

**You cannot say** which rung is correct in general. You can name the rung, say
why, and check whether the answer survives being asked one rung up.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Break the regression differently.** `eco_precinct.csv` has a `votes`
column. Re-run Goodman's regression *unweighted*, then weighted, then on the
largest 500 precincts only, then on the smallest 500. Report the four estimates
of the Democratic share of mail voters. Which one is closest to
`r pc(TRUE_MAIL)`%, and does
that tell you anything you could have known without the answer key?

**Option B — Find the level where the answer flips.** Aggregate `eco_precinct.csv`
into groups of increasing size — 500 groups, 159, 50, 10 — and plot the Goodman
estimate against the number of units. Where does it cross 100%? Say what that
curve means for anyone choosing a unit of analysis.

**Option C — Pick a question, defend a rung.** Choose a research question that is
*not* in `decision.csv`. Say which rung you would use, what you would lose from
the rung below, and what you would risk at the rung above. Then find the number
in this lab that best supports your choice.

**Option D — Your own question.**
`r length(list.files("data", pattern = "[.]csv$"))` files,
`r nrow(ld)` levels, three elections.

# What this lab cannot tell you

- **How often Goodman fails this badly.** One state, one contest, one variable.
  We know it *can* be `r pc(EE("precinct", "error_by_mail"))` points wrong and
  give no warning; we do not know the distribution.
- **What a cast vote record would show for an ordinary election.** Alaska's
  exists because the count is ranked. Nothing here says what one would reveal
  about a plurality race.
- **Whether King's method fixes it.** Not run here. `rpv` discusses it;
  neither lab runs it.
- **Anything about individual voters.** Every file here except the Alaska
  sequences is already an aggregate, and the Alaska file was aggregated to
  ranking sequences before it was committed.
- **Where the units are.** No boundaries, no maps. `precinct-geography` and
  `census-geography` are the labs about the shapes.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- The secret ballot destroys individual-level data on purpose. Should cast vote
  records be public anyway?
- Goodman's regression has been used as evidence in federal voting-rights cases
  since the 1980s, and here it returns 194%. What should a court do with that?
- Name a claim you have read in the news this term and say which rung it was
  made at. Was it the right one?

***

### Where the data came from

Every input is a CSV already committed to another lab in this course; nothing was
downloaded for this one. `data/build-data.R` re-derives every number above.

- **Alaska cast vote record, 2024 US House** — `cast-vote-records/data/sequences.csv`,
  `r n(V(cf, "ballots in the file"))` ballots collapsed to
  `r n(V(cf, "distinct ranking sequences"))` ranking sequences by that lab's build.
- **Georgia precinct returns, 2020 and 2024, with the vote-method breakdown** —
  `ga-precinct-returns/data/`, from the Secretary of State.
- **County returns for every state, 2024** — `data-sources/data/`.
- **Presidential results by congressional district, 2024** —
  `redistricting/data/`.
- **National and state results, 1864–2024** — `historical-campaigns/data/`.
- **Unit counts** (`r n(ld$units_nationally[ld$level == "precinct"])` precincts,
  `r n(ld$units_nationally[ld$level == "election jurisdiction"])` jurisdictions,
  `r n(ld$units_nationally[ld$level == "ballot"])` voters) —
  `eavs/data/`, the EAC's Election Administration and Voting Survey.
- **Houston County, Georgia primary with reported race** — `rpv/data/`.
