---
title: "Seventy-Five Seats"
subtitle: "84-355 Democracy's Data | Population and House seats by region, 1910–2020"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.0,
                      dpi = 120, fig.retina = 1)
options(scipen = 999)

# Render every data.frame in this document as a table rather than as console
# output. The `envir` argument is required: without it the method registers in
# the wrong namespace and knitr silently ignores it.
knit_print.data.frame <- function(x, ...) {
  n <- names(x); n <- gsub("_", " ", n)
  n <- sub("^(.)", "\\U\\1", n, perl = TRUE)
  knitr::knit_print(knitr::kable(x, col.names = n, row.names = FALSE), ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

Everything here runs as given. Your job is to read what comes out and say what
it means. Base R only, no packages.

> **Read `regional-shift-brief.Rmd` first.** This lab runs the same numbered
> steps with the code exposed. The brief tells you what the question is; this
> shows you where each number came from.

> **And read `apportionment` first if you have not.** That lab takes a
> *single* reapportionment apart — Huntington–Hill, and New York missing the
> 435th seat by 89 people. **This lab is about sixty years of them.** The
> arithmetic is assumed here, not repeated.

***

# Where did that number come from?

> *"Texas gained two seats and New York lost one."*

That is a real headline from April 2021, and it is accurate. It is also the
smallest true thing you can say about what happened.

The United States redistributes political power every ten years without anybody
voting on it. People move for jobs and weather; the census counts them; 435 seats
and 538 electoral votes follow. **The question in this lab is what that adds up
to when you stop looking one decade at a time.**

# Steps 1–3 — One record, the file, and why we start at 1960

```{r load}
st  <- read.csv("data/states.csv",     stringsAsFactors = FALSE)
reg <- read.csv("data/regions.csv",    stringsAsFactors = FALSE)
nat <- read.csv("data/nation.csv",     stringsAsFactors = FALSE)
sc  <- read.csv("data/seatchange.csv", stringsAsFactors = FALSE)
sd  <- read.csv("data/southdefs.csv",  stringsAsFactors = FALSE)

c(state_years = nrow(st), states = length(unique(st$name)),
  first = min(st$year), last = max(st$year), censuses = length(unique(st$year)))
```

One row per state per census. Here is one state, twice:

```{r one-record}
st[st$name == "Florida" & st$year %in% c(1960, 2020),
   c("name", "year", "pop", "reps", "repchg", "region")]
```

**Three things the build script had to survive**, all documented in
`data/build-data.R`:

1. Every number in the source file is a **quoted string with commas** —
   `"4,951,560"`. `as.numeric()` returns `NA` for all of them, silently.
2. **Puerto Rico is a `State` row and is not in the national total.** The
   **District of Columbia** is in the national total and has no representative.
3. Alaska and Hawaii are in the file from 1910 with **no representatives**:

```{r ak-hi}
st[st$name %in% c("Alaska", "Hawaii") & st$year %in% c(1950, 1960),
   c("name", "year", "pop", "reps")]
```

They were territories. **The fifty-state House begins in 1960**, which is why
every seat figure below runs 1960–2020. Population runs back to 1910 and is
labelled as such.

One more thing, and it is unusual enough to be worth a line. The Census Bureau
publishes its **own** region totals in the same file, so the state-to-region
mapping used here is checkable rather than assumed:

```{r check}
mine <- aggregate(pop ~ region + year, st, sum)
c(region_years_checked = nrow(mine),
  national_totals_match = all(tapply(st$pop, st$year, sum) ==
                              nat$pop[order(nat$year)]))
```

`build-data.R` reruns that comparison and **stops** if any region-year
disagrees. That is the only real way to know a lookup table is right.

# Step 4 — One decade at a time

```{r decade-2020}
d20 <- st[st$year == 2020 & !is.na(st$repchg) & st$repchg != 0,
          c("name", "region", "reps", "repchg")]
d20

c(seats_that_moved = sum(d20$repchg[d20$repchg > 0]),
  states_affected  = nrow(d20),
  largest_single_move = max(abs(d20$repchg)))
```

Seven seats, thirteen states, none moving more than two. And that is typical:

```{r decades}
dec <- st[st$year > 1960 & st$name != "District of Columbia" & !is.na(st$repchg), ]
moved_by_decade <- tapply(dec$repchg, dec$year, function(x) sum(x[x > 0]))

rbind(seats_changing_states = moved_by_decade,
      states_affected       = tapply(dec$repchg, dec$year, function(x) sum(x != 0)),
      pct_of_the_house      = round(100 * moved_by_decade / 435, 1))
```

Under five percent of the House even in the busiest decade. Taken alone, each
reapportionment reads as maintenance.

> **Stop here and write down two numbers before you run the next chunk.**
>
> New York's worst single decade in this period was **5 seats**, and most decades
> it lost one or two. **How many seats do you think New York lost in total
> between 1960 and 2020?**
>
> And, adding up the top row above: **how many seats do you think have changed
> states altogether?**

# Step 5 — The pivot: sixty years at once

```{r pivot}
c(seats_that_changed_states = sum(sc$change[sc$change > 0]),
  pct_of_the_house = round(100 * sum(sc$change[sc$change > 0]) / 435, 1),
  states_that_gained = sum(sc$change > 0),
  states_that_lost   = sum(sc$change < 0),
  states_unchanged   = sum(sc$change == 0))

sc[order(-abs(sc$change)), c("name", "region", "reps_1960", "reps_2020", "change")][1:10, ]
```

**Seventy-five seats.** Seventeen percent of the House of Representatives sits in
a different state than it did in 1960, and no single decade ever announced it.

New York lost **15**. Put that next to the size of everybody else's delegation:

```{r ny}
ny <- sc[sc$name == "New York", ]
c(new_york_1960 = ny$reps_1960, new_york_2020 = ny$reps_2020,
  lost = abs(ny$change),
  states_smaller_than_that_loss = sum(sc$reps_2020 <= abs(ny$change)))
```

**44 of the 50 states do not have 15 seats in total.**

```{r bars, fig.height=5.2}
m <- sc[sc$change != 0, ]; m <- m[order(m$change), ]
cols <- c(Northeast = "#2c7fb8", Midwest = "#4d9221",
          South = "#C41230", West = "#e08214")

par(mar = c(4.2, 8.4, 1, 2))
barplot(m$change, horiz = TRUE, las = 1, names.arg = m$name,
        col = cols[m$region], border = NA, cex.names = 0.7, xlim = c(-19, 19),
        xlab = "seats gained or lost, 1960 to 2020")
abline(v = 0, col = "#666")
legend("bottomright", names(cols), fill = cols, border = NA, bty = "n", cex = 0.8)
```

# Step 6 — Regions, and whether the movement ever reverses

```{r regions}
blk <- ifelse(sc$region %in% c("Northeast", "Midwest"), "NE+MW", "S+W")

rbind(seats_1960 = tapply(sc$reps_1960, sc$region, sum),
      seats_2020 = tapply(sc$reps_2020, sc$region, sum),
      change     = tapply(sc$change,    sc$region, sum))

rbind(seats_1960 = tapply(sc$reps_1960, blk, sum),
      seats_2020 = tapply(sc$reps_2020, blk, sum),
      pct_2020   = round(100 * tapply(sc$reps_2020, blk, sum) / 435, 1))

# of the seats that changed states, how many left their region altogether?
netreg <- tapply(sc$change, sc$region, sum)
c(seats_that_changed_states = sum(sc$change[sc$change > 0]),
  crossed_a_regional_border = sum(netreg[netreg > 0]),
  moved_within_one_region   = sum(sc$change[sc$change > 0]) - sum(netreg[netreg > 0]))
```

**In 1960 the Northeast and Midwest together held a majority of the House** —
233 of 435 seats. They now hold 167, or 38.4%. And 66 of the 75 seats that moved
at all left their region entirely — this is not states trading with neighbours.

Now: is this a trend or a fluctuation? Compare the sum of the per-decade columns
in Step 4 with the cumulative total in Step 5, then check each state's path.

```{r ratchet}
z  <- st[st$year >= 1960 & st$name != "District of Columbia", ]
sp <- split(z[order(z$year), ], z$name[order(z$year)])
mono <- sapply(sp, function(x) { d <- diff(x$reps[order(x$year)])
                                 all(d >= 0) || all(d <= 0) })

c(seat_gains_summed_by_decade = sum(dec$repchg[dec$repchg > 0]),
  net_gains_1960_to_2020      = sum(sc$change[sc$change > 0]),
  states_that_only_went_one_way = sum(mono))
names(mono)[!mono]
```

**47 of 50 states moved in one direction only.** This is not a cycle. It is a
ratchet.

# Step 7 — The long line

Seats are lumpy; a state has to gain a whole one. Population is not.

```{r shares}
wide <- reshape(reg[, c("region", "year", "share")], idvar = "year",
                timevar = "region", direction = "wide")
round(wide[order(wide$year), ], 1)
```

```{r lines, fig.height=4.0}
par(mar = c(4, 4.4, 1, 6.6))
plot(NA, xlim = c(1910, 2020), ylim = c(5, 42), las = 1, xaxt = "n",
     xlab = "", ylab = "% of U.S. resident population")
axis(1, at = seq(1910, 2020, 10), cex.axis = 0.85)
abline(v = 1960, col = "#999", lty = 3)
text(1960, 41.5, " 50 states from here", col = "#666", cex = 0.75, pos = 4)
ltys <- c(Northeast = 1, Midwest = 2, South = 1, West = 4)
for (r in names(cols)) {
  q <- reg[reg$region == r, ]; q <- q[order(q$year), ]
  lines(q$year, q$share, col = cols[r], lwd = 2.4, lty = ltys[r])
  text(2021, q$share[q$year == 2020], paste0(" ", r), col = cols[r],
       cex = 0.8, pos = 4, xpd = NA)
}
```

Four things in that picture.

- **The Northeast falls the whole way**, 28.0% to 17.4%, with no reversal.
- **The West passes the Northeast in 1990** and the Midwest in 2010.
- **The South passes the Midwest in 1940** and never gives the lead back.
- **And the South's rise starts after 1960, not before.** Between 1910 and 1960
  the Census South's share of the country *fell*.

# Step 8 — Does the finding depend on what "the South" means?

`house-competition` and `historical-campaigns` define the South as the
eleven former Confederate states. The Census Bureau's South is sixteen states
plus DC, adding Maryland, Delaware, Oklahoma, West Virginia and Kentucky.

```{r southdefs}
o <- sd[sd$year %in% c(1960, 2020), ]
w <- reshape(o[, c("definition", "year", "share")], idvar = "definition",
             timevar = "year", direction = "wide")
w$change <- round(w$share.2020 - w$share.1960, 1)
w$share.1960 <- round(w$share.1960, 1); w$share.2020 <- round(w$share.2020, 1)
w
```

**The narrower definition gives the larger rise**, which is only possible if the
extra states are a drag — and they are. The border states went from 6.4% of the
country to 5.5%. **So the finding is not an artefact of the definition, and the
broad version understates it.**

# Step 9 — What a seat is worth now

The House has been fixed at 435 by statute since 1929. The country was not.

```{r perseat}
nat[, c("year", "pop", "reps", "pop_per_rep")]

c(per_rep_1910 = nat$pop_per_rep[nat$year == 1910],
  per_rep_2020 = nat$pop_per_rep[nat$year == 2020],
  multiple     = round(nat$pop_per_rep[nat$year == 2020] /
                       nat$pop_per_rep[nat$year == 1910], 2),
  population_multiple = round(nat$pop[nat$year == 2020] /
                              nat$pop[nat$year == 1910], 2))
```

```{r step-chart, fig.height=3.4}
par(mar = c(4, 5, 1, 2))
plot(nat$year, nat$pop_per_rep / 1000, type = "s", lwd = 2.6, col = "#8856a7",
     xlim = c(1910, 2024), ylim = c(0, 830), las = 1, xaxt = "n", xlab = "",
     ylab = "thousands of people per representative")
axis(1, at = seq(1910, 2020, 10), cex.axis = 0.85)
points(nat$year, nat$pop_per_rep / 1000, pch = 19, cex = 0.8, col = "#8856a7")
abline(h = nat$pop_per_rep[nat$year == 1910] / 1000, col = "#999", lty = 2)
text(1965, nat$pop_per_rep[nat$year == 1910] / 1000 + 30, "the 1910 district",
     col = "#666", cex = 0.8)
```

**A member of the House now speaks for 3.62 times as many people as in 1910**,
and the country grew 3.59 times. Those two multiples being nearly identical *is*
the mechanism: the denominator was frozen in 1929.

But bigger districts and *less equal* districts are different complaints:

```{r spread}
ppr <- st[st$year >= 1960 & st$name != "District of Columbia" & !is.na(st$reps), ]
ppr$per <- ppr$pop / ppr$reps
do.call(rbind, lapply(split(ppr, ppr$year), function(x)
  data.frame(year = x$year[1],
             smallest = x$name[which.min(x$per)],
             largest  = x$name[which.max(x$per)],
             ratio    = round(max(x$per) / min(x$per), 2))))
```

**Interstate inequality got slightly better while district size nearly doubled.**

# Step 10 — Electoral votes moved, and nobody voted

A state's electors equal its House seats plus two.

```{r ec}
ev60 <- tapply(sc$reps_1960, blk, sum) + 2 * table(blk)
ev20 <- tapply(sc$reps_2020, blk, sum) + 2 * table(blk)
rbind(electoral_votes_1960 = ev60, electoral_votes_2020 = ev20,
      pct_1960 = round(100 * ev60 / sum(ev60), 1),
      pct_2020 = round(100 * ev20 / sum(ev20), 1))
```

**In 1960 the Northeast and Midwest controlled a majority of the electoral
votes** — 275 of the 535 cast by the fifty states. By 2020 they controlled 39.1%.

DC is excluded from both columns: it had none in 1960 and got three by
constitutional amendment in 1961, which is a rule change rather than a
population change. Including them lowers both shares slightly and changes
nothing.

# Step 11 — What you can now say

**You can say** that 75 House seats changed states between 1960 and 2020, that
66 of them crossed a regional boundary, and that the Northeast and Midwest went
from a majority of the House to 38.4% of it — with nothing enacted, proposed or
voted on.

**You can say it more sharply**: no single decade did this. The largest
reapportionment moved 19 seats and looked like housekeeping at the time.

**You cannot say why people moved.** This file has population and seats and
nothing else. **The `labs/migration` lab measures the moving; this one shows what
the moving does.**

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

***

### Option A — Move the window

```{r opt-a}
START <- 1960          # <- change this. Try 1970, 1980, 1990, 2000.
END   <- 2020

a <- st[st$year == START & st$name != "District of Columbia", c("name", "region", "reps")]
b <- st[st$year == END   & st$name != "District of Columbia", c("name", "reps")]
w <- merge(a, b, by = "name", suffixes = c("_start", "_end"))
w$change <- w$reps_end - w$reps_start

c(window = paste(START, "to", END), decades = (END - START) / 10,
  seats_that_changed_states = sum(w$change[w$change > 0]))
tapply(w$change, w$region, sum)
head(w[order(-abs(w$change)), ], 6)
```

**Write about:** does the total scale with the length of the window, or faster or
slower? Compare 1960–2020 with 1990–2020 — a window half as long. If the number
is more than half, the shift is accelerating; if less, it is slowing. Say which,
and then say honestly how confident three or six observations let you be. Does
the *ranking* of gainers and losers change, or only the size?

***

### Option B — Redefine the South

```{r opt-b, fig.height=3.4}
MY_SOUTH <- c("Alabama", "Arkansas", "Florida", "Georgia", "Louisiana",
              "Mississippi", "North Carolina", "South Carolina", "Tennessee",
              "Texas", "Virginia")     # <- add or remove states here

sel <- st[st$name %in% MY_SOUTH, ]
tot <- tapply(st$pop, st$year, sum)
share <- 100 * tapply(sel$pop, sel$year, sum) / tot

round(share, 1)
c(n_states = length(MY_SOUTH),
  change_1960_to_2020 = round(share["2020"] - share["1960"], 2),
  change_1910_to_1960 = round(share["1960"] - share["1910"], 2))

plot(as.numeric(names(share)), share, type = "b", pch = 19, col = "#C41230",
     ylim = c(0, 45), xlab = "", ylab = "% of U.S. population")
abline(v = 1960, lty = 3, col = "#999")
```

**Write about:** try adding Maryland, then Oklahoma, then Kentucky and West
Virginia, then DC. Which additions make the rise *smaller*? Those states are
being called Southern by the Census Bureau and are not behaving that way. Then
the harder question: **the variable encodes an argument.** What claim about what
the South *is* does the eleven-state list make, and what different claim does the
Census list make? Which one would you defend, and does the finding survive
either way?

***

### Option C — Growth is not seats

```{r opt-c, fig.height=4.0}
nat_growth <- 100 * (nat$pop[nat$year == 2020] / nat$pop[nat$year == 1960] - 1)

plot(sc$pop_growth, sc$change, pch = 19, col = cols[sc$region], cex = 0.9,
     xlab = "% population growth, 1960-2020", ylab = "seats gained or lost",
     xlim = c(-50, 1050), ylim = c(-17, 18))
abline(h = 0, col = "#999"); abline(v = nat_growth, lty = 2, col = "#999")
text(nat_growth, 17, " national growth", pos = 4, cex = 0.8, col = "#666")
big <- sc[abs(sc$change) >= 4 | sc$pop_growth > 400, ]
text(big$pop_growth, big$change, big$name, pos = 4, cex = 0.7)

c(correlation = round(cor(sc$pop_growth, sc$change), 2),
  national_growth_pct = round(nat_growth, 1))

g <- sc[, c("name", "region", "pop_growth", "reps_1960", "reps_2020", "change")]
g$pop_growth <- round(g$pop_growth)

# the two states the paragraph below is about
g[g$name %in% c("Nevada", "Florida"), ]

# states that beat the national growth rate and gained nothing
g[sc$pop_growth > nat_growth & sc$change <= 0, ]
```

**Write about:** Nevada's population grew **988%** and it gained 3 seats. Florida
grew 335% and gained 16. The correlation between growth and seats is only 0.45.
Explain why — two reasons, and both are in the plot. Then look at the seven
states that grew faster than the country and gained nothing at all, and say what
they have in common. What does this tell you about the difference between
*growing* and *growing faster than everybody else*?

***

### Option D — Ask your own question

612 state-years, 12 censuses, 110 years. The 1920s are worth a look: Congress
refused to reapportion at all after the 1920 census, the only time that has
happened, and this file shows you the country that produced the refusal.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code.

# What this lab cannot tell you

- **Why anyone moved.** No jobs, wages, climate, housing costs, births, deaths
  or immigration. This file cannot distinguish someone arriving from Michigan
  from someone arriving from abroad, or either from families being larger.
- **Anything between censuses.** Twelve observations, ten years apart. A
  hurricane, a plant closure or a pandemic is invisible unless it survives to the
  next April.
- **Anything partisan.** No election results, no party of any member. The
  electoral-vote figures in Step 10 are arithmetic about states, not about who
  won.
- **How accurate the count is.** Resident population is printed as an exact
  integer. The Bureau publishes coverage error separately and it is not in this
  file. The margins here are far larger than plausible error — 15 seats is not an
  error bar — but `apportionment` shows a case where it decides the
  answer.
- **What a different rule would do.** Counting citizens only, or counting
  incarcerated people at their home address, would change these numbers. Neither
  can be computed here.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Seventy-five seats moved between states and nobody voted on any of it. Is that
  a feature of the constitutional design or a problem with it?
- A representative today speaks for 3.6 times as many people as in 1910. What, if
  anything, does that do to representation — and why has no Congress fixed it?
- The Census Bureau's "South" includes Maryland and Oklahoma. Whose definition of
  a region should a federal statistical agency use, and what happens to a finding
  when the category is contested?

***

### Where the data came from

**U.S. Census Bureau**, apportionment time series 1910–2020 — a single 40 KB CSV,
no key required
(<https://www2.census.gov/programs-surveys/decennial/2020/data/apportionment/apportionment.csv>),
listed at
<https://www.census.gov/data/tables/time-series/dec/apportionment-data-text.html>.

Three things `data/build-data.R` had to handle and documents. **Numbers arrive as
quoted strings with thousands separators** and become `NA` if you convert them
without stripping the commas. **Puerto Rico is a `State` row that is not in the
national total, and DC is in the national total with no representative.** And
**Alaska and Hawaii have no representatives before 1960**, which is why the seat
window starts there.

The build script does one thing worth copying: the Bureau publishes its own
four-region totals in the same file, so the script **rebuilds those rows from the
state rows and stops if a single region-year disagrees.** A lookup table you
wrote by hand should always be checked against something you did not.

Regions and divisions follow the Census Bureau's *Statistical Groupings of States
and Counties* (Geographic Areas Reference Manual, ch. 6).
