---
title: "Key: Seventy-Five Seats"
subtitle: "84-355 Democracy's Data | Regional shift 1960–2020 | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.0,
                      dpi = 108, fig.retina = 1)
options(scipen = 999)

st  <- read.csv("data/states.csv",     stringsAsFactors = FALSE)
reg <- read.csv("data/regions.csv",    stringsAsFactors = FALSE)
nat <- read.csv("data/nation.csv",     stringsAsFactors = FALSE)
sc  <- read.csv("data/seatchange.csv", stringsAsFactors = FALSE)
sd  <- read.csv("data/southdefs.csv",  stringsAsFactors = FALSE)

Y1 <- 1960; Y2 <- 2020; HOUSE <- sum(sc$reps_2020)
n  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
dv <- function(d, y) sd$share[sd$definition == d & sd$year == y]

DEF_CEN <- "Census South (16 states + DC)"
DEF_CON <- "Confederate South (11 states)"
DEF_BOR <- "Border South only (DC, DE, KY, MD, OK, WV)"
DEF_SUN <- "Sun Belt (South + West)"

gross  <- sum(sc$change[sc$change > 0])
netreg <- tapply(sc$change, sc$region, sum)
crossed <- sum(netreg[netreg > 0])
ny     <- sc[sc$name == "New York", ]
blk    <- ifelse(sc$region %in% c("Northeast", "Midwest"), "NE+MW", "S+W")
b60    <- tapply(sc$reps_1960, blk, sum); b20 <- tapply(sc$reps_2020, blk, sum)
ev60   <- b60 + 2 * table(blk);           ev20 <- b20 + 2 * table(blk)

dec    <- st[st$year > Y1 & st$name != "District of Columbia" & !is.na(st$repchg), ]
byd    <- tapply(dec$repchg, dec$year, function(x) sum(x[x > 0]))
sumdec <- sum(byd)

z    <- st[st$year >= Y1 & st$name != "District of Columbia", ]
sp   <- split(z[order(z$year), ], z$name[order(z$year)])
mono <- sapply(sp, function(x) { d <- diff(x$reps[order(x$year)])
                                 all(d >= 0) || all(d <= 0) })
NONMONO <- names(mono)[!mono]

P1 <- nat$pop_per_rep[nat$year == 1910]; P2 <- nat$pop_per_rep[nat$year == Y2]
NG <- 100 * (nat$pop[nat$year == Y2] / nat$pop[nat$year == Y1] - 1)

# interstate spread in people per seat: largest state district / smallest
ppr <- st[st$year >= Y1 & st$name != "District of Columbia" & !is.na(st$reps), ]
ppr$per <- ppr$pop / ppr$reps
SPREAD <- tapply(ppr$per, ppr$year, function(v) max(v) / min(v))

# ---- the window sweep behind Option A --------------------------------------
win <- function(S) {
  a <- st[st$year == S  & st$name != "District of Columbia", c("name", "region", "reps")]
  b <- st[st$year == Y2 & st$name != "District of Columbia", c("name", "reps")]
  w <- merge(a, b, by = "name", suffixes = c("_s", "_e"))
  w$ch <- w$reps_e - w$reps_s
  w
}
WSTART <- seq(Y1, 2010, 10)

# ---- the South sweep behind Option B ---------------------------------------
CONFED <- c("Alabama", "Arkansas", "Florida", "Georgia", "Louisiana",
            "Mississippi", "North Carolina", "South Carolina", "Tennessee",
            "Texas", "Virginia")
BORDER <- c("Maryland", "Delaware", "Oklahoma", "West Virginia", "Kentucky",
            "District of Columbia")
TOT    <- tapply(st$pop, st$year, sum)
share_of <- function(v) {
  s <- st[st$name %in% v, ]
  100 * tapply(s$pop, s$year, sum) / TOT
}

# ---- render data.frames as tables; envir is required or knitr ignores it ----
knit_print.data.frame <- function(x, ...) {
  nm <- names(x); nm <- gsub("_", " ", nm)
  nm <- sub("^(.)", "\\U\\1", nm, perl = TRUE)
  knitr::knit_print(knitr::kable(x, col.names = nm, row.names = FALSE), ...)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame,
                 envir = asNamespace("knitr"))
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`regional-shift-brief.Rmd`) is now a chapter rather
than a parallel copy of the lab: it is organised by argument, under idea
headings, with four numbered figures and no numbered steps. The lab
(`regional-shift.Rmd`) still runs Steps 1–11 with the code exposed.
**Students should read the brief first.** **Step numbers below refer to the
lab**; the chapter is cited by figure number or section title.

**This session is built around a single reveal, and the chapter now protects it
too.** The brief holds the total back for several pages and stops the reader
with its own prediction box — write down two numbers, New York's total loss and
the seats that changed states altogether — immediately before *What the
maintenance adds up to*, which is where it reveals `r n(gross)`. A student who
read the chapter properly has therefore already guessed, while the guess was
still worth something. The lab repeats the structure: Steps 1–4 are deliberately
boring — they show one reapportionment, it moves
`r n(byd[[as.character(Y2)]])` seats, and the room concludes that
reapportionment is administrative housekeeping — and Step 5 adds six of them
together. **Get the prediction out of the room before Step 5 runs**: ask for the
number they wrote down while reading, and collect a fresh guess from anyone who
did not read. Everything else in the hour depends on that gap.

Every number below was recomputed from the built files at knit time.

***

# The session in one sentence

Between `r Y1` and `r Y2` the United States moved `r n(gross)` House seats —
`r pc(100 * gross / HOUSE)`% of the chamber, and with them
`r n(crossed)` electoral votes — from the Northeast and Midwest to the South and
West, without a single vote being taken on any of it, and no individual decade
ever looked like more than paperwork.

***

# The findings

```{r headline}
c(seats_that_changed_states = gross,
  pct_of_the_house          = round(100 * gross / HOUSE, 1),
  crossed_a_regional_border = crossed,
  largest_single_decade     = max(byd),
  smallest_single_decade    = min(byd),
  new_york_loss             = abs(ny$change),
  states_smaller_than_that  = sum(sc$reps_2020 <= abs(ny$change)),
  ne_mw_seats_1960          = b60[["NE+MW"]],
  ne_mw_seats_2020          = b20[["NE+MW"]],
  ne_mw_pct_2020            = round(100 * b20[["NE+MW"]] / HOUSE, 1),
  ne_mw_ev_pct_1960         = round(100 * ev60[["NE+MW"]] / sum(ev60), 1),
  ne_mw_ev_pct_2020         = round(100 * ev20[["NE+MW"]] / sum(ev20), 1),
  one_way_states            = sum(mono),
  people_per_seat_multiple  = round(P2 / P1, 2),
  population_multiple       = round(nat$pop[nat$year == Y2] /
                                    nat$pop[nat$year == 1910], 2))
```

- **`r n(gross)` seats changed states, `r n(crossed)` of them left their region
  entirely.** The shift is not neighbours trading; it is one half of the country
  moving to the other.
- **No decade did it.** The largest reapportionment in the window moved
  `r n(max(byd))` seats, the smallest `r n(min(byd))`. The finding is entirely a
  product of addition.
- **New York lost `r n(abs(ny$change))`**, and `r n(sum(sc$reps_2020 <= abs(ny$change)))`
  of the `r nrow(sc)` states do not have `r n(abs(ny$change))` seats in total.
- **The Northeast and Midwest went from a House majority
  (`r n(b60[["NE+MW"]])` of `r n(HOUSE)`) to `r pc(100 * b20[["NE+MW"]] / HOUSE)`%**,
  and from `r pc(100 * ev60[["NE+MW"]] / sum(ev60))`% of the fifty states'
  electoral votes to `r pc(100 * ev20[["NE+MW"]] / sum(ev20))`%.
- **`r sum(mono)` of `r length(mono)` states moved in one direction only.** A
  ratchet, not a cycle.
- **A representative now speaks for `r pc(P2 / P1, 2)`× as many people as in
  1910**, and the country grew `r pc(nat$pop[nat$year == Y2] / nat$pop[nat$year == 1910], 2)`×.
  The near-identity is the mechanism: the denominator was frozen in 1929.

**One arithmetic detail worth having ready**, because a sharp student will ask
why the per-decade gains do not sum to the cumulative total:

```{r reconcile}
c(sum_of_per_decade_gains = sumdec,
  cumulative_1960_to_2020 = gross,
  difference              = sumdec - gross,
  states_that_reversed    = length(NONMONO))
NONMONO
```

The gap is exactly the number of states that moved in both directions
(`r paste(NONMONO, collapse = ", ")`), each of which gained a seat it later gave
back or gave one back it later regained. **Gross movement exceeds net movement by
precisely `r sumdec - gross`, in sixty years.** That is the ratchet stated as an
identity rather than as a metaphor.

***

# The three things to get said

**1. The unit of observation is doing the work.** The same file, honestly
reported, supports "reapportionment is minor administrative housekeeping" and
"reapportionment moved a sixth of the House." Neither is a lie. The difference is
whether you look at one decade or six. **Make somebody say why a news cycle
produces the first framing** — because reporting is organised around events, and
this is not an event. It is a rate. The course's recurring question, *what
window is your claim about?*, has no cleaner example.

**2. Nobody decided this, and that is the interesting part.** Students arrive
expecting political change to have an author. Ask who is responsible for the
transfer of `r n(crossed)` electoral votes between regions. There is no answer:
no bill, no vote, no litigant, no agency. The Constitution specifies a procedure
and the procedure ran. **A student who says "so it is nobody's fault" has
half of it; push to the other half — the rules that convert population into power
were chosen, and choosing not to revisit them is also a choice.** The House has
been `r n(HOUSE)` since 1929.

**3. Step 8 is the methodological spine, and it is the step that is easiest to
skip.** "The South" has no agreed definition, the lab computes the finding under
several, and the answer moves — but never reverses. That combination is what an
honest sensitivity analysis looks like, and students almost never see one. See
the dedicated section below; **do not let this become a two-minute aside.**

**The line for the board:** power moved without anybody moving it, and the only
reason it looks like nothing happened is that we keep checking every ten years.

***

# Walking the steps

| Step | What it does | Min | Watch for |
|---|---|---|---|
| **1–3** One record, the file, why 1960 | Alaska and Hawaii have no seats before `r Y1` | 7 | **Do not skip the de-comma trap.** `"4,951,560"` becomes `NA` silently. It is the cheapest data-literacy lesson in the course. |
| **4** One decade at a time | `r n(byd[[as.character(Y2)]])` seats, `r n(sum(st$year == Y2 & !is.na(st$repchg) & st$repchg != 0))` states in `r Y2` | 8 | **Let them conclude it is trivial.** The step is supposed to underwhelm. |
| **—** The prediction | Written down, before Step 5 | **4** | **Non-negotiable.** Collect guesses aloud. Typical answers are 20–30; the answer is `r n(gross)`. |
| **5** The pivot | `r n(gross)` seats, `r pc(100 * gross / HOUSE)`% of the House | **10** | **The session.** Sit in the silence. Then the New York comparison: `r n(sum(sc$reps_2020 <= abs(ny$change)))` of `r nrow(sc)` states are smaller than what one state lost. |
| **6** Regions and the ratchet | `r n(b60[["NE+MW"]])` → `r n(b20[["NE+MW"]])`; `r sum(mono)` of `r length(mono)` states one-way | 8 | The monotonicity check is what upgrades "a shift" to "a trend". Students read it as a technicality. |
| **7** The long line | Population share, 1910–`r Y2` | 8 | **The South's rise begins after `r Y1`, not before.** Between 1910 and `r Y1` its share *fell*. Nearly everyone assumes otherwise. |
| **8** Defining the South | Four definitions, one direction | **10** | **The methods step.** See below. Take a vote on which definition is "right" before showing the table. |
| **9** What a seat is worth | `r n(P1)` → `r n(P2)` people per seat | 6 | Separate *bigger* districts from *less equal* districts — the ratio actually improved (`r pc(SPREAD[[as.character(Y1)]], 2)`→`r pc(SPREAD[[as.character(Y2)]], 2)`). |
| **10** Electoral votes | `r pc(100 * ev60[["NE+MW"]] / sum(ev60))`% → `r pc(100 * ev20[["NE+MW"]] / sum(ev20))`% | 6 | Keep it arithmetic. The file has no election results; do not let the room infer a partisan story it cannot support. |
| **11** What you can say | Including what you cannot | 5 | The "why" is not in this file. Name `labs/migration` out loud. |

**If you are short on time**, protect the prediction, Step 5 and Step 8. The
underwhelming decade, the sixty-year total, and the demonstration that the total
does not depend on how you drew the category are the whole argument. Steps 9 and
10 are extensions.

***

# The South-definition sensitivity, and how to use it

**This is the most transferable thing in the lab and it needs ten minutes, not
two.** The lab hands students the finding "the South rose"; Step 8 asks whether
that sentence survives a change in who counts as Southern.

Run the vote first. Ask the room: *is Maryland Southern? Oklahoma? Delaware? West
Virginia? the District of Columbia?* You will get disagreement in under a minute,
and the disagreement is the point — these are not edge cases invented for a
seminar. The Census Bureau counts all five as South. `house-competition` and
`historical-campaigns`, both in this course, count none of them.

Then show what it does to the answer.

```{r sens}
defs <- list(
  "11 Confederate states"        = CONFED,
  "  + Maryland"                 = c(CONFED, "Maryland"),
  "  + Oklahoma"                 = c(CONFED, "Oklahoma"),
  "  + Kentucky, West Virginia"  = c(CONFED, "Kentucky", "West Virginia"),
  "  + District of Columbia"     = c(CONFED, "District of Columbia"),
  "Census South (16 + DC)"       = c(CONFED, BORDER),
  "Sun Belt (South + West)"      = c(CONFED, BORDER,
                                     unique(st$name[st$region == "West"])))

out <- do.call(rbind, lapply(names(defs), function(k) {
  s <- share_of(defs[[k]])
  data.frame(definition = k, states = length(defs[[k]]),
             share_1910 = round(s[["1910"]], 2),
             share_1960 = round(s[[as.character(Y1)]], 2),
             share_2020 = round(s[[as.character(Y2)]], 2),
             rise_1960_2020 = round(s[[as.character(Y2)]] - s[[as.character(Y1)]], 2),
             drift_1910_1960 = round(s[[as.character(Y1)]] - s[["1910"]], 2))
}))
out
```

**Four things to draw out of that table.**

**The direction never changes.** Every definition rises, and rises by at least
`r pc(min(out$rise_1960_2020), 1)` points. **A finding that survives every
plausible operationalisation of its central category is a robust finding**, and
this is the sentence to make them say. It is not the same as a finding that is
big — it is a finding that is not an artefact of a coding decision.

**The magnitude does change, and by a defensible amount.** From
`r pc(min(out$rise_1960_2020), 2)` points to
`r pc(max(out$rise_1960_2020[out$definition != DEF_SUN]), 2)`
across the South definitions — a spread of
`r pc(max(out$rise_1960_2020[out$definition != DEF_SUN]) - min(out$rise_1960_2020), 2)`
points, or
`r pc(100 * (max(out$rise_1960_2020[out$definition != DEF_SUN]) - min(out$rise_1960_2020)) / max(out$rise_1960_2020[out$definition != DEF_SUN]))`%
of the largest estimate. **So a student who reports the
number to two decimals without saying which South they used has overstated their
precision**, even though they have not overstated their conclusion.

**The narrow definition gives the larger answer.** This is counter-intuitive and
worth pausing on: adding states to a growing category makes the growth look
*smaller*. It can only mean the added states are a drag, and they are. The border
bloc as a whole went from `r pc(dv(DEF_BOR, Y1))`% of the country to
`r pc(dv(DEF_BOR, Y2))`%.

**But not all of them — and this is the reward for a student who checks.**
Two of the six additions make the rise bigger, and Maryland is much the larger of
the two:

```{r border-detail}
b <- do.call(rbind, lapply(BORDER, function(v) {
  s <- share_of(v)
  data.frame(state = v, share_1960 = round(s[[as.character(Y1)]], 3),
             share_2020 = round(s[[as.character(Y2)]], 3),
             change = round(s[[as.character(Y2)]] - s[[as.character(Y1)]], 3))
}))
b[order(-b$change), ]
```

Maryland and Delaware grew slightly faster than the country;
`r paste(head(b$state[b$change < 0], -1), collapse = ", ")` and
`r tail(b$state[b$change < 0], 1)` did not. **The
Census South is a mixture of two demographic regimes wearing one label**, which
is a better criticism of the category than "it includes Maryland and Maryland is
not really Southern." The first is measurable; the second is an argument about
identity that this file cannot adjudicate.

**Where to land it.** The Census Bureau's regions are not wrong — they are a
statistical convenience with a legal and historical basis, and the Bureau
publishes its own totals so you can check your mapping against them, which
`data/build-data.R` does. The lesson is that **a category boundary is a
hypothesis, and a finding is only as strong as its behaviour when you move the
boundary.** Students should leave able to ask, of any regional or demographic
claim: *who is in the category, who decided, and does the answer move?*

***

# Option A — Move the window

## Verified output

```{r a}
a <- do.call(rbind, lapply(WSTART, function(S) {
  w <- win(S)
  data.frame(window = paste(S, "to", Y2), decades = (Y2 - S) / 10,
             seats_moved = sum(w$ch[w$ch > 0]),
             per_decade = round(sum(w$ch[w$ch > 0]) / ((Y2 - S) / 10), 2),
             NE_MW = sum(w$ch[w$region %in% c("Northeast", "Midwest")]),
             S_W   = sum(w$ch[w$region %in% c("South", "West")]))
}))
a$pct_of_1960_total <- round(100 * a$seats_moved / a$seats_moved[1], 1)
a
```

```{r a-check}
half <- a[a$window == paste(1990, "to", Y2), ]
c(full_window_seats   = a$seats_moved[1],
  half_window_seats   = half$seats_moved,
  half_of_full        = a$seats_moved[1] / 2,
  ratio               = round(half$seats_moved / a$seats_moved[1], 3))
```

**The lab asks a specific question and it has a specific answer.** A window half
as long (`r n(half$decades)` decades instead of `r n(a$decades[1])`) yields
`r n(half$seats_moved)` seats, not the `r pc(a$seats_moved[1] / 2, 1)` a
constant rate would give. The rate is
**falling**: `r pc(a$per_decade[1], 1)` seats per decade over the full window,
`r pc(a$per_decade[a$window == paste(2010, "to", Y2)], 1)` in the most recent one.

**3/4:** Changes `START`, reports that a shorter window moves fewer seats, and
gets the direction right — it is less than half, so the shift is decelerating
rather than accelerating.

**4/4:** Computes the per-decade rate rather than eyeballing the total, and
notices that the ranking of gainers and losers barely moves while the magnitudes
shrink — Texas, Florida and Arizona lead every window; New York, Ohio,
Pennsylvania, Illinois and Michigan trail every window. **The composition is
stable; only the speed changes.** That distinction is the answer.

**4/4, best:** Takes the confidence question seriously, which the prompt asks for
and most students skip. With `r n(a$decades[1])` observations of a per-decade
rate, "decelerating" is a description of six numbers, not an estimate with a
standard error — and the most recent point (`r n(min(byd))` seats in `r Y2`) is
the smallest in the series and therefore the most tempting and least reliable
place to draw a trend. A student who says *this could be a slowdown, or it could
be that the 2020 census undercounted, or it could be one decade of noise, and
this file cannot separate them* has understood what six observations buy you.
**Reward explicitly.**

***

# Option B — Redefine the South

## Verified output

The full sweep is in the sensitivity section above. Students will typically run
three or four variants; these are the ones the prompt names.

```{r b}
out[, c("definition", "states", "share_1960", "share_2020", "rise_1960_2020")]
```

**3/4:** Edits `MY_SOUTH`, reports that the share changes, and notices the rise
happens regardless.

**4/4:** Identifies *which* additions shrink the rise and says why — the added
states grew slower than the country, so they dilute a growing category. Connects
that back to the finding: the Census definition **understates** the shift, so
using it is the conservative choice.

**4/4, best:** Answers the harder half of the prompt — that the variable encodes
an argument. The eleven-state list defines the South by a political event of the
1860s; the Census list defines it by a line on a map maintained for statistical
convenience. **These are claims about what a region *is*, and they are not
interchangeable even though they produce similar answers here.** The strongest
version notes that the two happen to agree in this application and would not
agree in others — a lab about Reconstruction-era voting rights and a lab about
Sun Belt housing markets should not use the same South — and then says which one
this lab should use and why. Any defensible choice earns the mark; **the
reasoning is the assignment.**

The Maryland exception (`r pc(b$change[b$state == "Maryland"], 3)` points, the
larger of the two border additions that make the rise *larger*) is now priced out in the
chapter itself, under *Whether the South is eleven states or seventeen*, so
naming it no longer earns anything on its own. **Reward instead the student who
uses it:** who sees that the six border units are a drag in aggregate while two
of them are not, and concludes that "the border states behave like the
Northeast" is a statement about the aggregate that at least one member of the
aggregate contradicts. That is the same reading skill the sensitivity table
teaches, applied one level down.

***

# Option C — Growth is not seats

## Verified output

```{r c}
c(correlation_all_states = round(cor(sc$pop_growth, sc$change), 2),
  national_growth_pct    = round(NG, 1),
  fastest_growing_state  = round(max(sc$pop_growth)),
  its_seat_gain          = sc$change[which.max(sc$pop_growth)])

big <- sc[sc$reps_1960 >= 10, ]
c(states_with_10plus_seats_in_1960 = nrow(big),
  correlation_among_those          = round(cor(big$pop_growth, big$change), 2),
  correlation_with_size_x_relative_growth =
    round(cor(sc$change, sc$reps_1960 * (sc$pop_growth - NG)), 2))

x <- sc[sc$pop_growth > NG & sc$change <= 0,
        c("name", "region", "pop_growth", "reps_1960", "change")]
x$pop_growth <- round(x$pop_growth, 1)
x[order(x$reps_1960), ]
c(median_1960_delegation_of_those = median(x$reps_1960),
  median_1960_delegation_all      = median(sc$reps_1960))
```

**The two reasons the prompt asks for, both visible in the scatter:**

1. **Percentage growth ignores base size.** Nevada grew
   `r n(sc$pop_growth[sc$name == "Nevada"])`% from a delegation of
   `r n(sc$reps_1960[sc$name == "Nevada"])`; Florida grew
   `r n(sc$pop_growth[sc$name == "Florida"])`% from a delegation of
   `r n(sc$reps_1960[sc$name == "Florida"])`, and gained `r n(sc$change[sc$name == "Florida"])` seats to
   Nevada's `r n(sc$change[sc$name == "Nevada"])`. Seats follow *people*, and a
   large percentage of a small number is a small number.
2. **Apportionment is relative, not absolute.** The national benchmark is
   `r pc(NG)`% growth. Growing is not enough; you have to grow faster than
   everyone else, and by enough to cross a whole-seat threshold.

Put those together and the correlation stops being disappointing:
`r pc(cor(sc$change, sc$reps_1960 * (sc$pop_growth - NG)), 2)` between seat
change and *1960 delegation size × growth above the national rate*. **The raw
0.45 is not weak measurement; it is the wrong quantity.**

**3/4:** Reports the correlation, names Nevada and Florida, and gets at least one
of the two reasons.

**4/4:** Gets both reasons and identifies what the
`r nrow(x)` no-gain fast-growers have in common — they are small
(median `r n(median(x$reps_1960))` seats in `r Y1` against
`r n(median(sc$reps_1960))` for all states) or they beat the national rate only
narrowly (`r paste(x$name[x$pop_growth < 100], collapse = " and ")`). Whole seats
are a coarse instrument, and a state with one seat cannot be given nine tenths of
another.

**4/4, best:** Reframes the correlation as a measurement question rather than a
finding — asks what quantity *would* correlate with seat change, and constructs
it, or gets close in words. A student who says *the right predictor is share of
national population, not growth in own population* has stated the apportionment
rule from scratch, which is the whole content of `apportionment` arrived
at independently. **This is the strongest response the lab can produce; mark it
as such.**

***

# Option D — Ask your own question

Ungraded on content, graded on whether the question is answerable with this file.
The 1920s prompt is the one most students take, so here is what it contains:

```{r d}
nat[nat$year %in% c(1910, 1920, 1930), c("year", "pop", "reps", "pop_per_rep")]

chg <- st[st$year == 1920 & !is.na(st$repchg) & st$repchg != 0,
          c("name", "year", "reps", "repchg")]
chg
c(states_with_any_change_in_1920 = nrow(chg),
  states_in_1970_for_comparison  = sum(st$year == 1970 & !is.na(st$repchg) &
                                       st$repchg != 0))
```

**Congress refused to reapportion after the 1920 census — the only time it has
happened — and the file records the refusal as an absence.** Every state's change
is zero except Arizona and New Mexico, which are not reapportionment at all: they
became states in 1912 and enter the seat columns for the first time. **The
absence of an event is the observation**, which is a genuinely hard thing to see
in a data file and worth pointing out to anyone who takes this option.

The population columns show the country that produced the refusal: the 1920
census was the first to find a majority of Americans living in urban areas, and
reapportioning would have moved seats from rural states to cities. The file
shows the population change and is completely silent on the motive. **A student
who asserts the motive from this file has made exactly the error the lab is
about;** a student who says *the file shows the seats did not move and cannot
tell me why* has got it right.

Two smaller quirks in the source, in case they surface:

- `Change in Number of Representatives` is the Bureau's own column and disagrees
  with the difference of `Number of Representatives` in exactly two rows, both
  Arizona and New Mexico in 1930, both outside the `r Y1`–`r Y2` seat window.
- 1910 apportioned `r n(nat$reps[nat$year == 1910])` seats, not `r n(HOUSE)`;
  Arizona and New Mexico brought it to `r n(HOUSE)` after statehood, where the
  Permanent Apportionment Act of 1929 froze it.

***

# Grading

Four-point scale from the syllabus. Two or three paragraphs, one chosen chunk.

- **Never penalise code.** It is given, and several options require changing one
  literal — a student who changed it correctly has done everything asked.
- **The move that earns a 4** is treating a methodological choice — the window,
  the category boundary, the quantity correlated — as a substantive decision with
  consequences, and saying what those consequences are.
- **A 2** is "the South grew and the Northeast shrank." True, in the brief, and
  requires nothing of the student.
- **Also a 2**: any claim about *why* people moved. The file has population and
  seats. Jobs, weather, housing and air conditioning are not in it, and the brief
  says so twice.
- **Also a 2**: treating `r n(gross)` as large or small without a comparison. The
  number means nothing until it is put next to something — the size of a
  delegation, a decade's movement, the margin in the House.
- **Reward the reversal.** A student who notices that the same data supports two
  opposite-sounding framings, and explains what makes each one honest, has the
  point of the session.
- **Reward anyone who asks for uncertainty.** There is none in this file:
  populations print as exact integers, coverage error is published elsewhere.
  Noticing its absence is the most sophisticated available response, and it is
  the same instinct `apportionment` rewards for a case where an 89-person
  margin decided a seat.

## Anticipated questions

- *"Why not go back to 1910? The population data is there."* — Because the seat
  data would be comparing different countries. Alaska and Hawaii are in the file
  from 1910 with populations and no representatives. The lab uses 1910 population
  where it is honest to (Step 7, labelled) and 1960 seats everywhere else. **This
  question comes up most sessions and the answer is a good one to have crisp.**
- *"Isn't this just California and Texas?"* — No, and the bar chart answers it:
  `r n(sum(sc$change != 0))` states changed size and
  `r n(sum(sc$change < 0))` lost seats. But the follow-up is real — the top four
  gainers account for `r n(sum(sort(sc$change, decreasing = TRUE)[1:4]))` of the
  `r n(gross)` seats. Concentration and breadth are both true.
- *"Does the South include Texas?"* — In every definition used here, yes. Texas
  is `r n(sc$change[sc$name == "Texas"])` seats of the South's
  `r n(netreg[["South"]])` net gain, so a student who wants to argue Texas is its
  own thing is making a real point. Let them test it: remove Texas from
  `MY_SOUTH` in Option B and see what is left.
- *"Do the electoral-vote numbers mean Republicans gained?"* — Not from this
  file. The regions that gained have voted differently in different decades, and
  there is no election result anywhere in the data. **This is worth being firm
  about**; it is the single most likely place for a student to overreach.
- *"Will it keep going?"* — `r n(min(byd))` seats moved in `r Y2`, the smallest
  reapportionment in the window. Either a slowdown or noise. Twelve observations
  cannot tell you, and Option A is where a student can see that for themselves.
- *"Why is the House still `r n(HOUSE)`?"* — Statute, not the Constitution;
  Permanent Apportionment Act of 1929. `apportionment` runs that argument
  properly and this lab deliberately does not.

## Connections

- **`apportionment`** — the same machinery over one decade instead of six.
  That lab takes a single reapportionment apart, down to New York missing the
  435th seat by 89 people; this one never opens the formula and asks only what
  sixty years of it add up to. **Run `apportionment` first.** The
  arithmetic is assumed here, not taught.
- **`migration`** — the mechanism this lab can measure the consequence of and
  cannot explain. Every "why" question raised in this session belongs there, and
  saying so out loud is what keeps Step 11 honest.
- **`house-competition`** and **`historical-campaigns`** — both use the
  eleven-state Confederate South. Step 8 is the place to say that the course
  itself uses two different Souths and that this is a defensible choice rather
  than an inconsistency, provided each is stated.
- **`precinct-geography`** — the other lab where two defensible operational
  choices give different numbers from the same data. There it was areal versus
  population weighting; here it is which states are Southern. Same lesson, and
  students who have done both should be asked to name it.
