---
title: "Data Session 5 Key, part 3: What the Country Was Looking At"
subtitle: "84-355 Democracy's Data | Session 5 — Media attention (demo) | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
w <- read.csv("data/wiki_attention_2024.csv", stringsAsFactors = FALSE)
w$date <- as.Date(w$date)
g <- function(a) w[w$article == a, ]
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

Public attention to politics is compressed into a handful of days, arrives in
response to events rather than accumulating steadily, responds to false claims
exactly as readily as true ones — and bears no reliable relationship to who wins.

***

# The findings

```{r headline}
for (a in c("2024_United_States_presidential_election", "Kamala_Harris",
            "JD_Vance", "Tim_Walz", "Springfield,_Ohio")) {
  s <- g(a)
  cat(sprintf("%-42s median %7.0f  peak %9.0f  (%s)  ratio %5.0f\n",
              a, median(s$views), max(s$views),
              format(s$date[which.max(s$views)]),
              max(s$views) / median(s$views)))
}
```

- **Election article:** median 43,122, peak **4,260,085** — a **99×** ratio.
- **Harris:** median 12,417; **2,532,334** on 22 July, the day after Biden
  withdrew. **204×**.
- **Vance:** median **23 views a day** as a sitting US senator.
- **Walz:** **3,811,244** on 6 August — the largest single-day figure for any
  person in the dataset.
- **Springfield, Ohio:** median 242, peak **57,339** on 11 September. **237×**.

***

# The three ideas to teach

**1. Attention is event-driven and extremely concentrated.** Nothing accumulates.
There is a flat baseline and a few enormous spikes.

**2. Misinformation and truth are indistinguishable in the attention data.** The
Springfield claim was false and produced a 237-fold surge, plus a spillover to
`Haitian_Americans` (median 174, peak 3,984 on 13 September). Agenda-setting
does not require accuracy. This is the strongest illustration of Chapter 8's
central mechanism you will get from any dataset students can hold.

**3. Attention is not support.**

```{r peaks}
pk <- aggregate(views ~ article, data = w[w$article %in%
  c("Donald_Trump","Kamala_Harris","JD_Vance","Tim_Walz"), ], FUN = max)
pk[order(-pk$views), ]
```

Walz's peak exceeds Trump's. His ticket lost. If attention converted to votes,
that ordering would be a forecast; it is a record of what surprised people.

The lab has students write this ranking down before the table prints; expect most
to lead with Trump and finish with a running mate, and use the wrong guess rather
than the table as the way into the discussion.

***

# Option A — Pick an article and date its spikes

## Verified output

```{r a}
for (a in c("Project_2025", "Swing_state")) {
  s <- g(a); top <- s[order(-s$views), ][1:4, ]
  cat("\n", a, " (median ", median(s$views), ")\n", sep = "")
  print(top[order(top$date), ])
}
```

**Project 2025** peaks at 722,256 on 6 November, but its earlier spikes —
**482,218 on 5 July** and 478,842 on 6 July — do *not* correspond to any event in
`campaign_events_2024.csv`. That is the point of the option. Those days follow a
deliberate Democratic campaign to publicise a Republican-aligned policy document,
amplified heavily online.

**Swing_state** (median 1,346, peak 252,647) rises only very late.

## What a good answer says

**3/4:** Names the spike dates and checks them against `ev`, noticing that some
line up and some do not.

**4/4:** Reasons about the July spikes having no corresponding set-piece event,
and identifies that a campaign can *manufacture* attention to something the
other side would rather nobody read. That is a different mechanism from a debate
or a withdrawal — it is agenda-setting as strategy rather than as by-product.

**4/4, best:** notices that `Swing_state` stays flat until the final weeks and
draws the implication — ordinary people engage with the *mechanics* of elections
only at the very end, long after the campaigns have made their decisions about
where to compete and where the money went (the campaign-finance lab, this same
session).

**Common stumble:** assuming every spike must match a listed event. The events
file is deliberately incomplete; unexplained spikes are findings, not errors.

***

# Option B — How concentrated is attention?

## Verified output

```{r b}
for (a in c("Donald_Trump", "Kamala_Harris", "Opinion_poll",
            "Electoral_College_(United_States)")) {
  s <- g(a); s <- s[order(-s$views), ]; s$cum <- cumsum(s$views)/sum(s$views)*100
  cat(sprintf("%-40s top10 %5.1f%%  top30 %5.1f%%  days to half %3d\n",
              a, s$cum[10], s$cum[30], which(s$cum >= 50)[1]))
}
```

| article | top 10 days | top 30 days | days to reach half |
|---|---|---|---|
| Donald Trump | 27.0% | 41.5% | **53** |
| Kamala Harris | 39.3% | 60.1% | **18** |
| Opinion poll | 6.9% | 16.0% | 132 |
| Electoral College | 22.3% | 35.1% | 74 |

*(Three of these four rows were transcribed wrongly in an earlier version of
this key — Harris, Opinion poll and Electoral College. The chunk above was
always right; the markdown under it was not. Run the chunk, do not trust a
table.)*

## What a good answer says

**3/4:** Reports the numbers and observes that people are spikier than topics.

**4/4:** Explains *why* Harris is nearly three times as concentrated as Trump —
18 days versus 53 to reach half her attention. Trump was continuously famous
throughout 2024; Harris became the nominee overnight in July. Concentration here
is measuring **novelty**, not importance. A student who says "this statistic is
really about how recently someone became interesting" has read it correctly.

**4/4, best:** takes the strategic question seriously. If half a candidate's
annual attention lands in 18 days, then manufacturing such a day is worth more
than months of steady campaigning — which is an argument for exactly the kind of
event-driven, spectacle-oriented campaigning Chapter 8 describes, and a reason to
expect more of it.

***

# Option C — Two candidates side by side

## Verified output

```{r c}
h <- g("Kamala_Harris"); t <- g("Donald_Trump")
mg <- merge(h[, c("date","views")], t[, c("date","views")], by = "date",
            suffixes = c("_h","_t"))
mg$phase <- ifelse(mg$date < as.Date("2024-07-21"), "before Jul 21", "after Jul 21")
table(mg$phase, ifelse(mg$views_h > mg$views_t, "Harris", "Trump"))
```

Before 21 July: Trump leads **194 days to 8**. After: Harris leads **107 to 57**.

## What a good answer says

**3/4:** Reports the reversal and dates it to the withdrawal.

**4/4:** Refuses the word "won." Harris led attention on roughly two-thirds of
the days after July and lost the election by 1.5 points nationally. The honest
conclusion is that leading in pageviews indicates salience, not preference — and
that a candidate can dominate attention precisely *because* people do not yet
know who she is.

**4/4, best:** connects to Part 6 and to Session 9's polling material — attention
is a measure with no theory of the electorate behind it, whereas a poll at least
attempts to represent one. Asks what a campaign should actually want: attention,
or something else?

***

# Option D — Their own question

`Opinion_poll` and `Electoral_College_(United_States)` reward comparison against
the personalities. Unexplained spikes are legitimate findings.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is refusing to let a measure mean more than it
  does. This lab is built so that the most striking numbers (Walz's peak, Harris
  leading after July) point *away* from the conclusion a careless reader would
  draw. Students who notice that tension are doing the course's central skill.
- **Reward "I cannot tell from this."** Pageviews genuinely cannot separate
  curiosity from approval, or belief from disbelief.

## Anticipated questions

- *"Isn't Wikipedia a weird proxy?"* — It is a narrow one, and that is a virtue
  for teaching: it is publicly measured, consistently defined, and unarguably
  behavioural. Compare it to a survey question about media consumption, where the
  respondent has to remember and report accurately.
- *"Aren't a lot of these bots?"* — The file uses Wikimedia's `user` filter,
  which excludes known automated traffic. Worth saying, because it is the first
  objection a sharp student raises.
- *"Why is Vance's baseline 23 and Walz's 1,242?"* — Walz was a sitting governor
  of a state with a large media market; Vance was a first-term senator. Both are
  astonishingly small. The real lesson is how little attention national political
  figures command before they are elevated.
- *"Did the Springfield claim change any votes?"* — Unknown, and this data cannot
  say. It can say that it moved a great many people to go and read about
  Springfield. Distinguishing those two is the whole difficulty of media-effects
  research, and it is worth being explicit that political science has struggled
  with it for eighty years.

## Connections forward

- **Session 9 (polls, models, and markets)** picks up directly: attention is not
  preference, and polls are the instrument built to measure preference. Ask what
  each one can do that the other cannot.
- **Session 10 (live returns)** — students can pull the equivalent 2026
  series live from the same API; `data/build-data.R` needs only its date range
  changed. That is a good in-class demonstration on 5 November.
- **`campaign-finance`, this same session** — Project 2025's July spike is money and organisation converted into
  attention. The two labs describe the same machine from different ends.
- The assigned readings this week (Eady et al. on the ideology of news sharing,
  and the MSNBC ratings piece) are both about **supply** — who publishes what.
  This lab is about **demand**. Ask on Tuesday which one is easier to measure,
  and which one Chapter 8's argument actually needs.
