---
title: "Lab 9: What the Country Was Looking At"
subtitle: "84-355 Democracy's Data | Week 9"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Chapter 8 asks what the media does to voters. That question is hard to answer
because the two things you would need — what people were exposed to, and what
they came to believe — are both very difficult to measure.

Today we measure something narrower and much more tractable: **what people went
looking for.** Every time someone reads an English Wikipedia article, that
request is counted. The counts are public, daily, and go back years.

This is not exposure and it is certainly not persuasion. It is *curiosity*, and
curiosity leaves a sharper trace than you might expect. By the end of this lab
you will be able to date the turning points of the 2024 campaign to within a day
without looking at a single news story.

Everything is base R. Work in order.

***

# Part 1: A year of attention

```{r read}
w  <- read.csv("data/wiki_attention_2024.csv", stringsAsFactors = FALSE)
ev <- read.csv("data/campaign_events_2024.csv", stringsAsFactors = FALSE)
w$date  <- as.Date(w$date)
ev$date <- as.Date(ev$date)
dim(w)
```

Twelve articles, every day of 2024, in **long format** — one row per article per
day.

```{r glimpse}
head(w, 3)
table(w$article)
```

***

# Part 2: Attention is spiky, not steady

Start with the article for the election itself.

```{r election-article, fig.width = 8.5, fig.height = 5}
e <- w[w$article == "2024_United_States_presidential_election", ]

plot(e$date, e$views / 1000, type = "l", lwd = 1.8, col = "#54278F",
     xlab = "", ylab = "Pageviews (thousands)",
     main = "English Wikipedia: '2024 United States presidential election'")
```

```{r election-summary}
c(median = median(e$views), peak = max(e$views),
  ratio = round(max(e$views) / median(e$views)))
```

> A typical day drew about **43,000** readers. The peak drew **4,260,085** —
> **ninety-nine times** the median. Almost everything that happens in public
> attention happens on a handful of days.

That is the single most important fact about the media environment, and it is
visible before we have named a single event.

***

# Part 3: The day the race changed

Now the two principals, plus Biden.

```{r three, fig.width = 8.5, fig.height = 5.5}
cands <- c("Joe_Biden", "Donald_Trump", "Kamala_Harris")
cols  <- c("#999999", "#B2182B", "#2166AC")

sub <- w[w$article %in% cands & w$date >= as.Date("2024-06-01") &
         w$date <= as.Date("2024-09-01"), ]

plot(NULL, xlim = range(sub$date), ylim = c(0, max(sub$views) / 1000),
     xlab = "", ylab = "Pageviews (thousands)",
     main = "Summer 2024")
for (i in seq_along(cands)) {
  s <- sub[sub$article == cands[i], ]
  lines(s$date, s$views / 1000, lwd = 2.2, col = cols[i])
}
abline(v = as.Date(c("2024-06-27", "2024-07-13", "2024-07-21")),
       lty = 2, col = "grey40")
legend("topleft", gsub("_", " ", cands), col = cols, lwd = 2.2,
       bty = "n", cex = 0.9)
```

Three dashed lines: the June debate, the assassination attempt, and Biden's
withdrawal. Look at what each one does.

```{r july}
h <- w[w$article == "Kamala_Harris", ]
h[h$date >= as.Date("2024-07-19") & h$date <= as.Date("2024-07-24"), ]
```

```{r july-ratio}
round(max(h$views) / median(h$views))
```

> Harris drew about **12,400** readers on a typical day of 2024. On **22 July**,
> the day after Biden withdrew, she drew **2,532,334** — roughly **200 times**
> her own baseline.
>
> Two and a half million people went to look up who she was.

Check the events file rather than trusting the eye:

```{r events}
ev
```

***

# Part 4: From nobody to nominee in a day

The running mates are the clearest case in the whole dataset, because before
their selection almost nobody was looking for them at all.

```{r vp-baseline}
vp <- w[w$article %in% c("JD_Vance", "Tim_Walz"), ]
aggregate(views ~ article, data = vp, FUN = median)
```

> **The median day of 2024 sent 23 people to J.D. Vance's Wikipedia article.**
> Twenty-three. He was a sitting United States senator.

```{r vp-plot, fig.width = 8.5, fig.height = 5}
plot(NULL, xlim = as.Date(c("2024-07-01", "2024-11-30")),
     ylim = c(0, 4000), xlab = "", ylab = "Pageviews (thousands)",
     main = "Two running mates")
for (i in seq_along(c("JD_Vance", "Tim_Walz"))) {
  a <- c("JD_Vance", "Tim_Walz")[i]
  s <- vp[vp$article == a, ]
  lines(s$date, s$views / 1000, lwd = 2.2, col = c("#B2182B", "#2166AC")[i])
}
abline(v = as.Date(c("2024-07-15", "2024-08-06", "2024-10-01")), lty = 2,
       col = "grey40")
legend("topright", c("J.D. Vance", "Tim Walz"), col = c("#B2182B", "#2166AC"),
       lwd = 2.2, bty = "n")
```

```{r vp-peaks}
for (a in c("JD_Vance", "Tim_Walz")) {
  s <- vp[vp$article == a, ]
  cat(a, "- peak", format(max(s$views), big.mark = ","),
      "on", format(s$date[which.max(s$views)]), "\n")
}
```

> Walz's announcement day drew **`r format(max(vp$views[vp$article == "Tim_Walz"]), big.mark = ",")`** readers,
> against a median day of `r format(median(vp$views[vp$article == "Tim_Walz"]), big.mark = ",")`
> across the year. Vance's own peak is printed above too.
>
> Hold on to both numbers. We come back to them in Part 6.

***

# Part 5: Misinformation leaves a footprint

On 10 September 2024, during the presidential debate, a false claim was made
that Haitian immigrants in Springfield, Ohio were eating their neighbours' pets.
The claim was investigated by local officials and found to be baseless.

The claim was false. The attention was not.

```{r springfield, fig.width = 8.5, fig.height = 5}
sp <- w[w$article == "Springfield,_Ohio", ]

plot(sp$date, sp$views, type = "l", lwd = 1.8, col = "#D95F02",
     xlab = "", ylab = "Pageviews",
     main = "English Wikipedia: 'Springfield, Ohio'")
abline(v = as.Date("2024-09-10"), lty = 2, col = "grey30")
```

```{r springfield-numbers}
c(median_day = median(sp$views),
  peak       = max(sp$views),
  peak_date  = format(sp$date[which.max(sp$views)]))

round(max(sp$views) / median(sp$views))
```

> A town of 60,000 people drew **242** readers on a normal day and **57,339** the
> day after the debate — **237 times** its baseline.

The effect is not confined to the town:

```{r haitian}
ha <- w[w$article == "Haitian_Americans", ]
c(median_day = median(ha$views), peak = max(ha$views),
  peak_date = format(ha$date[which.max(ha$views)]))
```

> **This is what Chapter 8 means by an agenda-setting effect, caught in the
> act.** The media did not tell these hundreds of thousands of people what to
> think about Springfield. It told them that Springfield was a thing to think
> about at all — and a false statement did that just as efficiently as a true one
> would have.

***

# Part 6: What attention is not

Three cautions, and the third is the one that matters.

**Pageviews are not exposure.** Far more people saw a headline about Springfield
than looked it up. This measures the fraction curious enough to act.

**Pageviews are not belief.** Someone looking up Springfield may have been
checking whether the claim was true, appalled by it, or looking for an address.
The number cannot tell them apart.

**And attention is not support.**

> **Before you run the next chunk, write a ranking down.** Four people stood on
> the two national tickets in 2024: Harris, Trump, Vance, Walz. Put all four in
> order by each one's **single largest day** of readership — not the total for
> the year, not how the election came out, just each person's one biggest day.
> **Who is first, and who is fourth?** Three of those four numbers have already
> gone past you in this lab. Commit to the whole order of four, on paper, before
> you run the chunk.

```{r attention-vs-result}
peaks <- aggregate(views ~ article, data = w[w$article %in%
  c("Donald_Trump", "Kamala_Harris", "JD_Vance", "Tim_Walz"), ], FUN = max)
peaks[order(-peaks$views), ]
```

> **Tim Walz has the largest single-day attention figure of any candidate in
> 2024** — larger than Trump's peak the day after the election. His ticket lost.
>
> Now check your ranking. Hardly anyone puts Walz first and Vance fourth,
> because the order most people have in mind is an order of fame, or of how the
> election turned out. This table is neither. It ranks the four by how many
> strangers, on one day, needed to find out who they were — and three of the
> four numbers had already gone past you before you guessed.
>
> If attention converted into votes, that table would be a forecast. It is not
> one. It is a record of what surprised people — and novelty is not popularity.
> A country looking something up is a country that does not already know.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change, or nothing at all.

Pick **one**.

***

### Option A — Pick an article and date its spikes

```{r opt-a, fig.width = 8.5, fig.height = 5}
this_article <- "Project_2025"     # <- change to any article in the data

s <- w[w$article == this_article, ]

plot(s$date, s$views / 1000, type = "l", lwd = 1.8, col = "#54278F",
     xlab = "", ylab = "Pageviews (thousands)",
     main = gsub("_", " ", this_article))
abline(v = ev$date, lty = 3, col = "grey65")

top <- s[order(-s$views), ][1:6, ]
top[order(top$date), ]
c(median = median(s$views), peak = max(s$views),
  ratio = round(max(s$views) / median(s$views)))
```

**Write about:** name the days. Do the biggest spikes line up with the events in
`ev`, or with something the events file does not contain? Project 2025 is
interesting precisely because its spikes are *not* all set-piece campaign
moments — some are the product of one side deliberately drawing attention to the
other side's document. What does it mean for "media effects" when a campaign
succeeds in making people look something up?

Try `Swing_state` too, and say what its shape tells you about when ordinary
people start paying attention to the mechanics of an election.

***

### Option B — How concentrated is attention?

```{r opt-b, fig.width = 8, fig.height = 5}
this_article <- "Donald_Trump"     # <- change this

s <- w[w$article == this_article, ]
s <- s[order(-s$views), ]
s$cum <- cumsum(s$views) / sum(s$views) * 100

plot(seq_along(s$cum), s$cum, type = "l", lwd = 2.5, col = "#B2182B",
     xlab = "Days of 2024, busiest first", ylab = "% of the year's attention",
     main = paste("Concentration of attention:", gsub("_", " ", this_article)))
abline(h = 50, lty = 3); abline(v = c(10, 30), lty = 3, col = "grey60")

c(top_10_days = round(s$cum[10], 1),
  top_30_days = round(s$cum[30], 1),
  days_for_half = which(s$cum >= 50)[1])
```

**Write about:** how many days of the year account for half of all the attention
this article received? Compare a principal (Trump, Harris) against a topic
(`Opinion_poll`, `Electoral_College_(United_States)`). Which is spikier, and
why would that be? Then the question worth arguing: if public attention to
politics is compressed into a handful of days, what does that imply about a
campaign event's value — and about the strategy of *creating* such days?

***

### Option C — Two candidates, side by side

```{r opt-c, fig.width = 8.5, fig.height = 5.5}
a1 <- "Kamala_Harris"              # <- change either of these
a2 <- "Donald_Trump"

s1 <- w[w$article == a1, ]; s2 <- w[w$article == a2, ]

plot(s1$date, s1$views / 1000, type = "l", lwd = 2, col = "#2166AC",
     ylim = c(0, max(c(s1$views, s2$views)) / 1000),
     xlab = "", ylab = "Pageviews (thousands)",
     main = paste(gsub("_", " ", a1), "vs", gsub("_", " ", a2)))
lines(s2$date, s2$views / 1000, lwd = 2, col = "#B2182B")
abline(v = ev$date, lty = 3, col = "grey70")
legend("topleft", gsub("_", " ", c(a1, a2)), col = c("#2166AC", "#B2182B"),
       lwd = 2, bty = "n", cex = 0.9)

# Who led on how many days, before and after Biden withdrew?
mg <- merge(s1[, c("date","views")], s2[, c("date","views")], by = "date",
            suffixes = c("_1", "_2"))
mg$phase <- ifelse(mg$date < as.Date("2024-07-21"), "before Jul 21", "after Jul 21")
table(mg$phase, ifelse(mg$views_1 > mg$views_2, a1, a2))
```

**Write about:** who "won" the attention battle, and does the answer change
depending on which half of the year you look at? Be careful with the word won —
say explicitly what you think leading in pageviews does and does not indicate.
Then connect it to Part 6: Harris led on many days after July and lost the
election. Is attention worth anything at all to a campaign, and if so, what?

***

### Option D — Ask your own question

Any article in the file is fair game. Compare a person to an institution. Look
for a day with a spike nobody has explained yet. Follow whatever nags.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a curve of public attention and say something
careful about what it does and does not show.

***

## Where the data came from

**Wikimedia Foundation Pageviews API**
(<https://wikimedia.org/api/rest_v1/#/Pageviews%20data>), daily counts for
English Wikipedia, 1 January to 31 December 2024. Wikimedia publishes these
counts under CC0, which is why this lab can ship its own copy of them.

Two choices in how the data was pulled, both of which matter:

- **Human traffic only.** The API can return all requests including known bots
  and crawlers. This file uses the `user` filter, which excludes them. Raw
  all-agent counts are badly contaminated and would show spikes that no person
  ever caused.
- **English Wikipedia only.** A large share of the world's interest in an
  American election arrives in other languages, and none of it is here.

The deeper limit is the one Part 6 is about. A pageview is a person choosing to
look something up. It is a real behaviour, publicly and consistently measured,
and it is *not* the thing Chapter 8 is ultimately asking about. Nothing in this
file tells you what anyone believed before or afterwards.

**Campaign events** in `campaign_events_2024.csv` were compiled by hand from the
public record and are deliberately kept separate from the pageview file — they
are the independent yardstick the spikes get checked against, not something
derived from the spikes themselves.

The script that fetched, checked and wrote all of this is `data/build-data.R`.
