---
title: "Key: Where Candidates Actually Go"
subtitle: "84-355 Democracy's Data | Campaign visits — retired, no slot | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
v  <- read.csv("data/campaign_visits_2024.csv", stringsAsFactors = FALSE)
v$date <- as.Date(v$date)
st <- read.csv("data/pres2024_states.csv", stringsAsFactors = FALSE)
vis <- as.data.frame(table(v$state)); names(vis) <- c("state","visits")
vis$state <- as.character(vis$state)
m <- merge(st, vis, by = "state", all.x = TRUE); m$visits[is.na(m$visits)] <- 0
m$absm <- abs(m$margin); m$per_ev <- round(m$visits / m$ev, 2)
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

Candidates allocate their scarcest resource — their own time — by multiplying a
state's size by its competitiveness, which means 25 states get nothing at all,
and the Electoral College's logic is visible in a travel schedule.

***

# The findings

```{r headline}
c(total_stops = nrow(v), states_visited = length(unique(v$state)),
  states_with_zero = sum(m$visits == 0))
head(m[order(-m$visits), c("state","ev","absm","visits","per_ev")], 7)
round(100 * sum(head(sort(m$visits, decreasing = TRUE), 7)) / sum(m$visits), 1)
```

- **377 stops, 26 states, 25 with none.**
- **Top seven states: 83.8%** of the entire campaign.
- **Pennsylvania 80** stops; California, with nearly three times the electoral
  votes, got **2**.

## The interaction — this is the teaching moment

```{r interaction}
c(all_states_visits_vs_ev      = round(cor(m$visits, m$ev), 3),
  all_states_visits_vs_margin  = round(cor(m$visits, m$absm), 3))

cl <- m[m$absm < 5, ];  lg <- m[m$ev >= 15, ]
c(among_8_closest_visits_vs_ev     = round(cor(cl$visits, cl$ev), 3),
  among_10_largest_visits_vs_margin = round(cor(lg$visits, lg$absm), 3))
```

| | correlation |
|---|---|
| visits ~ electoral votes, **all states** | +0.21 |
| visits ~ margin, **all states** | −0.41 |
| visits ~ electoral votes, **among the 8 closest** | **+0.74** |
| visits ~ margin, **among the 10 largest** | **−0.83** |

Marginally, neither variable explains much. Conditionally, each is powerful. That
is the signature of an interaction, and students do not need the word to see the
thing: **a state has to be both big and close to be worth a candidate's day.**

***

# Option A — Inside one state

## Verified output

```{r a}
head(sort(table(v$city[v$state == "Pennsylvania"]), decreasing = TRUE), 6)
c(stops = sum(v$state == "Pennsylvania"),
  distinct_cities = length(unique(v$city[v$state == "Pennsylvania"])))
```

Philadelphia 11, Pittsburgh 10, Harrisburg 6, Erie 5, then a long tail. Eighty
stops spread across 36 towns.

## What a good answer says

**3/4:** Notices the spread and connects it to persuadable voters outside the
big cities.

**4/4:** Gets the media logic. A stop in Erie buys the Erie evening news, which
reaches people no Philadelphia rally touches. Campaigns are not buying the room;
they are buying the coverage of the room. A student who distinguishes the
audience present from the audience reached has understood modern campaigning
better than the chapter states it.

**4/4, best:** notices the *mobilisation versus persuasion* split. Philadelphia
visits turn out existing Democrats; Erie and Johnstown chase people who might
switch. Both are rational and they are different strategies, visible in the same
column.

**Common stumble:** "they go where the most people are." Philadelphia has eleven
stops and Pittsburgh ten, despite Philadelphia being nearly four times larger.
Population alone does not fit.

***

# Option B — Attention per voter

## Verified output

```{r b}
head(m[order(-m$per_ev), c("state","ev","absm","visits","per_ev")], 8)
```

Wisconsin **5.00** stops per electoral vote, then Pennsylvania 4.21, Michigan
4.20 and Nevada **4.17** in a near-tie. DC appears sixth at 2.67 and is an
artefact (see below). Note that the top four are now separated by less than a
stop per electoral vote — do not let a student's ordering of them decide a grade.

The most ignored large states:

```{r b2}
m[m$ev >= 10 & m$visits <= 2, c("state","ev","absm","visits")]
```

Missouri and Washington: **zero**. California: 2, for 54 electoral votes.

## What a good answer says

**3/4:** Identifies the over- and under-attended states correctly.

**4/4:** Answers the Electoral College question with precision rather than
slogan. The defence — that it forces attention onto small states — is **half
true and half false, and the data says which half**. Nevada (6 EV) and New
Hampshire (4 EV) do get attention out of proportion to size. But Wyoming,
Vermont, the Dakotas, Delaware and Alaska got *nothing*. The system does not
reward smallness; it rewards **closeness**, and a handful of small states happen
to be close. A small safe state is the most ignored place in America.

**4/4, best:** turns it around — under a national popular vote, California's 54
electoral votes' worth of voters would be worth campaigning for, and they
currently are not. Which system ignores more people is an empirical question, and
this table is evidence in it.

**Common stumble:** treating DC's 2.67 as a real finding. It is fundraisers and
proximity, not campaigning. Part 6 warns about it; reward students who exclude it
and note why.

***

# Option C — The shape of the calendar

## Verified output

```{r c}
v$month <- format(v$date, "%Y-%m")
table(v$month)
round(100 * sum(v$date >= as.Date("2024-09-01")) / nrow(v), 1)
```

**67.4%** of all logged stops fall in September, October and the first four days
of November. Monthly counts: March 8, April 13, May 15, June 11, July 26, August
50, September 61, October 144, November 49.

## What a good answer says

**3/4:** Describes the acceleration and reports the share.

**4/4:** Handles both distortions honestly, and *weighs* them rather than
reciting them. The tracker **stops on 4 November**, the day before the election,
so only election day itself is missing — the true late concentration is still
higher than 67.4%, but barely. This is the point to press: the distortion that
used to dominate this question is now small, and the *other* one is not. The
Democratic candidate changed in late July, so the summer dip is partly a real
pause in one campaign and partly an artefact of who counted as a candidate. A
student who says "truncation is now a rounding error and the July trough is not
purely strategic" has read the data properly.

**4/4, best:** notices that this file already shows what the earlier version had
to be argued around — 49 stops in the first four days of November, on a pace of
roughly twelve a day against an eight-month average of about 1.6. The endgame
is in the data now, not inferred from its absence.

**Note for grading:** if a student is working from an older knit or an older copy
of the tracker, they will report **62%** and a 1 November end date. That is the
correct answer to the previous version of the file, not an error of reasoning.
Ask which version they read — that is Part 6's lesson landing.

***

# Option D — Their own question

Ticket-versus-ticket maps are the most productive. Note the totals are nearly
identical — **188 Democratic stops to 189 Republican** — which surprises students
who expect an obvious resource asymmetry.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is reasoning about a *conditional* relationship —
  recognising that a variable can matter enormously within a subgroup and barely
  at all overall. This is the single most transferable idea in the lab and most
  students will not have met it before. Name it in feedback when you see it.
- **Reward exclusion of DC.** Noticing that a row satisfies your filter but does
  not belong to your question is exactly the judgement Session 1 was training.

## Anticipated questions

- *"Why is DC in there at all?"* — Fundraisers, studios, and the fact that two of
  the five candidates worked there. Not campaigning for votes.
- *"Doesn't this prove the Electoral College is unfair?"* — It proves attention
  is concentrated. Whether that is unfair depends on the alternative, and a
  national popular vote would concentrate attention differently, not evenly. Push
  students to say what they think the counterfactual looks like.
- *"Is one visit really comparable to another?"* — No, and the lab says so. A
  stadium rally and a diner stop are both one row. This is the biggest weakness
  of the dataset.
- *"Why does the data stop on November 4?"* — AP's choice, and it is a choice
  that moved. Nobody was obliged to publish any of this; unlike the FEC filings
  last week, this is a news organisation's log, with no due date and no moment
  at which it is finished.
- *"Why does my copy have 324 rows?"* — They have version 50 of the tracker.
  This lab is built on version 54, which runs four days further and also
  backfilled 15 events into the window version 50 already covered. Both are
  still served. This is the single best live illustration of Part 6 the course
  has; use it.

## Connections forward

- **Session 4 (returns over time)** is the direct setup: 13 states within ten
  points in 2024 — only **8** within five — down from 33 in 1960. This lab is
  that fact's consequence. Say so explicitly, and keep the two thresholds
  straight: the lab's interaction section uses the 8.
- **Session 5 (money, message, and attention)** — money and attention concentrate on the same small set of contests.
  Worth asking whether the causation runs in a circle: money follows
  competitiveness, competitiveness attracts visits, visits attract coverage,
  coverage attracts money.
- **Session 10 (live returns, Nov 5)** — students can build the equivalent map for
  the midterms and compare. The 2026 Senate map is a very different shape.
- **Sessions 2, 3 and 14, and the Dec 3 wrap** — the Electoral College logic here becomes the
  apportionment and districting material. The question "whose vote is worth
  campaigning for?" is the same question as "whose vote is worth counting equally?"
