---
title: "Everyone Guesses Forty"
subtitle: "What Americans think each party is made of, how wrong that is, and the one thing they are most wrong about"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("perception-gap-code.R")
```

```{r setup, include=FALSE}
```

Out of every hundred Republicans, how many do you think earn more than $250,000 a year?

The average American answer is **`r p1(fn("worst_perceived"))`**. The real figure is **`r p1(fn("worst_actual"))`**.

That is the most quoted number from a study of what people believe the two parties are made of, and it is one of eight. This chapter shows all eight, then shows the thing the famous figure cannot: **who was doing the guessing**, and how much that changes the answer.

## The study

In March 2015 YouGov put eight questions to `r n(fn("n_resp"))` Americans. For each party it named four groups that party is stereotyped around, and asked what percentage of that party's supporters belong to it. Four Democratic groups, four Republican ones, `r n(fn("n_answers"))` answers in all.

The study was run for two researchers, Douglas Ahler and Gaurav Sood, who published the results in 2018. **The survey itself is theirs and YouGov's, and nobody else has it.** What is public is a copy of part of it, which they deposited so their results could be checked. That is a different object from the study, and the last section of this chapter measures the difference.

## Everyone is wrong, and by a lot

```{r items-table}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. The eight items.** The dark dot is the true share, the light dot is the average guess, and the bar joining them is the gap.

Every single group is overestimated. The average gap is **`r p1(fn("gap_mean"))` points**.

The sizes are not alike. Earning $250K+ is out by **`r p1(fn("worst_gap"))` points**, which is the number everybody quotes. Southern is out by **`r p1(fn("best_gap"))`**, which nobody quotes. They are in the same figure.

**The pattern in those sizes is worth having before anything else.** The groups people are most wrong about are the rarest ones. Lesbian, gay or bisexual Democrats are `r p1(it$actual_pct[it$group == "LGB"])`% of the party and people said `r p1(it$perceived_pct[it$group == "LGB"])`%. Southerners are `r p1(it$actual_pct[it$group == "Southern"])`% of the Republican party and people said `r p1(it$perceived_pct[it$group == "Southern"])`%. Where a group really is common, the guess is close.

## And you are worse about the party you are not in

This is the study's actual argument, and Figure 1 cannot show it. Everyone in that figure is averaged together: Democrats, Republicans and people who are neither, all guessing about both parties at once.

The file records each person's own party, so the same eight items can be asked twice.

```{r party-prep, include=FALSE}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The same eight items, split by who answered.** The blue dot is what Democratic respondents said, the red dot what Republican respondents said, and the grey tick is the truth. The top four rows ask about groups of Democrats and the bottom four about groups of Republicans.

**The far dot changes colour halfway down**, and that is the finding.

```{r party-table}
```

Asked about their own party, people overestimate by **`r p1(fn("own_gap"))` points**. Asked about the other one, **`r p1(fn("other_gap"))`**. The worst single cell in the study is `r f("worst_cell_who")` estimating how many `r f("worst_cell_about_word")` are `r f("worst_cell_asked")`: out by **`r p1(fn("worst_cell_gap"))` points**.

You cannot meet a political party. You learn what is in the other one from people describing it to you, and what you end up with is a picture assembled out of the descriptions.

**Counting leaners as partisans is a decision here**, and it is worth saying what it costs. People who call themselves independent and then admit to leaning are counted as partisans above, which is the [party-id chapter](../party-id/party-id-brief.html)'s subject. Push them back out and the two numbers become `r p1(fn("own_gap_strict"))` and `r p1(fn("other_gap_strict"))`. A smaller gap, the same direction, the same conclusion.

> **Before you read on, write two numbers down.** The average gap in Figure 1 is
> about twenty points. **First: of the `r n(fn("n_answers"))` answers behind it,
> what share do you think were round numbers ending in a zero?** **Second: if
> you swapped the average guess for the middle guess, the value half the
> answers fall below, how much of that twenty-point gap do you think would
> survive?** Commit to both before turning the page.

## What the average is an average of

Each point in Figure 1 is a thousand people reduced to one number. Here is what the thousand look like.

```{r fig3-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r fig3-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Every answer, at one-point bins.** In the browser you can click any single item and see its own thousand answers against its own true share.

**`r p1(fn("mult10"))`% of the answers end in a zero.** `r p1(fn("mult5"))`% end in a zero or a five. Only `r p1(fn("not_mult5"))`% use any of the other eighty values on the scale, and exactly `r n(fn("nonint"))` answer in the whole study is not a whole number: somebody typed 0.5.

```{r heap-table}
```

And `r p1(fn("eq50"))`% of all answers are exactly **50**. No group here has a true share anywhere near a half. Fifty is what a person types when they do not know and the box will not let them say so.

**That pile is lumpy and it leans.** A few people answering 90 pull the average up and nothing pulls it back, so the average sits well above the middle of the answers.

```{r fig4-static, eval=!knitr::is_html_output(), fig.height=4.6}
```

```{r gap-table}
```

**Figure 4. The same eight gaps, twice.** Measured from the average guess, and from the middle one.

**Measured off the middle, the average gap is `r p1(fn("gap_median"))` points rather than `r p1(fn("gap_mean"))`.** Seven of the eight shrink and one grows. Neither number is the wrong one. The average answers *what does the country believe on average*, which is what the study asked. The middle answers *what does a typical person believe*. They are `r p1(fn("gap_mean") - fn("gap_median"))` points apart and the difference is never mentioned.

## Everyone guesses forty

The true shares span **`r p1(fn("span_actual"))` points** across these eight groups. The average guesses span **`r p1(fn("span_perceived"))`**.

People are not answering at random. The guesses do line up with reality, at a correlation of `r f("r_actual_perc")`, on a scale where 1 is a perfect straight line. But they are squeezed into a band, and the band sits high: six of the eight average guesses fall between 38 and 42.

That is the chapter's title and it is also a warning about the measure. Anyone answering near forty about a group that is really at `r p1(fn("worst_actual"))`% is `r p1(fn("worst_gap"))` points wrong automatically. **A list of these gaps sorted from largest to smallest is very close to a list of the groups sorted from rarest to commonest.** The ranking is telling you about the groups as much as about the people guessing.

## The survey, and the copy of it we have

Everything above comes from a file of `r n(fn("n_carried"))` columns. The survey it was taken from had many more.

That is not a complaint. It is the ordinary condition of published research data: a researcher deposits what their results rest on. The questionnaire as administered, the panel the thousand people were drawn from, everyone who started and gave up — none of that is deposited, and none of it is anywhere a reader can go. **YouGov does not publish this study, so the deposit is the only route to any of it.**

The interesting part is that the deposit ships its own codebook, so the two can be set against each other.

```{r coverage-table}
```

`r n(fn("n_unshipped"))` columns are described in the codebook and are not in the file. `r n(fn("n_undoc"))` are in the file and are not in the codebook, and one of those is `pid7`, **the column the whole of Figure 2 rests on**. The codebook says as much in its own first line: it is not comprehensive.

**The rule of thumb.** When a finding rests on a public file, the file is not the study. Ask what was collected, then ask what was deposited, and do not assume the second is the first with fewer rows.

## If you ran this in a room

Here is the block, ready to append to the class survey. Put it after the questions that ask who *you* are, because these ask who you think everybody else is, and that order keeps the second set from colouring the first.

```{r form-table}
```

> The last eight questions ask you to estimate. Nobody expects you to know
> these — guess. Give a whole number between 0 and 100. Please do not look
> anything up: a wrong guess is useful data and a correct look-up is not.

Set each item as a short answer, validated as a number between 0 and 100, shuffled within the block, all eight required. Two things there are decisions rather than formatting, and this chapter is the evidence for both.

**The wording asks for a count out of a hundred, not a percentage.** The original asked for a percentage and `r p1(fn("mult10"))`% of the answers came back ending in a zero. Asking somebody to count out of a hundred will not remove that, but it invites a finer answer from anyone who does not think in percentages.

**Score the room against two things, not one.** Against the true shares is whether the room is wrong. Against the national guesses is the better question: is a room of students less wrong than the country, or wrong in a different direction?

**And the actuals are a decade old.** They describe the parties as of 2015. Recomputing them from a current survey is a real exercise, and the distance between the 2015 figure and today's is a finding rather than a correction.

## What this chapter cannot tell you

**No room has answered these questions yet.** Everything here is the national study. The class comparison is set up and not run, and nothing here is a result about any group of students.

**Why people answer near forty is not settled.** It could be that they picture the party through its stereotypes, which is the study's argument. It could be that somebody with no information gives a number in the middle of the scale, which would produce the same pattern without anyone being misled about anything. This chapter can show the compression. It cannot say what causes it.

**The true shares are estimates too.** Each was computed by the authors from the American National Election Studies, the Current Population Survey or a Pew report, and each carries a standard error. The widest is `r f("se_group")`: `r p1(fn("se_actual"))`%, honestly somewhere between `r p1(fn("se_lo"))`% and `r p1(fn("se_hi"))`%. For most items that is small against a twenty-point gap. It still means the gap is a distance between two estimates, and "the true share" is doing work the word "estimate" would not do.

**The panel is not the country.** YouGov recruits people who volunteer to take surveys online, and the case weight is what stands between that and a national statement. The [surveys-source chapter](../surveys-source/surveys-source-brief.html) sets out what a sample like this can establish, and the [ces-class chapter](../ces-class/ces-class-brief.html) shows what happens when the weights are turned off.

**Eight groups is a small and chosen list.** The authors picked groups each party is stereotyped around, which is the right list for their question and a list built to produce large gaps. A different eight, say Democrats who are veterans or Republicans who rent, would give different numbers, and this chapter has no way to say how different.

## Sources

- The data is a **YouGov survey of `r n(fn("n_resp"))` American adults, fielded
  in March 2015** for Douglas J. Ahler and Gaurav Sood. YouGov does not publish
  it.
- The copy read here is their replication deposit for "The Parties in Our
  Heads: Misperceptions About Party Composition and Their Consequences," *The
  Journal of Politics* 80(3), 2018: 964–981, on Harvard Dataverse,
  doi:10.7910/DVN/CLMQ8E. `r n(fn("n_carried"))` columns of the study, plus the
  eight true shares and the published averages.
  <https://doi.org/10.7910/DVN/CLMQ8E>
- The true shares in that deposit were computed by the authors from the
  American National Election Studies, the Current Population Survey and a Pew
  Research Center report on religion.
- The averages recomputed here from the `r n(fn("n_answers"))` individual
  answers reproduce the published ones to five decimal places, which is the
  check this chapter runs before it writes anything else.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`by_party.csv`](data/derived/by_party.csv), [`coverage.csv`](data/derived/coverage.csv), [`dist.csv`](data/derived/dist.csv), [`heaping.csv`](data/derived/heaping.csv), [`items.csv`](data/derived/items.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `items.csv` puts `actual_pct` beside `perceived_pct` for every group asked
  about. Which group is misjudged by the widest margin, and is it guessed too
  high or too low?
- `by_party.csv` splits every guess by who is doing the guessing. Do people
  judge their own party more accurately than the other one? Check every item
  before you answer.
- `heaping.csv` counts the answers that are multiples of ten, multiples of
  five, and exactly 50. Work out what share are multiples of five. What does
  that tell you about how precise a survey percentage really is?
- `items.csv` gives both a mean and a median for the same question, and they
  are far apart. Look at `dist.csv` to see why. Which one would you report?

```{r ai-prompt, results='asis'}
```
