---
title: "Confidence in Institutions, GSS 1973-2024"
type: brief
subtitle: "Fifty-one years of one question, and the summary that does not survive it"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("gss-confidence-code.R")
```

```{r setup, include=FALSE}
```

Americans have lost faith in everything. You have read that sentence, or one of its cousins: trust has collapsed, institutions are in crisis, nobody believes in anything anymore. It is repeated so often that it reads as a fact rather than a claim. Is it true — did confidence fall in every institution, in which ones, and since when?

Answering needs something rare: the same question, asked the same way, for long enough to watch. Survey wording usually drifts, sponsors lose interest, and a question asked in 1975 is rarely asked in 2024 in a form anybody can compare. One battery of questions survived half a century intact, and most of what this brief can do rests on that fact.

## The test

The test is the [General Social Survey](https://gss.norc.org/), run by NORC at the University of Chicago and funded largely by the National Science Foundation. Since `r Y1` its interviewers have read the same sentence to a national sample of American adults, `r length(unique(cf$year))` times across `r Y2 - Y1` years:

> *"I am going to name some institutions in this country. As far as the people
> running these institutions are concerned, would you say you have a great deal
> of confidence, only some confidence, or hardly any confidence at all in
> them?"*

Nothing about the question is clever; its entire value is that nobody changed it. No election ever grades an answer like this — there is no true national figure for confidence, which is the standing condition of the [Public Opinion cluster's data chapter](../anes/anes-brief.html). When nothing outside the survey can grade the answer, the method is the whole credential, and this survey's method held still. A survey whose product is the comparison across time has every incentive to leave its wording alone, and the GSS acted on that incentive for fifty years.

Two features of the wording should be fixed in mind before any number arrives. **It asks about "the people running" the institution**, not the institution itself, so somebody who reveres the Supreme Court and distrusts its current members is recorded as distrusting. And **there are only three options with no neutral middle**, so a lukewarm respondent has to choose "only some." That is why the middle box is the largest almost everywhere, and why the interesting movement is at the two ends.

The battery also asks about `r nrow(it)` institutions in one interview, of the same respondents. That makes it an unusually fair test of the claim in the title: if confidence moved as a bloc, these `r nrow(it)` series should mostly move together.

## The data

The cumulative file is 570 megabytes of Stata holding 75,699 respondents and about six thousand variables. One row is one person interviewed, once, in one year. Here are the first five people interviewed in `r Y1`, in six of those columns:

| year | id | conmedic | conlegis | conarmy | wtssps |
|---:|---:|---:|---:|---:|---:|
| 1973 | 1 | 2 | 2 | 1 | 0.7229722 |
| 1973 | 2 | 1 | 1 | 1 | 1.4825919 |
| 1973 | 3 | 2 | 2 | 2 | 0.7921651 |
| 1973 | 4 | 2 | 2 | 1 | 0.7814606 |
| 1973 | 5 | 2 | 2 | 2 | 0.7412959 |

The meaning lives in the codebook, not the file. `conlegis` is confidence in Congress, coded 1 for a great deal, 2 for only some, 3 for hardly any; `wtssps` is the survey weight the respondent carries. **A `2` is not a quantity.** It is the middle of three ordered boxes, so every number below is a share of people who picked a box, never an average of the codes. Alongside the three answers sit twelve distinct ways of not answering: don't know, refused, inapplicable, skipped on web, and more. The build turns all twelve into the same blank.

The build reduces all of that to `r n(nrow(cf))` rows. Here is what those 1,504 people become:

```{r clean-gss}
```

One row per institution per survey year, in `confidence.csv`: the institution, its GSS `variable` name, the `year`, the number who `answered`, and the weighted shares `great_deal`, `only_some` and `hardly_any`, which sum to 100. A second table, `institutions.csv`, gives each institution its `first_year`, `last_year`, `survey_years` and the `change` across the series. What one row stands for has changed, from a person to an institution in a year, and there is no going back. Nothing below can ask whether the people who distrusted Congress also distrusted the press, because the rows that knew are gone.

Four decisions sit inside that one row. **The denominator is who answered, not who was asked.** The denominator is the bottom number, the figure the percentage is divided by: `r n(gg("Congress", Y1, "n"))` of the 1,504 people interviewed in `r Y1` answered the Congress item, and the rest are simply not in the percentage. **Every percentage is weighted**, and weighting means counting some answers more than others, to make a sample that does not look like the country look like it. The build uses `wtssps` throughout, the only weight spanning every year in the file, so no seam between two weights lands inside the argument.

**An institution-year with fewer than fifty answers is dropped**, not shown small. And **twenty columns of six thousand were kept**: everything else about these respondents stays in the source file and cannot be recovered from these tables.

Two irregularities are the file's own. The survey ran yearly into the 1990s and every other year after that, so the longest gap between readings is `r max(diff(sort(unique(cf$year))))` years, across the pandemic. And one institution, `r tolower(it$institution[it$first_year != Y1])`, joined the battery in `r min(it$first_year[it$first_year != Y1])`, so its series is shorter than the rest.

> **Before you read on, commit to three answers.** Of the `r nrow(it)`
> institutions in the battery, how many fell between their first reading and
> their last — all `r nrow(it)`? Ten? Seven? How many rose? And which single
> institution fell the furthest? Write the three down. A guess you have not
> written down is one you will quietly revise once you see the answer.

## What it says about democracy

Start with the series everybody quotes. Confidence in the people running Congress fell from `r pc(gg("Congress",Y1,"great_deal"))`% with a great deal of confidence in `r Y1` to `r pc(gg("Congress",Y2,"great_deal"))`% in `r Y2`. The more striking number is at the other end: **"hardly any" went from `r pc(gg("Congress",Y1,"hardly_any"))`% to `r pc(gg("Congress",Y2,"hardly_any"))`%**, from about one respondent in seven to a clear majority.

```{r bands-d3, results='asis', eval=knitr::is_html_output()}
```

```{r bands-static, eval=!knitr::is_html_output(), fig.height=3.9}
```

**Figure 1. Confidence in the people running Congress, drawn as a composition rather than a level.** The three answer shares at every survey year, stacked to 100% — the right form here, because every respondent picks exactly one of three boxes, so what moves is not a level but a mix. "Hardly any" overtook "a great deal" in `r CROSS` and passed half of all respondents in `r MAJ`, marked by the dashed line. Even the lukewarm middle fell, from `r pc(MID1)`% to `r pc(MID2)`%.

At this point the summary is sitting there waiting to be written: *Americans have lost confidence in their institutions.* But the claim is about institutions in general, and only one has been looked at. The file holds `r nrow(it)` of them, asked of the same respondents. Here they are, ranked by how far each moved between its first reading and its last:

```{r changes}
```

The largest decline is not a political institution at all. **Medicine is down `r pc(abs(min(it$change)))` points**, further than Congress. Organized religion, the press, television, major companies, banks and education all fall by double digits. And the **military rose `r pc(it$change[it$institution=="Military"])` points**, the only substantial increase anywhere in the file. Most readers name Congress for the steepest fall, and most put the number of risers at zero; neither holds.

A ranked list of thirteen differences invites a second mistake, which is reading the small ones as movements at all. Every number in that table is an estimate from a finite sample, and the difference between two estimates carries the uncertainty of both.

```{r dumb-d3, results='asis', eval=knitr::is_html_output()}
```

```{r dumb-static, eval=!knitr::is_html_output(), fig.height=4.8}
```

**Figure 2. First reading against last, with the uncertainty on each.** Share with a great deal of confidence in `r min(STARTS)` (hollow circle; `r max(STARTS)` for banks) against `r Y2` (filled), one row per institution, with a 95% interval on each single reading. Rows whose change does not clear the uncertainty of its two readings are drawn pale and marked *ns*. The intervals treat each year as a simple random sample; the GSS is clustered, so the true intervals are wider, which makes the small changes less secure rather than more.

Written out honestly: **`r wd(NFELL)` institutions fell, `r wd(NROSE)` rose, and `r wd(NSIG)` did not move.** Organized labor's `r pc(it$change[it$institution == "Organized labor"])` points sits inside its own uncertainty, so the right description is *no detectable change*. Sampling error is the wobble that comes from asking a sample rather than everyone. Everything else moved by more than it can explain, including the scientific community's `r pc(abs(it$change[it$institution == "Scientific community"]))`-point fall, the smallest real decline in the set.

The exception deserves its story, because a single rise could be a fluke of the two endpoints Figure 2 happens to compare. Follow the military's actual series:

```{r military}
```

It is not a steady climb. It peaks at `r pc(max(mil$great_deal))`% in `r mil$year[which.max(mil$great_deal)]`, the year of the Gulf War, falls back, then rises again. That cuts directly against the long-slow-decline framing the Congress series invites: **confidence responds to events**, and it went up sharply when an institution did something visible and widely regarded at the time as successful. Whatever drives the other twelve series is capable of moving in both directions.

Where everything stands now completes the picture:

```{r current}
```

Sorted by the net column, a pattern appears that the individual series conceal. The institutions people still trust do something technical and visible — the military, the scientific community, medicine. The ones at the bottom are the ones whose entire function is contestation: Congress, the press, television, the executive branch. Congress now stands at `r pc(gg("Congress",Y2,"great_deal"))`%, the lowest figure for anything asked about, and the Supreme Court at `r pc(gg("U.S. Supreme Court",Y2,"great_deal"))`%.

```{r lines-static, eval=!knitr::is_html_output(), fig.height=5}
```

```{r d3-lines, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. All `r nrow(it)` institutions, every survey year from `r Y1` to `r Y2`.** Share with a great deal of confidence in the people running each. The band narrows as it moves: institutions that began between `r pc(min(cf$great_deal[cf$year == Y1]), 0)`% and `r pc(max(cf$great_deal[cf$year == Y1]), 0)`% in `r Y1` end between `r pc(min(cf$great_deal[cf$year == Y2]), 0)`% and `r pc(max(cf$great_deal[cf$year == Y2]), 0)`%, with the military the only line finishing above where it started.

So the sentence this brief opened with is not right as written. Something did happen to confidence in American institutions. But it is not uniform, it is not confined to government or even to politics, and at least one institution moved the other way. The largest loss belongs to medicine and the smallest to science, neither of which is a branch of government. That is a strange result for a story usually told about politics, and the unevenness is the finding.

## What this brief cannot tell you

Whether institutions became less trustworthy, or people became less trusting, or the phrase "a great deal of confidence" drifted in meaning across fifty years. The question records a report, not a relationship, and no amount of this series settles which changed. Citrin and Stoker (2018) add that trust measures move with partisanship: some of what looks like erosion about Congress is people disliking the party currently running it, and those people flip when control changes.

Whether the decline was warranted. The `r Y1` baseline sits immediately after Watergate and Vietnam, and institutions went on to earn skepticism through visible failures. Low confidence in a body with a twenty percent approval rating may be accurate perception rather than civic decay, and the data cannot tell those apart.

Whether Americans changed their minds, or the Americans who held the old view died and were replaced by people who never held it. A cross-section samples new people each round; only a panel that re-asks the same people could separate the two, and this design confuses them.

Who moved. The file built here carries no breakdowns by party, age, race or education, which is where most of the interesting variation in this literature lives. And the GSS changed how it collected answers in 2021, moving to web and mail during the pandemic. That is exactly the kind of break that can look like a trend, and it sits directly underneath every `r Y2` reading above.

## What you should have learned

- One unchanged question, asked `r length(unique(cf$year))` times across `r Y2 - Y1` years, is what makes the claim testable at all; almost no other social data holds still that long.
- The fall is real but wildly uneven: `r wd(NFELL)` of `r nrow(it)` institutions fell by more than sampling error can explain, `r wd(NROSE)` rose, and `r wd(NSIG)` did not move.
- The steepest loss is medicine's `r pc(abs(min(it$change)))` points, not Congress's — the decline is not confined to government, or even to politics.
- The military's series peaks at `r pc(max(mil$great_deal))`% in `r mil$year[which.max(mil$great_deal)]` and rises overall, so whatever drives the declines is not an irreversible drift in national character.
- A percentage is a stack of decisions: who counts as answering, which weight, which cells are too thin to print. Every one of them was somebody's choice.

## Extensions

- `institutions.csv` gives every institution a `change` across the series. Sort by it. Which institutions lost the most confidence, and which gained?
- `institutions.csv` also has `first_year`, `last_year` and `survey_years`. Find the institutions measured for fewer years than the rest. Does a shorter series make a bigger change easier to claim?
- `confidence.csv` keeps `great_deal`, `only_some` and `hardly_any` separately. Pick an institution whose `great_deal` fell. Did those people move to the middle, or all the way to the bottom?
- Find a year in `confidence.csv` where many institutions move together, using the `year` column. Then go and read what happened that year. Is the survey measuring the institutions, or the mood?
- **Stretch:** Gallup has asked its own confidence battery since 1973, with different wording and a fourth answer option. Pull its Congress series from <https://news.gallup.com/> and set it beside this one. Where the two disagree, which difference — wording, options, sample — would you blame first?
- **Stretch:** the GSS Data Explorer at <https://gssdataexplorer.norc.org/> can cross any confidence item by party identification. Pick one institution from Figure 3 and find out whether its decline belongs to one party or both.

## Sources

**Data**

- General Social Survey cumulative file, NORC at the University of Chicago,
  `r Y1`–`r Y2`. <https://gss.norc.org/> — `data/build-data.R` here
  reduces the cumulative file to the `r n(nrow(cf))` institution-years read
  here, using the `wtssps` weight throughout.

**Academic literature**

- Citrin, J. and Stoker, L. (2018). "Political Trust in a Cynical Age." *Annual
  Review of Political Science* 21: 49–70.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`confidence.csv`](data/derived/confidence.csv), [`institutions.csv`](data/derived/institutions.csv)

```{r on-mark-halo, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```
