---
title: "Lab 2: Eighty-Nine People"
subtitle: "84-355 Democracy's Data | Week 2 — The census"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

**This is the second family: the census.**

Last week's returns had no producer. Fifty-one jurisdictions certified
separately, somebody scraped them together, and the file inherited every
inconsistency underneath. The census is the opposite in every respect — **one
agency, one number, a constitutional mandate and a legal deadline.** There is
exactly one right answer and the Bureau publishes it.

Watch what that does and does not buy you. By the end of the hour you will have
a number produced with more authority than anything else in this course, and you
will be able to show that it is not certain enough to have decided what it
decided.

***

The Constitution says the House shall be apportioned among the states "according
to their respective Numbers." It does not say how. Dividing 435 seats among 50
states will almost never come out in whole numbers, and every method of handling
the remainder is a choice that helps somebody.

Congress has changed its mind about that choice five times. Since 1941 the
method has been **Huntington–Hill**, and today you are going to run it — the
actual arithmetic, on the actual 2020 numbers, and get the actual House of
Representatives out the other end.

Then we will look at how close it was.

Everything is base R. Work in order.

***

# Part 1: The numbers that decide

```{r read}
ap <- read.csv("data/apportionment_2020.csv", stringsAsFactors = FALSE)
dim(ap)
head(ap, 5)
```

Fifty states. The columns:

- `app_pop` — the **apportionment population**, which is what the arithmetic uses
- `seats` — what each state actually got
- `overseas` — federal employees serving abroad, assigned back to a home state
- `resident_pop` — everybody actually living in the state

```{r totals}
c(states = nrow(ap), seats = sum(ap$seats))
```

Note that apportionment population is **not** the population of the state:

```{r overseas}
head(ap[order(-ap$overseas), c("state", "resident_pop", "overseas", "app_pop")], 5)
```

> Soldiers stationed in Germany are counted in the state they left. That is a
> policy decision, it moves tens of thousands of people between states, and we
> will come back to it in Part 5 — because it turns out to be larger than the
> margin that decided the last seat in the House.

***

# Part 2: The obvious method, and why it fails

The natural approach: work out each state's fair share and round it.

```{r quota}
ap$quota <- 435 * ap$app_pop / sum(ap$app_pop)
head(ap[order(-ap$quota), c("state", "app_pop", "quota", "seats")], 6)
```

New York's fair share is 26.56 seats. It cannot have 0.56 of a seat, and
somebody has to decide which way that goes.

So round it:

```{r naive}
ap$rounded <- round(ap$quota)
c(seats_if_we_just_round = sum(ap$rounded), target = 435)
```

The total comes out right. That is luck, not design — nothing forces the
fractions to cancel, and in other decades they have not. But look at *which*
House rounding produces:

```{r naive-diff}
ap[ap$rounded != ap$seats, c("state", "quota", "rounded", "seats")]
```

> **Four seats sit in different states.** Rounding hands an extra seat to New
> York and Ohio, and takes one each from Rhode Island and Montana.
>
> That is not a rounding artefact, it is a direction. Big states have big
> fractions to round up. Small states have small ones. A rule that looks
> perfectly neutral — "round to the nearest whole seat" — systematically moves
> seats from small states to large ones.

Hold on to those four names. You will meet them again in Part 7.

There is also a floor. Three states are not entitled to a full seat at all:

```{r wyoming}
ap[ap$quota < 1, c("state", "app_pop", "quota", "seats")]
```

The Constitution guarantees every state at least one Representative, so Wyoming's
0.76 of a seat becomes a whole one. Somebody else has to give that up.

***

# Part 3: Huntington–Hill, in eleven lines

The method used since 1941. Give every state one seat, then hand out the
remaining 385 one at a time, each to whichever state has the highest **priority
value**:

$$\text{priority} = \frac{\text{population}}{\sqrt{n(n+1)}}$$

where $n$ is the number of seats that state already holds. A state with few
seats and many people has a high priority; each seat it wins lowers its own
priority for the next one.

```{r hh}
pop   <- setNames(ap$app_pop, ap$state)
seats <- setNames(rep(1L, nrow(ap)), ap$state)   # every state starts with one
winners <- character(0)

for (i in 1:385) {
  priority <- pop / sqrt(seats * (seats + 1))
  w <- names(which.max(priority))
  winners[i] <- w
  seats[w] <- seats[w] + 1L
}

sum(seats)
```

That is the House. Did we get it right?

```{r check}
identical(as.integer(seats[ap$state]), as.integer(ap$seats))
```

> **TRUE.** Eleven lines of arithmetic reproduce the United States House of
> Representatives, seat for seat. There is no fudge factor and no committee.
> Once the census is counted, the House allocates itself.

***

# Part 4: The last seat

The seats were handed out in order. Look at the end of the queue.

```{r last-seats}
tail(data.frame(seat = 51:435, went_to = winners), 6)
```

Seat **435** — the last one — went to **Minnesota**.

Now: who was next in line, and how close were they?

```{r near-miss}
priority <- pop / sqrt(seats * (seats + 1))
sort(priority, decreasing = TRUE)[1:3]
```

New York was the state that would have received seat 436, had one existed.

```{r how-close}
mn <- ap[ap$state == "Minnesota", ]
ny <- ap[ap$state == "New York", ]

winning_priority <- mn$app_pop / sqrt(mn$seats * (mn$seats - 1))
needed <- winning_priority * sqrt(ny$seats * (ny$seats + 1))

c(new_york_had    = ny$app_pop,
  new_york_needed = ceiling(needed),
  short_by        = ceiling(needed - ny$app_pop))
```

> **Eighty-nine people.**
>
> New York needed 20,215,840 residents to take that seat from Minnesota. It was
> counted at 20,215,751. Eighty-nine more people — a large lecture hall — and
> the House of Representatives, and the Electoral College, would have been
> different for ten years.

***

# Part 5: What eighty-nine people means

Three things sit behind that number, and each one is bigger than it.

**The census is an estimate.** The Bureau's own post-enumeration survey concluded
the 2020 count missed some states and double-counted others, by margins in the
hundreds of thousands. The 89 is smaller than the error bar on the thing that
produced it.

That sentence is the whole family in one line. A single authoritative producer
removes the problems you met last week — no stitching, no mismatched keys, no
county that changed meaning — and it does not make the number true. **Authority
and certainty are different things**, and a file can have all of the first and
not enough of the second.

**The overseas adjustment.**

```{r overseas-vs-margin}
c(new_york_overseas  = ny$overseas,
  minnesota_overseas = mn$overseas,
  the_margin         = 89)
```

> New York had **14,502** people counted from abroad; Minnesota **3,258**. The
> rule that assigns overseas federal employees back to a home state moved a
> hundred and sixty times more people than the margin that decided the seat.
> Change that rule and you change the House.

**And a seat is not a small thing.**

```{r what-a-seat-is}
ap$people_per_seat <- round(ap$app_pop / ap$seats)
rbind(head(ap[order(ap$people_per_seat), c("state","seats","people_per_seat")], 3),
      head(ap[order(-ap$people_per_seat), c("state","seats","people_per_seat")], 3))
```

Even after all this arithmetic, districts are not equal. A seat is also a vote in
the Electoral College, which is why the same 89 people would have moved a
presidential elector too.

***

# Part 6: What this does not settle

**Huntington–Hill is one choice among several.** Jefferson's method favours large
states, Adams's favours small ones, Hamilton's produced the notorious *Alabama
paradox* — where adding seats to the House cost Alabama one. Congress adopted
equal proportions in 1941 after a decade of argument, and the argument was
partisan.

**435 is not in the Constitution.** It has been fixed by statute since 1929. The
House grew with the country for 140 years and then stopped.

**And the count is not the population.** It is a very good estimate of it,
produced under a deadline, by an agency that publishes its own uncertainty. That
is Week 12's subject.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — Resize the House

```{r opt-a}
house_size <- 435            # <- change this. Try 436, 450, 535, 600.

s <- setNames(rep(1L, nrow(ap)), ap$state)
for (i in seq_len(house_size - nrow(ap))) {
  p <- pop / sqrt(s * (s + 1)); s[names(which.max(p))] <- s[names(which.max(p))] + 1L
}

now <- data.frame(state = ap$state, actual = ap$seats,
                  alt = as.integer(s[ap$state]))
now$change <- now$alt - now$actual
now[now$change != 0, ]

c(house_size = house_size, largest_district = max(round(pop/s)),
  smallest_district = min(round(pop/s)))
```

**Write about:** who gains and who loses as the House grows? Try 436 first — one
extra seat — and say who gets it. Then try 600. The "Wyoming Rule" would set the
House so that no district is smaller than the least populous state; find roughly
what size that implies and say what it would cost. Then the real question: the
House has been frozen at 435 since 1929, while the country has more than tripled.
Who benefits from it staying there?

***

### Option B — How close was everybody else?

```{r opt-b, fig.width = 7.5, fig.height = 5.5}
n_show <- 12                 # <- change this

final <- pop / sqrt(seats * (seats + 1))          # priority for one more seat
gap <- sort(final, decreasing = TRUE)[1:n_show]

par(mar = c(5, 10, 4, 2))
barplot(rev(gap), names.arg = rev(names(gap)), horiz = TRUE, las = 1,
        col = "#54278F", border = NA,
        xlab = "Priority value for the next seat",
        main = paste("The", n_show, "states closest to another seat"))
abline(v = max(final), lty = 2)

round(gap)
```

**Write about:** New York missed by 89 people. How much would the *next* state
have needed? Work out roughly how many people separate second place from third,
and from fourth. Does the near-miss look like a freak accident or like something
that happens every decade? Then say what a state government should do about it —
and whether "count harder" is a real answer.

***

### Option C — Who would a different method favour?

```{r opt-c}
# Webster's method: the same machinery, a different divisor.
# Huntington-Hill uses sqrt(n(n+1)) — the geometric mean.
# Webster uses n + 0.5 — the arithmetic mean.
divisor <- function(n) n + 0.5      # <- change to sqrt(n*(n+1)) for Huntington-Hill

s <- setNames(rep(1L, nrow(ap)), ap$state)
for (i in 1:385) {
  p <- pop / divisor(s); s[names(which.max(p))] <- s[names(which.max(p))] + 1L
}

cmp <- data.frame(state = ap$state, huntington_hill = ap$seats,
                  this_method = as.integer(s[ap$state]))
cmp$change <- cmp$this_method - cmp$huntington_hill
cmp[cmp$change != 0, ]
```

**Write about:** which states move, and in which direction? You should recognise
all four from Part 2 — and that is the point. **Rounding the quota *is* Webster's
method**, arrived at by accident: rounding at the half is exactly what a divisor
of $n + 0.5$ does. The obvious thing you tried at the start turns out to be one
of the named contenders Congress rejected.

Huntington–Hill's geometric mean is very slightly kinder to small states than
Webster's arithmetic mean; check whether the states that change fit that
description. Then the question Congress actually argued about in the 1930s: if
two defensible methods give different answers, how should a country choose
between them, knowing in advance who each one helps?

***

### Option D — Ask your own question

Follow whatever nags. The 1920s are worth a look: Congress failed to reapportion
at all after the 1920 census, the only time that has happened.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a piece of arithmetic and see the political
decision inside it.

***

# What today was really about

Two families down, and they fail in opposite ways.

| | **Returns** (Week 1) | **The census** (this week) |
|---|---|---|
| Who makes it | nobody in particular | one agency, by law |
| How it breaks | stitching fifty-one sources together | a single number carrying uncertainty it cannot show you |
| What to check first | does it match the certified count? | what was counted, on what date, with what error? |
| What you saw | a scraped file 7,781 votes short | a seat decided by 89 people, inside an error bar of hundreds of thousands |

Neither family is more reliable than the other. They are **unreliable in
different directions**, and the failure of one is invisible using the habits that
catch the other. Checking a census file against a second source will not help
you — there is no second source. Reading the Bureau's own uncertainty will.

Thursday's reading is six pages from the Bureau on exactly that. It is the most
useful methods writing in the course, and you will use it again in Week 13, when
a legal threshold gets applied to an estimate that has a margin of error printed
next to it.

***

***

## Where the data came from

**U.S. Census Bureau**, 2020 Census Apportionment Results
(<https://www.census.gov/data/tables/2020/dec/2020-apportionment-data.html>),
Table 1 for apportionment population and seats, Table 3 for the overseas
population folded into each state's count.

This is the official file. The seats in it are the seats that exist.

Two things worth registering. **Apportionment population is not resident
population** — it includes federal employees serving overseas, assigned to a home
state under a rule that has itself been litigated. And the whole calculation runs
on a **count**, which is an estimate with published uncertainty; the margin that
decided the 435th seat is far smaller than that uncertainty.

The script that downloaded, parsed and checked this file is `data/build-data.R`.
It re-runs Huntington–Hill against the Bureau's own seat column and stops if the
two disagree — which is the only real way to know a file like this arrived
intact.
