---
title: "Data Session 2 Key, part 1: Eighty-Nine People"
subtitle: "84-355 Democracy's Data | Session 2 — The decennial census | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
ap  <- read.csv("data/apportionment_2020.csv", stringsAsFactors = FALSE)
pop <- setNames(ap$app_pop, ap$state)
hh <- function(size) {
  s <- setNames(rep(1L, length(pop)), names(pop))
  for (i in seq_len(size - length(pop))) {
    p <- pop / sqrt(s * (s + 1)); w <- names(which.max(p)); s[w] <- s[w] + 1L
  }
  s
}
base <- hh(435)
```

All Part 7 code is given to students. Grade the interpretation.

**Framing.** This is the **census** lab among the four families of democracy
data, and it is deliberately the mirror of Session 1. Returns fail by being
stitched from fifty-one sources; the census fails by being a single
authoritative number that carries uncertainty it cannot show you. The closing
table sets them side by side.

The line to land is **authority and certainty are different things**. The 89-vote
margin sits inside an error bar of hundreds of thousands, and no amount of
institutional standing fixes that. Students who reach for "so the census is
unreliable" have missed it — the census is excellent and still not precise
enough for what was asked of it.

Watch for the corollary in discussion: **you cannot check a census against a
second source, because there is no second source.** The habit that caught last
week's problem is unavailable here; reading the Bureau's own published
uncertainty is what replaces it. That is Thursday's reading.

Every number below was run against the data file.

***

# The lab in one sentence

The House allocates itself by arithmetic once the census is counted — and in
2020 that arithmetic turned on **89 people**, a margin smaller than every source
of error feeding into it.

***

# The findings

```{r headline}
c(states = nrow(ap), seats = sum(ap$seats),
  reproduces_official = identical(as.integer(base[ap$state]), as.integer(ap$seats)))

mn <- ap[ap$state == "Minnesota", ]; ny <- ap[ap$state == "New York", ]
wp <- mn$app_pop / sqrt(mn$seats * (mn$seats - 1))
need <- wp * sqrt(ny$seats * (ny$seats + 1))
c(ny_had = ny$app_pop, ny_needed = ceiling(need), short_by = ceiling(need - ny$app_pop))
```

- Huntington–Hill on the Bureau's own population column **reproduces the
  official 435 exactly**. If that ever returns FALSE, the data file is wrong —
  stop and rebuild it.
- **Seat 435 → Minnesota. New York missed by 89 people.**
- Overseas population: New York **14,502**, Minnesota **3,258** — the rule
  assigning federal employees abroad to a home state moved ~160× the margin.

***

# The two ideas to teach

**1. A neutral-sounding rule has a direction.** Part 2 has students round the
quota. It totals 435 — by luck — but produces a different House:

```{r rounding}
ap$quota <- 435 * ap$app_pop / sum(ap$app_pop)
ap$rounded <- round(ap$quota)
c(total = sum(ap$rounded))
ap[ap$rounded != ap$seats, c("state", "quota", "rounded", "seats")]
```

New York and Ohio gain, Rhode Island and Montana lose. Large states have large
fractions to round up. "Round to the nearest whole seat" is not neutral.

**2. Those four states come back in Option C**, because rounding the quota *is*
Webster's method. The obvious thing students tried in Part 2 turns out to be one
of the named contenders Congress rejected. If nobody notices, point at it.

***

# Option A — Resize the House

## Verified output

```{r a}
for (sz in c(436, 440, 450, 500, 600)) {
  s <- hh(sz); d <- s[ap$state] - base[ap$state]
  cat(sprintf("%4d seats (+%3d): %s\n", sz, sz - 435,
              paste(names(sort(d[d > 0], decreasing = TRUE))[1:6], collapse = ", ")))
}
```

| House | gains go to |
|---|---|
| **436** | **New York** — the seat it missed by 89 |
| 440 | TX, FL, NY, OH, AZ |
| 450 | CA, TX, FL, NY, PA, OH |
| 600 | CA, TX, FL, NY, PA, IL |

**The Wyoming Rule.** Set the average district to the least populous state:

```{r a-wyoming}
c(total_pop = sum(pop), wyoming = pop[["Wyoming"]],
  implied_house = round(sum(pop) / pop[["Wyoming"]]))
```

**573 seats.** But it does *not* equalise districts:

```{r a-wyoming-check}
for (sz in c(435, 573)) {
  s <- hh(sz); per <- round(pop / s)
  cat(sprintf("House %3d: districts %s to %s, ratio %.2f\n", sz,
      format(min(per), big.mark = ","), format(max(per), big.mark = ","),
      max(per) / min(per)))
}
```

Largest-to-smallest ratio falls only from **1.83 to 1.76**. At 435 the extremes
are Delaware (990,837, one seat) and Montana (542,704); at 573 they become North
Dakota (779,702) and South Dakota (443,885). Rounding still bites at the bottom.

## What a good answer says

**3/4:** Reports who gains and identifies the Wyoming Rule as implying roughly
570–575 seats.

**4/4:** Notices that **436 gives the seat to New York** and connects it back to
Part 4 — the near-miss is not an abstraction, it is one seat away. And answers
the "who benefits from 435" question with the mechanism rather than a slogan:
freezing the House caps the size of large-state delegations while every small
state keeps its floor of one, so the freeze is a small-state subsidy that
compounds every decade the country grows.

**4/4, best:** catches that the Wyoming Rule barely improves equality (1.83 →
1.76). It is a reform aimed at the *average* district while the problem lives in
the *extremes*, which are produced by the one-seat floor and by rounding, not by
House size. A student who works that out has understood why the obvious fix is
not much of a fix.

**Common stumble:** assuming a bigger House helps small states. It is the
reverse — every added seat goes to a large state until the House is very large
indeed.

***

# Option B — How close was everybody else?

## Verified output

```{r b}
pv <- pop / sqrt(base * (base + 1))
top <- sort(pv, decreasing = TRUE)[1:6]
for (nm in names(top)) {
  k <- ap$seats[ap$state == nm]
  cat(sprintf("  %-12s priority %9.1f   short by %s people\n", nm, top[[nm]],
      format(ceiling(wp * sqrt(k * (k + 1)) - pop[[nm]]), big.mark = ",")))
}
```

| priority rank | state | short by |
|---|---|---|
| 1 | New York | **89** |
| 2 | Ohio | 11,462 |
| 3 | Texas | 189,645 |
| 4 | Florida | 171,561 |
| 5 | Arizona | 79,509 |
| 6 | California | 478,806 |

*(Rows are in **priority order**, which is what the chunk prints and what
decides the next seat. An earlier version of this table sorted by shortfall
instead while still labelling the column "rank" — putting Arizona third and
Texas fifth, and contradicting the chunk directly above it. The shortfalls
themselves were always right. The note below is why the two orders differ.)*

## What a good answer says

**3/4:** Reports the ladder and notices the enormous jump from first to second.

**4/4:** Draws the right conclusion from the **129-fold gap between 89 and
11,462** — New York's near-miss was genuinely freakish, not the normal state of
affairs. Second place missed by a small city. A student who says "the headline
makes this look like a routine knife-edge, and the second row shows it wasn't"
is reading the table rather than the story about it.

**4/4, best:** takes on "count harder." A state cannot manufacture 89 people,
but it *can* fund outreach, run its own count-review, and challenge the Bureau's
figures — New York did in fact pursue this. The honest answer is that the margin
was smaller than the census's own uncertainty, so "count harder" is not a
strategy, it is a lottery ticket. That is the same point `census` makes about
differential privacy — this same session — and Session 3 makes about §203 thresholds: a sharp legal line
across an estimated quantity.

**Note on Texas vs Florida.** Texas has a *higher* priority value than Florida
yet needs *more* people. That is not an error: Texas has 38 seats to Florida's
28, so its divisor is larger and each unit of priority costs more population.
Reward anyone who spots it.

***

# Option C — Who would a different method favour?

## Verified output

```{r c}
webster <- function() {
  s <- setNames(rep(1L, length(pop)), names(pop))
  for (i in 1:385) { p <- pop / (s + 0.5); w <- names(which.max(p)); s[w] <- s[w] + 1L }
  s
}
w <- webster()
d <- data.frame(state = ap$state, huntington_hill = ap$seats,
                webster = as.integer(w[ap$state]))
d[d$webster != d$huntington_hill, ]
```

**New York +1, Ohio +1, Rhode Island −1, Montana −1** — the same four states as
naive rounding in Part 2, and for the same reason.

## What a good answer says

**3/4:** Reports the four states and the direction — Webster favours the large.

**4/4:** Makes the connection the lab set up: rounding at the half *is* a divisor
of $n + 0.5$, so Part 2's "obvious method" was Webster all along. Huntington–Hill
uses the geometric mean $\sqrt{n(n+1)}$, which is always slightly *below* the
arithmetic mean, and that small difference is exactly what protects Rhode Island
and Montana.

**4/4, best:** answers the choosing question without pretending it is technical.
Both methods are defensible; the 1930s debate was resolved by a National Academy
panel, and Congress adopted the method that happened to favour the states it
favoured. A student who says "there is no neutral place to stand, so the honest
move is to state the criterion first and pick the method second" has the
argument.

***

# Option D — Their own question

The 1920s are the richest thread: Congress simply **failed to reapportion** after
the 1920 census — the only time in American history — because the count showed a
country that had become urban and immigrant, and the sitting rural majority
declined to give up its seats. Ten years of a House that did not reflect the
census. That is worth a student's essay, and it is the answer to "does the
arithmetic decide?" No. Congress does; the arithmetic only decides once Congress
agrees to run it.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is seeing the political decision inside a piece of
  arithmetic — that "round to the nearest seat" has a direction, that the
  Wyoming Rule targets the wrong statistic, that 435 is a statute rather than a
  law of nature.
- **Watch for fatalism.** Some students conclude that because it is arithmetic
  it is neutral and nothing can be done. Part 2 and Option C exist to break that.

## Anticipated questions

- *"Why is DC not in the file?"* — DC has no apportioned Representative. Fifty
  rows, not fifty-one. Worth a moment; it is the same question §203 and the
  Electoral College raise from other angles.
- *"Is 435 in the Constitution?"* — No. Fixed by the Permanent Apportionment Act
  of 1929. The House grew with the country for 140 years and then stopped.
- *"Could New York have sued?"* — It explored challenges; the Bureau's count
  stands absent proven error. There is no recount for a census.
- *"What is the difference between apportionment population and the census
  count?"* — Overseas federal employees, assigned to a home state. Table 3 in the
  source. Students who ask this have read Part 1 properly.
- *"Does this happen every decade?"* — The 89 is extraordinary; see Option B's
  second row. Whether *some* state is always within a few thousand is a genuinely
  open question here, and answering it needs the 1990/2000/2010 files, which this
  lab does not ship. Say that rather than guessing.

## Connections forward

- **`census`, the other lab in this session,** is where the count itself comes under scrutiny. This
  lab is the reason it matters: a margin of 89 against an instrument with error
  bars in the hundreds of thousands.
- **Session 3 (the ACS and §203)** — the same structure, a sharp legal threshold applied to an
  estimate. Say "this is the apportionment problem again" when you get there.
- **The Dec 3 wrap (redistricting)** — apportionment decides how many seats a state
  gets; districting decides where the lines go. Students routinely conflate them,
  and having run the arithmetic here makes the distinction stick.
- **Chapter 2** — the Electoral College is apportionment plus two. Every seat
  moved here moves an elector, which is the cleanest way to explain why census
  litigation is partisan.
