---
title: "Data Session 2, part 3: Three Front Doors"
subtitle: "84-355 Democracy's Data | Getting census data out of the Census Bureau"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

Everything here runs as given, **offline**. Nothing in this file needs a key or
a working connection. Your job is to read what comes out and say what it means.

***

# The question

You have spent this session with census numbers. Where do you actually *get*
them?

There are three ways in, they return the same population counts, and they are
not interchangeable. **Which door you use decides what you can do.**

| Door | What it is | Key? |
|---|---|---|
| **data.census.gov** | the website: search, click, download | no |
| **www2.census.gov** | the raw published files, as released | no |
| **the API** | a URL that returns data | **yes, since 2026** |

The apportionment table you used earlier came through the second door. The
county file in the race section came through the second door too. Today we look
at the third one, because it is the only one you can automate — and it is the
only one that is gated.

# Part 1 — The API is two things, and only one is open

Ask the API *what exists* and it answers immediately, no key:

```
https://api.census.gov/data/2020/dec/pl/variables.json
https://api.census.gov/data/2020/dec/pl/geography.json
```

Ask it for *an actual number* and you get this:

```
https://api.census.gov/data/2020/dec/pl?get=NAME,P1_001N&for=state:42
```

> **Missing Key**

Not an error code. Not JSON. **An HTML page** — which means a program asking for
data without a key does not crash, it silently receives a web page where it
expected numbers. Hold that thought; it comes back in Part 4.

**The catalogue is public. The data is not.** We cached the two open files so
this lab runs without a network:

```{r load}
v <- read.csv("data/api_variables.csv", stringsAsFactors = FALSE)
g <- read.csv("data/api_geography.csv", stringsAsFactors = FALSE)

c(variables = nrow(v), geography_levels = nrow(g))
```

# Part 2 — What the 2020 census actually released

**338 variables.** Not 338 questions — the 2020 census asked about ten. These
are 338 *published cross-tabulations* of the answers.

```{r concepts}
sort(table(v$concept), decreasing = TRUE)[1:6]
```

Look at the top two: the same table, run twice, once for everyone and once for
**the population 18 years and over**. That is not curiosity. **The Voting Rights
Act is about voters**, so every racial cross-tabulation is published a second
time restricted to adults. Seventy-three variables exist because of a statute.

```{r race-vars}
p1 <- v[grepl("^P1_00", v$variable), c("variable", "label")]
p1$label <- gsub("!!", " > ", p1$label)
head(p1, 8)
```

The `!!` are nesting. `P1_003N` is *Total > Population of one race > White
alone* — the count you would casually call "the white population," and it is
three levels down inside a hierarchy, and it excludes anyone who ticked two
boxes.

# Part 3 — How small it goes

```{r geo}
nrow(g)
g$name[1:14]
```

**Ninety-six geography levels**, and `block` is one of them. A census block is
often a single city block; many contain nobody at all.

This is the whole reason the API matters. The website will hand you states and
counties comfortably. It will not hand you **eight million blocks** — but a
program can ask for them one state at a time, which is exactly what the
redistricting file in Session 14 is built from.

```{r geo-tail}
g$name[grepl("legislative|congress|school", g$name)]
```

Every one of these is a *political* geography — the Census Bureau publishes
population inside district lines that the census itself was used to draw. That
circularity is not a mistake; it is the machinery Session 14 takes apart.

# Part 4 — Reading a URL

```
https://api.census.gov/data/2020/dec/pl?get=NAME,P1_001N&for=state:42
└──────── host ────────┘└─ dataset ─┘ └── what ──┘ └── where ──┘
```

- `2020/dec/pl` — the 2020 **dec**ennial census, **P**ublic **L**aw 94-171 file
- `get=NAME,P1_001N` — the name of the place, and total population
- `for=state:42` — for state FIPS 42, which is Pennsylvania

Change `for=state:42` to `for=county:*&in=state:42` and you get all 67
Pennsylvania counties. That is the whole idea: **the URL is the query.**

```{r urls}
build <- function(vars, where) sprintf(
  "https://api.census.gov/data/2020/dec/pl?get=NAME,%s&for=%s", vars, where)

cat(build("P1_001N", "state:42"), "\n")
cat(build("P1_001N", "county:*&in=state:42"), "\n")
cat(build("P1_003N,P1_004N", "tract:*&in=state:42+county:003"), "\n")
```

**And every one of those, run today without a key, returns a web page saying
"Missing Key."** Not an error your code would notice.

**This is the practical lesson of the session.** A gate that returns HTML
instead of failing loudly is worse than a gate that returns an error, because
the failure only shows up later as an inexplicable number — or as no number at
all, in a script nobody re-reads.

# Part 5 — Getting a key, if you want one

Free, instant, and **entirely optional for this course**.

1. Go to <https://api.census.gov/data/key_signup.html>
2. Give an email and an organisation ("Carnegie Mellon University")
3. The key arrives by email — a long hexadecimal string

Then append `&key=YOUR_KEY_HERE` to any of the URLs above.

> **A key is a password.** It identifies you. Do not put one in a file you share,
> a public repository, or a handout. This is not hypothetical: a live key sat in
> a handout for this course's F25 section for a year.

**Nothing in this course requires one.** Every lab ships its data. The key is
for what you do afterwards.

# Part 6 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Find the adult version.** Every race table has a twin restricted to
18-and-over. Find the concept, count its variables, and work out which `P` table
number it uses (`grepl` on `v$concept`). Why does that twin exist, and what
would you be unable to do without it?

**Option B — Build a query.** Use `build()` to write the URL that would return
the Hispanic population of every county in Georgia. You will need to search
`v$label` for the right variable. You cannot run it without a key — say what you
expect it would return, and how many rows.

**Option C — The smallest unit.** How many of the 96 geography levels are
*smaller* than a county? What does the existence of that many levels tell you
about who the Census Bureau's users actually are?

**Option D — Your own question.** 338 variables, 96 geographies, one cached
catalogue.

# What this cannot tell you

- **We never ran a live query.** Everything here is the catalogue plus a URL
  grammar. If the API changes its rules again — it just did — this lab describes
  a system that no longer behaves that way.
- **338 variables is one file.** The decennial PL file is the smallest thing the
  Bureau publishes. The ACS has tens of thousands of variables, which is the
  subject of the next session.
- **The API returns what was published**, and publication involves decisions —
  suppression, rounding, and since 2020 the deliberate noise of differential
  privacy. The number in the JSON is not the number the enumerator wrote down.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- 73 of the 338 variables exist because a law needs them. What does that say
  about where data comes from?
- A missing key returns a *web page*, not an error. What breaks, and when would
  anyone find out?
- The Bureau publishes population inside congressional districts, which were
  drawn using that population. Is that circular, and does it matter?

***

### Where the data came from

`api_variables_2020pl.json` and `api_geography_2020pl.json` were fetched from
the Census API's open metadata endpoints in August 2026 and are committed here
unmodified. `build-api-data.R` converts them to the two CSVs this lab reads,
using base R only.

**The data endpoints were not used**, because they now require a key. The
population figures elsewhere in this session came from the published PL 94-171
files at `www2.census.gov`, which need no key and no account.
