---
title: "Data Session 2 Key, part 3: Three Front Doors"
subtitle: "84-355 Democracy's Data | Session 2 — The Census API | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
v <- read.csv("data/api_variables.csv", stringsAsFactors = FALSE)
g <- read.csv("data/api_geography.csv", stringsAsFactors = FALSE)
```

All code is given to students. Grade the interpretation.

**Where this sits.** The third part of Session 2, after apportionment and the
race classification. It is the shortest piece of the session and the one that
travels furthest — every later session gets its data from somewhere, and this is
where students learn that "somewhere" has a door, a lock, and a key.

**It runs entirely offline.** The two metadata files are cached in `data/`.
Nothing here needs a network or an API key, deliberately.

***

# The session in one sentence

The Census API answers "what exists?" to anybody and "what is the number?" only
to registered users, and when you ask without a key it hands you a **web page**
instead of an error.

***

# The findings

```{r headline}
c(variables        = nrow(v),
  geography_levels = nrow(g),
  vra_18plus_pair  = sum(v$concept == "HISPANIC OR LATINO, AND NOT HISPANIC OR LATINO BY RACE FOR THE POPULATION 18 YEARS AND OVER"),
  all_18plus       = sum(grepl("18 YEARS AND OVER", v$concept)))
```

- **338 variables** from a census that asked about ten things.
- **145 of them — 43% — are the 18-and-over versions** of other tables.
- **96 geography levels**, including `block` and `voting district`.
- A keyless data request returns an **HTML page titled "Missing Key."**

***

# The three things to get said

**1. A census with ten questions publishes 338 variables.** The gap between
*what was asked* and *what is released* is the whole point. Nobody was asked
"are you a non-Hispanic Asian adult living in a group quarters" — that is a
cross-tabulation, produced by the Bureau after the fact, and someone decided it
was worth producing.

**2. Forty-three percent of the file exists because of the Voting Rights Act.**
Every racial table is published twice, once for everyone and once for adults,
because **the VRA is about voters and voters are adults**. That is a statute
visible in the shape of a data file. **Ask the room to name another dataset
whose column list was determined by a law** — Session 6's voter file, whose race
column exists for the same reason, is the answer you want.

**3. A locked door that returns HTML is worse than one that returns an error.**
A script asking without a key does not crash — it receives a web page where it
expected numbers. Ask what happens next: either an obvious explosion downstream,
or, worse, a parser that silently produces nothing and a table of zeroes nobody
questions. **This is the same failure shape as Session 11's −99 codes and
Session 6's missing 2020 voters**: the data did not announce its own absence.

**The line for the board:** the catalogue is public, the data is not, and the
lock is quiet.

***

# On the key policy — decide before Sep 3

The API **now rejects keyless requests**; the old allowance of a few hundred
queries a day is gone. Three options, and the lab is built for the first:

1. **Cached metadata plus optional key** *(what this lab does)*. Everything runs
   offline. Students who want a live query register their own key in about a
   minute. Nobody is blocked by wifi or by a queue at the signup page.
2. Require every student to register. Adds a failure mode on day one for no
   pedagogical gain.
3. Share one key. **Do not.** See below.

**Revoke the F25 key.** A live Census API key sits in plain text in
`F25/Assignments/labs/surnames/surnames.md`, which was distributed to students a
year ago. Revoke at <https://api.census.gov/data/key_signup.html>. The warning in
Part 5 of the lab refers to this without identifying it; if a student asks, the
honest answer is that it was mine and it was a mistake, which is a better lesson
than a hypothetical.

***

# Option A — Find the adult version

## Verified output

```{r a}
ad <- grepl("18 YEARS AND OVER", v$concept)
table(sub("_.*", "", v$variable[ad]))
c(count = sum(ad), pct_of_file = round(100 * mean(ad)))
```

The 18+ tables are **P3** (race) and **P4** (Hispanic origin by race) — the
adult twins of P1 and P2. **145 variables, 43% of the file.** (`GEO` appears in
the count too; it is the geography identifier, not a substantive table.)

**3/4:** Finds the concept, counts the variables, identifies P3/P4.

**4/4:** Explains *why* — Section 2 vote-dilution claims turn on whether a
minority group is large enough to form a majority in a district, and the
relevant population is voting-age, not total. Without P3 and P4 you cannot
compute the thing the litigation is about.

**4/4, best:** Notices that even 18+ is not the right denominator, because it
includes non-citizens and people with felony disenfranchisement. **The Bureau
publishes the closest thing it can measure, not the thing the law wants**, and
courts have argued about that gap for decades. A student who gets here has
independently found CVAP.

***

# Option B — Build a query

## Verified output

```{r b}
h <- v[grepl("Hispanic or Latino$", v$label) & grepl("^P2_", v$variable), ]
h[, c("variable", "label")]
cat("\nhttps://api.census.gov/data/2020/dec/pl?get=NAME,P2_002N&for=county:*&in=state:13\n")
```

The variable is **`P2_002N`** and Georgia is state FIPS **13**. The query would
return **159 rows** — one per Georgia county — of which one is Houston County,
which students will meet again in Sessions 6, 7 and 14. (Georgia has 159
counties, second only to Texas. If a student says 160, that is worth catching:
the row count is a checkable prediction, which is the point of the exercise.)

**3/4:** Finds a plausible variable and writes a syntactically correct URL.

**4/4:** Picks `P2_002N` rather than `P4_002N` and can say why — P2 is total
population, P4 is adults only — and states the expected row count.

**4/4, best:** Notes that they cannot verify any of this without a key, and that
**writing a query you cannot run is the ordinary condition of working with
gated data.** Being able to state precisely what you would get is the skill.

***

# Option C — The smallest unit

## Verified output

```{r c}
small <- c("block", "block group", "tract", "county subdivision",
           "subminor civil division", "place", "voting district")
g$name[g$name %in% small]
c(below_county = sum(g$name %in% small), total_levels = nrow(g))
```

Seven levels sit below the county, and **`voting district` is one of them** —
the Census publishes population by precinct.

**3/4:** Counts them and observes the API goes very small.

**4/4:** Answers the question actually asked — who the users are. Ninety-six
levels exist because school boards, water districts, city planners, marketers
and litigators all need different geographies. **The Census Bureau's users are
overwhelmingly not researchers**, and the file is shaped for administration.

**4/4, best:** Spots `voting district` and works out what it implies — the
Bureau publishes population inside precinct lines, which is precisely the join
Session 14 performs to run racially polarized voting analysis. **The data
required to challenge a districting plan is published by the government that
drew it.** Say this is deliberate: PL 94-171 exists to enable exactly that.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is reading the file's shape as evidence about who
  made it and why. Anyone who writes "this variable exists because…" and
  finishes correctly has the session.
- **A 2** is describing what the API does without asking why it is arranged that
  way.
- **This is a short piece.** It follows apportionment and race classification in
  the same 80 minutes; do not let it run long. Twenty minutes is right.

## Anticipated questions

- *"Why do I need a key for public data?"* — Rate limiting and abuse prevention,
  officially. The effect is that the automatable door is the gated one, while
  the point-and-click door stays open. Both facts are true and worth stating.
- *"Is the API data different from the website's?"* — No, same published tables.
  What differs is scale: a person can click through fifty counties, not eight
  million blocks.
- *"Are these the real numbers?"* — They are the *published* numbers. Since 2020
  the Bureau injects noise for differential privacy, so block-level counts are
  deliberately not exact. Flag it; it is the wrap's census-privacy topic.
- *"Can I use this for the data journalism project?"* — Yes, and this is the
  session where somebody realises they can.

## Connections

- **Session 3 (ACS)** — the same API, tens of thousands of variables, and every
  one with a margin of error.
- **Session 6 (voter files)** — the other dataset whose columns were written by
  a statute.
- **Session 11 (EAVS)** — the other place where the data quietly fails to
  announce its own absence.
- **Session 14 (precincts)** — `voting district`, which is where that join comes
  from.
