---
title: "Part II — Surveys"
type: chapter
subtitle: "Asking people, when nobody is obliged to answer"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-2-surveys-code.R")
```

```{r setup, include=FALSE}
```

Part I already contains a survey. The American Community Survey asks a sample of about 3.5 million addresses a year and publishes estimates rather than counts. It sits in the counting part because answering it is required by law.

**Everything in this part is the other kind.** Nobody has to answer any of it. A researcher decided a question was worth asking, wrote it a particular way, found the money to field it, and asked some people — most of whom said no. If any step in that chain had not happened, the number would simply not exist, and no amount of searching would recover it.

That is the weakness of the instrument and its entire justification at once.

## The one thing no record contains

There is no register of opinions in the United States. Nothing in any government file records whether you trust Congress, which party you think of yourself as belonging to, or whether you approve of how an election was run. There has never been such a file and a free country is unlikely to build one.

So if you want to know whether trust in government fell after 1964, there is exactly one way to find out: somebody had to have asked people at the time. Somebody did, which is the only reason the question is answerable now.

The price is that a survey cannot tell you how many people live on a block, and this is arithmetic rather than a lack of effort. Spread even a large national survey across the country and it places roughly nobody in any particular place. **There is no sample size at which a sample becomes a count.**

## Two halves, and the order matters

This part deliberately runs in two stages.

**First you use the surveys.** Open the file and see what one person looks like as a row. Then ask real questions of it. How has party identification moved? What does the public think about abortion? Has confidence in institutions fallen?

**Then you turn on them.** The last chapters check the survey rather than using it. They compare what people said they did against records of what they actually did. They show what **weighting** repairs and what it cannot: weighting means counting some answers more than others, to make a sample that does not look like the country behave as though it did. And they show what a **margin of error** does not cover. It tells you how far off an estimate could be because you asked a sample instead of everyone, and says nothing about the other ways an estimate goes wrong.

The order is not decorative. A reader who meets the audit before ever reading a survey file has been handed the verdict before meeting the evidence.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

## Where this part sits

```{r reusetab}
```

Look at the middle row, which counts the chapters elsewhere in the book that read this part's files: **`r nn(RU("Chapters elsewhere that read this part\'s files"))`.** That is not a disappointment and it is not an accident. Survey data answers questions about what people think, and those answers do not become inputs to anything else in this book. This part is a destination rather than a supply line.

It reads `r nn(RU("Chapters here that read another part\'s files"))` file from elsewhere, and that one matters — it is how a survey gets checked against something that is not another survey.

## How to read this part

Start with [what a survey can establish](../surveys-source/surveys-source-brief.html). That is the argument for the whole method. Then read [whether you can get one at all](../survey-access/survey-access-brief.html), a short chapter with an uncomfortable finding about how many archives will answer a program.

If you are short of time, read the source chapter and then skip to the block at the end that turns on the instrument. Those are the ones that will change how you read a poll in a newspaper.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens.
- The instruments themselves are introduced, with their documentation and their
  terms of use, in the chapters that use them. None of them is redistributed
  here; several are gated downloads, and the chapters say so where that is true.
