---
title: "The Same Votes, Redrawn"
subtitle: "Seventeen elections, thirty-eight districts, and a mid-decade map that never comes out behind"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("mid-decade-code.R")
```

```{r setup, include=FALSE}
```

## A claim worth checking

In August 2026 a Texas congressman told *POLITICO* that his redrawn district might favor him more than the one it replaced (**Benson 2026**). Henry Cuellar represents the 28th district, which the 2025 mid-decade map made about three points more Republican at the top of the ballot. His staff found races further down the 2024 ballot where the new lines looked better for his party, and he carried the numbers to the Speaker of the House.

The claim is testable, and this chapter tests it. Texas replaced its congressional map in 2025, between censuses, with the stated aim of electing five more Republicans. The votes already cast in seventeen statewide elections can be re-counted inside both sets of lines. Same ballots, same totals; the only thing that changes is which district each precinct belongs to.

That re-count is the whole method here. If the new lines really were friendly to the congressman's party, some election among the seventeen should elect more Democrats under them. You are about to see whether any does.

## Nothing federal stops a state doing this

Redrawing a map between censuses is legal, and the reason is that no federal rule says otherwise. The Constitution orders a count every ten years. It does not say how often the lines may move, and Congress has never said either. **Once a decade is a habit, not a requirement.**

What limits it is each state's own law, and a few states have written the limit down. Missouri's constitution bars its commission from changing boundaries between cycles. North Carolina's constitution prohibits mid-decade redistricting. Tennessee forbids mid-decade changes to congressional lines by statute. Texas has no such provision, which is why 2025 was available to it.

**Texas has done this before, and the case it produced is the leading one.** After Republicans took the legislature, Texas replaced its congressional map in 2003, mid-decade, in circumstances the chapter above should sound familiar against. The plaintiffs asked the Supreme Court to hold that a mid-decade redraw is presumptively a purely partisan act. In *League of United Latin American Citizens v. Perry* (2006) the Court declined. It required proof of an actual burden on representation, measured by a reliable standard, whatever the timing.

The plaintiffs did win on other ground, and it was racial rather than partisan. The 2003 map had reworked District 23 so that Latino voters were a majority of its adults but no longer a majority of its *citizen* adults. As compensation the legislature offered a new District 25, stitched together from two distant Latino communities. The Court held that a district reaching out to grab scattered communities does not make up for dismantling a compact one.

**Both halves of that outcome still matter, and one of them has since been closed.** *Rucho v. Common Cause* (2019) held that partisan gerrymandering claims cannot be heard in federal court at all, so the burden test *LULAC* laid down has nothing left to apply to. What survives is the racial route, and the [vote-dilution chapter](../../06-putting-data-together/vote-dilution/vote-dilution-brief.html) follows what happened to that in 2026.

So the question this chapter asks is not whether the 2025 map was allowed. It was. The question is what it did.

## Where the data comes from, and what it is for

The two spreadsheets behind this chapter come from Dave's Redistricting, a free public site. It rebuilds American elections at the precinct level, the few thousand voters who share one polling place, and re-sums them inside any district map you load. One file carries the map Texas enacted in 2021; the other carries the 2025 replacement. Each holds the same seventeen statewide elections from 2016 through 2024.

**Who is in it, and how they got there.** Every voter who cast a countable vote for a Democratic or Republican candidate in those seventeen races. Their ballots entered the file through the official county canvasses, which the site's volunteers collect and reconcile precinct by precinct.

**Who it was made for.** People who draw and judge district maps. The site exists so that anyone proposing a plan can see, before any election is held, how the votes already on record would fall inside it.

**What it costs to get.** Nothing but an account and patience. There is no address a script can call; each election is exported by hand from a browser with the right map open, which is why this chapter ships its copies.

**What has already been done to it.** Three things. Precincts that straddle a district line were split, with votes shared out in proportion to population. Each district's votes were then reduced to a two-party share — the Democratic vote divided by the Democratic and Republican votes together, third parties set aside. And the same arithmetic was run twice, once per map, which is what makes the comparison possible.

> **Before you scroll, make two predictions.** Across seventeen statewide
> elections, in how many does the new map elect more Republicans than the old
> one — and in how many does it elect fewer? Write both numbers down.

## Watch the lines move

The map below carries both sets of lines. Every district's outline was traced with the same number of points under each plan, so the old shape can slide into the new one. Press the buttons and the lines themselves show you what the legislature did. Color is the Democratic share of the two-party vote in whichever election you pick.

```{r mm-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r mm-panel-static, eval=!knitr::is_html_output(), fig.height=4.8}
```

```{r mm-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Thirty-eight districts, two maps, one set of votes, and any one district's full history.** The morph button slides every boundary from its 2021 position to its 2025 position; color re-counts whichever statewide race you select inside the lines shown. Click any district, or pick one from the District menu, to load all seventeen of its elections below the map, under both plans at once, its outline overlaid in the inset. In print, the two plans sit side by side, colored by the 2024 presidential race, with district 28 below as the worked example.

## Seventeen elections, one direction

The seat count settles the congressman's claim in aggregate. Count the districts where the Democratic candidate's two-party share tops one-half, election by election, under each map. Every one of the seventeen races elects more Republicans under the new lines: at least `r fx("rep_gain_min")` more seats, at most `r fx("rep_gain_max")`, never zero, never negative.

```{r sb-static, eval=!knitr::is_html_output(), fig.height=5.4}
```

```{r sb-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The gain, race by race.** Each row re-counts one statewide election under both maps. The dot never moves left. One row keeps the collection's own shorthand: the 2022 race labeled `trea` arrives that way in the source export, and this chapter will not guess at an office name the file does not state.

The size of the gain moves with the year. The 2018 races, a strong Democratic year nationally, hand the new map its smallest gains; the 2024 races hand it some of its largest. A map drawn from recent returns is tuned to recent returns, and you are watching the tuning.

## Every district, every election

Seat counts hide the districts that never flip. The scatter below keeps all of them: `r nrow(long)` points, one per district per election, with the old share across and the new share up. A district the swap left alone sits on the diagonal. Below the diagonal, the new lines made it more Republican; the shaded corners hold the districts where the winning party changes.

```{r xy-static, eval=!knitr::is_html_output(), fig.height=5.6}
```

```{r xy-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. What the swap did to every district at once.** Gray points barely move: most districts were already safe and stayed safe. The colored points are the exchange the map was drawn to make — a handful of Democratic-held districts pushed across the line, and almost nothing pushed back. In the online version, pick a district from the menu above the chart, or from any of the other district figures, to pick its seventeen points out of all of them.

## Where each district's people came from

The seat counts and the scatter both treat a district as a fixed thing that merely got redder or bluer. It is not. A mid-decade map moves people, and a district that gains five points can do it by trading a quarter of its residents for someone else's. The figure below tracks that directly. Every 2020 census block group carries its population into whichever district holds it under each plan. Each ribbon is then a group of Texans, leaving the district that held them for the one that holds them now.

Nineteen of the thirty-eight districts keep essentially everyone they had. The rest were rebuilt to varying degrees, and a few were rebuilt almost completely: the new 9th keeps only 3% of the people who lived in the old 9th, and the new 35th keeps 10%. Those are the seats the map was drawn to move, and this is what moving one actually looks like.

```{r ov-prep}
```

```{r ov-static, eval=!knitr::is_html_output(), fig.height=8.8}
```

```{r ov-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 4. Where one district's people came from.** Every ribbon is a group of residents, leaving the 2021 district that held them and landing in the 2025 district that holds them now. One district is opened up to fill the frame while the other thirty-seven collapse to slivers, because thirty-eight districts at equal size leaves each one too thin to read. Opened here is district 28, which keeps 53% of its own people and imports 42% from the old 15th. Online, click any district or ribbon and the diagram re-opens around it.

## The district the congressman is looking at

Both sides of the argument over district 28 are reading real numbers. In the 2024 presidential race the new lines are worse for a Democrat: Trump's margin grows by about three points, from `r 100 - fx("d28_pres2024_old_pct")`% to `r 100 - fx("d28_pres2024_new_pct")`% of the two-party vote. But in `r fx("d28_bluer_elections")` of the seventeen races, all from 2016 through 2022, the new district is the bluer one. Figure 1's panel carries this pattern directly, with district 28 as its default view: the new district runs bluer in the older races, redder in the newest ones.

Figure 4 says where those voters came from. The new 28th keeps 53% of the people who lived in the old 28th and imports 42% from the old 15th. Nearly half the district he is arguing about is territory he has never represented. A claim about how it will behave is partly a claim about people who were somebody else's constituents until last year.

So the congressman's staff and the map's drafter can both be right, because they are looking at different elections. The new 28th holds voters who backed Democrats reliably before 2024 and swung hard toward Trump in 2024. Whether it favors its incumbent depends on which of those electorates shows up next. That is the bet the whole gerrymander rests on, in his district and in the four others the map was drawn to move.

Notice what the claim has become. It began as "the new map might elect more Democrats" and survives only as "one district might swing back if 2024 proves unusual." The seat counts in Figure 2 already answered the larger question, in every election the data can ask.

## What re-run elections cannot tell you

These seventeen races are statewide contests re-counted inside district lines, not district contests. A House race has its own candidates, its own money, and an incumbent voters know. The congressman's own evidence came from House returns, which this data does not carry, and the map's drafter answered him with statewide returns, which it does. You have been reading the drafter's kind of evidence, and should say so when you quote it.

The re-count also holds turnout fixed. It assumes the people who voted in 2016 through 2024 are the people who will vote in 2026, in the same numbers, with the same loyalties. Districts full of voters who switched sides recently, like the ones along the border, are exactly where that assumption is weakest. The map's authors bet five seats on 2024 being the new normal; the data here can describe the bet, but cannot settle it.

## Sources

**Data**

- District-level results for seventeen Texas statewide elections, 2016–2024,
  under the 2021 and 2025 congressional plans, from Dave's Redistricting
  (<https://davesredistricting.org>), which rests on reconstructed
  precinct-level results from the official county canvasses. District
  boundary files for both plans come from the same site's exports.
- 2020 Census centers of population by block group (`CenPop2020_Mean_BG48`,
  U.S. Census Bureau), used only for Figure 4, to carry each block group's
  population into whichever district holds it under each plan. A block group
  split by a district line is charged whole to the district containing its
  population center, which reproduces district totals to within about 1.5%;
  read that figure's shares to the point, not the decimal. The companion
  Florida chapter runs the same idea on individual census blocks, which it
  needs because three of its districts were renumbered and Texas's were not.
  <https://www2.census.gov/geo/docs/reference/cenpop2020/blkgrp/CenPop2020_Mean_BG48.txt>

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`district_shares.csv`](data/derived/district_shares.csv), [`district_summary.csv`](data/derived/district_summary.csv), [`seats_by_election.csv`](data/derived/seats_by_election.csv), [`tx-districts-tween.json`](data/derived/tx-districts-tween.json), [`tx_overlap_pop.csv`](data/derived/tx_overlap_pop.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `seats_by_election.csv` replays every election on both maps and reports `rep_gain`. Count the elections where the new map does worse for its authors. What does a column with one sign mean?
- `district_summary.csv` has `mean_old` and `mean_new` for each district. Sort by the difference. Which districts absorbed the change, and which were left alone?
- `tx_overlap_pop.csv` gives the population shared between each old and new district, with a `weight`. Find the districts that were essentially rebuilt from scratch.
- Pick one election in `district_shares.csv` and compare `share_old` against `share_new` across all districts. Do the shares move together, or does the map take from some and give to others?

**Cases and state law**

- *League of United Latin American Citizens v. Perry*, 548 U.S. 399 (2006) —
  the challenge to Texas's 2003 mid-decade map: no presumption against
  mid-decade redistricting, and District 23 struck under Section 2 of the
  Voting Rights Act.
- *Rucho v. Common Cause*, 588 U.S. 684 (2019).
- National Conference of State Legislatures, *Redistricting Law 2020*
  (Denver: NCSL, October 2019), ch. 10 for the states that bar mid-decade
  redistricting and ch. 6 for what *LULAC* settled. Read there rather than
  from each state's code, and frozen at 2019: a state could have added or
  repealed such a provision since.

**Press**

- Benson, S. (2026). "Democrats see a path through Texas' gerrymander."
  *POLITICO*, August 16, 2026. The source of the congressman's claim tested
  above.

**Academic literature**

- Grofman, B. and King, G. (2007). "The Future of Partisan Symmetry as a
  Judicial Test for Partisan Gerrymandering after *LULAC v. Perry*."
  *Election Law Journal* 6(1): 2–35.
- Cervas, J. and Grofman, B. (2020). "Tools for identifying partisan
  gerrymandering with an application to congressional districting in
  Pennsylvania." *Political Geography* 76: 102069.

```{r ai-prompt, echo=FALSE, results="asis"}
```
