---
title: "Lab 12: Who Counts, and How"
subtitle: "84-355 Democracy's Data | Week 12"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

The Constitution orders a count. Article I requires an "actual Enumeration"
every ten years, and everything downstream — how many seats your state gets, how
districts are drawn, where §203 language assistance is required — runs off that
number.

But a count requires categories, and categories are made by people. James Scott's
argument, which you read for Tuesday, is that states must simplify in order to
govern: to administer a population you first have to render it legible, and the
rendering is never neutral. Today you will hold the instrument that does the
rendering for American democracy.

The file is called **P.L. 94-171**. A 1975 statute requires the Census Bureau to
deliver it to every state within a year of Census Day, specifically so that
legislatures can draw districts. It is the legal foundation of every
redistricting cycle, including the ones you will study in Weeks 14 and 15.

Everything is base R. Work in order.

***

# Part 1: The file the law requires

```{r read}
d <- read.csv("data/pl94171_counties.csv", stringsAsFactors = FALSE,
              colClasses = c(fips = "character"))
dim(d)
table(d$state)
```

Six states, 503 counties. Start at home:

```{r allegheny}
a <- d[d$fips == "42003", ]
a[, c("county", "state", "total", "white", "black", "asian", "two_or_more")]
```

Allegheny County had **1,250,578** people on 1 April 2020. That is not an
estimate and not a projection; it is the number the Census Bureau certified, and
Pennsylvania's districts were drawn from it.

***

# Part 2: The first question — race

Table P1 asks one question and produces seventy-one columns. We have kept the
main ones:

```{r p1}
a[, c("total", "one_race", "white", "black", "aian", "asian", "nhpi",
      "other_race", "two_or_more")]
```

The seven "alone" categories plus everyone reporting two or more races. It has to
add up, and it does:

```{r p1-check}
with(a, one_race + two_or_more == total)
```

The full table goes further than this. Because a person may check any combination
of six race boxes, the Census publishes **63 distinct race categories** — every
combination, from "White alone" through "White; Black; American Indian; Asian;
Pacific Islander; Some Other Race." That is what the other sixty columns hold.

> Sixty-three categories is not the Census Bureau being fussy. It is what happens
> when you take an idea that people experience as simple and try to make it
> administrable.

***

# Part 3: The second question

Here is the thing that catches everyone out. **The Census does not treat Hispanic
or Latino origin as a race.** It is a separate question, asked separately, and a
Hispanic respondent also answers the race question.

So there are two different ways to ask how white a place is.

```{r two-answers}
st <- aggregate(cbind(total, white, nh_white, hispanic) ~ state,
                data = d, FUN = sum)
st$white_pct   <- round(100 * st$white / st$total, 1)
st$nhwhite_pct <- round(100 * st$nh_white / st$total, 1)
st$gap         <- round(st$white_pct - st$nhwhite_pct, 1)

st[order(-st$gap), c("state", "total", "white_pct", "nhwhite_pct", "gap")]
```

```{r two-answers-plot, fig.width = 7.5, fig.height = 5}
s <- st[order(st$white_pct), ]
par(mar = c(5, 10, 4, 2))
bp <- barplot(rbind(s$nhwhite_pct, s$white_pct - s$nhwhite_pct),
        horiz = TRUE, las = 1, names.arg = s$state, beside = FALSE,
        col = c("#2166AC", "#92C5DE"), border = NA, xlim = c(0, 80),
        xlab = "% white", main = "Two answers to the same question")
legend("bottomright", c("Non-Hispanic white", "Hispanic and white"),
       fill = c("#2166AC", "#92C5DE"), border = NA, bty = "n", cex = 0.85)
```

> **New Mexico is 51.0% white by the race question and 36.5% white once you
> account for the ethnicity question — a gap of 14.5 points.**
>
> By one measure New Mexico is a majority-white state. By the other it is not
> close. Both numbers are correct. They are answers to different questions, and
> almost nobody reporting a figure like this tells you which one they used.

Texas shows the same thing at 10.4 points. Mississippi shows almost none (0.6),
because very few Mississippians are both Hispanic and white.

The gap is not a flaw in the data. It is the visible seam between two ways of
classifying people that the United States has never fully reconciled.

***

# Part 4: "None of these"

Look at the `other_race` column — people who answered the race question by
rejecting every box offered.

```{r other-race}
st2 <- aggregate(cbind(total, other_race, two_or_more) ~ state, data = d, FUN = sum)
st2$other_pct <- round(100 * st2$other_race / st2$total, 1)
st2$two_pct   <- round(100 * st2$two_or_more / st2$total, 1)
st2[order(-st2$other_pct), c("state", "other_pct", "two_pct")]
```

> **Fifteen percent of New Mexicans, and 13.6% of Texans, looked at the race
> question and chose "Some Other Race."**

Who are they? The file answers this directly, because it reports each race
crossed with Hispanic origin.

```{r other-is-hispanic}
tot_other  <- sum(d$other_race)
nh_other   <- sum(d$nh_other)
hisp_other <- tot_other - nh_other

cat("Chose 'Some Other Race':", format(tot_other, big.mark = ","), "people
")
cat("  Hispanic:    ", format(hisp_other, big.mark = ","),
    sprintf("(%.1f%%)
", 100 * hisp_other / tot_other))
cat("  non-Hispanic:", format(nh_other, big.mark = ","),
    sprintf("(%.1f%%)
", 100 * nh_other / tot_other))
```

> **94.5% of the people who rejected every race box are Hispanic.**
>
> Read that again, because it is the finding of this lab. The residual category
> on the race question — the box for people who fit nowhere else — is, in
> practice, an ethnicity box. Seven million people were asked their race, found
> that the form did not offer one they recognised, and the overwhelming majority
> of them were the same seven million the *next* question was about.
>
> The form asks a question a large minority of respondents do not think is well
> posed, and then records their refusal as if it were an answer.

And the multiracial column varies enormously — 25.3% in Hawaii against 3.7% in
Mississippi. Some of that is real difference in who lives where. Some of it is
that the 2020 form invited people to elaborate in ways the 2010 form did not, and
millions of people answered differently to a question about themselves that had
not changed.

***

# Part 5: Counted where you sleep

The Census counts people at their "usual residence" — where they sleep most of
the time. For most people that is uncontroversial. For some it is not.

```{r prisons}
d$corr_pct <- round(100 * d$gq_correctional / d$total, 1)
head(d[order(-d$corr_pct),
       c("county", "state", "total", "gq_correctional", "corr_pct")], 8)
```

> **Thirty-eight percent of Forest County, Pennsylvania is in a state
> correctional facility.** 2,653 people out of 6,973.
>
> Those 2,653 people cannot vote in Pennsylvania. They are counted as residents
> of Forest County for the purpose of drawing districts, and they are represented
> by a legislator they had no part in choosing — while the communities they came
> from, mostly urban, lose the population that would have counted toward their
> own representation.

This is what "prison gerrymandering" means, and it is a consequence of a
residence rule rather than of anybody drawing a map badly. Some states now
reallocate incarcerated people to their home addresses before redistricting.
Pennsylvania began doing so in 2021.

The same rule catches students:

```{r colleges}
d$coll_pct <- round(100 * d$gq_college / d$total, 1)
head(d[order(-d$coll_pct),
       c("county", "state", "total", "gq_college", "coll_pct")], 5)
```

Centre County is 9.0% dormitory. Tompkins County, New York — Cornell — is 12.4%.
On 1 April 2020 those students were mostly not there, because of the pandemic,
and the Bureau had to decide what to do about it.

***

# Part 6: What this file will not tell you

**The numbers have noise in them, on purpose.** For 2020 the Census Bureau
applied *differential privacy*: small random adjustments designed so that no
individual's answers can be reconstructed from published tables. State totals are
untouched and county figures are close, but the smaller the geography the more
the injected noise matters — and districts are drawn from blocks, not counties.
Whether the protection was worth the distortion was litigated by states, and
remains contested.

**The categories are not stable.** Every decade the race question changes, and
comparisons across decades compare instruments as much as populations. The 2030
question is expected to change again, most likely by combining race and Hispanic
origin into one question — which would erase, at a stroke, the gap you measured
in Part 3.

**A count is not a census of citizens.** P.L. 94-171 counts residents. That is
what apportionment uses, and attempts to change it have failed in court so far.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — The gap, county by county

```{r opt-a, fig.width = 8, fig.height = 5.5}
this_state <- "Texas"          # <- change to any state in the data

s <- d[d$state == this_state, ]
s$white_pct   <- 100 * s$white / s$total
s$nhwhite_pct <- 100 * s$nh_white / s$total
s$gap <- s$white_pct - s$nhwhite_pct

plot(s$white_pct, s$nhwhite_pct, pch = 19, col = "#2166AC88", cex = 1.2,
     xlim = c(0, 100), ylim = c(0, 100),
     xlab = "% white (race question only)",
     ylab = "% non-Hispanic white (both questions)",
     main = paste("Two measures of white population:", this_state))
abline(0, 1, lty = 2, col = "grey50")

big <- head(s[order(-s$gap), ], 6)
big[, c("county", "total", "white_pct", "nhwhite_pct", "gap")]
```

**Write about:** every point sits below the diagonal, and the vertical distance
is the gap. Which counties have the largest gaps, and what do they have in
common? Then the question that matters for this course: a news story reports
that a county is "60% white." Depending on which measure was used, that same
county might be 60% or 35%. Which measure would you want a redistricting case to
use, and does the answer depend on what the case is about?

***

### Option B — Who is warehoused where

```{r opt-b, fig.width = 8, fig.height = 5.5}
kind <- "gq_correctional"      # <- try gq_college, gq_nursing, gq_military

d$share <- 100 * d[[kind]] / d$total
top <- head(d[order(-d$share), ], 12)
top <- top[order(top$share), ]

par(mar = c(5, 13, 4, 2))
barplot(top$share, names.arg = paste0(sub(" County","",top$county), ", ",
        substr(top$state, 1, 2)), horiz = TRUE, las = 1,
        col = "#B2182B", border = NA, xlab = "% of county population",
        main = paste("Group quarters:", sub("gq_", "", kind)))

top[order(-top$share), c("county", "state", "total", kind, "share")]
```

**Write about:** describe the counties at the top. Are they large or small, and
why does that matter for the percentage? Then take the representation question
seriously. For prisons, the people counted cannot vote where they are counted.
For colleges, they mostly can. For nursing homes, they can. Does the objection to
prison gerrymandering rest on the *counting*, or on the *disenfranchisement*? If
Pennsylvania restored the vote to incarcerated people tomorrow, would the problem
go away?

***

### Option C — How much does the question shape the answer?

```{r opt-c}
metrics <- data.frame(
  state = st$state,
  race_question_white = st$white_pct,
  both_questions_white = st$nhwhite_pct,
  hispanic_pct = round(100 * st$hispanic / st$total, 1))
metrics$ratio <- round(metrics$race_question_white /
                       metrics$both_questions_white, 2)
metrics[order(-metrics$ratio), ]

# Is the gap just a function of how Hispanic a state is?
round(cor(metrics$hispanic_pct,
          metrics$race_question_white - metrics$both_questions_white), 3)
```

**Write about:** the gap is strongly related to how Hispanic a state is — which
is unsurprising, since the gap is by construction the Hispanic-and-white
population. The interesting part is the *ratio* column: in New Mexico the race
question inflates the white share by a factor of 1.40. Then push further: Hawaii
has a small gap but the lowest white share of any state here, and a quarter of
its people report two or more races. Does the race question serve Hawaii well? A
good answer says what a better question would look like and what it would cost.

***

### Option D — Ask your own question

Look up your home county. Compare Mississippi's counties to Hawaii's. Find the
county with the largest multiracial share and work out why.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at an official count and see the decisions inside it.

***

## Where the data came from

**U.S. Census Bureau**, 2020 Census Redistricting Data (P.L. 94-171) Summary
File, legacy format
(<https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/>),
for Pennsylvania, Texas, New York, Hawaii, Mississippi and New Mexico. Tables P1
(race), P2 (Hispanic or Latino origin by race) and P5 (group quarters).

Six states rather than fifty, because each state ships about 300 MB and these six
pull the race and ethnicity questions apart in usefully different directions.

County names are joined from the Census Gazetteer rather than the file's own
geographic header, which populates the name field inconsistently — Pennsylvania
carries "Allegheny County" and Hawaii leaves it blank. That is a small example of
the thing this whole course keeps finding: even an authoritative file needs
checking before you trust a column.

**On the design of this lab.** The structure — race, then ethnicity as a second
dimension, then the question of what the categories do — follows Moon Duchin's
PUBPOL 2130 at Cornell, *Data and the State*, adapted from Python to R. The
readings for Tuesday (Scott, and Nobles on race and censuses) are hers as well.

The script that downloaded, parsed and checked these files is
`data/build-data.R`. It verifies that P1 and P2 each decompose exactly, and that
Allegheny County's total matches the published figure, before writing anything.
