---
title: "Data Session 2 Key, part 2: Who Counts, and How"
subtitle: "84-355 Democracy's Data | Session 2 — The decennial census | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
d <- read.csv("data/pl94171_counties.csv", stringsAsFactors = FALSE,
              colClasses = c(fips = "character"))
st <- aggregate(cbind(total, white, nh_white, hispanic, other_race,
                      two_or_more) ~ state, data = d, FUN = sum)
st$white_pct   <- round(100 * st$white / st$total, 1)
st$nhwhite_pct <- round(100 * st$nh_white / st$total, 1)
st$gap         <- round(st$white_pct - st$nhwhite_pct, 1)
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

The census is not a photograph of the country; it is an instrument with
categories chosen by people, and both the categories and the residence rules
change who ends up represented.

***

# The findings

```{r headline}
st[order(-st$gap), c("state", "total", "white_pct", "nhwhite_pct", "gap")]
st$other_pct <- round(100 * st$other_race / st$total, 1)
st$two_pct   <- round(100 * st$two_or_more / st$total, 1)
st[order(-st$other_pct), c("state", "other_pct", "two_pct")]
```

1. **Two answers to one question.** New Mexico is **51.0%** white by the race
   question and **36.5%** non-Hispanic white — a **14.5-point gap**, a ratio of
   **1.40**. Texas 10.4. Mississippi 0.6.
2. **Refusal.** **15.0%** of New Mexicans and **13.6%** of Texans chose "Some
   Other Race."
3. **Multiracial reporting varies six-fold** — Hawaii 25.3%, Mississippi 3.7%.
4. **Forest County, Pennsylvania is 38.0% incarcerated** — 2,653 of 6,973.

***

# The three things to teach

**The gap is a seam, not an error.** Race and Hispanic origin are separate
questions because the United States has never settled whether "Hispanic" names an
ethnicity, a race, a language group, or a national origin. Students should leave
able to ask "which measure?" whenever they see a figure for white population.

**"Some Other Race" is a finding, not a residual — and Part 4 now measures it.**
Of the 7,072,769 people in these six states who rejected every race box,
**6,684,740 are Hispanic: 94.5%.** The residual category on the race question is,
in practice, an ethnicity box.

That number is the headline of the lab and the sharpest evidence in the course
for Scott's argument: the state must simplify, the simplification does not fit
the people, and the form records their refusal as though it were an answer.
Connect it explicitly to *Seeing Like a State*, and to Tuesday's debate —
**this is what "constructed rather than discovered" looks like as a measurement
rather than a slogan.**

Two cautions when you put it on the board. It is **six states**, chosen precisely
because race and ethnicity pull apart in them, so it is not a national figure.
And the causation runs both ways: the box is used by Hispanic respondents
*because* the form separates race from origin, so this is evidence about the
instrument, not about the people.

**Residence rules are redistricting rules.** Forest County is the sharpest case.
The counting is lawful, the map-drawing is lawful, and the result is a county
whose population is more than a third people who cannot vote there.

***

# Option A — The gap, county by county

## Verified output

```{r a}
s <- d[d$state == "Texas", ]
s$white_pct   <- 100 * s$white / s$total
s$nhwhite_pct <- 100 * s$nh_white / s$total
s$gap <- s$white_pct - s$nhwhite_pct
head(s[order(-s$gap), c("county", "total", "white_pct", "nhwhite_pct", "gap")], 6)
```

The largest Texas gaps are all border and South Texas counties — the same
counties students met in Session 4's swing analysis (Maverick, Webb, Starr).

## What a good answer says

**3/4:** Identifies that every point falls below the diagonal, names the
high-gap counties, and connects them to large Hispanic populations.

**4/4:** Answers the redistricting question with the right distinction. For a
Voting Rights Act §2 claim about Latino vote dilution, the relevant population is
Hispanic — so the race-only measure is actively misleading and non-Hispanic white
is the right denominator. For a story about, say, the racial composition of a
school district, the answer may differ. The measure follows the question, and a
figure reported without saying which measure was used cannot be checked.

**4/4, best:** notices that this is the same county set as Session 4's largest
Republican swings, and asks whether a population that the census struggles to
classify is also one that political analysis struggles to predict.

***

# Option B — Who is warehoused where

## Verified output

```{r b}
for (k in c("gq_correctional", "gq_college", "gq_nursing")) {
  d$share <- 100 * d[[k]] / d$total
  cat("\n", k, "\n"); print(head(d[order(-d$share),
    c("county", "state", "total", k, "share")], 4))
}
```

Correctional: **Forest PA 38.0%**, Issaquena MS 36.8%, Hudspeth TX 31.1%.
College: Tompkins NY 12.4%, Centre PA 9.0%.

Note that every top county is small. The percentage is driven as much by the
denominator as by the facility.

## What a good answer says

**3/4:** Observes that the top counties are small and rural, and states the
representation problem.

**4/4:** Separates the two objections cleanly, which is what the question asks.
Prison gerrymandering combines (a) counting people where they are held and
(b) denying them the vote there. Colleges have (a) without (b), and almost nobody
objects. So the objection rests on the *combination* — and a student who works
that out has done real analytical work.

**4/4, best:** answers the counterfactual honestly. If incarcerated people could
vote where held, the transfer of representation from urban to rural counties
would remain, but the people transferred would at least have a say in it. That is
better and not obviously fine, because the community they came from still loses
the count. The problem is genuinely about *both* things.

**Common stumble:** assuming these counties chose to host prisons in order to
gain representation. Some sought the jobs; the representation effect is mostly a
by-product. Do not let students over-claim intent.

***

# Option C — How much does the question shape the answer?

## Verified output

```{r c}
m <- data.frame(state = st$state, race_only = st$white_pct,
                both = st$nhwhite_pct,
                hispanic_pct = round(100 * st$hispanic / st$total, 1))
m$ratio <- round(m$race_only / m$both, 2)
m[order(-m$ratio), ]
round(cor(m$hispanic_pct, m$race_only - m$both), 3)
```

Correlation between Hispanic share and the gap: **0.982**. Nearly deterministic,
as it must be — the gap *is* the Hispanic-and-white population.

## What a good answer says

**3/4:** Reports the ratio, notes New Mexico at 1.40, and recognises the
correlation is near-mechanical.

**4/4:** Does not treat 0.98 as a discovery. The interesting question is the one
about Hawaii: the smallest gap of the six states and the lowest white share,
because Hawaii's diversity is Asian and Pacific Islander rather than Hispanic —
so the race/ethnicity seam simply does not run through it. A quarter of the state
reports two or more races, which the categories handle by multiplying into 63
combinations rather than by asking a better question.

**4/4, best:** proposes what a combined race-and-ethnicity question would gain
and lose. It would end the two-answers problem and make trend comparison to
earlier decades impossible. That is the actual trade the Bureau faces for 2030,
and a student who states it fairly has understood that there is no option without
a cost.

***

# Option D — Their own question

Home counties work well. Hawaii's five counties versus Mississippi's eighty-two
is a productive comparison.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is treating a category as a decision rather than a
  fact. Students arrive believing that race is something the census *measures*;
  the lab works if they leave believing it is something the census *constructs*,
  without sliding into thinking the numbers are therefore meaningless.
- **Reward students who ask "which measure?"** That habit is the most portable
  thing here.

## Anticipated questions

- *"Which number is right, 51% or 36.5%?"* — Both. They answer different
  questions. This will come up immediately and is worth the whole first ten
  minutes.
- *"Why is Hispanic not a race?"* — Because OMB Directive 15 says so, and because
  no alternative has commanded agreement. The 2030 question may merge them.
- *"Is the noise a problem?"* — At county level, small. At block level, where
  districts are actually drawn, larger and contested. Alabama sued over it and
  lost. Do not overstate: state totals are exact.
- *"Can prisoners vote anywhere?"* — Varies by state; in Pennsylvania,
  incarcerated felons cannot vote, though people held pre-trial can. Maine and
  Vermont allow voting from prison.
- *"Why only six states?"* — Each state file is about 300 MB. Six was enough to
  make the seam visible. `data/build-data.R` will add more if you want them.

## Connections forward

- **Session 3 (the ACS and §203)** runs directly off this file's categories — the language
  minority groups are defined in these terms, and the thresholds are applied to
  ACS estimates of the same populations.
- **Session 7 (surnames and BISG)** uses exactly the six categories students met in Part 2,
  and the same suppression logic.
- **The Dec 3 wrap (redistricting)** is where the group-quarters material pays off. Ask
  what Forest County does to a Pennsylvania state senate district.
- **`apportionment`, the other lab in this same session** — students computed how seats are allocated. This is
  the count those seats are allocated from.

## Provenance note

The structure of this lab — race, then ethnicity as a second dimension, then the
question of what the categories accomplish — is adapted from Moon Duchin's
PUBPOL 2130 (*Data and the State*, Cornell), ported from Python to base R. The
Tuesday readings (Scott; Nobles) are hers.
