---
title: "Cast Vote Records and Ranked-Choice Counting"
type: brief
subtitle: "A congressional election recounted from the ballots themselves, and what the winner's majority is a majority of"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("cast-vote-records-code.R")
```

```{r setup, include=FALSE}
```

An election result is a set of totals, because the ballot is secret and a total is what remains once the ballots are put away. Alaska does not put them away. Since adopting ranked-choice voting, the state has published its **cast vote record**: one row per scanned ballot, showing every ranking the voter marked. No name and no address, but the actual choices, one ballot at a time. Nothing else in American election data comes this close to the individual act.

The claim you have heard about ranked-choice voting is its promise: rank as many candidates as you like, support your real first choice, and the eventual winner will hold a majority. Every part of that is checkable against the ballots of one congressional election. **What does a ballot-level record let you see that no total can — and when the winner's majority arrives, what is it a majority of?**

## The test

The contest is the November 2024 race for Alaska's single seat in the U.S. House: Mary Peltola, the Democratic incumbent, against Nick Begich, with two minor candidates. The data is the State of Alaska, Division of Elections cast vote record for that election, the JSON export the state's voting equipment writes, published at a stable address (<https://www.elections.alaska.gov/results/24GENR/>). Beside it sits the state's official round-by-round report, a PDF, typed into tables by hand and used below to grade a recount.

Why this source, of all sources: it is the bottom rung of the ladder of election data, the only public American election file in which a row is one ballot. [The levels-of-aggregation chapter](../levels-of-aggregation/levels-of-aggregation-brief.html) walks that ladder rung by rung. Everything this brief measures (how many rankings voters used, which candidates shared a ballot, where eliminated candidates' votes went) is a property of single pieces of paper. A total has already destroyed it.

## The data

```{r ballot-photo, out.width="100%", fig.align="center"}
```

**Figure 1. The instrument: Alaska's own demonstration ranked-choice ballot**, cropped to the instructions and the House contest. Candidates run down the side, ranking columns across, and the offer sits at the top: rank as many or as few as you like. Every row of the file below is one of these grids, flattened. State of Alaska, Division of Elections, retrieved 10 August 2026.

The export is large and raw: `r format(SH("cvr_files"), big.mark = ",")` ballot files, `r pc(SH("bytes_uncompressed") / 1e9, 2)` GB uncompressed, one session per scanned ballot, with marks nested inside contests nested inside cards. One session is kept verbatim in this brief's data, with four identifying fields masked. Its envelope:

```{r cvr-envelope}
```

The ballot carries `r ncon` contests and this brief wants one. Here it is, complete:

```{r cvr-contest}
```

Nothing in that record is readable on its own. `CandidateId` is a number whose meaning lives in a separate file, `CandidateManifest.json`, listing `r nrow(cm)` candidates across every contest:

```{r cvr-lookup}
```

Turning `r format(SH("cvr_files"), big.mark = ",")` such files into a table took five judgments, each a rule somebody chose. **Keep one contest** of the `r ncon` on each ballot. **Believe `IsVote`, not the ink**: each mark carries its darkness (`MarkDensity`) and the scanner's verdict, and the build takes the verdict rather than second-guessing it with its own threshold. **Decide what two marks in one column mean**: the build writes `OVERVOTE` at that rank, then sorts the ranks and closes single gaps. Alaska's tabulator applies further rules about skipped ranks that this reconstruction does not, and that honest difference is measured below. **Translate numbers into names** through the manifest.

**Then destroy what is left.** Ballots with identical rankings are collapsed to one row with a count, and precinct and tabulator identifiers are dropped before anything reaches disk. That last step is for privacy, not tidiness. A record with no name on it is not the same as a record nobody can identify; a distinctive ranking in a small precinct can narrow down who cast it. The collapse cannot be run backwards, which is the point of it.

What ships is three tables. `sequences.csv` has one row per distinct ranking: a `ballots` count, then `rank1` through `rank4`, one candidate name per column the voter used. `official_rounds.csv` holds the state's `votes` per `candidate` per `round`, and `official_transfers.csv` records where each `eliminated` candidate's ballots were `transferred_to` — both typed from the PDF report. The single ballot above becomes one line among `r n(SEQ)`:

```{r cvr-clean}
```

And the commonest lines in the file are the shortest ones:

```{r one-row}
```

```{r ballot-static, eval=!knitr::is_html_output(), fig.height=4.3}
```

```{r ballot-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Two ballots, and the two records they become.** On the left, the commonest marked grid in the election: one ranking used, `r n(b_one$ballots)` ballots identical to it. On the right, the commonest grid that used all `r length(RK)` columns, cast `r n(b_all$ballots)` times. Everything a cast vote record knows about a voter is the few characters underneath each grid.

> **Before you read on, commit to two answers.** Alaska's ballot offered
> `r length(RK)` rankings. First: **what share of the voters in this race used
> exactly one?** Second: **whose supporters stop after one ranking more
> often** — the front-runner's, or the smallest candidate's? Write both down.

## What it says about democracy

```{r shape}
```

**`r n(BAL)` ballots collapse to `r n(SEQ)` distinct patterns**, and most voters used one of the first few:

```{r ranks}
```

Set aside `r n(BLANK)` ballots that left this contest blank. Of the `r n(VOTED)` that voted in it, `r n(ONE)` ranked exactly one candidate: `r pc(BULLET)`%. **Given `r length(RK)` rankings, two-thirds of Alaskans used one.** Neither the system's advocates nor its opponents advertise that number. Hold the obvious explanation, a confusing ballot, because the file can test it, and the test comes out the other way.

First, the count. Instant-runoff fits in three sentences. Count everyone's top continuing choice. If nobody has a majority, eliminate the lowest candidate and read those ballots' next choice. Repeat.

```{r our-count}
```

**`r names(fin)[1]` wins with `r pc(WINPCT, 2)`%** — a congressional election recounted from the ballots as this page renders. Now check it against the state:

```{r official}
```

```{r gap}
```

The percentages match Alaska's official result to two decimal places. The counts do not: this recount carries `r n(GAP)` more ballots into the final round, because the state's tabulator applies printed rules about write-ins, skipped ranks and overvotes that the three-sentence version does not. **The rules are not in the data. They live in a separate document, and they change the count** — here by `r pc(100 * GAP / ACTIVEF, 3)`%, in a closer race by the winner.

```{r transfers}
```

When John Wayne Howe was eliminated, his `r n(HOWE)` ballots split, and the largest single destination was **nowhere**: `r n(EXH)` of them, `r pc(100 * EXH / HOWE)`%, were exhausted, because those voters had ranked nobody else. Their ballots left the count before the decisive round. That is the trade at the heart of ranked-choice voting: it lets you express more than one preference, and it does not make you.

```{r flow-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r flow-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Alaska's official count, all three rounds drawn to one scale.** A flow of bands, because the count is a process: each column is a round, each band is ballots moving to the next. Every band is drawn to the same scale, so the gray one is exactly as wide as it truly is. The gray band is the `r n(OUT3)` ballots that leave: `r pc(100 * OUT3 / TOT)`% of the `r n(TOT)` that started, made of `r n(OUT3 - OVR)` exhausted ballots plus `r n(OVR)` overvotes.

Two smaller things only ballot-level data can show. `r n(sum(cross$ballots))` ballots ranked one major candidate first and the other second — `r pc(100*sum(cross$ballots)/VOTED)`% of everyone who voted in the race, one voter in twenty taking the cross-divide offer the system is supposed to reward:

```{r crossover}
```

And `r n(sum(dup$ballots))` ballots named the same candidate twice; the tabulator skips the duplicate, the ballots count normally, and no total would ever show the behavior:

```{r dupes}
```

Now the prediction. If two-thirds of voters stopping after one ranking is confusion, the rate should be roughly flat across candidates, because confusion has no reason to prefer a front-runner. Split the bullet voters by who they ranked first:

```{r bullet-table}
```

**`r pc(bul$bullet[1])`% of `r sub(",.*", "", bul$candidate[1])`'s voters ranked nobody else, against `r pc(bul$bullet[nrow(bul)])`% of `r sub(",.*", "", bul$candidate[nrow(bul)])`'s**, and the rate falls cleanly with candidate size. That is not confusion's flat signature. It is arithmetic voters appear to have done. If your candidate is the front-runner, a second ranking will almost never be read. If your candidate polls at one percent, the second ranking is the only part of your ballot that will do anything.

A second reading, which this file cannot rule out, is campaign instruction — "rank one and stop" is a documented tactic. Either way the conclusion is the same: **these ballots show the campaigns' strategies, not just the voters' preferences.**

```{r d3-bullet, results='asis', eval=knitr::is_html_output()}
```

```{r bullet-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

**Figure 4. Bullet voting is not spread evenly across candidates.** One bar per candidate, ordered by first-preference votes (`r n(bul$voters[1])` down to `r n(bul$voters[nrow(bul)])`), because the claim is a gradient and ordered bars make it visible at a glance. The share who ranked nobody else falls from `r pc(bul$bullet[1])`% to `r pc(bul$bullet[nrow(bul)])`%.

Which leaves the finding in the subtitle:

```{r denominator}
```

The winner's `r pc(WINPCT, 2)`% is a majority of the `r n(ACTIVEF)` ballots still in the count, not of the `r n(BAL)` people who cast one. Measured against every ballot, the figure is `r pc(100 * fin[1] / BAL, 1)`% — a plurality. The majority is partly manufactured by shrinking the denominator, the bottom number the winner's votes are divided by, and Burnett and Kogan (2015) found the same shrinkage across four earlier ranked-choice elections. The fair answer is a question: what would the alternative denominator mean? Under plurality the winner's share here would also have sat under half, with no second round for anyone's ballot to survive into, and a voter who declines to rank anyone else has made a choice. Figure 4 says those voters were disproportionately the ones with the least to gain from a second ranking.

## What this brief cannot tell you

**It is not the certified count.** Alaska says so plainly: ballots counted by hand are not in the export and write-in names are not resolved, so `r n(BAL)` is not this contest's official turnout.

**It holds choices, not reasons.** Whether bullet voting is a calculation voters made or an instruction they followed cannot be separated from these rows; that would take survey data, and the ballot's anonymity is deliberate. Where surveys exist, Donovan, Tolbert and Gracey (2016, 2019) find ranked-choice voters mostly say they understand the system, which also cuts against the confusion reading. No demographics attach to any ranking.

**It is one race, in one state, in one year** — a state that adopted ranked-choice voting by initiative, whose voters are not a sample of anything. Whether the system changes who runs and how they campaign is a question about candidates, not ballots.

**And the file cannot say who cast what.** The published collapse to `r n(SEQ)` sequences was chosen so that it cannot be undone, which protects voters and closes some questions permanently.

## What you should have learned

- A cast vote record is the only American election data where a row is a ballot: `r n(BAL)` ballots collapse to `r n(SEQ)` distinct ranking patterns, and `r pc(BULLET)`% of voters used exactly one of `r length(RK)` offered rankings.
- The ballots do not count themselves. A three-sentence recount reproduces Alaska's official percentages to two decimals while differing by `r n(GAP)` ballots, because the binding rules live in a document, not in the data.
- `r n(OUT3)` ballots (`r pc(100 * OUT3 / TOT)`% of the count) left before the decisive round: `r n(OUT3 - OVR)` exhausted plus `r n(OVR)` overvotes.
- Bullet voting runs from `r pc(bul$bullet[1])`% among the front-runner's voters to `r pc(bul$bullet[nrow(bul)])`% among the smallest candidate's: a gradient that fits strategy or instruction, not confusion.
- The winner's `r pc(WINPCT, 2)`% majority is a majority of continuing ballots; of all ballots cast, it is `r pc(100 * fin[1] / BAL, 1)`%.

## Extensions

None of these is answered above, and every one can be answered from the tables the Sources section hands over.

- `sequences.csv` has one row per distinct ranking with a `ballots` count. Count the ballots whose `rank1` and `rank2` name different parties' candidates. How porous was the partisan divide, really?
- `sequences.csv` again: find the sequences containing `OVERVOTE`. How many ballots carried one, and at which rank do overvotes cluster?
- `official_transfers.csv` records where each eliminated candidate's ballots went. Compare the destinations against what the parties involved would predict. Were any transfers surprising?
- `official_rounds.csv` shows the count at each stage. Work out who led in round one and who won. Then decide whether the first-round leader has a complaint.

Two stretch ideas point past the spreadsheet. Maine and several cities publish cast vote records for their own ranked-choice races; fetch one and compute its exhaustion rate against Alaska's `r pc(100 * OUT3 / TOT)`%. And compare both against the four elections in Burnett and Kogan (2015): is exhaustion shrinking as voters gain experience with the system?

## Sources

**Data**

- State of Alaska, Division of Elections, 2024 General Election **cast vote
  record**, published as a Dominion JSON export.
  <https://www.elections.alaska.gov/results/24GENR/>
- Official round-by-round results and transfer detail from the state's **RCV
  Detailed Report**, <https://www.elections.alaska.gov/results/24GENR/RCV-USRep.pdf>
- The ballot in Figure 1 is the State of Alaska's own **sample ranked-choice
  ballot**, retrieved 10 August 2026.
  <https://www.elections.alaska.gov/doc/GenRCVsampleBallot11.9.21.pdf>
- The committed file contains no individual ballots: `r n(BAL)` records were
  collapsed to `r n(SEQ)` ranking sequences with counts, and precinct and
  tabulator identifiers were dropped, before anything was saved.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`official_rounds.csv`](data/derived/official_rounds.csv), [`official_transfers.csv`](data/derived/official_transfers.csv), [`sequences.csv`](data/derived/sequences.csv)

**Academic literature**

- Burnett, C. M. and Kogan, V. (2015). "Ballot (and Voter) 'Exhaustion' Under
  Instant Runoff Voting: An Examination of Four Ranked-Choice Elections."
  *Electoral Studies* 37: 41–49.
- Donovan, T., Tolbert, C. and Gracey, K. (2016). "Campaign Civility Under
  Preferential and Plurality Voting." *Electoral Studies* 42: 157–163.
- Donovan, T., Tolbert, C. and Gracey, K. (2019). "Self-Reported Understanding
  of Ranked-Choice Voting." *Social Science Quarterly* 100(5): 1768–1776.

```{r on-mark, results='asis', echo=FALSE, eval=knitr::is_html_output()}
```

```{r ai-prompt, echo=FALSE, results="asis"}
```
