---
title: "Data Session 8 Key, part 2: What People Say They Did"
subtitle: "84-355 Democracy's Data | Session 8 — Surveys, two ways | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
cps <- read.csv("data/cps_turnout.csv", stringsAsFactors = FALSE)
cps$gap_people <- cps$reported_voters - cps$actual_votes
cps$gap_pct <- 100 * cps$gap_people / cps$actual_votes
```

All Part 5 code is given to students. Grade the interpretation.

**Framing.** This is the **survey** lab among the four families of democracy
data, and it is the afternoon's second half — pair it with `ces-class`,
which takes one survey up close while this one follows one question over time.
The lab opens by distinguishing a survey from two families students already
have — returns (Session 1) and the census (Session 2) both try to count everyone
and fall short; a survey declines to try, asks a few thousand, and multiplies —
and ends with a table setting all three failures side by side.

The point of that table is that **none of the three failed through carelessness.**
Each failed in the way its method makes it liable to fail: returns by being
stitched, the census by carrying invisible uncertainty, surveys by the people who
answered differing from those who did not. Students who leave able to name the
question that belongs to each family have got the point.

The payoff line is in the closing: the fourth family, **administrative records**,
students already have — campaign finance in Session 5 and voter files in Session
6 — and its question is *what was this collected for?* Ask it backwards here.

Every number below was run against the data file.

***

# The lab in one sentence

This is the one day all term when a survey can be graded against the truth, and
it fails — by ten million people in 1972, by almost nothing in 2020, and the
change between those two is more interesting than the failure.

***

# The findings

```{r headline}
era <- cut(cps$year, c(1963, 1979, 1999, 2030),
           labels = c("1964-1976", "1980-1996", "2000-2024"))
round(tapply(cps$gap_pct, era, mean), 1)
c(worst_year = cps$year[which.max(cps$gap_pct)],
  worst_pct  = round(max(cps$gap_pct), 1),
  gap_2016   = round(cps$gap_people[cps$year == 2016]))
```

- **2016: 8,642,778 people** said they voted who do not appear in the count.
- **1972 is the peak: the survey overstated the electorate by 13.4%.**
- By era, over-report as a share of actual votes: **10.1% → 10.7% → 3.4%**.
- On the rate measure (1980 on) the mean gap is **3.7 points**, and **2020 is
  negative** — the survey slightly understated turnout.

***

# The idea to teach

**Not "people lie."** That is where students will stop, and it is the least
interesting reading. Three things are mixed together in the gap and the data
cannot separate them:

1. respondents misremembering or overclaiming,
2. the Bureau's weights stretching respondents' answers to cover non-respondents,
3. our own denominator and our use of presidential votes rather than all ballots.

**The finding that earns the hour is the trend.** A 10% over-report became a 3%
over-report without anyone setting out to fix it. Ask what else changed —
actual turnout rose sharply, survey response rates collapsed, and the Bureau
altered how it treats people who do not answer the voting question. Any of those
could produce this, and students should feel the difficulty of telling which.

**The transferable point** is in the closing section: voting is nearly the only
survey question with a public record to check against. Every poll they read
after today is asking for faith in exactly the thing they just watched miss.

***

# Option A — Go back further, and hit a wall

## Verified output

```{r a}
head(cps[, c("year", "cps_rate", "actual_rate", "gap_pct")], 16)[13:16, ]
```

The rate columns are **`NA` for 1964–1976**. Not a bug: the Bureau did not
publish a citizen-population basis before 1980, so no rate can be computed.
The counts still work, and on counts the pre-1980 mean over-report is **10.1%**.

## What a good answer says

**3/4:** Notices the blanks, identifies the cause, and redoes the comparison on
counts.

**4/4:** Answers the second half of the prompt — what you would have concluded
had you quietly dropped the blank years. You would have seen a series starting
in 1980 at its historical peak and falling ever since, and you would probably
have told a story about the 1980s as a turning point. **The pre-1980 data says
the level was already high and roughly flat.** Dropping missing rows is not
neutral; it silently chose a start date, which is the same error the midterm-loss
lab (Tue Oct 20) flags from the other direction.

**4/4, best:** notices that the missingness is *informative*. The Bureau began
publishing a citizen basis when non-citizens became numerous enough that the
distinction mattered. The gap in the data is itself a fact about the country.

***

# Option B — Midterms

No output; graded on reasoning.

**3/4:** Picks a direction and gives a reason.

**4/4:** Argues from the mechanism rather than intuition. The strongest case is
that over-reporting should be **larger** in midterms, because the social
pressure to have voted does not fall as much as actual turnout does — so the
same impulse operates over a smaller true base. A student who says that, and
then notes we would need midterm ballot counts (which are messier, since there
is no single national race), has done the work.

**Do not accept** "it would be the same" without an argument.

***

# Option C — Break the denominator

## Verified output

```{r c}
o <- cps[cps$year == 2024, ]
c(citizen_basis = round(100 * o$actual_votes / o$citizen_vap, 1),
  total_vap_basis = round(100 * o$actual_votes / (o$vap_thousands * 1000), 1))
```

**2024 turnout is 64.4% or 58.5%** depending on the denominator. Same ballots,
same year, **5.9 points apart**.

## What a good answer says

**3/4:** Reports both and says the denominator matters.

**4/4:** Says which is right *for what question*. Citizen VAP is right for "what
share of eligible people voted." Total VAP answers "what share of adults
voted," which includes people legally barred from voting and is therefore the
wrong denominator for a claim about civic engagement — but the right one if you
are asking how representative the electorate is of everyone living here.

**4/4, best:** connects it to the lab's own caveat. Our citizen VAP is derived
from the CPS, so the "actual" rate is a ballot count divided by a survey
estimate. **Neither number in the comparison is fully independent of the
survey.** That is uncomfortable and correct.

**This is the same "per what?" lesson as the midterm-loss lab**, which ran two
days ago on Tue Oct 20. Say so — the repetition two days apart is the point.

***

# Part 4 — turnout by race

2024 reported turnout: **White non-Hispanic 70.5%, Black 59.6%, Asian 57.1%,
Hispanic 50.6%.** Twenty points between the top and bottom.

Two cautions worth making explicit in class. These are **reported** figures and
carry the same over-report, which need not be equal across groups. And the
Asian and Hispanic estimates rest on much smaller subsamples, so they are less
precise — the table publishes no margin of error, which is exactly what the ACS
handbook reading (Session 3) said to look for and not find.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is refusing the easy story. "People lie" is a 2.
  Naming three candidate mechanisms and saying the data cannot separate them
  is a 4.
- **Reward anybody who is disturbed by the closing point** — that this is the
  only checkable survey question they will meet. That discomfort is the correct
  response and should be validated, not resolved.

## Anticipated questions

- *"So are surveys useless?"* — No. The survey got the *level* roughly right and
  the *trends between groups* very likely right. It was wrong by a few points on
  a socially loaded question. That is a real limitation and not a fatal one.
- *"Why does the Bureau publish a number that is wrong?"* — Because it is
  publishing what people said, accurately. The error is in our reading, not in
  their reporting. Worth five minutes; it is the difference between a bad
  measurement and a measurement of something other than what you assumed.
- *"Why not use total ballots instead of presidential votes?"* — Because there
  is no clean national total ballots series. The lab says so. The effect is
  small and makes our gap slightly too large.
- *"Is 60,000 households enough?"* — Yes, for national estimates; that is a
  much larger sample than any public poll. **Sample size is not the problem
  here**, which is the point students should take into Session 9 next Thursday.

## Connections

- **Sessions 2 and 3 (the census and the ACS)** — same agency, same survey apparatus, different question.
  The ACS margin-of-error reading pays off directly in Part 4.
- **Session 9 (polls, models, and markets)** — that lab simulates a poll failing; this one shows a
  real one failing. Run this first and refer back.
- **The midterm-loss lab (Tue Oct 20)** — the "per what?" lesson, second appearance.
- **Ch. 12 (participation), on the November Tuesdays** — the turnout gaps by race in Part 4 are
  the chapter's subject, and students will already have met them here with the
  caveat attached.
