---
title: "Lab 3: What People Say They Did"
subtitle: "84-355 Democracy's Data | Week 3 — Surveys"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# The question

**This is the third family: the survey.** Somebody asks people something, and we
treat the answers as facts about the country.

You have now seen two ways a data source can be built. Returns are assembled
from fifty-one separate authorities and nobody owns the whole. The census has
exactly one owner, a legal mandate, and uncertainty it cannot show you on the
face of the number. A survey is a third thing again: it does not try to reach
everybody, and it says so.

**That is the interesting part.** Returns and the census are both attempts to
count everyone and both fall short. A survey gives up on that by design, asks a
few thousand people, and then multiplies. Whether the multiplication is honest
is the question of the week.

That is an extraordinary thing to do when you think about it. Nobody checks. The
whole apparatus rests on people telling the truth to a stranger, and on the
people who answered resembling the people who did not.

Almost always we cannot test that. Today we can — because there is one survey
question whose true answer is a matter of public record.

**The Census Bureau asks about 60,000 households, after every federal election,
whether they voted.** And we know how many ballots were counted.

# Part 1 — The data

```{r load}
cps <- read.csv("data/cps_turnout.csv", stringsAsFactors = FALSE)
str(cps)
```

One row per presidential election, 1964 to 2024. Two quantities matter:

- `cps_rate` — turnout among citizens according to the **survey**.
- `actual_rate` — turnout among citizens according to the **ballots**.

```{r head}
head(cps[, c("year", "reported_voters", "actual_votes", "cps_rate", "actual_rate")], 5)
```

# Part 2 — What the survey says, and what happened

```{r gap-2016}
one <- cps[cps$year == 2016, ]
cat("2016\n")
cat("  Said they voted:", format(round(one$reported_voters), big.mark = ","), "\n")
cat("  Ballots cast:   ", format(one$actual_votes, big.mark = ","), "\n")
cat("  Difference:     ",
    format(round(one$reported_voters - one$actual_votes), big.mark = ","), "\n")
```

Roughly **eight and a half million people** told the Census Bureau they had
voted in an election in which they did not appear at the polls.

Some of that is people misremembering. Some is people saying the thing they
think they should have done. Some is that the survey's weights are imperfect,
so the answers of those who responded are being stretched to cover those who
did not. Nothing in the data tells you which.

# Part 3 — The gap over time

```{r gap-table}
recent <- cps[cps$year >= 1980, ]
data.frame(year        = recent$year,
           survey_says = round(recent$cps_rate, 1),
           ballots_say = round(recent$actual_rate, 1),
           gap_points  = round(recent$over_report_pp, 1))
```

```{r gap-plot}
plot(recent$year, recent$over_report_pp, type = "b", pch = 19, lwd = 2,
     col = "#C41230", ylim = c(-1, 8),
     xlab = "", ylab = "Survey minus ballots (percentage points)", main = "")
abline(h = 0, lwd = 1)
grid(col = "grey88", lty = 3)
```

## Part 3b — The same thing, in people

Rates need a denominator the Bureau only started publishing in 1980. Counts do
not, so this version reaches all the way back.

```{r counts-all}
cps$gap_people <- cps$reported_voters - cps$actual_votes
cps$gap_pct    <- 100 * cps$gap_people / cps$actual_votes

data.frame(year      = cps$year,
           gap_millions = round(cps$gap_people / 1e6, 1),
           over_by_pct  = round(cps$gap_pct, 1))
```

```{r counts-era}
era <- cut(cps$year, c(1963, 1979, 1999, 2030),
           labels = c("1964-1976", "1980-1996", "2000-2024"))
round(tapply(cps$gap_pct, era, mean), 1)
```

**In 1972 the survey overstated the electorate by 13%** — more than ten million
people. In the last three elections the average is under 3%.

**This is the finding of the lab, and it is not the one you would expect.**

In 1988 the survey overstated turnout by **7 points**. By 2020 the gap had gone
— the survey slightly *under*stated it. The problem people warn you about, the
one where respondents inflate their own civic virtue, was real and large and has
mostly disappeared over forty years.

Nobody decided to fix it. Ask what else changed.

# Part 4 — Who was asked

A survey's answer is only as good as its coverage. The same table reports
turnout by race, and those numbers carry the same caveat as everything above:
they are what people **said**.

```{r by-race}
r <- cps[cps$year >= 2008, c("year", "pct_white_nh", "pct_black",
                             "pct_asian", "pct_hispanic")]
names(r) <- c("year", "White (NH)", "Black", "Asian", "Hispanic")
r
```

Two things to notice. The gaps between groups are large and persistent. And the
Asian and Hispanic figures rest on far smaller subsamples than the White figure,
so they are less precise — the table does not say by how much, which is itself
worth noticing.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Go back further, and hit a wall.** Change `cps$year >= 1980` in
Part 3 to `cps$year >= 1964` and re-run. The rate columns come back **empty**
for the first four elections. That is not a bug: **the Bureau did not publish a
citizen-population basis before 1980**, so the rate cannot be computed. The
counts in Part 3b still work. Redo the comparison on counts for all sixteen
elections and say whether the story survives — and then say what you would have
concluded if you had quietly dropped the years that came back blank.

**Option B — Midterms.** This file has presidential years only, because that is
what we could match to counted ballots. Would you expect the over-report to be
larger or smaller in a midterm, and why? Say what data you would need.

**Option C — Break the denominator.** Every rate here divides by *citizen*
voting-age population, which is itself estimated by this same survey. Recompute
using total voting-age population instead — `100 * cps$actual_votes / (cps$vap_thousands * 1000)`
— and compare. How much of the "turnout rate" is a choice about the denominator?

**Option D — Your own question.** Sixteen elections, one row each.

# What this lab cannot tell you

Say these out loud before you use any of these numbers.

- **"Actual votes" are votes for president**, not ballots cast. Some people vote
  and skip the top of the ticket, so real turnout is slightly higher than shown
  and the gap slightly smaller.
- **The denominator comes from the survey itself.** We are checking a survey
  partly against arithmetic derived from that survey.
- **This is one question.** Voting is unusual: it is socially desirable, and it
  is verifiable. Most survey questions — what you believe, whom you trust, how
  you will vote next month — have no public record to check against at all.

That last point is the one to carry forward. **Today is the only day this term
you will be able to grade a survey against the truth.** Every poll you read for
the rest of the semester is asking you to take on faith the thing you just
watched fail by seven points.

***

# What today was really about

Three families down. Each is unreliable in its own direction, and the habit that
catches one is useless against the others.

| | **Returns** (Wk 1) | **Census** (Wk 2) | **Surveys** (Wk 3) |
|---|---|---|---|
| Who makes it | nobody in particular | one agency, by law | a pollster or an agency |
| Ambition | count every ballot | count every person | ask a few thousand and multiply |
| How it breaks | stitching separate sources together | one number carrying hidden uncertainty | the people who answered are not the people who did not |
| Check first | does it match the certified count? | what was counted, and with what error? | who was asked, how many refused, how were the rest weighted? |
| What you saw | a file **7,781 votes** short | a seat decided by **89 people**, inside an error bar of hundreds of thousands | **8.6 million** people who said they voted and did not |

Notice what is *not* on that list: none of the three failed because somebody was
careless. Every one of them failed in the way its method makes it liable to
fail. That is why the family is worth knowing before the file is.

**The fourth family — administrative records — arrives in Week 5**, when you open
campaign finance filings. Ask the question that belongs to it: *what was this
collected for?* Nobody files with the FEC in order to help you study elections,
and what a form does not ask for is not in the data.

# On the discussion board

By **Friday, 11:59 p.m.** — that evening or the next day, while it is fresh.
No length requirement; a sentence or two is enough, so long as it shows you
engaged with what we did. Any *one* of the usual four: something that surprised
you, something we missed, something you do not understand, or an answer to a
classmate.

Three that would make good posts:

- The over-report shrank from 7 points to nothing. Give one explanation, and say
  what evidence would test it.
- If a survey is wrong by 7 points about something checkable, what should you
  assume about its answers on things nobody can check?
- Which of the three caveats above bothers you most, and why?

***

### Where these numbers come from

U.S. Census Bureau, Current Population Survey, *Voting and Registration
Supplement*, Table A-1 (November 1964 to 2024),
<https://www.census.gov/data/tables/time-series/demo/voting-and-registration/voting-historical-time-series.html>.
Counted ballots are the national presidential totals from the Week 4 lab.

The Bureau publishes percentages, not counts. The voter counts here are derived
from its published figures — the arithmetic is in `data/build-data.R`, and it
is worth reading if you want to see exactly how much of a "finding" is a
calculation somebody chose to do.
