---
title: "Hispanic Origin Is Not a Race on the Census Form"
type: brief
subtitle: "Two questions about one person on the 2020 census form, and the two boxes people use to answer neither"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 2
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("census-race-code.R")
```

```{r setup, include=FALSE}
```

The census, people say, tells us what race Americans are. On the 2020 form, question 8 asked whether the person was of Hispanic, Latino or Spanish origin. Question 9 asked what the person's race was. Between them the Bureau printed a note telling respondents to answer both, and explaining why:

> For this census, Hispanic origins are not races.

That sentence is a decision, made by the federal government, about the relationship between two things people are routinely asked to be one thing about. So what does the census actually measure when it measures race? And what happens when the form's categories and people's self-understanding pull apart? Both questions can be answered from the record, because the form that separates race from origin also publishes the two answers together.

## The test

The data is the [2020 Census Redistricting Data (P.L. 94-171) Summary File](https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/), the counts certified as of 1 April 2020, read here for every county in the United States. A 1975 statute obliges the Bureau to hand this file to every state within a year of Census Day so that legislatures can draw districts from it.^[Public Law 94-171, now 13 U.S.C. § 141(c): <https://www.law.cornell.edu/uscode/text/13/141>.] The form's seven questions, and how one of them becomes most of the published file, are the subject of the [decennial chapter](../census-decennial/census-decennial-brief.html).

It is the right file for two reasons. First, it is the operative one. Every legal claim about racial fairness in American elections is stated in its categories. A vote-dilution claim under §2 of the Voting Rights Act, the claim that a district map spreads a minority group's voters so thinly that they can elect no one, turns on whether the group could form a majority in a district, a number computed from this file.^[52 U.S.C. § 10301: <https://www.law.cornell.edu/uscode/text/52/10301>.] Language-assistance coverage under §203, which obliges some counties to print ballots in a second language, is decided in the same terms.^[52 U.S.C. § 10503: <https://www.law.cornell.edu/uscode/text/52/10503>.] If the categories fit people badly, the misfit does not stay in the data; it propagates into the districts, the lawsuits and the money.

Second, the file reports race *crossed with* Hispanic origin, so it can be used to interrogate its own categories. The American Community Survey asks the same two questions in more detail, but it uses the same boxes, so it inherits the misfit rather than measuring it.

The boxes themselves are not the Bureau's. They come from the **Office of Management and Budget**, a White House office whose Statistical Policy Directive No. 15 binds every federal agency. The version governing 2020, written in 1997,^[Office of Management and Budget, "Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity," *Federal Register* 62 (30 October 1997): 58782, <https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf>.] set five minimum race categories and two ethnicity categories, required that ethnicity be asked first as a separate question, and allowed more than one race. It also said what the categories are not: socio-political constructs, in its own words, not scientific or anthropological ones.

The same text defined White as "a person having origins in any of the original peoples of Europe, the Middle East or North Africa". A respondent whose family came from Lebanon or Iran had no other place to go. In March 2024 the directive was rewritten again — a single combined question, a new Middle Eastern or North African category, compliance by 2029.^[Office of Management and Budget, "Revisions to OMB's Statistical Policy Directive No. 15," *Federal Register* 89 (29 March 2024): 22182, <https://www.federalregister.gov/documents/2024/03/29/2024-06469/revisions-to-ombs-statistical-policy-directive-no-15-standards-for-maintaining-collecting-and-presenting-federal-data-on-race-and-ethnicity>.]

## The data

The file has six tables, and this brief reads three of them, named as the Bureau names them: P1 (race), P2 (Hispanic origin by race) and P5 (group quarters, the people counted in prisons, dormitories, nursing homes and the like). The other three repeat P1 and P2 for people 18 and over, and count housing units. The rows here are counties. The Bureau publishes the same tables for every census tract and block, the neighbourhood-sized and block-sized areas that districts are actually drawn from, and those are left aside. Each county is identified by its FIPS code, the five-digit number the federal government assigns every county (`01001` is Autauga County, Alabama). It is kept as text rather than as a number so that the leading zero survives: a code that has quietly become 1,001 no longer matches the same county in any other table. County names come from the Census Gazetteer, the Bureau's list of places and their codes, matched on that code.

```{r file}
```

One clean row per county, `r n(nrow(d))` of them: every county and county equivalent (Louisiana's parishes, Alaska's boroughs, Virginia's independent cities) in the fifty states and the District of Columbia. The variables, by name: `total`; the six race-alone counts (`white`, `black`, `aian`, `asian`, `nhpi`, `other_race`) plus `two_or_more`; and the origin pair `hispanic` and `not_hispanic`. An `nh_` version of every race column counts only non-Hispanic respondents, and the group-quarters columns run `gq_correctional` through `gq_military`. The rows sum to `r n(sum(d$total))` people, which is the 2020 resident population of the United States exactly. Every share below is a national one.

Two things to know about the numbers before reading them. The Bureau injected privacy noise: small deliberate adjustments so that no individual's answers can be reconstructed, exact at the state level, close at the county level.^[The Bureau's own account of the method: *Disclosure Avoidance for the 2020 Census: An Introduction* (2021), <https://www.census.gov/library/publications/2021/decennial/2020-census-disclosure-avoidance-handbook.html>.] And the file's arithmetic was checked before anything was drawn from it:

```{r checks}
```

Nothing that follows is a data-quality problem. The count is exact, it breaks apart without error, and the counties roll up to the Bureau's own published state and national totals to the person. It is the categories that are in question.

> **Before you read on, commit to two numbers.** In the 2020 census,
> `r n(tot_other)` people answered the race question by choosing "Some Other
> Race", and `r n(tot_two)` reported two or more races. The form asked
> about Hispanic origin separately, one question earlier. **What share of
> each of those two groups do you think said yes to it?** Write both
> percentages down before you read on.

## What it says about democracy

Because Hispanic origin is asked separately, a person can be, and millions are, both Hispanic and white. Which means "how white is this state?" is a question with two correct answers.

```{r two-static, eval=!knitr::is_html_output(), fig.height=7.6}
```

```{r two-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Two correct answers to "how white is this state?", in all fifty states and DC.** Each state appears once with two readings joined by a rule: the share answering the race question "white", and the share who are white *and* not Hispanic. The distance between the dots is the point of the figure: it is the people who are both, and it is why the form is a dumbbell rather than a bar. The dashed line is 50%, and `r wd(nrow(flip))` states have one reading on each side of it. States are ordered here by the race-question reading; in the HTML edition the buttons above the figure reorder them by the size of the gap, which is a different argument, or alphabetically, which is none.

**New Mexico is `r pc(S("New Mexico","white_pct"))`% white by the race question and `r pc(S("New Mexico","nhwhite_pct"))`% white once the origin question is counted.** By one measure it is a majority-white state; by the other it is not close. Both numbers are correct, they answer different questions, and almost nobody reporting a figure like this says which one they used.

New Mexico is the widest case but not a lonely one. `r Wd(nrow(flip))` states — `r paste(flip$state, collapse = ", ")` — are majority white by the race question and not majority white once the origin question is counted, so the choice of definition decides whether they get described as majority-minority, the usual label for a state where non-Hispanic whites are no longer a majority, at all. Across all `r nrow(st)` states the gap tracks the Hispanic share almost mechanically. The correlation, a measure of how closely two quantities move together in which 1 means perfect lockstep and 0 means no relationship at all, is `r pc(cor_gap, 3)`: `r pc(S("New Mexico","gap"))` points in New Mexico, `r pc(S("Mississippi","gap"))` in Mississippi, where few people are both.

So far this is an accounting consequence of asking two questions. It becomes something else in the people who answered the race question by rejecting every box on it. `r pc(S("New Mexico","other_pct"))`% of New Mexicans and `r pc(S("Texas","other_pct"))`% of Texans chose "Some Other Race" — `r pc(100 * tot_other / sum(d$total))`% of the country — and the file says who they are. It says so by subtraction: P1 counts everyone who chose that box, `r n(tot_other)` people; P2 counts how many of them were not Hispanic, `r n(nh_other)`; the difference, `r n(hisp_other)`, is everyone who was.

```{r other-who}
```

**`r pc(pct_other_h)`% of the people who rejected every race box are Hispanic.** The leftover category on the race question is, in practice, an ethnicity box. The form asks a question a large minority of respondents do not think is well posed, then records their refusal as though it were an answer.

That would be a contained problem if it were the only such box. The second unusual column is `two_or_more`, normally read as a story about mixed-race families, with Hawaii as the obvious place to look. Most readers' written-down guess for it is in the teens or twenties.

```{r two-race-who}
```

**`r pc(pct_two_h)`% of everyone in the country reporting two or more races is Hispanic.** Put every race answer beside the origin answer and both hatches show at once:

```{r sankey-static, eval=!knitr::is_html_output(), fig.height=5.2}
```

**Figure 2. Every race answer in the 2020 census, split by the answer to the separate origin question.** All `r n(sum(rc$tot))` people counted in the fifty states and DC. Bands are ordered by how Hispanic each race answer turned out to be. The two at the top are the two leftover categories: "`r rc$race[1]`" at `r pc(rc$share[1])`% Hispanic and "`r rc$race[2]`" at `r pc(rc$share[2])`%. The next answer down is `r pc(rc$share[3])`%, and everything below it behaves nothing like the top two. In the HTML edition, hovering a ribbon gives both of the percentages a cell of a cross-tabulation has: the share of that race answer, and the share of that origin answer.

```{r sankey-d3, results='asis', eval=knitr::is_html_output()}
```

The counties say the same thing more sharply. **`r Wd(top_two_tx)` of the ten most multiracial counties in the country are in Texas**, strung along the Rio Grande; the tenth is Miami-Dade. Every one of them is between `r pc(min(top_two$hisp_pct))`% and `r pc(max(top_two$hisp_pct))`% Hispanic. Hawaii's most multiracial county, `r hi_top$county` at `r pc(hi_top$two_pct)`%, ranks `r hi_rank[1]`th of `r n(nrow(d))` and does not make the list.

```{r top-two}
```

```{r scatter-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r scatter-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. The multiracial share of a county is largely a function of its Hispanic share.** One dot per county, all `r n(nrow(d))` of them; a scatter, because the claim is a relationship between two shares. The correlation is `r pc(cor_two, 3)`. Hawaii's five counties are picked out because Hawaii is where most people locate multiracial America. They do sit high — `r pc(hi_state_two)`% of the state reports two or more races, and its counties rank between `r hi_rank[1]`th and `r hi_rank[2]`th of `r n(nrow(d))` — but they sit high while barely Hispanic at all, which puts them off the trend rather than at the top of it. They are the exception that a national scatter can show and a six-state one could not.

So the obvious reading is wrong twice over. "Two or more races" is not primarily mixed-race families, and it is not primarily Hawaii. **It is the second escape hatch from the same badly-posed question, used by the same population.** The mechanism is on the form itself. The 2020 questionnaire added write-in lines under the race boxes, and Hispanic respondents who had previously ticked "Some Other Race" alone often wrote origins the Bureau coded as two races. The multiracial count on that border jumped accordingly.^[The Bureau describes the redesign and what it did to the coding in "Improvements to the 2020 Census Race and Hispanic Origin Question Designs, Data Processing, and Coding Procedures" (August 2021): <https://www.census.gov/newsroom/blogs/random-samplings/2021/08/improvements-to-2020-census-race-hispanic-origin-question-designs.html>.]

That is the finding about democracy, and it is worth stating as a rule. **A trend line that crosses a category rewrite partly measures the rewrite.** The 2010-to-2020 rise in multiracial America compares instruments as much as people. The same rule now points forward: the 2024 revision merges race and Hispanic origin into one question by 2029. That change will end the two-answers problem in Figure 1 and make Figures 2 and 3 impossible to extend.

The categories these charts measure are federal policy, revised in 1997 and 2024. Every downstream series is stated in the categories as they stood that decade: districts, §2 claims, funding formulas. The population the 2020 form fit worst is precisely the population those protections most concern.

## What this brief cannot tell you

**A count is not a rate.** Every share above is of the resident population as enumerated, so places with large group-quarters populations — a prison, a campus, a base — carry people counted where the facility is rather than where they live. The `gq_` columns are in the file so that this can be checked rather than assumed.

**What people meant is not in the file.** The write-in text behind "Some Other Race", where respondents said in their own words what the boxes could not hold, is coded into categories before publication and never released. This brief can show that a hatch was used; it cannot show what was written in it.

**The noise is unquantified.** The privacy adjustments were applied before release and the file carries no column saying by how much. These are counties; the blocks that districts are actually built from are smaller by orders of magnitude, and that is where the noise bites hardest.^[Measured for redistricting by Christopher T. Kenny et al., "The Use of Differential Privacy for Census Data and Its Impact on Redistricting: The Case of the 2020 U.S. Census," *Science Advances* 7, no. 41 (2021): eabk3283, <https://pmc.ncbi.nlm.nih.gov/articles/PMC8500496/>.]

**Decades do not compare.** The 2020 form added write-in areas and examples the 2010 form lacked, and the coding changed with them, so a 2010-to-2020 change in these columns is not a change in people alone.

**Residents, not citizens.** The file counts everyone living in a county, which is what apportionment uses; it says nothing about citizenship.

**Counties, not districts.** Every number here is a county, and counties are not the units districts are built from. The relationships hold across `r n(nrow(d))` counties; whether they hold across blocks is a different question, and the privacy noise is worse there.

## What you should have learned

- Race and Hispanic origin are two separate questions by federal rule, so
  "how white is New Mexico?" has two correct answers:
  `r pc(S("New Mexico","white_pct"))`% and
  `r pc(S("New Mexico","nhwhite_pct"))`%. In `r wd(nrow(flip))` states the two
  answers fall on opposite sides of 50%.
- The race question has two leftover boxes, and one population uses both:
  nationally `r pc(pct_other_h)`% of "Some Other Race" and
  `r pc(pct_two_h)`% of "two or more races" respondents are Hispanic.
- Multiracial geography follows the seam, not Hawaii: the correlation between
  a county's Hispanic share and its multiracial share is
  `r pc(cor_two, 3)` across all `r n(nrow(d))` counties, and `r wd(top_two_tx)` of
  the ten most multiracial counties are on the Texas border.
- The categories are set by the Office of Management and Budget, rewritten in
  1997 and 2024, so a trend crossing a rewrite partly measures the rewrite.
- The counts themselves are exact and internally consistent: both
  decompositions hold in all `r n(nrow(d))` counties, and the counties sum to
  the published national figure of `r n(sum(d$total))` to the person — which is
  how a file can be arithmetically perfect and still measure its own
  categories' misfit.

## Extensions

None of these is answered above, and every one can be answered by sorting or
grouping in a spreadsheet.

- `pl94171_counties.csv` carries both `white` and `nh_white`. Take the difference for every county, divide by `total`, and sort. Where does the choice between the two definitions change the answer most?
- `other_race` and `two_or_more` are the boxes people reach for when none of the offered ones fit. Compute their combined share of `total`, county by county. Which places does the form fit worst?
- `hispanic` and `not_hispanic` should add to `total` in every row. Check it, then explain to somebody who has not read this brief why Hispanic origin is missing from the race columns.
- `gq_correctional` counts people in prison, counted where the facility is. Divide it by `total` and sort. What would the top counties' shares do to a district drawn from this file?

Two stretch ideas, beyond the spreadsheet. Sort the file to Wayne County, Michigan and its neighbours, where the Middle Eastern and North African population is largest, and ask whether the White column behaves like a box that fits — it is the population the 2024 revision creates a category for. Or pull the 2010 P.L. 94-171 file and compare `two_or_more` county by county, to put a number on how much of the multiracial rise is the instrument rather than the people.

## Sources

**Data**

- U.S. Census Bureau, *2020 Census Redistricting Data (P.L. 94-171) Summary
  File*, tables P1, P2 and P5, for every county in the fifty states and the
  District of Columbia. Collected by the Bureau in the 2020 count and published
  two ways: as one national dataset, `2020/dec/pl`
  (<https://api.census.gov/data/2020/dec/pl>), which needs a free key to
  download, and as one archive per state at
  <https://www2.census.gov/programs-surveys/decennial/2020/data/01-Redistricting_File--PL_94-171/>;
  the numbers here are checked against both. County names come from the
  Census Gazetteer
  (<https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html>),
  except
  Connecticut's, which the Gazetteer now carries as nine planning regions
  against this file's eight counties — two vintages of a boundary, each
  correct on its own date.
- U.S. Census Bureau, *2020 Census Questionnaire*. The two questions at the
  head of this brief are its questions 8 and 9, and the sentence between them
  is the note the Bureau printed to explain why both had to be answered.
  <https://www2.census.gov/programs-surveys/decennial/2020/technical-documentation/questionnaires-and-instructions/questionnaires/2020-informational-questionnaire.pdf>

**Literature and law**

- Office of Management and Budget (1997). "Revisions to the Standards for the
  Classification of Federal Data on Race and Ethnicity." Statistical Policy
  Directive No. 15, *Federal Register* 62: 58782 (30 October 1997) — the
  standard governing the 2020 census.
  <https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf>
- Office of Management and Budget (2024). "Revisions to OMB's Statistical
  Policy Directive No. 15." *Federal Register* 89: 22182 (29 March 2024) —
  the combined question and the Middle Eastern or North African category.
  <https://www.federalregister.gov/documents/2024/03/29/2024-06469/revisions-to-ombs-statistical-policy-directive-no-15-standards-for-maintaining-collecting-and-presenting-federal-data-on-race-and-ethnicity>
- Voting Rights Act § 2, 52 U.S.C. § 10301, and § 203, 52 U.S.C. § 10503 —
  the enforcement the race and Hispanic-origin questions exist to serve.
  <https://www.law.cornell.edu/uscode/text/52/10301> and
  <https://www.law.cornell.edu/uscode/text/52/10503>
- Nobles, M. (2000). *Shades of Citizenship: Race and the Census in Modern
  Politics.* Stanford: Stanford University Press.
- Prewitt, K. (2013). *What Is Your Race? The Census and Our Flawed Efforts to
  Classify Americans.* Princeton: Princeton University Press.
- Kenny, C. T., Kuriwaki, S., McCartan, C., Rosenman, E. T. R., Simko, T. and
  Imai, K. (2021). "The Use of Differential Privacy for Census Data and Its
  Impact on Redistricting: The Case of the 2020 U.S. Census." *Science
  Advances* 7(41): eabk3283.
  <https://pmc.ncbi.nlm.nih.gov/articles/PMC8500496/>
- *Alabama v. U.S. Department of Commerce*, 546 F. Supp. 3d 1057 (M.D. Ala.
  2021) — Alabama's challenge to the privacy noise, dismissed.

**The data itself**

Every figure above is drawn from this table. It is plain CSV, it opens in a
spreadsheet, and it is yours to take further.

[`pl94171_counties.csv`](data/derived/pl94171_counties.csv)

```{r ai-prompt, echo=FALSE, results="asis"}
```
