---
title: "What the Roll Call Record Contains"
subtitle: "Every vote since 1789, and no note of what most of them were about"
author: "84-355 Democracy's Data"
date: "`r format(Sys.Date(), '%B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("rollcalls-source-code.R")
```

```{r setup, include=FALSE}
```

An election return is complete: every ballot counted lands in some total. A roll call looks equally official and is a different kind of object, because **the Constitution does not require one**.

Article I, Section 5 provides that the yeas and nays are entered on the Journal "at the Desire of one fifth of those Present." A recorded vote happens when somebody wants one. Everything else leaves no record of who was on which side: voice votes, unanimous consent, and the great bulk of legislative business.

So the roll call record is a sample of decisions, and the sampling was done by the people being recorded. That is the first thing to know about it. The second is that for most of American history, the record does not say what the vote was about.

## What the instrument is good for

`r nn(TOTAL)` recorded votes, House and Senate, from the 1st Congress to the `r LASTC`th. For each one, how every member voted, by name.

Nothing else in this book is like that. Election returns are anonymous by constitutional design. Surveys ask a sample and get self-reports. Voter files record whether you voted and never how. This is the only source where the behaviour of named individuals is recorded completely, by law, over two centuries. That is why the entire study of legislative coalitions, party discipline and ideological scaling rests on it.

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.2}
```

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. Recorded votes per Congress.** Hover for both chambers at any Congress. The marked rise after the 91st is not a change in how much Congress did. It is a change in how much of it was recorded, following rules that made recorded votes easier to demand.

The shape of Figure 1 is worth pausing on, because it is easy to read as a story about Congress and is mostly a story about recording. The jump after the 91st Congress follows procedural changes that made recorded votes easier to obtain in the House. Members did not suddenly start deciding more things. More of what they decided started leaving a trace.

**A count of roll calls is a measure of the recording rule, not of legislative activity.** Any series built on "votes per Congress" inherits that.

## The file does not say what most votes were about

Voteview's roll call file has a field for the question being voted on. Here is how often it is filled in.

```{r contab}
```

**`r nn(NOQ)` of `r nn(TOTAL)` roll calls have no recorded question**, or `r p1(100 - COVER)`% of them. And the gap is not scattered thinly across the series. It is *everything before the 101st Congress*: zero per cent recorded for two hundred years, then 98 and 100 per cent for the last four decades.

For the first two centuries, then, the record is complete about **behaviour** and empty about **content**. You can see exactly how every member voted on `r nn(NOQ)` occasions where nobody wrote down what the occasion was.

This is worth holding next to what the file is used for. Ideological scaling works from the pattern of agreement and disagreement alone. That means DW-NOMINATE and its relatives, the subject of the [dw-nominate chapter](../dw-nominate/dw-nominate-brief.html). It does not read the bills, and for most of the series **there is nothing to read**. That is not a criticism of the method; it is a description of what the method had available. But it means a sentence like "this member voted conservatively in the 1870s" is a statement about who they voted *with*, and cannot be a statement about what they voted *for*.

> **Before you read on, commit to a view.** Two members of the 45th Congress
> agree on 90% of roll calls. **What do you actually know about them**, and
> what would you need in order to say they agreed about anything in
> particular?

## Not every recorded vote is the same kind of event

Where the question *is* recorded, this is what it is.

```{r qtab}
```

**Final passage is `r p1(PASS)`% of all roll calls.** The rest are amendments, procedural motions, cloture, tabling, nominations, and motions to suspend the rules.

These are not interchangeable. A member may vote against a bill on final passage having voted for every amendment that shaped it. Or vote to end debate on a bill they intend to oppose. Or vote to table a measure they support, because the timing is wrong. A procedural vote can be the substantive one, as cloture in the Senate frequently is. And a substantive vote can be a free vote whose outcome was never in doubt.

A method that treats all `r nn(TOTAL)` alike is making an assumption. It is often a reasonable one, and it is never a fact.

## Most recorded votes are not close

```{r martab}
```

```{r fig2-static, eval=!knitr::is_html_output(), fig.height=3.4}
```

```{r fig2-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. Roll calls by how large the losing side was.** Votes at the top carry almost no information about who disagrees with whom, which is the raw material every scaling method depends on. Hovering also gives the running total down the bands, which is the quantity that decides how much of the record a scaling method can actually use: `r p1(sum(mar$pct[1:2]))`% of recorded votes are unanimous or near-unanimous.

`r p1(UNAN)`% of recorded votes are unanimous and `r p1(LOP)`% have a losing side under a tenth. A 430–0 vote is a real event with real meaning in the chamber. It tells you nothing whatever about which members differ from which others, and that is the only thing a scaling method can use.

Against that, `r p1(CLOSE)`% are genuinely close. The informative half of the record is roughly half of it, and which half you are looking at depends entirely on the question you asked.

> **A second prediction.** Suppose a reform made recorded votes automatic for
> every question, including those now decided by voice. **Would measured
> polarization rise or fall?** Say why, in terms of Figure 2.

## What this chapter cannot tell you

**Nothing here measures what was never voted on.** A bill that dies in committee, an amendment ruled out of order, a question the leadership declines to schedule: none of it appears. No analysis of this file can recover it. The agenda is set by people with an interest in the outcome, and it is invisible in exactly the way a survey's unasked questions are invisible. This is the largest limitation of the source and the one least often stated.

The file is **researcher-assembled**, not a government publication. Poole, Rosenthal, Lewis and colleagues (1997) built it from the Journals for the early Congresses and from Clerk and Senate feeds for the modern ones. It is enormously convenient, meticulously done, and under no legal obligation to anybody. Its gaps, including the missing questions, are inherited along with the convenience.

This chapter also does not distinguish sincere from strategic voting, because nothing in the record can. A recorded vote is an act, not a preference, and the gap between them is where much of legislative politics lives.

Finally, "how every member voted" is not quite complete either. Members are absent, are paired, vote "present", or are announced without voting. Those outcomes are recorded and are not the same as a yea or a nay, and how a particular analysis treats them is a decision that this chapter has not made for you.

## Sources

- Lewis, Jeffrey B., Keith Poole, Howard Rosenthal, Adam Boche, Aaron Rudkin
  and Luke Sonnet. *Voteview: Congressional Roll-Call Votes Database.*
  `HSall_rollcalls.csv`. <https://voteview.com/>
- U.S. Constitution, Article I, Section 5, clause 3.
- Poole, Keith T. and Howard Rosenthal (1997). *Congress: A Political-Economic
  History of Roll Call Voting.* Oxford University Press.
- Roberts, Jason M. (2007). "The Statistical Analysis of Roll-Call Data: A
  Cautionary Tale." *Legislative Studies Quarterly* 32(3): 341–360.
- Clinton, Joshua D. (2012). "Using Roll Call Estimates to Test Models of
  Politics." *Annual Review of Political Science* 15: 79–99.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`content.csv`](data/derived/content.csv), [`margins.csv`](data/derived/margins.csv), [`questions.csv`](data/derived/questions.csv), [`volume.csv`](data/derived/volume.csv)

**Your turn**

None of these is answered above, and every one can be answered from the
tables you just opened.

- `content.csv` gives `question_recorded_pct` by era. Find when the record starts saying what votes were about. What can you not ask about the years before that?
- `margins.csv` bands the votes by how close they were. Work out the share that were unanimous or nearly so. What does that do to a study of contested votes?
- `questions.csv` gives the most common questions, with `pct_of_recorded` and `pct_of_all`. Compare the two columns. Why are they so different?
- `volume.csv` counts roll calls per Congress for both chambers. Plot it. Is a rise in recorded votes a rise in disagreement, or a change in procedure?

```{r ai-prompt, echo=FALSE, results="asis"}
```
