---
title: "Data Session 8 Key, part 1: You Are the Sample"
subtitle: "84-355 Democracy's Data | Session 8 — Surveys, two ways | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
b <- read.csv("data/ces2024_benchmarks.csv", stringsAsFactors = FALSE)
b$shift <- b$pct_weighted - b$pct_unweighted
```

All code is given to students. Grade the interpretation.

***

# ⚠ Before you teach this

**Replace `data/class_responses_EXAMPLE.csv`.** Export the first-day CES
replication from Google Forms, recode to CES numeric codes (listed in
`data/build-data.R`), and save as `data/class_responses.csv` with two columns,
`variable` and `code`, one row per student per question. The lab picks up the
real file automatically and prints a warning while the placeholder is in use.

**The placeholder numbers are invented.** Everything in the CES columns is real;
nothing in the class column is, until you replace the file.

***

**Where this sits.** Session 8 runs two surveys side by side: this one (a single
instrument, up close) and the CPS turnout lab (one question, sixty years). Part 1
closes the loop opened on the first day of class — students took the CES
replication before they knew what a survey was for, and now they meet the
instrument it was copied from.

Every CES number below was run against the 2024 Common Content, n = 60,000.

***

# The session in one sentence

The class is not the country, most of the reasons are boring, the interesting
one cannot be separated from the boring ones — and the survey we are comparing
against has itself been adjusted by nineteen points on the question that matters
most.

***

# The findings

```{r headline}
head(b[order(-abs(b$shift)),
       c("variable", "category", "pct_unweighted", "pct_weighted", "shift")], 5)
```

- **Voter registration moves 91.5% → 72.5% when weighted. Nineteen points**, the
  largest adjustment in the file.
- Strong Democrats **27.6% → 21.9%**.
- Following news "most of the time" **51.2% → 46.4%**.
- With n ≈ 19, the class's own margin of error is about **±22 points.**

***

# The three things to get said

**1. The weighted and unweighted columns are the same 60,000 people.** Students
skate past this. Stop and make somebody say what a weight *is*: a number
attached to a respondent saying how many people like them the sample should have
had. **Every CES figure ever quoted at them has been through this**, and the raw
column is what the researchers actually collected.

**2. What the nineteen points are correcting is self-selection, not error.**
Nobody lied about being registered. People who agree to join an online political
panel are simply far more likely to be registered than people who do not. **The
adjustment is a confession about the sample**, and it is published openly, which
is more than most polls do.

**3. Nineteen people cannot detect anything.** Do the arithmetic on the board:
±22 points. Then connect it to the other half of the session — the CPS has
60,000 respondents and *still* got turnout wrong, because its error was bias,
not sampling. **Sample size and bias are different problems and neither fixes
the other.** If students leave with one sentence, this is the one.

**The line for the board:** a bigger sample makes you more precise about
whatever your sample happens to be.

***

# Handling the room

**Do not let this become a session about how unusual CMU students are.** That is
true, boring, and unfalsifiable here — we have no other classroom. The
interesting content is entirely about *the instrument*.

**Anticipate the good question:** "if the class is unrepresentative, why bother?"
The answer is that the class is not the object of study — the **comparison** is.
Students learn what a benchmark is by having their own answers benchmarked, and
they will read the next poll differently for it.

***

# Option A — A different variable

Grade on whether the student sorts the difference into the right bin:

- **Structural**: education, age. Certain, uninteresting, requires no data.
- **Political**: party, ideology, news attention. Interesting, confounded.
- **Noise**: anything under about twenty points, given n ≈ 19.

**4/4** requires naming which bin and why. **A student who reports a
fifteen-point gap as a finding has failed Part 4** and should be marked
accordingly however fluent the writing.

***

# Option B — Collapse the categories

## Verified output

```{r bb}
p   <- b[b$variable == "pid7", ]
grp <- ifelse(p$code %in% 1:3, "Democrat",
       ifelse(p$code == 4,     "Independent",
       ifelse(p$code %in% 5:7, "Republican", "Not sure")))
aggregate(cbind(pct_unweighted, pct_weighted) ~ grp, data = cbind(p, grp), sum)
```

**Collapsed, the raw CES sample is D+13.5. Weighted it is D+1.9.** The
adjustment moves the headline partisan balance by more than eleven points.

**3/4:** Collapses correctly and reports that the class's cells were too small
to read at seven categories.

**4/4:** Says what collapsing hid. Seven-point party ID separates **leaners**
from independents, and leaners vote like partisans — so the "Independent" column
shrinks from 13.2% to a much smaller genuinely-unaffiliated group once you know
the breakdown. Collapsing makes the number stable and throws away the
distinction the seven-point scale exists to capture.

**4/4, best:** Notices this is a general trade — smaller cells are noisier and
more informative — and connects it to the class's n. With nineteen students you
*must* collapse to say anything, and collapsing is exactly what makes the
comparison less interesting. **There is no sample size at which this design
works well**, which is a real conclusion.

***

# Option C — Which column is the right comparison?

No single correct answer; grade the argument.

**The case for unweighted:** the class survey has no weights, so comparing it to
a weighted estimate compares an adjusted number to an unadjusted one. Like to
like means raw to raw.

**The case for weighted:** the question students actually care about is "how
does this room differ from the country," and the weighted column is the best
available estimate of the country. The unweighted column is an estimate of
nothing — it describes CES panellists.

**3/4:** Picks a side with a reason.

**4/4:** Recognises that the two comparisons answer **different questions** —
raw-to-raw asks "how do our respondents differ from their respondents," and
class-to-weighted asks "how does this room differ from America." Both are
legitimate; they are not interchangeable.

**4/4, best:** Points out that neither is available for the class, because a
class of nineteen **cannot be weighted to anything** — there is no target
population it is a sample of. The class is a census of itself. That reframing is
the most sophisticated answer available.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is refusing to treat a class-vs-country difference
  as a finding without addressing the margin of error.
- **A 2** is a list of differences with no account of which are real.
- **Reward anyone who asks whether the class answered honestly** in a room where
  the instructor could see their screen. That is a genuine mode effect and it is
  not in the lab.

## Anticipated questions

- *"Is the CES a random sample?"* — No. It is an opt-in online panel matched to a
  target sample and then weighted. Say this plainly; students assume "60,000"
  implies rigour, and size is not design.
- *"Why is registration weighted down so hard?"* — Because panel members are
  unusually engaged. The weights push the sample back toward CPS/ACS benchmarks
  for the adult population.
- *"Could we weight our class?"* — To what? There is no population of which this
  room is a sample. Good question, useful answer.
- *"Are our answers anonymous?"* — Say what is actually true of your Google Form
  setup **before** the first day, not now.

## Connections

- **Session 3 (ACS margins)** — the first appearance of published uncertainty.
- **Session 8 part 2 (CPS)** — same session: a huge probability sample that is
  still wrong. Teach these two back to back and name the contrast.
- **Session 9 (polls, models, markets)** — *next week*, and this is the
  groundwork: students who understand weighting will understand house effects.
- **Session 6 (voter files)** — the CES asks whether you are registered; the
  voter file *records* it. Self-report against administrative record, again.
