---
title: "Data Session 8, part 1: You Are the Sample"
subtitle: "84-355 Democracy's Data | This room against the Cooperative Election Study"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"Forty-two percent of Americans say…"*

On the first day of class you filled in a survey. It was not an icebreaker. It
was a **replication of the Cooperative Election Study** — the same questions, in
the same words, with the same answer options, because that is the only way a
comparison means anything.

Today you meet the real thing. The **CES 2024** interviewed **60,000 people**,
which makes it one of the largest political surveys in the world, and it is free
for anyone to download.

You are about to find out how much this room resembles the country. **The
interesting part is not that it doesn't.** It is *which* differences are real,
which are your age, and which are an artefact of how both surveys were built.

# Part 1 — The two samples

```{r load}
b <- read.csv("data/ces2024_benchmarks.csv", stringsAsFactors = FALSE)

f <- if (file.exists("data/class_responses.csv"))
       "data/class_responses.csv" else "data/class_responses_EXAMPLE.csv"
cls <- read.csv(f, stringsAsFactors = FALSE)

cat("class file in use:", f, "\n")
if (grepl("EXAMPLE", f))
  cat("\n*** THIS IS THE PLACEHOLDER FILE. The numbers below are invented.\n",
      "*** Real class responses have not been loaded yet.\n")

n_class <- max(table(cls$variable))   # one row per student per question
c(students = n_class,
  questions = length(unique(cls$variable)),
  rows = nrow(cls),
  ces_respondents = 60000,
  ratio = round(60000 / n_class))
```

The CES has roughly **three thousand times** as many people as this room. Keep
that number in mind for Part 4, when it turns out to matter much less than you
would think.

# Part 2 — Side by side

```{r compare}
share <- function(v) {
  x <- cls$code[cls$variable == v]
  bb <- b[b$variable == v, ]
  cnt <- as.vector(table(factor(x, levels = bb$code)))
  data.frame(category = bb$category,
             class_pct = round(100 * cnt / sum(cnt), 1),
             ces_unweighted = bb$pct_unweighted,
             ces_weighted   = bb$pct_weighted,
             stringsAsFactors = FALSE)
}

share("pid7")
```

```{r compare-ideo}
share("ideo5")
```

```{r plot-pid}
p <- share("pid7")
m <- rbind(class = p$class_pct, CES = p$ces_weighted)
barplot(m, beside = TRUE, names.arg = sub(" ", "\n", p$category),
        col = c("#C41230", "#2c7fb8"), las = 2, cex.names = 0.6,
        ylab = "percent", legend.text = c("this room", "CES 2024 (weighted)"))
```

Read the party-ID and ideology tables together and say what you see before
reading on.

# Part 3 — What weighting does

Look again at the last two columns of Part 2. They are **the same 60,000
people**, counted twice.

```{r weights}
b$shift <- b$pct_weighted - b$pct_unweighted
head(b[order(-abs(b$shift)),
       c("variable", "category", "pct_unweighted", "pct_weighted", "shift")], 6)
```

**The largest adjustment in the entire file is voter registration: 91.5% of CES
respondents say they are registered, and the weights cut that to 72.5%.**
Nineteen points.

That is not a correction for a bad question. It is a correction for **who agreed
to be in an online panel in the first place.** People who opt into political
surveys are much more likely to be registered, more likely to follow the news —
`newsint` moves nearly five points — and more likely to be strong partisans.
Strong Democrats fall from 27.6% to 21.9%.

**Every CES number you have ever seen quoted has been through this.** The
weighted column is what gets reported. The unweighted column is what the
researchers actually collected.

```{r weights-plot}
v <- b[b$variable == "votereg", ]
m <- rbind(raw = v$pct_unweighted, weighted = v$pct_weighted)
barplot(m, beside = TRUE, names.arg = v$category,
        col = c("grey65", "#2c7fb8"), ylab = "percent",
        legend.text = c("what they collected", "what they report"))
```

Now the uncomfortable question: **our class survey has no weights at all.**
Nobody built them, because you cannot weight a sample of nineteen to a
population. So the class column in Part 2 is the equivalent of the grey bars —
raw, uncorrected, and describing whoever happened to be in the room.

# Part 4 — What the class comparison can and cannot show

```{r educ}
share("educ")
```

Some differences here are **certain, structural, and boring**. You are all
students, so education looks nothing like the country's. That tells you nothing
about politics; it tells you that a class of undergraduates is a class of
undergraduates.

```{r age-note}
share("newsint")
```

Others are **genuinely interesting but confounded**. If the class follows news
less than the country, is that because you are 20, or because you are at
Carnegie Mellon, or because our survey was taken in class where nobody had
prepared for it? **The design cannot separate these.** One room, one moment, no
comparison group.

**And here is the part worth sitting with.** Suppose the class differs from the
CES by fifteen points on some question. With nineteen people, the margin of
error on our own estimate is roughly:

```{r moe}
n <- n_class
round(100 * 1.96 * sqrt(0.25 / n), 1)
```

**About twenty-two points.** So a fifteen-point gap is not a finding. It is
noise, and it would take a much bigger room before it stopped being noise.

This is the same lesson as part 2 of today's session from the other end: there,
a survey of 60,000 got turnout wrong by systematic bias that no sample size
could fix. Here, a survey of nineteen cannot detect anything, however true it
is. **Sample size and bias are different problems, and neither one fixes the
other.**

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — A different variable.** `share()` works for any of the seven:
gender4, educ, race, pid7, ideo5, newsint, votereg. Pick one nobody discussed,
report the comparison, and say whether the difference is structural, political,
or noise.

**Option B — Collapse the categories.** Seven-point party ID has eight
categories and nineteen people, so most cells have one or two students. Collapse
to three (Democrat / Independent / Republican) by summing codes 1–3, 4, and 5–7,
and redo the comparison. Does the picture change, and what did collapsing hide?

**Option C — Which column is the right comparison?** Part 2 gives you both the
weighted and unweighted CES. For comparing against *this class*, which one
should you use, and why? There is a real argument on both sides — make it.

**Option D — Your own question.** Seven variables, 39 benchmark categories,
one room.

# What this lab cannot tell you

- **Whether the class is unusual for a class.** We have no other classroom to
  compare against. Every difference gets attributed to "being students," which
  is untestable here.
- **Anything with confidence.** Nineteen people. See Part 4.
- **Whether people answered honestly.** A survey taken in a room where the
  instructor can see your screen is not a private survey. The CES is taken alone
  at home, which is a different instrument even when the words are identical.
- **What the CES itself gets wrong.** It is an opt-in online panel matched to a
  target sample, not a random sample of Americans. Weighting adjusts for what
  the researchers knew to adjust for. Part 2 of today's session shows the CPS —
  a genuine probability sample, ten times the size — still missing turnout.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- 91.5% of CES respondents say they are registered; the weights cut it to 72.5%.
  What is that adjustment actually doing, and would you trust it?
- Our class differs from the country on something. Pick the difference you find
  most interesting and say whether you think it is real.
- You answered these questions before you knew what they were for. Would you
  answer differently now?

***

### Where the data came from

**CES 2024** — Cooperative Election Study Common Content, 2024, Data Release 2
(Ansolabehere, Schaffner and Pope), Harvard Dataverse,
<https://doi.org/10.7910/DVN/X11EP6>. 60,000 respondents, 694 variables,
184 MB. The committed benchmark file is derived from it; the weight is
`commonweight`.

**Value labels were verified against the official Guide to the 2024 CES**, not
assumed — which mattered, because the guide lists categories in a different
order than their numeric codes.

**Class responses** come from the replication survey you took on the first day
of class, recoded to CES codes. If the file above says `EXAMPLE`, the class
numbers on this page are invented placeholders and only the CES columns are
real.
