---
title: "Data Session 5 Key, part 1: Following the Money"
subtitle: "84-355 Democracy's Data | Session 5 — Money, message, and attention | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
fec <- read.csv("data/fec_candidates_2024.csv", stringsAsFactors = FALSE)
```

```{r vals, include=FALSE}
# Every figure quoted in this key is computed here and inserted inline, using
# the same definitions as the brief, so the two cannot drift apart.
.h   <- fec[fec$office == "House", ]
.hs  <- .h[.h$ici %in% c("I", "C", "O"), ]
.ser <- .hs[.hs$ttl_receipts > 50000, ]
.ser$pac_share <- 100 * .ser$pac_contrib / .ser$ttl_receipts
.sen <- fec[fec$office == "Senate" & fec$ici %in% c("I", "C", "O") &
            fec$ttl_receipts > 50000, ]
.sen$pac_share <- 100 * .sen$pac_contrib / .sen$ttl_receipts

MEDH  <- median(.h$ttl_receipts)
MNH   <- mean(.h$ttl_receipts)
.r    <- sort(.h$ttl_receipts, decreasing = TRUE)
TOP10 <- 100 * sum(.r[1:round(0.10 * length(.r))]) / sum(.r)
BOT50 <- 100 * sum(.r[(round(0.5 * length(.r)) + 1):length(.r)]) / sum(.r)

MED  <- function(k) median(.hs$ttl_receipts[.hs$ici == k])
PACH <- function(k) median(.ser$pac_share[.ser$ici == k])
PACS <- function(k) median(.sen$pac_share[.sen$ici == k])
RATIO <- MED("I") / MED("C")

SRC <- sapply(c("C", "O", "I"), function(k) {
  x <- .ser[.ser$ici == k, ]
  c(Individuals = sum(x$indiv_contrib), PACs = sum(x$pac_contrib),
    Party = sum(x$party_contrib), Self = sum(x$self_funding)) /
    sum(x$ttl_receipts) * 100
})

BH    <- fec$ttl_receipts[fec$cand_name == "HARRIS, KAMALA"][1]
BROWN <- fec$ttl_receipts[grepl("BROWN, SHERROD", fec$cand_name)][1]
MEDSI <- median(.ser$ttl_receipts[.ser$ici == "I"])

# duplicate groups above the $1m floor used in Option C
.dg  <- local({
  b <- fec[fec$ttl_receipts > 1e6, ]
  tb <- table(b$ttl_receipts); shv <- as.numeric(names(tb[tb > 1]))
  sub <- b[b$ttl_receipts %in% shv, ]
  split(sub, sub$ttl_receipts)
})
NDUP  <- length(.dg)
NHS   <- sum(sapply(.dg, function(g) all(c("House", "Senate") %in% g$office)))
NCOIN <- sum(sapply(.dg, function(g) length(unique(sub(",.*", "", g$cand_name))) > 1))
GF    <- fec$ttl_receipts[fec$cand_name == "GOODLANDER, MAGGIE"][1]

pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(round(x), big.mark = ",", trim = TRUE)
dl <- function(x) paste0("$", nn(x))
dm <- function(x) paste0("$", pc(x / 1e6), " million")
```

All Part 8 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

Organised money in American politics is not distributed by competitiveness,
ideology, or merit — it is distributed by **who currently holds the seat** — and
the ledger that proves this is also incapable of telling you who won.

***

# The three findings

```{r headline}
h  <- fec[fec$office == "House", ]
hs <- h[h$ici %in% c("I", "C", "O"), ]

c(median = round(median(h$ttl_receipts)), mean = round(mean(h$ttl_receipts)))

aggregate(ttl_receipts ~ ici, data = hs, FUN = median)

ser <- hs[hs$ttl_receipts > 50000, ]
ser$pac_share <- 100 * ser$pac_contrib / ser$ttl_receipts
aggregate(pac_share ~ ici, data = ser, FUN = function(x) round(median(x), 1))
```

1. **Concentration.** Median House candidate raises **`r dl(MEDH)`**; mean is
   **`r dl(MNH)`**, ~`r round(MNH / MEDH)`× higher. Top 10% of candidates raise
   **`r pc(TOP10)`%** of all House money; bottom 50% raise **`r pc(BOT50)`%**.
2. **Incumbency.** Median incumbent **`r dl(MED("I"))`**; median challenger
   **`r dl(MED("C"))`**. A ratio of **`r round(RATIO)`:1**.
3. **PAC money follows office, not contests.** Among serious House campaigns
   (>$50k), PACs supply **`r pc(PACH("I"))`%** of the median incumbent's money,
   **`r pc(PACH("C"))`%** of a challenger's, **`r pc(PACH("O"))`%** in an open seat.

Finding 3 is the one to spend class time on. Finding 2 is ambiguous between "money
makes incumbents" and "incumbents attract money." Finding 3 is not: an open-seat
race is *maximally* competitive and PACs fund it at `r pc(PACH("O"))`%. They are
not buying influence over outcomes; they are buying access to people who already
have power.

***

# Option A — Does the Senate work the same way?

## Verified output

```{r a}
o <- fec[fec$office == "Senate" & fec$ici %in% c("I","C","O") &
         fec$ttl_receipts > 50000, ]
o$pac_share <- 100 * o$pac_contrib / o$ttl_receipts
aggregate(pac_share ~ ici, data = o, FUN = function(x) round(median(x), 1))
aggregate(ttl_receipts ~ ici, data = o, FUN = function(x) round(median(x)))
table(o$ici)
```

Senate PAC shares: incumbents **`r pc(PACS("I"))`%**, challengers
**`r pc(PACS("C"))`%**, open seats **`r pc(PACS("O"))`%**. Same direction, much
smaller magnitude. Sample is small — `r sum(.sen$ici == "I")` incumbents,
`r sum(.sen$ici == "C")` challengers, `r sum(.sen$ici == "O")` open-seat
candidates above the threshold.

## What a good answer says

**3/4:** Reports that the pattern holds but is weaker, and notices the small
sample.

**4/4:** Gets the mechanism. PAC contribution limits are *fixed per committee per
election* while Senate campaigns are enormous. The same $10,000 that is a
meaningful slice of the `r dm(MEDSI)` a typical serious House incumbent raises is
a rounding error in Sherrod Brown's `r dm(BROWN)`. PAC money has not gone away in
Senate races; it has been
**diluted by individual money** flowing in at a scale House races never see. A
student who reasons from "the denominator changed" rather than "PACs care less
about senators" has understood it.

**4/4 bonus:** notes that this makes the House the place where PAC money has the
most *leverage*, which is the opposite of where most public attention goes.

**Common stumble:** "PACs prefer the House." The absolute dollars say otherwise;
it is the share that falls.

***

# Option B — Where does the money come from?

## Verified output

```{r b}
s <- fec[fec$office == "House" & fec$ici %in% c("I","C","O") &
         fec$ttl_receipts > 50000, ]
src <- sapply(c("C","O","I"), function(k) {
  x <- s[s$ici == k, ]
  c(Individuals = sum(x$indiv_contrib), PACs = sum(x$pac_contrib),
    Party = sum(x$party_contrib), Self = sum(x$self_funding)) /
    sum(x$ttl_receipts) * 100
})
colnames(src) <- c("Challenger", "Open seat", "Incumbent")
round(src, 1)
```

| source | Challenger | Open seat | Incumbent |
|---|---|---|---|
| Individuals | `r pc(SRC["Individuals","C"])` | `r pc(SRC["Individuals","O"])` | `r pc(SRC["Individuals","I"])` |
| PACs | **`r pc(SRC["PACs","C"])`** | `r pc(SRC["PACs","O"])` | **`r pc(SRC["PACs","I"])`** |
| Party | `r pc(SRC["Party","C"])` | `r pc(SRC["Party","O"])` | `r pc(SRC["Party","I"])` |
| Self | **`r pc(SRC["Self","C"])`** | `r pc(SRC["Self","O"])` | **`r pc(SRC["Self","I"])`** |

## What a good answer says

**3/4:** Notices the aggregate PAC gap (`r pc(SRC["PACs","C"])` vs
`r pc(SRC["PACs","I"])`) is much narrower than the median gap
(`r pc(PACH("C"))` vs `r pc(PACH("I"))`), and can explain why: the aggregate is dominated by a
handful of enormous challenger campaigns that *do* attract PAC money, while the
typical challenger gets none at all.

**4/4:** Identifies **self-funding as what fills the gap** —
`r pc(SRC["Self","C"])`% for challengers and `r pc(SRC["Self","O"])`% in open
seats, against `r pc(SRC["Self","I"])`% for incumbents. Then draws the
conclusion that matters: if you want to challenge a sitting member of Congress
and you are not independently wealthy, the organised money will not come, and
the shortfall is made up out of the candidate's own pocket or not at all. That is
a barrier to entry with a dollar figure attached.

**4/4, best:** answers the "which summary is fairer" question honestly — neither;
they answer different questions. The median describes the typical candidate's
experience, the aggregate describes where the money in the system actually sits.
A journalist wanting a dramatic number takes `r pc(PACH("C"))`% vs
`r pc(PACH("I"))`%. A journalist wanting an accurate picture of the flow of funds
takes both.

**Common stumble:** treating the two sets of numbers as a contradiction, or
assuming one must be an error. Address this directly — it is the most
transferable statistical idea in the lab.

***

# Option C — How big is the double-counting problem?

## Verified output

```{r c}
for (fl in c(1000000, 100000, 10000)) {
  b <- fec[fec$ttl_receipts > fl, ]
  d <- table(b$ttl_receipts); sh <- as.numeric(names(d[d > 1]))
  cat("floor $", format(fl, big.mark = ","), ": ", nrow(b), " candidates, ",
      sum(b$ttl_receipts %in% sh), " rows sharing a total\n", sep = "")
}
```

At the $1,000,000 floor the chunk finds **`r NDUP`** shared totals, not a
handful, and nearly all of them are genuine. Biden/Harris carry
**`r dl(BH)`** on two records. **`r NHS`** of the groups pair a House filing with
a Senate one — members who moved committee money across when they ran for the
Senate in 2024, of whom **Gallego, Trone, Schiff** and **Rosen** are only the
four largest — and **Hovde** appears under two Senate IDs in the same race.

Only **`r NCOIN`** groups at this floor put two different surnames on the same
total, and one of those is Biden/Harris. The other is worth knowing before a
student finds it: **Goodlander** (NH-02) and the write-in presidential candidate
**Fishback** share `r dl(GF)` — and, on inspection, *every other financial column
in the row as well*. That is a defect in the source file rather than a
coincidence, and it is the sharpest thing in this option.

## What a good answer says

**3/4:** Reports that lowering the floor finds more apparent duplicates, and
suspects some are coincidences.

**4/4:** Reasons about *why* coincidences appear — at low dollar values, round
numbers cluster. Two candidates who each raised exactly $12,500, or exactly
$15,000, are not the same person; both of those collisions are in the file at the
$10,000 floor. The method is most trustworthy where the amounts are large and
precise to the cent, which is where it was used in Part 6.

**Full credit, and give it readily:** a student who finds the Goodlander/Fishback
pair and concludes that *size and precision are not sufficient either* has beaten
the key's own argument with the data. Do not mark that wrong.

**4/4, best:** answers the "what would you actually need" question. The right
key is not the dollar amount but the **candidate's committee ID or their name
plus state** — and better still, the FEC's own linkage between candidate IDs and
committee IDs, which exists in a different bulk file. A student who says "I would
join on identity, not on a value that happens to be equal" has learned Session 1's
lesson properly.

***

# Option D — Their own question

Reward curiosity. Debt, cash-on-hand, and single-state comparisons all work.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The central intellectual move** is distinguishing "money causes winning" from
  "winning attracts money." Part 4 is designed so that the open-seat column
  settles it. Students who use that column in their argument have done the real
  work; students who only compare incumbents to challengers have not.
- **Reward students who notice a number is doing less than it appears to.** The
  median-versus-aggregate split in Option B is the sharpest instance.

## Anticipated questions

- *"Doesn't this prove money buys elections?"* — No, and the lab is careful not
  to claim it. It shows money *follows* office. Whether it also causes winning is
  the question the FEC file cannot answer, which is Part 7.
- *"Where is the super PAC money?"* — Not here. This file covers candidates'
  own committees only. Independent expenditures are reported separately and are
  enormous post-*Citizens United*. Worth saying out loud, because students will
  otherwise conclude they have seen all the money. They have seen perhaps half.
- *"Why does the FEC not record who won?"* — Because it regulates finance, not
  elections. This is the same structural fact as Session 1: no national election
  administrator exists. Connect the two explicitly.
- *"Is `r dl(MED("C"))` really the median challenger?"* — Yes. Most challengers are not
  serious candidates and raise almost nothing. This is itself a finding about how
  few congressional races are genuinely contested — and it points straight back
  at Session 4's **13 competitive states**.

## Connections forward

- **Session 4 (returns over time)** — the vanishing battleground and the `r round(RATIO)`:1 funding gap are the same
  phenomenon measured two ways. Uncontested seats do not attract money.
- **`campaign-visits`** (retired, on disk) — money and attention both concentrate on the same
  small set of races. Ask whether the causation runs in a circle.
- **`media-attention`, the demo in this same session** — most of this money is spent on advertising, which is where
  the media unit picks the thread up.
- **Session 1 (election returns)** — the FEC file is authoritative and *still* needs deduplication and
  still lacks outcomes. Authority is not the same as completeness.
- The assigned readings this week (the Trump donor op-ed and the Harris donors
  interactive) both describe the **individual** side of the ledger. Part 4 is the
  organised side. Ask on Tuesday which one the coverage over-weights, and why.
