---
title: "Lab 5: Following the Money"
subtitle: "84-355 Democracy's Data | Week 5"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Chapter 4 asks who pays for campaigns and what they get for it. Today we open the
actual ledger.

Every federal candidate must report to the Federal Election Commission: what came
in, what went out, and — crucially — **who it came from**. That last column is
where this lab lives. It is easy to ask how much money is in politics. It is much
more interesting to ask *whose* money goes to *whom*, and today you can answer
that precisely.

Everything is base R. Work in order.

***

# Part 1: The ledger

```{r read}
fec <- read.csv("data/fec_candidates_2024.csv", stringsAsFactors = FALSE)
dim(fec)
```

Nearly four thousand candidates for federal office in the 2023–24 cycle. One row
each.

```{r glimpse}
head(fec[, c("cand_name", "office", "office_st", "ici", "party",
             "ttl_receipts", "indiv_contrib", "pac_contrib")], 4)
```

The columns that matter today:

- `ttl_receipts` — everything the campaign took in
- `indiv_contrib` — money from individual people
- `pac_contrib` — money from political action committees
- `party_contrib` — money from the party
- `self_funding` — money the candidate gave or lent their own campaign
- `ici` — **I**ncumbent, **C**hallenger, or **O**pen seat

That last one is the most important variable in the file, and you will see why
shortly.

```{r offices}
table(fec$office)
table(fec$ici)
```

***

# Part 2: The shape of the money

Start with the House, because there are enough candidates for the pattern to be
clear.

```{r house}
h <- fec[fec$office == "House", ]
nrow(h)
```

What does a typical House candidate raise?

```{r typical}
round(median(h$ttl_receipts))
round(mean(h$ttl_receipts))
```

> **Stop and look at those two numbers.** The median House candidate raised about
> $22,600. The average House candidate raised about $782,000 — **nearly
> thirty-five times more**. Whenever a mean is that far above a median, a small
> number of cases are carrying almost everything.

How far does it go?

```{r concentration}
r <- sort(h$ttl_receipts, decreasing = TRUE)
top10 <- round(0.10 * length(r))

round(100 * sum(r[1:top10]) / sum(r), 1)                     # top 10% of candidates
round(100 * sum(r[(round(0.5 * length(r)) + 1):length(r)]) / sum(r), 1)   # bottom 50%
```

The best-funded tenth of House candidates raised **72.7%** of all House money.
The bottom half raised **0.2%** — two tenths of one percent, split between more
than fifteen hundred people.

```{r skew-plot, fig.width = 7.5, fig.height = 5}
plot(seq_along(r) / length(r) * 100, cumsum(r) / sum(r) * 100, type = "l",
     lwd = 2.5, col = "#2166AC",
     xlab = "Candidates, best-funded first (%)",
     ylab = "Share of all House money raised (%)",
     main = "Almost all the money is with almost none of the candidates")
abline(0, 1, lty = 2, col = "grey60")
abline(v = 10, lty = 3, col = "grey40")
```

The dashed diagonal is what perfect equality would look like. We are nowhere near
it.

***

# Part 3: Incumbents and everyone else

Now split by whether the candidate already holds the seat.

```{r by-ici}
hs <- h[h$ici %in% c("I", "C", "O"), ]
aggregate(ttl_receipts ~ ici, data = hs, FUN = median)
```

```{r by-ici-plot, fig.width = 7, fig.height = 5}
meds <- aggregate(ttl_receipts ~ ici, data = hs, FUN = median)
meds <- meds[match(c("C", "O", "I"), meds$ici), ]

bp <- barplot(meds$ttl_receipts / 1000,
        names.arg = c("Challenger", "Open seat", "Incumbent"),
        col = c("#B2182B", "#999999", "#2166AC"), border = NA,
        ylab = "Median receipts ($ thousands)",
        main = "What a typical House candidate raises")
text(bp, meds$ttl_receipts / 1000, labels = paste0("$", round(meds$ttl_receipts / 1000)),
     pos = 3, xpd = TRUE, cex = 0.9)
```

> The median incumbent raised **$1,537,105**. The median challenger raised
> **$9,107**. That is a ratio of about **169 to 1**.

Before concluding anything, note what this cannot distinguish. Incumbents might
raise more *because* incumbency attracts money, or the people who attract money
might be the ones who become incumbents. Both stories fit this chart.

The next part separates them.

***

# Part 4: Whose money?

Here is the question the ledger can actually settle. Take every serious House
campaign — say, everyone who raised more than $50,000 — and ask what **share** of
their money came from PACs rather than individuals.

```{r pac-share}
ser <- hs[hs$ttl_receipts > 50000, ]
ser$pac_share <- 100 * ser$pac_contrib / ser$ttl_receipts

aggregate(pac_share ~ ici, data = ser, FUN = function(x) round(median(x), 1))
```

```{r pac-plot, fig.width = 7, fig.height = 5}
ps <- aggregate(pac_share ~ ici, data = ser, FUN = median)
ps <- ps[match(c("C", "O", "I"), ps$ici), ]

bp <- barplot(ps$pac_share, names.arg = c("Challenger", "Open seat", "Incumbent"),
        col = c("#B2182B", "#999999", "#2166AC"), border = NA, ylim = c(0, 45),
        ylab = "PAC share of receipts (%), median",
        main = "Where PAC money goes")
text(bp, ps$pac_share, labels = paste0(round(ps$pac_share, 1), "%"),
     pos = 3, xpd = TRUE, cex = 0.95)
```

> **This is the finding of the lab.** For a typical serious House incumbent,
> **37.6%** of all money raised came from PACs. For a challenger running just as
> serious a campaign, the figure is **1.1%**. For an open seat, **1.4%**.
>
> Individuals give to candidates they like. PACs give to people who already hold
> power — and they are not subtle about it. An open-seat candidate has just as
> much chance of winning as an incumbent's challenger does of losing, and PACs
> fund neither.

This is why the incumbency gap in Part 3 is not simply "good candidates attract
money." Organised money is choosing by *office held*, not by quality, competitive­
ness, or ideology.

***

# Part 5: Buying your own campaign

Some candidates skip donors entirely.

```{r self-fund}
big <- fec[fec$ttl_receipts > 100000, ]
big$self_pct <- 100 * big$self_funding / big$ttl_receipts

sum(big$self_pct > 50)
nrow(big)

head(fec[order(-fec$self_funding),
         c("cand_name", "office", "party", "ttl_receipts", "self_funding")], 6)
```

Of the 1,421 candidates who raised over $100,000, **225 — about one in six —**
put up more than half of it themselves.

***

# Part 6: The trap in this file

Look at the two best-funded candidates in the entire dataset:

```{r biden-harris}
fec[grepl("HARRIS, KAMALA|BIDEN, JOSEPH", fec$cand_name),
    c("cand_id", "cand_name", "ttl_receipts", "indiv_contrib")]
```

**Those are the same dollars.** Biden's presidential committee became Harris's
when he withdrew in July 2024. The FEC carries two candidate records — different
IDs, different names, one pile of money — and reports the full
**$1,175,189,365** against each.

It is not only the presidential race:

```{r dup-hunt}
b <- fec[fec$ttl_receipts > 1e6, ]
d <- table(b$ttl_receipts)
shared <- as.numeric(names(d[d > 1]))

head(b[b$ttl_receipts %in% shared,
       c("cand_id", "cand_name", "office", "office_st", "ttl_receipts")], 10)
```

Ruben Gallego, David Trone, Adam Schiff and Jacky Rosen each appear **twice** —
once as a House candidate and once as a Senate candidate — because each ran for
the Senate in 2024 and moved their committee money across. Eric Hovde appears
twice under two candidate IDs for the same Senate race.

How much does this matter?

```{r dup-cost}
round(sum(fec$ttl_receipts) / 1e6)

nodup <- fec[!duplicated(fec$ttl_receipts) | fec$ttl_receipts == 0, ]
round(sum(nodup$ttl_receipts) / 1e6)
```

> Add up the column naively and you get **$6.85 billion** raised in the 2024
> cycle. Strip out the repeated totals and you get about **$5.27 billion**. The
> difference — roughly **$1.6 billion** — is money that does not exist. It is the
> same money, counted twice.
>
> Note that the second number is a rough correction, not a right answer: dropping
> every repeated value would also drop two genuinely different candidates who
> happened to raise exactly the same amount. It is good enough to show you the
> *size* of the problem, and not good enough to publish. Knowing the difference
> between those two standards is most of what this course is about.

***

# Part 7: What this file cannot tell you

The FEC's file layout has a column for whether the candidate won the general
election. Here is how much of it is filled in:

```{r no-outcomes}
"gen_election" %in% names(fec)
```

The column is not in our file **because it is empty in the source** — all 3,856
rows of it. The FEC collects money, not results. It is a regulator of finance,
not an election administrator, and as you learned in Week 2, the United States
does not have one of those.

> So the most obvious question in this whole subject — **does the candidate who
> raises more money win?** — cannot be answered with campaign finance data alone.
> You would have to go and find election results from fifty-one separate
> authorities and join them on, and then you would be doing Week 2's lab again,
> with all of Week 2's problems.
>
> This is not a gap in our dataset. It is a fact about how the American state
> keeps its books.

What you *can* say from this file is nonetheless substantial: the money is
extraordinarily concentrated, it flows overwhelmingly to people who already hold
office, and the organised part of it flows almost nowhere else.

***

# Part 8: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change, or nothing at all.

Pick **one**.

***

### Option A — Does the Senate work the same way?

Everything above used the House. Change one word and find out.

```{r opt-a}
which_office <- "Senate"          # <- change to "House" or "President"

o <- fec[fec$office == which_office & fec$ici %in% c("I", "C", "O"), ]
o <- o[o$ttl_receipts > 50000, ]
o$pac_share <- 100 * o$pac_contrib / o$ttl_receipts

aggregate(pac_share ~ ici, data = o, FUN = function(x) round(median(x), 1))
aggregate(ttl_receipts ~ ici, data = o, FUN = function(x) round(median(x)))
table(o$ici)
```

**Write about:** the same pattern appears, but the PAC gap is much smaller in the
Senate — around 15% for incumbents against under 1% for challengers, rather than
37.6% against 1.1%. Suggest why. Senate races are far more expensive and far more
often competitive; does either fact help explain a *smaller* PAC share? Think
about what a PAC's maximum contribution is worth as a fraction of a $90 million
campaign versus a $2 million one.

***

### Option B — Where does the money come from?

```{r opt-b, fig.width = 7.5, fig.height = 5}
min_raised <- 50000               # <- change this threshold

s <- fec[fec$office == "House" & fec$ici %in% c("I","C","O") &
         fec$ttl_receipts > min_raised, ]

src <- sapply(c("C","O","I"), function(k) {
  x <- s[s$ici == k, ]
  c(Individuals = sum(x$indiv_contrib), PACs = sum(x$pac_contrib),
    Party = sum(x$party_contrib), Self = sum(x$self_funding)) / sum(x$ttl_receipts) * 100
})
colnames(src) <- c("Challenger", "Open seat", "Incumbent")
round(src, 1)

barplot(src, beside = FALSE, col = c("#2166AC", "#B2182B", "#999999", "#F4A582"),
        ylab = "% of all money raised", main = "Sources of House campaign money",
        legend.text = TRUE, args.legend = list(x = "topright", bty = "n", cex = 0.8))
```

**Write about:** these are shares of the *total* raised by each group, which is a
different question from Part 4's median candidate. Do the two tell the same
story? Which is the fairer summary, and why might a journalist prefer one? Note
what fills the gap for challengers where PAC money is absent — and what that
implies about who can afford to run against a sitting member.

***

### Option C — How big is the double-counting problem?

```{r opt-c}
floor_amt <- 1000000              # <- change this

b <- fec[fec$ttl_receipts > floor_amt, ]
d <- table(b$ttl_receipts)
shared <- as.numeric(names(d[d > 1]))

c(candidates_above_floor = nrow(b),
  rows_sharing_a_total   = sum(b$ttl_receipts %in% shared),
  dollars_double_counted = round(sum(shared) / 1e6))

b[b$ttl_receipts %in% shared, c("cand_name", "office", "office_st", "ttl_receipts")]
```

**Write about:** lower the floor to $100,000 and then to $10,000. Does the number
of apparent duplicates rise faster than you would expect? At some point matching
on dollar amount stops finding real duplicates and starts finding coincidences —
two candidates who each raised exactly $250,000, say. Where do you think that
point is, and how would you tell? What would you actually need in order to find
duplicates *properly*?

***

### Option D — Ask your own question

Pick a state and look at everyone who ran in it. Compare the parties. Find the
candidate with the most debt, or the most cash left over. Follow whatever nags.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a ledger and say something true about who
finances American politics.

***

## Where the data came from

**Federal Election Commission**, bulk download, "All Candidates" file for the
2023–24 cycle
(<https://www.fec.gov/files/bulk-downloads/2024/weball24.zip>), covering every
candidate committee's activity through 31 December 2024.

This is as close to an official source as anything in this course. It is a
mandatory legal filing to a federal regulator, not a scrape or a compilation.
That is worth registering after Week 2 — and it is also worth registering how
much this authoritative file still cannot do. It does not say who won. It does
not deduplicate candidates who change office mid-cycle. It does not distinguish
money spent *for* a candidate by outside groups, which since *Citizens United*
(2010) has become a very large share of what is actually spent on a race and
appears nowhere in these rows.

Independent expenditures, super PACs and so-called dark money are all real, all
enormous, and all outside this file. What you are looking at is the money that
passed through the candidates' own hands.

The script that downloaded, checked and trimmed this file is `data/build-data.R`.
It prints its own sanity checks, including a count of the duplicate records that
Part 6 is about — those are left in deliberately.
