---
title: "Data Session 5 Key, part 2: The Money Nobody Controls"
subtitle: "84-355 Democracy's Data | Session 5 — Independent expenditures | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
s  <- read.csv("data/ie_summary.csv",   stringsAsFactors = FALSE)
o  <- read.csv("data/ie_outliers.csv",  stringsAsFactors = FALSE)
cd <- read.csv("data/ie_candidates.csv", stringsAsFactors = FALSE)
cm <- read.csv("data/ie_committees.csv", stringsAsFactors = FALSE)
```

```{r vals, include=FALSE}
# Same definitions as the brief, so key and brief cannot drift apart.
# ROWS is the source file's row count. The 19 MB raw FEC file is not committed;
# this figure is the count recorded by data/build-ie-data.R when it ran, and is
# the one number in this key that cannot be re-derived from data/.
ROWS <- 73449

get  <- function(v, k) s$dollars[s$version == v & s$side == k]
RAWS <- get("raw", "support");     RAWO <- get("raw", "oppose")
CLNS <- get("cleaned", "support"); CLNO <- get("cleaned", "oppose")
RAW  <- RAWS + RAWO;               CLN  <- CLNS + CLNO
OUT  <- sum(o$amount); NOUT <- nrow(o)
SHARE <- 100 * OUT / RAW

FW    <- o[o$spender == "Food & Water Action", ]
FWN   <- nrow(FW); FWAMT <- sum(FW$amount)
WW    <- o[o$spender == "Gus Associates", ]
OUTMIN <- min(o$amount)

cd$off <- c(P = "President", S = "Senate", H = "House")[cd$office]
PA <- function(k) 100 * sum(cd$opposing[cd$off == k]) /
  sum(cd$supporting[cd$off == k] + cd$opposing[cd$off == k])
CASEY <- cd$pct_against[cd$candidate == "CASEY, ROBERT"]

# Re-derive any threshold at or above the $50m cut used to build ie_outliers.csv.
# Below that cut the removed rows are not in data/, so those table rows stay typed.
TH <- function(t) {
  drop <- o$amount > t
  sup  <- RAWS - sum(o$amount[drop & o$sup_opp == "S"])
  opp  <- RAWO - sum(o$amount[drop & o$sup_opp == "O"])
  c(removed = sum(drop), total = sup + opp,
    pct_sup = 100 * sup / (sup + opp), pct_opp = 100 * opp / (sup + opp))
}
THR <- function(t) unname(TH(t)["removed"])
THT <- function(t) bnk(unname(TH(t)["total"]), 2)
THP <- function(t) paste0(pc(unname(TH(t)["pct_sup"])), " / ",
                          pc(unname(TH(t)["pct_opp"])))

pc  <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
bnk <- function(x, k = 2) paste0("$", pc(x / 1e9, k), "bn")
dm  <- function(x) paste0("$", nn(x / 1e6), "M")
```

All code is given to students. Grade the interpretation.

**Where this sits.** The second half of Session 5, after the candidates' own
FEC filings. **It replaced a planned lab on the Meta and Google ad libraries,
which turned out to be closed** — Meta needs a token and identity verification;
Google publishes only through BigQuery, in spend ranges rather than figures.
The syllabus now spends a minute on that absence, and it is worth making: the
most detailed record of political persuasion in America is not public.

**This is the dirtiest dataset in the course**, and it is the point.

Every number below was run against the data files.

***

# The session in one sentence

The `r NOUT` rows in `data/ie_outliers.csv`, out of `r nn(ROWS)` in the source
file, inflate the federal record of outside spending `r round(RAW / CLN)`-fold
and reverse its most important finding, and they are still there.

***

# The findings

```{r headline}
c(rows_total     = ROWS,          # 73,449; see setup — the raw file is not committed
  outlier_rows   = nrow(o),
  raw_billions   = round(sum(s$dollars[s$version == "raw"]) / 1e9, 1),
  clean_billions = round(sum(s$dollars[s$version == "cleaned"]) / 1e9, 2),
  raw_pct_oppose   = s$pct[s$version == "raw"     & s$side == "oppose"],
  clean_pct_oppose = s$pct[s$version == "cleaned" & s$side == "oppose"])
```

- **`r bnk(RAW, 1)` raw, `r bnk(CLN)` real.** `r NOUT` rows carry
  **`r pc(SHARE)`%** of the raw total.
- **Support/oppose flips: `r pc(100 * RAWS / RAW)`/`r pc(100 * RAWO / RAW)` raw →
  `r pc(100 * CLNS / CLN)`/`r pc(100 * CLNO / CLN)` cleaned.**
- **Senate races are `r pc(PA("Senate"))`% negative, House
  `r pc(PA("House"))`%, President `r pc(PA("President"))`%.**
- Bob Casey: **`r pc(CASEY)`%** of the outside money around him was spent
  against him.

***

# The three things to get said

**1. There are two different problems here and students must separate them.**
The junk filings (*THE COURT OF DIVINE JUSTICE*, *`r dm(WW$amount)` for Walter
White*) are **people lying to the government**. The `r FWN`-fold Food & Water duplicate is
**the government's own export repeating a row**. One is fraud, one is plumbing,
and they need different fixes. **Do not let the room lump them together as
"bad data."**

**2. The flip is the lesson, not the total.** Anyone can see that `r bnk(RAW, 1)` is
too big. The dangerous error is the one that looks fine: `r pc(100 * RAWS / RAW)`% supportive is a
perfectly plausible number that a student could put in a paper. **Wrongness that
announces itself is the safe kind.** Ask what else they have read that had a
quiet version of this problem.

**3. "The FEC publishes what is filed."** It is a disclosure agency, not an
auditor. Ask what checking would cost and who would do it — and note that the
alternative, an agency that decides which filings are real, has obvious
problems of its own. **There is no free version of this.**

**The line for the board:** official does not mean checked.

***

# Option A — Argue with the threshold

## Verified output

| cut at | rows removed | total | support / oppose |
|---|---|---|---|
| none (raw) | 0 | `r bnk(RAW, 1)` | `r pc(100 * RAWS / RAW)` / `r pc(100 * RAWO / RAW)` |
| $1M | 897 | $2.39bn | 47.7 / 52.3 |
| $5M | 102 | $3.92bn | 39.8 / 60.2 |
| $10M | 42 | $4.34bn | 40.9 / 59.1 |
| **$50M** | **`r THR(5e7)`** | **`r THT(5e7)`** | **`r THP(5e7)`** |
| **$100M** | **`r THR(1e8)`** | **`r THT(1e8)`** | **`r THP(1e8)`** |
| $500M | `r THR(5e8)` | `r THT(5e8)` | `r THP(5e8)` |

The three middle rows are the only figures in this key that are **not**
re-derived from `data/`. `ie_outliers.csv` holds the rows above the $50M cut, so
every threshold from $50M up recomputes here; the $1M, $5M and $10M rows were
recorded from the uncommitted 19 MB raw file when `data/build-ie-data.R` ran.

**3/4:** Runs two or three thresholds and reports that the answer moves.

**4/4:** Sees that **the flip survives everything from $1M to $100M** — every
threshold in that range gives roughly 40/60 — so the finding is robust across
two orders of magnitude. That is a much stronger claim than "it worked at the
number I picked."

**4/4, best:** Works out **why $500M breaks it.** Raising the cut that far lets
`r NOUT - THR(5e8)` rows back in, every one of them on the *supporting* side: the
`r FWN` identical Food & Water rows, `r bnk(FWAMT)` of duplicated money, together
with the `r dm(WW$amount)` Walter White filing. So the sensitivity is not
vagueness in the data; it is two specific artifacts re-entering — one a
duplicate, one a fabrication — and a student who diagnoses either has done real
work.

Also worth noticing: **$50M and $100M give identical answers**, because no row
sits between them — the smallest row removed at either cut is `r dm(OUTMIN)`. A
threshold with slack around it is more defensible than one tuned to the data.

***

# Option B — A better rule

No single answer; grade the design.

**3/4:** Proposes a plausible rule — flag committees that filed once, or amounts
above the cycle's known total.

**4/4:** Says what the rule would wrongly remove. A single-filing rule kills
every small genuine committee that spent once. A "purpose text looks fake" rule
requires somebody to judge text, which is the discretion problem from Session 11
in a new place.

**4/4, best:** Notices that the honest fix is not a rule at all but
**publication of the `r NOUT` rows alongside the total**, so a reader can decide.
That is what this lab does, and a student who arrives at it independently should
be told so.

***

# Option C — Negative by office

## Verified output

```{r c}
cd$off <- c(P = "President", S = "Senate", H = "House")[cd$office]
aggregate(cbind(supporting, opposing) ~ off, data = cd, sum)
```

Across the committed file: **Senate `r pc(PA("Senate"))`% against, House
`r pc(PA("House"))`%, President `r pc(PA("President"))`%.**

*(An earlier version of this key reported 71.5 / 69.1 / 42.6. Those were
computed on the full cleaned FEC extract, which is not what ships in `data/` —
and 42.6 was Harris's own `pct_against`, not the presidential aggregate. The
figures above re-derive from `ie_candidates.csv` and are what the chunk prints.
Every number in a key should reproduce from the committed data; these now do.)*

**3/4:** Reports the ranking.

**4/4:** Explains the presidential exception. Both 2024 presidential candidates
had very large *supportive* super PACs; down-ballot, outside groups mostly
attack, because attacking an unknown candidate is cheaper than introducing your
own. **Name recognition is the mechanism.**

**4/4, best:** Notes the strategic asymmetry — a negative ad about a little-known
Senate challenger can define them before they define themselves, which is not
available against a sitting president everybody already has an opinion about.
**The spending pattern is a function of how much the electorate already knows.**

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is separating fraud from plumbing, and noticing
  that the plausible error is more dangerous than the absurd one.
- **A 2** is "you can't trust government data." That is both lazy and wrong —
  the data is fine once you look at it, and looking took `r NOUT` rows.
- **Reward anyone who asks what the FEC is for.** Disclosure and verification
  are different jobs and Congress only funded one of them.

## Anticipated questions

- *"Is filing a fake report a crime?"* — Knowingly false FEC filings can be.
  Whether these were prosecuted is not in this file; do not speculate.
- *"How did nobody catch this?"* — Somebody probably did. The bulk file is the
  raw record; the FEC's own summary pages apply filters this file does not.
  **That is worth saying: the same agency publishes both a clean and a dirty
  version, and the dirty one is the one you download.**
- *"Who is Nader Akhlaghy?"* — A minor candidate named in the junk filings. We
  do not know why, and the file cannot say.
- *"Why is 'independent' spending independent?"* — Because *Buckley* and
  *Citizens United* permit unlimited spending only if uncoordinated. It is a
  legal category, and whether it describes reality is what enforcement fights
  are about. This connects straight to Tuesday's Ch. 4 debate.

## Connections

- **Session 5 part 1 (FEC candidate filings)** — same agency, same cycle, and a
  file that is clean. **Contrast them explicitly; it shows the problem is the
  form, not the FEC.**
- **Session 11 (EAVS)** — the other self-reported dataset, and the other place
  where the reporting entity is the entity being measured.
- **Session 1 (returns)** — the first "two official sources disagree" moment.
- **Session 14 (RPV)** — where a defensible-looking methodological choice
  changes a legal conclusion. Same shape as the threshold here.
