---
title: "Data Session 5, part 2: The Money Nobody Controls"
subtitle: "84-355 Democracy's Data | Independent expenditures, 2024"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Where did that number come from?

> *"Outside groups spent $4 billion on the 2024 election."*

You have just seen what candidates raise. This is the other pot: an
**independent expenditure** is money spent for or against a candidate by
somebody the candidate does not control — and, by law, may not coordinate with.

It is reported to the FEC separately, it is frequently larger than the
campaign's own spending, and it is the part of the disclosure system that works
least well. Today you find out how badly.

**Start by reading a number off the raw file.**

```{r first-look}
s <- read.csv("data/ie_summary.csv", stringsAsFactors = FALSE)
s[s$version == "raw", ]
```

**$49.7 billion**, of which 94.5% supports candidates and 5.5% opposes them.

Both of those numbers are wrong. One is wrong by a factor of ten. The other is
backwards.

# Part 1 — Sixteen rows

```{r outliers}
o <- read.csv("data/ie_outliers.csv", stringsAsFactors = FALSE)
nrow(o)
o[, c("spender", "cand_name", "amount", "pur")]
```

That is the entire problem. **Sixteen rows out of 73,449 — 0.02% of the file —
carry 90.4% of the dollars.** Read who filed them.

**"THE COMMITTEE OF 300."** **"THE COURT OF DIVINE JUSTICE."** **"The Masonic
Illuminati Eye."** Nearly ten billion dollars each, almost all naming one
obscure candidate, for purposes including **"DEPOSIT FOR WINNING IN ELECTION"**.
One committee reports **$255,000,000** spent on a candidate named **Walter
White**, for "funding for waltuh white's campaign."

These are not errors. They are **junk filings**, and they are sitting in the
federal government's official record of campaign spending, published exactly as
submitted.

**The FEC publishes what is filed. It does not check whether it is true.**

```{r dupes}
o[o$spender == "Food & Water Action", c("spender", "amount", "file_num")]
```

And then a completely different problem. Nine rows, all $114,056,874, all the
same committee, all supporting the same candidate — and **all carrying the same
file number.** One filing, exported nine times. Not a prank; a plumbing failure.

**Two kinds of bad data in one table**, and only one of them was anybody's
fault.

# Part 2 — What happens when you take them out

```{r cleaned}
s
```

```{r flip-plot}
m <- matrix(c(s$pct[s$version == "raw"], s$pct[s$version == "cleaned"]), nrow = 2)
barplot(m, beside = FALSE, names.arg = c("raw file", "16 rows removed"),
        col = c("#2c7fb8", "#C41230"), ylab = "percent of dollars",
        legend.text = c("supporting a candidate", "opposing a candidate"))
```

**The total falls from $49.7 billion to $4.75 billion** — which is the real
figure, and matches what watchdog groups report.

**And the headline reverses.** In the raw file, 94.5% of outside money supports
candidates. Once sixteen rows are gone, **57.6% of it is spent attacking
somebody.**

That is not a rounding change. It is the single most important substantive fact
about independent expenditure — *most of this money is negative* — and the raw
file tells you the opposite.

**How the cleaning was done, so you can argue with it:** every row above $100
million was dropped. That is a **threshold, not a diagnosis**. It happens to
catch exactly the sixteen rows above and nothing else, but it is crude, and a
real analyst would have to defend it. Part 5 asks you to.

# Part 3 — Who is being attacked

```{r candidates}
cd <- read.csv("data/ie_candidates.csv", stringsAsFactors = FALSE)
head(cd[, c("candidate", "office", "supporting", "opposing", "pct_against")], 10)
```

```{r cand-plot}
top <- head(cd[cd$office == "S", ], 6)
barplot(rbind(top$supporting, top$opposing) / 1e6, beside = TRUE,
        names.arg = sub(",.*", "", top$candidate), las = 2, cex.names = 0.7,
        col = c("#2c7fb8", "#C41230"), ylab = "$ millions",
        legend.text = c("supporting", "opposing"))
```

Look at the Senate races. **Sherrod Brown: 81.7% of the outside money aimed at
him was spent against him. Bob Casey: 87.6%.**

For the presidential candidates the split is closer to even — Harris 42.6%
against, Trump 55.0% — because both had large supportive super PACs. **Senate
challengers and incumbents mostly get attacked.**

Notice what this does to the phrase "money spent on a candidate." Money spent
*on* Sherrod Brown was overwhelmingly money spent trying to defeat him.

# Part 4 — Who is doing it

```{r committees}
cm <- read.csv("data/ie_committees.csv", stringsAsFactors = FALSE)
head(cm[, c("spender", "amount", "pct_against")], 12)
```

Some committees exist almost entirely to attack. Others almost entirely to
promote. **The same legal instrument does both**, and the disclosure form does
not distinguish between them beyond a single letter — `S` or `O` — in a column
most people never read.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Argue with the threshold.** The $100 million cut is arbitrary. Try
$50 million and $500 million (`o$amount > 5e7`). Does the support/oppose flip
survive? At what threshold does it stop working, and what does that tell you
about how robust the finding is?

**Option B — A better rule.** Rather than a dollar threshold, you could drop
rows whose *purpose* text looks fake, or whose committee filed only once, or
whose amount exceeds the entire cycle's real spending. Pick one, say how you
would implement it, and say what it would wrongly remove.

**Option C — Negative by office.** Compare `pct_against` for presidential,
Senate and House candidates in `ie_candidates.csv`. Which office attracts the
most negative money, and what about those races would explain it?

**Option D — Your own question.** 73,449 filings, 16 of them nonsense.

# What this lab cannot tell you

- **Whether the junk filings were prosecuted.** The FEC has an enforcement
  process; this file records what was filed, not what happened next.
- **Whether an ad worked.** We have dollars and a direction, not persuasion.
  The obvious next dataset would be the ad libraries — which are closed. Meta
  requires an access token and identity verification; Google reports spend in
  **ranges** rather than figures, through a cloud service. **The most detailed
  record of political persuasion in America is not public.**
- **Coordination.** "Independent" is a legal claim, not an observed fact. The
  data cannot show you whether a super PAC and a campaign were actually working
  together; that is what investigations are for.
- **Everything spent.** Independent expenditures are one category. Electioneering
  communications, party spending and dark-money groups that never file at all
  are elsewhere or nowhere.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- Sixteen rows out of 73,449 changed the headline finding from "mostly
  supportive" to "mostly negative." What does that do to your confidence in
  numbers quoted from official data?
- The FEC published "THE COURT OF DIVINE JUSTICE" spending $9.9 billion. Should
  it have? What would checking cost, and who would pay?
- 87.6% of the outside money around Bob Casey was spent against him. Is that a
  problem, and for whom?

***

### Where the data came from

Federal Election Commission bulk downloads, **independent expenditures, 2024
cycle** — 73,449 filings.
<https://www.fec.gov/files/bulk-downloads/2024/independent_expenditure_2024.csv>

The build is in `data/build-ie-data.R`. It documents all three data problems
found in this file: the junk filings, the nine-fold duplicate, and the
unnormalised candidate names that split "Harris, Kamala" from "HARRIS, KAMALA"
until they were merged. **Every row removed by the cleaning rule is committed in
`ie_outliers.csv`**, so you can check the decision rather than trust it.
