---
title: "Key: Two Systems, One Map"
subtitle: "84-355 Democracy's Data | Census geography | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, dpi = 96, fig.retina = 1,
                      fig.width = 7.2)
options(scipen = 999)
lv <- read.csv("data/levels.csv",  stringsAsFactors = FALSE)
n  <- read.csv("data/nesting.csv", stringsAsFactors = FALSE)

# The county maps. GEOIDs are CHARACTER -- read as numbers a 15-digit block id
# becomes 1.3153e+14 and leading zeros vanish.
D   <- "data"
fc  <- read.csv(file.path(D, "facts.csv"),  colClasses = c(value = "character"))
sp  <- read.csv(file.path(D, "splits.csv"), stringsAsFactors = FALSE)
mu  <- read.csv(file.path(D, "map_units.csv"), stringsAsFactors = FALSE)
tsp <- read.csv(file.path(D, "tract_split.csv"), colClasses = c(geoid = "character"))
FV <- function(k) fc$value[fc$name == k]
SP <- function(o, cc) sp[[cc]][sp$overlay == o]
nn <- function(x) format(as.numeric(x), big.mark = ",")

onepath <- function(X, Y) {
  dx <- X[-1] - X[-length(X)]; dy <- Y[-1] - Y[-length(Y)]
  seg <- paste0(dx, ifelse(dy < 0, "", ","), dy)
  sep <- c("", ifelse(substr(seg[-1], 1, 1) == "-", "", " "))
  paste0("M", X[1], ",", Y[1], "l", paste0(sep, seg, collapse = ""), "Z")
}
ringpaths <- function(d, sx, sy) {
  if (!nrow(d)) return(character(0))
  k <- interaction(d$uid, d$part, drop = TRUE)
  p <- vapply(split(d, k), function(z) {
    X <- round(sx(z$x)); Y <- round(sy(z$y))
    keep <- c(TRUE, X[-1] != X[-length(X)] | Y[-1] != Y[-length(Y)])
    X <- X[keep]; Y <- Y[keep]
    if (length(X) < 3) return(paste0("M", X[1], ",", Y[1], "h1v1h-1Z"))
    onepath(X, Y)
  }, character(1))
  p[nzchar(p)]
}
drawrings <- function(d, col = NA, border = "#999", lwd = 0.4) {
  if (!nrow(d)) return(invisible())
  k <- interaction(d$uid, d$part, drop = TRUE)
  for (l in levels(k)) {
    z <- d[k == l, ]
    polygon(z$x, z$y, col = col, border = border, lwd = lwd)
  }
}
jstr <- function(x) paste0("[", paste0('"', x, '"', collapse = ","), "]")
G <- function(lev, uid = NULL) {
  d <- mu[mu$lev == lev, ]
  if (!is.null(uid)) d <- d[d$uid %in% uid, ]
  d
}
win <- function(d, pad = 0.06) {
  rx <- range(d$x); ry <- range(d$y)
  s <- max(diff(rx), diff(ry)) * (1 + pad) / 2
  list(x = mean(rx) + c(-s, s), y = mean(ry) + c(-s, s), s = 2 * s)
}
```

All code is given to students. Grade the interpretation.

**Structure, as of the rebuild.** The brief (`census-geography-brief.Rmd`) and the lab (`census-geography.Rmd`)
run the *same numbered steps* — the brief in prose with D3 figures, the lab with
the code exposed. **Students should read the brief first.** Step numbers below
refer to both.


**Teach this EARLY.** It is the foundation five later labs assume and none of
them supplies — precincts against blocks, HOLC areas against tracts, EAVS
jurisdictions, §203 thresholds, BISG. Run before those and each of them gets a
sentence shorter and a great deal clearer. Run after, and it works as the
session where everything that went wrong is explained at once.

**It is also the shortest lab in the course.** Six small tables, three maps of
one county, no modeling. Use the time it saves on the discussion.

**The maps are all `r FV("county_name")`** — the same county the
precinct-geography lab uses, so students meet the ground twice. Say that out
loud; the second encounter is worth much more if they know it is the second.

Every number below was run against the data files.

***

# The session in one sentence

American geography is two systems — one where the identifier contains the
hierarchy and every join is exact, and one where it does not and every join is
an assumption — and almost every data problem in this course lives on the seam.

***

# The findings

```{r headline}
c(counties = lv$count[lv$geography == "county"],
  tracts   = lv$count[lv$geography == "census tract"],
  zctas    = lv$count[lv$geography == "ZCTA (ZIP approximation)"],
  tracts_nesting_cleanly = n$n_pass[1],
  zcta_file_has_a_state  = n$result[3])
```

- **85,396 of 85,396 tracts** match a real county by ID prefix. Zero orphans.
- **The ZCTA file has no state column and no state in the ID.**
- A median county is **604 sq mi**; a median tract **1.74**; a median ZCTA
  **34.89** — twenty tracts' worth.

And the same two claims drawn, on `r FV("county_name")`
(`r nn(FV("n_blocks"))` blocks, `r nn(FV("n_bg"))` block groups,
`r nn(FV("n_tracts"))` tracts):

```{r headline-map, echo=FALSE, results='asis', eval=knitr::is_html_output()}
PW <- 214L; GAP <- 22L; TOPN <- 40L; BOTN <- 30L
Wn <- 3L * PW + 2L * GAP; Hn <- TOPN + PW + BOTN
w <- win(G("c"), pad = 0.04); s <- PW / w$s
OV <- c("g", "p", "z"); OCL <- c("#2c7fb8", "#C41230", "#e08214")
FLG <- c("split_bg", "split_place", "split_zcta")
ROW <- c("block group", "census place (city limits)", "ZCTA (ZIP approximation)")
mk <- function(i) {
  off <- (i - 1L) * (PW + GAP)
  fx <- function(x) off + (x - w$x[1]) * s
  fy <- function(y) TOPN + (w$y[2] - y) * s
  paste0("[", jstr(ringpaths(G("t"), fx, fy)), ",",
         jstr(ringpaths(G("t", tsp$uid[tsp[[FLG[i]]] == 1]), fx, fy)), ",",
         jstr(ringpaths(G(OV[i]), fx, fy)), ",", off, "]")
}
PAN <- vapply(1:3, mk, character(1))
cnt <- sprintf("%s of %s tracts cut",
               vapply(ROW, function(r) nn(SP(r, "tracts_split")), character(1)),
               nn(FV("n_tracts")))
cat(paste0('
<div id="nst" style="margin:1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const P=[', paste(PAN, collapse = ","), '],T=', jstr(c("block groups",
  "city limits", "ZCTAs")), ',C=', jstr(cnt), ',OC=', jstr(OCL), ';
const W=', Wn, ',H=', Hn, ',PW=', PW, ',TOP=', TOPN, ';
const svg=d3.select("#nst").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const defs=svg.append("defs");
P.forEach((p,i)=>{
  defs.append("clipPath").attr("id","kc"+i).append("rect")
    .attr("x",p[3]).attr("y",TOP).attr("width",PW).attr("height",PW);
  const g=svg.append("g").attr("clip-path","url(#kc"+i+")");
  g.selectAll("path.t").data(p[0]).join("path").attr("d",d=>d)
    .attr("fill","#f7f7f7").attr("stroke","#b6b6b6").attr("stroke-width",0.6);
  g.selectAll("path.c").data(p[1]).join("path").attr("d",d=>d)
    .attr("fill","#f4d9dd").attr("stroke","#b6b6b6").attr("stroke-width",0.6);
  g.selectAll("path.o").data(p[2]).join("path").attr("d",d=>d)
    .attr("fill","none").attr("stroke",OC[i]).attr("stroke-width",1.4);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",18).attr("text-anchor","middle")
    .attr("font-size","13px").attr("font-weight","600").attr("fill",OC[i])
    .text(T[i]);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",TOP+PW+18)
    .attr("text-anchor","middle").attr("font-size","13px")
    .attr("font-weight","600").attr("fill",i?"#C41230":"#2c7fb8").text(C[i]);
});
})();
</script>
'))
```

```{r headline-map-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.6}
OV <- c("g", "p", "z"); OCL <- c("#2c7fb8", "#C41230", "#e08214")
FLG <- c("split_bg", "split_place", "split_zcta")
ROW <- c("block group", "census place (city limits)", "ZCTA (ZIP approximation)")
ttl <- c("block groups", "city limits", "ZCTAs")
w <- win(G("c"), pad = 0.04)
layout(matrix(1:3, 1, 3))
op <- par(mar = c(2.4, 0.3, 2.4, 0.3), xpd = FALSE)
for (i in 1:3) {
  plot(NA, xlim = w$x, ylim = w$y, asp = 1, axes = FALSE, ann = FALSE)
  drawrings(G("t"), col = "#f7f7f7", border = "#b6b6b6", lwd = 0.5)
  drawrings(G("t", tsp$uid[tsp[[FLG[i]]] == 1]), col = "#f4d9dd",
            border = "#b6b6b6", lwd = 0.5)
  drawrings(G(OV[i]), col = NA, border = OCL[i], lwd = 1)
  title(ttl[i], cex.main = 1, col.main = OCL[i], line = 0.6)
  mtext(paste0(nn(SP(ROW[i], "tracts_split")), " of ", nn(FV("n_tracts")),
               " tracts cut"), side = 1, line = 0.9, cex = 0.72, font = 2,
        col = if (i == 1) "#2c7fb8" else "#C41230")
}
par(op); layout(1)
```

- **A block group boundary cuts `r nn(SP("block group", "tracts_split"))` of the
  county's `r nn(FV("n_tracts"))` tracts. A city limit cuts
  `r nn(SP("census place (city limits)", "tracts_split"))` of them, and a ZCTA
  boundary cuts `r nn(SP("ZCTA (ZIP approximation)", "tracts_split"))`.**
- Warner Robins is in **`r FV("wr_counties")` counties**, and
  `r nn(FV("zctas_multi_county"))` of the county's `r nn(FV("n_zctas"))` ZCTAs
  are in more than one county. That is the identifier's missing county
  component, on the ground.

***

# The three things to get said

**1. The GEOID is the join.** Students think of an ID as an arbitrary key. Write
the block identifier from Step 1 on the board — `131530211233013` — and take it
apart: state, county, tract, block group, block. **Once they see that the first
five digits *are* the county, the nesting spine stops being a fact to memorise
and becomes something they can read off any file.** The five-panel figure in
Step 1 is the same five pieces as places; put them side by side.

**2. Nesting is by construction, not by luck.** The Census Bureau draws small
units inside large ones on purpose, so tracts nest perfectly and always will.
Places, ZIP codes and school districts were drawn by other people, for other
reasons, at other times. **Nothing forces them to line up, and they don't.** The
three panels in Step 7 are the whole argument in one row of pictures: the
geography that nests cuts zero tracts, the two that do not cut
`r nn(SP("census place (city limits)", "tracts_split"))` and
`r nn(SP("ZCTA (ZIP approximation)", "tracts_split"))`.

**3. The ZIP code is the one they will actually use.** Every student will one
day be handed data "by ZIP." Make sure they leave knowing it is a **mail route,
not an area** — a set of delivery stops the Postal Service changes when routing
changes. The Census's ZCTA is an approximation of something that is not a shape.
**Point at the file: no state column, because some ZIPs are in two states.**

**The line for the board:** if the ID contains the hierarchy, the join is free.
Otherwise somebody has to decide.

***

# The move that makes this session land

**Ask the room where their ZIP code is.** They took the census replication survey
on the first day and typed one in. Then show them Step 8.

It reframes the whole thing from an abstraction about identifiers into something
they personally got wrong two months ago, and it costs ninety seconds.

***

***

# Walking the steps

| Steps | What they do | Min | Watch for |
|---|---|---|---|
| **1-3** | One GEOID; the five maps it names; the gazetteer; how many of each | 9 | The identifier is built by concatenation. That claim is testable, and the zoom sequence is the same claim as geography. |
| **4** | Test it: all 85,396 tracts match a county by prefix; the county at four scales | 9 | Zero orphans. Nesting is not documentation, it is arithmetic — and no line is drawn twice. |
| **5-7** | The two tests that could not be run; what each ID encodes; the three overlays counted | 14 | **The pivot.** An empty result column because the file has no such field, then 0 / 32 / 22 tracts cut. |
| **8** | ZIP codes | 8 | Five digits encoding nothing — and the first-day survey asked for one. |
| **9-10** | Why this bit four other labs; what you can say | 10 | Precincts, HOLC, EAVS jurisdictions, Cook County. One cause. |

**Step 5 is the pivot.** A test that *cannot be run* is a stronger result than one that fails: the Bureau declined to record the relationship, because the relationship does not exist.

**If you are short on time**, protect the pivot and the closing step. The numbers are what make the argument concrete; the sequence is the argument.


# Option A — Build the hierarchy

## Verified output

```{r a}
lv[, c("geography", "nests_into", "id_length")]
```

The tree: **state → county → { census tract → block group → block, county
subdivision }**. Outside it: places, ZCTAs, both kinds of school district.

**3/4:** Draws the tree correctly and puts the crosscutting units to one side.

**4/4:** Says what the outsiders have in common — **they were all drawn by
somebody other than the Census Bureau, for a purpose other than counting
people.** Municipalities incorporate, school districts are created by states, ZIP
codes serve mail. The Bureau records them; it does not design them.

**4/4, best:** Notices that "nests into state only" is doing sneaky work. A place
ID contains a state, so places *do* nest into states — but that is nearly
useless, because the join anybody wants is to a county or a tract. **A hierarchy
that only reaches the top level is not a hierarchy you can use.**

***

# Option B — How wrong is a ZIP?

## Verified output

```{r b}
z <- lv$median_sq_mi[lv$geography == "ZCTA (ZIP approximation)"]
t <- lv$median_sq_mi[lv$geography == "census tract"]
c(median_zcta = z, median_tract = t, ratio = round(z / t, 1))
```

**About twenty tracts to a ZCTA by area.**

**3/4:** Reports the ratio and says ZIP-level statistics average over a lot.

**4/4:** Makes it concrete — a ZIP-level median income is one number for an area
that might contain both the richest and poorest tracts in a city, and the
averaging is not random, it is *spatial*. Neighbourhoods differ block to block;
that is the entire finding of the redlining lab.

**4/4, best:** Notices the comparison is by **area**, not population, and that
these are very different in a city — a dense urban ZCTA might contain many more
than twenty tracts while a rural one contains a fraction of one. **The
twenty-times figure is itself an average that hides its own variance**, which is
the lesson eating its own tail, appropriately.

***

# Option C — The school district trap

No verified output beyond the counts; grade the reasoning.

**1,952 elementary districts, 10,893 unified.**

**3/4:** Works out that a student in an elementary district must attend a
secondary district for high school, so the two overlap.

**4/4:** Draws the consequence for data — **the same child is inside two
districts at once**, so "students per district" double-counts unless you know
which system you are in, and a map of "school districts" is either incomplete or
overlapping depending on which files you loaded.

**4/4, best:** Connects it back to the main lesson: this is not a quirk of
schools, it is what happens whenever two authorities draw boundaries over the
same ground for different purposes. **Congressional districts, precincts, water
boards, fire districts — the voter file in Session 6 had twenty-one such columns
for exactly this reason.**

***

# Option D — Read the middle panel

## Verified output

```{r d}
sp[, c("overlay", "role", "units", "tracts_total", "tracts_split")]
```

The question asks what you must assume to report a tract-level statistic for a
city whose limits cut `r nn(SP("census place (city limits)", "tracts_split"))`
of `r nn(FV("n_tracts"))` tracts.

**3/4:** Says you would have to split the cut tracts somehow, and names area or
population weighting.

**4/4:** States the assumption in its strong form — **that the people inside the
city limits look like the people outside them, within the same tract** — and
notices that this is exactly the assumption the tract was drawn to avoid making,
since tracts are meant to be internally similar and a city limit is not.

**4/4, best:** Names a case where it fails hard and says why. Annexation is
selective: cities annex subdivisions and shopping centres and leave trailer
parks and farmland outside, so the line between incorporated and unincorporated
land inside one tract is frequently a line between different kinds of household.
**The error is not random, it is correlated with exactly the thing you are
measuring.** Anyone who gets there has understood the lab.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is explaining *why* the nesting units nest — by
  construction — rather than treating it as a happy accident.
- **A 2** is "you should use tracts instead of ZIP codes" with no account of
  why, or of what to do when the data only comes by ZIP.
- **Reward anyone who reconsiders an earlier lab** in light of this one. That is
  the whole point of running it.

## Anticipated questions

- *"Can I convert ZIP to tract?"* — Crosswalks exist (HUD publishes one) and
  they are **allocations, not translations** — each ZIP maps to several tracts
  with weights. Fine if you say so; wrong if you pretend it is a lookup.
- *"Why does anyone publish by ZIP then?"* — Because it is the only geography
  respondents know. That is a real answer, not a dodge, and it is why the first-
  day survey asked for one.
- *"Do blocks nest into precincts?"* — **No, and that is Session 14's whole
  problem.** Good moment to forward-reference it.
- *"What about congressional districts?"* — They nest into states and nothing
  else, and they are redrawn every decade, so they do not even nest into their
  own past. The gazetteer's CD file was unavailable when this was built.
- *"Is a 'place' the same as a city?"* — Roughly, plus census-designated places,
  which are unincorporated areas the Bureau treats as places for counting. A
  place can exist for statistics and not for government. The county mapped here
  has three incorporated cities and one CDP, **Robins AFB**, which is the base:
  a place for counting, not a municipality.
- *"Why this county?"* — Because the precinct-geography lab uses it too, so the
  ground is already familiar by the second time they see it, and because it is
  small enough that all `r nn(FV("n_blocks"))` of its blocks fit on a page.
  Nothing about the argument depends on the choice; the numbers do.
- *"Is 32 of 38 typical?"* — **Unknown from this lab**, and worth saying so. It
  is one county with a heavily annexed city in it. The direction of the result
  is structural; the size of it is local.
- *"Why do blocks nest into everything?"* — Because the Bureau draws them last,
  against every boundary it knows about, which is why a block is the atom the
  crosswalks in later labs are built on. **Blocks nest into places and ZCTAs;
  tracts do not.** That is the sentence to leave them with.

## Connections

**This lab is a prerequisite for five others**, and each should point back:

- **Session 6 (voter files)** — 21 of 53 columns are overlapping districts.
- **Session 11 (EAVS)** — "jurisdiction" varies by state.
- **Session 14 (precincts)** — precincts against blocks, the same seam.
- **Redlining** — 1937 areas against 2020 tracts, joined by centroid.
- **Session 7 (BISG)** — works *because* blocks nest. The easy case, and worth
  saying so, since students otherwise never notice when a join went right.
