---
title: "Two Systems, One Map"
subtitle: "84-355 Democracy's Data | What census geography nests, and what doesn't"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2,
                      fig.height = 4.4, dpi = 96, fig.retina = 1,
                      message = FALSE, warning = FALSE)
options(scipen = 999)

# ---- the maps -------------------------------------------------------------
# Read here so the visible chunks below stay about the argument. GEOIDs are
# CHARACTER: read as numbers, a 15-digit block identifier turns into
# 1.3153e+14 and a leading zero disappears for good.
D   <- "data"
fc  <- read.csv(file.path(D, "facts.csv"),  colClasses = c(value = "character"))
sp  <- read.csv(file.path(D, "splits.csv"), stringsAsFactors = FALSE)
ch  <- read.csv(file.path(D, "chain.csv"),
                colClasses = c(geoid = "character", digits = "character",
                               lev = "character"))
mu  <- read.csv(file.path(D, "map_units.csv"), stringsAsFactors = FALSE)
mi  <- read.csv(file.path(D, "map_ids.csv"),  colClasses = c(geoid = "character"))
tsp <- read.csv(file.path(D, "tract_split.csv"), colClasses = c(geoid = "character"))

FV <- function(k) fc$value[fc$name == k]
SP <- function(o, cc) sp[[cc]][sp$overlay == o]
CHV <- function(l, cc) ch[[cc]][ch$level == l]
nn <- function(x) format(as.numeric(x), big.mark = ",")
LEVN <- c("state", "county", "tract", "block group", "block")
LEVK <- c("s", "c", "t", "g", "b")

onepath <- function(X, Y) {
  dx <- X[-1] - X[-length(X)]; dy <- Y[-1] - Y[-length(Y)]
  seg <- paste0(dx, ifelse(dy < 0, "", ","), dy)
  sep <- c("", ifelse(substr(seg[-1], 1, 1) == "-", "", " "))
  paste0("M", X[1], ",", Y[1], "l", paste0(sep, seg, collapse = ""), "Z")
}
ringpaths <- function(d, sx, sy) {
  if (!nrow(d)) return(character(0))
  k <- interaction(d$uid, d$part, drop = TRUE)
  p <- vapply(split(d, k), function(z) {
    X <- round(sx(z$x)); Y <- round(sy(z$y))
    keep <- c(TRUE, X[-1] != X[-length(X)] | Y[-1] != Y[-length(Y)])
    X <- X[keep]; Y <- Y[keep]
    if (length(X) < 3) return(paste0("M", X[1], ",", Y[1], "h1v1h-1Z"))
    onepath(X, Y)
  }, character(1))
  p[nzchar(p)]
}
drawrings <- function(d, col = NA, border = "#999", lwd = 0.4) {
  if (!nrow(d)) return(invisible())
  k <- interaction(d$uid, d$part, drop = TRUE)
  for (l in levels(k)) {
    z <- d[k == l, ]
    polygon(z$x, z$y, col = col, border = border, lwd = lwd)
  }
}
jstr <- function(x) paste0("[", paste0('"', x, '"', collapse = ","), "]")
G <- function(lev, uid = NULL) {
  d <- mu[mu$lev == lev, ]
  if (!is.null(uid)) d <- d[d$uid %in% uid, ]
  d
}
win <- function(d, pad = 0.06) {
  rx <- range(d$x); ry <- range(d$y)
  s <- max(diff(rx), diff(ry)) * (1 + pad) / 2
  list(x = mean(rx) + c(-s, s), y = mean(ry) + c(-s, s), s = 2 * s)
}
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

> **Read `census-geography-brief.Rmd` first.** This lab runs the same numbered steps with the
> code exposed. The brief tells you what the question is and why it is hard;
> this shows you where each number came from.

***

# Where did that number come from?

> *"The data is by ZIP code."*

Half the joins you will ever attempt fail, and they nearly all fail for the same
reason. Somebody had data about one kind of area and needed it for a different
kind of area, and assumed the two fit together.

**They usually don't.** American geography is two systems laid on top of each
other. One nests perfectly and the other cuts across it — and knowing which is
which is the difference between an analysis and a mess.

This is the lab that everything else in the course has been quietly assuming.

# Steps 1-3 — One identifier, the source, and how many of each

```{r load}
lv <- read.csv("data/levels.csv", stringsAsFactors = FALSE)
lv[, c("geography", "count", "median_sq_mi")]
```

Seven kinds of area, from the Census Bureau's own gazetteer. **3,222 counties.
85,396 tracts. 33,791 ZCTAs. 32,333 places.**

The sizes tell you these are not variations on one idea. A median county is
**604 square miles**; a median tract is **1.74**. They are different tools.

# Step 4 — Test the claim the identifier makes

> **The identifier makes a testable claim.** If an 11-digit tract GEOID really
> is state(2) + county(3) + tract(6), then slicing off the first five digits
> should give a county that exists — for every tract in the country, with no
> exceptions. **Predict how many of the 85,396 will fail.**

Here is the thing nobody tells you, and it is the most useful fact in the
course.

```{r ids}
lv[, c("geography", "id_length", "id_encodes", "nests_into")]
```

**A census GEOID is not an arbitrary key. It is an address in a hierarchy.**

The 15-digit block ID mapped below reads:

```{r idtree, echo=FALSE, results='asis'}
# Drawn from chain.csv rather than typed, so the diagram cannot drift away
# from the block the maps actually show.
w   <- nchar(ch$digits) + 3           # three spaces between the groups
col <- cumsum(c(0, head(w, -1)))
lab <- c("state", "county within that state", "tract within that county",
         "block group within that tract", "block within that block group")
lab[1] <- paste0("state (", ch$digits[1], " = ", ch$name[1], ")")
lab[2] <- paste0("county (", ch$digits[2], " = ",
                 sub(" County.*", "", ch$name[2]), ")")
out <- c(paste(ch$digits, collapse = "   "), "")
mark <- rep(" ", max(col) + 2)
for (i in seq_along(col)) mark[col[i] + 1] <- "|"
out[2] <- paste(mark, collapse = "")
for (i in rev(seq_along(col))) {
  m <- mark; m[col[i] + 1] <- "+"
  for (j in (col[i] + 2):(max(col) + 2)) m[j] <- "-"
  out <- c(out, paste0(paste(m, collapse = ""), " ", lab[i]))
  mark[col[i] + 1] <- " "
}
cat("```\n", paste(out, collapse = "\n"), "\n```\n", sep = "")
```

So the first five digits of any tract ID **are** a county ID. You do not need a
lookup table; the join is already in the number.

```{r nest}
n <- read.csv("data/nesting.csv", stringsAsFactors = FALSE)
n
```

**All 85,396 tracts match a real county by prefix. Zero orphans.** Same for all
36,421 county subdivisions. The nesting spine — block → block group → tract →
county → state — is exact, and it is exact *by construction*, because the Census
Bureau draws the small ones inside the big ones.

That is a claim about strings. Here is the same claim about ground:
`r FV("county_name")` drawn four times, with one block and each of the units
containing it picked out in red.

```{r scales-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
PW <- 168L; GAP <- 10L; TOPS <- 40L; BOT <- 24L
Ws <- 4L * PW + 3L * GAP; Hs <- TOPS + PW + BOT
KEY <- c("b", "g", "t", "c")
w <- win(G("c"), pad = 0.04); s <- PW / w$s
mk <- function(i) {
  off <- (i - 1L) * (PW + GAP)
  fx <- function(x) off + (x - w$x[1]) * s
  fy <- function(y) TOPS + (w$y[2] - y) * s
  hit <- CHV(if (KEY[i] == "c") "county" else LEVN[match(KEY[i], LEVK)], "uid")
  paste0("[", jstr(ringpaths(G("c"), fx, fy)), ",",
         jstr(ringpaths(G(KEY[i]), fx, fy)), ",",
         jstr(ringpaths(G(KEY[i], hit), fx, fy)), ",", off, "]")
}
PAN <- vapply(seq_along(KEY), mk, character(1))
ttl <- c(paste(nn(FV("n_blocks")), "blocks"),
         paste(nn(FV("n_bg")), "block groups"),
         paste(nn(FV("n_tracts")), "tracts"), "1 county")
sub <- paste(c(15, 12, 11, 5), "digits in the ID")
cat(paste0('
<div id="scl" style="margin:1.1em 0"></div>
<script src="https://d3js.org/d3.v7.min.js"></script>
<script>
(function(){
const P=[', paste(PAN, collapse = ","), '],T=', jstr(ttl), ',S=', jstr(sub), ';
const LW=[0.25,0.6,0.8,1.4];
const W=', Ws, ',H=', Hs, ',PW=', PW, ',TOP=', TOPS, ';
const svg=d3.select("#scl").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
P.forEach((p,i)=>{
  const g=svg.append("g");
  g.selectAll("path.b").data(p[0]).join("path").attr("d",d=>d)
    .attr("fill","#eef1f4").attr("stroke","none");
  g.selectAll("path.u").data(p[1]).join("path").attr("d",d=>d)
    .attr("fill","none").attr("stroke","#4d6a86").attr("stroke-width",LW[i]);
  g.selectAll("path.h").data(p[2]).join("path").attr("d",d=>d)
    .attr("fill","#C41230").attr("fill-opacity",0.9)
    .attr("stroke","#C41230").attr("stroke-width",1.2);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",16).attr("text-anchor","middle")
    .attr("font-size","12.5px").attr("font-weight","600").attr("fill","#222")
    .text(T[i]);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",28).attr("text-anchor","middle")
    .attr("font-size","9.5px").attr("fill","#888").text(S[i]);
});
})();
</script>
'))
```

```{r scales-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=2.9}
KEY <- c("b", "g", "t", "c")
ttl <- c(paste(nn(FV("n_blocks")), "blocks"),
         paste(nn(FV("n_bg")), "block groups"),
         paste(nn(FV("n_tracts")), "tracts"), "1 county")
sub <- paste(c(15, 12, 11, 5), "digits in the ID")
lwd <- c(0.12, 0.3, 0.45, 0.8)
w <- win(G("c"), pad = 0.04)
layout(matrix(1:4, 1, 4))
op <- par(mar = c(0.4, 0.2, 2.6, 0.2), xpd = FALSE)
for (i in seq_along(KEY)) {
  plot(NA, xlim = w$x, ylim = w$y, asp = 1, axes = FALSE, ann = FALSE)
  drawrings(G("c"), col = "#eef1f4", border = NA)
  drawrings(G(KEY[i]), col = NA, border = "#4d6a86", lwd = lwd[i])
  hit <- CHV(if (KEY[i] == "c") "county" else LEVN[match(KEY[i], LEVK)], "uid")
  drawrings(G(KEY[i], hit), col = "#C41230", border = "#C41230", lwd = 1)
  title(ttl[i], cex.main = 0.9, line = 1.1)
  mtext(sub[i], side = 3, line = 0.25, cex = 0.55, col = "#888")
}
par(op); layout(1)
```

`r nn(FV("n_blocks"))` blocks become `r nn(FV("n_bg"))` block groups become
`r nn(FV("n_tracts"))` tracts become one county, and **no boundary is drawn
twice.** Every line in a later panel was already a line in the first one. That
is what "nests" means when you can see it.

# Steps 5-7 — The tests that could not be run, and what each ID encodes

```{r other}
lv[lv$nests_into != "county" & lv$nests_into != "state",
   c("geography", "count", "id_length", "id_encodes", "nests_into")]
```

**A place ID is seven digits: state, then place.** No county. That is not an
oversight — **places cross county lines**, so there is no county to record. A
city is a legal entity that grew where it grew; counties were drawn separately.

Warner Robins, the city in the county we are mapping, is the example: of its
`r FV("wr_sq_km_total")` square kilometres, `r FV("wr_sq_km_in_county")` are in
Houston County and the rest are in the next county over. **There is no county to
put in the identifier because the city is in two of them.**

Here is what that does to the tracts underneath. Same `r nn(FV("n_tracts"))`
tracts in all three panels; only the boundary laid over them changes. A tract is
**cut** when part of it is inside a unit of the other geography and part is not.

```{r nest-d3, results='asis', echo=FALSE, eval=knitr::is_html_output()}
PW <- 214L; GAP <- 22L; TOPN <- 40L; BOTN <- 30L
Wn <- 3L * PW + 2L * GAP; Hn <- TOPN + PW + BOTN
w <- win(G("c"), pad = 0.04); s <- PW / w$s
OV <- c("g", "p", "z"); OCL <- c("#2c7fb8", "#C41230", "#e08214")
FLG <- c("split_bg", "split_place", "split_zcta")
ROW <- c("block group", "census place (city limits)", "ZCTA (ZIP approximation)")
mk <- function(i) {
  off <- (i - 1L) * (PW + GAP)
  fx <- function(x) off + (x - w$x[1]) * s
  fy <- function(y) TOPN + (w$y[2] - y) * s
  paste0("[", jstr(ringpaths(G("t"), fx, fy)), ",",
         jstr(ringpaths(G("t", tsp$uid[tsp[[FLG[i]]] == 1]), fx, fy)), ",",
         jstr(ringpaths(G(OV[i]), fx, fy)), ",", off, "]")
}
PAN <- vapply(1:3, mk, character(1))
ttl <- c("block groups", "city limits", "ZCTAs")
cnt <- sprintf("%s of %s tracts cut",
               vapply(ROW, function(r) nn(SP(r, "tracts_split")), character(1)),
               nn(FV("n_tracts")))
rol <- vapply(ROW, function(r) SP(r, "role"), character(1))
cat(paste0('
<div id="nst" style="margin:1.1em 0"></div>
<script>
(function(){
const P=[', paste(PAN, collapse = ","), '],T=', jstr(ttl), ',C=', jstr(cnt), ';
const R=', jstr(rol), ',OC=', jstr(OCL), ';
const W=', Wn, ',H=', Hn, ',PW=', PW, ',TOP=', TOPN, ';
const svg=d3.select("#nst").append("svg").attr("viewBox","0 0 "+W+" "+H)
  .attr("style","max-width:100%;height:auto;font:12px inherit");
const defs=svg.append("defs");
P.forEach((p,i)=>{
  defs.append("clipPath").attr("id","lc"+i).append("rect")
    .attr("x",p[3]).attr("y",TOP).attr("width",PW).attr("height",PW);
  const g=svg.append("g").attr("clip-path","url(#lc"+i+")");
  g.selectAll("path.t").data(p[0]).join("path").attr("d",d=>d)
    .attr("fill","#f7f7f7").attr("stroke","#b6b6b6").attr("stroke-width",0.6);
  g.selectAll("path.c").data(p[1]).join("path").attr("d",d=>d)
    .attr("fill","#f4d9dd").attr("stroke","#b6b6b6").attr("stroke-width",0.6);
  g.selectAll("path.o").data(p[2]).join("path").attr("d",d=>d)
    .attr("fill","none").attr("stroke",OC[i]).attr("stroke-width",1.4);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",16).attr("text-anchor","middle")
    .attr("font-size","13px").attr("font-weight","600").attr("fill",OC[i])
    .text(T[i]);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",28).attr("text-anchor","middle")
    .attr("font-size","9.5px").attr("fill","#888").text(R[i]);
  svg.append("text").attr("x",p[3]+PW/2).attr("y",TOP+PW+18)
    .attr("text-anchor","middle").attr("font-size","13px")
    .attr("font-weight","600").attr("fill",i?"#C41230":"#2c7fb8").text(C[i]);
});
})();
</script>
'))
```

```{r nest-static, echo=FALSE, eval=!knitr::is_html_output(), fig.height=3.6}
OV <- c("g", "p", "z"); OCL <- c("#2c7fb8", "#C41230", "#e08214")
FLG <- c("split_bg", "split_place", "split_zcta")
ROW <- c("block group", "census place (city limits)", "ZCTA (ZIP approximation)")
ttl <- c("block groups", "city limits", "ZCTAs")
w <- win(G("c"), pad = 0.04)
layout(matrix(1:3, 1, 3))
op <- par(mar = c(2.4, 0.3, 3.0, 0.3), oma = c(1.2, 0, 0, 0), xpd = FALSE)
for (i in 1:3) {
  plot(NA, xlim = w$x, ylim = w$y, asp = 1, axes = FALSE, ann = FALSE)
  drawrings(G("t"), col = "#f7f7f7", border = "#b6b6b6", lwd = 0.5)
  drawrings(G("t", tsp$uid[tsp[[FLG[i]]] == 1]), col = "#f4d9dd",
            border = "#b6b6b6", lwd = 0.5)
  drawrings(G(OV[i]), col = NA, border = OCL[i], lwd = 1)
  title(ttl[i], cex.main = 1, col.main = OCL[i], line = 1.4)
  mtext(SP(ROW[i], "role"), side = 3, line = 0.5, cex = 0.55, col = "#888")
  mtext(paste0(nn(SP(ROW[i], "tracts_split")), " of ", nn(FV("n_tracts")),
               " tracts cut"), side = 1, line = 0.9, cex = 0.72, font = 2,
        col = if (i == 1) "#2c7fb8" else "#C41230")
}
mtext("shaded: a tract with a boundary running through it", side = 1,
      outer = TRUE, line = 0, cex = 0.6, col = "#777")
par(op); layout(1)
```

The counts are not eyeballed off the picture. They come from the Census
Bureau's own block-to-place assignment and its own ZCTA-to-tract relationship
file, and they are in `splits.csv`:

```{r splits}
s <- read.csv("data/splits.csv", stringsAsFactors = FALSE)
s[, c("overlay", "role", "units", "tracts_total", "tracts_split")]
```

**Zero, `r nn(SP("census place (city limits)", "tracts_split"))`,
`r nn(SP("ZCTA (ZIP approximation)", "tracts_split"))`.** Block groups are
carved out of tracts, so they cannot cross one. Cities and ZIP codes were drawn
by other people for other reasons, and they cross constantly.

Same for school districts. Same, most importantly, for the one everybody uses.

# Step 8 — The one everybody uses works worst

```{r zcta}
lv[lv$geography == "ZCTA (ZIP approximation)", ]
n[n$test == "ZCTA file records a state", ]
```

**The ZCTA file has no state column, and the identifier is five digits with
nothing else in it.** You cannot determine what state a ZCTA is in from its ID.

That is not sloppiness. It is because **some ZIP codes are in two states.**

And it gets worse, because a ZIP code is not an area at all. **It is a set of
mail delivery stops** — a route, maintained by the Postal Service, changed
whenever routing changes. Some ZIP codes are a single building. Some cover no
residents. They were never meant to describe where people live, and the Census
Bureau's ZCTA is a best-effort approximation of a thing that is not a shape.

```{r zcta-size}
c(zctas = lv$count[lv$geography == "ZCTA (ZIP approximation)"],
  tracts = lv$count[lv$geography == "census tract"],
  median_zcta_sq_mi = lv$median_sq_mi[lv$geography == "ZCTA (ZIP approximation)"],
  median_tract_sq_mi = lv$median_sq_mi[lv$geography == "census tract"])
```

A median ZCTA is **34.89 square miles** — twenty times a median tract, and
wildly uneven.

**You gave me your ZIP code on the first day of class.** It was the right thing
to ask for, because it is the only geography anybody knows off the top of their
head. It is also the one that will not join cleanly to anything else in this
course.

# Step 9 — Why this keeps happening in other labs

Every one of these was the same problem:

| Where | What did not nest |
|---|---|
| **Precinct returns** | precincts are drawn by counties and ignore census blocks |
| **Redlining** | 1937 HOLC areas against 2020 tracts |
| **Election administration** | "jurisdiction" is a township in Wisconsin, a county in Texas |
| **Section 203** | thresholds applied to jurisdictions, estimates published for others |
| **BISG** | works because blocks *do* nest — that one was the easy case |

```{r moral}
c(nesting_geographies = sum(lv$nests_into %in% c("state", "county")),
  crosscutting = sum(!lv$nests_into %in% c("state", "county")))
```

**When two geographies nest, the join is free and exact. When they don't, every
join is an assumption**, and somebody has to decide how to split the overlap.
There is no correct answer, only conventions — and the convention is usually
buried in a methods appendix nobody reads.

# Step 10 — What you can now say

**You can say** that American geography has a **spine** and a set of
**crosscutting** systems; that the spine's identifiers encode the nesting, so
aggregation needs no assumptions; and that everything off the spine requires a
crosswalk resting on an assumption somebody chose.

**You can say which of these seven geographies are which** — and you can say it
because the nesting was *tested*, not read from documentation.

**You cannot move data between crosscutting geographies** without deciding how
people are distributed within units, and no data will check that decision for
you.

***

# Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Build the hierarchy.** Using only `levels.csv`, draw the nesting
tree: which units contain which. Then mark the ones that cannot be placed in the
tree at all, and say what they have in common.

**Option B — How wrong is a ZIP?** A median ZCTA is 34.89 square miles and a
median tract 1.74 — roughly twenty tracts to a ZCTA. If someone reports a
statistic "by ZIP code," how much variation are they averaging away? Make the
argument with numbers.

**Option C — The school district trap.** There are 1,952 elementary districts
and 10,893 unified districts. A unified district covers all grades; an
elementary one does not. What happens to a student living in an elementary
district when they reach high school, and what does that do to anyone trying to
count "students per district"?

**Option D — Read the middle panel.** Look at the three maps again. The city
limits cut `r nn(SP("census place (city limits)", "tracts_split"))` of
`r nn(FV("n_tracts"))` tracts and the block groups cut none. Suppose you have
median income by tract and you are asked for median income *in Warner Robins*.
Say exactly what you would have to assume, and name one kind of neighbourhood
where the assumption would fail badly.

**Option E — Your own question.** Seven geographies, 200,000-odd units, and one
county drawn at four scales.

# What this lab cannot tell you

- **Whether one county is typical.** The maps are `r FV("county_name")`, and
  `r nn(SP("census place (city limits)", "tracts_split"))` of
  `r nn(FV("n_tracts"))` is a fact about it, not a national rate. A denser
  county or an emptier one could be worse or better, and nothing here says
  which.
- **How to fix a broken join.** The standard answer is area-weighted or
  population-weighted allocation, and it is an approximation with its own
  errors. The lab shows you the problem; the fix is a decision, not a lookup.
- **Which ZIP codes cross state lines.** We can see that
  `r nn(FV("zctas_multi_county"))` of this county's `r nn(FV("n_zctas"))` ZCTAs
  cross a *county* line; the state-line cases need the same check run
  nationally.
- **Almost anything about people.** The county maps carry block population, but
  the seven national geographies are counted and measured only as containers.

# On the discussion board

By **Friday, 11:59 p.m.** A sentence or two is enough.

Three that would make good posts:

- The tract ID contains the county ID. What else in this course would have been
  easier if every identifier worked that way?
- A ZIP code is a mail route, not a place. Why is so much data published by ZIP
  anyway?
- Which of the joins we have done this term do you now trust less?

***

### Where the data came from

**U.S. Census Bureau Gazetteer Files, 2024** — counties, tracts, places, ZCTAs,
county subdivisions, and elementary and unified school districts.
<https://www2.census.gov/geo/docs/maps-data/data/gazetteer/2024_Gazetteer/>

**U.S. Census Bureau TIGER/Line Shapefiles, 2020** — every boundary on the maps
above: the `r nn(FV("n_blocks"))` blocks, `r nn(FV("n_bg"))` block groups,
`r nn(FV("n_tracts"))` tracts, `r nn(FV("n_places"))` places and
`r nn(FV("n_zctas"))` ZCTAs of `r FV("county_name")`.
<https://www2.census.gov/geo/tiger/TIGER2020/>

**U.S. Census Bureau 2020 Block Assignment Files and ZCTA relationship files** —
which block is in which municipality, and how ZCTA land divides among tracts.
The cut counts come from these, not from our own overlay.

Everything is public and needs no key. The builds are `data/build-data.py` (the
national counts) and `data/build-maps.R` (the geometry and the cut counts); the
nesting tests in `nesting.csv` were run against the full files, not asserted.
