---
title: "Lab 13: A Threshold in the Law"
subtitle: "84-355 Democracy's Data | Week 13"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
```

# Before we start

Every lab so far has asked what data can tell us. Today the data *is* the
decision.

Section 203 of the Voting Rights Act says that if enough voters in a
jurisdiction speak a language other than English and have limited English
proficiency, that jurisdiction **must** print its ballots in that language. Not
should — must. It is federal law, and the Director of the Census decides who is
covered by running an arithmetic test on survey estimates.

The statute sets sharp lines. A jurisdiction is covered for a language group if
**either**:

- more than **10,000** voting-age citizens in that group have limited English
  proficiency, **or**
- more than **5%** of the jurisdiction's voting-age citizens are in that group
  and have limited English proficiency,

**and** that group's illiteracy rate exceeds the national rate. Alaska Native and
American Indian areas have additional provisions we will come to.

Today you have the Census Bureau's own file — the actual arithmetic behind the
2021 determinations. Everything is base R. Work in order.

***

# Part 1: Who is covered

```{r read}
det <- read.csv("data/sect203_determined.csv", stringsAsFactors = FALSE)
dim(det)
```

Every jurisdiction the Bureau made a determination about in 2021.

```{r covered}
cov <- det[det$FLAG_COV == 1, ]
nrow(cov)
length(unique(cov$S203_GEOID))
```

**334 jurisdictions**, some covered for more than one language.

```{r languages}
sort(table(cov$LANGUAGE), decreasing = TRUE)[1:10]
```

Spanish — recorded here as "Hispanic" — dominates, at 235 of 391 coverage
determinations. But look further down the list.

```{r rare-languages}
tail(sort(table(cov$LANGUAGE), decreasing = TRUE), 12)
```

> Choctaw. Navajo. Yup'ik. Ute. Apache. Inupiat. Hopi. Paiute. Aleut. Seminole.
> Shoshone.
>
> Somewhere in America a county must print its ballots in Shoshone, because
> enough Shoshone speakers live there and Congress decided in 1975 that this was
> a right rather than a courtesy.

***

# Part 2: The line

Now the file that makes the analysis possible: every county in the country,
every language group with any limited-English population at all.

```{r counties}
cty <- read.csv("data/sect203_counties.csv", stringsAsFactors = FALSE)
dim(cty)
sum(cty$FLAG_COV == 1)
```

Twenty-three thousand county-language pairs. Two hundred and ninety-six covered.

The columns that matter:

- `VACLEP` — voting-age citizens in the group with limited English proficiency
- `MVACLEP` — **the margin of error on that estimate**
- `LEPPCT` — the same thing as a percentage of all voting-age citizens
- `FLAG10`, `FLAG5`, `FLAG_EDU` — which statutory test the jurisdiction passed
- `FLAG_COV` — covered or not

Let us watch the 10,000 rule work:

```{r rule-works}
near <- cty[cty$VACLEP >= 9000 & cty$VACLEP <= 11500, ]
near <- near[order(-near$VACLEP), ]
near[, c("NAMELSAD", "LANGUAGE", "VACLEP", "FLAG10", "FLAG_COV")]
```

The rule does exactly what it says. Above 10,000, `FLAG10` is 1. Below, it is 0.

***

# Part 3: The line is not as sharp as the law thinks

Every one of those numbers is an **estimate from a survey**, not a count. The
American Community Survey asks a sample. The Census Bureau publishes, alongside
each estimate, how uncertain it is — and that column has been sitting in the file
the whole time.

```{r moe}
near[, c("NAMELSAD", "LANGUAGE", "VACLEP", "MVACLEP", "FLAG10", "FLAG_COV")]
```

Look at the margins. They are typically 300 to 500 — larger than the distance
many of these counties sit from the threshold.

```{r straddle}
cty$lo <- cty$VACLEP - cty$MVACLEP
cty$hi <- cty$VACLEP + cty$MVACLEP

s10 <- cty[cty$lo < 10000 & cty$hi > 10000, ]
nrow(s10)
s10[order(-s10$VACLEP),
    c("NAMELSAD", "LANGUAGE", "VACLEP", "MVACLEP", "FLAG10", "FLAG_COV")]
```

> **Eight counties sit close enough to the threshold that the survey cannot tell
> which side of it they are on.** Four are covered, four are not — and one of the
> four covered, Kings County, got there through the 5% test rather than this one.
>
> **Philadelphia County** is covered for Chinese: the estimate is 10,150, which
> clears 10,000 by 150. The margin of error is **450**.
>
> **Solano County, California** is not covered for Spanish: the estimate is
> 9,943, which misses by 57. The margin of error is **377**.
>
> On the evidence in this file, those two counties are statistically
> indistinguishable. One must print Chinese ballots. The other need not print
> Spanish ones. The law does not have a category for "we cannot tell."

```{r straddle-plot, fig.width = 8, fig.height = 5.5}
s <- s10[order(s10$VACLEP), ]
lab <- paste0(sub(" County", "", s$NAMELSAD), " (",
              substr(s$LANGUAGE, 1, 10), ")")

par(mar = c(5, 12, 4, 2))
plot(s$VACLEP, seq_len(nrow(s)), xlim = range(c(s$lo, s$hi)),
     pch = 19, col = ifelse(s$FLAG_COV == 1, "#2166AC", "#B2182B"),
     yaxt = "n", ylab = "", xlab = "Voting-age citizens with limited English",
     main = "Eight counties the survey cannot place")
axis(2, at = seq_len(nrow(s)), labels = lab, las = 1, cex.axis = 0.8)
segments(s$lo, seq_len(nrow(s)), s$hi, seq_len(nrow(s)),
         col = ifelse(s$FLAG_COV == 1, "#2166AC", "#B2182B"), lwd = 2)
abline(v = 10000, lty = 2, lwd = 2)
legend("topright", c("covered", "not covered"), col = c("#2166AC", "#B2182B"),
       pch = 19, bty = "n", cex = 0.85)
```

Every horizontal bar crosses the dashed line. The colour — the legal
consequence — is decided by where the dot happens to fall.

***

# Part 4: The percentage test is worse

```{r pct-straddle}
p5 <- cty[!is.na(cty$LEPPCT) &
          cty$LEPPCT - cty$MLEPPCT < 5 &
          cty$LEPPCT + cty$MLEPPCT > 5, ]
nrow(p5)
```

**Fifty-one counties** cannot be placed on the 5% line by their own survey
estimate.

```{r pct-examples}
head(p5[order(-p5$VACLEP),
        c("NAMELSAD", "LANGUAGE", "VACIT", "VACLEP", "LEPPCT", "MLEPPCT",
          "FLAG5", "FLAG_COV")], 8)
```

Small counties are the problem. A percentage estimated from a small sample is
badly behaved, and the smaller the county the wider the margin — which means the
rule is least reliable exactly where a handful of voters would notice a
translated ballot most.

***

# Part 5: The exception that proves the rule

Some jurisdictions are covered despite failing every numeric test:

```{r exceptions}
odd <- cty[cty$FLAG_COV == 1 &
           cty$FLAG10 == 0 & cty$FLAG5 == 0, ]
nrow(odd)
head(odd[, c("NAMELSAD", "LANGUAGE", "VACLEP", "LEPPCT",
             "FLAG10", "FLAG5", "FLAG_AIAN", "FLAG_ANRC", "FLAG_COV")], 8)
```

Bristol Bay Borough is covered for Yup'ik with an estimated **3** limited-English
voting-age citizens. Kodiak Island Borough with **5**. Lake and Peninsula Borough
with **9**.

These are the Alaska Native and American Indian provisions. For Native areas the
determination is made on the **reservation or Native village**, not on the county
that contains it — and where the population is tiny, the statute stops trying to
measure and simply covers.

> Congress built an escape hatch into the arithmetic for exactly the places where
> a sample survey cannot measure anything reliably. Whether the rest of the
> statute should have one too is the question this lab leaves you with.

***

# Part 6: What this file does and does not settle

**It does not tell you whether coverage works.** Nothing here measures turnout,
registration, or whether a translated ballot changed anyone's vote.

**It does not tell you what compliance looks like.** A covered jurisdiction is
legally obliged to provide materials. Whether it does, and how well, is a
separate question and a live subject of litigation.

**It does tell you that a sharp legal line has been drawn across a fuzzy
measurement**, and it lets you name the counties where that matters.

***

# Part 7: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change, or nothing at all.

Pick **one**.

***

### Option A — Move the threshold

```{r opt-a}
threshold <- 10000            # <- change this. Try 7500, 9000, 12500, 15000.

elig <- cty[cty$VACLEP > threshold & cty$FLAG_EDU == 1, ]
c(threshold = threshold, counties_qualifying = nrow(elig))

# Who is added or dropped relative to the real 10,000 rule?
real <- cty$VACLEP > 10000 & cty$FLAG_EDU == 1
now  <- cty$VACLEP > threshold & cty$FLAG_EDU == 1
added   <- cty[now & !real, c("NAMELSAD","LANGUAGE","VACLEP")]
dropped <- cty[real & !now, c("NAMELSAD","LANGUAGE","VACLEP")]
c(added = nrow(added), dropped = nrow(dropped))
head(added[order(-added$VACLEP), ], 8)
```

**Write about:** how sensitive is the map of covered jurisdictions to a number
Congress chose? Try a few thresholds and report how many counties change status.
Then the harder question: 10,000 is a round number in a statute written in 1975
and reauthorised since. Is there anything in this data that would justify 10,000
rather than 9,000 or 12,000 — or is the number simply a decision that had to be
made somehow? Does that make it a bad rule?

***

### Option B — Who is closest to the line?

```{r opt-b, fig.width = 8, fig.height = 5.5}
band <- 1500                  # <- change this window around 10,000

b <- cty[abs(cty$VACLEP - 10000) <= band, ]
b$dist <- b$VACLEP - 10000
b <- b[order(b$dist), ]

plot(b$dist, seq_len(nrow(b)), pch = 19,
     col = ifelse(b$FLAG_COV == 1, "#2166AC", "#B2182B"),
     xlab = "Distance from the 10,000 threshold", ylab = "",
     yaxt = "n", main = paste("Counties within", band, "of the line"))
abline(v = 0, lty = 2, lwd = 2)
segments(b$dist - b$MVACLEP, seq_len(nrow(b)),
         b$dist + b$MVACLEP, seq_len(nrow(b)),
         col = ifelse(b$FLAG_COV == 1, "#2166AC88", "#B2182B88"))

nrow(b)
sum(b$FLAG_COV == 1)
b[, c("NAMELSAD","LANGUAGE","VACLEP","MVACLEP","FLAG_COV")]
```

**Write about:** name the counties on each side and the languages at stake. How
many of the error bars cross zero? Then think like a county clerk: you are 57
people short, the margin of error is 377, and translating your ballots would cost
real money. What would you do, and what would you want the law to say?

***

### Option C — Which languages sit closest to disappearing?

```{r opt-c}
langs <- aggregate(cbind(VACLEP, MVACLEP) ~ LANGUAGE,
                   data = cty[cty$FLAG_COV == 1, ], FUN = median)
counts <- as.data.frame(table(cty$LANGUAGE[cty$FLAG_COV == 1]))
names(counts) <- c("LANGUAGE", "counties")
lt <- merge(counts, langs, by = "LANGUAGE")
lt$moe_pct <- round(100 * lt$MVACLEP / lt$VACLEP, 1)

lt[order(lt$counties), c("LANGUAGE","counties","VACLEP","MVACLEP","moe_pct")]
```

**Write about:** the languages covered in only one or two counties are mostly
Native American. Look at `moe_pct` — the margin of error as a share of the
estimate itself. For which languages is the underlying measurement least
precise, and why would that be? Then connect it to Part 5: if the ACS cannot
measure small populations well, is the Alaska Native provision a special
favour, or a correction for a known defect in the instrument?

***

### Option D — Ask your own question

Look up your own county, or Allegheny County. Compare states. Find the largest
limited-English population that is *not* covered and work out why.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can look at a legal threshold applied to uncertain data and
say something precise about what that costs and who it falls on.

***

## Where the data came from

**U.S. Census Bureau**, Redistricting and Voting Rights Data Office, *2021
Section 203 Determinations — Public Use File*
(<https://www.census.gov/programs-surveys/decennial-census/about/voting-rights/voting-rights-determination-file.html>).
Determinations published 8 December 2021 under 52 U.S.C. §10503, using American
Community Survey estimates for 2015–2019 and Census 2020 population counts.

This is the arithmetic itself, published by the agency that performed it. It is
the most directly consequential dataset in this course: a `1` in the `FLAG_COV`
column is a legal obligation on a county government.

Three things to hold on to:

- **The estimates are survey estimates.** The ACS samples about 3.5 million
  addresses a year. Every number here carries the uncertainty of a sample, and
  the Bureau publishes that uncertainty honestly in the `M`-prefixed columns.
  Very few consequential datasets do.
- **Determinations are made every five years.** The next is due in 2026. The
  jurisdictions on the wrong side of a margin of error today may be on the other
  side of it then, for reasons that have nothing to do with anybody moving.
- **Coverage is not compliance.** This file records who is obliged, not who
  complies.

The script that downloaded, filtered and checked all of this is
`data/build-data.R`. The full public use file is about 120 MB; the committed
extracts are the county level and the determined jurisdictions.
