---
title: "Data Session 3 Key: A Threshold in the Law"
subtitle: "84-355 Democracy's Data | Session 3 — The ACS and its margins | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
det <- read.csv("data/sect203_determined.csv", stringsAsFactors = FALSE)
cty <- read.csv("data/sect203_counties.csv", stringsAsFactors = FALSE)
cty$lo <- cty$VACLEP - cty$MVACLEP
cty$hi <- cty$VACLEP + cty$MVACLEP
```

All Part 7 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

Section 203 draws a sharp legal line across an estimate that is not sharp, and
the Census Bureau's own file tells you exactly which counties fall in the gap.

***

# The findings

```{r headline}
cv <- det[det$FLAG_COV == 1, ]
c(coverage_rows = nrow(cv), jurisdictions = length(unique(cv$S203_GEOID)),
  spanish_rows = sum(cv$LANGUAGE == "Hispanic"))

s10 <- cty[cty$lo < 10000 & cty$hi > 10000, ]
s10[order(-s10$VACLEP),
    c("NAMELSAD","LANGUAGE","VACLEP","MVACLEP","FLAG10","FLAG5","FLAG_COV")]
```

- **334 jurisdictions**, 391 coverage determinations; **235** are for Spanish.
- **Eight counties** have a margin of error straddling the 10,000 threshold.
  **Four covered, four not.**
- **51 counties** straddle the 5% threshold.

## The pair to teach

**Philadelphia County**, Chinese: estimate **10,150**, margin **±450** →
**covered**, clearing the line by 150.

**Solano County, California**, Spanish: estimate **9,943**, margin **±377** →
**not covered**, missing by 57.

Both intervals contain 10,000. The survey cannot distinguish these counties. The
law treats one as legally obliged to translate its ballots and the other as free
not to.

Note for class: **Kings County, California** is in the straddle list and *is*
covered, but through the 5% test (`FLAG5 == 1`), not the 10,000 test. Students who
notice this have read the flags carefully and should be praised for it — it is
the difference between "covered" and "covered by the rule we are discussing."

***

# Option A — Move the threshold

## Verified output

```{r a}
real <- cty$VACLEP > 10000 & cty$FLAG_EDU == 1
for (th in c(7500, 9000, 10000, 12500, 15000)) {
  now <- cty$VACLEP > th & cty$FLAG_EDU == 1
  cat(sprintf("threshold %6d: qualifying %4d   added %3d   dropped %3d\n",
              th, sum(now), sum(now & !real), sum(real & !now)))
}
```

| threshold | qualifying | added vs. real rule | dropped |
|---|---|---|---|
| 7,500 | 163 | 33 | 0 |
| 9,000 | 145 | 15 | 0 |
| **10,000** | **130** | — | — |
| 12,500 | 101 | 0 | 29 |
| 15,000 | 86 | 0 | 44 |

*(This table was transcribed from an earlier build that still included
non-county rows, and read 194 / 175 / 161 / 139 / 122. The chunk above always
printed the right numbers. Run the chunk; do not trust a typed table.)*

## What a good answer says

**3/4:** Reports that the count moves substantially and that a modest change in
the number changes coverage for dozens of counties.

**4/4:** Confronts the arbitrariness honestly. Nothing in the data justifies
10,000 over 9,000. It is a round number, chosen by Congress in 1975, applied
nationally to counties ranging from a few thousand people to ten million. A
student who says "there is no natural break in this distribution at 10,000" and
checks that claim has done the work.

**4/4, best:** resists the easy conclusion that arbitrary means bad. Any
bright-line rule is arbitrary at the margin; the alternative — case-by-case
judgement about who deserves translated ballots — has its own and probably worse
failure modes. The real criticism is not that the line is arbitrary but that the
statute treats an *estimate* as if it were a *count*. That is a fixable defect;
arbitrariness is not.

***

# Option B — Who is closest to the line?

## Verified output

```{r b}
b <- cty[abs(cty$VACLEP - 10000) <= 1500, ]
b <- b[order(b$VACLEP - 10000), ]
c(counties_in_band = nrow(b), covered = sum(b$FLAG_COV == 1),
  error_bars_crossing_the_line = sum(b$lo < 10000 & b$hi > 10000))
b[, c("NAMELSAD","LANGUAGE","VACLEP","MVACLEP","FLAG_COV")]
```

## What a good answer says

**3/4:** Names the counties and languages and counts the crossing intervals.

**4/4:** Takes the clerk's position seriously. The honest answer is that a
county 57 short with a margin of 377 has no way to know whether it is legally
obliged, and the statute gives it no mechanism to find out. Translating ballots
voluntarily costs money it may not have; not translating risks litigation. A
student who identifies that the *uncertainty itself* imposes a cost on
jurisdictions — separate from the cost of translation — has found something
genuinely worth writing about.

**4/4, best:** proposes a fix and reasons about its cost. Options include
covering any jurisdiction whose confidence interval crosses the threshold
(expands coverage, more expensive, errs toward access), or covering only those
whose *lower bound* clears it (contracts coverage, errs toward cost). Naming the
direction of the error is the point.

***

# Option C — Which languages sit closest to disappearing?

## Verified output

```{r c}
langs <- aggregate(cbind(VACLEP, MVACLEP) ~ LANGUAGE,
                   data = cty[cty$FLAG_COV == 1, ], FUN = median)
counts <- as.data.frame(table(cty$LANGUAGE[cty$FLAG_COV == 1]))
names(counts) <- c("LANGUAGE", "counties")
lt <- merge(counts, langs, by = "LANGUAGE")
lt$moe_pct <- round(100 * lt$MVACLEP / lt$VACLEP, 1)
head(lt[order(lt$counties), c("LANGUAGE","counties","VACLEP","MVACLEP","moe_pct")], 12)
```

## What a good answer says

**3/4:** Notices the single-county languages are overwhelmingly Native American
and that their margins are proportionally larger.

**4/4:** Explains the mechanism — the ACS margin of error shrinks with sample
size, so small populations get proportionally huge uncertainty. The languages
with the fewest speakers are measured worst. Then draws the consequence: the
groups whose language rights are most precarious are also the ones the
instrument can least reliably detect.

**4/4, best:** answers the Alaska question the right way round. The AIAN and ANRC
provisions are not a favour; they are a **correction for a known defect in the
measurement**. Congress knew the ACS could not resolve a village of 400, so it
stopped asking it to. A student who frames the special provision as
methodological rather than political has understood both the statute and the
survey.

***

# Option D — Their own question

Allegheny County is worth a look. So is the largest uncovered limited-English
population in the file — students are often surprised at how large it is.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **This lab rewards precision about consequences.** The move that earns a 4 is
  connecting a statistical fact (a margin of error) to a legal fact (an
  obligation on a county) without overstating either. Students who write "so
  Section 203 is broken" without saying what specifically is broken have not
  done it.
- **Reward students who distinguish the tests.** `FLAG10`, `FLAG5` and
  `FLAG_EDU` are separate conditions and several counties qualify on one and not
  another.

## Anticipated questions

- *"Is this a scandal?"* — Not exactly. Every bright-line rule has a margin. The
  defensible criticism is narrower and more interesting: the statute is written
  as if the input were a census count when it is a survey estimate, and the file
  publishes the uncertainty that the rule then ignores.
- *"Why 'Hispanic' rather than 'Spanish'?"* — The statute defines language
  minority groups by heritage categories, not by language spoken. That is a
  genuine oddity worth a minute; the covered obligation is to provide Spanish
  materials.
- *"Bristol Bay is covered with three people?"* — Yes, via the Alaska Native
  provisions, where determination is made on the Native village rather than the
  borough. Part 5.
- *"When does this change?"* — Determinations are every five years; **the next
  is due in 2026**, using more recent ACS data. Counties in the straddle band may
  flip for purely statistical reasons. This is live, not historical.
- *"Does anyone challenge these?"* — Yes. Coverage determinations and compliance
  are both litigated. Worth mentioning your own expert-witness work here if you
  want the room's attention.

## Connections forward

- **Session 2 (the decennial census)** is the direct prerequisite: this lab is what the ACS
  is *for*, legally speaking. The margins of error students meet abstractly in
  Session 2 acquire a legal consequence here.
- **Session 7 (surnames and BISG)** — another case of inferring group membership
  from imperfect data, with the same tension between usefulness and error.
- **Session 14 (RPV and precinct maps)** — VRA §2 litigation depends on estimates
  of racially polarized voting that carry their own uncertainty, and courts
  treat those estimates as facts too. Same structural problem, higher stakes.
- **Session 1 (election returns)** — the provenance lesson, now with legal force. A `1` in a column is
  an obligation on a county government.
