---
title: "Part VI — Putting Data Together"
type: chapter
subtitle: "Where the interesting questions and nearly all the errors live"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("part-6-putting-data-together-code.R")
```

```{r setup, include=FALSE}
```

Every part before this one worked with one source at a time. That was a simplification, and this part drops it.

**Nothing here introduces new data.** Every file these `r NCH` chapters touch comes from elsewhere in this book, not from a new fetch. What is new is the *combining*. A name and an address joined to a guess about race. A precinct result joined to the demographics of the precinct. A district map graded against a rule somebody wrote down.

That is where the questions people actually care about live. It is also where nearly all the wrong answers come from, and the two facts are related.

## Nobody recorded any of these numbers

Start with what these chapters produce, because it is a genuinely different kind of object from anything so far.

There is no file anywhere that records a voter's race, if the voter's state does not ask. There is no file that records how Black voters in a county voted, because ballots are secret. There is no file that records whether a district was drawn to dilute somebody's vote. **Every one of those numbers is manufactured** — computed from other numbers, by somebody, using a method with assumptions in it.

Once it is computed it looks exactly like the measured numbers around it. It has the same number of decimal places. It goes in the same column. The assumptions do not come with it.

## So every chapter here is a test

Because these numbers cannot be checked against a file, they have to be checked another way: run the method on a case where the true answer is already known, and see whether it gets it.

That is what this part does, over and over. Guess people's race from their names and addresses, then compare against a state that records race and see how often the guess is right. Estimate how a group voted from precinct totals, then compare against ballots that were actually counted separately. Search for a fair map with an algorithm, then check whether the map it could not find has been law for years.

The answers are not reassuring, and that is the point of putting them last. **You now know enough about where data comes from to understand why the methods fail where they do** — which is a different and more useful thing than being told not to trust them.

## The shape of this part

Every part of this book is built out of the same few moves, though no part uses all of them and the order changes when the material demands it. Here is what this one runs, and what each move is for:

```{r beattab}
```

The rule is cumulative: **nothing in a later move is worked out from a source you have not already met.** If a chapter uses a file, the chapter that introduced that file came earlier.

```{r chaptab}
```

There is only one row in that first table, and it is the honest description of this part: every chapter takes a number somebody built and grades it.

## Where this part sits

```{r reusetab}
```

Read those two numbers together, because they are the shape of a final part.

**This part reads files from `r nn(RU("Chapters here that read another part\'s files"))` chapters elsewhere in the book, and no chapter anywhere reads its files.** It imports more than any other part and exports nothing. That is what a destination looks like: everything flows in, nothing flows out, and the book stops here because there is nowhere further to carry it.

## How to read this part

**Read it in order.** This is the one part of the book where that matters. The chapters build: uncertainty sets up what a margin does not cover, the race imputation chapter supplies a method, the voting-estimate chapter uses it, and the districting chapters put the whole apparatus against real law.

If you only read one, read [the racially polarized voting chapter](../rpv/rpv-brief.html) — a method used in federal voting-rights litigation, run against an answer key, getting an impossible number.

## Sources

- The corpus itself. The chapter list, its order and each chapter's subject are
  read from this book's own `INDEX.md`, so this page cannot disagree with the
  part it opens.
- No new data is acquired in this part. Every file is one introduced earlier in
  the book, and each chapter names the chapter it borrowed from.
