---
title: "Data Session 7 Key, part 1: Guessing Who People Are"
subtitle: "84-355 Democracy's Data | Session 7 — Surnames and BISG | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
sn <- read.csv("data/census_surnames.csv", stringsAsFactors = FALSE)
co <- read.csv("data/county_race.csv", stringsAsFactors = FALSE,
               colClasses = c(fips = "character"))
grp  <- c("pctwhite","pctblack","pctapi","pctaian","pct2prace","pcthispanic")
grp6 <- c("white","black","api","aian","twoplus","hispanic")
natl <- colSums(co[, grp6] * co$pop) / sum(co$pop)
look <- function(nm) sn[sn$name == toupper(nm), ]
bisg <- function(name, fips) {
  s <- look(name)
  p <- c(s$pctwhite,s$pctblack,s$pctapi,s$pctaian,s$pct2prace,s$pcthispanic)/100
  p[is.na(p)] <- 0
  g <- as.numeric(co[co$fips == fips, grp6])
  u <- p * g / natl
  setNames(round(100 * u / sum(u), 1), grp6)
}
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(x, big.mark = ",", trim = TRUE)
big <- sn[sn$count > 1000, ]
n90 <- sapply(grp, function(v) sum(big[[v]] > 90, na.rm = TRUE))
# how decisive is a name, for how many people
mx <- suppressWarnings(apply(sn[, grp], 1, max, na.rm = TRUE))
mx[!is.finite(mx)] <- NA
tot_ppl  <- sum(sn$count)
dec_90   <- 100 * sum(sn$count[!is.na(mx) & mx > 90]) / tot_ppl
amb_n    <- sum(sn$count[!is.na(mx) & mx < 60])
amb_one  <- tot_ppl / amb_n
# suppression by bucket
sn$bucket <- cut(sn$count, breaks = c(0, 200, 500, 1000, 5000, 1e7),
                 labels = c("<200", "200-500", "500-1k", "1k-5k", ">5k"))
supp     <- 100 * tapply(is.na(sn$pctblack), sn$bucket, mean)
supp_pk  <- max(supp); supp_pk_b <- names(supp)[which.max(supp)]
supp_all <- 100 * mean(is.na(as.matrix(sn[, grp])))
supp_col <- sapply(grp, function(v) 100 * tapply(is.na(sn[[v]]), sn$bucket, mean))
pk_bkt   <- apply(supp_col, 2, function(x) rownames(supp_col)[which.max(x)])
# Bronx rescue
bronx <- function(nm) {
  s <- look(nm)
  before <- max(c(s$pctwhite, s$pctblack, s$pctapi, s$pcthispanic), na.rm = TRUE)
  after  <- max(bisg(nm, "36005"))
  c(before = before, after = after, gain = after - before)
}
BR <- sapply(c("SMITH","WILLIAMS","JOHNSON","LEE","NGUYEN","GARCIA"), bronx)
# Williams across the six counties
WPL <- c("Allegheny PA" = "42003", "Bronx NY" = "36005", "Honolulu HI" = "15003",
         "El Paso TX" = "48141", "Los Angeles CA" = "06037", "Manhattan NY" = "36061")
WM  <- t(sapply(WPL, function(f) bisg("WILLIAMS", f)))
rownames(WM) <- names(WPL)
w_blk_lo <- min(WM[, "black"]); w_blk_hi <- max(WM[, "black"])
w_blk_lo_nm <- rownames(WM)[which.min(WM[, "black"])]
w_blk_hi_nm <- rownames(WM)[which.max(WM[, "black"])]
```

All Part 6 code is given to students. Grade the interpretation.

Every number below was run against the data file.

***

# The lab in one sentence

BISG works — well enough that courts rely on it — and it works by attaching a
population-level probability to an individual person, which is both its utility
and the thing that should never be forgotten about it.

***

# The findings

```{r headline}
big <- sn[sn$count > 1000, ]
sapply(grp, function(v) sum(big[[v]] > 90, na.rm = TRUE))
rbind(look("WASHINGTON"), look("JEFFERSON"), look("LINCOLN"))
rbind(SMITH_Allegheny = bisg("SMITH","42003"), SMITH_Bronx = bisg("SMITH","36005"))
rbind(LEE_Allegheny = bisg("LEE","42003"), LEE_Honolulu = bisg("LEE","15003"))
```

1. **The asymmetry.** Of surnames held by more than 1,000 people,
   **`r nn(n90[["pctwhite"]])`** are more than 90% white;
   **`r nn(n90[["pctblack"]])`** are more than 90% Black. A
   `r round(n90[["pctwhite"]] / n90[["pctblack"]], -1)`-fold gap for a group that
   is an eighth of the country.
2. **Smith flips.** `r pc(look("SMITH")$pctwhite)`% white nationally →
   **`r pc(bisg("SMITH","36005")[["black"]])`% Black** in the Bronx.
3. **Lee flips the other way.** `r pc(look("LEE")$pctapi, 0)`% API by name →
   `r pc(bisg("LEE","42003")[["api"]])`% API in Allegheny,
   **`r pc(bisg("LEE","15003")[["api"]], 0)`%** in Honolulu.
4. **A third of the file is suppressed** — `r pc(supp_all)`% of cells, and the
   Black column is withheld `r pc(supp[["<200"]])`% of the time among names held
   by fewer than 200 people. **The gradient is not monotone**: suppression
   *peaks* at `r pc(supp_pk)`% in the `r supp_pk_b` bucket and only then
   collapses, to `r pc(supp[[">5k"]])`% above five thousand bearers.

***

# The three things to get across

**The asymmetry is history, not a data defect.** Enslaved people took or were
given the surnames of enslavers. Black and white Americans have shared a surname
stock for 160 years, which is exactly why surname-only inference fails for Black
Americans and works well for recent immigrant groups. Students often assume the
method is *biased against* Black Americans in a design sense; it is more precise
to say the signal is genuinely absent, for reasons that are themselves the
subject of the Voting Rights Act.

**Washington and Jefferson are the exceptions, and Lincoln is not.** Black share
of the name: `r pc(look("WASHINGTON")$pctblack)`%,
`r pc(look("JEFFERSON")$pctblack)`%, and `r pc(look("LINCOLN")$pctblack)`%.
Worth pausing on: nobody organised this. It is millions of
individual decisions after emancipation, still legible in a 2010 file.

**Geography does most of the work.** This is the point students most often miss.
For ambiguous names — which is most American names — the county is carrying the
inference, not the surname. That has an uncomfortable implication: BISG is
substantially a method for guessing race from address.

***

# Option A — One name, many places

## Verified output

```{r a}
pl <- c("42003","36005","15003","48141","06037","36061")
o <- t(sapply(pl, function(f) bisg("WILLIAMS", f)))
rownames(o) <- c("Allegheny PA","Bronx NY","Honolulu HI","El Paso TX",
                 "Los Angeles CA","Manhattan NY")
o
```

Williams — `r pc(look("WILLIAMS")$pctwhite)`% white /
`r pc(look("WILLIAMS")$pctblack)`% Black by name alone — ranges across these six
counties from **`r pc(w_blk_lo)`% Black in `r w_blk_lo_nm`** to
**`r pc(w_blk_hi)`% Black in `r w_blk_hi_nm`**, with Allegheny at
`r pc(WM["Allegheny PA", "black"])`%. In El Paso it becomes
`r pc(WM["El Paso TX", "hispanic"])`% Hispanic; in Honolulu it lands at
`r pc(WM["Honolulu HI", "twoplus"])`% two-or-more-races, which is a real
feature of Hawaii's population and not an error.

**The scale of the problem, now computed in Part 1.** For **`r pc(dec_90)`% of
Americans** a surname is nearly decisive — one group above 90%. For
**`r nn(amb_n)` people, about one in `r round(amb_one)`**, no group reaches even
60%, so the name settles nothing. Williams is the
`r c("first","second","third")[look("WILLIAMS")$rank]` most common surname in the
country, held by `r pc(look("WILLIAMS")$count / 1e6, 2)` million people; this is
not an edge case.

Say the consequence plainly: BISG is not uniformly good or uniformly bad. It is
excellent for Nguyen, useless for Williams, and **the people it fails on are not
a random sample of the country** — they are concentrated among exactly the
surnames that are shared across Black and white America for historical reasons
this course has already covered.

## What a good answer says

**3/4:** Reports the range and identifies which counties move it most.

**4/4:** Answers the court question honestly. For Williams, nearly all the
discrimination between outcomes is coming from the address, because the name
itself is a coin flip. A student who says "if I told you only the name you would
learn almost nothing, and if I told you only the county you would learn most of
what the model outputs" has understood what BISG actually is.

**4/4, best:** notices the discomfort and states it precisely — the method infers
race from residential geography, and residential geography in America is
segregated. The estimate is accurate *because* of segregation. That is worth
saying out loud, and it is a fair thing for a court to be told.

**Common stumble:** treating a high probability as a certainty about a named
individual. Push on this in feedback every time.

***

# Option B — Which names does geography rescue?

## Verified output

```{r b}
for (nm in c("SMITH","WILLIAMS","JOHNSON","LEE","NGUYEN","GARCIA","WASHINGTON")) {
  s <- look(nm)
  before <- max(c(s$pctwhite,s$pctblack,s$pctapi,s$pcthispanic), na.rm = TRUE)
  after  <- max(bisg(nm, "36005"))
  cat(sprintf("%-12s %5.1f%% -> %5.1f%%  (%+.1f)\n", nm, before, after, after - before))
}
```

In the Bronx the large gains are Williams
`r sprintf("%+.1f", BR["gain","WILLIAMS"])` and Johnson
`r sprintf("%+.1f", BR["gain","JOHNSON"])`. **Read Smith carefully**: the
leading share only moves `r sprintf("%+.1f", BR["gain","SMITH"])`, but the
*group* it belongs to flips, from `r pc(BR["before","SMITH"])`% white nationally to
`r pc(BR["after","SMITH"])`% Black in the Bronx — the number is nearly unchanged
and the answer is completely different. Nguyen
(`r sprintf("%+.1f", BR["gain","NGUYEN"])`) and Garcia
(`r sprintf("%+.1f", BR["gain","GARCIA"])`, into an already-decisive
`r pc(BR["after","GARCIA"])`%) are not rescued, because neither needed rescuing.

## What a good answer says

**3/4:** Reports that ambiguous names gain and decisive names do not.

**4/4:** Explains it as Bayes' rule behaving correctly — a sharp prior is hard to
move, a flat prior is easy. Then makes the inference that matters: **the names
that need help are the common ones**, and common names cover most Americans.
Nguyen and Garcia are precise but describe a minority of the population; Smith,
Johnson and Williams are three of the five most common surnames in the country
and are all essentially uninformative on their own.

**4/4, best:** connects to Session 3 (§203). Both labs are about a method that works worst
for the population it most needs to serve.

***

# Option C — The cost of suppression

## Verified output

```{r c}
sn$bucket <- cut(sn$count, breaks = c(0,200,500,1000,5000,1e7),
                 labels = c("<200","200-500","500-1k","1k-5k",">5k"))
round(100 * tapply(is.na(sn$pctblack), sn$bucket, mean), 1)
c(names_under_1000 = sum(sn$count < 1000),
  share = round(100 * mean(sn$count < 1000), 1))
```

Suppression of the Black column runs **`r pc(supp[["<200"]])`%** for names held
by fewer than 200 people, then **rises** to `r pc(supp[["200-500"]])`% and peaks
at **`r pc(supp_pk)`% in the `r supp_pk_b` bucket** before collapsing to
`r pc(supp[["1k-5k"]])`% and `r pc(supp[[">5k"]])`%. **It is not a smooth
decline, and the printed table above says so** — do not let the shape be
described as one, in your own comments or a student's.
**`r pc(100 * mean(sn$count < 1000))`%** of all surnames in the file are held by
fewer than 1,000 people.

If a student asks *why* the hump is there, the honest answer is that the file
does not say. The Bureau publishes the suppression flag, not the rule that
generated it, and published cells include exact zeros — so the pattern cannot be
reverse-engineered from what we have. What *is* checkable, and worth showing, is
that the peak sits in a different bucket for each column: the Hispanic column is
worst at `r pk_bkt[["pcthispanic"]]`, the Black column at
`r pk_bkt[["pctblack"]]`, the American Indian column at `r pk_bkt[["pctaian"]]`.
Whatever the rule is, it does not act on how rare the name is alone.

## What a good answer says

**3/4:** Reads the bucket table and notes that most names in the file are rare.
Reporting the `<200` figure alone is a 3, not a 4 — it is the headline number
and it is not the worst one.

**4/4:** Answers the neutrality question properly. Suppression is applied by a
neutral rule — small cells get hidden — but it does not *land* neutrally. Recent
immigrant communities, small national-origin groups, and Native American families
are disproportionately the holders of rare surnames. A rule can be neutral in
form and unequal in effect, which is a distinction this course has now met three
times (the Electoral College, §203's coverage thresholds in Session 3, and here).

**4/4, best:** proposes what a court should be told — not "the method is
unreliable," but specifically that the *precision varies systematically, and not
monotonically, with the rarity of the name*, and that any estimate should carry
that caveat by subgroup.

***

# Option D — Their own question

Students looking up family names is the most engaged this lab gets. Expect
questions about names that "feel wrong"; usually the answer is that the name is
rare and half its columns are suppressed.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is your code.
- **The move that earns a 4** is holding utility and limitation together. A paper
  that concludes "this is racist pseudoscience" and one that concludes "this is
  a validated method" are both 2s. The method is genuinely relied upon in
  litigation *and* is genuinely a guess; the assignment is to say both.
- **Watch for individual-level claims.** "This tells us Mr Smith is Black" is
  wrong and should be corrected explicitly.

## Anticipated questions

- *"Isn't this racial profiling?"* — Take the question seriously; it is the right
  question. The answer is that the alternative is not neutrality but blindness:
  without some estimate of race, no §2 vote-dilution claim can be brought and no
  lending discrimination can be detected. The method exists because the records
  that would answer honestly do not exist. Whether that trade is acceptable is
  worth ten minutes of class.
- *"Why not just ask people?"* — Some states do collect race on registration.
  Coverage is partial, non-response is high and non-random, and most states
  collect nothing. Worth noting which states do.
- *"Is my name's estimate right?"* — For rare names, often not, and the
  suppressed columns are why.
- *"Does the real method do better?"* — Yes. `wru` adds first names, age, sex,
  party, and block-level geography. Better in degree, not in kind.
- *"Why didn't we use `wru`?"* — Because the arithmetic is three lines and worth
  seeing. Also because the package needs a Census API key, and this course does
  not ask students to obtain credentials.

## Connections forward

- **Session 14 (RPV and ecological inference)** is the payoff. BISG estimates *who
  the voters are*; ecological inference estimates *how groups voted*. Chained
  together they are the standard evidentiary basis of a §2 case — and the errors
  compound. Say so explicitly on 1 December.
- **Session 3 (§203)** — both labs are about legal consequences resting on
  statistical estimates, and both have the same structure: a method least
  reliable exactly where it matters most.
- **Session 2 (the decennial census)** — the suppression rules students meet here are
  disclosure avoidance, the same concern that produced differential privacy in
  the 2020 Census.
