---
title: "Lab 14: Guessing Who People Are"
subtitle: "84-355 Democracy's Data | Week 14"
author: "Your name here"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
```

# Before we start

Voter files do not record race. In most states nobody is asked, and in the states
that do ask, many people decline.

But the Voting Rights Act is about race. To argue in court that a districting
plan dilutes Black or Latino voting strength, somebody has to estimate how many
Black or Latino voters there are and how they voted — from records that do not
say. The method the courts have largely accepted is the one you will build today:
**Bayesian Improved Surname Geocoding**, from Imai and Khanna's 2016 paper, which
you read for Tuesday.

It works by combining two things the government does publish: what a surname
suggests, and where the person lives.

You are going to build it yourself, in three lines of arithmetic, and then spend
the rest of the lab on the harder question of when it should be trusted.

Everything is base R. Work in order.

***

# Part 1: What a surname knows

```{r read}
sn <- read.csv("data/census_surnames.csv", stringsAsFactors = FALSE)
dim(sn)
```

Every surname held by **100 or more people** at the 2010 Census — 162,254 of
them — with the racial and ethnic composition of the people who hold it.

```{r head}
head(sn, 3)
```

The six columns are percentages of the people with that name, and they are
mutually exclusive: non-Hispanic White, non-Hispanic Black, non-Hispanic
Asian/Pacific Islander, non-Hispanic American Indian/Alaska Native, non-Hispanic
two or more races, and Hispanic of any race.

Look one up:

```{r lookup}
look <- function(nm) sn[sn$name == toupper(nm), ]

look("NGUYEN")
look("GARCIA")
```

> Ninety-six percent of Nguyens are Asian or Pacific Islander. Ninety-two percent
> of Garcias are Hispanic. For some names, the surname alone is nearly decisive.

Now try a few others:

```{r lookup2}
rbind(look("SMITH"), look("JOHNSON"), look("WILLIAMS"))
```

> Williams is **45.8% white and 47.7% Black.** The name is a coin flip.

Williams is not an obscure case: it is the **third most common surname in the
country**, held by 1.63 million people. So how much of America does the method
fail on?

```{r ambiguous-scale}
grp6 <- c("pctwhite","pctblack","pctapi","pctaian","pct2prace","pcthispanic")
sn$top <- apply(sn[, grp6], 1, function(x) max(x, na.rm = TRUE))
ok <- is.finite(sn$top) & !is.na(sn$count)

cat("People whose surname is >90% one group:",
    format(sum(sn$count[ok & sn$top > 90]), big.mark = ","),
    sprintf("(%.1f%%)\n", 100 * sum(sn$count[ok & sn$top > 90]) / sum(sn$count[ok])))
cat("People whose surname never reaches 60%:",
    format(sum(sn$count[ok & sn$top < 60]), big.mark = ","),
    sprintf("(%.1f%%)\n", 100 * sum(sn$count[ok & sn$top < 60]) / sum(sn$count[ok])))
```

> **For 37.5% of Americans a surname is nearly decisive. For 27.5 million — one
> in eleven — it settles nothing at all.**
>
> That is the shape of the whole method. BISG is not uniformly good or uniformly
> bad; it is excellent for Nguyen, useless for Williams, and the people it fails
> on are not a random sample of the country.

**Try your own.** Change the name below and re-run. Your name is in this file if
at least a hundred Americans shared it in 2010.

```{r your-name}
look("CERVAS")
```

***

# Part 2: The asymmetry

How often does a surname actually settle the question? Count the names —
restricting to those held by more than a thousand people, so we are not looking
at noise — where one group makes up more than 90% of bearers.

```{r asymmetry}
big <- sn[sn$count > 1000, ]
nrow(big)

grp <- c("pctwhite", "pctblack", "pctapi", "pctaian", "pct2prace", "pcthispanic")
sapply(grp, function(v) sum(big[[v]] > 90, na.rm = TRUE))
```

```{r asymmetry-plot, fig.width = 7.5, fig.height = 5}
n90 <- sapply(grp, function(v) sum(big[[v]] > 90, na.rm = TRUE))
names(n90) <- c("White", "Black", "API", "AIAN", "Two+", "Hispanic")
n90 <- sort(n90)

par(mar = c(5, 8, 4, 2))
barplot(n90, horiz = TRUE, las = 1, col = "#4D4D4D", border = NA,
        xlab = "Number of surnames that are >90% this group",
        main = "How often does a surname give the answer away?")
```

> **12,022 surnames are more than 90% white. Only 122 are more than 90% Black.**
>
> There are a hundred times more names that identify a white American than
> identify a Black American — and Black Americans are about an eighth of the
> country, not a hundredth.

This is not a quirk of the data. It is history sitting in a spreadsheet. Enslaved
people were given, or took, the surnames of enslavers, and after emancipation
those names stayed. Black and white Americans have shared a common stock of
surnames for a hundred and sixty years.

The exceptions are worth seeing:

```{r blackest}
b <- big[!is.na(big$pctblack), ]
head(b[order(-b$pctblack), c("name", "count", "pctblack", "pctwhite")], 8)
```

```{r presidents}
rbind(look("WASHINGTON"), look("JEFFERSON"), look("LINCOLN"))
```

> **Washington is the most identifiably Black common surname in America** —
> 87.5% of the 177,386 people who carry it. Jefferson is 74.2%.
>
> These names were chosen. After emancipation many freed people took the names of
> presidents associated with liberty, and the pattern is still legible in the
> 2010 Census.
>
> Notice that Lincoln is only 14.7% Black — far below the other two. The
> adoption was not uniform, and nobody planned it. Whatever produced this
> pattern was millions of separate decisions, and it is still measurable a
> century and a half later.

***

# Part 3: Where you live

The second ingredient.

```{r counties}
co <- read.csv("data/county_race.csv", stringsAsFactors = FALSE,
               colClasses = c(fips = "character"))
dim(co)
head(co[, c("fips", "county", "state", "pop", "white", "black", "hispanic")], 3)
```

Every county, in the same six categories. Two examples:

```{r county-examples}
co[co$fips %in% c("42003", "36005"),
   c("county", "state", "pop", "white", "black", "api", "hispanic")]
```

Allegheny County is 76% white and 13% Black. The Bronx is 9% white, 30% Black,
and 55% Hispanic. A name means something different in each.

***

# Part 4: Bayes' rule, by hand

Here is the whole method.

We want $P(\text{race} \mid \text{surname}, \text{county})$. We have
$P(\text{race} \mid \text{surname})$ from the surname file and
$P(\text{race} \mid \text{county})$ from the county file. Bayes' rule combines
them:

$$P(r \mid s, g) \;\propto\; \frac{P(r \mid s)\;P(r \mid g)}{P(r)}$$

where $P(r)$ is the national share of group $r$. Dividing by it prevents
double-counting the fact that some groups are large everywhere.

```{r natl}
grp6 <- c("white", "black", "api", "aian", "twoplus", "hispanic")
natl <- colSums(co[, grp6] * co$pop) / sum(co$pop)
round(natl, 4)
```

That is the entire model:

```{r bisg}
bisg <- function(name, fips) {
  s <- sn[sn$name == toupper(name), ]
  p_s <- c(s$pctwhite, s$pctblack, s$pctapi,
           s$pctaian, s$pct2prace, s$pcthispanic) / 100
  p_s[is.na(p_s)] <- 0                      # suppressed cells contribute nothing

  p_g <- as.numeric(co[co$fips == fips, grp6])

  u <- p_s * p_g / natl                     # <- Bayes' rule, one line
  setNames(round(100 * u / sum(u), 1), grp6)
}
```

Three lines. Now watch what it does.

```{r smith-bronx}
bisg("SMITH", "42003")     # Allegheny County, PA
bisg("SMITH", "36005")     # Bronx County, NY
```

> **Smith is 70.9% white nationally. A Smith in the Bronx is 73.7% Black.**
>
> The name did not change. What changed is that we stopped ignoring where the
> person lives.

```{r lee}
bisg("LEE", "42003")       # Allegheny County, PA
bisg("LEE", "15003")       # Honolulu County, HI
```

> Lee is 42% Asian and 36% white by name alone. In Allegheny County it resolves
> to white; in Honolulu it resolves to 90% Asian. The same four letters support
> opposite conclusions.

***

# Part 5: When it fails

**A third of the surname file is missing.**

```{r suppression}
cells <- nrow(sn) * length(grp)
miss  <- sum(is.na(sn[, grp]))
c(cells = cells, suppressed = miss, pct = round(100 * miss / cells, 1))
```

The Census suppresses any cell small enough to risk identifying someone. Those
appear as `NA`, and in the function above they contribute zero — which quietly
assumes the true value is zero when what we actually know is *small and
unknown*. For rare names this matters.

```{r rare}
bucket <- cut(sn$count, breaks = c(0, 200, 500, 1000, 5000, 1e7),
              labels = c("<200", "200-500", "500-1k", "1k-5k", ">5k"))
supp <- 100 * tapply(is.na(sn$pctblack), bucket, mean)
round(supp, 1)
```

> For surnames held by fewer than two hundred people, the Black percentage is
> suppressed `r pc(supp[["<200"]])`% of the time. **That is the headline number,
> and it is not the worst one.** Suppression *rises* to
> `r pc(supp[["200-500"]])`% and peaks at `r pc(max(supp))`% in the
> `r names(supp)[which.max(supp)]` bucket, then falls away to
> `r pc(supp[["1k-5k"]])`% and `r pc(supp[[">5k"]])`%. **The method is not
> simply least informative for the rarest names** — the loss of information
> is not a smooth decline at all. And rarity is not evenly distributed across
> groups.

**And a probability is not a person.** The output of `bisg()` is a statement
about a population: *of all the Smiths in the Bronx, roughly 74% are Black.* It
is not a statement about any individual Smith. Every use of this method involves
taking a number that is true on average and attaching it to somebody in
particular.

That matters because of what it is used for. BISG is used to estimate racial
voting patterns in Voting Rights Act litigation — the evidence on which
districts are drawn or struck down. It is used by the Consumer Financial
Protection Bureau to detect lending discrimination. In both cases a real decision
about real people rests on an inference from a name and a zip code.

The alternative is not a better measurement. The alternative is **no
measurement**, and therefore no ability to detect discrimination at all. That is
the trade Imai and Khanna's paper is really about, and it is worth being clear
that it is a trade rather than a solution.

***

# Part 6: On your own

This is the part you turn in. **You are not being asked to write code.** Each
option is a working chunk with one thing to change.

Pick **one**.

***

### Option A — One name, many places

```{r opt-a}
my_name <- "WILLIAMS"        # <- change this. Try your own name.

places <- c("42003", "36005", "15003", "48141", "06037", "36061")
labels <- c("Allegheny PA", "Bronx NY", "Honolulu HI",
            "El Paso TX", "Los Angeles CA", "Manhattan NY")

out <- t(sapply(places, function(f) bisg(my_name, f)))
rownames(out) <- labels
out

look(my_name)
```

**Write about:** how much does the estimate move across these six counties? Which
county changes it most, and why? Then the question that matters: if this name
were on a voter file and a court needed to know the voter's race, how much of the
answer would come from the name and how much from the address? Would you be
comfortable with the result being used as evidence?

***

### Option B — Which names does geography rescue?

```{r opt-b}
test_names <- c("SMITH", "WILLIAMS", "JOHNSON", "LEE",
                "NGUYEN", "GARCIA", "WASHINGTON")   # <- edit this list

fips_here <- "36005"                                # <- and this county

for (nm in test_names) {
  s <- look(nm)
  before <- max(c(s$pctwhite, s$pctblack, s$pctapi, s$pcthispanic), na.rm = TRUE)
  after  <- max(bisg(nm, fips_here))
  cat(sprintf("%-12s strongest guess: %5.1f%% -> %5.1f%%  (%+.1f)\n",
              nm, before, after, after - before))
}
co[co$fips == fips_here, c("county", "state", "white", "black", "api", "hispanic")]
```

**Write about:** for which names does adding geography sharpen the estimate, and
for which does it barely help? You should find that names which were already
decisive (Nguyen, Garcia) gain little, while ambiguous ones (Smith, Williams)
gain a great deal. Explain why that is exactly what Bayes' rule should do. Then
say which kind of name is more common in America, and what that implies about how
much of BISG's accuracy is really coming from the address rather than the name.

***

### Option C — The cost of suppression

```{r opt-c, fig.width = 7.5, fig.height = 5}
cutoff <- 1000               # <- change this

sn$bucket <- cut(sn$count, breaks = c(0, 200, 500, 1000, 5000, 1e7),
                 labels = c("<200", "200-500", "500-1k", "1k-5k", ">5k"))
supp <- tapply(is.na(sn$pctblack), sn$bucket, mean)
round(100 * supp, 1)

barplot(100 * supp, col = "#B2182B", border = NA,
        ylab = "% of names with Black share suppressed",
        xlab = "How many people hold the name",
        main = "How often the Black share is suppressed, by how common the name is")

c(names_below_cutoff = sum(sn$count < cutoff),
  share_of_all_names = round(100 * mean(sn$count < cutoff), 1))
```

**Write about:** describe the pattern, and read the printed numbers before you
do — suppression does not simply fall away as names get more common, and the
rarest bucket is not the worst one. Now think about who has a rare surname in
the United States — recent immigrants, small national-origin groups, Native
American families. Is a method whose precision depends this much on how common
your name is neutral with respect to who it works worst *for*? Say what you
would want a court to be told about this before the estimate was admitted.

***

### Option D — Ask your own question

Look up your family's names, or the names of people in the news. Find a name
whose estimate you think is wrong and work out why. Compare two counties you
know.

***

**What to submit.** Knit to HTML or PDF and upload to Canvas, with your chosen
chunk and **two or three paragraphs** of interpretation.

The writing is the assignment. I am not grading your code — it is my code. I am
grading whether you can hold two things at once: that this method is genuinely
useful, and that it is genuinely a guess.

***

## Where the data came from

**Surnames** — U.S. Census Bureau, *Frequently Occurring Surnames from the 2010
Census* (<https://www2.census.gov/topics/genealogy/2010surnames/names.zip>).
Every surname held by 100 or more people, with the racial composition of
bearers. Cells small enough to risk identifying an individual are suppressed.

**County composition** — U.S. Census Bureau, Population Estimates Program,
county characteristics file for 2024, collapsed to the same six categories.

**The method** — Kosuke Imai & Kabir Khanna, "Improving Ecological Inference by
Predicting Individual Ethnicity from Voter Registration Records," *Political
Analysis* 24 (2016): 263–72, which you read for Tuesday.

Two notes on what we did *not* use. The published `wru` package implements this
method properly, including first names, age, sex, and party registration, and
geography down to the census block. We have implemented the county-level core by
hand instead, so that the arithmetic is visible rather than hidden inside a
function call — and so that this lab needs no packages and no API key.

The real thing is better than what you built today. It is better in degree, not
in kind: it is still a guess about a person, assembled from facts about
populations.

The script that downloaded, checked and wrote these files is `data/build-data.R`.
