---
title: "Validated Turnout Against Self-Reported Turnout"
type: brief
subtitle: "Counted ballots against what people said they did, 1964-2024"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("validated-turnout-code.R")
```

```{r setup, include=FALSE}
```

Almost every published claim about who takes part in American elections rests on a survey. The Census Bureau asks about 60,000 households after each federal election whether they voted, and nobody checks the answers. The states, separately, count the ballots.

That makes turnout a rare thing: a survey question with an outside answer key. Do the two numbers agree? The question matters to a citizen because the survey is the only source that reports turnout by race, age and education. Those figures are quoted in voting-rights litigation and in nearly every argument about unequal participation. If the survey is wrong it is wrong in a direction, and the published gaps between groups are wrong with it.

## The test

This brief joins one source from each of two earlier sections of this book. The survey half is the U.S. Census Bureau's Current Population Survey *Voting and Registration Supplement*, historical Table A-1, covering November `r Y1` through `r Y2` (<https://www.census.gov/data/tables/time-series/demo/voting-and-registration/voting-historical-time-series.html>). The record half is the count of ballots cast for president in the same `r nrow(cps)` elections, from Jason Timm's *PresElectionResults*, which compiles the official state canvasses (<https://github.com/jaytimm/PresElectionResults>).

Those two sources can answer the question because they measure the same event and share nothing else. The Bureau's interviewers never see a ballot, and the state canvassers never see the survey. Their only common column is the election year, and that is the key the two files join on. A year is about the best-behaved join key in this book. [Join Keys and Crosswalks](../data-sources/data-sources-brief.html) is the chapter on what a join key can and cannot promise. It also covers the crosswalk, a table that translates one set of identifiers into another when two files share no key at all.

One thing about the record half is worth saying before the numbers start. No federal agency records a national popular vote, because the United States does not hold one. The counted-ballot column here is a compilation of the state canvasses, assembled by people rather than published by a government. The word "actual" does slightly more work than the record behind it pays for.

## The data

The table is `cps_turnout.csv`: one row per presidential election, `r nrow(cps)` rows from `r Y1` through `r Y2`. `year` is the join key. `vap_thousands` is the voting-age population the Bureau published, `citizen_vap` the same population counting citizens only, and `pct_citizen_basis` the share of citizens who said they voted. `reported_voters` is how many people the survey implies said they voted, `actual_votes` the ballots cast for president. `cps_rate`, `actual_rate` and `over_report_pp` hold the two rates and the distance between them. Four columns carry reported turnout by group: `pct_white_nh`, `pct_black`, `pct_asian` and `pct_hispanic`.

```{r one-record}
```

The survey half arrives as `hst_vote01.xlsx`, a workbook with no header row. It has a header *region* spread over four rows, formatted for a person to read on a page. The build finds the stub heading `Total, Percent Voted`, starts one row below it, and takes every value by fixed column position.

Three decisions in that build settle what the table can be asked.

**The count of self-reported voters is derived, not published.** Table A-1 prints percentages and one population total. `reported_voters` is the voting-age population multiplied by the published percentage, which is arithmetic on the Bureau's own figures rather than an estimate of anybody's.

**Ballots for president are not the whole electorate.** People vote and skip the top of the ticket, so the true ballot count is a little higher than the column used here. Every gap below is a small overstatement for that reason.

**The rate columns are empty before 1980.** The Bureau began publishing a citizen basis only when the difference between residents and citizens grew large enough to matter. Quietly dropping those blank rows would move the series' start date to 1980, which happens to be its historical peak.

> **Before you read on, commit to a direction.** Over the six decades in this
> file, does the gap between what people say and what the ballots show get
> **larger**, stay **flat**, or get **smaller**? Write down one of the three.

## What it says about democracy

More people say they voted than there are ballots, in `r sum(cps$gap_people > 0)` of `r nrow(cps)` elections. The average excess is `r mn(mean(cps$gap_people))` million people, and in 2016 it was `r cnt(g(2016, "gap_people"))`. That is the finding at its plainest: the survey describes an electorate the states never counted.

```{r band-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r band-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 1. What people said, and what was counted, `r Y1` to `r Y2`.** The upper line is how many Americans the survey implies said they voted; the lower line is ballots cast for president, and the shaded wedge between them is the gap. Two lines with the space between them filled in is the right form here because the question is about a difference rather than a level. Both series roughly doubled over sixty years; the wedge did not. The widest year is `r cps$year[which.max(cps$gap_people)]`, at `r mn(max(cps$gap_people))` million people.

The obvious explanation arrives first and almost everybody writes it down: people lie about voting. It is not an absurd reading. Voting is socially desirable, the interviewer represents the federal government, and admitting you stayed home is mildly embarrassing. Over-reporting of that kind is real and well documented.

But notice what the explanation predicts. If social pressure drives the gap, the gap should hold roughly steady, or grow as voting becomes more moralised in a polarised era. It does neither.

```{r gap-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r gap-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 2. The gap, election by election, `r Y1` to `r Y2`.** The vertical axis is how far the survey electorate exceeded the ballot count, as a percentage of ballots. The dashed segments are era averages: `r pc(em[1])`% through the 1960s and 1970s, `r pc(em[2])`% through the 1980s and 1990s, and `r pc(em[3])`% since 2000.

**The gap shrank by roughly three quarters, and nobody repaired it.** There was no reform, no announced correction, no working group. Measured in percentage points of turnout, the survey overstated the rate by `r pc(g(1988, "over_report_pp"))` points in 1988 and by `r pc(g(2024, "over_report_pp"), 2)` points in `r Y2`.

The exceptions are informative rather than noisy. `r paste(cps$year[cps$gap_people < 0], collapse = ", ")` is the only election in the file where the survey came in *below* the ballot count, by `r cnt(abs(g(2020, "gap_people")))` people. It is also the highest-turnout election in the series, which is what you would expect if part of the gap is people who did not vote being under-represented among those who answer surveys. The largest proportional gap, `r pc(max(cps$gap_pct))`% in `r cps$year[which.max(cps$gap_pct)]`, sits at the other end of the same story.

That decline breaks the explanation everyone reaches for first. Human vanity did not fall by three quarters between 1972 and 2020. Whatever the gap is made of, a constant of human nature is not it.

Now the part that is actually used. The Bureau's by-group table is the source of nearly every published claim about unequal participation, and every line in it is a self-report.

```{r by-race}
```

```{r slope-static, eval=!knitr::is_html_output(), fig.height=4.4}
```

```{r slope-d3, results='asis', eval=knitr::is_html_output()}
```

**Figure 3. Reported turnout by race, five elections.** One line per group, as the Bureau publishes them: the share of citizens in each group who told the survey they voted. The spread between the highest and lowest group was `r pc(rg_sp(min(rg$year)))` points in `r min(rg$year)` and `r pc(rg_sp(max(rg$year)))` points in `r max(rg$year)`.

In `r Y2` reported turnout ran from `r pc(g(Y2, "pct_white_nh"))`% among White non-Hispanic citizens to `r pc(g(Y2, "pct_hispanic"))`% among Hispanic citizens. Everything established above applies to each of those numbers, and applies **unequally**. There is no reason to assume the over-report is the same size in every group, and if it is not, the gaps between the columns are wrong by an unknown amount in an unknown direction.

So the answer to the opening question depends on which instrument you trust. The survey says one thing about the size of the American electorate and the record says another, and the two disagree by millions of people in most years. Which one you reach for decides the answer, and only one of them can be checked.

## What this brief cannot tell you

**It cannot say why the gap exists.** At least three mechanisms live inside one difference, and the file separates none of them.

```{r mechanisms}
```

Consider the second row seriously. Survey response rates collapsed over exactly the period the gap shrank. If the people who still answer a government survey are disproportionately voters, the raw sample drifts toward voters on its own. The Bureau's weighting, which means counting some answers more than others so the sample matches the country, corrects for age, sex, race and region. It does not correct for how civic-minded somebody is, because nobody holds a population total for that.

**It cannot tell you whether validation studies measure lying or measure record linkage.** Ansolabehere and Hersh (2012) matched a large survey to a commercial voter file and concluded that self-reports are badly inflated. Berent, Krosnick and Lupia (2016) answered that a failed match looks exactly like a liar, because names change, addresses change and records get purged. Every audit needs its own instrument, so the data cannot arbitrate.

**It cannot give the by-group figures a margin of error.** The published table carries nothing saying how far off each percentage could be, and the Asian and Hispanic figures rest on the smallest subsamples in it.

**And it covers presidential years only.** Midterms are absent from this file, and there is every reason to expect the gap behaves differently in them.

## What you should have learned

- Self-reported turnout has exceeded the counted ballots in `r sum(cps$gap_people > 0)` of `r nrow(cps)` presidential elections, by `r mn(mean(cps$gap_people))` million people on average and by `r cnt(g(2016, "gap_people"))` in 2016.
- The excess fell from about `r pc(em[1])`% of the ballot count in the 1960s and 1970s to about `r pc(em[3])`% since 2000, with no correction announced by anybody, which rules out any explanation that is a constant of human nature.
- `r paste(cps$year[cps$gap_people < 0], collapse = ", ")` is the one election where the survey came in below the count, and it is also the highest-turnout election in the file.
- Both halves of this table are constructed: `reported_voters` is a percentage multiplied by a population, and `actual_votes` is a compilation of state canvasses rather than a federal record.
- The by-group figures carry every caveat above, unequally, so a published `r pc(g(Y2,"pct_white_nh") - g(Y2,"pct_hispanic"))`-point spread between two groups in `r Y2` is wrong by an unknown amount.

## Extensions

None of these is answered above, and every one can be answered from the table the Sources section hands over.

- `cps_turnout.csv` carries `cps_rate` and `actual_rate` side by side with `over_report_pp` between them. Sort by `over_report_pp` and read the top and bottom five years. Do the worst years share anything?
- The same file has `vap_thousands` and `citizen_vap`. Compute `actual_votes` divided by each of them for every year since 1980, and see how far apart the two turnout rates run.
- Compare `pct_white_nh`, `pct_black`, `pct_asian` and `pct_hispanic` across the years present. Which group's reported turnout moved most between 2016 and 2020, and which moved least?
- Take `reported_voters` minus `actual_votes` for one year and divide it by the state populations you can find. How many states' worth of people are in that difference?
- **Stretch.** The Bureau publishes the same supplement for midterm years. Download two of them, set the reported voters against the ballots counted, and see whether the gap is larger in a midterm than in a presidential year.
- **Stretch.** Several states publish a voter file recording whether each registrant voted. Take one state's file and compare its turnout to the survey's figure for that state, which is a join on a much worse key than a year.

## Sources

**Data**

- U.S. Census Bureau, Current Population Survey, *Voting and Registration
  Supplement*, Table A-1, November `r Y1`–`r Y2`.
  <https://www.census.gov/data/tables/time-series/demo/voting-and-registration/voting-historical-time-series.html>
- Counted ballots: national presidential vote totals, `r Y1`–`r Y2`, from Jason
  Timm, *PresElectionResults* (<https://github.com/jaytimm/PresElectionResults>),
  which compiles the per-election encyclopedia articles that rest on Dave Leip's
  *Atlas of U.S. Presidential Elections* and, beneath that, the official state
  canvasses.

**The data itself**

Every figure above is drawn from this table. It is plain CSV, it opens in a
spreadsheet, and it is yours to take further.

[`cps_turnout.csv`](data/derived/cps_turnout.csv)

**Academic literature**

- Ansolabehere, S. and Hersh, E. (2012). "Validation: What Big Data Reveal About
  Survey Misreporting and the Real Electorate." *Political Analysis* 20(4):
  437–459.
- Berent, M. K., Krosnick, J. A. and Lupia, A. (2016). "Measuring Voter
  Registration and Turnout in Surveys: Do Official Government Records Yield More
  Accurate Assessments?" *Public Opinion Quarterly* 80(3): 597–621.
- Groves, R. M. (2006). "Nonresponse Rates and Nonresponse Bias in Household
  Surveys." *Public Opinion Quarterly* 70(5): 646–675.
- Hur, A. and Achen, C. H. (2013). "Coding Voter Turnout Responses in the Current
  Population Survey." *Public Opinion Quarterly* 77(4): 985–993.
- McDonald, M. P. and Popkin, S. L. (2001). "The Myth of the Vanishing Voter."
  *American Political Science Review* 95(4): 963–974.

```{r ai-prompt, echo=FALSE, results="asis"}
```
