---
title: "Mapping Votes Against Land Area"
type: brief
subtitle: "One election, three thousand counties, and the difference between coloring land and counting people"
author: "Democracy's Data"
date: "`r format(max(file.mtime(list.files(pattern = '[.]Rmd$|[.]R$'))), 'Last updated %e %B %Y')`"
toc_depth: 3
knit: (function(input, ...) { source(file.path(dirname(input), "../../../../../_syllabus-template/render-brief.R")); render_brief(input) })
---

```{r knitr-setup, include=FALSE}
source("../../../../../_syllabus-template/read-chunks.R")
dd_read_chunks("mapping-code.R")
```

```{r setup, include=FALSE}
```

San Bernardino County, California covers `r nn(big$aland_km2)` square kilometers. It is larger than `r n_small` of the `r length(starea)` jurisdictions this brief maps, larger than Costa Rica, and on any map of the United States it is an unmissable slab of the desert southwest. In November 2024 it cast `r nn(big$two)` votes for the two major presidential candidates. New York County, New York, the island of Manhattan, covers `r pc(sml$aland_km2, 1)` square kilometers, `r nn(g("pair_area_ratio"))` times less, and cast `r nn(sml$two)` votes.

Now recall the map you saw on election night: nearly the whole country red, with blue at the edges and in specks. It is probably the single most reproduced graphic in American politics, and it is the reason a great many Americans believe the 2024 election was a landslide. The two candidates finished `r pc(g("gop_share_two"), 1)` to `r pc(g("dem_share_two"), 1)` in the two-party vote, a gap of `r pc(g("national_margin"), 1)` points. So the question this brief asks is the one the map plants in every viewer: **when a county map is almost all one color, what exactly has been measured?**

## The test

The map needs three things in the same rows: votes, boundaries, and land. The votes are county-level 2024 presidential returns from Tony McGovern's compilation (<https://github.com/tonmcg/US_County_Level_Election_Results_08-24>). **The federal government does not publish county-level presidential returns**, so every national county map rests on somebody's private reconciliation of fifty-one state publications, and this compiler's own README says the numbers are not authoritative. Why that is the ordinary condition of American election data is the [returns-source chapter's](../returns-source/returns-source-brief.html) subject. The boundaries and the land areas are federal: the Census Bureau's 2024 cartographic boundary file, whose `ALAND` field is the Bureau's own measurement in square meters, stored beside the outline rather than derived from it.

Counties are the right unit because they are the only one that works. States are too few and too internally mixed to show that color tracks acreage; precincts are not published nationally, and no federal file measures their land. Land, votes and boundaries meet at the county or nowhere. And because the [county-returns brief](../county-returns/county-returns-brief.html) has since assembled the same election from the fifty-one chief election offices, the compilation can now be measured against the certified record rather than merely doubted.

## The data

The build joins the returns to the boundaries on the five-digit county FIPS code, read as **text, never as a number**. About a seventh of the codes begin with a zero, and a numeric read turns Autauga County, Alabama into a county in Georgia. Three places refuse the join, and each was settled by a documented decision:

```{r units}
```

The District is the dangerous one, because it fails silently. `11001` is present in both files, meaning Ward 1 in the returns and the whole District in the geography, so the join *succeeds* and quietly discards `r nn(g("dc_ward_error"))` votes. The repair, summing the `r nn(g("dc_wards"))` wards back into one District, is one line. Knowing the line is needed is the whole job. Alaska cannot be repaired that way, so it is dropped whole, with Hawaii, which is outside the frame:

```{r excluded}
```

Every excluded unit is named individually, `r nn(nrow(ex))` rows carrying `r nn(sum(ex$votes))` votes, so the size of the hole sits on the page rather than in a footnote. What survives is `r nn(nrow(d))` rows in `counties.csv`, one per county of the `r nn(NST)` contiguous states plus the District. Here is one:

```{r clean-mapping}
```

Each row carries the identifiers, the two vote columns and their total, the `winner`, the `margin`, the Bureau's `aland_km2`, and `density`, votes per square kilometer. A `pop` column from the Bureau's 2020 population file covers `r nn(g("pop_units"))` of the `r nn(nrow(d))` units, because that file predates Connecticut's `r nn(g("ct_units"))` planning regions. Every headline number below is land, counties or votes, which exist for all of them. The outlines were thinned to `r pc(g("points_pct"), 1)`% of their `r nn(g("points_before"))` corner points for drawing; every share quoted below was recomputed from the thinned outlines and moved by `r max(abs(pa$difference))`, because no headline number ever reads a boundary.

One more check, because the vote column is a press reconstruction. Put the mapped units against the certified assembly in the [county-returns brief](../county-returns/county-returns-brief.html): `r nn(n_chk)` of `r nn(nrow(d))` can be compared, and `r nn(n_diff)` disagree on a major-party count. One of those is a boundary dispute about Kansas City rather than an error. The ratio this brief turns on moves from `r pc(ink_cmp, 2)` to `r pc(ink_off, 2)` depending on which file you believe, which is to say it does not move. The doubt is measured now, not merely declared.

> **Before you read on, commit to a number.** Trump carried
> `r nn(g("n_R"))` of the `r nn(nrow(d))` counties on the map below --
> `r pc(g("pct_counties_R"), 1)`% of them, and
> `r pc(g("pct_land_R"), 1)`% of the land. **What share of the votes on this
> map were cast inside those `r nn(g("n_R"))` counties?** Not what share Trump
> won: what share of all the two-party ballots on the map were cast anywhere
> inside a county he carried. Write down a percentage before you read on.

## What it says about democracy

```{r fig1-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig1-static, eval=!knitr::is_html_output(), fig.height=4.7}
```

**Figure 1. The 2024 presidential result, one color per county.** A choropleth, a map that fills each area with a color, fits the question *who carried the ground you are standing on*, which is the question this figure answers and the only one. All `r nn(nrow(d))` counties are filled with the color of the candidate who carried them, on an equal-area projection, so a square centimeter of paper is the same number of square kilometers everywhere on the page. Trump carried `r nn(g("n_R"))` counties, `r pc(g("pct_counties_R"), 1)`%, covering `r pc(g("pct_land_R"), 1)`% of the land; Harris carried `r nn(g("n_D"))`, covering `r pc(g("pct_land_D"), 1)`%.

**Every county in Figure 1 is the correct color.** No boundary is wrong, no result is misreported, and no projection trick has been played. The mechanism is not subtle once it is stated. A map that fills areas with color encodes **area**, and the eye adds it up whether or not you asked it to. Area is a property of the ground, not of the people standing on it. Here is the same election as a ledger instead of a picture:

```{r ledger}
```

Read the second and third rows together. Trump's counties are `r pc(g("pct_land_R"), 1)`% of the land and hold `r pc(g("pct_pop_R"), 1)`% of the people. The map is not misdrawn; **it is drawing the land row while the reader is reading the people row.** Put it as a rate: every vote Trump won sits on an average of `r pc(g("km2_per_vote_R"), 3)` square kilometers of map, and every vote Harris won sits on `r pc(g("km2_per_vote_D"), 3)`. **A Trump vote is drawn `r pc(g("ink_ratio"), 1)` times larger than a Harris vote.** No chart anywhere may distort the size of its marks by `r pc(g("ink_ratio"), 1)` to 1 without saying so in the legend, and this one never says so.

The answer to the prompt above follows from where people live. Sorting the counties from emptiest to fullest and adding as you go:

```{r conc}
```

The emptier half of the counties hold `r pc(g("land_at_half_counties"), 1)`% of the country's land and `r pc(g("votes_at_half_counties"), 1)`% of its votes. `r nn(g("counties_for_half"))` counties cast half of every ballot on this map, on `r pc(g("land_for_half"), 1)`% of its surface. So the answer is: `r pc(g("pct_counties_R"), 1)`% of the counties, `r pc(g("pct_land_R"), 1)`% of the land — and **`r pc(g("pct_votes_R_counties"), 1)`% of the votes.** Most readers write down something in the seventies or eighties, and the gap between that guess and `r pc(g("pct_votes_R_counties"), 1)` is this brief's entire subject. The `r nn(g("n_D"))` blue counties are `r pc(g("pct_votes_D_counties"), 1)`% of the electorate sitting on `r pc(g("pct_land_D"), 1)`% of the ground.

Now draw the same table so that the ink tells the truth about votes. Each county becomes a circle whose **area is proportional to the votes cast there**; circles keep their positions as nearly as they can and are pushed apart until they stop overlapping. This is a Dorling cartogram — a cartogram is a map that sizes places by a quantity instead of by their ground. Not one vote has been added, removed or recounted.

```{r fig4-d3, results='asis', eval=knitr::is_html_output()}
```

```{r fig4-static, eval=!knitr::is_html_output(), fig.height=4.7}
```

**Figure 2. The same `r nn(nrow(d))` counties, with area proportional to votes.** The radius is one constant times the square root of the county's two-party vote, so one square unit of circle is `r nn(g("dorling_votes_per_unit2"))` votes anywhere on the page. The largest circle is `r gs("dorling_biggest")` County, `r gs("dorling_biggest_state")`, at `r nn(max(d$two))` votes. Red now covers `r pc(g("gop_share_two"), 1)`% of the ink because Trump won `r pc(g("gop_share_two"), 1)`% of the two-party vote. **You cannot look at this map and think the election was a landslide, because the ink is reporting the result rather than the terrain.**

Figure 2 is not a better map than Figure 1. It is a map of a different quantity, and the quantity happens to be the one an election is decided by. The price is real and measurable: the nudging moved the middle county `r pc(g("dorling_shift_median_km"), 1)` km from its true position and the furthest `r pc(g("dorling_shift_max_km"), 0)` km, and it stopped with `r nn(g("dorling_overlaps"))` of `r nn(g("dorling_pairs"))` tested pairs still touching. Shape, adjacency and the ability to find your own county are gone. And the eye judges area badly, which is a century-old finding in graphical perception. Figure 2 is defensible only because the alternative is an encoding that is off by a factor of `r pc(g("ink_ratio"), 1)`, not because area is a good channel.

**Neither map is the repair for the other.** A cartogram does not remove distortion; it chooses a different one. All the drawings are true, and none of them is *the* map. Which one gets published is a decision about what the reader will believe, and in most newsrooms it is made by whoever happens to own the template.

## What this brief cannot tell you

**It cannot tell you the choropleth is a lie.** Figure 1 answers *who carried this ground*, and for county politics or the geography of a party's base that is exactly the right question. The failure is in reading it as an answer to *how did the country vote*, which is a different question with a different bottom number underneath it.

**It cannot certify any particular county.** The vote column is a press reconstruction; `r nn(n_diff)` of the `r nn(n_chk)` checkable counties disagree with the state canvasses on a major-party count. The aggregate survives, because a county would have to be wrong by the size of a city to move it. A sentence about one county should come from that state's certified return instead.

**It cannot say counties are the right unit.** They vary by a factor of `r nn(max(d$two) / min(d$two))` in electorate, they were drawn for administrative reasons, and the same election added up over different boundaries would give a different picture. That is the modifiable areal unit problem, the answer depending on how you cut the country up, and the [areal-units brief](../../01-census-bureau/areal-units/areal-units-brief.html) takes it apart properly.

**And it is not how the president is chosen.** Electors are allocated by state, the level the [electoral-map brief](../electoral-map/electoral-map-brief.html) works at. The fill here is also binary, painting a `r pc(min(abs(d$margin)), 2)`-point county the same shade as a `r pc(max(abs(d$margin)), 0)`-point one. Alaska and Hawaii, `r pc(100 * (g("ak_votes") + g("hi_votes")) / g("returns_votes_all"), 1)`% of the national ballots, are absent from every figure and every number, so this is a map of the contiguous states. And one election is one election: the concentration measured above is a fact about November 2024, not a law of American geography.

## What you should have learned

- A county choropleth of the 2024 election is accurate in every county and still misleads, because it encodes land: Trump's counties cover `r pc(g("pct_land_R"), 1)`% of the land and cast `r pc(g("pct_votes_R_counties"), 1)`% of the votes.
- On Figure 1, one vote for the winner of empty counties gets about `r pc(g("ink_ratio"), 1)` times the ink of a vote for the winner of full ones — a distortion no legend admits to.
- The electorate is concentrated: `r nn(g("counties_for_half"))` of `r nn(nrow(d))` counties cast half the ballots, on `r pc(g("land_for_half"), 1)`% of the land.
- An honest encoding costs something else: sizing counties by votes reports the `r pc(g("gop_share_two"), 1)`&ndash;`r pc(g("dem_share_two"), 1)` result in the ink, and destroys shape, adjacency and findability to do it.
- There is no federal county-level presidential file; the map's votes are a volunteer compilation, and checking it against the certified state canvasses is what turns that doubt into a measured, small number.

## Extensions

- `lorenz.csv` accumulates `cum_land` and `cum_votes` together, county by county. Find the point where half the land has been covered. What share of the votes is in it?
- `counties.csv` has `aland_km2`, `votes_tot` and `density`. Sort by `density` and compare the extremes. How many of the emptiest counties add up to one Manhattan?
- `excluded.csv` lists the units that could not be mapped, each with a `reason` and its `votes`. Add them up. How many real voters vanish from a map of the country?
- `parity.csv` compares, in `difference`, the headline numbers before and after the outlines were thinned. Nothing moved. Work out what a difference there would have meant.
- **Stretch.** The compilation carries every election back to 2008. Fetch 2020, rebuild the ledger, and see whether the land-versus-votes gap is growing.
- **Stretch.** Redraw Figure 1 with a diverging margin scale instead of the binary fill, and decide which of this brief's sentences would need rewriting.

## Sources

**Data**

- County-level 2024 presidential returns: Tony McGovern, *US County Level
  Election Results 2008-2024*,
  <https://github.com/tonmcg/US_County_Level_Election_Results_08-24>, held in
  this book's `pres2024_counties.csv`
  (`r nn(g("returns_rows"))` rows). Compiled from Fox News, Politico and New
  York Times reporting; **the compilation's own README states that the figures
  are not authoritative.** There is no federal publisher of county-level
  presidential returns. This is the file every figure in this brief is drawn
  from.
- Official county-level presidential returns, as published by the
  `r nn(length(unique(of$state_name)))` state chief election offices:
  `pres2024_counties_official.csv` and its 2020 companion, `r nn(nrow(of))`
  rows each, assembled by the
  [county-returns brief](../county-returns/county-returns-brief.html) with
  one row per jurisdiction recording the URL, the format, and whether the
  state calls the figures certified. Used here as a ruler and not as a source
  of paint: no figure is drawn from it.
- Associated Press, certified state-level presidential totals for 2024, held as
  `crosscheck_ap_2024.csv` (`r nn(nrow(ap))` rows). Used only to score the two
  county files against a third that neither was built from.
- U.S. Census Bureau, 2024 cartographic boundary file, counties, 1:5,000,000:
  `r gs("geo_url")` — retrieved `r gs("fetch_date")`,
  `r nn(g("geo_bytes"))` bytes, `r nn(g("geo_features"))` features. Supplies
  both the outlines and the `ALAND` land areas.
- U.S. Census Bureau, 2020 centers of population by county:
  `r gs("cen_url")` — retrieved `r gs("fetch_date")`,
  `r nn(g("cen_rows"))` rows. Supplies the population column only.

**The data itself**

Every figure above is drawn from these tables. They are plain CSV, they
open in a spreadsheet, and they are yours to take further.

[`counties.csv`](data/derived/counties.csv), [`county_rings.csv`](data/derived/county_rings.csv), [`dorling.csv`](data/derived/dorling.csv), [`excluded.csv`](data/derived/excluded.csv), [`lorenz.csv`](data/derived/lorenz.csv), [`parity.csv`](data/derived/parity.csv), [`state_rings.csv`](data/derived/state_rings.csv)

**Method**

- Dorling, D. (1996). *Area Cartograms: Their Use and Creation.* Concepts and
  Techniques in Modern Geography 59. University of East Anglia. — the circle
  cartogram of Figure 2.
- Monmonier, M. (1991). *How to Lie with Maps.* University of Chicago Press.
- Cleveland, W. S. and McGill, R. (1984). "Graphical Perception: Theory,
  Experimentation, and Application to the Development of Graphical Methods."
  *Journal of the American Statistical Association* 79(387): 531-554. — the
  finding that area is a poor channel, cited against Figure 2 above.

```{r ai-prompt, echo=FALSE, results="asis"}
```
