---
title: "Lab 14b: Grading the Guess"
subtitle: "84-355 Democracy's Data | BISG against the truth, Houston County, Georgia"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", fig.width = 7.2, fig.height = 4.4)
options(scipen = 999)
pc <- function(x, k = 1) formatC(x, format = "f", digits = k)
nn <- function(x) format(x, big.mark = ",", trim = TRUE)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# The question

This morning you learned how BISG works: take a surname, take a place, combine
them, and out comes a probability that a voter is Black, or white, or Hispanic.
Courts use it. Expert witnesses use it. The Voting Rights Act cases you read for
Week 13 rest on it.

And normally you can never check it, because the whole point is that you do not
know the answer. That is what makes this an *inference* rather than a
measurement.

**Today we check it.** Georgia is one of a handful of states that records race on
the voter registration form — collected for Voting Rights Act compliance, not for
anybody's research. So for Houston County we have almost fifty thousand people
whose race the method is trying to guess and whose race we already know.

You are about to find out how good the guess is. It is not what the overall
number will tell you.

# Part 1 — The data

```{r load}
v <- read.csv("data/houston_voters.csv", colClasses = c(GEOID20 = "character"))
g <- read.csv("data/houston_blocks.csv",  colClasses = c(GEOID20 = "character"))
sn <- read.csv("../surnames/data/census_surnames.csv", stringsAsFactors = FALSE)

str(v)
sort(table(v$race), decreasing = TRUE)
```

Three columns: a surname, the race the voter reported, and the census block they
live in. No names, no addresses — the model only ever gets the surname and the
block, so that is all the file carries.

```{r blocks}
nrow(g)
colSums(g[, -1])
```

The county is `r nn(sum(g$white))` white residents and `r nn(sum(g$black))`
Black. That second file is the **geographic prior**: for each of
`r nn(nrow(g))` blocks, who lives there.

```{r setup-matrices}
R  <- c("white", "black", "hispanic", "asian", "aian")
sc <- c(white = "pctwhite", black = "pctblack", hispanic = "pcthispanic",
        asian = "pctapi",   aian  = "pctaian")

i <- match(v$surname, sn$name)
cat("surnames found in the Census list:", sum(!is.na(i)),
    sprintf("(%.1f%%)\n", 100 * mean(!is.na(i))))

v <- v[!is.na(i), ]
S <- as.matrix(sn[i[!is.na(i)], sc]); S[is.na(S)] <- 0
keep <- rowSums(S) > 0
v <- v[keep, ]; S <- S[keep, , drop = FALSE]
S <- S / rowSums(S)          # P(race | surname)
```

# Part 2 — The surname alone

```{r surname-only}
sur <- R[max.col(S, ties.method = "first")]

cat("SURNAME ONLY\n")
cat(sprintf("  overall accuracy: %.1f%%\n\n", 100 * mean(sur == v$race)))
for (r in R)
  cat(sprintf("  %-9s n = %6d   correct %5.1f%%\n",
              r, sum(v$race == r), 100 * mean(sur[v$race == r] == r)))
```

**Stop and look at that before going on.** Overall accuracy is
`r pc(100 * mean(sur == v$race))`%, which sounds like a method that works. Now
read down the column.

It gets white voters right `r pc(100 * mean(sur[v$race == "white"] == "white"))`%
of the time and Black voters right
**`r pc(100 * mean(sur[v$race == "black"] == "black"))`%** of the time. The
overall figure is not a summary of how the method performs — it is a summary of
how many white people live in Houston County. A method that simply guessed
"white" for everybody would score `r pc(100 * mean(v$race == "white"))`%.

# Part 3 — Adding the place

Surname alone is not BISG. The method combines the surname with where the person
lives: if a name is ambiguous but the block is 90% Black, the block should carry
the inference.

```{r bisg}
P <- as.matrix(g[, R]); rownames(P) <- g$GEOID20
M <- sweep(P, 2, colSums(P), "/")[v$GEOID20, , drop = FALSE]   # P(block | race)

# Two places the method has no answer, and what we do about it.
empty <- rowSums(M) == 0                 # block with no recorded population
M[empty, ] <- matrix(colSums(P) / sum(P), sum(empty), 5, byrow = TRUE)

post <- S * M                            # P(race | surname) x P(block | race)
degenerate <- rowSums(post) == 0         # surname and block flatly contradict
post[degenerate, ] <- S[degenerate, ]

post <- post / rowSums(post)
bis  <- R[max.col(post, ties.method = "first")]

cat("voters in a block with no census population:", sum(empty), "\n")
cat("voters where surname and block cancel out:  ", sum(degenerate), "\n")

cat("BISG: SURNAME + BLOCK\n")
cat(sprintf("  overall accuracy: %.1f%%\n\n", 100 * mean(bis == v$race)))
for (r in R)
  cat(sprintf("  %-9s n = %6d   correct %5.1f%%\n",
              r, sum(v$race == r), 100 * mean(bis[v$race == r] == r)))
```

Overall accuracy rises from `r pc(100 * mean(sur == v$race))`% to
`r pc(100 * mean(bis == v$race))`%. That is the number a methods section would
report.

**Notice the two lines of housekeeping in that chunk**, because they are not
housekeeping. Some `r nn(sum(empty))` voters live in a block the census records
as having nobody in it — new construction, mostly — so the geographic term is
empty and we fall back to the county. And for **`r nn(sum(degenerate))` voters
the method simply has no answer**: their surname is 100% one race and their
block has zero residents of that race, so the multiplication gives zero for
every category. BISG is undefined for them.

Without those two lines R would not have complained. `max.col` picks the first
column when every value is zero, so those `r nn(sum(empty | degenerate))` people
would have been silently predicted **white**, and
`r nn(sum(v$race[empty | degenerate] == "black"))` of them are Black. A quiet
default is worse than an error.

# Part 4 — Who it helps, and who it costs

```{r change}
chg <- sapply(R, function(r)
  100 * (mean(bis[v$race == r] == r) - mean(sur[v$race == r] == r)))
round(chg, 1)
```

```{r change-plot}
barplot(chg, col = ifelse(chg > 0, "#2c7fb8", "#C41230"),
        ylab = "Change in accuracy from adding block (points)",
        ylim = c(-12, 26))
abline(h = 0)
```

**Adding geography is not a free improvement.** It is a trade:

- **Black voters: `r sprintf("%+.1f", chg[["black"]])` points.** From
  `r pc(100 * mean(sur[v$race == "black"] == "black"))`% to
  `r pc(100 * mean(bis[v$race == "black"] == "black"))`%. This is the method
  working, and it works because Black residents of Houston County are
  geographically concentrated, so the block genuinely carries information.
- **Asian voters: `r sprintf("%+.1f", chg[["asian"]])` points. Hispanic:
  `r sprintf("%+.1f", chg[["hispanic"]])`.** The block makes them *worse*.
  These populations are spread thinly across the county, so no block is
  distinctively Asian or Hispanic. The geographic term drags the estimate toward
  the local majority and overrides a surname signal that was already decent.
- **White voters: `r sprintf("%+.1f", chg[["white"]])` points**, for the same
  reason in reverse.

The overall `r sprintf("%+.1f", 100 * (mean(bis == v$race) - mean(sur == v$race)))`
hides all of this. One group gained twenty-four points, two groups lost.

**And the number that should stay with you:** even with everything the method
has, BISG identifies Black voters correctly
**`r pc(100 * mean(bis[v$race == "black"] == "black"))`% of the time**. It is
wrong about them more often than it is right — in a county that is
`r pc(100 * sum(g$black) / sum(g[, R]))`% Black, on a question that Voting
Rights Act litigation turns on.

# Part 5 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Weight the classes.** `max.col` picks the largest posterior, which
favours big groups. Instead of the winner, look at the *probability* BISG assigns
to the truth: `mean(post[cbind(1:nrow(post), match(v$race, R))] / rowSums(post))`.
Compare that across groups. Is the method badly calibrated, or just badly
decided?

**Option B — A different geography.** Replace the block prior with a county-level
one — set every row of `M` to `colSums(P)/sum(P)` — so geography carries no
information. What does accuracy fall to, and how close is that to Part 2? What
does the answer say about how much of BISG is really the surname?

**Option C — Only the ambiguous names.** Restrict to voters whose surname gives
no group above 60% (`keep <- apply(S, 1, max) < 0.6`) and re-run both methods.
These are the people the method is supposed to be *for*. Does it do better or
worse on them?

**Option D — Your own question.** `r nn(nrow(v))` voters,
`r nn(length(unique(v$GEOID20)))` blocks.

# What this lab cannot tell you

- **One county.** Houston County's population is
  `r pc(100 * sum(g$white) / sum(g[, R]))`% white and
  `r pc(100 * sum(g$black) / sum(g[, R]))`% Black, and moderately
  segregated. Somewhere more integrated, the block would carry less; somewhere
  more segregated, more. Do not generalise the numbers, only the shape.
- **The race on the form is also a measurement.** A voter picks one box. People
  of two races, and Hispanic people of any race, are being flattened by the
  registration form exactly as Week 12 showed the census flattening them.
- **"Any-part Black."** We counted Black residents as Black alone *or in
  combination*, which is Voting Rights Act practice. The stricter definition —
  Black alone, not Hispanic — gives 51,992 for this county instead of
  `r nn(sum(g$black))`, a difference of 4,528 people. Every number above would
  shift.
- **The surname list is from 2010.** The county is not.

# On the discussion board

By **Friday, 11:59 p.m.** No length requirement; a sentence or two is enough, so
long as it shows you engaged with what we did. Any *one* of the usual four.

Three that would make good posts:

- A method that is right `r pc(100 * mean(bis == v$race))`% of the time overall
  and `r pc(100 * mean(bis[v$race == "black"] == "black"))`% of the time for
  Black voters is used in federal voting-rights cases. Is that acceptable? Say
  what would change your mind.
- Adding geography helped one group and hurt two. Would you use it?
- American Indian voters — `r nn(sum(v$race == "aian"))` of them — are identified
  correctly `r pc(100 * mean(sur[v$race == "aian"] == "aian"))`% of the time by
  the surname alone and `r pc(100 * mean(bis[v$race == "aian"] == "aian"))`% of
  the time by BISG. What happens to a group the method cannot see?

***

### Where the data came from

Houston County, Georgia voter registration with self-reported race, and the
geocoded block assignment, from working data for the 2026 Houston County
redistricting matter. Block population by race is 2020 Census PL 94-171 / DHC
tables. Surname probabilities are the 2010 Census surname list you used this
morning.

**Rows where race had been filled in by BISG rather than reported by the voter
were dropped** before anything else. Scoring BISG against BISG would have
produced an excellent and completely meaningless result. The check is in
`data/build-data.R`.
