---
title: "Key: The District That Was Not Drawn"
subtitle: "84-355 Democracy's Data | Vote dilution, Houston County, Georgia | Instructor copy — do not distribute"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
options(scipen = 999)

b  <- read.csv("data/blocks.csv", colClasses = c(GEOID20 = "character"))
a  <- read.csv("data/adjacency.csv")
d  <- read.csv("data/districts.csv",  stringsAsFactors = FALSE)
tr <- read.csv("data/tradeoff.csv",   stringsAsFactors = FALSE)
sw <- read.csv("data/seed_sweep.csv", stringsAsFactors = FALSE)
cu <- read.csv("data/ga_county_units.csv", colClasses = c(fips = "character"))
sg <- read.csv("data/segregation.csv", stringsAsFactors = FALSE)

# grow() and score() are the lab's own, copied verbatim, plus a num/den argument
# so Option C can be run without editing the body.
nb <- split(c(a$a, a$b), c(a$b, a$a))
NB <- vector("list", nrow(b)); NB[as.integer(names(nb))] <- nb
grow <- function(seed, target, reach = Inf, num = "vap_black", den = "vap") {
  inn <- logical(nrow(b)); inn[seed] <- TRUE
  P <- b$pop[seed]; N <- b[[num]][seed]; D <- b[[den]][seed]
  cx <- b$x_km[seed]; cy <- b$y_km[seed]; fr <- NB[[seed]]
  while (P < target && length(fr)) {
    if (is.finite(reach) && reach < length(fr)) {
      dd <- (b$x_km[fr] - cx)^2 + (b$y_km[fr] - cy)^2
      cand <- fr[order(dd)[seq_len(reach)]]
    } else cand <- fr
    pick <- cand[which.max((N + b[[num]][cand]) / pmax(D + b[[den]][cand], 1))]
    inn[pick] <- TRUE
    P <- P + b$pop[pick]; N <- N + b[[num]][pick]; D <- D + b[[den]][pick]
    w <- b$pop[inn] + 1
    cx <- sum(b$x_km[inn] * w) / sum(w); cy <- sum(b$y_km[inn] * w) / sum(w)
    fr <- setdiff(unique(c(fr, NB[[pick]])), which(inn))
  }
  which(inn)
}
score <- function(i) c(population = 100 * sum(b$pop_black[i]) / sum(b$pop[i]),
                       adults     = 100 * sum(b$vap_black[i]) / sum(b$vap[i]),
                       registered = 100 * sum(b$reg_black[i]) / sum(b$reg[i]))

# Rounding, so that a live re-run and the stored tables never disagree in the
# last digit. build-data.R writes its percentages rounded to three decimals, so
# snap to three first; then round half-UP, because R's round() is half-to-even
# and would print 41.26 where every other document in this folder prints 41.27.
r2 <- function(v) floor(100 * round(v, 3) + 0.5) / 100
pc <- function(x, k = 2)
  formatC(floor(10^k * round(x, 3) + 0.5) / 10^k, format = "f", digits = k)
nn <- function(x) format(round(x), big.mark = ",", trim = TRUE)

POP <- sum(b$pop); VAP <- sum(b$vap); REG <- sum(b$reg)
SH_VAP <- 100 * sum(b$vap_black) / VAP; SH_REG <- 100 * sum(b$reg_black) / REG
SH_POP <- 100 * sum(b$pop_black) / POP
ag <- aggregate(pop ~ boe5, b, sum); IDEAL <- mean(ag$pop); IDEAL4 <- POP / 4
DEV <- 100 * (ag$pop - IDEAL) / IDEAL; DEV_RANGE <- max(DEV) - min(DEV)
TIMES <- sum(b$vap_black) / (VAP / 5 / 2)
DISS <- sg$value[1]
CU_RAT <- max(cu$people_per_unit) / min(cu$people_per_unit)
o <- cu[order(cu$people_per_unit), ]
K <- which(cumsum(o$units) >= floor(sum(cu$units) / 2) + 1)[1]
K_SHR <- 100 * sum(o$pop[seq_len(K)]) / sum(cu$pop)

t5 <- tr[tr$districts == 5, ]; t4 <- tr[tr$districts == 4, ]
R1  <- t5[t5$reach == 1, ]; RI <- t5[!is.finite(t5$reach), ]
CROSS <- t5[t5$vap_black_pct > 50, ][1, ]
WOB <- t5[which.min(t5$vap_black_pct), ]
f4 <- t4[is.finite(t4$reach) & t4$reach <= 100, ]
E4 <- d[d$plan == "Enacted BOE district 4", ]
BEAT   <- sum(t5$vap_black_pct >= E4$vap_black_pct & t5$pp >= E4$pp)
BLACKER <- sum(t5$vap_black_pct >= E4$vap_black_pct)
ROUNDER <- sum(t5$pp >= E4$pp)
CEIL <- d[d$plan == "Ceiling, 5 districts", ]

seed <- which.max(b$vap_black); ord <- order(-b$vap_black)

# Option B: the same 306 seeds the build script swept, re-run at reach = 1.
HI  <- which(b$vap >= 20 & b$vap_black / pmax(b$vap, 1) > 0.5)
R1SW <- vapply(HI, function(s) score(grow(s, IDEAL, 1))[2], numeric(1))
R1SW_SPREAD <- max(R1SW) - min(R1SW)
SW_SPREAD   <- max(sw$d5_vap_black_pct) - min(sw$d5_vap_black_pct)

# Option C: the two denominators, at both ends of the dial.
CA1 <- grow(seed, IDEAL, 1);   CR1 <- grow(seed, IDEAL, 1,   "reg_black", "reg")
CAI <- grow(seed, IDEAL, Inf); CRI <- grow(seed, IDEAL, Inf, "reg_black", "reg")
JAC <- 100 * length(intersect(CAI, CRI)) / length(union(CAI, CRI))

# Option D
D41 <- score(grow(seed, IDEAL4, 1)); D4I <- score(grow(seed, IDEAL4, Inf))

# Option A: the value most likely to be tried, computed once and cited twice.
A150 <- score(grow(seed, IDEAL, 150))[2]
R100 <- t5$vap_black_pct[t5$reach == 100]
```

All code is given to students. Grade the interpretation.

**Structure.** The brief (`vote-dilution-brief.Rmd`) and the lab
(`vote-dilution.Rmd`) run the same sequence — the brief in prose with D3 figures,
the lab with the code exposed. **Students should read the brief first.** The
brief numbers its Steps 1–10; the lab numbers the same material Parts 1–10. They
correspond one to one.

**This is the lab where a confident conclusion is produced, held for two
sections, and then demolished by a fact that was on the table the whole time.**
The demolition is the point. Do not let Part 7 arrive as a footnote.

**The law under it moved on 29 April 2026**, and unlike the companion RPV lab,
here the change lands on the *evidence-production* step rather than the
estimation step. *Louisiana v. Callais*, 608 U.S. \_\_\_ (2026), left *Gingles*
standing and "update[d]" it. Precondition one now carries two conditions: a
plaintiff "cannot use race as a districting criterion" in an illustrative map,
and the map "must meet all the State's legitimate districting objectives,
including traditional districting criteria and the State's specified political
goals," achieving them "just as well" as the State's own map (slip op., at 29).
**The lab's dial is a machine for violating the first condition, deliberately,
and it cannot even be evaluated against the second.** Both facts are marked in
the lab and the brief. See *Callais and what it does to this lab*, below.

Every number below was computed from the built files in `data/`.

***

# The session in one sentence

A claim about a district that does not exist can only be settled by building
one — so the evidence is manufactured by the party that needs it, every choice
the builder makes moves the answer, and a search that fails proves nothing at
all.

***

# The findings

```{r headline}
c(blocks            = nrow(b),
  population        = POP,
  black_adults_pct  = r2(SH_VAP),
  black_reg_pct     = r2(SH_REG),
  deviation_spread  = r2(DEV_RANGE),
  large_enough_by   = r2(TIMES),
  dissimilarity     = DISS,
  reach1_adults_pct = r2(R1$vap_black_pct),
  reachInf_adults   = r2(RI$vap_black_pct),
  enacted_d4_adults = r2(E4$vap_black_pct),
  enacted_d4_pp     = round(E4$pp, 3),
  algo_beating_d4   = BEAT)
```

- **The enacted plan spans `r pc(DEV_RANGE)` points against a 10-point
  allowance.** Malapportionment is over in one line. That is the setup.
- **"Sufficiently large" is satisfied `r pc(TIMES)` times over** —
  `r nn(sum(b$vap_black))` Black adults where a bare majority of one district
  needs `r nn(VAP / 5 / 2)`.
- **The dial runs from `r pc(R1$vap_black_pct)`% Black adults at Polsby-Popper
  `r pc(R1$pp, 3)` to `r pc(RI$vap_black_pct)`% at `r pc(RI$pp, 3)`.** Buying
  `r pc(RI$vap_black_pct - R1$vap_black_pct, 1)` points of Black share costs a
  factor of `r pc(R1$pp / RI$pp, 0)` in compactness.
- **Enacted District 4 is `r pc(E4$vap_black_pct)`% Black among adults at
  Polsby-Popper `r pc(E4$pp, 3)`.** `r BLACKER` of `r nrow(t5)` algorithm
  districts are more Black; `r ROUNDER` are more compact; **`r BEAT` are both.**
- **Where you start barely matters — until race stops driving.** At unlimited
  reach, `r nrow(sw)` starting blocks span `r pc(SW_SPREAD, 2)` points. At
  `reach = 1` the same `r nrow(sw)` starting blocks span
  **`r pc(R1SW_SPREAD, 2)` points.** This is new to the key and it is the best
  thing in it — see Option B.

***

# The three things to get said

**1. Failing to find something is not evidence that it is not there.** Part 6
produces an honest, careful, quantified conclusion — *a majority-Black district
here requires abandoning compactness* — and Part 7 exhibits a district that is
both more Black and more compact than anything the search produced, and that has
been enacted law for years. **Make somebody articulate the asymmetry**: exhibiting
a map settles the question in the affirmative permanently; failing to find one
settles nothing, because no search of this kind is ever exhaustive. Precondition
one is provable and not disprovable, and that structural fact is what the fight
over "what counts as a demonstration map" is really about.

**2. The choice that moves the answer is the one that looks like a setting.**
Part 9 is deliberately anticlimactic: at the dial's far end the seed does almost
nothing (`r pc(SW_SPREAD, 2)` points across `r nrow(sw)` restarts) and the dial
does everything. `reach` is not a statistical parameter with a correct value — it
encodes *what you think a district is allowed to look like*, which is a legal
judgment wearing the clothes of a tuning knob. **The lesson is not "analyst
choices matter." It is that they matter unequally, and the important one was
disguised.**

Say the scope out loud, because Option B breaks it: *at unlimited reach* the seed
is nearly irrelevant, and at `reach = 1` it is worth `r pc(R1SW_SPREAD, 2)`
points. **Which knob matters is itself contingent on where the other knob is
set**, so "we checked robustness to the starting block" is a claim about one
configuration and not about the method.

**3. "Majority" has three denominators and they disagree.** Part 8 puts up
`r nrow(f4)` four-district configurations in which **every single one is
majority-Black among registered voters and not one is majority-Black among
voting-age population.** Same blocks, same lines, same people. *Bartlett v.
Strickland* says a real numerical majority is required and does not say of what.
**Ask the room to pick before revealing that both answers are defensible.**

**The line for the board:** the map that does not exist is the one the case turns
on, and somebody has to draw it before anyone can argue about it.

***

# Walking the steps

| Part | What it does | Min | Watch for |
|---|---|---|---|
| **1** What you observe | Three Black shares, three denominators | 5 | Point at the three numbers now. Part 8 cashes them. |
| **2** Dilution by subtraction | `r pc(DEV_RANGE)` points against 10 | 5 | Students expect this to be the lab. It is the setup. |
| **3** What that rule replaced | County unit rule on 2020 population | 5 | `r pc(CU_RAT, 0)`-to-1. The rule worked so well it stopped being the argument. |
| **4** The half you cannot subtract | Large enough `r pc(TIMES)`× over; dissimilarity `r DISS` | 7 | **Large and spread evenly beats small and clustered.** Not intuitive. |
| **5** One algorithm, one dial | `grow()`, and the seed | 8 | Read the seed line aloud. Race picked the starting block. |
| **6** Turning the dial | The frontier, and the confident conclusion | **11** | **Take the prediction before the table.** Let them commit. |
| **7** The map already exists | Enacted D4 dominates everything | **10** | **The pivot.** Do not preview it. Do not rush it. |
| **8** "Majority" of what | `r nrow(f4)` districts, two answers | 6 | Take a vote on the denominator first. |
| **9** Does the seed matter | `r pc(SW_SPREAD, 2)` points across `r nrow(sw)` restarts | 4 | Anticlimax on purpose. Then Option B breaks it. |
| **10** Your turn | Five options | 4 | Steer strong students to B or C. |

**Part 6 deserves the prediction, and it is the one prediction in this course
that almost everybody gets wrong in the same direction.** Before the table, ask
the room for two numbers: the highest Black share a *contiguous* district of
`r nn(IDEAL)` people can reach, and how compact that district will be. The county
is `r pc(SH_VAP)`% Black among adults and holds `r pc(TIMES)` times the Black
adults a majority district needs, so most students guess something comfortably
over 50% at a respectable shape. The honest answer at a respectable shape is
`r pc(R1$vap_black_pct)`%, and the first setting that clears 50% —
`reach = `r CROSS$reach`` — does it at Polsby-Popper `r pc(CROSS$pp, 3)`,
about `r pc(R1$pp / CROSS$pp, 0)` times worse than where it started.

**If you are short on time**, protect Parts 6, 7 and 8. The confident conclusion,
its demolition, and the denominator. Parts 2 and 3 can be compressed to four
minutes together with no loss.

***

# The pivot, and how to run it

**Do not tell them the enacted map is coming.** The lab is written so Part 6
reads as a finding and Part 7 reads as a correction, and the sequence only works
if the room has actually accepted Part 6. Three moves:

1. **Make them write the sentence.** At the end of Part 6, have someone say out
   loud what an expert would put in a report: *in Houston County a majority-Black
   district can be constructed only by abandoning compactness.* Nothing in the
   analysis up to that point is wrong. The arithmetic is right, the frontier is
   real, the compactness measures are standard.
2. **Then run Part 7 without commentary.** Print the table. Let them find
   District 4 in it. `r pc(E4$vap_black_pct)`% at Polsby-Popper
   `r pc(E4$pp, 3)` — better than every setting on both axes at once.
3. **Then ask what went wrong.** The productive answers are all about the search
   and none about the county: it grows from a single seed, it never backtracks,
   it is greedy, and it only ever builds one district. The unproductive answer is
   "the algorithm was bad." It was a perfectly ordinary algorithm, and the
   frontier it traced is a **lower bound** on what a mapmaker can do — which is
   all any such search can ever produce.

**The wobble is corroborating evidence and worth two minutes.** At
`reach = `r WOB$reach`` the district *drops* to `r pc(WOB$vap_black_pct)`%,
below several tighter settings. A greedy search is not guaranteed to be monotone
in its own parameter. **A student running Option A at `k = 150` will find
`r pc(A150)`%** — also below `reach = 100`'s
`r pc(R100)`%. So it is not one bad point; it is
a whole region where reaching *further* finds *less*. The curve is a property of
the search, not of the county, and that is exactly the point Part 7 makes
geometrically.

**Say what Part 7 does not show.** It is not a claim about how the enacted plan
was drawn or with what in front of whom — that question is not in this file. And
it is not a comparison of two illustrative maps: an enacted plan and a
plaintiff's demonstration exhibit are different objects judged by different
rules. The claim is about evidence, and only about evidence.

***

# *Callais* and what it does to this lab

**The lab got better, not worse, and it is worth saying why.** Before 29 April
2026 the dial was a nice illustration of researcher degrees of freedom. Now
every setting except the leftmost describes a practice the Court has
disqualified, so the dial is also a dated artefact — and students can watch a
methodological choice become a legal one without any of the underlying numbers
changing.

Three things to have straight before the room asks.

- **The first condition bites on the dial.** Plaintiffs "cannot use race as a
  districting criterion." Every point on the Part 6 frontier except `reach = 1`
  was produced by letting race choose the next block.
- **The seed is not exempt, and this is the good question.** `reach = 1` takes
  race out of the *annexation rule* and leaves it in the *starting block*, which
  is `which.max(b$vap_black)`. If race may not be a districting criterion, a
  race-chosen seed is a problem too. **Option B is where a student can measure
  what that choice is worth**, and the answer is: at `reach = 1`, nearly
  everything.
- **The second condition is the one nothing here can meet.** An illustrative map
  must satisfy all the State's legitimate objectives — traditional criteria and
  "the State's specified political goals" — "just as well" as the enacted map.
  That needs a complete partition of the county, a statement of Georgia's stated
  objectives, and election results. **This folder has none of the three.** If a
  student concludes that `reach = 1` is therefore "the admissible answer," that
  is a 3, not a 4: `reach = 1` clears the first condition and is untested against
  the second.

**And the sequel is worth thirty seconds.** Five weeks after *Callais*, the Court
stayed the renewed injunction in the *Milligan* litigation itself, on the ground
that the plaintiffs' map had not performed "just as well" on Alabama's stated
criteria of keeping the Gulf Coast together and avoiding incumbent pairings.
*Allen v. Milligan*, 608 U.S. \_\_\_ (2026) (per curiam) (June 2, 2026). **The
plaintiffs who won in 2023 no longer have their district.** Students who read
*Milligan* as the stable point of this area of law should hear that.

***

# Handling the room

- **Nobody writes code here.** Say it at the top, as always. The `grow()`
  function is fifteen lines and is printed so that students can see there is no
  magic in it, not so they can write one.
- **Somebody will say the exercise is racist.** Take it seriously rather than
  deflecting, because the observation is correct and the lab is built on it: to
  find out whether Black voters *could* have a district, the algorithm sorted
  every block in the county by race. Section 2 remedies require sorting voters by
  race in order to stop voters being sorted by race. That tension is the subject,
  it is why *Shaw* and *Miller* exist, and it is what *Callais* claims to
  resolve. **There is no race-blind way to ask this question**, and saying so
  plainly is better than any answer that pretends otherwise.
- **The live case needs one sentence, early.** The instructor is retained by
  plaintiffs in *Driver v. Houston County*, No. 5:25-cv-00025 (M.D. Ga.). No
  expert-report material is used, every non-enacted district here was drawn by
  the printed algorithm, and the lab reaches no legal conclusion. Say it once, at
  the start, and move on.
- **Compactness scores are not facts about virtue.** A district bounded by a
  winding river scores badly for reasons no mapmaker chose, and any district can
  be made to score better by smoothing a boundary without moving a single voter.

***

# Option A — Move the dial yourself

## Verified output

```{r a}
t(sapply(c(5, 30, 150), function(k) {
  g <- grow(seed, IDEAL, k)
  c(reach = k, blocks = length(g), r2(score(g)))
}))
```

For comparison, the settings on either side, from `tradeoff.csv`:

```{r a2}
print(t5[t5$reach %in% c(4, 6, 24, 40, 100, 200),
         c("reach", "vap_black_pct", "reg_black_pct", "pp", "reock")],
      row.names = FALSE)
```

**3/4:** Runs a value of `k`, reports the Black share, and places it correctly
between its neighbours on the Part 6 curve.

**4/4:** Answers the second half of the question — could you defend that `k` to a
judge who asked why you chose it? — and concludes that you could not, because
`reach` has no meaning outside this program. There is no `k` that corresponds to
a districting principle, which means the whole curve is a family of arbitrary
maps and the analyst picks a point on it.

**4/4, best:** Picks `k = 150` (or any value in the 100–200 range) and notices
that it comes out at **`r pc(A150)`%, below
`reach = 100`'s `r pc(R100)`%** — so the wobble
the lab flags at `reach = `r WOB$reach`` is not an isolated point but a region.
A student who works out *why* a greedy search can do worse with more options
(early annexations foreclose later ones; the centroid drifts and changes what
counts as "near") has understood that the frontier is an artefact of the
procedure. That is the same lesson as Part 7, reached independently.

***

# Option B — Start somewhere else

**This is the most valuable option in the lab and the one to steer good students
toward.** It produces a result that contradicts the lab's own Part 9, and the
contradiction is correct.

## Verified output

```{r b}
t(sapply(c(1, 10, 50), function(i) {
  s <- ord[i]
  c(rank = i, seed_black_adults = b$vap_black[s],
    reach1_adults   = unname(r2(score(grow(s, IDEAL, 1))[2])),
    reachInf_adults = unname(r2(score(grow(s, IDEAL, Inf))[2])))
}))
```

The full sweep — the same `r nrow(sw)` starting blocks the build script used,
re-run at `reach = 1`:

```{r b2}
c(seeds             = length(R1SW),
  reach1_min        = r2(min(R1SW)),
  reach1_max        = r2(max(R1SW)),
  reach1_sd         = r2(sd(R1SW)),
  reach1_spread     = r2(R1SW_SPREAD),
  reachInf_spread   = r2(SW_SPREAD),
  reachInf_sd       = r2(sd(sw$d5_vap_black_pct)),
  reach1_over_50    = sum(R1SW > 50))
```

**Part 9 says the seed barely matters. Part 9 is measuring the race-aware
algorithm.** At unlimited reach the seed is nearly irrelevant —
`r pc(SW_SPREAD, 2)` points across `r nrow(sw)` restarts — because the search
reaches far enough to find the same Black population wherever it starts. At
`reach = 1` the same `r nrow(sw)` restarts span **`r pc(R1SW_SPREAD, 2)`
points**, from `r pc(min(R1SW))`% to `r pc(max(R1SW))`%, a standard deviation of
`r pc(sd(R1SW))` against `r pc(sd(sw$d5_vap_black_pct))`.

**3/4:** Runs the three seeds, reports that the answers differ more at
`reach = 1` than the lab's Part 9 suggested.

**4/4:** Explains the mechanism. Unlimited reach can go and get the Black
population from anywhere in the county, so the starting block washes out; at
`reach = 1` the district is whatever happens to be adjacent, so it is a
description of one neighbourhood rather than a search. **The constraint that
makes the map legally admissible is the same constraint that makes it depend
almost entirely on an arbitrary choice.**

**4/4, best:** Closes the loop on *Callais*. The seeds were ranked by
`order(-b$vap_black)` — chosen by race. So the one race-based decision left in
the `reach = 1` construction is now the decision that determines the answer, and
it moves it `r pc(R1SW_SPREAD, 2)` points. A student who asks whether a
`reach = 1` map grown from a race-chosen seed is really a race-blind map, and
notices that **none of the `r length(R1SW)` race-blind-growth districts clears
50% anyway** (the best is `r pc(max(R1SW))`%), has produced a better argument
than the lab makes on its own. Give it a 4 and use it next year.

***

# Option C — Change the denominator

## Verified output

```{r c}
rbind(
  `reach 1, max adults`       = r2(score(CA1)),
  `reach 1, max registered`   = r2(score(CR1)),
  `reach Inf, max adults`     = r2(score(CAI)),
  `reach Inf, max registered` = r2(score(CRI)))
```

```{r c2}
identical(sort(CA1), sort(CR1))          # reach 1: the very same blocks
r2(JAC)                                  # reach Inf: % of blocks shared
```

**The `reach = 1` rows are identical, and that is the finding.** With one
candidate on the frontier there is no choice to make, so the objective function
never runs. **The denominator only matters when the dial lets it matter** — which
means the two analyst choices in this lab are not independent, and a student who
sees that has spotted something the brief does not say.

At unlimited reach they separate hard. Maximising adults gives
`r pc(score(CAI)[2])`% of adults and `r pc(score(CAI)[3])`% of registrants;
maximising registrants gives `r pc(score(CRI)[3])`% of registrants and only
`r pc(score(CRI)[2])`% of adults. The two districts share
`r pc(JAC, 1)`% of their blocks.

**3/4:** Swaps the columns, reports that the district changes, and notes the
Black share is higher on the registered measure.

**4/4:** Answers the normative half. Courts generally use voting-age or
citizen-voting-age population, and the case for it is that registration is
*endogenous to the wrong being alleged* — if a group is discouraged from
registering, using registration as the denominator lets the discouragement
manufacture the majority. The case for registration is that it measures the
actual electorate, and *Bartlett*'s concern was whether a group can elect its
candidate, which registrants do and non-registered adults do not.

**4/4, best:** Notices the `reach = 1` identity above and draws the general
lesson: **a sensitivity analysis that varies one choice at a time will report
that the denominator does not matter, because at the defensible setting of the
other knob it genuinely does not.** That is a real methodological trap and it is
worth naming for the whole room if a student finds it.

***

# Option D — Four districts instead of five

## Verified output

```{r d}
rbind(`reach 1`   = r2(D41),
      `reach Inf` = r2(D4I))
```

```{r d2}
print(f4[, c("reach", "pop_black_pct", "vap_black_pct", "reg_black_pct")],
      row.names = FALSE, digits = 4)
```

Larger districts, and the arithmetic gets easier: the ideal rises from
`r nn(IDEAL)` to `r nn(IDEAL4)` people. But **the answer depends entirely on the
denominator, which is the trap the option is built around.** Across all
`r nrow(f4)` finite settings at or below `reach = 100`, the Black share of
registered voters never falls below `r pc(min(f4$reg_black_pct))`% and the Black
share of adults never rises above `r pc(max(f4$vap_black_pct))`%. Every one of
them is majority-Black on one measure and not the other.

**3/4:** Re-runs with `ideal4`, reports that the shares are higher than in the
five-district case, and concludes a majority-Black district is easier.

**4/4:** Refuses to answer "easier or harder" without naming the denominator,
because on registered voters the four-district configuration clears 50% at every
finite setting including `reach = 1`, and on voting-age population it clears 50%
at none of them.

**4/4, best:** Notices that fewer, larger districts is not a neutral change —
it is a change in the *unit of representation*, and it makes proportionality
coarser in a way that cuts both directions. A group at `r pc(SH_VAP)`% of adults
is entitled by proportionality to `r pc(5 * SH_VAP / 100, 2)` of five seats and
`r pc(4 * SH_VAP / 100, 2)` of four. **Shrinking the board makes each seat worth
more and makes the group's claim to a whole one weaker**, even as it makes the
district easier to draw. Reward anyone who gets both halves.

***

# Option E — Your own question

Grade on whether the question could be answered with what is in the folder.

Good ones that have come up:

- *Does the seed's own Black share predict the district's?* It does at
  `reach = 1` and barely at all at unlimited reach — a clean restatement of
  Option B.
- *How many blocks does each district use?* `reach = 1` takes `r R1$blocks`
  blocks to reach the ideal population; unlimited reach takes `r RI$blocks` —
  `r pc(RI$blocks / R1$blocks, 1)` times as many for the same people. That ratio
  is what a tendril looks like numerically.
- *What if contiguity is dropped entirely?* That is the ceiling already in the
  lab: `r pc(CEIL$vap_black_pct)`% at Polsby-Popper `r pc(CEIL$pp, 3)`. A
  student who asks why the ceiling is not simply the answer has understood that
  "could a district be drawn" means a district somebody could enact.

Questions the folder cannot answer, which should be redirected rather than
penalised: anything about how anyone voted (no election results are here),
anything about citizenship (registration is not citizenship), and anything
requiring a complete five-district plan.

***

# Grading

Four-point scale from the syllabus.

- **Never penalise code.** It is given, and it runs.
- **The move that earns a 4** is treating the failed search in Part 6 as
  information about the search rather than about the county. Everything else in
  the lab is downstream of that distinction.
- **A 2** is "the algorithm was bad" or "you can make data say anything." Both
  skip the work. The algorithm was ordinary and the data were not manipulated;
  that is why the result is disturbing.
- **Also a 2:** reporting a number off the Part 6 frontier as though it were a
  fact about Houston County. The frontier is a lower bound.
- **A 3, not a 4:** concluding that `reach = 1` is "the answer after *Callais*."
  It satisfies the first condition and is untested against the second, and its
  own value swings `r pc(R1SW_SPREAD, 2)` points on the seed.
- **Reward anyone who asks what a *complete plan* would have shown.** The
  algorithm builds one district and stops; a real demonstration plan assigns
  every block and holds all five inside the population band at once. Noticing
  that the lab answers an easier question than a court would ask is the most
  sophisticated available response, and it is not stated anywhere in the lab
  except in the limitations.

## Anticipated questions

- *"Why not just look at the enacted map from the start?"* — Because in a real
  case the map you need does not exist; that is the entire difficulty. The
  enacted plan is available here as an answer key, and answer keys are what this
  course uses to grade methods that are normally unfalsifiable. Same move as the
  BISG and RPV labs.
- *"Isn't Polsby-Popper just a bad measure?"* — It has real defects, and Duchin
  and Tenner (2024) argue the whole contour-based family shares them. But the
  qualitative result here does not depend on the measure: Reock moves the same
  way, from `r pc(R1$reock, 3)` to `r pc(RI$reock, 3)`. Two contour measures are
  one limitation, not two independent checks — say that too.
- *"Could a second majority-Black district be drawn?"* — Unknown, and the lab
  cannot say. The algorithm builds one district and stops. A complete plan can
  fail for reasons having nothing to do with the group, such as running out of
  population in the right places.
- *"Is `reach = 1` what an expert would actually file now?"* — No. It is the only
  end of *this* dial that survives the first *Callais* condition, which is not
  the same as being an admissible demonstration map. Nobody yet knows what the
  procedure is; the Court required the showing and did not specify how to make
  it.
- *"The registration file has race on it — why is the census used at all?"* —
  Because registration is endogenous to the claim and the census is not, and
  because *Bartlett* is generally read against voting-age population. Part 8 is
  the demonstration that this is a choice and not a technicality.
- *"Did we prove Houston County dilutes anyone's vote?"* — No, and do not let
  this become a finding about the county. Nothing here touches preconditions two
  and three or the totality of the circumstances. **Precondition one is a
  threshold, not a verdict.**

## Connections

- **`rpv`** — preconditions two and three, same county, same registration
  file. Run it first; this lab assumes the room already knows what a *Gingles*
  precondition is. The two labs also split *Callais* cleanly between them: the
  RPV lab gets the party-control requirement, this one gets the illustrative-map
  conditions.
- **`bisg-check` and `surnames`** — where the self-reported race
  on the registration file comes from, and what the any-part-Black definition
  costs.
- **`areal-units`** — districting is the zoning effect deployed on purpose.
  `r nn(nrow(b))` blocks re-bundled into five districts give different summary
  numbers depending only on where the lines fall; that lab is the general case
  and this one is the consequential instance.
- **`precinct-geography`** — the other lab where two defensible methods on the
  same data disagree about something basic. Pair the two if you want the
  methodological point to land twice.
- **`apportionment`** — the other place a technical rule with no obvious
  politics turns out to allocate something people care about, and the other place
  equal population is the whole argument.
