---
title: "Lab: The District That Was Not Drawn"
subtitle: "84-355 Democracy's Data | Vote dilution, Houston County, Georgia"
author: "Jonathan Cervas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>", message = FALSE,
                      warning = FALSE, fig.width = 7.2, fig.height = 4.6)
options(scipen = 999)
```

Everything here runs as given. Your job is to read what comes out and say what
it means.

***

# Before we start

**Vote dilution** is the claim that your vote is worth less than somebody
else's. American law recognises two versions and they are not alike.

**Malapportionment** is districts of unequal population. *Reynolds v. Sims*, 377
U.S. 533 (1964) held the Equal Protection Clause forbids it. You settle it by
counting people.

**Racial vote dilution** under Section 2 of the Voting Rights Act, 52 U.S.C.
§ 10301, is districts drawn so that a cohesive minority group, voted against by
a cohesive majority, cannot elect anyone. *Thornburg v. Gingles*, 478 U.S. 30,
50–51 (1986) made a plaintiff show three things first: the group is
**sufficiently large and geographically compact** to be a majority in a
single-member district; it is **politically cohesive**; and the majority **votes
as a bloc** usually to defeat its preferred candidate.

**Last week did preconditions two and three.** The RPV lab estimated whether
Black and white voters preferred different candidates, and graded the estimate
against records that happen to exist.

**Today does precondition one, and it is a different kind of claim.** The other
two are about an election that happened. This one is about **a district that
does not exist** — that a map *could* have been drawn in which the group would
have been a majority. Nobody drew it. So we have to.

> **The law changed four months ago.** *Louisiana v. Callais*, 608 U.S. \_\_\_
> (2026), decided 29 April 2026, left the three preconditions standing and
> changed what satisfies them. Precondition one now carries **two** conditions.
> A plaintiff "cannot use race as a districting criterion" in drawing an
> illustrative map; and the map "must meet all the State's legitimate districting
> objectives, including traditional districting criteria and the State's
> specified political goals," achieving them **"just as well"** as the State's own
> map (slip op., at 29). Part 5 builds a district with a dial that optimises race
> and nothing else, because that was standard practice until this spring. Part 6
> says which end of the dial survives the first condition — and why nothing in
> this lab survives the second.

For Parts 5 and 6, **pretend the enacted map does not exist.** That is the
position of an expert asked whether a district *could* be drawn.

# Part 1 — What you observe

```{r load}
b <- read.csv("data/blocks.csv", colClasses = c(GEOID20 = "character"))
a <- read.csv("data/adjacency.csv")

cat("census blocks:      ", nrow(b), "\n")
cat("population:         ", sum(b$pop), "\n")
cat("voting-age pop:     ", sum(b$vap), "\n")
cat("registered voters:  ", sum(b$reg), "\n\n")
cat(sprintf("Black share of population:      %.2f%%\n",
            100 * sum(b$pop_black) / sum(b$pop)))
cat(sprintf("Black share of voting-age pop:  %.2f%%\n",
            100 * sum(b$vap_black) / sum(b$vap)))
cat(sprintf("Black share of registered:      %.2f%%\n",
            100 * sum(b$reg_black) / sum(b$reg)))
```

One row per **2020 census block**, the atom every districting plan is built
from. Houston County's Board of Education is elected from five single-member
districts under a plan enacted by the Georgia General Assembly; column `boe5`
says which district each block is in.

**Read those last three numbers again.** The Black share of this county is one
of three different things depending on which population you mean. Part 7 is
about what that does.

# Part 2 — The dilution you settle by subtraction

```{r deviation}
ag <- aggregate(cbind(pop, vap, vap_black) ~ boe5, b, sum)
ideal <- mean(ag$pop)
ag$deviation <- 100 * (ag$pop - ideal) / ideal

print(ag[, c("boe5", "pop", "deviation")], row.names = FALSE, digits = 3)

cat(sprintf("\nideal district: %.0f people\n", ideal))
cat(sprintf("total spread:   %.2f points\n", max(ag$deviation) - min(ag$deviation)))
cat(sprintf("a vote in the smallest district is worth %.2f%% more than in the largest\n",
            100 * (max(ag$pop) / min(ag$pop) - 1)))
```

```{r dev-plot, fig.height=3.4}
par(mar = c(4.4, 5.6, 2.0, 1.4))
y <- 5:1
plot(NA, xlim = c(-5.6, 5.6), ylim = c(0.5, 5.5), yaxt = "n", bty = "n", las = 1,
     xlab = "deviation from equal population (%)", ylab = "")
rect(-5, 0.3, 5, 5.7, col = "#f0f5ea", border = NA)
abline(v = 0, col = "grey45"); abline(v = c(-5, 5), lty = 2, col = "#4d9221")
axis(2, at = y, labels = paste("District", ag$boe5), las = 1, tick = FALSE)
segments(0, y, ag$deviation, y, col = "#2c7fb8", lwd = 3)
points(ag$deviation, y, pch = 19, cex = 1.5, col = "#2c7fb8")
mtext("green band is the 10% a legislative plan is allowed", 3, line = 0.3,
      cex = 0.82, col = "#555")
```

*Brown v. Thomson*, 462 U.S. 835, 842 (1983) collected the rule that a state or
local plan under a **10%** maximum deviation is a minor deviation that does not
by itself make out a case. **This plan uses a sixth of that allowance.** The
malapportionment inquiry is over.

If it is this easy, what was it for?

# Part 3 — What that rule replaced

Until 1963 Georgia nominated statewide officers under the **county unit
system**: counties cast unit votes, not people — six each for the eight largest,
four for the next thirty, two for the remaining hundred and twenty-one — and a
candidate needed a majority of units. *Gray v. Sanders*, 372 U.S. 368 (1963)
struck it down, and it is where Justice Douglas wrote, at 381, that political
equality "can mean only one thing — one person, one vote."

That rule, unchanged, on Georgia's 2020 population:

```{r county-units}
cu <- read.csv("data/ga_county_units.csv", colClasses = c(fips = "character"))

cat(sprintf("people per unit vote, largest county:  %8.0f\n",
            max(cu$people_per_unit)))
cat(sprintf("people per unit vote, smallest county: %8.0f\n",
            min(cu$people_per_unit)))
cat(sprintf("ratio:                                 %8.0f to 1\n",
            max(cu$people_per_unit) / min(cu$people_per_unit)))

o <- cu[order(cu$people_per_unit), ]
k <- which(cumsum(o$units) >= floor(sum(cu$units) / 2) + 1)[1]
cat(sprintf("\n%d of 159 counties could cast a unit-vote majority,\n", k))
cat(sprintf("and they hold %.1f%% of Georgians.\n",
            100 * sum(o$pop[seq_len(k)]) / sum(cu$pop)))
```

Those are the numbers *Reynolds* and *Gray* were written against. **The
equal-population rule worked so completely that it stopped being where the
argument is.**

# Part 4 — The dilution you cannot settle by subtraction

The first *Gingles* precondition has two halves. Only the first is arithmetic.

```{r sufficiently-large}
d_vap   <- sum(b$vap) / 5          # voting-age people per district
need_bl <- d_vap / 2               # Black adults a bare majority would take
have_bl <- sum(b$vap_black)

cat(sprintf("voting-age people per district: %8.0f\n", d_vap))
cat(sprintf("Black adults a majority needs:  %8.0f\n", need_bl))
cat(sprintf("Black adults the county has:    %8.0f\n", have_bl))
cat(sprintf("\n'sufficiently large' is satisfied %.2f times over\n",
            have_bl / need_bl))
```

*Bartlett v. Strickland*, 556 U.S. 1 (2009), a plurality, read the precondition
to require a real numerical majority — greater than 50 percent of voting-age
population, at 19–20. A district the group could win only with crossover help
"§ 2 does not require," at 25.

That leaves **"geographically compact."** The Court's current phrase is a
**"reasonably configured"** district, from *Wisconsin Legislature v. Wisconsin
Elections Commission*, 595 U.S. 398, 402 (2022); *Allen v. Milligan*, 599 U.S. 1,
18 (2023) glossed it as one that "comports with traditional districting
criteria." It is a fact about where people live.

```{r map, fig.height=6.2, fig.width=6.4}
bs <- read.csv("data/block_shapes.csv", colClasses = c(id = "character"))
ps <- read.csv("data/plan_shapes.csv")

share <- ifelse(b$vap > 0, b$vap_black / b$vap, NA)
brk   <- c(0, .10, .25, .40, .50, .65, 1)
col   <- c("#f7f7f7", "#d9e6f2", "#a9c9e2", "#6fa8d0", "#2c7fb8", "#17527a")
fill  <- ifelse(is.na(share), "#f4f4f4", col[cut(share, brk, include.lowest = TRUE)])

par(mar = c(0.4, 0.4, 2.0, 0.4))
plot(NA, xlim = range(bs$x), ylim = range(bs$y), asp = 1, axes = FALSE,
     xlab = "", ylab = "")
for (r in split(bs, interaction(bs$id, bs$part, drop = TRUE)))
  polygon(r$x, r$y, col = fill[match(r$id[1], b$GEOID20)], border = NA)
for (dd in 1:5) {
  z <- ps[ps$id == paste("Enacted BOE district", dd), ]
  for (r in split(z, z$part)) lines(r$x, r$y, col = "#C41230", lwd = 1.4)
}
legend("topleft", c("over 65%", "50-65%", "40-50%", "25-40%", "10-25%",
                    "under 10%", "no adults"),
       fill = c(rev(col), "#f4f4f4"), border = "grey70", bty = "n", cex = 0.7,
       title = "Black share of a block's adults", title.adj = 0)
mtext("red lines are the five enacted districts", 3, line = 0.4, cex = 0.82,
      col = "#555")
```

```{r segregation}
sg <- read.csv("data/segregation.csv")
print(sg, row.names = FALSE)

hi <- b$vap >= 20 & b$vap_black / pmax(b$vap, 1) > 0.5
cat(sprintf("\nblocks where Black adults are already a majority: %d of %d\n",
            sum(hi), nrow(b)))
cat(sprintf("share of the county's Black adults living in them: %.1f%%\n",
            100 * sum(b$vap_black[hi]) / have_bl))
```

The **index of dissimilarity** is the share of Black adults who would have to
move blocks for the two groups to be spread identically. This is moderate
segregation, not severe — and **a group can be plenty large and still be spread
too evenly to be gathered into a majority anywhere.**

# Part 5 — One algorithm with one dial

To find out whether the district exists, build it. Start at one block and annex
neighbours until the district holds the ideal population. At each step the
mapmaker looks at the **`reach`** nearest available blocks on the frontier and
takes whichever leaves the Black share of adults highest.

```{r grow}
# adjacency list: which blocks touch which. Row numbers of `b`.
nb <- split(c(a$a, a$b), c(a$b, a$a))
NB <- vector("list", nrow(b))
NB[as.integer(names(nb))] <- nb

grow <- function(seed, target, reach = Inf) {
  inn <- logical(nrow(b)); inn[seed] <- TRUE
  P <- b$pop[seed]; N <- b$vap_black[seed]; D <- b$vap[seed]
  cx <- b$x_km[seed]; cy <- b$y_km[seed]
  fr <- NB[[seed]]
  while (P < target && length(fr)) {
    if (is.finite(reach) && reach < length(fr)) {
      dd <- (b$x_km[fr] - cx)^2 + (b$y_km[fr] - cy)^2
      cand <- fr[order(dd)[seq_len(reach)]]
    } else cand <- fr
    pick <- cand[which.max((N + b$vap_black[cand]) / pmax(D + b$vap[cand], 1))]
    inn[pick] <- TRUE
    P <- P + b$pop[pick]; N <- N + b$vap_black[pick]; D <- D + b$vap[pick]
    w  <- b$pop[inn] + 1
    cx <- sum(b$x_km[inn] * w) / sum(w); cy <- sum(b$y_km[inn] * w) / sum(w)
    fr <- setdiff(unique(c(fr, NB[[pick]])), which(inn))
  }
  which(inn)
}

# how Black is a set of blocks, three ways?
score <- function(i) c(
  population = 100 * sum(b$pop_black[i]) / sum(b$pop[i]),
  adults     = 100 * sum(b$vap_black[i]) / sum(b$vap[i]),
  registered = 100 * sum(b$reg_black[i]) / sum(b$reg[i]))

seed <- which.max(b$vap_black)          # the block with most Black adults
cat("seed block:", b$GEOID20[seed], "with", b$vap_black[seed], "Black adults\n")
```

- **`reach = 1`.** One candidate is ever considered, so race never enters the
  *annexation rule*. The roundest district the geography allows — and after
  *Callais*, the only end of this dial that could be offered to a court.
- **`reach = Inf`.** Every frontier block is a candidate, however far out. One
  objective, no other constraint.

**Look at the line above the bullets before you accept that.** `seed` is
`which.max(b$vap_black)` — the block picked *because* it holds the most Black
adults. Race chose the starting point at every setting of the dial, `reach = 1`
included. Turning the dial down takes race out of how the district grows and
leaves it in where the district begins. Option B is where you can test what that
is worth.

```{r two-ends}
round(rbind(`reach = 1`   = score(grow(seed, ideal, 1)),
            `reach = Inf` = score(grow(seed, ideal, Inf))), 2)
```

A third construction abandons contiguity: take the most heavily Black blocks in
order, wherever they are. Nobody could enact it. It is a ceiling.

```{r ceiling}
o    <- order(b$vap_black / pmax(b$vap, 1), b$vap_black, decreasing = TRUE)
ceil <- o[seq_len(sum(cumsum(b$pop[o]) < ideal) + 1)]
round(score(ceil), 2)
```

# Part 6 — Turning the dial

```{r tradeoff}
tr <- read.csv("data/tradeoff.csv")
t5 <- tr[tr$districts == 5, ]
print(t5[, c("reach", "vap_black_pct", "reg_black_pct", "pp", "reock")],
      row.names = FALSE, digits = 4)
```

**Polsby-Popper** is `4*pi*A / P^2`: one for a circle, near zero for anything
with a long ragged edge. **Reock** is the district's area over the area of the
smallest circle containing it. Both are computed on the unsimplified geometry by
`data/build-data.R`, because base R cannot do it.

```{r frontier-plot, fig.height=4.4}
d  <- read.csv("data/districts.csv")
e4 <- d[d$plan == "Enacted BOE district 4", ]
cl <- d[d$plan == "Ceiling, 5 districts", ]

par(mar = c(4.4, 4.6, 2.0, 1.2))
plot(NA, xlim = c(0, 0.36), ylim = c(38, 70), las = 1,
     xlab = "Polsby-Popper compactness (1 is a circle)",
     ylab = "Black share of the district's adults (%)")
rect(-1, 50, 1, 100, col = "#f0f5ea", border = NA)
abline(h = 50, lty = 2, col = "#4d9221", lwd = 1.6)
z <- t5[order(t5$pp), ]
lines(z$pp, z$vap_black_pct, col = "#2c7fb8", lwd = 2)
points(t5$pp, t5$vap_black_pct, pch = 19, col = "#2c7fb8")
points(cl$pp, cl$vap_black_pct, pch = 17, cex = 1.3, col = "#8856a7")
text(cl$pp, cl$vap_black_pct, " ceiling: contiguity dropped", pos = 4,
     cex = 0.72, col = "#8856a7")
text(0.355, 51.4, "majority-Black ", pos = 2, cex = 0.74, col = "#4d9221")
mtext("each point is one setting of the dial", 3, line = 0.3, cex = 0.82,
      col = "#555")
```

**The dial buys Black share and pays in compactness, at a brutal rate.** At
`reach = 1` the district is a respectable shape and nowhere near a majority. The
first setting that clears 50% does so with a Polsby-Popper roughly nine times
worse.

One setting is out of order — the district at `reach = 200` drops below several
tighter settings. **A greedy algorithm is not guaranteed to be monotone in its
own parameter.** The curve an expert report shows you is a property of the
search, not of the county.

And read it with the calendar in mind. Every point except the leftmost was
produced by letting race decide which block to annex. **After *Callais* a
plaintiff "cannot use race as a districting criterion,"** which leaves
`reach = 1`.

That is the generous reading. The second condition asks more than this chart can
answer: the illustrative map must meet **all** the State's legitimate objectives
— traditional criteria and "the State's specified political goals" — **"just as
well"** as the enacted map. Nothing here is a complete plan, nothing here knows
what Georgia's objectives were, and no election result is in the folder. **On the
current standard the whole chart is inadmissible, not just its right-hand side.**

The honest write-up: *in Houston County a majority-Black district can be
constructed only by abandoning compactness.*

That is wrong.

# Part 7 — The map already exists

```{r pivot}
print(d[, c("plan", "pop", "vap_black_pct", "reg_black_pct", "pp", "reock")],
      row.names = FALSE, digits = 4)
```

```{r pivot-map, fig.height=3.4, fig.width=7.2}
par(mfrow = c(1, 3), mar = c(0.3, 0.3, 2.0, 0.3))
draw <- function(nm, colr, ttl) {
  plot(NA, xlim = range(bs$x), ylim = range(bs$y), asp = 1, axes = FALSE,
       xlab = "", ylab = "")
  z <- ps[ps$id == "COUNTY", ]
  for (r in split(z, z$part)) polygon(r$x, r$y, col = "#f2f2f2", border = "grey65")
  z <- ps[ps$id == nm, ]
  for (r in split(z, z$part)) polygon(r$x, r$y, col = colr, border = colr)
  s <- d[d$plan == nm, ]
  mtext(ttl, 3, line = 0.85, cex = 0.74)
  mtext(sprintf("%.1f%% Black adults, PP %.3f", s$vap_black_pct, s$pp), 3,
        line = 0.05, cex = 0.64, col = "#666")
}
draw("Reach 1, 5 districts", "#2c7fb8", "Algorithm, reach 1")
draw("Reach unlimited, 5 districts", "#e08214", "Algorithm, reach unlimited")
draw("Enacted BOE district 4", "#C41230", "Enacted District 4")
```

```{r dominance}
cat(sprintf("enacted District 4: %.2f%% Black adults, Polsby-Popper %.3f\n",
            e4$vap_black_pct, e4$pp))
cat(sprintf("algorithm districts that beat it on BOTH measures: %d of %d\n",
            sum(t5$vap_black_pct >= e4$vap_black_pct & t5$pp >= e4$pp), nrow(t5)))
```

**The algorithm's failure was a fact about the algorithm.** A district that
already existed, drawn by hand out of the whole county, dominated everything it
produced.

This does **not** show how the enacted plan was drawn, and it is not a
comparison of two illustrative maps. What it shows is about evidence.
**Exhibiting a map settles the question in the affirmative for good. Failing to
find one settles nothing**, because the search was never exhaustive and never
could be. The first *Gingles* precondition is provable and not disprovable, and
that is exactly where *Callais* intervened.

# Part 8 — "Majority" of what?

*Bartlett* requires a majority. Of what? Take the **four-district**
configuration, where the three denominators disagree.

```{r denominator}
t4 <- tr[tr$districts == 4, ]
f4 <- t4[is.finite(t4$reach) & t4$reach <= 100, ]
o  <- f4[, c("reach", "pop_black_pct", "vap_black_pct", "reg_black_pct")]
o$majority_on_adults     <- f4$vap_black_pct > 50
o$majority_on_registered <- f4$reg_black_pct > 50
print(o, row.names = FALSE, digits = 4)
```

**Every one of those districts is majority-Black among registered voters. Not
one of them is majority-Black among voting-age population.** Same blocks, same
lines, same people, two answers to the question *Bartlett* says is dispositive.

Courts generally work in voting-age or citizen-voting-age population — the more
conservative of the two here. But the choice is a choice, and it decides the
question.

# Part 9 — Does the starting block matter?

```{r sweep}
sw <- read.csv("data/seed_sweep.csv")
cat(sprintf("starting blocks tried: %d\n", nrow(sw)))
cat(sprintf("results ran %.2f%% to %.2f%%, sd %.2f points\n",
            min(sw$d5_vap_black_pct), max(sw$d5_vap_black_pct),
            sd(sw$d5_vap_black_pct)))
cat(sprintf("\nfor comparison, the dial moved the answer %.1f points\n",
            max(t5$vap_black_pct) - min(t5$vap_black_pct[t5$reach == 1])))
```

**Where you start barely matters. How far you are willing to reach matters
enormously.** Not every analyst choice moves the answer — and the one that does
is the one encoding what you think a district is allowed to look like.

# Part 10 — Your turn

Change **one** thing, re-run, and write about what changed. Pick one.

**Option A — Move the dial yourself.** Run `grow(seed, ideal, k)` for a value of
`k` not in the table — try 5, 30, or 150 — and `score()` the result. Where does
your district sit against the curve in Part 6, and could you defend that value
of `k` to a judge who asked why you chose it?

**Option B — Start somewhere else.** Part 9 says the seed barely matters for the
unlimited-reach algorithm. Test whether that holds at `reach = 1`, where race is
out of the annexation rule: pick three different seeds
(`order(-b$vap_black)[c(1, 10, 50)]`) and `score(grow(s, ideal, 1))` each. Does
the answer depend on the starting block more or less than the race-aware
version does, and why would that be? Then the harder half: those three seeds
were themselves ranked by Black adults. **If race may not be a districting
criterion, is `reach = 1` from a race-chosen seed a race-blind map or not?**

**Option C — Change the denominator.** `grow()` maximises the Black share of
**adults**. Rewrite it to maximise the share of **registered voters** instead
(swap `b$vap_black`/`b$vap` for `b$reg_black`/`b$reg` inside the function).
Compare the district you get to the one in Part 5. Which denominator should an
expert use, and what is the argument for the other one?

**Option D — Four districts instead of five.** Set `ideal4 <- sum(b$pop) / 4`
and re-run Parts 5 and 6. A smaller board means larger districts. Does that make
a majority-Black district easier or harder, and does your answer depend on which
denominator you count?

**Option E — Your own question.** `r format(nrow(b), big.mark = ",")` blocks, an adjacency list, three
populations, and an enacted map to check yourself against.

# What this lab cannot tell you

- **Whether anyone's vote is diluted.** Nothing here touches preconditions two
  and three, or the totality of the circumstances, which is where these cases
  are decided. **Precondition one is a threshold, not a verdict.**
- **Whether a *second* majority-Black district could be drawn.** The algorithm
  builds one district and stops. A real demonstration plan must assign every
  block and hold all five inside the population band at once — a harder problem
  that can fail for reasons having nothing to do with the group.
- **Whether any of this would be admissible.** *Callais*'s second condition
  wants a complete plan that meets all of Georgia's stated objectives, political
  ones included, "just as well" as the enacted map. That needs a full partition,
  a statement of what those objectives were, and election data. This folder has
  none of the three. **We measure the first condition and are silent on the
  second**, which is the harder one.
- **What is actually achievable.** Part 7 is the proof: our search failed and a
  better map existed. The frontier in Part 6 is a lower bound on what a mapmaker
  could do, and nothing more.
- **Anything about compactness scores as facts.** A district bounded by a
  winding river scores badly for reasons no mapmaker chose, and one can be made
  to score better by smoothing a boundary without moving a voter.
- **How people vote.** No election results are in this file. Whether a district
  *performs* is a separate question.
- And notice what the exercise required: to ask whether Black voters *could*
  have a district, the algorithm had to sort every block by race. **There is no
  race-blind way to ask this question.**

# On the discussion board

By **Friday, 11:59 p.m.** No length requirement; a sentence or two is enough, so
long as it shows you engaged with what we did. Any *one* of the usual four.

Three that would make good posts:

- Part 6 concluded that no compact majority-Black district could be drawn, and
  Part 7 showed one that already existed. What is the general lesson about
  evidence produced by a search?
- The same district is majority-Black on one denominator and not on another.
  If you were the judge, which would you ask for, and what would you say to the
  side that wanted the other one?
- Section 2 remedies require sorting voters by race in order to stop voters
  being sorted by race, and *Callais* has now said a plaintiff "cannot use race
  as a districting criterion" in the map that proves the claim. Is there a
  coherent way out of that, or is the statute asking for something impossible?

***

### Where the data came from

**2020 Census, P.L. 94-171** redistricting file at block level for Houston
County, Georgia: population and voting-age population, both by race. **Black is
any-part Black** (alone or in combination), matching Voting Rights Act practice;
the surnames lab documents what the alternative definition costs.

**2020 TIGER/Line tabulation blocks** for Georgia supply geometry, the adjacency
list, and both compactness measures. TIGER's independent `POP20` field agrees
with the census tables at `r format(sum(b$pop), big.mark = ",")` people.

**The Houston County Board of Education plan**, as enacted by the Georgia
General Assembly and in force, as a block-equivalency file.

**The Georgia voter registration file** for Houston County: **self-reported**
race and census block. Rows whose race had been filled in by inference rather
than reported by the voter were dropped upstream — see
`bisg-check/data/build-data.R`.

**Every district here other than the enacted five was drawn by the `grow()`
function printed in Part 5.** The assembly, including the compactness geometry
base R cannot compute, is in `data/build-data.R`.

Houston County's method of electing its Board of Commissioners is the subject of
pending litigation, *Driver v. Houston County*, No. 5:25-cv-00025 (M.D. Ga.), in
which your instructor has been retained by plaintiffs. **No material from any
expert report is used here, and this lab reaches no legal conclusion about that
case or any other.**
